### 2013-02-12 cordillera-climate.tex
Initial commit
Created an embryonic paper draft with abstract, sections and figures.

% Copernicus manuscript
\documentclass[tc, ms, noauthor]{copernicus}

% Copernicus final print
%\documentclass[tc]{copernicus}

% Copernicus discussion paper
%\documentclass[tcd, hvmath]{copernicus_discussions}

% Extra packages
%\usepackage{color}
%\usepackage{hyperref}

% All floats at the end of the manuscripts
\usepackage{endfloat}

% Add to-do notes
\usepackage{todonotes}
\newcommand{\missingref}[0]{\todo[color=cyan]{Need ref.}}

% Set figures repository
\graphicspath{{figures/}}

% ----------------------------------------------------------------------
\begin{document}

\title{Choosing climate forcing for numerical simulations of the Cordilleran ice sheet}
\runningtitle{Cordilleran ice sheet climate forcing}

%\author{Julien Seguinot et al.}
%\affil{Department of Physical Geography and Quaternary Geology, Stockholm University, S-106 91 Stockholm, Sweden}
\runningauthor{J. Seguinot et al.}
\correspondence{J. Seguinot\\ (julien.seguinot@natgeo.su.se)}

\received{}
\pubdiscuss{}
\revised{}
\accepted{}
\published{}

\maketitle

% ----------------------------------------------------------------------
\begin{abstract}
We present an ensemble of numerical simulations of the former Cordilleran ice sheet forced by present-day climatologies from five different datasets. Surface air temperature and precipitation rates from the NCEP/NCAR reanalysis, the Climate Forecast System Reanalysis, the ERA-Interim reanalysis and the North American Regional Reanalysis are used to force the ice sheet model PISM. Glacial conditions are mimicked by applying constant temperature offsets over the entire modelling domain, and a positive degree-day model is used to compute surface mass balance. The modelled ice sheet appears highly sensitive to the choice of climate forcing. Best results were obtained from the North American Regional Reanalysis, a high-resolution regional reanalysis which assimilates precipitation.
\end{abstract}
% ----------------------------------------------------------------------

% ----------------------------------------------------------------------
\input{sections/intro}
\input{sections/climate}
\input{sections/model}
\input{sections/results}
\input{sections/discuss}
% ----------------------------------------------------------------------

% ----------------------------------------------------------------------
\conclusions  %% \conclusions[modified heading if necessary]
\label{sec:conclusion}
% ----------------------------------------------------------------------

% ----------------------------------------------------------------------
%\begin{acknowledgements}
%\end{acknowledgements}
% ----------------------------------------------------------------------

\bibliographystyle{copernicus}
\bibliography{/home/julien/work/biblio/ref}
\end{document}

### 2013-02-12 sections/climate.tex
Initial commit
Created an embryonic paper draft with abstract, sections and figures.

% ----------------------------------------------------------------------
\section{Climate forcing}
\label{sec:climate}
% ----------------------------------------------------------------------

\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-temp}
	\end{center}
	\caption{Summer temperature maps from the five datasets used in this study and winter temperature map from the WorldClim dataset.}
	\label{fig:temp}
\end{figure}

\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-prec}
	\end{center}
	\caption{Winter precipitation rate maps from the five datasets used in this study and summer precipitation map from the WorldClim dataset.}
	\label{fig:prec}
\end{figure}

\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-topo}
	\end{center}
	\caption{Reference topography used for temperature lapse-rate corrections from the five climate datasets used in the study and ETOPO1 topography used as basal condition for the ice flow model.}
	\label{fig:topo}
\end{figure}

### 2013-02-12 sections/discuss.tex
Initial commit
Created an embryonic paper draft with abstract, sections and figures.

--- 2013-02-12 sections/climate.tex
@@
 % ----------------------------------------------------------------------
-\section{Climate forcing}
-\label{sec:climate}
+\section{Discussion}
+\label{sec:discussion}
 % ----------------------------------------------------------------------
-
-\begin{figure}[t]
-	\vspace*{2mm}
-	\begin{center}
-		\includegraphics[width=13cm]{cordillera-climate-temp}
-	\end{center}
-	\caption{Summer temperature maps from the five datasets used in this study and winter temperature map from the WorldClim dataset.}
-	\label{fig:temp}
-\end{figure}
-
-\begin{figure}[t]
-	\vspace*{2mm}
-	\begin{center}
-		\includegraphics[width=13cm]{cordillera-climate-prec}
-	\end{center}
-	\caption{Winter precipitation rate maps from the five datasets used in this study and summer precipitation map from the WorldClim dataset.}
-	\label{fig:prec}
-\end{figure}
-
-\begin{figure}[t]
-	\vspace*{2mm}
-	\begin{center}
-		\includegraphics[width=13cm]{cordillera-climate-topo}
-	\end{center}
-	\caption{Reference topography used for temperature lapse-rate corrections from the five climate datasets used in the study and ETOPO1 topography used as basal condition for the ice flow model.}
-	\label{fig:topo}
-\end{figure}

### 2013-02-12 sections/intro.tex
Initial commit
Created an embryonic paper draft with abstract, sections and figures.

% ----------------------------------------------------------------------
\introduction
\label{sec:intro}
% ----------------------------------------------------------------------

\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=8cm]{cordillera-climate-locmap}
	\end{center}
	\caption{Shaded relief map of northern North America. The frame delimits the modelling domain. After ETOPO1 \citep{data:etopo1} and Natural Earth Data \citep{data:naturalearth}.}
	\label{fig:locmap}
\end{figure}

### 2013-02-12 sections/results.tex
Initial commit
Created an embryonic paper draft with abstract, sections and figures.

% ----------------------------------------------------------------------
\section{Results}
\label{sec:results}
% ----------------------------------------------------------------------

\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-cool06}
	\end{center}
	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a climate 6\,\unit{\degree C} colder than present for each climate forcing.}
	\label{fig:cool06}
\end{figure}

\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-extent}
	\end{center}
	\caption{Extent of ice cover after 10\,kyr as a function of applied temperature offsets for each climate forcing.}
	\label{fig:extent}
\end{figure}

\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=9cm]{cordillera-climate-ivolarea}
	\end{center}
	\caption{Total ice volume and glaciated area after 10\,kyr as a function of temperature offset and climate forcing.}
	\label{fig:ivolarea}
\end{figure}

\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=13cm]{cordillera-climate-best}
	\end{center}
	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr using temperature offsets that lead to similar areas of ice cover for each climate forcing.}
	\label{fig:best}
\end{figure}

### 2013-02-12 sections/climate.tex
Added subsections in climate section
Added subsection 2.1 Observation data
Mostly typed in from paper draft with a few changes.

@@
 % ----------------------------------------------------------------------
+
+\subsection{Observation data}
+
+WorldClim \citep{data:worldclim} is a high resolution climate dataset interpolated between from weather station data over global land areas.
+The interpolation uses 30\,arc\,s aggregated, hole-filled SRTM\needref data \todo{check WorldClim methods}, which provides a resolution much higher than attained by general circulation models.
+
+However, the density of weather stations used by WorldClim in the Northern American Cordillera is highly inhomogeneous, and if good coverage exists in the southern parts of our modelling domain, several hundred kilometers can separate nearby stations in the north \citep{data:worldclim}.
+
+A second drawback of the dataset in an ice sheet modelling view is the lack of data on marine surfaces, particularly on the Pacific continental shelf where glaciers have advanced during the last glacial cycle\needref.
+
+% ----------------------------------------------------------------------
+
+\subsection{Reanalysis data}
+
+% ----------------------------------------------------------------------
+
+\subsection{Cordillera climates}
 
@@
 
+% ----------------------------------------------------------------------
+
+\subsection{Preprocessing and lapse-rate corrections}
+
 \begin{figure}[t]

### 2013-02-12 sections/climate.tex
Added subsection 2.2 Reanalysis data

@@
 \subsection{Reanalysis data}
+
+In addition to the WorldClim data, we use three global and one regional climate reanalysis to force the ice sheet model.
+Monthly climatologies from the NCEP/NCAR reanalysis\needref and the North American Regional Reanalysis (NARR)\needref were obtained from the NOAA ESRL's PSD website\needref, and monthly climatologies from the ERA-Interim\needref and Climate System Forecast Reanalyses (CFSR)\needref were computed from their monthly mean timeseries.
+As a mixed product of observation and circulation model, we believe that reanalysis may perform best in poorly monitored regions such as the northernmost American Cordillera.
+Additional information on the data used is gatherd in table \ref{tab:reanalyses}\todo{fill the table}.
+
+\begin{table}[t]
+	\caption{Characteristic of reanalysis data used as a forcing for the ice sheet model.}
+	\label{tab:reanalyses}
+	\vskip4mm
+	\centering
+	\begin{tabular}{lllll}
+		\tophline
+		Reanalysis& coverage& time period&  resolution& reference\\
+		\middlehline
+		NCEP/NCAR&  global&   1981 -- 2010& ??& ??\\
+		ERA-Interim&global&   1979 -- 2011& ??& ??\\
+		CFSR&       global&   1979 -- 2010& ??& ??\\
+		NARR&       global&   1979 -- 2000& ??& ??\\
+		\bottomhline
+	\end{tabular}
+\end{table}
 

### 2013-02-13 sections/climate.tex
Added subsection 2.3 Cordilleran climates

@@
 
-\subsection{Cordillera climates}
+\subsection{Cordilleran climates}
 
@@
 
+Figure~\ref{fig:temp} shows the spatial distribution of summer (JJA) air surface temperatures from the five climate datasets used in the study, and winter (DJF) air surface temperatures from the WorldClim interpolated observation data.
+Summer temperatures are most relevant to the glacier model as they drive summer melt.
+
+As can be expected, temperatures decrease with latitude, and regions further inland experience colder winters.
+It should be noted that the temperature gradient is much stronger in the winter than it is in the summer, and that at low elevations, temperatures get well above zero during the summer months in the entire modelling domain.
+In other words, there is a strong seasonality contrast beween coastal and inland regions, and regions where mean annual temperatures are well below freezing do experience warm summers.
+
 \begin{figure}[t]
@@
 \end{figure}
+
+Figure~\ref{fig:prec} shows the spatial distribution of winter (DJF) precipitation rates from the five climate datasets used in the study, and summer (JJA) precipitation rates from the WorldClim interpolated observation data.
+Winter precipitation is most relevant to the glacier model as it drives winter accumulation.
+
+Once again expectedly, coastal regions recieve much more precipitation than inland ones due to the unborken topographical barrier of the Boundary Ranges / Coast Mountains / Cascadia.
+It should be noted that from a glacier mass-balance view, this contrast is made even stronger by the difference of timing of the precipitation peak through the year.
+Although coastal regions experience most precipitation in the accumulation season, inland regions experience dry winters and most of the precipitation falls as rain during the summer months.
+
+In regions like Northern Yukon and Alaska, dry winters and warm summers prevent snow accumulation and glacial inception despite of mean annual temperatures being well below zero at present.
+In order to account for these strong gradients in seasonality, we use monthly means to drive the ice sheet model.
+Previous runs made with mean annual temperature and precipitation forcing and not presented here lead to unrealistic ice cover in Northern Yukon and Alaska even under present-day climate while leaving the Coast Mountains free of ice.
 

### 2013-02-13 sections/climate.tex
Added subsection 2.4 Preprocessing and lapse-rates

@@
 
+Air surface temperature and precipitation rates from the NCEP/NCAR reanalysis, the ERA-Interim reanalysis, the CFSR, the NARR and WorldClim data were reprojected to Canadian Atlas Lambert (EPSG code XXXX\todo{add EPSG code}) conformal conic projection.
+The data were bilinearly interpolated to the model resolution of 10\,km, which is not shown by figures~\ref{fig:temp} and~\ref{fig:prec}.
+
+For the WorldClim data, data holes in oceanic areas were filled by a nearest-neighbour extrapolation to allow ice to extend to these regions.
+
+For the CFSR, an alternative forcing was prepared by smoothing the precipitation fields to avoid artifacts seen in figure~\ref{fig:prec}.
+This was done by averaging data locally in a circular neighbourhood of 7 pixels in diameter prior to reprojection.
+However the effect of smoothin the precipitation field is very limited as we will see further.
+All the preprocessing steps were made in GRASS~GIS using scripts made available on the first author's website.
+
 \begin{figure}[t]
@@
 \end{figure}
+
+Throughout the simulation, the numerical model dynamically applies temperature lapse-rate corrections that account for both differences between the climate reference topography and the ice flow model topography, and the thickness of overlying ice thickness.
+The reference topography from the five forcing datasets as well as the ETOPO1\citep{data:etopo1} topography used to force the ice flow model are shown in figure~\ref{fig:topo}.
+A lapse-rate of 6\unit{\degree C\,km^{-1}} was used in all simulations.
+No lapse-rate corrections are applied to precipitation rate.

### 2013-02-13 sections/climate.tex
Added reanalysis data references

@@
 
-In addition to the WorldClim data, we use three global and one regional climate reanalysis to force the ice sheet model.
-Monthly climatologies from the NCEP/NCAR reanalysis\needref and the North American Regional Reanalysis (NARR)\needref were obtained from the NOAA ESRL's PSD website\needref, and monthly climatologies from the ERA-Interim\needref and Climate System Forecast Reanalyses (CFSR)\needref were computed from their monthly mean timeseries.
-As a mixed product of observation and circulation model, we believe that reanalysis may perform best in poorly monitored regions such as the northernmost American Cordillera.
-Additional information on the data used is gatherd in table \ref{tab:reanalyses}\todo{fill the table}.
+In addition to the WorldClim data, we use surface air temperature and precipitation rate data from three global and one regional climate reanalysis to force the ice sheet model: the NCEP/NCAR reanalysis, the ERA-Interim reanalysis, the Climate System Forecast Reanalysis (CFSR) and the North American Regional Reanalysis (NARR).
+Monthly climatologies from NCEP/NCAR and NARR reanalysis were provided by the NOAA/OAR/ESRL PSD, Boulder, Colorado, USA, from their Web site at \url{http://www.esrl.noaa.gov/psd/} and monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean timeseries.
+Further information on the data used is gatherd in Table \ref{tab:reanalyses}.
+As a mixed product from observations and a circulation model, we believe that reanalysis may perform best in poorly monitored regions such as the northernmost American Cordillera.
 
 \begin{table}[t]
-	\caption{Characteristic of reanalysis data used as a forcing for the ice sheet model.}
+	\caption{Characteristic of reanalysis climatologies used to force the ice sheet model.}
 	\label{tab:reanalyses}
@@
 		\tophline
-		Reanalysis& coverage& time period&  resolution& reference\\
+		Reanalysis& Spatial coverage& Averaging period& Resolution& Description\\
 		\middlehline
-		NCEP/NCAR&  global&   1981 -- 2010& ??& ??\\
-		ERA-Interim&global&   1979 -- 2011& ??& ??\\
-		CFSR&       global&   1979 -- 2010& ??& ??\\
-		NARR&       global&   1979 -- 2000& ??& ??\\
+		NCEP/NCAR&  global&     1981 -- 2010& 1.875\degree& \citet{data:ncar}\\
+		ERA-Interim&global&     1979 -- 2011& 1.000\degree& \citet{data:erai}\\
+		CFSR&       global&     1979 -- 2010& 0.325\degree& \citet{data:cfsr}\\
+		NARR&       North America& 1979 -- 2000& 32\,km& \citet{data:narr}\\
 		\bottomhline

### 2013-02-13 sections/climate.tex
Improved WorldClim data description

@@
 
-WorldClim \citep{data:worldclim} is a high resolution climate dataset interpolated between from weather station data over global land areas.
-The interpolation uses 30\,arc\,s aggregated, hole-filled SRTM\needref data \todo{check WorldClim methods}, which provides a resolution much higher than attained by general circulation models.
+WorldClim \citep{data:worldclim} is a high resolution climate dataset built from observation data.
+These climate surfaces were generated by spatial interpolation between a large set of weather stations using a 30\,arc\,s-aggregated flavour of hole-filled SRTM data over global land areas.
+This provides a resolution much higher than attained by general circulation models.
 

### 2013-02-14 sections/climate.tex
Improved climate section text

@@
 
-As can be expected, temperatures decrease with latitude, and regions further inland experience colder winters.
-It should be noted that the temperature gradient is much stronger in the winter than it is in the summer, and that at low elevations, temperatures get well above zero during the summer months in the entire modelling domain.
+Temperatures generally decrease from south to north and regions further inland experience colder winters.
+It should be noted that the temperature gradient is much stronger in the winter than it is in the summer.
+At low elevations temperature get well above zero during the summer months over the entire modelling domain.
 In other words, there is a strong seasonality contrast beween coastal and inland regions, and regions where mean annual temperatures are well below freezing do experience warm summers.
@@
 
-Once again expectedly, coastal regions recieve much more precipitation than inland ones due to the unborken topographical barrier of the Boundary Ranges / Coast Mountains / Cascadia.
-It should be noted that from a glacier mass-balance view, this contrast is made even stronger by the difference of timing of the precipitation peak through the year.
-Although coastal regions experience most precipitation in the accumulation season, inland regions experience dry winters and most of the precipitation falls as rain during the summer months.
+Coastal regions generally recieve much more precipitation than inland ones as a results of the continuous topographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascadia.
+In a glacier mass-balance view, this contrast is made even stronger by the difference in timing of the precipitation peak through the year.
+Although coastal regions experience most precipitation during the accumulation season, inland regions experience dry winters and most of the precipitation falls as rain during the summer months.
 
-In regions like Northern Yukon and Alaska, dry winters and warm summers prevent snow accumulation and glacial inception despite of mean annual temperatures being well below zero at present.
+In regions like Northern Yukon and Alaska, dry winters and warm summers are unfavourable conditions to accumulation and glacial inception despite of the strongly negative mean annual temperatures at present.
 In order to account for these strong gradients in seasonality, we use monthly means to drive the ice sheet model.
-Previous runs made with mean annual temperature and precipitation forcing and not presented here lead to unrealistic ice cover in Northern Yukon and Alaska even under present-day climate while leaving the Coast Mountains free of ice.
+Previous runs forced by annual means of temperature and precipitation not presented here systematically conducted to unrealistic ice build-up in northern Yukon and Alaska under present-day climate while leaving the Coast Mountains free of ice.
 
@@
 
-Air surface temperature and precipitation rates from the NCEP/NCAR reanalysis, the ERA-Interim reanalysis, the CFSR, the NARR and WorldClim data were reprojected to Canadian Atlas Lambert (EPSG code XXXX\todo{add EPSG code}) conformal conic projection.
-The data were bilinearly interpolated to the model resolution of 10\,km, which is not shown by figures~\ref{fig:temp} and~\ref{fig:prec}.
+Air surface temperature and precipitation rates from the NCEP/NCAR, ERA-Interim, CFSR and NARR reanalyses and WorldClim data were reprojected to Canadian Atlas Lambert conformal conic projection (EPSG code~3978) and bilinearly interpolated to the model resolution of 10\,km.
+In addition, WorldClim data was extrapolated to marine surfaces using a nearest-neighbour approach.
+These last two steps are not presented on figures~\ref{fig:temp} and~\ref{fig:prec}.
 
-For the WorldClim data, data holes in oceanic areas were filled by a nearest-neighbour extrapolation to allow ice to extend to these regions.
-
-For the CFSR, an alternative forcing was prepared by smoothing the precipitation fields to avoid artifacts seen in figure~\ref{fig:prec}.
+For the CFSR, an alternative forcing was prepared by smoothing the precipitation field in order to correct for artifacts seen in figure~\ref{fig:prec}.
 This was done by averaging data locally in a circular neighbourhood of 7 pixels in diameter prior to reprojection.
-However the effect of smoothin the precipitation field is very limited as we will see further.
-All the preprocessing steps were made in GRASS~GIS using scripts made available on the first author's website.
+Yet we will show that this smoothing has very limited effect on the ice sheet model outcome.
+All the preprocessing steps described above were made in GRASS~GIS using scripts available on the first author's website.
 
@@
 
-Throughout the simulation, the numerical model dynamically applies temperature lapse-rate corrections that account for both differences between the climate reference topography and the ice flow model topography, and the thickness of overlying ice thickness.
-The reference topography from the five forcing datasets as well as the ETOPO1\citep{data:etopo1} topography used to force the ice flow model are shown in figure~\ref{fig:topo}.
-A lapse-rate of 6\unit{\degree C\,km^{-1}} was used in all simulations.
-No lapse-rate corrections are applied to precipitation rate.
+Throughout the simulation, the numerical model dynamically applies temperature lapse-rate corrections that account for differences between the climate forcing reference topography and the ice flow model topography on the one hand, and the evolution of ice thickness on the over hand.
+For reanalysis data, the circulation model surface topography (or surface geopotential height) was used as reference topography.
+For the WorldClim data, the aggregated hole-filled SRTM altitude field is used as reference topography.
+As a basal condition for the ice flow model, we use the ETOPO1\citep{data:etopo1} combined topography and bathymetry dataset.
+The reference topographies from the five forcing datasets, alongside the ETOPO1 data are shown in figure~\ref{fig:topo}.
+A lapse-rate of 6\unit{\degree C\,km^{-1}} is used in all simulations and no lapse-rate corrections apply to precipitation rates.

### 2013-02-14 sections/model.tex
Initial commit
Created an embryonic paper draft with abstract, sections and figures.
Added techincal note for Constantine

--- 2013-02-12 sections/intro.tex
@@
 % ----------------------------------------------------------------------
-\introduction
-\label{sec:intro}
+\section{Model setup}
+\label{sec:setup}
 % ----------------------------------------------------------------------
 
-\begin{figure}[t]
-	\vspace*{2mm}
-	\begin{center}
-		\includegraphics[width=8cm]{cordillera-climate-locmap}
-	\end{center}
-	\caption{Shaded relief map of northern North America. The frame delimits the modelling domain. After ETOPO1 \citep{data:etopo1} and Natural Earth Data \citep{data:naturalearth}.}
-	\label{fig:locmap}
-\end{figure}
+\begin{figure}\begin{framed}
+	\textbf{Technical note for Constantine}
+	\todo{Remove this note}
+
+	All my simulations use the same set of options except of a different \verb|-atmosphere_given_file| (climate forcing) and \verb|-atmosphere_delta_T_file| (temperature offset).
+Here comes the options that I used; I think you saw them a couple of times now.
+I simplified paths and filenames and remove output options for readability
+
+	\begin{verbatim}
+	mpprun pismr -boot_file boot.nc -Mx 150 -My 300
+	-Mz 51 -Lz 5000 -Mbz 11 -Lbz 1000 -ys 0 -ye 10000
+	-topg_to_phi 10,30,0.0,200.0
+	-atmosphere given,lapse_rate,delta_T -atmosphere_given_file atm.nc
+	-atmosphere_given_period 1
+	-temp_lapse_rate 6 -atmosphere_lapse_rate_file atm.nc
+	-atmosphere_delta_T_file step-cool08.nc
+	-surface pdd -pdd_annualize
+	-ocean constant,delta_SL -ocean_delta_SL_file step-low120.nc
+	-ssa_sliding
+	-config_override config.nc
+	\end{verbatim}
+
+	And here comes the relevant contents of \verb|config.nc|:
+
+	\begin{verbatim}
+	netcdf config \{
+	variables:
+	  byte pism_overrides ;
+	    pism_overrides:bed_deformation_model = "lc" ;
+	    pism_overrides:do_thickness_calving = "true" ;
+	    pism_overrides:calving_at_thickness = 200. ;
+	    pism_overrides:calving_front_stress_boundary_condition = "yes" ;
+	    pism_overrides:part_grid = "yes" ;
+	    pism_overrides:part_redist = "yes" ;
+	    pism_overrides:kill_icebergs = "yes" ;
+	    pism_overrides:bootstrapping_geothermal_flux_value_no_var = 0.07 ;
+	    pism_overrides:pdd_factor_snow = 0.003 ;
+	    pism_overrides:pdd_factor_ice = 0.008 ;
+	    pism_overrides:pdd_refreeze = 0.6 ;
+	    pism_overrides:pdd_std_dev = 5 ;
+	\}
+
+	\end{verbatim}
+
+\end{framed}\end{figure}

### 2013-02-17 sections/model.tex
Included text from the old draft as comments
Added subsection 3.1 Surface mass-balance

@@
 	\textbf{Technical note for Constantine}
-	\todo{Remove this note}
 
@@
 \end{framed}\end{figure}
+
+\subsection{Surface mass-balance}
+
+\begin{table}[t]
+	\caption{Values of the positive degree-day model parameters.}
+	\label{tab:pdd}
+	\vskip4mm
+	\centering
+	\begin{tabular}{ll}
+		\tophline
+		Parameter& Value\\
+		\middlehline
+		degree-day factor for snow & 3\,\unit{mm\,K^{-1}\,day^{-1}}\\
+		degree-day factor for ice & 8\,\unit{mm\,K^{-1}\,day^{-1}}\\
+		standard deviation of temperatures & 5\,K\\
+		refreezing faction & 0.6 \\
+		\bottomhline
+	\end{tabular}
+\end{table}
+
+Monthly temperatures and precipitation rates are converted to surface accumulation and ablation using a temperature-index (positive degree-day) model\needref.
+Accumulation is computed as the amount of precipitation that falls at times when temperature is below freezing \todo{Check PISM doc}.
+Ablation is a transfer function of the number of positive degree-day, defined as the sum of mean daily temperatures through the year.
+Positive degree-day are computed by the \citet{calov-greve-2011} integrand to account for day-to-day variability which is not represented in the monthly climate forcing.
+The ablation model use a different coefficient for snow and ice and accounts for refreezing.
+Model parameters are explicited in table~\ref{tab:pdd}
+
+Although a positive degree-day model is a very crude approximation of processes occurring at the glacier surface, we will show that it captures reasonable regional patterns of accumulation and ablation when using accurate climate data.
+
+\todo[inline]{
+	\textbf{Include the \citet{calov-greve-2011} integrand formula?}
+
+	The number of positive degree-days, defined as the integrand of positive temperatures over time, is computed annually using a formula derived by \citet{calov-greve-2011} that accounts for day-to-day variability of temperature by assuming a normal distribution of variance $\sigma=5\degree C$ along a central (input) value $T$:
+
+	\begin{equation}
+		\mathrm{PDD} = \int_{t_0}^{t_1} dt \left[
+			\frac{\sigma}{\sqrt{2\pi}}
+			\exp\left({-\frac{T(t)^2}{2\sigma^2}}\right)
+			+\frac{T(t)}{2}
+			\mathrm{erfc} \left(-\frac{T(t)}{\sqrt{2}\sigma}\right)
+		\right]
+	\end{equation}
+
+Where $[t_0; t_1]$ is one year and $\mathrm{erfc}$ is the complementary error function.}
+
+%\subsection{Ice flow dynamics}
+%To resolve ice dynamics, PISM specifically uses a heuristic combination of two classical and well-based zeroth-order shallow-ice approximations through a weighting function \citep{bueler-brown-2009}.
+The Shallow Shelf Approximation \citep[SSA,][]{morland-1987,weis-etal-1999} is used as a ``sliding law'' for the Shallow Ice Approximation \citep[SSA,][]{hutter-1983}.
+Depth-averaged velocities $\vec{v}$ are computed as a weighted sum of SIA velocities $\vec{v_{\mathrm{SIA}}}$ and SSA velocities $\vec{v_{\mathrm{SSA}}}$ by
+%
+%\begin{equation}
+%	\vec{v} = f(|\vec{v_{\mathrm{SSA}}}|)\vec{v_{\mathrm{SIA}}}
+%	+ (1-f(|\vec{v_{\mathrm{SSA}}}|))\vec{v_{\mathrm{SSA}}},
+%\end{equation}
+%
+%with
+%
+%\begin{equation}
+%	f(|\vec{v_{\mathrm{SSA}}}|) = 1
+%	- \frac{2}{\pi}\arctan{\frac{|\vec{v_{\mathrm{SSA}}}|^2}{100^2}}.
+%\end{equation}
+%
+%This approach, although using a heuristic, allows realistic representation of horizontal shear and longitudinal stresses where they matter \citep{bueler-brown-2009}.
+%
+%
+%The SSA velocities are determined by a a pseudo-plastic sliding law in which the yield stress is depends on the availability of basal water and a prescibed till friction angle.
+We apply lower friction angles at lower elevation to model more slippery beds due to the presence of marine sediments.
+%
+%\begin{equation}
+%	\phi = \left\{\begin{array}{llc}
+%		10\,\unit{\degree C} & \mathrm{for} &               z<  0\,\unit{m} \\
+%		z/10 + 10°           & \mathrm{for} &   0\,\unit{m}<z<200\,\unit{m} \\
+%		30\,\unit{\degree C} & \mathrm{for} & 200\,\unit{m}<z               \\
+%	\end{array}\right.
+%\end{equation}
+%
+%\subsection{Glacier bed interface}
+%
+%The ice-sheet bed topography responds to the ice-loads following a regional isostasy model with flexure and relaxation.
+%
+%Geothermal heat flux is prescribed to 70\,\unit{W\,m^{-2}}
+%
+%\subsection{Simulation design and numerical implementation}
+%
+%The simulations were run with PISM, a parallel ice sheet model (version stable 0.5.11).
+Documentation on the code can be found in several publications \citep[e.g.,][]{bueler-brown-2009,aschwanden-etal-2012} and more extensively online \citep[{\url{http://www.pism-docs.org}},][]{web:pism}.
+
+%Model equations are discretized horizontally on a 10~km-resolution, 150 by 300 points grid, and vertically within the ice on 21 irregularly-spaced layers.
+Time-stepping is adaptive, but a maximum allowed time step of 0.08 is used to ensure monthly capture of climate data by the surface mass-balance model.
+%
+%The simulations were run in parallel on 9~computation nodes of 8~cores each, for a total of 72~cores, on the supercomputer ekman at the PDC center for high performance computing of Kungliga Tekniska högskolan in Stockholm.
+Each simulation completed in about 10~hours.
+
+%In this study we model glacial inception and growth of the Cordilleran ice sheet to a state nearing its last glacial maximum.
+For each present-day climate forcing we run a series of six simulations by applying a constant, homogeneous temperature offset of 4 to 9 \unit{\degree C} over the entire model domain.
+Each simulation initializes from ice-free conditions and is run for 10~000~years, a time interval we think to be representative of the rapid growth of the ice sheet from nearly ice-free conditions to last glacial maximum extent \needref.
+No precipitation corrections were used.
+%

### 2013-02-17 sections/model.tex
Added subsection 3.2 Ice dynamics

@@
 
-%\subsection{Ice flow dynamics}
-%To resolve ice dynamics, PISM specifically uses a heuristic combination of two classical and well-based zeroth-order shallow-ice approximations through a weighting function \citep{bueler-brown-2009}.
-The Shallow Shelf Approximation \citep[SSA,][]{morland-1987,weis-etal-1999} is used as a ``sliding law'' for the Shallow Ice Approximation \citep[SSA,][]{hutter-1983}.
+\subsection{Ice dynamics}
+
+Ice flow velocities are computed by a heuristic where the Shallow Shelf Approximation (SSA) is used as a ``sliding law'' to a polythermal Shallow Ice Approximation (SIA) \citep{bueler-brown-2009}.
+
+\todo[inline]{
+	\textbf{Include SSA+SIA \citet{bueler-brown-2009} formula}
+
 	Depth-averaged velocities $\vec{v}$ are computed as a weighted sum of SIA velocities $\vec{v_{\mathrm{SIA}}}$ and SSA velocities $\vec{v_{\mathrm{SSA}}}$ by
-%
-%\begin{equation}
-%	\vec{v} = f(|\vec{v_{\mathrm{SSA}}}|)\vec{v_{\mathrm{SIA}}}
-%	+ (1-f(|\vec{v_{\mathrm{SSA}}}|))\vec{v_{\mathrm{SSA}}},
-%\end{equation}
-%
-%with
-%
-%\begin{equation}
-%	f(|\vec{v_{\mathrm{SSA}}}|) = 1
-%	- \frac{2}{\pi}\arctan{\frac{|\vec{v_{\mathrm{SSA}}}|^2}{100^2}}.
-%\end{equation}
-%
-%This approach, although using a heuristic, allows realistic representation of horizontal shear and longitudinal stresses where they matter \citep{bueler-brown-2009}.
-%
-%
-%The SSA velocities are determined by a a pseudo-plastic sliding law in which the yield stress is depends on the availability of basal water and a prescibed till friction angle.
-We apply lower friction angles at lower elevation to model more slippery beds due to the presence of marine sediments.
-%
-%\begin{equation}
-%	\phi = \left\{\begin{array}{llc}
-%		10\,\unit{\degree C} & \mathrm{for} &               z<  0\,\unit{m} \\
-%		z/10 + 10°           & \mathrm{for} &   0\,\unit{m}<z<200\,\unit{m} \\
-%		30\,\unit{\degree C} & \mathrm{for} & 200\,\unit{m}<z               \\
-%	\end{array}\right.
-%\end{equation}
-%
+
+	\begin{equation}
+		\vec{v} = f(|\vec{v_{\mathrm{SSA}}}|)\vec{v_{\mathrm{SIA}}}
+	  + (1-f(|\vec{v_{\mathrm{SSA}}}|))\vec{v_{\mathrm{SSA}}},
+	\end{equation}
+
+	with
+
+	\begin{equation}
+		f(|\vec{v_{\mathrm{SSA}}}|) = 1
+		- \frac{2}{\pi}\arctan{\frac{|\vec{v_{\mathrm{SSA}}}|^2}{100^2}}.
+	\end{equation}
+}
+
+SSA velocities are determined by a a pseudo-plastic sliding law where the yield stress depends on the availability of basal water and a prescibed till friction angle$\phi$.
+We apply lower friction angles at low elevation to mimick the more slippery beds associated to the presence of marine sediments:
+
+\begin{equation}
+	\phi = \left\{\begin{array}{llc}
+		10\,\unit{\degree C} & \mathrm{for} &               z<  0\,\unit{m} \\
+		z/10 + 10°           & \mathrm{for} &   0\,\unit{m}<z<200\,\unit{m} \\
+		30\,\unit{\degree C} & \mathrm{for} & 200\,\unit{m}<z               \\
+	\end{array}\right.
+\end{equation}
+
 %\subsection{Glacier bed interface}

### 2013-02-17 sections/model.tex
Added subsection 3.4 Simulation design and num...

@@
 % ----------------------------------------------------------------------
+
+%The simulations were run with PISM, a parallel ice sheet model (version stable 0.5.11).
+Documentation on the code can be found in several publications \citep[e.g.,][]{bueler-brown-2009,aschwanden-etal-2012} and more extensively online \citep[{\url{http://www.pism-docs.org}},][]{web:pism}.
 
@@
 %
-%\subsection{Simulation design and numerical implementation}
-%
-%The simulations were run with PISM, a parallel ice sheet model (version stable 0.5.11).
-Documentation on the code can be found in several publications \citep[e.g.,][]{bueler-brown-2009,aschwanden-etal-2012} and more extensively online \citep[{\url{http://www.pism-docs.org}},][]{web:pism}.
+\subsection{Simulation design and numerical implementation}
 
-%Model equations are discretized horizontally on a 10~km-resolution, 150 by 300 points grid, and vertically within the ice on 21 irregularly-spaced layers.
-Time-stepping is adaptive, but a maximum allowed time step of 0.08 is used to ensure monthly capture of climate data by the surface mass-balance model.
-%
-%The simulations were run in parallel on 9~computation nodes of 8~cores each, for a total of 72~cores, on the supercomputer ekman at the PDC center for high performance computing of Kungliga Tekniska högskolan in Stockholm.
-Each simulation completed in about 10~hours.
+In this study we model glacial inception and growth of the Cordilleran ice sheet to a state approaching its last glacial maximum.
+We mimick glacial conditions by applying constant temperature offsets homogeneously over the entire domain.
+Each simulation starts from ice-free conditions and is run for 10,000 years, a time interval which we think to be representative of the rapid build-up of the last Cordilleran Ice Sheet from nearly ice-free condition to last glacial maximum extent \needref.
 
-%In this study we model glacial inception and growth of the Cordilleran ice sheet to a state nearing its last glacial maximum.
-For each present-day climate forcing we run a series of six simulations by applying a constant, homogeneous temperature offset of 4 to 9 \unit{\degree C} over the entire model domain.
-Each simulation initializes from ice-free conditions and is run for 10~000~years, a time interval we think to be representative of the rapid growth of the ice sheet from nearly ice-free conditions to last glacial maximum extent \needref.
-No precipitation corrections were used.
-%
+Our computational domain consists of 10~km-resolution, 150 by 300 points horizontal grid.
+Equations that involve a vertical coordinate are solved on 51 irregularly-spaced layers in ice and 11 regularly spaced layers in rock\todo{Check PISM doc}.
+The simulations were run in parallel on 32 cores, on NSC's supercomputer triolith\needref.

### 2013-02-18 sections/model.tex
Added subsection 3.3 Glacier bed interface

@@
 
-%\subsection{Glacier bed interface}
-%
-%The ice-sheet bed topography responds to the ice-loads following a regional isostasy model with flexure and relaxation.
-%
-%Geothermal heat flux is prescribed to 70\,\unit{W\,m^{-2}}
-%
+\subsection{Glacier bed interface}
+
+The glacier bed topography response to the ice load is computed by a regional isostasy model with flexure and relaxation \citep{lingle-clark-1985}.
+To compute basal temperatures with higher accuracy, the thermal model is extended 1\,km into the bedrock.
+An homogeneous geothermal heat flux of 70\,\unit{W\,m^{-2}} is provided as basal condition to this model.
+
 \subsection{Simulation design and numerical implementation}

### 2013-02-18 sections/model.tex
Added section 3 Model setup foreword

@@
 
-%The simulations were run with PISM, a parallel ice sheet model (version stable 0.5.11).
+We use PISM, a parallel ice sheet model (version stable 0.5.11).
 Documentation on the code can be found in several publications \citep[e.g.,][]{bueler-brown-2009,aschwanden-etal-2012} and more extensively online \citep[{\url{http://www.pism-docs.org}},][]{web:pism}.
@@
 
-The glacier bed topography response to the ice load is computed by a regional isostasy model with flexure and relaxation \citep{lingle-clark-1985}.
+The glacier bed topography response to the ice load is computed by a regional isostasy model with flexure and relaxation \citep{lingle-clark-1985}\todo{check ref.}.
 To compute basal temperatures with higher accuracy, the thermal model is extended 1\,km into the bedrock.

### 2013-02-18 sections/model.tex
Moved technical note to separate file

@@
 Documentation on the code can be found in several publications \citep[e.g.,][]{bueler-brown-2009,aschwanden-etal-2012} and more extensively online \citep[{\url{http://www.pism-docs.org}},][]{web:pism}.
-
-\begin{figure}\begin{framed}
-	\textbf{Technical note for Constantine}
-
-	All my simulations use the same set of options except of a different \verb|-atmosphere_given_file| (climate forcing) and \verb|-atmosphere_delta_T_file| (temperature offset).
-Here comes the options that I used; I think you saw them a couple of times now.
-I simplified paths and filenames and remove output options for readability
-
-	\begin{verbatim}
-	mpprun pismr -boot_file boot.nc -Mx 150 -My 300
-	-Mz 51 -Lz 5000 -Mbz 11 -Lbz 1000 -ys 0 -ye 10000
-	-topg_to_phi 10,30,0.0,200.0
-	-atmosphere given,lapse_rate,delta_T -atmosphere_given_file atm.nc
-	-atmosphere_given_period 1
-	-temp_lapse_rate 6 -atmosphere_lapse_rate_file atm.nc
-	-atmosphere_delta_T_file step-cool08.nc
-	-surface pdd -pdd_annualize
-	-ocean constant,delta_SL -ocean_delta_SL_file step-low120.nc
-	-ssa_sliding
-	-config_override config.nc
-	\end{verbatim}
-
-	And here comes the relevant contents of \verb|config.nc|:
-
-	\begin{verbatim}
-	netcdf config \{
-	variables:
-	  byte pism_overrides ;
-	    pism_overrides:bed_deformation_model = "lc" ;
-	    pism_overrides:do_thickness_calving = "true" ;
-	    pism_overrides:calving_at_thickness = 200. ;
-	    pism_overrides:calving_front_stress_boundary_condition = "yes" ;
-	    pism_overrides:part_grid = "yes" ;
-	    pism_overrides:part_redist = "yes" ;
-	    pism_overrides:kill_icebergs = "yes" ;
-	    pism_overrides:bootstrapping_geothermal_flux_value_no_var = 0.07 ;
-	    pism_overrides:pdd_factor_snow = 0.003 ;
-	    pism_overrides:pdd_factor_ice = 0.008 ;
-	    pism_overrides:pdd_refreeze = 0.6 ;
-	    pism_overrides:pdd_std_dev = 5 ;
-	\}
-
-	\end{verbatim}
-
-\end{framed}\end{figure}
 

### 2013-02-18 sections/technote.tex
Moved technical note to separate file

\begin{figure}\begin{framed}
	\textbf{Technical note for Constantine}

	All my simulations use the same set of options except of a different \verb|-atmosphere_given_file| (climate forcing) and \verb|-atmosphere_delta_T_file| (temperature offset). Here comes the options that I used; I think you saw them a couple of times now. I simplified paths and filenames and remove output options for readability

	\begin{verbatim}
	mpprun pismr -boot_file boot.nc -Mx 150 -My 300
	-Mz 51 -Lz 5000 -Mbz 11 -Lbz 1000 -ys 0 -ye 10000
	-topg_to_phi 10,30,0.0,200.0
	-atmosphere given,lapse_rate,delta_T -atmosphere_given_file atm.nc
	-atmosphere_given_period 1
	-temp_lapse_rate 6 -atmosphere_lapse_rate_file atm.nc
	-atmosphere_delta_T_file step-cool08.nc
	-surface pdd -pdd_annualize
	-ocean constant,delta_SL -ocean_delta_SL_file step-low120.nc
	-ssa_sliding
	-config_override config.nc
	\end{verbatim}

	And here comes the relevant contents of \verb|config.nc|:

	\begin{verbatim}
	netcdf config \{
	variables:
	  byte pism_overrides ;
	    pism_overrides:bed_deformation_model = "lc" ;
	    pism_overrides:do_thickness_calving = "true" ;
	    pism_overrides:calving_at_thickness = 200. ;
	    pism_overrides:calving_front_stress_boundary_condition = "yes" ;
	    pism_overrides:part_grid = "yes" ;
	    pism_overrides:part_redist = "yes" ;
	    pism_overrides:kill_icebergs = "yes" ;
	    pism_overrides:bootstrapping_geothermal_flux_value_no_var = 0.07 ;
	    pism_overrides:pdd_factor_snow = 0.003 ;
	    pism_overrides:pdd_factor_ice = 0.008 ;
	    pism_overrides:pdd_refreeze = 0.6 ;
	    pism_overrides:pdd_std_dev = 5 ;
	\}

	\end{verbatim}

\end{framed}\end{figure}

### 2013-02-18 sections/model.tex
Added missing references

@@
 We use PISM, a parallel ice sheet model (version stable 0.5.11).
-Documentation on the code can be found in several publications \citep[e.g.,][]{bueler-brown-2009,aschwanden-etal-2012} and more extensively online \citep[{\url{http://www.pism-docs.org}},][]{web:pism}.
+Documentation on the code can be found in several publications \citep[e.g.,][]{bueler-brown-2009,winkelmann-etal-2011,aschwanden-etal-2012} and more extensively online \citep[{\url{http://www.pism-docs.org}},][]{web:pism}.
 

### 2013-02-19 sections/model.tex
Improved surface mass-balance model description

@@
 
-Monthly temperatures and precipitation rates are converted to surface accumulation and ablation using a temperature-index (positive degree-day) model\needref.
-Accumulation is computed as the amount of precipitation that falls at times when temperature is below freezing \todo{Check PISM doc}.
-Ablation is a transfer function of the number of positive degree-day, defined as the sum of mean daily temperatures through the year.
-Positive degree-day are computed by the \citet{calov-greve-2011} integrand to account for day-to-day variability which is not represented in the monthly climate forcing.
-The ablation model use a different coefficient for snow and ice and accounts for refreezing.
-Model parameters are explicited in table~\ref{tab:pdd}
-
-Although a positive degree-day model is a very crude approximation of processes occurring at the glacier surface, we will show that it captures reasonable regional patterns of accumulation and ablation when using accurate climate data.
+Monthly temperatures and precipitation rates are converted to surface accumulation and ablation using a temperature-index (positive degree-day) model.
+Accumulation is computed as the amount of precipitation that falls when temperature is below freezing.\askconstantine{Is this true?
+What is the current threshold in stable 0.5?} Ablation is computed from the number of positive degree-days, defined as the integrand of temperatures above 0\,\degree\,C in one year.
+Positive degree-days are computed by the \citet{calov-greve-2011} formula to account for day-to-day variability unrepresented in the monthly climate forcing.
+The ablation model use different degree-day factors for snow and ice and accounts for refreezing.
+Complete description of the ablation model is available in PISM documentation \citep[{\url{http://www.pism-docs.org}},][]{web:pism} and model parameters are explicited in table~\ref{tab:pdd}.
 
@@
 
+A positive degree-day model is a very crude approximation of physical processes occurring at a glacier's surface.
+Yet we believe that it captures regional patterns of accumulation and ablation resonable enough for our study if accurate climate data is used, as we will show further.\todo{Is this paragraph really needed?}
+
 \subsection{Ice dynamics}
@@
 \todo[inline]{
-	\textbf{Include SSA+SIA \citet{bueler-brown-2009} formula}
+	\textbf{Include SSA+SIA \citet{bueler-brown-2009} formula?}
 

### 2013-02-19 sections/model.tex
Merged ice dynamics and bed interface

@@
 
-\subsection{Ice dynamics}
+\subsection{Ice thermodynamics}
 
@@
 
+SIA and SSA velocities are computed by finite difference methods on a 10km-resolution horizontal grid of 300 by 150 points.
+SIA velocities follow temperature-dependant creep.
+Temperatures are derived from an enthalpy formulation \citep{aschwanden-etal-2012} and computed in three dimensions on 51 irregularly spaced layers in ice and 11 regularly spaced layers in rock.
+An homogeneous geothermal heat flux of 70\,\unit{W\,m^{-2}} is provided as lower boundary condition to the thermal model, and surface air temperature from the various climate forcing is used as upper boundary condition.
+
 SSA velocities are determined by a a pseudo-plastic sliding law where the yield stress depends on the availability of basal water and a prescibed till friction angle$\phi$.
@@
 
-\subsection{Glacier bed interface}
-
-The glacier bed topography response to the ice load is computed by a regional isostasy model with flexure and relaxation \citep{lingle-clark-1985}\todo{check ref.}.
-To compute basal temperatures with higher accuracy, the thermal model is extended 1\,km into the bedrock.
-An homogeneous geothermal heat flux of 70\,\unit{W\,m^{-2}} is provided as basal condition to this model.
+The glacier bed topography responds to the ice load following a regional isostasy model with flexure and relaxation \citep{lingle-clark-1985}\todo{check ref.}.
 

### 2013-02-19 sections/model.tex
Numerics are now merged with thermodynamics

@@
 
-\subsection{Simulation design and numerical implementation}
+\subsection{Simulation design}
 
@@
 Each simulation starts from ice-free conditions and is run for 10,000 years, a time interval which we think to be representative of the rapid build-up of the last Cordilleran Ice Sheet from nearly ice-free condition to last glacial maximum extent \needref.
-
-Our computational domain consists of 10~km-resolution, 150 by 300 points horizontal grid.
-Equations that involve a vertical coordinate are solved on 51 irregularly-spaced layers in ice and 11 regularly spaced layers in rock\todo{Check PISM doc}.
-The simulations were run in parallel on 32 cores, on NSC's supercomputer triolith\needref.
+Our simulations were run in parallel on 32 cores at the Swedish National Supercomputing Center.

### 2013-02-19 sections/model.tex
Fancier todo notes

@@
 Monthly temperatures and precipitation rates are converted to surface accumulation and ablation using a temperature-index (positive degree-day) model.
-Accumulation is computed as the amount of precipitation that falls when temperature is below freezing.\askconstantine{Is this true?
+Accumulation is computed as the amount of precipitation that falls when temperature is below freezing.\ask{Is this true?
 What is the current threshold in stable 0.5?} Ablation is computed from the number of positive degree-days, defined as the integrand of temperatures above 0\,\degree\,C in one year.
@@
 
-\todo[inline]{
-	\textbf{Include the \citet{calov-greve-2011} integrand formula?}
+\ask[inline]{\emph{Should I include the \citet{calov-greve-2011} integrand formula?}
 
@@
 A positive degree-day model is a very crude approximation of physical processes occurring at a glacier's surface.
-Yet we believe that it captures regional patterns of accumulation and ablation resonable enough for our study if accurate climate data is used, as we will show further.\todo{Is this paragraph really needed?}
+Yet we believe that it captures regional patterns of accumulation and ablation resonable enough for our study if accurate climate data is used, as we will show further.\ask{Is this paragraph really needed?}
 
@@
 
-\todo[inline]{
-	\textbf{Include SSA+SIA \citet{bueler-brown-2009} formula?}
+\ask[inline]{\emph{Should I include the SSA+SIA \citet{bueler-brown-2009} formula?}
 
@@
 
-The glacier bed topography responds to the ice load following a regional isostasy model with flexure and relaxation \citep{lingle-clark-1985}\todo{check ref.}.
+The glacier bed topography responds to the ice load following a regional isostasy model with flexure and relaxation \citep{lingle-clark-1985}\todo{Check \citet{lingle-clark-1985}}.
 

### 2013-02-20 sections/results.tex
Added todonotes in results section figures.
Added section 4 results (first shot)

@@
 % ----------------------------------------------------------------------
+
+Using climate forcing from WordClim data and the NCEP/NCAR, ERA-Iterim, CFSR and NARR reanalyses described in section~\ref{sec:climate} and the ice sheet model described in section~\ref{sec:model}, we run simulations of glacial inception and growth of the Cordilleran ice sheet using different temperature offsets.
 
@@
 	\end{center}
-	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a climate 6\,\unit{\degree C} colder than present for each climate forcing.}
+	\todo[inline]{Change panel order to match climate section}
+	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a climate 6\degC colder than present for each climate forcing.}
 	\label{fig:cool06}
 \end{figure}
+
+Figure~\ref{fig:cool06} shows the effect of a 6\degC temperature offset on the six different climate forcings.
+It apperas that the magnitude of glaciation differs highly between forcing datasets.
+NCEP/NCAR and CFSR forcings produce large ice sheets covering parts of Yukon and Alaska which are thought to have remained unglaciated for at least several glacial cycles.
+\needref The smoothing of CFSR precipitation has limited effect on the resulting ice sheet geometry.
+The WorldClim forcing results to a glaciation mode limited to major mountain ranges, whereas ERA-Interim and NARR produce medium-sized ice sheets with a difference in shape.
 
@@
 	\end{center}
+	\todo[inline]{Change panel order to match climate section}
+	\qiong[inline]{Move ticks on the colorbar. The 1 is confusing.}
 	\caption{Extent of ice cover after 10\,kyr as a function of applied temperature offsets for each climate forcing.}
@@
 \end{figure}
+
+As we aim to model an ice sheet approaching LGM size, yet do not know how cold was the climate shaping it, we apply temperature offsets ranging from 2 to 9\degC for each climate forcing.
+Figure~\ref{fig:extent} shows the extent of ice cover at the end of each of theses 40~simulations, grouped by climate forcing.
+Again it is visible that different forcings lead to very different final ice cover, generally.
+Notably, for the lowest temperature offset of 2\degC, NCAR and CFSR simulatinos lead to a significant ice-sheet, whereas WorldClim, NARR and ERAI produce mountain ice caps, not more extensive than present for WorldClim.
+It should be noted that several of the NCAR and CFSR simulations producve oversised ice-sheets whose extent is primirily bounded by the model domain boundary conditions rather than any physical process.
+Where different climate forcings lead to similar ice sheet sizes, differences in pattern can be noted.
+For instance ERAI ice seets appear more northernly-centered than NARR ice sheets.
 
@@
 	\end{center}
+	\todo[inline]{Use black-and-white-proof markers}
+	\todo[inline]{Highlight the ``best'' runs.}
 	\caption{Total ice volume and glaciated area after 10\,kyr as a function of temperature offset and climate forcing.}
@@
 \end{figure}
+
+Figure~\ref{fig:ivolarea} shows the final ice volume and glaciated area at the end of each of the 48~simulations.
+Differences in ice sheet size can be quantified.
+ERAI and NARR lead to similar ice volumes and glaciated areas, yet patterns are different as shown by Figure~\ref{fig:extent}.
 
@@
 	\end{center}
+	\todo[inline]{Change panel order to match climate section}
 	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr using temperature offsets that lead to similar areas of ice cover for each climate forcing.}
@@
 \end{figure}
+
+To be able to compare runs of similar magnitudes, and as we consider temperature offset as an unknown in this study, we select for each climate forcing the simulation that lead to most realistic results in terms of glaciated area.
+This was done by using the LGM extent countour from \needref shown in Figure~\ref{fig:locmap}, which covers an area of XXX \todo{do the maths}.
+These so called ``best'' runs are presented in Figure~\ref{fig:best} along with the temperature offset values associated.

### 2013-02-20 sections/results.tex
Improved section 4 Results

--- 2013-02-12 sections/results.tex
@@
 	\end{center}
-	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a climate 6\,\unit{\degree C} colder than present for each climate forcing.}
+	\todo[inline]{Change panel order to match climate section}
+	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a climate 6\degC colder than present for each climate forcing.}
 	\label{fig:cool06}
 \end{figure}
+
+Figure~\ref{fig:cool06} shows the outcome simulations run with a 6\degC temperature offset for each of the six climate forcings.
+In all simulations, ice accumulate and glaciers form as a result of the artificially cooled climate.
+Yet the magnitude of glaciation appears very different from one climate forcing to the next.
+Whereas NCAR and CFSR forcings produce a large ice sheet that covers most of the model domain land, WorldClim, ERA-Interim and NARR forcings lead to more restrictive ice cover, bounded to mountain ice caps in the case of WorldClim data.
+The smoothing of CFSR precipitation data has limited visual effect on the resulting ice sheet geometry.
 
@@
 \end{figure}
+
+As we aim to model an ice sheet approaching LGM size, and consider temperature offset as an unknown parameter in this study, we ran simulations using offset values ranging from 2 to 9\degC for each climate forcing.
+
+Figure~\ref{fig:extent} shows the extent of ice cover at the end of each of these (48) simulations, grouped by climate forcing.
+It again visible that different forcings generally lead to very different final ice cover Notably, for the smallest temperature offset value of 2\degC that was used, NCAR and CFSR simulations lead to an ice-sheet, whereas WorldClim, NARR and ERAI produced local ice caps only, restricted to the Wrangell-St.~Elias mountains in the case of WorldClim, an area presently glaciated.
+Several of the NCAR and CFSR simulations produced oversized ice-sheets whose extent is primirily bounded by the model domain boundary conditions rather than physical processes.
+Differences in regional patterns can also be noted.
+For instance ERAI ice seets generally appear more northernly-centered than NARR ice sheets.
 
@@
 \end{figure}
+
+Final ice volume and final ice coverage at the end of all simulations is quantified in Figure~\ref{fig:ivolarea}.
+NCEP/NCAR and CFSR forcing lead to much larger and more extensive glaciation than ERAI, NARR and WorldClim focrings.
+ERAI and NARR lead to similar ice volumes and glaciated areas, yet regional patterns are different, as shown by Figure~\ref{fig:extent}.
 
@@
 \end{figure}
+
+For each climate forcing, we selected the simulation that lead to a glaciated area closest to geomorphological reconstruction.
+This was done using the LGM extent countour from \needref shown in Figure~\ref{fig:locmap}, which covers an area of XXX \todo{Compute geomorpho-based ice cover area}.
+These ``best'' runs are presented in Figure~\ref{fig:best} along with the associated emperature offset values.
+We observe than although similar in size, the selected ice sheets are different in shape and flow patterns.
+More particularly, NCEP/NCAR, CFSR and ERA-Interim ice sheets cover large territories in Yukon and Alaska that where proven to have remain ice-free for at least several glacial cycles.\needref

### 2013-02-20 cordillera-climate.tex
Renamed \missingref to \needref
Enlarge right margin for todonotes
Moved floats into text
Commented out empty sections
Added dumb conclusion
Moved references to local bibtex file
Added techincal note for Constantine
Added Qiong suggestions
Added coloured hyperlinks
Moved technical note to separate file
Improved surface mass-balance model description
Fancier todo notes
Added todonotes in results section figures.
Added section 4 results (first shot)
Improved section 4 Results
Starred figures and tables for two-columns output
Added title and abstract notes

@@
 % Copernicus final print
-%\documentclass[tc]{copernicus}
+%\documentclass[tc, noauthor]{copernicus}
 
@@
 % Extra packages
-%\usepackage{color}
-%\usepackage{hyperref}
+\usepackage[colorlinks,citecolor=blue]{hyperref}
+\usepackage{framed}
 
 % All floats at the end of the manuscripts
-\usepackage{endfloat}
+%\usepackage{endfloat}
 
-% Add to-do notes
+% Add to-do notes and enlarge right margin
 \usepackage{todonotes}
-\newcommand{\missingref}[0]{\todo[color=cyan]{Need ref.}}
+\usepackage[left=54pt, top=81pt, textwidth=398pt, marginparsep=11pt, marginparwidth=93pt, textheight=625pt, footskip=30pt]{geometry}
+\newcommand{\needref}[0]{\todo[color=cyan]{Reference needed}}
+\newcommand{\arjen}[2][]{\todo[color=lime,{#1}]{\textbf{Arjen suggests:} #2}}
+\newcommand{\constantine}[2][]{\todo[color=yellow,{#1}]{\textbf{Constantine suggests:} #2}}
+\newcommand{\qiong}[2][]{\todo[color=pink,{#1}]{\textbf{Qiong suggests:} #2}}
+\newcommand{\ask}[2][]{\todo[color=lightgray,{#1}]{\textbf{Julien asks:} #2}}
+
+% Personal shortcuts
+\newcommand{\degC}[0]{\,\unit{\degree C} }
 
@@
 \maketitle
+\ask[inline]{\emph{Should the title be changed?}
+
+	\emph{I am open for discussion. As I see it, this paper has two major lines of thought, maybe not so easily combined:
+	\begin{enumerate}
+		\item Choosing climate forcing for palæo-ice sheet modelling.
+This is the main point of the paper if only.
+I am choosing climate forcing for the rest of my PhD as far as I am concern.
+Choosing climate forcing for other palæo-ice sheet modelling exercises as the reader is (potentially) concerned.
+		\item Numerically modelling the Cordilleran ice sheet. This is a point in itself as it has not quite being done so far as far as I am aware of.
+	\end{enumerate}
+	So I tried my best to include both ideas in the title. Yet this is preliminary...}}
 
@@
 \begin{abstract}
+\todo{This abstract is preliminary. I will rewrite it when I am done with the rest}
 We present an ensemble of numerical simulations of the former Cordilleran ice sheet forced by present-day climatologies from five different datasets.
@@
 \input{sections/results}
-\input{sections/discuss}
+%\input{sections/discuss}
+\input{sections/technote}
 % ----------------------------------------------------------------------
@@
 \label{sec:conclusion}
+\emph{NARR is the best!}\todo{elaborate upon this...}
 % ----------------------------------------------------------------------
@@
 \bibliographystyle{copernicus}
-\bibliography{/home/julien/work/biblio/ref}
+\bibliography{references}
 \end{document}

### 2013-02-20 sections/climate.tex
Added Qiong suggestions
Starred figures and tables for two-columns output
Restructured climate section

@@
 % ----------------------------------------------------------------------
+
+\qiong[inline]{Include foreword. What is the forcing for?}
 
@@
 
-However, the density of weather stations used by WorldClim in the Northern American Cordillera is highly inhomogeneous, and if good coverage exists in the southern parts of our modelling domain, several hundred kilometers can separate nearby stations in the north \citep{data:worldclim}.
+\begin{figure*}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics[width=13cm]{cordillera-climate-temp}
+	\end{center}
+	\qiong[inline]{Use same order in text and figures}
+	\caption{Summer temperature maps from the five datasets used in this study and winter temperature map from the WorldClim dataset.}
+	\label{fig:temp}
+\end{figure*}
 
+Figure~\ref{fig:temp} shows the spatial distribution of winter (DJF) and summer (JJA) air surface temperatures in WorldClim.
+Temperatures generally decrease from south to north and regions further inland experience colder winters.
+It should be noted that the temperature gradient is much stronger in the winter than it is in the summer.
+At low elevations temperature get well above zero during the summer months over the entire modelling domain.
+In other words, there is a strong seasonality contrast beween coastal and inland regions, and regions where mean annual temperatures are well below freezing do experience warm summers.
+
+\begin{figure*}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics[width=13cm]{cordillera-climate-prec}
+	\end{center}
+	\qiong[inline]{Use same order in text and figures}
+	\caption{Winter precipitation rate maps from the five datasets used in this study and summer precipitation map from the WorldClim dataset.}
+	\label{fig:prec}
+\end{figure*}
+
+Figure~\ref{fig:prec} shows the spatial distribution of winter (DJF) and summer (JJA) precipitation rates in WorldClim.
+Coastal regions generally recieve much more precipitation than inland ones as a results of the continuous topographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascadia.
+In a glacier mass-balance view, this contrast is made even stronger by the difference in timing of the precipitation peak through the year.
+Although coastal regions experience most precipitation during the accumulation season, inland regions experience dry winters and most of the precipitation falls as rain during the summer months.
+
+In regions like Northern Yukon and Alaska, dry winters and warm summers are unfavourable conditions to accumulation and glacial inception despite of the strongly negative mean annual temperatures at present.
+In order to account for these strong gradients in seasonality, we use monthly means to drive the ice sheet model.
+Previous runs forced by annual means of temperature and precipitation not presented here systematically conducted to unrealistic ice build-up in northern Yukon and Alaska under present-day climate while leaving the Coast Mountains free of ice.
+
+The density of weather stations used by WorldClim in the Northern American Cordillera is highly inhomogeneous.
+If good coverage exists in the southern parts of our modelling domain, several hundred kilometers can separate nearby stations in the north \citep{data:worldclim}.
 A second drawback of the dataset in an ice sheet modelling view is the lack of data on marine surfaces, particularly on the Pacific continental shelf where glaciers have advanced during the last glacial cycle\needref.
@@
 
-In addition to the WorldClim data, we use surface air temperature and precipitation rate data from three global and one regional climate reanalysis to force the ice sheet model: the NCEP/NCAR reanalysis, the ERA-Interim reanalysis, the Climate System Forecast Reanalysis (CFSR) and the North American Regional Reanalysis (NARR).
-Monthly climatologies from NCEP/NCAR and NARR reanalysis were provided by the NOAA/OAR/ESRL PSD, Boulder, Colorado, USA, from their Web site at \url{http://www.esrl.noaa.gov/psd/} and monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean timeseries.
-Further information on the data used is gatherd in Table \ref{tab:reanalyses}.
-As a mixed product from observations and a circulation model, we believe that reanalysis may perform best in poorly monitored regions such as the northernmost American Cordillera.
-
-\begin{table}[t]
+\begin{table*}[t]
 	\caption{Characteristic of reanalysis climatologies used to force the ice sheet model.}
@@
 	\end{tabular}
-\end{table}
+\end{table*}
 
-% ----------------------------------------------------------------------
+In addition to the WorldClim data, we use surface air temperature and precipitation rate data from three global and one regional climate reanalyses to force the ice sheet model: the NCEP/NCAR reanalysis, the ERA-Interim reanalysis, the Climate System Forecast Reanalysis (CFSR) and the North American Regional Reanalysis (NARR).
+Monthly climatologies from NCEP/NCAR and NARR reanalysis were provided by the NOAA/OAR/ESRL PSD, Boulder, Colorado, USA, from their website at \url{http://www.esrl.noaa.gov/psd/} and monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean timeseries.
+Further information on the data used is gathered in Table \ref{tab:reanalyses}.
 
-\subsection{Cordilleran climates}
+The spatial distribution of summer (JJA) air surface temperatures and winter (DJF) precipitation rates from the four reanalyses datasets are shown in Figures~\ref{fig:temp} and~\ref{fig:prec}.
+Summer temperature and are winter precipitation are most relevant to the glacier model as they respectively drive summer melt and winter accumulation.
 
-\begin{figure}[t]
-	\vspace*{2mm}
-	\begin{center}
-		\includegraphics[width=13cm]{cordillera-climate-temp}
-	\end{center}
-	\caption{Summer temperature maps from the five datasets used in this study and winter temperature map from the WorldClim dataset.}
-	\label{fig:temp}
-\end{figure}
-
-Figure~\ref{fig:temp} shows the spatial distribution of summer (JJA) air surface temperatures from the five climate datasets used in the study, and winter (DJF) air surface temperatures from the WorldClim interpolated observation data.
-Summer temperatures are most relevant to the glacier model as they drive summer melt.
-
-Temperatures generally decrease from south to north and regions further inland experience colder winters.
-It should be noted that the temperature gradient is much stronger in the winter than it is in the summer.
-At low elevations temperature get well above zero during the summer months over the entire modelling domain.
-In other words, there is a strong seasonality contrast beween coastal and inland regions, and regions where mean annual temperatures are well below freezing do experience warm summers.
-
-\begin{figure}[t]
-	\vspace*{2mm}
-	\begin{center}
-		\includegraphics[width=13cm]{cordillera-climate-prec}
-	\end{center}
-	\caption{Winter precipitation rate maps from the five datasets used in this study and summer precipitation map from the WorldClim dataset.}
-	\label{fig:prec}
-\end{figure}
-
-Figure~\ref{fig:prec} shows the spatial distribution of winter (DJF) precipitation rates from the five climate datasets used in the study, and summer (JJA) precipitation rates from the WorldClim interpolated observation data.
-Winter precipitation is most relevant to the glacier model as it drives winter accumulation.
-
-Coastal regions generally recieve much more precipitation than inland ones as a results of the continuous topographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascadia.
-In a glacier mass-balance view, this contrast is made even stronger by the difference in timing of the precipitation peak through the year.
-Although coastal regions experience most precipitation during the accumulation season, inland regions experience dry winters and most of the precipitation falls as rain during the summer months.
-
-In regions like Northern Yukon and Alaska, dry winters and warm summers are unfavourable conditions to accumulation and glacial inception despite of the strongly negative mean annual temperatures at present.
-In order to account for these strong gradients in seasonality, we use monthly means to drive the ice sheet model.
-Previous runs forced by annual means of temperature and precipitation not presented here systematically conducted to unrealistic ice build-up in northern Yukon and Alaska under present-day climate while leaving the Coast Mountains free of ice.
+As a mixed product from observations and a circulation model, we believe that reanalysis may perform best in poorly monitored regions such as the northernmost American Cordillera.
 
@@
 
-\begin{figure}[t]
+\begin{figure*}[t]
 	\vspace*{2mm}
@@
 	\end{center}
+	\qiong[inline]{Use same order in text and figures}
 	\caption{Reference topography used for temperature lapse-rate corrections from the five climate datasets used in the study and ETOPO1 topography used as basal condition for the ice flow model.}
 	\label{fig:topo}
-\end{figure}
+\end{figure*}
 

### 2013-02-21 sections/climate.tex
Added climate section foreword
Removed figure todonotes (update coming soon)
Fixed typos, language and style

@@
 
-\qiong[inline]{Include foreword. What is the forcing for?}
+We force an ice sheet model with temperature and precipitation data from one observational dataset, three global and one regional reanalyses.
+Here we present this climate data and describe how we use it as a forcing.
 
@@
 
-WorldClim \citep{data:worldclim} is a high resolution climate dataset built from observation data.
+WorldClim \citep{data:worldclim} is a high-resolution climate dataset built from observation data.
 These climate surfaces were generated by spatial interpolation between a large set of weather stations using a 30\,arc\,s-aggregated flavour of hole-filled SRTM data over global land areas.
@@
 	\end{center}
-	\qiong[inline]{Use same order in text and figures}
 	\caption{Summer temperature maps from the five datasets used in this study and winter temperature map from the WorldClim dataset.}
@@
 
-Figure~\ref{fig:temp} shows the spatial distribution of winter (DJF) and summer (JJA) air surface temperatures in WorldClim.
+Figure~\ref{fig:temp} shows the spatial distribution of summer (JJA) and winter (DJF) air surface temperatures in WorldClim.
 Temperatures generally decrease from south to north and regions further inland experience colder winters.
 It should be noted that the temperature gradient is much stronger in the winter than it is in the summer.
-At low elevations temperature get well above zero during the summer months over the entire modelling domain.
-In other words, there is a strong seasonality contrast beween coastal and inland regions, and regions where mean annual temperatures are well below freezing do experience warm summers.
+At low elevations, temperature get well above zero during the summer months over the entire modelling domain.
+In other words, there is a strong contrast in seasonality between coastal and inland regions, and regions where mean annual temperatures are well below freezing do experience warm summers.
 
@@
 	\end{center}
-	\qiong[inline]{Use same order in text and figures}
 	\caption{Winter precipitation rate maps from the five datasets used in this study and summer precipitation map from the WorldClim dataset.}
@@
 Figure~\ref{fig:prec} shows the spatial distribution of winter (DJF) and summer (JJA) precipitation rates in WorldClim.
-Coastal regions generally recieve much more precipitation than inland ones as a results of the continuous topographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascadia.
+Coastal regions generally receive much more precipitation than inland ones as a result of the continuous topographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascades.
 In a glacier mass-balance view, this contrast is made even stronger by the difference in timing of the precipitation peak through the year.
@@
 
-In regions like Northern Yukon and Alaska, dry winters and warm summers are unfavourable conditions to accumulation and glacial inception despite of the strongly negative mean annual temperatures at present.
+In regions like Northern Yukon and Alaska, dry winters and warm summers are unfavourable conditions to accumulation and glacial inception despite of the strongly negative present mean annual temperatures.
 In order to account for these strong gradients in seasonality, we use monthly means to drive the ice sheet model.
-Previous runs forced by annual means of temperature and precipitation not presented here systematically conducted to unrealistic ice build-up in northern Yukon and Alaska under present-day climate while leaving the Coast Mountains free of ice.
+Previous runs forced by annual means of temperature and precipitation, not presented here systematically conducted to unrealistic ice build-up in northern Yukon and Alaska under present-day climate while leaving the Coast Mountains free of ice.
 
 The density of weather stations used by WorldClim in the Northern American Cordillera is highly inhomogeneous.
-If good coverage exists in the southern parts of our modelling domain, several hundred kilometers can separate nearby stations in the north \citep{data:worldclim}.
+If good coverage exists in the southern parts of our modelling domain, several hundred kilometres can separate nearby stations in the north \citep{data:worldclim}.
 A second drawback of the dataset in an ice sheet modelling view is the lack of data on marine surfaces, particularly on the Pacific continental shelf where glaciers have advanced during the last glacial cycle\needref.
@@
 
-In addition to the WorldClim data, we use surface air temperature and precipitation rate data from three global and one regional climate reanalyses to force the ice sheet model: the NCEP/NCAR reanalysis, the ERA-Interim reanalysis, the Climate System Forecast Reanalysis (CFSR) and the North American Regional Reanalysis (NARR).
-Monthly climatologies from NCEP/NCAR and NARR reanalysis were provided by the NOAA/OAR/ESRL PSD, Boulder, Colorado, USA, from their website at \url{http://www.esrl.noaa.gov/psd/} and monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean timeseries.
+Aside WorldClim data, we use surface air temperatures and precipitation rates from three global and one regional climate reanalyses to force the ice sheet model: the NCEP/NCAR reanalysis, the ERA-Interim reanalysis, the Climate System Forecast Reanalysis (CFSR) and the North American Regional Reanalysis (NARR).
+Monthly climatologies from NCEP/NCAR and NARR reanalysis were provided by the NOAA/OAR/ESRL PSD, Boulder, Colorado, USA, from their website at \url{http://www.esrl.noaa.gov/psd/} whereas monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean time series.
 Further information on the data used is gathered in Table \ref{tab:reanalyses}.
@@
 The spatial distribution of summer (JJA) air surface temperatures and winter (DJF) precipitation rates from the four reanalyses datasets are shown in Figures~\ref{fig:temp} and~\ref{fig:prec}.
-Summer temperature and are winter precipitation are most relevant to the glacier model as they respectively drive summer melt and winter accumulation.
+Summer temperature and winter precipitation are most relevant to the glacier model as they respectively drive summer melt and winter accumulation.
 
@@
 
-Air surface temperature and precipitation rates from the NCEP/NCAR, ERA-Interim, CFSR and NARR reanalyses and WorldClim data were reprojected to Canadian Atlas Lambert conformal conic projection (EPSG code~3978) and bilinearly interpolated to the model resolution of 10\,km.
+Air surface temperature and precipitation rates from the WorldClim data, the NCEP/NCAR, ERA-Interim, CFSR and NARR reanalyses were re-projected to Canadian Atlas Lambert conformal conic projection (EPSG code~3978) and bilinearly interpolated to the model resolution of 10\,km.
 In addition, WorldClim data was extrapolated to marine surfaces using a nearest-neighbour approach.
@@
 
-For the CFSR, an alternative forcing was prepared by smoothing the precipitation field in order to correct for artifacts seen in figure~\ref{fig:prec}.
-This was done by averaging data locally in a circular neighbourhood of 7 pixels in diameter prior to reprojection.
+For the CFSR, an alternative forcing was prepared by smoothing the precipitation field in order to correct for artefacts seen in figure~\ref{fig:prec}.
+This was done by averaging data locally in a circular neighbourhood of 7\,pixels in diameter prior to re-projection.
 Yet we will show that this smoothing has very limited effect on the ice sheet model outcome.
-All the preprocessing steps described above were made in GRASS~GIS using scripts available on the first author's website.
+This preprocessing steps were made in GRASS~GIS using scripts available on the first author's website.
 
@@
 	\end{center}
-	\qiong[inline]{Use same order in text and figures}
 	\caption{Reference topography used for temperature lapse-rate corrections from the five climate datasets used in the study and ETOPO1 topography used as basal condition for the ice flow model.}
@@
 The reference topographies from the five forcing datasets, alongside the ETOPO1 data are shown in figure~\ref{fig:topo}.
-A lapse-rate of 6\unit{\degree C\,km^{-1}} is used in all simulations and no lapse-rate corrections apply to precipitation rates.
+A lapse-rate of 6\unit{\degree C\,km^{-1}} is used in all simulations.
+No lapse-rate corrections apply to precipitation rates.

### 2013-02-21 sections/model.tex
Added section 4 results (first shot)
Fixed typos, language and style

@@
 \section{Model setup}
-\label{sec:setup}
+\label{sec:model}
 % ----------------------------------------------------------------------
@@
 We use PISM, a parallel ice sheet model (version stable 0.5.11).
-Documentation on the code can be found in several publications \citep[e.g.,][]{bueler-brown-2009,winkelmann-etal-2011,aschwanden-etal-2012} and more extensively online \citep[{\url{http://www.pism-docs.org}},][]{web:pism}.
+Documentation on the code can be found in several publications \citep[e.g.,][]{bueler-brown-2009,winkelmann-etal-2011,aschwanden-etal-2012} and more extensively on-line \citep[{\url{http://www.pism-docs.org}},][]{web:pism}.
 
@@
 
-Monthly temperatures and precipitation rates are converted to surface accumulation and ablation using a temperature-index (positive degree-day) model.
+Monthly temperature and precipitation rate are converted to surface accumulation and ablation by a temperature-index (positive degree-day) model.
 Accumulation is computed as the amount of precipitation that falls when temperature is below freezing.\ask{Is this true?
 What is the current threshold in stable 0.5?} Ablation is computed from the number of positive degree-days, defined as the integrand of temperatures above 0\,\degree\,C in one year.
-Positive degree-days are computed by the \citet{calov-greve-2011} formula to account for day-to-day variability unrepresented in the monthly climate forcing.
+Positive degree-days are computed by the \citet{calov-greve-2011} formula to account for day-to-day variability, which is not represented in the monthly climate forcing.
 The ablation model use different degree-day factors for snow and ice and accounts for refreezing.
-Complete description of the ablation model is available in PISM documentation \citep[{\url{http://www.pism-docs.org}},][]{web:pism} and model parameters are explicited in table~\ref{tab:pdd}.
+Complete description of the ablation model is available in PISM documentation \citep[{\url{http://www.pism-docs.org}},][]{web:pism} and model parameters are expressed in table~\ref{tab:pdd}.
 
@@
 
-	The number of positive degree-days, defined as the integrand of positive temperatures over time, is computed annually using a formula derived by \citet{calov-greve-2011} that accounts for day-to-day variability of temperature by assuming a normal distribution of variance $\sigma=5\degree C$ along a central (input) value $T$:
+	The number of positive degree-days is computed annually using a formula derived by \citet{calov-greve-2011} that accounts for day-to-day variability of temperature by assuming a normal distribution of variance $\sigma=5\degree C$ along a central (input) value $T$:
 
@@
 A positive degree-day model is a very crude approximation of physical processes occurring at a glacier's surface.
-Yet we believe that it captures regional patterns of accumulation and ablation resonable enough for our study if accurate climate data is used, as we will show further.\ask{Is this paragraph really needed?}
+Yet we believe that it captures regional patterns of accumulation and ablation that are reasonable enough for our study if accurate climate data is used, as we will show further.\ask{Is this paragraph really needed?}
 
@@
 Temperatures are derived from an enthalpy formulation \citep{aschwanden-etal-2012} and computed in three dimensions on 51 irregularly spaced layers in ice and 11 regularly spaced layers in rock.
-An homogeneous geothermal heat flux of 70\,\unit{W\,m^{-2}} is provided as lower boundary condition to the thermal model, and surface air temperature from the various climate forcing is used as upper boundary condition.
+An homogeneous geothermal heat flux of 70\,\unit{W\,m^{-2}} is provided as lower boundary condition to the thermal model, and the surface air temperature from the various climate forcing is used as upper boundary condition.
 
-SSA velocities are determined by a a pseudo-plastic sliding law where the yield stress depends on the availability of basal water and a prescibed till friction angle$\phi$.
-We apply lower friction angles at low elevation to mimick the more slippery beds associated to the presence of marine sediments:
+SSA velocities are determined by a pseudo-plastic sliding law where the yield stress depends on the availability of basal water and a prescribed till friction angle$\phi$.
+We apply lower friction angles at low elevation to mimic the more slippery beds associated to the presence of marine sediments:
 
@@
 In this study we model glacial inception and growth of the Cordilleran ice sheet to a state approaching its last glacial maximum.
-We mimick glacial conditions by applying constant temperature offsets homogeneously over the entire domain.
-Each simulation starts from ice-free conditions and is run for 10,000 years, a time interval which we think to be representative of the rapid build-up of the last Cordilleran Ice Sheet from nearly ice-free condition to last glacial maximum extent \needref.
+We mimic glacial conditions by applying constant temperature offsets homogeneously over the entire domain.
+Each simulation starts from ice-free conditions and is run for 10,000 years, a time interval representative of the rapid build-up of the last Cordilleran Ice Sheet from nearly ice-free condition to last glacial maximum extent \needref.
 Our simulations were run in parallel on 32 cores at the Swedish National Supercomputing Center.

### 2013-02-21 sections/results.tex
Starred figures and tables for two-columns output
Added more todonotes in figures
Removed figure todonotes (update coming soon)
Fixed typos, language and style

--- 2013-02-12 sections/results.tex
@@
 
-\begin{figure}[t]
+Using climate forcing from WordClim data and the NCEP/NCAR, ERA-Iterim, CFSR and NARR reanalyses described in section~\ref{sec:climate} and the ice sheet model described in section~\ref{sec:model}, we run simulations of glacial inception and growth of the Cordilleran ice sheet using different temperature offsets.
+
+\begin{figure*}[t]
 	\vspace*{2mm}
@@
 	\end{center}
-	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a climate 6\,\unit{\degree C} colder than present for each climate forcing.}
+	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a climate 6\degC colder than present for each climate forcing.}
 	\label{fig:cool06}
-\end{figure}
+\end{figure*}
 
-\begin{figure}[t]
+Figure~\ref{fig:cool06} shows the outcome of simulations run with a 6\degC temperature offset for each of the 6 climate forcings.
+In all simulations, ice accumulates and glaciers form as a result of the artificially cooled climate.
+Yet the magnitude of glaciation appears very different from one climate forcing to another.
+Whereas the NCAR and CFSR forcings produce a large ice sheet that covers most of the model domain land, the WorldClim, ERA-Interim and NARR forcings lead to more restrictive ice cover, bounded to mountain ice caps in the case of WorldClim data.
+The smoothing of CFSR precipitation data has limited visual effect on the resulting ice sheet geometry.
+
+\begin{figure*}[t]
 	\vspace*{2mm}
@@
 	\end{center}
+	\qiong[inline]{Move ticks on the colorbar. The 1 is confusing.}
 	\caption{Extent of ice cover after 10\,kyr as a function of applied temperature offsets for each climate forcing.}
 	\label{fig:extent}
-\end{figure}
+\end{figure*}
+
+As we aim to model an ice sheet approaching LGM size, and consider temperature offset as an unknown parameter in this study, we ran simulations using offset values ranging from 2 to 9\degC for each climate forcing.
+
+Figure~\ref{fig:extent} shows the extent of ice cover at the end of each of these~(48) simulations, grouped by climate forcing.
+It is again visible that different forcings generally lead to very different final ice cover.
+Notably, for the smallest temperature offset value of 2\degC that was used, NCAR and CFSR simulations lead to an ice-sheet, whereas WorldClim, NARR and ERA-Interim produced local ice caps only, restricted to the Wrangell-St.~Elias mountains in the case of WorldClim, an area presently glaciated.
+Several of the NCAR and CFSR simulations produced oversized ice-sheets whose extent is primirily bounded by the model domain boundary conditions rather than physical processes.
+Differences in regional patterns can also be noted.
+For instance ERA-Interim ice sheets generally appear more northernly-centered than NARR ice sheets.
 
@@
 	\end{center}
+	\todo[inline]{Use black-and-white-proof markers and highlight the ``best'' runs.}
 	\caption{Total ice volume and glaciated area after 10\,kyr as a function of temperature offset and climate forcing.}
@@
 
-\begin{figure}[t]
+Final ice volume and final ice coverage at the end of all simulations is quantified in Figure~\ref{fig:ivolarea}.
+The NCEP/NCAR and CFSR forcing lead to a much larger and more extensive glaciation than the ERA-Interim, NARR and WorldClim forcings.
+The ERA-Interim and NARR lead to similar ice volumes and glaciated areas, yet regional patterns are different, as shown by Figure~\ref{fig:extent}.
+
+\begin{figure*}[t]
 	\vspace*{2mm}
@@
 	\end{center}
+	\todo[inline]{The must would be to overlay geomorphic ice extent.}
 	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr using temperature offsets that lead to similar areas of ice cover for each climate forcing.}
 	\label{fig:best}
-\end{figure}
+\end{figure*}
+
+For each climate forcing, we selected the simulation that lead to a glaciated area closest to geomorphological reconstruction.
+This was done using the LGM extent countour from \needref shown in Figure~\ref{fig:locmap}, which covers an area of XXX \todo{Compute geomorpho-based ice cover area}.
+These ``best'' runs are presented in Figure~\ref{fig:best} along with the associated temperature offset values.
+We observe that although similar in size, the selected ice sheets are different in shape and flow patterns.
+More particularly, NCEP/NCAR, CFSR and ERA-Interim ice sheets cover large territories in Yukon and Alaska that where shown to have remained ice-free for at least several glacial cycles.\needref

### 2013-02-21 sections/results.tex
Uneasy area calculation becomes arbitrary choice

@@
 
-For each climate forcing, we selected the simulation that lead to a glaciated area closest to geomorphological reconstruction.
-This was done using the LGM extent countour from \needref shown in Figure~\ref{fig:locmap}, which covers an area of XXX \todo{Compute geomorpho-based ice cover area}.
-These ``best'' runs are presented in Figure~\ref{fig:best} along with the associated temperature offset values.
+To compare modelled ice sheets of similar size, we selected for each climate forcing the simulation that lead to a glaciated area closest to an arbitrary target of $4\,\times10^6\,\unit{km^2}$.
+These qualitative ``best'' runs are presented in Figure~\ref{fig:best} along the associated temperature offset values.
 We observe that although similar in size, the selected ice sheets are different in shape and flow patterns.
-More particularly, NCEP/NCAR, CFSR and ERA-Interim ice sheets cover large territories in Yukon and Alaska that where shown to have remained ice-free for at least several glacial cycles.\needref
+More specifically, NCEP/NCAR, CFSR and ERA-Interim ice sheets cover large territories in Yukon and Alaska that where shown to have remained ice-free for at least several glacial cycles.\needref

### 2013-02-21 sections/results.tex
Fixed a typo

@@
 
-Using climate forcing from WordClim data and the NCEP/NCAR, ERA-Iterim, CFSR and NARR reanalyses described in section~\ref{sec:climate} and the ice sheet model described in section~\ref{sec:model}, we run simulations of glacial inception and growth of the Cordilleran ice sheet using different temperature offsets.
+Using climate forcing from WorldClim data, the NCEP/NCAR, ERA-Iterim, CFSR and NARR reanalyses described in section~\ref{sec:climate} and the ice sheet model described in section~\ref{sec:model}, we run simulations of glacial inception and growth of the Cordilleran ice sheet using different temperature offsets.
 

### 2013-03-21 sections/climate.tex
Implement Arjen comments
Trivial comments implemented in the text; others as todo notes.

@@
 
-We force an ice sheet model with temperature and precipitation data from one observational dataset, three global and one regional reanalyses.
+We force PISM with temperature and precipitation data from one observational dataset, three global and one regional reanalyses.
 Here we present this climate data and describe how we use it as a forcing.
@@
 
-WorldClim \citep{data:worldclim} is a high-resolution climate dataset built from observation data.
-These climate surfaces were generated by spatial interpolation between a large set of weather stations using a 30\,arc\,s-aggregated flavour of hole-filled SRTM data over global land areas.
+WorldClim is a high-resolution climate dataset built from observation data \citep{data:worldclim}.
+These climate surfaces were generated by spatial interpolation between a large set of weather stations over global land areas, using 30\,arc\,s-aggregated hole-filled SRTM elevation data.
 This provides a resolution much higher than attained by general circulation models.
@@
 It should be noted that the temperature gradient is much stronger in the winter than it is in the summer.
-At low elevations, temperature get well above zero during the summer months over the entire modelling domain.
-In other words, there is a strong contrast in seasonality between coastal and inland regions, and regions where mean annual temperatures are well below freezing do experience warm summers.
+Temperature reach well above zero during the summer months over the entire modelling domain, except of the highest peaks.
+In other words, there is a strong contrast in seasonality between coastal and inland regions, even and regions where mean annual temperatures are well below freezing do experience warm summers.
 
@@
 
-Figure~\ref{fig:prec} shows the spatial distribution of winter (DJF) and summer (JJA) precipitation rates in WorldClim.
-Coastal regions generally receive much more precipitation than inland ones as a result of the continuous topographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascades.
-In a glacier mass-balance view, this contrast is made even stronger by the difference in timing of the precipitation peak through the year.
-Although coastal regions experience most precipitation during the accumulation season, inland regions experience dry winters and most of the precipitation falls as rain during the summer months.
+\arjen{It is an unusual style to start each sentence with Figure X shows...
+We should rephrase this later-on.} Figure~\ref{fig:prec} shows the spatial distribution of DJF and JJA precipitation rates in WorldClim.
+Coastal regions generally receive much more precipitation than inland ones as a result of the continuous topographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascades\arjen{orographic precip!}.
+From a glacier mass-balance point of view, this contrast is made even stronger by the difference in timing of the precipitation peak through the year.
+Although coastal regions experience most precipitation as snow during the accumulation season, inland regions experience dry winters and most of the precipitation falls as rain during the summer months.
 
-In regions like Northern Yukon and Alaska, dry winters and warm summers are unfavourable conditions to accumulation and glacial inception despite of the strongly negative present mean annual temperatures.
-In order to account for these strong gradients in seasonality, we use monthly means to drive the ice sheet model.
-Previous runs forced by annual means of temperature and precipitation, not presented here systematically conducted to unrealistic ice build-up in northern Yukon and Alaska under present-day climate while leaving the Coast Mountains free of ice.
+In regions like Northern Yukon and Alaska, dry winters and warm summers are unfavourable conditions to ice accumulation and glacial inception despite of the strongly negative present mean annual temperatures.
+In order to account for these strong gradients in seasonality, we use monthly averages of temperature and precipitation to drive the ice sheet model.
 
 The density of weather stations used by WorldClim in the Northern American Cordillera is highly inhomogeneous.
-If good coverage exists in the southern parts of our modelling domain, several hundred kilometres can separate nearby stations in the north \citep{data:worldclim}.
-A second drawback of the dataset in an ice sheet modelling view is the lack of data on marine surfaces, particularly on the Pacific continental shelf where glaciers have advanced during the last glacial cycle\needref.
+If good coverage exists in the southern parts of our modelling domain, several hundred kilometres can separate neighbouring stations in the north \citep{data:worldclim}.
+A second drawback of the dataset in the scope of this study is the lack of data across oceans, particularly on the Pacific continental shelf where glaciers have advanced during the last glacial cycle\needref.
 
@@
 
-Aside WorldClim data, we use surface air temperatures and precipitation rates from three global and one regional climate reanalyses to force the ice sheet model: the NCEP/NCAR reanalysis, the ERA-Interim reanalysis, the Climate System Forecast Reanalysis (CFSR) and the North American Regional Reanalysis (NARR).
-Monthly climatologies from NCEP/NCAR and NARR reanalysis were provided by the NOAA/OAR/ESRL PSD, Boulder, Colorado, USA, from their website at \url{http://www.esrl.noaa.gov/psd/} whereas monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean time series.
+Apart from the WorldClim data, we use surface air temperatures and precipitation rates from three global and one regional climate reanalyses to force the ice sheet model: the NCEP/NCAR reanalysis, the ERA-Interim reanalysis, the Climate System Forecast Reanalysis (CFSR) and the North American Regional Reanalysis (NARR).
+Monthly climatologies from NCEP/NCAR and NARR reanalysis were provided by the NOAA/OAR/ESRL Physical Sciences Division, Boulder, Colorado, USA, from their website at \url{http://www.esrl.noaa.gov/psd/} whereas monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean time series.
 Further information on the data used is gathered in Table \ref{tab:reanalyses}.
 
-The spatial distribution of summer (JJA) air surface temperatures and winter (DJF) precipitation rates from the four reanalyses datasets are shown in Figures~\ref{fig:temp} and~\ref{fig:prec}.
+The spatial distribution of JJA air surface temperatures and DJF precipitation rates from the four reanalyses datasets are shown in Figures~\ref{fig:temp} and~\ref{fig:prec}.
 Summer temperature and winter precipitation are most relevant to the glacier model as they respectively drive summer melt and winter accumulation.
 
-As a mixed product from observations and a circulation model, we believe that reanalysis may perform best in poorly monitored regions such as the northernmost American Cordillera.
+As a mixed product from observations and circulation modelling, we believe that reanalysis may perform best in sparsely monitored regions such as the northernmost American Cordillera.
 
@@
 
-Air surface temperature and precipitation rates from the WorldClim data, the NCEP/NCAR, ERA-Interim, CFSR and NARR reanalyses were re-projected to Canadian Atlas Lambert conformal conic projection (EPSG code~3978) and bilinearly interpolated to the model resolution of 10\,km.
-In addition, WorldClim data was extrapolated to marine surfaces using a nearest-neighbour approach.
+Air surface temperature and precipitation rates \arjen{do you really mean rate as in mm/time unit?} from the WorldClim data, the NCEP/NCAR, ERA-Interim, CFSR and NARR reanalyses were re-projected to Canadian Atlas Lambert conformal conic projection (EPSG code~3978) \arjen{is this general knowledge?
+I am lost...} and bilinearly interpolated to the model resolution of 10\,km.
+In addition, WorldClim data was extrapolated to oceans using a nearest-neighbour approach.
 These last two steps are not presented on figures~\ref{fig:temp} and~\ref{fig:prec}.
@@
 Yet we will show that this smoothing has very limited effect on the ice sheet model outcome.
-This preprocessing steps were made in GRASS~GIS using scripts available on the first author's website.
+This preprocessing steps were made in GRASS~GIS using scripts available on the first author's website\arjen{I am unsure if this is the best way t do it?
+Were you using principles described by others then you should refer to the litterature instead.}.
 

### 2013-03-21 sections/intro.tex
Empty sections coming soon...
Added more todonotes in figures
Provided vector location map instead of bitmap
Fixed typos, language and style
Implement Arjen comments
Trivial comments implemented in the text; others as todo notes.

@@
 	\end{center}
-	\caption{Shaded relief map of northern North America. The frame delimits the modelling domain. After ETOPO1 \citep{data:etopo1} and Natural Earth Data \citep{data:naturalearth}.}
+	\caption{Shaded relief map of northern North America.
+The frame delimits this study's modelling domain.
+The outlines of the former ice sheets are from \citet{kleman-etal-2010}.
+The background map was made with ETOPO1 \citep{data:etopo1} and Natural Earth Data \citep{data:naturalearth}.}
 	\label{fig:locmap}
 \end{figure}
+
+\emph{Coming soon...}
+
+% Define PISM = Parallel Ice Sheet Model

### 2013-03-21 sections/model.tex
Implement Arjen comments
Trivial comments implemented in the text; others as todo notes.

@@
 \begin{table}[t]
-	\caption{Values of the positive degree-day model parameters.}
+	\caption{Positive degree-day model parameters.}
 	\label{tab:pdd}
@@
 
-Monthly temperature and precipitation rate are converted to surface accumulation and ablation by a temperature-index (positive degree-day) model.
-Accumulation is computed as the amount of precipitation that falls when temperature is below freezing.\ask{Is this true?
-What is the current threshold in stable 0.5?} Ablation is computed from the number of positive degree-days, defined as the integrand of temperatures above 0\,\degree\,C in one year.
-Positive degree-days are computed by the \citet{calov-greve-2011} formula to account for day-to-day variability, which is not represented in the monthly climate forcing.
-The ablation model use different degree-day factors for snow and ice and accounts for refreezing.
-Complete description of the ablation model is available in PISM documentation \citep[{\url{http://www.pism-docs.org}},][]{web:pism} and model parameters are expressed in table~\ref{tab:pdd}.
-
-\ask[inline]{\emph{Should I include the \citet{calov-greve-2011} integrand formula?}
-
-	The number of positive degree-days is computed annually using a formula derived by \citet{calov-greve-2011} that accounts for day-to-day variability of temperature by assuming a normal distribution of variance $\sigma=5\degree C$ along a central (input) value $T$:
+Monthly temperature and precipitation are converted to surface accumulation and ablation by a temperature-index (positive degree-day) model.
+Accumulation is computed as the amount of precipitation that falls when temperature is below freezing\ask{Is this true?
+What is the current threshold in stable 0.5?}.
+Ablation is computed from the number of positive degree-days, defined as the integral of temperatures above 0\,\degree\,C in one year.
+Positive degree-days are calculated by the \citet{calov-greve-2011} formula to accounts for day-to-day temperature variability by assuming a normal distribution of variance $\sigma=5\degree C$ along a central (input) value $T$:
 
@@
 
-Where $[t_0; t_1]$ is one year and $\mathrm{erfc}$ is the complementary error function.}
+Where $[t_0; t_1]$ is one year and $\mathrm{erfc}$ is the complementary error function.
+The ablation model employs different degree-day factors for snow and ice and accounts for refreezing (table~\ref{tab:pdd}).
 
-A positive degree-day model is a very crude approximation of physical processes occurring at a glacier's surface.
-Yet we believe that it captures regional patterns of accumulation and ablation that are reasonable enough for our study if accurate climate data is used, as we will show further.\ask{Is this paragraph really needed?}
+A positive degree-day model is a crude parametrization of physical processes occurring at a glacier's surface.
+Yet we believe that it captures reasonably accurate patterns of accumulation and ablation for our study area considering the employed climate data.
 
@@
 
-Ice flow velocities are computed by a heuristic where the Shallow Shelf Approximation (SSA) is used as a ``sliding law'' to a polythermal Shallow Ice Approximation (SIA) \citep{bueler-brown-2009}.
+Ice flow velocities are computed by a heuristic \arjen{not sure what this expresses!!} where the Shallow Shelf Approximation (SSA) is used as a ``sliding law'' to a polythermal Shallow Ice Approximation (SIA) \citep{bueler-brown-2009}.
 
@@
 
+\arjen[inline]{unless variations in these determine your best fit input dataset, I would not include them in this paper.}
+
 SIA and SSA velocities are computed by finite difference methods on a 10km-resolution horizontal grid of 300 by 150 points.
@@
 \begin{equation}
-	\phi = \left\{\begin{array}{llc}
-		10\,\unit{\degree C} & \mathrm{for} &               z<  0\,\unit{m} \\
-		z/10 + 10°           & \mathrm{for} &   0\,\unit{m}<z<200\,\unit{m} \\
-		30\,\unit{\degree C} & \mathrm{for} & 200\,\unit{m}<z               \\
+	\phi = \left\{\begin{array}{llrll}
+		10      & \mathrm{for} &      &z&<  0 \\
+		z/10+10 & \mathrm{for} &   0 <&z&<200 \\
+		30      & \mathrm{for} & 200 <&z&     \\
 	\end{array}\right.
@@
 
+With $\phi$ in degrees and $z$ in meters above present sea level.
 The glacier bed topography responds to the ice load following a regional isostasy model with flexure and relaxation \citep{lingle-clark-1985}\todo{Check \citet{lingle-clark-1985}}.
@@
 Our simulations were run in parallel on 32 cores at the Swedish National Supercomputing Center.
+
+\arjen[inline]{Check Stroeven et al.
+(2010): "Radiocarbon ages from wood and fossils underlying McConnell till indicate that the McConnell glaciation began after 29,600  300 14C yr BP (Matthews et al., 1990), but glaciers did not advance into major valleys before 26,350  280 14C yr BP (Jackson and Harington, 1991), and a fully developed CIS did not exist until after 23,900  1140 14C yr BP (Lowdon and Blake,1981).
+Published radiocarbon ages indicate that the St.
+Elias lobe of the CIS was in retreat by 13,660  180 14C yr BP (Rampton, 1971) and the Selwyn lobe by 12,590  120 14C yr BP (Ward, 1989) (Fig.
+2)." and references therein.}

### 2013-03-21 sections/results.tex
Implement Arjen comments
Trivial comments implemented in the text; others as todo notes.

@@
 	\label{fig:cool06}
+  \arjen[inline]{Why do you inconsistently use K and degrees C?}
 \end{figure*}
@@
 	\qiong[inline]{Move ticks on the colorbar. The 1 is confusing.}
+  \arjen[inline]{I think just removing "1" is enough. No other number are in need of moving I think.}
 	\caption{Extent of ice cover after 10\,kyr as a function of applied temperature offsets for each climate forcing.}
@@
 	\todo[inline]{Use black-and-white-proof markers and highlight the ``best'' runs.}
+  \arjen[inline]{The shape of these are modulated by the modelling domain.
+It looks as if the curves are at maximum difference for 5-6 degrees C and are closer again 7-9 degrees C, but perhaps the latter is an artefact from the fact that the ice sheets cannot grow beyond the domain boundaries?}
 	\caption{Total ice volume and glaciated area after 10\,kyr as a function of temperature offset and climate forcing.}

### 2013-06-04 sections/intro.tex
Add introduction first shot
Or a ridiculous uninspired fifteen lines.

@@
 % ----------------------------------------------------------------------
+
+The numerical modelling of past glaciers and ice sheets allows indirect confrontation between glaciological theories and geomorphological reconstructions, challenging the validity of both.
+In this perspective, the Cordilleran ice sheet in north-western North America is one of the least studied paleo-ice sheets on Earth, where nethertheless significant geomorphological data is available\needref.
+
+To our knowledge, the Cordilleran ice sheet has not been the central target of a numerical modelling study since the one by \citep{roberts-1991}, who focused specifically on its southern margin.
+However, the Cordilleran ice sheet is often modelled as part of the North-american ice sheet complex\needref, or even of a larger-scale Arctic domain\needref.
+If these studies reproduce well the magnitude of the North American glaciation at Last Glacial Maximum, they commonly predict excessive ice cover in northern Yukon Territory and Alaska, regions that have not been covered by ice in hundreds of thousands of years\needref.
+
+Here we use a Parallel Ice Sheet Model (PISM) to model the entire Cordilleran ice sheet at the Last Glacial Maximum.
+We force our model with multiple climate datasets and compare our results to the mapped LGM ice sheet margin.
+To avoid circular dependance on prescribed ice sheet geometries, we use present-day climate data and given temperature offsets rather than LGM climate model runs.
 
@@
 \end{figure}
-
-\emph{Coming soon...}
-
-% Define PISM = Parallel Ice Sheet Model

### 2013-06-11 sections/intro.tex
Throw in all the modelling references
And expand the introduction.

--- 2013-03-21 sections/intro.tex
@@
 % ----------------------------------------------------------------------
+
+The numerical modelling of past glaciers and ice sheets allows indirect confrontation between glaciological theories and geomorphological reconstructions, challenging the validity of both.
+Yet a major obstacle in this exercise resides in the large uncertainties that affect climate forcing of the numerical ice sheet model \citep{hebeler-etal-2008}.
+This includes the uncertainty on Earth's present climate, and the larger uncertainty on its changes in the past.
+
+Our area of focus for the present study is the Cordilleran Ice Sheet in north-western North America.
+%In a numerical modelling perspective it is one of the least studied paleo-ice sheets on Earth, where nethertheless significant geomorphological data is available \citep{jackson-clague-1991}.
+The Cordilleran Ice Sheet has been previously modelled as part of the North-american ice sheet complex \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-2004,bintanja-etal-2005,gregoire-etal-2012}, and within a larger-scale Arctic or Global numerical domain \citep{huybrechts-tsiobbel-1996,charbit-etal-2002,johnson-fastook-2002,zweck-huybrechts-2005,abeouchi-etal-2007,charbit-etal-2013}.
+In those studies, various methods have been used to provide climate forcing to ice-sheet models, ranging from simplistic parametrizations to coupling to General Circulation Models (GCM).
+
+\begin{itemize}
+
+  \item \citet{johnson-fastook-2002} used a parametrized climate forcing based on elevation and distance to a climatic north pole \citep{fastook-prentice-1994}.
+
+  \item \citet{bintanja-etal-2005} used present-day reanalysis data as a reference climate to which they applied time-dependant temperature offsets to recreate glacial conditions.
+
+  \item \citet{huybrechts-tsiobbel-1996} use palæo-climate simulation output from GCM to force an ice-sheet model at steady-state.
+Such GCMs are typically too computationally demanding to simulate climate evolution at the time-scales of ice sheet growth and decay.
+Therefore a common method to produce time-dependant climate forcing consist in interpolating in time between different GCM runs, either linearly \citep{charbit-etal-2002} or more commonly using a wieghting function based on ice-core record data \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}.
+GCM palæo-climate simulations typically rely on ice-sheet reconstructions of the type of ICE-4G for their surface topography boundary condition.
+In turn, ice-sheet models based on these simulations generally reproduce LGM ice extents closely consistent with the reconstruction.
+
+  \item Finally \citet{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013} couple their ice-sheet models to GCM of coarser resolution and reduced complexity.
+
+\end{itemize}
+
+%TODO tarasov-peltier-1997
+
+If these studies reproduce well the magnitude of the North American glaciation at Last Glacial Maximum, there is a general tendency for studies that do not undirectly rely on reconstructions of the type of ICE-4G to predict excessive ice cover in northern Yukon Territory and Alaska, regions that have not been covered by ice in hundreds of thousands of years \citep{johnson-fastook-2002,bintanja-etal-2005,huybrechts-tsiobbel-1996,charbit-etal-2002,marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}.
+
+Here we use a Parallel Ice Sheet Model (PISM) \citep{web:pism} to simulate the entire Cordilleran Ice Sheet at the Last Glacial Maximum (LGM).
+We force our model with multiple climate datasets and compare our results to the mapped LGM ice sheet margin from \citet{kleman-etal-2010}.
+See \citet{quiquet-etal-2012} for a similar comparison on Greenland.
+To avoid circular dependance on prescribed ice sheet geometries, and allow high spatial resolution in our climate forcing, we use present-day climate data and given temperature offsets in a manner similar to \citet{bintanja-etal-2005}.
+To our knowledge, this is the first modelling study which specifically focus on the Cordilleran Ice Sheet since the one by \citep{robert-1991}, who focused specifically on its southern margin.
 

### 2013-06-11 sections/model.tex
Throw in all the modelling references
And expand the introduction.

@@
 Ablation is computed from the number of positive degree-days, defined as the integral of temperatures above 0\,\degree\,C in one year.
-Positive degree-days are calculated by the \citet{calov-greve-2011} formula to accounts for day-to-day temperature variability by assuming a normal distribution of variance $\sigma=5\degree C$ along a central (input) value $T$:
+Positive degree-days are calculated by the \citet{calov-greve-2005} formula to accounts for day-to-day temperature variability by assuming a normal distribution of variance $\sigma=5\degree C$ along a central (input) value $T$:
 

### 2013-06-11 cordillera-climate.tex
Empty sections coming soon...
Added more todonotes in figures
Implement Arjen comments
Trivial comments implemented in the text; others as todo notes.
Throw in all the modelling references
And expand the introduction.
Remove technical note
Change the title

--- 2013-02-12 cordillera-climate.tex
@@
 % Extra packages
-%\usepackage{color}
-%\usepackage{hyperref}
+\usepackage[colorlinks,citecolor=blue]{hyperref}
+\usepackage{framed}
+
+% Recognize utf-8 special characters
+\usepackage[utf8]{inputenc}
 
 % All floats at the end of the manuscripts
-\usepackage{endfloat}
+%\usepackage{endfloat}
 
-% Add to-do notes
+% Add to-do notes and enlarge right margin
 \usepackage{todonotes}
-\newcommand{\missingref}[0]{\todo[color=cyan]{Need ref.}}
+\usepackage[left=54pt, top=81pt, textwidth=398pt, marginparsep=11pt, marginparwidth=93pt, textheight=625pt, footskip=30pt]{geometry}
+\newcommand{\needref}[0]{\todo[color=cyan]{Reference needed}}
+\newcommand{\arjen}[2][]{\todo[color=lime,{#1}]{\textbf{Arjen:} #2}}
+\newcommand{\constantine}[2][]{\todo[color=yellow,{#1}]{\textbf{Constantine:} #2}}
+\newcommand{\qiong}[2][]{\todo[color=pink,{#1}]{\textbf{Qiong:} #2}}
+\newcommand{\ask}[2][]{\todo[color=lightgray,{#1}]{\textbf{Julien:} #2}}
+
+% Personal shortcuts
+\newcommand{\degC}[0]{\,\unit{\degree C} }
+\newcommand{\masl}[0]{\,\unit{m a.s.l.} }
 
@@
 
-\title{Choosing climate forcing for numerical simulations of the Cordilleran ice sheet}
-\runningtitle{Cordilleran ice sheet climate forcing}
+\title{The effect of climate forcing on numerical simulations of the Cordilleran Ice Sheet at the last glacial maximum}
+\runningtitle{Climate forcing for Cordilleran ice sheet simulations}
 
@@
 \label{sec:conclusion}
+\emph{Coming soon...}
 % ----------------------------------------------------------------------

### 2013-06-12 sections/intro.tex
Imrove introduction, add references

--- 2013-03-21 sections/intro.tex
@@
 % ----------------------------------------------------------------------
+
+The numerical modelling of past glaciers and ice sheets allows indirect confrontation between glaciological theories and geomorphological reconstructions, challenging the validity of both.
+Yet a major obstacle in this exercise resides in the large uncertainties that affect climate forcing of numerical glacier models \citep{hebeler-etal-2008}.
+This includes the uncertainty on Earth's present climate, and the larger uncertainty on its changes in the past.
+
+Our area of focus for the present study is the Cordilleran Ice Sheet in north-western North America.
+In a numerical modelling perspective, it is one of the least studied paleo-ice sheets on Earth, where nethertheless significant geomorphological data is available \citep{jackson-clague-1991}.
+The Cordilleran Ice Sheet has been previously modelled as part of the North American ice sheet complex \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-2004,bintanja-etal-2005,gregoire-etal-2012} and within a larger-scale Arctic or Global numerical domain \citep{huybrechts-tsiobbel-1996,charbit-etal-2002,johnson-fastook-2002,zweck-huybrechts-2005,abeouchi-etal-2007,charbit-etal-2013}.
+In those studies, various climate forcing methods have been used, ranging from simplistic parametrizations to coupling to General Circulation Models (GCM).
+
+\begin{itemize}
+
+  \item \citet{johnson-fastook-2002} used a parametrized climate forcing based on elevation and distance from a climatic pole initiallly developped for Antarctica \citep{fastook-prentice-1994}.
+
+  \item \citet{bintanja-etal-2005} used present-day reanalysis data as a reference climate to which they applied time-dependant temperature offsets to recreate glacial conditions.
+While this method ensures a more accurate representation of present climate, the approach to past climate change remain simplistic.
+
+  \item \citet{huybrechts-tsiobbel-1996} use GCM output from a palæo-climate simulation to force an ice-sheet model to steady-state.
+Due to GCMs' computational expense, palæo-climate simulations were typically run for discrete ages of reasonably known ice-sheet configuration, such as the LGM.
+To produce time-dependant climate forcing, glacier modellers interpolated GCM output in time between different simulations, either linearly \citep{charbit-etal-2002} or more commonly using a wieghting function based on ice-core records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}.
+GCM palæo-climate simulations typically rely on ice-sheet reconstructions of the type of ICE-4G \citep{peltier-1994} for their surface topography boundary condition.
+In turn, ice-sheet models based on these simulations generally reproduce LGM ice extents closely consistent with the reconstruction.
+
+  \item Finally \citet{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013} couple their ice-sheet models to GCM of coarser resolution and reduced complexity.
+
+\end{itemize}
+
+%TODO tarasov-peltier-1997
+
+If these studies reproduce well the magnitude of the North American glaciation at Last Glacial Maximum, there is a general tendency for studies that do not undirectly rely on reconstructions of the type of ICE-4G\citep{johnson-fastook-2002,bintanja-etal-2005,huybrechts-tsiobbel-1996,charbit-etal-2002,marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}  to predict excessive ice cover in northern Yukon Territory and Alaska, regions that have not been covered by ice in hundreds of thousands of years \citep{dukrodkin-1999,kaufman-manley-2004}.
+
+Here we use a Parallel Ice Sheet Model (PISM) \citep{web:pism} to simulate the entire Cordilleran Ice Sheet at the Last Glacial Maximum (LGM).
+We force our model with multiple climate datasets and compare our results to the mapped LGM ice sheet margin from \citet{kleman-etal-2010}.
+See \citet{quiquet-etal-2012} for a similar comparison on Greenland.
+To avoid circular dependance on prescribed ice sheet geometries, and allow high spatial resolution in our climate forcing, we use present-day climate data and given temperature offsets in a manner similar to \citet{bintanja-etal-2005}.
+To our knowledge, this is the first modelling study which specifically focus on the Cordilleran Ice Sheet since the one by \citep{robert-1991}, who focused specifically on its southern margin.
 

### 2013-06-12 sections/intro.tex
Restructure the introduction

@@
 
-Our area of focus for the present study is the Cordilleran Ice Sheet in north-western North America.
-In a numerical modelling perspective, it is one of the least studied paleo-ice sheets on Earth, where nethertheless significant geomorphological data is available \citep{jackson-clague-1991}.
-The Cordilleran Ice Sheet has been previously modelled as part of the North American ice sheet complex \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-2004,bintanja-etal-2005,gregoire-etal-2012} and within a larger-scale Arctic or Global numerical domain \citep{huybrechts-tsiobbel-1996,charbit-etal-2002,johnson-fastook-2002,zweck-huybrechts-2005,abeouchi-etal-2007,charbit-etal-2013}.
-In those studies, various climate forcing methods have been used, ranging from simplistic parametrizations to coupling to General Circulation Models (GCM).
+In numerical models of the former North American ice sheets, various climate forcing methods have been used, ranging from simplistic parametrizations to coupling to General Circulation Models (GCM).
+Arguably the most physical way to force an ice sheet model in the past is to couple it with a GCM \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013}.
+However due to the high computational cost of this exercise only GCMs of coarser resolution and reduced complexity can achieve it.
 
-\begin{itemize}
-
-  \item \citet{johnson-fastook-2002} used a parametrized climate forcing based on elevation and distance from a climatic pole initiallly developped for Antarctica \citep{fastook-prentice-1994}.
-
-  \item \citet{bintanja-etal-2005} used present-day reanalysis data as a reference climate to which they applied time-dependant temperature offsets to recreate glacial conditions.
-While this method ensures a more accurate representation of present climate, the approach to past climate change remain simplistic.
-
-  \item \citet{huybrechts-tsiobbel-1996} use GCM output from a palæo-climate simulation to force an ice-sheet model to steady-state.
-Due to GCMs' computational expense, palæo-climate simulations were typically run for discrete ages of reasonably known ice-sheet configuration, such as the LGM.
-To produce time-dependant climate forcing, glacier modellers interpolated GCM output in time between different simulations, either linearly \citep{charbit-etal-2002} or more commonly using a wieghting function based on ice-core records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}.
-GCM palæo-climate simulations typically rely on ice-sheet reconstructions of the type of ICE-4G \citep{peltier-1994} for their surface topography boundary condition.
+More commonly GCM output from palæo-climate simulations is used as a forcing \citep{huybrechts-tsiobbel-1996}.
+To produce time-dependant forcing, GCM is often interpolated in time between different palæo-climate simulations, either linearly \citep{charbit-etal-2002} or more commonly using a wieghting function based on ice-core records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}.
+GCM palæo-climate simulations typically rely on ice-sheet reconstructions of the type of the ICE-4G \citep{peltier-1994} for their surface topography boundary condition.
 In turn, ice-sheet models based on these simulations generally reproduce LGM ice extents closely consistent with the reconstruction.
 
-  \item Finally \citet{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013} couple their ice-sheet models to GCM of coarser resolution and reduced complexity.
-
-\end{itemize}
+Other methods include parametrization of climate forcing based on elevation \citep{robert-1991}, geographical coordinates \citep{fastook-prentice-1994,johnson-fastook-2002} and temperature offsets applied to a present-day reference climate \citep{bintanja-etal-2005}.
 
@@
 
-If these studies reproduce well the magnitude of the North American glaciation at Last Glacial Maximum, there is a general tendency for studies that do not undirectly rely on reconstructions of the type of ICE-4G\citep{johnson-fastook-2002,bintanja-etal-2005,huybrechts-tsiobbel-1996,charbit-etal-2002,marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}  to predict excessive ice cover in northern Yukon Territory and Alaska, regions that have not been covered by ice in hundreds of thousands of years \citep{dukrodkin-1999,kaufman-manley-2004}.
+Our area of focus for the present study is the Cordilleran Ice Sheet in north-western North America.
+In a numerical modelling perspective, it is one of the least studied paleo-ice sheets on Earth, where nethertheless significant geomorphological data is available \citep{jackson-clague-1991}.
+Although generally not the primary target of these simulations, the Cordilleran Ice Sheet has been previously modelled as part of the North American ice sheet complex \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-2004,bintanja-etal-2005,gregoire-etal-2012} and within a larger-scale Arctic or Global numerical domain \citep{huybrechts-tsiobbel-1996,charbit-etal-2002,johnson-fastook-2002,zweck-huybrechts-2005,abeouchi-etal-2007,charbit-etal-2013}.
+
+If these studies reproduce well the magnitude of the North American glaciation at Last Glacial Maximum, there is a general tendency for studies that do not undirectly rely on reconstructions of the type of ICE-4G \citep{johnson-fastook-2002,bintanja-etal-2005,huybrechts-tsiobbel-1996,charbit-etal-2002,marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012} to predict excessive ice cover in northern Yukon Territory and Alaska, regions that have not been covered by ice sheets throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.
 

### 2013-06-13 sections/intro.tex
Polish the introduction

@@
 
-In numerical models of the former North American ice sheets, various climate forcing methods have been used, ranging from simplistic parametrizations to coupling to General Circulation Models (GCM).
-Arguably the most physical way to force an ice sheet model in the past is to couple it with a GCM \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013}.
-However due to the high computational cost of this exercise only GCMs of coarser resolution and reduced complexity can achieve it.
+Numerical models of the former North American ice sheets have used various climate forcing methods, ranging from simplistic parametrizations to coupling to General Circulation Models (GCM).
+The latter is arguably the most physically sound way to force an ice sheet model in the past and has proven effective \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013}.
+However the computational demand of GCMs is so that only models of reduced complexity and coarser resolution can run on time-scales of ice sheet growth and decay.
 
-More commonly GCM output from palæo-climate simulations is used as a forcing \citep{huybrechts-tsiobbel-1996}.
-To produce time-dependant forcing, GCM is often interpolated in time between different palæo-climate simulations, either linearly \citep{charbit-etal-2002} or more commonly using a wieghting function based on ice-core records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}.
-GCM palæo-climate simulations typically rely on ice-sheet reconstructions of the type of the ICE-4G \citep{peltier-1994} for their surface topography boundary condition.
-In turn, ice-sheet models based on these simulations generally reproduce LGM ice extents closely consistent with the reconstruction.
-
-Other methods include parametrization of climate forcing based on elevation \citep{robert-1991}, geographical coordinates \citep{fastook-prentice-1994,johnson-fastook-2002} and temperature offsets applied to a present-day reference climate \citep{bintanja-etal-2005}.
+Another method consist in using atmospheric fields produced in uncoupled palæo-climate simulations \citep{huybrechts-tsiobbel-1996}.
+To produce a time-dependant forcing, the GCM output is interpolated in time between different simulations \citep{charbit-etal-2002} or weighted based on ice-core records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}.
+These palæo-climate simulations use ice sheet reconstructions such as the ICE-4G \citep{peltier-1994} as topographic boundary condition, which results in subsequently modelled ice cover consistent with these reconstructions.
+\ask[noline]{I am thinking whether I should extend this literature list to papers not concerning North America?}
 
@@
 
-Our area of focus for the present study is the Cordilleran Ice Sheet in north-western North America.
-In a numerical modelling perspective, it is one of the least studied paleo-ice sheets on Earth, where nethertheless significant geomorphological data is available \citep{jackson-clague-1991}.
-Although generally not the primary target of these simulations, the Cordilleran Ice Sheet has been previously modelled as part of the North American ice sheet complex \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-2004,bintanja-etal-2005,gregoire-etal-2012} and within a larger-scale Arctic or Global numerical domain \citep{huybrechts-tsiobbel-1996,charbit-etal-2002,johnson-fastook-2002,zweck-huybrechts-2005,abeouchi-etal-2007,charbit-etal-2013}.
-
-If these studies reproduce well the magnitude of the North American glaciation at Last Glacial Maximum, there is a general tendency for studies that do not undirectly rely on reconstructions of the type of ICE-4G \citep{johnson-fastook-2002,bintanja-etal-2005,huybrechts-tsiobbel-1996,charbit-etal-2002,marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012} to predict excessive ice cover in northern Yukon Territory and Alaska, regions that have not been covered by ice sheets throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.
-
-Here we use a Parallel Ice Sheet Model (PISM) \citep{web:pism} to simulate the entire Cordilleran Ice Sheet at the Last Glacial Maximum (LGM).
-We force our model with multiple climate datasets and compare our results to the mapped LGM ice sheet margin from \citet{kleman-etal-2010}.
-See \citet{quiquet-etal-2012} for a similar comparison on Greenland.
-To avoid circular dependance on prescribed ice sheet geometries, and allow high spatial resolution in our climate forcing, we use present-day climate data and given temperature offsets in a manner similar to \citet{bintanja-etal-2005}.
-To our knowledge, this is the first modelling study which specifically focus on the Cordilleran Ice Sheet since the one by \citep{robert-1991}, who focused specifically on its southern margin.
+Avoiding this circular dependence unfortunately goes with simplistic assumptions on Earth's past climate.
+\citet{robert-1991} use a parametrization of glacier surface mass balance based on elevation.
+\citet{fastook-prentice-1994} use a parametrization of climate forcing based on elevation and distance from a climatic pole initially prepared for Antarctica \citep{johnson-fastook-2002}.
+Finally, \citet{bintanja-etal-2005} apply time-dependant temperature offsets to a reference present-day climate obtained from climate reanalysis data.
+We use a similar method.
 
@@
 \end{figure}
+
+Our area of focus for the present study is the Cordilleran Ice Sheet in north-western North America (Figure 1).
+In a numerical modelling perspective, it is one of the least studied palæo-ice sheets of the Northern Hemisphere, where nevertheless significant geomorphological data is available \citep{jackson-clague-1991}.
+Although generally not the primary target of these simulations, the Cordilleran Ice Sheet has been previously modelled as part of the North American ice sheet complex \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-2004,bintanja-etal-2005,gregoire-etal-2012} and within a larger-scale Arctic or Global numerical domain \citep{huybrechts-tsiobbel-1996,charbit-etal-2002,johnson-fastook-2002,zweck-huybrechts-2005,abeouchi-etal-2007,charbit-etal-2013}.
+
+While these studies reproduce well the magnitude of the North American glaciation at Last Glacial Maximum, they tend to predict excessive ice cover in parts of northern Yukon Territory and Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.
+
+Here we use a Parallel Ice Sheet Model \citep[PISM,][]{web:pism} to simulate the entire Cordilleran Ice Sheet at the Last Glacial Maximum (LGM).
+We force our model with multiple climate datasets and compare our results to the mapped LGM ice sheet margin from \citet{kleman-etal-2010}.
+See \citet{quiquet-etal-2012} for a similar comparison on Greenland.
+To avoid any dependence on ice-sheet reconstructions and allow high spatial resolution in our climate forcing, we use present-day climate data and given temperature offsets in a manner similar to \citet{bintanja-etal-2005}.
+To our knowledge, this is the first modelling study which specifically focus on the Cordilleran Ice Sheet since the one by \citep{robert-1991}, who focused specifically on its southern margin.

### 2013-06-13 cordillera-climate.tex
General clean-up in the todonotes

@@
 \newcommand{\qiong}[2][]{\todo[color=pink,{#1}]{\textbf{Qiong:} #2}}
-\newcommand{\ask}[2][]{\todo[color=lightgray,{#1}]{\textbf{Julien:} #2}}
+\newcommand{\julien}[2][]{\todo[color=lightgray,{#1}]{\textbf{Julien:} #2}}
 
@@
 \begin{abstract}
-\todo{This abstract is preliminary. I will rewrite it when I am done with the rest}
+\julien[noline]{The abstract is preliminary.}
 We present an ensemble of numerical simulations of the former Cordilleran ice sheet forced by present-day climatologies from five different datasets.

### 2013-06-13 sections/climate.tex
General clean-up in the todonotes

@@
 
-\arjen{It is an unusual style to start each sentence with Figure X shows...
-We should rephrase this later-on.} Figure~\ref{fig:prec} shows the spatial distribution of DJF and JJA precipitation rates in WorldClim.
-Coastal regions generally receive much more precipitation than inland ones as a result of the continuous topographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascades\arjen{orographic precip!}.
+Figure~\ref{fig:prec} shows the spatial distribution of DJF and JJA precipitation rates in WorldClim.
+Coastal regions generally receive much more precipitation than inland ones as a result of the continuous orographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascades.
 From a glacier mass-balance point of view, this contrast is made even stronger by the difference in timing of the precipitation peak through the year.
@@
 If good coverage exists in the southern parts of our modelling domain, several hundred kilometres can separate neighbouring stations in the north \citep{data:worldclim}.
-A second drawback of the dataset in the scope of this study is the lack of data across oceans, particularly on the Pacific continental shelf where glaciers have advanced during the last glacial cycle\needref.
+A second drawback of the dataset in the scope of this study is the lack of data across oceans, particularly on the Pacific continental shelf where glaciers have advanced during the last glacial cycle \citep{jackson-clague-1991}.
 
@@
 
-Air surface temperature and precipitation rates \arjen{do you really mean rate as in mm/time unit?} from the WorldClim data, the NCEP/NCAR, ERA-Interim, CFSR and NARR reanalyses were re-projected to Canadian Atlas Lambert conformal conic projection (EPSG code~3978) \arjen{is this general knowledge?
-I am lost...} and bilinearly interpolated to the model resolution of 10\,km.
+Air surface temperature and precipitation from the WorldClim data, the NCEP/NCAR, ERA-Interim, CFSR and NARR reanalyses were re-projected to Canadian Atlas Lambert conformal conic projection (EPSG code~3978) and bilinearly interpolated to the model resolution of 10\,km.
 In addition, WorldClim data was extrapolated to oceans using a nearest-neighbour approach.
@@
 Yet we will show that this smoothing has very limited effect on the ice sheet model outcome.
-This preprocessing steps were made in GRASS~GIS using scripts available on the first author's website\arjen{I am unsure if this is the best way t do it?
-Were you using principles described by others then you should refer to the litterature instead.}.
+These preprocessing steps were made in GRASS~GIS.
 

### 2013-06-13 sections/intro.tex
General clean-up in the todonotes

@@
 These palæo-climate simulations use ice sheet reconstructions such as the ICE-4G \citep{peltier-1994} as topographic boundary condition, which results in subsequently modelled ice cover consistent with these reconstructions.
-\ask[noline]{I am thinking whether I should extend this literature list to papers not concerning North America?}
+\julien[noline]{Should this literature list be extended to studies not covering North America?}
 
@@
 Here we use a Parallel Ice Sheet Model \citep[PISM,][]{web:pism} to simulate the entire Cordilleran Ice Sheet at the Last Glacial Maximum (LGM).
-We force our model with multiple climate datasets and compare our results to the mapped LGM ice sheet margin from \citet{kleman-etal-2010}.
+We force our model with multiple climate datasets and compare our results to the mapped LGM ice sheet margin from \julien{I will try to find an outline that is not cropped over Alaska instead.}\citet{kleman-etal-2010}.
 See \citet{quiquet-etal-2012} for a similar comparison on Greenland.

### 2013-06-13 sections/model.tex
General clean-up in the todonotes

@@
 Monthly temperature and precipitation are converted to surface accumulation and ablation by a temperature-index (positive degree-day) model.
-Accumulation is computed as the amount of precipitation that falls when temperature is below freezing\ask{Is this true?
-What is the current threshold in stable 0.5?}.
-Ablation is computed from the number of positive degree-days, defined as the integral of temperatures above 0\,\degree\,C in one year.
-Positive degree-days are calculated by the \citet{calov-greve-2005} formula to accounts for day-to-day temperature variability by assuming a normal distribution of variance $\sigma=5\degree C$ along a central (input) value $T$:
+Ice accumulation is equal to precipitation when temperature is below 0\,\degC, and decreases linearly with temperature between 0 and 2\,\degC.
+Ice ablation is computed from the number of positive degree-days, defined as the integral of temperatures above 0\,\degC in one year.
+Positive degree-days are calculated by the \citet{calov-greve-2005} formula to accounts for day-to-day temperature variability by assuming a normal distribution of variance $\sigma=5\degC$ along a central (input) value $T$:
 
@@
 
-Ice flow velocities are computed by a heuristic \arjen{not sure what this expresses!!} where the Shallow Shelf Approximation (SSA) is used as a ``sliding law'' to a polythermal Shallow Ice Approximation (SIA) \citep{bueler-brown-2009}.
-
-\ask[inline]{\emph{Should I include the SSA+SIA \citet{bueler-brown-2009} formula?}
-
+Ice flow velocities are computed by a heuristic the Shallow Shelf Approximation (SSA) is used as a ``sliding law'' to a polythermal Shallow Ice Approximation (SIA) \citep{bueler-brown-2009}.
 Depth-averaged velocities $\vec{v}$ are computed as a weighted sum of SIA velocities $\vec{v_{\mathrm{SIA}}}$ and SSA velocities $\vec{v_{\mathrm{SSA}}}$ by
@@
 \end{equation}
-}
-
-\arjen[inline]{unless variations in these determine your best fit input dataset, I would not include them in this paper.}
 
@@
 With $\phi$ in degrees and $z$ in meters above present sea level.
-The glacier bed topography responds to the ice load following a regional isostasy model with flexure and relaxation \citep{lingle-clark-1985}\todo{Check \citet{lingle-clark-1985}}.
+The glacier bed topography responds to the ice load following a regional isostasy model with flexure and relaxation \citep{lingle-clark-1985}.
 
@@
 We mimic glacial conditions by applying constant temperature offsets homogeneously over the entire domain.
-Each simulation starts from ice-free conditions and is run for 10,000 years, a time interval representative of the rapid build-up of the last Cordilleran Ice Sheet from nearly ice-free condition to last glacial maximum extent \needref.
+Each simulation starts from ice-free conditions and is run for 10,000 years, a time interval representative of the rapid build-up of the last Cordilleran Ice Sheet from nearly ice-free condition to last glacial maximum extent \citep{clague-1989}.
 Our simulations were run in parallel on 32 cores at the Swedish National Supercomputing Center.
-
-\arjen[inline]{Check Stroeven et al.
-(2010): "Radiocarbon ages from wood and fossils underlying McConnell till indicate that the McConnell glaciation began after 29,600  300 14C yr BP (Matthews et al., 1990), but glaciers did not advance into major valleys before 26,350  280 14C yr BP (Jackson and Harington, 1991), and a fully developed CIS did not exist until after 23,900  1140 14C yr BP (Lowdon and Blake,1981).
-Published radiocarbon ages indicate that the St.
-Elias lobe of the CIS was in retreat by 13,660  180 14C yr BP (Rampton, 1971) and the Selwyn lobe by 12,590  120 14C yr BP (Ward, 1989) (Fig.
-2)." and references therein.}

### 2013-06-13 sections/results.tex
General clean-up in the todonotes

--- 2013-02-21 sections/results.tex
@@
 	\end{center}
-	\qiong[inline]{Move ticks on the colorbar. The 1 is confusing.}
 	\caption{Extent of ice cover after 10\,kyr as a function of applied temperature offsets for each climate forcing.}
@@
 	\end{center}
-	\todo[inline]{Use black-and-white-proof markers and highlight the ``best'' runs.}
+	\julien[inline]{I will make the figure readable in black-and-white and highlight markers corresponding to the ``best'' runs.}
 	\caption{Total ice volume and glaciated area after 10\,kyr as a function of temperature offset and climate forcing.}
@@
 	\end{center}
-	\todo[inline]{The must would be to overlay geomorphic ice extent.}
+	\julien[inline]{I will to overlay the geomorphologic LGM ice margin so that it becomes easier to see which runs match it better.}
 	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr using temperature offsets that lead to similar areas of ice cover for each climate forcing.}
@@
 We observe that although similar in size, the selected ice sheets are different in shape and flow patterns.
-More specifically, NCEP/NCAR, CFSR and ERA-Interim ice sheets cover large territories in Yukon and Alaska that where shown to have remained ice-free for at least several glacial cycles.\needref
+More specifically, NCEP/NCAR, CFSR and ERA-Interim ice sheets cover large territories in Yukon and Alaska that where shown to have remained ice-free for at least several glacial cycles\citep{dukrodkin-1999,kaufman-manley-2004}.

### 2013-06-13 sections/model.tex
Update the mass balance parameters

@@
 \subsection{Surface mass-balance}
-
-\begin{table}[t]
-	\caption{Positive degree-day model parameters.}
-	\label{tab:pdd}
-	\vskip4mm
-	\centering
-	\begin{tabular}{ll}
-		\tophline
-		Parameter& Value\\
-		\middlehline
-		degree-day factor for snow & 3\,\unit{mm\,K^{-1}\,day^{-1}}\\
-		degree-day factor for ice & 8\,\unit{mm\,K^{-1}\,day^{-1}}\\
-		standard deviation of temperatures & 5\,K\\
-		refreezing faction & 0.6 \\
-		\bottomhline
-	\end{tabular}
-\end{table}
 
@@
 		\mathrm{erfc} \left(-\frac{T(t)}{\sqrt{2}\sigma}\right)
-	\right]
+	\right],
 \end{equation}
 
-Where $[t_0; t_1]$ is one year and $\mathrm{erfc}$ is the complementary error function.
-The ablation model employs different degree-day factors for snow and ice and accounts for refreezing (table~\ref{tab:pdd}).
+where the integration interval $[t_0; t_1]$ is one week and $\mathrm{erfc}$ is the complementary error function.
+The ablation model uses degree-day factors of 3.04~\unit{mm\,K^{-1}\,day^{-1}} for snow and 4.59~\unit{mm\,K^{-1}\,day^{-1}} for ice, as derived from mass-balance measurements on present glaciers from the Coast and Rocky Mountains in British Columbia \citep{shea-etal-2009}.
 

### 2013-06-13 sections/results.tex
Update the mass balance parameters

@@
 We observe that although similar in size, the selected ice sheets are different in shape and flow patterns.
-More specifically, NCEP/NCAR, CFSR and ERA-Interim ice sheets cover large territories in Yukon and Alaska that where shown to have remained ice-free for at least several glacial cycles\citep{dukrodkin-1999,kaufman-manley-2004}.
+More specifically, NCEP/NCAR, CFSR and ERA-Interim ice sheets cover large territories in Yukon and Alaska that where shown to have remained ice-free for at least several glacial cycles \citep{dukrodkin-1999,kaufman-manley-2004}.

### 2013-06-19 sections/intro.tex
Use Dyke 2004 LGM ice margin

@@
 The frame delimits this study's modelling domain.
-The outlines of the former ice sheets are from \citet{kleman-etal-2010}.
+The outlines of the former ice sheets depict ice cover at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) after \citet{dyke-2004}.
 The background map was made with ETOPO1 \citep{data:etopo1} and Natural Earth Data \citep{data:naturalearth}.}
@@
 Here we use a Parallel Ice Sheet Model \citep[PISM,][]{web:pism} to simulate the entire Cordilleran Ice Sheet at the Last Glacial Maximum (LGM).
-We force our model with multiple climate datasets and compare our results to the mapped LGM ice sheet margin from \julien{I will try to find an outline that is not cropped over Alaska instead.}\citet{kleman-etal-2010}.
+We force our model with multiple climate datasets and compare our results to the mapped LGM ice sheet margin from \citet{dyke-2004}.
 See \citet{quiquet-etal-2012} for a similar comparison on Greenland.

### 2013-06-19 sections/results.tex
Add Dyke LGM margin to best runs figure

@@
 	\end{center}
-	\julien[inline]{I will to overlay the geomorphologic LGM ice margin so that it becomes easier to see which runs match it better.}
-	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr using temperature offsets that lead to similar areas of ice cover for each climate forcing.}
+	\julien[inline]{I think there is too much information in this figure. I will simplify it or divide it in two figures instead.}
+	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr using temperature offsets that lead to similar areas of ice cover for each climate forcing.
+14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) ice margin from \citet{dyke-2004}.}
 	\label{fig:best}

### 2013-06-19 data/get.sh

@@ -6,7 +6,7 @@ for clim in wcnn erai narr cfsrs7 cfsr ncar; do
     atm=cordillera-$clim-10km-bl
     run=pism/output/$atm/stepcool${cool}sll120+ccli+till1030/y0010000
     if [ ! -f $clim-$cool.nc ]; then
-      ssh scooter "ncks -O -v topg,thk,usurf,csurf $run.nc /tmp/tmp.nc"
+      ssh scooter "ncks -O -v lat,lon,topg,thk,usurf,csurf $run.nc /tmp/tmp.nc"
       scp scooter:/tmp/tmp.nc $clim-$cool.nc
     fi
     if [ ! -f $clim-$cool-ts.nc ]; then

### 2013-06-19 figures/plot.py

@@ -2,9 +2,12 @@
 # coding: utf-8
 
 import numpy as np
+import shapefile
 from netCDF4 import Dataset
 from matplotlib import pyplot as plt
 from matplotlib import colors as mcolors
+from matplotlib.collections import LineCollection
+from mpl_toolkits.basemap import Basemap
 from iceplot import plot as iplt
 from iceplot import cm as icm
 
@@ -16,6 +19,7 @@ rc('mathtext', default='regular')
 ### Globals ###
 
 pismdir = '/home/julien/work/code/pism'
+shpfile = '/home/julien/work/data/dyke-deglaciation/16.8-ka.shp'
 mapsize = (30., 60.)
 labels = {
   'etopo':  'ETOPO1',
@@ -29,7 +33,7 @@ labels = {
 
 ### Base function ###
 
-def _plot_maps(climates, clabel, output, func, ci=-1, cticks=None):
+def _plot_maps(climates, clabel, output, func, ci=-1, cticks=None, margin=False):
     """Base function to plot maps"""
 
     # initialize figure
@@ -40,6 +44,26 @@ def _plot_maps(climates, clabel, output, func, ci=-1, cticks=None):
     for i, clim in enumerate(climates):
       ax = plt.axes(fig.grid[i])
       sm.append(func(i, clim))
+
+    # add LGM ice margin
+    if margin is True:
+      nc = Dataset('../data/%s-00.nc' % (climates[0]))
+      m = Basemap(projection='lcc',
+        lat_1=49, lat_2=77, lat_0=49, lon_0=-95,
+        llcrnrlat=nc.variables['lat'][0,0],
+        urcrnrlat=nc.variables['lat'][-1,-1],
+        llcrnrlon=nc.variables['lon'][0,0],
+        urcrnrlon=nc.variables['lon'][-1,-1])
+      sf = shapefile.Reader(shpfile)
+      for record, shape in zip(sf.records(),sf.shapes()):
+        lons,lats = zip(*shape.points)
+        data = np.array(m(lons, lats)).T/10000
+        #lines.set_facecolors('blue')
+        #lines.set_edgecolors('k')
+        #lines.set_linewidth(0.1)
+      for ax in fig.grid:
+        lines = LineCollection([data,], antialiaseds=(1,))
+        ax.add_collection(lines)
 
     # add colorbar
     cb = fig.colorbar(sm[ci], fig.grid.cbar_axes[0], ticks=cticks)
@@ -119,15 +143,15 @@ def topo():
 
[…]

### 2013-06-21 sections/results.tex
Simplify best runs figure

--- 2013-06-13 sections/results.tex
@@
 	\end{center}
-	\julien[inline]{I will to overlay the geomorphologic LGM ice margin so that it becomes easier to see which runs match it better.}
-	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr using temperature offsets that lead to similar areas of ice cover for each climate forcing.}
+	\caption{Ice surface topography (black contours every 1000\,m) after 10\,kyr using temperature offsets that lead to similar areas of ice cover for each climate forcing.
+14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) ice margin (blue line) from \citet{dyke-2004}.}
 	\label{fig:best}

### 2013-06-21 figures/plot.py

@@ -185,14 +185,17 @@ def best():
     # initialize figure
     climates = ['wcnn', 'erai', 'narr', 'cfsrs7', 'cfsr', 'ncar']
     offsets  = ['09', '07', '08', '05', '05', '04']
-    fig = iplt.gridfigure(mapsize, (2, len(climates)/2), cbar_mode='single')
+    fig = iplt.gridfigure(mapsize, (2, len(climates)/2), cbar_mode='none')
 
     # loop on climate datasets
     for i, clim in enumerate(climates):
       ax = plt.axes(fig.grid[i])
       nc = Dataset('../data/%s-%s.nc' % (clim, offsets[i]))
       plt.title('%s - %s K' % (labels[clim], offsets[i]))
-      im = iplt.icemap(nc)
+      iplt.bedtopoimage(nc)
+      iplt.icemargincontour(nc, linecolors=None, colors='white', alpha=0.5)
+      iplt.icemargincontour(nc)
+      cs = iplt.surftopocontour(nc)
 
     # add LGM ice margin
     m = Basemap(projection='lcc',
@@ -205,15 +208,13 @@ def best():
     for record, shape in zip(sf.records(),sf.shapes()):
       lons,lats = zip(*shape.points)
       data = np.array(m(lons, lats)).T/10000
-      #lines.set_facecolors('blue')
-      #lines.set_edgecolors('k')
-      #lines.set_linewidth(0.1)
     for ax in fig.grid:
-      lines = LineCollection([data,], antialiaseds=(1,))
+      lines = LineCollection([data,], antialiaseds=(1,), colors='#000080')
+      lines.zorder=0.5
       ax.add_collection(lines)
 
     # add colorbar and save
-    cb = fig.colorbar(im, fig.grid.cbar_axes[0])
+    cb = fig.colorbar(cs, fig.grid.cbar_axes[0])
     cb.set_label(u'ice surface velocity (m/s)')
     _savefig('cordillera-climate-best')
 

### 2013-06-23 sections/discuss.tex
Empty sections coming soon...
Add discussion draft

@@
 % ----------------------------------------------------------------------
+
+\subsection{Differences with reconstructed LGM margin}
+
+We compared our modelled ice sheet geometries to a reconstruction of the ice margin based on glacial geomorphology and radiocarbon ages \citep{dyke-2004}.
+We use the 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) time-slice, which in this reconstruction corresponds to the LGM extent in most parts of the Cordilleran ice sheet, known to have occurred earlier than for the Laurentide ice sheet.
+Our model results deviates from this reconstruction in different ways, some of which are controlled by climate forcing.
+
+First, the model produces too extensive ice cover in the east.
+There is at least two ways to interpret this.
+In our simulations, a continuous ice cover forms rapidly over the western ranges, which receive more precipitation than the rest of the domain.
+As individual ice caps merged to form a continuous ice cover, they may have enhanced the topographical barrier already formed by the western ranges, which would lead to further reduction of precipitation and warmer temperature in the interior.
+This effect, which we do not include, has shown to have determinant effects in Greenland.
+
+Secondly, simulations produces anomalous accumulation in parts of the Arctic Ocean.
+We interpret this as resulting from or simplistic application of homogeneous temperature offsets over the whole modelling domain.
+Theses regions are marine today, and as a result have lower summer temperature than the adjacent land (Figures~\ref{fig:temp}).
+In our simulation, they become land as we lower sea-level, while retaining the imprint of present-day lower temperature.
+The same effect occur over Great Bear Lake in some simulations.
+
+These digressions taken apart, we can compare the results of different simulations to the mapped LGM ice sheet margin. NARR performs best.
+
+\subsection{Sensitivity to climate forcing}
+
+The outcome of our numerical simulations is strongly dependant to which dataset is used as climate forcing (Figures~\ref{fig:cool06}-\ref{fig:best}).
+These differences are to be attributed to the variations in distributions of temperature and precipitation (Figures~\ref{fig:temp}-\ref{fig:prec}) between forcing datasets.
+Concerning reanalysis dataset, we interpret these differences as mainly the results of different GCM resolutions (Figures~\ref{fig:topo}, Table~\ref{tab:reanalyses}).
+
+As described in section~\ref{sec:climate}, the topography of the Northern American Cordillera is such that its western ranges form a continuous orographic barrier, causing high precipitation along the Pacific coast while leaving much of the interior arid.
+However the capability of a GCM to reproduce these precipitation contrasts is bound to its resolution.
+In a model of coarser resolution such as used in the NCEP/NCAR reanalysis, these high mountain ranges are reduced to smoothed hills and the modelled distribution of precipitation is also smoother.
+
+Even under present-day conditions, our run that was forced by the NCEP/NCAR reanalysis produced an ice-sheet, which is unrealistic.
+
+\subsection{Potential method improvements}
+
+Our simplistic temperature offset is certainly a crude simplification of past climate changes over the region.
+Although it is clear that these temperature changes were neither homogeneous nor constant in time, and probably associated with precipitation changes, their patterns are not trivial and potential strongly inter-dependant with the Cordilleran and Laurentide ice sheet evolution.
+A more correct approach would be to use coupling to a GCM of intermediate complexity.
+
+Some simplifications were made in the surface mass-balance model.
+Our PDD model do not include refreezing.
+Refreezing of melted snow and ice, however, can form a large proportion of the surface mass-balance of an ice sheet.\needref Furthermore, we model temperature variability using a constant, homogeneous value of standard deviation.
+This approaches implies large biases of surface mass-balance\needref, particularly over a region with such various climates as the Northern American Cordillera.

### 2013-06-25 sections/discuss.tex
Improve discussion and add references

--- 2013-02-12 sections/discuss.tex
@@
 % ----------------------------------------------------------------------
+
+\subsection{Differences with reconstructed LGM margin}
+
+\julien[noline]{This sentence may be better placed in the results section.}
+We compare the numerically modelled ice sheet geometries to a reconstruction of the ice margin at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) by \citet{dyke-2004}, based on glacial geomorphology and radiocarbon dating.
+This corresponds to the LGM extent of most of the Cordilleran ice sheet, which occurred later than in the Laurentide ice sheet \citep{dyke-2004}.
+Although the outcome of our numerical simulations is strongly dependant to which dataset is used as climate forcing, common discrepancies between the modelled ice sheet geometry and the ice margin from \citet{dyke-2004} can be observed (Figures~\ref{fig:cool06}-\ref{fig:best}).
+
+For all climate forcing used, the modelled eastern margin of the ice sheet extends further east than the reconstructed boundary between the Cordilleran and the Laurentide ice sheets.
+This can be explained by the potential effects of the growing ice-sheet on regional climate not included in our temperature offset method.
+The Cordilleran ice sheet initiated from the junction of mountain ice caps over the major reliefs.
+In our simulations, a continuous ice cover quickly forms over the western ranges, where precipitation rates are higher than in the rest of the domain.
+This continuous ice cover may have enhanced the topographical barrier already formed by the western ranges, resulting in fewer precipitation and warmer air in the interior.
+\citet{langen-etal-2012} demonstrated that this process, which they refer as a ``self-inhibiting growth'', potentially prevents a regrowth of the Greenland Ice Sheet to its present size.
+Alternatively, the Laurentide Ice Sheet, not included in our model, may have formed a buttress against the smaller Cordilleran Ice Sheet and stopped its advance onto the Canadian Prairies.
+
+Our simulations produces anomalous ice cover in parts on the continental shelf in the Arctic Ocean.
+These regions experience a marine climate, including lower summer temperature than on the adjacent land (Figures~\ref{fig:temp}).
+However sea-level is lowered by 120~m, turning large parts of the low-sloping continental shelf into land.
+A similar effect occurs over Great Bear Lake in some simulations, particularly those driven by the NCEP/NCAR forcing (Figures~\ref{fig:extent}).
+This anomalous ice cover is to be interpreted as an artefact from our simplistic temperature offset method, and has little effect on the rest of the results.
+
+\subsection{Sensitivity to climate forcing}
+
+The outcome of our numerical simulations is strongly dependant to which dataset is used for climate forcing (Figures~\ref{fig:cool06}-\ref{fig:best}).
+This results from temperature and precipitation differences (Figures~\ref{fig:temp}-\ref{fig:prec}) between forcing datasets, to which the glacier model is highly sensitive.
+
+Differences between modelled ice sheet geometries are notably visible in northern Yukon Territory and Alaska, where previous modelling studies commonly produced too extensive ice cover, whereas geomorphological data shows that ice cover have been sparse.
+Concerning reanalysis datasets, we interpret theses differences as mainly the results of different GCM resolutions (Figures~\ref{fig:topo}, Table~\ref{tab:reanalyses}).
+
+As described in section~\ref{sec:climate}, the topography of the Northern American Cordillera is such that its western ranges form a continuous orographic barrier, causing high precipitation along the Pacific coast while leaving much of the interior arid.
+However the capability of a GCM to reproduce these precipitation contrasts is bound to its resolution.
+In a model of coarser resolution such as used in the NCEP/NCAR reanalysis, these high mountain ranges are reduced to smoothed hills and the modelled distribution of precipitation is also smoother.
+
+\subsection{Potential method improvements}
+
+Our simplistic temperature offset is certainly a crude simplification of past climate changes over the region.
+Although it is clear that these temperature changes were neither homogeneous nor constant in time, and probably associated with precipitation changes, their patterns are not trivial and potential strongly inter-dependant with the Cordilleran and Laurentide ice sheet evolution.
+A more correct approach would be to use coupling to a GCM of intermediate complexity \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013}.
+
+Some simplifications were made in the surface mass-balance model.
+Our PDD model do not include refreezing.
+Refreezing of melted snow and ice, however, can form a large proportion of the surface mass-balance of an ice sheet\needref.
+Furthermore, we model temperature variability using a constant, homogeneous value of standard deviation.
+This approaches implies large biases of surface mass-balance \citep{charbit-etal-2013,rau-rogozhina-2013,seguinot-inreview}, particularly over a region with such various climates as the Northern American Cordillera.

### 2013-07-01 sections/intro.tex
Rework the introduction
Using from Arjen, Qiong and Irina. Hopefully this is close to last version.

@@
 
-The numerical modelling of past glaciers and ice sheets allows indirect confrontation between glaciological theories and geomorphological reconstructions, challenging the validity of both.
-Yet a major obstacle in this exercise resides in the large uncertainties that affect climate forcing of numerical glacier models \citep{hebeler-etal-2008}.
-This includes the uncertainty on Earth's present climate, and the larger uncertainty on its changes in the past.
+At the Last Glacial Maximum (LGM), glaciers of a size comparable to the present Greenland and Antarctic ice sheets covered parts of Northern America (Laurentide, Cordilleran and Innuitian ice sheets) and Northern Eurasia (Fennoscandian Ice Sheet).
+Numerical modelling of these former ice masses allows for a comparison between glaciological theories embedded in the models and geomorphological traces underpinning palaeo-glaciological reconstructions.
+Yet, a major obstacle in this exercise resides in large uncertainties concerning climate forcing, typically temperature and precipitation data, needed by numerical glacier models \citep{hebeler-etal-2008}.
+This includes uncertainty in representation of Earth's present climate in regions of poor station coverage, and even larger uncertainty concerning accurate reconstructions of past climate change.
 
-Numerical models of the former North American ice sheets have used various climate forcing methods, ranging from simplistic parametrizations to coupling to General Circulation Models (GCM).
-The latter is arguably the most physically sound way to force an ice sheet model in the past and has proven effective \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013}.
-However the computational demand of GCMs is so that only models of reduced complexity and coarser resolution can run on time-scales of ice sheet growth and decay.
+Arguably the most physically sound way to force an ice sheet model in the past is to couple it with a General Circulation Model (GCM) \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013}.
+However the computational demand of GCMs is such that only models of intermediate complexity can run on the time-scales of tens of thousands of years characteristic of ice sheet growth and decay.
 
-Another method consist in using atmospheric fields produced in uncoupled palæo-climate simulations \citep{huybrechts-tsiobbel-1996}.
-To produce a time-dependant forcing, the GCM output is interpolated in time between different simulations \citep{charbit-etal-2002} or weighted based on ice-core records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}.
-These palæo-climate simulations use ice sheet reconstructions such as the ICE-4G \citep{peltier-1994} as topographic boundary condition, which results in subsequently modelled ice cover consistent with these reconstructions.
+%Irina  I think it would be reasonable to discuss the limitations of GCM in this context
+
+Climatologies obtained from uncoupled GCM palaeo-climate simulations such as produced within the PMIP project \citep{joussaume-taylor-1995} provide a more accurate representation of past climate and are commonly used to force ice sheet models.
+Because they are only available for specific periods of time, they require either an assumption of steady-state \citep{huybrechts-tsiobbel-1996}, or interpolation through time between climatologies from different periods, which can be linear \citep{charbit-etal-2002}, or modulated by a ``glacial index'' weighting function derived from ice core $\delta^{18}$\,O records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}.
+An important drawback in this approach is that palaeo-climate simulations themselves rely on global ice sheet reconstructions such as the ICE-4G \citep{peltier-1994} for their surface topographic boundary condition.
+This indirectly controls subsequently modelled ice sheet geometries, whose spatial extent tend to conform to these reconstructions.
 \julien[noline]{Should this literature list be extended to studies not covering North America?}
 
-%TODO tarasov-peltier-1997
+The other side of the coin is that avoiding this circular dependence unfortunately goes hand-in-hand with a need for simplifying assumptions on Earth's past climate.
+They include energy balance modelling approaches \citep{tarasov-peltier-1997} and geographic parametrizations of surface mass balance \citep{robert-1991} or climate forcing \citep{johnson-fastook-2002}.
+Standing on a middle ground, temperature offset methods \citep{greve-etal-1999,bintanja-etal-2005} make use of the high level of detail available in present climate datasets such as gridded observation datasets, GCM output or reanalyses, while using simplifying representations of past climate deviations from this present state.
 
-Avoiding this circular dependence unfortunately goes with simplistic assumptions on Earth's past climate.
-\citet{robert-1991} use a parametrization of glacier surface mass balance based on elevation.
-\citet{fastook-prentice-1994} use a parametrization of climate forcing based on elevation and distance from a climatic pole initially prepared for Antarctica \citep{johnson-fastook-2002}.
-Finally, \citet{bintanja-etal-2005} apply time-dependant temperature offsets to a reference present-day climate obtained from climate reanalysis data.
-We use a similar method.
+Here, we propose to address some of the uncertainties concerning climate forcing of numerical glacier models by evaluating their response, in terms of glacier extent, to inputs from several climate datasets.
+Few studies of this kind are presently available.
+\citet{quiquet-etal-2012} assessed the sensitivity of a Greenland ice sheet model to various atmospheric forcing, including a regional parametrization \citep{fausto-etal-2009}, output from several GCMs and a climate reanalysis.
+\citet{rodgers-etal-2004,charbit-etal-2007} tested the sensitivity of a model of the Northern Hemisphere ice sheets to different PMIP LGM simulations.
+Here, to limit degrees of freedom in our model, and obtain results independent from palaeo-ice sheet reconstructions such as the ICE-4G, we use a simple temperature offset approach similar to \citep{greve-etal-1999,bintanja-etal-2005} and assess ice sheet model sensitivity to the choice of present-day climate data.
+Rather than using GCM output, we force our model with climate reanalysis data, which include observational information through data assimilation \citep{bengtsson-etal-2007}.
+Furthermore, we focus our study regionally to the former Cordilleran Ice Sheet in Northern America.
 
@@
 
-Our area of focus for the present study is the Cordilleran Ice Sheet in north-western North America (Figure 1).
-In a numerical modelling perspective, it is one of the least studied palæo-ice sheets of the Northern Hemisphere, where nevertheless significant geomorphological data is available \citep{jackson-clague-1991}.
-Although generally not the primary target of these simulations, the Cordilleran Ice Sheet has been previously modelled as part of the North American ice sheet complex \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-2004,bintanja-etal-2005,gregoire-etal-2012} and within a larger-scale Arctic or Global numerical domain \citep{huybrechts-tsiobbel-1996,charbit-etal-2002,johnson-fastook-2002,zweck-huybrechts-2005,abeouchi-etal-2007,charbit-etal-2013}.
+The Cordilleran Ice Sheet (Fig.~\ref{fig:locmap}) covered an area that presently experiences strong regional variations in climate.
+In a numerical modelling perspective, it is one of the least studied palaeo-ice sheets of the Northern Hemisphere, despite the fact that significant geomorphological data is available to constrain its extent \citep{jackson-clague-1991,dukrodkin-1999,kaufman-manley-2004,kleman-etal-2010,margold-etal-2011}.
+Although the Cordilleran Ice Sheet was not the primary target of these simulations, it has previously been modelled as part of efforts to model ice sheets in North America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} or world-wide \citep{yoshimori-etal-2001}.
+While these studies reproduce the magnitude of North American glaciation at the Last Glacial Maximum (LGM) reasonably well, there exist a tendency in simulations independent from ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and continental Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.
 
-While these studies reproduce well the magnitude of the North American glaciation at Last Glacial Maximum, they tend to predict excessive ice cover in parts of northern Yukon Territory and Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.
-
-Here we use a Parallel Ice Sheet Model \citep[PISM,][]{web:pism} to simulate the entire Cordilleran Ice Sheet at the Last Glacial Maximum (LGM).
+Here we use a Parallel Ice Sheet Model \citep[PISM,][]{web:pism} \julien{Recommended by \url{http://www.pism-docs.org/wiki/doku.php?id=citing_pism}} to simulate the extent and thickness of Cordilleran Ice Sheet at the LGM.
 We force our model with multiple climate datasets and compare our results to the mapped LGM ice sheet margin from \citet{dyke-2004}.
-See \citet{quiquet-etal-2012} for a similar comparison on Greenland.
-To avoid any dependence on ice-sheet reconstructions and allow high spatial resolution in our climate forcing, we use present-day climate data and given temperature offsets in a manner similar to \citet{bintanja-etal-2005}.
-To our knowledge, this is the first modelling study which specifically focus on the Cordilleran Ice Sheet since the one by \citep{robert-1991}, who focused specifically on its southern margin.
+To our knowledge, this is the first modelling study which specifically focus on the Cordilleran Ice Sheet since the one by \citet{robert-1991}.
+While \citet{robert-1991} modelling domain covered only the southern part of the ice sheet, our model also includes higher resolution and an improved treatment of ice thermodynamics and bedrock response due to large advances made in ice sheet modelling since then.
+Model set-up is presented in section~\ref{sec:model} and climate forcing in section~\ref{sec:climate}.
+Results are exposed in section~\ref{sec:results} and discussed in section~\ref{sec:discussion}.

### 2013-07-01 cordillera-climate.tex
Put floats to the end for review
Rework the introduction
Using from Arjen, Qiong and Irina. Hopefully this is close to last version.
Physically moved floats to the end

@@
+% cordillera-climate.tex
+% ----------------------------------------------------------------------
+
 % Copernicus manuscript
@@
 
-% Extra packages
+% Coloured hyperlinks
 \usepackage[colorlinks,citecolor=blue]{hyperref}
-\usepackage{framed}
 
@@
 \usepackage[utf8]{inputenc}
-
-% All floats at the end of the manuscripts
-%\usepackage{endfloat}
 
@@
 \begin{document}
+% ----------------------------------------------------------------------
 
-\title{The effect of climate forcing on numerical simulations of the Cordilleran Ice Sheet at the last glacial maximum}
+% Title and authors
+\title{The effect of climate forcing on numerical simulations of the Cordilleran Ice Sheet at the Last Glacial Maximum}
 \runningtitle{Climate forcing for Cordilleran ice sheet simulations}
-
 %\author{Julien Seguinot et al.}
@@
 
+% For Copernicus
 \received{}
@@
 
+% Title
 \maketitle
 
-% ----------------------------------------------------------------------
-\begin{abstract}
-\julien[noline]{The abstract is preliminary.}
-We present an ensemble of numerical simulations of the former Cordilleran ice sheet forced by present-day climatologies from five different datasets.
-Surface air temperature and precipitation rates from the NCEP/NCAR reanalysis, the Climate Forecast System Reanalysis, the ERA-Interim reanalysis and the North American Regional Reanalysis are used to force the ice sheet model PISM.
-Glacial conditions are mimicked by applying constant temperature offsets over the entire modelling domain, and a positive degree-day model is used to compute surface mass balance.
-The modelled ice sheet appears highly sensitive to the choice of climate forcing.
-Best results were obtained from the North American Regional Reanalysis, a high-resolution regional reanalysis which assimilates precipitation.
-\end{abstract}
-% ----------------------------------------------------------------------
-
-% ----------------------------------------------------------------------
+% Sections
+\input{sections/abstract}
 \input{sections/intro}
@@
 \input{sections/discuss}
-% ----------------------------------------------------------------------
+\input{sections/concl}
 
-% ----------------------------------------------------------------------
-\conclusions  %% \conclusions[modified heading if necessary]
-\label{sec:conclusion}
-\emph{Coming soon...}
-% ----------------------------------------------------------------------
-
-% ----------------------------------------------------------------------
+% Acknowledgements
 %\begin{acknowledgements}
 %\end{acknowledgements}
-% ----------------------------------------------------------------------
 
+% References
 \bibliographystyle{copernicus}
 \bibliography{references}
+
+% Floats
+\input{floats/tables}
+\input{floats/figures}
 \end{document}

### 2013-07-01 floats/figures.tex
Physically moved floats to the end

--- 2013-06-21 sections/results.tex
@@
-% ----------------------------------------------------------------------
-\section{Results}
-\label{sec:results}
+% floats/figures.tex
 % ----------------------------------------------------------------------
 
-Using climate forcing from WorldClim data, the NCEP/NCAR, ERA-Iterim, CFSR and NARR reanalyses described in section~\ref{sec:climate} and the ice sheet model described in section~\ref{sec:model}, we run simulations of glacial inception and growth of the Cordilleran ice sheet using different temperature offsets.
+% fig:locmap
+\begin{figure}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics[width=8cm]{cordillera-climate-locmap}
+	\end{center}
+	\caption{Shaded relief map of northern North America.
+The frame delimits this study's modelling domain.
+The outlines of the former ice sheets depict ice cover at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) after \cite{dyke-2004}.
+The background map was made with ETOPO1 (\cite{data:etopo1}) and Natural Earth Data (\cite{data:naturalearth}).}
+	\label{fig:locmap}
+\end{figure}
 
+% fig:temp
+\begin{figure*}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics[width=13cm]{cordillera-climate-temp}
+	\end{center}
+	\caption{Summer temperature maps from the five datasets used in this study and winter temperature map from the WorldClim dataset.}
+	\label{fig:temp}
+\end{figure*}
+
+% fig:prec
+\begin{figure*}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics[width=13cm]{cordillera-climate-prec}
+	\end{center}
+	\caption{Winter precipitation rate maps from the five datasets used in this study and summer precipitation map from the WorldClim dataset.}
+	\label{fig:prec}
+\end{figure*}
+
+% fig:topo
+\begin{figure*}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics[width=13cm]{cordillera-climate-topo}
+	\end{center}
+	\caption{Reference topography used for temperature lapse-rate corrections from the five climate datasets used in the study and ETOPO1 topography used as basal condition for the ice flow model.}
+	\label{fig:topo}
+\end{figure*}
+
+% fig:cool06
 \begin{figure*}[t]
@@
 
-Figure~\ref{fig:cool06} shows the outcome of simulations run with a 6\degC temperature offset for each of the 6 climate forcings.
-In all simulations, ice accumulates and glaciers form as a result of the artificially cooled climate.
-Yet the magnitude of glaciation appears very different from one climate forcing to another.
-Whereas the NCAR and CFSR forcings produce a large ice sheet that covers most of the model domain land, the WorldClim, ERA-Interim and NARR forcings lead to more restrictive ice cover, bounded to mountain ice caps in the case of WorldClim data.
-The smoothing of CFSR precipitation data has limited visual effect on the resulting ice sheet geometry.
-
+% fig:extent
 \begin{figure*}[t]
@@
 
-As we aim to model an ice sheet approaching LGM size, and consider temperature offset as an unknown parameter in this study, we ran simulations using offset values ranging from 2 to 9\degC for each climate forcing.
-
-Figure~\ref{fig:extent} shows the extent of ice cover at the end of each of these~(48) simulations, grouped by climate forcing.
-It is again visible that different forcings generally lead to very different final ice cover.
-Notably, for the smallest temperature offset value of 2\degC that was used, NCAR and CFSR simulations lead to an ice-sheet, whereas WorldClim, NARR and ERA-Interim produced local ice caps only, restricted to the Wrangell-St.~Elias mountains in the case of WorldClim, an area presently glaciated.
-Several of the NCAR and CFSR simulations produced oversized ice-sheets whose extent is primirily bounded by the model domain boundary conditions rather than physical processes.
-Differences in regional patterns can also be noted.
-For instance ERA-Interim ice sheets generally appear more northernly-centered than NARR ice sheets.
-
+% fig:ivolarea
 \begin{figure}[t]
@@
 
-Final ice volume and final ice coverage at the end of all simulations is quantified in Figure~\ref{fig:ivolarea}.
-The NCEP/NCAR and CFSR forcing lead to a much larger and more extensive glaciation than the ERA-Interim, NARR and WorldClim forcings.
-The ERA-Interim and NARR lead to similar ice volumes and glaciated areas, yet regional patterns are different, as shown by Figure~\ref{fig:extent}.
-
+% fig:best
 \begin{figure*}[t]
@@
 \end{figure*}
-
-To compare modelled ice sheets of similar size, we selected for each climate forcing the simulation that lead to a glaciated area closest to an arbitrary target of $4\,\times10^6\,\unit{km^2}$.
-These qualitative ``best'' runs are presented in Figure~\ref{fig:best} along the associated temperature offset values.
-We observe that although similar in size, the selected ice sheets are different in shape and flow patterns.
-More specifically, NCEP/NCAR, CFSR and ERA-Interim ice sheets cover large territories in Yukon and Alaska that where shown to have remained ice-free for at least several glacial cycles \citep{dukrodkin-1999,kaufman-manley-2004}.

### 2013-07-01 floats/tables.tex
Physically moved floats to the end

--- 2013-06-13 sections/climate.tex
@@
-% ----------------------------------------------------------------------
-\section{Climate forcing}
-\label{sec:climate}
-% ----------------------------------------------------------------------
+% floats/tables.tex
 
-We force PISM with temperature and precipitation data from one observational dataset, three global and one regional reanalyses.
-Here we present this climate data and describe how we use it as a forcing.
-
-\subsection{Observation data}
-
-WorldClim is a high-resolution climate dataset built from observation data \citep{data:worldclim}.
-These climate surfaces were generated by spatial interpolation between a large set of weather stations over global land areas, using 30\,arc\,s-aggregated hole-filled SRTM elevation data.
-This provides a resolution much higher than attained by general circulation models.
-
-\begin{figure*}[t]
-	\vspace*{2mm}
-	\begin{center}
-		\includegraphics[width=13cm]{cordillera-climate-temp}
-	\end{center}
-	\caption{Summer temperature maps from the five datasets used in this study and winter temperature map from the WorldClim dataset.}
-	\label{fig:temp}
-\end{figure*}
-
-Figure~\ref{fig:temp} shows the spatial distribution of summer (JJA) and winter (DJF) air surface temperatures in WorldClim.
-Temperatures generally decrease from south to north and regions further inland experience colder winters.
-It should be noted that the temperature gradient is much stronger in the winter than it is in the summer.
-Temperature reach well above zero during the summer months over the entire modelling domain, except of the highest peaks.
-In other words, there is a strong contrast in seasonality between coastal and inland regions, even and regions where mean annual temperatures are well below freezing do experience warm summers.
-
-\begin{figure*}[t]
-	\vspace*{2mm}
-	\begin{center}
-		\includegraphics[width=13cm]{cordillera-climate-prec}
-	\end{center}
-	\caption{Winter precipitation rate maps from the five datasets used in this study and summer precipitation map from the WorldClim dataset.}
-	\label{fig:prec}
-\end{figure*}
-
-Figure~\ref{fig:prec} shows the spatial distribution of DJF and JJA precipitation rates in WorldClim.
-Coastal regions generally receive much more precipitation than inland ones as a result of the continuous orographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascades.
-From a glacier mass-balance point of view, this contrast is made even stronger by the difference in timing of the precipitation peak through the year.
-Although coastal regions experience most precipitation as snow during the accumulation season, inland regions experience dry winters and most of the precipitation falls as rain during the summer months.
-
-In regions like Northern Yukon and Alaska, dry winters and warm summers are unfavourable conditions to ice accumulation and glacial inception despite of the strongly negative present mean annual temperatures.
-In order to account for these strong gradients in seasonality, we use monthly averages of temperature and precipitation to drive the ice sheet model.
-
-The density of weather stations used by WorldClim in the Northern American Cordillera is highly inhomogeneous.
-If good coverage exists in the southern parts of our modelling domain, several hundred kilometres can separate neighbouring stations in the north \citep{data:worldclim}.
-A second drawback of the dataset in the scope of this study is the lack of data across oceans, particularly on the Pacific continental shelf where glaciers have advanced during the last glacial cycle \citep{jackson-clague-1991}.
-
-% ----------------------------------------------------------------------
-
-\subsection{Reanalysis data}
-
+% tab:reanalyses
 \begin{table*}[t]
@@
 \end{table*}
-
-Apart from the WorldClim data, we use surface air temperatures and precipitation rates from three global and one regional climate reanalyses to force the ice sheet model: the NCEP/NCAR reanalysis, the ERA-Interim reanalysis, the Climate System Forecast Reanalysis (CFSR) and the North American Regional Reanalysis (NARR).
-Monthly climatologies from NCEP/NCAR and NARR reanalysis were provided by the NOAA/OAR/ESRL Physical Sciences Division, Boulder, Colorado, USA, from their website at \url{http://www.esrl.noaa.gov/psd/} whereas monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean time series.
-Further information on the data used is gathered in Table \ref{tab:reanalyses}.
-
-The spatial distribution of JJA air surface temperatures and DJF precipitation rates from the four reanalyses datasets are shown in Figures~\ref{fig:temp} and~\ref{fig:prec}.
-Summer temperature and winter precipitation are most relevant to the glacier model as they respectively drive summer melt and winter accumulation.
-
-As a mixed product from observations and circulation modelling, we believe that reanalysis may perform best in sparsely monitored regions such as the northernmost American Cordillera.
-
-% ----------------------------------------------------------------------
-
-\subsection{Preprocessing and lapse-rate corrections}
-
-Air surface temperature and precipitation from the WorldClim data, the NCEP/NCAR, ERA-Interim, CFSR and NARR reanalyses were re-projected to Canadian Atlas Lambert conformal conic projection (EPSG code~3978) and bilinearly interpolated to the model resolution of 10\,km.
-In addition, WorldClim data was extrapolated to oceans using a nearest-neighbour approach.
-These last two steps are not presented on figures~\ref{fig:temp} and~\ref{fig:prec}.
-
-For the CFSR, an alternative forcing was prepared by smoothing the precipitation field in order to correct for artefacts seen in figure~\ref{fig:prec}.
-This was done by averaging data locally in a circular neighbourhood of 7\,pixels in diameter prior to re-projection.
-Yet we will show that this smoothing has very limited effect on the ice sheet model outcome.
-These preprocessing steps were made in GRASS~GIS.
-
-\begin{figure*}[t]
-	\vspace*{2mm}
-	\begin{center}
-		\includegraphics[width=13cm]{cordillera-climate-topo}
-	\end{center}
-	\caption{Reference topography used for temperature lapse-rate corrections from the five climate datasets used in the study and ETOPO1 topography used as basal condition for the ice flow model.}
-	\label{fig:topo}
-\end{figure*}
-
-Throughout the simulation, the numerical model dynamically applies temperature lapse-rate corrections that account for differences between the climate forcing reference topography and the ice flow model topography on the one hand, and the evolution of ice thickness on the over hand.
-For reanalysis data, the circulation model surface topography (or surface geopotential height) was used as reference topography.
-For the WorldClim data, the aggregated hole-filled SRTM altitude field is used as reference topography.
-As a basal condition for the ice flow model, we use the ETOPO1\citep{data:etopo1} combined topography and bathymetry dataset.
-The reference topographies from the five forcing datasets, alongside the ETOPO1 data are shown in figure~\ref{fig:topo}.
-A lapse-rate of 6\unit{\degree C\,km^{-1}} is used in all simulations.
-No lapse-rate corrections apply to precipitation rates.

### 2013-07-01 sections/abstract.tex
Physically moved floats to the end

% section/abstract.tex
% ----------------------------------------------------------------------

\begin{abstract}
\julien[noline]{The abstract is preliminary.}
We present an ensemble of numerical simulations of the former Cordilleran ice sheet forced by present-day climatologies from five different datasets. Surface air temperature and precipitation rates from the NCEP/NCAR reanalysis, the Climate Forecast System Reanalysis, the ERA-Interim reanalysis and the North American Regional Reanalysis are used to force the ice sheet model PISM. Glacial conditions are mimicked by applying constant temperature offsets over the entire modelling domain, and a positive degree-day model is used to compute surface mass balance. The modelled ice sheet appears highly sensitive to the choice of climate forcing. Best results were obtained from the North American Regional Reanalysis, a high-resolution regional reanalysis which assimilates precipitation.
\end{abstract}

### 2013-07-01 sections/climate.tex
Physically moved floats to the end

@@
+% sections/climate.tex
 % ----------------------------------------------------------------------
@@
 
-\begin{figure*}[t]
-	\vspace*{2mm}
-	\begin{center}
-		\includegraphics[width=13cm]{cordillera-climate-temp}
-	\end{center}
-	\caption{Summer temperature maps from the five datasets used in this study and winter temperature map from the WorldClim dataset.}
-	\label{fig:temp}
-\end{figure*}
-
 Figure~\ref{fig:temp} shows the spatial distribution of summer (JJA) and winter (DJF) air surface temperatures in WorldClim.
@@
 In other words, there is a strong contrast in seasonality between coastal and inland regions, even and regions where mean annual temperatures are well below freezing do experience warm summers.
-
-\begin{figure*}[t]
-	\vspace*{2mm}
-	\begin{center}
-		\includegraphics[width=13cm]{cordillera-climate-prec}
-	\end{center}
-	\caption{Winter precipitation rate maps from the five datasets used in this study and summer precipitation map from the WorldClim dataset.}
-	\label{fig:prec}
-\end{figure*}
 
@@
 \subsection{Reanalysis data}
-
-\begin{table*}[t]
-	\caption{Characteristic of reanalysis climatologies used to force the ice sheet model.}
-	\label{tab:reanalyses}
-	\vskip4mm
-	\centering
-	\begin{tabular}{lllll}
-		\tophline
-		Reanalysis& Spatial coverage& Averaging period& Resolution& Description\\
-		\middlehline
-		NCEP/NCAR&  global&     1981 -- 2010& 1.875\degree& \citet{data:ncar}\\
-		ERA-Interim&global&     1979 -- 2011& 1.000\degree& \citet{data:erai}\\
-		CFSR&       global&     1979 -- 2010& 0.325\degree& \citet{data:cfsr}\\
-		NARR&       North America& 1979 -- 2000& 32\,km& \citet{data:narr}\\
-		\bottomhline
-	\end{tabular}
-\end{table*}
 
@@
 
-\begin{figure*}[t]
-	\vspace*{2mm}
-	\begin{center}
-		\includegraphics[width=13cm]{cordillera-climate-topo}
-	\end{center}
-	\caption{Reference topography used for temperature lapse-rate corrections from the five climate datasets used in the study and ETOPO1 topography used as basal condition for the ice flow model.}
-	\label{fig:topo}
-\end{figure*}
-
 Throughout the simulation, the numerical model dynamically applies temperature lapse-rate corrections that account for differences between the climate forcing reference topography and the ice flow model topography on the one hand, and the evolution of ice thickness on the over hand.

### 2013-07-01 sections/concl.tex
Physically moved floats to the end

--- 2013-07-01 sections/abstract.tex
@@
-% section/abstract.tex
+% section/concl.tex
+% ----------------------------------------------------------------------
+\conclusions
+\label{sec:concl}
 % ----------------------------------------------------------------------
 
-\begin{abstract}
-\julien[noline]{The abstract is preliminary.}
-We present an ensemble of numerical simulations of the former Cordilleran ice sheet forced by present-day climatologies from five different datasets.
-Surface air temperature and precipitation rates from the NCEP/NCAR reanalysis, the Climate Forecast System Reanalysis, the ERA-Interim reanalysis and the North American Regional Reanalysis are used to force the ice sheet model PISM.
-Glacial conditions are mimicked by applying constant temperature offsets over the entire modelling domain, and a positive degree-day model is used to compute surface mass balance.
-The modelled ice sheet appears highly sensitive to the choice of climate forcing.
-Best results were obtained from the North American Regional Reanalysis, a high-resolution regional reanalysis which assimilates precipitation.
-\end{abstract}
+\emph{Coming soon...}

### 2013-07-01 sections/intro.tex
Physically moved floats to the end

@@
+% section/intro.tex
 % ----------------------------------------------------------------------
@@
 
-\begin{figure}[t]
-	\vspace*{2mm}
-	\begin{center}
-		\includegraphics[width=8cm]{cordillera-climate-locmap}
-	\end{center}
-	\caption{Shaded relief map of northern North America.
-The frame delimits this study's modelling domain.
-The outlines of the former ice sheets depict ice cover at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) after \citet{dyke-2004}.
-The background map was made with ETOPO1 \citep{data:etopo1} and Natural Earth Data \citep{data:naturalearth}.}
-	\label{fig:locmap}
-\end{figure}
-
 The Cordilleran Ice Sheet (Fig.~\ref{fig:locmap}) covered an area that presently experiences strong regional variations in climate.

### 2013-07-01 sections/results.tex
Physically moved floats to the end

@@
+% sections/results.tex
 % ----------------------------------------------------------------------
@@
 
-\begin{figure*}[t]
-	\vspace*{2mm}
-	\begin{center}
-		\includegraphics[width=13cm]{cordillera-climate-cool06}
-	\end{center}
-	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a climate 6\degC colder than present for each climate forcing.}
-	\label{fig:cool06}
-\end{figure*}
-
 Figure~\ref{fig:cool06} shows the outcome of simulations run with a 6\degC temperature offset for each of the 6 climate forcings.
@@
 The smoothing of CFSR precipitation data has limited visual effect on the resulting ice sheet geometry.
-
-\begin{figure*}[t]
-	\vspace*{2mm}
-	\begin{center}
-		\includegraphics[width=13cm]{cordillera-climate-extent}
-	\end{center}
-	\caption{Extent of ice cover after 10\,kyr as a function of applied temperature offsets for each climate forcing.}
-	\label{fig:extent}
-\end{figure*}
 
@@
 
-\begin{figure}[t]
-	\vspace*{2mm}
-	\begin{center}
-		\includegraphics[width=9cm]{cordillera-climate-ivolarea}
-	\end{center}
-	\julien[inline]{I will make the figure readable in black-and-white and highlight markers corresponding to the ``best'' runs.}
-	\caption{Total ice volume and glaciated area after 10\,kyr as a function of temperature offset and climate forcing.}
-	\label{fig:ivolarea}
-\end{figure}
-
 Final ice volume and final ice coverage at the end of all simulations is quantified in Figure~\ref{fig:ivolarea}.
@@
 The ERA-Interim and NARR lead to similar ice volumes and glaciated areas, yet regional patterns are different, as shown by Figure~\ref{fig:extent}.
-
-\begin{figure*}[t]
-	\vspace*{2mm}
-	\begin{center}
-		\includegraphics[width=13cm]{cordillera-climate-best}
-	\end{center}
-	\caption{Ice surface topography (black contours every 1000\,m) after 10\,kyr using temperature offsets that lead to similar areas of ice cover for each climate forcing.
-14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) ice margin (blue line) from \citet{dyke-2004}.}
-	\label{fig:best}
-\end{figure*}
 

### 2013-07-01 sections/intro.tex
Some commenting and small changes in the intro

@@
 However the computational demand of GCMs is such that only models of intermediate complexity can run on the time-scales of tens of thousands of years characteristic of ice sheet growth and decay.
-
-%Irina  I think it would be reasonable to discuss the limitations of GCM in this context
+\julien{Add words on limitations of GCM in this context?}
 
@@
 This indirectly controls subsequently modelled ice sheet geometries, whose spatial extent tend to conform to these reconstructions.
-\julien[noline]{Should this literature list be extended to studies not covering North America?}
+\julien[noline]{Should I include references not covering the Cordilleran Ice Sheet?}
 
@@
 Although the Cordilleran Ice Sheet was not the primary target of these simulations, it has previously been modelled as part of efforts to model ice sheets in North America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} or world-wide \citep{yoshimori-etal-2001}.
-While these studies reproduce the magnitude of North American glaciation at the Last Glacial Maximum (LGM) reasonably well, there exist a tendency in simulations independent from ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and continental Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.
+While these studies reproduce the magnitude of North American glaciation at LGM reasonably well, there exist a tendency in simulations independent from ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and continental Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.
 
-Here we use a Parallel Ice Sheet Model \citep[PISM,][]{web:pism} \julien{Recommended by \url{http://www.pism-docs.org/wiki/doku.php?id=citing_pism}} to simulate the extent and thickness of Cordilleran Ice Sheet at the LGM.
+Here we use a PISM, a Parallel Ice Sheet Model \citep{web:pism} \julien{Citation recommended by \url{http://www.pism-docs.org/wiki/doku.php?id=citing_pism}} to simulate the extent and thickness of Cordilleran Ice Sheet at the LGM.
 We force our model with multiple climate datasets and compare our results to the mapped LGM ice sheet margin from \citet{dyke-2004}.
 To our knowledge, this is the first modelling study which specifically focus on the Cordilleran Ice Sheet since the one by \citet{robert-1991}.
-While \citet{robert-1991} modelling domain covered only the southern part of the ice sheet, our model also includes higher resolution and an improved treatment of ice thermodynamics and bedrock response due to large advances made in ice sheet modelling since then.
+While the modelling domain of \citet{robert-1991} covered only the southern part of the ice sheet, our model also includes higher resolution and an improved treatment of ice thermodynamics and bedrock response due to large advances made in ice sheet modelling since then.
 Model set-up is presented in section~\ref{sec:model} and climate forcing in section~\ref{sec:climate}.

### 2013-07-08 sections/intro.tex
Apply Irina's corrections in the intro

@@
 However the computational demand of GCMs is such that only models of intermediate complexity can run on the time-scales of tens of thousands of years characteristic of ice sheet growth and decay.
-\julien{Add words on limitations of GCM in this context?}
 
@@
 An important drawback in this approach is that palaeo-climate simulations themselves rely on global ice sheet reconstructions such as the ICE-4G \citep{peltier-1994} for their surface topographic boundary condition.
-This indirectly controls subsequently modelled ice sheet geometries, whose spatial extent tend to conform to these reconstructions.
+This exerts indirect influence on subsequently modelled ice sheet geometries, whose spatial extent tends to conform to these reconstructions.
 \julien[noline]{Should I include references not covering the Cordilleran Ice Sheet?}
@@
 
-Here, we propose to address some of the uncertainties concerning climate forcing of numerical glacier models by evaluating their response, in terms of glacier extent, to inputs from several climate datasets.
+Here, we propose to address some of the uncertainties concerning climate forcing of numerical glacier models by evaluating their responses, in terms of glacier extent, to inputs from several climate datasets.
 Few studies of this kind are presently available.
@@
 \citet{rodgers-etal-2004,charbit-etal-2007} tested the sensitivity of a model of the Northern Hemisphere ice sheets to different PMIP LGM simulations.
-Here, to limit degrees of freedom in our model, and obtain results independent from palaeo-ice sheet reconstructions such as the ICE-4G, we use a simple temperature offset approach similar to \citep{greve-etal-1999,bintanja-etal-2005} and assess ice sheet model sensitivity to the choice of present-day climate data.
+Here, to limit degrees of freedom in our model, and obtain results independent of palaeo-ice sheet reconstructions such as the ICE-4G, we use a simple temperature offset approach similar to \citep{greve-etal-1999,bintanja-etal-2005} and assess ice sheet model sensitivity to the choice of present-day climate data.
 Rather than using GCM output, we force our model with climate reanalysis data, which include observational information through data assimilation \citep{bengtsson-etal-2007}.
-Furthermore, we focus our study regionally to the former Cordilleran Ice Sheet in Northern America.
+Furthermore, we focus our study regionally on the former Cordilleran Ice Sheet in Northern America.
 
@@
 Although the Cordilleran Ice Sheet was not the primary target of these simulations, it has previously been modelled as part of efforts to model ice sheets in North America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} or world-wide \citep{yoshimori-etal-2001}.
-While these studies reproduce the magnitude of North American glaciation at LGM reasonably well, there exist a tendency in simulations independent from ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and continental Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.
+While these studies reproduce the magnitude of North American glaciation at LGM reasonably well, there exists a tendency in the simulations that are independent of ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and continental Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.
 
-Here we use a PISM, a Parallel Ice Sheet Model \citep{web:pism} \julien{Citation recommended by \url{http://www.pism-docs.org/wiki/doku.php?id=citing_pism}} to simulate the extent and thickness of Cordilleran Ice Sheet at the LGM.
+Here we use a PISM, a Parallel Ice Sheet Model \citep{web:pism} \julien{Citation recommended by \url{http://www.pism-docs.org/wiki/doku.php?id=citing_pism}}, to simulate the extent and thickness of Cordilleran Ice Sheet at the LGM.
 We force our model with multiple climate datasets and compare our results to the mapped LGM ice sheet margin from \citet{dyke-2004}.
-To our knowledge, this is the first modelling study which specifically focus on the Cordilleran Ice Sheet since the one by \citet{robert-1991}.
+To our knowledge, this is the first modelling study that specifically focuses on the Cordilleran Ice Sheet since the one by \citet{robert-1991}.
 While the modelling domain of \citet{robert-1991} covered only the southern part of the ice sheet, our model also includes higher resolution and an improved treatment of ice thermodynamics and bedrock response due to large advances made in ice sheet modelling since then.

### 2013-08-20 sections/intro.tex
Small style corrections in introduction

@@
 Climatologies obtained from uncoupled GCM palaeo-climate simulations such as produced within the PMIP project \citep{joussaume-taylor-1995} provide a more accurate representation of past climate and are commonly used to force ice sheet models.
-Because they are only available for specific periods of time, they require either an assumption of steady-state \citep{huybrechts-tsiobbel-1996}, or interpolation through time between climatologies from different periods, which can be linear \citep{charbit-etal-2002}, or modulated by a ``glacial index'' weighting function derived from ice core $\delta^{18}$\,O records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}.
+Because such climatologies are only available for specific periods of time, this method requires either an assumption of steady-state \citep{huybrechts-tsiobbel-1996}, or interpolation through time between climatologies from different periods, which can be linear \citep{charbit-etal-2002}, or modulated by a ``glacial index'' weighting function derived from ice core $\delta^{18}$\,O records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}.
 An important drawback in this approach is that palaeo-climate simulations themselves rely on global ice sheet reconstructions such as the ICE-4G \citep{peltier-1994} for their surface topographic boundary condition.
@@
 \citet{quiquet-etal-2012} assessed the sensitivity of a Greenland ice sheet model to various atmospheric forcing, including a regional parametrization \citep{fausto-etal-2009}, output from several GCMs and a climate reanalysis.
-\citet{rodgers-etal-2004,charbit-etal-2007} tested the sensitivity of a model of the Northern Hemisphere ice sheets to different PMIP LGM simulations.
-Here, to limit degrees of freedom in our model, and obtain results independent of palaeo-ice sheet reconstructions such as the ICE-4G, we use a simple temperature offset approach similar to \citep{greve-etal-1999,bintanja-etal-2005} and assess ice sheet model sensitivity to the choice of present-day climate data.
+\citet{rodgers-etal-2004} and \citet{charbit-etal-2007} tested the sensitivity of a model of the Northern Hemisphere ice sheets to different PMIP LGM simulations.
+Here, to limit degrees of freedom in our model, and obtain results independent of palaeo-ice sheet reconstructions such as the ICE-4G, we use a simple temperature offset approach similar to \citep{greve-etal-1999} and \citet{bintanja-etal-2005} and assess ice sheet model sensitivity to the choice of present-day climate data.
 Rather than using GCM output, we force our model with climate reanalysis data, which include observational information through data assimilation \citep{bengtsson-etal-2007}.
@@
 
-Here we use a PISM, a Parallel Ice Sheet Model \citep{web:pism} \julien{Citation recommended by \url{http://www.pism-docs.org/wiki/doku.php?id=citing_pism}}, to simulate the extent and thickness of Cordilleran Ice Sheet at the LGM.
+Here we use a PISM, a Parallel Ice Sheet Model \citep{web:pism}\julien{Citation recommended by \url{http://www.pism-docs.org/wiki/doku.php?id=citing_pism}}, to simulate the extent and thickness of Cordilleran Ice Sheet at the LGM.
 We force our model with multiple climate datasets and compare our results to the mapped LGM ice sheet margin from \citet{dyke-2004}.
 To our knowledge, this is the first modelling study that specifically focuses on the Cordilleran Ice Sheet since the one by \citet{robert-1991}.
-While the modelling domain of \citet{robert-1991} covered only the southern part of the ice sheet, our model also includes higher resolution and an improved treatment of ice thermodynamics and bedrock response due to large advances made in ice sheet modelling since then.
+While the modelling domain of \citet{robert-1991} covered only the southern part of the ice sheet, our model also includes higher resolution and an improved treatment of ice thermodynamics and bedrock response following the large advances made in ice sheet modelling since then.
 Model set-up is presented in section~\ref{sec:model} and climate forcing in section~\ref{sec:climate}.

### 2013-08-20 sections/model.tex
Physically moved floats to the end
Updated model part for standard deviation

@@
+% sections/model.tex
 % ----------------------------------------------------------------------
@@
 
-We use PISM, a parallel ice sheet model (version stable 0.5.11).
+We use PISM, a Parallel Ice Sheet Model (version stable 0.5.11).
 Documentation on the code can be found in several publications \citep[e.g.,][]{bueler-brown-2009,winkelmann-etal-2011,aschwanden-etal-2012} and more extensively on-line \citep[{\url{http://www.pism-docs.org}},][]{web:pism}.
@@
 Ice ablation is computed from the number of positive degree-days, defined as the integral of temperatures above 0\,\degC in one year.
-Positive degree-days are calculated by the \citet{calov-greve-2005} formula to accounts for day-to-day temperature variability by assuming a normal distribution of variance $\sigma=5\degC$ along a central (input) value $T$:
+Positive degree-days are calculated by the \citet{calov-greve-2005} formula to accounts for day-to-day temperature variability by assuming a normal distribution along a central (input) value $T$:
 
@@
 where the integration interval $[t_0; t_1]$ is one week and $\mathrm{erfc}$ is the complementary error function.
+The temperature standard deviation $\sigma$  is a constant model parameter and was assigned a value of 3.068094\degC which corresponds to the summer (JJA) mean, model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from NARR data similarly to \citet{seguinot-inreview}.
 The ablation model uses degree-day factors of 3.04~\unit{mm\,K^{-1}\,day^{-1}} for snow and 4.59~\unit{mm\,K^{-1}\,day^{-1}} for ice, as derived from mass-balance measurements on present glaciers from the Coast and Rocky Mountains in British Columbia \citep{shea-etal-2009}.

### 2013-08-20 sections/model.tex
Implement Irina's corrections in model part

@@
 Ice ablation is computed from the number of positive degree-days, defined as the integral of temperatures above 0\,\degC in one year.
-Positive degree-days are calculated by the \citet{calov-greve-2005} formula to accounts for day-to-day temperature variability by assuming a normal distribution along a central (input) value $T$:
+Positive degree-days are calculated using the \citet{calov-greve-2005} formula to account for a day-to-day temperature variability by assuming a normal distribution along a central (input) value $T$:
 
@@
 A positive degree-day model is a crude parametrization of physical processes occurring at a glacier's surface.
-Yet we believe that it captures reasonably accurate patterns of accumulation and ablation for our study area considering the employed climate data.
+Yet we believe that it captures reasonably accurately patterns of accumulation and ablation for our study area considering the employed climate data.
 
@@
 
-Ice flow velocities are computed by a heuristic the Shallow Shelf Approximation (SSA) is used as a ``sliding law'' to a polythermal Shallow Ice Approximation (SIA) \citep{bueler-brown-2009}.
-Depth-averaged velocities $\vec{v}$ are computed as a weighted sum of SIA velocities $\vec{v_{\mathrm{SIA}}}$ and SSA velocities $\vec{v_{\mathrm{SSA}}}$ by
+Depth-averaged velocities $\vec{v}$ are computed as a weighted sum of Shallow Ice Approximation (SIA) velocities $\vec{v_{\mathrm{SIA}}}$ and Shallow Shelf Approximation (SSA) velocities $\vec{v_{\mathrm{SSA}}}$ \citep{bueler-brown-2009} by
 
@@
 SIA and SSA velocities are computed by finite difference methods on a 10km-resolution horizontal grid of 300 by 150 points.
-SIA velocities follow temperature-dependant creep.
-Temperatures are derived from an enthalpy formulation \citep{aschwanden-etal-2012} and computed in three dimensions on 51 irregularly spaced layers in ice and 11 regularly spaced layers in rock.
-An homogeneous geothermal heat flux of 70\,\unit{W\,m^{-2}} is provided as lower boundary condition to the thermal model, and the surface air temperature from the various climate forcing is used as upper boundary condition.
+SIA velocities follow temperature-dependent creep.
+Temperatures are derived from an enthalpy formulation \citep{aschwanden-etal-2012} and computed in three dimensions in 51 irregularly spaced layers in ice and 11 regularly spaced layers in bedrock.
+A homogeneous geothermal heat flux of 70\,\unit{W\,m^{-2}} is provided as a lower boundary condition to this composite thermal model, and the surface air temperature from various climate forcing is used as an upper boundary condition.
 
-SSA velocities are determined by a pseudo-plastic sliding law where the yield stress depends on the availability of basal water and a prescribed till friction angle$\phi$.
+SSA velocities are determined by a pseudo-plastic sliding law where the yield stress depends on the availability of basal water and a prescribed till friction angle $\phi$.
 We apply lower friction angles at low elevation to mimic the more slippery beds associated to the presence of marine sediments:
@@
 
-With $\phi$ in degrees and $z$ in meters above present sea level.
+Where $\phi$ is given in degrees and $z$ in meters above the present sea level.
 The glacier bed topography responds to the ice load following a regional isostasy model with flexure and relaxation \citep{lingle-clark-1985}.

### 2013-08-21 sections/model.tex
Arjen's corrections in model section

@@
 
-We use PISM, a Parallel Ice Sheet Model (version stable 0.5.11).
-Documentation on the code can be found in several publications \citep[e.g.,][]{bueler-brown-2009,winkelmann-etal-2011,aschwanden-etal-2012} and more extensively on-line \citep[{\url{http://www.pism-docs.org}},][]{web:pism}.
+We use PISM, a Parallel Ice Sheet Model (version stable 0.5.11, e.g., \citet{bueler-brown-2009,winkelmann-etal-2011,aschwanden-etal-2012,web:pism}.
 
@@
 
-Monthly temperature and precipitation are converted to surface accumulation and ablation by a temperature-index (positive degree-day) model.
-Ice accumulation is equal to precipitation when temperature is below 0\,\degC, and decreases linearly with temperature between 0 and 2\,\degC.
+Monthly mean surface air temperature and monthly precipitation are converted to ice surface accumulation and ablation by a temperature-index (positive degree-day) model\needref.
+Ice accumulation is equal to precipitation when temperature is below 0\,\degC, and decreases to zero linearly with temperature between 0 and 2\,\degC.
 Ice ablation is computed from the number of positive degree-days, defined as the integral of temperatures above 0\,\degC in one year.
-Positive degree-days are calculated using the \citet{calov-greve-2005} formula to account for a day-to-day temperature variability by assuming a normal distribution along a central (input) value $T$:
+Positive degree-days are calculated using a expression by \citet{calov-greve-2005} which account for temperature variability assuming a normal distribution along a central (input) value $T$:
 
@@
 The temperature standard deviation $\sigma$  is a constant model parameter and was assigned a value of 3.068094\degC which corresponds to the summer (JJA) mean, model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from NARR data similarly to \citet{seguinot-inreview}.
-The ablation model uses degree-day factors of 3.04~\unit{mm\,K^{-1}\,day^{-1}} for snow and 4.59~\unit{mm\,K^{-1}\,day^{-1}} for ice, as derived from mass-balance measurements on present glaciers from the Coast and Rocky Mountains in British Columbia \citep{shea-etal-2009}.
-
-A positive degree-day model is a crude parametrization of physical processes occurring at a glacier's surface.
-Yet we believe that it captures reasonably accurately patterns of accumulation and ablation for our study area considering the employed climate data.
+The ablation model incorporates degree-day factors of 3.04~\unit{mm\,K^{-1}\,day^{-1}} for snow and 4.59~\unit{mm\,K^{-1}\,day^{-1}} for ice, as derived from mass-balance measurements on contemporary glaciers from the Coast Mountains and Rocky Mountains in British Columbia \citep{shea-etal-2009}.
 
@@
 
-Depth-averaged velocities $\vec{v}$ are computed as a weighted sum of Shallow Ice Approximation (SIA) velocities $\vec{v_{\mathrm{SIA}}}$ and Shallow Shelf Approximation (SSA) velocities $\vec{v_{\mathrm{SSA}}}$ \citep{bueler-brown-2009} by
+The depth-averaged ice velocity $\vec{v}$ is computed as a weighted sum of the Shallow Ice Approximation (SIA) velocity $\vec{v_{\mathrm{SIA}}}$ and the Shallow Shelf Approximation (SSA) velocity $\vec{v_{\mathrm{SSA}}}$ \citep{bueler-brown-2009} by
 
@@
 
-with
+where
 
@@
 
-SIA and SSA velocities are computed by finite difference methods on a 10km-resolution horizontal grid of 300 by 150 points.
-SIA velocities follow temperature-dependent creep.
-Temperatures are derived from an enthalpy formulation \citep{aschwanden-etal-2012} and computed in three dimensions in 51 irregularly spaced layers in ice and 11 regularly spaced layers in bedrock.
-A homogeneous geothermal heat flux of 70\,\unit{W\,m^{-2}} is provided as a lower boundary condition to this composite thermal model, and the surface air temperature from various climate forcing is used as an upper boundary condition.
+SIA and SSA velocities are computed by finite difference methods on a 10km-resolution horizontal grid of 300 by 150 points (the modelling domain).
+SIA velocities follow temperature-dependent creep\needref.
+Ice temperatures are derived from an enthalpy formulation \citep{aschwanden-etal-2012} and computed in three dimensions in up to 51 irregularly spaced layers in ice and in 11 regularly spaced layers in bedrock.
+A homogeneous geothermal heat flux of 70\,\unit{W\,m^{-2}} provides the lower boundary condition to this composite thermal model, and surface air temperature from various climate forcing provides the upper boundary condition.
 
 SSA velocities are determined by a pseudo-plastic sliding law where the yield stress depends on the availability of basal water and a prescribed till friction angle $\phi$.
-We apply lower friction angles at low elevation to mimic the more slippery beds associated to the presence of marine sediments:
+We apply a range of till friction angles between 10 and 30\,\degree, where the lower friction angles occur at low elevation to mimic the more deformable beds associated with presence of marine sediments:
+\julien[inline]{phi here is independent from basal water which is computed by the model (and sliding depends on both) so the 10 to 30 degree range is meant to reflect different bedrock/sediment surfaces rather than dry vs wet beds.}
 
@@
 
-Where $\phi$ is given in degrees and $z$ in meters above the present sea level.
-The glacier bed topography responds to the ice load following a regional isostasy model with flexure and relaxation \citep{lingle-clark-1985}.
+Where $\phi$ is given in degrees and $z$ in meters above contemporary sea level.
+The elevation of the basal topography responds to ice load following a regional isostasy model that includes lithospheric flexure and mantle relaxation \citep{lingle-clark-1985}.
 
@@
 
-In this study we model glacial inception and growth of the Cordilleran ice sheet to a state approaching its last glacial maximum.
-We mimic glacial conditions by applying constant temperature offsets homogeneously over the entire domain.
-Each simulation starts from ice-free conditions and is run for 10,000 years, a time interval representative of the rapid build-up of the last Cordilleran Ice Sheet from nearly ice-free condition to last glacial maximum extent \citep{clague-1989}.
-Our simulations were run in parallel on 32 cores at the Swedish National Supercomputing Center.
+In this study we model glacial inception and growth of the Cordilleran Ice Sheet towards a configuration as last attained during the LGM.
+We mimic LGM conditions by applying constant temperature offsets homogeneously over the modelling domain.
+Each simulation starts from ice-free conditions and runs for 10\,kyr, a time interval representative of the rapid build-up of the last Cordilleran Ice Sheet from nearly ice-free to full glacial conditions \citep{clague-1989}.
+Our simulations were performed in parallel on 32 cores at the Swedish National Supercomputing Center.

### 2013-08-21 sections/results.tex
Update figures to new simu and extent offset range

@@
 
-As we aim to model an ice sheet approaching LGM size, and consider temperature offset as an unknown parameter in this study, we ran simulations using offset values ranging from 2 to 9\degC for each climate forcing.
+As we aim to model an ice sheet approaching LGM size, and consider temperature offset as an unknown parameter in this study, we ran simulations using offset values ranging from 0 to 15\degC for each climate forcing.
 
-Figure~\ref{fig:extent} shows the extent of ice cover at the end of each of these~(48) simulations, grouped by climate forcing.
+Figure~\ref{fig:extent} shows the extent of ice cover at the end of each of these~(96) simulations, grouped by climate forcing.
 It is again visible that different forcings generally lead to very different final ice cover.

### 2013-08-21 data/get.sh

@@ -2,9 +2,9 @@
 
 # download and strip data files from scooter
 for clim in wcnn erai narr cfsrs7 cfsr ncar; do
-  for cool in $(seq -f %02g 00 09); do
+  for cool in $(seq -f %02g 00 15); do
     atm=cordillera-$clim-10km-bl
-    run=pism/output/$atm/stepcool${cool}sll120+ccli+till1030/y0010000
+    run=pism/output/$atm/stepcool${cool}sll120+ccli+pp+till1030/y0010000
     if [ ! -f $clim-$cool.nc ]; then
       ssh scooter "ncks -O -v lat,lon,topg,thk,usurf,csurf $run.nc /tmp/tmp.nc"
       scp scooter:/tmp/tmp.nc $clim-$cool.nc

### 2013-08-21 figures/plot.py

@@ -184,7 +184,7 @@ def best():
 
     # initialize figure
     climates = ['wcnn', 'erai', 'narr', 'cfsrs7', 'cfsr', 'ncar']
-    offsets  = ['09', '07', '08', '05', '05', '04']
+    offsets  = ['11', '10', '11', '06', '06', '07']
     fig = iplt.gridfigure(mapsize, (2, len(climates)/2), cbar_mode='none')
 
     # loop on climate datasets
@@ -270,27 +270,26 @@ def extent():
 
     # initialize figure
     climates = ['wcnn', 'erai', 'narr', 'cfsrs7', 'cfsr', 'ncar']
-    offsets  = [cool]*6
     fig = iplt.gridfigure(mapsize, (2, len(climates)/2), cbar_mode='single')
 
     # loop on climate datasets
     for i, clim in enumerate(climates):
       ax = plt.axes(fig.grid[i])
       plt.title(labels[clim])
-      icecover = 10
-      for dt in range(9,-1,-1):
+      icecover = 16
+      for dt in range(15,-1,-1):
         nc = Dataset('../data/%s-%02g.nc' % (clim, dt))
         icecover = np.where(nc.variables['thk'][0] > 1, dt, icecover)
       #cs = plt.contour(icecover.T,
-      #  levels = range(0, 10),
+      #  levels = range(0, 16),
       #  colors = 'black',
       #  linewidths = 0.2)
       cs = plt.contourf(icecover.T,
-        levels = range(-1, 10),
+        levels = range(-1, 16),
         cmap   = plt.cm.Blues_r)
 
     # add colorbar and save
-    cb = fig.colorbar(cs, fig.grid.cbar_axes[0], ticks=range(10))
+    cb = fig.colorbar(cs, fig.grid.cbar_axes[0], ticks=range(16))
     cb.set_label(u'temperature offset (°C)')
     _savefig('cordillera-climate-extent')
 
@@ -311,7 +310,7 @@ def ivolarea():
 
     # select input data
     climates = ['wcnn', 'erai', 'narr', 'cfsrs7', 'cfsr', 'ncar']
-    coolings = range(10)
+    coolings = range(16)
 
     # read input data
     for clim in climates:

### 2013-08-21 sections/climate.tex
Irina's corrections in climate section

@@
 We force PISM with temperature and precipitation data from one observational dataset, three global and one regional reanalyses.
-Here we present this climate data and describe how we use it as a forcing.
+Here we present the climate datasets and describe how we use them as a forcing.
 
-\subsection{Observation data}
+\subsection{Observational data}
 
-WorldClim is a high-resolution climate dataset built from observation data \citep{data:worldclim}.
-These climate surfaces were generated by spatial interpolation between a large set of weather stations over global land areas, using 30\,arc\,s-aggregated hole-filled SRTM elevation data.
-This provides a resolution much higher than attained by general circulation models.
+WorldClim is a high-resolution climate dataset built from observational data \citep{data:worldclim}.
+These climate surfaces were generated by spatial interpolation between a large set of measurements taken at weather stations over global land areas, using 30\,arc\,s-aggregated hole-filled SRTM elevation data.
+This provides a resolution much higher than attained by general circulation models in areas where sufficient amount of weather stations exist.
 
 Figure~\ref{fig:temp} shows the spatial distribution of summer (JJA) and winter (DJF) air surface temperatures in WorldClim.
-Temperatures generally decrease from south to north and regions further inland experience colder winters.
+In the study area, temperatures generally decrease from south to north and regions further inland experience colder winters.
 It should be noted that the temperature gradient is much stronger in the winter than it is in the summer.
-Temperature reach well above zero during the summer months over the entire modelling domain, except of the highest peaks.
-In other words, there is a strong contrast in seasonality between coastal and inland regions, even and regions where mean annual temperatures are well below freezing do experience warm summers.
+Temperature rises well above zero during the summer months over the entire modelling domain, except for the highest mountain peaks.
+In other words, there is a strong contrast in seasonality between coastal and inland regions, and regions where mean annual temperatures are well below freezing do experience warm summers.
 
@@
 From a glacier mass-balance point of view, this contrast is made even stronger by the difference in timing of the precipitation peak through the year.
-Although coastal regions experience most precipitation as snow during the accumulation season, inland regions experience dry winters and most of the precipitation falls as rain during the summer months.
+Whereas coastal regions experience most precipitation as snow during the accumulation season, inland regions experience dry winters and most of the precipitation falls as rain during the summer months.
 
@@
 The reference topographies from the five forcing datasets, alongside the ETOPO1 data are shown in figure~\ref{fig:topo}.
-A lapse-rate of 6\unit{\degree C\,km^{-1}} is used in all simulations.
+An annual lapse-rate of 6\unit{\degree C\,km^{-1}} is used in all simulations.
 No lapse-rate corrections apply to precipitation rates.

### 2013-08-21 sections/climate.tex
Arjen's correction in climate section

@@
 
-We force PISM with temperature and precipitation data from one observational dataset, three global and one regional reanalyses.
-Here we present the climate datasets and describe how we use them as a forcing.
+We drive the mass balance routine of PISM with temperature and precipitation data from one observational dataset and climatologies from three global and one regional reanalyses.
+We here present the different climate datasets used in this study.
 
-\subsection{Observational data}
+\subsection{Observational data: WorldClim}
 
-WorldClim is a high-resolution climate dataset built from observational data \citep{data:worldclim}.
-These climate surfaces were generated by spatial interpolation between a large set of measurements taken at weather stations over global land areas, using 30\,arc\,s-aggregated hole-filled SRTM elevation data.
-This provides a resolution much higher than attained by general circulation models in areas where sufficient amount of weather stations exist.
+WorldClim is a high-resolution climate dataset built from meteorological observations \citep{data:worldclim}.
+These climate surfaces were generated by spatial interpolation, using 30\,arc\,s-aggregated hole-filled SRTM elevation data \citep{jarvis-etal-2008}, between a large set of measurements taken at weather stations over global land areas, using .
+This provides a resolution much higher than attained through GCM in areas where sufficient amount of weather stations exist.
 
-Figure~\ref{fig:temp} shows the spatial distribution of summer (JJA) and winter (DJF) air surface temperatures in WorldClim.
-In the study area, temperatures generally decrease from south to north and regions further inland experience colder winters.
-It should be noted that the temperature gradient is much stronger in the winter than it is in the summer.
-Temperature rises well above zero during the summer months over the entire modelling domain, except for the highest mountain peaks.
-In other words, there is a strong contrast in seasonality between coastal and inland regions, and regions where mean annual temperatures are well below freezing do experience warm summers.
+ In the study area, the spatial distributions of summer (JJA) and winter (DJF) air surface temperatures in WorldClim generally show a decrease from south to north (JJA) and from southwest to northeast (DJF), such that regions further inland experience colder winters (Fig.~\ref{fig:temp}).
+It should be noted that temperature gradients are reversed (colder inland than at the coast) and much stronger in winter than in summer.
+Temperatures rises well above zero during the summer months over the entire modelling domain, except for the highest mountain peaks, and even regions where mean annual temperatures are well below freezing do experience warm summers.
+In other words, there is a strong contrast in seasonality between coastal temperate-climate and inland continental-climate regions.
 
-Figure~\ref{fig:prec} shows the spatial distribution of DJF and JJA precipitation rates in WorldClim.
-Coastal regions generally receive much more precipitation than inland ones as a result of the continuous orographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascades.
-From a glacier mass-balance point of view, this contrast is made even stronger by the difference in timing of the precipitation peak through the year.
-Whereas coastal regions experience most precipitation as snow during the accumulation season, inland regions experience dry winters and most of the precipitation falls as rain during the summer months.
+The spatial distribution of JJA and DJF precipitation rates in WorldClim reveal a strong precipitation decline beyond the coastal regions (Fig.~\ref{fig:prec}), primarily as a result of the continuous orographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascades.
+From an ice sheet mass balance point of view, this contrast becomes even stronger by the difference in timing of the precipitation peak through the year.
+Whereas coastal regions experience most precipitation as snow during the accumulation season (DJF), inland regions experience dry winters and most of the annual precipitation consequently falls as rain during the ablation season (JJA).
 
-In regions like Northern Yukon and Alaska, dry winters and warm summers are unfavourable conditions to ice accumulation and glacial inception despite of the strongly negative present mean annual temperatures.
-In order to account for these strong gradients in seasonality, we use monthly averages of temperature and precipitation to drive the ice sheet model.
+In regions such as the northern Yukon Territory and interior Alaska, dry winters and warm summers prohibit ice to accumulate and glaciers to form, despite strongly negative mean annual temperatures.
+In order to account for these strong gradients in seasonality, we use monthly rather than annual averages of temperature and precipitation to drive the ice sheet model.
 
-The density of weather stations used by WorldClim in the Northern American Cordillera is highly inhomogeneous.
-If good coverage exists in the southern parts of our modelling domain, several hundred kilometres can separate neighbouring stations in the north \citep{data:worldclim}.
-A second drawback of the dataset in the scope of this study is the lack of data across oceans, particularly on the Pacific continental shelf where glaciers have advanced during the last glacial cycle \citep{jackson-clague-1991}.
+There are two problems inherent to the use of WorldClim data in our region of study.
+Firstly, the density of weather stations used by WorldClim in the Northern American Cordillera is highly inhomogeneous.
+Although good coverage exists for the southern part of our modelling domain, several hundred kilometres can separate neighbouring stations in the north \citep{data:worldclim}.
+Secondly, WorldClim lacks of data offshore, which would have been particularly useful for the on the Pacific continental shelf which was partly covered by ice during the LGM \citep{jackson-clague-1991}.
 
@@
 
-\subsection{Reanalysis data}
+\subsection{Reanalysis data: NCEP/NCAR, ERA-Interim, CFSR and NARR}
 
-Apart from the WorldClim data, we use surface air temperatures and precipitation rates from three global and one regional climate reanalyses to force the ice sheet model: the NCEP/NCAR reanalysis, the ERA-Interim reanalysis, the Climate System Forecast Reanalysis (CFSR) and the North American Regional Reanalysis (NARR).
-Monthly climatologies from NCEP/NCAR and NARR reanalysis were provided by the NOAA/OAR/ESRL Physical Sciences Division, Boulder, Colorado, USA, from their website at \url{http://www.esrl.noaa.gov/psd/} whereas monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean time series.
-Further information on the data used is gathered in Table \ref{tab:reanalyses}.
+In addition to the WorldClim data, we use surface air temperatures and precipitation rates from three global climate reanalyses and one regional climate reanalysis to inform the mass balance routine of PISM: the NCEP/NCAR reanalysis, the ERA-Interim reanalysis, the Climate System Forecast Reanalysis (CFSR), and the North American Regional Reanalysis (NARR).
+Monthly climatologies from NCEP/NCAR and NARR reanalysis were provided by the NOAA/OAR/ESRL Physical Sciences Division, Boulder, Colorado, USA, from their website at \url{http://www.esrl.noaa.gov/psd/} whereas monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean time series (Table \ref{tab:reanalyses}).
 
-The spatial distribution of JJA air surface temperatures and DJF precipitation rates from the four reanalyses datasets are shown in Figures~\ref{fig:temp} and~\ref{fig:prec}.
+The spatial distributions of JJA air surface temperatures and DJF precipitation rates from the four reanalyses datasets are shown in Figures~\ref{fig:temp} and~\ref{fig:prec}.
 Summer temperature and winter precipitation are most relevant to the glacier model as they respectively drive summer melt and winter accumulation.
 
-As a mixed product from observations and circulation modelling, we believe that reanalysis may perform best in sparsely monitored regions such as the northernmost American Cordillera.
+Because reanalysis are a product combinaing observations and circulation modelling, we believe that the input from reanalyses may perform better than WorldClim in sparsely monitored regions such as the northernmost American Cordillera.
 
@@
 
-Air surface temperature and precipitation from the WorldClim data, the NCEP/NCAR, ERA-Interim, CFSR and NARR reanalyses were re-projected to Canadian Atlas Lambert conformal conic projection (EPSG code~3978) and bilinearly interpolated to the model resolution of 10\,km.
-In addition, WorldClim data was extrapolated to oceans using a nearest-neighbour approach.
+WorldClim, NCEP/NCAR, ERA-Interim, CFSR and NARR climatologies were re-projected to Canadian Atlas Lambert conformal conic projection (EPSG code~3978) and bilinearly interpolated to a 10\,km model grid (see section~\ref{sec:model}).
+In addition, WorldClim data was extrapolated to cover oceans grid points using a nearest-neighbour approach.
 These last two steps are not presented on figures~\ref{fig:temp} and~\ref{fig:prec}.
 
-For the CFSR, an alternative forcing was prepared by smoothing the precipitation field in order to correct for artefacts seen in figure~\ref{fig:prec}.
-This was done by averaging data locally in a circular neighbourhood of 7\,pixels in diameter prior to re-projection.
-Yet we will show that this smoothing has very limited effect on the ice sheet model outcome.
-These preprocessing steps were made in GRASS~GIS.
+An alternative forcing was prepared for CFSR to correct for artefacts seen in its precipitation field (Fig.~\ref{fig:prec}).
+A smoothing of the precipitation field was achieved by averaging data locally in a circular kernel of 7\,pixels in diameter prior to re-projection.
+These preprocessing steps were performed in GRASS~GIS.
 
 Throughout the simulation, the numerical model dynamically applies temperature lapse-rate corrections that account for differences between the climate forcing reference topography and the ice flow model topography on the one hand, and the evolution of ice thickness on the over hand.
-For reanalysis data, the circulation model surface topography (or surface geopotential height) was used as reference topography.
+For reanalysis data, the GCM surface topography (or surface geopotential height) was used as reference topography.
 For the WorldClim data, the aggregated hole-filled SRTM altitude field is used as reference topography.

### 2013-08-21 sections/results.tex
Irina's corrections in result section

@@
 
-As we aim to model an ice sheet approaching LGM size, and consider temperature offset as an unknown parameter in this study, we ran simulations using offset values ranging from 0 to 15\degC for each climate forcing.
+As we aim to model an ice sheet approaching LGM size, and consider temperature offset as an unknown parameter in this study, we run simulations using offset values ranging from 0 to 15\degC for each climate forcing.
 
@@
 
-Final ice volume and final ice coverage at the end of all simulations is quantified in Figure~\ref{fig:ivolarea}.
-The NCEP/NCAR and CFSR forcing lead to a much larger and more extensive glaciation than the ERA-Interim, NARR and WorldClim forcings.
+Final ice volume and final ice coverage is quantified in Figure~\ref{fig:ivolarea}.
+The NCEP/NCAR and CFSR forcing leads to a much larger and more extensive glaciation than the ERA-Interim, NARR and WorldClim forcings.
 The ERA-Interim and NARR lead to similar ice volumes and glaciated areas, yet regional patterns are different, as shown by Figure~\ref{fig:extent}.
 
-To compare modelled ice sheets of similar size, we selected for each climate forcing the simulation that lead to a glaciated area closest to an arbitrary target of $4\,\times10^6\,\unit{km^2}$.
-These qualitative ``best'' runs are presented in Figure~\ref{fig:best} along the associated temperature offset values.
+To compare modelled ice sheets of similar size, we selected for each climate forcing the simulation that leads to a glaciated area closest to an arbitrary target of $2\,\times10^6\,\unit{km^2}$.
+These qualitative ``best'' runs are presented in Figure~\ref{fig:best} along with the associated temperature offset values.
 We observe that although similar in size, the selected ice sheets are different in shape and flow patterns.

### 2013-08-21 sections/results.tex
Arjen's corrections in result part
(Further reworking needed)

@@
 
-Using climate forcing from WorldClim data, the NCEP/NCAR, ERA-Iterim, CFSR and NARR reanalyses described in section~\ref{sec:climate} and the ice sheet model described in section~\ref{sec:model}, we run simulations of glacial inception and growth of the Cordilleran ice sheet using different temperature offsets.
+Using WorldClim, NCEP/NCAR, ERA-Interim, CFSR (smoothed and original datasets) and NARR input climatologies (section~\ref{sec:climate}), PISM (section~\ref{sec:model}) was run using 0 to 15\degC  temperature offsets for 10~kyr to mimic glacial inception and the growth of the Cordilleran ice sheet towards its LGM configuration.
+Important differences in patterns of ice sheet growth, final glacial extent, and volume were attained (Fig.
+\ref{fig:cool06}-\ref{fig:best}).
 
-Figure~\ref{fig:cool06} shows the outcome of simulations run with a 6\degC temperature offset for each of the 6 climate forcings.
-In all simulations, ice accumulates and glaciers form as a result of the artificially cooled climate.
-Yet the magnitude of glaciation appears very different from one climate forcing to another.
-Whereas the NCAR and CFSR forcings produce a large ice sheet that covers most of the model domain land, the WorldClim, ERA-Interim and NARR forcings lead to more restrictive ice cover, bounded to mountain ice caps in the case of WorldClim data.
-The smoothing of CFSR precipitation data has limited visual effect on the resulting ice sheet geometry.
+To contrast the six different input climatologies, we plot simulations run with a 6\degC temperature offset (~\ref{fig:cool06}).
+Clearly, the magnitude of glaciation differs dramatically from one climate forcing to another.
+In fact, differences between least extent and most extent differ by a factor above 3 and least volume and most volume by a factor above 6 (Fig.
+\ref{fig:ivolarea}).
+Whereas the NCEP/NCAR and CFSR climatologies produce a large ice sheet that covers most of the model domain (lower panels, Fig.-\ref{fig:cool06}), WorldClim, ERA-Interim and NARR climatologies lead to more restrictive ice cover (upper panels, Fig.~\ref{fig:cool06}).
+The WorldClim run only resulted in the growth of mountain ice caps bounded to the highest topography.
+The smoothing of CFSR precipitation yielded an insignificant effect on the resulting ice sheet geometry.
 
-As we aim to model an ice sheet approaching LGM size, and consider temperature offset as an unknown parameter in this study, we run simulations using offset values ranging from 0 to 15\degC for each climate forcing.
-
-Figure~\ref{fig:extent} shows the extent of ice cover at the end of each of these~(96) simulations, grouped by climate forcing.
-It is again visible that different forcings generally lead to very different final ice cover.
-Notably, for the smallest temperature offset value of 2\degC that was used, NCAR and CFSR simulations lead to an ice-sheet, whereas WorldClim, NARR and ERA-Interim produced local ice caps only, restricted to the Wrangell-St.~Elias mountains in the case of WorldClim, an area presently glaciated.
-Several of the NCAR and CFSR simulations produced oversized ice-sheets whose extent is primirily bounded by the model domain boundary conditions rather than physical processes.
+As illustrated in Figure \ref{fig:cool06}, spatial patterns and rates of ice sheet growth differ tremendously between the input climatologies, given a certain temperature depression.
+Because we performed a total of 8 simulations for each input climatology, we derived an ensemble of 96 simulations (Fig.
+\ref{fig:extent}).
+Despite temperature depressions as large as 15 °C to the input reference climatologies it is again visible that each leads to a different final ice sheet extent and volume (Fig.
+\ref{fig:extent} and \ref{fig:ivolarea}).
+Notably, for the smallest temperature offset value of 0\degC, NCAR and CFSR simulations lead to an ice-sheet, whereas WorldClim, NARR and ERA-Interim produce local ice caps only.
+In the case of WorldClim, ice is restricted to the Wrangell-St.~Elias mountains, an area presently glaciated.
+Several of the NCAR and CFSR simulations produce oversized ice sheets whose extent is primarily restricted by the domain boundaries.
 Differences in regional patterns can also be noted.
@@
 
-Final ice volume and final ice coverage is quantified in Figure~\ref{fig:ivolarea}.
-The NCEP/NCAR and CFSR forcing leads to a much larger and more extensive glaciation than the ERA-Interim, NARR and WorldClim forcings.
+Final ice volume and final ice coverage are shown in Figure~\ref{fig:ivolarea}.
+NCEP/NCAR and CFSR climatologies lead to much larger and more extensive glaciation than the ERA-Interim, NARR and WorldClim forcings.
 The ERA-Interim and NARR lead to similar ice volumes and glaciated areas, yet regional patterns are different, as shown by Figure~\ref{fig:extent}.
 
-To compare modelled ice sheets of similar size, we selected for each climate forcing the simulation that leads to a glaciated area closest to an arbitrary target of $2\,\times10^6\,\unit{km^2}$.
-These qualitative ``best'' runs are presented in Figure~\ref{fig:best} along with the associated temperature offset values.
+To compare modelled ice sheets of similar size, we selected for each climate forcing the simulation that leads to a glaciated area closest to the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$.
+These qualitative ``best'' runs are presented in Figure~\ref{fig:best} along with their associated temperature depressions.
 We observe that although similar in size, the selected ice sheets are different in shape and flow patterns.
-More specifically, NCEP/NCAR, CFSR and ERA-Interim ice sheets cover large territories in Yukon and Alaska that where shown to have remained ice-free for at least several glacial cycles \citep{dukrodkin-1999,kaufman-manley-2004}.
+More specifically, NCEP/NCAR, CFSR and ERA-Interim ice sheets cover parts of Yukon Territory and Alaska that were shown to have remained ice-free for at least several glacial cycles \citep{dukrodkin-1999,kaufman-manley-2004}.

### 2013-08-21 sections/discuss.tex
Physically moved floats to the end
Irina's corrections to discussion

@@
+% sections/discuss.tex
 % ----------------------------------------------------------------------
@@
 This corresponds to the LGM extent of most of the Cordilleran ice sheet, which occurred later than in the Laurentide ice sheet \citep{dyke-2004}.
-Although the outcome of our numerical simulations is strongly dependant to which dataset is used as climate forcing, common discrepancies between the modelled ice sheet geometry and the ice margin from \citet{dyke-2004} can be observed (Figures~\ref{fig:cool06}-\ref{fig:best}).
+Although the outcome of our numerical simulations strongly depends on which dataset is used as a climate forcing, common discrepancies between the modelled ice sheet geometry and the ice margin mapped by \citet{dyke-2004} can be observed (Figures~\ref{fig:cool06}-\ref{fig:best}).
 
@@
 This can be explained by the potential effects of the growing ice-sheet on regional climate not included in our temperature offset method.
-The Cordilleran ice sheet initiated from the junction of mountain ice caps over the major reliefs.
+The modelled Cordilleran ice sheet initiates from the junction of mountain ice caps over the major reliefs.
 In our simulations, a continuous ice cover quickly forms over the western ranges, where precipitation rates are higher than in the rest of the domain.
 This continuous ice cover may have enhanced the topographical barrier already formed by the western ranges, resulting in fewer precipitation and warmer air in the interior.
-\citet{langen-etal-2012} demonstrated that this process, which they refer as a ``self-inhibiting growth'', potentially prevents a regrowth of the Greenland Ice Sheet to its present size.
+\citet{langen-etal-2012} demonstrated that this process, which they refer as a ``self-inhibiting growth'', may potentially prevent a regrowth of the Greenland Ice Sheet to its present size.
 Alternatively, the Laurentide Ice Sheet, not included in our model, may have formed a buttress against the smaller Cordilleran Ice Sheet and stopped its advance onto the Canadian Prairies.
 
-Our simulations produces anomalous ice cover in parts on the continental shelf in the Arctic Ocean.
+Our simulations produce anomalous ice cover in parts on the continental shelf in the Arctic Ocean.
 These regions experience a marine climate, including lower summer temperature than on the adjacent land (Figures~\ref{fig:temp}).
@@
 A similar effect occurs over Great Bear Lake in some simulations, particularly those driven by the NCEP/NCAR forcing (Figures~\ref{fig:extent}).
-This anomalous ice cover is to be interpreted as an artefact from our simplistic temperature offset method, and has little effect on the rest of the results.
+This anomalous ice cover is to be interpreted as an artefact arising from our simplistic temperature offset method, and has little effect on the rest of the results.
 
@@
 
-The outcome of our numerical simulations is strongly dependant to which dataset is used for climate forcing (Figures~\ref{fig:cool06}-\ref{fig:best}).
-This results from temperature and precipitation differences (Figures~\ref{fig:temp}-\ref{fig:prec}) between forcing datasets, to which the glacier model is highly sensitive.
+The outcome of our numerical simulations strongly depends on which dataset is used for climate forcing (Figures~\ref{fig:cool06}-\ref{fig:best}).
+This results from differences in temperature and precipitation (Figures~\ref{fig:temp}-\ref{fig:prec}) between forcing datasets, to which the glacier model is highly sensitive.
 
-Differences between modelled ice sheet geometries are notably visible in northern Yukon Territory and Alaska, where previous modelling studies commonly produced too extensive ice cover, whereas geomorphological data shows that ice cover have been sparse.
-Concerning reanalysis datasets, we interpret theses differences as mainly the results of different GCM resolutions (Figures~\ref{fig:topo}, Table~\ref{tab:reanalyses}).
+Difference between modelled ice sheet geometries is notably visible in northern Yukon Territory and Alaska, where previous modelling studies commonly produced too extensive ice cover, whereas geomorphological data shows that ice cover was sparse.
+Concerning reanalysis datasets, we interpret these differences as mainly the result of different GCM resolutions (Figures~\ref{fig:topo}, Table~\ref{tab:reanalyses}).
 
 As described in section~\ref{sec:climate}, the topography of the Northern American Cordillera is such that its western ranges form a continuous orographic barrier, causing high precipitation along the Pacific coast while leaving much of the interior arid.
-However the capability of a GCM to reproduce these precipitation contrasts is bound to its resolution.
+However the ability of a GCM to reproduce these contrasts in precipitation is bound to its resolution.
 In a model of coarser resolution such as used in the NCEP/NCAR reanalysis, these high mountain ranges are reduced to smoothed hills and the modelled distribution of precipitation is also smoother.
@@
 Our simplistic temperature offset is certainly a crude simplification of past climate changes over the region.
-Although it is clear that these temperature changes were neither homogeneous nor constant in time, and probably associated with precipitation changes, their patterns are not trivial and potential strongly inter-dependant with the Cordilleran and Laurentide ice sheet evolution.
+Although it is clear that these temperature changes were neither homogeneous nor constant in time, and probably associated with precipitation changes, their patterns are not trivial and may potentially display a strongly inter-dependende with the evolution of the Cordilleran and Laurentide ice sheets.
 A more correct approach would be to use coupling to a GCM of intermediate complexity \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013}.
@@
 Our PDD model do not include refreezing.
-Refreezing of melted snow and ice, however, can form a large proportion of the surface mass-balance of an ice sheet\needref.
+Refreezing of melted snow and ice, however, can form a large proportion of the surface mass-balance of an ice sheet \citep{janssens-huybrechts-2000}.
 Furthermore, we model temperature variability using a constant, homogeneous value of standard deviation.
-This approaches implies large biases of surface mass-balance \citep{charbit-etal-2013,rau-rogozhina-2013,seguinot-inreview}, particularly over a region with such various climates as the Northern American Cordillera.
+This approach implies large biases of surface mass-balance \citep{charbit-etal-2013,rau-rogozhina-2013,seguinot-inreview}, particularly over regions with such various climates as the Northern American Cordillera.

### 2013-08-22 floats/figures.tex
Arjen's corrections in float captions

@@
 	\end{center}
-	\caption{Shaded relief map of northern North America.
-The frame delimits this study's modelling domain.
-The outlines of the former ice sheets depict ice cover at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) after \cite{dyke-2004}.
-The background map was made with ETOPO1 (\cite{data:etopo1}) and Natural Earth Data (\cite{data:naturalearth}).}
+	\caption{Shaded relief map of northern North America with the extent of the Cordilleran (CIS), Laurentide (LIS) and Innuitian (IIS) ice sheets at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) \citep{dyke-2004}.
+The background map consists of ETOPO1 (\cite{data:etopo1}) and Natural Earth Data (\cite{data:naturalearth}).}
 	\label{fig:locmap}
@@
 	\end{center}
-	\caption{Summer temperature maps from the five datasets used in this study and winter temperature map from the WorldClim dataset.}
+	\caption{Summer (JJA) and winter (DJF) temperature maps from CFSR (Climate System Forecast Reanalysis), ERA-Interim reanalysis, NARR (North American Regional Reanalysis), and NCEP/NCAR reanalysis climatologies.}
 	\label{fig:temp}
@@
 	\end{center}
-	\caption{Winter precipitation rate maps from the five datasets used in this study and summer precipitation map from the WorldClim dataset.}
+	\caption{Summer (JJA) and winter (DJF) precipitation maps from CFSR (Climate System Forecast Reanalysis), ERA-Interim reanalysis, NARR (North American Regional Reanalysis), and NCEP/NCAR reanalysis climatologies.}
 	\label{fig:prec}

### 2013-08-22 floats/figures.tex
Improve intro and model sections

@@
 \end{figure}
+
+% fig:topo
+\begin{figure*}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics[width=13cm]{cordillera-climate-topo}
+	\end{center}
+	\caption{Topography maps from CFSR (Climate System Forecast Reanalysis), ERA-Interim reanalysis, NARR (North American Regional Reanalysis), and NCEP/NCAR reanalysis data used as a reference for temperature lapse-rate corrections, and topography map from ETOPO1 used as basal topography.}
+	\label{fig:topo}
+\end{figure*}
 
@@
 	\label{fig:prec}
-\end{figure*}
-
-% fig:topo
-\begin{figure*}[t]
-	\vspace*{2mm}
-	\begin{center}
-		\includegraphics[width=13cm]{cordillera-climate-topo}
-	\end{center}
-	\caption{Reference topography used for temperature lapse-rate corrections from the five climate datasets used in the study and ETOPO1 topography used as basal condition for the ice flow model.}
-	\label{fig:topo}
 \end{figure*}

### 2013-08-22 sections/intro.tex
Improve intro and model sections

@@
 Climatologies obtained from uncoupled GCM palaeo-climate simulations such as produced within the PMIP project \citep{joussaume-taylor-1995} provide a more accurate representation of past climate and are commonly used to force ice sheet models.
-Because such climatologies are only available for specific periods of time, this method requires either an assumption of steady-state \citep{huybrechts-tsiobbel-1996}, or interpolation through time between climatologies from different periods, which can be linear \citep{charbit-etal-2002}, or modulated by a ``glacial index'' weighting function derived from ice core $\delta^{18}$\,O records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}.
+Because such climatologies are only available for specific periods of time, this method requires either an assumption of steady-state \citep{huybrechts-tsiobbel-1996}, or interpolation through time between climatologies from different periods, which can be linear \citep{charbit-etal-2002}, or modulated by a ``glacial index'' weighting function derived from ice core $\delta^{18}$O records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}.
 An important drawback in this approach is that palaeo-climate simulations themselves rely on global ice sheet reconstructions such as the ICE-4G \citep{peltier-1994} for their surface topographic boundary condition.
@@
 They include energy balance modelling approaches \citep{tarasov-peltier-1997} and geographic parametrizations of surface mass balance \citep{robert-1991} or climate forcing \citep{johnson-fastook-2002}.
-Standing on a middle ground, temperature offset methods \citep{greve-etal-1999,bintanja-etal-2005} make use of the high level of detail available in present climate datasets such as gridded observation datasets, GCM output or reanalyses, while using simplifying representations of past climate deviations from this present state.
+Standing on a middle ground, temperature offset methods \citet{greve-etal-1999,bintanja-etal-2005} make use of the high level of detail available in present climate datasets such as gridded observation datasets, GCM output or reanalyses, while using simplifying representations of past climate deviations from this present state.
 
@@
 Although the Cordilleran Ice Sheet was not the primary target of these simulations, it has previously been modelled as part of efforts to model ice sheets in North America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} or world-wide \citep{yoshimori-etal-2001}.
-While these studies reproduce the magnitude of North American glaciation at LGM reasonably well, there exists a tendency in the simulations that are independent of ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and continental Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.
+While these studies reproduce the magnitude of North American glaciation at LGM reasonably well, there exists a tendency in the simulations that are independent of ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and interior Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.
 
-Here we use a PISM, a Parallel Ice Sheet Model \citep{web:pism}\julien{Citation recommended by \url{http://www.pism-docs.org/wiki/doku.php?id=citing_pism}}, to simulate the extent and thickness of Cordilleran Ice Sheet at the LGM.
+Here we use a PISM, a Parallel Ice Sheet Model \citep{web:pism}\julien{Citation recommended by \url{http://www.pism-docs.org/wiki/doku.php?id=citing_pism}}, to simulate the extent and thickness of the Cordilleran Ice Sheet at the LGM.
 We force our model with multiple climate datasets and compare our results to the mapped LGM ice sheet margin from \citet{dyke-2004}.
 To our knowledge, this is the first modelling study that specifically focuses on the Cordilleran Ice Sheet since the one by \citet{robert-1991}.
-While the modelling domain of \citet{robert-1991} covered only the southern part of the ice sheet, our model also includes higher resolution and an improved treatment of ice thermodynamics and bedrock response following the large advances made in ice sheet modelling since then.
+While the modelling domain of \citet{robert-1991} covered only the southern part of the ice sheet, ours encompasses most of the LGM ice cover and adjacent non-glaciated parts of northern Yukon Territory and interior Alaska (Fig.~\ref{fig:locmap}).
+Our model also includes higher resolution and an improved treatment of ice thermodynamics and bedrock response following the large advances made in scientific computing and ice sheet modelling since then.
 Model set-up is presented in section~\ref{sec:model} and climate forcing in section~\ref{sec:climate}.

### 2013-08-22 sections/model.tex
Improve intro and model sections

@@
 We use PISM, a Parallel Ice Sheet Model (version stable 0.5.11, e.g., \citet{bueler-brown-2009,winkelmann-etal-2011,aschwanden-etal-2012,web:pism}.
-
-\subsection{Surface mass-balance}
-
-Monthly mean surface air temperature and monthly precipitation are converted to ice surface accumulation and ablation by a temperature-index (positive degree-day) model\needref.
-Ice accumulation is equal to precipitation when temperature is below 0\,\degC, and decreases to zero linearly with temperature between 0 and 2\,\degC.
-Ice ablation is computed from the number of positive degree-days, defined as the integral of temperatures above 0\,\degC in one year.
-Positive degree-days are calculated using a expression by \citet{calov-greve-2005} which account for temperature variability assuming a normal distribution along a central (input) value $T$:
-
-\begin{equation}
-	\mathrm{PDD} = \int_{t_0}^{t_1} dt \left[
-		\frac{\sigma}{\sqrt{2\pi}}
-		\exp\left({-\frac{T(t)^2}{2\sigma^2}}\right)
-		+\frac{T(t)}{2}
-		\mathrm{erfc} \left(-\frac{T(t)}{\sqrt{2}\sigma}\right)
-	\right],
-\end{equation}
-
-where the integration interval $[t_0; t_1]$ is one week and $\mathrm{erfc}$ is the complementary error function.
-The temperature standard deviation $\sigma$  is a constant model parameter and was assigned a value of 3.068094\degC which corresponds to the summer (JJA) mean, model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from NARR data similarly to \citet{seguinot-inreview}.
-The ablation model incorporates degree-day factors of 3.04~\unit{mm\,K^{-1}\,day^{-1}} for snow and 4.59~\unit{mm\,K^{-1}\,day^{-1}} for ice, as derived from mass-balance measurements on contemporary glaciers from the Coast Mountains and Rocky Mountains in British Columbia \citep{shea-etal-2009}.
+Given basal topography and climate forcing, the model computes the thermal and dynamic state of an ice sheet and the associated lithospheric response.
+Our modelling domain encompasses most of the area covered by the Cordilleran Ice Sheet at LGM with the exception of western Alaska and includes non-glaciated regions in northern Yukon Territory and interior Alaska.
+Our simulations are performed in parallel on 32 cores at the Swedish National Supercomputing Center.
 
@@
 SIA and SSA velocities are computed by finite difference methods on a 10km-resolution horizontal grid of 300 by 150 points (the modelling domain).
-SIA velocities follow temperature-dependent creep\needref.
-Ice temperatures are derived from an enthalpy formulation \citep{aschwanden-etal-2012} and computed in three dimensions in up to 51 irregularly spaced layers in ice and in 11 regularly spaced layers in bedrock.
-A homogeneous geothermal heat flux of 70\,\unit{W\,m^{-2}} provides the lower boundary condition to this composite thermal model, and surface air temperature from various climate forcing provides the upper boundary condition.
+Ice rheology depends on temperature and water content through an enthalpy formulation \citep{aschwanden-blatter-2009,aschwanden-etal-2012}.
+Enthalpy is computed in three dimensions in up to 51 irregularly spaced layers in ice and temperature is computed further in 11 regularly spaced layers in bedrock.
+A homogeneous geothermal heat flux of 70\,\unit{W\,m^{-2}} provides the lower boundary condition to the bedrock thermal model, and surface air temperature from the climate forcing provides the upper boundary condition to the ice enthalpy model.
 
@@
 We apply a range of till friction angles between 10 and 30\,\degree, where the lower friction angles occur at low elevation to mimic the more deformable beds associated with presence of marine sediments:
-\julien[inline]{phi here is independent from basal water which is computed by the model (and sliding depends on both) so the 10 to 30 degree range is meant to reflect different bedrock/sediment surfaces rather than dry vs wet beds.}
+\julien{phi here is independent from basal water which is computed by the model (and sliding depends on both) so the 10 to 30 degree range is meant to reflect different bedrock/sediment surfaces rather than dry vs wet beds.}
 
@@
 Where $\phi$ is given in degrees and $z$ in meters above contemporary sea level.
-The elevation of the basal topography responds to ice load following a regional isostasy model that includes lithospheric flexure and mantle relaxation \citep{lingle-clark-1985}.
+Basal topography is derived from the ETOPO1\citep{data:etopo1} combined topography and bathymetry dataset (Fig.~\ref{fig:topo}).
+It responds to ice load following a regional isostasy model that includes lithospheric flexure and mantle relaxation \citep{lingle-clark-1985}.
 
-\subsection{Simulation design}
+\subsection{Surface mass-balance}
 
-In this study we model glacial inception and growth of the Cordilleran Ice Sheet towards a configuration as last attained during the LGM.
-We mimic LGM conditions by applying constant temperature offsets homogeneously over the modelling domain.
-Each simulation starts from ice-free conditions and runs for 10\,kyr, a time interval representative of the rapid build-up of the last Cordilleran Ice Sheet from nearly ice-free to full glacial conditions \citep{clague-1989}.
-Our simulations were performed in parallel on 32 cores at the Swedish National Supercomputing Center.
+Ice surface accumulation and ablation are computed from monthly mean surface air temperature and monthly precipitation by a temperature-index (positive degree-day) model\citep{hock-2003}.
+Ice accumulation is equal to precipitation when temperature is below 0\,\degC, and decreases to zero linearly with temperature between 0 and 2\,\degC.
+Ice ablation is computed from the number of positive degree-days, defined as the integral of temperatures above 0\,\degC in one year.
+Positive degree-days are calculated using a expression by \citet{calov-greve-2005} that accounts for temperature variability assuming a normal distribution along a central (input) value $T$:
+
+\begin{equation}
+	\mathrm{PDD} = \int_{t_0}^{t_1} dt \left[
+		\frac{\sigma}{\sqrt{2\pi}}
+		\exp\left({-\frac{T(t)^2}{2\sigma^2}}\right)
+		+\frac{T(t)}{2}
+		\mathrm{erfc} \left(-\frac{T(t)}{\sqrt{2}\sigma}\right)
+	\right],
+\end{equation}
+
+where the integration interval $[t_0; t_1]$ is one week and $\mathrm{erfc}$ is the complementary error function.
+The temperature standard deviation $\sigma$ is a constant model parameter and was assigned a value of 3.068094\degC which is the summer (JJA) mean, model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from NARR data similarly to \citet{seguinot-inreview}.
+The ablation model incorporates degree-day factors of 3.04~\unit{mm\,K^{-1}\,day^{-1}} for snow and 4.59~\unit{mm\,K^{-1}\,day^{-1}} for ice, as derived from mass-balance measurements on contemporary glaciers from the Coast Mountains and Rocky Mountains in British Columbia \citep{shea-etal-2009}.
+
+\subsection{Atmospheric corrections}
+
+Prior to surface-mass balance computation, the model dynamically applies a lapse-rate correction to surface air temperature.
+This correction account for evolution of ice thickness on the one hand, and differences between the climate forcing reference topography and the ice flow model topography on the over hand.
+It uses a reference topography distinct from the basal topography and specific to each climate forcing dataset (Fig.~\ref{fig:topo}).
+An annual lapse-rate of 6\unit{\degree C\,km^{-1}} is used in all simulations.
+No lapse-rate corrections applies to precipitation.
+
+As we aim to model glacial inception and growth of the Cordilleran Ice Sheet towards a configuration as last attained during the LGM, we mimic LGM conditions by applying constant temperature offsets homogeneously over the modelling domain.
+
+Each simulation starts from ice-free conditions and runs for 10\,kyr, a time interval representative of the rapid build-up of the last Cordilleran Ice Sheet from nearly ice-free to full glacial conditions \citep{clague-1989,stroeven-etal-2010}.
+Precipitation is corrected using the SeaRISE-Greenland formula for paleo-precipitation correction \citep{huybrechts-2002}, which implies a 7.3\% decrease of precipitation per degree Celsius of temperature lowering.

### 2013-08-23 sections/climate.tex
Improve climate part and add missing references

@@
 
-We drive the mass balance routine of PISM with temperature and precipitation data from one observational dataset and climatologies from three global and one regional reanalyses.
-We here present the different climate datasets used in this study.
+Our atmospheric forcing consists of monthly climatologies of surface air temperature and precipitation obtained from one observational dataset, three global reanalysis and one regional reanalysis.
 
@@
 WorldClim is a high-resolution climate dataset built from meteorological observations \citep{data:worldclim}.
-These climate surfaces were generated by spatial interpolation, using 30\,arc\,s-aggregated hole-filled SRTM elevation data \citep{jarvis-etal-2008}, between a large set of measurements taken at weather stations over global land areas, using .
-This provides a resolution much higher than attained through GCM in areas where sufficient amount of weather stations exist.
+It was built by spatial interpolation over global land areas, using 30\,arc\,s-aggregated hole-filled SRTM elevation data \citep{data:srtm} and GTOPO30 elevation data\citep{data:gtopo30}, between a large set of measurements taken at weather stations worldwide.
+This provides a resolution much higher than attained through GCM in areas where sufficient amount of observations exist.
 
- In the study area, the spatial distributions of summer (JJA) and winter (DJF) air surface temperatures in WorldClim generally show a decrease from south to north (JJA) and from southwest to northeast (DJF), such that regions further inland experience colder winters (Fig.~\ref{fig:temp}).
+Within our modelling domain, the spatial distributions of summer (JJA) and winter (DJF) air surface temperatures in WorldClim generally show a decrease from south to north (JJA) and from south-west to north-east (DJF), such that regions further inland experience colder winters (Fig.~\ref{fig:temp}).
 It should be noted that temperature gradients are reversed (colder inland than at the coast) and much stronger in winter than in summer.
@@
 
-There are two problems inherent to the use of WorldClim data in our region of study.
+There are two problems inherent to the use of WorldClim data in our study area.
 Firstly, the density of weather stations used by WorldClim in the Northern American Cordillera is highly inhomogeneous.
 Although good coverage exists for the southern part of our modelling domain, several hundred kilometres can separate neighbouring stations in the north \citep{data:worldclim}.
-Secondly, WorldClim lacks of data offshore, which would have been particularly useful for the on the Pacific continental shelf which was partly covered by ice during the LGM \citep{jackson-clague-1991}.
+Secondly, WorldClim lacks of data offshore, which would have been particularly useful over the Pacific continental shelf which was partly covered by ice during the LGM \citep{jackson-clague-1991}.
 
@@
 
-In addition to the WorldClim data, we use surface air temperatures and precipitation rates from three global climate reanalyses and one regional climate reanalysis to inform the mass balance routine of PISM: the NCEP/NCAR reanalysis, the ERA-Interim reanalysis, the Climate System Forecast Reanalysis (CFSR), and the North American Regional Reanalysis (NARR).
-Monthly climatologies from NCEP/NCAR and NARR reanalysis were provided by the NOAA/OAR/ESRL Physical Sciences Division, Boulder, Colorado, USA, from their website at \url{http://www.esrl.noaa.gov/psd/} whereas monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean time series (Table \ref{tab:reanalyses}).
+In addition to the WorldClim data, we use surface air temperatures and precipitation rates from three global climate reanalyses and one regional climate reanalysis to inform the mass balance routine of PISM: the NCEP/NCAR reanalysis\citep{data:ncar}, the ERA-Interim reanalysis\citep{data:erai}, the Climate System Forecast Reanalysis (CFSR, \citet{data:cfsr}), and the North American Regional Reanalysis (NARR, \citet{data:narr}).
+Monthly climatologies from NCEP/NCAR and NARR reanalysis were provided by the \citet{web:psd} whereas monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean time series (Table \ref{tab:reanalyses}).
 
-The spatial distributions of JJA air surface temperatures and DJF precipitation rates from the four reanalyses datasets are shown in Figures~\ref{fig:temp} and~\ref{fig:prec}.
+The spatial distributions of JJA air surface temperatures and DJF precipitation rates from the four reanalyses climatologies are shown in Figures~\ref{fig:temp} and~\ref{fig:prec}.
 Summer temperature and winter precipitation are most relevant to the glacier model as they respectively drive summer melt and winter accumulation.
 
-Because reanalysis are a product combinaing observations and circulation modelling, we believe that the input from reanalyses may perform better than WorldClim in sparsely monitored regions such as the northernmost American Cordillera.
+Because reanalyses include observational information through data assimilation, they are more accurate than simple GCMs in densely monitored regions, while offering physically based output in sparsely monitored regions such as the northern part of our modelling domain (Bengtsson et al., 2007).
 
@@
 
-WorldClim, NCEP/NCAR, ERA-Interim, CFSR and NARR climatologies were re-projected to Canadian Atlas Lambert conformal conic projection (EPSG code~3978) and bilinearly interpolated to a 10\,km model grid (see section~\ref{sec:model}).
+WorldClim, NCEP/NCAR, ERA-Interim, CFSR and NARR climatologies were re-projected to Canadian Atlas Lambert conformal conic projection (EPSG code~3978) and bilinearly interpolated to the 10\,km-resolution model grid.
 In addition, WorldClim data was extrapolated to cover oceans grid points using a nearest-neighbour approach.
-These last two steps are not presented on figures~\ref{fig:temp} and~\ref{fig:prec}.
+Note that we choose to present here (Fig.~\ref{fig:temp} and~\ref{fig:prec}) the original rather than interpolated data to emphasize differences in spatial resolution among datasets.
 
@@
 A smoothing of the precipitation field was achieved by averaging data locally in a circular kernel of 7\,pixels in diameter prior to re-projection.
-These preprocessing steps were performed in GRASS~GIS.
-
-Throughout the simulation, the numerical model dynamically applies temperature lapse-rate corrections that account for differences between the climate forcing reference topography and the ice flow model topography on the one hand, and the evolution of ice thickness on the over hand.
-For reanalysis data, the GCM surface topography (or surface geopotential height) was used as reference topography.
-For the WorldClim data, the aggregated hole-filled SRTM altitude field is used as reference topography.
-As a basal condition for the ice flow model, we use the ETOPO1\citep{data:etopo1} combined topography and bathymetry dataset.
-The reference topographies from the five forcing datasets, alongside the ETOPO1 data are shown in figure~\ref{fig:topo}.
-An annual lapse-rate of 6\unit{\degree C\,km^{-1}} is used in all simulations.
-No lapse-rate corrections apply to precipitation rates.
+These preprocessing steps were performed in GRASS~GIS using Python packages Numpy, Scipy and NetCDF4-Python\needref.

### 2013-08-26 sections/results.tex
Remove precip corrections and shrink offset range

@@
 
-Using WorldClim, NCEP/NCAR, ERA-Interim, CFSR (smoothed and original datasets) and NARR input climatologies (section~\ref{sec:climate}), PISM (section~\ref{sec:model}) was run using 0 to 15\degC  temperature offsets for 10~kyr to mimic glacial inception and the growth of the Cordilleran ice sheet towards its LGM configuration.
+Using WorldClim, NCEP/NCAR, ERA-Interim, CFSR (smoothed and original datasets) and NARR input climatologies (section~\ref{sec:climate}), PISM (section~\ref{sec:model}) was run using 0 to 10\degC  temperature offsets for 10~kyr to mimic glacial inception and the growth of the Cordilleran ice sheet towards its LGM configuration.
 Important differences in patterns of ice sheet growth, final glacial extent, and volume were attained (Fig.
@@
 As illustrated in Figure \ref{fig:cool06}, spatial patterns and rates of ice sheet growth differ tremendously between the input climatologies, given a certain temperature depression.
-Because we performed a total of 8 simulations for each input climatology, we derived an ensemble of 96 simulations (Fig.
+Because we performed a total of 11 simulations for each input climatology, we derived an ensemble of 66 simulations (Fig.
 \ref{fig:extent}).
-Despite temperature depressions as large as 15 °C to the input reference climatologies it is again visible that each leads to a different final ice sheet extent and volume (Fig.
+Despite temperature depressions as large as 10\degC to the input reference climatologies it is again visible that each leads to a different final ice sheet extent and volume (Fig.
 \ref{fig:extent} and \ref{fig:ivolarea}).

### 2013-08-26 data/get.sh

@@ -2,9 +2,9 @@
 
 # download and strip data files from scooter
 for clim in wcnn erai narr cfsrs7 cfsr ncar; do
-  for cool in $(seq -f %02g 00 15); do
+  for cool in $(seq -f %02g 00 10); do
     atm=cordillera-$clim-10km-bl
-    run=pism/output/$atm/stepcool${cool}sll120+ccli+pp+till1030/y0010000
+    run=pism/output/$atm/stepcool${cool}sll120+ccli+till1030/y0010000
     if [ ! -f $clim-$cool.nc ]; then
       ssh scooter "ncks -O -v lat,lon,topg,thk,usurf,csurf $run.nc /tmp/tmp.nc"
       scp scooter:/tmp/tmp.nc $clim-$cool.nc

### 2013-08-26 figures/plot.py

@@ -184,7 +184,7 @@ def best():
 
     # initialize figure
     climates = ['wcnn', 'erai', 'narr', 'cfsrs7', 'cfsr', 'ncar']
-    offsets  = ['11', '10', '11', '06', '06', '07']
+    offsets  = ['08', '06', '07', '04', '04', '04']
     fig = iplt.gridfigure(mapsize, (2, len(climates)/2), cbar_mode='none')
 
     # loop on climate datasets
@@ -276,20 +276,20 @@ def extent():
     for i, clim in enumerate(climates):
       ax = plt.axes(fig.grid[i])
       plt.title(labels[clim])
-      icecover = 16
-      for dt in range(15,-1,-1):
+      icecover = 11
+      for dt in range(10,-1,-1):
         nc = Dataset('../data/%s-%02g.nc' % (clim, dt))
         icecover = np.where(nc.variables['thk'][0] > 1, dt, icecover)
       #cs = plt.contour(icecover.T,
-      #  levels = range(0, 16),
+      #  levels = range(0, 11),
       #  colors = 'black',
       #  linewidths = 0.2)
       cs = plt.contourf(icecover.T,
-        levels = range(-1, 16),
+        levels = range(-1, 11),
         cmap   = plt.cm.Blues_r)
 
     # add colorbar and save
-    cb = fig.colorbar(cs, fig.grid.cbar_axes[0], ticks=range(16))
+    cb = fig.colorbar(cs, fig.grid.cbar_axes[0], ticks=range(11))
     cb.set_label(u'temperature offset (°C)')
     _savefig('cordillera-climate-extent')
 
@@ -310,7 +310,7 @@ def ivolarea():
 
     # select input data
     climates = ['wcnn', 'erai', 'narr', 'cfsrs7', 'cfsr', 'ncar']
-    coolings = range(16)
+    coolings = range(11)
 
     # read input data
     for clim in climates:

### 2013-08-27 sections/climate.tex
Polish intro, model and climate sections

@@
 WorldClim is a high-resolution climate dataset built from meteorological observations \citep{data:worldclim}.
-It was built by spatial interpolation over global land areas, using 30\,arc\,s-aggregated hole-filled SRTM elevation data \citep{data:srtm} and GTOPO30 elevation data\citep{data:gtopo30}, between a large set of measurements taken at weather stations worldwide.
+It was built by spatial interpolation over global land areas, using 30\,arc\,s-aggregated hole-filled SRTM elevation data \citep{data:srtm} and GTOPO30 elevation data \citep{data:gtopo30}, between a large set of measurements taken at weather stations worldwide.
 This provides a resolution much higher than attained through GCM in areas where sufficient amount of observations exist.
@@
 
-In addition to the WorldClim data, we use surface air temperatures and precipitation rates from three global climate reanalyses and one regional climate reanalysis to inform the mass balance routine of PISM: the NCEP/NCAR reanalysis\citep{data:ncar}, the ERA-Interim reanalysis\citep{data:erai}, the Climate System Forecast Reanalysis (CFSR, \citet{data:cfsr}), and the North American Regional Reanalysis (NARR, \citet{data:narr}).
+In addition to the WorldClim data, we use surface air temperatures and precipitation rates from three global climate reanalyses and one regional climate reanalysis to inform the mass balance routine of PISM: the NCEP/NCAR reanalysis \citep{data:ncar}, the ERA-Interim reanalysis \citep{data:erai}, the Climate System Forecast Reanalysis \citep[CFSR,][]{data:cfsr}, and the North American Regional Reanalysis \citep[NARR,][]{data:narr}.
 Monthly climatologies from NCEP/NCAR and NARR reanalysis were provided by the \citet{web:psd} whereas monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean time series (Table \ref{tab:reanalyses}).
@@
 Summer temperature and winter precipitation are most relevant to the glacier model as they respectively drive summer melt and winter accumulation.
-
-Because reanalyses include observational information through data assimilation, they are more accurate than simple GCMs in densely monitored regions, while offering physically based output in sparsely monitored regions such as the northern part of our modelling domain (Bengtsson et al., 2007).
+Because reanalyses include observational information through data assimilation, they are more accurate than simple GCMs in densely monitored regions, while offering physically based output in sparsely monitored regions such as the northern part of our modelling domain \citep{bengtsson-etal-2007}.
 
@@
 
-WorldClim, NCEP/NCAR, ERA-Interim, CFSR and NARR climatologies were re-projected to Canadian Atlas Lambert conformal conic projection (EPSG code~3978) and bilinearly interpolated to the 10\,km-resolution model grid.
-In addition, WorldClim data was extrapolated to cover oceans grid points using a nearest-neighbour approach.
-Note that we choose to present here (Fig.~\ref{fig:temp} and~\ref{fig:prec}) the original rather than interpolated data to emphasize differences in spatial resolution among datasets.
+WorldClim, NCEP/NCAR, ERA-Interim, CFSR and NARR climatologies were re-projected to Canadian Atlas Lambert conformal conic projection (EPSG code~3978) and bilinearly interpolated to the 10\,km-resolution model grid using the module r.proj from GRASS~GIS \citep{soft:grass}.
+In addition, WorldClim data was extrapolated in a nearest-neighbour fashion to cover oceans grid points using the interpolate.griddata method from Scipy \citep{soft:scipy}.
+Note that we choose to present here (Fig.~\ref{fig:temp} and~\ref{fig:prec}) the original rather than interpolated data in order to highlight differences related to spatial resolution among datasets.
 
+\julien{Qiong, could you help me with one or two sentences on why we smooth CFSR data?}
 An alternative forcing was prepared for CFSR to correct for artefacts seen in its precipitation field (Fig.~\ref{fig:prec}).
-A smoothing of the precipitation field was achieved by averaging data locally in a circular kernel of 7\,pixels in diameter prior to re-projection.
-These preprocessing steps were performed in GRASS~GIS using Python packages Numpy, Scipy and NetCDF4-Python\needref.
+A smoothing of the precipitation field was achieved by averaging data locally in a circular kernel of 7\,pixels in diameter prior to re-projection using the module r.neigbors from GRASS~GIS \citep{soft:grass}.
+Finally, input files for PISM were prepared using \citet{web:nc4py}.

### 2013-08-27 sections/intro.tex
Polish intro, model and climate sections

@@
 They include energy balance modelling approaches \citep{tarasov-peltier-1997} and geographic parametrizations of surface mass balance \citep{robert-1991} or climate forcing \citep{johnson-fastook-2002}.
-Standing on a middle ground, temperature offset methods \citet{greve-etal-1999,bintanja-etal-2005} make use of the high level of detail available in present climate datasets such as gridded observation datasets, GCM output or reanalyses, while using simplifying representations of past climate deviations from this present state.
+Standing on a middle ground, temperature offset methods \citep{greve-etal-1999,bintanja-etal-2005} make use of the high level of detail available in present climate datasets such as gridded observation datasets, GCM output or reanalyses, while using simplifying representations of past climate deviations from this present state.
 
@@
 \citet{quiquet-etal-2012} assessed the sensitivity of a Greenland ice sheet model to various atmospheric forcing, including a regional parametrization \citep{fausto-etal-2009}, output from several GCMs and a climate reanalysis.
-\citet{rodgers-etal-2004} and \citet{charbit-etal-2007} tested the sensitivity of a model of the Northern Hemisphere ice sheets to different PMIP LGM simulations.
-Here, to limit degrees of freedom in our model, and obtain results independent of palaeo-ice sheet reconstructions such as the ICE-4G, we use a simple temperature offset approach similar to \citep{greve-etal-1999} and \citet{bintanja-etal-2005} and assess ice sheet model sensitivity to the choice of present-day climate data.
+\citet{rodgers-etal-2004} and \citet{charbit-etal-2007} tested the sensitivity of a model of the Northern Hemisphere ice sheets to climate forcing from different PMIP LGM simulations.
+Here, to limit degrees of freedom in our model, and obtain results independent of palaeo-ice sheet reconstructions such as the ICE-4G, we use a simple temperature offset approach similar to \citet{greve-etal-1999} and \citet{bintanja-etal-2005} and assess ice sheet model sensitivity to the choice of present-day climate data.
 Rather than using GCM output, we force our model with climate reanalysis data, which include observational information through data assimilation \citep{bengtsson-etal-2007}.
@@
 In a numerical modelling perspective, it is one of the least studied palaeo-ice sheets of the Northern Hemisphere, despite the fact that significant geomorphological data is available to constrain its extent \citep{jackson-clague-1991,dukrodkin-1999,kaufman-manley-2004,kleman-etal-2010,margold-etal-2011}.
-Although the Cordilleran Ice Sheet was not the primary target of these simulations, it has previously been modelled as part of efforts to model ice sheets in North America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} or world-wide \citep{yoshimori-etal-2001}.
+Although the Cordilleran Ice Sheet was not the primary target of these simulations, it has previously been modelled as part of efforts to model ice sheets in Northern America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} or world-wide \citep{yoshimori-etal-2001}.
 While these studies reproduce the magnitude of North American glaciation at LGM reasonably well, there exists a tendency in the simulations that are independent of ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and interior Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.

### 2013-08-27 sections/model.tex
Improve climate part and add missing references
Remove precip corrections and shrink offset range
Polish intro, model and climate sections

@@
 
-We use PISM, a Parallel Ice Sheet Model (version stable 0.5.11, e.g., \citet{bueler-brown-2009,winkelmann-etal-2011,aschwanden-etal-2012,web:pism}.
+We use PISM, a Parallel Ice Sheet Model, version stable~0.5.12 \citep[e.g.,][]{bueler-brown-2009,winkelmann-etal-2011,aschwanden-etal-2012,web:pism}.
 Given basal topography and climate forcing, the model computes the thermal and dynamic state of an ice sheet and the associated lithospheric response.
-Our modelling domain encompasses most of the area covered by the Cordilleran Ice Sheet at LGM with the exception of western Alaska and includes non-glaciated regions in northern Yukon Territory and interior Alaska.
+Our modelling domain encompasses most of the area covered by the Cordilleran Ice Sheet at LGM except of the western Alaskan ranges and includes non-glaciated regions in northern Yukon Territory and interior Alaska.
 Our simulations are performed in parallel on 32 cores at the Swedish National Supercomputing Center.
@@
 Ice rheology depends on temperature and water content through an enthalpy formulation \citep{aschwanden-blatter-2009,aschwanden-etal-2012}.
-Enthalpy is computed in three dimensions in up to 51 irregularly spaced layers in ice and temperature is computed further in 11 regularly spaced layers in bedrock.
+Enthalpy is computed in three dimensions in up to 51 irregularly spaced layers in ice and temperature is further computed in 11 regularly spaced layers in bedrock.
 A homogeneous geothermal heat flux of 70\,\unit{W\,m^{-2}} provides the lower boundary condition to the bedrock thermal model, and surface air temperature from the climate forcing provides the upper boundary condition to the ice enthalpy model.
@@
 Where $\phi$ is given in degrees and $z$ in meters above contemporary sea level.
-Basal topography is derived from the ETOPO1\citep{data:etopo1} combined topography and bathymetry dataset (Fig.~\ref{fig:topo}).
-It responds to ice load following a regional isostasy model that includes lithospheric flexure and mantle relaxation \citep{lingle-clark-1985}.
+Basal topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography and bathymetry dataset \citep{data:etopo1}.
+\julien{NaturalEarth data are vectors (coastline, rivers, etc.) used in Fig.
+1, not model input.} It responds to ice load following a regional isostasy model that includes lithospheric flexure and mantle relaxation \citep{lingle-clark-1985}.
+
+% ----------------------------------------------------------------------
 
@@
 
-Ice surface accumulation and ablation are computed from monthly mean surface air temperature and monthly precipitation by a temperature-index (positive degree-day) model\citep{hock-2003}.
+Ice surface accumulation and ablation are computed from monthly mean surface air temperature and monthly precipitation by a temperature-index (positive degree-day) model \citep{hock-2003}.
 Ice accumulation is equal to precipitation when temperature is below 0\,\degC, and decreases to zero linearly with temperature between 0 and 2\,\degC.
@@
 where the integration interval $[t_0; t_1]$ is one week and $\mathrm{erfc}$ is the complementary error function.
-The temperature standard deviation $\sigma$ is a constant model parameter and was assigned a value of 3.068094\degC which is the summer (JJA) mean, model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from NARR data similarly to \citet{seguinot-inreview}.
+The temperature standard deviation $\sigma$ is a constant model parameter and was assigned a value of 3.068094\degC, which is the summer (JJA) mean, model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from North American Regional Reanalysis data \citep{data:narr} similarly to \citet{seguinot-inreview}.
 The ablation model incorporates degree-day factors of 3.04~\unit{mm\,K^{-1}\,day^{-1}} for snow and 4.59~\unit{mm\,K^{-1}\,day^{-1}} for ice, as derived from mass-balance measurements on contemporary glaciers from the Coast Mountains and Rocky Mountains in British Columbia \citep{shea-etal-2009}.
+
+% ----------------------------------------------------------------------
 
@@
 
-Prior to surface-mass balance computation, the model dynamically applies a lapse-rate correction to surface air temperature.
-This correction account for evolution of ice thickness on the one hand, and differences between the climate forcing reference topography and the ice flow model topography on the over hand.
+Prior to surface mass balance computation, the model dynamically applies a lapse-rate correction to surface air temperature.
+This correction accounts for evolution of ice thickness on the one hand, and differences between the climate forcing reference topography and the ice flow model basal topography on the over hand.
 It uses a reference topography distinct from the basal topography and specific to each climate forcing dataset (Fig.~\ref{fig:topo}).
@@
 As we aim to model glacial inception and growth of the Cordilleran Ice Sheet towards a configuration as last attained during the LGM, we mimic LGM conditions by applying constant temperature offsets homogeneously over the modelling domain.
-
 Each simulation starts from ice-free conditions and runs for 10\,kyr, a time interval representative of the rapid build-up of the last Cordilleran Ice Sheet from nearly ice-free to full glacial conditions \citep{clague-1989,stroeven-etal-2010}.
-Precipitation is corrected using the SeaRISE-Greenland formula for paleo-precipitation correction \citep{huybrechts-2002}, which implies a 7.3\% decrease of precipitation per degree Celsius of temperature lowering.

### 2013-08-30 floats/figures.tex
Rework results section and figures

@@
 
-% fig:cool06
+% fig:cool05
 \begin{figure*}[t]
@@
 	\begin{center}
-		\includegraphics[width=13cm]{cordillera-climate-cool06}
+		\includegraphics[width=13cm]{cordillera-climate-cool05}
 	\end{center}
-	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a climate 6\degC colder than present for each climate forcing.}
-	\label{fig:cool06}
+	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a climate 5\degC colder than present for each climate forcing.}
+	\label{fig:cool05}
 \end{figure*}

### 2013-08-30 sections/results.tex
Rework results section and figures

--- 2013-08-21 sections/results.tex
@@
 
-Using WorldClim, NCEP/NCAR, ERA-Interim, CFSR (smoothed and original datasets) and NARR input climatologies (section~\ref{sec:climate}), PISM (section~\ref{sec:model}) was run using 0 to 15\degC  temperature offsets for 10~kyr to mimic glacial inception and the growth of the Cordilleran ice sheet towards its LGM configuration.
+Using WorldClim, NCEP/NCAR, ERA-Interim, CFSR (smoothed and original datasets) and NARR input climatologies as a climate forcing (section~\ref{sec:climate}) for the ice sheet model PISM (section~\ref{sec:model}), we run 66~simulations of glacial inception and growth of the Cordilleran ice sheet over 10~kyr using temperature offsets ranging from 0 to 10\degC.
 Important differences in patterns of ice sheet growth, final glacial extent, and volume were attained (Fig.
-\ref{fig:cool06}-\ref{fig:best}).
+\ref{fig:cool05}-\ref{fig:best}).
 
-To contrast the six different input climatologies, we plot simulations run with a 6\degC temperature offset (~\ref{fig:cool06}).
-Clearly, the magnitude of glaciation differs dramatically from one climate forcing to another.
-In fact, differences between least extent and most extent differ by a factor above 3 and least volume and most volume by a factor above 6 (Fig.
-\ref{fig:ivolarea}).
-Whereas the NCEP/NCAR and CFSR climatologies produce a large ice sheet that covers most of the model domain (lower panels, Fig.-\ref{fig:cool06}), WorldClim, ERA-Interim and NARR climatologies lead to more restrictive ice cover (upper panels, Fig.~\ref{fig:cool06}).
+To contrast the six different input climatologies, we plot simulations run with the a 5\degC temperature offset (Fig.~\ref{fig:cool05}).
+The magnitude of glaciation differs dramatically from one climate forcing to another.
+Whereas the NCEP/NCAR and CFSR climatologies produce a large ice sheet that covers most of the model domain (Fig.~\ref{fig:cool05}, lower panels), WorldClim, ERA-Interim and NARR climatologies lead to more restrictive ice cover (Fig.~\ref{fig:cool05}, upper panels).
 The WorldClim run only resulted in the growth of mountain ice caps bounded to the highest topography.
@@
 
-As illustrated in Figure \ref{fig:cool06}, spatial patterns and rates of ice sheet growth differ tremendously between the input climatologies, given a certain temperature depression.
-Because we performed a total of 8 simulations for each input climatology, we derived an ensemble of 96 simulations (Fig.
-\ref{fig:extent}).
-Despite temperature depressions as large as 15 °C to the input reference climatologies it is again visible that each leads to a different final ice sheet extent and volume (Fig.
-\ref{fig:extent} and \ref{fig:ivolarea}).
-Notably, for the smallest temperature offset value of 0\degC, NCAR and CFSR simulations lead to an ice-sheet, whereas WorldClim, NARR and ERA-Interim produce local ice caps only.
-In the case of WorldClim, ice is restricted to the Wrangell-St.~Elias mountains, an area presently glaciated.
-Several of the NCAR and CFSR simulations produce oversized ice sheets whose extent is primarily restricted by the domain boundaries.
-Differences in regional patterns can also be noted.
-For instance ERA-Interim ice sheets generally appear more northernly-centered than NARR ice sheets.
+Across the range of temperature offsets used, final glaciated area and final ice volume differs widely between different input climatologies (Fig.~\ref{fig:ivolarea}).
+Using a 5\degC temperature offset (Fig.~\ref{fig:cool05}) final glaciated area differs between WorldClim and CFSR forcing by a factor of 6 and final volume by a factor of 12 (Fig.~\ref{fig:ivolarea}).
+For particular temperature offset values, given input climatologies result in similar final glaciated areas and ice volume.
+Yet regional patterns are different (Fig.~\ref{fig:extent}).
 
-Final ice volume and final ice coverage are shown in Figure~\ref{fig:ivolarea}.
-NCEP/NCAR and CFSR climatologies lead to much larger and more extensive glaciation than the ERA-Interim, NARR and WorldClim forcings.
-The ERA-Interim and NARR lead to similar ice volumes and glaciated areas, yet regional patterns are different, as shown by Figure~\ref{fig:extent}.
+Under present climate (0\degC offset), WorldClim, NARR and ERA-Interim forcing produce local ice caps in areas presently glaciated.
+This results are fairly consistent with contemporary ice coverage, although our 10~km resolution, and the occurrence of glaciers and ice caps in the ETOPO1 input basal topography do not allow for a detailed comparison.
+CFSR forcing, on the other hand, yields continuous ice cover over the northern Coast Mountains, and NCEP/NCAR forcing results in ice-sheet growth, which is largely inconsistent with the present state of glaciation.
+At the opposite side of the temperature offset range, it can be noted that NCAR and CFSR simulations produce oversized ice sheets whose extent is primarily restricted by the domain boundaries.
 
 To compare modelled ice sheets of similar size, we selected for each climate forcing the simulation that leads to a glaciated area closest to the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$.
-These qualitative ``best'' runs are presented in Figure~\ref{fig:best} along with their associated temperature depressions.
+These qualitative ``best'' runs are presented in Figure~\ref{fig:best} along with their associated temperature depressions and a reconstruction of the ice margin at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) by \citet{dyke-2004}.
 We observe that although similar in size, the selected ice sheets are different in shape and flow patterns.
-More specifically, NCEP/NCAR, CFSR and ERA-Interim ice sheets cover parts of Yukon Territory and Alaska that were shown to have remained ice-free for at least several glacial cycles \citep{dukrodkin-1999,kaufman-manley-2004}.
+More specifically, ice sheets modelled using NCEP/NCAR, CFSR and ERA-Interim input climatologies cover parts of Yukon Territory and Alaska that were shown to have remained ice-free for at least several glacial cycles \citep{dukrodkin-1999,kaufman-manley-2004}.

### 2013-08-30 figures/plot.py

@@ -286,7 +286,7 @@ def extent():
       #  linewidths = 0.2)
       cs = plt.contourf(icecover.T,
         levels = range(-1, 11),
-        cmap   = plt.cm.Blues_r)
+        cmap   = plt.cm.Spectral)
 
     # add colorbar and save
     cb = fig.colorbar(cs, fig.grid.cbar_axes[0], ticks=range(11))

### 2013-08-30 sections/discuss.tex
Rework discussion and add one figure

@@
 This corresponds to the LGM extent of most of the Cordilleran ice sheet, which occurred later than in the Laurentide ice sheet \citep{dyke-2004}.
-Although the outcome of our numerical simulations strongly depends on which dataset is used as a climate forcing, common discrepancies between the modelled ice sheet geometry and the ice margin mapped by \citet{dyke-2004} can be observed (Figures~\ref{fig:cool06}-\ref{fig:best}).
+Although the outcome of our numerical simulations strongly depends on which dataset is used as a climate forcing, common discrepancies between the modelled ice sheet geometry and the ice margin mapped by \citet{dyke-2004} can be observed (Fig.~\ref{fig:cool05}-\ref{fig:best}).
 
-For all climate forcing used, the modelled eastern margin of the ice sheet extends further east than the reconstructed boundary between the Cordilleran and the Laurentide ice sheets.
-This can be explained by the potential effects of the growing ice-sheet on regional climate not included in our temperature offset method.
-The modelled Cordilleran ice sheet initiates from the junction of mountain ice caps over the major reliefs.
+For all climate forcing used, the modelled eastern margin of the ice sheet extends further east than the reconstructed boundary between the Cordilleran and the Laurentide ice sheets (Fig.~\ref{fig:best}).
+This can be explained by the potential effects of the growing ice-sheet on regional climate not included in our model, or by buttressing effects against the Laurentide Ice Sheet.
+The Cordilleran ice sheet initiated from the junction of mountain ice caps over the major reliefs~\citep{clague-1989}.
 In our simulations, a continuous ice cover quickly forms over the western ranges, where precipitation rates are higher than in the rest of the domain.
 This continuous ice cover may have enhanced the topographical barrier already formed by the western ranges, resulting in fewer precipitation and warmer air in the interior.
-\citet{langen-etal-2012} demonstrated that this process, which they refer as a ``self-inhibiting growth'', may potentially prevent a regrowth of the Greenland Ice Sheet to its present size.
+\citet{langen-etal-2012} demonstrated that this process, which they refer to as a ``self-inhibiting growth'', may have been limiting on the build-up of the Greenland Ice Sheet.
 Alternatively, the Laurentide Ice Sheet, not included in our model, may have formed a buttress against the smaller Cordilleran Ice Sheet and stopped its advance onto the Canadian Prairies.
 
-Our simulations produce anomalous ice cover in parts on the continental shelf in the Arctic Ocean.
-These regions experience a marine climate, including lower summer temperature than on the adjacent land (Figures~\ref{fig:temp}).
-However sea-level is lowered by 120~m, turning large parts of the low-sloping continental shelf into land.
-A similar effect occurs over Great Bear Lake in some simulations, particularly those driven by the NCEP/NCAR forcing (Figures~\ref{fig:extent}).
+Our simulations produce anomalous ice cover on parts of the continental shelf in the Arctic Ocean.
+These regions experience a marine climate, including lower summer temperature than on the adjacent land (Fig.~\ref{fig:temp}).
+However in our simulations sea-level is lowered by 120~m, turning large parts of the low-sloping continental shelf into land.
+A similar effect occurs over Great Bear Lake in some simulations, particularly those driven by the NCEP/NCAR forcing (Fig.~\ref{fig:extent}).
 This anomalous ice cover is to be interpreted as an artefact arising from our simplistic temperature offset method, and has little effect on the rest of the results.
@@
 
-The outcome of our numerical simulations strongly depends on which dataset is used for climate forcing (Figures~\ref{fig:cool06}-\ref{fig:best}).
-This results from differences in temperature and precipitation (Figures~\ref{fig:temp}-\ref{fig:prec}) between forcing datasets, to which the glacier model is highly sensitive.
+The outcome of our numerical simulations strongly depends on which dataset is used for climate forcing (Fig.~\ref{fig:cool05}-\ref{fig:best}).
+This results from differences in temperature and precipitation (Fig.~\ref{fig:temp}-\ref{fig:prec}) between forcing datasets, to which the glacier model is highly sensitive.
 
 Difference between modelled ice sheet geometries is notably visible in northern Yukon Territory and Alaska, where previous modelling studies commonly produced too extensive ice cover, whereas geomorphological data shows that ice cover was sparse.
-Concerning reanalysis datasets, we interpret these differences as mainly the result of different GCM resolutions (Figures~\ref{fig:topo}, Table~\ref{tab:reanalyses}).
+Concerning reanalysis datasets, we interpret these differences as mainly the result of different GCM resolutions (Fig.~\ref{fig:topo}, Table~\ref{tab:reanalyses}).
 
@@
 However the ability of a GCM to reproduce these contrasts in precipitation is bound to its resolution.
-In a model of coarser resolution such as used in the NCEP/NCAR reanalysis, these high mountain ranges are reduced to smoothed hills and the modelled distribution of precipitation is also smoother.
+In a model of coarser resolution such as used in the NCEP/NCAR reanalysis, these high mountain ranges are reduced to smoothed hills and the modelled distribution of precipitation is also smoother (Fig.~\ref{fig:oroprecip}).
+This interpretation does not apply to the CFSR, which exhibits high inland precipitation rates despite of its high resolution (Table~\ref{tab:reanalyses}).
 
@@
 
-Our simplistic temperature offset is certainly a crude simplification of past climate changes over the region.
-Although it is clear that these temperature changes were neither homogeneous nor constant in time, and probably associated with precipitation changes, their patterns are not trivial and may potentially display a strongly inter-dependende with the evolution of the Cordilleran and Laurentide ice sheets.
-A more correct approach would be to use coupling to a GCM of intermediate complexity \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013}.
+Our simplistic temperature offset method is certainly a crude simplification of past climate changes over the region.
+Although it is clear that these temperature changes were neither homogeneous nor constant in time, and probably associated with precipitation changes, their patterns are not trivial and may potentially display a strong inter-dependence with the evolution of the Cordilleran and Laurentide ice sheets.
+A more correct, yet more complex approach would be to use coupling to a GCM of intermediate complexity \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013}.
 
@@
 Our PDD model do not include refreezing.
-Refreezing of melted snow and ice, however, can form a large proportion of the surface mass-balance of an ice sheet \citep{janssens-huybrechts-2000}.
-Furthermore, we model temperature variability using a constant, homogeneous value of standard deviation.
+Refreezing of melted snow and ice and its retention in the snow pack, however, can large alter mass-balance at the surface of an ice sheet \citep{janssens-huybrechts-2000}.
+Furthermore, we model temperature variability using a constant, homogeneous value of temperature standard deviation.
 This approach implies large biases of surface mass-balance \citep{charbit-etal-2013,rau-rogozhina-2013,seguinot-inreview}, particularly over regions with such various climates as the Northern American Cordillera.

### 2013-08-31 floats/figures.tex
Rework discussion and add one figure
General reread and polish

@@
 	\end{center}
+    \irina[inline]{include all geographic names}
+    \arjen[inline]{elevation scale white-end confused with glaciers}
+    \arjen[inline]{remove Arctic circle and miles scale}
+    \arjen[inline]{use CIS, LIS and IIS for ice sheet names}
+    \arjen[inline]{use N, S, E and W instead of minus, add 75\degree N label}
 	\caption{Shaded relief map of northern North America with the extent of the Cordilleran (CIS), Laurentide (LIS) and Innuitian (IIS) ice sheets at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) \citep{dyke-2004}.
-The background map consists of ETOPO1 (\cite{data:etopo1}) and Natural Earth Data (\cite{data:naturalearth}).}
+The background map consists of ETOPO1 \citep{data:etopo1} and Natural Earth Data \citep{data:naturalearth}.
+Assembled with GRASS~GIS \citep{soft:grass} and Inkscape \citep{web:inkscape}}
 	\label{fig:locmap}
@@
 	\end{center}
-	\caption{Topography maps from CFSR (Climate System Forecast Reanalysis), ERA-Interim reanalysis, NARR (North American Regional Reanalysis), and NCEP/NCAR reanalysis data used as a reference for temperature lapse-rate corrections, and topography map from ETOPO1 used as basal topography.}
+	\caption{Topography maps from CFSR (Climate System Forecast Reanalysis), ERA-Interim reanalysis, NARR (North American Regional Reanalysis), and NCEP/NCAR reanalysis data used as a reference for temperature lapse-rate corrections, and topography map from ETOPO1 used as basal topography.
+Drawn using Matplotlib \citep{soft:mpl}.}
 	\label{fig:topo}
@@
 	\end{center}
-	\caption{Summer (JJA) and winter (DJF) temperature maps from CFSR (Climate System Forecast Reanalysis), ERA-Interim reanalysis, NARR (North American Regional Reanalysis), and NCEP/NCAR reanalysis climatologies.}
+	\caption{Summer (JJA) and winter (DJF) temperature maps from CFSR (Climate System Forecast Reanalysis), ERA-Interim reanalysis, NARR (North American Regional Reanalysis), and NCEP/NCAR reanalysis climatologies.
+Drawn using Matplotlib \citep{soft:mpl}.}
 	\label{fig:temp}
@@
 	\end{center}
-	\caption{Summer (JJA) and winter (DJF) precipitation maps from CFSR (Climate System Forecast Reanalysis), ERA-Interim reanalysis, NARR (North American Regional Reanalysis), and NCEP/NCAR reanalysis climatologies.}
+	\caption{Summer (JJA) and winter (DJF) precipitation maps from CFSR (Climate System Forecast Reanalysis), ERA-Interim reanalysis, NARR (North American Regional Reanalysis), and NCEP/NCAR reanalysis climatologies.
+Drawn using Matplotlib \citep{soft:mpl}.}
 	\label{fig:prec}
@@
 	\end{center}
-	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a climate 5\degC colder than present for each climate forcing.}
+	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a climate 5\degC colder than present for each climate forcing.
+Drawn using Matplotlib \citep{soft:mpl}.}
 	\label{fig:cool05}
@@
 	\end{center}
-	\caption{Extent of ice cover after 10\,kyr as a function of applied temperature offsets for each climate forcing.}
+	\caption{Extent of ice cover after 10\,kyr as a function of applied temperature offsets for each climate forcing. Drawn using Matplotlib \citep{soft:mpl}.}
 	\label{fig:extent}
@@
 	\julien[inline]{I will make the figure readable in black-and-white and highlight markers corresponding to the ``best'' runs.}
-	\caption{Total ice volume and glaciated area after 10\,kyr as a function of temperature offset and climate forcing.}
+	\caption{Total ice volume and glaciated area after 10\,kyr as a function of temperature offset and climate forcing. Drawn using Matplotlib \citep{soft:mpl}.}
 	\label{fig:ivolarea}
@@
 	\caption{Ice surface topography (black contours every 1000\,m) after 10\,kyr using temperature offsets that lead to similar areas of ice cover for each climate forcing.
-14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) ice margin (blue line) from \citet{dyke-2004}.}
+14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) ice margin (blue line) from \citet{dyke-2004}.
+Drawn using Matplotlib \citep{soft:mpl}.}
 	\label{fig:best}
 \end{figure*}
+
+% fig:oroprecip
+\begin{figure}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics[width=8cm]{cordillera-climate-oroprecip}
+	\end{center}
+	\caption{Schematic representation of the orographic precipitation effect over the Coast Mountains.
+In a GCM of lower resolution, the precipitation peak appear smoother and the precipitation shadow is not as pronounced.
+Drawn using Inkscape \citep{web:inkscape}.}
+	\label{fig:oroprecip}
+\end{figure}

### 2013-08-31 sections/climate.tex
General reread and polish

@@
 It was built by spatial interpolation over global land areas, using 30\,arc\,s-aggregated hole-filled SRTM elevation data \citep{data:srtm} and GTOPO30 elevation data \citep{data:gtopo30}, between a large set of measurements taken at weather stations worldwide.
-This provides a resolution much higher than attained through GCM in areas where sufficient amount of observations exist.
+This provides a resolution much higher than attained by GCMs in areas where sufficient amount of observations exists.
 
@@
 
-In addition to the WorldClim data, we use surface air temperatures and precipitation rates from three global climate reanalyses and one regional climate reanalysis to inform the mass balance routine of PISM: the NCEP/NCAR reanalysis \citep{data:ncar}, the ERA-Interim reanalysis \citep{data:erai}, the Climate System Forecast Reanalysis \citep[CFSR,][]{data:cfsr}, and the North American Regional Reanalysis \citep[NARR,][]{data:narr}.
+In addition to the WorldClim data, we use surface air temperatures and precipitation rates from three global climate reanalyses and one regional climate reanalysis to inform the mass balance routine of PISM: the NCEP/NCAR reanalysis \citep{data:ncar}, the ERA-Interim reanalysis \citep{data:erai}, the Climate System Forecast Reanalysis (CFSR) \citep{data:cfsr}, and the North American Regional Reanalysis (NARR) \citep{data:narr}.
 Monthly climatologies from NCEP/NCAR and NARR reanalysis were provided by the \citet{web:psd} whereas monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean time series (Table \ref{tab:reanalyses}).
@@
 In addition, WorldClim data was extrapolated in a nearest-neighbour fashion to cover oceans grid points using the interpolate.griddata method from Scipy \citep{soft:scipy}.
-Note that we choose to present here (Fig.~\ref{fig:temp} and~\ref{fig:prec}) the original rather than interpolated data in order to highlight differences related to spatial resolution among datasets.
+Note that we choose to present here the original rather than interpolated data in order to highlight differences related to spatial resolution among datasets (Fig.~\ref{fig:temp} and~\ref{fig:prec}).
 

### 2013-08-31 sections/discuss.tex
General reread and polish

@@
 
-\julien[noline]{This sentence may be better placed in the results section.}
 We compare the numerically modelled ice sheet geometries to a reconstruction of the ice margin at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) by \citet{dyke-2004}, based on glacial geomorphology and radiocarbon dating.
@@
 This continuous ice cover may have enhanced the topographical barrier already formed by the western ranges, resulting in fewer precipitation and warmer air in the interior.
-\citet{langen-etal-2012} demonstrated that this process, which they refer to as a ``self-inhibiting growth'', may have been limiting on the build-up of the Greenland Ice Sheet.
+\citet{langen-etal-2012} demonstrated that this process, which they refer to as a ``self-inhibiting growth'', may have been limiting during the build-up of the Greenland Ice Sheet.
 Alternatively, the Laurentide Ice Sheet, not included in our model, may have formed a buttress against the smaller Cordilleran Ice Sheet and stopped its advance onto the Canadian Prairies.
@@
 These regions experience a marine climate, including lower summer temperature than on the adjacent land (Fig.~\ref{fig:temp}).
-However in our simulations sea-level is lowered by 120~m, turning large parts of the low-sloping continental shelf into land.
+However in our simulations, sea-level is lowered by 120~m, turning large parts of the low-sloping continental shelf into land.
 A similar effect occurs over Great Bear Lake in some simulations, particularly those driven by the NCEP/NCAR forcing (Fig.~\ref{fig:extent}).
@@
 
-Difference between modelled ice sheet geometries is notably visible in northern Yukon Territory and Alaska, where previous modelling studies commonly produced too extensive ice cover, whereas geomorphological data shows that ice cover was sparse.
+Difference between modelled ice sheet geometries is notably visible in northern Yukon Territory and Alaska, where previous modelling studies commonly produced too extensive ice cover, whereas geomorphological data shows that the ice cover was sparse.
 Concerning reanalysis datasets, we interpret these differences as mainly the result of different GCM resolutions (Fig.~\ref{fig:topo}, Table~\ref{tab:reanalyses}).
@@
 In a model of coarser resolution such as used in the NCEP/NCAR reanalysis, these high mountain ranges are reduced to smoothed hills and the modelled distribution of precipitation is also smoother (Fig.~\ref{fig:oroprecip}).
-This interpretation does not apply to the CFSR, which exhibits high inland precipitation rates despite of its high resolution (Table~\ref{tab:reanalyses}).
+This interpretation does not apply to the CFSR climatology, which exhibits high inland precipitation rates despite of the high model resolution (Table~\ref{tab:reanalyses}).
 
@@
 
-Some simplifications were made in the surface mass-balance model.
+Furthermore, some simplifications were made in the surface mass balance model.
 Our PDD model do not include refreezing.
-Refreezing of melted snow and ice and its retention in the snow pack, however, can large alter mass-balance at the surface of an ice sheet \citep{janssens-huybrechts-2000}.
-Furthermore, we model temperature variability using a constant, homogeneous value of temperature standard deviation.
+Refreezing of melted snow and ice and retention in the snow pack, however, can largely alter mass balance at the surface of an ice sheet \citep{janssens-huybrechts-2000}.
+Additionally, we model temperature variability using a constant, homogeneous value of temperature standard deviation.
 This approach implies large biases of surface mass-balance \citep{charbit-etal-2013,rau-rogozhina-2013,seguinot-inreview}, particularly over regions with such various climates as the Northern American Cordillera.

### 2013-08-31 sections/intro.tex
General reread and polish

@@
 In a numerical modelling perspective, it is one of the least studied palaeo-ice sheets of the Northern Hemisphere, despite the fact that significant geomorphological data is available to constrain its extent \citep{jackson-clague-1991,dukrodkin-1999,kaufman-manley-2004,kleman-etal-2010,margold-etal-2011}.
-Although the Cordilleran Ice Sheet was not the primary target of these simulations, it has previously been modelled as part of efforts to model ice sheets in Northern America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} or world-wide \citep{yoshimori-etal-2001}.
+Although the Cordilleran Ice Sheet was not the primary target of these simulations, it has previously been modelled as part of efforts to simulate ice sheets in Northern America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} or world-wide \citep{yoshimori-etal-2001}.
 While these studies reproduce the magnitude of North American glaciation at LGM reasonably well, there exists a tendency in the simulations that are independent of ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and interior Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.
@@
 To our knowledge, this is the first modelling study that specifically focuses on the Cordilleran Ice Sheet since the one by \citet{robert-1991}.
-While the modelling domain of \citet{robert-1991} covered only the southern part of the ice sheet, ours encompasses most of the LGM ice cover and adjacent non-glaciated parts of northern Yukon Territory and interior Alaska (Fig.~\ref{fig:locmap}).
+While the modelling domain of \citet{robert-1991} covered only the southern part of the ice sheet, ours encompasses most of the LGM ice extent and adjacent non-glaciated parts of northern Yukon Territory and interior Alaska (Fig.~\ref{fig:locmap}).
 Our model also includes higher resolution and an improved treatment of ice thermodynamics and bedrock response following the large advances made in scientific computing and ice sheet modelling since then.

### 2013-08-31 sections/model.tex
General reread and polish

@@
 
-We use PISM, a Parallel Ice Sheet Model, version stable~0.5.12 \citep[e.g.,][]{bueler-brown-2009,winkelmann-etal-2011,aschwanden-etal-2012,web:pism}.
+We use PISM, a Parallel Ice Sheet Model, version stable~0.5.12 (e.g.,) \citep{bueler-brown-2009,winkelmann-etal-2011,aschwanden-etal-2012,web:pism}.
 Given basal topography and climate forcing, the model computes the thermal and dynamic state of an ice sheet and the associated lithospheric response.
@@
 We apply a range of till friction angles between 10 and 30\,\degree, where the lower friction angles occur at low elevation to mimic the more deformable beds associated with presence of marine sediments:
-\julien{phi here is independent from basal water which is computed by the model (and sliding depends on both) so the 10 to 30 degree range is meant to reflect different bedrock/sediment surfaces rather than dry vs wet beds.}
+\needref
+\julien{phi is independent from basal water which is computed by the model (and sliding depends on both) so the 10 to 30 degree range is meant to reflect different bedrock/sediment surfaces rather than dry vs wet beds.}
 
@@
 where the integration interval $[t_0; t_1]$ is one week and $\mathrm{erfc}$ is the complementary error function.
-The temperature standard deviation $\sigma$ is a constant model parameter and was assigned a value of 3.068094\degC, which is the summer (JJA) mean, model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from North American Regional Reanalysis data \citep{data:narr} similarly to \citet{seguinot-inreview}.
+The temperature standard deviation $\sigma$ is a constant model parameter and was assigned a value of 3.068094\degC, which corresponds to the summer (JJA) mean, model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from North American Regional Reanalysis data \citep{data:narr} in a manner similar to \citet{seguinot-inreview}.
 The ablation model incorporates degree-day factors of 3.04~\unit{mm\,K^{-1}\,day^{-1}} for snow and 4.59~\unit{mm\,K^{-1}\,day^{-1}} for ice, as derived from mass-balance measurements on contemporary glaciers from the Coast Mountains and Rocky Mountains in British Columbia \citep{shea-etal-2009}.
@@
 It uses a reference topography distinct from the basal topography and specific to each climate forcing dataset (Fig.~\ref{fig:topo}).
-An annual lapse-rate of 6\unit{\degree C\,km^{-1}} is used in all simulations.
+An annual lapse-rate of 6\,\unit{\degree C\,km^{-1}} is used in all simulations.
 No lapse-rate corrections applies to precipitation.
 
-As we aim to model glacial inception and growth of the Cordilleran Ice Sheet towards a configuration as last attained during the LGM, we mimic LGM conditions by applying constant temperature offsets homogeneously over the modelling domain.
+As we aim to model glacial inception and growth of the Cordilleran Ice Sheet towards a configuration as last attained during the LGM, we mimic palaeo-climatic conditions by applying constant temperature offsets homogeneously over the modelling domain.
 Each simulation starts from ice-free conditions and runs for 10\,kyr, a time interval representative of the rapid build-up of the last Cordilleran Ice Sheet from nearly ice-free to full glacial conditions \citep{clague-1989,stroeven-etal-2010}.

### 2013-08-31 sections/results.tex
General reread and polish

@@
 Using a 5\degC temperature offset (Fig.~\ref{fig:cool05}) final glaciated area differs between WorldClim and CFSR forcing by a factor of 6 and final volume by a factor of 12 (Fig.~\ref{fig:ivolarea}).
-For particular temperature offset values, given input climatologies result in similar final glaciated areas and ice volume.
-Yet regional patterns are different (Fig.~\ref{fig:extent}).
+For particular temperature offset values, given input climatologies result in similar final glaciated areas and ice volume, yet regional patterns are different (Fig.~\ref{fig:extent}).
 
-Under present climate (0\degC offset), WorldClim, NARR and ERA-Interim forcing produce local ice caps in areas presently glaciated.
-This results are fairly consistent with contemporary ice coverage, although our 10~km resolution, and the occurrence of glaciers and ice caps in the ETOPO1 input basal topography do not allow for a detailed comparison.
+Under present climate (0\degC offset), WorldClim, NARR and ERA-Interim forcing produce local ice caps in areas presently glaciated (Fig.~\ref{fig:extent}).
+This results are fairly consistent with contemporary ice coverage, although both the 10~km horizontal resolution, and the occurrence of glaciers and ice caps in the ETOPO1 input basal topography prevent a detailed comparison.
 CFSR forcing, on the other hand, yields continuous ice cover over the northern Coast Mountains, and NCEP/NCAR forcing results in ice-sheet growth, which is largely inconsistent with the present state of glaciation.
-At the opposite side of the temperature offset range, it can be noted that NCAR and CFSR simulations produce oversized ice sheets whose extent is primarily restricted by the domain boundaries.
+At the opposite side of the temperature offset range, it can be noted that NCEP/NCAR and CFSR forcing produce oversized ice sheets whose extent is primarily restricted by the domain boundaries (Fig.~\ref{fig:extent}).
 

### 2013-09-02 sections/concl.tex
Add conclusions

@@
 
-\emph{Coming soon...}
+Our study shows a strong dependency of the modelled ice sheet to the choice of climate forcing data.
+When using input climatologies from climate reanalyses, spatial resolution of the GCM appears critical for providing the ice sheet model with an accurate precipitation field.
+This confirms results obtained by \citet{quiquet-etal-2012} from various GCM forcing over the Greenland Ice Sheet.
+
+For the Cordilleran Ice Sheet, we achieve a best fit to the mapped LGM margin by \citet{dyke-2004} using climate forcing from the high-resolution interpolated observational data WorldClim \citep{data:worldclim} and the North American Regional Reanalysis (NARR) \citep{data:narr}.
+The latter dataset is preferable in our case due to the lack of WorldClim data offshore.
+
+One must keep in mind, however, that these results are limited to our choices of ice-sheet model (PISM), surface mass balance (PDD) model, study area (the Cordilleran Ice Sheet), and more importantly palaeo-climate representation (temperature offsets).
+At LGM, the presence of ice sheets most probably significantly affected circulation patterns and distribution of rainfall.
+At the cost of greater computational expense, a more accurate representation of palaeo-climate may allow for a better fit between model results and the mapped LGM margin.

### 2013-09-02 sections/abstract.tex
Rework abstract

@@
 \begin{abstract}
-\julien[noline]{The abstract is preliminary.}
-We present an ensemble of numerical simulations of the former Cordilleran ice sheet forced by present-day climatologies from five different datasets.
-Surface air temperature and precipitation rates from the NCEP/NCAR reanalysis, the Climate Forecast System Reanalysis, the ERA-Interim reanalysis and the North American Regional Reanalysis are used to force the ice sheet model PISM.
-Glacial conditions are mimicked by applying constant temperature offsets over the entire modelling domain, and a positive degree-day model is used to compute surface mass balance.
-The modelled ice sheet appears highly sensitive to the choice of climate forcing.
-Best results were obtained from the North American Regional Reanalysis, a high-resolution regional reanalysis which assimilates precipitation.
+We present an ensemble of numerical simulations of the Cordilleran Ice Sheet during the Last Glacial Maximum performed with the Parallel Ice Sheet Model, using temperature offsets from present-day climatologies from five different datasets.
+Monthly mean surface air temperature and monthly precipitation from the NCEP/NCAR reanalysis, the Climate Forecast System Reanalysis, the ERA-Interim reanalysis and the North American Regional Reanalysis are used to compute surface mass balance using a positive degree-day model.
+Modelled ice sheet outlines and volumes appear highly sensitive to the choice of climate forcing.
+Best results were obtained using the North American Regional Reanalysis, a high-resolution regional climate reanalysis.
 \end{abstract}

### 2013-09-02 floats/figures.tex
Get rid of personal shortcuts

@@
 	\end{center}
-	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a climate 5\degC colder than present for each climate forcing.
+	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a climate 5\,\unit{\degree C} colder than present for each climate forcing.
 Drawn using Matplotlib \citep{soft:mpl}.}

### 2013-09-02 sections/model.tex
Get rid of personal shortcuts

@@
 Ice surface accumulation and ablation are computed from monthly mean surface air temperature and monthly precipitation by a temperature-index (positive degree-day) model \citep{hock-2003}.
-Ice accumulation is equal to precipitation when temperature is below 0\,\degC, and decreases to zero linearly with temperature between 0 and 2\,\degC.
-Ice ablation is computed from the number of positive degree-days, defined as the integral of temperatures above 0\,\degC in one year.
+Ice accumulation is equal to precipitation when temperature is below 0\,\unit{\degree C}, and decreases to zero linearly with temperature between 0 and 2\,\unit{\degree C}.
+Ice ablation is computed from the number of positive degree-days, defined as the integral of temperatures above 0\,\unit{\degree C} in one year.
 Positive degree-days are calculated using a expression by \citet{calov-greve-2005} that accounts for temperature variability assuming a normal distribution along a central (input) value $T$:
@@
 where the integration interval $[t_0; t_1]$ is one week and $\mathrm{erfc}$ is the complementary error function.
-The temperature standard deviation $\sigma$ is a constant model parameter and was assigned a value of 3.068094\degC, which corresponds to the summer (JJA) mean, model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from North American Regional Reanalysis data \citep{data:narr} in a manner similar to \citet{seguinot-inreview}.
+The temperature standard deviation $\sigma$ is a constant model parameter and was assigned a value of 3.068094\,\unit{\degree C}, which corresponds to the summer (JJA) mean, model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from North American Regional Reanalysis data \citep{data:narr} in a manner similar to \citet{seguinot-inreview}.
 The ablation model incorporates degree-day factors of 3.04~\unit{mm\,K^{-1}\,day^{-1}} for snow and 4.59~\unit{mm\,K^{-1}\,day^{-1}} for ice, as derived from mass-balance measurements on contemporary glaciers from the Coast Mountains and Rocky Mountains in British Columbia \citep{shea-etal-2009}.

### 2013-09-02 sections/results.tex
Get rid of personal shortcuts

@@
 
-Using WorldClim, NCEP/NCAR, ERA-Interim, CFSR (smoothed and original datasets) and NARR input climatologies as a climate forcing (section~\ref{sec:climate}) for the ice sheet model PISM (section~\ref{sec:model}), we run 66~simulations of glacial inception and growth of the Cordilleran ice sheet over 10~kyr using temperature offsets ranging from 0 to 10\degC.
+Using WorldClim, NCEP/NCAR, ERA-Interim, CFSR (smoothed and original datasets) and NARR input climatologies as a climate forcing (section~\ref{sec:climate}) for the ice sheet model PISM (section~\ref{sec:model}), we run 66~simulations of glacial inception and growth of the Cordilleran ice sheet over 10~kyr using temperature offsets ranging from 0 to 10\,\unit{\degree C}.
 Important differences in patterns of ice sheet growth, final glacial extent, and volume were attained (Fig.
@@
 
-To contrast the six different input climatologies, we plot simulations run with the a 5\degC temperature offset (Fig.~\ref{fig:cool05}).
+To contrast the six different input climatologies, we plot simulations run with the a 5\,\unit{\degree C} temperature offset (Fig.~\ref{fig:cool05}).
 The magnitude of glaciation differs dramatically from one climate forcing to another.
@@
 Across the range of temperature offsets used, final glaciated area and final ice volume differs widely between different input climatologies (Fig.~\ref{fig:ivolarea}).
-Using a 5\degC temperature offset (Fig.~\ref{fig:cool05}) final glaciated area differs between WorldClim and CFSR forcing by a factor of 6 and final volume by a factor of 12 (Fig.~\ref{fig:ivolarea}).
+Using a 5\,\unit{\degree C} temperature offset (Fig.~\ref{fig:cool05}) final glaciated area differs between WorldClim and CFSR forcing by a factor of 6 and final volume by a factor of 12 (Fig.~\ref{fig:ivolarea}).
 For particular temperature offset values, given input climatologies result in similar final glaciated areas and ice volume, yet regional patterns are different (Fig.~\ref{fig:extent}).
 
-Under present climate (0\degC offset), WorldClim, NARR and ERA-Interim forcing produce local ice caps in areas presently glaciated (Fig.~\ref{fig:extent}).
+Under present climate (0\,\unit{\degree C} offset), WorldClim, NARR and ERA-Interim forcing produce local ice caps in areas presently glaciated (Fig.~\ref{fig:extent}).
 This results are fairly consistent with contemporary ice coverage, although both the 10~km horizontal resolution, and the occurrence of glaciers and ice caps in the ETOPO1 input basal topography prevent a detailed comparison.

### 2013-09-02 floats/figures.tex
Merge latex2rtf and pdflatex main files
Add a copernicus_rtf input file and replace to-do notes by regular footnotes. Using a copernicus_rtf package or class file did not seem to work with latex2rtf.

@@
 % ----------------------------------------------------------------------
+
+\emph{See to-do notes on Figure \ref{fig:locmap}
+    \irina[inline]{include all geographic names}
+    \arjen[inline]{elevation scale white-end confused with glaciers}
+    \arjen[inline]{remove Arctic circle and miles scale}
+    \arjen[inline]{use CIS, LIS and IIS for ice sheet names}
+    \arjen[inline]{use N, S, E and W instead of minus, add 75\degree N label}
+and \ref{fig:ivolarea}
+    \julien[inline]{Make the figure print-proof and highlight markers corresponding to the ``best'' runs.}
+}
 
@@
 	\end{center}
-    \irina[inline]{include all geographic names}
-    \arjen[inline]{elevation scale white-end confused with glaciers}
-    \arjen[inline]{remove Arctic circle and miles scale}
-    \arjen[inline]{use CIS, LIS and IIS for ice sheet names}
-    \arjen[inline]{use N, S, E and W instead of minus, add 75\degree N label}
 	\caption{Shaded relief map of northern North America with the extent of the Cordilleran (CIS), Laurentide (LIS) and Innuitian (IIS) ice sheets at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) \citep{dyke-2004}.
@@
 	\end{center}
-	\julien[inline]{I will make the figure readable in black-and-white and highlight markers corresponding to the ``best'' runs.}
 	\caption{Total ice volume and glaciated area after 10\,kyr as a function of temperature offset and climate forcing. Drawn using Matplotlib \citep{soft:mpl}.}

### 2013-09-03 floats/figures.tex
General reread and polish

@@
 \emph{See to-do notes on Figure \ref{fig:locmap}
-    \irina[inline]{include all geographic names}
-    \arjen[inline]{elevation scale white-end confused with glaciers}
-    \arjen[inline]{remove Arctic circle and miles scale}
-    \arjen[inline]{use CIS, LIS and IIS for ice sheet names}
-    \arjen[inline]{use N, S, E and W instead of minus, add 75\degree N label}
+    \irina{include all geographic names}
+    \arjen{elevation scale white-end confused with glaciers}
+    \arjen{remove Arctic circle and miles scale}
+    \arjen{use CIS, LIS and IIS for ice sheet names}
+    \arjen{use N, S, E and W instead of minus, add 75\degree N label}
 and \ref{fig:ivolarea}
-    \julien[inline]{Make the figure print-proof and highlight markers corresponding to the ``best'' runs.}
+    \julien{Make the figure print-proof and highlight markers corresponding to the ``best'' runs.}
 }

### 2013-09-03 sections/abstract.tex
General reread and polish

@@
 \begin{abstract}
-We present an ensemble of numerical simulations of the Cordilleran Ice Sheet during the Last Glacial Maximum performed with the Parallel Ice Sheet Model, using temperature offsets from present-day climatologies from five different datasets.
+We present an ensemble of numerical simulations of the Cordilleran Ice Sheet during the Last Glacial Maximum performed with the Parallel Ice Sheet Model (PISM), using temperature offsets from present-day climatologies from five different datasets.
 Monthly mean surface air temperature and monthly precipitation from the NCEP/NCAR reanalysis, the Climate Forecast System Reanalysis, the ERA-Interim reanalysis and the North American Regional Reanalysis are used to compute surface mass balance using a positive degree-day model.
 Modelled ice sheet outlines and volumes appear highly sensitive to the choice of climate forcing.
-Best results were obtained using the North American Regional Reanalysis, a high-resolution regional climate reanalysis.
+We assess model performance against a reconstruction of the ice margin at the Last Glacial Maximum in continental regions of northern Yukon Territory and interior Alaska.
+The closest match between model output and the reconstructed ice margin was obtained using climate forcing from the North American Regional Reanalysis, a high-resolution regional climate reanalysis.
 \end{abstract}

### 2013-09-03 sections/climate.tex
General reread and polish

@@
 
+An alternative forcing was prepared for CFSR to correct for artefacts seen in its precipitation field (Fig.~\ref{fig:prec}).
+A smoothing of the precipitation field was achieved by averaging data locally in a circular kernel of 7\,pixels in diameter prior to re-projection using the module r.neighbors from GRASS~GIS \citep{soft:grass}.
+Finally, input files for PISM were prepared using \citet{web:nc4py}.
 \julien{Qiong, could you help me with one or two sentences on why we smooth CFSR data?}
-An alternative forcing was prepared for CFSR to correct for artefacts seen in its precipitation field (Fig.~\ref{fig:prec}).
-A smoothing of the precipitation field was achieved by averaging data locally in a circular kernel of 7\,pixels in diameter prior to re-projection using the module r.neigbors from GRASS~GIS \citep{soft:grass}.
-Finally, input files for PISM were prepared using \citet{web:nc4py}.

### 2013-09-03 sections/concl.tex
General reread and polish

@@
 The latter dataset is preferable in our case due to the lack of WorldClim data offshore.
+NCEP/NCAR forcing, on the other hand, produces an anomalously large ice sheet even under present-day conditions.
 
 One must keep in mind, however, that these results are limited to our choices of ice-sheet model (PISM), surface mass balance (PDD) model, study area (the Cordilleran Ice Sheet), and more importantly palaeo-climate representation (temperature offsets).
-At LGM, the presence of ice sheets most probably significantly affected circulation patterns and distribution of rainfall.
+At LGM, it is most likely that the presence of ice sheets significantly affected circulation patterns and distribution of rainfall.
 At the cost of greater computational expense, a more accurate representation of palaeo-climate may allow for a better fit between model results and the mapped LGM margin.

### 2013-09-03 sections/intro.tex
General reread and polish

@@
 This exerts indirect influence on subsequently modelled ice sheet geometries, whose spatial extent tends to conform to these reconstructions.
-\julien[noline]{Should I include references not covering the Cordilleran Ice Sheet?}
 
@@
 In a numerical modelling perspective, it is one of the least studied palaeo-ice sheets of the Northern Hemisphere, despite the fact that significant geomorphological data is available to constrain its extent \citep{jackson-clague-1991,dukrodkin-1999,kaufman-manley-2004,kleman-etal-2010,margold-etal-2011}.
-Although the Cordilleran Ice Sheet was not the primary target of these simulations, it has previously been modelled as part of efforts to simulate ice sheets in Northern America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} or world-wide \citep{yoshimori-etal-2001}.
-While these studies reproduce the magnitude of North American glaciation at LGM reasonably well, there exists a tendency in the simulations that are independent of ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and interior Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.
+Although the Cordilleran Ice Sheet was not the primary target of these simulations, it has previously been modelled as part of efforts to simulate ice sheets in Northern America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} and world-wide \citep{yoshimori-etal-2001}.
+While these studies generally reproduce the magnitude of North American glaciation at LGM reasonably well, there exists a tendency in the simulations that are independent of ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and interior Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.
 

### 2013-09-03 sections/model.tex
General reread and polish

@@
 SSA velocities are determined by a pseudo-plastic sliding law where the yield stress depends on the availability of basal water and a prescribed till friction angle $\phi$.
-We apply a range of till friction angles between 10 and 30\,\degree, where the lower friction angles occur at low elevation to mimic the more deformable beds associated with presence of marine sediments:
+We apply a range of till friction angles between 10 and 30\degree, where the lower friction angles occur at low elevation to mimic the more deformable beds associated with presence of marine sediments:
 \needref
@@
 Where $\phi$ is given in degrees and $z$ in meters above contemporary sea level.
-Basal topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography and bathymetry dataset \citep{data:etopo1}.
-\julien{NaturalEarth data are vectors (coastline, rivers, etc.) used in Fig.
+Basal topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography and bathymetry dataset \citep{data:etopo1}.\julien{NaturalEarth data are vectors (coastline, rivers, etc.) used in Fig.
 1, not model input.} It responds to ice load following a regional isostasy model that includes lithospheric flexure and mantle relaxation \citep{lingle-clark-1985}.
@@
 The temperature standard deviation $\sigma$ is a constant model parameter and was assigned a value of 3.068094\,\unit{\degree C}, which corresponds to the summer (JJA) mean, model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from North American Regional Reanalysis data \citep{data:narr} in a manner similar to \citet{seguinot-inreview}.
-The ablation model incorporates degree-day factors of 3.04~\unit{mm\,K^{-1}\,day^{-1}} for snow and 4.59~\unit{mm\,K^{-1}\,day^{-1}} for ice, as derived from mass-balance measurements on contemporary glaciers from the Coast Mountains and Rocky Mountains in British Columbia \citep{shea-etal-2009}.
+The ablation model incorporates degree-day factors of 3.04\,\unit{mm\,K^{-1}\,day^{-1}} for snow and 4.59\,\unit{mm\,K^{-1}\,day^{-1}} for ice, as derived from mass-balance measurements on contemporary glaciers from the Coast Mountains and Rocky Mountains in British Columbia \citep{shea-etal-2009}.
 

### 2013-09-03 sections/results.tex
General reread and polish

@@
 Across the range of temperature offsets used, final glaciated area and final ice volume differs widely between different input climatologies (Fig.~\ref{fig:ivolarea}).
-Using a 5\,\unit{\degree C} temperature offset (Fig.~\ref{fig:cool05}) final glaciated area differs between WorldClim and CFSR forcing by a factor of 6 and final volume by a factor of 12 (Fig.~\ref{fig:ivolarea}).
+Using a 5\,\unit{\degree C} temperature offset (Fig.~\ref{fig:cool05}), final glaciated area differs between WorldClim and CFSR forcing by a factor of~6 and final volume by a factor of~12 (Fig.~\ref{fig:ivolarea}).
 For particular temperature offset values, given input climatologies result in similar final glaciated areas and ice volume, yet regional patterns are different (Fig.~\ref{fig:extent}).

### 2013-09-03 figures/plot.py
Customize markers on volume and area plot

@@ -310,10 +310,11 @@ def ivolarea():
 
     # select input data
     climates = ['wcnn', 'erai', 'narr', 'cfsrs7', 'cfsr', 'ncar']
+    styles   = ['s-', 'D-', 'o-', '^-', 'v-', 'h-']
     coolings = range(11)
 
     # read input data
-    for clim in climates:
+    for clim, style in zip(climates, styles):
       iarea = []
       ivol  = []
       for cool in coolings:
@@ -323,8 +324,8 @@ def ivolarea():
         ivol.append(nc.variables['ivol'][-1]/1e15)
 
       # plot
-      ax1.plot(coolings, iarea, '.-')
-      ax2.plot(coolings, ivol, '.-')
+      ax1.plot(coolings, iarea, style, mew=0, ms=3)
+      ax2.plot(coolings, ivol, style, mew=0, ms=3)
 
     # add legend
     ax2.legend(

### 2013-09-03 sections/intro.tex
Various changes on location map

@@
 
-The Cordilleran Ice Sheet (Fig.~\ref{fig:locmap}) covered an area that presently experiences strong regional variations in climate.
+The Cordilleran Ice Sheet (Fig.~\ref{fig:locmap})\irina{include all geographic names mentioned in the text}\arjen{elevation scale white-end can be confused with glaciers} covered an area that presently experiences strong regional variations in climate.
 In a numerical modelling perspective, it is one of the least studied palaeo-ice sheets of the Northern Hemisphere, despite the fact that significant geomorphological data is available to constrain its extent \citep{jackson-clague-1991,dukrodkin-1999,kaufman-manley-2004,kleman-etal-2010,margold-etal-2011}.

### 2013-09-13 sections/climate.tex
Implement corrections from Constantine

@@
 WorldClim, NCEP/NCAR, ERA-Interim, CFSR and NARR climatologies were re-projected to Canadian Atlas Lambert conformal conic projection (EPSG code~3978) and bilinearly interpolated to the 10\,km-resolution model grid using the module r.proj from GRASS~GIS \citep{soft:grass}.
-In addition, WorldClim data was extrapolated in a nearest-neighbour fashion to cover oceans grid points using the interpolate.griddata method from Scipy \citep{soft:scipy}.
-Note that we choose to present here the original rather than interpolated data in order to highlight differences related to spatial resolution among datasets (Fig.~\ref{fig:temp} and~\ref{fig:prec}).
+In addition, WorldClim data was extrapolated to cover grid points in the ocean using the nearest-neighbour algorithm \citep{soft:scipy}.
+Note that here we choose to present original rather than interpolated data, in order to highlight differences between datasets related to spatial resolution (Fig.~\ref{fig:temp} and~\ref{fig:prec}).
 

### 2013-09-13 sections/discuss.tex
General reread and polish
Implement corrections from Constantine

@@
 
-\subsection{Differences with reconstructed LGM margin}
+\subsection{Comparison to reconstructed LGM margin}
 
@@
 Furthermore, some simplifications were made in the surface mass balance model.
-Our PDD model do not include refreezing.
-Refreezing of melted snow and ice and retention in the snow pack, however, can largely alter mass balance at the surface of an ice sheet \citep{janssens-huybrechts-2000}.
+Our PDD model does not include refreezing.
+Refreezing of melted snow and ice and retention in the snow pack, however, can greatly alter mass balance at the surface of an ice sheet \citep{janssens-huybrechts-2000}.
 Additionally, we model temperature variability using a constant, homogeneous value of temperature standard deviation.

### 2013-09-13 sections/model.tex
Implement corrections from Constantine

@@
 
-The depth-averaged ice velocity $\vec{v}$ is computed as a weighted sum of the Shallow Ice Approximation (SIA) velocity $\vec{v_{\mathrm{SIA}}}$ and the Shallow Shelf Approximation (SSA) velocity $\vec{v_{\mathrm{SSA}}}$ \citep{bueler-brown-2009} by
+The Shallow Shelf Approximation (SSA) is used as a ``sliding law'' for the Shallow Ice Approximation (SIA) \citep{winkelmann-etal-2011}.
+SIA and SSA velocities are computed by finite difference methods on a 10km-resolution horizontal grid of 300 by 150 points (the modelling domain).
+Ice softness depends on temperature and water content through an enthalpy formulation \citep{aschwanden-blatter-2009,aschwanden-etal-2012}.
+Enthalpy is computed in three dimensions in up to 51 irregularly spaced layers in ice and temperature is further computed in 11 regularly spaced layers in bedrock.
+A uniform geothermal heat flux of 70\,\unit{W\,m^{-2}} provides the lower boundary condition to the bedrock thermal model, and surface air temperature from the climate forcing provides the upper boundary condition to the ice enthalpy model.
 
-\begin{equation}
-	\vec{v} = f(|\vec{v_{\mathrm{SSA}}}|)\vec{v_{\mathrm{SIA}}}
-  + (1-f(|\vec{v_{\mathrm{SSA}}}|))\vec{v_{\mathrm{SSA}}},
-\end{equation}
-
-where
-
-\begin{equation}
-	f(|\vec{v_{\mathrm{SSA}}}|) = 1
-	- \frac{2}{\pi}\arctan{\frac{|\vec{v_{\mathrm{SSA}}}|^2}{100^2}}.
-\end{equation}
-
-SIA and SSA velocities are computed by finite difference methods on a 10km-resolution horizontal grid of 300 by 150 points (the modelling domain).
-Ice rheology depends on temperature and water content through an enthalpy formulation \citep{aschwanden-blatter-2009,aschwanden-etal-2012}.
-Enthalpy is computed in three dimensions in up to 51 irregularly spaced layers in ice and temperature is further computed in 11 regularly spaced layers in bedrock.
-A homogeneous geothermal heat flux of 70\,\unit{W\,m^{-2}} provides the lower boundary condition to the bedrock thermal model, and surface air temperature from the climate forcing provides the upper boundary condition to the ice enthalpy model.
-
-SSA velocities are determined by a pseudo-plastic sliding law where the yield stress depends on the availability of basal water and a prescribed till friction angle $\phi$.
-We apply a range of till friction angles between 10 and 30\degree, where the lower friction angles occur at low elevation to mimic the more deformable beds associated with presence of marine sediments:
+A pseudo-plastic sliding law \citep{aschwanden-etal-2013} (supplement) relates the bed-parallel shear stress and the sliding velocity.
+The yield stress is modelled using the Mohr-Coulomb criterion.
+The till friction angle $\phi$ varies from 10 to 30\degree\ and is a function of contemporary bed elevation, with lowest values occurring at low elevations.
 \needref
@@
 Where $\phi$ is given in degrees and $z$ in meters above contemporary sea level.
+This models weakening of the till associated with the presence of marine sediments.
 Basal topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography and bathymetry dataset \citep{data:etopo1}.\julien{NaturalEarth data are vectors (coastline, rivers, etc.) used in Fig.
@@
 Ice ablation is computed from the number of positive degree-days, defined as the integral of temperatures above 0\,\unit{\degree C} in one year.
-Positive degree-days are calculated using a expression by \citet{calov-greve-2005} that accounts for temperature variability assuming a normal distribution along a central (input) value $T$:
 
-\begin{equation}
-	\mathrm{PDD} = \int_{t_0}^{t_1} dt \left[
-		\frac{\sigma}{\sqrt{2\pi}}
-		\exp\left({-\frac{T(t)^2}{2\sigma^2}}\right)
-		+\frac{T(t)}{2}
-		\mathrm{erfc} \left(-\frac{T(t)}{\sqrt{2}\sigma}\right)
-	\right],
-\end{equation}
-
-where the integration interval $[t_0; t_1]$ is one week and $\mathrm{erfc}$ is the complementary error function.
-The temperature standard deviation $\sigma$ is a constant model parameter and was assigned a value of 3.068094\,\unit{\degree C}, which corresponds to the summer (JJA) mean, model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from North American Regional Reanalysis data \citep{data:narr} in a manner similar to \citet{seguinot-inreview}.
+The positive degree-day integral \citep{calov-greve-2005} is numerically approximated using week-long sub-intervals.
+It accounts for temperature variability assuming a normal distribution along a central (input) value.
+The temperature standard deviation is a constant model parameter and was assigned a value of 3.068094\,\unit{\degree C}, which corresponds to the summer (JJA) mean, model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from North American Regional Reanalysis data \citep{data:narr} in a manner similar to \citet{seguinot-inreview}.
 The ablation model incorporates degree-day factors of 3.04\,\unit{mm\,K^{-1}\,day^{-1}} for snow and 4.59\,\unit{mm\,K^{-1}\,day^{-1}} for ice, as derived from mass-balance measurements on contemporary glaciers from the Coast Mountains and Rocky Mountains in British Columbia \citep{shea-etal-2009}.
@@
 This correction accounts for evolution of ice thickness on the one hand, and differences between the climate forcing reference topography and the ice flow model basal topography on the over hand.
-It uses a reference topography distinct from the basal topography and specific to each climate forcing dataset (Fig.~\ref{fig:topo}).
+It uses a reference topography distinct from the modelled basal topography and specific to each climate forcing dataset (Fig.~\ref{fig:topo}).
 An annual lapse-rate of 6\,\unit{\degree C\,km^{-1}} is used in all simulations.

### 2013-09-13 sections/model.tex
Add references for topg_to_phi (from Constantine)

@@
 The yield stress is modelled using the Mohr-Coulomb criterion.
-The till friction angle $\phi$ varies from 10 to 30\degree\ and is a function of contemporary bed elevation, with lowest values occurring at low elevations.
-\needref
+The till friction angle $\phi$ varies from 10 to 30\degree\ and is a function of contemporary bed elevation, with lowest values occurring at low elevations,
 \julien{phi is independent from basal water which is computed by the model (and sliding depends on both) so the 10 to 30 degree range is meant to reflect different bedrock/sediment surfaces rather than dry vs wet beds.}
@@
 
-Where $\phi$ is given in degrees and $z$ in meters above contemporary sea level.
-This models weakening of the till associated with the presence of marine sediments.
+where $\phi$ is given in degrees and $z$ in meters above contemporary sea level.
+This models weakening of the till associated with the presence of marine sediments \citep{martin-etal-2011,aschwanden-etal-2013}.
 Basal topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography and bathymetry dataset \citep{data:etopo1}.\julien{NaturalEarth data are vectors (coastline, rivers, etc.) used in Fig.

### 2013-09-14 floats/figures.tex
Implement corrections from Qiong

@@
 	\end{center}
-	\caption{Topography maps from CFSR (Climate System Forecast Reanalysis), ERA-Interim reanalysis, NARR (North American Regional Reanalysis), and NCEP/NCAR reanalysis data used as a reference for temperature lapse-rate corrections, and topography map from ETOPO1 used as basal topography.
-Drawn using Matplotlib \citep{soft:mpl}.}
+	\caption{Topography maps from WorldClim, ERA-Interim reanalysis, NARR (North American Regional Reanalysis), CFSR (Climate System Forecast Reanalysis), and NCEP/NCAR reanalysis climatologies used as a reference for temperature lapse-rate corrections, and topography map from ETOPO1 used as basal topography.
+Fig.~\ref{fig:topo}-\ref{fig:best} drawn using Matplotlib \citep{soft:mpl}.}
 	\label{fig:topo}
@@
 	\end{center}
-	\caption{Summer (JJA) and winter (DJF) temperature maps from CFSR (Climate System Forecast Reanalysis), ERA-Interim reanalysis, NARR (North American Regional Reanalysis), and NCEP/NCAR reanalysis climatologies.
-Drawn using Matplotlib \citep{soft:mpl}.}
+	\caption{Summer (JJA) and winter (DJF) temperature maps from WorldClim, and summer (JJA) temperature maps from ERA-Interim reanalysis, NARR (North American Regional Reanalysis), CFSR (Climate System Forecast Reanalysis), and NCEP/NCAR reanalysis climatologies.}
 	\label{fig:temp}
@@
 	\end{center}
-	\caption{Summer (JJA) and winter (DJF) precipitation maps from CFSR (Climate System Forecast Reanalysis), ERA-Interim reanalysis, NARR (North American Regional Reanalysis), and NCEP/NCAR reanalysis climatologies.
-Drawn using Matplotlib \citep{soft:mpl}.}
+	\caption{Winter (DJF) and summer (JJA) precipitation maps from WorldClim, and winter (DFJ) temperature maps from ERA-Interim reanalysis, NARR (North American Regional Reanalysis), CFSR (Climate System Forecast Reanalysis), and NCEP/NCAR reanalysis climatologies.}
 	\label{fig:prec}
@@
 	\end{center}
-	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a climate 5\,\unit{\degree C} colder than present for each climate forcing.
-Drawn using Matplotlib \citep{soft:mpl}.}
+	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a climate 5\,\unit{\degree C} colder than present for each climate forcing.}
 	\label{fig:cool05}
@@
 	\end{center}
-	\caption{Extent of ice cover after 10\,kyr as a function of applied temperature offsets for each climate forcing. Drawn using Matplotlib \citep{soft:mpl}.}
+	\caption{Extent of ice cover after 10\,kyr as a function of applied temperature offsets for each climate forcing.}
 	\label{fig:extent}
@@
 	\end{center}
-	\caption{Total ice volume and glaciated area after 10\,kyr as a function of temperature offset and climate forcing. Drawn using Matplotlib \citep{soft:mpl}.}
+	\caption{Total glaciated area and ice volume after 10\,kyr as a function of temperature offset and climate forcing.}
 	\label{fig:ivolarea}
@@
 	\caption{Ice surface topography (black contours every 1000\,m) after 10\,kyr using temperature offsets that lead to similar areas of ice cover for each climate forcing.
-14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) ice margin (blue line) from \citet{dyke-2004}.
-Drawn using Matplotlib \citep{soft:mpl}.}
+14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) ice margin (blue line) from \citet{dyke-2004}.}
 	\label{fig:best}
@@
 	\caption{Schematic representation of the orographic precipitation effect over the Coast Mountains.
-In a GCM of lower resolution, the precipitation peak appear smoother and the precipitation shadow is not as pronounced.
+In a GCM of lower resolution, the precipitation peak appears smoother and the precipitation shadow is not as pronounced.
 Drawn using Inkscape \citep{web:inkscape}.}

### 2013-09-14 sections/abstract.tex
Implement corrections from Qiong

@@
 We present an ensemble of numerical simulations of the Cordilleran Ice Sheet during the Last Glacial Maximum performed with the Parallel Ice Sheet Model (PISM), using temperature offsets from present-day climatologies from five different datasets.
-Monthly mean surface air temperature and monthly precipitation from the NCEP/NCAR reanalysis, the Climate Forecast System Reanalysis, the ERA-Interim reanalysis and the North American Regional Reanalysis are used to compute surface mass balance using a positive degree-day model.
+Monthly mean surface air temperature and precipitation from the NCEP/NCAR reanalysis, the Climate Forecast System Reanalysis, the ERA-Interim reanalysis and the North American Regional Reanalysis are used to compute surface mass balance by using a positive degree-day model.
 Modelled ice sheet outlines and volumes appear highly sensitive to the choice of climate forcing.
 We assess model performance against a reconstruction of the ice margin at the Last Glacial Maximum in continental regions of northern Yukon Territory and interior Alaska.
-The closest match between model output and the reconstructed ice margin was obtained using climate forcing from the North American Regional Reanalysis, a high-resolution regional climate reanalysis.
+The best match between model output and the reconstructed ice margin is obtained using climate forcing from the North American Regional Reanalysis, which is a regional climate reanalysis with the highest resolution.
 \end{abstract}

### 2013-09-14 sections/climate.tex
Implement corrections from Qiong

@@
 It should be noted that temperature gradients are reversed (colder inland than at the coast) and much stronger in winter than in summer.
-Temperatures rises well above zero during the summer months over the entire modelling domain, except for the highest mountain peaks, and even regions where mean annual temperatures are well below freezing do experience warm summers.
+Temperatures rises well above zero during the summer months over the entire modelling domain, except for the highest mountain peaks, and even regions where mean annual temperatures are well below freezing point do experience warm summers.
 In other words, there is a strong contrast in seasonality between coastal temperate-climate and inland continental-climate regions.
 
-The spatial distribution of JJA and DJF precipitation rates in WorldClim reveal a strong precipitation decline beyond the coastal regions (Fig.~\ref{fig:prec}), primarily as a result of the continuous orographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascades.
+The spatial distribution of JJA and DJF mean precipitation rates in WorldClim reveal a strong precipitation decline beyond the coastal regions (Fig.~\ref{fig:prec})\qiong{Change unit to mm per month.}, primarily as a result of the continuous orographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascades.
 From an ice sheet mass balance point of view, this contrast becomes even stronger by the difference in timing of the precipitation peak through the year.
@@
 
-In addition to the WorldClim data, we use surface air temperatures and precipitation rates from three global climate reanalyses and one regional climate reanalysis to inform the mass balance routine of PISM: the NCEP/NCAR reanalysis \citep{data:ncar}, the ERA-Interim reanalysis \citep{data:erai}, the Climate System Forecast Reanalysis (CFSR) \citep{data:cfsr}, and the North American Regional Reanalysis (NARR) \citep{data:narr}.
+In addition to the WorldClim data, we use surface air temperatures and precipitation rates from three global atmospheric reanalyses and one regional atmospheric reanalysis to inform the mass balance routine of PISM: the NCEP/NCAR reanalysis \citep{data:ncar}, the ERA-Interim reanalysis \citep{data:erai}, the Climate System Forecast Reanalysis (CFSR) \citep{data:cfsr}, and the North American Regional Reanalysis (NARR) \citep{data:narr}.
 Monthly climatologies from NCEP/NCAR and NARR reanalysis were provided by the \citet{web:psd} whereas monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean time series (Table \ref{tab:reanalyses}).
@@
 Summer temperature and winter precipitation are most relevant to the glacier model as they respectively drive summer melt and winter accumulation.
-Because reanalyses include observational information through data assimilation, they are more accurate than simple GCMs in densely monitored regions, while offering physically based output in sparsely monitored regions such as the northern part of our modelling domain \citep{bengtsson-etal-2007}.
+Because reanalyses include observational information through data assimilation, they are closer to observations than simple GCMs in densely monitored regions, while offering physically based output in sparsely monitored regions such as the northern part of our modelling domain \citep{bengtsson-etal-2007}.
 
@@
 
-An alternative forcing was prepared for CFSR to correct for artefacts seen in its precipitation field (Fig.~\ref{fig:prec}).
-A smoothing of the precipitation field was achieved by averaging data locally in a circular kernel of 7\,pixels in diameter prior to re-projection using the module r.neighbors from GRASS~GIS \citep{soft:grass}.
+The CFSR climatology presents wave-like artefacts in its precipitation field (Fig.~\ref{fig:prec}).
+These are a common feature in GCM simulations due to the failure of spectral models to resolve the very local nature of topography-induced precipitation events, which becomes particularly visible in the high-resolution, less spatially smoothed data.
+Therefore, an alternative forcing was prepared for CFSR by smoothing its precipitation field.
+This was achieved by averaging data locally in a circular kernel of 7\,pixels in diameter prior to re-projection using the module r.neighbors from GRASS~GIS \citep{soft:grass}.
 Finally, input files for PISM were prepared using \citet{web:nc4py}.
-\julien{Qiong, could you help me with one or two sentences on why we smooth CFSR data?}

### 2013-09-14 sections/discuss.tex
Implement corrections from Qiong

@@
 In our simulations, a continuous ice cover quickly forms over the western ranges, where precipitation rates are higher than in the rest of the domain.
-This continuous ice cover may have enhanced the topographical barrier already formed by the western ranges, resulting in fewer precipitation and warmer air in the interior.
+This continuous ice cover may have enhanced the topographical barrier already formed by the western ranges, resulting in less precipitation and warmer air in the interior.
 \citet{langen-etal-2012} demonstrated that this process, which they refer to as a ``self-inhibiting growth'', may have been limiting during the build-up of the Greenland Ice Sheet.
-Alternatively, the Laurentide Ice Sheet, not included in our model, may have formed a buttress against the smaller Cordilleran Ice Sheet and stopped its advance onto the Canadian Prairies.
+Alternatively, the Laurentide Ice Sheet, which is not included in our model, may have formed a buttress against the smaller Cordilleran Ice Sheet and stopped its advance onto the Canadian Prairies.
 
@@
 
-Difference between modelled ice sheet geometries is notably visible in northern Yukon Territory and Alaska, where previous modelling studies commonly produced too extensive ice cover, whereas geomorphological data shows that the ice cover was sparse.
-Concerning reanalysis datasets, we interpret these differences as mainly the result of different GCM resolutions (Fig.~\ref{fig:topo}, Table~\ref{tab:reanalyses}).
+Difference between modelled ice sheet geometries is notably visible in northern Yukon Territory and Alaska, where previous modelling studies commonly produced too extensive ice cover, whereas geomorphological data show that the ice cover was sparse.
+Concerning reanalysis datasets, we interpret these differences as mainly the result of different GCM resolutions and related model physics (Fig.~\ref{fig:topo}, Table~\ref{tab:reanalyses}).
 
@@
 Refreezing of melted snow and ice and retention in the snow pack, however, can greatly alter mass balance at the surface of an ice sheet \citep{janssens-huybrechts-2000}.
-Additionally, we model temperature variability using a constant, homogeneous value of temperature standard deviation.
+Additionally, we simulate temperature variability by using a constant, uniform value of temperature standard deviation.
 This approach implies large biases of surface mass-balance \citep{charbit-etal-2013,rau-rogozhina-2013,seguinot-inreview}, particularly over regions with such various climates as the Northern American Cordillera.

### 2013-09-14 sections/intro.tex
Implement corrections from Qiong

@@
 Numerical modelling of these former ice masses allows for a comparison between glaciological theories embedded in the models and geomorphological traces underpinning palaeo-glaciological reconstructions.
-Yet, a major obstacle in this exercise resides in large uncertainties concerning climate forcing, typically temperature and precipitation data, needed by numerical glacier models \citep{hebeler-etal-2008}.
+Yet, a major obstacle in this exercise resides in large uncertainties concerning climate forcing, typically temperature and precipitation data, that are needed by numerical glacier models \citep{hebeler-etal-2008}.
 This includes uncertainty in representation of Earth's present climate in regions of poor station coverage, and even larger uncertainty concerning accurate reconstructions of past climate change.
@@
 
-Climatologies obtained from uncoupled GCM palaeo-climate simulations such as produced within the PMIP project \citep{joussaume-taylor-1995} provide a more accurate representation of past climate and are commonly used to force ice sheet models.
+Climatologies obtained from GCM palaeo-climate simulations such as produced within the Paleoclimate Modelling Intercomparison Project (PMIP) \citep{joussaume-taylor-1995} provide a more reasonable representation of past climate and are commonly used to force ice sheet models.
 Because such climatologies are only available for specific periods of time, this method requires either an assumption of steady-state \citep{huybrechts-tsiobbel-1996}, or interpolation through time between climatologies from different periods, which can be linear \citep{charbit-etal-2002}, or modulated by a ``glacial index'' weighting function derived from ice core $\delta^{18}$O records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}.
@@
 Few studies of this kind are presently available.
-\citet{quiquet-etal-2012} assessed the sensitivity of a Greenland ice sheet model to various atmospheric forcing, including a regional parametrization \citep{fausto-etal-2009}, output from several GCMs and a climate reanalysis.
-\citet{rodgers-etal-2004} and \citet{charbit-etal-2007} tested the sensitivity of a model of the Northern Hemisphere ice sheets to climate forcing from different PMIP LGM simulations.
-Here, to limit degrees of freedom in our model, and obtain results independent of palaeo-ice sheet reconstructions such as the ICE-4G, we use a simple temperature offset approach similar to \citet{greve-etal-1999} and \citet{bintanja-etal-2005} and assess ice sheet model sensitivity to the choice of present-day climate data.
+\citet{quiquet-etal-2012} assessed the sensitivity of a Greenland ice sheet model to various atmospheric forcing, including a regional parametrization \citep{fausto-etal-2009}, output from several GCMs and an atmospheric reanalysis.
+\citet{rodgers-etal-2004} and \citet{charbit-etal-2007} tested the sensitivity of a model of the Northern Hemisphere ice sheets to climate forcing from different PMIP LGM simulations\qiong{Detail in a few lines the conclusions of theses studies}.
+Here, in order to limit degrees of freedom in our model, and obtain results independent of palaeo-ice sheet reconstructions such as the ICE-4G, we use a simple temperature offset approach similar to \citet{greve-etal-1999} and \citet{bintanja-etal-2005} and assess ice sheet model sensitivity to the choice of present-day climate data.
 Rather than using GCM output, we force our model with climate reanalysis data, which include observational information through data assimilation \citep{bengtsson-etal-2007}.

### 2013-09-14 sections/model.tex
Implement corrections from Qiong

@@
 Basal topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography and bathymetry dataset \citep{data:etopo1}.\julien{NaturalEarth data are vectors (coastline, rivers, etc.) used in Fig.
-1, not model input.} It responds to ice load following a regional isostasy model that includes lithospheric flexure and mantle relaxation \citep{lingle-clark-1985}.
+1, not model input.} Sea level is lowered by 120\,m and basal topography responds to ice load following a regional isostasy model that includes lithospheric flexure and mantle relaxation \citep{lingle-clark-1985}.
 
@@
 It accounts for temperature variability assuming a normal distribution along a central (input) value.
-The temperature standard deviation is a constant model parameter and was assigned a value of 3.068094\,\unit{\degree C}, which corresponds to the summer (JJA) mean, model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from North American Regional Reanalysis data \citep{data:narr} in a manner similar to \citet{seguinot-inreview}.
+The temperature standard deviation is a constant model parameter and was assigned a value of 3.07\,\unit{\degree C}, which corresponds to the summer (JJA) mean, model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from North American Regional Reanalysis data \citep{data:narr} in a manner similar to \citet{seguinot-inreview}.
 The ablation model incorporates degree-day factors of 3.04\,\unit{mm\,K^{-1}\,day^{-1}} for snow and 4.59\,\unit{mm\,K^{-1}\,day^{-1}} for ice, as derived from mass-balance measurements on contemporary glaciers from the Coast Mountains and Rocky Mountains in British Columbia \citep{shea-etal-2009}.
@@
 Prior to surface mass balance computation, the model dynamically applies a lapse-rate correction to surface air temperature.
-This correction accounts for evolution of ice thickness on the one hand, and differences between the climate forcing reference topography and the ice flow model basal topography on the over hand.
+This correction accounts for evolution of ice thickness on the one hand, and differences between the climate forcing reference topography and the ice flow model basal topography on the other hand.
 It uses a reference topography distinct from the modelled basal topography and specific to each climate forcing dataset (Fig.~\ref{fig:topo}).

### 2013-09-15 sections/results.tex
Implement corrections from Qiong
Change figure order in results (Qiong)

--- 2013-09-02 sections/results.tex
@@
 Important differences in patterns of ice sheet growth, final glacial extent, and volume were attained (Fig.
-\ref{fig:cool05}-\ref{fig:best}).
-
-To contrast the six different input climatologies, we plot simulations run with the a 5\,\unit{\degree C} temperature offset (Fig.~\ref{fig:cool05}).
-The magnitude of glaciation differs dramatically from one climate forcing to another.
-Whereas the NCEP/NCAR and CFSR climatologies produce a large ice sheet that covers most of the model domain (Fig.~\ref{fig:cool05}, lower panels), WorldClim, ERA-Interim and NARR climatologies lead to more restrictive ice cover (Fig.~\ref{fig:cool05}, upper panels).
-The WorldClim run only resulted in the growth of mountain ice caps bounded to the highest topography.
-The smoothing of CFSR precipitation yielded an insignificant effect on the resulting ice sheet geometry.
+\ref{fig:ivolarea}-\ref{fig:cool05}).
 
 Across the range of temperature offsets used, final glaciated area and final ice volume differs widely between different input climatologies (Fig.~\ref{fig:ivolarea}).
-Using a 5\,\unit{\degree C} temperature offset (Fig.~\ref{fig:cool05}) final glaciated area differs between WorldClim and CFSR forcing by a factor of 6 and final volume by a factor of 12 (Fig.~\ref{fig:ivolarea}).
+For instance, using a 5\,\unit{\degree C} temperature offset, final glaciated area differs between WorldClim and CFSR forcing by a factor of~6 and final volume by a factor of~12 \ref{fig:ivolarea}).
 For particular temperature offset values, given input climatologies result in similar final glaciated areas and ice volume, yet regional patterns are different (Fig.~\ref{fig:extent}).
@@
 At the opposite side of the temperature offset range, it can be noted that NCEP/NCAR and CFSR forcing produce oversized ice sheets whose extent is primarily restricted by the domain boundaries (Fig.~\ref{fig:extent}).
+The smoothing of CFSR precipitation yielded an insignificant effect on the resulting ice sheet geometry.
+
+To contrast the six different input climatologies, we plot simulations run with the a 5\,\unit{\degree C} temperature offset (Fig.~\ref{fig:cool05}).
+The magnitude of glaciation differs dramatically from one climate forcing to another.
+Whereas the NCEP/NCAR and CFSR climatologies produce a large ice sheet that covers most of the model domain (Fig.~\ref{fig:cool05}, lower panels), WorldClim, ERA-Interim and NARR climatologies lead to more restrictive ice cover (Fig.~\ref{fig:cool05}, upper panels).
+The WorldClim run only resulted in the growth of mountain ice caps bounded to the highest topography.
 

### 2013-09-19 sections/climate.tex
Implement corrections from Irina
Mostly as to-do notes for now

@@
 
-Our atmospheric forcing consists of monthly climatologies of surface air temperature and precipitation obtained from one observational dataset, three global reanalysis and one regional reanalysis.
+Our atmospheric forcing consists of monthly climatologies of surface air temperature and precipitation obtained from one observational dataset, three global reanalyses and one regional reanalysis.
 
@@
 
-Within our modelling domain, the spatial distributions of summer (JJA) and winter (DJF) air surface temperatures in WorldClim generally show a decrease from south to north (JJA) and from south-west to north-east (DJF), such that regions further inland experience colder winters (Fig.~\ref{fig:temp}).
+Within our modelling domain, the spatial distributions of summer (JJA) and winter (DJF) air surface temperatures in WorldClim generally show a decrease of temperature from south to north (JJA) and from south-west to north-east (DJF), such that regions further inland experience colder winters (Fig.~\ref{fig:temp}).
 It should be noted that temperature gradients are reversed (colder inland than at the coast) and much stronger in winter than in summer.
@@
 The spatial distribution of JJA and DJF mean precipitation rates in WorldClim reveal a strong precipitation decline beyond the coastal regions (Fig.~\ref{fig:prec})\qiong{Change unit to mm per month.}, primarily as a result of the continuous orographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascades.
-From an ice sheet mass balance point of view, this contrast becomes even stronger by the difference in timing of the precipitation peak through the year.
+From an ice sheet mass balance point of view, this contrast becomes even stronger due to the difference in timing of the precipitation peak through the year.
 Whereas coastal regions experience most precipitation as snow during the accumulation season (DJF), inland regions experience dry winters and most of the annual precipitation consequently falls as rain during the ablation season (JJA).
@@
 Although good coverage exists for the southern part of our modelling domain, several hundred kilometres can separate neighbouring stations in the north \citep{data:worldclim}.
-Secondly, WorldClim lacks of data offshore, which would have been particularly useful over the Pacific continental shelf which was partly covered by ice during the LGM \citep{jackson-clague-1991}.
+Secondly, WorldClim lacks data offshore, which would have been particularly useful over the Pacific continental shelf which was partly covered by ice during the LGM \citep{jackson-clague-1991}.
 
@@
 In addition to the WorldClim data, we use surface air temperatures and precipitation rates from three global atmospheric reanalyses and one regional atmospheric reanalysis to inform the mass balance routine of PISM: the NCEP/NCAR reanalysis \citep{data:ncar}, the ERA-Interim reanalysis \citep{data:erai}, the Climate System Forecast Reanalysis (CFSR) \citep{data:cfsr}, and the North American Regional Reanalysis (NARR) \citep{data:narr}.
-Monthly climatologies from NCEP/NCAR and NARR reanalysis were provided by the \citet{web:psd} whereas monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean time series (Table \ref{tab:reanalyses}).
+Monthly climatologies from NCEP/NCAR and NARR reanalyses were provided by the \citet{web:psd} whereas monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean time series (Table \ref{tab:reanalyses}).
 

### 2013-09-19 sections/concl.tex
Implement corrections from Qiong
Implement corrections from Irina
Mostly as to-do notes for now

--- 2013-07-01 sections/concl.tex
@@
 
-\emph{Coming soon...}
+Our study shows a strong dependency of the modelled ice sheet on the choice of climate forcing data.
+When using input climatologies from climate reanalyses, spatial resolution of the GCM appears critical for providing the ice sheet model with an accurate precipitation field.
+This confirms results obtained by \citet{quiquet-etal-2012} from various GCM forcing over the Greenland Ice Sheet.
+
+For the Cordilleran Ice Sheet, we achieve the best fit to the mapped LGM margin by \citet{dyke-2004} using climate forcing from the high-resolution interpolated observational data WorldClim \citep{data:worldclim} and the North American Regional Reanalysis (NARR) \citep{data:narr}.
+The latter dataset is preferable in our case due to the lack of WorldClim data offshore.
+NCEP/NCAR forcing, on the other hand, produces an anomalously large ice sheet even under present-day conditions.
+
+One must keep in mind, however, that these results are biased by our choices of ice-sheet model (PISM), surface mass balance (PDD) model, study area (the Cordilleran Ice Sheet), and more importantly palaeo-climate representation (temperature offsets).
+At LGM, it is most likely that the presence of ice sheets significantly affected circulation patterns and distribution of precipitation.
+At the cost of greater computational expense, a more accurate representation of palaeo-climate may allow for a better fit between model results and the mapped LGM margin.
+\irina{I have this feeling that the manuscript contains a number of repetitive ideas and arguments. I will try to track them throughout the text.}

### 2013-09-19 sections/discuss.tex
Implement corrections from Irina
Mostly as to-do notes for now

@@
 
-Our simulations produce anomalous ice cover on parts of the continental shelf in the Arctic Ocean.
+Our simulations produce anomalous\irina{Any evidence that there was no ice cover there?
+Refer to the source.} ice cover on parts of the continental shelf in the Arctic Ocean.
 These regions experience a marine climate, including lower summer temperature than on the adjacent land (Fig.~\ref{fig:temp}).
@@
 As described in section~\ref{sec:climate}, the topography of the Northern American Cordillera is such that its western ranges form a continuous orographic barrier, causing high precipitation along the Pacific coast while leaving much of the interior arid.
-However the ability of a GCM to reproduce these contrasts in precipitation is bound to its resolution.
+However the ability of a GCM to reproduce these contrasts in precipitation is bounded by resolution.
 In a model of coarser resolution such as used in the NCEP/NCAR reanalysis, these high mountain ranges are reduced to smoothed hills and the modelled distribution of precipitation is also smoother (Fig.~\ref{fig:oroprecip}).

### 2013-09-19 sections/model.tex
Implement corrections from Irina
Mostly as to-do notes for now

@@
 We use PISM, a Parallel Ice Sheet Model, version stable~0.5.12 (e.g.,) \citep{bueler-brown-2009,winkelmann-etal-2011,aschwanden-etal-2012,web:pism}.
-Given basal topography and climate forcing, the model computes the thermal and dynamic state of an ice sheet and the associated lithospheric response.
-Our modelling domain encompasses most of the area covered by the Cordilleran Ice Sheet at LGM except of the western Alaskan ranges and includes non-glaciated regions in northern Yukon Territory and interior Alaska.
+Given basal topography\irina{geothermal heat flux and sea level} and climate forcing, the model computes the thermal and dynamic state of an ice sheet and the associated lithospheric response.
+Our modelling domain encompasses most of the area covered by the Cordilleran Ice Sheet at LGM except from the western Alaskan ranges and includes non-glaciated regions in northern Yukon Territory and interior Alaska.
 Our simulations are performed in parallel on 32 cores at the Swedish National Supercomputing Center.
@@
 The Shallow Shelf Approximation (SSA) is used as a ``sliding law'' for the Shallow Ice Approximation (SIA) \citep{winkelmann-etal-2011}.
-SIA and SSA velocities are computed by finite difference methods on a 10km-resolution horizontal grid of 300 by 150 points (the modelling domain).
+SIA and SSA velocities are computed by finite difference methods on a 10\,km-resolution horizontal grid of 300 by 150 points (the modelling domain).
 Ice softness depends on temperature and water content through an enthalpy formulation \citep{aschwanden-blatter-2009,aschwanden-etal-2012}.
-Enthalpy is computed in three dimensions in up to 51 irregularly spaced layers in ice and temperature is further computed in 11 regularly spaced layers in bedrock.
+Enthalpy is computed in three dimensions in up to 51 irregularly spaced layers in ice, and temperature is further computed in 11 regularly spaced layers in bedrock.
 A uniform geothermal heat flux of 70\,\unit{W\,m^{-2}} provides the lower boundary condition to the bedrock thermal model, and surface air temperature from the climate forcing provides the upper boundary condition to the ice enthalpy model.
@@
 An annual lapse-rate of 6\,\unit{\degree C\,km^{-1}} is used in all simulations.
-No lapse-rate corrections applies to precipitation.
+No lapse-rate corrections apply to precipitation.
 

### 2013-09-19 sections/results.tex
Implement corrections from Irina
Mostly as to-do notes for now

@@
 Using WorldClim, NCEP/NCAR, ERA-Interim, CFSR (smoothed and original datasets) and NARR input climatologies as a climate forcing (section~\ref{sec:climate}) for the ice sheet model PISM (section~\ref{sec:model}), we run 66~simulations of glacial inception and growth of the Cordilleran ice sheet over 10~kyr using temperature offsets ranging from 0 to 10\,\unit{\degree C}.
-Important differences in patterns of ice sheet growth, final glacial extent, and volume were attained (Fig.
+Significant differences in patterns of ice sheet growth, final glacial extent, and volume were attained (Fig.
 \ref{fig:ivolarea}-\ref{fig:cool05}).
 
-Across the range of temperature offsets used, final glaciated area and final ice volume differs widely between different input climatologies (Fig.~\ref{fig:ivolarea}).
-For instance, using a 5\,\unit{\degree C} temperature offset, final glaciated area differs between WorldClim and CFSR forcing by a factor of~6 and final volume by a factor of~12 \ref{fig:ivolarea}).
-For particular temperature offset values, given input climatologies result in similar final glaciated areas and ice volume, yet regional patterns are different (Fig.~\ref{fig:extent}).
+Across the range of temperature offsets used, final glaciated area and final ice volume differ widely between different input climatologies (Fig.~\ref{fig:ivolarea}).
+For instance, using a 5\,\unit{\degree C} temperature offset, final glaciated area differs between WorldClim and CFSR forcing by a factor of~6 and final volume by a factor of~12 (Fig.~\ref{fig:ivolarea}).
+For particular temperature offset values, given input climatologies result in similar final glaciated areas and ice volumes, yet regional patterns are different (Fig.~\ref{fig:extent}).
 
@@
 CFSR forcing, on the other hand, yields continuous ice cover over the northern Coast Mountains, and NCEP/NCAR forcing results in ice-sheet growth, which is largely inconsistent with the present state of glaciation.
-At the opposite side of the temperature offset range, it can be noted that NCEP/NCAR and CFSR forcing produce oversized ice sheets whose extent is primarily restricted by the domain boundaries (Fig.~\ref{fig:extent}).
-The smoothing of CFSR precipitation yielded an insignificant effect on the resulting ice sheet geometry.
+At the opposite side of the temperature offset range, it can be noted that both the NCEP/NCAR and CFSR forcing produce oversized ice sheets whose extent are primarily restricted by the domain boundaries (Fig.~\ref{fig:extent}).
+The smoothing of CFSR precipitation yields an insignificant effect on the resulting ice sheet geometry.
 
-To contrast the six different input climatologies, we plot simulations run with the a 5\,\unit{\degree C} temperature offset (Fig.~\ref{fig:cool05}).
+To contrast the six different input climatologies, we plot simulations run with the 5\,\unit{\degree C} temperature offset (Fig.~\ref{fig:cool05}).
 The magnitude of glaciation differs dramatically from one climate forcing to another.
@@
 
-To compare modelled ice sheets of similar size, we selected for each climate forcing the simulation that leads to a glaciated area closest to the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$.
+To compare modelled ice sheets of similar size, we selected for each climate forcing the simulation that leads to a glaciated area closest to the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$\irina{justify}.
 These qualitative ``best'' runs are presented in Figure~\ref{fig:best} along with their associated temperature depressions and a reconstruction of the ice margin at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) by \citet{dyke-2004}.

### 2013-10-28 floats/figures.tex
Change figure order in results (Qiong)
Implement minor corrections from Arjen

@@
 	\end{center}
-	\caption{Winter (DJF) and summer (JJA) precipitation maps from WorldClim, and winter (DFJ) temperature maps from ERA-Interim reanalysis, NARR (North American Regional Reanalysis), CFSR (Climate System Forecast Reanalysis), and NCEP/NCAR reanalysis climatologies.}
+	\caption{Winter (DJF) and summer (JJA) precipitation maps from WorldClim, and winter (DFJ) temperature maps from ERA-Interim reanalysis, NARR (North American Regional Reanalysis), CFSR (Climate System Forecast Reanalysis), and NCEP/NCAR reanalysis climatologies.
+	Additional forcing data was prepared to correct for wave-like precipitation artefacts in CFSR (section~\ref{sec:climate}).}
 	\label{fig:prec}
+\end{figure*}
+
+% fig:ivolarea
+\begin{figure}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics[width=9cm]{cordillera-climate-ivolarea}
+	\end{center}
+	\caption{Total glaciated area and ice volume after 10\,kyr as a function of temperature offset and climate forcing.}
+	\label{fig:ivolarea}
+\end{figure}
+
+% fig:extent
+\begin{figure*}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics[width=13cm]{cordillera-climate-extent}
+	\end{center}
+	\caption{Extent of ice cover after 10\,kyr as a function of applied temperature offsets for each climate forcing.}
+	\label{fig:extent}
 \end{figure*}
@@
 \end{figure*}
-
-% fig:extent
-\begin{figure*}[t]
-	\vspace*{2mm}
-	\begin{center}
-		\includegraphics[width=13cm]{cordillera-climate-extent}
-	\end{center}
-	\caption{Extent of ice cover after 10\,kyr as a function of applied temperature offsets for each climate forcing.}
-	\label{fig:extent}
-\end{figure*}
-
-% fig:ivolarea
-\begin{figure}[t]
-	\vspace*{2mm}
-	\begin{center}
-		\includegraphics[width=9cm]{cordillera-climate-ivolarea}
-	\end{center}
-	\caption{Total glaciated area and ice volume after 10\,kyr as a function of temperature offset and climate forcing.}
-	\label{fig:ivolarea}
-\end{figure}
 

### 2013-10-28 sections/abstract.tex
Implement corrections from Irina
Mostly as to-do notes for now
Implement minor corrections from Arjen

@@
 We assess model performance against a reconstruction of the ice margin at the Last Glacial Maximum in continental regions of northern Yukon Territory and interior Alaska.
-The best match between model output and the reconstructed ice margin is obtained using climate forcing from the North American Regional Reanalysis, which is a regional climate reanalysis with the highest resolution.
+The best match between model output and the reconstructed ice margin is obtained using the North American Regional Reanalysis, which is a regional climate reanalysis with the highest resolution.
+\irina{Maybe more on conclusions.}
 \end{abstract}

### 2013-10-28 sections/climate.tex
Implement minor corrections from Arjen

@@
 
-WorldClim is a high-resolution climate dataset built from meteorological observations \citep{data:worldclim}.
-It was built by spatial interpolation over global land areas, using 30\,arc\,s-aggregated hole-filled SRTM elevation data \citep{data:srtm} and GTOPO30 elevation data \citep{data:gtopo30}, between a large set of measurements taken at weather stations worldwide.
-This provides a resolution much higher than attained by GCMs in areas where sufficient amount of observations exists.
+WorldClim is a high resolution global climate dataset built from meteorological observations \citep{data:worldclim}.
+It was built by applying a spatial interpolation between a large set of measurements taken at terrestrial weather stations worldwide, using SRTM \citep{data:srtm} and GTOPO30 elevation data \citep{data:gtopo30}.
+This dataset provides a resolution much higher than attained by circulation models, however its reliability is variable to the highly inhomogeneous distribution of measurements \citep{data:worldclim}.
 
@@
 The spatial distribution of JJA and DJF mean precipitation rates in WorldClim reveal a strong precipitation decline beyond the coastal regions (Fig.~\ref{fig:prec})\qiong{Change unit to mm per month.}, primarily as a result of the continuous orographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascades.
-From an ice sheet mass balance point of view, this contrast becomes even stronger due to the difference in timing of the precipitation peak through the year.
+A less pronounced precipitation peak is also visible over the interior ranges.
+From an ice sheet mass balance point of view, these heterogeneities become even stronger due to the difference in timing of the precipitation peak through the year.
 Whereas coastal regions experience most precipitation as snow during the accumulation season (DJF), inland regions experience dry winters and most of the annual precipitation consequently falls as rain during the ablation season (JJA).
@@
 Summer temperature and winter precipitation are most relevant to the glacier model as they respectively drive summer melt and winter accumulation.
-Because reanalyses include observational information through data assimilation, they are closer to observations than simple GCMs in densely monitored regions, while offering physically based output in sparsely monitored regions such as the northern part of our modelling domain \citep{bengtsson-etal-2007}.
+Because reanalyses include observational information through data assimilation, they are closer to observations than pure GCMs in densely monitored regions, such as the southern part of our modelling domain \citep{bengtsson-etal-2007}.
+On the other hand, the atmospheric physics embedded in the models provide more realistic output than obtained by spatial interpolation in sparsely monitored regions, such as the northern part of our modelling domain.
 

### 2013-10-28 sections/discuss.tex
Implement minor corrections from Arjen

@@
 Additionally, we simulate temperature variability by using a constant, uniform value of temperature standard deviation.
-This approach implies large biases of surface mass-balance \citep{charbit-etal-2013,rau-rogozhina-2013,seguinot-inreview}, particularly over regions with such various climates as the Northern American Cordillera.
+This approach implies large biases of surface mass-balance \citep{charbit-etal-2013,rau-rogozhina-2013,seguinot-inpress}, particularly over regions with such various climates as the Northern American Cordillera.

### 2013-10-28 sections/intro.tex
Implement minor corrections from Arjen

--- 2013-09-03 sections/intro.tex
@@
 
-At the Last Glacial Maximum (LGM), glaciers of a size comparable to the present Greenland and Antarctic ice sheets covered parts of Northern America (Laurentide, Cordilleran and Innuitian ice sheets) and Northern Eurasia (Fennoscandian Ice Sheet).
+At the Last Glacial Maximum (LGM), glaciers of a size comparable to the present Greenland and Antarctic ice sheets covered parts of Northern America (Laurentide, Cordilleran and Innuitian ice sheets) and northern Eurasia (Fennoscandian Ice Sheet).
 Numerical modelling of these former ice masses allows for a comparison between glaciological theories embedded in the models and geomorphological traces underpinning palaeo-glaciological reconstructions.
-Yet, a major obstacle in this exercise resides in large uncertainties concerning climate forcing, typically temperature and precipitation data, needed by numerical glacier models \citep{hebeler-etal-2008}.
-This includes uncertainty in representation of Earth's present climate in regions of poor station coverage, and even larger uncertainty concerning accurate reconstructions of past climate change.
+Yet, a major obstacle in this exercise resides in large uncertainties concerning the climate forcing, typically atmospheric temperature and precipitation, required as input to numerical glacier models \citep{hebeler-etal-2008}.
+This includes uncertainty in the representation of Earth's present climate in regions of poor station coverage, and even larger uncertainty concerning accurate reconstructions of past climate change.
 
-Arguably the most physically sound way to force an ice sheet model in the past is to couple it with a General Circulation Model (GCM) \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013}.
+Arguably the most physically sound way to force an ice sheet model for simulations of glacial history is to couple it with a General Circulation Model (GCM) \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013}.
 However the computational demand of GCMs is such that only models of intermediate complexity can run on the time-scales of tens of thousands of years characteristic of ice sheet growth and decay.
 
-Climatologies obtained from uncoupled GCM palaeo-climate simulations such as produced within the PMIP project \citep{joussaume-taylor-1995} provide a more accurate representation of past climate and are commonly used to force ice sheet models.
-Because such climatologies are only available for specific periods of time, this method requires either an assumption of steady-state \citep{huybrechts-tsiobbel-1996}, or interpolation through time between climatologies from different periods, which can be linear \citep{charbit-etal-2002}, or modulated by a ``glacial index'' weighting function derived from ice core $\delta^{18}$O records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}.
-An important drawback in this approach is that palaeo-climate simulations themselves rely on global ice sheet reconstructions such as the ICE-4G \citep{peltier-1994} for their surface topographic boundary condition.
-This exerts indirect influence on subsequently modelled ice sheet geometries, whose spatial extent tends to conform to these reconstructions.
+Climatologies obtained from GCM palaeo-climate simulations such as produced within the Paleoclimate Modelling Intercomparison Project (PMIP) \citep{joussaume-taylor-1995} provide perhaps a more reasonable representation of past climate.
+However, as such climatologies are only available for specific periods of time, using them as climate forcing for an ice sheet model requires either an assumption of steady-state ice sheet response to climatic fluctuations \citep{huybrechts-tsiobbel-1996}, or interpolation through time between climatologies from different periods, which can be linear \citep{charbit-etal-2002}, or modulated by a ``glacial index'' weighting function derived from ice core $\delta^{18}$O records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}.
+An important drawback in this approach is that GCM palaeo-climate simulations themselves rely on an accurate description of the global surface topography and therefore include global ice sheet reconstructions such as ICE-4G \citep{peltier-1994} for their surface topographic boundary condition.
+This condition potentially exerts influence on subsequently modelled ice sheet geometries, and particularly glacier extent.
+\arjen{Has this been shown? Add ref.}
 
-The other side of the coin is that avoiding this circular dependence unfortunately goes hand-in-hand with a need for simplifying assumptions on Earth's past climate.
-They include energy balance modelling approaches \citep{tarasov-peltier-1997} and geographic parametrizations of surface mass balance \citep{robert-1991} or climate forcing \citep{johnson-fastook-2002}.
-Standing on a middle ground, temperature offset methods \citep{greve-etal-1999,bintanja-etal-2005} make use of the high level of detail available in present climate datasets such as gridded observation datasets, GCM output or reanalyses, while using simplifying representations of past climate deviations from this present state.
+The other side of the coin is that avoiding this circular dependence unfortunately goes hand-in-hand with a need for simplifying assumptions regarding Earth's past climate.
+Such studies include energy balance modelling approaches \citep{tarasov-peltier-1997} and geographic parametrizations of surface mass balance \citep{robert-1991} or climate forcing \citep{johnson-fastook-2002}.
+Standing on a middle ground, temperature offset methods \citep{greve-etal-1999,bintanja-etal-2005} make use of the high level of detail available in present climate datasets such as gridded observation datasets, GCM output or reanalyses, while using simplifying representations of past climate deviations from the present state.
 
-Here, we propose to address some of the uncertainties concerning climate forcing of numerical glacier models by evaluating their responses, in terms of glacier extent, to inputs from several climate datasets.
-Few studies of this kind are presently available.
-\citet{quiquet-etal-2012} assessed the sensitivity of a Greenland ice sheet model to various atmospheric forcing, including a regional parametrization \citep{fausto-etal-2009}, output from several GCMs and a climate reanalysis.
-\citet{rodgers-etal-2004} and \citet{charbit-etal-2007} tested the sensitivity of a model of the Northern Hemisphere ice sheets to climate forcing from different PMIP LGM simulations.
-Here, to limit degrees of freedom in our model, and obtain results independent of palaeo-ice sheet reconstructions such as the ICE-4G, we use a simple temperature offset approach similar to \citet{greve-etal-1999} and \citet{bintanja-etal-2005} and assess ice sheet model sensitivity to the choice of present-day climate data.
-Rather than using GCM output, we force our model with climate reanalysis data, which include observational information through data assimilation \citep{bengtsson-etal-2007}.
-Furthermore, we focus our study regionally on the former Cordilleran Ice Sheet in Northern America.
+Here, we propose to address some of the uncertainties concerning climate forcing of numerical glacier models by evaluating the responses of a numerical model, in terms of glacier extent, to inputs fromo several climate datasets.
+This is not entirely a new approach as a few studies of this kind are presently available.
+\citet{quiquet-etal-2012}, for example, assessed the sensitivity of a Greenland ice sheet model to various atmospheric forcing, including a regional parametrization by \citet{fausto-etal-2009}, output from several GCMs and an atmospheric reanalysis.
+\citet{rodgers-etal-2004} and \citet{charbit-etal-2007} tested the sensitivity of a model of the Northern Hemisphere ice sheet model to climate forcing from different PMIP LGM simulations\qiong{Detail in a few lines the conclusions of theses studies}.
+In our study, to limit the degrees of freedom in our model and obtain results independent of palaeo-ice sheet reconstructions such as ICE-4G, we use a simple temperature offset similar to the approaches by \citet{greve-etal-1999} and \citet{bintanja-etal-2005} and assess ice sheet model sensitivity to the choice of present-day climate data.
+Hence, rather than using GCM output, we force our model with climate reanalysis data, which builds on observational information through data assimilation \citep{bengtsson-etal-2007}.
+Furthermore, we focus our study regionally on the former Cordilleran Ice Sheet in western Northern America.
 
 The Cordilleran Ice Sheet (Fig.~\ref{fig:locmap})\irina{include all geographic names mentioned in the text}\arjen{elevation scale white-end can be confused with glaciers} covered an area that presently experiences strong regional variations in climate.
-In a numerical modelling perspective, it is one of the least studied palaeo-ice sheets of the Northern Hemisphere, despite the fact that significant geomorphological data is available to constrain its extent \citep{jackson-clague-1991,dukrodkin-1999,kaufman-manley-2004,kleman-etal-2010,margold-etal-2011}.
-Although the Cordilleran Ice Sheet was not the primary target of these simulations, it has previously been modelled as part of efforts to simulate ice sheets in Northern America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} and world-wide \citep{yoshimori-etal-2001}.
+From a numerical modelling perspective, it is one of the least studied palaeo-ice sheets of the Northern Hemisphere, despite the fact that significant stratigraphical and geomorphological data is available to constrain its extent \citep{jackson-clague-1991,dukrodkin-1999,kaufman-manley-2004,kleman-etal-2010,margold-etal-2011}.
+Although the Cordilleran Ice Sheet was not the primary target of these simulations, it has previously been modelled as part of efforts to simulate ice sheets in North America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} and world-wide \citep{yoshimori-etal-2001}.
 While these studies generally reproduce the magnitude of North American glaciation at LGM reasonably well, there exists a tendency in the simulations that are independent of ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and interior Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.
 
-Here we use a PISM, a Parallel Ice Sheet Model \citep{web:pism}\julien{Citation recommended by \url{http://www.pism-docs.org/wiki/doku.php?id=citing_pism}}, to simulate the extent and thickness of the Cordilleran Ice Sheet at the LGM.
-We force our model with multiple climate datasets and compare our results to the mapped LGM ice sheet margin from \citet{dyke-2004}.
-To our knowledge, this is the first modelling study that specifically focuses on the Cordilleran Ice Sheet since the one by \citet{robert-1991}.
-While the modelling domain of \citet{robert-1991} covered only the southern part of the ice sheet, ours encompasses most of the LGM ice extent and adjacent non-glaciated parts of northern Yukon Territory and interior Alaska (Fig.~\ref{fig:locmap}).
-Our model also includes higher resolution and an improved treatment of ice thermodynamics and bedrock response following the large advances made in scientific computing and ice sheet modelling since then.
+Here we use PISM, a Parallel Ice Sheet Model \citep{web:pism}, to simulate the extent and thickness of the Cordilleran Ice Sheet at the LGM (Fig.~\ref{fig:locmap}).
+We force our model with multiple climate datasets and compare the modelled ice extent to a geomorphological reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.
+Thereby, we aim to determine the climate dataset which is most suited for simulation of the Cordilleran Ice Sheet and to use as input for future, transient studies over a glacial cycle.
 Model set-up is presented in section~\ref{sec:model} and climate forcing in section~\ref{sec:climate}.
 Results are exposed in section~\ref{sec:results} and discussed in section~\ref{sec:discussion}.
+To our knowledge, this is the first modelling study that specifically focuses on the Cordilleran Ice Sheet since \citet{robert-1991}.
+
+%While the modelling domain of \citet{robert-1991} covered only the southern part of the ice sheet, ours encompasses most of the LGM ice extent and adjacent non-glaciated parts of northern Yukon Territory and interior Alaska (Fig.~\ref{fig:locmap}).
+%Our model also includes higher resolution and an improved treatment of ice thermodynamics and bedrock response following the large advances made in scientific computing and ice sheet modelling since then.

### 2013-10-28 sections/model.tex
Implement minor corrections from Arjen

@@
 
-We use PISM, a Parallel Ice Sheet Model, version stable~0.5.12 (e.g.,) \citep{bueler-brown-2009,winkelmann-etal-2011,aschwanden-etal-2012,web:pism}.
-Given basal topography\irina{geothermal heat flux and sea level} and climate forcing, the model computes the thermal and dynamic state of an ice sheet and the associated lithospheric response.
-Our modelling domain encompasses most of the area covered by the Cordilleran Ice Sheet at LGM except from the western Alaskan ranges and includes non-glaciated regions in northern Yukon Territory and interior Alaska.
+Given basal topography, sea level, geothermal heat flux and climate forcing, the model computes ice extent and thickness, its thermal and dynamic state, and the associated lithospheric response.
+Our modelling domain encompasses the entire area covered by the Cordilleran Ice Sheet at LGM including independent ice growing on the western Alaskan ranges and the Brooks Range and intermediated expanses of ice-free terrain in northern Yukon Territory and interior Alaska (Fig.~\ref{fig:locmap}).
+
+As we aim to model glacial inception and growth of the Cordilleran Ice Sheet towards a configuration as last attained during the LGM, we mimic palaeo-climatic conditions by applying constant temperature offsets homogeneously over the modelling domain.
+Each simulation starts from ice-free conditions and runs for 10\,kyr, a time interval representative of the rapid build-up of the last Cordilleran Ice Sheet from nearly ice-free to full glacial conditions \citep{clague-1989,stroeven-etal-2010}.
 Our simulations are performed in parallel on 32 cores at the Swedish National Supercomputing Center.
@@
 
-The Shallow Shelf Approximation (SSA) is used as a ``sliding law'' for the Shallow Ice Approximation (SIA) \citep{winkelmann-etal-2011}.
+The central part of an ice sheet model consist of the computation of flow velocity which itself depends on temperature.
+PISM is a shallow model, which implies that the balance of stresses is approximated based on their predominant components.
+On the other hand, the model is polythermal: it accounts for differences in temperature and softness within the ice column.
+
+The Shallow Shelf Approximation (SSA) is used as a ``sliding law'' for the Shallow Ice Approximation (SIA) \citep{bueler-brown-2009,winkelmann-etal-2011}.
 SIA and SSA velocities are computed by finite difference methods on a 10\,km-resolution horizontal grid of 300 by 150 points (the modelling domain).
 Ice softness depends on temperature and water content through an enthalpy formulation \citep{aschwanden-blatter-2009,aschwanden-etal-2012}.
-Enthalpy is computed in three dimensions in up to 51 irregularly spaced layers in ice, and temperature is further computed in 11 regularly spaced layers in bedrock.
+Enthalpy is computed in three dimensions in up to 51 irregularly spaced layers within the ice, and temperature is additionally computed in 11 regularly spaced layers in bedrock to a depth of 1\,km.
 A uniform geothermal heat flux of 70\,\unit{W\,m^{-2}} provides the lower boundary condition to the bedrock thermal model, and surface air temperature from the climate forcing provides the upper boundary condition to the ice enthalpy model.
@@
 The yield stress is modelled using the Mohr-Coulomb criterion.
-The till friction angle $\phi$ varies from 10 to 30\degree\ and is a function of contemporary bed elevation, with lowest values occurring at low elevations,
+The till friction angle $\phi$ varies from 10 to 30\degree\ and is a function of current bed elevation, with lowest values occurring at low elevations,
 \julien{phi is independent from basal water which is computed by the model (and sliding depends on both) so the 10 to 30 degree range is meant to reflect different bedrock/sediment surfaces rather than dry vs wet beds.}
@@
 
-where $\phi$ is given in degrees and $z$ in meters above contemporary sea level.
+where $\phi$ is given in degrees and $z$ in meters above current sea level.
 This models weakening of the till associated with the presence of marine sediments \citep{martin-etal-2011,aschwanden-etal-2013}.
-Basal topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography and bathymetry dataset \citep{data:etopo1}.\julien{NaturalEarth data are vectors (coastline, rivers, etc.) used in Fig.
+Basal topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography and bathymetry dataset with a resolution of 1\,arc-minute \citep{data:etopo1}.\julien{NaturalEarth data are vectors (coastline, rivers, etc.) used in Fig.
 1, not model input.} Sea level is lowered by 120\,m and basal topography responds to ice load following a regional isostasy model that includes lithospheric flexure and mantle relaxation \citep{lingle-clark-1985}.
@@
 It accounts for temperature variability assuming a normal distribution along a central (input) value.
-The temperature standard deviation is a constant model parameter and was assigned a value of 3.07\,\unit{\degree C}, which corresponds to the summer (JJA) mean, model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from North American Regional Reanalysis data \citep{data:narr} in a manner similar to \citet{seguinot-inreview}.
+The temperature standard deviation is a constant model parameter and was assigned a value of 3.07\,\unit{\degree C}, which corresponds to the mean summer (JJA), model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from North American Regional Reanalysis data \citep{data:narr} in a manner similar to \citet{seguinot-inpress}.
 The ablation model incorporates degree-day factors of 3.04\,\unit{mm\,K^{-1}\,day^{-1}} for snow and 4.59\,\unit{mm\,K^{-1}\,day^{-1}} for ice, as derived from mass-balance measurements on contemporary glaciers from the Coast Mountains and Rocky Mountains in British Columbia \citep{shea-etal-2009}.
@@
 Prior to surface mass balance computation, the model dynamically applies a lapse-rate correction to surface air temperature.
-This correction accounts for evolution of ice thickness on the one hand, and differences between the climate forcing reference topography and the ice flow model basal topography on the other hand.
+This correction accounts for the evolution of ice thickness on the one hand, and differences between the climate forcing reference topography and the ice flow model basal topography on the other hand.
 It uses a reference topography distinct from the modelled basal topography and specific to each climate forcing dataset (Fig.~\ref{fig:topo}).
-An annual lapse-rate of 6\,\unit{\degree C\,km^{-1}} is used in all simulations.
-No lapse-rate corrections apply to precipitation.
-
-As we aim to model glacial inception and growth of the Cordilleran Ice Sheet towards a configuration as last attained during the LGM, we mimic palaeo-climatic conditions by applying constant temperature offsets homogeneously over the modelling domain.
-Each simulation starts from ice-free conditions and runs for 10\,kyr, a time interval representative of the rapid build-up of the last Cordilleran Ice Sheet from nearly ice-free to full glacial conditions \citep{clague-1989,stroeven-etal-2010}.
+An annual lapse rate of 6\,\unit{\degree C\,km^{-1}} is used in all simulations.
+No corrections apply to precipitation changes with elevation.

### 2013-10-28 sections/results.tex
Implement minor corrections from Arjen

@@
 
-Using WorldClim, NCEP/NCAR, ERA-Interim, CFSR (smoothed and original datasets) and NARR input climatologies as a climate forcing (section~\ref{sec:climate}) for the ice sheet model PISM (section~\ref{sec:model}), we run 66~simulations of glacial inception and growth of the Cordilleran ice sheet over 10~kyr using temperature offsets ranging from 0 to 10\,\unit{\degree C}.
+Using WorldClim, NCEP/NCAR, ERA-Interim, CFSR (smoothed and original datasets) and NARR input climatologies as climate forcing, we run 66~simulations of glacial inception and growth of the Cordilleran ice sheet over 10~kyr using temperature offsets ranging from 0 to 10\,\unit{\degree C}.
 Significant differences in patterns of ice sheet growth, final glacial extent, and volume were attained (Fig.
@@
 Under present climate (0\,\unit{\degree C} offset), WorldClim, NARR and ERA-Interim forcing produce local ice caps in areas presently glaciated (Fig.~\ref{fig:extent}).
-This results are fairly consistent with contemporary ice coverage, although both the 10~km horizontal resolution, and the occurrence of glaciers and ice caps in the ETOPO1 input basal topography prevent a detailed comparison.
-CFSR forcing, on the other hand, yields continuous ice cover over the northern Coast Mountains, and NCEP/NCAR forcing results in ice-sheet growth, which is largely inconsistent with the present state of glaciation.
+These results are fairly consistent with contemporary ice distribution, although both the 10~km horizontal resolution and the occurrence of glaciers and ice caps surface topographies in the ETOPO1 dataset used for basal topography prevent a meaningful detailed comparison.
+CFSR forcing, on the other hand, yields continuous ice cover over the northern Coast Mountains, and NCEP/NCAR forcing results in ice sheet growth, both of which are largely inconsistent with the present state of glaciation.
 At the opposite side of the temperature offset range, it can be noted that both the NCEP/NCAR and CFSR forcing produce oversized ice sheets whose extent are primarily restricted by the domain boundaries (Fig.~\ref{fig:extent}).
@@
 
-To contrast the six different input climatologies, we plot simulations run with the 5\,\unit{\degree C} temperature offset (Fig.~\ref{fig:cool05}).
+To contrast the effect of the six different input climatologies, we plot simulations using a 5\,\unit{\degree C} temperature offset (Fig.~\ref{fig:cool05}).
 The magnitude of glaciation differs dramatically from one climate forcing to another.
@@
 To compare modelled ice sheets of similar size, we selected for each climate forcing the simulation that leads to a glaciated area closest to the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$\irina{justify}.
-These qualitative ``best'' runs are presented in Figure~\ref{fig:best} along with their associated temperature depressions and a reconstruction of the ice margin at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) by \citet{dyke-2004}.
-We observe that although similar in size, the selected ice sheets are different in shape and flow patterns.
-More specifically, ice sheets modelled using NCEP/NCAR, CFSR and ERA-Interim input climatologies cover parts of Yukon Territory and Alaska that were shown to have remained ice-free for at least several glacial cycles \citep{dukrodkin-1999,kaufman-manley-2004}.
+These qualitative ``best'' runs are presented in Figure~\ref{fig:best} along with their associated temperature depressions and a reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.
+We observe that although similar in size, the selected ice sheets are different in shape and, therefore, flow patterns.
+More specifically, ice sheets modelled using NCEP/NCAR, CFSR and ERA-Interim input climatologies cover parts of Yukon Territory and Alaska that were shown to have remained ice free \citep{dukrodkin-1999,kaufman-manley-2004}.

### 2013-10-30 sections/discuss.tex
Resection the discussion and add new figures

@@
 % ----------------------------------------------------------------------
+
+\subsection{Comparison of forcing climatologies}
+\emph{Fig.
+topo-prec: there are large differences in climate forcing.
+(1) Temperature diff are due to topography, except for cold NCAR.
+(2) Precipitation diff related to orographic effects, except for CFSR (Fig.~\ref{fig:tempheatmap}-\ref{fig:oroprecip}).}
+
+\subsection{Sensitivity to climate forcing}
+\emph{Most differences are due to precip, except for NCAR (Fig.~\ref{fig:biatm}).}
+
+The outcome of our numerical simulations strongly depends on which dataset is used for climate forcing (Fig.~\ref{fig:cool05}-\ref{fig:best}).
+This results from differences in temperature and precipitation (Fig.~\ref{fig:temp}-\ref{fig:prec}) between forcing datasets, to which the glacier model is highly sensitive.
+
+Difference between modelled ice sheet geometries is notably visible in northern Yukon Territory and Alaska, where previous modelling studies commonly produced too extensive ice cover, whereas geomorphological data show that the ice cover was sparse.
+Concerning reanalysis datasets, we interpret these differences as mainly the result of different GCM resolutions and related model physics (Fig.~\ref{fig:topo}, Table~\ref{tab:reanalyses}).
+
+As described in section~\ref{sec:climate}, the topography of the Northern American Cordillera is such that its western ranges form a continuous orographic barrier, causing high precipitation along the Pacific coast while leaving much of the interior arid.
+However the ability of a GCM to reproduce these contrasts in precipitation is bounded by resolution.
+In a model of coarser resolution such as used in the NCEP/NCAR reanalysis, these high mountain ranges are reduced to smoothed hills and the modelled distribution of precipitation is also smoother (Fig.~\ref{fig:oroprecip}).
+This interpretation does not apply to the CFSR climatology, which exhibits high inland precipitation rates despite of the high model resolution (Table~\ref{tab:reanalyses}).
 
@@
 
-\subsection{Sensitivity to climate forcing}
+\subsection{Sensitivity to glaciation duration}
+\emph{Glaciation duration has upper bound, but it may have been shorter.
+Shorter gives better results, so 10Ka may be too much, or at least it shows that our approach may be too simplistic.
+Longer gives worst result: the LGM CIS is not at steady-state (Fig.~\ref{fig:durationstack}).}
 
-The outcome of our numerical simulations strongly depends on which dataset is used for climate forcing (Fig.~\ref{fig:cool05}-\ref{fig:best}).
-This results from differences in temperature and precipitation (Fig.~\ref{fig:temp}-\ref{fig:prec}) between forcing datasets, to which the glacier model is highly sensitive.
-
-Difference between modelled ice sheet geometries is notably visible in northern Yukon Territory and Alaska, where previous modelling studies commonly produced too extensive ice cover, whereas geomorphological data show that the ice cover was sparse.
-Concerning reanalysis datasets, we interpret these differences as mainly the result of different GCM resolutions and related model physics (Fig.~\ref{fig:topo}, Table~\ref{tab:reanalyses}).
-
-As described in section~\ref{sec:climate}, the topography of the Northern American Cordillera is such that its western ranges form a continuous orographic barrier, causing high precipitation along the Pacific coast while leaving much of the interior arid.
-However the ability of a GCM to reproduce these contrasts in precipitation is bounded by resolution.
-In a model of coarser resolution such as used in the NCEP/NCAR reanalysis, these high mountain ranges are reduced to smoothed hills and the modelled distribution of precipitation is also smoother (Fig.~\ref{fig:oroprecip}).
-This interpretation does not apply to the CFSR climatology, which exhibits high inland precipitation rates despite of the high model resolution (Table~\ref{tab:reanalyses}).
-
-\subsection{Potential method improvements}
+\subsection{Other potential method improvements}
 

### 2013-10-31 sections/discuss.tex
New discussion first draft

--- 2013-10-28 sections/discuss.tex
@@
 % ----------------------------------------------------------------------
+
+The outcome of our numerical simulations shows a very large sensitivity to the choice of climate forcing data (Fig.~\ref{fig:ivolarea}-\ref{fig:cool05}).To understand the origin of these discrepancies, we compare the distributions of surface air temperature and precipitation rate among between datasets and test their effect on modelled ice sheet geometries.
+We then compare our findings to a reconstruction of the LGM Cordilleran Ice Sheet margin, and discuss some of the key assumptions of this study.
+
+\subsection{Comparison of forcing climatologies}
+
+The PDD model that we use is potentially highly sensitive to temperature and precipitation forcing.
+More particularly, winter (DJF) precipitation and summer (JJA) temperature respectively predominantly drive accumulation and melt at the surface of the modelled ice sheet.
+As illustrated (Fig.~\ref{fig:topo}-\ref{fig:prec}), significant differences between datasets exist in their distribution exist.
+
+Surface air temperature from the WorldClim observational data is fairly well resolved by reanalysis data (Fig.~\ref{fig:tempheatmap}), apart of the NCAR reanalysis which shows a large spread around the WorldClim values.
+For ERA-Interim, NARR and CFSR reanalyses, it seems that most of the differences are due to unresolved local topographic detail (Fig.~\ref{fig:tempdiff}).
+NCAR shows a significant cold anomaly over the entire region.
+It is known from literature that NCAR produces too cold temperature at high latitudes.
+Finally NARR data seem to be a bit too warm.
+
+Regarding precipitation data, more significant differences exist between datasets.
+All reanalysis data used seem to overestimate precipitation rates over most of the modelling domain (Fig.~\ref{fig:precheatmap}).
+The overestimation is the largest for CFSR data.
+Again, NCAR shows the largest discrepancy and this is probably due to the coarse model resolution.
+When looking at the spatial distribution of precipitation anomalies, it appears that precipitation is underestimated on most of the major reliefs, while it is overestimated on the lowlands and plateaux located beyond these reliefs (Fig.~\ref{fig:precdiff}).
+We think this is an orographic precipitation effect linked to horizontal GCM resolution (Fig.~\ref{fig:oroprecip}).
+Precipitation rate, however, is generally and largely overestimated in CFSR data, despite of high model resolution.
+
+\subsection{Sensitivity to climate forcing}
+
+The outcome of our numerical simulations strongly depends on which dataset is used for climate forcing (Fig.~\ref{fig:cool05}-\ref{fig:best}).
+This results from differences in temperature and precipitation (Fig.~\ref{fig:temp}-\ref{fig:prec}) between forcing datasets, to which the glacier model is highly sensitive.
+
+Difference between modelled ice sheet geometries is notably visible in northern Yukon Territory and Alaska, where previous modelling studies commonly produced too extensive ice cover, whereas geomorphological data show that the ice cover was sparse.
+Concerning reanalysis datasets, we interpret these differences as mainly the result of different GCM resolutions and related model physics (Fig.~\ref{fig:topo}, Table~\ref{tab:reanalyses}).
+
+As described in section~\ref{sec:climate}, the topography of the Northern American Cordillera is such that its western ranges form a continuous orographic barrier, causing high precipitation along the Pacific coast while leaving much of the interior arid.
+However the ability of a GCM to reproduce these contrasts in precipitation is bounded by resolution.
+In a model of coarser resolution such as used in the NCEP/NCAR reanalysis, these high mountain ranges are reduced to smoothed hills and the modelled distribution of precipitation is also smoother (Fig.~\ref{fig:oroprecip}).
+This interpretation does not apply to the CFSR climatology, which exhibits high inland precipitation rates despite of the high model resolution (Table~\ref{tab:reanalyses}).
+
+To distinguish the role of temperature and precipitation biases outlined above in forcing the ice sheet model, we run a series of additional experiments using artificial, “hybrid” atmospheres that consist of temperature data from WorldClim, and precipitation data from each of the four reanalysis, and the other way around.
+These eight simulations were run using a temperature offset value of 5\,K, and temperature lapse rates were computed using the reference topography from the appropriate data.
+
+Although only temperature or precipitation changes are applied, this experiment again resulted in a largely different modelled ice sheet geometries (Fig.~\ref{fig:biatm}).
+Varying temperature forcing alone however, had limited effect, expect in the case of NCAR, due to its cold bias previously mentioned.
+For ERA-Interim, NARR and CFSR data, variations in precipitations dominate the ice sheet response, whereas for NCAR data, both temperature and precipitation are responsible for the anomalously large ice sheets produced (Fig.~\ref{fig:biatmbars}).
 
@@
 
-\subsection{Sensitivity to climate forcing}
+\subsection{Sensitivity to englaciation duration}
 
-The outcome of our numerical simulations strongly depends on which dataset is used for climate forcing (Fig.~\ref{fig:cool05}-\ref{fig:best}).
-This results from differences in temperature and precipitation (Fig.~\ref{fig:temp}-\ref{fig:prec}) between forcing datasets, to which the glacier model is highly sensitive.
+Among the most decisive assumptions for the present study figures certainly the hypothesis that the Cordilleran Ice Sheet built up from ice free conditions to LGM conditions in 10Ka, which motivates our choice of simulation length.
+Although geomorphological data shows that this period could not have been much longer, they provide no lower bound for it.
+The growth of the Cordilleran Ice Sheet to full glacial condition could have lasted less than 10Ka.
 
-Difference between modelled ice sheet geometries is notably visible in northern Yukon Territory and Alaska, where previous modelling studies commonly produced too extensive ice cover, whereas geomorphological data show that the ice cover was sparse.
-Concerning reanalysis datasets, we interpret these differences as mainly the result of different GCM resolutions and related model physics (Fig.~\ref{fig:topo}, Table~\ref{tab:reanalyses}).
+In our simulations, using NARR data and temperatures more than 7 degrees below present, the modelled ice sheet reaches near-LGM ice cover in less than 10Ka (Fig.~\ref{fig:durationstack}).
+A shorter glaciation period lead to fewer ice cover in the eastern part of the modelling domain, where precipitation is limiting the rate of growth, but further glaciation along the Pacific margin to the south, where warm temperatures normally stop glacier advance.
+This shows that our choice of 10Ka may be somewhat too much.
+On the other hand our representation of climate history is voluntarily simplistic, and the effects of a more complex palaeo-climate history on the Cordilleran Ice Sheet will need to be studied further.
 
-As described in section~\ref{sec:climate}, the topography of the Northern American Cordillera is such that its western ranges form a continuous orographic barrier, causing high precipitation along the Pacific coast while leaving much of the interior arid.
-However the ability of a GCM to reproduce these contrasts in precipitation is bounded by resolution.
-In a model of coarser resolution such as used in the NCEP/NCAR reanalysis, these high mountain ranges are reduced to smoothed hills and the modelled distribution of precipitation is also smoother (Fig.~\ref{fig:oroprecip}).
-This interpretation does not apply to the CFSR climatology, which exhibits high inland precipitation rates despite of the high model resolution (Table~\ref{tab:reanalyses}).
-
-\subsection{Potential method improvements}
+\subsection{Other potential method improvements}
 

### 2013-11-05 sections/discuss.tex
New discussion second draft

--- 2013-10-28 sections/discuss.tex
@@
 
+The outcome of our numerical simulations shows a very large sensitivity to the choice of climate forcing data (Fig.~\ref{fig:ivolarea}--\ref{fig:cool05}).To understand the origin of these discrepancies, we quantify surface air temperature and precipitation rate differences between input climatologies and test their effect on modelled ice sheet geometries.
+We then compare our findings to a reconstruction of the LGM Cordilleran Ice Sheet margin, and discuss some of the key assumptions of this study.
+
+\subsection{Comparison of forcing climatologies}
+
+The PDD model used in this study is highly sensitive to surface air temperature and precipitation forcing.
+As previously outlined, winter (DJF) precipitation and summer (JJA) temperature are the predominant controls on accumulation and melt at the modelled ice sheet surface.
+Significant differences in their distribution exist between datasets (Fig.~\ref{fig:topo}--\ref{fig:prec}).
+
+In this section we use the WorldClim climatology, which is derived from observations, as a reference with which to compare other climate forcing data on land.
+ERA-Interim, CFSR and NARR climatologies exhibit temperature distributions fairly consistent with WorldClim data (Fig.~\ref{fig:tempheatmap}).
+When examining the spatial distribution of temperature differences to WorldClim data, it appears that most of disparity between reanalyses and observation data is caused by topographical detail at scales unresolved by the atmospheric reanalyses (Fig.~\ref{fig:tempdiff}).
+Surface air temperature data from the NCAR climatology shows a large disparity with that of WorldClim, with a significant cold anomaly over most of the modelling domain (Fig.~\ref{fig:tempheatmap}--\ref{fig:tempdiff}).
+
+Important differences in precipitation rate exist between input climatologies.
+All four reanalyses used in this study exhibit higher precipitation than WorldClim data (Fig.~\ref{fig:precheatmap}).
+However the magnitude of this anomaly differs significantly between datasets.
+The spatial distributions of precipitation differences between reanalyses and WorldClim data show that negative precipitation anomalies are generally constrained to the windward slope of the major mountain ranges, while positive precipitation anomalies are found on the leeward slope of these ranges and extend downwind to interior plateaux and lowlands (Fig.~\ref{fig:precdiff}).
+Most likely, this is the signature of an orographic precipitation effect, whose reproduction by atmosphere reanalyses is limited by horizontal GCM resolution (Fig.~\ref{fig:oroprecip}).
+However this effect alone can not explain the widespread and pronounced positive precipitation anomalies of CFSR data (Fig.~\ref{fig:precheatmap}--\ref{fig:precdiff}).
+
+\subsection{Model sensitivity to climate forcing}
+
+To distinguish the effects of temperature and precipitation biases in the ice sheet model, we run a series of eight additional simulations.
+These new simulations use an “hybrid” atmosphere forcing that consist of temperature data from WorldClim, combined with precipitation data from each of the four reanalyses, and vice-versa.
+A single temperature offset value of 5\,K is used.
+This value was chosen to allow comparison with previous results (Fig.~\ref{fig:cool05}).
+In each case, temperature lapse rates are computed using the reference topography from the corresponding dataset.
+
+Although only temperature or precipitation changes are applied, this experiment results once more in large differences in modelled ice sheet geometries (Fig.~\ref{fig:biatm}) as compared to the reference WorldClim 5\,K run (Fig.~\ref{fig:cool05}, top-left panel).
+For ERA-Interim, CFSR and NARR climatologies, precipitation anomalies clearly dominate the differences in ice sheet response, whereas temperature anomalies have relatively little effect (Fig.~\ref{fig:biatmbars}).
+However in the case of NCAR forcing, both the negative temperature bias (Fig.~\ref{fig:tempheatmap}--\ref{fig:tempdiff}) and the important precipitation differences (Fig.~\ref{fig:precheatmap}--\ref{fig:precdiff}) contribute to produce oversized ice sheets (Fig.~\ref{fig:biatmbars}).
+
 \subsection{Comparison to reconstructed LGM margin}
@@
 This corresponds to the LGM extent of most of the Cordilleran ice sheet, which occurred later than in the Laurentide ice sheet \citep{dyke-2004}.
-Although the outcome of our numerical simulations strongly depends on which dataset is used as a climate forcing, common discrepancies between the modelled ice sheet geometry and the ice margin mapped by \citet{dyke-2004} can be observed (Fig.~\ref{fig:cool05}-\ref{fig:best}).
+
+To compare modelled ice sheets of similar size, we selected for each climate forcing the simulation that leads to a final glaciated area closest to the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$\irina{justify}.
+These runs are presented in Figure~\ref{fig:best} along with their associated temperature depressions and a reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.
+
+Temperature depression necessary to reach the LGM extent differ significantly between input climatologies.
+Also, although similar in size, the selected ice sheets are different in shape.
+More specifically, ice sheets modelled using ERA-Interim, CFSR,  and NCEP/NCAR and  input climatologies cover parts of Yukon Territory and Alaska that were shown to have remained ice free \citep{dukrodkin-1999,kaufman-manley-2004}.
+
+Although the outcome of our numerical simulations strongly depends on which dataset is used as a climate forcing, common discrepancies between the modelled ice sheet geometry and the ice margin mapped by \citet{dyke-2004} can be observed (Fig.~\ref{fig:best}).
 
@@
 
-\subsection{Sensitivity to climate forcing}
-
-The outcome of our numerical simulations strongly depends on which dataset is used for climate forcing (Fig.~\ref{fig:cool05}-\ref{fig:best}).
-This results from differences in temperature and precipitation (Fig.~\ref{fig:temp}-\ref{fig:prec}) between forcing datasets, to which the glacier model is highly sensitive.
-
 Difference between modelled ice sheet geometries is notably visible in northern Yukon Territory and Alaska, where previous modelling studies commonly produced too extensive ice cover, whereas geomorphological data show that the ice cover was sparse.
@@
 
-\subsection{Potential method improvements}
+\subsection{Model sensitivity to englaciation period}
+
+A key assumption for the present study is our choice of 10\,Ka for the length of all our simulations.
+Although geomorphological data shows that the period of growth of the Cordilleran Ice Sheet from nearly ice-free to full glacial conditions could not have been much longer than 10\,Ka, they provide no lower bound to this value.
+
+To test the effect of a shorter englaciation period on model results, we use simulations run under NARR climate forcing with temperatures offsets of 7 to 11\,K, and compare modelled ice sheet extent when glaciated area reach the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$.
+
+This experiment shows that a shorter englaciation period leads to more restricted ice cover in the eastern part of the modelling domain, where precipitation rate control ice advance, but further glaciation along the Pacific margin to the south, where temperature is the limiting factor (Fig.~\ref{fig:durationstack}).
+Reducing simulation length lead to a closer match with the LGM ice sheet margin by \citet{dyke-2004} in some parts of the modelling domain, but higher discrepancy in other parts.
+
+Therefore, our choice of 10\,Ka simulation length could certainly be discussed further.
+However, our representation of climate history by constant temperature offsets is voluntarily simplistic.
+To better understand the transient character of the LGM Cordilleran Ice Sheet, time-dependent palaeo-climate forcing would be needed, and this is left for future research.
+
+\subsection{Other potential method improvements}
 

### 2013-11-05 sections/results.tex
New discussion second draft

@@
 The WorldClim run only resulted in the growth of mountain ice caps bounded to the highest topography.
-
-To compare modelled ice sheets of similar size, we selected for each climate forcing the simulation that leads to a glaciated area closest to the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$\irina{justify}.
-These qualitative ``best'' runs are presented in Figure~\ref{fig:best} along with their associated temperature depressions and a reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.
-We observe that although similar in size, the selected ice sheets are different in shape and, therefore, flow patterns.
-More specifically, ice sheets modelled using NCEP/NCAR, CFSR and ERA-Interim input climatologies cover parts of Yukon Territory and Alaska that were shown to have remained ice free \citep{dukrodkin-1999,kaufman-manley-2004}.

### 2013-11-05 sections/discuss.tex
Rework comparison to LGM margin subsection

@@
 
-The outcome of our numerical simulations shows a very large sensitivity to the choice of climate forcing data (Fig.~\ref{fig:ivolarea}--\ref{fig:cool05}).To understand the origin of these discrepancies, we quantify surface air temperature and precipitation rate differences between input climatologies and test their effect on modelled ice sheet geometries.
+The outcome of our numerical simulations shows a very large sensitivity to the choice of climate forcing data (Fig.~\ref{fig:ivolarea}--\ref{fig:cool05}).
+To understand the origin of these discrepancies, we quantify surface air temperature and precipitation rate differences between input climatologies and test their effect on modelled ice sheet geometries.
 We then compare our findings to a reconstruction of the LGM Cordilleran Ice Sheet margin, and discuss some of the key assumptions of this study.
@@
 The spatial distributions of precipitation differences between reanalyses and WorldClim data show that negative precipitation anomalies are generally constrained to the windward slope of the major mountain ranges, while positive precipitation anomalies are found on the leeward slope of these ranges and extend downwind to interior plateaux and lowlands (Fig.~\ref{fig:precdiff}).
-Most likely, this is the signature of an orographic precipitation effect, whose reproduction by atmosphere reanalyses is limited by horizontal GCM resolution (Fig.~\ref{fig:oroprecip}).
-However this effect alone can not explain the widespread and pronounced positive precipitation anomalies of CFSR data (Fig.~\ref{fig:precheatmap}--\ref{fig:precdiff}).
+
+Most likely, this is the signature of an orographic precipitation effect.
+As described in section~\ref{sec:climate}, the topography of the Northern American Cordillera is such that its western ranges form a continuous orographic barrier, causing high precipitation along the Pacific coast while leaving much of the interior arid.
+However the ability of a GCM to reproduce these contrasts in precipitation is bounded by resolution.
+In a model of coarser resolution, these high mountain ranges are reduced to smoothed hills and the modelled distribution of precipitation is also smoother (Fig.~\ref{fig:oroprecip}).
+However this effect alone can not explain the widespread and pronounced positive precipitation anomalies observed in CFSR data (Fig.~\ref{fig:precheatmap}--\ref{fig:precdiff}) despite of the high model resolution (Table~\ref{tab:reanalyses}).
 
@@
 
-To compare modelled ice sheets of similar size, we selected for each climate forcing the simulation that leads to a final glaciated area closest to the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$\irina{justify}.
-These runs are presented in Figure~\ref{fig:best} along with their associated temperature depressions and a reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.
+Considering weaknesses of the CFSR and NCEP/NCAR data previously identified (Fig.~\ref{fig:tempheatmap}--\ref{fig:biatmbars}), we bound our analysis to the  WorldClim, ERA-Interim and NARR input climatologies.
+In order to compare modelled ice sheets of a similar size, we select for each climate forcing the simulation that leads to a final glaciated area closest to the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$\irina{justify}.
+This correspond to temperature offset values of 8 (WorldClim), 6 (ERA-Interim) and 7\,K (NARR).
+These qualitative “best” runs are presented in Figure~\ref{fig:best} along with their associated temperature depressions and a reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.
 
-Temperature depression necessary to reach the LGM extent differ significantly between input climatologies.
-Also, although similar in size, the selected ice sheets are different in shape.
-More specifically, ice sheets modelled using ERA-Interim, CFSR,  and NCEP/NCAR and  input climatologies cover parts of Yukon Territory and Alaska that were shown to have remained ice free \citep{dukrodkin-1999,kaufman-manley-2004}.
+Although this results in similar-sized ice sheets, noticeable differences in shape exist.
+The ERA-Interim simulation produced a northerly-centred ice sheet with too much ice in the North and too little in the South.
+This may partly reflect unresolved orographic precipitation effects (Fig.~\ref{fig:oroprecip}) due to coarser GCM resolution (Table~\ref{tab:reanalyses}).
 
-Although the outcome of our numerical simulations strongly depends on which dataset is used as a climate forcing, common discrepancies between the modelled ice sheet geometry and the ice margin mapped by \citet{dyke-2004} can be observed (Fig.~\ref{fig:best}).
+The LGM ice sheet margin by \citet{dyke-2004} is best reproduced by simulations driven by input climatologies from WorldClim and NARR data.
+However, common discrepancies between the modelled ice sheet geometry and the geomorphological reconstruction can still be observed (Fig.~\ref{fig:best}).
 
-For all climate forcing used, the modelled eastern margin of the ice sheet extends further east than the reconstructed boundary between the Cordilleran and the Laurentide ice sheets (Fig.~\ref{fig:best}).
-This can be explained by the potential effects of the growing ice-sheet on regional climate not included in our model, or by buttressing effects against the Laurentide Ice Sheet.
-The Cordilleran ice sheet initiated from the junction of mountain ice caps over the major reliefs~\citep{clague-1989}.
+Firstly, the modelled eastern margin of the ice sheet extends further east than where one would expect junction between the Cordilleran and the Laurentide ice sheets (Fig.~\ref{fig:best}).
+However the Laurentide Ice Sheet, which is not included in our model, may have formed a buttress against the smaller Cordilleran Ice Sheet and stopped its advance onto the Canadian Prairies.
+In addition, potential effects of the growing ice-sheet on regional climate are not included in our model.
+The Cordilleran ice sheet initiated from the junction of mountain ice caps over the major reliefs \citep{clague-1989}.
 In our simulations, a continuous ice cover quickly forms over the western ranges, where precipitation rates are higher than in the rest of the domain.
@@
 \citet{langen-etal-2012} demonstrated that this process, which they refer to as a ``self-inhibiting growth'', may have been limiting during the build-up of the Greenland Ice Sheet.
-Alternatively, the Laurentide Ice Sheet, which is not included in our model, may have formed a buttress against the smaller Cordilleran Ice Sheet and stopped its advance onto the Canadian Prairies.
 
-Our simulations produce anomalous\irina{Any evidence that there was no ice cover there?
+Secondly, our simulations produce anomalous\irina{Any evidence that there was no ice cover there?
 Refer to the source.} ice cover on parts of the continental shelf in the Arctic Ocean.
@@
 This anomalous ice cover is to be interpreted as an artefact arising from our simplistic temperature offset method, and has little effect on the rest of the results.
-
-Difference between modelled ice sheet geometries is notably visible in northern Yukon Territory and Alaska, where previous modelling studies commonly produced too extensive ice cover, whereas geomorphological data show that the ice cover was sparse.
-Concerning reanalysis datasets, we interpret these differences as mainly the result of different GCM resolutions and related model physics (Fig.~\ref{fig:topo}, Table~\ref{tab:reanalyses}).
-
-As described in section~\ref{sec:climate}, the topography of the Northern American Cordillera is such that its western ranges form a continuous orographic barrier, causing high precipitation along the Pacific coast while leaving much of the interior arid.
-However the ability of a GCM to reproduce these contrasts in precipitation is bounded by resolution.
-In a model of coarser resolution such as used in the NCEP/NCAR reanalysis, these high mountain ranges are reduced to smoothed hills and the modelled distribution of precipitation is also smoother (Fig.~\ref{fig:oroprecip}).
-This interpretation does not apply to the CFSR climatology, which exhibits high inland precipitation rates despite of the high model resolution (Table~\ref{tab:reanalyses}).
 

### 2013-11-05 floats/figures.tex
Resection the discussion and add new figures
New discussion first draft
Add new figure captions

@@
 	\begin{center}
-		\includegraphics[width=13cm]{cordillera-climate-topo}
+		\includegraphics{cordillera-climate-topo}
 	\end{center}
@@
 	\begin{center}
-		\includegraphics[width=13cm]{cordillera-climate-temp}
+		\includegraphics{cordillera-climate-temp}
 	\end{center}
@@
 	\begin{center}
-		\includegraphics[width=13cm]{cordillera-climate-prec}
+		\includegraphics{cordillera-climate-prec}
 	\end{center}
@@
 	\begin{center}
-		\includegraphics[width=9cm]{cordillera-climate-ivolarea}
+		\includegraphics{cordillera-climate-ivolarea}
 	\end{center}
@@
 	\begin{center}
-		\includegraphics[width=13cm]{cordillera-climate-extent}
+		\includegraphics{cordillera-climate-extent}
 	\end{center}
@@
 	\begin{center}
-		\includegraphics[width=13cm]{cordillera-climate-cool05}
+		\includegraphics{cordillera-climate-cool05}
 	\end{center}
@@
 	\label{fig:cool05}
+\end{figure*}
+
+% fig:tempheatmap
+\begin{figure}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics{cordillera-climate-tempheatmap}
+	\end{center}
+	\caption{Logarithmically coloured density maps, showing a comparison of JJA surface air temperature data from the WorldClim climatology, against that of each atmospheric reanalysis.
+Apparent horizontal lines are an artefact of different horizontal resolutions.
+Note the cold bias of NCEP/NCAR data.}
+	\label{fig:tempheatmap}
+\end{figure}
+
+% fig:tempdiff
+\begin{figure}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics{cordillera-climate-tempdiff}
+	\end{center}
+	\caption{JJA surface air temperature difference maps against WorldClim data, after bi-linear spatial interpolation.
+Note the cold bias of NCEP/NCAR data and temperature anomalies due to unresolved topographic detail.}
+	\label{fig:tempdiff}
+\end{figure}
+
+% fig:precheatmap
+\begin{figure}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics{cordillera-climate-precheatmap}
+	\end{center}
+	\caption{Logarithmically coloured density maps, showing a comparison of DJF precipitation rate from the WorldClim climatology, against that of each atmospheric reanalysis.
+Apparent horizontal lines are an artefact of different horizontal resolutions.
+Apparent vertical lines at low values are an artefact of WorldClim data precision.
+Note the wet bias of all reanalysis data.}
+	\label{fig:precheatmap}
+\end{figure}
+
+% fig:precdiff
+\begin{figure}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics{cordillera-climate-precdiff}
+	\end{center}
+	\caption{DJF precipitation rate relative difference maps against WorldClim data, after bi-linear spatial interpolation.
+Note the wet bias of all reanalysis data and the large anomalies of CFSR and NCEP/NCAR data.}
+	\label{fig:precdiff}
+\end{figure}
+
+% fig:oroprecip
+\begin{figure}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics{cordillera-climate-oroprecip}
+	\end{center}
+	\caption{Schematic representation of the orographic precipitation effect over the Coast Mountains.
+In a GCM of lower resolution, the precipitation peak appears smoother and the precipitation shadow is not as pronounced.
+Drawn using Inkscape \citep{web:inkscape}.}
+	\label{fig:oroprecip}
+\end{figure}
+
+% fig:biatm
+\begin{figure*}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics{cordillera-climate-biatm}
+	\end{center}
+	\caption{“Hybrid” atmosphere simulations using precipitation rate from WorldClim and surface air temperature from each reanalysis (upper row), and vice-versa (lower row).
+In other words, the upper row shows the effect of temperature anomalies, and the lower row the effect of precipitation anomalies, for each reanalysis, relative to WorldClim data.
+Each simulation use a 5\,\unit{\degree C} offset.
+Compare with figure~\ref{fig:cool05}}
+	\label{fig:biatm}
+\end{figure*}
+
+% fig:biatmbars
+\begin{figure*}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics{cordillera-climate-biatmbars}
+	\end{center}
+	\caption{Effect of temperature and precipitation anomalies of each reanalysis, and both, on modelled final ice volume, using the WorldClim 5\,\unit{\degree C} offset simulation as a reference.
+Corresponding ice sheet geometries are presented in figures~\ref{fig:cool05} and~\ref{fig:biatm}}
+	\label{fig:biatmbars}
 \end{figure*}
@@
 	\begin{center}
-		\includegraphics[width=13cm]{cordillera-climate-best}
+		\includegraphics{cordillera-climate-best}
 	\end{center}
@@
 
-% fig:oroprecip
+% fig:durationstack
 \begin{figure}[t]
@@
 	\begin{center}
-		\includegraphics[width=8cm]{cordillera-climate-oroprecip}
+		\includegraphics{cordillera-climate-durationstack}
 	\end{center}
-	\caption{Schematic representation of the orographic precipitation effect over the Coast Mountains.
-In a GCM of lower resolution, the precipitation peak appears smoother and the precipitation shadow is not as pronounced.
-Drawn using Inkscape \citep{web:inkscape}.}
-	\label{fig:oroprecip}
+	\caption{Model sensitivity to simulation length.
+Modelled glaciated area, using NARR forcing data, and temperature offsets from 0 to 15\,\unit{\degree C}, solid lines corresponding to values from 7 to 11\,\unit{\degree C} (left).
+Corresponding modelled ice margin, compared to the LGM ice sheet margin by \citep{dyke-2004} (right).}
+	\label{fig:durationstack}
 \end{figure}

### 2013-11-06 cordillera-climate.tex
Change section order
Small style corrections in introduction
General reread and polish
Get rid of personal shortcuts
Merge latex2rtf and pdflatex main files
Add a copernicus_rtf input file and replace to-do notes by regular footnotes. Using a copernicus_rtf package or class file did not seem to work with latex2rtf.
Implement corrections from Constantine
Implement corrections from Qiong
Add authors

@@
 % Copernicus manuscript
-\documentclass[tc, ms, noauthor]{copernicus}
+\documentclass[tc, ms]{copernicus}
 
 % Copernicus final print
-%\documentclass[tc, noauthor]{copernicus}
+%\documentclass[tc]{copernicus}
 
@@
 %\documentclass[tcd, hvmath]{copernicus_discussions}
+
+% Copernicus-like latex2rtf compatible
+%\input{copernicus_rtf}
 
@@
 % Recognize utf-8 special characters
+\usepackage[T1]{fontenc}
 \usepackage[utf8]{inputenc}
 
-% Add to-do notes and enlarge right margin
-\usepackage{todonotes}
-\usepackage[left=54pt, top=81pt, textwidth=398pt, marginparsep=11pt, marginparwidth=93pt, textheight=625pt, footskip=30pt]{geometry}
-\newcommand{\needref}[0]{\todo[color=cyan]{Reference needed}}
-\newcommand{\arjen}[2][]{\todo[color=lime,{#1}]{\textbf{Arjen:} #2}}
-\newcommand{\constantine}[2][]{\todo[color=yellow,{#1}]{\textbf{Constantine:} #2}}
-\newcommand{\qiong}[2][]{\todo[color=pink,{#1}]{\textbf{Qiong:} #2}}
-\newcommand{\julien}[2][]{\todo[color=lightgray,{#1}]{\textbf{Julien:} #2}}
-
-% Personal shortcuts
-\newcommand{\degC}[0]{\,\unit{\degree C} }
-\newcommand{\masl}[0]{\,\unit{m a.s.l.} }
+% To-do footnotes
+\newcommand{\needref}[0]{\footnote{Reference needed}}
+\newcommand{\julien}[1]{\footnote{\textbf{Julien:} #1}}
+\newcommand{\arjen}[1]{\footnote{\textbf{Arjen:} #1}}
+\newcommand{\qiong}[1]{\footnote{\textbf{Qiong:} #1}}
+\newcommand{\irina}[1]{\footnote{\textbf{Irina:} #1}}
 
@@
 
-% Title and authors
+% Title
 \title{The effect of climate forcing on numerical simulations of the Cordilleran Ice Sheet at the Last Glacial Maximum}
 \runningtitle{Climate forcing for Cordilleran ice sheet simulations}
-%\author{Julien Seguinot et al.}
-%\affil{Department of Physical Geography and Quaternary Geology, Stockholm University, S-106 91 Stockholm, Sweden}
+
+% Authors
+\author[1]{Julien Seguinot}
+\author[2]{Constantine Khroulev}
+\author[3]{Irina Rogozhina}
+\author[1]{Arjen Stroeven}
+\author[1]{Qiong Zhang}
 \runningauthor{J. Seguinot et al.}
 \correspondence{J. Seguinot\\ (julien.seguinot@natgeo.su.se)}
+
+% Affiliations
+\affil[1]{Department of Physical Geography and Quaternary Geology, Stockholm University, Stockholm, Sweden}
+\affil[2]{Geophysical Institute, University of Alaska Fairbanks, Fairbanks, AK, USA}
+\affil[3]{Helmholtz Centre Potsdam, GFZ German Research Centre for Geosciences, Potsdam, Germany}
 
@@
 \input{sections/intro}
+\input{sections/model}
 \input{sections/climate}
-\input{sections/model}
 \input{sections/results}
@@
 % Floats
+\newpage
 \input{floats/tables}
 \input{floats/figures}
+
+% ----------------------------------------------------------------------
 \end{document}
+% ----------------------------------------------------------------------

### 2013-11-07 sections/abstract.tex
New minor corrections from Irina

--- 2013-09-14 sections/abstract.tex
@@
 \begin{abstract}
-We present an ensemble of numerical simulations of the Cordilleran Ice Sheet during the Last Glacial Maximum performed with the Parallel Ice Sheet Model (PISM), using temperature offsets from present-day climatologies from five different datasets.
+We present an ensemble of numerical simulations of the Cordilleran Ice Sheet during the Last Glacial Maximum performed with the Parallel Ice Sheet Model (PISM), applying temperature offsets to the present-day climatologies from five different datasets.
 Monthly mean surface air temperature and precipitation from the NCEP/NCAR reanalysis, the Climate Forecast System Reanalysis, the ERA-Interim reanalysis and the North American Regional Reanalysis are used to compute surface mass balance by using a positive degree-day model.
@@
 We assess model performance against a reconstruction of the ice margin at the Last Glacial Maximum in continental regions of northern Yukon Territory and interior Alaska.
-The best match between model output and the reconstructed ice margin is obtained using climate forcing from the North American Regional Reanalysis, which is a regional climate reanalysis with the highest resolution.
+The best match between model output and the reconstructed ice margin is obtained using the North American Regional Reanalysis, which is a regional climate reanalysis with the highest resolution.\irina{WorldClim?}\irina{Maybe more on conclusions, a bit weak at the moment.}
 \end{abstract}

### 2013-11-07 sections/concl.tex
New minor corrections from Irina

@@
 The latter dataset is preferable in our case due to the lack of WorldClim data offshore.
-NCEP/NCAR forcing, on the other hand, produces an anomalously large ice sheet even under present-day conditions.
+NCEP/NCAR forcing, on the other hand, produces an anomalously large ice sheet even under present-day conditions.\irina{More explanations about the prec./temp.
+role here}\irina{What about the other datasets?}
 

### 2013-11-07 sections/discuss.tex
New minor corrections from Irina

@@
 We then compare our findings to a reconstruction of the LGM Cordilleran Ice Sheet margin, and discuss some of the key assumptions of this study.
-
 \subsection{Comparison of forcing climatologies}
@@
 
-Important differences in precipitation rate exist between input climatologies.
-All four reanalyses used in this study exhibit higher precipitation than WorldClim data (Fig.~\ref{fig:precheatmap}).
+All four reanalyses used in this study exhibit higher precipitation rates than WorldClim data (Fig.~\ref{fig:precheatmap}).
 However the magnitude of this anomaly differs significantly between datasets.
@@
 To distinguish the effects of temperature and precipitation biases in the ice sheet model, we run a series of eight additional simulations.
-These new simulations use an “hybrid” atmosphere forcing that consist of temperature data from WorldClim, combined with precipitation data from each of the four reanalyses, and vice-versa.
+These new simulations use a “hybrid” atmosphere forcing that consist of temperature data from WorldClim, combined with precipitation data from each of the four reanalyses, and vice-versa.
 A single temperature offset value of 5\,K is used.
@@
 
-We compare the numerically modelled ice sheet geometries to a reconstruction of the ice margin at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) by \citet{dyke-2004}, based on glacial geomorphology and radiocarbon dating.
-This corresponds to the LGM extent of most of the Cordilleran ice sheet, which occurred later than in the Laurentide ice sheet \citep{dyke-2004}.
+We compare numerically modelled ice sheet geometries to a reconstruction of the ice margin at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) by \citet{dyke-2004}, based on glacial geomorphology and radiocarbon dating.
+This corresponds to the LGM extent of most of the Cordilleran ice sheet, which occurred later than for the Laurentide ice sheet \citep{dyke-2004}.
 
@@
 In order to compare modelled ice sheets of a similar size, we select for each climate forcing the simulation that leads to a final glaciated area closest to the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$\irina{justify}.
-This correspond to temperature offset values of 8 (WorldClim), 6 (ERA-Interim) and 7\,K (NARR).
+This corresponds to temperature offset values of 8 (WorldClim), 6 (ERA-Interim) and 7\,K (NARR).
 These qualitative “best” runs are presented in Figure~\ref{fig:best} along with their associated temperature depressions and a reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.
 
-Although this results in similar-sized ice sheets, noticeable differences in shape exist.
-The ERA-Interim simulation produced a northerly-centred ice sheet with too much ice in the North and too little in the South.
+Although this selection result in similar-sized ice sheets, noticeable differences in shape exist.
+The ERA-Interim simulation produces a northerly-centred ice sheet with too much ice in the North and too little in the South.
 This may partly reflect unresolved orographic precipitation effects (Fig.~\ref{fig:oroprecip}) due to coarser GCM resolution (Table~\ref{tab:reanalyses}).
@@
 However in our simulations, sea-level is lowered by 120~m, turning large parts of the low-sloping continental shelf into land.
-A similar effect occurs over Great Bear Lake in some simulations, particularly those driven by the NCEP/NCAR forcing (Fig.~\ref{fig:extent}).
-This anomalous ice cover is to be interpreted as an artefact arising from our simplistic temperature offset method, and has little effect on the rest of the results.
+This anomalous ice cover is to be interpreted as an artefact arising from our simplistic temperature offset method, and has little effect on the rest of the results.\irina{You have to discuss the shortcoming of WorldClim to justify the choice of NARR in the next section (ocean).}
 
@@
 
-To test the effect of a shorter englaciation period on model results, we use simulations run under NARR climate forcing with temperatures offsets of 7 to 11\,K, and compare modelled ice sheet extent when glaciated area reach the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$.
+To test the effect of a shorter englaciation period on model results, we use simulations run under NARR climate forcing with temperatures offsets of 7 to 11\,K, and compare modelled ice sheet extent when glaciated area reaches the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$.
 
-This experiment shows that a shorter englaciation period leads to more restricted ice cover in the eastern part of the modelling domain, where precipitation rate control ice advance, but further glaciation along the Pacific margin to the south, where temperature is the limiting factor (Fig.~\ref{fig:durationstack}).
-Reducing simulation length lead to a closer match with the LGM ice sheet margin by \citet{dyke-2004} in some parts of the modelling domain, but higher discrepancy in other parts.
+This experiment shows that a shorter englaciation period leads to more restricted ice cover in the eastern part of the modelling domain, where precipitation rate controls ice advance, but further glaciation along the Pacific margin to the south, where temperature is the limiting factor (Fig.~\ref{fig:durationstack}).
+Reducing simulation length leads to a closer match with the LGM ice sheet margin by \citet{dyke-2004} in some parts of the modelling domain, but higher discrepancy in other parts.
 
@@
 
-Our simplistic temperature offset method is certainly a crude simplification of past climate changes over the region.
+Our simplistic temperature offset method is certainly a crude simplification\irina{repetitive} of past climate changes over the region.
 Although it is clear that these temperature changes were neither homogeneous nor constant in time, and probably associated with precipitation changes, their patterns are not trivial and may potentially display a strong inter-dependence with the evolution of the Cordilleran and Laurentide ice sheets.

### 2013-11-07 sections/results.tex
New minor corrections from Irina

@@
 Using WorldClim, NCEP/NCAR, ERA-Interim, CFSR (smoothed and original datasets) and NARR input climatologies as climate forcing, we run 66~simulations of glacial inception and growth of the Cordilleran ice sheet over 10~kyr using temperature offsets ranging from 0 to 10\,\unit{\degree C}.
-Significant differences in patterns of ice sheet growth, final glacial extent, and volume were attained (Fig.
+Significant differences in patterns of ice sheet growth, final glacial extent, and volume are attained (Fig.
 \ref{fig:ivolarea}-\ref{fig:cool05}).
@@
 CFSR forcing, on the other hand, yields continuous ice cover over the northern Coast Mountains, and NCEP/NCAR forcing results in ice sheet growth, both of which are largely inconsistent with the present state of glaciation.
-At the opposite side of the temperature offset range, it can be noted that both the NCEP/NCAR and CFSR forcing produce oversized ice sheets whose extent are primarily restricted by the domain boundaries (Fig.~\ref{fig:extent}).
+At the opposite side of the temperature offset range, it can be noted that both the NCEP/NCAR and CFSR forcing produce oversized ice sheets whose extents are primarily restricted by the domain boundaries (Fig.~\ref{fig:extent}).
 The smoothing of CFSR precipitation yields an insignificant effect on the resulting ice sheet geometry.

### 2013-11-07 sections/discuss.tex
Arjen corrections and reread in the discussion

@@
 
-The outcome of our numerical simulations shows a very large sensitivity to the choice of climate forcing data (Fig.~\ref{fig:ivolarea}--\ref{fig:cool05}).
-To understand the origin of these discrepancies, we quantify surface air temperature and precipitation rate differences between input climatologies and test their effect on modelled ice sheet geometries.
-We then compare our findings to a reconstruction of the LGM Cordilleran Ice Sheet margin, and discuss some of the key assumptions of this study.
+These simulations of the LGM Cordilleran Ice Sheet show a very large sensitivity to the choice of climate forcing data used (Fig.~\ref{fig:ivolarea}--\ref{fig:cool05}).
+To understand the origin of the discrepancies between model results, we quantify differences in surface air temperature and precipitation rate between the input climatologies and test their effect on the ice sheet model.
+We then compare our findings to a geomorphological reconstruction of the LGM Cordilleran Ice Sheet margin, and discuss some of the key assumptions of this study.
+
 \subsection{Comparison of forcing climatologies}
 
-The PDD model used in this study is highly sensitive to surface air temperature and precipitation forcing.
-As previously outlined, winter (DJF) precipitation and summer (JJA) temperature are the predominant controls on accumulation and melt at the modelled ice sheet surface.
-Significant differences in their distribution exist between datasets (Fig.~\ref{fig:topo}--\ref{fig:prec}).
-
+As previously outlined, significant differences in the distribution of winter (DJF) precipitation and summer (JJA) temperature, which are the predominant controls on accumulation and melt at the modelled ice sheet surface, exist between datasets (Fig.~\ref{fig:topo}--\ref{fig:prec}).
 In this section we use the WorldClim climatology, which is derived from observations, as a reference with which to compare other climate forcing data on land.
 ERA-Interim, CFSR and NARR climatologies exhibit temperature distributions fairly consistent with WorldClim data (Fig.~\ref{fig:tempheatmap}).
-When examining the spatial distribution of temperature differences to WorldClim data, it appears that most of disparity between reanalyses and observation data is caused by topographical detail at scales unresolved by the atmospheric reanalyses (Fig.~\ref{fig:tempdiff}).
+When examining the spatial distribution of temperature differences between reanalyses and WorldClim data, it appears that most of the disparity between reanalyses and observation data is caused by topographical detail at scales unresolved by the atmospheric reanalyses (Fig.~\ref{fig:tempdiff}).
 Surface air temperature data from the NCAR climatology shows a large disparity with that of WorldClim, with a significant cold anomaly over most of the modelling domain (Fig.~\ref{fig:tempheatmap}--\ref{fig:tempdiff}).
@@
 As described in section~\ref{sec:climate}, the topography of the Northern American Cordillera is such that its western ranges form a continuous orographic barrier, causing high precipitation along the Pacific coast while leaving much of the interior arid.
-However the ability of a GCM to reproduce these contrasts in precipitation is bounded by resolution.
-In a model of coarser resolution, these high mountain ranges are reduced to smoothed hills and the modelled distribution of precipitation is also smoother (Fig.~\ref{fig:oroprecip}).
+However the ability of a GCM to reproduce these contrasts in precipitation is bounded by horizontal resolution.
+In a model of coarser resolution, these high mountain ranges are reduced to smooth hills and the modelled distribution of precipitation is also smoother (Fig.~\ref{fig:oroprecip}).
 However this effect alone can not explain the widespread and pronounced positive precipitation anomalies observed in CFSR data (Fig.~\ref{fig:precheatmap}--\ref{fig:precdiff}) despite of the high model resolution (Table~\ref{tab:reanalyses}).
@@
 
-To distinguish the effects of temperature and precipitation biases in the ice sheet model, we run a series of eight additional simulations.
+To distinguish the effects of temperature and precipitation biases on the ice sheet model, we run a series of eight additional simulations.
 These new simulations use a “hybrid” atmosphere forcing that consist of temperature data from WorldClim, combined with precipitation data from each of the four reanalyses, and vice-versa.
-A single temperature offset value of 5\,K is used.
+A single temperature offset of 5\,K is used.
 This value was chosen to allow comparison with previous results (Fig.~\ref{fig:cool05}).
@@
 
-\subsection{Comparison to reconstructed LGM margin}
+\subsection{Comparison to geomorphological LGM margin}
 
@@
 
-Considering weaknesses of the CFSR and NCEP/NCAR data previously identified (Fig.~\ref{fig:tempheatmap}--\ref{fig:biatmbars}), we bound our analysis to the  WorldClim, ERA-Interim and NARR input climatologies.
+Considering previously identified weaknesses of the CFSR and NCEP/NCAR data (Fig.~\ref{fig:tempheatmap}--\ref{fig:biatmbars}), we bound our analysis to the  WorldClim, ERA-Interim and NARR input climatologies.
 In order to compare modelled ice sheets of a similar size, we select for each climate forcing the simulation that leads to a final glaciated area closest to the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$\irina{justify}.
@@
 
-Although this selection result in similar-sized ice sheets, noticeable differences in shape exist.
-The ERA-Interim simulation produces a northerly-centred ice sheet with too much ice in the North and too little in the South.
+Although this selection result in similarly sized ice sheets, noticeable differences in shape exist.
+The ERA-Interim simulation produces a ore northerly-centred ice sheet with too much ice in the North and too little in the South.
 This may partly reflect unresolved orographic precipitation effects (Fig.~\ref{fig:oroprecip}) due to coarser GCM resolution (Table~\ref{tab:reanalyses}).
@@
 
-Firstly, the modelled eastern margin of the ice sheet extends further east than where one would expect junction between the Cordilleran and the Laurentide ice sheets (Fig.~\ref{fig:best}).
-However the Laurentide Ice Sheet, which is not included in our model, may have formed a buttress against the smaller Cordilleran Ice Sheet and stopped its advance onto the Canadian Prairies.
+Firstly, the modelled eastern margin of the ice sheet extends further east than where one would expect junction between the Cordilleran and the Laurentide ice sheets.
+This can be seen for instance in the south-eastern part of our modelling domain, where the modelled ice sheets cover an ice-free corridor of the LGM reconstruction (Fig.~\ref{fig:best}).
+However the Laurentide Ice Sheet, absent from our simulations, likely formed a buttress against the smaller Cordilleran Ice Sheet and stopped its advance onto the Canadian Prairies.
 In addition, potential effects of the growing ice-sheet on regional climate are not included in our model.
@@
 This continuous ice cover may have enhanced the topographical barrier already formed by the western ranges, resulting in less precipitation and warmer air in the interior.
-\citet{langen-etal-2012} demonstrated that this process, which they refer to as a ``self-inhibiting growth'', may have been limiting during the build-up of the Greenland Ice Sheet.
+This process of ``self-inhibiting growth'' was demonstrated in a similar setting over Greenland, where it may have been limiting the advance of the ice sheet during its build-up \citep{langen-etal-2012}.
 
@@
 However in our simulations, sea-level is lowered by 120~m, turning large parts of the low-sloping continental shelf into land.
-This anomalous ice cover is to be interpreted as an artefact arising from our simplistic temperature offset method, and has little effect on the rest of the results.\irina{You have to discuss the shortcoming of WorldClim to justify the choice of NARR in the next section (ocean).}
+This anomalous ice cover is to be interpreted as an artefact arising from our simplistic temperature offset method, and has little effect on the rest of the results.
+
+Considering the lack of WorldClim data offshore, we consider the NARR climate forcing as best suited for simulations of the Cordilleran Ice Sheet among those tested, and under the assumptions of the present study.
 
@@
 
-Therefore, our choice of 10\,Ka simulation length could certainly be discussed further.
+Our choice of 10\,Ka simulation length could certainly be discussed further.
 However, our representation of climate history by constant temperature offsets is voluntarily simplistic.
-To better understand the transient character of the LGM Cordilleran Ice Sheet, time-dependent palaeo-climate forcing would be needed, and this is left for future research.
+To better understand the transient character of the LGM Cordilleran Ice Sheet, time-dependent palaeo-climate forcing would be needed.
+This is left for future research.
 

### 2013-11-07 floats/figures.tex
Arjen corrections in figure captions

@@
 	\caption{Shaded relief map of northern North America with the extent of the Cordilleran (CIS), Laurentide (LIS) and Innuitian (IIS) ice sheets at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) \citep{dyke-2004}.
-The background map consists of ETOPO1 \citep{data:etopo1} and Natural Earth Data \citep{data:naturalearth}.
-Assembled with GRASS~GIS \citep{soft:grass} and Inkscape \citep{web:inkscape}}
+While this age denotes the LIS after retreat from its LGM, it closely corresponds to the LGM extent of most of the Cordilleran ice sheet \citep{porter-swanson-1998,dyke-2004,stroeven-etal-2010}.
+The rectangular box denotes the modelling domain of the Cordilleran ice sheet of 1500 by 3000\,km.
+The background map consists of ETOPO1 \citep{data:etopo1} and Natural Earth Data \citep{data:naturalearth} and was assembled with GRASS~GIS \citep{soft:grass}.}
 	\label{fig:locmap}
@@
 	\end{center}
-	\caption{Topography maps from WorldClim, ERA-Interim reanalysis, NARR (North American Regional Reanalysis), CFSR (Climate System Forecast Reanalysis), and NCEP/NCAR reanalysis climatologies used as a reference for temperature lapse-rate corrections, and topography map from ETOPO1 used as basal topography.
+	\caption{Topography maps from WorldClim \citep{data:worldclim}, ERA-Interim reanalysis \citep{data:erai}, North American Regional Reanalysis (NARR) \citep{data:narr}, ETOPO1 \citep{data:etopo1}, Climate System Forecast Reanalysis (CFSR) \citep{data:cfsr}, and NCEP/NCAR reanalysis \citep{data:ncar}.
+Whereas ETOPO1 is used as basal topography for the ice sheet model, all others are used data used as a reference for temperature lapse-rate corrections.
 Fig.~\ref{fig:topo}-\ref{fig:best} drawn using Matplotlib \citep{soft:mpl}.}
@@
 	\end{center}
-	\caption{Summer (JJA) and winter (DJF) temperature maps from WorldClim, and summer (JJA) temperature maps from ERA-Interim reanalysis, NARR (North American Regional Reanalysis), CFSR (Climate System Forecast Reanalysis), and NCEP/NCAR reanalysis climatologies.}
+	\caption{Summer (JJA) and winter (DJF) temperature maps from WorldClim \citep{data:worldclim}, and summer (JJA) temperature maps from ERA-Interim reanalysis \citep{data:erai}, North American Regional Reanalysis (NARR) \citep{data:narr}, Climate System Forecast Reanalysis (CFSR) \citep{data:cfsr}, and NCEP/NCAR reanalysis \citep{data:ncar} climatologies.}
 	\label{fig:temp}
@@
 	\end{center}
-	\caption{Winter (DJF) and summer (JJA) precipitation maps from WorldClim, and winter (DFJ) temperature maps from ERA-Interim reanalysis, NARR (North American Regional Reanalysis), CFSR (Climate System Forecast Reanalysis), and NCEP/NCAR reanalysis climatologies.
+	\caption{Winter (DJF) and summer (JJA) precipitation maps from WorldClim \citep{data:worldclim}, and winter (DFJ) temperature maps from ERA-Interim reanalysis \citep{data:erai}, North American Regional Reanalysis (NARR) \citep{data:narr}, Climate System Forecast Reanalysis (CFSR) \citep{data:cfsr}, and NCEP/NCAR reanalysis \citep{data:ncar} climatologies.
 Additional forcing data was prepared to correct for wave-like precipitation artefacts in CFSR (section~\ref{sec:climate}).}
@@
 	\end{center}
-	\caption{Total glaciated area and ice volume after 10\,kyr as a function of temperature offset and climate forcing.}
+	\caption{Total glaciated area and ice volume after 10\,kyr as a function of temperature offset and climate forcing used.}
 	\label{fig:ivolarea}
@@
 	\end{center}
-	\caption{Extent of ice cover after 10\,kyr as a function of applied temperature offsets for each climate forcing.}
+	\caption{Extent of ice cover after 10\,kyr as a function of applied temperature offsets for each climate forcing used.}
 	\label{fig:extent}
@@
 	\end{center}
-	\caption{Ice surface topography (black contours every 1000\,m) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a climate 5\,\unit{\degree C} colder than present for each climate forcing.}
+	\caption{Ice surface topography (1 km contours) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a climate 5\,\unit{\degree C} colder than present for each climate forcing used.}
 	\label{fig:cool05}
@@
 	\end{center}
-	\caption{Schematic representation of the orographic precipitation effect over the Coast Mountains.
-In a GCM of lower resolution, the precipitation peak appears smoother and the precipitation shadow is not as pronounced.
-Drawn using Inkscape \citep{web:inkscape}.}
+	\caption{Schematic representation of the orographic precipitation effect over a mountain range.
+In a GCM of low resolution, the precipitation peak appears downwind-shifted and smoother and the precipitation shadow is less pronounced than in a high resolution GCM.}
 	\label{fig:oroprecip}
@@
 	\caption{“Hybrid” atmosphere simulations using precipitation rate from WorldClim and surface air temperature from each reanalysis (upper row), and vice-versa (lower row).
-In other words, the upper row shows the effect of temperature anomalies, and the lower row the effect of precipitation anomalies, for each reanalysis, relative to WorldClim data.
+(1 km contours)In other words, the upper row shows the effect of temperature anomalies, and the lower row the effect of precipitation anomalies, for each reanalysis, relative to WorldClim data.
 Each simulation use a 5\,\unit{\degree C} offset.
@@
 % fig:biatmbars
-\begin{figure*}[t]
+\begin{figure}[t]
 	\vspace*{2mm}
@@
 	\label{fig:biatmbars}
-\end{figure*}
+\end{figure}
 
@@
 	\end{center}
-	\caption{Ice surface topography (black contours every 1000\,m) after 10\,kyr using temperature offsets that lead to similar areas of ice cover for each climate forcing.
-14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) ice margin (blue line) from \citet{dyke-2004}.}
+	\caption{Ice surface topography (1\,km contours) after 10\,kyr using temperature offsets resulting in glaciated areas of circa $2\,\times10^6\,\unit{km^2}$.
+A reconstructed LGM ice sheet margin by \citet{dyke-2004} (Fig.~\ref{fig:locmap}) is given for reference (blue line).}
 	\label{fig:best}

### 2013-11-07 sections/discuss.tex
Arjen corrections in figure captions

@@
 We compare numerically modelled ice sheet geometries to a reconstruction of the ice margin at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) by \citet{dyke-2004}, based on glacial geomorphology and radiocarbon dating.
-This corresponds to the LGM extent of most of the Cordilleran ice sheet, which occurred later than for the Laurentide ice sheet \citep{dyke-2004}.
+This corresponds to the LGM extent of most of the Cordilleran ice sheet, which occurred later than for the Laurentide ice sheet \citep{porter-swanson-1998,dyke-2004,stroeven-etal-2010}.
 

### 2013-11-07 sections/abstract.tex
Rework the abstract

@@
 We present an ensemble of numerical simulations of the Cordilleran Ice Sheet during the Last Glacial Maximum performed with the Parallel Ice Sheet Model (PISM), applying temperature offsets to the present-day climatologies from five different datasets.
-Monthly mean surface air temperature and precipitation from the NCEP/NCAR reanalysis, the Climate Forecast System Reanalysis, the ERA-Interim reanalysis and the North American Regional Reanalysis are used to compute surface mass balance by using a positive degree-day model.
+Monthly mean surface air temperature and precipitation from WorldClim, the NCEP/NCAR reanalysis, the ERA-Interim reanalysis, the Climate Forecast System Reanalysis and the North American Regional Reanalysis are used to compute surface mass balance in a positive degree-day model.
 Modelled ice sheet outlines and volumes appear highly sensitive to the choice of climate forcing.
-We assess model performance against a reconstruction of the ice margin at the Last Glacial Maximum in continental regions of northern Yukon Territory and interior Alaska.
-The best match between model output and the reconstructed ice margin is obtained using the North American Regional Reanalysis, which is a regional climate reanalysis with the highest resolution.\irina{WorldClim?}\irina{Maybe more on conclusions, a bit weak at the moment.}
+For three of the four reanalysis dataset used, differences in precipitation is the major source for discrepancies between model results.
+We assess model performance against a geomorphological reconstruction of the ice margin at the Last Glacial Maximum, and suggest that part of the mismatch is due to unresolved orographic precipitation effects caused by the coarse resolution of reanalysis data.
+The best match between model output and the reconstructed ice margin is obtained using the high-resolution North American Regional Reanalysis, which we retain for future simulations of the Cordilleran Ice Sheet.
 \end{abstract}

### 2013-11-07 sections/concl.tex
Rework the conclusion

--- 2013-09-19 sections/concl.tex
@@
 
-Our study shows a strong dependency of the modelled ice sheet on the choice of climate forcing data.
-When using input climatologies from climate reanalyses, spatial resolution of the GCM appears critical for providing the ice sheet model with an accurate precipitation field.
-This confirms results obtained by \citet{quiquet-etal-2012} from various GCM forcing over the Greenland Ice Sheet.
+Our study shows a strong dependency of ice sheet model results on the choice of climate forcing data.
+For three of the four reanalysis datasets used, precipitation rate, over surface air temperature, causes much of the discrepancy between modelled ice sheet outlines and volumes.
+Furthermore, the spatial resolution of input data appears critical for providing the ice sheet model with an accurate precipitation field, confirming results obtained by \citet{quiquet-etal-2012} from various GCM forcing over the Greenland Ice Sheet.
 
@@
 The latter dataset is preferable in our case due to the lack of WorldClim data offshore.
-NCEP/NCAR forcing, on the other hand, produces an anomalously large ice sheet even under present-day conditions.
+Climate forcing from CFSR and the NCEP/NCAR reanalysis produces largely oversized ice sheets due to too high precipitation rates, and in the second case, too low surface air temperature.
+The ERA-Interim data used in this study produces more reasonable results, but it may be too coarse to accurately resolve the spatial distribution of orographic precipitation associated with the rugged topography of the northern American Cordillera, resulting in a misplaced ice cover in regard to geomorphological data.
+Therefore, we retain NARR data for forcing future simulations of the Cordilleran Ice Sheet.
 
@@
 At the cost of greater computational expense, a more accurate representation of palaeo-climate may allow for a better fit between model results and the mapped LGM margin.
-\irina{I have this feeling that the manuscript contains a number of repetitive ideas and arguments. I will try to track them throughout the text.}

### 2013-11-07 floats/figures.tex
Check unit consistency (K -> degC, yr -> a)

@@
 Whereas ETOPO1 is used as basal topography for the ice sheet model, all others are used data used as a reference for temperature lapse-rate corrections.
-Fig.~\ref{fig:topo}-\ref{fig:best} drawn using Matplotlib \citep{soft:mpl}.}
+Fig.~\ref{fig:topo}-\ref{fig:durationstack} drawn using Matplotlib \citep{soft:mpl}.}
 	\label{fig:topo}
@@
 	\end{center}
-	\caption{Total glaciated area and ice volume after 10\,kyr as a function of temperature offset and climate forcing used.}
+	\caption{Total glaciated area and ice volume after 10\,ka as a function of temperature offset and climate forcing used.}
 	\label{fig:ivolarea}
@@
 	\end{center}
-	\caption{Extent of ice cover after 10\,kyr as a function of applied temperature offsets for each climate forcing used.}
+	\caption{Extent of ice cover after 10\,ka as a function of applied temperature offsets for each climate forcing used.}
 	\label{fig:extent}
@@
 	\end{center}
-	\caption{Ice surface topography (1 km contours) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a climate 5\,\unit{\degree C} colder than present for each climate forcing used.}
+	\caption{Ice surface topography (1 km contours) and velocity (\unit{m\,a^{-1}}) after 10\,ka under a climate 5\,\unit{\degree C} colder than present for each climate forcing used.}
 	\label{fig:cool05}
@@
 	\end{center}
-	\caption{Ice surface topography (1\,km contours) after 10\,kyr using temperature offsets resulting in glaciated areas of circa $2\,\times10^6\,\unit{km^2}$.
+	\caption{Ice surface topography (1\,km contours) after 10\,ka using temperature offsets resulting in glaciated areas of circa $2\,\times10^6\,\unit{km^2}$.
 A reconstructed LGM ice sheet margin by \citet{dyke-2004} (Fig.~\ref{fig:locmap}) is given for reference (blue line).}

### 2013-11-07 sections/discuss.tex
Check unit consistency (K -> degC, yr -> a)

@@
 These new simulations use a “hybrid” atmosphere forcing that consist of temperature data from WorldClim, combined with precipitation data from each of the four reanalyses, and vice-versa.
-A single temperature offset of 5\,K is used.
+A single temperature offset of 5\,\unit{\degree C} is used.
 This value was chosen to allow comparison with previous results (Fig.~\ref{fig:cool05}).
@@
 
-Although only temperature or precipitation changes are applied, this experiment results once more in large differences in modelled ice sheet geometries (Fig.~\ref{fig:biatm}) as compared to the reference WorldClim 5\,K run (Fig.~\ref{fig:cool05}, top-left panel).
+Although only temperature or precipitation changes are applied, this experiment results once more in large differences in modelled ice sheet geometries (Fig.~\ref{fig:biatm}) as compared to the reference WorldClim 5\,\unit{\degree C} run (Fig.~\ref{fig:cool05}, top-left panel).
 For ERA-Interim, CFSR and NARR climatologies, precipitation anomalies clearly dominate the differences in ice sheet response, whereas temperature anomalies have relatively little effect (Fig.~\ref{fig:biatmbars}).
@@
 In order to compare modelled ice sheets of a similar size, we select for each climate forcing the simulation that leads to a final glaciated area closest to the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$\irina{justify}.
-This corresponds to temperature offset values of 8 (WorldClim), 6 (ERA-Interim) and 7\,K (NARR).
+This corresponds to temperature offset values of 8 (WorldClim), 6 (ERA-Interim) and 7\,\unit{\degree C} (NARR).
 These qualitative “best” runs are presented in Figure~\ref{fig:best} along with their associated temperature depressions and a reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.
@@
 
-A key assumption for the present study is our choice of 10\,Ka for the length of all our simulations.
-Although geomorphological data shows that the period of growth of the Cordilleran Ice Sheet from nearly ice-free to full glacial conditions could not have been much longer than 10\,Ka, they provide no lower bound to this value.
+A key assumption for the present study is our choice of 10\,ka for the length of all our simulations.
+Although geomorphological data shows that the period of growth of the Cordilleran Ice Sheet from nearly ice-free to full glacial conditions could not have been much longer than 10\,ka, they provide no lower bound to this value.
 
-To test the effect of a shorter englaciation period on model results, we use simulations run under NARR climate forcing with temperatures offsets of 7 to 11\,K, and compare modelled ice sheet extent when glaciated area reaches the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$.
+To test the effect of a shorter englaciation period on model results, we use simulations run under NARR climate forcing with temperatures offsets of 7 to 11\,\unit{\degree C}, and compare modelled ice sheet extent when glaciated area reaches the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$.
 
@@
 
-Our choice of 10\,Ka simulation length could certainly be discussed further.
+Our choice of 10\,ka simulation length could certainly be discussed further.
 However, our representation of climate history by constant temperature offsets is voluntarily simplistic.

### 2013-11-07 sections/model.tex
Check unit consistency (K -> degC, yr -> a)

@@
 As we aim to model glacial inception and growth of the Cordilleran Ice Sheet towards a configuration as last attained during the LGM, we mimic palaeo-climatic conditions by applying constant temperature offsets homogeneously over the modelling domain.
-Each simulation starts from ice-free conditions and runs for 10\,kyr, a time interval representative of the rapid build-up of the last Cordilleran Ice Sheet from nearly ice-free to full glacial conditions \citep{clague-1989,stroeven-etal-2010}.
+Each simulation starts from ice-free conditions and runs for 10\,ka, a time interval representative of the rapid build-up of the last Cordilleran Ice Sheet from nearly ice-free to full glacial conditions \citep{clague-1989,stroeven-etal-2010}.
 Our simulations are performed in parallel on 32 cores at the Swedish National Supercomputing Center.

### 2013-11-07 sections/results.tex
Check unit consistency (K -> degC, yr -> a)

@@
 
-Using WorldClim, NCEP/NCAR, ERA-Interim, CFSR (smoothed and original datasets) and NARR input climatologies as climate forcing, we run 66~simulations of glacial inception and growth of the Cordilleran ice sheet over 10~kyr using temperature offsets ranging from 0 to 10\,\unit{\degree C}.
+Using WorldClim, NCEP/NCAR, ERA-Interim, CFSR (smoothed and original datasets) and NARR input climatologies as climate forcing, we run 66~simulations of glacial inception and growth of the Cordilleran ice sheet over 10~ka using temperature offsets ranging from 0 to 10\,\unit{\degree C}.
 Significant differences in patterns of ice sheet growth, final glacial extent, and volume are attained (Fig.

### 2013-11-07 figures/plot.py

@@ -208,7 +208,7 @@ def prec():
 
     # add colorbar and save
     cb = fig.colorbar(im, fig.grid.cbar_axes[0])
-    cb.set_label(u'precipitation rate (m/yr)')
+    cb.set_label(u'precipitation rate (m/a)')
     _savefig('cordillera-climate-prec')
 
 def precdiff():
@@ -290,7 +290,7 @@ def precheatmap():
       print (data-refdata).mean()
 
     # save
-    fig.suptitle('DJF mean precipitation rate (m/yr)')
+    fig.suptitle('DJF mean precipitation rate (m/a)')
     _savefig('cordillera-climate-precheatmap', pdf=True)
 
 def topo():
@@ -551,7 +551,7 @@ def durationstack():
     ax.set_xlim((0, 10))
     ax.set_ylim((0, 4))
     ax.set_yticks(range(5))
-    ax.set_xlabel('Model time (Ka)')
+    ax.set_xlabel('Model time (ka)')
     ax.set_ylabel(u'Glaciated area ($\mathsf{10^6~km^2}$)')
 
     # plot maps

### 2013-11-08 sections/climate.tex
General footnotes clear-up

@@
 
-The spatial distribution of JJA and DJF mean precipitation rates in WorldClim reveal a strong precipitation decline beyond the coastal regions (Fig.~\ref{fig:prec})\qiong{Change unit to mm per month.}, primarily as a result of the continuous orographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascades.
+The spatial distribution of JJA and DJF mean precipitation rates in WorldClim reveal a strong precipitation decline beyond the coastal regions (Fig.~\ref{fig:prec}), primarily as a result of the continuous orographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascades.
 A less pronounced precipitation peak is also visible over the interior ranges.

### 2013-11-08 sections/discuss.tex
General footnotes clear-up

@@
 Considering previously identified weaknesses of the CFSR and NCEP/NCAR data (Fig.~\ref{fig:tempheatmap}--\ref{fig:biatmbars}), we bound our analysis to the  WorldClim, ERA-Interim and NARR input climatologies.
-In order to compare modelled ice sheets of a similar size, we select for each climate forcing the simulation that leads to a final glaciated area closest to the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$\irina{justify}.
+In order to compare modelled ice sheets of a similar size, we select for each climate forcing the simulation that leads to a final glaciated area closest to the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$.
 This corresponds to temperature offset values of 8 (WorldClim), 6 (ERA-Interim) and 7\,\unit{\degree C} (NARR).
@@
 
-Secondly, our simulations produce anomalous\irina{Any evidence that there was no ice cover there?
-Refer to the source.} ice cover on parts of the continental shelf in the Arctic Ocean.
+Secondly, our simulations produce anomalous ice cover on parts of the continental shelf in the Arctic Ocean (Fig.~\ref{fig:best}).
 These regions experience a marine climate, including lower summer temperature than on the adjacent land (Fig.~\ref{fig:temp}).
@@
 
+\subsection{Other potential method improvements}
+
 Our choice of 10\,ka simulation length could certainly be discussed further.
@@
 To better understand the transient character of the LGM Cordilleran Ice Sheet, time-dependent palaeo-climate forcing would be needed.
-This is left for future research.
-
-\subsection{Other potential method improvements}
-
-Our simplistic temperature offset method is certainly a crude simplification\irina{repetitive} of past climate changes over the region.
-Although it is clear that these temperature changes were neither homogeneous nor constant in time, and probably associated with precipitation changes, their patterns are not trivial and may potentially display a strong inter-dependence with the evolution of the Cordilleran and Laurentide ice sheets.
+Although it is clear that past temperature changes over the region were neither homogeneous nor constant in time, and probably associated with precipitation changes, their patterns are not trivial and may potentially display a strong inter-dependence with the evolution of the Cordilleran and Laurentide ice sheets.
 A more correct, yet more complex approach would be to use coupling to a GCM of intermediate complexity \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013}.

### 2013-11-08 sections/intro.tex
General footnotes clear-up

@@
 An important drawback in this approach is that GCM palaeo-climate simulations themselves rely on an accurate description of the global surface topography and therefore include global ice sheet reconstructions such as ICE-4G \citep{peltier-1994} for their surface topographic boundary condition.
-This condition potentially exerts influence on subsequently modelled ice sheet geometries, and particularly glacier extent.
-\arjen{Has this been shown? Add ref.}
+This condition could potentially exert influence on subsequently modelled ice sheet geometries, and particularly glacier extent.
 
@@
 \citet{quiquet-etal-2012}, for example, assessed the sensitivity of a Greenland ice sheet model to various atmospheric forcing, including a regional parametrization by \citet{fausto-etal-2009}, output from several GCMs and an atmospheric reanalysis.
-\citet{rodgers-etal-2004} and \citet{charbit-etal-2007} tested the sensitivity of a model of the Northern Hemisphere ice sheet model to climate forcing from different PMIP LGM simulations\qiong{Detail in a few lines the conclusions of theses studies}.
+\citet{rodgers-etal-2004} and \citet{charbit-etal-2007} tested the sensitivity of a model of the Northern Hemisphere ice sheet model to climate forcing from different PMIP LGM simulations.
+Although using different set-ups, all three studies demonstrate the very large sensitivity of ice sheet model to the choice of climate forcing data used.
 In our study, to limit the degrees of freedom in our model and obtain results independent of palaeo-ice sheet reconstructions such as ICE-4G, we use a simple temperature offset similar to the approaches by \citet{greve-etal-1999} and \citet{bintanja-etal-2005} and assess ice sheet model sensitivity to the choice of present-day climate data.
@@
 
-The Cordilleran Ice Sheet (Fig.~\ref{fig:locmap})\irina{include all geographic names mentioned in the text}\arjen{elevation scale white-end can be confused with glaciers} covered an area that presently experiences strong regional variations in climate.
+The Cordilleran Ice Sheet (Fig.~\ref{fig:locmap}) covered an area that presently experiences strong regional variations in climate.
 From a numerical modelling perspective, it is one of the least studied palaeo-ice sheets of the Northern Hemisphere, despite the fact that significant stratigraphical and geomorphological data is available to constrain its extent \citep{jackson-clague-1991,dukrodkin-1999,kaufman-manley-2004,kleman-etal-2010,margold-etal-2011}.
@@
 To our knowledge, this is the first modelling study that specifically focuses on the Cordilleran Ice Sheet since \citet{robert-1991}.
-
-%While the modelling domain of \citet{robert-1991} covered only the southern part of the ice sheet, ours encompasses most of the LGM ice extent and adjacent non-glaciated parts of northern Yukon Territory and interior Alaska (Fig.~\ref{fig:locmap}).
-%Our model also includes higher resolution and an improved treatment of ice thermodynamics and bedrock response following the large advances made in scientific computing and ice sheet modelling since then.

### 2013-11-08 sections/model.tex
General footnotes clear-up

@@
 The till friction angle $\phi$ varies from 10 to 30\degree\ and is a function of current bed elevation, with lowest values occurring at low elevations,
-\julien{phi is independent from basal water which is computed by the model (and sliding depends on both) so the 10 to 30 degree range is meant to reflect different bedrock/sediment surfaces rather than dry vs wet beds.}
 
@@
 This models weakening of the till associated with the presence of marine sediments \citep{martin-etal-2011,aschwanden-etal-2013}.
-Basal topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography and bathymetry dataset with a resolution of 1\,arc-minute \citep{data:etopo1}.\julien{NaturalEarth data are vectors (coastline, rivers, etc.) used in Fig.
-1, not model input.} Sea level is lowered by 120\,m and basal topography responds to ice load following a regional isostasy model that includes lithospheric flexure and mantle relaxation \citep{lingle-clark-1985}.
+Basal topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography and bathymetry dataset with a resolution of 1\,arc-minute \citep{data:etopo1}.
+Sea level is lowered by 120\,m and basal topography responds to ice load following a regional isostasy model that includes lithospheric flexure and mantle relaxation \citep{lingle-clark-1985}.
 

### 2013-11-08 floats/figures.tex
Fix parentheses in references

@@
 	\end{center}
-	\caption{Topography maps from WorldClim \citep{data:worldclim}, ERA-Interim reanalysis \citep{data:erai}, North American Regional Reanalysis (NARR) \citep{data:narr}, ETOPO1 \citep{data:etopo1}, Climate System Forecast Reanalysis (CFSR) \citep{data:cfsr}, and NCEP/NCAR reanalysis \citep{data:ncar}.
+	\caption{Topography maps from WorldClim \citep{data:worldclim}, ERA-Interim reanalysis \citep{data:erai}, North American Regional Reanalysis \citep[NARR;][]{data:narr}, ETOPO1 \citep{data:etopo1}, Climate System Forecast Reanalysis \citep[CFSR;][]{data:cfsr}, and NCEP/NCAR reanalysis \citep{data:ncar}.
 Whereas ETOPO1 is used as basal topography for the ice sheet model, all others are used data used as a reference for temperature lapse-rate corrections.
@@
 	\end{center}
-	\caption{Summer (JJA) and winter (DJF) temperature maps from WorldClim \citep{data:worldclim}, and summer (JJA) temperature maps from ERA-Interim reanalysis \citep{data:erai}, North American Regional Reanalysis (NARR) \citep{data:narr}, Climate System Forecast Reanalysis (CFSR) \citep{data:cfsr}, and NCEP/NCAR reanalysis \citep{data:ncar} climatologies.}
+	\caption{Summer (JJA) and winter (DJF) temperature maps from WorldClim \citep{data:worldclim}, and summer (JJA) temperature maps from ERA-Interim reanalysis \citep{data:erai}, North American Regional Reanalysis \citep[NARR;][]{data:narr}, Climate System Forecast Reanalysis \citep[CFSR;][]{data:cfsr}, and NCEP/NCAR reanalysis \citep{data:ncar} climatologies.}
 	\label{fig:temp}
@@
 	\end{center}
-	\caption{Winter (DJF) and summer (JJA) precipitation maps from WorldClim \citep{data:worldclim}, and winter (DFJ) temperature maps from ERA-Interim reanalysis \citep{data:erai}, North American Regional Reanalysis (NARR) \citep{data:narr}, Climate System Forecast Reanalysis (CFSR) \citep{data:cfsr}, and NCEP/NCAR reanalysis \citep{data:ncar} climatologies.
+	\caption{Winter (DJF) and summer (JJA) precipitation maps from WorldClim \citep{data:worldclim}, and winter (DFJ) temperature maps from ERA-Interim reanalysis \citep{data:erai}, North American Regional Reanalysis \citep[NARR;][]{data:narr}, Climate System Forecast Reanalysis \citep[CFSR;][]{data:cfsr}, and NCEP/NCAR reanalysis \citep{data:ncar} climatologies.
 Additional forcing data was prepared to correct for wave-like precipitation artefacts in CFSR (section~\ref{sec:climate}).}

### 2013-11-08 sections/climate.tex
Fix parentheses in references

@@
 
-In addition to the WorldClim data, we use surface air temperatures and precipitation rates from three global atmospheric reanalyses and one regional atmospheric reanalysis to inform the mass balance routine of PISM: the NCEP/NCAR reanalysis \citep{data:ncar}, the ERA-Interim reanalysis \citep{data:erai}, the Climate System Forecast Reanalysis (CFSR) \citep{data:cfsr}, and the North American Regional Reanalysis (NARR) \citep{data:narr}.
+In addition to the WorldClim data, we use surface air temperatures and precipitation rates from three global atmospheric reanalyses and one regional atmospheric reanalysis to inform the mass balance routine of PISM: the NCEP/NCAR reanalysis \citep{data:ncar}, the ERA-Interim reanalysis \citep{data:erai}, the Climate System Forecast Reanalysis \citep[CFSR;][]{data:cfsr}, and the North American Regional Reanalysis \citep[NARR;][]{data:narr}.
 Monthly climatologies from NCEP/NCAR and NARR reanalyses were provided by the \citet{web:psd} whereas monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean time series (Table \ref{tab:reanalyses}).

### 2013-11-08 sections/concl.tex
Fix parentheses in references

@@
 
-For the Cordilleran Ice Sheet, we achieve the best fit to the mapped LGM margin by \citet{dyke-2004} using climate forcing from the high-resolution interpolated observational data WorldClim \citep{data:worldclim} and the North American Regional Reanalysis (NARR) \citep{data:narr}.
+For the Cordilleran Ice Sheet, we achieve the best fit to the mapped LGM margin by \citet{dyke-2004} using climate forcing from the high-resolution interpolated observational data WorldClim \citep{data:worldclim} and the North American Regional Reanalysis \citep[NARR;][]{data:narr}.
 The latter dataset is preferable in our case due to the lack of WorldClim data offshore.

### 2013-11-08 sections/intro.tex
Fix parentheses in references

@@
 
-Arguably the most physically sound way to force an ice sheet model for simulations of glacial history is to couple it with a General Circulation Model (GCM) \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013}.
+Arguably the most physically sound way to force an ice sheet model for simulations of glacial history is to couple it with a General Circulation Model \citep[GCM;][]{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013}.
 However the computational demand of GCMs is such that only models of intermediate complexity can run on the time-scales of tens of thousands of years characteristic of ice sheet growth and decay.
 
-Climatologies obtained from GCM palaeo-climate simulations such as produced within the Paleoclimate Modelling Intercomparison Project (PMIP) \citep{joussaume-taylor-1995} provide perhaps a more reasonable representation of past climate.
+Climatologies obtained from GCM palaeo-climate simulations such as produced within the Paleoclimate Modelling Intercomparison Project \citep[PMIP;][]{joussaume-taylor-1995} provide perhaps a more reasonable representation of past climate.
 However, as such climatologies are only available for specific periods of time, using them as climate forcing for an ice sheet model requires either an assumption of steady-state ice sheet response to climatic fluctuations \citep{huybrechts-tsiobbel-1996}, or interpolation through time between climatologies from different periods, which can be linear \citep{charbit-etal-2002}, or modulated by a ``glacial index'' weighting function derived from ice core $\delta^{18}$O records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}.

### 2013-11-08 floats/figures.tex
Use SymLogNorm in precipitation difference map

@@
 	\end{center}
-	\caption{Logarithmically coloured density maps, showing a comparison of JJA surface air temperature data from the WorldClim climatology, against that of each atmospheric reanalysis.
+	\caption{Logarithmically coloured density maps, showing a comparison of summer (JJA) surface air temperature data from the WorldClim climatology, against that of each atmospheric reanalysis.
 Apparent horizontal lines are an artefact of different horizontal resolutions.
@@
 	\end{center}
-	\caption{JJA surface air temperature difference maps against WorldClim data, after bi-linear spatial interpolation.
+	\caption{Summer (JJA) surface air temperature difference maps against WorldClim data, after bi-linear spatial interpolation.
 Note the cold bias of NCEP/NCAR data and temperature anomalies due to unresolved topographic detail.}
@@
 	\end{center}
-	\caption{Logarithmically coloured density maps, showing a comparison of DJF precipitation rate from the WorldClim climatology, against that of each atmospheric reanalysis.
+	\caption{Logarithmically coloured density maps, showing a comparison of winter (DJF) precipitation rate from the WorldClim climatology, against that of each atmospheric reanalysis.
 Apparent horizontal lines are an artefact of different horizontal resolutions.
@@
 	\end{center}
-	\caption{DJF precipitation rate relative difference maps against WorldClim data, after bi-linear spatial interpolation.
+	\caption{Winter (DJF) precipitation rate difference maps against WorldClim data, after bi-linear spatial interpolation.
 Note the wet bias of all reanalysis data and the large anomalies of CFSR and NCEP/NCAR data.}

### 2013-11-08 figures/plot.py

@@ -221,7 +221,7 @@ def precdiff():
     # read WorldClim data as reference
     nc = Dataset('../data/wc-nn.nc')
     var = nc.variables['precipitation']
-    ref = _seasonmean(var, 'djf')
+    ref = _seasonmean(var, 'djf')*1000/12.
 
     # loop on climate datasets
     for i, clim in enumerate(climates):
@@ -229,14 +229,15 @@ def precdiff():
       nc = Dataset('../data/%s-bl.nc' % clim)
       _annotate(labels[clim])
       var = nc.variables['precipitation']
-      data = _seasonmean(var, 'djf')
-      im = plt.imshow(data/ref-1 ,
+      data = _seasonmean(var, 'djf')*1000/12.
+      im = plt.imshow(data - ref,
         cmap = plt.cm.PuOr,
-        norm = mcolors.Normalize(-2, 2))
+        norm = mcolors.SymLogNorm(10., 1.0, -300., 300.))
 
     # add colorbar and save
-    cb = fig.colorbar(im, fig.grid.cbar_axes[0], ticks=None)
-    cb.set_label(u'normalized DJF precipitation rate difference to WorldClim data')
+    cb = fig.colorbar(im, fig.grid.cbar_axes[0],
+      ticks=[-300, -200, -100, -50, -10, 0, 10, 50, 100, 200, 300])
+    cb.set_label('DJF precipitation rate difference to WorldClim data (mm/month)')
     _savefig('cordillera-climate-precdiff')
 
 def precheatmap():

### 2013-11-12 sections/abstract.tex
More small corrections from Irina

@@
 Modelled ice sheet outlines and volumes appear highly sensitive to the choice of climate forcing.
-For three of the four reanalysis dataset used, differences in precipitation is the major source for discrepancies between model results.
+For three of the four reanalysis datasets used, differences in precipitation are the major source for discrepancies between model results.
 We assess model performance against a geomorphological reconstruction of the ice margin at the Last Glacial Maximum, and suggest that part of the mismatch is due to unresolved orographic precipitation effects caused by the coarse resolution of reanalysis data.
-The best match between model output and the reconstructed ice margin is obtained using the high-resolution North American Regional Reanalysis, which we retain for future simulations of the Cordilleran Ice Sheet.
+The best match between model output and the reconstructed ice margin is obtained using the high-resolution North American Regional Reanalysis, which we retain for simulations of the Cordilleran Ice Sheet in the future.
 \end{abstract}

### 2013-11-12 sections/climate.tex
More small corrections from Irina

@@
 It was built by applying a spatial interpolation between a large set of measurements taken at terrestrial weather stations worldwide, using SRTM \citep{data:srtm} and GTOPO30 elevation data \citep{data:gtopo30}.
-This dataset provides a resolution much higher than attained by circulation models, however its reliability is variable to the highly inhomogeneous distribution of measurements \citep{data:worldclim}.
+This dataset provides a resolution much higher than attained by circulation models, however its reliability is variable due to the highly inhomogeneous distribution of measurements \citep{data:worldclim}.
 
 Within our modelling domain, the spatial distributions of summer (JJA) and winter (DJF) air surface temperatures in WorldClim generally show a decrease of temperature from south to north (JJA) and from south-west to north-east (DJF), such that regions further inland experience colder winters (Fig.~\ref{fig:temp}).
-It should be noted that temperature gradients are reversed (colder inland than at the coast) and much stronger in winter than in summer.
-Temperatures rises well above zero during the summer months over the entire modelling domain, except for the highest mountain peaks, and even regions where mean annual temperatures are well below freezing point do experience warm summers.
+It should be noted that temperature gradients are much stronger in winter than in summer.
+Temperature rises well above zero during the summer months over the entire modelling domain, except for the highest mountain peaks, and even regions where mean annual temperatures are well below freezing point do experience warm summers.
 In other words, there is a strong contrast in seasonality between coastal temperate-climate and inland continental-climate regions.
 
-The spatial distribution of JJA and DJF mean precipitation rates in WorldClim reveal a strong precipitation decline beyond the coastal regions (Fig.~\ref{fig:prec}), primarily as a result of the continuous orographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascades.
+The spatial distribution of JJA and DJF mean precipitation rates in WorldClim reveals a strong precipitation decline beyond the coastal regions (Fig.~\ref{fig:prec}), primarily as a result of the continuous orographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascades.
 A less pronounced precipitation peak is also visible over the interior ranges.
@@
 
-The spatial distributions of JJA air surface temperatures and DJF precipitation rates from the four reanalyses climatologies are shown in Figures~\ref{fig:temp} and~\ref{fig:prec}.
+The spatial distributions of JJA air surface temperatures and DJF precipitation rates from the four reanalysis climatologies are shown in Figures~\ref{fig:temp} and~\ref{fig:prec}.
 Summer temperature and winter precipitation are most relevant to the glacier model as they respectively drive summer melt and winter accumulation.

### 2013-11-12 sections/concl.tex
More small corrections from Irina

@@
 The latter dataset is preferable in our case due to the lack of WorldClim data offshore.
-Climate forcing from CFSR and the NCEP/NCAR reanalysis produces largely oversized ice sheets due to too high precipitation rates, and in the second case, too low surface air temperature.
+Climate forcing from CFSR and the NCEP/NCAR reanalysis produce largely oversized ice sheets due to too high precipitation rates, and in the second case, too low surface air temperature.
 The ERA-Interim data used in this study produces more reasonable results, but it may be too coarse to accurately resolve the spatial distribution of orographic precipitation associated with the rugged topography of the northern American Cordillera, resulting in a misplaced ice cover in regard to geomorphological data.
-Therefore, we retain NARR data for forcing future simulations of the Cordilleran Ice Sheet.
+Therefore, we retain NARR data for forcing simulations of the Cordilleran Ice Sheet in the future.
 

### 2013-11-12 sections/discuss.tex
More small corrections from Irina

@@
 
-Although this selection result in similarly sized ice sheets, noticeable differences in shape exist.
-The ERA-Interim simulation produces a ore northerly-centred ice sheet with too much ice in the North and too little in the South.
+Although this selection results in similarly sized ice sheets, noticeable differences in shape exist.
+The ERA-Interim simulation produces a more northerly-centred ice sheet with too much ice in the North and too little in the South.
 This may partly reflect unresolved orographic precipitation effects (Fig.~\ref{fig:oroprecip}) due to coarser GCM resolution (Table~\ref{tab:reanalyses}).
@@
 However the Laurentide Ice Sheet, absent from our simulations, likely formed a buttress against the smaller Cordilleran Ice Sheet and stopped its advance onto the Canadian Prairies.
-In addition, potential effects of the growing ice-sheet on regional climate are not included in our model.
+In addition, potential effects of the growing ice-sheet on atmospheric circulation and precipitation changes are not included in our model.
 The Cordilleran ice sheet initiated from the junction of mountain ice caps over the major reliefs \citep{clague-1989}.
@@
 
-Considering the lack of WorldClim data offshore, we consider the NARR climate forcing as best suited for simulations of the Cordilleran Ice Sheet among those tested, and under the assumptions of the present study.
+Given the lack of WorldClim data offshore, we consider the NARR climate forcing as best suited for simulations of the Cordilleran Ice Sheet among those tested, and under the assumptions of the present study.
 
@@
 
-To test the effect of a shorter englaciation period on model results, we use simulations run under NARR climate forcing with temperatures offsets of 7 to 11\,\unit{\degree C}, and compare modelled ice sheet extent when glaciated area reaches the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$.
+To test the effect of a shorter englaciation period on model results, we use simulations run under NARR climate forcing with temperature offsets of 7 to 11\,\unit{\degree C}, and compare modelled ice sheet extent when glaciated area reaches the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$.
 
@@
 Refreezing of melted snow and ice and retention in the snow pack, however, can greatly alter mass balance at the surface of an ice sheet \citep{janssens-huybrechts-2000}.
-Additionally, we simulate temperature variability by using a constant, uniform value of temperature standard deviation.
+Additionally, we simulate temperature variability by using a constant, uniform value of daily temperature standard deviation.
 This approach implies large biases of surface mass-balance \citep{charbit-etal-2013,rau-rogozhina-2013,seguinot-inpress}, particularly over regions with such various climates as the Northern American Cordillera.

### 2013-11-12 sections/intro.tex
More small corrections from Irina

@@
 
-Here, we propose to address some of the uncertainties concerning climate forcing of numerical glacier models by evaluating the responses of a numerical model, in terms of glacier extent, to inputs fromo several climate datasets.
+Here, we propose to address some of the uncertainties concerning climate forcing of numerical glacier models by evaluating the responses of a numerical model, in terms of glacier extent, to inputs from several climate datasets.
 This is not entirely a new approach as a few studies of this kind are presently available.
@@
 From a numerical modelling perspective, it is one of the least studied palaeo-ice sheets of the Northern Hemisphere, despite the fact that significant stratigraphical and geomorphological data is available to constrain its extent \citep{jackson-clague-1991,dukrodkin-1999,kaufman-manley-2004,kleman-etal-2010,margold-etal-2011}.
-Although the Cordilleran Ice Sheet was not the primary target of these simulations, it has previously been modelled as part of efforts to simulate ice sheets in North America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} and world-wide \citep{yoshimori-etal-2001}.
+The Cordilleran Ice Sheet has previously been modelled as part of efforts to simulate ice sheets in North America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} and world-wide \citep{yoshimori-etal-2001}.
 While these studies generally reproduce the magnitude of North American glaciation at LGM reasonably well, there exists a tendency in the simulations that are independent of ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and interior Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.

### 2013-11-12 sections/model.tex
Fix parentheses in references
More small corrections from Irina

@@
 Given basal topography, sea level, geothermal heat flux and climate forcing, the model computes ice extent and thickness, its thermal and dynamic state, and the associated lithospheric response.
-Our modelling domain encompasses the entire area covered by the Cordilleran Ice Sheet at LGM including independent ice growing on the western Alaskan ranges and the Brooks Range and intermediated expanses of ice-free terrain in northern Yukon Territory and interior Alaska (Fig.~\ref{fig:locmap}).
+Our modelling domain encompasses the entire area covered by the Cordilleran Ice Sheet at LGM including independent ice growing on the western Alaskan ranges and the Brooks Range and intermediate expanses of ice-free terrain in northern Yukon Territory and interior Alaska (Fig.~\ref{fig:locmap}).
 
@@
 
-The central part of an ice sheet model consist of the computation of flow velocity which itself depends on temperature.
+The central part of an ice sheet model consists of the computation of flow velocity which itself depends on temperature.
 PISM is a shallow model, which implies that the balance of stresses is approximated based on their predominant components.
@@
 
-A pseudo-plastic sliding law \citep{aschwanden-etal-2013} (supplement) relates the bed-parallel shear stress and the sliding velocity.
+A pseudo-plastic sliding law \citep[supplement]{aschwanden-etal-2013} relates the bed-parallel shear stress and the sliding velocity.
 The yield stress is modelled using the Mohr-Coulomb criterion.

### 2013-11-12 sections/results.tex
More small corrections from Irina

@@
 The magnitude of glaciation differs dramatically from one climate forcing to another.
-Whereas the NCEP/NCAR and CFSR climatologies produce a large ice sheet that covers most of the model domain (Fig.~\ref{fig:cool05}, lower panels), WorldClim, ERA-Interim and NARR climatologies lead to more restrictive ice cover (Fig.~\ref{fig:cool05}, upper panels).
+Whereas the NCEP/NCAR and CFSR climatologies produce a large ice sheet that covers most of the model domain (Fig.~\ref{fig:cool05}, lower panels), WorldClim, ERA-Interim and NARR climatologies lead to a more restrictive ice cover (Fig.~\ref{fig:cool05}, upper panels).
 The WorldClim run only resulted in the growth of mountain ice caps bounded to the highest topography.

### 2013-11-13 floats/figures.tex
Improvements in figure captions (Irina)

@@
 	\caption{Topography maps from WorldClim \citep{data:worldclim}, ERA-Interim reanalysis \citep{data:erai}, North American Regional Reanalysis \citep[NARR;][]{data:narr}, ETOPO1 \citep{data:etopo1}, Climate System Forecast Reanalysis \citep[CFSR;][]{data:cfsr}, and NCEP/NCAR reanalysis \citep{data:ncar}.
-Whereas ETOPO1 is used as basal topography for the ice sheet model, all others are used data used as a reference for temperature lapse-rate corrections.
-Fig.~\ref{fig:topo}-\ref{fig:durationstack} drawn using Matplotlib \citep{soft:mpl}.}
+Whereas ETOPO1 is used as basal topography for the ice sheet model, all other data are used as a reference for temperature lapse-rate corrections.
+Fig.~\ref{fig:topo}-\ref{fig:durationstack} are drawn using Matplotlib \citep{soft:mpl}.}
 	\label{fig:topo}
@@
 	\end{center}
-	\caption{Total glaciated area and ice volume after 10\,ka as a function of temperature offset and climate forcing used.}
+	\caption{Total glaciated area and ice volume after 10\,ka as a function of temperature offset for each climate forcing used.}
 	\label{fig:ivolarea}
@@
 	\end{center}
-	\caption{“Hybrid” atmosphere simulations using precipitation rate from WorldClim and surface air temperature from each reanalysis (upper row), and vice-versa (lower row).
-(1 km contours)In other words, the upper row shows the effect of temperature anomalies, and the lower row the effect of precipitation anomalies, for each reanalysis, relative to WorldClim data.
-Each simulation use a 5\,\unit{\degree C} offset.
-Compare with figure~\ref{fig:cool05}}
+	\caption{Ice surface topography (1 km contours) and velocity (\unit{m\,a^{-1}}) after 10\,ka using “hybrid” atmosphere simulations with precipitation rate from WorldClim and surface air temperature from each reanalysis (upper row), and vice-versa (lower row).
+In other words, the upper row shows the effect of temperature anomalies, and the lower row the effect of precipitation anomalies, for each reanalysis, relative to WorldClim data.
+Each simulation uses a 5\,\unit{\degree C} offset for comparison with figure~\ref{fig:cool05}}
 	\label{fig:biatm}
@@
 	\end{center}
-	\caption{Effect of temperature and precipitation anomalies of each reanalysis, and both, on modelled final ice volume, using the WorldClim 5\,\unit{\degree C} offset simulation as a reference.
+	\caption{Effect of temperature and precipitation anomalies (separately and jointly) from each reanalysis on modelled final ice volume relative to the result of the WorldClim 5\,\unit{\degree C} offset simulation.
 Corresponding ice sheet geometries are presented in figures~\ref{fig:cool05} and~\ref{fig:biatm}}
@@
 	\end{center}
-	\caption{Model sensitivity to simulation length.
+	\caption{Left panel: model sensitivity to simulation length.
 Modelled glaciated area, using NARR forcing data, and temperature offsets from 0 to 15\,\unit{\degree C}, solid lines corresponding to values from 7 to 11\,\unit{\degree C} (left).
-Corresponding modelled ice margin, compared to the LGM ice sheet margin by \citep{dyke-2004} (right).}
+Right panel: modelled ice margin corresponding to temperature offsets from 7 to 11\,\unit{\degree C} when total glaciated area reach the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$.
+A reconstructed LGM ice sheet margin by \citet{dyke-2004} (Fig.~\ref{fig:locmap}) is given for reference (grey shading).
+Note that shorter (and colder) simulations lead to more restrictive glaciation of the continental eastern margin but further ice extent in the maritime south-western part of the modelling domain.}
 	\label{fig:durationstack}

### 2013-11-18 floats/figures.tex
Final comments from Arjen

@@
 	\end{center}
-	\caption{Topography maps from WorldClim \citep{data:worldclim}, ERA-Interim reanalysis \citep{data:erai}, North American Regional Reanalysis \citep[NARR;][]{data:narr}, ETOPO1 \citep{data:etopo1}, Climate System Forecast Reanalysis \citep[CFSR;][]{data:cfsr}, and NCEP/NCAR reanalysis \citep{data:ncar}.
+	\caption{Topography maps from WorldClim \citep{data:worldclim}, ERA-Interim reanalysis \citep{data:erai}, North American Regional Reanalysis \citep[NARR;][]{data:narr}, ETOPO1 \citep{data:etopo1}, Climate Forecast System Reanalysis \citep[CFSR;][]{data:cfsr}, and NCEP/NCAR reanalysis \citep{data:ncar}.
 Whereas ETOPO1 is used as basal topography for the ice sheet model, all other data are used as a reference for temperature lapse-rate corrections.
@@
 	\end{center}
-	\caption{Summer (JJA) and winter (DJF) temperature maps from WorldClim \citep{data:worldclim}, and summer (JJA) temperature maps from ERA-Interim reanalysis \citep{data:erai}, North American Regional Reanalysis \citep[NARR;][]{data:narr}, Climate System Forecast Reanalysis \citep[CFSR;][]{data:cfsr}, and NCEP/NCAR reanalysis \citep{data:ncar} climatologies.}
+	\caption{Summer (JJA) and winter (DJF) temperature maps from WorldClim \citep{data:worldclim}, and summer (JJA) temperature maps from ERA-Interim reanalysis \citep{data:erai}, North American Regional Reanalysis \citep[NARR;][]{data:narr}, Climate Forecast System Reanalysis \citep[CFSR;][]{data:cfsr}, and NCEP/NCAR reanalysis \citep{data:ncar} climatologies.}
 	\label{fig:temp}
@@
 	\end{center}
-	\caption{Winter (DJF) and summer (JJA) precipitation maps from WorldClim \citep{data:worldclim}, and winter (DFJ) temperature maps from ERA-Interim reanalysis \citep{data:erai}, North American Regional Reanalysis \citep[NARR;][]{data:narr}, Climate System Forecast Reanalysis \citep[CFSR;][]{data:cfsr}, and NCEP/NCAR reanalysis \citep{data:ncar} climatologies.
+	\caption{Winter (DJF) and summer (JJA) precipitation maps from WorldClim \citep{data:worldclim}, and winter (DFJ) temperature maps from ERA-Interim reanalysis \citep{data:erai}, North American Regional Reanalysis \citep[NARR;][]{data:narr}, Climate Forecast System Reanalysis \citep[CFSR;][]{data:cfsr}, and NCEP/NCAR reanalysis \citep{data:ncar} climatologies.
 Additional forcing data was prepared to correct for wave-like precipitation artefacts in CFSR (section~\ref{sec:climate}).}
@@
 	\end{center}
-	\caption{Logarithmically coloured density maps, showing a comparison of summer (JJA) surface air temperature data from the WorldClim climatology, against that of each atmospheric reanalysis.
+	\caption{Logarithmically coloured density maps, showing a comparison of summer (JJA) surface air temperature data from the WorldClim climatology, against that of each reanalysis.
 Apparent horizontal lines are an artefact of different horizontal resolutions.
@@
 	\end{center}
-	\caption{Logarithmically coloured density maps, showing a comparison of winter (DJF) precipitation rate from the WorldClim climatology, against that of each atmospheric reanalysis.
+	\caption{Logarithmically coloured density maps, showing a comparison of winter (DJF) precipitation rate from the WorldClim climatology, against that of each reanalysis.
 Apparent horizontal lines are an artefact of different horizontal resolutions.
@@
 	\end{center}
-	\caption{Ice surface topography (1 km contours) and velocity (\unit{m\,a^{-1}}) after 10\,ka using “hybrid” atmosphere simulations with precipitation rate from WorldClim and surface air temperature from each reanalysis (upper row), and vice-versa (lower row).
+	\caption{Ice surface topography (1 km contours) and velocity (\unit{m\,a^{-1}}) after 10\,ka using “hybrid” climate forcing with precipitation rate from WorldClim and surface air temperature from each reanalysis (upper row), and surface air temperature from WorldClim and precipitation rate from each reanalysis (lower row).
 In other words, the upper row shows the effect of temperature anomalies, and the lower row the effect of precipitation anomalies, for each reanalysis, relative to WorldClim data.

### 2013-11-18 sections/climate.tex
Final comments from Arjen

@@
 
-In addition to the WorldClim data, we use surface air temperatures and precipitation rates from three global atmospheric reanalyses and one regional atmospheric reanalysis to inform the mass balance routine of PISM: the NCEP/NCAR reanalysis \citep{data:ncar}, the ERA-Interim reanalysis \citep{data:erai}, the Climate System Forecast Reanalysis \citep[CFSR;][]{data:cfsr}, and the North American Regional Reanalysis \citep[NARR;][]{data:narr}.
+In addition to the WorldClim data, we use surface air temperatures and precipitation rates from three global atmospheric reanalyses and one regional atmospheric reanalysis to inform the mass balance routine of PISM: the NCEP/NCAR reanalysis \citep{data:ncar}, the ERA-Interim reanalysis \citep{data:erai}, the Climate Forecast System Reanalysis \citep[CFSR;][]{data:cfsr}, and the North American Regional Reanalysis \citep[NARR;][]{data:narr}.
 Monthly climatologies from NCEP/NCAR and NARR reanalyses were provided by the \citet{web:psd} whereas monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean time series (Table \ref{tab:reanalyses}).
@@
 In addition, WorldClim data was extrapolated to cover grid points in the ocean using the nearest-neighbour algorithm \citep{soft:scipy}.
-Note that here we choose to present original rather than interpolated data, in order to highlight differences between datasets related to spatial resolution (Fig.~\ref{fig:temp} and~\ref{fig:prec}).
+Note that here we choose to present original rather than interpolated data, in order to highlight differences between datasets related to spatial resolution (Figs.~\ref{fig:temp} and~\ref{fig:prec}).
 

### 2013-11-18 sections/concl.tex
Final comments from Arjen

@@
 Our study shows a strong dependency of ice sheet model results on the choice of climate forcing data.
-For three of the four reanalysis datasets used, precipitation rate, over surface air temperature, causes much of the discrepancy between modelled ice sheet outlines and volumes.
-Furthermore, the spatial resolution of input data appears critical for providing the ice sheet model with an accurate precipitation field, confirming results obtained by \citet{quiquet-etal-2012} from various GCM forcing over the Greenland Ice Sheet.
+For three of the four reanalysis datasets used, precipitation rate, over surface air temperature, causes much of the variation in modelled ice sheet outlines and volumes.
+Furthermore, the spatial resolution of input data appears critical for providing the ice sheet model with an accurate precipitation field, confirming results obtained by \citet{quiquet-etal-2012} from various GCM forcing datasets over the Greenland Ice Sheet.
 
-For the Cordilleran Ice Sheet, we achieve the best fit to the mapped LGM margin by \citet{dyke-2004} using climate forcing from the high-resolution interpolated observational data WorldClim \citep{data:worldclim} and the North American Regional Reanalysis \citep[NARR;][]{data:narr}.
+For the Cordilleran Ice Sheet, we achieve the best fit to the mapped LGM margin by \citet{dyke-2004} using climate forcing from the high-resolution interpolated observational dataset WorldClim \citep{data:worldclim} and the North American Regional Reanalysis \citep[NARR;][]{data:narr}.
 The latter dataset is preferable in our case due to the lack of WorldClim data offshore.
 Climate forcing from CFSR and the NCEP/NCAR reanalysis produce largely oversized ice sheets due to too high precipitation rates, and in the second case, too low surface air temperature.
-The ERA-Interim data used in this study produces more reasonable results, but it may be too coarse to accurately resolve the spatial distribution of orographic precipitation associated with the rugged topography of the northern American Cordillera, resulting in a misplaced ice cover in regard to geomorphological data.
+The ERA-Interim data used in this study produces more reasonable results, but it may be too coarse to accurately resolve the spatial distribution of orographic precipitation associated with the rugged topography of the study area, resulting in a misplaced ice coverage in regard to geomorphological data.
 Therefore, we retain NARR data for forcing simulations of the Cordilleran Ice Sheet in the future.
 
-One must keep in mind, however, that these results are biased by our choices of ice-sheet model (PISM), surface mass balance (PDD) model, study area (the Cordilleran Ice Sheet), and more importantly palaeo-climate representation (temperature offsets).
+One must keep in mind, however, that these results are only accurate for our choices of ice-sheet model (PISM), surface mass balance model (PDD), study object (the Cordilleran Ice Sheet), and, most importantly, palaeo-climate representation (temperature offsets).
 At LGM, it is most likely that the presence of ice sheets significantly affected circulation patterns and distribution of precipitation.

### 2013-11-18 sections/discuss.tex
Final comments from Arjen

@@
 
-These simulations of the LGM Cordilleran Ice Sheet show a very large sensitivity to the choice of climate forcing data used (Fig.~\ref{fig:ivolarea}--\ref{fig:cool05}).
+These simulations of the LGM Cordilleran Ice Sheet show a very large sensitivity to the choice of climate forcing data used (Figs.~\ref{fig:ivolarea}--\ref{fig:cool05}).
 To understand the origin of the discrepancies between model results, we quantify differences in surface air temperature and precipitation rate between the input climatologies and test their effect on the ice sheet model.
@@
 
-As previously outlined, significant differences in the distribution of winter (DJF) precipitation and summer (JJA) temperature, which are the predominant controls on accumulation and melt at the modelled ice sheet surface, exist between datasets (Fig.~\ref{fig:topo}--\ref{fig:prec}).
+As previously outlined, significant differences in the distribution of winter (DJF) precipitation and summer (JJA) temperature, which are the predominant controls on accumulation and melt at the modelled ice sheet surface, exist between datasets (Figs.~\ref{fig:topo}--\ref{fig:prec}).
 In this section we use the WorldClim climatology, which is derived from observations, as a reference with which to compare other climate forcing data on land.
 ERA-Interim, CFSR and NARR climatologies exhibit temperature distributions fairly consistent with WorldClim data (Fig.~\ref{fig:tempheatmap}).
-When examining the spatial distribution of temperature differences between reanalyses and WorldClim data, it appears that most of the disparity between reanalyses and observation data is caused by topographical detail at scales unresolved by the atmospheric reanalyses (Fig.~\ref{fig:tempdiff}).
-Surface air temperature data from the NCAR climatology shows a large disparity with that of WorldClim, with a significant cold anomaly over most of the modelling domain (Fig.~\ref{fig:tempheatmap}--\ref{fig:tempdiff}).
+When examining the spatial distribution of temperature differences between reanalyses and WorldClim data, it appears that most of the disparity between reanalyses and observation data is caused by topographical detail at scales unresolved in the reanalysis datasets (Fig.~\ref{fig:tempdiff}).
+Surface air temperature data from the NCEP/NCAR climatology shows the largest disparity with that of WorldClim, with a significant cold anomaly over most of the modelling domain (Figs.~\ref{fig:tempheatmap}--\ref{fig:tempdiff}).
 
@@
 As described in section~\ref{sec:climate}, the topography of the Northern American Cordillera is such that its western ranges form a continuous orographic barrier, causing high precipitation along the Pacific coast while leaving much of the interior arid.
-However the ability of a GCM to reproduce these contrasts in precipitation is bounded by horizontal resolution.
+However the ability of a GCM to reproduce these contrasts in precipitation is bounded by spatial resolution.
 In a model of coarser resolution, these high mountain ranges are reduced to smooth hills and the modelled distribution of precipitation is also smoother (Fig.~\ref{fig:oroprecip}).
-However this effect alone can not explain the widespread and pronounced positive precipitation anomalies observed in CFSR data (Fig.~\ref{fig:precheatmap}--\ref{fig:precdiff}) despite of the high model resolution (Table~\ref{tab:reanalyses}).
+However this effect alone cannot explain the widespread and pronounced positive precipitation anomalies observed in CFSR data (Figs.~\ref{fig:precheatmap}--\ref{fig:precdiff}) despite of its high spatial resolution (Table~\ref{tab:reanalyses}).
 
@@
 To distinguish the effects of temperature and precipitation biases on the ice sheet model, we run a series of eight additional simulations.
-These new simulations use a “hybrid” atmosphere forcing that consist of temperature data from WorldClim, combined with precipitation data from each of the four reanalyses, and vice-versa.
+These new simulations use a “hybrid” climate forcing that consists of temperature data from WorldClim, combined with precipitation data from each of the four reanalyses, and vice-versa.
 A single temperature offset of 5\,\unit{\degree C} is used.
@@
 For ERA-Interim, CFSR and NARR climatologies, precipitation anomalies clearly dominate the differences in ice sheet response, whereas temperature anomalies have relatively little effect (Fig.~\ref{fig:biatmbars}).
-However in the case of NCAR forcing, both the negative temperature bias (Fig.~\ref{fig:tempheatmap}--\ref{fig:tempdiff}) and the important precipitation differences (Fig.~\ref{fig:precheatmap}--\ref{fig:precdiff}) contribute to produce oversized ice sheets (Fig.~\ref{fig:biatmbars}).
+However in the case of NCAR forcing, both the negative temperature bias (Figs.~\ref{fig:tempheatmap}--\ref{fig:tempdiff}) and the important precipitation differences (Figs.~\ref{fig:precheatmap}--\ref{fig:precdiff}) contribute approximately equally to produce oversized ice sheets (Fig.~\ref{fig:biatmbars}).
 
@@
 
-We compare numerically modelled ice sheet geometries to a reconstruction of the ice margin at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) by \citet{dyke-2004}, based on glacial geomorphology and radiocarbon dating.
-This corresponds to the LGM extent of most of the Cordilleran ice sheet, which occurred later than for the Laurentide ice sheet \citep{porter-swanson-1998,dyke-2004,stroeven-etal-2010}.
+We compare numerically modelled ice sheet geometries to a reconstruction of the ice margin at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) by \citet{dyke-2004}, based on glacial geomorphology and radiocarbon dating (Fig.~\ref{fig:locmap}).
+This corresponds to the LGM extent of most of the Cordilleran ice sheet, which occurred later than for the Laurentide ice sheet \citep{porter-swanson-1998,dyke-2004,stroeven-etal-2010,stroeven-etal-inpress}.
 
-Considering previously identified weaknesses of the CFSR and NCEP/NCAR data (Fig.~\ref{fig:tempheatmap}--\ref{fig:biatmbars}), we bound our analysis to the  WorldClim, ERA-Interim and NARR input climatologies.
+Considering previously identified weaknesses of the CFSR and NCEP/NCAR data (Figs.~\ref{fig:tempheatmap}--\ref{fig:biatmbars}), we restricted our analysis to the WorldClim, ERA-Interim and NARR input climatologies.
 In order to compare modelled ice sheets of a similar size, we select for each climate forcing the simulation that leads to a final glaciated area closest to the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$.
-This corresponds to temperature offset values of 8 (WorldClim), 6 (ERA-Interim) and 7\,\unit{\degree C} (NARR).
+This corresponds to temperature offset values of 8\,\unit{\degree C} (WorldClim), 6\,\unit{\degree C} (ERA-Interim) and 7\,\unit{\degree C} (NARR).
 These qualitative “best” runs are presented in Figure~\ref{fig:best} along with their associated temperature depressions and a reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.
@@
 Although this selection results in similarly sized ice sheets, noticeable differences in shape exist.
-The ERA-Interim simulation produces a more northerly-centred ice sheet with too much ice in the North and too little in the South.
-This may partly reflect unresolved orographic precipitation effects (Fig.~\ref{fig:oroprecip}) due to coarser GCM resolution (Table~\ref{tab:reanalyses}).
-
+The ERA-Interim simulation produces a more northerly-centred ice sheet with too much ice in the north and too little in the south.
+This may partly reflect unresolved orographic precipitation effects (Fig.~\ref{fig:oroprecip}) due to the coarser GCM resolution (Table~\ref{tab:reanalyses}).
 The LGM ice sheet margin by \citet{dyke-2004} is best reproduced by simulations driven by input climatologies from WorldClim and NARR data.
@@
 
-Firstly, the modelled eastern margin of the ice sheet extends further east than where one would expect junction between the Cordilleran and the Laurentide ice sheets.
+Firstly, the modelled eastern margin of the ice sheet extends further east than the inferred position of the junction between the Cordilleran and the Laurentide ice sheets.
 This can be seen for instance in the south-eastern part of our modelling domain, where the modelled ice sheets cover an ice-free corridor of the LGM reconstruction (Fig.~\ref{fig:best}).
-However the Laurentide Ice Sheet, absent from our simulations, likely formed a buttress against the smaller Cordilleran Ice Sheet and stopped its advance onto the Canadian Prairies.
+However, such comparisons have been hampered by the fact that we have not considered the buttressing effect of the Laurentide Ice Sheet, which would have inhibited the Cordilleran Ice Sheet from advancing onto the Canadian Prairies.
 In addition, potential effects of the growing ice-sheet on atmospheric circulation and precipitation changes are not included in our model.
-The Cordilleran ice sheet initiated from the junction of mountain ice caps over the major reliefs \citep{clague-1989}.
+The Cordilleran ice sheet initiated from the formation of mountain ice caps over the high mountain ranges \citep{clague-1989}.
 In our simulations, a continuous ice cover quickly forms over the western ranges, where precipitation rates are higher than in the rest of the domain.
-This continuous ice cover may have enhanced the topographical barrier already formed by the western ranges, resulting in less precipitation and warmer air in the interior.
-This process of ``self-inhibiting growth'' was demonstrated in a similar setting over Greenland, where it may have been limiting the advance of the ice sheet during its build-up \citep{langen-etal-2012}.
+In reality, this continuous ice cover enhanced the topographical barrier already formed by the western ranges, and would have resulted in less precipitation and warmer air in the interior.
+This process of ``self-inhibiting growth'' is not captured in our model but was demonstrated to be important in a similar setting over Greenland, where it may have been limiting the advance of the ice sheet during its build-up \citep{langen-etal-2012}.
 
@@
 
-\subsection{Model sensitivity to englaciation period}
+\subsection{Model sensitivity to the duration of ice sheet inception}
 
-A key assumption for the present study is our choice of 10\,ka for the length of all our simulations.
-Although geomorphological data shows that the period of growth of the Cordilleran Ice Sheet from nearly ice-free to full glacial conditions could not have been much longer than 10\,ka, they provide no lower bound to this value.
+A key assumption for the present study is our choice of 10\,ka for the duration of all our simulations.
+Although geomorphological data shows that the period of Cordilleran Ice Sheet inception, from nearly ice-free to full glacial conditions, could not have been much longer than 10\,ka, they provide no lower bound to this value.
 
-To test the effect of a shorter englaciation period on model results, we use simulations run under NARR climate forcing with temperature offsets of 7 to 11\,\unit{\degree C}, and compare modelled ice sheet extent when glaciated area reaches the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$.
+To test the effect of an inception period of shorter duration on model results, we use simulations run under NARR climate forcing with temperature offsets of 7 to 11\,\unit{\degree C}, and compare modelled ice sheet extent when glaciated area reaches the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$.
 
-This experiment shows that a shorter englaciation period leads to more restricted ice cover in the eastern part of the modelling domain, where precipitation rate controls ice advance, but further glaciation along the Pacific margin to the south, where temperature is the limiting factor (Fig.~\ref{fig:durationstack}).
-Reducing simulation length leads to a closer match with the LGM ice sheet margin by \citet{dyke-2004} in some parts of the modelling domain, but higher discrepancy in other parts.
+This experiment shows that a shorter start-up period leads to more restricted ice cover in the eastern part of the modelling domain, where precipitation rate controls ice advance, but more extensive glaciation in the south-western domain along the Pacific coast, where temperature is the limiting factor (Fig.~\ref{fig:durationstack}).
+Hence, reducing simulation duration leads to a closer match with the LGM ice sheet margin by \citet{dyke-2004} in some parts of the modelling domain, but higher discrepancy in other parts.
 
@@
 
-Our choice of 10\,ka simulation length could certainly be discussed further.
-However, our representation of climate history by constant temperature offsets is voluntarily simplistic.
+Our representation of climate history by constant temperature offsets is voluntarily simplistic.
 To better understand the transient character of the LGM Cordilleran Ice Sheet, time-dependent palaeo-climate forcing would be needed.
@@
 Additionally, we simulate temperature variability by using a constant, uniform value of daily temperature standard deviation.
-This approach implies large biases of surface mass-balance \citep{charbit-etal-2013,rau-rogozhina-2013,seguinot-inpress}, particularly over regions with such various climates as the Northern American Cordillera.
+This approach implies large biases of surface mass-balance \citep{charbit-etal-2013,rau-rogozhina-2013,seguinot-2013}, particularly over regions with such various climates as experienced within our model domain.

### 2013-11-18 sections/intro.tex
Final comments from Arjen

@@
 This is not entirely a new approach as a few studies of this kind are presently available.
-\citet{quiquet-etal-2012}, for example, assessed the sensitivity of a Greenland ice sheet model to various atmospheric forcing, including a regional parametrization by \citet{fausto-etal-2009}, output from several GCMs and an atmospheric reanalysis.
-\citet{rodgers-etal-2004} and \citet{charbit-etal-2007} tested the sensitivity of a model of the Northern Hemisphere ice sheet model to climate forcing from different PMIP LGM simulations.
-Although using different set-ups, all three studies demonstrate the very large sensitivity of ice sheet model to the choice of climate forcing data used.
+\citet{quiquet-etal-2012}, for example, assessed the sensitivity of a Greenland ice sheet model to various climate forcing fields, including a regional parametrization by \citet{fausto-etal-2009}, output from several GCMs and an atmospheric reanalysis.
+\citet{rodgers-etal-2004} and \citet{charbit-etal-2007} tested the sensitivity of a model of Northern Hemisphere ice sheets to climate forcing from different PMIP LGM simulations.
+Although using different set-ups, all three studies demonstrate the very large sensitivity of ice sheet models to the choice of climate forcing data used.
 In our study, to limit the degrees of freedom in our model and obtain results independent of palaeo-ice sheet reconstructions such as ICE-4G, we use a simple temperature offset similar to the approaches by \citet{greve-etal-1999} and \citet{bintanja-etal-2005} and assess ice sheet model sensitivity to the choice of present-day climate data.
 Hence, rather than using GCM output, we force our model with climate reanalysis data, which builds on observational information through data assimilation \citep{bengtsson-etal-2007}.
-Furthermore, we focus our study regionally on the former Cordilleran Ice Sheet in western Northern America.
+Furthermore, we focus our study regionally on the former Cordilleran Ice Sheet in western North America.
 
 The Cordilleran Ice Sheet (Fig.~\ref{fig:locmap}) covered an area that presently experiences strong regional variations in climate.
-From a numerical modelling perspective, it is one of the least studied palaeo-ice sheets of the Northern Hemisphere, despite the fact that significant stratigraphical and geomorphological data is available to constrain its extent \citep{jackson-clague-1991,dukrodkin-1999,kaufman-manley-2004,kleman-etal-2010,margold-etal-2011}.
+From a numerical modelling perspective, it is one of the least studied palaeo-ice sheets of the Northern Hemisphere, despite the fact that significant stratigraphical, geomorphological, and chronological data is available to constrain its extent \citep{jackson-clague-1991,porter-swanson-1998,dukrodkin-1999,dyke-2004,kaufman-manley-2004,kleman-etal-2010,stroeven-etal-2010,stroeven-etal-inpress,margold-etal-2011}.
 The Cordilleran Ice Sheet has previously been modelled as part of efforts to simulate ice sheets in North America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} and world-wide \citep{yoshimori-etal-2001}.
@@
 Thereby, we aim to determine the climate dataset which is most suited for simulation of the Cordilleran Ice Sheet and to use as input for future, transient studies over a glacial cycle.
-Model set-up is presented in section~\ref{sec:model} and climate forcing in section~\ref{sec:climate}.
-Results are exposed in section~\ref{sec:results} and discussed in section~\ref{sec:discussion}.
 To our knowledge, this is the first modelling study that specifically focuses on the Cordilleran Ice Sheet since \citet{robert-1991}.

### 2013-11-18 sections/model.tex
Final comments from Arjen

@@
 
-A pseudo-plastic sliding law \citep[supplement]{aschwanden-etal-2013} relates the bed-parallel shear stress and the sliding velocity.
+A pseudo-plastic sliding law \citep{aschwanden-etal-2013} relates the bed-parallel shear stress and the sliding velocity.
 The yield stress is modelled using the Mohr-Coulomb criterion.
@@
 where $\phi$ is given in degrees and $z$ in meters above current sea level.
-This models weakening of the till associated with the presence of marine sediments \citep{martin-etal-2011,aschwanden-etal-2013}.
+This model thus accounts for a weakening of till associated with the presence of marine sediments \citep{martin-etal-2011,aschwanden-etal-2013}.
 Basal topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography and bathymetry dataset with a resolution of 1\,arc-minute \citep{data:etopo1}.
@@
 It accounts for temperature variability assuming a normal distribution along a central (input) value.
-The temperature standard deviation is a constant model parameter and was assigned a value of 3.07\,\unit{\degree C}, which corresponds to the mean summer (JJA), model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from North American Regional Reanalysis data \citep{data:narr} in a manner similar to \citet{seguinot-inpress}.
+The temperature standard deviation is a constant model parameter and was assigned a value of 3.07\,\unit{\degree C}, which corresponds to the mean summer (JJA), model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from North American Regional Reanalysis data \citep{data:narr} in a manner similar to \citet{seguinot-2013}.
 The ablation model incorporates degree-day factors of 3.04\,\unit{mm\,K^{-1}\,day^{-1}} for snow and 4.59\,\unit{mm\,K^{-1}\,day^{-1}} for ice, as derived from mass-balance measurements on contemporary glaciers from the Coast Mountains and Rocky Mountains in British Columbia \citep{shea-etal-2009}.

### 2013-11-18 floats/figures.tex
Qiong final comments

@@
 	\end{center}
-	\caption{Logarithmically coloured density maps, showing a comparison of summer (JJA) surface air temperature data from the WorldClim climatology, against that of each reanalysis.
+	\caption{Density maps showing a comparison of summer (JJA) surface air temperature data from the WorldClim climatology, against that of each reanalysis.
 Apparent horizontal lines are an artefact of different horizontal resolutions.
-Note the cold bias of NCEP/NCAR data.}
+Colour mapping is based on a logarythmic scale.
+Note the cold bias of NCEP/NCAR data relative to WorldClim data.}
 	\label{fig:tempheatmap}
@@
 	\end{center}
-	\caption{Logarithmically coloured density maps, showing a comparison of winter (DJF) precipitation rate from the WorldClim climatology, against that of each reanalysis.
+	\caption{Density maps showing a comparison of winter (DJF) precipitation rate from the WorldClim climatology, against that of each reanalysis.
 Apparent horizontal lines are an artefact of different horizontal resolutions.
-Apparent vertical lines at low values are an artefact of WorldClim data precision.
-Note the wet bias of all reanalysis data.}
+Apparent vertical lines at low values are an artefact of the low precision of WorldClim data.
+Colour mapping is based on a logarythmic scale.
+Note the wet bias of all reanalysis data relative to WorldClim data.}
 	\label{fig:precheatmap}
@@
 	\caption{Left panel: model sensitivity to simulation length.
-Modelled glaciated area, using NARR forcing data, and temperature offsets from 0 to 15\,\unit{\degree C}, solid lines corresponding to values from 7 to 11\,\unit{\degree C} (left).
+Modelled glaciated area, using NARR forcing data, and temperature offsets from 0 to 15\,\unit{\degree C}, solid lines corresponding to values from 7 to 11\,\unit{\degree C}.
 Right panel: modelled ice margin corresponding to temperature offsets from 7 to 11\,\unit{\degree C} when total glaciated area reach the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$.

### 2013-11-18 figures/plot.py

@@ -436,7 +436,7 @@ def biatmbars():
     # plot bars
     barkwargs = {'linewidth': 0.2, 'alpha': 0.5}
     ax.bar(pos, both-ref, w, color='#333333',
-      label='cumulative', **barkwargs)
+      label='both effects', **barkwargs)
     ax.bar(pos-w/2, temp-ref, w, color='#ff3333',
       label='temperature', **barkwargs)
     ax.bar(pos+w/2, prec-ref, w, color='#3333ff',

### 2013-11-18 sections/climate.tex
Update location map
Add geographical names and update colormap.

@@
 
-The spatial distribution of JJA and DJF mean precipitation rates in WorldClim reveals a strong precipitation decline beyond the coastal regions (Fig.~\ref{fig:prec}), primarily as a result of the continuous orographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascades.
+The spatial distribution of JJA and DJF mean precipitation rates in WorldClim reveals a strong precipitation decline beyond the coastal regions (Fig.~\ref{fig:prec}), primarily as a result of the continuous orographic barrier formed by the Wrangell and St.-Elias mountains, the Coast Mountains and the Cascades (Fig.~\ref{fig:locmap}).
 A less pronounced precipitation peak is also visible over the interior ranges.

### 2013-11-18 sections/model.tex
Add reference for geothermal heat flux

@@
 Enthalpy is computed in three dimensions in up to 51 irregularly spaced layers within the ice, and temperature is additionally computed in 11 regularly spaced layers in bedrock to a depth of 1\,km.
-A uniform geothermal heat flux of 70\,\unit{W\,m^{-2}} provides the lower boundary condition to the bedrock thermal model, and surface air temperature from the climate forcing provides the upper boundary condition to the ice enthalpy model.
+A uniform geothermal heat flux of 70\,\unit{W\,m^{-2}}, representative of the thick crust underlying the study area \citep[e.g.,][]{blackwell-richards-2004}, provides the lower boundary condition to the bedrock thermal model, and surface air temperature from the climate forcing provides the upper boundary condition to the ice enthalpy model.
 

### 2013-11-18 sections/climate.tex
Some small changes for homogeneity

@@
 
-Our atmospheric forcing consists of monthly climatologies of surface air temperature and precipitation obtained from one observational dataset, three global reanalyses and one regional reanalysis.
+Our climate forcing consists of monthly climatologies of surface air temperature and precipitation obtained from one observational dataset, three global reanalyses and one regional reanalysis.
 

### 2013-11-18 sections/results.tex
Final comments from Arjen
Some small changes for homogeneity

@@
 Using WorldClim, NCEP/NCAR, ERA-Interim, CFSR (smoothed and original datasets) and NARR input climatologies as climate forcing, we run 66~simulations of glacial inception and growth of the Cordilleran ice sheet over 10~ka using temperature offsets ranging from 0 to 10\,\unit{\degree C}.
-Significant differences in patterns of ice sheet growth, final glacial extent, and volume are attained (Fig.
+Significant differences in patterns of ice sheet growth, final glacial extent, and volume are attained (Figs.
 \ref{fig:ivolarea}-\ref{fig:cool05}).
@@
 
-Under present climate (0\,\unit{\degree C} offset), WorldClim, NARR and ERA-Interim forcing produce local ice caps in areas presently glaciated (Fig.~\ref{fig:extent}).
+Under present climate (0\,\unit{\degree C} offset), WorldClim, ERA-Interim and NARR forcing produce local ice caps in areas presently glaciated (Fig.~\ref{fig:extent}).
 These results are fairly consistent with contemporary ice distribution, although both the 10~km horizontal resolution and the occurrence of glaciers and ice caps surface topographies in the ETOPO1 dataset used for basal topography prevent a meaningful detailed comparison.

### 2013-11-19 cordillera-climate.tex
General footnotes clear-up
Final comments from Arjen
Add acknowledgements

@@
 
-% To-do footnotes
-\newcommand{\needref}[0]{\footnote{Reference needed}}
-\newcommand{\julien}[1]{\footnote{\textbf{Julien:} #1}}
-\newcommand{\arjen}[1]{\footnote{\textbf{Arjen:} #1}}
-\newcommand{\qiong}[1]{\footnote{\textbf{Qiong:} #1}}
-\newcommand{\irina}[1]{\footnote{\textbf{Irina:} #1}}
-
 % Set figures repository
@@
 \author[3]{Irina Rogozhina}
-\author[1]{Arjen Stroeven}
+\author[1]{Arjen P. Stroeven}
 \author[1]{Qiong Zhang}
@@
 % Acknowledgements
-%\begin{acknowledgements}
-%\end{acknowledgements}
+\begin{acknowledgements}
+We thank Ed Bueler and Andy Aschwanden for providing constant support on PISM and patiently answering our questions on the code internals, Laurent Brodeau and Soon-Heum Ko for assistance concerning software installations on supercomputing resources, and all PISM authors for providing the model open-source.
+Funding was provided by the Swedish Research Council (VR) to A.
+Stroeven (No.
+2008-3449) and Stockholm University.
+Computer resources were provided by the Swedish National Infrastructure for Computing (SNIC) at the National Supercomputing Center (NSC) to A.
+Stroeven (No.
+2013/1-145).
+
+% Author contributions
+\textit{Author contributions.}
+J.
+Seguinot ran the simulations and wrote most of the paper; C.
+Khroulev implemented necessary code changes in PISM; I.
+Rogozhina helped with model set-up and experiment design; A.
+Stroeven initiated the Cordilleran Ice Sheet modelling project; and Q.
+Zhang prepared some of the forcing data.
+All authors contributed to the redaction.
+\end{acknowledgements}
 

### 2013-11-20 sections/model.tex
Some small changes for homogeneity
Change geothermal reference (Irina)

--- 2013-11-18 sections/model.tex
@@
 Enthalpy is computed in three dimensions in up to 51 irregularly spaced layers within the ice, and temperature is additionally computed in 11 regularly spaced layers in bedrock to a depth of 1\,km.
-A uniform geothermal heat flux of 70\,\unit{W\,m^{-2}} provides the lower boundary condition to the bedrock thermal model, and surface air temperature from the climate forcing provides the upper boundary condition to the ice enthalpy model.
+A uniform geothermal heat flux of 70\,\unit{W\,m^{-2}}, characteristic of active tectonic margins on Earth \citep[e.g.,][]{artemieva-mooney-2001}, provides the lower boundary condition to the bedrock thermal model, and surface air temperature from the climate forcing provides the upper boundary condition to the ice enthalpy model.
 
@@
 
-\subsection{Atmospheric corrections}
+\subsection{Air temperature corrections}
 

### 2013-11-20 cordillera-climate.tex
Update acknowledgements (Arjen and Constantine)

@@
 We thank Ed Bueler and Andy Aschwanden for providing constant support on PISM and patiently answering our questions on the code internals, Laurent Brodeau and Soon-Heum Ko for assistance concerning software installations on supercomputing resources, and all PISM authors for providing the model open-source.
-Funding was provided by the Swedish Research Council (VR) to A.
-Stroeven (No.
-2008-3449) and Stockholm University.
-Computer resources were provided by the Swedish National Infrastructure for Computing (SNIC) at the National Supercomputing Center (NSC) to A.
-Stroeven (No.
-2013/1-145).
+This work was supported by the Swedish Research Council (VR) grant \#2008-3449 to A.
+P.
+Stroeven and by NASA grant \#NNX13AM16G to E.
+Bueler.
+Computer resources were provided by the Swedish National Infrastructure for Computing (SNIC) allocation \#2013/1-145 to A.
+P.
+Stroeven at the National Supercomputing Center (NSC).
 
@@
 Rogozhina helped with model set-up and experiment design; A.
+P.
 Stroeven initiated the Cordilleran Ice Sheet modelling project; and Q.
 Zhang prepared some of the forcing data.
-All authors contributed to the redaction.
+All authors contributed to paper writing.
 \end{acknowledgements}

### 2013-12-05 abstract.tex
Add separate abstract (needed for submission)

\documentclass[12pt]{article}
\usepackage[margin=25mm]{geometry}
\renewcommand{\familydefault}{\sfdefault}
\setlength\parindent{0pt}
\parskip=2ex
\thispagestyle{empty}
\begin{document}

\textbf{The effect of climate forcing on numerical simulations of the Cordilleran Ice Sheet at the Last Glacial Maximum}

J. Seguinot\textsuperscript{1},
C. Khroulev\textsuperscript{2},
I. Rogozhina\textsuperscript{3},
A. P. Stroeven\textsuperscript{1},
and Q. Zhang\textsuperscript{1}

\textsuperscript{1}Department of Physical Geography and Quaternary Geology, Stockholm University, Stockholm, Sweden \\
\textsuperscript{2}Geophysical Institute, University of Alaska Fairbanks, Fairbanks, AK, USA \\
\textsuperscript{3}Helmholtz Centre Potsdam, GFZ German Research Centre for Geosciences, Potsdam, Germany

\input{sections/abstract.tex}

\end{document}

### 2013-12-05 reviews/initial-response.tex
Address minor editor comments

\documentclass{letter}

\signature{Julien Seguinot}
\address{
  julien.seguinot@natgeo.su.se\\\\
  Stockholm University\\
  SE-106 91 Stockholm\\
  SWEDEN}

\usepackage[utf8]{inputenc}
\usepackage{geometry}
\usepackage{xcolor}
%\usepackage{enumitem}
%\usepackage{letterbib}

\newcommand{\rev}[0]{\color{blue!50!black}\it}
\newcommand{\textrev}[1]{{\rev``#1''}}
\newcommand{\revpoint}[1]{{\rev\item``#1''}}
\newcommand{\todo}[1]{\textcolor{red!50!black}{Todo: #1}}
\newcommand{\done}[1]{\textcolor{green!50!black}{Done: #1}}

% ----------------------------------------------------------------------

\begin{document}
\begin{letter}{Reply to Editor Initial Decision}

\opening{Dear Editor,}

Many thanks for accepting our paper for publication in TCD. We address your comments in the updated manuscript and detail the changes made below.

% ----------------------------------------------------------------------

\begin{itemize}

\revpoint{Line 43: please rephrase this sentence: 'other side of the coin' and 'hand-in-hand' into something of higher scientific value.}

This sentence was rephrased to: 'However, avoiding this circular dependence requires simplifying assumptions regarding Earth's past climate.'

\revpoint{Line 47: idem: Standing on middle ground is not a good beginning of the sentence. What is actually meant by this. Be more precise.}

Rephrased to 'Alternatively.'

\revpoint{Line 59-60: remove 'In our study'. just start sentence by: 'To limit the degrees of freedom ...'}

Rephrased accordingly.

\revpoint{Line 85: remove 'Thereby'. Just start with 'We aim to determine ...'}

Rephrased accordingly.

\revpoint{Line 87-88: Remove this sentence. It may be the first study to focus on the Cordilleran ice sheet since Robert (1991), but isn't that at the same time a weakness, since the model is not taking into account the buttressing effect of the Laurentide, which is done in other model studies that take into account the whole North American continent?}

This sentence was removed.

\revpoint{Line 113-116: Is this realistic to take a constant value of 70 $mW/m^2$? Aren't there direct measurements in the region to support this. Since it is an active margin, there could well be quite some spatial variability. Especially since one of the co-authors (I. Roghozina) is specialized in geothermal heat flow underneath ice sheets. Some evidence as to the choice of this value should be given as well as the eventual consequences of this choice.}

\todo{Develop this point.}

\revpoint{Line 199-200: 'Summer temperature and winter precipitation are most relevant to the glacier model as they drive summer melt and winter accumulation, respectively'.}

Moved 'respectively' to the end of the sentence.

\revpoint{Line 206: 're-projected to the Canadian ...'}

Added 'the'.

\revpoint{Line 216: spatially-smoothed}

Added a dash.

\revpoint{Line 276: However,}

Added a coma.

\revpoint{Line 284: to allow for comparison}

Added 'for'.

\revpoint{Line 294: compare modelled ice sheet ...}

Removed 'numerically.'

\revpoint{Line 362-367: This is a simple parameterization in Janssens and Huybrechts. If it would make such a difference, what is the rationale behind the fact that this has not been done? Please state.}

\todo{Develop this point.}

\revpoint{Figures: would it be possible to draw the limits of the Cordilleran ice sheets (as shown in Fig.~1) to the figures that show the simulated ice extents? It would make visual inspection easier.}

\todo{Develop this point.}

\end{itemize}

% ----------------------------------------------------------------------

\closing{Best regards,}

\end{letter}
\end{document}

### 2013-12-05 sections/climate.tex
Address minor editor comments

@@
 The spatial distributions of JJA air surface temperatures and DJF precipitation rates from the four reanalysis climatologies are shown in Figures~\ref{fig:temp} and~\ref{fig:prec}.
-Summer temperature and winter precipitation are most relevant to the glacier model as they respectively drive summer melt and winter accumulation.
+Summer temperature and winter precipitation are most relevant to the glacier model as they drive summer melt and winter accumulation, respectively.
 Because reanalyses include observational information through data assimilation, they are closer to observations than pure GCMs in densely monitored regions, such as the southern part of our modelling domain \citep{bengtsson-etal-2007}.
@@
 
-WorldClim, NCEP/NCAR, ERA-Interim, CFSR and NARR climatologies were re-projected to Canadian Atlas Lambert conformal conic projection (EPSG code~3978) and bilinearly interpolated to the 10\,km-resolution model grid using the module r.proj from GRASS~GIS \citep{soft:grass}.
+WorldClim, NCEP/NCAR, ERA-Interim, CFSR and NARR climatologies were re-projected to the Canadian Atlas Lambert conformal conic projection (EPSG code~3978) and bilinearly interpolated to the 10\,km-resolution model grid using the module r.proj from GRASS~GIS \citep{soft:grass}.
 In addition, WorldClim data was extrapolated to cover grid points in the ocean using the nearest-neighbour algorithm \citep{soft:scipy}.
@@
 The CFSR climatology presents wave-like artefacts in its precipitation field (Fig.~\ref{fig:prec}).
-These are a common feature in GCM simulations due to the failure of spectral models to resolve the very local nature of topography-induced precipitation events, which becomes particularly visible in the high-resolution, less spatially smoothed data.
+These are a common feature in GCM simulations due to the failure of spectral models to resolve the very local nature of topography-induced precipitation events, which becomes particularly visible in the high-resolution, less spatially-smoothed data.
 Therefore, an alternative forcing was prepared for CFSR by smoothing its precipitation field.

### 2013-12-05 sections/discuss.tex
Address minor editor comments

@@
 In a model of coarser resolution, these high mountain ranges are reduced to smooth hills and the modelled distribution of precipitation is also smoother (Fig.~\ref{fig:oroprecip}).
-However this effect alone cannot explain the widespread and pronounced positive precipitation anomalies observed in CFSR data (Figs.~\ref{fig:precheatmap}--\ref{fig:precdiff}) despite of its high spatial resolution (Table~\ref{tab:reanalyses}).
+However, this effect alone cannot explain the widespread and pronounced positive precipitation anomalies observed in CFSR data (Figs.~\ref{fig:precheatmap}--\ref{fig:precdiff}) despite of its high spatial resolution (Table~\ref{tab:reanalyses}).
 
@@
 A single temperature offset of 5\,\unit{\degree C} is used.
-This value was chosen to allow comparison with previous results (Fig.~\ref{fig:cool05}).
+This value was chosen to allow for comparison with previous results (Fig.~\ref{fig:cool05}).
 In each case, temperature lapse rates are computed using the reference topography from the corresponding dataset.
@@
 
-We compare numerically modelled ice sheet geometries to a reconstruction of the ice margin at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) by \citet{dyke-2004}, based on glacial geomorphology and radiocarbon dating (Fig.~\ref{fig:locmap}).
+We compare modelled ice sheet geometries to a reconstruction of the ice margin at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) by \citet{dyke-2004}, based on glacial geomorphology and radiocarbon dating (Fig.~\ref{fig:locmap}).
 This corresponds to the LGM extent of most of the Cordilleran ice sheet, which occurred later than for the Laurentide ice sheet \citep{porter-swanson-1998,dyke-2004,stroeven-etal-2010,stroeven-etal-inpress}.

### 2013-12-05 sections/intro.tex
Address minor editor comments

@@
 
-The other side of the coin is that avoiding this circular dependence unfortunately goes hand-in-hand with a need for simplifying assumptions regarding Earth's past climate.
+However, avoiding this circular dependence requires simplifying assumptions regarding Earth's past climate.
 Such studies include energy balance modelling approaches \citep{tarasov-peltier-1997} and geographic parametrizations of surface mass balance \citep{robert-1991} or climate forcing \citep{johnson-fastook-2002}.
-Standing on a middle ground, temperature offset methods \citep{greve-etal-1999,bintanja-etal-2005} make use of the high level of detail available in present climate datasets such as gridded observation datasets, GCM output or reanalyses, while using simplifying representations of past climate deviations from the present state.
+Alternatively, temperature offset methods \citep{greve-etal-1999,bintanja-etal-2005} make use of the high level of detail available in present climate datasets such as gridded observation datasets, GCM output or reanalyses, while using simplifying representations of past climate deviations from the present state.
 
@@
 Although using different set-ups, all three studies demonstrate the very large sensitivity of ice sheet models to the choice of climate forcing data used.
-In our study, to limit the degrees of freedom in our model and obtain results independent of palaeo-ice sheet reconstructions such as ICE-4G, we use a simple temperature offset similar to the approaches by \citet{greve-etal-1999} and \citet{bintanja-etal-2005} and assess ice sheet model sensitivity to the choice of present-day climate data.
+To limit the degrees of freedom in our model and obtain results independent of palaeo-ice sheet reconstructions such as ICE-4G, we use a simple temperature offset similar to the approaches by \citet{greve-etal-1999} and \citet{bintanja-etal-2005} and assess ice sheet model sensitivity to the choice of present-day climate data.
 Hence, rather than using GCM output, we force our model with climate reanalysis data, which builds on observational information through data assimilation \citep{bengtsson-etal-2007}.
@@
 We force our model with multiple climate datasets and compare the modelled ice extent to a geomorphological reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.
-Thereby, we aim to determine the climate dataset which is most suited for simulation of the Cordilleran Ice Sheet and to use as input for future, transient studies over a glacial cycle.
-To our knowledge, this is the first modelling study that specifically focuses on the Cordilleran Ice Sheet since \citet{robert-1991}.
+We aim to determine the climate dataset which is most suited for simulation of the Cordilleran Ice Sheet and to use as input for future, transient studies over a glacial cycle.

### 2013-12-09 floats/figures.tex
Update location map
Add geographical names and update colormap.
Address editor comment on figures

@@
 The rectangular box denotes the modelling domain of the Cordilleran ice sheet of 1500 by 3000\,km.
+Major mountain ranges covered by the ice sheet include the Wrangell and St.-Elias mountains (W-SE), the Selwyn and MacKenzie mountains (S-MK), the Coast Mountains and the Rocky Mountains.
 The background map consists of ETOPO1 \citep{data:etopo1} and Natural Earth Data \citep{data:naturalearth} and was assembled with GRASS~GIS \citep{soft:grass}.}
@@
 	\end{center}
-	\caption{Extent of ice cover after 10\,ka as a function of applied temperature offsets for each climate forcing used.}
+	\caption{Extent of ice cover after 10\,ka as a function of applied temperature offsets for each climate forcing used.
+A reconstructed LGM ice sheet margin by \citet{dyke-2004} (Fig.~\ref{fig:locmap}) is given for reference (black line).}
 	\label{fig:extent}

### 2013-12-09 reviews/initial-response.tex
Address editor comment on figures

@@
 \usepackage{xcolor}
+\usepackage{graphicx}
 %\usepackage{enumitem}
 %\usepackage{letterbib}
+\graphicspath{{../figures/}}
 
+\newcommand{\unit}[1]{\ensuremath{\mathrm{#1}}}
+\newcommand{\degree}[0]{\ensuremath{^{\circ}}}
 \newcommand{\rev}[0]{\color{blue!50!black}\it}
@@
 
-\todo{Develop this point.}
+The LGM ice margin reconstruction visible on Figs.~1 and~15 was added to Fig.~6, which shows the extent of modelled ice cover for each climate forcing used.
+However we did not add it to Figs.~7 and 13 as we felt the resulting figures contained too much information (Fig.~a).
 
 \end{itemize}
+
+\begin{center}
+	\includegraphics{cordillera-climate-cool05+lgm.pdf}\\
+	\textbf{Fig. a.} Paper figure~7 with added LGM ice margin, not included in the new version.
+\end{center}
 

### 2013-12-09 figures/plot.py

@@ -379,7 +379,7 @@ def bestdiff():
     cb.set_label(u'ice surface elevation difference to %s %s K simulation (m)' % (labels[climates[0]], offsets[0]))
     _savefig('cordillera-climate-bestdiff')
 
-def biatm():
+def biatm(draw_lgm_on_icemap=False):
     """Plot hybrid atmosphere output"""
 
     # initialize figure
@@ -395,10 +395,12 @@ def biatm():
         nc = Dataset('../data/%s-05.nc' % biclim)
         im = iplt.icemap(nc)
 
-    # add colorbar and save
+    # add LGM margin, colorbar and save
+    if draw_lgm_on_icemap: _drawlgm(nc, fig.grid, ec='b', fc='none')
     cb = fig.colorbar(im, fig.grid.cbar_axes[0])
     cb.set_label(r'ice surface velocity ($m\,a^{-1}$)')
-    _savefig('cordillera-climate-biatm')
+    suffix = ('+lgm' if draw_lgm_on_icemap else '')
+    _savefig('cordillera-climate-biatm' + suffix)
 
 def biatmbars():
     """Plot hybrid atmosphere bar chart"""
@@ -458,7 +460,7 @@ def biatmbars():
     # save
     _savefig('cordillera-climate-biatmbars')
 
-def cool(cool):
+def cool(cool, draw_lgm_on_icemap=False):
     """Plot icemaps for given temperature offset"""
 
     # initialize figure
@@ -473,10 +475,12 @@ def cool(cool):
       _annotate(labels[clim])
       im = iplt.icemap(nc)
 
-    # add colorbar and save
+    # add LGM margin, colorbar and save
+    if draw_lgm_on_icemap: _drawlgm(nc, fig.grid, ec='b', fc='none')
     cb = fig.colorbar(im, fig.grid.cbar_axes[0])
     cb.set_label(r'ice surface velocity ($m\,a^{-1}$)')
-    _savefig('cordillera-climate-cool' + cool)
+    suffix = ('+lgm' if draw_lgm_on_icemap else '')
+    _savefig('cordillera-climate-cool' + cool + suffix)
 
 def duration():
     """Plot effect of glaciation duration on multiple maps"""
@@ -595,7 +599,8 @@ def extent():
         levels = range(-1, 11),
         cmap   = plt.cm.Spectral)
 
-    # add colorbar and save
+    # add LGM ice margin, colorbar and save
[…]

### 2013-12-09 reviews/initial-response.tex
Adress editor comment on geothermal heat flux

@@
 
-\todo{Develop this point.}
+Thank you for raising this point.
+A justification was added in the manuscript.
+In this study, we focus mainly on glacier extent, on which geothermal heat flux is not the major control.
+However we will be happy to add a sensitivity experiment to the paper if this point is risen again at the discussion stage.
 

### 2013-12-09 sections/model.tex
Adress editor comment on geothermal heat flux

--- 2013-11-18 sections/model.tex
@@
 Enthalpy is computed in three dimensions in up to 51 irregularly spaced layers within the ice, and temperature is additionally computed in 11 regularly spaced layers in bedrock to a depth of 1\,km.
-A uniform geothermal heat flux of 70\,\unit{W\,m^{-2}} provides the lower boundary condition to the bedrock thermal model, and surface air temperature from the climate forcing provides the upper boundary condition to the ice enthalpy model.
+Surface air temperature from the climate forcing provides the upper boundary condition to the ice enthalpy model, and a geothermal heat flux of 70\,\unit{mW\,m^{-2}} provides the lower boundary condition to the bedrock thermal model.
+Although this uniform value does not account for the high spatial geothermal variability of the region \citep{blackwell-richards-2004}, it is on average representative of available heat flow measurements \citep{artemieva-mooney-2001,blackwell-richards-2004}.
+In addition, whereas spatial variations of geothermal heat flux may significantly affect basal conditions of ice sheets \citep{pattyn-2010}, this study focuses on reconstructing the former extent of ice cover, on which it has relatively little effect \citep{rogozhina-etal-2012}.
 

### 2013-12-09 reviews/initial-response.tex
Adress editor comment on refreezing

@@
 In this study, we focus mainly on glacier extent, on which geothermal heat flux is not the major control.
-However we will be happy to add a sensitivity experiment to the paper if this point is risen again at the discussion stage.
+However we are currently working on improving this point and will be happy to add a sensitivity experiment to the paper if this point is risen again at the discussion stage.
 
@@
 
-\todo{Develop this point.}
+There exist different refreezing parametrizations in the literature.
+We developped this point in the manuscript.
+As we have not assessed which parametrization scheme, if any, would be appropriate in our set-up, we preferred to disregard refreezing rather than adding a new source of uncertainty in our model.
+Again, we could consider adding a sensitivity experiment at the review stage if you think this is needed.
 

### 2013-12-09 sections/discuss.tex
Adress editor comment on refreezing

@@
 
-Furthermore, some simplifications were made in the surface mass balance model.
-Our PDD model does not include refreezing.
-Refreezing of melted snow and ice and retention in the snow pack, however, can greatly alter mass balance at the surface of an ice sheet \citep{janssens-huybrechts-2000}.
-Additionally, we simulate temperature variability by using a constant, uniform value of daily temperature standard deviation.
+Furthermore, the surface mass balance model crudely simplifies of processes taking place at the glacier surface.
+Refreezing of melted snow and ice and retention in the snow pack was shown to alter surface mass balance of glaciers \citep{trabant-mayo-1985}, yet it is not accounted for in our PDD model.
+Based on scarce observations over the Greenland Ice Sheet, different parametrizations of this effect have been developped \citep{janssens-huybrechts-2000,fausto-etal-2009b}.
+However, these parametrizations were shown to disagree with more complex models \citep{reijmer-etal-2012}, and their portability to a different region and time period has not been proven.
+In addition, we simulate temperature variability by using a constant, uniform value of daily temperature standard deviation.
 This approach implies large biases of surface mass-balance \citep{charbit-etal-2013,rau-rogozhina-2013,seguinot-2013}, particularly over regions with such various climates as experienced within our model domain.

### 2013-12-09 sections/intro.tex
Adress editor comment on refreezing

@@
 This is not entirely a new approach as a few studies of this kind are presently available.
-\citet{quiquet-etal-2012}, for example, assessed the sensitivity of a Greenland ice sheet model to various climate forcing fields, including a regional parametrization by \citet{fausto-etal-2009}, output from several GCMs and an atmospheric reanalysis.
+\citet{quiquet-etal-2012}, for example, assessed the sensitivity of a Greenland ice sheet model to various climate forcing fields, including a regional parametrization by \citet{fausto-etal-2009a}, output from several GCMs and an atmospheric reanalysis.
 \citet{rodgers-etal-2004} and \citet{charbit-etal-2007} tested the sensitivity of a model of Northern Hemisphere ice sheets to climate forcing from different PMIP LGM simulations.

### 2013-12-09 reviews/initial-response.tex
Small corrections (Irina)

--- 2013-12-09 reviews/initial-response.tex
@@
 
-\opening{Dear Editor,}
+\opening{Dear Frank Pattyn,}
 
@@
 % ----------------------------------------------------------------------
-
 \begin{itemize}
@@
 In this study, we focus mainly on glacier extent, on which geothermal heat flux is not the major control.
-However we will be happy to add a sensitivity experiment to the paper if this point is risen again at the discussion stage.
+However we are currently working on an improved geothermal forcing which will be included in transient simulations of the Cordilleran Ice Sheet in the future.
 
@@
 
-\todo{Develop this point.}
+There exist different refreezing parametrizations in the literature.
+We developed this point in the manuscript.
+As we have not assessed which parametrization scheme, if any, would be appropriate for our set-up, we preferred to disregard refreezing rather than adding a new source of uncertainty in our model.
 
@@
 The LGM ice margin reconstruction visible on Figs.~1 and~15 was added to Fig.~6, which shows the extent of modelled ice cover for each climate forcing used.
-However we did not add it to Figs.~7 and 13 as we felt the resulting figures contained too much information (Fig.~a).
+However we did not add it to Figs.~7 and 13 as we felt that the resulting figures contained too much information (Fig.~a).
 

### 2013-12-09 sections/discuss.tex
Small corrections (Irina)

@@
 
-Furthermore, the surface mass balance model crudely simplifies of processes taking place at the glacier surface.
+Furthermore, the surface mass balance model crudely simplifies the processes taking place at the glacier surface.
 Refreezing of melted snow and ice and retention in the snow pack was shown to alter surface mass balance of glaciers \citep{trabant-mayo-1985}, yet it is not accounted for in our PDD model.
-Based on scarce observations over the Greenland Ice Sheet, different parametrizations of this effect have been developped \citep{janssens-huybrechts-2000,fausto-etal-2009b}.
+Based on scarce observations over the Greenland Ice Sheet, different parametrizations of this effect have been developed \citep{janssens-huybrechts-2000,fausto-etal-2009b}.
 However, these parametrizations were shown to disagree with more complex models \citep{reijmer-etal-2012}, and their portability to a different region and time period has not been proven.

### 2013-12-09 sections/model.tex
Small corrections (Irina)

@@
 Surface air temperature from the climate forcing provides the upper boundary condition to the ice enthalpy model, and a geothermal heat flux of 70\,\unit{mW\,m^{-2}} provides the lower boundary condition to the bedrock thermal model.
-Although this uniform value does not account for the high spatial geothermal variability of the region \citep{blackwell-richards-2004}, it is on average representative of available heat flow measurements \citep{artemieva-mooney-2001,blackwell-richards-2004}.
-In addition, whereas spatial variations of geothermal heat flux may significantly affect basal conditions of ice sheets \citep{pattyn-2010}, this study focuses on reconstructing the former extent of ice cover, on which it has relatively little effect \citep{rogozhina-etal-2012}.
+Although this uniform value does not account for the high spatial geothermal variability in the region, it is on average representative of available heat flow measurements \citep{artemieva-mooney-2001,blackwell-richards-2004}.
+Whereas spatial variations in geothermal heat flux may significantly affect basal conditions of ice sheets \citep{pattyn-2010}, this study focuses on reconstructing the former extent of ice cover, on which it has relatively little effect \citep{rogozhina-etal-2012}.
 

### 2013-12-11 sections/discuss.tex
Comments from Arjen and Constantine

@@
 Furthermore, the surface mass balance model crudely simplifies the processes taking place at the glacier surface.
-Refreezing of melted snow and ice and retention in the snow pack was shown to alter surface mass balance of glaciers \citep{trabant-mayo-1985}, yet it is not accounted for in our PDD model.
+Refreezing of melted snow and ice and water retention in the snow pack was shown to influence surface mass balance of glaciers \citep{trabant-mayo-1985}, yet it is not accounted for in our PDD model.
 Based on scarce observations over the Greenland Ice Sheet, different parametrizations of this effect have been developed \citep{janssens-huybrechts-2000,fausto-etal-2009b}.
-However, these parametrizations were shown to disagree with more complex models \citep{reijmer-etal-2012}, and their portability to a different region and time period has not been proven.
+However, these parametrizations were shown to disagree with more complex models \citep{reijmer-etal-2012}, and it remains undemonstrated whether their application to a different region and time period is sound.
 In addition, we simulate temperature variability by using a constant, uniform value of daily temperature standard deviation.

### 2013-12-11 sections/model.tex
Comments from Arjen and Constantine

@@
 Although this uniform value does not account for the high spatial geothermal variability in the region, it is on average representative of available heat flow measurements \citep{artemieva-mooney-2001,blackwell-richards-2004}.
-Whereas spatial variations in geothermal heat flux may significantly affect basal conditions of ice sheets \citep{pattyn-2010}, this study focuses on reconstructing the former extent of ice cover, on which it has relatively little effect \citep{rogozhina-etal-2012}.
+Whereas spatial variations in the geothermal heat flux may significantly affect basal conditions \citep{pattyn-2010}, this study focuses on reconstructing the former extent of the Cordilleran Ice Sheet, on which it has relatively little effect \citep{rogozhina-etal-2012}.
 

### 2014-03-19 cordillera-climate.tex
Flatten sections (no actual changes)

@@
 
-% Sections
-\input{sections/abstract}
-\input{sections/intro}
-\input{sections/model}
-\input{sections/climate}
-\input{sections/results}
-\input{sections/discuss}
-\input{sections/concl}
+% Abstract
+\begin{abstract}
+We present an ensemble of numerical simulations of the Cordilleran Ice Sheet during the Last Glacial Maximum performed with the Parallel Ice Sheet Model (PISM), applying temperature offsets to the present-day climatologies from five different datasets.
+Monthly mean surface air temperature and precipitation from WorldClim, the NCEP/NCAR reanalysis, the ERA-Interim reanalysis, the Climate Forecast System Reanalysis and the North American Regional Reanalysis are used to compute surface mass balance in a positive degree-day model.
+Modelled ice sheet outlines and volumes appear highly sensitive to the choice of climate forcing.
+For three of the four reanalysis datasets used, differences in precipitation are the major source for discrepancies between model results.
+We assess model performance against a geomorphological reconstruction of the ice margin at the Last Glacial Maximum, and suggest that part of the mismatch is due to unresolved orographic precipitation effects caused by the coarse resolution of reanalysis data.
+The best match between model output and the reconstructed ice margin is obtained using the high-resolution North American Regional Reanalysis, which we retain for simulations of the Cordilleran Ice Sheet in the future.
+\end{abstract}
+
+% ----------------------------------------------------------------------
+\introduction
+\label{sec:intro}
+% ----------------------------------------------------------------------
+
+At the Last Glacial Maximum (LGM), glaciers of a size comparable to the present Greenland and Antarctic ice sheets covered parts of Northern America (Laurentide, Cordilleran and Innuitian ice sheets) and northern Eurasia (Fennoscandian Ice Sheet).
+Numerical modelling of these former ice masses allows for a comparison between glaciological theories embedded in the models and geomorphological traces underpinning palaeo-glaciological reconstructions.
+Yet, a major obstacle in this exercise resides in large uncertainties concerning the climate forcing, typically atmospheric temperature and precipitation, required as input to numerical glacier models \citep{hebeler-etal-2008}.
+This includes uncertainty in the representation of Earth's present climate in regions of poor station coverage, and even larger uncertainty concerning accurate reconstructions of past climate change.
+
+Arguably the most physically sound way to force an ice sheet model for simulations of glacial history is to couple it with a General Circulation Model \citep[GCM;][]{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013}.
+However the computational demand of GCMs is such that only models of intermediate complexity can run on the time-scales of tens of thousands of years characteristic of ice sheet growth and decay.
+
+Climatologies obtained from GCM palaeo-climate simulations such as produced within the Paleoclimate Modelling Intercomparison Project \citep[PMIP;][]{joussaume-taylor-1995} provide perhaps a more reasonable representation of past climate.
+However, as such climatologies are only available for specific periods of time, using them as climate forcing for an ice sheet model requires either an assumption of steady-state ice sheet response to climatic fluctuations \citep{huybrechts-tsiobbel-1996}, or interpolation through time between climatologies from different periods, which can be linear \citep{charbit-etal-2002}, or modulated by a ``glacial index'' weighting function derived from ice core $\delta^{18}$O records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}.
+An important drawback in this approach is that GCM palaeo-climate simulations themselves rely on an accurate description of the global surface topography and therefore include global ice sheet reconstructions such as ICE-4G \citep{peltier-1994} for their surface topographic boundary condition.
+This condition could potentially exert influence on subsequently modelled ice sheet geometries, and particularly glacier extent.
+
+However, avoiding this circular dependence requires simplifying assumptions regarding Earth's past climate.
+Such studies include energy balance modelling approaches \citep{tarasov-peltier-1997} and geographic parametrizations of surface mass balance \citep{robert-1991} or climate forcing \citep{johnson-fastook-2002}.
+Alternatively, temperature offset methods \citep{greve-etal-1999,bintanja-etal-2005} make use of the high level of detail available in present climate datasets such as gridded observation datasets, GCM output or reanalyses, while using simplifying representations of past climate deviations from the present state.
+
+Here, we propose to address some of the uncertainties concerning climate forcing of numerical glacier models by evaluating the responses of a numerical model, in terms of glacier extent, to inputs from several climate datasets.
+This is not entirely a new approach as a few studies of this kind are presently available.
+\citet{quiquet-etal-2012}, for example, assessed the sensitivity of a Greenland ice sheet model to various climate forcing fields, including a regional parametrization by \citet{fausto-etal-2009a}, output from several GCMs and an atmospheric reanalysis.
+\citet{rodgers-etal-2004} and \citet{charbit-etal-2007} tested the sensitivity of a model of Northern Hemisphere ice sheets to climate forcing from different PMIP LGM simulations.
+Although using different set-ups, all three studies demonstrate the very large sensitivity of ice sheet models to the choice of climate forcing data used.
+To limit the degrees of freedom in our model and obtain results independent of palaeo-ice sheet reconstructions such as ICE-4G, we use a simple temperature offset similar to the approaches by \citet{greve-etal-1999} and \citet{bintanja-etal-2005} and assess ice sheet model sensitivity to the choice of present-day climate data.
+Hence, rather than using GCM output, we force our model with climate reanalysis data, which builds on observational information through data assimilation \citep{bengtsson-etal-2007}.
+Furthermore, we focus our study regionally on the former Cordilleran Ice Sheet in western North America.
+
+The Cordilleran Ice Sheet (Fig.~\ref{fig:locmap}) covered an area that presently experiences strong regional variations in climate.
+From a numerical modelling perspective, it is one of the least studied palaeo-ice sheets of the Northern Hemisphere, despite the fact that significant stratigraphical, geomorphological, and chronological data is available to constrain its extent \citep{jackson-clague-1991,porter-swanson-1998,dukrodkin-1999,dyke-2004,kaufman-manley-2004,kleman-etal-2010,stroeven-etal-2010,stroeven-etal-inpress,margold-etal-2011}.
+The Cordilleran Ice Sheet has previously been modelled as part of efforts to simulate ice sheets in North America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} and world-wide \citep{yoshimori-etal-2001}.
+While these studies generally reproduce the magnitude of North American glaciation at LGM reasonably well, there exists a tendency in the simulations that are independent of ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and interior Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.
+
+Here we use PISM, a Parallel Ice Sheet Model \citep{web:pism}, to simulate the extent and thickness of the Cordilleran Ice Sheet at the LGM (Fig.~\ref{fig:locmap}).
+We force our model with multiple climate datasets and compare the modelled ice extent to a geomorphological reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.
+We aim to determine the climate dataset which is most suited for simulation of the Cordilleran Ice Sheet and to use as input for future, transient studies over a glacial cycle.
+
+% ----------------------------------------------------------------------
+\section{Model setup}
+\label{sec:model}
+% ----------------------------------------------------------------------
+
+Given basal topography, sea level, geothermal heat flux and climate forcing, the model computes ice extent and thickness, its thermal and dynamic state, and the associated lithospheric response.
+Our modelling domain encompasses the entire area covered by the Cordilleran Ice Sheet at LGM including independent ice growing on the western Alaskan ranges and the Brooks Range and intermediate expanses of ice-free terrain in northern Yukon Territory and interior Alaska (Fig.~\ref{fig:locmap}).
+
+As we aim to model glacial inception and growth of the Cordilleran Ice Sheet towards a configuration as last attained during the LGM, we mimic palaeo-climatic conditions by applying constant temperature offsets homogeneously over the modelling domain.
+Each simulation starts from ice-free conditions and runs for 10\,ka, a time interval representative of the rapid build-up of the last Cordilleran Ice Sheet from nearly ice-free to full glacial conditions \citep{clague-1989,stroeven-etal-2010}.
+Our simulations are performed in parallel on 32 cores at the Swedish National Supercomputing Center.
+
+\subsection{Ice thermodynamics}
+
+The central part of an ice sheet model consists of the computation of flow velocity which itself depends on temperature.
+PISM is a shallow model, which implies that the balance of stresses is approximated based on their predominant components.
+On the other hand, the model is polythermal: it accounts for differences in temperature and softness within the ice column.
+
+The Shallow Shelf Approximation (SSA) is used as a ``sliding law'' for the Shallow Ice Approximation (SIA) \citep{bueler-brown-2009,winkelmann-etal-2011}.
+SIA and SSA velocities are computed by finite difference methods on a 10\,km-resolution horizontal grid of 300 by 150 points (the modelling domain).
+Ice softness depends on temperature and water content through an enthalpy formulation \citep{aschwanden-blatter-2009,aschwanden-etal-2012}.
+Enthalpy is computed in three dimensions in up to 51 irregularly spaced layers within the ice, and temperature is additionally computed in 11 regularly spaced layers in bedrock to a depth of 1\,km.
+Surface air temperature from the climate forcing provides the upper boundary condition to the ice enthalpy model, and a geothermal heat flux of 70\,\unit{mW\,m^{-2}} provides the lower boundary condition to the bedrock thermal model.
+Although this uniform value does not account for the high spatial geothermal variability in the region, it is on average representative of available heat flow measurements \citep{artemieva-mooney-2001,blackwell-richards-2004}.
+Whereas spatial variations in the geothermal heat flux may significantly affect basal conditions \citep{pattyn-2010}, this study focuses on reconstructing the former extent of the Cordilleran Ice Sheet, on which it has relatively little effect \citep{rogozhina-etal-2012}.
+
+A pseudo-plastic sliding law \citep{aschwanden-etal-2013} relates the bed-parallel shear stress and the sliding velocity.
+The yield stress is modelled using the Mohr-Coulomb criterion.
+The till friction angle $\phi$ varies from 10 to 30\degree\ and is a function of current bed elevation, with lowest values occurring at low elevations,
+
+\begin{equation}
+	\phi = \left\{\begin{array}{llrll}
+		10      & \mathrm{for} &      &z&<  0 \\
+		z/10+10 & \mathrm{for} &   0 <&z&<200 \\
+		30      & \mathrm{for} & 200 <&z&     \\
+	\end{array}\right.
+\end{equation}
+
+where $\phi$ is given in degrees and $z$ in meters above current sea level.
+This model thus accounts for a weakening of till associated with the presence of marine sediments \citep{martin-etal-2011,aschwanden-etal-2013}.
+Basal topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography and bathymetry dataset with a resolution of 1\,arc-minute \citep{data:etopo1}.
+Sea level is lowered by 120\,m and basal topography responds to ice load following a regional isostasy model that includes lithospheric flexure and mantle relaxation \citep{lingle-clark-1985}.
+
+\subsection{Surface mass-balance}
+
+Ice surface accumulation and ablation are computed from monthly mean surface air temperature and monthly precipitation by a temperature-index (positive degree-day) model \citep{hock-2003}.
+Ice accumulation is equal to precipitation when temperature is below 0\,\unit{\degree C}, and decreases to zero linearly with temperature between 0 and 2\,\unit{\degree C}.
+Ice ablation is computed from the number of positive degree-days, defined as the integral of temperatures above 0\,\unit{\degree C} in one year.
+
+The positive degree-day integral \citep{calov-greve-2005} is numerically approximated using week-long sub-intervals.
+It accounts for temperature variability assuming a normal distribution along a central (input) value.
+The temperature standard deviation is a constant model parameter and was assigned a value of 3.07\,\unit{\degree C}, which corresponds to the mean summer (JJA), model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from North American Regional Reanalysis data \citep{data:narr} in a manner similar to \citet{seguinot-2013}.
+The ablation model incorporates degree-day factors of 3.04\,\unit{mm\,K^{-1}\,day^{-1}} for snow and 4.59\,\unit{mm\,K^{-1}\,day^{-1}} for ice, as derived from mass-balance measurements on contemporary glaciers from the Coast Mountains and Rocky Mountains in British Columbia \citep{shea-etal-2009}.
+
+\subsection{Air temperature corrections}
+
+Prior to surface mass balance computation, the model dynamically applies a lapse-rate correction to surface air temperature.
+This correction accounts for the evolution of ice thickness on the one hand, and differences between the climate forcing reference topography and the ice flow model basal topography on the other hand.
+It uses a reference topography distinct from the modelled basal topography and specific to each climate forcing dataset (Fig.~\ref{fig:topo}).
+An annual lapse rate of 6\,\unit{\degree C\,km^{-1}} is used in all simulations.
+No corrections apply to precipitation changes with elevation.
+
+% ----------------------------------------------------------------------
+\section{Climate forcing}
+\label{sec:climate}
+% ----------------------------------------------------------------------
+
+Our climate forcing consists of monthly climatologies of surface air temperature and precipitation obtained from one observational dataset, three global reanalyses and one regional reanalysis.
+
+\subsection{Observational data: WorldClim}
+
+WorldClim is a high resolution global climate dataset built from meteorological observations \citep{data:worldclim}.
+It was built by applying a spatial interpolation between a large set of measurements taken at terrestrial weather stations worldwide, using SRTM \citep{data:srtm} and GTOPO30 elevation data \citep{data:gtopo30}.
+This dataset provides a resolution much higher than attained by circulation models, however its reliability is variable due to the highly inhomogeneous distribution of measurements \citep{data:worldclim}.
+
+Within our modelling domain, the spatial distributions of summer (JJA) and winter (DJF) air surface temperatures in WorldClim generally show a decrease of temperature from south to north (JJA) and from south-west to north-east (DJF), such that regions further inland experience colder winters (Fig.~\ref{fig:temp}).
+It should be noted that temperature gradients are much stronger in winter than in summer.
+Temperature rises well above zero during the summer months over the entire modelling domain, except for the highest mountain peaks, and even regions where mean annual temperatures are well below freezing point do experience warm summers.
+In other words, there is a strong contrast in seasonality between coastal temperate-climate and inland continental-climate regions.
+
+The spatial distribution of JJA and DJF mean precipitation rates in WorldClim reveals a strong precipitation decline beyond the coastal regions (Fig.~\ref{fig:prec}), primarily as a result of the continuous orographic barrier formed by the Wrangell and St.-Elias mountains, the Coast Mountains and the Cascades (Fig.~\ref{fig:locmap}).
+A less pronounced precipitation peak is also visible over the interior ranges.
+From an ice sheet mass balance point of view, these heterogeneities become even stronger due to the difference in timing of the precipitation peak through the year.
+Whereas coastal regions experience most precipitation as snow during the accumulation season (DJF), inland regions experience dry winters and most of the annual precipitation consequently falls as rain during the ablation season (JJA).
+
+In regions such as the northern Yukon Territory and interior Alaska, dry winters and warm summers prohibit ice to accumulate and glaciers to form, despite strongly negative mean annual temperatures.
+In order to account for these strong gradients in seasonality, we use monthly rather than annual averages of temperature and precipitation to drive the ice sheet model.
+
+There are two problems inherent to the use of WorldClim data in our study area.
+Firstly, the density of weather stations used by WorldClim in the Northern American Cordillera is highly inhomogeneous.
+Although good coverage exists for the southern part of our modelling domain, several hundred kilometres can separate neighbouring stations in the north \citep{data:worldclim}.
+Secondly, WorldClim lacks data offshore, which would have been particularly useful over the Pacific continental shelf which was partly covered by ice during the LGM \citep{jackson-clague-1991}.
+
+\subsection{Reanalysis data: NCEP/NCAR, ERA-Interim, CFSR and NARR}
+
+In addition to the WorldClim data, we use surface air temperatures and precipitation rates from three global atmospheric reanalyses and one regional atmospheric reanalysis to inform the mass balance routine of PISM: the NCEP/NCAR reanalysis \citep{data:ncar}, the ERA-Interim reanalysis \citep{data:erai}, the Climate Forecast System Reanalysis \citep[CFSR;][]{data:cfsr}, and the North American Regional Reanalysis \citep[NARR;][]{data:narr}.
+Monthly climatologies from NCEP/NCAR and NARR reanalyses were provided by the \citet{web:psd} whereas monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean time series (Table \ref{tab:reanalyses}).
+
+The spatial distributions of JJA air surface temperatures and DJF precipitation rates from the four reanalysis climatologies are shown in Figures~\ref{fig:temp} and~\ref{fig:prec}.
+Summer temperature and winter precipitation are most relevant to the glacier model as they drive summer melt and winter accumulation, respectively.
+Because reanalyses include observational information through data assimilation, they are closer to observations than pure GCMs in densely monitored regions, such as the southern part of our modelling domain \citep{bengtsson-etal-2007}.
+On the other hand, the atmospheric physics embedded in the models provide more realistic output than obtained by spatial interpolation in sparsely monitored regions, such as the northern part of our modelling domain.
+
+\subsection{Preprocessing and lapse-rate corrections}
+
+WorldClim, NCEP/NCAR, ERA-Interim, CFSR and NARR climatologies were re-projected to the Canadian Atlas Lambert conformal conic projection (EPSG code~3978) and bilinearly interpolated to the 10\,km-resolution model grid using the module r.proj from GRASS~GIS \citep{soft:grass}.
+In addition, WorldClim data was extrapolated to cover grid points in the ocean using the nearest-neighbour algorithm \citep{soft:scipy}.
+Note that here we choose to present original rather than interpolated data, in order to highlight differences between datasets related to spatial resolution (Figs.~\ref{fig:temp} and~\ref{fig:prec}).
+
+The CFSR climatology presents wave-like artefacts in its precipitation field (Fig.~\ref{fig:prec}).
+These are a common feature in GCM simulations due to the failure of spectral models to resolve the very local nature of topography-induced precipitation events, which becomes particularly visible in the high-resolution, less spatially-smoothed data.
+Therefore, an alternative forcing was prepared for CFSR by smoothing its precipitation field.
+This was achieved by averaging data locally in a circular kernel of 7\,pixels in diameter prior to re-projection using the module r.neighbors from GRASS~GIS \citep{soft:grass}.
+Finally, input files for PISM were prepared using \citet{web:nc4py}.
+
+% ----------------------------------------------------------------------
+\section{Results}
+\label{sec:results}
+% ----------------------------------------------------------------------
+
+Using WorldClim, NCEP/NCAR, ERA-Interim, CFSR (smoothed and original datasets) and NARR input climatologies as climate forcing, we run 66~simulations of glacial inception and growth of the Cordilleran ice sheet over 10~ka using temperature offsets ranging from 0 to 10\,\unit{\degree C}.
+Significant differences in patterns of ice sheet growth, final glacial extent, and volume are attained (Figs.
+\ref{fig:ivolarea}-\ref{fig:cool05}).
+
+Across the range of temperature offsets used, final glaciated area and final ice volume differ widely between different input climatologies (Fig.~\ref{fig:ivolarea}).
+For instance, using a 5\,\unit{\degree C} temperature offset, final glaciated area differs between WorldClim and CFSR forcing by a factor of~6 and final volume by a factor of~12 (Fig.~\ref{fig:ivolarea}).
+For particular temperature offset values, given input climatologies result in similar final glaciated areas and ice volumes, yet regional patterns are different (Fig.~\ref{fig:extent}).
+
+Under present climate (0\,\unit{\degree C} offset), WorldClim, ERA-Interim and NARR forcing produce local ice caps in areas presently glaciated (Fig.~\ref{fig:extent}).
+These results are fairly consistent with contemporary ice distribution, although both the 10~km horizontal resolution and the occurrence of glaciers and ice caps surface topographies in the ETOPO1 dataset used for basal topography prevent a meaningful detailed comparison.
+CFSR forcing, on the other hand, yields continuous ice cover over the northern Coast Mountains, and NCEP/NCAR forcing results in ice sheet growth, both of which are largely inconsistent with the present state of glaciation.
+At the opposite side of the temperature offset range, it can be noted that both the NCEP/NCAR and CFSR forcing produce oversized ice sheets whose extents are primarily restricted by the domain boundaries (Fig.~\ref{fig:extent}).
+The smoothing of CFSR precipitation yields an insignificant effect on the resulting ice sheet geometry.
+
+To contrast the effect of the six different input climatologies, we plot simulations using a 5\,\unit{\degree C} temperature offset (Fig.~\ref{fig:cool05}).
+The magnitude of glaciation differs dramatically from one climate forcing to another.
+Whereas the NCEP/NCAR and CFSR climatologies produce a large ice sheet that covers most of the model domain (Fig.~\ref{fig:cool05}, lower panels), WorldClim, ERA-Interim and NARR climatologies lead to a more restrictive ice cover (Fig.~\ref{fig:cool05}, upper panels).
+The WorldClim run only resulted in the growth of mountain ice caps bounded to the highest topography.
+
+% ----------------------------------------------------------------------
+\section{Discussion}
+\label{sec:discussion}
+% ----------------------------------------------------------------------
+
+These simulations of the LGM Cordilleran Ice Sheet show a very large sensitivity to the choice of climate forcing data used (Figs.~\ref{fig:ivolarea}--\ref{fig:cool05}).
+To understand the origin of the discrepancies between model results, we quantify differences in surface air temperature and precipitation rate between the input climatologies and test their effect on the ice sheet model.
+We then compare our findings to a geomorphological reconstruction of the LGM Cordilleran Ice Sheet margin, and discuss some of the key assumptions of this study.
+
+\subsection{Comparison of forcing climatologies}
+
+As previously outlined, significant differences in the distribution of winter (DJF) precipitation and summer (JJA) temperature, which are the predominant controls on accumulation and melt at the modelled ice sheet surface, exist between datasets (Figs.~\ref{fig:topo}--\ref{fig:prec}).
+In this section we use the WorldClim climatology, which is derived from observations, as a reference with which to compare other climate forcing data on land.
+ERA-Interim, CFSR and NARR climatologies exhibit temperature distributions fairly consistent with WorldClim data (Fig.~\ref{fig:tempheatmap}).
+When examining the spatial distribution of temperature differences between reanalyses and WorldClim data, it appears that most of the disparity between reanalyses and observation data is caused by topographical detail at scales unresolved in the reanalysis datasets (Fig.~\ref{fig:tempdiff}).
+Surface air temperature data from the NCEP/NCAR climatology shows the largest disparity with that of WorldClim, with a significant cold anomaly over most of the modelling domain (Figs.~\ref{fig:tempheatmap}--\ref{fig:tempdiff}).
+
+All four reanalyses used in this study exhibit higher precipitation rates than WorldClim data (Fig.~\ref{fig:precheatmap}).
+However the magnitude of this anomaly differs significantly between datasets.
+The spatial distributions of precipitation differences between reanalyses and WorldClim data show that negative precipitation anomalies are generally constrained to the windward slope of the major mountain ranges, while positive precipitation anomalies are found on the leeward slope of these ranges and extend downwind to interior plateaux and lowlands (Fig.~\ref{fig:precdiff}).
+
+Most likely, this is the signature of an orographic precipitation effect.
+As described in section~\ref{sec:climate}, the topography of the Northern American Cordillera is such that its western ranges form a continuous orographic barrier, causing high precipitation along the Pacific coast while leaving much of the interior arid.
+However the ability of a GCM to reproduce these contrasts in precipitation is bounded by spatial resolution.
+In a model of coarser resolution, these high mountain ranges are reduced to smooth hills and the modelled distribution of precipitation is also smoother (Fig.~\ref{fig:oroprecip}).
+However, this effect alone cannot explain the widespread and pronounced positive precipitation anomalies observed in CFSR data (Figs.~\ref{fig:precheatmap}--\ref{fig:precdiff}) despite of its high spatial resolution (Table~\ref{tab:reanalyses}).
+
+\subsection{Model sensitivity to climate forcing}
+
+To distinguish the effects of temperature and precipitation biases on the ice sheet model, we run a series of eight additional simulations.
+These new simulations use a “hybrid” climate forcing that consists of temperature data from WorldClim, combined with precipitation data from each of the four reanalyses, and vice-versa.
+A single temperature offset of 5\,\unit{\degree C} is used.
+This value was chosen to allow for comparison with previous results (Fig.~\ref{fig:cool05}).
+In each case, temperature lapse rates are computed using the reference topography from the corresponding dataset.
+
+Although only temperature or precipitation changes are applied, this experiment results once more in large differences in modelled ice sheet geometries (Fig.~\ref{fig:biatm}) as compared to the reference WorldClim 5\,\unit{\degree C} run (Fig.~\ref{fig:cool05}, top-left panel).
+For ERA-Interim, CFSR and NARR climatologies, precipitation anomalies clearly dominate the differences in ice sheet response, whereas temperature anomalies have relatively little effect (Fig.~\ref{fig:biatmbars}).
+However in the case of NCAR forcing, both the negative temperature bias (Figs.~\ref{fig:tempheatmap}--\ref{fig:tempdiff}) and the important precipitation differences (Figs.~\ref{fig:precheatmap}--\ref{fig:precdiff}) contribute approximately equally to produce oversized ice sheets (Fig.~\ref{fig:biatmbars}).
+
+\subsection{Comparison to geomorphological LGM margin}
+
+We compare modelled ice sheet geometries to a reconstruction of the ice margin at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) by \citet{dyke-2004}, based on glacial geomorphology and radiocarbon dating (Fig.~\ref{fig:locmap}).
+This corresponds to the LGM extent of most of the Cordilleran ice sheet, which occurred later than for the Laurentide ice sheet \citep{porter-swanson-1998,dyke-2004,stroeven-etal-2010,stroeven-etal-inpress}.
+
+Considering previously identified weaknesses of the CFSR and NCEP/NCAR data (Figs.~\ref{fig:tempheatmap}--\ref{fig:biatmbars}), we restricted our analysis to the WorldClim, ERA-Interim and NARR input climatologies.
+In order to compare modelled ice sheets of a similar size, we select for each climate forcing the simulation that leads to a final glaciated area closest to the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$.
+This corresponds to temperature offset values of 8\,\unit{\degree C} (WorldClim), 6\,\unit{\degree C} (ERA-Interim) and 7\,\unit{\degree C} (NARR).
+These qualitative “best” runs are presented in Figure~\ref{fig:best} along with their associated temperature depressions and a reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.
+
+Although this selection results in similarly sized ice sheets, noticeable differences in shape exist.
+The ERA-Interim simulation produces a more northerly-centred ice sheet with too much ice in the north and too little in the south.
+This may partly reflect unresolved orographic precipitation effects (Fig.~\ref{fig:oroprecip}) due to the coarser GCM resolution (Table~\ref{tab:reanalyses}).
+The LGM ice sheet margin by \citet{dyke-2004} is best reproduced by simulations driven by input climatologies from WorldClim and NARR data.
+However, common discrepancies between the modelled ice sheet geometry and the geomorphological reconstruction can still be observed (Fig.~\ref{fig:best}).
+
+Firstly, the modelled eastern margin of the ice sheet extends further east than the inferred position of the junction between the Cordilleran and the Laurentide ice sheets.
+This can be seen for instance in the south-eastern part of our modelling domain, where the modelled ice sheets cover an ice-free corridor of the LGM reconstruction (Fig.~\ref{fig:best}).
+However, such comparisons have been hampered by the fact that we have not considered the buttressing effect of the Laurentide Ice Sheet, which would have inhibited the Cordilleran Ice Sheet from advancing onto the Canadian Prairies.
+In addition, potential effects of the growing ice-sheet on atmospheric circulation and precipitation changes are not included in our model.
+The Cordilleran ice sheet initiated from the formation of mountain ice caps over the high mountain ranges \citep{clague-1989}.
+In our simulations, a continuous ice cover quickly forms over the western ranges, where precipitation rates are higher than in the rest of the domain.
+In reality, this continuous ice cover enhanced the topographical barrier already formed by the western ranges, and would have resulted in less precipitation and warmer air in the interior.
+This process of ``self-inhibiting growth'' is not captured in our model but was demonstrated to be important in a similar setting over Greenland, where it may have been limiting the advance of the ice sheet during its build-up \citep{langen-etal-2012}.
+
+Secondly, our simulations produce anomalous ice cover on parts of the continental shelf in the Arctic Ocean (Fig.~\ref{fig:best}).
+These regions experience a marine climate, including lower summer temperature than on the adjacent land (Fig.~\ref{fig:temp}).
+However in our simulations, sea-level is lowered by 120~m, turning large parts of the low-sloping continental shelf into land.
+This anomalous ice cover is to be interpreted as an artefact arising from our simplistic temperature offset method, and has little effect on the rest of the results.
+
+Given the lack of WorldClim data offshore, we consider the NARR climate forcing as best suited for simulations of the Cordilleran Ice Sheet among those tested, and under the assumptions of the present study.
+
+\subsection{Model sensitivity to the duration of ice sheet inception}
+
+A key assumption for the present study is our choice of 10\,ka for the duration of all our simulations.
+Although geomorphological data shows that the period of Cordilleran Ice Sheet inception, from nearly ice-free to full glacial conditions, could not have been much longer than 10\,ka, they provide no lower bound to this value.
+
+To test the effect of an inception period of shorter duration on model results, we use simulations run under NARR climate forcing with temperature offsets of 7 to 11\,\unit{\degree C}, and compare modelled ice sheet extent when glaciated area reaches the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$.
+
+This experiment shows that a shorter start-up period leads to more restricted ice cover in the eastern part of the modelling domain, where precipitation rate controls ice advance, but more extensive glaciation in the south-western domain along the Pacific coast, where temperature is the limiting factor (Fig.~\ref{fig:durationstack}).
+Hence, reducing simulation duration leads to a closer match with the LGM ice sheet margin by \citet{dyke-2004} in some parts of the modelling domain, but higher discrepancy in other parts.
+
+\subsection{Other potential method improvements}
+
+Our representation of climate history by constant temperature offsets is voluntarily simplistic.
+To better understand the transient character of the LGM Cordilleran Ice Sheet, time-dependent palaeo-climate forcing would be needed.
+Although it is clear that past temperature changes over the region were neither homogeneous nor constant in time, and probably associated with precipitation changes, their patterns are not trivial and may potentially display a strong inter-dependence with the evolution of the Cordilleran and Laurentide ice sheets.
+A more correct, yet more complex approach would be to use coupling to a GCM of intermediate complexity \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013}.
+
+Furthermore, the surface mass balance model crudely simplifies the processes taking place at the glacier surface.
+Refreezing of melted snow and ice and water retention in the snow pack was shown to influence surface mass balance of glaciers \citep{trabant-mayo-1985}, yet it is not accounted for in our PDD model.
+Based on scarce observations over the Greenland Ice Sheet, different parametrizations of this effect have been developed \citep{janssens-huybrechts-2000,fausto-etal-2009b}.
+However, these parametrizations were shown to disagree with more complex models \citep{reijmer-etal-2012}, and it remains undemonstrated whether their application to a different region and time period is sound.
+In addition, we simulate temperature variability by using a constant, uniform value of daily temperature standard deviation.
+This approach implies large biases of surface mass-balance \citep{charbit-etal-2013,rau-rogozhina-2013,seguinot-2013}, particularly over regions with such various climates as experienced within our model domain.
+
+% ----------------------------------------------------------------------
+\conclusions
+\label{sec:concl}
+% ----------------------------------------------------------------------
+
+Our study shows a strong dependency of ice sheet model results on the choice of climate forcing data.
+For three of the four reanalysis datasets used, precipitation rate, over surface air temperature, causes much of the variation in modelled ice sheet outlines and volumes.
+Furthermore, the spatial resolution of input data appears critical for providing the ice sheet model with an accurate precipitation field, confirming results obtained by \citet{quiquet-etal-2012} from various GCM forcing datasets over the Greenland Ice Sheet.
+
+For the Cordilleran Ice Sheet, we achieve the best fit to the mapped LGM margin by \citet{dyke-2004} using climate forcing from the high-resolution interpolated observational dataset WorldClim \citep{data:worldclim} and the North American Regional Reanalysis \citep[NARR;][]{data:narr}.
+The latter dataset is preferable in our case due to the lack of WorldClim data offshore.
+Climate forcing from CFSR and the NCEP/NCAR reanalysis produce largely oversized ice sheets due to too high precipitation rates, and in the second case, too low surface air temperature.
+The ERA-Interim data used in this study produces more reasonable results, but it may be too coarse to accurately resolve the spatial distribution of orographic precipitation associated with the rugged topography of the study area, resulting in a misplaced ice coverage in regard to geomorphological data.
+Therefore, we retain NARR data for forcing simulations of the Cordilleran Ice Sheet in the future.
+
+One must keep in mind, however, that these results are only accurate for our choices of ice-sheet model (PISM), surface mass balance model (PDD), study object (the Cordilleran Ice Sheet), and, most importantly, palaeo-climate representation (temperature offsets).
+At LGM, it is most likely that the presence of ice sheets significantly affected circulation patterns and distribution of precipitation.
+At the cost of greater computational expense, a more accurate representation of palaeo-climate may allow for a better fit between model results and the mapped LGM margin.
 

### 2014-03-19 cordillera-climate.tex
Flatten floats (no actual changes)

@@
 
+% ----------------------------------------------------------------------
 % Floats
+% ----------------------------------------------------------------------
 \newpage
-\input{floats/tables}
-\input{floats/figures}
+
+% tab:reanalyses
+\begin{table*}[t]
+	\caption{Characteristics of reanalysis climatologies used to inform the mass balance component of PISM.}
+	\label{tab:reanalyses}
+	\vskip4mm
+	\centering
+	\begin{tabular}{lllll}
+		\tophline
+		Reanalysis& Spatial coverage& Averaging period& Resolution& Description\\
+		\middlehline
+		NCEP/NCAR&  global&     1981 -- 2010& 1.875\degree& \citet{data:ncar}\\
+		ERA-Interim&global&     1979 -- 2011& 1.000\degree& \citet{data:erai}\\
+		CFSR&       global&     1979 -- 2010& 0.325\degree& \citet{data:cfsr}\\
+		NARR&       North America& 1979 -- 2000& 32\,km& \citet{data:narr}\\
+		\bottomhline
+	\end{tabular}
+\end{table*}
+
+% fig:locmap
+\begin{figure}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics[width=8cm]{cordillera-climate-locmap}
+	\end{center}
+	\caption{Shaded relief map of northern North America with the extent of the Cordilleran (CIS), Laurentide (LIS) and Innuitian (IIS) ice sheets at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) \citep{dyke-2004}.
+While this age denotes the LIS after retreat from its LGM, it closely corresponds to the LGM extent of most of the Cordilleran ice sheet \citep{porter-swanson-1998,dyke-2004,stroeven-etal-2010}.
+The rectangular box denotes the modelling domain of the Cordilleran ice sheet of 1500 by 3000\,km.
+Major mountain ranges covered by the ice sheet include the Wrangell and St.-Elias mountains (W-SE), the Selwyn and MacKenzie mountains (S-MK), the Coast Mountains and the Rocky Mountains.
+The background map consists of ETOPO1 \citep{data:etopo1} and Natural Earth Data \citep{data:naturalearth} and was assembled with GRASS~GIS \citep{soft:grass}.}
+	\label{fig:locmap}
+\end{figure}
+
+% fig:topo
+\begin{figure*}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics{cordillera-climate-topo}
+	\end{center}
+	\caption{Topography maps from WorldClim \citep{data:worldclim}, ERA-Interim reanalysis \citep{data:erai}, North American Regional Reanalysis \citep[NARR;][]{data:narr}, ETOPO1 \citep{data:etopo1}, Climate Forecast System Reanalysis \citep[CFSR;][]{data:cfsr}, and NCEP/NCAR reanalysis \citep{data:ncar}.
+Whereas ETOPO1 is used as basal topography for the ice sheet model, all other data are used as a reference for temperature lapse-rate corrections.
+Fig.~\ref{fig:topo}-\ref{fig:durationstack} are drawn using Matplotlib \citep{soft:mpl}.}
+	\label{fig:topo}
+\end{figure*}
+
+% fig:temp
+\begin{figure*}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics{cordillera-climate-temp}
+	\end{center}
+	\caption{Summer (JJA) and winter (DJF) temperature maps from WorldClim \citep{data:worldclim}, and summer (JJA) temperature maps from ERA-Interim reanalysis \citep{data:erai}, North American Regional Reanalysis \citep[NARR;][]{data:narr}, Climate Forecast System Reanalysis \citep[CFSR;][]{data:cfsr}, and NCEP/NCAR reanalysis \citep{data:ncar} climatologies.}
+	\label{fig:temp}
+\end{figure*}
+
+% fig:prec
+\begin{figure*}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics{cordillera-climate-prec}
+	\end{center}
+	\caption{Winter (DJF) and summer (JJA) precipitation maps from WorldClim \citep{data:worldclim}, and winter (DFJ) temperature maps from ERA-Interim reanalysis \citep{data:erai}, North American Regional Reanalysis \citep[NARR;][]{data:narr}, Climate Forecast System Reanalysis \citep[CFSR;][]{data:cfsr}, and NCEP/NCAR reanalysis \citep{data:ncar} climatologies.
+Additional forcing data was prepared to correct for wave-like precipitation artefacts in CFSR (section~\ref{sec:climate}).}
+	\label{fig:prec}
+\end{figure*}
+
+% fig:ivolarea
+\begin{figure}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics{cordillera-climate-ivolarea}
+	\end{center}
+	\caption{Total glaciated area and ice volume after 10\,ka as a function of temperature offset for each climate forcing used.}
+	\label{fig:ivolarea}
+\end{figure}
+
+% fig:extent
+\begin{figure*}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics{cordillera-climate-extent}
+	\end{center}
+	\caption{Extent of ice cover after 10\,ka as a function of applied temperature offsets for each climate forcing used.
+A reconstructed LGM ice sheet margin by \citet{dyke-2004} (Fig.~\ref{fig:locmap}) is given for reference (black line).}
+	\label{fig:extent}
+\end{figure*}
+
+% fig:cool05
+\begin{figure*}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics{cordillera-climate-cool05}
+	\end{center}
+	\caption{Ice surface topography (1 km contours) and velocity (\unit{m\,a^{-1}}) after 10\,ka under a climate 5\,\unit{\degree C} colder than present for each climate forcing used.}
+	\label{fig:cool05}
+\end{figure*}
+
+% fig:tempheatmap
+\begin{figure}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics{cordillera-climate-tempheatmap}
+	\end{center}
+	\caption{Density maps showing a comparison of summer (JJA) surface air temperature data from the WorldClim climatology, against that of each reanalysis.
+Apparent horizontal lines are an artefact of different horizontal resolutions.
+Colour mapping is based on a logarythmic scale.
+Note the cold bias of NCEP/NCAR data relative to WorldClim data.}
+	\label{fig:tempheatmap}
+\end{figure}
+
+% fig:tempdiff
+\begin{figure}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics{cordillera-climate-tempdiff}
+	\end{center}
+	\caption{Summer (JJA) surface air temperature difference maps against WorldClim data, after bi-linear spatial interpolation.
+Note the cold bias of NCEP/NCAR data and temperature anomalies due to unresolved topographic detail.}
+	\label{fig:tempdiff}
+\end{figure}
+
+% fig:precheatmap
+\begin{figure}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics{cordillera-climate-precheatmap}
+	\end{center}
+	\caption{Density maps showing a comparison of winter (DJF) precipitation rate from the WorldClim climatology, against that of each reanalysis.
+Apparent horizontal lines are an artefact of different horizontal resolutions.
+Apparent vertical lines at low values are an artefact of the low precision of WorldClim data.
+Colour mapping is based on a logarythmic scale.
+Note the wet bias of all reanalysis data relative to WorldClim data.}
+	\label{fig:precheatmap}
+\end{figure}
+
+% fig:precdiff
+\begin{figure}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics{cordillera-climate-precdiff}
+	\end{center}
+	\caption{Winter (DJF) precipitation rate difference maps against WorldClim data, after bi-linear spatial interpolation.
+Note the wet bias of all reanalysis data and the large anomalies of CFSR and NCEP/NCAR data.}
+	\label{fig:precdiff}
+\end{figure}
+
+% fig:oroprecip
+\begin{figure}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics{cordillera-climate-oroprecip}
+	\end{center}
+	\caption{Schematic representation of the orographic precipitation effect over a mountain range.
+In a GCM of low resolution, the precipitation peak appears downwind-shifted and smoother and the precipitation shadow is less pronounced than in a high resolution GCM.}
+	\label{fig:oroprecip}
+\end{figure}
+
+% fig:biatm
+\begin{figure*}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics{cordillera-climate-biatm}
+	\end{center}
+	\caption{Ice surface topography (1 km contours) and velocity (\unit{m\,a^{-1}}) after 10\,ka using “hybrid” climate forcing with precipitation rate from WorldClim and surface air temperature from each reanalysis (upper row), and surface air temperature from WorldClim and precipitation rate from each reanalysis (lower row).
+In other words, the upper row shows the effect of temperature anomalies, and the lower row the effect of precipitation anomalies, for each reanalysis, relative to WorldClim data.
+Each simulation uses a 5\,\unit{\degree C} offset for comparison with figure~\ref{fig:cool05}}
+	\label{fig:biatm}
+\end{figure*}
+
+% fig:biatmbars
+\begin{figure}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics{cordillera-climate-biatmbars}
+	\end{center}
+	\caption{Effect of temperature and precipitation anomalies (separately and jointly) from each reanalysis on modelled final ice volume relative to the result of the WorldClim 5\,\unit{\degree C} offset simulation.
+Corresponding ice sheet geometries are presented in figures~\ref{fig:cool05} and~\ref{fig:biatm}}
+	\label{fig:biatmbars}
+\end{figure}
+
+% fig:best
+\begin{figure*}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics{cordillera-climate-best}
+	\end{center}
+	\caption{Ice surface topography (1\,km contours) after 10\,ka using temperature offsets resulting in glaciated areas of circa $2\,\times10^6\,\unit{km^2}$.
+A reconstructed LGM ice sheet margin by \citet{dyke-2004} (Fig.~\ref{fig:locmap}) is given for reference (blue line).}
+	\label{fig:best}
+\end{figure*}
+
+% fig:durationstack
+\begin{figure}[t]
+	\vspace*{2mm}
+	\begin{center}
+		\includegraphics{cordillera-climate-durationstack}
+	\end{center}
+	\caption{Left panel: model sensitivity to simulation length.
+Modelled glaciated area, using NARR forcing data, and temperature offsets from 0 to 15\,\unit{\degree C}, solid lines corresponding to values from 7 to 11\,\unit{\degree C}.
+Right panel: modelled ice margin corresponding to temperature offsets from 7 to 11\,\unit{\degree C} when total glaciated area reach the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$.
+A reconstructed LGM ice sheet margin by \citet{dyke-2004} (Fig.~\ref{fig:locmap}) is given for reference (grey shading).
+Note that shorter (and colder) simulations lead to more restrictive glaciation of the continental eastern margin but further ice extent in the maritime south-western part of the modelling domain.}
+	\label{fig:durationstack}
+\end{figure}
 

### 2014-03-19 cordillera-climate.tex
Typesetting: unbreak post-article space

@@
 We present an ensemble of numerical simulations of the Cordilleran Ice Sheet during the Last Glacial Maximum performed with the Parallel Ice Sheet Model (PISM), applying temperature offsets to the present-day climatologies from five different datasets.
-Monthly mean surface air temperature and precipitation from WorldClim, the NCEP/NCAR reanalysis, the ERA-Interim reanalysis, the Climate Forecast System Reanalysis and the North American Regional Reanalysis are used to compute surface mass balance in a positive degree-day model.
+Monthly mean surface air temperature and precipitation from WorldClim, the NCEP/NCAR reanalysis, the ERA-Interim reanalysis, the Climate Forecast System Reanalysis and the North American Regional Reanalysis are used to compute surface mass balance in a~positive degree-day model.
 Modelled ice sheet outlines and volumes appear highly sensitive to the choice of climate forcing.
 For three of the four reanalysis datasets used, differences in precipitation are the major source for discrepancies between model results.
-We assess model performance against a geomorphological reconstruction of the ice margin at the Last Glacial Maximum, and suggest that part of the mismatch is due to unresolved orographic precipitation effects caused by the coarse resolution of reanalysis data.
+We assess model performance against a~geomorphological reconstruction of the ice margin at the Last Glacial Maximum, and suggest that part of the mismatch is due to unresolved orographic precipitation effects caused by the coarse resolution of reanalysis data.
 The best match between model output and the reconstructed ice margin is obtained using the high-resolution North American Regional Reanalysis, which we retain for simulations of the Cordilleran Ice Sheet in the future.
@@
 
-At the Last Glacial Maximum (LGM), glaciers of a size comparable to the present Greenland and Antarctic ice sheets covered parts of Northern America (Laurentide, Cordilleran and Innuitian ice sheets) and northern Eurasia (Fennoscandian Ice Sheet).
-Numerical modelling of these former ice masses allows for a comparison between glaciological theories embedded in the models and geomorphological traces underpinning palaeo-glaciological reconstructions.
-Yet, a major obstacle in this exercise resides in large uncertainties concerning the climate forcing, typically atmospheric temperature and precipitation, required as input to numerical glacier models \citep{hebeler-etal-2008}.
+At the Last Glacial Maximum (LGM), glaciers of a~size comparable to the present Greenland and Antarctic ice sheets covered parts of Northern America (Laurentide, Cordilleran and Innuitian ice sheets) and northern Eurasia (Fennoscandian Ice Sheet).
+Numerical modelling of these former ice masses allows for a~comparison between glaciological theories embedded in the models and geomorphological traces underpinning palaeo-glaciological reconstructions.
+Yet, a~major obstacle in this exercise resides in large uncertainties concerning the climate forcing, typically atmospheric temperature and precipitation, required as input to numerical glacier models \citep{hebeler-etal-2008}.
 This includes uncertainty in the representation of Earth's present climate in regions of poor station coverage, and even larger uncertainty concerning accurate reconstructions of past climate change.
 
-Arguably the most physically sound way to force an ice sheet model for simulations of glacial history is to couple it with a General Circulation Model \citep[GCM;][]{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013}.
+Arguably the most physically sound way to force an ice sheet model for simulations of glacial history is to couple it with a~General Circulation Model \citep[GCM;][]{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013}.
 However the computational demand of GCMs is such that only models of intermediate complexity can run on the time-scales of tens of thousands of years characteristic of ice sheet growth and decay.
 
-Climatologies obtained from GCM palaeo-climate simulations such as produced within the Paleoclimate Modelling Intercomparison Project \citep[PMIP;][]{joussaume-taylor-1995} provide perhaps a more reasonable representation of past climate.
-However, as such climatologies are only available for specific periods of time, using them as climate forcing for an ice sheet model requires either an assumption of steady-state ice sheet response to climatic fluctuations \citep{huybrechts-tsiobbel-1996}, or interpolation through time between climatologies from different periods, which can be linear \citep{charbit-etal-2002}, or modulated by a ``glacial index'' weighting function derived from ice core $\delta^{18}$O records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}.
+Climatologies obtained from GCM palaeo-climate simulations such as produced within the Paleoclimate Modelling Intercomparison Project \citep[PMIP;][]{joussaume-taylor-1995} provide perhaps a~more reasonable representation of past climate.
+However, as such climatologies are only available for specific periods of time, using them as climate forcing for an ice sheet model requires either an assumption of steady-state ice sheet response to climatic fluctuations \citep{huybrechts-tsiobbel-1996}, or interpolation through time between climatologies from different periods, which can be linear \citep{charbit-etal-2002}, or modulated by a~``glacial index'' weighting function derived from ice core $\delta^{18}$O records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}.
 An important drawback in this approach is that GCM palaeo-climate simulations themselves rely on an accurate description of the global surface topography and therefore include global ice sheet reconstructions such as ICE-4G \citep{peltier-1994} for their surface topographic boundary condition.
@@
 
-Here, we propose to address some of the uncertainties concerning climate forcing of numerical glacier models by evaluating the responses of a numerical model, in terms of glacier extent, to inputs from several climate datasets.
-This is not entirely a new approach as a few studies of this kind are presently available.
-\citet{quiquet-etal-2012}, for example, assessed the sensitivity of a Greenland ice sheet model to various climate forcing fields, including a regional parametrization by \citet{fausto-etal-2009a}, output from several GCMs and an atmospheric reanalysis.
-\citet{rodgers-etal-2004} and \citet{charbit-etal-2007} tested the sensitivity of a model of Northern Hemisphere ice sheets to climate forcing from different PMIP LGM simulations.
+Here, we propose to address some of the uncertainties concerning climate forcing of numerical glacier models by evaluating the responses of a~numerical model, in terms of glacier extent, to inputs from several climate datasets.
+This is not entirely a~new approach as a~few studies of this kind are presently available.
+\citet{quiquet-etal-2012}, for example, assessed the sensitivity of a~Greenland ice sheet model to various climate forcing fields, including a~regional parametrization by \citet{fausto-etal-2009a}, output from several GCMs and an atmospheric reanalysis.
+\citet{rodgers-etal-2004} and \citet{charbit-etal-2007} tested the sensitivity of a~model of Northern Hemisphere ice sheets to climate forcing from different PMIP LGM simulations.
 Although using different set-ups, all three studies demonstrate the very large sensitivity of ice sheet models to the choice of climate forcing data used.
-To limit the degrees of freedom in our model and obtain results independent of palaeo-ice sheet reconstructions such as ICE-4G, we use a simple temperature offset similar to the approaches by \citet{greve-etal-1999} and \citet{bintanja-etal-2005} and assess ice sheet model sensitivity to the choice of present-day climate data.
+To limit the degrees of freedom in our model and obtain results independent of palaeo-ice sheet reconstructions such as ICE-4G, we use a~simple temperature offset similar to the approaches by \citet{greve-etal-1999} and \citet{bintanja-etal-2005} and assess ice sheet model sensitivity to the choice of present-day climate data.
 Hence, rather than using GCM output, we force our model with climate reanalysis data, which builds on observational information through data assimilation \citep{bengtsson-etal-2007}.
@@
 The Cordilleran Ice Sheet (Fig.~\ref{fig:locmap}) covered an area that presently experiences strong regional variations in climate.
-From a numerical modelling perspective, it is one of the least studied palaeo-ice sheets of the Northern Hemisphere, despite the fact that significant stratigraphical, geomorphological, and chronological data is available to constrain its extent \citep{jackson-clague-1991,porter-swanson-1998,dukrodkin-1999,dyke-2004,kaufman-manley-2004,kleman-etal-2010,stroeven-etal-2010,stroeven-etal-inpress,margold-etal-2011}.
+From a~numerical modelling perspective, it is one of the least studied palaeo-ice sheets of the Northern Hemisphere, despite the fact that significant stratigraphical, geomorphological, and chronological data is available to constrain its extent \citep{jackson-clague-1991,porter-swanson-1998,dukrodkin-1999,dyke-2004,kaufman-manley-2004,kleman-etal-2010,stroeven-etal-2010,stroeven-etal-inpress,margold-etal-2011}.
 The Cordilleran Ice Sheet has previously been modelled as part of efforts to simulate ice sheets in North America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} and world-wide \citep{yoshimori-etal-2001}.
-While these studies generally reproduce the magnitude of North American glaciation at LGM reasonably well, there exists a tendency in the simulations that are independent of ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and interior Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.
+While these studies generally reproduce the magnitude of North American glaciation at LGM reasonably well, there exists a~tendency in the simulations that are independent of ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and interior Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.
 
-Here we use PISM, a Parallel Ice Sheet Model \citep{web:pism}, to simulate the extent and thickness of the Cordilleran Ice Sheet at the LGM (Fig.~\ref{fig:locmap}).
-We force our model with multiple climate datasets and compare the modelled ice extent to a geomorphological reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.
-We aim to determine the climate dataset which is most suited for simulation of the Cordilleran Ice Sheet and to use as input for future, transient studies over a glacial cycle.
+Here we use PISM, a~Parallel Ice Sheet Model \citep{web:pism}, to simulate the extent and thickness of the Cordilleran Ice Sheet at the LGM (Fig.~\ref{fig:locmap}).
+We force our model with multiple climate datasets and compare the modelled ice extent to a~geomorphological reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.
+We aim to determine the climate dataset which is most suited for simulation of the Cordilleran Ice Sheet and to use as input for future, transient studies over a~glacial cycle.
 
@@
 
-As we aim to model glacial inception and growth of the Cordilleran Ice Sheet towards a configuration as last attained during the LGM, we mimic palaeo-climatic conditions by applying constant temperature offsets homogeneously over the modelling domain.
-Each simulation starts from ice-free conditions and runs for 10\,ka, a time interval representative of the rapid build-up of the last Cordilleran Ice Sheet from nearly ice-free to full glacial conditions \citep{clague-1989,stroeven-etal-2010}.
+As we aim to model glacial inception and growth of the Cordilleran Ice Sheet towards a~configuration as last attained during the LGM, we mimic palaeo-climatic conditions by applying constant temperature offsets homogeneously over the modelling domain.
+Each simulation starts from ice-free conditions and runs for 10\,ka, a~time interval representative of the rapid build-up of the last Cordilleran Ice Sheet from nearly ice-free to full glacial conditions \citep{clague-1989,stroeven-etal-2010}.
 Our simulations are performed in parallel on 32 cores at the Swedish National Supercomputing Center.
@@
 The central part of an ice sheet model consists of the computation of flow velocity which itself depends on temperature.
-PISM is a shallow model, which implies that the balance of stresses is approximated based on their predominant components.
+PISM is a~shallow model, which implies that the balance of stresses is approximated based on their predominant components.
 On the other hand, the model is polythermal: it accounts for differences in temperature and softness within the ice column.
 
-The Shallow Shelf Approximation (SSA) is used as a ``sliding law'' for the Shallow Ice Approximation (SIA) \citep{bueler-brown-2009,winkelmann-etal-2011}.
-SIA and SSA velocities are computed by finite difference methods on a 10\,km-resolution horizontal grid of 300 by 150 points (the modelling domain).
+The Shallow Shelf Approximation (SSA) is used as a~``sliding law'' for the Shallow Ice Approximation (SIA) \citep{bueler-brown-2009,winkelmann-etal-2011}.
+SIA and SSA velocities are computed by finite difference methods on a~10\,km-resolution horizontal grid of 300 by 150 points (the modelling domain).
 Ice softness depends on temperature and water content through an enthalpy formulation \citep{aschwanden-blatter-2009,aschwanden-etal-2012}.
-Enthalpy is computed in three dimensions in up to 51 irregularly spaced layers within the ice, and temperature is additionally computed in 11 regularly spaced layers in bedrock to a depth of 1\,km.
-Surface air temperature from the climate forcing provides the upper boundary condition to the ice enthalpy model, and a geothermal heat flux of 70\,\unit{mW\,m^{-2}} provides the lower boundary condition to the bedrock thermal model.
+Enthalpy is computed in three dimensions in up to 51 irregularly spaced layers within the ice, and temperature is additionally computed in 11 regularly spaced layers in bedrock to a~depth of 1\,km.
+Surface air temperature from the climate forcing provides the upper boundary condition to the ice enthalpy model, and a~geothermal heat flux of 70\,\unit{mW\,m^{-2}} provides the lower boundary condition to the bedrock thermal model.
 Although this uniform value does not account for the high spatial geothermal variability in the region, it is on average representative of available heat flow measurements \citep{artemieva-mooney-2001,blackwell-richards-2004}.
@@
 
-A pseudo-plastic sliding law \citep{aschwanden-etal-2013} relates the bed-parallel shear stress and the sliding velocity.
+A~pseudo-plastic sliding law \citep{aschwanden-etal-2013} relates the bed-parallel shear stress and the sliding velocity.
 The yield stress is modelled using the Mohr-Coulomb criterion.
-The till friction angle $\phi$ varies from 10 to 30\degree\ and is a function of current bed elevation, with lowest values occurring at low elevations,
+The till friction angle $\phi$ varies from 10 to 30\degree\ and is a~function of current bed elevation, with lowest values occurring at low elevations,
 
@@
 where $\phi$ is given in degrees and $z$ in meters above current sea level.
-This model thus accounts for a weakening of till associated with the presence of marine sediments \citep{martin-etal-2011,aschwanden-etal-2013}.
-Basal topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography and bathymetry dataset with a resolution of 1\,arc-minute \citep{data:etopo1}.
-Sea level is lowered by 120\,m and basal topography responds to ice load following a regional isostasy model that includes lithospheric flexure and mantle relaxation \citep{lingle-clark-1985}.
+This model thus accounts for a~weakening of till associated with the presence of marine sediments \citep{martin-etal-2011,aschwanden-etal-2013}.
+Basal topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography and bathymetry dataset with a~resolution of 1\,arc-minute \citep{data:etopo1}.
+Sea level is lowered by 120\,m and basal topography responds to ice load following a~regional isostasy model that includes lithospheric flexure and mantle relaxation \citep{lingle-clark-1985}.
 
@@
 
-Ice surface accumulation and ablation are computed from monthly mean surface air temperature and monthly precipitation by a temperature-index (positive degree-day) model \citep{hock-2003}.
+Ice surface accumulation and ablation are computed from monthly mean surface air temperature and monthly precipitation by a~temperature-index (positive degree-day) model \citep{hock-2003}.
 Ice accumulation is equal to precipitation when temperature is below 0\,\unit{\degree C}, and decreases to zero linearly with temperature between 0 and 2\,\unit{\degree C}.
@@
 The positive degree-day integral \citep{calov-greve-2005} is numerically approximated using week-long sub-intervals.
-It accounts for temperature variability assuming a normal distribution along a central (input) value.
-The temperature standard deviation is a constant model parameter and was assigned a value of 3.07\,\unit{\degree C}, which corresponds to the mean summer (JJA), model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from North American Regional Reanalysis data \citep{data:narr} in a manner similar to \citet{seguinot-2013}.
+It accounts for temperature variability assuming a~normal distribution along a~central (input) value.
+The temperature standard deviation is a~constant model parameter and was assigned a~value of 3.07\,\unit{\degree C}, which corresponds to the mean summer (JJA), model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from North American Regional Reanalysis data \citep{data:narr} in a~manner similar to \citet{seguinot-2013}.
 The ablation model incorporates degree-day factors of 3.04\,\unit{mm\,K^{-1}\,day^{-1}} for snow and 4.59\,\unit{mm\,K^{-1}\,day^{-1}} for ice, as derived from mass-balance measurements on contemporary glaciers from the Coast Mountains and Rocky Mountains in British Columbia \citep{shea-etal-2009}.
@@
 
-Prior to surface mass balance computation, the model dynamically applies a lapse-rate correction to surface air temperature.
+Prior to surface mass balance computation, the model dynamically applies a~lapse-rate correction to surface air temperature.
 This correction accounts for the evolution of ice thickness on the one hand, and differences between the climate forcing reference topography and the ice flow model basal topography on the other hand.
-It uses a reference topography distinct from the modelled basal topography and specific to each climate forcing dataset (Fig.~\ref{fig:topo}).
+It uses a~reference topography distinct from the modelled basal topography and specific to each climate forcing dataset (Fig.~\ref{fig:topo}).
 An annual lapse rate of 6\,\unit{\degree C\,km^{-1}} is used in all simulations.
@@
 
-WorldClim is a high resolution global climate dataset built from meteorological observations \citep{data:worldclim}.
-It was built by applying a spatial interpolation between a large set of measurements taken at terrestrial weather stations worldwide, using SRTM \citep{data:srtm} and GTOPO30 elevation data \citep{data:gtopo30}.
-This dataset provides a resolution much higher than attained by circulation models, however its reliability is variable due to the highly inhomogeneous distribution of measurements \citep{data:worldclim}.
+WorldClim is a~high resolution global climate dataset built from meteorological observations \citep{data:worldclim}.
+It was built by applying a~spatial interpolation between a~large set of measurements taken at terrestrial weather stations worldwide, using SRTM \citep{data:srtm} and GTOPO30 elevation data \citep{data:gtopo30}.
+This dataset provides a~resolution much higher than attained by circulation models, however its reliability is variable due to the highly inhomogeneous distribution of measurements \citep{data:worldclim}.
 
-Within our modelling domain, the spatial distributions of summer (JJA) and winter (DJF) air surface temperatures in WorldClim generally show a decrease of temperature from south to north (JJA) and from south-west to north-east (DJF), such that regions further inland experience colder winters (Fig.~\ref{fig:temp}).
+Within our modelling domain, the spatial distributions of summer (JJA) and winter (DJF) air surface temperatures in WorldClim generally show a~decrease of temperature from south to north (JJA) and from south-west to north-east (DJF), such that regions further inland experience colder winters (Fig.~\ref{fig:temp}).
 It should be noted that temperature gradients are much stronger in winter than in summer.
 Temperature rises well above zero during the summer months over the entire modelling domain, except for the highest mountain peaks, and even regions where mean annual temperatures are well below freezing point do experience warm summers.
-In other words, there is a strong contrast in seasonality between coastal temperate-climate and inland continental-climate regions.
+In other words, there is a~strong contrast in seasonality between coastal temperate-climate and inland continental-climate regions.
 
-The spatial distribution of JJA and DJF mean precipitation rates in WorldClim reveals a strong precipitation decline beyond the coastal regions (Fig.~\ref{fig:prec}), primarily as a result of the continuous orographic barrier formed by the Wrangell and St.-Elias mountains, the Coast Mountains and the Cascades (Fig.~\ref{fig:locmap}).
-A less pronounced precipitation peak is also visible over the interior ranges.
+The spatial distribution of JJA and DJF mean precipitation rates in WorldClim reveals a~strong precipitation decline beyond the coastal regions (Fig.~\ref{fig:prec}), primarily as a~result of the continuous orographic barrier formed by the Wrangell and St.-Elias mountains, the Coast Mountains and the Cascades (Fig.~\ref{fig:locmap}).
+A~less pronounced precipitation peak is also visible over the interior ranges.
 From an ice sheet mass balance point of view, these heterogeneities become even stronger due to the difference in timing of the precipitation peak through the year.
@@
 The CFSR climatology presents wave-like artefacts in its precipitation field (Fig.~\ref{fig:prec}).
-These are a common feature in GCM simulations due to the failure of spectral models to resolve the very local nature of topography-induced precipitation events, which becomes particularly visible in the high-resolution, less spatially-smoothed data.
+These are a~common feature in GCM simulations due to the failure of spectral models to resolve the very local nature of topography-induced precipitation events, which becomes particularly visible in the high-resolution, less spatially-smoothed data.
 Therefore, an alternative forcing was prepared for CFSR by smoothing its precipitation field.
-This was achieved by averaging data locally in a circular kernel of 7\,pixels in diameter prior to re-projection using the module r.neighbors from GRASS~GIS \citep{soft:grass}.
+This was achieved by averaging data locally in a~circular kernel of 7\,pixels in diameter prior to re-projection using the module r.neighbors from GRASS~GIS \citep{soft:grass}.
 Finally, input files for PISM were prepared using \citet{web:nc4py}.
@@
 Across the range of temperature offsets used, final glaciated area and final ice volume differ widely between different input climatologies (Fig.~\ref{fig:ivolarea}).
-For instance, using a 5\,\unit{\degree C} temperature offset, final glaciated area differs between WorldClim and CFSR forcing by a factor of~6 and final volume by a factor of~12 (Fig.~\ref{fig:ivolarea}).
+For instance, using a~5\,\unit{\degree C} temperature offset, final glaciated area differs between WorldClim and CFSR forcing by a~factor of~6 and final volume by a~factor of~12 (Fig.~\ref{fig:ivolarea}).
 For particular temperature offset values, given input climatologies result in similar final glaciated areas and ice volumes, yet regional patterns are different (Fig.~\ref{fig:extent}).
@@
 Under present climate (0\,\unit{\degree C} offset), WorldClim, ERA-Interim and NARR forcing produce local ice caps in areas presently glaciated (Fig.~\ref{fig:extent}).
-These results are fairly consistent with contemporary ice distribution, although both the 10~km horizontal resolution and the occurrence of glaciers and ice caps surface topographies in the ETOPO1 dataset used for basal topography prevent a meaningful detailed comparison.
+These results are fairly consistent with contemporary ice distribution, although both the 10~km horizontal resolution and the occurrence of glaciers and ice caps surface topographies in the ETOPO1 dataset used for basal topography prevent a~meaningful detailed comparison.
 CFSR forcing, on the other hand, yields continuous ice cover over the northern Coast Mountains, and NCEP/NCAR forcing results in ice sheet growth, both of which are largely inconsistent with the present state of glaciation.
@@
 
-To contrast the effect of the six different input climatologies, we plot simulations using a 5\,\unit{\degree C} temperature offset (Fig.~\ref{fig:cool05}).
+To contrast the effect of the six different input climatologies, we plot simulations using a~5\,\unit{\degree C} temperature offset (Fig.~\ref{fig:cool05}).
 The magnitude of glaciation differs dramatically from one climate forcing to another.
-Whereas the NCEP/NCAR and CFSR climatologies produce a large ice sheet that covers most of the model domain (Fig.~\ref{fig:cool05}, lower panels), WorldClim, ERA-Interim and NARR climatologies lead to a more restrictive ice cover (Fig.~\ref{fig:cool05}, upper panels).
+Whereas the NCEP/NCAR and CFSR climatologies produce a~large ice sheet that covers most of the model domain (Fig.~\ref{fig:cool05}, lower panels), WorldClim, ERA-Interim and NARR climatologies lead to a~more restrictive ice cover (Fig.~\ref{fig:cool05}, upper panels).
 The WorldClim run only resulted in the growth of mountain ice caps bounded to the highest topography.
@@
 
-These simulations of the LGM Cordilleran Ice Sheet show a very large sensitivity to the choice of climate forcing data used (Figs.~\ref{fig:ivolarea}--\ref{fig:cool05}).
+These simulations of the LGM Cordilleran Ice Sheet show a~very large sensitivity to the choice of climate forcing data used (Figs.~\ref{fig:ivolarea}--\ref{fig:cool05}).
 To understand the origin of the discrepancies between model results, we quantify differences in surface air temperature and precipitation rate between the input climatologies and test their effect on the ice sheet model.
-We then compare our findings to a geomorphological reconstruction of the LGM Cordilleran Ice Sheet margin, and discuss some of the key assumptions of this study.
+We then compare our findings to a~geomorphological reconstruction of the LGM Cordilleran Ice Sheet margin, and discuss some of the key assumptions of this study.
 
@@
 As previously outlined, significant differences in the distribution of winter (DJF) precipitation and summer (JJA) temperature, which are the predominant controls on accumulation and melt at the modelled ice sheet surface, exist between datasets (Figs.~\ref{fig:topo}--\ref{fig:prec}).
-In this section we use the WorldClim climatology, which is derived from observations, as a reference with which to compare other climate forcing data on land.
+In this section we use the WorldClim climatology, which is derived from observations, as a~reference with which to compare other climate forcing data on land.
 ERA-Interim, CFSR and NARR climatologies exhibit temperature distributions fairly consistent with WorldClim data (Fig.~\ref{fig:tempheatmap}).
 When examining the spatial distribution of temperature differences between reanalyses and WorldClim data, it appears that most of the disparity between reanalyses and observation data is caused by topographical detail at scales unresolved in the reanalysis datasets (Fig.~\ref{fig:tempdiff}).
-Surface air temperature data from the NCEP/NCAR climatology shows the largest disparity with that of WorldClim, with a significant cold anomaly over most of the modelling domain (Figs.~\ref{fig:tempheatmap}--\ref{fig:tempdiff}).
+Surface air temperature data from the NCEP/NCAR climatology shows the largest disparity with that of WorldClim, with a~significant cold anomaly over most of the modelling domain (Figs.~\ref{fig:tempheatmap}--\ref{fig:tempdiff}).
 
@@
 Most likely, this is the signature of an orographic precipitation effect.
-As described in section~\ref{sec:climate}, the topography of the Northern American Cordillera is such that its western ranges form a continuous orographic barrier, causing high precipitation along the Pacific coast while leaving much of the interior arid.
-However the ability of a GCM to reproduce these contrasts in precipitation is bounded by spatial resolution.
-In a model of coarser resolution, these high mountain ranges are reduced to smooth hills and the modelled distribution of precipitation is also smoother (Fig.~\ref{fig:oroprecip}).
+As described in section~\ref{sec:climate}, the topography of the Northern American Cordillera is such that its western ranges form a~continuous orographic barrier, causing high precipitation along the Pacific coast while leaving much of the interior arid.
+However the ability of a~GCM to reproduce these contrasts in precipitation is bounded by spatial resolution.
+In a~model of coarser resolution, these high mountain ranges are reduced to smooth hills and the modelled distribution of precipitation is also smoother (Fig.~\ref{fig:oroprecip}).
 However, this effect alone cannot explain the widespread and pronounced positive precipitation anomalies observed in CFSR data (Figs.~\ref{fig:precheatmap}--\ref{fig:precdiff}) despite of its high spatial resolution (Table~\ref{tab:reanalyses}).
@@
 
-To distinguish the effects of temperature and precipitation biases on the ice sheet model, we run a series of eight additional simulations.
-These new simulations use a “hybrid” climate forcing that consists of temperature data from WorldClim, combined with precipitation data from each of the four reanalyses, and vice-versa.
-A single temperature offset of 5\,\unit{\degree C} is used.
+To distinguish the effects of temperature and precipitation biases on the ice sheet model, we run a~series of eight additional simulations.
+These new simulations use a~“hybrid” climate forcing that consists of temperature data from WorldClim, combined with precipitation data from each of the four reanalyses, and vice-versa.
+A~single temperature offset of 5\,\unit{\degree C} is used.
 This value was chosen to allow for comparison with previous results (Fig.~\ref{fig:cool05}).
@@
 
-We compare modelled ice sheet geometries to a reconstruction of the ice margin at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) by \citet{dyke-2004}, based on glacial geomorphology and radiocarbon dating (Fig.~\ref{fig:locmap}).
+We compare modelled ice sheet geometries to a~reconstruction of the ice margin at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) by \citet{dyke-2004}, based on glacial geomorphology and radiocarbon dating (Fig.~\ref{fig:locmap}).
 This corresponds to the LGM extent of most of the Cordilleran ice sheet, which occurred later than for the Laurentide ice sheet \citep{porter-swanson-1998,dyke-2004,stroeven-etal-2010,stroeven-etal-inpress}.
@@
 Considering previously identified weaknesses of the CFSR and NCEP/NCAR data (Figs.~\ref{fig:tempheatmap}--\ref{fig:biatmbars}), we restricted our analysis to the WorldClim, ERA-Interim and NARR input climatologies.
-In order to compare modelled ice sheets of a similar size, we select for each climate forcing the simulation that leads to a final glaciated area closest to the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$.
+In order to compare modelled ice sheets of a~similar size, we select for each climate forcing the simulation that leads to a~final glaciated area closest to the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$.
 This corresponds to temperature offset values of 8\,\unit{\degree C} (WorldClim), 6\,\unit{\degree C} (ERA-Interim) and 7\,\unit{\degree C} (NARR).
-These qualitative “best” runs are presented in Figure~\ref{fig:best} along with their associated temperature depressions and a reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.
+These qualitative “best” runs are presented in Figure~\ref{fig:best} along with their associated temperature depressions and a~reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.
 
 Although this selection results in similarly sized ice sheets, noticeable differences in shape exist.
-The ERA-Interim simulation produces a more northerly-centred ice sheet with too much ice in the north and too little in the south.
+The ERA-Interim simulation produces a~more northerly-centred ice sheet with too much ice in the north and too little in the south.
 This may partly reflect unresolved orographic precipitation effects (Fig.~\ref{fig:oroprecip}) due to the coarser GCM resolution (Table~\ref{tab:reanalyses}).
@@
 The Cordilleran ice sheet initiated from the formation of mountain ice caps over the high mountain ranges \citep{clague-1989}.
-In our simulations, a continuous ice cover quickly forms over the western ranges, where precipitation rates are higher than in the rest of the domain.
+In our simulations, a~continuous ice cover quickly forms over the western ranges, where precipitation rates are higher than in the rest of the domain.
 In reality, this continuous ice cover enhanced the topographical barrier already formed by the western ranges, and would have resulted in less precipitation and warmer air in the interior.
-This process of ``self-inhibiting growth'' is not captured in our model but was demonstrated to be important in a similar setting over Greenland, where it may have been limiting the advance of the ice sheet during its build-up \citep{langen-etal-2012}.
+This process of ``self-inhibiting growth'' is not captured in our model but was demonstrated to be important in a~similar setting over Greenland, where it may have been limiting the advance of the ice sheet during its build-up \citep{langen-etal-2012}.
 
 Secondly, our simulations produce anomalous ice cover on parts of the continental shelf in the Arctic Ocean (Fig.~\ref{fig:best}).
-These regions experience a marine climate, including lower summer temperature than on the adjacent land (Fig.~\ref{fig:temp}).
+These regions experience a~marine climate, including lower summer temperature than on the adjacent land (Fig.~\ref{fig:temp}).
 However in our simulations, sea-level is lowered by 120~m, turning large parts of the low-sloping continental shelf into land.
@@
 
-A key assumption for the present study is our choice of 10\,ka for the duration of all our simulations.
+A~key assumption for the present study is our choice of 10\,ka for the duration of all our simulations.
 Although geomorphological data shows that the period of Cordilleran Ice Sheet inception, from nearly ice-free to full glacial conditions, could not have been much longer than 10\,ka, they provide no lower bound to this value.
@@
 
-This experiment shows that a shorter start-up period leads to more restricted ice cover in the eastern part of the modelling domain, where precipitation rate controls ice advance, but more extensive glaciation in the south-western domain along the Pacific coast, where temperature is the limiting factor (Fig.~\ref{fig:durationstack}).
-Hence, reducing simulation duration leads to a closer match with the LGM ice sheet margin by \citet{dyke-2004} in some parts of the modelling domain, but higher discrepancy in other parts.
+This experiment shows that a~shorter start-up period leads to more restricted ice cover in the eastern part of the modelling domain, where precipitation rate controls ice advance, but more extensive glaciation in the south-western domain along the Pacific coast, where temperature is the limiting factor (Fig.~\ref{fig:durationstack}).
+Hence, reducing simulation duration leads to a~closer match with the LGM ice sheet margin by \citet{dyke-2004} in some parts of the modelling domain, but higher discrepancy in other parts.
 
@@
 To better understand the transient character of the LGM Cordilleran Ice Sheet, time-dependent palaeo-climate forcing would be needed.
-Although it is clear that past temperature changes over the region were neither homogeneous nor constant in time, and probably associated with precipitation changes, their patterns are not trivial and may potentially display a strong inter-dependence with the evolution of the Cordilleran and Laurentide ice sheets.
-A more correct, yet more complex approach would be to use coupling to a GCM of intermediate complexity \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013}.
+Although it is clear that past temperature changes over the region were neither homogeneous nor constant in time, and probably associated with precipitation changes, their patterns are not trivial and may potentially display a~strong inter-dependence with the evolution of the Cordilleran and Laurentide ice sheets.
+A~more correct, yet more complex approach would be to use coupling to a~GCM of intermediate complexity \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013}.
 
@@
 Based on scarce observations over the Greenland Ice Sheet, different parametrizations of this effect have been developed \citep{janssens-huybrechts-2000,fausto-etal-2009b}.
-However, these parametrizations were shown to disagree with more complex models \citep{reijmer-etal-2012}, and it remains undemonstrated whether their application to a different region and time period is sound.
-In addition, we simulate temperature variability by using a constant, uniform value of daily temperature standard deviation.
+However, these parametrizations were shown to disagree with more complex models \citep{reijmer-etal-2012}, and it remains undemonstrated whether their application to a~different region and time period is sound.
+In addition, we simulate temperature variability by using a~constant, uniform value of daily temperature standard deviation.
 This approach implies large biases of surface mass-balance \citep{charbit-etal-2013,rau-rogozhina-2013,seguinot-2013}, particularly over regions with such various climates as experienced within our model domain.
@@
 
-Our study shows a strong dependency of ice sheet model results on the choice of climate forcing data.
+Our study shows a~strong dependency of ice sheet model results on the choice of climate forcing data.
 For three of the four reanalysis datasets used, precipitation rate, over surface air temperature, causes much of the variation in modelled ice sheet outlines and volumes.
@@
 Climate forcing from CFSR and the NCEP/NCAR reanalysis produce largely oversized ice sheets due to too high precipitation rates, and in the second case, too low surface air temperature.
-The ERA-Interim data used in this study produces more reasonable results, but it may be too coarse to accurately resolve the spatial distribution of orographic precipitation associated with the rugged topography of the study area, resulting in a misplaced ice coverage in regard to geomorphological data.
+The ERA-Interim data used in this study produces more reasonable results, but it may be too coarse to accurately resolve the spatial distribution of orographic precipitation associated with the rugged topography of the study area, resulting in a~misplaced ice coverage in regard to geomorphological data.
 Therefore, we retain NARR data for forcing simulations of the Cordilleran Ice Sheet in the future.
@@
 At LGM, it is most likely that the presence of ice sheets significantly affected circulation patterns and distribution of precipitation.
-At the cost of greater computational expense, a more accurate representation of palaeo-climate may allow for a better fit between model results and the mapped LGM margin.
+At the cost of greater computational expense, a~more accurate representation of palaeo-climate may allow for a~better fit between model results and the mapped LGM margin.
 
@@
 	\caption{Topography maps from WorldClim \citep{data:worldclim}, ERA-Interim reanalysis \citep{data:erai}, North American Regional Reanalysis \citep[NARR;][]{data:narr}, ETOPO1 \citep{data:etopo1}, Climate Forecast System Reanalysis \citep[CFSR;][]{data:cfsr}, and NCEP/NCAR reanalysis \citep{data:ncar}.
-Whereas ETOPO1 is used as basal topography for the ice sheet model, all other data are used as a reference for temperature lapse-rate corrections.
+Whereas ETOPO1 is used as basal topography for the ice sheet model, all other data are used as a~reference for temperature lapse-rate corrections.
 Fig.~\ref{fig:topo}-\ref{fig:durationstack} are drawn using Matplotlib \citep{soft:mpl}.}
@@
 	\end{center}
-	\caption{Total glaciated area and ice volume after 10\,ka as a function of temperature offset for each climate forcing used.}
+	\caption{Total glaciated area and ice volume after 10\,ka as a~function of temperature offset for each climate forcing used.}
 	\label{fig:ivolarea}
@@
 	\end{center}
-	\caption{Extent of ice cover after 10\,ka as a function of applied temperature offsets for each climate forcing used.
-A reconstructed LGM ice sheet margin by \citet{dyke-2004} (Fig.~\ref{fig:locmap}) is given for reference (black line).}
+	\caption{Extent of ice cover after 10\,ka as a~function of applied temperature offsets for each climate forcing used.
+A~reconstructed LGM ice sheet margin by \citet{dyke-2004} (Fig.~\ref{fig:locmap}) is given for reference (black line).}
 	\label{fig:extent}
@@
 	\end{center}
-	\caption{Ice surface topography (1 km contours) and velocity (\unit{m\,a^{-1}}) after 10\,ka under a climate 5\,\unit{\degree C} colder than present for each climate forcing used.}
+	\caption{Ice surface topography (1 km contours) and velocity (\unit{m\,a^{-1}}) after 10\,ka under a~climate 5\,\unit{\degree C} colder than present for each climate forcing used.}
 	\label{fig:cool05}
@@
 	\end{center}
-	\caption{Density maps showing a comparison of summer (JJA) surface air temperature data from the WorldClim climatology, against that of each reanalysis.
+	\caption{Density maps showing a~comparison of summer (JJA) surface air temperature data from the WorldClim climatology, against that of each reanalysis.
 Apparent horizontal lines are an artefact of different horizontal resolutions.
-Colour mapping is based on a logarythmic scale.
+Colour mapping is based on a~logarythmic scale.
 Note the cold bias of NCEP/NCAR data relative to WorldClim data.}
@@
 	\end{center}
-	\caption{Density maps showing a comparison of winter (DJF) precipitation rate from the WorldClim climatology, against that of each reanalysis.
+	\caption{Density maps showing a~comparison of winter (DJF) precipitation rate from the WorldClim climatology, against that of each reanalysis.
 Apparent horizontal lines are an artefact of different horizontal resolutions.
 Apparent vertical lines at low values are an artefact of the low precision of WorldClim data.
-Colour mapping is based on a logarythmic scale.
+Colour mapping is based on a~logarythmic scale.
 Note the wet bias of all reanalysis data relative to WorldClim data.}
@@
 	\end{center}
-	\caption{Schematic representation of the orographic precipitation effect over a mountain range.
-In a GCM of low resolution, the precipitation peak appears downwind-shifted and smoother and the precipitation shadow is less pronounced than in a high resolution GCM.}
+	\caption{Schematic representation of the orographic precipitation effect over a~mountain range.
+In a~GCM of low resolution, the precipitation peak appears downwind-shifted and smoother and the precipitation shadow is less pronounced than in a~high resolution GCM.}
 	\label{fig:oroprecip}
@@
 In other words, the upper row shows the effect of temperature anomalies, and the lower row the effect of precipitation anomalies, for each reanalysis, relative to WorldClim data.
-Each simulation uses a 5\,\unit{\degree C} offset for comparison with figure~\ref{fig:cool05}}
+Each simulation uses a~5\,\unit{\degree C} offset for comparison with figure~\ref{fig:cool05}}
 	\label{fig:biatm}
@@
 	\caption{Ice surface topography (1\,km contours) after 10\,ka using temperature offsets resulting in glaciated areas of circa $2\,\times10^6\,\unit{km^2}$.
-A reconstructed LGM ice sheet margin by \citet{dyke-2004} (Fig.~\ref{fig:locmap}) is given for reference (blue line).}
+A~reconstructed LGM ice sheet margin by \citet{dyke-2004} (Fig.~\ref{fig:locmap}) is given for reference (blue line).}
 	\label{fig:best}
@@
 Right panel: modelled ice margin corresponding to temperature offsets from 7 to 11\,\unit{\degree C} when total glaciated area reach the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$.
-A reconstructed LGM ice sheet margin by \citet{dyke-2004} (Fig.~\ref{fig:locmap}) is given for reference (grey shading).
+A~reconstructed LGM ice sheet margin by \citet{dyke-2004} (Fig.~\ref{fig:locmap}) is given for reference (grey shading).
 Note that shorter (and colder) simulations lead to more restrictive glaciation of the continental eastern margin but further ice extent in the maritime south-western part of the modelling domain.}

### 2014-03-20 abstract.tex
Typesetting: title and authors

@@
 
-\textbf{The effect of climate forcing on numerical simulations of the Cordilleran Ice Sheet at the Last Glacial Maximum}
+\textbf{The effect of climate forcing on numerical simulations of the Cordilleran ice sheet at the Last Glacial Maximum}
 
-J. Seguinot\textsuperscript{1},
-C. Khroulev\textsuperscript{2},
-I. Rogozhina\textsuperscript{3},
-A. P. Stroeven\textsuperscript{1},
-and Q. Zhang\textsuperscript{1}
+J.~Seguinot\textsuperscript{1},
+C.~Khroulev\textsuperscript{2},
+I.~Rogozhina\textsuperscript{3},
+A.~P.~Stroeven\textsuperscript{1},
+and Q.~Zhang\textsuperscript{1}
 

### 2014-03-20 cordillera-climate.tex
Typesetting: title and authors

@@
 % ----------------------------------------------------------------------
-\begin{document}
+\begin{document}\hack{\sloppy}
 % ----------------------------------------------------------------------
@@
 % Title
-\title{The effect of climate forcing on numerical simulations of the Cordilleran Ice Sheet at the Last Glacial Maximum}
+\title{The effect of climate forcing on numerical simulations of the Cordilleran ice sheet at the Last Glacial Maximum}
 \runningtitle{Climate forcing for Cordilleran ice sheet simulations}
@@
 % Authors
-\author[1]{Julien Seguinot}
-\author[2]{Constantine Khroulev}
-\author[3]{Irina Rogozhina}
-\author[1]{Arjen P. Stroeven}
-\author[1]{Qiong Zhang}
-\runningauthor{J. Seguinot et al.}
-\correspondence{J. Seguinot\\ (julien.seguinot@natgeo.su.se)}
+\author[1]{J.~Seguinot}
+\author[2]{C.~Khroulev}
+\author[3]{I.~Rogozhina}
+\author[1]{A.~P.~Stroeven}
+\author[1]{Q.~Zhang}
+\runningauthor{J.~Seguinot et~al.}
+\correspondence{J.~Seguinot (julien.seguinot@natgeo.su.se)}
 
@@
 % Title
+\firstpage{1}
 \maketitle

### 2014-03-20 cordillera-climate.tex
Typesetting: decapitalize "ice sheet"

@@
 \begin{abstract}
-We present an ensemble of numerical simulations of the Cordilleran Ice Sheet during the Last Glacial Maximum performed with the Parallel Ice Sheet Model (PISM), applying temperature offsets to the present-day climatologies from five different datasets.
+We present an ensemble of numerical simulations of the Cordilleran ice sheet during the Last Glacial Maximum performed with the Parallel Ice Sheet Model (PISM), applying temperature offsets to the present-day climatologies from five different datasets.
 Monthly mean surface air temperature and precipitation from WorldClim, the NCEP/NCAR reanalysis, the ERA-Interim reanalysis, the Climate Forecast System Reanalysis and the North American Regional Reanalysis are used to compute surface mass balance in a~positive degree-day model.
@@
 We assess model performance against a~geomorphological reconstruction of the ice margin at the Last Glacial Maximum, and suggest that part of the mismatch is due to unresolved orographic precipitation effects caused by the coarse resolution of reanalysis data.
-The best match between model output and the reconstructed ice margin is obtained using the high-resolution North American Regional Reanalysis, which we retain for simulations of the Cordilleran Ice Sheet in the future.
+The best match between model output and the reconstructed ice margin is obtained using the high-resolution North American Regional Reanalysis, which we retain for simulations of the Cordilleran ice sheet in the future.
 \end{abstract}
@@
 
-At the Last Glacial Maximum (LGM), glaciers of a~size comparable to the present Greenland and Antarctic ice sheets covered parts of Northern America (Laurentide, Cordilleran and Innuitian ice sheets) and northern Eurasia (Fennoscandian Ice Sheet).
+At the Last Glacial Maximum (LGM), glaciers of a~size comparable to the present Greenland and Antarctic ice sheets covered parts of Northern America (Laurentide, Cordilleran and Innuitian ice sheets) and northern Eurasia (Fennoscandian ice sheet).
 Numerical modelling of these former ice masses allows for a~comparison between glaciological theories embedded in the models and geomorphological traces underpinning palaeo-glaciological reconstructions.
@@
 Hence, rather than using GCM output, we force our model with climate reanalysis data, which builds on observational information through data assimilation \citep{bengtsson-etal-2007}.
-Furthermore, we focus our study regionally on the former Cordilleran Ice Sheet in western North America.
+Furthermore, we focus our study regionally on the former Cordilleran ice sheet in western North America.
 
-The Cordilleran Ice Sheet (Fig.~\ref{fig:locmap}) covered an area that presently experiences strong regional variations in climate.
+The Cordilleran ice sheet (Fig.~\ref{fig:locmap}) covered an area that presently experiences strong regional variations in climate.
 From a~numerical modelling perspective, it is one of the least studied palaeo-ice sheets of the Northern Hemisphere, despite the fact that significant stratigraphical, geomorphological, and chronological data is available to constrain its extent \citep{jackson-clague-1991,porter-swanson-1998,dukrodkin-1999,dyke-2004,kaufman-manley-2004,kleman-etal-2010,stroeven-etal-2010,stroeven-etal-inpress,margold-etal-2011}.
-The Cordilleran Ice Sheet has previously been modelled as part of efforts to simulate ice sheets in North America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} and world-wide \citep{yoshimori-etal-2001}.
+The Cordilleran ice sheet has previously been modelled as part of efforts to simulate ice sheets in North America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} and world-wide \citep{yoshimori-etal-2001}.
 While these studies generally reproduce the magnitude of North American glaciation at LGM reasonably well, there exists a~tendency in the simulations that are independent of ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and interior Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.
 
-Here we use PISM, a~Parallel Ice Sheet Model \citep{web:pism}, to simulate the extent and thickness of the Cordilleran Ice Sheet at the LGM (Fig.~\ref{fig:locmap}).
+Here we use PISM, a~Parallel Ice Sheet Model \citep{web:pism}, to simulate the extent and thickness of the Cordilleran ice sheet at the LGM (Fig.~\ref{fig:locmap}).
 We force our model with multiple climate datasets and compare the modelled ice extent to a~geomorphological reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.
-We aim to determine the climate dataset which is most suited for simulation of the Cordilleran Ice Sheet and to use as input for future, transient studies over a~glacial cycle.
+We aim to determine the climate dataset which is most suited for simulation of the Cordilleran ice sheet and to use as input for future, transient studies over a~glacial cycle.
 
@@
 Given basal topography, sea level, geothermal heat flux and climate forcing, the model computes ice extent and thickness, its thermal and dynamic state, and the associated lithospheric response.
-Our modelling domain encompasses the entire area covered by the Cordilleran Ice Sheet at LGM including independent ice growing on the western Alaskan ranges and the Brooks Range and intermediate expanses of ice-free terrain in northern Yukon Territory and interior Alaska (Fig.~\ref{fig:locmap}).
+Our modelling domain encompasses the entire area covered by the Cordilleran ice sheet at LGM including independent ice growing on the western Alaskan ranges and the Brooks Range and intermediate expanses of ice-free terrain in northern Yukon Territory and interior Alaska (Fig.~\ref{fig:locmap}).
 
-As we aim to model glacial inception and growth of the Cordilleran Ice Sheet towards a~configuration as last attained during the LGM, we mimic palaeo-climatic conditions by applying constant temperature offsets homogeneously over the modelling domain.
-Each simulation starts from ice-free conditions and runs for 10\,ka, a~time interval representative of the rapid build-up of the last Cordilleran Ice Sheet from nearly ice-free to full glacial conditions \citep{clague-1989,stroeven-etal-2010}.
+As we aim to model glacial inception and growth of the Cordilleran ice sheet towards a~configuration as last attained during the LGM, we mimic palaeo-climatic conditions by applying constant temperature offsets homogeneously over the modelling domain.
+Each simulation starts from ice-free conditions and runs for 10\,ka, a~time interval representative of the rapid build-up of the last Cordilleran ice sheet from nearly ice-free to full glacial conditions \citep{clague-1989,stroeven-etal-2010}.
 Our simulations are performed in parallel on 32 cores at the Swedish National Supercomputing Center.
@@
 Although this uniform value does not account for the high spatial geothermal variability in the region, it is on average representative of available heat flow measurements \citep{artemieva-mooney-2001,blackwell-richards-2004}.
-Whereas spatial variations in the geothermal heat flux may significantly affect basal conditions \citep{pattyn-2010}, this study focuses on reconstructing the former extent of the Cordilleran Ice Sheet, on which it has relatively little effect \citep{rogozhina-etal-2012}.
+Whereas spatial variations in the geothermal heat flux may significantly affect basal conditions \citep{pattyn-2010}, this study focuses on reconstructing the former extent of the Cordilleran ice sheet, on which it has relatively little effect \citep{rogozhina-etal-2012}.
 
@@
 
-These simulations of the LGM Cordilleran Ice Sheet show a~very large sensitivity to the choice of climate forcing data used (Figs.~\ref{fig:ivolarea}--\ref{fig:cool05}).
+These simulations of the LGM Cordilleran ice sheet show a~very large sensitivity to the choice of climate forcing data used (Figs.~\ref{fig:ivolarea}--\ref{fig:cool05}).
 To understand the origin of the discrepancies between model results, we quantify differences in surface air temperature and precipitation rate between the input climatologies and test their effect on the ice sheet model.
-We then compare our findings to a~geomorphological reconstruction of the LGM Cordilleran Ice Sheet margin, and discuss some of the key assumptions of this study.
+We then compare our findings to a~geomorphological reconstruction of the LGM Cordilleran ice sheet margin, and discuss some of the key assumptions of this study.
 
@@
 Considering previously identified weaknesses of the CFSR and NCEP/NCAR data (Figs.~\ref{fig:tempheatmap}--\ref{fig:biatmbars}), we restricted our analysis to the WorldClim, ERA-Interim and NARR input climatologies.
-In order to compare modelled ice sheets of a~similar size, we select for each climate forcing the simulation that leads to a~final glaciated area closest to the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$.
+In order to compare modelled ice sheets of a~similar size, we select for each climate forcing the simulation that leads to a~final glaciated area closest to the approximate size of the LGM Cordilleran ice sheet of $2\,\times10^6\,\unit{km^2}$.
 This corresponds to temperature offset values of 8\,\unit{\degree C} (WorldClim), 6\,\unit{\degree C} (ERA-Interim) and 7\,\unit{\degree C} (NARR).
@@
 This can be seen for instance in the south-eastern part of our modelling domain, where the modelled ice sheets cover an ice-free corridor of the LGM reconstruction (Fig.~\ref{fig:best}).
-However, such comparisons have been hampered by the fact that we have not considered the buttressing effect of the Laurentide Ice Sheet, which would have inhibited the Cordilleran Ice Sheet from advancing onto the Canadian Prairies.
+However, such comparisons have been hampered by the fact that we have not considered the buttressing effect of the Laurentide ice sheet, which would have inhibited the Cordilleran ice sheet from advancing onto the Canadian Prairies.
 In addition, potential effects of the growing ice-sheet on atmospheric circulation and precipitation changes are not included in our model.
@@
 
-Given the lack of WorldClim data offshore, we consider the NARR climate forcing as best suited for simulations of the Cordilleran Ice Sheet among those tested, and under the assumptions of the present study.
+Given the lack of WorldClim data offshore, we consider the NARR climate forcing as best suited for simulations of the Cordilleran ice sheet among those tested, and under the assumptions of the present study.
 
@@
 A~key assumption for the present study is our choice of 10\,ka for the duration of all our simulations.
-Although geomorphological data shows that the period of Cordilleran Ice Sheet inception, from nearly ice-free to full glacial conditions, could not have been much longer than 10\,ka, they provide no lower bound to this value.
+Although geomorphological data shows that the period of Cordilleran ice sheet inception, from nearly ice-free to full glacial conditions, could not have been much longer than 10\,ka, they provide no lower bound to this value.
 
-To test the effect of an inception period of shorter duration on model results, we use simulations run under NARR climate forcing with temperature offsets of 7 to 11\,\unit{\degree C}, and compare modelled ice sheet extent when glaciated area reaches the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$.
+To test the effect of an inception period of shorter duration on model results, we use simulations run under NARR climate forcing with temperature offsets of 7 to 11\,\unit{\degree C}, and compare modelled ice sheet extent when glaciated area reaches the approximate size of the LGM Cordilleran ice sheet of $2\,\times10^6\,\unit{km^2}$.
 
@@
 Our representation of climate history by constant temperature offsets is voluntarily simplistic.
-To better understand the transient character of the LGM Cordilleran Ice Sheet, time-dependent palaeo-climate forcing would be needed.
+To better understand the transient character of the LGM Cordilleran ice sheet, time-dependent palaeo-climate forcing would be needed.
 Although it is clear that past temperature changes over the region were neither homogeneous nor constant in time, and probably associated with precipitation changes, their patterns are not trivial and may potentially display a~strong inter-dependence with the evolution of the Cordilleran and Laurentide ice sheets.
@@
 Refreezing of melted snow and ice and water retention in the snow pack was shown to influence surface mass balance of glaciers \citep{trabant-mayo-1985}, yet it is not accounted for in our PDD model.
-Based on scarce observations over the Greenland Ice Sheet, different parametrizations of this effect have been developed \citep{janssens-huybrechts-2000,fausto-etal-2009b}.
+Based on scarce observations over the Greenland ice sheet, different parametrizations of this effect have been developed \citep{janssens-huybrechts-2000,fausto-etal-2009b}.
 However, these parametrizations were shown to disagree with more complex models \citep{reijmer-etal-2012}, and it remains undemonstrated whether their application to a~different region and time period is sound.
@@
 For three of the four reanalysis datasets used, precipitation rate, over surface air temperature, causes much of the variation in modelled ice sheet outlines and volumes.
-Furthermore, the spatial resolution of input data appears critical for providing the ice sheet model with an accurate precipitation field, confirming results obtained by \citet{quiquet-etal-2012} from various GCM forcing datasets over the Greenland Ice Sheet.
+Furthermore, the spatial resolution of input data appears critical for providing the ice sheet model with an accurate precipitation field, confirming results obtained by \citet{quiquet-etal-2012} from various GCM forcing datasets over the Greenland ice sheet.
 
-For the Cordilleran Ice Sheet, we achieve the best fit to the mapped LGM margin by \citet{dyke-2004} using climate forcing from the high-resolution interpolated observational dataset WorldClim \citep{data:worldclim} and the North American Regional Reanalysis \citep[NARR;][]{data:narr}.
+For the Cordilleran ice sheet, we achieve the best fit to the mapped LGM margin by \citet{dyke-2004} using climate forcing from the high-resolution interpolated observational dataset WorldClim \citep{data:worldclim} and the North American Regional Reanalysis \citep[NARR;][]{data:narr}.
 The latter dataset is preferable in our case due to the lack of WorldClim data offshore.
@@
 The ERA-Interim data used in this study produces more reasonable results, but it may be too coarse to accurately resolve the spatial distribution of orographic precipitation associated with the rugged topography of the study area, resulting in a~misplaced ice coverage in regard to geomorphological data.
-Therefore, we retain NARR data for forcing simulations of the Cordilleran Ice Sheet in the future.
+Therefore, we retain NARR data for forcing simulations of the Cordilleran ice sheet in the future.
 
-One must keep in mind, however, that these results are only accurate for our choices of ice-sheet model (PISM), surface mass balance model (PDD), study object (the Cordilleran Ice Sheet), and, most importantly, palaeo-climate representation (temperature offsets).
+One must keep in mind, however, that these results are only accurate for our choices of ice-sheet model (PISM), surface mass balance model (PDD), study object (the Cordilleran ice sheet), and, most importantly, palaeo-climate representation (temperature offsets).
 At LGM, it is most likely that the presence of ice sheets significantly affected circulation patterns and distribution of precipitation.
@@
 P.
-Stroeven initiated the Cordilleran Ice Sheet modelling project; and Q.
+Stroeven initiated the Cordilleran ice sheet modelling project; and Q.
 Zhang prepared some of the forcing data.
@@
 Modelled glaciated area, using NARR forcing data, and temperature offsets from 0 to 15\,\unit{\degree C}, solid lines corresponding to values from 7 to 11\,\unit{\degree C}.
-Right panel: modelled ice margin corresponding to temperature offsets from 7 to 11\,\unit{\degree C} when total glaciated area reach the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$.
+Right panel: modelled ice margin corresponding to temperature offsets from 7 to 11\,\unit{\degree C} when total glaciated area reach the approximate size of the LGM Cordilleran ice sheet of $2\,\times10^6\,\unit{km^2}$.
 A~reconstructed LGM ice sheet margin by \citet{dyke-2004} (Fig.~\ref{fig:locmap}) is given for reference (grey shading).

### 2014-03-20 cordillera-climate.tex
Typesetting: unit formatting

@@
 Climatologies obtained from GCM palaeo-climate simulations such as produced within the Paleoclimate Modelling Intercomparison Project \citep[PMIP;][]{joussaume-taylor-1995} provide perhaps a~more reasonable representation of past climate.
-However, as such climatologies are only available for specific periods of time, using them as climate forcing for an ice sheet model requires either an assumption of steady-state ice sheet response to climatic fluctuations \citep{huybrechts-tsiobbel-1996}, or interpolation through time between climatologies from different periods, which can be linear \citep{charbit-etal-2002}, or modulated by a~``glacial index'' weighting function derived from ice core $\delta^{18}$O records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}.
+However, as such climatologies are only available for specific periods of time, using them as climate forcing for an ice sheet model requires either an assumption of steady-state ice sheet response to climatic fluctuations \citep{huybrechts-tsiobbel-1996}, or interpolation through time between climatologies from different periods, which can be linear \citep{charbit-etal-2002}, or modulated by a~``glacial index'' weighting function derived from ice core \chem{\delta^{18}O} records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}.
 An important drawback in this approach is that GCM palaeo-climate simulations themselves rely on an accurate description of the global surface topography and therefore include global ice sheet reconstructions such as ICE-4G \citep{peltier-1994} for their surface topographic boundary condition.
@@
 As we aim to model glacial inception and growth of the Cordilleran ice sheet towards a~configuration as last attained during the LGM, we mimic palaeo-climatic conditions by applying constant temperature offsets homogeneously over the modelling domain.
-Each simulation starts from ice-free conditions and runs for 10\,ka, a~time interval representative of the rapid build-up of the last Cordilleran ice sheet from nearly ice-free to full glacial conditions \citep{clague-1989,stroeven-etal-2010}.
+Each simulation starts from ice-free conditions and runs for 10\,kyr, a~time interval representative of the rapid build-up of the last Cordilleran ice sheet from nearly ice-free to full glacial conditions \citep{clague-1989,stroeven-etal-2010}.
 Our simulations are performed in parallel on 32 cores at the Swedish National Supercomputing Center.
@@
 The Shallow Shelf Approximation (SSA) is used as a~``sliding law'' for the Shallow Ice Approximation (SIA) \citep{bueler-brown-2009,winkelmann-etal-2011}.
-SIA and SSA velocities are computed by finite difference methods on a~10\,km-resolution horizontal grid of 300 by 150 points (the modelling domain).
+SIA and SSA velocities are computed by finite difference methods on a~10\,km resolution horizontal grid of 300 by 150 points (the modelling domain).
 Ice softness depends on temperature and water content through an enthalpy formulation \citep{aschwanden-blatter-2009,aschwanden-etal-2012}.
@@
 This model thus accounts for a~weakening of till associated with the presence of marine sediments \citep{martin-etal-2011,aschwanden-etal-2013}.
-Basal topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography and bathymetry dataset with a~resolution of 1\,arc-minute \citep{data:etopo1}.
+Basal topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography and bathymetry dataset with a~resolution of 1\,arc-min \citep{data:etopo1}.
 Sea level is lowered by 120\,m and basal topography responds to ice load following a~regional isostasy model that includes lithospheric flexure and mantle relaxation \citep{lingle-clark-1985}.
@@
 Ice surface accumulation and ablation are computed from monthly mean surface air temperature and monthly precipitation by a~temperature-index (positive degree-day) model \citep{hock-2003}.
-Ice accumulation is equal to precipitation when temperature is below 0\,\unit{\degree C}, and decreases to zero linearly with temperature between 0 and 2\,\unit{\degree C}.
-Ice ablation is computed from the number of positive degree-days, defined as the integral of temperatures above 0\,\unit{\degree C} in one year.
+Ice accumulation is equal to precipitation when temperature is below 0\,\unit{{\degree}C}, and decreases to zero linearly with temperature between 0 and 2\,\unit{{\degree}C}.
+Ice ablation is computed from the number of positive degree-days, defined as the integral of temperatures above 0\,\unit{{\degree}C} in one year.
 
@@
 It accounts for temperature variability assuming a~normal distribution along a~central (input) value.
-The temperature standard deviation is a~constant model parameter and was assigned a~value of 3.07\,\unit{\degree C}, which corresponds to the mean summer (JJA), model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from North American Regional Reanalysis data \citep{data:narr} in a~manner similar to \citet{seguinot-2013}.
+The temperature standard deviation is a~constant model parameter and was assigned a~value of 3.07\,\unit{{\degree}C}, which corresponds to the mean summer (JJA), model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from North American Regional Reanalysis data \citep{data:narr} in a~manner similar to \citet{seguinot-2013}.
 The ablation model incorporates degree-day factors of 3.04\,\unit{mm\,K^{-1}\,day^{-1}} for snow and 4.59\,\unit{mm\,K^{-1}\,day^{-1}} for ice, as derived from mass-balance measurements on contemporary glaciers from the Coast Mountains and Rocky Mountains in British Columbia \citep{shea-etal-2009}.
@@
 It uses a~reference topography distinct from the modelled basal topography and specific to each climate forcing dataset (Fig.~\ref{fig:topo}).
-An annual lapse rate of 6\,\unit{\degree C\,km^{-1}} is used in all simulations.
+An annual lapse rate of 6\,\unit{{\degree}C\,km^{-1}} is used in all simulations.
 No corrections apply to precipitation changes with elevation.
@@
 
-WorldClim, NCEP/NCAR, ERA-Interim, CFSR and NARR climatologies were re-projected to the Canadian Atlas Lambert conformal conic projection (EPSG code~3978) and bilinearly interpolated to the 10\,km-resolution model grid using the module r.proj from GRASS~GIS \citep{soft:grass}.
+WorldClim, NCEP/NCAR, ERA-Interim, CFSR and NARR climatologies were re-projected to the Canadian Atlas Lambert conformal conic projection (EPSG code~3978) and bilinearly interpolated to the 10\,km resolution model grid using the module r.proj from GRASS~GIS \citep{soft:grass}.
 In addition, WorldClim data was extrapolated to cover grid points in the ocean using the nearest-neighbour algorithm \citep{soft:scipy}.
@@
 
-Using WorldClim, NCEP/NCAR, ERA-Interim, CFSR (smoothed and original datasets) and NARR input climatologies as climate forcing, we run 66~simulations of glacial inception and growth of the Cordilleran ice sheet over 10~ka using temperature offsets ranging from 0 to 10\,\unit{\degree C}.
+Using WorldClim, NCEP/NCAR, ERA-Interim, CFSR (smoothed and original datasets) and NARR input climatologies as climate forcing, we run 66~simulations of glacial inception and growth of the Cordilleran ice sheet over 10\,kyr using temperature offsets ranging from 0 to 10\,\unit{{\degree}C}.
 Significant differences in patterns of ice sheet growth, final glacial extent, and volume are attained (Figs.
@@
 Across the range of temperature offsets used, final glaciated area and final ice volume differ widely between different input climatologies (Fig.~\ref{fig:ivolarea}).
-For instance, using a~5\,\unit{\degree C} temperature offset, final glaciated area differs between WorldClim and CFSR forcing by a~factor of~6 and final volume by a~factor of~12 (Fig.~\ref{fig:ivolarea}).
+For instance, using a~5\,\unit{{\degree}C} temperature offset, final glaciated area differs between WorldClim and CFSR forcing by a~factor of~6 and final volume by a~factor of~12 (Fig.~\ref{fig:ivolarea}).
 For particular temperature offset values, given input climatologies result in similar final glaciated areas and ice volumes, yet regional patterns are different (Fig.~\ref{fig:extent}).
 
-Under present climate (0\,\unit{\degree C} offset), WorldClim, ERA-Interim and NARR forcing produce local ice caps in areas presently glaciated (Fig.~\ref{fig:extent}).
-These results are fairly consistent with contemporary ice distribution, although both the 10~km horizontal resolution and the occurrence of glaciers and ice caps surface topographies in the ETOPO1 dataset used for basal topography prevent a~meaningful detailed comparison.
+Under present climate (0\,\unit{{\degree}C} offset), WorldClim, ERA-Interim and NARR forcing produce local ice caps in areas presently glaciated (Fig.~\ref{fig:extent}).
+These results are fairly consistent with contemporary ice distribution, although both the 10\,\unit{km} horizontal resolution and the occurrence of glaciers and ice caps surface topographies in the ETOPO1 dataset used for basal topography prevent a~meaningful detailed comparison.
 CFSR forcing, on the other hand, yields continuous ice cover over the northern Coast Mountains, and NCEP/NCAR forcing results in ice sheet growth, both of which are largely inconsistent with the present state of glaciation.
@@
 
-To contrast the effect of the six different input climatologies, we plot simulations using a~5\,\unit{\degree C} temperature offset (Fig.~\ref{fig:cool05}).
+To contrast the effect of the six different input climatologies, we plot simulations using a~5\,\unit{{\degree}C} temperature offset (Fig.~\ref{fig:cool05}).
 The magnitude of glaciation differs dramatically from one climate forcing to another.
@@
 These new simulations use a~“hybrid” climate forcing that consists of temperature data from WorldClim, combined with precipitation data from each of the four reanalyses, and vice-versa.
-A~single temperature offset of 5\,\unit{\degree C} is used.
+A~single temperature offset of 5\,\unit{{\degree}C} is used.
 This value was chosen to allow for comparison with previous results (Fig.~\ref{fig:cool05}).
@@
 
-Although only temperature or precipitation changes are applied, this experiment results once more in large differences in modelled ice sheet geometries (Fig.~\ref{fig:biatm}) as compared to the reference WorldClim 5\,\unit{\degree C} run (Fig.~\ref{fig:cool05}, top-left panel).
+Although only temperature or precipitation changes are applied, this experiment results once more in large differences in modelled ice sheet geometries (Fig.~\ref{fig:biatm}) as compared to the reference WorldClim 5\,\unit{{\degree}C} run (Fig.~\ref{fig:cool05}, top-left panel).
 For ERA-Interim, CFSR and NARR climatologies, precipitation anomalies clearly dominate the differences in ice sheet response, whereas temperature anomalies have relatively little effect (Fig.~\ref{fig:biatmbars}).
@@
 
-We compare modelled ice sheet geometries to a~reconstruction of the ice margin at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) by \citet{dyke-2004}, based on glacial geomorphology and radiocarbon dating (Fig.~\ref{fig:locmap}).
+We compare modelled ice sheet geometries to a~reconstruction of the ice margin at 14\,\unit{\chem{^{14}C}\,kyr\,BP} (16.8\,cal\,kyr\,BP) by \citet{dyke-2004}, based on glacial geomorphology and radiocarbon dating (Fig.~\ref{fig:locmap}).
 This corresponds to the LGM extent of most of the Cordilleran ice sheet, which occurred later than for the Laurentide ice sheet \citep{porter-swanson-1998,dyke-2004,stroeven-etal-2010,stroeven-etal-inpress}.
@@
 Considering previously identified weaknesses of the CFSR and NCEP/NCAR data (Figs.~\ref{fig:tempheatmap}--\ref{fig:biatmbars}), we restricted our analysis to the WorldClim, ERA-Interim and NARR input climatologies.
-In order to compare modelled ice sheets of a~similar size, we select for each climate forcing the simulation that leads to a~final glaciated area closest to the approximate size of the LGM Cordilleran ice sheet of $2\,\times10^6\,\unit{km^2}$.
-This corresponds to temperature offset values of 8\,\unit{\degree C} (WorldClim), 6\,\unit{\degree C} (ERA-Interim) and 7\,\unit{\degree C} (NARR).
+In order to compare modelled ice sheets of a~similar size, we select for each climate forcing the simulation that leads to a~final glaciated area closest to the approximate size of the LGM Cordilleran ice sheet of $2\times 10^6\,\unit{km^2}$.
+This corresponds to temperature offset values of 8\,\unit{{\degree}C} (WorldClim), 6\,\unit{{\degree}C} (ERA-Interim) and 7\,\unit{{\degree}C} (NARR).
 These qualitative “best” runs are presented in Figure~\ref{fig:best} along with their associated temperature depressions and a~reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.
@@
 These regions experience a~marine climate, including lower summer temperature than on the adjacent land (Fig.~\ref{fig:temp}).
-However in our simulations, sea-level is lowered by 120~m, turning large parts of the low-sloping continental shelf into land.
+However in our simulations, sea-level is lowered by 120\,\unit{m}, turning large parts of the low-sloping continental shelf into land.
 This anomalous ice cover is to be interpreted as an artefact arising from our simplistic temperature offset method, and has little effect on the rest of the results.
@@
 
-A~key assumption for the present study is our choice of 10\,ka for the duration of all our simulations.
-Although geomorphological data shows that the period of Cordilleran ice sheet inception, from nearly ice-free to full glacial conditions, could not have been much longer than 10\,ka, they provide no lower bound to this value.
+A~key assumption for the present study is our choice of 10\,kyr for the duration of all our simulations.
+Although geomorphological data shows that the period of Cordilleran ice sheet inception, from nearly ice-free to full glacial conditions, could not have been much longer than 10\,kyr, they provide no lower bound to this value.
 
-To test the effect of an inception period of shorter duration on model results, we use simulations run under NARR climate forcing with temperature offsets of 7 to 11\,\unit{\degree C}, and compare modelled ice sheet extent when glaciated area reaches the approximate size of the LGM Cordilleran ice sheet of $2\,\times10^6\,\unit{km^2}$.
+To test the effect of an inception period of shorter duration on model results, we use simulations run under NARR climate forcing with temperature offsets of 7 to 11\,\unit{{\degree}C}, and compare modelled ice sheet extent when glaciated area reaches the approximate size of the LGM Cordilleran ice sheet of $2\times 10^6\,\unit{km^2}$.
 
@@
 		\middlehline
-		NCEP/NCAR&  global&     1981 -- 2010& 1.875\degree& \citet{data:ncar}\\
-		ERA-Interim&global&     1979 -- 2011& 1.000\degree& \citet{data:erai}\\
-		CFSR&       global&     1979 -- 2010& 0.325\degree& \citet{data:cfsr}\\
-		NARR&       North America& 1979 -- 2000& 32\,km& \citet{data:narr}\\
+    NCEP/NCAR   & global        & 1981--2010 & 1.875{\degree} & \citet{data:ncar} \\
+    ERA-Interim & global        & 1979--2011 & 1.000{\degree} & \citet{data:erai} \\
+    CFSR        & global        & 1979--2010 & 0.325{\degree} & \citet{data:cfsr} \\
+    NARR        & North America & 1979--2000 & 32\,km       & \citet{data:narr} \\
 		\bottomhline
@@
 	\end{center}
-	\caption{Shaded relief map of northern North America with the extent of the Cordilleran (CIS), Laurentide (LIS) and Innuitian (IIS) ice sheets at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) \citep{dyke-2004}.
+	\caption{Shaded relief map of northern North America with the extent of the Cordilleran (CIS), Laurentide (LIS) and Innuitian (IIS) ice sheets at 14\,\unit{\chem{^{14}C}\,kyr\,BP} (16.8\,cal\,kyr\,BP) \citep{dyke-2004}.
 While this age denotes the LIS after retreat from its LGM, it closely corresponds to the LGM extent of most of the Cordilleran ice sheet \citep{porter-swanson-1998,dyke-2004,stroeven-etal-2010}.
@@
 	\end{center}
-	\caption{Total glaciated area and ice volume after 10\,ka as a~function of temperature offset for each climate forcing used.}
+	\caption{Total glaciated area and ice volume after 10\,kyr as a~function of temperature offset for each climate forcing used.}
 	\label{fig:ivolarea}
@@
 	\end{center}
-	\caption{Extent of ice cover after 10\,ka as a~function of applied temperature offsets for each climate forcing used.
+	\caption{Extent of ice cover after 10\,kyr as a~function of applied temperature offsets for each climate forcing used.
 A~reconstructed LGM ice sheet margin by \citet{dyke-2004} (Fig.~\ref{fig:locmap}) is given for reference (black line).}
@@
 	\end{center}
-	\caption{Ice surface topography (1 km contours) and velocity (\unit{m\,a^{-1}}) after 10\,ka under a~climate 5\,\unit{\degree C} colder than present for each climate forcing used.}
+	\caption{Ice surface topography (1\,\unit{km} contours) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a~climate 5\,\unit{{\degree}C} colder than present for each climate forcing used.}
 	\label{fig:cool05}
@@
 	\end{center}
-	\caption{Ice surface topography (1 km contours) and velocity (\unit{m\,a^{-1}}) after 10\,ka using “hybrid” climate forcing with precipitation rate from WorldClim and surface air temperature from each reanalysis (upper row), and surface air temperature from WorldClim and precipitation rate from each reanalysis (lower row).
+	\caption{Ice surface topography (1\,\unit{km} contours) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr using “hybrid” climate forcing with precipitation rate from WorldClim and surface air temperature from each reanalysis (upper row), and surface air temperature from WorldClim and precipitation rate from each reanalysis (lower row).
 In other words, the upper row shows the effect of temperature anomalies, and the lower row the effect of precipitation anomalies, for each reanalysis, relative to WorldClim data.
-Each simulation uses a~5\,\unit{\degree C} offset for comparison with figure~\ref{fig:cool05}}
+Each simulation uses a~5\,\unit{{\degree}C} offset for comparison with figure~\ref{fig:cool05}}
 	\label{fig:biatm}
@@
 	\end{center}
-	\caption{Effect of temperature and precipitation anomalies (separately and jointly) from each reanalysis on modelled final ice volume relative to the result of the WorldClim 5\,\unit{\degree C} offset simulation.
+	\caption{Effect of temperature and precipitation anomalies (separately and jointly) from each reanalysis on modelled final ice volume relative to the result of the WorldClim 5\,\unit{{\degree}C} offset simulation.
 Corresponding ice sheet geometries are presented in figures~\ref{fig:cool05} and~\ref{fig:biatm}}
@@
 	\end{center}
-	\caption{Ice surface topography (1\,km contours) after 10\,ka using temperature offsets resulting in glaciated areas of circa $2\,\times10^6\,\unit{km^2}$.
+	\caption{Ice surface topography (1\,km contours) after 10\,kyr using temperature offsets resulting in glaciated areas of circa $2 \times 10^6\,\unit{km^2}$.
 A~reconstructed LGM ice sheet margin by \citet{dyke-2004} (Fig.~\ref{fig:locmap}) is given for reference (blue line).}
@@
 	\caption{Left panel: model sensitivity to simulation length.
-Modelled glaciated area, using NARR forcing data, and temperature offsets from 0 to 15\,\unit{\degree C}, solid lines corresponding to values from 7 to 11\,\unit{\degree C}.
-Right panel: modelled ice margin corresponding to temperature offsets from 7 to 11\,\unit{\degree C} when total glaciated area reach the approximate size of the LGM Cordilleran ice sheet of $2\,\times10^6\,\unit{km^2}$.
+Modelled glaciated area, using NARR forcing data, and temperature offsets from 0 to 15\,\unit{{\degree}C}, solid lines corresponding to values from 7 to 11\,\unit{{\degree}C}.
+Right panel: modelled ice margin corresponding to temperature offsets from 7 to 11\,\unit{{\degree}C} when total glaciated area reach the approximate size of the LGM Cordilleran ice sheet of $2 \times 10^6\,\unit{km^2}$.
 A~reconstructed LGM ice sheet margin by \citet{dyke-2004} (Fig.~\ref{fig:locmap}) is given for reference (grey shading).

### 2014-03-20 cordillera-climate.tex
Typesetting: Fix Fig., Table and Sect. references

@@
 In addition to the WorldClim data, we use surface air temperatures and precipitation rates from three global atmospheric reanalyses and one regional atmospheric reanalysis to inform the mass balance routine of PISM: the NCEP/NCAR reanalysis \citep{data:ncar}, the ERA-Interim reanalysis \citep{data:erai}, the Climate Forecast System Reanalysis \citep[CFSR;][]{data:cfsr}, and the North American Regional Reanalysis \citep[NARR;][]{data:narr}.
-Monthly climatologies from NCEP/NCAR and NARR reanalyses were provided by the \citet{web:psd} whereas monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean time series (Table \ref{tab:reanalyses}).
+Monthly climatologies from NCEP/NCAR and NARR reanalyses were provided by the \citet{web:psd} whereas monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean time series (Table~\ref{tab:reanalyses}).
 
-The spatial distributions of JJA air surface temperatures and DJF precipitation rates from the four reanalysis climatologies are shown in Figures~\ref{fig:temp} and~\ref{fig:prec}.
+The spatial distributions of JJA air surface temperatures and DJF precipitation rates from the four reanalysis climatologies are shown in Figs.~\ref{fig:temp} and~\ref{fig:prec}.
 Summer temperature and winter precipitation are most relevant to the glacier model as they drive summer melt and winter accumulation, respectively.
@@
 Using WorldClim, NCEP/NCAR, ERA-Interim, CFSR (smoothed and original datasets) and NARR input climatologies as climate forcing, we run 66~simulations of glacial inception and growth of the Cordilleran ice sheet over 10\,kyr using temperature offsets ranging from 0 to 10\,\unit{{\degree}C}.
-Significant differences in patterns of ice sheet growth, final glacial extent, and volume are attained (Figs.
-\ref{fig:ivolarea}-\ref{fig:cool05}).
+Significant differences in patterns of ice sheet growth, final glacial extent, and volume are attained (Figs.~\ref{fig:ivolarea}--\ref{fig:cool05}).
 
@@
 When examining the spatial distribution of temperature differences between reanalyses and WorldClim data, it appears that most of the disparity between reanalyses and observation data is caused by topographical detail at scales unresolved in the reanalysis datasets (Fig.~\ref{fig:tempdiff}).
-Surface air temperature data from the NCEP/NCAR climatology shows the largest disparity with that of WorldClim, with a~significant cold anomaly over most of the modelling domain (Figs.~\ref{fig:tempheatmap}--\ref{fig:tempdiff}).
+Surface air temperature data from the NCEP/NCAR climatology shows the largest disparity with that of WorldClim, with a~significant cold anomaly over most of the modelling domain (Figs.~\ref{fig:tempheatmap} and \ref{fig:tempdiff}).
 
@@
 Most likely, this is the signature of an orographic precipitation effect.
-As described in section~\ref{sec:climate}, the topography of the Northern American Cordillera is such that its western ranges form a~continuous orographic barrier, causing high precipitation along the Pacific coast while leaving much of the interior arid.
+As described in Sect.~\ref{sec:climate}, the topography of the Northern American Cordillera is such that its western ranges form a~continuous orographic barrier, causing high precipitation along the Pacific coast while leaving much of the interior arid.
 However the ability of a~GCM to reproduce these contrasts in precipitation is bounded by spatial resolution.
 In a~model of coarser resolution, these high mountain ranges are reduced to smooth hills and the modelled distribution of precipitation is also smoother (Fig.~\ref{fig:oroprecip}).
-However, this effect alone cannot explain the widespread and pronounced positive precipitation anomalies observed in CFSR data (Figs.~\ref{fig:precheatmap}--\ref{fig:precdiff}) despite of its high spatial resolution (Table~\ref{tab:reanalyses}).
+However, this effect alone cannot explain the widespread and pronounced positive precipitation anomalies observed in CFSR data (Figs.~\ref{fig:precheatmap} and \ref{fig:precdiff}) despite of its high spatial resolution (Table~\ref{tab:reanalyses}).
 
@@
 For ERA-Interim, CFSR and NARR climatologies, precipitation anomalies clearly dominate the differences in ice sheet response, whereas temperature anomalies have relatively little effect (Fig.~\ref{fig:biatmbars}).
-However in the case of NCAR forcing, both the negative temperature bias (Figs.~\ref{fig:tempheatmap}--\ref{fig:tempdiff}) and the important precipitation differences (Figs.~\ref{fig:precheatmap}--\ref{fig:precdiff}) contribute approximately equally to produce oversized ice sheets (Fig.~\ref{fig:biatmbars}).
+However in the case of NCAR forcing, both the negative temperature bias (Figs.~\ref{fig:tempheatmap} and \ref{fig:tempdiff}) and the important precipitation differences (Figs.~\ref{fig:precheatmap} and \ref{fig:precdiff}) contribute approximately equally to produce oversized ice sheets (Fig.~\ref{fig:biatmbars}).
 
@@
 This corresponds to temperature offset values of 8\,\unit{{\degree}C} (WorldClim), 6\,\unit{{\degree}C} (ERA-Interim) and 7\,\unit{{\degree}C} (NARR).
-These qualitative “best” runs are presented in Figure~\ref{fig:best} along with their associated temperature depressions and a~reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.
+These qualitative “best” runs are presented in Fig.~\ref{fig:best} along with their associated temperature depressions and a~reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.
 
@@
 Whereas ETOPO1 is used as basal topography for the ice sheet model, all other data are used as a~reference for temperature lapse-rate corrections.
-Fig.~\ref{fig:topo}-\ref{fig:durationstack} are drawn using Matplotlib \citep{soft:mpl}.}
+Figs.~\ref{fig:topo}--\ref{fig:durationstack} are drawn using Matplotlib \citep{soft:mpl}.}
 	\label{fig:topo}
@@
 	\caption{Winter (DJF) and summer (JJA) precipitation maps from WorldClim \citep{data:worldclim}, and winter (DFJ) temperature maps from ERA-Interim reanalysis \citep{data:erai}, North American Regional Reanalysis \citep[NARR;][]{data:narr}, Climate Forecast System Reanalysis \citep[CFSR;][]{data:cfsr}, and NCEP/NCAR reanalysis \citep{data:ncar} climatologies.
-Additional forcing data was prepared to correct for wave-like precipitation artefacts in CFSR (section~\ref{sec:climate}).}
+Additional forcing data was prepared to correct for wave-like precipitation artefacts in CFSR (Sect.~\ref{sec:climate}).}
 	\label{fig:prec}
@@
 In other words, the upper row shows the effect of temperature anomalies, and the lower row the effect of precipitation anomalies, for each reanalysis, relative to WorldClim data.
-Each simulation uses a~5\,\unit{{\degree}C} offset for comparison with figure~\ref{fig:cool05}}
+Each simulation uses a~5\,\unit{{\degree}C} offset for comparison with Fig.~\ref{fig:cool05}.}
 	\label{fig:biatm}
@@
 	\caption{Effect of temperature and precipitation anomalies (separately and jointly) from each reanalysis on modelled final ice volume relative to the result of the WorldClim 5\,\unit{{\degree}C} offset simulation.
-Corresponding ice sheet geometries are presented in figures~\ref{fig:cool05} and~\ref{fig:biatm}}
+Corresponding ice sheet geometries are presented in Figs.~\ref{fig:cool05} and~\ref{fig:biatm}.}
 	\label{fig:biatmbars}

### 2014-03-20 cordillera-climate.tex
Typesetting: fix quotes and preceding space

@@
 Climatologies obtained from GCM palaeo-climate simulations such as produced within the Paleoclimate Modelling Intercomparison Project \citep[PMIP;][]{joussaume-taylor-1995} provide perhaps a~more reasonable representation of past climate.
-However, as such climatologies are only available for specific periods of time, using them as climate forcing for an ice sheet model requires either an assumption of steady-state ice sheet response to climatic fluctuations \citep{huybrechts-tsiobbel-1996}, or interpolation through time between climatologies from different periods, which can be linear \citep{charbit-etal-2002}, or modulated by a~``glacial index'' weighting function derived from ice core \chem{\delta^{18}O} records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}.
+However, as such climatologies are only available for specific periods of time, using them as climate forcing for an ice sheet model requires either an assumption of steady-state ice sheet response to climatic fluctuations \citep{huybrechts-tsiobbel-1996}, or interpolation through time between climatologies from different periods, which can be linear \citep{charbit-etal-2002}, or modulated by a ``glacial index'' weighting function derived from ice core \chem{\delta^{18}O} records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}.
 An important drawback in this approach is that GCM palaeo-climate simulations themselves rely on an accurate description of the global surface topography and therefore include global ice sheet reconstructions such as ICE-4G \citep{peltier-1994} for their surface topographic boundary condition.
@@
 
-The Shallow Shelf Approximation (SSA) is used as a~``sliding law'' for the Shallow Ice Approximation (SIA) \citep{bueler-brown-2009,winkelmann-etal-2011}.
+The Shallow Shelf Approximation (SSA) is used as a ``sliding law'' for the Shallow Ice Approximation (SIA) \citep{bueler-brown-2009,winkelmann-etal-2011}.
 SIA and SSA velocities are computed by finite difference methods on a~10\,km resolution horizontal grid of 300 by 150 points (the modelling domain).
@@
 To distinguish the effects of temperature and precipitation biases on the ice sheet model, we run a~series of eight additional simulations.
-These new simulations use a~“hybrid” climate forcing that consists of temperature data from WorldClim, combined with precipitation data from each of the four reanalyses, and vice-versa.
+These new simulations use a ``hybrid'' climate forcing that consists of temperature data from WorldClim, combined with precipitation data from each of the four reanalyses, and vice-versa.
 A~single temperature offset of 5\,\unit{{\degree}C} is used.
@@
 This corresponds to temperature offset values of 8\,\unit{{\degree}C} (WorldClim), 6\,\unit{{\degree}C} (ERA-Interim) and 7\,\unit{{\degree}C} (NARR).
-These qualitative “best” runs are presented in Fig.~\ref{fig:best} along with their associated temperature depressions and a~reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.
+These qualitative ``best'' runs are presented in Fig.~\ref{fig:best} along with their associated temperature depressions and a~reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.
 
@@
 	\end{center}
-	\caption{Ice surface topography (1\,\unit{km} contours) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr using “hybrid” climate forcing with precipitation rate from WorldClim and surface air temperature from each reanalysis (upper row), and surface air temperature from WorldClim and precipitation rate from each reanalysis (lower row).
+	\caption{Ice surface topography (1\,\unit{km} contours) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr using ``hybrid'' climate forcing with precipitation rate from WorldClim and surface air temperature from each reanalysis (upper row), and surface air temperature from WorldClim and precipitation rate from each reanalysis (lower row).
 In other words, the upper row shows the effect of temperature anomalies, and the lower row the effect of precipitation anomalies, for each reanalysis, relative to WorldClim data.

### 2014-03-20 cordillera-climate.tex
Typesetting: table and figure formatting
Typesetting: till friction equation
Typesetting: various

@@
 \title{The effect of climate forcing on numerical simulations of the Cordilleran ice sheet at the Last Glacial Maximum}
-\runningtitle{Climate forcing for Cordilleran ice sheet simulations}
 
@@
 
+% Running title
+\runningtitle{Climate forcing for Cordilleran ice sheet simulations}
+
 % Affiliations
@@
 % For Copernicus
-\received{}
-\pubdiscuss{}
-\revised{}
-\accepted{}
+\received{22~November 2013}
+\accepted{5~December 2013}
 \published{}
@@
 A~pseudo-plastic sliding law \citep{aschwanden-etal-2013} relates the bed-parallel shear stress and the sliding velocity.
-The yield stress is modelled using the Mohr-Coulomb criterion.
+The yield stress is modelled using the Mohr--Coulomb criterion.
 The till friction angle $\phi$ varies from 10 to 30\degree\ and is a~function of current bed elevation, with lowest values occurring at low elevations,
 
-\begin{equation}
-	\phi = \left\{\begin{array}{llrll}
-		10      & \mathrm{for} &      &z&<  0 \\
-		z/10+10 & \mathrm{for} &   0 <&z&<200 \\
-		30      & \mathrm{for} & 200 <&z&     \\
-	\end{array}\right.
-\end{equation}
+\begin{align}
+  \phi =
+  \begin{cases}
+    10      & \text{for }       z<  0 \\
+    z/10+10 & \text{for }    0 <z<200 \\
+    30      & \text{for }  200 <z     \\
+  \end{cases}
+\end{align}
 
@@
 We thank Ed Bueler and Andy Aschwanden for providing constant support on PISM and patiently answering our questions on the code internals, Laurent Brodeau and Soon-Heum Ko for assistance concerning software installations on supercomputing resources, and all PISM authors for providing the model open-source.
-This work was supported by the Swedish Research Council (VR) grant \#2008-3449 to A.
+This work was supported by the Swedish Research Council (VR) grant {\#}2008-3449 to A.
 P.
-Stroeven and by NASA grant \#NNX13AM16G to E.
+Stroeven and by NASA grant {\#}NNX13AM16G to E.
 Bueler.
-Computer resources were provided by the Swedish National Infrastructure for Computing (SNIC) allocation \#2013/1-145 to A.
+Computer resources were provided by the Swedish National Infrastructure for Computing (SNIC) allocation {\#}2013/1-145 to A.
 P.
-Stroeven at the National Supercomputing Center (NSC).
+Stroeven at the National Supercomputing Center (NSC).\\
 
 % Author contributions
-\textit{Author contributions.}
+\hack{\noindent}\textit{Author contributions.}
 J.
@@
 	\label{tab:reanalyses}
-	\vskip4mm
-	\centering
-	\begin{tabular}{lllll}
+	{\begin{tabular}{lcccc}
 		\tophline
@@
 		\bottomhline
-	\end{tabular}
+	\end{tabular}}
+  \belowtable{}
 \end{table*}
@@
 % fig:locmap
-\begin{figure}[t]
-	\vspace*{2mm}
-	\begin{center}
-		\includegraphics[width=8cm]{cordillera-climate-locmap}
-	\end{center}
+\begin{figure}
+	\includegraphics[width=80mm]{cordillera-climate-locmap}
 	\caption{Shaded relief map of northern North America with the extent of the Cordilleran (CIS), Laurentide (LIS) and Innuitian (IIS) ice sheets at 14\,\unit{\chem{^{14}C}\,kyr\,BP} (16.8\,cal\,kyr\,BP) \citep{dyke-2004}.
@@
 The rectangular box denotes the modelling domain of the Cordilleran ice sheet of 1500 by 3000\,km.
-Major mountain ranges covered by the ice sheet include the Wrangell and St.-Elias mountains (W-SE), the Selwyn and MacKenzie mountains (S-MK), the Coast Mountains and the Rocky Mountains.
+Major mountain ranges covered by the ice sheet include the Wrangell and St.-Elias mountains (W--SE), the Selwyn and MacKenzie mountains (S--MK), the Coast Mountains and the Rocky Mountains.
 The background map consists of ETOPO1 \citep{data:etopo1} and Natural Earth Data \citep{data:naturalearth} and was assembled with GRASS~GIS \citep{soft:grass}.}
@@
 % fig:topo
-\begin{figure*}[t]
-	\vspace*{2mm}
-	\begin{center}
+\begin{figure*}
 	\includegraphics{cordillera-climate-topo}
-	\end{center}
 	\caption{Topography maps from WorldClim \citep{data:worldclim}, ERA-Interim reanalysis \citep{data:erai}, North American Regional Reanalysis \citep[NARR;][]{data:narr}, ETOPO1 \citep{data:etopo1}, Climate Forecast System Reanalysis \citep[CFSR;][]{data:cfsr}, and NCEP/NCAR reanalysis \citep{data:ncar}.
@@
 % fig:temp
-\begin{figure*}[t]
-	\vspace*{2mm}
-	\begin{center}
+\begin{figure*}
 	\includegraphics{cordillera-climate-temp}
-	\end{center}
 	\caption{Summer (JJA) and winter (DJF) temperature maps from WorldClim \citep{data:worldclim}, and summer (JJA) temperature maps from ERA-Interim reanalysis \citep{data:erai}, North American Regional Reanalysis \citep[NARR;][]{data:narr}, Climate Forecast System Reanalysis \citep[CFSR;][]{data:cfsr}, and NCEP/NCAR reanalysis \citep{data:ncar} climatologies.}
@@
 % fig:prec
-\begin{figure*}[t]
-	\vspace*{2mm}
-	\begin{center}
+\begin{figure*}
 	\includegraphics{cordillera-climate-prec}
-	\end{center}
 	\caption{Winter (DJF) and summer (JJA) precipitation maps from WorldClim \citep{data:worldclim}, and winter (DFJ) temperature maps from ERA-Interim reanalysis \citep{data:erai}, North American Regional Reanalysis \citep[NARR;][]{data:narr}, Climate Forecast System Reanalysis \citep[CFSR;][]{data:cfsr}, and NCEP/NCAR reanalysis \citep{data:ncar} climatologies.
@@
 % fig:ivolarea
-\begin{figure}[t]
-	\vspace*{2mm}
-	\begin{center}
+\begin{figure}
 	\includegraphics{cordillera-climate-ivolarea}
-	\end{center}
 	\caption{Total glaciated area and ice volume after 10\,kyr as a~function of temperature offset for each climate forcing used.}
@@
 % fig:extent
-\begin{figure*}[t]
-	\vspace*{2mm}
-	\begin{center}
+\begin{figure*}
 	\includegraphics{cordillera-climate-extent}
-	\end{center}
 	\caption{Extent of ice cover after 10\,kyr as a~function of applied temperature offsets for each climate forcing used.
@@
 % fig:cool05
-\begin{figure*}[t]
-	\vspace*{2mm}
-	\begin{center}
+\begin{figure*}
 	\includegraphics{cordillera-climate-cool05}
-	\end{center}
 	\caption{Ice surface topography (1\,\unit{km} contours) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a~climate 5\,\unit{{\degree}C} colder than present for each climate forcing used.}
@@
 % fig:tempheatmap
-\begin{figure}[t]
-	\vspace*{2mm}
-	\begin{center}
+\begin{figure}
 	\includegraphics{cordillera-climate-tempheatmap}
-	\end{center}
 	\caption{Density maps showing a~comparison of summer (JJA) surface air temperature data from the WorldClim climatology, against that of each reanalysis.
@@
 % fig:tempdiff
-\begin{figure}[t]
-	\vspace*{2mm}
-	\begin{center}
+\begin{figure}
 	\includegraphics{cordillera-climate-tempdiff}
-	\end{center}
 	\caption{Summer (JJA) surface air temperature difference maps against WorldClim data, after bi-linear spatial interpolation.
@@
 % fig:precheatmap
-\begin{figure}[t]
-	\vspace*{2mm}
-	\begin{center}
+\begin{figure}
 	\includegraphics{cordillera-climate-precheatmap}
-	\end{center}
 	\caption{Density maps showing a~comparison of winter (DJF) precipitation rate from the WorldClim climatology, against that of each reanalysis.
@@
 % fig:precdiff
-\begin{figure}[t]
-	\vspace*{2mm}
-	\begin{center}
+\begin{figure}
 	\includegraphics{cordillera-climate-precdiff}
-	\end{center}
 	\caption{Winter (DJF) precipitation rate difference maps against WorldClim data, after bi-linear spatial interpolation.
@@
 % fig:oroprecip
-\begin{figure}[t]
-	\vspace*{2mm}
-	\begin{center}
+\begin{figure}
 	\includegraphics{cordillera-climate-oroprecip}
-	\end{center}
 	\caption{Schematic representation of the orographic precipitation effect over a~mountain range.
@@
 % fig:biatm
-\begin{figure*}[t]
-	\vspace*{2mm}
-	\begin{center}
+\begin{figure*}
 	\includegraphics{cordillera-climate-biatm}
-	\end{center}
 	\caption{Ice surface topography (1\,\unit{km} contours) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr using ``hybrid'' climate forcing with precipitation rate from WorldClim and surface air temperature from each reanalysis (upper row), and surface air temperature from WorldClim and precipitation rate from each reanalysis (lower row).
@@
 % fig:biatmbars
-\begin{figure}[t]
-	\vspace*{2mm}
-	\begin{center}
+\begin{figure}
 	\includegraphics{cordillera-climate-biatmbars}
-	\end{center}
 	\caption{Effect of temperature and precipitation anomalies (separately and jointly) from each reanalysis on modelled final ice volume relative to the result of the WorldClim 5\,\unit{{\degree}C} offset simulation.
@@
 % fig:best
-\begin{figure*}[t]
-	\vspace*{2mm}
-	\begin{center}
+\begin{figure*}
 	\includegraphics{cordillera-climate-best}
-	\end{center}
 	\caption{Ice surface topography (1\,km contours) after 10\,kyr using temperature offsets resulting in glaciated areas of circa $2 \times 10^6\,\unit{km^2}$.
@@
 % fig:durationstack
-\begin{figure}[t]
-	\vspace*{2mm}
-	\begin{center}
+\begin{figure}
 	\includegraphics{cordillera-climate-durationstack}
-	\end{center}
 	\caption{Left panel: model sensitivity to simulation length.
@@
 \end{document}
+\endinput
 % ----------------------------------------------------------------------

### 2014-03-24 abstract.tex
Fix broken abstract file

@@
 
-\input{sections/abstract.tex}
+\begin{abstract}
+  We present an ensemble of numerical simulations of the Cordilleran ice sheet during the Last Glacial Maximum performed with the Parallel Ice Sheet Model (PISM), applying temperature offsets to the present-day climatologies from five different datasets.
+Monthly mean surface air temperature and precipitation from WorldClim, the NCEP/NCAR reanalysis, the ERA-Interim reanalysis, the Climate Forecast System Reanalysis and the North American Regional Reanalysis are used to compute surface mass balance in a~positive degree-day model.
+Modelled ice sheet outlines and volumes appear highly sensitive to the choice of climate forcing.
+For three of the four reanalysis datasets used, differences in precipitation are the major source for discrepancies between model results.
+We assess model performance against a~geomorphological reconstruction of the ice margin at the Last Glacial Maximum, and suggest that part of the mismatch is due to unresolved orographic precipitation effects caused by the coarse resolution of reanalysis data.
+The best match between model output and the reconstructed ice margin is obtained using the high-resolution North American Regional Reanalysis, which we retain for simulations of the Cordilleran ice sheet in the future.
+\end{abstract}
 

### 2014-03-26 reviews/disc-ac1.tex
Add authors response to reviewers

\documentclass[10pt]{article}

\usepackage[T1]{fontenc}
\usepackage[pdftex]{graphicx}
\usepackage[pdftex]{color}
\usepackage{doi}

\sloppy
\setlength{\parskip}{1.2ex}
\setlength{\parindent}{0em}
\renewcommand\familydefault{\sfdefault}

\graphicspath{{../figures/}}
\definecolor{journalname}{rgb}{0.34,0.59,0.82}

\begin{document}

\textbf{Authors' response to Referee {\#}1}
\bigskip

% ----------------------------------------------------------------------
\def\referee#1{\bigskip\textcolor{journalname}{\textit{#1}}}

Dear Referee {\#}1,

Thank you very much for this supportive review. Your comments raise valid points about presentation and interpretation of the results. Please find our response to them below.

\referee{Section~5.1: Are the temperature climatologies compared after applying a lapse-rate correction when interpolating the reanalyses to the same present-day topography and resolution? This would be important to avoid artificial biases in temperature.}

In fact, no. In Figs.~8 and~9 of the discussion paper, temperature climatologies are compared without lapse-rate correction. Due to differences in resolution of surface topography between climatological datasets used, this caused appearance of numerous topographic artefacts on paper Fig.~9. Following your comment, we produced updated figures to include a lapse-rate correction of 6\,{$^\circ$}C\,km$^{-1}$, as applied in the simulations.

For paper Fig.~9 (temperature difference maps, Fig.~1 in this comment), lapse-rate correction results in smoother temperature difference maps, where effective temperature discrepancies are no longer overshadowed by local topographic anomalies (this comment's Fig.~2). Because temperature differences in this new map are much reduced, we adjusted colour scaling to emphasize smaller values (this comment's Fig.~3).

For paper Fig.~8 (temperature density maps), we now use bilinearly-interpolated temperature maps, rather than the original data (as presented in paper Fig.~2--4). This results in smoother density plots than those presented in the discussion paper (this comment's Fig.~4). In a second step, we apply a lapse-rate correction to project all reanalysis data onto the WorldClim (higher resolution) topography (this comment's Fig.~5). It now becomes clear that much of the discrepancies observed were due to lapse-rate effects.

For consistency, paper Fig.~10 (precipitation density maps) was also updated using bilinearly-interpolated data (this comment's Fig.~6). Note that for all density maps, colour maps were changed to allow for a higher level of detail.

While these new figures present an altered version of climate forcing data used in our study, they more closely reflect how this data is read in by the ice sheet model, which seems appropriate for the discussion section. Moreover, the new maps reveal more clearly the strengths and weakness of different datasets, and support more strongly our interpretation of the ``hybrid'' climate forcing experiments presented in Sect.~5.2. We will definitely include these new figures in the revised manuscript. Thank you very much for this constructive comment.

\referee{Discussion PaperPage 6184, line 2: ``despite of'' $=>$ ``despite''}

We will correct this.

\referee{Page 6184, line 8 (and elsewhere): ``A single temperature offset of 5\,{$^\circ$} is used.'' This would be clearer with a minus sign, since it is a negative temperature offset correct? Please check throughout the manuscript.}

To avoid confusion, we will use negative values in the revised manuscript.

\referee{Section~5.5: In this discussion, please add some sentences about the potential effect of elevation changes on the precipitation fields as the ice sheet evolves. In Section~2.3, it was mentioned that no correction was applied. However, I could imagine that as the dome of the ice sheet grows, a very distinct pattern of precipitation maximum could occur near the margins of the ice sheet. Perhaps, for example, including such a correction would actually make the ice sheets evolve to more similar states after 10ka.}

A short discussion of these effects is given in Sect.~5.3, in the context of comparison of simulation results to the geomorphological last glacial maximum ice margin. However you raise here an issue that we had not consider, regarding potential ice sheet evolution to more similar states when accounting for precipitation changes. In the East more particularly, where in most of our simulations, the ice margin do not attain a steady configuration, its position after 10\,kyr is largely determined by its rate of advance during the simulation length, which in turn depends on the amount of precipitation received. Therefore, precipitation corrections in this region could significantly slow down the ice margin advance, resulting in more similar configuration after 10\,kyr. We will include a discussion of this effect in Sect.~5.5.

% ----------------------------------------------------------------------

\clearpage

\begin{figure}
    \center
    \includegraphics[width=70mm]{cordillera-climate-tempdiff+tcdc}
    \caption{Summer temperature difference maps, as in paper Fig.~9.}
\end{figure}
\begin{figure}
    \center
    \includegraphics[width=70mm]{cordillera-climate-tempdiff+lr+tcdc}
    \caption{Summer temperature difference maps, after applying a lapse-rate correction of 6\,{$^\circ$}C\,km$^{-1}$.}
\end{figure}
\begin{figure}
    \center
    \includegraphics[width=70mm]{cordillera-climate-tempdiff+lr}
    \caption{Summer temperature difference maps, after applying a lapse-rate correction of 6\,{$^\circ$}C\,km$^{-1}$ and an updated colour mapping.}
\end{figure}
\begin{figure}
    \center
    \includegraphics[width=70mm]{cordillera-climate-tempheatmap}
    \caption{Summer temperature density maps, after bilinear interpolation.}
\end{figure}
\begin{figure}
    \center
    \includegraphics[width=70mm]{cordillera-climate-tempheatmap+lr}
    \caption{Summer temperature density maps, after bilinear interpolation and lapse-rate correction.}
\end{figure}
\begin{figure}
    \center
    \includegraphics[width=70mm]{cordillera-climate-precheatmap}
    \caption{Winter precipitation density maps, after bilinear interpolation.}
\end{figure}

\end{document}

### 2014-03-26 reviews/disc-ac2.tex
Add authors response to reviewers

\documentclass[10pt]{article}

\usepackage[T1]{fontenc}
\usepackage[pdftex]{graphicx}
\usepackage[pdftex]{color}
\usepackage[colorlinks,citecolor=blue]{hyperref}
\usepackage{natbib}
\usepackage{doi}

\sloppy
\setlength{\parskip}{1.2ex}
\setlength{\parindent}{0em}
\renewcommand\familydefault{\sfdefault}

\graphicspath{{../figures/}}
\definecolor{journalname}{rgb}{0.34,0.59,0.82}
%\providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}

\begin{document}

\textbf{Authors' response to Referee {\#}2}
\bigskip

% ----------------------------------------------------------------------
\def\referee#1{\bigskip\textcolor{journalname}{\textit{#1}}}

Dear Referee {\#}2,

\referee{Before publication, I do have one scientific concern about the way the experiments were carried out. The model setup (Section 2) includes only brief mention of isostatic effects. These are really significant in a run of this sort because the elevation of the ice surface could be as much as $\sim$1800 m too high without the effect. Given that lapse rates are used to re-scale the climate data to elevations that are altered by the presence of the ice sheet, and that the sliding relation is strongly controlled by bedrock elevation, I think a more detailed discussion of isostacy is warranted. Specifically, given that the goal is an LGM configuration, I'd like assurance that the bedrock elevation has come close to equilibrium at the 10ka point that is reported on. This assurance should be simple to provide with minor modifications to the model setup section.}

Thank you for your comment. We apologize for the lack of detail on these issues in the discussion paper, which possibly lead to confusion from your side.

The bedrock deformation model used in our study describes the flexure of an elastic lithosphere on top of an infinite half-space viscous mantle \citep{lingle-clark-1985}. It can be described by a single differential equation of the bed elevation $u(x, y, t)$,

\begin{equation}
    2 \nu |\nabla| \frac{\partial u}{\partial t}
    + \rho_r g u
    + D \nabla^4 u
    = \sigma_{zz},
\end{equation}

where $\nu=1\times10^{21}$\,Pa\,s is mantle viscosity, $\rho_r = 3300$\,kg\,m$^{-3}$ is lithosphere density, $D=5\times10^{24}$\,N\,m is the lithosphere's flexural rigidity, and $\sigma_{zz}$ corresponds to the ice load \citet{bueler-etal-2007}. In the left-hand part of this equation, the first term accounts for mantle relaxation, the second for point-wise isostasy, and the third for elastic flexure. Please refer to \citet{bueler-etal-2007} for a definition of the pseudo-differential operator $|\nabla|$. Because mantle viscosity $\nu$ is non-zero, there exist a time lag between ice sheet growth and the isostatic bedrock response.

We would like to emphasize that most of our simulations do not reach a climatic equilibrium. In fact, simulations ran to climatic equilibrium (not shown in the paper) produce excessive ice cover in continental regions regardless of the forcing climatology used. Moreover, geomorphological and palaeoclimatic records give support for a short-lived Cordilleran ice sheet. Because the modelled ice sheet is in constant expansion, and because of the first term in the above equation, bedrock elevation do not come to equilibrium with the ice sheet configuration in our runs.

Although we do not believe that bedrock elevation must have been at equilibrium with the ice sheet configuration at the LGM, we recognize that through our simplistic palaeoclimatic approach, isostatic effects are another significant source of error on modelled ice surface elevation, which we omitted to discuss in our manuscript.

Finally, we also would like to emphasize that the sliding relation is controlled by initial bedrock elevation, and is not altered by its evolution through the run. Your comments show that we need more details about computation of bedrock deformation and its effects in the manuscript. These will for sure be added in the revised version.

\referee{The only writing problem I found was in figure 5, `logarythmic' should be `logarithmic'. Many thanks to the authors for getting the wording right the first time.}

Thank you for noticing this mistake. We will correct it.

\begin{thebibliography}{69}

\bibitem[{Lingle and Clark(1985)}]{lingle-clark-1985}
Lingle,~C.~S. and Clark,~J.~A.: A~numerical model of interactions between a~marine ice sheet and the solid Earth: application to a~West Antarctic ice stream, J. Geophys. Res., 90, 1100--1114, \doi{10.1029/JC090iC01p01100}, 1985.

\bibitem[{Bueler et~al.(2007)}]{bueler-etal-2007}
Bueler,~E. and Lingle,~C.~S. and Brown,~J.: Fast computation of a viscoelastic deformable Earth model for ice-sheet simulations, Ann. Glaciol., 46, 97--105, \doi{10.3189/172756407782871567}, 2007

\end{thebibliography}

\end{document}

### 2014-03-26 reviews/disc-ac1.tex
Use different figure numbers in author comment

@@
 \begin{document}
-
 \textbf{Authors' response to Referee {\#}1}
@@
 % ----------------------------------------------------------------------
+% Interactive comment text begins
+% ----------------------------------------------------------------------
+
+\renewcommand\thefigure{AC\arabic{figure}}
 \def\referee#1{\bigskip\textcolor{journalname}{\textit{#1}}}
@@
 
-For paper Fig.~9 (temperature difference maps, Fig.~1 in this comment), lapse-rate correction results in smoother temperature difference maps, where effective temperature discrepancies are no longer overshadowed by local topographic anomalies (this comment's Fig.~2).
-Because temperature differences in this new map are much reduced, we adjusted colour scaling to emphasize smaller values (this comment's Fig.~3).
+For paper Fig.~9 (temperature difference maps, Fig.~AC1), lapse-rate correction results in smoother temperature difference maps, where effective temperature discrepancies are no longer overshadowed by local topographic anomalies (Fig.~AC2).
+Because temperature differences in this new map are much reduced, we adjusted colour scaling to emphasize smaller values (Fig.~AC3).
 
-For paper Fig.~8 (temperature density maps), we now use bilinearly-interpolated temperature maps, rather than the original data (as presented in paper Fig.~2--4).
-This results in smoother density plots than those presented in the discussion paper (this comment's Fig.~4).
-In a second step, we apply a lapse-rate correction to project all reanalysis data onto the WorldClim (higher resolution) topography (this comment's Fig.~5).
+For paper Fig.~8 (temperature density maps), we now use bilinearly-interpolated temperature maps, rather than the original data (as presented in Fig.~2--4).
+This results in smoother density plots than those presented in the discussion paper (Fig.~AC4).
+In a second step, we apply a lapse-rate correction to project all reanalysis data onto the WorldClim (higher resolution) topography (Fig.~AC5).
 It now becomes clear that much of the discrepancies observed were due to lapse-rate effects.
 
-For consistency, paper Fig.~10 (precipitation density maps) was also updated using bilinearly-interpolated data (this comment's Fig.~6).
+For consistency, paper Fig.~10 (precipitation density maps) was also updated using bilinearly-interpolated data (this comment's Fig.~AC6).
 Note that for all density maps, colour maps were changed to allow for a higher level of detail.
@@
 % ----------------------------------------------------------------------
+% Interactive comment text ends
+% ----------------------------------------------------------------------
 
@@
 \end{figure}
+
 \begin{figure}

### 2014-03-26 reviews/disc-ac1.tex
Add corrections from Irina

@@
 
-Thank you very much for this supportive review. Your comments raise valid points about presentation and interpretation of the results. Please find our response to them below.
+Thank you very much for reading our manuscript and writing this supportive review.
+Your comments raise valid points about presentation and interpretation of the results.
+Please find our response below, in which we discuss these points with support of updated figures.
 
@@
 Due to differences in resolution of surface topography between climatological datasets used, this caused appearance of numerous topographic artefacts on paper Fig.~9.
-Following your comment, we produced updated figures to include a lapse-rate correction of 6\,{$^\circ$}C\,km$^{-1}$, as applied in the simulations.
+Following your comment, we produced updated figures that include a lapse-rate correction of 6\,{$^\circ$}C\,km$^{-1}$, as applied in the simulations.
 
-For paper Fig.~9 (temperature difference maps, Fig.~AC1), lapse-rate correction results in smoother temperature difference maps, where effective temperature discrepancies are no longer overshadowed by local topographic anomalies (Fig.~AC2).
+The implementation of lapse-rate correction in Fig.~9 (temperature difference maps, Fig.~AC1)  results in smoother temperature difference maps, where effective temperature discrepancies are no longer overshadowed by local topographic anomalies (Fig.~AC2).
 Because temperature differences in this new map are much reduced, we adjusted colour scaling to emphasize smaller values (Fig.~AC3).
 
-For paper Fig.~8 (temperature density maps), we now use bilinearly-interpolated temperature maps, rather than the original data (as presented in Fig.~2--4).
+Also, the new Fig.~8 (temperature density maps) now use bilinearly-interpolated temperature maps, rather than the original data (as presented in Fig.~2--4).
 This results in smoother density plots than those presented in the discussion paper (Fig.~AC4).
@@
 
-For consistency, paper Fig.~10 (precipitation density maps) was also updated using bilinearly-interpolated data (this comment's Fig.~AC6).
-Note that for all density maps, colour maps were changed to allow for a higher level of detail.
+For consistency, Fig.~10 (precipitation density maps) was also updated using bilinearly-interpolated data (this comment's Fig.~AC6).
+Note that for all density maps, colour scales were changed to allow for a higher level of detail.
 
@@
 A short discussion of these effects is given in Sect.~5.3, in the context of comparison of simulation results to the geomorphological last glacial maximum ice margin.
-However you raise here an issue that we had not consider, regarding potential ice sheet evolution to more similar states when accounting for precipitation changes.
-In the East more particularly, where in most of our simulations, the ice margin do not attain a steady configuration, its position after 10\,kyr is largely determined by its rate of advance during the simulation length, which in turn depends on the amount of precipitation received.
-Therefore, precipitation corrections in this region could significantly slow down the ice margin advance, resulting in more similar configuration after 10\,kyr.
+However you raise here a valid concern regarding potential ice sheet evolution to more similar states when accounting for precipitation changes, not considered in the current version of our manuscript.
+In the East more particularly, where in most of our simulations, the ice margin does not attain a steady-state configuration, its position after 10\,kyr is largely determined by its rate of advance during the simulation length, which in turn depends on the amount of precipitation received.
+We agree that precipitation corrections in this region could slow down the ice margin advance, resulting in more similar configuration after 10\,kyr.
 We will include a discussion of this effect in Sect.~5.5.

### 2014-03-26 reviews/disc-ac2.tex
Add custom DOI command that works with TCD
Add corrections from Irina

@@
 \usepackage{natbib}
-\usepackage{doi}
 
@@
 \definecolor{journalname}{rgb}{0.34,0.59,0.82}
-%\providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}
 
@@
 % ----------------------------------------------------------------------
+% Interactive comment text begins
+% ----------------------------------------------------------------------
+
+\def\doi#1{doi:\allowbreak\href{http://dx.doi.org/#1}{#1}}
 \def\referee#1{\bigskip\textcolor{journalname}{\textit{#1}}}
@@
 Dear Referee {\#}2,
+
+Thank you very much for reading and positively commenting on our manuscript. We apologize for the lack of detail regarding isostatic issues, which we discuss below.
 
@@
 This assurance should be simple to provide with minor modifications to the model setup section.}
-
-Thank you for your comment. We apologize for the lack of detail on these issues in the discussion paper, which possibly lead to confusion from your side.
 
@@
 Please refer to \citet{bueler-etal-2007} for a definition of the pseudo-differential operator $|\nabla|$.
-Because mantle viscosity $\nu$ is non-zero, there exist a time lag between ice sheet growth and the isostatic bedrock response.
+Due to high mantle viscosity $\nu$, there exist a time lag between the ice sheet growth and the isostatic bedrock response.
 
@@
 Moreover, geomorphological and palaeoclimatic records give support for a short-lived Cordilleran ice sheet.
-Because the modelled ice sheet is in constant expansion, and because of the first term in the above equation, bedrock elevation do not come to equilibrium with the ice sheet configuration in our runs.
+Because the modelled ice sheet is in constant expansion, and because of the first term in the above equation, bedrock elevation does not come to equilibrium with the ice sheet configuration in our runs.
 
@@
 
-Finally, we also would like to emphasize that the sliding relation is controlled by initial bedrock elevation, and is not altered by its evolution through the run.
-Your comments show that we need more details about computation of bedrock deformation and its effects in the manuscript.
-These will for sure be added in the revised version.
+Through the till friction angle parameter $\phi$, sliding is controlled by bedrock elevation.
+This control is meant to reflect the presence of lubricating marine sediment at the ice sheet base.
+Such sediments are expected below current sea level, and, to a lesser extent, in areas that were submerged during former glacial periods, which is reflected in our simulation set-up by a linear increase of $\phi$ between 0 and 200 meters above sea level, where it reaches its maximum value.
+
+We want to clarify here that $\phi$ is computed only once and for the entire simulation length at the beginning of the run, based on modern bedrock elevation.
+Because the distribution of marine sediments is not likely to change much during a single phase of ice sheet advance and retreat, sliding is not affected by bedrock elevation changes due to isostatic effects in our runs.
+
+In general, it became clear to us through your comments that we need more details about computation of bedrock deformation and its effects in the manuscript.
+These will surely be added in the revised version.
 
@@
 
+% ----------------------------------------------------------------------
+% Interactive comment text ends
+% ----------------------------------------------------------------------
+
 \end{document}

### 2014-03-27 reviews/disc-ac1.tex
Minor changes for latex2rtf compatibility
Add more corrections from Irina

@@
 \usepackage[pdftex]{color}
-\usepackage{doi}
 
@@
 
-Also, the new Fig.~8 (temperature density maps) now use bilinearly-interpolated temperature maps, rather than the original data (as presented in Fig.~2--4).
+Also, the new Fig.~8 (temperature density maps) now uses bilinearly-interpolated temperature maps, rather than the original data (as presented in Fig.~2--4).
 This results in smoother density plots than those presented in the discussion paper (Fig.~AC4).
@@
 \begin{figure}
-    \center
+    \centering
     \includegraphics[width=70mm]{cordillera-climate-tempdiff+tcdc}
@@
 \begin{figure}
-    \center
+    \centering
     \includegraphics[width=70mm]{cordillera-climate-tempdiff+lr+tcdc}
@@
 \begin{figure}
-    \center
+    \centering
     \includegraphics[width=70mm]{cordillera-climate-tempdiff+lr}
@@
 \begin{figure}
-    \center
+    \centering
     \includegraphics[width=70mm]{cordillera-climate-tempheatmap}
@@
 \begin{figure}
-    \center
+    \centering
     \includegraphics[width=70mm]{cordillera-climate-tempheatmap+lr}
@@
 \begin{figure}
-    \center
+    \centering
     \includegraphics[width=70mm]{cordillera-climate-precheatmap}

### 2014-03-27 reviews/disc-ac2.tex
Add more corrections from Irina

@@
 Please refer to \citet{bueler-etal-2007} for a definition of the pseudo-differential operator $|\nabla|$.
-Due to high mantle viscosity $\nu$, there exist a time lag between the ice sheet growth and the isostatic bedrock response.
+Due to high mantle viscosity $\nu$, there exists a time lag between the ice sheet growth and the isostatic bedrock response.
 
@@
 
-Although we do not believe that bedrock elevation must have been at equilibrium with the ice sheet configuration at the LGM, we recognize that through our simplistic palaeoclimatic approach, isostatic effects are another significant source of error on modelled ice surface elevation, which we omitted to discuss in our manuscript.
-
-Through the till friction angle parameter $\phi$, sliding is controlled by bedrock elevation.
-This control is meant to reflect the presence of lubricating marine sediment at the ice sheet base.
-Such sediments are expected below current sea level, and, to a lesser extent, in areas that were submerged during former glacial periods, which is reflected in our simulation set-up by a linear increase of $\phi$ between 0 and 200 meters above sea level, where it reaches its maximum value.
-
-We want to clarify here that $\phi$ is computed only once and for the entire simulation length at the beginning of the run, based on modern bedrock elevation.
-Because the distribution of marine sediments is not likely to change much during a single phase of ice sheet advance and retreat, sliding is not affected by bedrock elevation changes due to isostatic effects in our runs.
+Although we do not believe that bedrock elevation must have been in equilibrium with the ice sheet configuration at the LGM, we recognize that through our simplistic palaeoclimatic approach, isostatic effects are another significant source of error on modelled ice surface elevation, which we omitted to discuss in our manuscript.
 

### 2014-03-31 reviews/disc-ac1.tex
Produce lapse-rate corrected figures by default

@@
     \centering
-    \includegraphics[width=70mm]{cordillera-climate-tempdiff+tcdc}
+    \includegraphics[width=70mm]{cordillera-climate-tempdiff+nolr+tcdc}
     \caption{Summer temperature difference maps, as in paper Fig.~9.}
@@
     \centering
-    \includegraphics[width=70mm]{cordillera-climate-tempdiff+lr+tcdc}
+    \includegraphics[width=70mm]{cordillera-climate-tempdiff+tcdc}
     \caption{Summer temperature difference maps, after applying a lapse-rate correction of 6\,{$^\circ$}C\,km$^{-1}$.}
@@
     \centering
-    \includegraphics[width=70mm]{cordillera-climate-tempdiff+lr}
+    \includegraphics[width=70mm]{cordillera-climate-tempdiff}
     \caption{Summer temperature difference maps, after applying a lapse-rate correction of 6\,{$^\circ$}C\,km$^{-1}$ and an updated colour mapping.}
@@
     \centering
-    \includegraphics[width=70mm]{cordillera-climate-tempheatmap}
+    \includegraphics[width=70mm]{cordillera-climate-tempheatmap+nolr}
     \caption{Summer temperature density maps, after bilinear interpolation.}
@@
     \centering
-    \includegraphics[width=70mm]{cordillera-climate-tempheatmap+lr}
+    \includegraphics[width=70mm]{cordillera-climate-tempheatmap}
     \caption{Summer temperature density maps, after bilinear interpolation and lapse-rate correction.}

### 2014-03-31 figures/plot.py

@@ -110,7 +110,7 @@ def tempbox():
     # save
     _savefig('cordillera-climate-tempbox')
 
-def tempdiff(tempdiff_lapse_rate=False, tcd_colors=False):
+def tempdiff(no_lapse_rate=False, tcd_colors=False):
     """Plot temperature difference maps"""
 
     # initialize figure
@@ -137,7 +137,7 @@ def tempdiff(tempdiff_lapse_rate=False, tcd_colors=False):
       var = nc.variables['air_temp']
       data = _seasonmean(var, 'jja')
       z = nc.variables['usurf'][:].T
-      if tempdiff_lapse_rate:
+      if not no_lapse_rate:
         data += 0.006*(z-zref)
       if tcd_colors:
         norm = mcolors.Normalize(-10.0, 10.0)
@@ -151,10 +151,10 @@ def tempdiff(tempdiff_lapse_rate=False, tcd_colors=False):
     cb = fig.colorbar(im, fig.grid.cbar_axes[0], ticks=None)
     cb.set_label(u'JJA surface air temperature difference to WorldClim data (°C)')
     _savefig('cordillera-climate-tempdiff'
-             + tempdiff_lapse_rate*'+lr'
+             + no_lapse_rate*'+nolr'
              + tcd_colors*'+tcdc')
 
-def tempheatmap(tempdiff_lapse_rate=False, tcd_colors=False):
+def tempheatmap(no_lapse_rate=False, tcd_colors=False):
     """Plot temperature heat maps"""
 
     from mpl_toolkits.axes_grid1 import ImageGrid
@@ -188,7 +188,7 @@ def tempheatmap(tempdiff_lapse_rate=False, tcd_colors=False):
       var = nc.variables['air_temp']
       data = _seasonmean(var, 'jja') - 273.15
       z = nc.variables['usurf'][:].T
-      if tempdiff_lapse_rate:
+      if not no_lapse_rate:
         data +=0.006*(z-zref)
       data = np.ma.array(data, mask=ref.mask).compressed()
 
@@ -215,7 +215,7 @@ def tempheatmap(tempdiff_lapse_rate=False, tcd_colors=False):
     cb.set_label(u'number of grid points')
     fig.text(37/85., 2/72., u'JJA mean surface air temperature (°C)', ha='center')
     _savefig('cordillera-climate-tempheatmap'
-             + tempdiff_lapse_rate*'+lr'
+             + no_lapse_rate*'+nolr'
              + tcd_colors*'+tcdc')
 
 def prec():
[…]

### 2014-03-31 cordillera-climate.tex
Tabs to spaces
Add formatted references
Adapt article text to lapse-rate corrected figures

@@
 
-% Recognize utf-8 special characters
+% Encoding
 \usepackage[T1]{fontenc}
-\usepackage[utf8]{inputenc}
 
-% Set figures repository
+% Figure directory
 \graphicspath{{figures/}}
@@
 In this section we use the WorldClim climatology, which is derived from observations, as a~reference with which to compare other climate forcing data on land.
+
+Because most of the disparity between reanalytic and observational temperature data is caused by topographical detail at scales unresolved in the reanalysis datasets, we apply a lapse-rate correction prior to this comparison, using the same lapse rate of 6\,\unit{{\degree}C\,km^{-1}} as in the simulations.
 ERA-Interim, CFSR and NARR climatologies exhibit temperature distributions fairly consistent with WorldClim data (Fig.~\ref{fig:tempheatmap}).
-When examining the spatial distribution of temperature differences between reanalyses and WorldClim data, it appears that most of the disparity between reanalyses and observation data is caused by topographical detail at scales unresolved in the reanalysis datasets (Fig.~\ref{fig:tempdiff}).
-Surface air temperature data from the NCEP/NCAR climatology shows the largest disparity with that of WorldClim, with a~significant cold anomaly over most of the modelling domain (Figs.~\ref{fig:tempheatmap} and \ref{fig:tempdiff}).
+The spatial distribution of temperature differences between reanalyses and WorldClim data shows partly consistent patterns, which may relate to weaknesses of WorldClim data in areas of sparse observational coverage (Fig.~\ref{fig:tempdiff}).
+However significant disparity among reanalyses exists, too.
+More specifically, surface air temperature data from the NCEP/NCAR climatology shows the largest disparity with that of WorldClim, with a~significant cold anomaly over most of the modelling domain (Figs.~\ref{fig:tempheatmap} and \ref{fig:tempdiff}).
 
@@
 % References
-\bibliographystyle{copernicus}
-\bibliography{references}
+\input{references.bbl}
+\newpage
 
@@
 % ----------------------------------------------------------------------
-\newpage
 
@@
   \caption{Density maps showing a~comparison of summer (JJA) surface air temperature data from the WorldClim climatology, against that of each reanalysis.
-Apparent horizontal lines are an artefact of different horizontal resolutions.
+Climate data are presented after bilinear spatial interpolation and correction for topographic differences to WorldClim data, using a lapse rate of 6\,\unit{{\degree}C\,km^{-1}}.
 Colour mapping is based on a~logarythmic scale.
@@
   \includegraphics{cordillera-climate-tempdiff}
-	\caption{Summer (JJA) surface air temperature difference maps against WorldClim data, after bi-linear spatial interpolation.
+  \caption{Summer (JJA) surface air temperature difference maps against WorldClim data, after bi-linear spatial interpolation and lapse-rate correction.
 Note the cold bias of NCEP/NCAR data and temperature anomalies due to unresolved topographic detail.}
@@
   \includegraphics{cordillera-climate-precheatmap}
-	\caption{Density maps showing a~comparison of winter (DJF) precipitation rate from the WorldClim climatology, against that of each reanalysis.
-Apparent horizontal lines are an artefact of different horizontal resolutions.
-Apparent vertical lines at low values are an artefact of the low precision of WorldClim data.
+  \caption{Density maps showing a~comparison of winter (DJF) precipitation rate from the WorldClim climatology, against that of each reanalysis, after bi-linear spatial interpolation.
 Colour mapping is based on a~logarythmic scale.

### 2014-03-31 cordillera-climate.tex
Correct three typos (reviewers)

@@
 In a~model of coarser resolution, these high mountain ranges are reduced to smooth hills and the modelled distribution of precipitation is also smoother (Fig.~\ref{fig:oroprecip}).
-However, this effect alone cannot explain the widespread and pronounced positive precipitation anomalies observed in CFSR data (Figs.~\ref{fig:precheatmap} and \ref{fig:precdiff}) despite of its high spatial resolution (Table~\ref{tab:reanalyses}).
+However, this effect alone cannot explain the widespread and pronounced positive precipitation anomalies observed in CFSR data (Figs.~\ref{fig:precheatmap} and \ref{fig:precdiff}) despite its high spatial resolution (Table~\ref{tab:reanalyses}).
 
@@
 Climate data are presented after bilinear spatial interpolation and correction for topographic differences to WorldClim data, using a lapse rate of 6\,\unit{{\degree}C\,km^{-1}}.
-Colour mapping is based on a~logarythmic scale.
+Colour mapping is based on a~logarithmic scale.
 Note the cold bias of NCEP/NCAR data relative to WorldClim data.}
@@
   \caption{Density maps showing a~comparison of winter (DJF) precipitation rate from the WorldClim climatology, against that of each reanalysis, after bi-linear spatial interpolation.
-Colour mapping is based on a~logarythmic scale.
+Colour mapping is based on a~logarithmic scale.
 Note the wet bias of all reanalysis data relative to WorldClim data.}

### 2014-03-31 cordillera-climate.tex
Clarify that phi is based on modern topography

@@
 The yield stress is modelled using the Mohr--Coulomb criterion.
-The till friction angle $\phi$ varies from 10 to 30\degree\ and is a~function of current bed elevation, with lowest values occurring at low elevations,
+The till friction angle $\phi$ varies from 10 to 30{\degree}.
+It is computed once at the beginning of the run, as a~function of modern bed elevation, with lowest values occurring at low elevations,
 
@@
 Sea level is lowered by 120\,m and basal topography responds to ice load following a~regional isostasy model that includes lithospheric flexure and mantle relaxation \citep{lingle-clark-1985}.
+Till friction angle $\phi$ is not affected by bed elevation changes.
 

### 2014-03-31 cordillera-climate.tex
Negate offsets and lapse rate in figs. and text

@@
 It uses a~reference topography distinct from the modelled basal topography and specific to each climate forcing dataset (Fig.~\ref{fig:topo}).
-An annual lapse rate of 6\,\unit{{\degree}C\,km^{-1}} is used in all simulations.
+An annual lapse rate of -6\,\unit{{\degree}C\,km^{-1}} is used in all simulations.
 No corrections apply to precipitation changes with elevation.
@@
 
-Using WorldClim, NCEP/NCAR, ERA-Interim, CFSR (smoothed and original datasets) and NARR input climatologies as climate forcing, we run 66~simulations of glacial inception and growth of the Cordilleran ice sheet over 10\,kyr using temperature offsets ranging from 0 to 10\,\unit{{\degree}C}.
+Using WorldClim, NCEP/NCAR, ERA-Interim, CFSR (smoothed and original datasets) and NARR input climatologies as climate forcing, we run 66~simulations of glacial inception and growth of the Cordilleran ice sheet over 10\,kyr using temperature offsets ranging from -10 to 0\,\unit{{\degree}C}.
 Significant differences in patterns of ice sheet growth, final glacial extent, and volume are attained (Figs.~\ref{fig:ivolarea}--\ref{fig:cool05}).
@@
 Across the range of temperature offsets used, final glaciated area and final ice volume differ widely between different input climatologies (Fig.~\ref{fig:ivolarea}).
-For instance, using a~5\,\unit{{\degree}C} temperature offset, final glaciated area differs between WorldClim and CFSR forcing by a~factor of~6 and final volume by a~factor of~12 (Fig.~\ref{fig:ivolarea}).
+For instance, using a~-5\,\unit{{\degree}C} temperature offset, final glaciated area differs between WorldClim and CFSR forcing by a~factor of~6 and final volume by a~factor of~12 (Fig.~\ref{fig:ivolarea}).
 For particular temperature offset values, given input climatologies result in similar final glaciated areas and ice volumes, yet regional patterns are different (Fig.~\ref{fig:extent}).
@@
 
-To contrast the effect of the six different input climatologies, we plot simulations using a~5\,\unit{{\degree}C} temperature offset (Fig.~\ref{fig:cool05}).
+To contrast the effect of the six different input climatologies, we plot simulations using a~-5\,\unit{{\degree}C} temperature offset (Fig.~\ref{fig:cool05}).
 The magnitude of glaciation differs dramatically from one climate forcing to another.
@@
 
-Because most of the disparity between reanalytic and observational temperature data is caused by topographical detail at scales unresolved in the reanalysis datasets, we apply a lapse-rate correction prior to this comparison, using the same lapse rate of 6\,\unit{{\degree}C\,km^{-1}} as in the simulations.
+Because most of the disparity between reanalytic and observational temperature data is caused by topographical detail at scales unresolved in the reanalysis datasets, we apply a lapse-rate correction prior to this comparison, using the same lapse rate of -6\,\unit{{\degree}C\,km^{-1}} as in the simulations.
 ERA-Interim, CFSR and NARR climatologies exhibit temperature distributions fairly consistent with WorldClim data (Fig.~\ref{fig:tempheatmap}).
@@
 These new simulations use a ``hybrid'' climate forcing that consists of temperature data from WorldClim, combined with precipitation data from each of the four reanalyses, and vice-versa.
-A~single temperature offset of 5\,\unit{{\degree}C} is used.
+A~single temperature offset of -5\,\unit{{\degree}C} is used.
 This value was chosen to allow for comparison with previous results (Fig.~\ref{fig:cool05}).
@@
 
-Although only temperature or precipitation changes are applied, this experiment results once more in large differences in modelled ice sheet geometries (Fig.~\ref{fig:biatm}) as compared to the reference WorldClim 5\,\unit{{\degree}C} run (Fig.~\ref{fig:cool05}, top-left panel).
+Although only temperature or precipitation changes are applied, this experiment results once more in large differences in modelled ice sheet geometries (Fig.~\ref{fig:biatm}) as compared to the reference WorldClim -5\,\unit{{\degree}C} run (Fig.~\ref{fig:cool05}, top-left panel).
 For ERA-Interim, CFSR and NARR climatologies, precipitation anomalies clearly dominate the differences in ice sheet response, whereas temperature anomalies have relatively little effect (Fig.~\ref{fig:biatmbars}).
@@
 In order to compare modelled ice sheets of a~similar size, we select for each climate forcing the simulation that leads to a~final glaciated area closest to the approximate size of the LGM Cordilleran ice sheet of $2\times 10^6\,\unit{km^2}$.
-This corresponds to temperature offset values of 8\,\unit{{\degree}C} (WorldClim), 6\,\unit{{\degree}C} (ERA-Interim) and 7\,\unit{{\degree}C} (NARR).
+This corresponds to temperature offset values of -8\,\unit{{\degree}C} (WorldClim), -6\,\unit{{\degree}C} (ERA-Interim) and -7\,\unit{{\degree}C} (NARR).
 These qualitative ``best'' runs are presented in Fig.~\ref{fig:best} along with their associated temperature depressions and a~reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.
@@
 
-To test the effect of an inception period of shorter duration on model results, we use simulations run under NARR climate forcing with temperature offsets of 7 to 11\,\unit{{\degree}C}, and compare modelled ice sheet extent when glaciated area reaches the approximate size of the LGM Cordilleran ice sheet of $2\times 10^6\,\unit{km^2}$.
+To test the effect of an inception period of shorter duration on model results, we use simulations run under NARR climate forcing with temperature offsets of -11 to -7\,\unit{{\degree}C}, and compare modelled ice sheet extent when glaciated area reaches the approximate size of the LGM Cordilleran ice sheet of $2\times 10^6\,\unit{km^2}$.
 
@@
   \includegraphics{cordillera-climate-cool05}
-  \caption{Ice surface topography (1\,\unit{km} contours) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a~climate 5\,\unit{{\degree}C} colder than present for each climate forcing used.}
+  \caption{Ice surface topography (1\,\unit{km} contours) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a~climate -5\,\unit{{\degree}C} colder than present for each climate forcing used.}
   \label{fig:cool05}
@@
   \caption{Density maps showing a~comparison of summer (JJA) surface air temperature data from the WorldClim climatology, against that of each reanalysis.
-Climate data are presented after bilinear spatial interpolation and correction for topographic differences to WorldClim data, using a lapse rate of 6\,\unit{{\degree}C\,km^{-1}}.
+Climate data are presented after bilinear spatial interpolation and correction for topographic differences to WorldClim data, using a lapse rate of -6\,\unit{{\degree}C\,km^{-1}}.
 Colour mapping is based on a~logarithmic scale.
@@
 In other words, the upper row shows the effect of temperature anomalies, and the lower row the effect of precipitation anomalies, for each reanalysis, relative to WorldClim data.
-Each simulation uses a~5\,\unit{{\degree}C} offset for comparison with Fig.~\ref{fig:cool05}.}
+Each simulation uses a~-5\,\unit{{\degree}C} offset for comparison with Fig.~\ref{fig:cool05}.}
   \label{fig:biatm}
@@
   \includegraphics{cordillera-climate-biatmbars}
-  \caption{Effect of temperature and precipitation anomalies (separately and jointly) from each reanalysis on modelled final ice volume relative to the result of the WorldClim 5\,\unit{{\degree}C} offset simulation.
+  \caption{Effect of temperature and precipitation anomalies (separately and jointly) from each reanalysis on modelled final ice volume relative to the result of the WorldClim -5\,\unit{{\degree}C} offset simulation.
 Corresponding ice sheet geometries are presented in Figs.~\ref{fig:cool05} and~\ref{fig:biatm}.}
@@
   \caption{Left panel: model sensitivity to simulation length.
-Modelled glaciated area, using NARR forcing data, and temperature offsets from 0 to 15\,\unit{{\degree}C}, solid lines corresponding to values from 7 to 11\,\unit{{\degree}C}.
-Right panel: modelled ice margin corresponding to temperature offsets from 7 to 11\,\unit{{\degree}C} when total glaciated area reach the approximate size of the LGM Cordilleran ice sheet of $2 \times 10^6\,\unit{km^2}$.
+Modelled glaciated area, using NARR forcing data, and temperature offsets from -15 to 0\,\unit{{\degree}C}, solid lines corresponding to values from -11 to -7\,\unit{{\degree}C}.
+Right panel: modelled ice margin corresponding to temperature offsets from -11 to -7\,\unit{{\degree}C} when total glaciated area reach the approximate size of the LGM Cordilleran ice sheet of $2 \times 10^6\,\unit{km^2}$.
 A~reconstructed LGM ice sheet margin by \citet{dyke-2004} (Fig.~\ref{fig:locmap}) is given for reference (grey shading).

### 2014-03-31 figures/plot.py

@@ -378,14 +378,14 @@ def best():
 
     # initialize figure
     climates = ['wcnn', 'erai', 'narr'] #['cfsrs7', 'cfsr', 'ncar']
-    offsets  = ['08', '06', '07'] #['04', '04', '04']
+    offsets  = [-8, -6, -7] #[-4, -4, -4]
     fig = iplt.gridfigure(mapsize, (1, 3), cbar_mode='none', **figkwa)
 
     # loop on climate datasets
     for i, clim in enumerate(climates):
       ax = plt.axes(fig.grid[i])
-      nc = Dataset('../data/%s-%s.nc' % (clim, offsets[i]))
-      _annotate('%s - %s K' % (labels[clim], offsets[i]))
+      nc = Dataset('../data/%s-%02i.nc' % (clim, -offsets[i]))
+      _annotate('%s, %s K' % (labels[clim], offsets[i]))
       iplt.bedtopoimage(nc)
       iplt.icemargincontour(nc, linecolors=None, colors='white', alpha=0.5)
       iplt.icemargincontour(nc)
@@ -530,8 +530,8 @@ def duration():
 
     # initialize figure
     clim = 'narr'
-    alloffsets = range(16)
-    mapoffsets = range(7, 12)
+    alloffsets = range(-15, 1)
+    mapoffsets = range(-11, -6)
     tarea = 2.1823
     lgmtimes = []
     figw, figh = 115., 135.
@@ -546,8 +546,8 @@ def duration():
 
     # plot glaciatied area curves
     ax = plt.axes(grid[0])
-    for dt in range(16):
-      nc = Dataset('../data/%s-%02g-extra.nc' % (clim, dt))
+    for dt in alloffsets:
+      nc = Dataset('../data/%s-%02g-extra.nc' % (clim, -dt))
       time = nc.variables['time'][1:] / 31556925974.7
       iarea = (nc.variables['thk'][1:]>10).sum(axis=(1, 2))/1e4
       plt.plot(time, iarea, ('b-' if dt in mapoffsets else 'b--'))
@@ -561,7 +561,7 @@ def duration():
       ax = plt.axes(grid[i+1])
       ax.get_xaxis().set_visible(False)
       ax.get_yaxis().set_visible(False)
-      nc = Dataset('../data/%s-%02g-extra.nc' % (clim, dt))
+      nc = Dataset('../data/%s-%02g-extra.nc' % (clim, -dt))
[…]

### 2014-03-31 cordillera-climate.tex
Add detail on bedrock deformation and its effects

@@
 Basal topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography and bathymetry dataset with a~resolution of 1\,arc-min \citep{data:etopo1}.
-Sea level is lowered by 120\,m and basal topography responds to ice load following a~regional isostasy model that includes lithospheric flexure and mantle relaxation \citep{lingle-clark-1985}.
-Till friction angle $\phi$ is not affected by bed elevation changes.
+
+Sea level is lowered by 120\,m and basal topography responds to ice load following a bedrock deformation model that includes point-wise isostasy, elastic lithosphere flexure and viscous mantle deformation in a~semi-infinite half-space \citep{lingle-clark-1985,bueler-etal-2007}.
+It uses a lithosphere density of 3300\,\unit{kg\,m^{-3}}, a flexural rigidity of $5 \times 10^{24}$\,\unit{N\,m} and a mantle viscosity of $1 \times 10^{21}$\,\unit{Pa\,s}.
+Due to the high mantle viscosity, there exists a time lag between ice sheet growth and isostatic bedrock response.
+Because the distribution of marine sediments is assumed constant throughout the run, the till friction angle $\phi$ is not affected by changes in bed elevation.
 
@@
 
-A~key assumption for the present study is our choice of 10\,kyr for the duration of all our simulations.
+Most of the simulations presented in this study, and more particularly those that reproduce the LGM ice margin more closely (Fig.~\ref{fig:best}) do not reach a steady-state.
+Instead, rates of growth remain high throughout the run, and bed elevation does not come close to equilibrium with the ice load.
+In turn, a~key assumption for the present study is our choice of 10\,kyr for the duration of all our simulations.
 Although geomorphological data shows that the period of Cordilleran ice sheet inception, from nearly ice-free to full glacial conditions, could not have been much longer than 10\,kyr, they provide no lower bound to this value.
@@
 This approach implies large biases of surface mass-balance \citep{charbit-etal-2013,rau-rogozhina-2013,seguinot-2013}, particularly over regions with such various climates as experienced within our model domain.
+
+The bedrock deformation model uses homogeneous lithospheric properties and a semi-infinite mantle of constant viscosity.
+Thus, it does not represent lateral variations of crustal and mantle properties characteristic of tectonically active margins such as the American Cordillera, nor does it include influence from the neighbouring Laurentide ice sheet.
+These limitations should be acknowledged when interpreting ice surface elevation and volume reconstructions presented in this study.
 

### 2014-03-31 cordillera-climate.tex
Add some sentences on precipitation changes

@@
 To better understand the transient character of the LGM Cordilleran ice sheet, time-dependent palaeo-climate forcing would be needed.
-Although it is clear that past temperature changes over the region were neither homogeneous nor constant in time, and probably associated with precipitation changes, their patterns are not trivial and may potentially display a~strong inter-dependence with the evolution of the Cordilleran and Laurentide ice sheets.
-A~more correct, yet more complex approach would be to use coupling to a~GCM of intermediate complexity \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013}.
+Although it is clear that past temperature changes over the region were neither homogeneous nor constant in time, their patterns are not trivial and may potentially display a~strong inter-dependence with the evolution of the Cordilleran and Laurentide ice sheets.
+As previously discussed, potential effects of the growing ice-sheet on precipitation changes are not included in our model.
+Although the ice sheet interior likely became dryer, precipitation may increased along part of the margin where the presence of ice imposed ascending winds.
+In this context, using a~GCM of intermediate complexity \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013} may help producing more westerly-centred ice configurations than those presented here (Fig.~\ref{fig:best}).
 

### 2014-03-31 reviews/disc-ac1.tex
Remove TCD colors tempdiff maps from responses

@@
 The implementation of lapse-rate correction in Fig.~9 (temperature difference maps, Fig.~AC1) results in smoother temperature difference maps, where effective temperature discrepancies are no longer overshadowed by local topographic anomalies (Fig.~AC2).
-Because temperature differences in this new map are much reduced, we adjusted colour scaling to emphasize smaller values (Fig.~AC3).
+Because temperature differences in this new map are much reduced, we adjusted colour scaling to emphasize smaller values.
 
 Also, the new Fig.~8 (temperature density maps) now uses bilinearly-interpolated temperature maps, rather than the original data (as presented in Fig.~2--4).
-This results in smoother density plots than those presented in the discussion paper (Fig.~AC4).
-In a second step, we apply a lapse-rate correction to project all reanalysis data onto the WorldClim (higher resolution) topography (Fig.~AC5).
+This results in smoother density plots than those presented in the discussion paper (Fig.~AC3).
+In a second step, we apply a lapse-rate correction to project all reanalysis data onto the WorldClim (higher resolution) topography (Fig.~AC4).
 It now becomes clear that much of the discrepancies observed were due to lapse-rate effects.
 
-For consistency, Fig.~10 (precipitation density maps) was also updated using bilinearly-interpolated data (this comment's Fig.~AC6).
+For consistency, Fig.~10 (precipitation density maps) was also updated using bilinearly-interpolated data (this comment's Fig.~AC5).
 Note that for all density maps, colour scales were changed to allow for a higher level of detail.
@@
     \centering
-    \includegraphics[width=70mm]{cordillera-climate-tempdiff+nolr+tcdc}
+    \includegraphics[width=70mm]{cordillera-climate-tempdiff+nolr}
     \caption{Summer temperature difference maps, as in paper Fig.~9.}
@@
     \centering
-    \includegraphics[width=70mm]{cordillera-climate-tempdiff+tcdc}
+    \includegraphics[width=70mm]{cordillera-climate-tempdiff}
     \caption{Summer temperature difference maps, after applying a lapse-rate correction of 6\,{$^\circ$}C\,km$^{-1}$.}
@@
     \centering
-    \includegraphics[width=70mm]{cordillera-climate-tempdiff}
-    \caption{Summer temperature difference maps, after applying a lapse-rate correction of 6\,{$^\circ$}C\,km$^{-1}$ and an updated colour mapping.}
-\end{figure}
-
-\begin{figure}
-    \centering
+    \caption{Summer temperature density maps, after bilinear interpolation.}
     \includegraphics[width=70mm]{cordillera-climate-tempheatmap+nolr}
-    \caption{Summer temperature density maps, after bilinear interpolation.}
 \end{figure}

### 2014-03-31 reviews/disc-ac1.tex
Add minor corretions from Arjen to responses

@@
 In Figs.~8 and~9 of the discussion paper, temperature climatologies are compared without lapse-rate correction.
-Due to differences in resolution of surface topography between climatological datasets used, this caused appearance of numerous topographic artefacts on paper Fig.~9.
+Due to differences in resolution of surface topography between climatological datasets used, this caused appearance of numerous topographic artefacts in Fig.~9.
 Following your comment, we produced updated figures that include a lapse-rate correction of 6\,{$^\circ$}C\,km$^{-1}$, as applied in the simulations.
 
-The implementation of lapse-rate correction in Fig.~9 (temperature difference maps, Fig.~AC1) results in smoother temperature difference maps, where effective temperature discrepancies are no longer overshadowed by local topographic anomalies (Fig.~AC2).
+The implementation of a lapse-rate correction in Fig.~9 (temperature difference maps, Fig.~AC1) results in smoother temperature difference maps, where effective temperature discrepancies are no longer overshadowed by local topographic anomalies (Fig.~AC2).
 Because temperature differences in this new map are much reduced, we adjusted colour scaling to emphasize smaller values.
 
-Also, the new Fig.~8 (temperature density maps) now uses bilinearly-interpolated temperature maps, rather than the original data (as presented in Fig.~2--4).
+Also, the new Fig.~8 (temperature density maps) now uses bilinearly-interpolated temperature maps, rather than the original data (as presented in Figs.~2--4).
 This results in smoother density plots than those presented in the discussion paper (Fig.~AC3).
@@
 
-For consistency, Fig.~10 (precipitation density maps) was also updated using bilinearly-interpolated data (this comment's Fig.~AC5).
+For consistency, Fig.~10 (precipitation density maps) was also updated using bilinearly-interpolated data (Fig.~AC5).
 Note that for all density maps, colour scales were changed to allow for a higher level of detail.
 
-While these new figures present an altered version of climate forcing data used in our study, they more closely reflect how this data is read in by the ice sheet model, which seems appropriate for the discussion section.
+These new figures present an altered version of climate forcing data used in our study, and they now more closely reflect how this data is read in by the ice sheet model, which seems appropriate for the discussion section.
 Moreover, the new maps reveal more clearly the strengths and weakness of different datasets, and support more strongly our interpretation of the ``hybrid'' climate forcing experiments presented in Sect.~5.2.
-We will definitely include these new figures in the revised manuscript.
+These new figures are now included in the revised manuscript.
 Thank you very much for this constructive comment.
@@
 
-We will correct this.
+We have corrected this.
 
@@
 
-To avoid confusion, we will use negative values in the revised manuscript.
+To avoid confusion, we have adopted negative temperature offset values in the revised manuscript and figures where appropriate.
 
@@
 
-A short discussion of these effects is given in Sect.~5.3, in the context of comparison of simulation results to the geomorphological last glacial maximum ice margin.
+A short discussion of these effects is given in Sect.~5.3, in the context of comparison of simulation results to the field evidence of the last glacial maximum ice margin.
 However you raise here a valid concern regarding potential ice sheet evolution to more similar states when accounting for precipitation changes, not considered in the current version of our manuscript.
-In the East more particularly, where in most of our simulations, the ice margin does not attain a steady-state configuration, its position after 10\,kyr is largely determined by its rate of advance during the simulation length, which in turn depends on the amount of precipitation received.
+More particularly, in the east, where in most of our simulations, the ice margin does not attain a steady-state configuration, its position after 10\,kyr is largely determined by its rate of advance during the simulation length, which in turn depends on the amount of precipitation received.
 We agree that precipitation corrections in this region could slow down the ice margin advance, resulting in more similar configuration after 10\,kyr.
-We will include a discussion of this effect in Sect.~5.5.
+We have included a discussion of this effect in Sect.~5.5.
 

### 2014-04-01 reviews/disc-ac2.tex
Add minor corretions from Arjen to responses
Add manuscript quotes (work in progress)

@@
 \def\referee#1{\bigskip\textcolor{journalname}{\textit{#1}}}
+\def\msquote#1{\begin{quote}\textit{#1}\end{quote}}
 
@@
 
-Although we do not believe that bedrock elevation must have been in equilibrium with the ice sheet configuration at the LGM, we recognize that through our simplistic palaeoclimatic approach, isostatic effects are another significant source of error on modelled ice surface elevation, which we omitted to discuss in our manuscript.
+To clarify our methods, we have added the following paragraph:
+
+\msquote{Sea level is lowered by 120\,m and basal topography responds to ice load following a bedrock deformation model that includes point-wise isostasy, elastic lithosphere flexure and viscous mantle deformation in a~semi-infinite half-space \citep{lingle-clark-1985,bueler-etal-2007}.
+It uses a lithosphere density of 3300\,kg\,m$^{-3}$, a flexural rigidity of $5 \times 10^{24}$\,N\,m and a mantle viscosity of $1 \times 10^{21}$\,Pa\,s.
+Due to the high mantle viscosity, there exists a time lag between ice sheet growth and isostatic bedrock response.
+Because the distribution of marine sediments is assumed constant throughout the run, the till friction angle $\phi$ is not affected by changes in bed elevation.}
+
+Although we do not believe that bedrock elevation must have been in equilibrium with the ice sheet configuration at the LGM, we recognize that through our simplistic palaeoclimatic approach, isostatic effects are another significant source of error on modelled ice surface elevation.
+
+\msquote{The bedrock deformation model uses homogeneous lithospheric properties and a semi-infinite mantle of constant viscosity.
+Thus, it does not represent lateral variations of crustal and mantle properties characteristic of tectonically active margins such as the American Cordillera, nor does it include influence from the neighbouring Laurentide ice sheet.
+These limitations should be acknowledged when interpreting ice surface elevation and volume reconstructions presented in this study.}
 
 In general, it became clear to us through your comments that we need more details about computation of bedrock deformation and its effects in the manuscript.
-These will surely be added in the revised version.
+These have been added in the revised version.
 
@@
 
-Thank you for noticing this mistake. We will correct it.
+Thank you for noticing this mistake. We have corrected it.
 

### 2014-04-01 cordillera-climate.tex
Rework discussion of precipitation changes

@@
 As previously discussed, potential effects of the growing ice-sheet on precipitation changes are not included in our model.
-Although the ice sheet interior likely became dryer, precipitation may increased along part of the margin where the presence of ice imposed ascending winds.
-In this context, using a~GCM of intermediate complexity \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013} may help producing more westerly-centred ice configurations than those presented here (Fig.~\ref{fig:best}).
+These changes likely consist of a reduction of precipitation in continental regions and in the ice sheet interior, and an increase of precipitation along part of the margin where the presence of ice imposed ascending winds, which could result in a thinner ice sheet than modelled here.
+In addition, the final position of the eastern ice margin is largely controlled by its advance rate through the run.
+Therefore, precipitation shadowing effects may have resulted in more similar ice-sheet configurations if they were included in the model.
+Although using a GCM of intermediate complexity \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013} may represent a first step towards including ice sheet feedback on climate, their spatial resolution does not allow for accurate resolution of orographic precipitation changes in a region as mountainous as the North American Cordillera.
 

### 2014-04-01 reviews/disc-ac1.tex
Add manuscript quotes (work in progress)
Add manuscript quotes (complete)

@@
 Moreover, the new maps reveal more clearly the strengths and weakness of different datasets, and support more strongly our interpretation of the ``hybrid'' climate forcing experiments presented in Sect.~5.2.
-These new figures are now included in the revised manuscript.
+These new figures are now included in the revised manuscript, and caption and body text was reworked accordingly.
 Thank you very much for this constructive comment.
@@
 
-A short discussion of these effects is given in Sect.~5.3, in the context of comparison of simulation results to the field evidence of the last glacial maximum ice margin.
 However you raise here a valid concern regarding potential ice sheet evolution to more similar states when accounting for precipitation changes, not considered in the current version of our manuscript.
@@
 We have included a discussion of this effect in Sect.~5.5.
+
+\msquote{As previously discussed, potential effects of the growing ice-sheet on precipitation changes are not included in our model.
+These changes likely consist of a reduction of precipitation in continental regions and in the ice sheet interior, and an increase of precipitation along part of the margin where the presence of ice imposed ascending winds, which could result in a thinner ice sheet than modelled here.
+In addition, the final position of the eastern ice margin is largely controlled by its advance rate through the run.
+Therefore, precipitation shadowing effects may have resulted in more similar ice-sheet configurations if they were included in the model.
+Although using a GCM of intermediate complexity may represent a first step towards including ice sheet feedback on climate, their spatial resolution does not allow for accurate resolution of orographic precipitation changes in a region as mountainous as the North American Cordillera.}
+
+A more detailed discussion of potential effects of the growing ice-sheet on regional climate is given in Sect.~5.3 of the discussion paper, in the context of comparison of simulation results to the field evidence of the last glacial maximum ice margin.
 

### 2014-04-01 reviews/disc-ac2.tex
Add manuscript quotes (complete)

@@
 
-To clarify our methods, we have added the following paragraph:
+To clarify our methods, we have reworked the last sentence of Sect.~2.1 into the following short paragraph:
 
@@
 
-Although we do not believe that bedrock elevation must have been in equilibrium with the ice sheet configuration at the LGM, we recognize that through our simplistic palaeoclimatic approach, isostatic effects are another significant source of error on modelled ice surface elevation.
+Although we do not believe that bedrock elevation must have been in equilibrium with the ice sheet configuration at the LGM, we recognize that through our simplistic palaeoclimatic approach, isostatic effects are another significant source of error on modelled ice surface elevation, as stated in a new paragraph of Sect.~5.5.
 
@@
 These limitations should be acknowledged when interpreting ice surface elevation and volume reconstructions presented in this study.}
-
-In general, it became clear to us through your comments that we need more details about computation of bedrock deformation and its effects in the manuscript.
-These have been added in the revised version.
 

### 2014-04-01 reviews/disc-ac1.tex
Polish the responses and add polite words

@@
 In Figs.~8 and~9 of the discussion paper, temperature climatologies are compared without lapse-rate correction.
-Due to differences in resolution of surface topography between climatological datasets used, this caused appearance of numerous topographic artefacts in Fig.~9.
+Due to differences in resolution of surface topography between climatological datasets used, this caused appearance of numerous topographic artefacts (Fig.~AC1).
 Following your comment, we produced updated figures that include a lapse-rate correction of 6\,{$^\circ$}C\,km$^{-1}$, as applied in the simulations.
 
-The implementation of a lapse-rate correction in Fig.~9 (temperature difference maps, Fig.~AC1) results in smoother temperature difference maps, where effective temperature discrepancies are no longer overshadowed by local topographic anomalies (Fig.~AC2).
+The implementation of a lapse-rate correction in Fig.~9 (temperature difference maps) results in smoother maps, where effective temperature discrepancies are no longer overshadowed by local topographic anomalies (Fig.~AC2).
 Because temperature differences in this new map are much reduced, we adjusted colour scaling to emphasize smaller values.
@@
 
-However you raise here a valid concern regarding potential ice sheet evolution to more similar states when accounting for precipitation changes, not considered in the current version of our manuscript.
-More particularly, in the east, where in most of our simulations, the ice margin does not attain a steady-state configuration, its position after 10\,kyr is largely determined by its rate of advance during the simulation length, which in turn depends on the amount of precipitation received.
+We think that you raise here a valid concern.
+In the east, where in most of our simulations, the ice margin does not attain a steady-state configuration, its position after 10\,kyr is largely determined by its rate of advance during the simulation length, which in turn depends on the amount of precipitation received.
 We agree that precipitation corrections in this region could slow down the ice margin advance, resulting in more similar configuration after 10\,kyr.
@@
 A more detailed discussion of potential effects of the growing ice-sheet on regional climate is given in Sect.~5.3 of the discussion paper, in the context of comparison of simulation results to the field evidence of the last glacial maximum ice margin.
+
+Thank you again for your positive comments and constructive input.
 
@@
     \centering
+    \includegraphics[width=70mm]{cordillera-climate-tempheatmap+nolr}
     \caption{Summer temperature density maps, after bilinear interpolation.}
-    \includegraphics[width=70mm]{cordillera-climate-tempheatmap+nolr}
 \end{figure}

### 2014-04-01 reviews/disc-ac2.tex
Polish the responses and add polite words

@@
 
-Thank you very much for reading and positively commenting on our manuscript. We apologize for the lack of detail regarding isostatic issues, which we discuss below.
+Thank you very much for reading our manuscript and positively commenting on it. We apologize for the lack of detail regarding isostatic issues, which we discuss below.
 
@@
 
+We have emphasized the fact that bed elevation does not reach equilibrium at the beginning of Sect.~5.4.
+
+\msquote{Most of the simulations presented in this study, and more particularly those that reproduce the LGM ice margin more closely (Fig.~15) do not reach a steady-state.
+Instead, rates of growth remain high throughout the run, and bed elevation does not come close to equilibrium with the ice load.}
+
 Although we do not believe that bedrock elevation must have been in equilibrium with the ice sheet configuration at the LGM, we recognize that through our simplistic palaeoclimatic approach, isostatic effects are another significant source of error on modelled ice surface elevation, as stated in a new paragraph of Sect.~5.5.
@@
 
-Thank you for noticing this mistake. We have corrected it.
+Thank you for noticing this mistake. We have corrected it (in Figs.~8 and~10).
+
+In general, thank you again for this supportive and constructive review.
 

### 2014-04-01 cordillera-climate.tex
Add corrections from Irina in manuscript

--- 2014-03-31 cordillera-climate.tex
@@
 This model thus accounts for a~weakening of till associated with the presence of marine sediments \citep{martin-etal-2011,aschwanden-etal-2013}.
+Because the distribution of marine sediments is assumed constant throughout the run, the till friction angle $\phi$ is not affected by changes in bed elevation.
 Basal topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography and bathymetry dataset with a~resolution of 1\,arc-min \citep{data:etopo1}.
@@
 Due to the high mantle viscosity, there exists a time lag between ice sheet growth and isostatic bedrock response.
-Because the distribution of marine sediments is assumed constant throughout the run, the till friction angle $\phi$ is not affected by changes in bed elevation.
 
@@
 It uses a~reference topography distinct from the modelled basal topography and specific to each climate forcing dataset (Fig.~\ref{fig:topo}).
-An annual lapse rate of -6\,\unit{{\degree}C\,km^{-1}} is used in all simulations.
+An annual lapse rate of 6\,\unit{{\degree}C\,km^{-1}} is used in all simulations.
 No corrections apply to precipitation changes with elevation.
@@
 
-Because most of the disparity between reanalytic and observational temperature data is caused by topographical detail at scales unresolved in the reanalysis datasets, we apply a lapse-rate correction prior to this comparison, using the same lapse rate of -6\,\unit{{\degree}C\,km^{-1}} as in the simulations.
+Because much of the disparity between reanalytic and observational temperature data is caused by topographical detail at scales unresolved in the reanalyses, we apply a lapse-rate correction prior to this comparison, using the same lapse rate of 6\,\unit{{\degree}C\,km^{-1}} as in the simulations.
 ERA-Interim, CFSR and NARR climatologies exhibit temperature distributions fairly consistent with WorldClim data (Fig.~\ref{fig:tempheatmap}).
-The spatial distribution of temperature differences between reanalyses and WorldClim data shows partly consistent patterns, which may relate to weaknesses of WorldClim data in areas of sparse observational coverage (Fig.~\ref{fig:tempdiff}).
-However significant disparity among reanalyses exists, too.
+The spatial distribution of temperature differences between reanalyses and WorldClim data shows partly consistent patterns, such as positive anomalies in the northernmost part of the modelling domain.
+These may relate to weaknesses of WorldClim data in areas of sparse observational coverage (Fig.~\ref{fig:tempdiff}).
+However, significant disparity among reanalyses exists, too.
 More specifically, surface air temperature data from the NCEP/NCAR climatology shows the largest disparity with that of WorldClim, with a~significant cold anomaly over most of the modelling domain (Figs.~\ref{fig:tempheatmap} and \ref{fig:tempdiff}).
@@
 
-Most of the simulations presented in this study, and more particularly those that reproduce the LGM ice margin more closely (Fig.~\ref{fig:best}) do not reach a steady-state.
+Most of the simulations presented in this study, and more particularly those that reproduce the geomorphological LGM margin more closely (Fig.~\ref{fig:best}) do not reach a steady-state.
 Instead, rates of growth remain high throughout the run, and bed elevation does not come close to equilibrium with the ice load.
@@
 As previously discussed, potential effects of the growing ice-sheet on precipitation changes are not included in our model.
-Although the ice sheet interior likely became dryer, precipitation may increased along part of the margin where the presence of ice imposed ascending winds.
-In this context, using a~GCM of intermediate complexity \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013} may help producing more westerly-centred ice configurations than those presented here (Fig.~\ref{fig:best}).
+These changes likely consist of a reduction of precipitation in continental regions and in the ice sheet interior, and an increase of precipitation along part of the margin where the presence of ice imposed ascending winds.
+They could result in a more westerly-centred ice sheet than modelled here, with lower ice thickness in its interior.
+In addition, the final position of the eastern ice margin is largely controlled by its advance rate through the run.
+Therefore, the precipitation shadowing effects may have resulted in more similar ice-sheet configurations if they were included in the model.
+Although using a GCM of intermediate complexity \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013} may represent a first step towards including ice sheet feedback on climate, their spatial resolution do not allow for accurate modelling of orographic precipitation changes in a region as mountainous as the North American Cordillera.
 
@@
 In addition, we simulate temperature variability by using a~constant, uniform value of daily temperature standard deviation.
-This approach implies large biases of surface mass-balance \citep{charbit-etal-2013,rau-rogozhina-2013,seguinot-2013}, particularly over regions with such various climates as experienced within our model domain.
+This approach implies large biases of surface mass-balance \citep{charbit-etal-2013,rogozhina-rau-2013,seguinot-2013}, particularly over regions with such various climates as experienced within our model domain.
 
 The bedrock deformation model uses homogeneous lithospheric properties and a semi-infinite mantle of constant viscosity.
-Thus, it does not represent lateral variations of crustal and mantle properties characteristic of tectonically active margins such as the American Cordillera, nor does it include influence from the neighbouring Laurentide ice sheet.
+Thus, it does not represent lateral variations of crustal and mantle properties characteristic of tectonically active margins such as the American Cordillera, nor does it include the influence from the neighbouring Laurentide ice sheet.
 These limitations should be acknowledged when interpreting ice surface elevation and volume reconstructions presented in this study.
@@
   \caption{Density maps showing a~comparison of summer (JJA) surface air temperature data from the WorldClim climatology, against that of each reanalysis.
-Climate data are presented after bilinear spatial interpolation and correction for topographic differences to WorldClim data, using a lapse rate of -6\,\unit{{\degree}C\,km^{-1}}.
+Climate data are presented after bilinear spatial interpolation and correction for topographic differences to WorldClim data, using a lapse rate of 6\,\unit{{\degree}C\,km^{-1}}.
 Colour mapping is based on a~logarithmic scale.
@@
   \includegraphics{cordillera-climate-tempdiff}
-  \caption{Summer (JJA) surface air temperature difference maps against WorldClim data, after bi-linear spatial interpolation and lapse-rate correction.
+  \caption{Summer (JJA) surface air temperature difference maps against WorldClim data, after bilinear spatial interpolation and lapse-rate correction.
 Note the cold bias of NCEP/NCAR data and temperature anomalies due to unresolved topographic detail.}
@@
   \includegraphics{cordillera-climate-precheatmap}
-  \caption{Density maps showing a~comparison of winter (DJF) precipitation rate from the WorldClim climatology, against that of each reanalysis, after bi-linear spatial interpolation.
+  \caption{Density maps showing a~comparison of winter (DJF) precipitation rate from the WorldClim climatology, against that of each reanalysis, after bilinear spatial interpolation.
 Colour mapping is based on a~logarithmic scale.
@@
   \includegraphics{cordillera-climate-precdiff}
-  \caption{Winter (DJF) precipitation rate difference maps against WorldClim data, after bi-linear spatial interpolation.
+  \caption{Winter (DJF) precipitation rate difference maps against WorldClim data, after bilinear spatial interpolation.
 Note the wet bias of all reanalysis data and the large anomalies of CFSR and NCEP/NCAR data.}

### 2014-04-01 cordillera-climate.tex
Correct date in Rogozhina and Rau, 2014

@@
 In addition, we simulate temperature variability by using a~constant, uniform value of daily temperature standard deviation.
-This approach implies large biases of surface mass-balance \citep{charbit-etal-2013,rogozhina-rau-2013,seguinot-2013}, particularly over regions with such various climates as experienced within our model domain.
+This approach implies large biases of surface mass-balance \citep{charbit-etal-2013,rogozhina-rau-2014,seguinot-2013}, particularly over regions with such various climates as experienced within our model domain.
 

### 2014-04-03 cordillera-climate.tex
Add comments from Arjen in manuscript

@@
 % Affiliations
-\affil[1]{Department of Physical Geography and Quaternary Geology, Stockholm University, Stockholm, Sweden}
+\affil[1]{Department of Physical Geography and Quaternary Geology and the Bolin Centre for Climate Research, Stockholm University, Stockholm, Sweden}
 \affil[2]{Geophysical Institute, University of Alaska Fairbanks, Fairbanks, AK, USA}
@@
 
-At the Last Glacial Maximum (LGM), glaciers of a~size comparable to the present Greenland and Antarctic ice sheets covered parts of Northern America (Laurentide, Cordilleran and Innuitian ice sheets) and northern Eurasia (Fennoscandian ice sheet).
+At the Last Glacial Maximum (LGM), glaciers of a~size comparable to the present Greenland and Antarctic ice sheets covered parts of Northern America (Laurentide, Cordilleran and Innuitian ice sheets) and northern Eurasia (Fennoscandian and Eurasian ice sheets).
 Numerical modelling of these former ice masses allows for a~comparison between glaciological theories embedded in the models and geomorphological traces underpinning palaeo-glaciological reconstructions.
@@
 
-Because much of the disparity between reanalytic and observational temperature data is caused by topographical detail at scales unresolved in the reanalyses, we apply a lapse-rate correction prior to this comparison, using the same lapse rate of 6\,\unit{{\degree}C\,km^{-1}} as in the simulations.
+Because much of the disparity between reanalyses and observational temperature data is caused by topographical detail at scales unresolved in the reanalyses, we apply a lapse-rate correction prior to this comparison, using the same lapse rate of 6\,\unit{{\degree}C\,km^{-1}} as in the simulations.
 ERA-Interim, CFSR and NARR climatologies exhibit temperature distributions fairly consistent with WorldClim data (Fig.~\ref{fig:tempheatmap}).
@@
 Most likely, this is the signature of an orographic precipitation effect.
-As described in Sect.~\ref{sec:climate}, the topography of the Northern American Cordillera is such that its western ranges form a~continuous orographic barrier, causing high precipitation along the Pacific coast while leaving much of the interior arid.
-However the ability of a~GCM to reproduce these contrasts in precipitation is bounded by spatial resolution.
+The topography of the Northern American Cordillera is such that its western ranges form a~continuous orographic barrier, causing high precipitation along the Pacific coast while leaving much of the interior arid.
+The ability of a~GCM to reproduce these contrasts in precipitation is bounded by spatial resolution.
 In a~model of coarser resolution, these high mountain ranges are reduced to smooth hills and the modelled distribution of precipitation is also smoother (Fig.~\ref{fig:oroprecip}).
@@
 For ERA-Interim, CFSR and NARR climatologies, precipitation anomalies clearly dominate the differences in ice sheet response, whereas temperature anomalies have relatively little effect (Fig.~\ref{fig:biatmbars}).
-However in the case of NCAR forcing, both the negative temperature bias (Figs.~\ref{fig:tempheatmap} and \ref{fig:tempdiff}) and the important precipitation differences (Figs.~\ref{fig:precheatmap} and \ref{fig:precdiff}) contribute approximately equally to produce oversized ice sheets (Fig.~\ref{fig:biatmbars}).
+However, in the case of NCAR forcing, both the negative temperature bias (Figs.~\ref{fig:tempheatmap} and \ref{fig:tempdiff}) and the important precipitation differences (Figs.~\ref{fig:precheatmap} and \ref{fig:precdiff}) contribute approximately equally to produce oversized ice sheets (Fig.~\ref{fig:biatmbars}).
 
-\subsection{Comparison to geomorphological LGM margin}
+\subsection{Comparison to geomorphological LGM margin reconstruction}
 
@@
 This experiment shows that a~shorter start-up period leads to more restricted ice cover in the eastern part of the modelling domain, where precipitation rate controls ice advance, but more extensive glaciation in the south-western domain along the Pacific coast, where temperature is the limiting factor (Fig.~\ref{fig:durationstack}).
-Hence, reducing simulation duration leads to a~closer match with the LGM ice sheet margin by \citet{dyke-2004} in some parts of the modelling domain, but higher discrepancy in other parts.
+Hence, reducing simulation duration leads to a~closer match with the LGM ice sheet margin by \citet{dyke-2004} in some sectors of the modelling domain, but higher discrepancy in other sectors.
 

### 2014-04-03 cordillera-climate.tex
Update manuscript quotes in responses

@@
 
-Most of the simulations presented in this study, and more particularly those that reproduce the geomorphological LGM margin more closely (Fig.~\ref{fig:best}) do not reach a steady-state.
+Most of the simulations presented in this study, and more particularly those that reproduce the LGM ice margin reconstruction more closely (Fig.~\ref{fig:best}) do not reach a steady-state.
 Instead, rates of growth remain high throughout the run, and bed elevation does not come close to equilibrium with the ice load.
@@
 As previously discussed, potential effects of the growing ice-sheet on precipitation changes are not included in our model.
-These changes likely consist of a reduction of precipitation in continental regions and in the ice sheet interior, and an increase of precipitation along part of the margin where the presence of ice imposed ascending winds.
+These changes likely consisted of a reduction of precipitation in continental regions and in the ice sheet interior, and an increase of precipitation along part of the margin where the presence of ice imposed ascending winds.
 They could result in a more westerly-centred ice sheet than modelled here, with lower ice thickness in its interior.

### 2014-04-03 reviews/disc-ac1.tex
Update manuscript quotes in responses

@@
 \msquote{As previously discussed, potential effects of the growing ice-sheet on precipitation changes are not included in our model.
-These changes likely consist of a reduction of precipitation in continental regions and in the ice sheet interior, and an increase of precipitation along part of the margin where the presence of ice imposed ascending winds, which could result in a thinner ice sheet than modelled here.
+These changes likely consisted of a reduction of precipitation in continental regions and in the ice sheet interior, and an increase of precipitation along part of the margin where the presence of ice imposed ascending winds.
+They could result in a more westerly-centred ice sheet than modelled here, with lower ice thickness in its interior.
 In addition, the final position of the eastern ice margin is largely controlled by its advance rate through the run.
-Therefore, precipitation shadowing effects may have resulted in more similar ice-sheet configurations if they were included in the model.
-Although using a GCM of intermediate complexity may represent a first step towards including ice sheet feedback on climate, their spatial resolution does not allow for accurate resolution of orographic precipitation changes in a region as mountainous as the North American Cordillera.}
+Therefore, the precipitation shadowing effects may have resulted in more similar ice-sheet configurations if they were included in the model.
+Although using a GCM of intermediate complexity may represent a first step towards including ice sheet feedback on climate, their spatial resolution do not allow for accurate modelling of orographic precipitation changes in a region as mountainous as the North American Cordillera.}
 

### 2014-04-03 reviews/disc-ac2.tex
Update manuscript quotes in responses

@@
 It uses a lithosphere density of 3300\,kg\,m$^{-3}$, a flexural rigidity of $5 \times 10^{24}$\,N\,m and a mantle viscosity of $1 \times 10^{21}$\,Pa\,s.
-Due to the high mantle viscosity, there exists a time lag between ice sheet growth and isostatic bedrock response.
-Because the distribution of marine sediments is assumed constant throughout the run, the till friction angle $\phi$ is not affected by changes in bed elevation.}
+Due to the high mantle viscosity, there exists a time lag between ice sheet growth and isostatic bedrock response.}
 
@@
 
-\msquote{Most of the simulations presented in this study, and more particularly those that reproduce the LGM ice margin more closely (Fig.~15) do not reach a steady-state.
+\msquote{Most of the simulations presented in this study, and more particularly those that reproduce the LGM ice margin reconstruction more closely (Fig.~15) do not reach a steady-state.
 Instead, rates of growth remain high throughout the run, and bed elevation does not come close to equilibrium with the ice load.}
@@
 \msquote{The bedrock deformation model uses homogeneous lithospheric properties and a semi-infinite mantle of constant viscosity.
-Thus, it does not represent lateral variations of crustal and mantle properties characteristic of tectonically active margins such as the American Cordillera, nor does it include influence from the neighbouring Laurentide ice sheet.
+Thus, it does not represent lateral variations of crustal and mantle properties characteristic of tectonically active margins such as the American Cordillera, nor does it include the influence from the neighbouring Laurentide ice sheet.
 These limitations should be acknowledged when interpreting ice surface elevation and volume reconstructions presented in this study.}

### 2014-04-11 cordillera-climate.tex
Consistently use degrees C troughout the paper

@@
 The temperature standard deviation is a~constant model parameter and was assigned a~value of 3.07\,\unit{{\degree}C}, which corresponds to the mean summer (JJA), model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from North American Regional Reanalysis data \citep{data:narr} in a~manner similar to \citet{seguinot-2013}.
-The ablation model incorporates degree-day factors of 3.04\,\unit{mm\,K^{-1}\,day^{-1}} for snow and 4.59\,\unit{mm\,K^{-1}\,day^{-1}} for ice, as derived from mass-balance measurements on contemporary glaciers from the Coast Mountains and Rocky Mountains in British Columbia \citep{shea-etal-2009}.
+The ablation model incorporates degree-day factors of 3.04\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for snow and 4.59\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for ice, as derived from mass-balance measurements on contemporary glaciers from the Coast Mountains and Rocky Mountains in British Columbia \citep{shea-etal-2009}.
 

### 2014-04-11 figures/plot.py

@@ -385,7 +385,7 @@ def best():
     for i, clim in enumerate(climates):
       ax = plt.axes(fig.grid[i])
       nc = Dataset('../data/%s-%02i.nc' % (clim, -offsets[i]))
-      _annotate('%s, %s K' % (labels[clim], offsets[i]))
+      _annotate(u'%s, %s °C' % (labels[clim], offsets[i]))
       iplt.bedtopoimage(nc)
       iplt.icemargincontour(nc, linecolors=None, colors='white', alpha=0.5)
       iplt.icemargincontour(nc)

### 2014-04-17 cordillera-climate.tex
Add authors' response.

@@
 Therefore, the precipitation shadowing effects may have resulted in more similar ice-sheet configurations if they were included in the model.
-Although using a GCM of intermediate complexity \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013} may represent a first step towards including ice sheet feedback on climate, their spatial resolution do not allow for accurate modelling of orographic precipitation changes in a region as mountainous as the North American Cordillera.
+Although using a GCM of intermediate complexity \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013} may represent a first step towards including ice sheet feedback on climate, their spatial resolution does not allow for accurate modelling of orographic precipitation changes in a region as mountainous as the North American Cordillera.
 

### 2014-05-13 cordillera-cycle.tex
Initial commit.
Add minimal article file and Makefile.

--- 2014-04-03 cordillera-climate.tex
@@
 % Coloured hyperlinks
-\usepackage[colorlinks,citecolor=blue]{hyperref}
+\usepackage[colorlinks]{hyperref}
 
@@
 % Title
-\title{The effect of climate forcing on numerical simulations of the Cordilleran ice sheet at the Last Glacial Maximum}
+\title{Numerical simulation of the Cordilleran ice sheet through the last glacial cycle}
 
 % Authors
-\author[1]{J.~Seguinot}
-\author[2]{C.~Khroulev}
-\author[3]{I.~Rogozhina}
-\author[1]{A.~P.~Stroeven}
-\author[1]{Q.~Zhang}
+\author[1,2]{J.~Seguinot}
 \runningauthor{J.~Seguinot et~al.}
@@
 \affil[1]{Department of Physical Geography and Quaternary Geology and the Bolin Centre for Climate Research, Stockholm University, Stockholm, Sweden}
-\affil[2]{Geophysical Institute, University of Alaska Fairbanks, Fairbanks, AK, USA}
-\affil[3]{Helmholtz Centre Potsdam, GFZ German Research Centre for Geosciences, Potsdam, Germany}
+\affil[2]{Helmholtz Centre Potsdam, GFZ German Research Centre for Geosciences, Potsdam, Germany}
 
 % For Copernicus
-\received{22~November 2013}
-\accepted{5~December 2013}
+\received{}
+\accepted{}
 \published{}
@@
 \begin{abstract}
-  We present an ensemble of numerical simulations of the Cordilleran ice sheet during the Last Glacial Maximum performed with the Parallel Ice Sheet Model (PISM), applying temperature offsets to the present-day climatologies from five different datasets.
-Monthly mean surface air temperature and precipitation from WorldClim, the NCEP/NCAR reanalysis, the ERA-Interim reanalysis, the Climate Forecast System Reanalysis and the North American Regional Reanalysis are used to compute surface mass balance in a~positive degree-day model.
-Modelled ice sheet outlines and volumes appear highly sensitive to the choice of climate forcing.
-For three of the four reanalysis datasets used, differences in precipitation are the major source for discrepancies between model results.
-We assess model performance against a~geomorphological reconstruction of the ice margin at the Last Glacial Maximum, and suggest that part of the mismatch is due to unresolved orographic precipitation effects caused by the coarse resolution of reanalysis data.
-The best match between model output and the reconstructed ice margin is obtained using the high-resolution North American Regional Reanalysis, which we retain for simulations of the Cordilleran ice sheet in the future.
 \end{abstract}
@@
 
-At the Last Glacial Maximum (LGM), glaciers of a~size comparable to the present Greenland and Antarctic ice sheets covered parts of Northern America (Laurentide, Cordilleran and Innuitian ice sheets) and northern Eurasia (Fennoscandian and Eurasian ice sheets).
-Numerical modelling of these former ice masses allows for a~comparison between glaciological theories embedded in the models and geomorphological traces underpinning palaeo-glaciological reconstructions.
-Yet, a~major obstacle in this exercise resides in large uncertainties concerning the climate forcing, typically atmospheric temperature and precipitation, required as input to numerical glacier models \citep{hebeler-etal-2008}.
-This includes uncertainty in the representation of Earth's present climate in regions of poor station coverage, and even larger uncertainty concerning accurate reconstructions of past climate change.
-
-Arguably the most physically sound way to force an ice sheet model for simulations of glacial history is to couple it with a~General Circulation Model \citep[GCM;][]{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013}.
-However the computational demand of GCMs is such that only models of intermediate complexity can run on the time-scales of tens of thousands of years characteristic of ice sheet growth and decay.
-
-Climatologies obtained from GCM palaeo-climate simulations such as produced within the Paleoclimate Modelling Intercomparison Project \citep[PMIP;][]{joussaume-taylor-1995} provide perhaps a~more reasonable representation of past climate.
-However, as such climatologies are only available for specific periods of time, using them as climate forcing for an ice sheet model requires either an assumption of steady-state ice sheet response to climatic fluctuations \citep{huybrechts-tsiobbel-1996}, or interpolation through time between climatologies from different periods, which can be linear \citep{charbit-etal-2002}, or modulated by a ``glacial index'' weighting function derived from ice core \chem{\delta^{18}O} records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}.
-An important drawback in this approach is that GCM palaeo-climate simulations themselves rely on an accurate description of the global surface topography and therefore include global ice sheet reconstructions such as ICE-4G \citep{peltier-1994} for their surface topographic boundary condition.
-This condition could potentially exert influence on subsequently modelled ice sheet geometries, and particularly glacier extent.
-
-However, avoiding this circular dependence requires simplifying assumptions regarding Earth's past climate.
-Such studies include energy balance modelling approaches \citep{tarasov-peltier-1997} and geographic parametrizations of surface mass balance \citep{robert-1991} or climate forcing \citep{johnson-fastook-2002}.
-Alternatively, temperature offset methods \citep{greve-etal-1999,bintanja-etal-2005} make use of the high level of detail available in present climate datasets such as gridded observation datasets, GCM output or reanalyses, while using simplifying representations of past climate deviations from the present state.
-
-Here, we propose to address some of the uncertainties concerning climate forcing of numerical glacier models by evaluating the responses of a~numerical model, in terms of glacier extent, to inputs from several climate datasets.
-This is not entirely a~new approach as a~few studies of this kind are presently available.
-\citet{quiquet-etal-2012}, for example, assessed the sensitivity of a~Greenland ice sheet model to various climate forcing fields, including a~regional parametrization by \citet{fausto-etal-2009a}, output from several GCMs and an atmospheric reanalysis.
-\citet{rodgers-etal-2004} and \citet{charbit-etal-2007} tested the sensitivity of a~model of Northern Hemisphere ice sheets to climate forcing from different PMIP LGM simulations.
-Although using different set-ups, all three studies demonstrate the very large sensitivity of ice sheet models to the choice of climate forcing data used.
-To limit the degrees of freedom in our model and obtain results independent of palaeo-ice sheet reconstructions such as ICE-4G, we use a~simple temperature offset similar to the approaches by \citet{greve-etal-1999} and \citet{bintanja-etal-2005} and assess ice sheet model sensitivity to the choice of present-day climate data.
-Hence, rather than using GCM output, we force our model with climate reanalysis data, which builds on observational information through data assimilation \citep{bengtsson-etal-2007}.
-Furthermore, we focus our study regionally on the former Cordilleran ice sheet in western North America.
-
-The Cordilleran ice sheet (Fig.~\ref{fig:locmap}) covered an area that presently experiences strong regional variations in climate.
-From a~numerical modelling perspective, it is one of the least studied palaeo-ice sheets of the Northern Hemisphere, despite the fact that significant stratigraphical, geomorphological, and chronological data is available to constrain its extent \citep{jackson-clague-1991,porter-swanson-1998,dukrodkin-1999,dyke-2004,kaufman-manley-2004,kleman-etal-2010,stroeven-etal-2010,stroeven-etal-inpress,margold-etal-2011}.
-The Cordilleran ice sheet has previously been modelled as part of efforts to simulate ice sheets in North America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} and world-wide \citep{yoshimori-etal-2001}.
-While these studies generally reproduce the magnitude of North American glaciation at LGM reasonably well, there exists a~tendency in the simulations that are independent of ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and interior Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.
-
-Here we use PISM, a~Parallel Ice Sheet Model \citep{web:pism}, to simulate the extent and thickness of the Cordilleran ice sheet at the LGM (Fig.~\ref{fig:locmap}).
-We force our model with multiple climate datasets and compare the modelled ice extent to a~geomorphological reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.
-We aim to determine the climate dataset which is most suited for simulation of the Cordilleran ice sheet and to use as input for future, transient studies over a~glacial cycle.
-
 % ----------------------------------------------------------------------
@@
 % ----------------------------------------------------------------------
-
-Given basal topography, sea level, geothermal heat flux and climate forcing, the model computes ice extent and thickness, its thermal and dynamic state, and the associated lithospheric response.
-Our modelling domain encompasses the entire area covered by the Cordilleran ice sheet at LGM including independent ice growing on the western Alaskan ranges and the Brooks Range and intermediate expanses of ice-free terrain in northern Yukon Territory and interior Alaska (Fig.~\ref{fig:locmap}).
-
-As we aim to model glacial inception and growth of the Cordilleran ice sheet towards a~configuration as last attained during the LGM, we mimic palaeo-climatic conditions by applying constant temperature offsets homogeneously over the modelling domain.
-Each simulation starts from ice-free conditions and runs for 10\,kyr, a~time interval representative of the rapid build-up of the last Cordilleran ice sheet from nearly ice-free to full glacial conditions \citep{clague-1989,stroeven-etal-2010}.
-Our simulations are performed in parallel on 32 cores at the Swedish National Supercomputing Center.
-
-\subsection{Ice thermodynamics}
-
-The central part of an ice sheet model consists of the computation of flow velocity which itself depends on temperature.
-PISM is a~shallow model, which implies that the balance of stresses is approximated based on their predominant components.
-On the other hand, the model is polythermal: it accounts for differences in temperature and softness within the ice column.
-
-The Shallow Shelf Approximation (SSA) is used as a ``sliding law'' for the Shallow Ice Approximation (SIA) \citep{bueler-brown-2009,winkelmann-etal-2011}.
-SIA and SSA velocities are computed by finite difference methods on a~10\,km resolution horizontal grid of 300 by 150 points (the modelling domain).
-Ice softness depends on temperature and water content through an enthalpy formulation \citep{aschwanden-blatter-2009,aschwanden-etal-2012}.
-Enthalpy is computed in three dimensions in up to 51 irregularly spaced layers within the ice, and temperature is additionally computed in 11 regularly spaced layers in bedrock to a~depth of 1\,km.
-Surface air temperature from the climate forcing provides the upper boundary condition to the ice enthalpy model, and a~geothermal heat flux of 70\,\unit{mW\,m^{-2}} provides the lower boundary condition to the bedrock thermal model.
-Although this uniform value does not account for the high spatial geothermal variability in the region, it is on average representative of available heat flow measurements \citep{artemieva-mooney-2001,blackwell-richards-2004}.
-Whereas spatial variations in the geothermal heat flux may significantly affect basal conditions \citep{pattyn-2010}, this study focuses on reconstructing the former extent of the Cordilleran ice sheet, on which it has relatively little effect \citep{rogozhina-etal-2012}.
-
-A~pseudo-plastic sliding law \citep{aschwanden-etal-2013} relates the bed-parallel shear stress and the sliding velocity.
-The yield stress is modelled using the Mohr--Coulomb criterion.
-The till friction angle $\phi$ varies from 10 to 30{\degree}.
-It is computed once at the beginning of the run, as a~function of modern bed elevation, with lowest values occurring at low elevations,
-
-\begin{align}
-  \phi =
-  \begin{cases}
-    10      & \text{for }       z<  0 \\
-    z/10+10 & \text{for }    0 <z<200 \\
-    30      & \text{for }  200 <z     \\
-  \end{cases}
-\end{align}
-
-where $\phi$ is given in degrees and $z$ in meters above current sea level.
-This model thus accounts for a~weakening of till associated with the presence of marine sediments \citep{martin-etal-2011,aschwanden-etal-2013}.
-Because the distribution of marine sediments is assumed constant throughout the run, the till friction angle $\phi$ is not affected by changes in bed elevation.
-Basal topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography and bathymetry dataset with a~resolution of 1\,arc-min \citep{data:etopo1}.
-
-Sea level is lowered by 120\,m and basal topography responds to ice load following a bedrock deformation model that includes point-wise isostasy, elastic lithosphere flexure and viscous mantle deformation in a~semi-infinite half-space \citep{lingle-clark-1985,bueler-etal-2007}.
-It uses a lithosphere density of 3300\,\unit{kg\,m^{-3}}, a flexural rigidity of $5 \times 10^{24}$\,\unit{N\,m} and a mantle viscosity of $1 \times 10^{21}$\,\unit{Pa\,s}.
-Due to the high mantle viscosity, there exists a time lag between ice sheet growth and isostatic bedrock response.
-
-\subsection{Surface mass-balance}
-
-Ice surface accumulation and ablation are computed from monthly mean surface air temperature and monthly precipitation by a~temperature-index (positive degree-day) model \citep{hock-2003}.
-Ice accumulation is equal to precipitation when temperature is below 0\,\unit{{\degree}C}, and decreases to zero linearly with temperature between 0 and 2\,\unit{{\degree}C}.
-Ice ablation is computed from the number of positive degree-days, defined as the integral of temperatures above 0\,\unit{{\degree}C} in one year.
-
-The positive degree-day integral \citep{calov-greve-2005} is numerically approximated using week-long sub-intervals.
-It accounts for temperature variability assuming a~normal distribution along a~central (input) value.
-The temperature standard deviation is a~constant model parameter and was assigned a~value of 3.07\,\unit{{\degree}C}, which corresponds to the mean summer (JJA), model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from North American Regional Reanalysis data \citep{data:narr} in a~manner similar to \citet{seguinot-2013}.
-The ablation model incorporates degree-day factors of 3.04\,\unit{mm\,K^{-1}\,day^{-1}} for snow and 4.59\,\unit{mm\,K^{-1}\,day^{-1}} for ice, as derived from mass-balance measurements on contemporary glaciers from the Coast Mountains and Rocky Mountains in British Columbia \citep{shea-etal-2009}.
-
-\subsection{Air temperature corrections}
-
-Prior to surface mass balance computation, the model dynamically applies a~lapse-rate correction to surface air temperature.
-This correction accounts for the evolution of ice thickness on the one hand, and differences between the climate forcing reference topography and the ice flow model basal topography on the other hand.
-It uses a~reference topography distinct from the modelled basal topography and specific to each climate forcing dataset (Fig.~\ref{fig:topo}).
-An annual lapse rate of 6\,\unit{{\degree}C\,km^{-1}} is used in all simulations.
-No corrections apply to precipitation changes with elevation.
-
-% ----------------------------------------------------------------------
-\section{Climate forcing}
-\label{sec:climate}
-% ----------------------------------------------------------------------
-
-Our climate forcing consists of monthly climatologies of surface air temperature and precipitation obtained from one observational dataset, three global reanalyses and one regional reanalysis.
-
-\subsection{Observational data: WorldClim}
-
-WorldClim is a~high resolution global climate dataset built from meteorological observations \citep{data:worldclim}.
-It was built by applying a~spatial interpolation between a~large set of measurements taken at terrestrial weather stations worldwide, using SRTM \citep{data:srtm} and GTOPO30 elevation data \citep{data:gtopo30}.
-This dataset provides a~resolution much higher than attained by circulation models, however its reliability is variable due to the highly inhomogeneous distribution of measurements \citep{data:worldclim}.
-
-Within our modelling domain, the spatial distributions of summer (JJA) and winter (DJF) air surface temperatures in WorldClim generally show a~decrease of temperature from south to north (JJA) and from south-west to north-east (DJF), such that regions further inland experience colder winters (Fig.~\ref{fig:temp}).
-It should be noted that temperature gradients are much stronger in winter than in summer.
-Temperature rises well above zero during the summer months over the entire modelling domain, except for the highest mountain peaks, and even regions where mean annual temperatures are well below freezing point do experience warm summers.
-In other words, there is a~strong contrast in seasonality between coastal temperate-climate and inland continental-climate regions.
-
-The spatial distribution of JJA and DJF mean precipitation rates in WorldClim reveals a~strong precipitation decline beyond the coastal regions (Fig.~\ref{fig:prec}), primarily as a~result of the continuous orographic barrier formed by the Wrangell and St.-Elias mountains, the Coast Mountains and the Cascades (Fig.~\ref{fig:locmap}).
-A~less pronounced precipitation peak is also visible over the interior ranges.
-From an ice sheet mass balance point of view, these heterogeneities become even stronger due to the difference in timing of the precipitation peak through the year.
-Whereas coastal regions experience most precipitation as snow during the accumulation season (DJF), inland regions experience dry winters and most of the annual precipitation consequently falls as rain during the ablation season (JJA).
-
-In regions such as the northern Yukon Territory and interior Alaska, dry winters and warm summers prohibit ice to accumulate and glaciers to form, despite strongly negative mean annual temperatures.
-In order to account for these strong gradients in seasonality, we use monthly rather than annual averages of temperature and precipitation to drive the ice sheet model.
-
-There are two problems inherent to the use of WorldClim data in our study area.
-Firstly, the density of weather stations used by WorldClim in the Northern American Cordillera is highly inhomogeneous.
-Although good coverage exists for the southern part of our modelling domain, several hundred kilometres can separate neighbouring stations in the north \citep{data:worldclim}.
-Secondly, WorldClim lacks data offshore, which would have been particularly useful over the Pacific continental shelf which was partly covered by ice during the LGM \citep{jackson-clague-1991}.
-
-\subsection{Reanalysis data: NCEP/NCAR, ERA-Interim, CFSR and NARR}
-
-In addition to the WorldClim data, we use surface air temperatures and precipitation rates from three global atmospheric reanalyses and one regional atmospheric reanalysis to inform the mass balance routine of PISM: the NCEP/NCAR reanalysis \citep{data:ncar}, the ERA-Interim reanalysis \citep{data:erai}, the Climate Forecast System Reanalysis \citep[CFSR;][]{data:cfsr}, and the North American Regional Reanalysis \citep[NARR;][]{data:narr}.
-Monthly climatologies from NCEP/NCAR and NARR reanalyses were provided by the \citet{web:psd} whereas monthly climatologies from the ERA-Interim and CFSR reanalyses were computed from their monthly mean time series (Table~\ref{tab:reanalyses}).
-
-The spatial distributions of JJA air surface temperatures and DJF precipitation rates from the four reanalysis climatologies are shown in Figs.~\ref{fig:temp} and~\ref{fig:prec}.
-Summer temperature and winter precipitation are most relevant to the glacier model as they drive summer melt and winter accumulation, respectively.
-Because reanalyses include observational information through data assimilation, they are closer to observations than pure GCMs in densely monitored regions, such as the southern part of our modelling domain \citep{bengtsson-etal-2007}.
-On the other hand, the atmospheric physics embedded in the models provide more realistic output than obtained by spatial interpolation in sparsely monitored regions, such as the northern part of our modelling domain.
-
-\subsection{Preprocessing and lapse-rate corrections}
-
-WorldClim, NCEP/NCAR, ERA-Interim, CFSR and NARR climatologies were re-projected to the Canadian Atlas Lambert conformal conic projection (EPSG code~3978) and bilinearly interpolated to the 10\,km resolution model grid using the module r.proj from GRASS~GIS \citep{soft:grass}.
-In addition, WorldClim data was extrapolated to cover grid points in the ocean using the nearest-neighbour algorithm \citep{soft:scipy}.
-Note that here we choose to present original rather than interpolated data, in order to highlight differences between datasets related to spatial resolution (Figs.~\ref{fig:temp} and~\ref{fig:prec}).
-
-The CFSR climatology presents wave-like artefacts in its precipitation field (Fig.~\ref{fig:prec}).
-These are a~common feature in GCM simulations due to the failure of spectral models to resolve the very local nature of topography-induced precipitation events, which becomes particularly visible in the high-resolution, less spatially-smoothed data.
-Therefore, an alternative forcing was prepared for CFSR by smoothing its precipitation field.
-This was achieved by averaging data locally in a~circular kernel of 7\,pixels in diameter prior to re-projection using the module r.neighbors from GRASS~GIS \citep{soft:grass}.
-Finally, input files for PISM were prepared using \citet{web:nc4py}.
 
@@
 
-Using WorldClim, NCEP/NCAR, ERA-Interim, CFSR (smoothed and original datasets) and NARR input climatologies as climate forcing, we run 66~simulations of glacial inception and growth of the Cordilleran ice sheet over 10\,kyr using temperature offsets ranging from -10 to 0\,\unit{{\degree}C}.
-Significant differences in patterns of ice sheet growth, final glacial extent, and volume are attained (Figs.~\ref{fig:ivolarea}--\ref{fig:cool05}).
-
-Across the range of temperature offsets used, final glaciated area and final ice volume differ widely between different input climatologies (Fig.~\ref{fig:ivolarea}).
-For instance, using a~-5\,\unit{{\degree}C} temperature offset, final glaciated area differs between WorldClim and CFSR forcing by a~factor of~6 and final volume by a~factor of~12 (Fig.~\ref{fig:ivolarea}).
-For particular temperature offset values, given input climatologies result in similar final glaciated areas and ice volumes, yet regional patterns are different (Fig.~\ref{fig:extent}).
-
-Under present climate (0\,\unit{{\degree}C} offset), WorldClim, ERA-Interim and NARR forcing produce local ice caps in areas presently glaciated (Fig.~\ref{fig:extent}).
-These results are fairly consistent with contemporary ice distribution, although both the 10\,\unit{km} horizontal resolution and the occurrence of glaciers and ice caps surface topographies in the ETOPO1 dataset used for basal topography prevent a~meaningful detailed comparison.
-CFSR forcing, on the other hand, yields continuous ice cover over the northern Coast Mountains, and NCEP/NCAR forcing results in ice sheet growth, both of which are largely inconsistent with the present state of glaciation.
-At the opposite side of the temperature offset range, it can be noted that both the NCEP/NCAR and CFSR forcing produce oversized ice sheets whose extents are primarily restricted by the domain boundaries (Fig.~\ref{fig:extent}).
-The smoothing of CFSR precipitation yields an insignificant effect on the resulting ice sheet geometry.
-
-To contrast the effect of the six different input climatologies, we plot simulations using a~-5\,\unit{{\degree}C} temperature offset (Fig.~\ref{fig:cool05}).
-The magnitude of glaciation differs dramatically from one climate forcing to another.
-Whereas the NCEP/NCAR and CFSR climatologies produce a~large ice sheet that covers most of the model domain (Fig.~\ref{fig:cool05}, lower panels), WorldClim, ERA-Interim and NARR climatologies lead to a~more restrictive ice cover (Fig.~\ref{fig:cool05}, upper panels).
-The WorldClim run only resulted in the growth of mountain ice caps bounded to the highest topography.
-
 % ----------------------------------------------------------------------
@@
 % ----------------------------------------------------------------------
-
-These simulations of the LGM Cordilleran ice sheet show a~very large sensitivity to the choice of climate forcing data used (Figs.~\ref{fig:ivolarea}--\ref{fig:cool05}).
-To understand the origin of the discrepancies between model results, we quantify differences in surface air temperature and precipitation rate between the input climatologies and test their effect on the ice sheet model.
-We then compare our findings to a~geomorphological reconstruction of the LGM Cordilleran ice sheet margin, and discuss some of the key assumptions of this study.
-
-\subsection{Comparison of forcing climatologies}
-
-As previously outlined, significant differences in the distribution of winter (DJF) precipitation and summer (JJA) temperature, which are the predominant controls on accumulation and melt at the modelled ice sheet surface, exist between datasets (Figs.~\ref{fig:topo}--\ref{fig:prec}).
-In this section we use the WorldClim climatology, which is derived from observations, as a~reference with which to compare other climate forcing data on land.
-
-Because much of the disparity between reanalyses and observational temperature data is caused by topographical detail at scales unresolved in the reanalyses, we apply a lapse-rate correction prior to this comparison, using the same lapse rate of 6\,\unit{{\degree}C\,km^{-1}} as in the simulations.
-ERA-Interim, CFSR and NARR climatologies exhibit temperature distributions fairly consistent with WorldClim data (Fig.~\ref{fig:tempheatmap}).
-The spatial distribution of temperature differences between reanalyses and WorldClim data shows partly consistent patterns, such as positive anomalies in the northernmost part of the modelling domain.
-These may relate to weaknesses of WorldClim data in areas of sparse observational coverage (Fig.~\ref{fig:tempdiff}).
-However, significant disparity among reanalyses exists, too.
-More specifically, surface air temperature data from the NCEP/NCAR climatology shows the largest disparity with that of WorldClim, with a~significant cold anomaly over most of the modelling domain (Figs.~\ref{fig:tempheatmap} and \ref{fig:tempdiff}).
-
-All four reanalyses used in this study exhibit higher precipitation rates than WorldClim data (Fig.~\ref{fig:precheatmap}).
-However the magnitude of this anomaly differs significantly between datasets.
-The spatial distributions of precipitation differences between reanalyses and WorldClim data show that negative precipitation anomalies are generally constrained to the windward slope of the major mountain ranges, while positive precipitation anomalies are found on the leeward slope of these ranges and extend downwind to interior plateaux and lowlands (Fig.~\ref{fig:precdiff}).
-
-Most likely, this is the signature of an orographic precipitation effect.
-The topography of the Northern American Cordillera is such that its western ranges form a~continuous orographic barrier, causing high precipitation along the Pacific coast while leaving much of the interior arid.
-The ability of a~GCM to reproduce these contrasts in precipitation is bounded by spatial resolution.
-In a~model of coarser resolution, these high mountain ranges are reduced to smooth hills and the modelled distribution of precipitation is also smoother (Fig.~\ref{fig:oroprecip}).
-However, this effect alone cannot explain the widespread and pronounced positive precipitation anomalies observed in CFSR data (Figs.~\ref{fig:precheatmap} and \ref{fig:precdiff}) despite its high spatial resolution (Table~\ref{tab:reanalyses}).
-
-\subsection{Model sensitivity to climate forcing}
-
-To distinguish the effects of temperature and precipitation biases on the ice sheet model, we run a~series of eight additional simulations.
-These new simulations use a ``hybrid'' climate forcing that consists of temperature data from WorldClim, combined with precipitation data from each of the four reanalyses, and vice-versa.
-A~single temperature offset of -5\,\unit{{\degree}C} is used.
-This value was chosen to allow for comparison with previous results (Fig.~\ref{fig:cool05}).
-In each case, temperature lapse rates are computed using the reference topography from the corresponding dataset.
-
-Although only temperature or precipitation changes are applied, this experiment results once more in large differences in modelled ice sheet geometries (Fig.~\ref{fig:biatm}) as compared to the reference WorldClim -5\,\unit{{\degree}C} run (Fig.~\ref{fig:cool05}, top-left panel).
-For ERA-Interim, CFSR and NARR climatologies, precipitation anomalies clearly dominate the differences in ice sheet response, whereas temperature anomalies have relatively little effect (Fig.~\ref{fig:biatmbars}).
-However, in the case of NCAR forcing, both the negative temperature bias (Figs.~\ref{fig:tempheatmap} and \ref{fig:tempdiff}) and the important precipitation differences (Figs.~\ref{fig:precheatmap} and \ref{fig:precdiff}) contribute approximately equally to produce oversized ice sheets (Fig.~\ref{fig:biatmbars}).
-
-\subsection{Comparison to geomorphological LGM margin reconstruction}
-
-We compare modelled ice sheet geometries to a~reconstruction of the ice margin at 14\,\unit{\chem{^{14}C}\,kyr\,BP} (16.8\,cal\,kyr\,BP) by \citet{dyke-2004}, based on glacial geomorphology and radiocarbon dating (Fig.~\ref{fig:locmap}).
-This corresponds to the LGM extent of most of the Cordilleran ice sheet, which occurred later than for the Laurentide ice sheet \citep{porter-swanson-1998,dyke-2004,stroeven-etal-2010,stroeven-etal-inpress}.
-
-Considering previously identified weaknesses of the CFSR and NCEP/NCAR data (Figs.~\ref{fig:tempheatmap}--\ref{fig:biatmbars}), we restricted our analysis to the WorldClim, ERA-Interim and NARR input climatologies.
-In order to compare modelled ice sheets of a~similar size, we select for each climate forcing the simulation that leads to a~final glaciated area closest to the approximate size of the LGM Cordilleran ice sheet of $2\times 10^6\,\unit{km^2}$.
-This corresponds to temperature offset values of -8\,\unit{{\degree}C} (WorldClim), -6\,\unit{{\degree}C} (ERA-Interim) and -7\,\unit{{\degree}C} (NARR).
-These qualitative ``best'' runs are presented in Fig.~\ref{fig:best} along with their associated temperature depressions and a~reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.
-
-Although this selection results in similarly sized ice sheets, noticeable differences in shape exist.
-The ERA-Interim simulation produces a~more northerly-centred ice sheet with too much ice in the north and too little in the south.
-This may partly reflect unresolved orographic precipitation effects (Fig.~\ref{fig:oroprecip}) due to the coarser GCM resolution (Table~\ref{tab:reanalyses}).
-The LGM ice sheet margin by \citet{dyke-2004} is best reproduced by simulations driven by input climatologies from WorldClim and NARR data.
-However, common discrepancies between the modelled ice sheet geometry and the geomorphological reconstruction can still be observed (Fig.~\ref{fig:best}).
-
-Firstly, the modelled eastern margin of the ice sheet extends further east than the inferred position of the junction between the Cordilleran and the Laurentide ice sheets.
-This can be seen for instance in the south-eastern part of our modelling domain, where the modelled ice sheets cover an ice-free corridor of the LGM reconstruction (Fig.~\ref{fig:best}).
-However, such comparisons have been hampered by the fact that we have not considered the buttressing effect of the Laurentide ice sheet, which would have inhibited the Cordilleran ice sheet from advancing onto the Canadian Prairies.
-In addition, potential effects of the growing ice-sheet on atmospheric circulation and precipitation changes are not included in our model.
-The Cordilleran ice sheet initiated from the formation of mountain ice caps over the high mountain ranges \citep{clague-1989}.
-In our simulations, a~continuous ice cover quickly forms over the western ranges, where precipitation rates are higher than in the rest of the domain.
-In reality, this continuous ice cover enhanced the topographical barrier already formed by the western ranges, and would have resulted in less precipitation and warmer air in the interior.
-This process of ``self-inhibiting growth'' is not captured in our model but was demonstrated to be important in a~similar setting over Greenland, where it may have been limiting the advance of the ice sheet during its build-up \citep{langen-etal-2012}.
-
-Secondly, our simulations produce anomalous ice cover on parts of the continental shelf in the Arctic Ocean (Fig.~\ref{fig:best}).
-These regions experience a~marine climate, including lower summer temperature than on the adjacent land (Fig.~\ref{fig:temp}).
-However in our simulations, sea-level is lowered by 120\,\unit{m}, turning large parts of the low-sloping continental shelf into land.
-This anomalous ice cover is to be interpreted as an artefact arising from our simplistic temperature offset method, and has little effect on the rest of the results.
-
-Given the lack of WorldClim data offshore, we consider the NARR climate forcing as best suited for simulations of the Cordilleran ice sheet among those tested, and under the assumptions of the present study.
-
-\subsection{Model sensitivity to the duration of ice sheet inception}
-
-Most of the simulations presented in this study, and more particularly those that reproduce the LGM ice margin reconstruction more closely (Fig.~\ref{fig:best}) do not reach a steady-state.
-Instead, rates of growth remain high throughout the run, and bed elevation does not come close to equilibrium with the ice load.
-In turn, a~key assumption for the present study is our choice of 10\,kyr for the duration of all our simulations.
-Although geomorphological data shows that the period of Cordilleran ice sheet inception, from nearly ice-free to full glacial conditions, could not have been much longer than 10\,kyr, they provide no lower bound to this value.
-
-To test the effect of an inception period of shorter duration on model results, we use simulations run under NARR climate forcing with temperature offsets of -11 to -7\,\unit{{\degree}C}, and compare modelled ice sheet extent when glaciated area reaches the approximate size of the LGM Cordilleran ice sheet of $2\times 10^6\,\unit{km^2}$.
-
-This experiment shows that a~shorter start-up period leads to more restricted ice cover in the eastern part of the modelling domain, where precipitation rate controls ice advance, but more extensive glaciation in the south-western domain along the Pacific coast, where temperature is the limiting factor (Fig.~\ref{fig:durationstack}).
-Hence, reducing simulation duration leads to a~closer match with the LGM ice sheet margin by \citet{dyke-2004} in some sectors of the modelling domain, but higher discrepancy in other sectors.
-
-\subsection{Other potential method improvements}
-
-Our representation of climate history by constant temperature offsets is voluntarily simplistic.
-To better understand the transient character of the LGM Cordilleran ice sheet, time-dependent palaeo-climate forcing would be needed.
-Although it is clear that past temperature changes over the region were neither homogeneous nor constant in time, their patterns are not trivial and may potentially display a~strong inter-dependence with the evolution of the Cordilleran and Laurentide ice sheets.
-As previously discussed, potential effects of the growing ice-sheet on precipitation changes are not included in our model.
-These changes likely consisted of a reduction of precipitation in continental regions and in the ice sheet interior, and an increase of precipitation along part of the margin where the presence of ice imposed ascending winds.
-They could result in a more westerly-centred ice sheet than modelled here, with lower ice thickness in its interior.
-In addition, the final position of the eastern ice margin is largely controlled by its advance rate through the run.
-Therefore, the precipitation shadowing effects may have resulted in more similar ice-sheet configurations if they were included in the model.
-Although using a GCM of intermediate complexity \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013} may represent a first step towards including ice sheet feedback on climate, their spatial resolution do not allow for accurate modelling of orographic precipitation changes in a region as mountainous as the North American Cordillera.
-
-Furthermore, the surface mass balance model crudely simplifies the processes taking place at the glacier surface.
-Refreezing of melted snow and ice and water retention in the snow pack was shown to influence surface mass balance of glaciers \citep{trabant-mayo-1985}, yet it is not accounted for in our PDD model.
-Based on scarce observations over the Greenland ice sheet, different parametrizations of this effect have been developed \citep{janssens-huybrechts-2000,fausto-etal-2009b}.
-However, these parametrizations were shown to disagree with more complex models \citep{reijmer-etal-2012}, and it remains undemonstrated whether their application to a~different region and time period is sound.
-In addition, we simulate temperature variability by using a~constant, uniform value of daily temperature standard deviation.
-This approach implies large biases of surface mass-balance \citep{charbit-etal-2013,rogozhina-rau-2014,seguinot-2013}, particularly over regions with such various climates as experienced within our model domain.
-
-The bedrock deformation model uses homogeneous lithospheric properties and a semi-infinite mantle of constant viscosity.
-Thus, it does not represent lateral variations of crustal and mantle properties characteristic of tectonically active margins such as the American Cordillera, nor does it include the influence from the neighbouring Laurentide ice sheet.
-These limitations should be acknowledged when interpreting ice surface elevation and volume reconstructions presented in this study.
 
@@
 
-Our study shows a~strong dependency of ice sheet model results on the choice of climate forcing data.
-For three of the four reanalysis datasets used, precipitation rate, over surface air temperature, causes much of the variation in modelled ice sheet outlines and volumes.
-Furthermore, the spatial resolution of input data appears critical for providing the ice sheet model with an accurate precipitation field, confirming results obtained by \citet{quiquet-etal-2012} from various GCM forcing datasets over the Greenland ice sheet.
-
-For the Cordilleran ice sheet, we achieve the best fit to the mapped LGM margin by \citet{dyke-2004} using climate forcing from the high-resolution interpolated observational dataset WorldClim \citep{data:worldclim} and the North American Regional Reanalysis \citep[NARR;][]{data:narr}.
-The latter dataset is preferable in our case due to the lack of WorldClim data offshore.
-Climate forcing from CFSR and the NCEP/NCAR reanalysis produce largely oversized ice sheets due to too high precipitation rates, and in the second case, too low surface air temperature.
-The ERA-Interim data used in this study produces more reasonable results, but it may be too coarse to accurately resolve the spatial distribution of orographic precipitation associated with the rugged topography of the study area, resulting in a~misplaced ice coverage in regard to geomorphological data.
-Therefore, we retain NARR data for forcing simulations of the Cordilleran ice sheet in the future.
-
-One must keep in mind, however, that these results are only accurate for our choices of ice-sheet model (PISM), surface mass balance model (PDD), study object (the Cordilleran ice sheet), and, most importantly, palaeo-climate representation (temperature offsets).
-At LGM, it is most likely that the presence of ice sheets significantly affected circulation patterns and distribution of precipitation.
-At the cost of greater computational expense, a~more accurate representation of palaeo-climate may allow for a~better fit between model results and the mapped LGM margin.
-
 % Acknowledgements
-\begin{acknowledgements}
-  We thank Ed Bueler and Andy Aschwanden for providing constant support on PISM and patiently answering our questions on the code internals, Laurent Brodeau and Soon-Heum Ko for assistance concerning software installations on supercomputing resources, and all PISM authors for providing the model open-source.
-This work was supported by the Swedish Research Council (VR) grant {\#}2008-3449 to A.
-P.
-Stroeven and by NASA grant {\#}NNX13AM16G to E.
-Bueler.
-Computer resources were provided by the Swedish National Infrastructure for Computing (SNIC) allocation {\#}2013/1-145 to A.
-P.
-Stroeven at the National Supercomputing Center (NSC).\\
-
+%\begin{acknowledgements}
   % Author contributions
-  \hack{\noindent}\textit{Author contributions.}
-J.
-Seguinot ran the simulations and wrote most of the paper; C.
-Khroulev implemented necessary code changes in PISM; I.
-Rogozhina helped with model set-up and experiment design; A.
-P.
-Stroeven initiated the Cordilleran ice sheet modelling project; and Q.
-Zhang prepared some of the forcing data.
-All authors contributed to paper writing.
-\end{acknowledgements}
+  %\hack{\noindent}\textit{Author contributions.}
+%\end{acknowledgements}
 
 % References
-\input{references.bbl}
-\newpage
+%\input{references.bbl}
+%\newpage
 
@@
 
-% tab:reanalyses
-\begin{table*}[t]
-  \caption{Characteristics of reanalysis climatologies used to inform the mass balance component of PISM.}
-  \label{tab:reanalyses}
-  {\begin{tabular}{lcccc}
-    \tophline
-    Reanalysis& Spatial coverage& Averaging period& Resolution& Description\\
-    \middlehline
-    NCEP/NCAR   & global        & 1981--2010 & 1.875{\degree} & \citet{data:ncar} \\
-    ERA-Interim & global        & 1979--2011 & 1.000{\degree} & \citet{data:erai} \\
-    CFSR        & global        & 1979--2010 & 0.325{\degree} & \citet{data:cfsr} \\
-    NARR        & North America & 1979--2000 & 32\,km       & \citet{data:narr} \\
-    \bottomhline
-  \end{tabular}}
-  \belowtable{}
-\end{table*}
-
 % fig:locmap
-\begin{figure}
-  \includegraphics[width=80mm]{cordillera-climate-locmap}
-  \caption{Shaded relief map of northern North America with the extent of the Cordilleran (CIS), Laurentide (LIS) and Innuitian (IIS) ice sheets at 14\,\unit{\chem{^{14}C}\,kyr\,BP} (16.8\,cal\,kyr\,BP) \citep{dyke-2004}.
-While this age denotes the LIS after retreat from its LGM, it closely corresponds to the LGM extent of most of the Cordilleran ice sheet \citep{porter-swanson-1998,dyke-2004,stroeven-etal-2010}.
-The rectangular box denotes the modelling domain of the Cordilleran ice sheet of 1500 by 3000\,km.
-Major mountain ranges covered by the ice sheet include the Wrangell and St.-Elias mountains (W--SE), the Selwyn and MacKenzie mountains (S--MK), the Coast Mountains and the Rocky Mountains.
-The background map consists of ETOPO1 \citep{data:etopo1} and Natural Earth Data \citep{data:naturalearth} and was assembled with GRASS~GIS \citep{soft:grass}.}
-  \label{fig:locmap}
-\end{figure}
-
-% fig:topo
-\begin{figure*}
-  \includegraphics{cordillera-climate-topo}
-  \caption{Topography maps from WorldClim \citep{data:worldclim}, ERA-Interim reanalysis \citep{data:erai}, North American Regional Reanalysis \citep[NARR;][]{data:narr}, ETOPO1 \citep{data:etopo1}, Climate Forecast System Reanalysis \citep[CFSR;][]{data:cfsr}, and NCEP/NCAR reanalysis \citep{data:ncar}.
-Whereas ETOPO1 is used as basal topography for the ice sheet model, all other data are used as a~reference for temperature lapse-rate corrections.
-Figs.~\ref{fig:topo}--\ref{fig:durationstack} are drawn using Matplotlib \citep{soft:mpl}.}
-  \label{fig:topo}
-\end{figure*}
-
-% fig:temp
-\begin{figure*}
-  \includegraphics{cordillera-climate-temp}
-  \caption{Summer (JJA) and winter (DJF) temperature maps from WorldClim \citep{data:worldclim}, and summer (JJA) temperature maps from ERA-Interim reanalysis \citep{data:erai}, North American Regional Reanalysis \citep[NARR;][]{data:narr}, Climate Forecast System Reanalysis \citep[CFSR;][]{data:cfsr}, and NCEP/NCAR reanalysis \citep{data:ncar} climatologies.}
-  \label{fig:temp}
-\end{figure*}
-
-% fig:prec
-\begin{figure*}
-  \includegraphics{cordillera-climate-prec}
-  \caption{Winter (DJF) and summer (JJA) precipitation maps from WorldClim \citep{data:worldclim}, and winter (DFJ) temperature maps from ERA-Interim reanalysis \citep{data:erai}, North American Regional Reanalysis \citep[NARR;][]{data:narr}, Climate Forecast System Reanalysis \citep[CFSR;][]{data:cfsr}, and NCEP/NCAR reanalysis \citep{data:ncar} climatologies.
-Additional forcing data was prepared to correct for wave-like precipitation artefacts in CFSR (Sect.~\ref{sec:climate}).}
-  \label{fig:prec}
-\end{figure*}
-
-% fig:ivolarea
-\begin{figure}
-  \includegraphics{cordillera-climate-ivolarea}
-  \caption{Total glaciated area and ice volume after 10\,kyr as a~function of temperature offset for each climate forcing used.}
-  \label{fig:ivolarea}
-\end{figure}
-
-% fig:extent
-\begin{figure*}
-  \includegraphics{cordillera-climate-extent}
-  \caption{Extent of ice cover after 10\,kyr as a~function of applied temperature offsets for each climate forcing used.
-A~reconstructed LGM ice sheet margin by \citet{dyke-2004} (Fig.~\ref{fig:locmap}) is given for reference (black line).}
-  \label{fig:extent}
-\end{figure*}
-
-% fig:cool05
-\begin{figure*}
-  \includegraphics{cordillera-climate-cool05}
-  \caption{Ice surface topography (1\,\unit{km} contours) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a~climate -5\,\unit{{\degree}C} colder than present for each climate forcing used.}
-  \label{fig:cool05}
-\end{figure*}
-
-% fig:tempheatmap
-\begin{figure}
-  \includegraphics{cordillera-climate-tempheatmap}
-  \caption{Density maps showing a~comparison of summer (JJA) surface air temperature data from the WorldClim climatology, against that of each reanalysis.
-Climate data are presented after bilinear spatial interpolation and correction for topographic differences to WorldClim data, using a lapse rate of 6\,\unit{{\degree}C\,km^{-1}}.
-Colour mapping is based on a~logarithmic scale.
-Note the cold bias of NCEP/NCAR data relative to WorldClim data.}
-  \label{fig:tempheatmap}
-\end{figure}
-
-% fig:tempdiff
-\begin{figure}
-  \includegraphics{cordillera-climate-tempdiff}
-  \caption{Summer (JJA) surface air temperature difference maps against WorldClim data, after bilinear spatial interpolation and lapse-rate correction.
-Note the cold bias of NCEP/NCAR data and temperature anomalies due to unresolved topographic detail.}
-  \label{fig:tempdiff}
-\end{figure}
-
-% fig:precheatmap
-\begin{figure}
-  \includegraphics{cordillera-climate-precheatmap}
-  \caption{Density maps showing a~comparison of winter (DJF) precipitation rate from the WorldClim climatology, against that of each reanalysis, after bilinear spatial interpolation.
-Colour mapping is based on a~logarithmic scale.
-Note the wet bias of all reanalysis data relative to WorldClim data.}
-  \label{fig:precheatmap}
-\end{figure}
-
-% fig:precdiff
-\begin{figure}
-  \includegraphics{cordillera-climate-precdiff}
-  \caption{Winter (DJF) precipitation rate difference maps against WorldClim data, after bilinear spatial interpolation.
-Note the wet bias of all reanalysis data and the large anomalies of CFSR and NCEP/NCAR data.}
-  \label{fig:precdiff}
-\end{figure}
-
-% fig:oroprecip
-\begin{figure}
-  \includegraphics{cordillera-climate-oroprecip}
-  \caption{Schematic representation of the orographic precipitation effect over a~mountain range.
-In a~GCM of low resolution, the precipitation peak appears downwind-shifted and smoother and the precipitation shadow is less pronounced than in a~high resolution GCM.}
-  \label{fig:oroprecip}
-\end{figure}
-
-% fig:biatm
-\begin{figure*}
-  \includegraphics{cordillera-climate-biatm}
-  \caption{Ice surface topography (1\,\unit{km} contours) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr using ``hybrid'' climate forcing with precipitation rate from WorldClim and surface air temperature from each reanalysis (upper row), and surface air temperature from WorldClim and precipitation rate from each reanalysis (lower row).
-In other words, the upper row shows the effect of temperature anomalies, and the lower row the effect of precipitation anomalies, for each reanalysis, relative to WorldClim data.
-Each simulation uses a~-5\,\unit{{\degree}C} offset for comparison with Fig.~\ref{fig:cool05}.}
-  \label{fig:biatm}
-\end{figure*}
-
-% fig:biatmbars
-\begin{figure}
-  \includegraphics{cordillera-climate-biatmbars}
-  \caption{Effect of temperature and precipitation anomalies (separately and jointly) from each reanalysis on modelled final ice volume relative to the result of the WorldClim -5\,\unit{{\degree}C} offset simulation.
-Corresponding ice sheet geometries are presented in Figs.~\ref{fig:cool05} and~\ref{fig:biatm}.}
-  \label{fig:biatmbars}
-\end{figure}
-
-% fig:best
-\begin{figure*}
-  \includegraphics{cordillera-climate-best}
-  \caption{Ice surface topography (1\,km contours) after 10\,kyr using temperature offsets resulting in glaciated areas of circa $2 \times 10^6\,\unit{km^2}$.
-A~reconstructed LGM ice sheet margin by \citet{dyke-2004} (Fig.~\ref{fig:locmap}) is given for reference (blue line).}
-  \label{fig:best}
-\end{figure*}
-
-% fig:durationstack
-\begin{figure}
-  \includegraphics{cordillera-climate-durationstack}
-  \caption{Left panel: model sensitivity to simulation length.
-Modelled glaciated area, using NARR forcing data, and temperature offsets from -15 to 0\,\unit{{\degree}C}, solid lines corresponding to values from -11 to -7\,\unit{{\degree}C}.
-Right panel: modelled ice margin corresponding to temperature offsets from -11 to -7\,\unit{{\degree}C} when total glaciated area reach the approximate size of the LGM Cordilleran ice sheet of $2 \times 10^6\,\unit{km^2}$.
-A~reconstructed LGM ice sheet margin by \citet{dyke-2004} (Fig.~\ref{fig:locmap}) is given for reference (grey shading).
-Note that shorter (and colder) simulations lead to more restrictive glaciation of the continental eastern margin but further ice extent in the maritime south-western part of the modelling domain.}
-  \label{fig:durationstack}
-\end{figure}
+%\begin{figure}
+%  \includegraphics{cordillera-climate-locmap}
+%  \caption{}
+%  \label{fig:locmap}
+%\end{figure}
 

### 2014-05-15 cordillera-cycle.tex
Add figures into manuscript.
Force 79-char line wrapping for prettier diffs.
Add first abstract.

@@
 
+% My commands
+\def\todo#1{\textcolor[rgb]{0.5,0,0}{\textbf{[TODO: #1]}}}
+\def\aref{\textcolor[rgb]{0.5,0,0}{\textbf{[REF]}}}
+
 % ----------------------------------------------------------------------
@@
 % Title
-\title{Numerical simulation of the Cordilleran ice sheet through the last glacial cycle}
+\title{Numerical simulation of the Cordilleran ice sheet
+       through the last glacial cycle}
 
@@
 % Affiliations
-\affil[1]{Department of Physical Geography and Quaternary Geology and the Bolin Centre for Climate Research, Stockholm University, Stockholm, Sweden}
-\affil[2]{Helmholtz Centre Potsdam, GFZ German Research Centre for Geosciences, Potsdam, Germany}
+\affil[1]{Department of Physical Geography and Quaternary Geology and the
+          Bolin Centre for Climate Research, Stockholm University,
+          Stockholm, Sweden}
+\affil[2]{Helmholtz Centre Potsdam, GFZ German Research Centre for Geosciences,
+          Potsdam, Germany}
 
@@
 \begin{abstract}
+
+  Despite more than a century of geological observations, the Cordilleran ice
+  sheet of North America remains poorly understood in terms of its former
+  extent, volume and dynamics. Although geomorphological evidence is abundant,
+  its complexity is such that whole ice-sheet reconstructions of advance and
+  retreat patterns are lacking. Here we use a numerical ice sheet model
+  calibrated by field-based evidence to attempt a quantitative reconstruction
+  of Cordilleran ice sheet history through the last glacial cycle. Surface mass
+  balance is computed by a positive degree-day model forced by time-dependent
+  temperature offsets from six proxy records located around the globe. Although
+  this approach reveals large variations in model response to atmospheric
+  forcing, all simulations produce two major glaciations events during MIS~4
+  (60\,\unit{kyr}) and MIS~2 (20\,\unit{kyr}). The timing of glaciation is
+  better reproduced using temperature forcing derived from Greenland and
+  Antarctic ice-core than from regional ocean sediment cores. During most of
+  the last glacial cycle, the modelled ice cover is discontinuous and
+  restricted to high mountain areas. However, widespread precipitation over the
+  Skeena mountains favours the persistence of a central ice dome throughout the
+  glacial cycle. It acts as a nucleation centre before last glacial maximum
+  (35\,\unit{kyr}) and hosts the last remain of Cordilleran ice during the
+  early Holocene (8\,\unit{kyr}).
+
+  \todo{Match dates to model results, ev. using a range.}
+
 \end{abstract}
@@
 
+Fig.~\ref{fig:locmap} about here.
+
 % ----------------------------------------------------------------------
@@
 
+Fig.~\ref{fig:atm} about here.
+
 % ----------------------------------------------------------------------
@@
 % ----------------------------------------------------------------------
+
+Figs.~\ref{fig:multirec-timeseries}, ~\ref{fig:multirec-snapshots},
+~\ref{fig:hires-timeseries}, ~\ref{fig:hires-snapshots},
+~\ref{fig:duration} and ~\ref{fig:deglac} about here.
 
@@
 % fig:locmap
-%\begin{figure}
-%  \includegraphics{cordillera-climate-locmap}
-%  \caption{}
-%  \label{fig:locmap}
-%\end{figure}
+\begin{figure}
+  \includegraphics{locmap}
+  \caption{Location map. In construction.}
+  \label{fig:locmap}
+\end{figure}
+
+% fig:atm
+\begin{figure}
+  \includegraphics{atm}
+  \caption{Focring climatology. In construction.}
+  \label{fig:atm}
+\end{figure}
+
+% fig:multirec-timeseries
+\begin{figure}
+  \includegraphics{multirec-timeseries}
+  \caption{Multi-record timeseries. In construction.}
+  \label{fig:multirec-timeseries}
+\end{figure}
+
+% fig:multirec-snapshots
+\begin{figure}
+  \includegraphics{multirec-snapshots}
+  \caption{Multi-record snapshots. In construction.}
+  \label{fig:multirec-snapshots}
+\end{figure}
+
+% fig:hires-timeseries
+\begin{figure}
+  \includegraphics{hires-timeseries}
+  \caption{High resolution timeseries. In construction.}
+  \label{fig:hires-timeseries}
+\end{figure}
+
+% fig:hires-snapshots
+\begin{figure}
+  \includegraphics{hires-snapshots}
+  \caption{High resolution snapshots. In construction.}
+  \label{fig:hires-snapshots}
+\end{figure}
+
+% fig:duration
+\begin{figure}
+  \includegraphics{duration}
+  \caption{Ice cover duration. In construction.}
+  \label{fig:duration}
+\end{figure}
+
+% fig:deglac
+\begin{figure}
+  \includegraphics{deglac}
+  \caption{Deglaciation age. In construction.}
+  \label{fig:deglac}
+\end{figure}
 

### 2014-05-15 cordillera-cycle.tex
Sketch paper outline.

@@
 
+\subsection{Ice thermodynamics}
+
+\subsection{Bedrock response}
+
+\subsection{Surface mass balance}
+
+\subsection{Atmospheric forcing}
+
 Fig.~\ref{fig:atm} about here.
@@
 
-Figs.~\ref{fig:multirec-timeseries}, ~\ref{fig:multirec-snapshots},
-~\ref{fig:hires-timeseries}, ~\ref{fig:hires-snapshots},
-~\ref{fig:duration} and ~\ref{fig:deglac} about here.
+\subsection{Palaeo-climate sensitivity test}
+
+Figs.~\ref{fig:multirec-timeseries} and ~\ref{fig:multirec-snapshots}.
+
+\subsection{High-resolution simulations}
+
+Figs. ~\ref{fig:hires-timeseries} and ~\ref{fig:hires-snapshots},
 
@@
 % ----------------------------------------------------------------------
+
+\subsection{Timing of glaciation}
+
+\subsection{Extent of glaciation}
+
+Fig~\ref{fig:duration} about here.
+
+\subsection{Deglaciation patterns}
+
+Fig ~\ref{fig:deglac} about here.
 

### 2014-05-16 cordillera-cycle.tex
Throw some ideas on figure improvement.
Add Irina's corrections in abstract.

@@
 % My commands
+\def\idea#1{\textcolor[rgb]{0,0.5,0}{\textbf{[IDEA: #1]}}}
 \def\todo#1{\textcolor[rgb]{0.5,0,0}{\textbf{[TODO: #1]}}}
@@
   retreat patterns are lacking. Here we use a numerical ice sheet model
-  calibrated by field-based evidence to attempt a quantitative reconstruction
-  of Cordilleran ice sheet history through the last glacial cycle. Surface mass
-  balance is computed by a positive degree-day model forced by time-dependent
-  temperature offsets from six proxy records located around the globe. Although
-  this approach reveals large variations in model response to atmospheric
+  calibrated against field-based evidence to attempt a quantitative
+  reconstruction of the Cordilleran ice sheet history through the last glacial
+  cycle. A series of simulations is driven by time-dependent temperature
+  offsets from six proxy records located around the globe. Although this
+  approach reveals large variations in model response to evolving atmospheric
   forcing, all simulations produce two major glaciations events during MIS~4
@@
   Skeena mountains favours the persistence of a central ice dome throughout the
-  glacial cycle. It acts as a nucleation centre before last glacial maximum
+  glacial cycle. It acts as a nucleation centre before the last glacial maximum
   (35\,\unit{kyr}) and hosts the last remain of Cordilleran ice during the
@@
   \includegraphics{atm}
-  \caption{Focring climatology. In construction.}
+  \caption{Focring climatology. In construction.
+           \idea{Show seasonal variability}}
   \label{fig:atm}
@@
   \includegraphics{multirec-timeseries}
-  \caption{Multi-record timeseries. In construction.}
+  \caption{Multi-record timeseries. In construction.
+           \todo{Add NGRIP}
+           \idea{Use same colors but different patterns for similar records}}
   \label{fig:multirec-timeseries}
@@
   \includegraphics{multirec-snapshots}
-  \caption{Multi-record snapshots. In construction.}
+  \caption{Multi-record snapshots. In construction.
+           \todo{Add NGRIP}}
   \label{fig:multirec-snapshots}
@@
   \includegraphics{hires-timeseries}
-  \caption{High resolution timeseries. In construction.}
+  \caption{High resolution timeseries. In construction.
+           \todo{Use same scaling at low and high resolution}
+           \idea{Compare to low-resolution}}
   \label{fig:hires-timeseries}
@@
   \includegraphics{hires-snapshots}
-  \caption{High resolution snapshots. In construction.}
+  \caption{High resolution snapshots. In construction.
+           \idea{Fewer snapshots, bigger figure}}
   \label{fig:hires-snapshots}
@@
   \includegraphics{duration}
-  \caption{Ice cover duration. In construction.}
+  \caption{Ice cover duration. In construction.
+           \todo{Add Antarctic-record high-resolution simulation}}
   \label{fig:duration}
@@
   \includegraphics{deglac}
-  \caption{Deglaciation age. In construction.}
+  \caption{Deglaciation age. In construction.
+           \todo{Add Antarctic-record high-resolution simulation}
+           \idea{Compare to Dyke deglaciation outlines}}
   \label{fig:deglac}

### 2014-05-17 cordillera-cycle.tex
Write down the main points.

@@
 
+The Cordilleran ice sheet has a long history of explorations and geological
+investigations. However, the complexity of the evidence is such that no ice-
+sheet scale reconstructions are yet possible. Here we use numerical modelling,
+calibrated with field evidence to attempt such a reconstruction.
+
 Fig.~\ref{fig:locmap} about here.
@@
 
+SSA+SIA, pseudo-plastic sliding, temperature-dependent flow.
+
 \subsection{Bedrock response}
+
+Thermal model and deformation model.
 
@@
 
+Positive degree-day model including spatially and seasonally variable standard
+deviation (Fig.~\ref{fig:atm}).
+
 \subsection{Atmospheric forcing}
 
-Fig.~\ref{fig:atm} about here.
+Input climatology from North American Regional Reanalysis according to
+sensitivity study from the previous paper (Fig.~\ref{fig:atm}).
 
@@
 
-Figs.~\ref{fig:multirec-timeseries} and ~\ref{fig:multirec-snapshots}.
+All six simulations consistently produce two major glaciations during MIS~4 and
+MIS~2 (Figs.~\ref{fig:multirec-timeseries} and~\ref{fig:multirec-snapshots}).
+However the timing is much different. Notably the oceanic core forcing result
+in much too early last glacial maximum.
 
@@
 
-Figs. ~\ref{fig:hires-timeseries} and ~\ref{fig:hires-snapshots},
+We re-run the GRIP (?) and Vostok (?) core simulations at higher resolution
+(Figs.~\ref{fig:hires-timeseries} and ~\ref{fig:hires-snapshots}). Results are
+consistent with lower resolution but differences appear due to small-scale
+topography better capturing mass-balance lapse-rate effects. In general the
+Greenland-based simulation show much more variability.
 
@@
 
+Timing of glaciation is better reproduced by Greenland and Antarctic cores than
+nearby oceanic records. This is because the oceanic cores are affected by
+variations in strength of a local current (the California current) associated
+with the presence of the ice sheet. The timing of the LGM is better reproduced
+by Antarctic than Greenland records.
+
 \subsection{Extent of glaciation}
 
-Fig~\ref{fig:duration} about here.
+During most of the glacial cycle, ice cover is restricted to high mountain
+areas. This is particularly true when using the Greenland-based forcing
+(Fig~\ref{fig:duration}). Most of the glacial maxima margin have a transient
+character, showing that the ice sheet is far out of balance with climate. There
+persist a central ice dome over the Skeena mountains over the entire glacial
+period. This ice dome serves as a nucleation centre at the initiation of the
+advance phase preceding the last glacial maximum. It is here because of more
+widespread precipitation over the Skeena mountains than across other East-West
+transects towards south or north of this region. We interpret this as an
+orographic precipitation effect resulting from a "window" through the sharp
+range of the Coast Mountains at this latitude (Fig?).
 
@@
 
-Fig ~\ref{fig:deglac} about here.
+Deglaciation happens first in the southern half and then in the northern half
+of the ice-covered area (Fig.~\ref{fig:deglac}). Retreat patterns are different
+than reconstructed over central BC, however they match Martin's reconstruction
+of Liard lowland "unzipping" pretty well. In the GRIP-driven simulation, the
+patterns of de-glaciation and Younger Dryas re-advance match with recent
+cosmogenic exposure dates.
 

### 2014-05-20 figures/timeseries.py
Isolate multi-record timeseries plotting script.

@@ -1,0 +1,64 @@
+#!/usr/bin/env python2
+# coding: utf-8
+
+import numpy as np
+from netCDF4 import Dataset
+from matplotlib import pyplot as mplt
+from paperglobals import *
+
+# simulations used
+res = '20km'
+records = ['grip', 'ngrip', 'epica', 'vostok', 'odp1012', 'odp1020']
+
+# initialize time-series figure
+figw, figh = 122.51, 80.01
+fig = mplt.figure(2, (figw*in2mm, figh*in2mm))
+ax1 = fig.add_axes([10/figw, 45/figh, 110/figw, 32.5/figh])
+ax2 = fig.add_axes([10/figw, 10/figh, 110/figw, 32.5/figh])
+mis_idces = np.zeros((len(records), 3), dtype=int)
+mis_times = np.zeros((len(records), 3), dtype=float)
+
+# loop on records[i]
+for i, rec in enumerate(records):
+    print 'plotting %s time series...' % rec
+
+    # get MIS times
+    mis_idces[i], mis_times[i] = get_mis_times(run_path % (res, rec) + '-ts.nc')
+
+    # load forcing time series
+    nc = Dataset(dt_file % rec)
+    dt_time = nc.variables['time'][:]*1e-3
+    dt_temp = nc.variables['delta_T'][:]
+    nc.close()
+
+    # load output time series
+    nc = Dataset(run_path % (res, rec) + '-ts.nc')
+    ts_time = nc.variables['time'][:]*s2ka
+    ts_ivol = nc.variables['ivol'][:]*1e-15
+    nc.close()
+
+    # plot time series
+    ax1.plot(dt_time, dt_temp, color=colors[rec])
+    ax2.plot(ts_time, ts_ivol, color=colors[rec])
+    ax2.plot(mis_times[i], ts_ivol[mis_idces[i]], ls=' ', mew=0.2, ms=4,
+             color=colors[rec], marker=markers[rec], label=labels[rec])
+
+# mark MIS stages
+mistmin = mis_times.min(axis=0)
+mistmax = mis_times.max(axis=0)
+for i in range(3):
+    print 'MIS %i between %.1f and %.1f kyr' % (4-i, -mistmin[i], -mistmax[i])
+    ax2.axvspan(mistmin[i], mistmax[i], color='0.75', lw=0.0)
+
+# set axes properties and save time series
[…]

### 2014-05-20 cordillera-cycle.tex
Isolate multi-record timeseries plotting script.
Remove hi-res snapshots and timeseries.
For the moment I don't see the use of them.

@@
 All six simulations consistently produce two major glaciations during MIS~4 and
-MIS~2 (Figs.~\ref{fig:multirec-timeseries} and~\ref{fig:multirec-snapshots}).
+MIS~2 (Figs.~\ref{fig:timeseries} and~\ref{fig:multirec}).
 However the timing is much different. Notably the oceanic core forcing result
@@
 
-We re-run the GRIP (?) and Vostok (?) core simulations at higher resolution
-(Figs.~\ref{fig:hires-timeseries} and ~\ref{fig:hires-snapshots}). Results are
+We re-run the GRIP (?) and Vostok (?) core simulations at higher resolution.
+Results are
 consistent with lower resolution but differences appear due to small-scale
@@
 \begin{figure}
-  \includegraphics{multirec-timeseries}
+  \includegraphics{timeseries}
   \caption{Multi-record timeseries. In construction.
@@
            \idea{Use same colors but different patterns for similar records}}
-  \label{fig:multirec-timeseries}
+  \label{fig:timeseries}
 \end{figure}
@@
 \begin{figure}
-  \includegraphics{multirec-snapshots}
+  \includegraphics{multirec}
   \caption{Multi-record snapshots. In construction.
            \todo{Add NGRIP}}
-  \label{fig:multirec-snapshots}
-\end{figure}
-
-% fig:hires-timeseries
-\begin{figure}
-  \includegraphics{hires-timeseries}
-  \caption{High resolution timeseries. In construction.
-           \todo{Use same scaling at low and high resolution}
-           \idea{Compare to low-resolution}}
-  \label{fig:hires-timeseries}
-\end{figure}
-
-% fig:hires-snapshots
-\begin{figure}
-  \includegraphics{hires-snapshots}
-  \caption{High resolution snapshots. In construction.
-           \idea{Fewer snapshots, bigger figure}}
-  \label{fig:hires-snapshots}
+  \label{fig:multirec}
 \end{figure}

### 2014-05-20 figures/hires.py

@@ -1,111 +1,0 @@
-#!/usr/bin/env python2
-# coding: utf-8
-
-import numpy as np
-from netCDF4 import Dataset
-from matplotlib import pyplot as mplt
-from mpl_toolkits.axes_grid1.axes_grid import ImageGrid
-from iceplot import plot as iplt
-from paperglobals import *
-
-# simulations used
-res = '20km'
-records = ['grip', 'epica']
-
-# initialize time-series figure
-figw, figh = 122.51, 80.01
-ts_fig = mplt.figure(2, (figw*in2mm, figh*in2mm))
-ts_ax1 = ts_fig.add_axes([10/figw, 45/figh, 110/figw, 32.5/figh])
-ts_ax2 = ts_fig.add_axes([10/figw, 10/figh, 110/figw, 32.5/figh])
-
-# initialize snapshots figure
-rect = [2.5/figw, 2.5/figh, 110/figw, 95/figh]
-sn_fig = iplt.gridfigure((25.0, 50.0), (len(records), 3), axes_pad=2.5*in2mm,
-                         cbar_mode='none', cbar_location='bottom',
-                         cbar_pad=2.5*in2mm, cbar_size=5*in2mm)
-sn_grid = sn_fig.grid
-
-# loop on records
-for i, rec in enumerate(records):
-
-    # load forcing time series
-    print 'reading %s temperature offset tme series...' % rec
-    nc = Dataset(dt_file % rec)
-    dt_time = nc.variables['time'][:]*1e-3
-    dt_temp = nc.variables['delta_T'][:]
-    nc.close()
-
-    # load output time series
-    print 'reading %s ice volume time series...' % rec
-    nc = Dataset(run_path % (res, rec) + '-ts.nc')
-    ts_time = nc.variables['time'][:]*s2ka
-    ts_ivol = nc.variables['ivol'][:]*1e-15
-    nc.close()
-
-    # locate snapshot times using time series
-    snapindexes = [
-        bounded_argmax(ts_ivol, ts_time, -80, -40),  # MIS4
-        bounded_argmin(ts_ivol, ts_time, -60, -20),  # MIS3
-        bounded_argmax(ts_ivol, ts_time, -40, -00)]  # MIS2
-    snaptimes = ts_time[snapindexes]
-
-    # plot time series
-    print 'plotting %s time series...' % rec
-    ts_ax1.plot(dt_time, dt_temp, color=colors[rec])
[…]

### 2014-05-21 figures/multirec.py
Update multi-record snapshots.

@@ -1,66 +1,0 @@
-#!/usr/bin/env python2
-# coding: utf-8
-
-import numpy as np
-from netCDF4 import Dataset
-from matplotlib import pyplot as mplt
-from mpl_toolkits.axes_grid1.axes_grid import ImageGrid
-from iceplot import plot as iplt
-from paperglobals import *
-
-# simulations used
-res = '20km'
-records = ['grip', 'ngrip', 'epica', 'vostok', 'odp1012', 'odp1020']
-
-# initialize snapshots figure
-figw, figh = 122.51, 80.01
-rect = [2.5/figw, 2.5/figh, 110/figw, 95/figh]
-sn_fig = iplt.gridfigure((17.5, 35.0), (3, len(records)), axes_pad=2.5*in2mm,
-                         cbar_mode='none', cbar_location='bottom',
-                         cbar_pad=2.5*in2mm, cbar_size=5*in2mm)
-sn_grid = sn_fig.grid
-
-# get MIS indexes from TS file
-mis_idces = np.zeros((len(records), 3), dtype=int)
-mis_times = np.zeros((len(records), 3), dtype=int)
-
-# loop on records[i]
-for i, rec in enumerate(records):
-    print 'reading %s extra output...' % rec
-
-    # get MIS times
-    mis_idces, mis_times = get_mis_times(run_path % (res, rec) + '-ts.nc')
-
-    # load extra output
-    nc = Dataset(run_path % (res, rec) + '-extra.nc')
-    time = nc.variables['time'][:]*s2ka
-    thk = nc.variables['thk']
-
-    # round snapshot times to nearest slice
-    mis_idces = [(np.abs(time[:]-t)).argmin() for t in mis_times]
-    mis_times = time[mis_idces]
-
-    # plot maps
-    for j, t in enumerate(mis_idces):
-        print 'plotting %s at %s...' % (rec, mis_times[j])
-        ax = sn_grid[i+j*len(records)]
-        mplt.sca(ax)
-        iplt.bedtopoimage(nc, t)
-        iplt.icemargincontour(nc, t)
-        iplt.surftopocontour(nc, t, levels=range(250, 5000, 250),
-                             linewidths=0.2)
-        iplt.surftopocontour(nc, t, levels=range(1000, 5000, 1000))
-        #im = iplt.surfvelimage(nc, t, alpha=0.5)
-        mplt.text(13/15., 28/30., '%s kyr' % (mis_times[j]),
[…]

### 2014-05-21 figures/paperglobals.py

@@ -4,11 +4,13 @@
 import numpy as np
 import brewer2mpl
 from netCDF4 import Dataset
-from matplotlib.colors import LinearSegmentedColormap, Normalize
+from matplotlib.colors import LogNorm, Normalize
+from iceplot.cm import velocity
 
 
 # unit conversion
 in2mm = 1/25.4
+pt2mm = 72*in2mm
 s2ka = 1/(365.0 * 24 * 60 * 60 * 1000)
 
 
@@ -26,12 +28,14 @@ dt = 5.8
 rec = 'grip'
 
 # colors
-bmap = brewer2mpl.get_map('Paired', 'qualitative', 6)
-cols = bmap.mpl_colors
-lightblue, darkblue = cols[0:2]
-lightgreen, darkgreen = cols[2:4]
-lightred, darkred = cols[4:6]
-
+palette = brewer2mpl.get_map('Paired', 'qualitative', 6).mpl_colors
+lightblue, darkblue = palette[0:2]
+lightgreen, darkgreen = palette[2:4]
+lightred, darkred = palette[4:6]
+topo_cmap = 'Greys'
+topo_norm = Normalize(-3000, 6000)
+vel_cmap = velocity
+vel_norm = LogNorm(1e1, 1e4)
 
 # alternative for controlled brightness
 #rbmap = brewer2mpl.get_map('Reds', 'sequential', 9)
@@ -88,6 +92,7 @@ def annotate(ax, s):
                    bbox=dict(ec='k', fc='w', alpha=1.0),
                    transform=ax.transAxes)
 
+
 def draw_boot_topo(grid, res):
     nc = Dataset(boot_file % res)
     x = nc.variables['x']
@@ -98,7 +103,7 @@ def draw_boot_topo(grid, res):
     n = (3*y[0]-y[1])/2
     s = (3*y[-1]-y[-2])/2
     for ax in grid:
-        im = ax.imshow(topg[:].T, cmap='Greys', norm=Normalize(-3000, 6000),
+        im = ax.imshow(topg[:].T, cmap=topo_cmap, norm=topo_norm,
                        extent=(w, e, n, s))
     nc.close()
     return im

### 2014-05-21 figures/snapshots.py

@@ -1,0 +1,78 @@
+#!/usr/bin/env python2
+# coding: utf-8
+
+import numpy as np
+from netCDF4 import Dataset, MFDataset
+from matplotlib import pyplot as plt
+from mpl_toolkits.axes_grid1.axes_grid import ImageGrid
+from iceplot import plot as iplt
+from paperglobals import *
+
+# simulations used
+res = '10km'
+
+# initialize snapshots figure
+figw, figh = 120.0, 102.5
+fig = plt.figure(0, (figw*in2mm, figh*in2mm))
+rect = [5.0/figw, 2.5/figh, 107.5/figw, 95.0/figh]
+grid = ImageGrid(fig, rect, (3, len(records)), axes_pad=2.5*in2mm,
+                 cbar_mode='single', cbar_location='right',
+                 cbar_pad=2.5*in2mm, cbar_size=2.5*in2mm)
+remove_ticks(grid)
+
+# get MIS indexes from TS file
+mis_idces = np.zeros((len(records), 3), dtype=int)
+mis_times = np.zeros((len(records), 3), dtype=int)
+
+# loop on records[i]
+for i, rec in enumerate(records):
+    print 'reading %s extra output...' % rec
+    this_run_path = run_path % (res, rec, offsets[i]*100)
+
+    # get MIS times
+    mis_idces, mis_times = get_mis_times(this_run_path + '-ts.nc')
+
+    # load extra output
+    nc = MFDataset(this_run_path + '-extra*')
+    time = nc.variables['time'][:]*s2ka
+    topg = nc.variables['topg']
+    usurf = nc.variables['uvelsurf']
+    vsurf = nc.variables['vvelsurf']
+    thk = nc.variables['thk']
+
+    # round snapshot times to nearest slice
+    mis_idces = [(np.abs(time[:]-t)).argmin() for t in mis_times]
+    mis_times = time[mis_idces]
+
+    # plot maps
+    for j, t in enumerate(mis_idces):
+        print 'plotting %s at %s...' % (rec, mis_times[j])
+        ax = grid[i+j*len(records)]
+        plt.sca(ax)
+        iplt.bedtopoimage(nc, t, cmap=topo_cmap, norm=topo_norm)
+        iplt.icemargincontour(nc, t)
+        iplt.surftopocontour(nc, t, levels=range(1000, 5000, 1000), lw=0.25)
+        csurf = (usurf[t].T**2 + vsurf[t].T**2)**0.5
[…]

### 2014-05-21 cordillera-cycle.tex
Update multi-record snapshots.
Add intro first shot.

@@
 
-The Cordilleran ice sheet has a long history of explorations and geological
-investigations. However, the complexity of the evidence is such that no ice-
-sheet scale reconstructions are yet possible. Here we use numerical modelling,
-calibrated with field evidence to attempt such a reconstruction.
+During the last glacial cycle, glaciers and ice caps of the North American
+Cordillera have been more extensive than today. At the peak of glaciation, a
+contiguous blanket of ice extended from Denali Peak in the north to the Puget
+Sound in the south. This blanket is known as the former Cordilleran ice sheet.
+
+For more than a century, exploration and geological investigation of the
+Cordillera lead to a large collection of evidence of the former ice cover. This
+evidence consists of mapped boundaries of the former ice extent, direction of
+past sliding velocities and location of former melt-water streams, as well as
+carbon dating and cosmogenic dating of glacier retreat.
+
+Field-based evidence allowed to reconstruct, with a high degree of confidence,
+the maximal extent attained by the Cordilleran ice sheet during the last
+glacial cycle. However, former ice thickness, and in turn the ice sheet's
+contribution to the last glacial maximum sea-level low-stand, remains poorly
+constrained. Moreover, our understanding of the deglaciation phase is
+restricted to a regional level, while little is known about the ice sheet
+history preceding the last glacial maximum extent. Although time-evolving,
+whole ice-sheet reconstructions of glacial advance and retreat patters are
+available for the Laurentide and Eurasian ice sheet, this is not the case for
+the Cordilleran ice sheet.
+
+The present study aims to use a numerical ice sheet model calibrated against
+field-based evidence to attempt a quantitative reconstruction of the
+Cordilleran ice sheet history through the last glacial cycle. Although
+numerical modelling has been identified as a tool to improve our understanding
+of Cordilleran ice sheet more than twenty years ago, the ubiquitously
+mountainous topography of the region has presented a major challenge to its
+application, which is only overcome by recent development of numerical ice
+sheet models and underlying scientific tools. In addition to the topographic
+complexity, strong climatic contrasts characteristic of the North American
+Cordillera require the use of high-resolution temperature and precipitation
+forcing as input data to the ice sheet model.
+
+Here we use the Parallel Ice Sheet Model (PISM), forced by a high-resolution
+climatology from the North American Regional Reanalysis, to model the
+Cordilleran ice sheet through the last glacial cycle. Because past climate
+is mostly unknown, palaeoclimate forcing is applied in the simplified form of
+six time-dependent temperature offset time-series. These time-series were
+obtained by scaling six different palaeo-temperature reconstructions from
+proxy records around the globe, including two \chem{\delta^{18}O} records from
+Greenland ice cores, two \chem{\delta^{18}O} records from Antarctic ice cores,
+and two alkenone unsaturation index records from Northwest Pacific oceanic
+sediment cores. The output from the model is compared to geomorphological
+evidence in terms of timing of glaciation, extent of glaciation and patterns
+of deglaciation.
+
+% Greenland Ice Core Project (GRIP)
+% North Greenland Ice Core Project (NGRIP)
+% European Project for Ice Coring in Antarctica (EPICA)
+% Lake Vostok (Vostok)
+% Ocean Drilling Program (ODP)
 
@@
 All six simulations consistently produce two major glaciations during MIS~4 and
-MIS~2 (Figs.~\ref{fig:timeseries} and~\ref{fig:multirec}).
+MIS~2 (Figs.~\ref{fig:timeseries} and~\ref{fig:snapshots}).
 However the timing is much different. Notably the oceanic core forcing result
@@
 
-% fig:multirec-timeseries
+% fig:timeseries
 \begin{figure}
@@
 
-% fig:multirec-snapshots
+% fig:snapshots
 \begin{figure}
-  \includegraphics{multirec}
+  \includegraphics{snapshots}
   \caption{Multi-record snapshots. In construction.
            \todo{Add NGRIP}}
-  \label{fig:multirec}
+  \label{fig:snapshots}
 \end{figure}

### 2014-05-22 cordillera-cycle.tex
Add precise dates in the abstract.

@@
   forcing, all simulations produce two major glaciations events during MIS~4
-  (60\,\unit{kyr}) and MIS~2 (20\,\unit{kyr}). The timing of glaciation is
+  (61.9--55.4\,\unit{kyr}) and MIS~2 (29.5--16.9\,\unit{kyr}). The timing of
+  glaciation is
   better reproduced using temperature forcing derived from Greenland and
@@
   glacial cycle. It acts as a nucleation centre before the last glacial maximum
-  (35\,\unit{kyr}) and hosts the last remain of Cordilleran ice during the
-  early Holocene (8\,\unit{kyr}).
-
-  \todo{Match dates to model results, ev. using a range.}
+  and hosts the last remain of Cordilleran ice during the
+  early Holocene (10.9--9.5\,\unit{kyr}).
 

### 2014-05-22 cordillera-cycle.tex
Minor corrections in the intro.

@@
 \author[1,2]{J.~Seguinot}
+\author[3]{M.~Margold}
+\author[2]{I.~Rogozhina}
+\author[1]{A.~Stroeven}
 \runningauthor{J.~Seguinot et~al.}
@@
           Potsdam, Germany}
+\affil[3]{Department of Geography, Durham University, UK}
 
@@
 contiguous blanket of ice extended from Denali Peak in the north to the Puget
-Sound in the south. This blanket is known as the former Cordilleran ice sheet.
+Sound in the south. This ice mass is known as the former Cordilleran ice sheet
+(Fig.~\ref{fig:locmap}).
 
@@
 whole ice-sheet reconstructions of glacial advance and retreat patters are
-available for the Laurentide and Eurasian ice sheet, this is not the case for
+available for the Laurentide and Eurasian ice sheet, such is not the case for
 the Cordilleran ice sheet.
@@
 The present study aims to use a numerical ice sheet model calibrated against
-field-based evidence to attempt a quantitative reconstruction of the
+field-based evidence to perform a quantitative reconstruction of the
 Cordilleran ice sheet history through the last glacial cycle. Although
@@
 Here we use the Parallel Ice Sheet Model (PISM), forced by a high-resolution
-climatology from the North American Regional Reanalysis, to model the
-Cordilleran ice sheet through the last glacial cycle. Because past climate
-is mostly unknown, palaeoclimate forcing is applied in the simplified form of
+climatology from the North American Regional Reanalysis (NARR), to model the
+Cordilleran ice sheet evolution through the last glacial cycle. Because past climate
+is mostly unknown, palaeoclimate forcing is simplistically applied as
 six time-dependent temperature offset time-series. These time-series were
@@
 and two alkenone unsaturation index records from Northwest Pacific oceanic
-sediment cores. The output from the model is compared to geomorphological
-evidence in terms of timing of glaciation, extent of glaciation and patterns
-of deglaciation.
+sediment cores. Model output is compared to geomorphological evidence in terms
+of timing of glaciation, extent of glaciation and patterns of deglaciation.
 
@@
 
-Fig.~\ref{fig:locmap} about here.
-
 % ----------------------------------------------------------------------
@@
 
+The simulations presented here were run using Parallel Ice Sheet Model (PISM,
+development version~11b0a7f), an open-source, finite-difference, shallow ice
+sheet model. The model takes basal topography, sea level, geothermal heat flux
+and climate forcing, as input fields and computes the evolution of ice extent
+and thickness in time, its thermal and dynamic state, and
+the associated lithospheric response. Our modelling domain encompasses the
+entire area covered by the Cordilleran ice sheet at the last glacial maximum.
+It includes independent ice growing on the western Alaskan ranges and the
+Brooks Range and intermediate expanses of ice-free terrain in northern Yukon
+Territory and interior Alaska (Fig.~\ref{fig:locmap}).
+
+To reconstruct the successive phases growth and decay of the last Cordilleran
+ice sheet, palaeo-climatic conditions of the last glacial cycle are mimicked
+by applying time-dependent temperature offsets derived from multiple
+palaeo-temperature proxy records. Each simulation starts from assumed ice-free
+conditions at -120\,kyr, and runs to present. These computations were
+performed on 16 to 128 cores at the Swedish National Supercomputing
+Center.
+
 \subsection{Ice thermodynamics}
 
-SSA+SIA, pseudo-plastic sliding, temperature-dependent flow.
+The central part of an ice sheet model consists of the computation of flow
+velocity which itself depends on temperature. PISM is a~shallow model, which
+implies that the balance of stresses is approximated based on their predominant
+components. On the other hand, the model is polythermal: it accounts for
+differences in temperature and softness within the ice column.
 
-\subsection{Bedrock response}
+The Shallow Shelf Approximation (SSA) is used as a ``sliding law'' for the
+Shallow Ice Approximation (SIA) \citep{bueler-brown-2009,winkelmann-etal-2011}.
+SIA and SSA velocities are computed by finite difference methods on a~10\,km
+resolution horizontal grid of 300 by 150 points (the modelling domain). Ice
+softness depends on temperature and water content through an enthalpy
+formulation \citep{aschwanden-blatter-2009,aschwanden-etal-2012}. Enthalpy is
+computed in three dimensions in up to 51 irregularly spaced layers within the
+ice, and temperature is additionally computed in 31 regularly spaced layers in
+bedrock to a~depth of 3\,km. Surface air temperature from the climate forcing
+provides the upper boundary condition to the ice enthalpy model, and
+a~geothermal heat flux of 70\,\unit{mW\,m^{-2}} provides the lower boundary
+condition to the bedrock thermal model. Although this uniform value does not
+account for the high spatial geothermal variability in the region, it is on
+average representative of available heat flow measurements
+\citep{artemieva-mooney-2001,blackwell-richards-2004}. Whereas spatial
+variations in the geothermal heat flux may significantly affect basal
+conditions \citep{pattyn-2010}, this study focuses on reconstructing the former
+extent of the Cordilleran ice sheet, on which it has relatively little effect
+\citep{rogozhina-etal-2012}.
 
-Thermal model and deformation model.
+\todo{Describe the new sliding law?}
 
-\subsection{Surface mass balance}
+A~pseudo-plastic sliding law \citep{aschwanden-etal-2013} relates the
+bed-parallel shear stress and the sliding velocity. The yield stress is
+modelled using the Mohr--Coulomb criterion. The till friction angle $\phi$
+varies from 15 to 45{\degree}. It is computed once at the beginning of the run,
+as a~function of modern bed elevation, with lowest values occurring at low
+elevations,
 
-Positive degree-day model including spatially and seasonally variable standard
-deviation (Fig.~\ref{fig:atm}).
+\begin{align}
+  \phi =
+  \begin{cases}
+    15      & \text{for }       z<  0 \\
+    z/15+15 & \text{for }    0 <z<200 \\
+    45      & \text{for }  200 <z     \\
+  \end{cases}
+\end{align}
+
+where $\phi$ is given in degrees and $z$ in meters above current sea level.
+This model thus accounts for a~weakening of till associated with the presence
+of marine sediments \citep{martin-etal-2011,aschwanden-etal-2013}. Because the
+distribution of marine sediments is assumed constant throughout the run, the
+till friction angle $\phi$ is not affected by changes in bed elevation. Basal
+topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography
+and bathymetry dataset with a~resolution of 1\,arc-min \citep{data:etopo1}.
+
+Sea level is lowered as a function of time according to the SPECMAP
+reconstruction, and basal topography responds to ice load
+following a bedrock deformation model that includes point-wise isostasy,
+elastic lithosphere flexure and viscous mantle deformation in a~semi-infinite
+half-space \citep{lingle-clark-1985,bueler-etal-2007}. It uses a lithosphere
+density of 3300\,\unit{kg\,m^{-3}}, a flexural rigidity of $5 \times
+10^{24}$\,\unit{N\,m} and a mantle viscosity of $1 \times
+10^{21}$\,\unit{Pa\,s}. Due to the high mantle viscosity, there exists a time
+lag between ice sheet growth and isostatic bedrock response.
+
+\subsection{Surface mass-balance}
+
+Ice surface accumulation and ablation are computed from monthly mean surface
+air temperature and monthly precipitation by a~temperature-index (positive
+degree-day) model \citep{hock-2003}. Ice accumulation is equal to precipitation
+when temperature is below 0\,\unit{{\degree}C}, and decreases to zero linearly
+with temperature between 0 and 2\,\unit{{\degree}C}. Ice ablation is computed
+from the number of positive degree-days, defined as the integral of
+temperatures above 0\,\unit{{\degree}C} in one year.
+
+The positive degree-day integral \citep{calov-greve-2005} is numerically
+approximated using week-long sub-intervals. It accounts for temperature
+variability assuming a~normal distribution along a~central (input) value. The
+temperature standard deviation is part of the forcing climatology and was
+computed from daily temperature values from the North American Regional
+Reanalysis \citep{data:narr} and excludes variability associated to the annual
+temperature cycle itself \citet{seguinot-rogozhina-2014}. The
+ablation model incorporates degree-day factors of
+3.04\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for snow and
+4.59\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for ice, as derived from
+mass-balance measurements on contemporary glaciers from the Coast Mountains and
+Rocky Mountains in British Columbia \citep{shea-etal-2009}.
 
@@
 
-Input climatology from North American Regional Reanalysis according to
-sensitivity study from the previous paper (Fig.~\ref{fig:atm}).
+Atmospheric forcing of the model consists of a present-day monthly climatology
+on the one hand, and time-dependent temperature offsets on the other hand. The
+present-day climatology was computed from the near-surface air temperature and
+precipitation rate fields of the North American Regional Reanalysis (NARR) over
+the period xxxx--xxxx.
+
+Prior to surface mass balance computation, the model dynamically applies
+a~lapse-rate correction to surface air temperature. This correction accounts
+for the evolution of ice thickness on the one hand, and differences between the
+climate forcing reference topography and the ice flow model basal topography on
+the other hand. It uses a~reference topography distinct from the modelled basal
+topography and specific to each climate forcing dataset (Fig.~\ref{fig:topo}).
+An annual lapse rate of 6\,\unit{{\degree}C\,km^{-1}} is used in all
+simulations. No corrections apply to precipitation changes with elevation.
 
@@
 
-\subsection{Palaeo-climate sensitivity test}
-
 All six simulations consistently produce two major glaciations during MIS~4 and
-MIS~2 (Figs.~\ref{fig:timeseries} and~\ref{fig:snapshots}).
-However the timing is much different. Notably the oceanic core forcing result
-in much too early last glacial maximum.
-
-\subsection{High-resolution simulations}
-
-We re-run the GRIP (?) and Vostok (?) core simulations at higher resolution.
-Results are
-consistent with lower resolution but differences appear due to small-scale
-topography better capturing mass-balance lapse-rate effects. In general the
-Greenland-based simulation show much more variability.
+MIS~2 (Figs.~\ref{fig:timeseries} and~\ref{fig:snapshots}). However the timing
+is much different. Notably the oceanic core forcing result in much too early
+last glacial maximum. In general the Greenland-based simulation show much more
+variability.
 
@@
 %\input{references.bbl}
-%\newpage
+\newpage
 
@@
 % ----------------------------------------------------------------------
+
+% tab:records
+\begin{table*}[t]
+  \caption{Palaeo-temperature proxy records and scaling parameters used to
+           prepare temperature offset time-series used to force the ice sheet
+           model through the last 120\,\unit{kyr}. $T_{[32;22]}$ refers to the
+           mean temperature anomaly during the period -32 to~-22~\unit{kyr} after
+           scaling.}
+  \label{tab:records}
+  {\begin{tabular}{lcccc}
+    \tophline
+    Record & Proxy & Scaling factor & $T_{[32;22]}$ & Source\\
+    \middlehline
+    GRIP     & \chem{\delta^{18}O} & ?\% & -5.8{\degree}C & \aref \\
+    NGRIP    & \chem{\delta^{18}O} & ?\% & -6.0{\degree}C & \aref \\
+    EPICA    & \chem{\delta^{18}O} & ?\% & -5.6{\degree}C & \aref \\
+    Vostok   & \chem{\delta^{18}O} & ?\% & -5.6{\degree}C & \aref \\
+    ODP~1012 & \chem{U^{K'}_{37}}  & ?\% & -5.8{\degree}C & \aref \\
+    ODP~1020 & \chem{U^{K'}_{37}}  & ?\% & -5.8{\degree}C & \aref \\
+    \bottomhline
+  \end{tabular}}
+  \belowtable{}
+\end{table*}
 
@@
   \includegraphics{atm}
-  \caption{Focring climatology. In construction.
-           \idea{Show seasonal variability}}
+  \caption{Monthly mean near-surface air temperature, precipitation and
+           standard deviation of daily mean temperature for January and July
+           months from the North American Regional Reanalysis (NARR)
+           climatology, used as input fields to the ice sheet model. Note the
+           strong contrasts in seasonality, timing of the precipitation peak,
+           and temperature variability over the model domain, notably between
+           the maritime and continental regions.}
   \label{fig:atm}
@@
   \includegraphics{timeseries}
-  \caption{Multi-record timeseries. In construction.
-           \todo{Add NGRIP}
-           \idea{Use same colors but different patterns for similar records}}
+  \caption{\textbf{(top)} temperature offset time-series from ice and sediment
+           core records used as palaeo-climate forcing for the ice sheet model
+           (Table~\ref{tab:records}). \textbf{(bottom)} Modelled ice volume
+           through the last 120\,\unit{kyr}, expressed in meters of sea-level
+           equivalent. The grey spans indicate the ranges of modelled times of
+           glacial extrema corresponding to MIS~4 (61.9--55.4\,\unit{kyr}), 3
+           (52.2--45.6\,\unit{kyr}), and 2 (last glacial maximum,
+           29.5--16.9\,\unit{kyr}).}
   \label{fig:timeseries}
@@
   \includegraphics{snapshots}
-  \caption{Multi-record snapshots. In construction.
-           \todo{Add NGRIP}}
+  \caption{Snapshots of modelled ice surface topography (1\,\unit{km} contours)
+           and velocity (\unit{m\,yr^{-1}}) at modelled times of glacial
+           extrema, corresponding to MIS~4, 3, and 2.}
   \label{fig:snapshots}
@@
   \includegraphics{duration}
-  \caption{Ice cover duration. In construction.
-           \todo{Add Antarctic-record high-resolution simulation}}
+  \caption{Modelled duration of ice cover during the last 120\,\unit{kyr}.
+           Aside the location of present-day ice caps, note the persistence of
+           a central ice dome over the Skeena mountains during most of the
+           simulation. On the other hand, the maximal extent of the ice sheet
+           corresponds to short durations of ice cover along most of the margin.}
   \label{fig:duration}
@@
   \includegraphics{deglac}
-  \caption{Deglaciation age. In construction.
-           \todo{Add Antarctic-record high-resolution simulation}
-           \idea{Compare to Dyke deglaciation outlines}}
+  \caption{Modelled age of the last deglaciation. Areas where the MIS~4 glacial
+           advance exceeded the last glacial maximum advanced are marked in
+           green. Hatches denote the Younger Dryas re-advance, which is most
+           significant in the GRIP-driven simulation.}
   \label{fig:deglac}

### 2014-05-23 cordillera-cycle.tex
Remove the hi-res stuff for now.
Add preliminary list of authors.
Copy-paste methods from the previous paper.

--- 2014-05-22 cordillera-cycle.tex
@@
 contiguous blanket of ice extended from Denali Peak in the north to the Puget
-Sound in the south. This ice mass is known as the former Cordilleran ice sheet
-(Fig.~\ref{fig:locmap}).
+Sound in the south. This blanket is known as the former Cordilleran ice sheet.
 
@@
 whole ice-sheet reconstructions of glacial advance and retreat patters are
-available for the Laurentide and Eurasian ice sheet, such is not the case for
+available for the Laurentide and Eurasian ice sheet, this is not the case for
 the Cordilleran ice sheet.
@@
 The present study aims to use a numerical ice sheet model calibrated against
-field-based evidence to perform a quantitative reconstruction of the
+field-based evidence to attempt a quantitative reconstruction of the
 Cordilleran ice sheet history through the last glacial cycle. Although
@@
 Here we use the Parallel Ice Sheet Model (PISM), forced by a high-resolution
-climatology from the North American Regional Reanalysis (NARR), to model the
-Cordilleran ice sheet evolution through the last glacial cycle. Because past climate
-is mostly unknown, palaeoclimate forcing is simplistically applied as
+climatology from the North American Regional Reanalysis, to model the
+Cordilleran ice sheet through the last glacial cycle. Because past climate
+is mostly unknown, palaeoclimate forcing is applied in the simplified form of
 six time-dependent temperature offset time-series. These time-series were
@@
 and two alkenone unsaturation index records from Northwest Pacific oceanic
-sediment cores. Model output is compared to geomorphological evidence in terms
-of timing of glaciation, extent of glaciation and patterns of deglaciation.
+sediment cores. The output from the model is compared to geomorphological
+evidence in terms of timing of glaciation, extent of glaciation and patterns
+of deglaciation.
 
@@
 
-The simulations presented here were run using Parallel Ice Sheet Model (PISM,
-development version~11b0a7f), an open-source, finite-difference, shallow ice
-sheet model. The model takes basal topography, sea level, geothermal heat flux
-and climate forcing, as input fields and computes the evolution of ice extent
-and thickness in time, its thermal and dynamic state, and
+Given basal topography, sea level, geothermal heat flux and climate forcing,
+the model computes ice extent and thickness, its thermal and dynamic state, and
 the associated lithospheric response. Our modelling domain encompasses the
-entire area covered by the Cordilleran ice sheet at the last glacial maximum.
-It includes independent ice growing on the western Alaskan ranges and the
-Brooks Range and intermediate expanses of ice-free terrain in northern Yukon
-Territory and interior Alaska (Fig.~\ref{fig:locmap}).
+entire area covered by the Cordilleran ice sheet at LGM including independent
+ice growing on the western Alaskan ranges and the Brooks Range and intermediate
+expanses of ice-free terrain in northern Yukon Territory and interior Alaska
+(Fig.~\ref{fig:locmap}).
 
-To reconstruct the successive phases growth and decay of the last Cordilleran
-ice sheet, palaeo-climatic conditions of the last glacial cycle are mimicked
-by applying time-dependent temperature offsets derived from multiple
-palaeo-temperature proxy records. Each simulation starts from assumed ice-free
-conditions at -120\,kyr, and runs to present. These computations were
-performed on 16 to 128 cores at the Swedish National Supercomputing
+As we aim to model glacial inception and growth of the Cordilleran ice sheet
+towards a~configuration as last attained during the LGM, we mimic
+palaeo-climatic conditions by applying constant temperature offsets
+homogeneously over the modelling domain. Each simulation starts from ice-free
+conditions and runs for 10\,kyr, a~time interval representative of the rapid
+build-up of the last Cordilleran ice sheet from nearly ice-free to full glacial
+conditions \citep{clague-1989,stroeven-etal-2010}. Our simulations are
+performed in parallel on 32 cores at the Swedish National Supercomputing
 Center.
@@
 \subsection{Ice thermodynamics}
+
+%SSA+SIA, pseudo-plastic sliding, temperature-dependent flow.
+%Thermal model and deformation model.
 
@@
 computed in three dimensions in up to 51 irregularly spaced layers within the
-ice, and temperature is additionally computed in 31 regularly spaced layers in
-bedrock to a~depth of 3\,km. Surface air temperature from the climate forcing
+ice, and temperature is additionally computed in 11 regularly spaced layers in
+bedrock to a~depth of 1\,km. Surface air temperature from the climate forcing
 provides the upper boundary condition to the ice enthalpy model, and
@@
 
-\todo{Describe the new sliding law?}
-
 A~pseudo-plastic sliding law \citep{aschwanden-etal-2013} relates the
@@
 modelled using the Mohr--Coulomb criterion. The till friction angle $\phi$
-varies from 15 to 45{\degree}. It is computed once at the beginning of the run,
+varies from 10 to 30{\degree}. It is computed once at the beginning of the run,
 as a~function of modern bed elevation, with lowest values occurring at low
@@
   \begin{cases}
-    15      & \text{for }       z<  0 \\
-    z/15+15 & \text{for }    0 <z<200 \\
-    45      & \text{for }  200 <z     \\
+    10      & \text{for }       z<  0 \\
+    z/10+10 & \text{for }    0 <z<200 \\
+    30      & \text{for }  200 <z     \\
   \end{cases}
@@
 
-Sea level is lowered as a function of time according to the SPECMAP
-reconstruction, and basal topography responds to ice load
+Sea level is lowered by 120\,m and basal topography responds to ice load
 following a bedrock deformation model that includes point-wise isostasy,
@@
 \subsection{Surface mass-balance}
+
+%Positive degree-day model including spatially and seasonally variable standard
+%deviation (Fig.~\ref{fig:atm}).
 
@@
 variability assuming a~normal distribution along a~central (input) value. The
-temperature standard deviation is part of the forcing climatology and was
-computed from daily temperature values from the North American Regional
-Reanalysis \citep{data:narr} and excludes variability associated to the annual
-temperature cycle itself \citet{seguinot-rogozhina-2014}. The
+temperature standard deviation is a~constant model parameter and was assigned
+a~value of 3.07\,\unit{{\degree}C}, which corresponds to the mean summer (JJA),
+model domain-averaged monthly standard deviation of daily mean temperature from
+monthly mean temperature, as computed from North American Regional Reanalysis
+data \citep{data:narr} in a~manner similar to \citet{seguinot-2013}. The
 ablation model incorporates degree-day factors of
@@
 
-Atmospheric forcing of the model consists of a present-day monthly climatology
-on the one hand, and time-dependent temperature offsets on the other hand. The
-present-day climatology was computed from the near-surface air temperature and
-precipitation rate fields of the North American Regional Reanalysis (NARR) over
-the period xxxx--xxxx.
+%Input climatology from North American Regional Reanalysis according to
+%sensitivity study from the previous paper (Fig.~\ref{fig:atm}).
 

### 2014-05-23 cordillera-cycle.tex
Adapt methods to this paper.

--- 2014-05-22 cordillera-cycle.tex
@@
 % Ocean Drilling Program (ODP)
+
+Fig.~\ref{fig:locmap} about here.
 

### 2014-05-23 cordillera-cycle.tex
Corrections from Irina in abstract and intro.

@@
 \def\aref{\textcolor[rgb]{0.5,0,0}{\textbf{[REF]}}}
+\def\citep#1{}
+\def\citet#1{}
 
@@
   approach reveals large variations in model response to evolving atmospheric
-  forcing, all simulations produce two major glaciations events during MIS~4
+  forcing, all simulations produce two major glaciation events during MIS~4
   (61.9--55.4\,\unit{kyr}) and MIS~2 (29.5--16.9\,\unit{kyr}). The timing of
   glaciation is
-  better reproduced using temperature forcing derived from Greenland and
-  Antarctic ice-core than from regional ocean sediment cores. During most of
+  better reproduced using temperature reconstructions from Greenland and
+  Antarctic ice cores than from regional ocean sediment cores. During most of
   the last glacial cycle, the modelled ice cover is discontinuous and
@@
   glacial cycle. It acts as a nucleation centre before the last glacial maximum
-  and hosts the last remain of Cordilleran ice during the
+  and hosts the last remains of Cordilleran ice during the
   early Holocene (10.9--9.5\,\unit{kyr}).
@@
 During the last glacial cycle, glaciers and ice caps of the North American
-Cordillera have been more extensive than today. At the peak of glaciation, a
+Cordillera were more extensive than today. During the last glacial maximum, a
 contiguous blanket of ice extended from Denali Peak in the north to the Puget
-Sound in the south. This blanket is known as the former Cordilleran ice sheet.
+Sound in the south. This ice mass is known as the former Cordilleran ice
+sheet (Fig.~\ref{fig:locmap}).
 
 For more than a century, exploration and geological investigation of the
-Cordillera lead to a large collection of evidence of the former ice cover. This
-evidence consists of mapped boundaries of the former ice extent, direction of
-past sliding velocities and location of former melt-water streams, as well as
-carbon dating and cosmogenic dating of glacier retreat.
+Cordillera have led to a large collection of evidence of the former ice cover. This
+evidence consists of mapped boundaries of the former ice extent, directions of
+past sliding velocities and locations of former melt-water streams, as well as
+the timing of glacier retreat identified from radiocarbon and cosmogenic dating.
 
 Field-based evidence allowed to reconstruct, with a high degree of confidence,
-the maximal extent attained by the Cordilleran ice sheet during the last
-glacial cycle. However, former ice thickness, and in turn the ice sheet's
-contribution to the last glacial maximum sea-level low-stand, remains poorly
-constrained. Moreover, our understanding of the deglaciation phase is
-restricted to a regional level, while little is known about the ice sheet
-history preceding the last glacial maximum extent. Although time-evolving,
-whole ice-sheet reconstructions of glacial advance and retreat patters are
-available for the Laurentide and Eurasian ice sheet, this is not the case for
-the Cordilleran ice sheet.
+the maximal extent attained by the Cordilleran ice sheet during the last glacial
+cycle. However, former ice thickness and the ice sheet's contribution to the
+last glacial maximum sea-level low-stand remain poorly constrained. Moreover,
+our understanding of the glaciation history is mainly restricted to the
+deglaciation phase and a regional level, while little is known about the ice
+sheet evolution prior to the last glacial maximum extent. Although
+time-evolving, whole ice-sheet reconstructions of glacial advance and retreat
+patters are available for the Laurentide and Eurasian ice sheets, such is not
+the case for the Cordilleran ice sheet, where complex patterns of ice flow
+emanating from multiple glaciation centres have been identified.
 
-The present study aims to use a numerical ice sheet model calibrated against
-field-based evidence to attempt a quantitative reconstruction of the
+The present study uses a numerical ice sheet model (PISM) calibrated against
+field-based evidence to perform a quantitative reconstruction of the
 Cordilleran ice sheet history through the last glacial cycle. Although
-numerical modelling has been identified as a tool to improve our understanding
-of Cordilleran ice sheet more than twenty years ago, the ubiquitously
+numerical modelling has been established as a useful tool to improve our
+understanding
+of the Cordilleran ice sheet more than twenty years ago \aref, the ubiquitously
 mountainous topography of the region has presented a major challenge to its
-application, which is only overcome by recent development of numerical ice
-sheet models and underlying scientific tools. In addition to the topographic
+application, which has only been overcome by recent development of numerical ice
+sheet models and underlying scientific computing tools. In addition to the topographic
 complexity, strong climatic contrasts characteristic of the North American
-Cordillera require the use of high-resolution temperature and precipitation
+Cordillera require the use of a high-resolution temperature and precipitation
 forcing as input data to the ice sheet model.
 
-Here we use the Parallel Ice Sheet Model (PISM), forced by a high-resolution
-climatology from the North American Regional Reanalysis, to model the
-Cordilleran ice sheet through the last glacial cycle. Because past climate
-is mostly unknown, palaeoclimate forcing is applied in the simplified form of
-six time-dependent temperature offset time-series. These time-series were
+Because past climate conditions are subject to considerable uncertainty, our
+palaeoclimate forcing is a simplistic approximation of time-evolved temperature
+and precipitation fields derived from a combination of a present-day
+atmospheric reanalysis (NARR), lapse-rate corrections and temperature offset
+time series. The latter are
 obtained by scaling six different palaeo-temperature reconstructions from
@@
 and two alkenone unsaturation index records from Northwest Pacific oceanic
-sediment cores. The output from the model is compared to geomorphological
-evidence in terms of timing of glaciation, extent of glaciation and patterns
-of deglaciation.
+sediment cores. Model output is compared to geomorphological evidence in terms
+of timing and extent of glaciation and patterns of deglaciation.
 
@@
 % Ocean Drilling Program (ODP)
-
-Fig.~\ref{fig:locmap} about here.
 

### 2014-05-23 cordillera-cycle.tex
Develop athosmpheric forcing method.

@@
 
-% Greenland Ice Core Project (GRIP)
-% North Greenland Ice Core Project (NGRIP)
-% European Project for Ice Coring in Antarctica (EPICA)
-% Lake Vostok (Vostok)
-% Ocean Drilling Program (ODP)
-
 % ----------------------------------------------------------------------
@@
 till friction angle $\phi$ is not affected by changes in bed elevation. Basal
-topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography
+topography is derived from the ETOPO1 combined topography
 and bathymetry dataset with a~resolution of 1\,arc-min \citep{data:etopo1}.
@@
 Atmospheric forcing of the model consists of a present-day monthly climatology
-on the one hand, and time-dependent temperature offsets on the other hand. The
+modified by a lapse-rate correction and time-dependent temperature offsets. The
 present-day climatology was computed from the near-surface air temperature and
 precipitation rate fields of the North American Regional Reanalysis (NARR) over
-the period xxxx--xxxx.
+the period xxxx--xxxx (Fig.~\ref{fig:atm}). Modern climate of the North
+American Cordillera is characterised by strong geographic variations of
+seasonality, timing of the precipitation peak and daily temperature variability.
+Our choice of data from the NARR is motivated by the need for an accurate,
+high-resolution precipitation forcing, as outlined in a previous sensitivity
+study.
+
+Temperature offset time-series are reconstructed from proxy records from
+the Greenland Ice Core Project (GRIP), the North Greenland Ice Core Project
+(NGRIP), the European Project for Ice Coring in Antarctica (EPICA), the Lake
+Vostok ice core, and Ocean Drilling Program (ODP) sites 1012 and 1020, both
+located off the Californian shore. Palaeo-temperatures from the GRIP and NGRIP
+records were calculated using a quadratic equation \aref:
+
+\todo{add equation here,}
+
+while temperature reconstructions from the Antarctic and oceanic cores was
+provided by their authors. In addition, all records were scaled linearly in
+order to simulate comparable ice volumes and realistic ice extents at the last
+glacial maximum (Table~\ref{tab:records}, Fig.~\ref{fig:timeseries}).
 
@@
 the other hand. It uses a~reference topography distinct from the modelled basal
-topography and specific to each climate forcing dataset (Fig.~\ref{fig:topo}).
+topography and specific to each climate forcing dataset.
 An annual lapse rate of 6\,\unit{{\degree}C\,km^{-1}} is used in all
-simulations. No corrections apply to precipitation changes with elevation.
+simulations.
 

### 2014-05-23 cordillera-cycle.tex
Develop results section.

@@
 
-All six simulations consistently produce two major glaciations during MIS~4 and
-MIS~2 (Figs.~\ref{fig:timeseries} and~\ref{fig:snapshots}). However the timing
-is much different. Notably the oceanic core forcing result in much too early
-last glacial maximum. In general the Greenland-based simulation show much more
-variability.
+Despite large differences in the input temperature offset time-series, the
+model output presents consistent features that can be observed across the range
+of forcing data used. In all simulations, ice cover is relatively restricted
+during most of the glacial cycle, while two major glaciation events occur
+during MIS~4 (61.9--55.4\,\unit{kyr}) and MIS~2 (29.5--16.9\,\unit{kyr},
+Fig.~\ref{fig:timeseries}). A ice volume minimum is attained in between these
+two events during MIS~3 (52.2--45.6\,\unit{kyr}).
+
+On the other hand, the magitude and timing of these events vary significantly
+between our simulations. For instance, GRIP and NGRIP forcing result in highest
+variability in modelled ice volume, while simulations driven by oceanic records
+and the NGRIP reach last glacial maximum ice volume prior to 20.0\,\unit{kyr}.
+
+Snapshots of model output show relatively consistent patterns of glaciation
+(Fig.~\ref{fig:snapshots}). In all simulations, a central ice cap persists over
+the Skeena mountains between the two glaciation events. However, the size of
+this ice cap, as well as the magnitude of the MIS~4, depends on the temperature
+offset time-series used to drive the model.
 

### 2014-05-23 cordillera-cycle.tex
Add notes for co-authors.

@@
 \def\idea#1{\textcolor[rgb]{0,0.5,0}{\textbf{[IDEA: #1]}}}
+\def\note#1{\textcolor[rgb]{0,0,0.5}{\textbf{[NOTE: #1]}}}
 \def\todo#1{\textcolor[rgb]{0.5,0,0}{\textbf{[TODO: #1]}}}
@@
 
+\note{Please read through the abstract, introduction and figure captions. All
+      other sections are very preliminary at this stage. I need some feedback
+      on the (currently embryonic) discussion part.}
+
 % ----------------------------------------------------------------------
@@
 % ----------------------------------------------------------------------
+
+\note{Currently, this section contains parts from the previous paper.}
 
@@
 
+\note{At this stage, this section is just a collection of thoughts that need to
+be developed into paragraphs. Please comment, criticize, add your own ideas.}
+
 \subsection{Timing of glaciation}
@@
 % ----------------------------------------------------------------------
+
+\note{Nothing here at the moment.}
 
@@
            (52.2--45.6\,\unit{kyr}), and 2 (last glacial maximum,
-           29.5--16.9\,\unit{kyr}).}
+           29.5--16.9\,\unit{kyr}).
+           \note{The dotted line represents an ongoing 5km-resolution
+           simulation, which may be eventually be used in
+           Figs.~\ref{fig:duration} and \ref{fig:deglac}.}}
   \label{fig:timeseries}

### 2014-05-23 cordillera-cycle.tex
Add latex2rtf header.
Add one note about deglaciation plot colours.

@@
 % My commands
-\def\idea#1{\textcolor[rgb]{0,0.5,0}{\textbf{[IDEA: #1]}}}
-\def\note#1{\textcolor[rgb]{0,0,0.5}{\textbf{[NOTE: #1]}}}
-\def\todo#1{\textcolor[rgb]{0.5,0,0}{\textbf{[TODO: #1]}}}
-\def\aref{\textcolor[rgb]{0.5,0,0}{\textbf{[REF]}}}
+\newcommand{\idea}[1]{\textbf{[IDEA: #1]}}
+\newcommand{\note}[1]{\textbf{[NOTE: #1]}}
+\newcommand{\todo}[1]{\textbf{[TODO: #1]}}
+\newcommand{\aref}[0]{\textbf{[REF]}}
 \def\citep#1{}
@@
            simulation. On the other hand, the maximal extent of the ice sheet
-           corresponds to short durations of ice cover along most of the margin.}
+           corresponds to short durations of ice cover along most of the
+           margin.
+           \note{I am trying to improve the readability of low ice cover
+           duration values on this figure.}}
   \label{fig:duration}

### 2014-05-26 cordillera-cycle.tex
Add corrections from Irina.

@@
 % Copernicus manuscript
-\documentclass[tc, ms]{copernicus}
+%\documentclass[tc, ms]{copernicus}
 
@@
 % Copernicus-like latex2rtf compatible
-%\input{copernicus_rtf}
+\input{copernicus_rtf}
 
@@
 our understanding of the glaciation history is mainly restricted to the
-deglaciation phase and a regional level, while little is known about the ice
+deglaciation phase and on regional scale, while little is known about the ice
 sheet evolution prior to the last glacial maximum extent. Although
 time-evolving, whole ice-sheet reconstructions of glacial advance and retreat
-patters are available for the Laurentide and Eurasian ice sheets, such is not
-the case for the Cordilleran ice sheet, where complex patterns of ice flow
+patterns are available for the Laurentide and Eurasian ice sheets, such is not
+the case for the Cordilleran ice sheet, where complex arrangements of ice flow
 emanating from multiple glaciation centres have been identified.
@@
 
-To reconstruct the successive phases growth and decay of the last Cordilleran
+To reconstruct the successive phases of growth and decay of the last Cordilleran
 ice sheet, palaeo-climatic conditions of the last glacial cycle are mimicked
@@
 
-The central part of an ice sheet model consists of the computation of flow
-velocity which itself depends on temperature. PISM is a~shallow model, which
-implies that the balance of stresses is approximated based on their predominant
-components. On the other hand, the model is polythermal: it accounts for
-differences in temperature and softness within the ice column.
-
+PISM is a~shallow model, which implies that the balance of stresses is
+approximated based on their predominant components.
 The Shallow Shelf Approximation (SSA) is used as a ``sliding law'' for the
@@
 computed from daily temperature values from the North American Regional
-Reanalysis \citep{data:narr} and excludes variability associated to the annual
-temperature cycle itself \citet{seguinot-rogozhina-2014}. The
+Reanalysis \citep{data:narr} and excludes variability associated with the annual
+temperature cycle itself \note{improvement} \citet{seguinot-rogozhina-2014}. The
 ablation model incorporates degree-day factors of
@@
 
-Temperature offset time-series are reconstructed from proxy records from
+Temperature offset time-series are derived from proxy records from
 the Greenland Ice Core Project (GRIP), the North Greenland Ice Core Project
@@
 
-while temperature reconstructions from the Antarctic and oceanic cores was
-provided by their authors. In addition, all records were scaled linearly in
+while the temperature reconstructions from the Antarctic and oceanic cores were
+provided by their authors. All records were scaled linearly in
 order to simulate comparable ice volumes and realistic ice extents at the last
@@
 climate forcing reference topography and the ice flow model basal topography on
-the other hand. It uses a~reference topography distinct from the modelled basal
-topography and specific to each climate forcing dataset.
+the other hand.
 An annual lapse rate of 6\,\unit{{\degree}C\,km^{-1}} is used in all
@@
 during MIS~4 (61.9--55.4\,\unit{kyr}) and MIS~2 (29.5--16.9\,\unit{kyr},
-Fig.~\ref{fig:timeseries}). A ice volume minimum is attained in between these
+Fig.~\ref{fig:timeseries}). A local ice minimum is consistently attained in between these
 two events during MIS~3 (52.2--45.6\,\unit{kyr}).
 
-On the other hand, the magitude and timing of these events vary significantly
-between our simulations. For instance, GRIP and NGRIP forcing result in highest
-variability in modelled ice volume, while simulations driven by oceanic records
-and the NGRIP reach last glacial maximum ice volume prior to 20.0\,\unit{kyr}.
+However, the magitude and timing of these events vary significantly
+between our simulations. For instance, the use of GRIP and NGRIP forcing
+results in highest variability in modelled ice volume. Simulations driven by
+the Antarctic and GRIP records reach a last glacial maximum in ice volume at
+19.5--16.9\,\unit{kyr}, while simulations driven by oceanic records
+and the NGRIP record attain it prior to 20.0\,\unit{kyr}.
 
-Snapshots of model output show relatively consistent patterns of glaciation
+Despite the different timing,
+snapshots of model output show relatively consistent patterns of glaciation
 (Fig.~\ref{fig:snapshots}). In all simulations, a central ice cap persists over
 the Skeena mountains between the two glaciation events. However, the size of
-this ice cap, as well as the magnitude of the MIS~4, depends on the temperature
+this ice cap, as well as the magnitude of the MIS~4, depends sensitively on the temperature
 offset time-series used to drive the model.
@@
 
-Timing of glaciation is better reproduced by Greenland and Antarctic cores than
+Timing of glaciation is better reproduced by Greenland and Antarctic records than
 nearby oceanic records. This is because the oceanic cores are affected by
@@
 areas. This is particularly true when using the Greenland-based forcing
-(Fig~\ref{fig:duration}). Most of the glacial maxima margin have a transient
+(Fig~\ref{fig:duration}). Most of the glacial maxima margin has a transient
 character, showing that the ice sheet is far out of balance with climate. There
-persist a central ice dome over the Skeena mountains over the entire glacial
+persists a central ice dome over the Skeena mountains over the entire glacial
 period. This ice dome serves as a nucleation centre at the initiation of the

### 2014-05-27 cordillera-cycle.tex
Improve methods part.

@@
 % Copernicus manuscript
-%\documentclass[tc, ms]{copernicus}
+\documentclass[tc, ms]{copernicus}
 
@@
 % Copernicus-like latex2rtf compatible
-\input{copernicus_rtf}
+%\input{copernicus_rtf}
 
@@
 \newcommand{\todo}[1]{\textbf{[TODO: #1]}}
-\newcommand{\aref}[0]{\textbf{[REF]}}
-\def\citep#1{}
-\def\citet#1{}
+\newcommand{\aref}[0]{\textbf{[ref.]}}
+\renewcommand{\citep}[1]{\aref}
+\renewcommand{\citet}[1]{\aref}
 
@@
 our understanding of the glaciation history is mainly restricted to the
-deglaciation phase and on regional scale, while little is known about the ice
+deglaciation phase and on a regional scale, while little is known about the ice
 sheet evolution prior to the last glacial maximum extent. Although
@@
 the case for the Cordilleran ice sheet, where complex arrangements of ice flow
-emanating from multiple glaciation centres have been identified.
+directions emanating from multiple glaciation centres have been identified.
 
@@
 the associated lithospheric response. Our modelling domain encompasses the
-entire area covered by the Cordilleran ice sheet at the last glacial maximum.
-It includes independent ice growing on the western Alaskan ranges and the
-Brooks Range and intermediate expanses of ice-free terrain in northern Yukon
-Territory and interior Alaska (Fig.~\ref{fig:locmap}).
+entire area covered by the Cordilleran ice sheet at the last glacial maximum
+(Fig.~\ref{fig:locmap}).
 
@@
 formulation \citep{aschwanden-blatter-2009,aschwanden-etal-2012}. Enthalpy is
-computed in three dimensions in up to 51 irregularly spaced layers within the
-ice, and temperature is additionally computed in 31 regularly spaced layers in
-bedrock to a~depth of 3\,km. Surface air temperature from the climate forcing
+computed in three dimensions in up to 51 layers within the
+ice, and temperature is additionally computed in 31 layers in bedrock to
+a~depth of 3\,km. Surface air temperature from the atmospheric forcing
 provides the upper boundary condition to the ice enthalpy model, and
@@
 average representative of available heat flow measurements
-\citep{artemieva-mooney-2001,blackwell-richards-2004}. Whereas spatial
-variations in the geothermal heat flux may significantly affect basal
-conditions \citep{pattyn-2010}, this study focuses on reconstructing the former
-extent of the Cordilleran ice sheet, on which it has relatively little effect
-\citep{rogozhina-etal-2012}.
-
-\todo{Describe the new sliding law?}
+\citep{artemieva-mooney-2001,blackwell-richards-2004}.
 
@@
 modelled using the Mohr--Coulomb criterion. The till friction angle $\phi$
-varies from 15 to 45{\degree}. It is computed once at the beginning of the run,
-as a~function of modern bed elevation, with lowest values occurring at low
-elevations,
+varies from 15 to 45{\degree}. It is taken as a~function of modern bed
+elevation, with lowest values occurring at low elevation, thus accounting
+for a~weakening of till associated with the presence of marine sediments
+\citep{martin-etal-2011,aschwanden-etal-2013}. The till effective pressure is
+related to the modelled amount of water in the till, using a formula derived
+from laboratory experiments with till extracted from an Antarctic ice stream.
+Basal topography is derived from the ETOPO1 combined topography
+and bathymetry dataset with a~resolution of 1\,arc-min \citep{data:etopo1}.
 
-\begin{align}
-  \phi =
-  \begin{cases}
-    15      & \text{for }       z<  0 \\
-    z/15+15 & \text{for }    0 <z<200 \\
-    45      & \text{for }  200 <z     \\
-  \end{cases}
-\end{align}
-
-where $\phi$ is given in degrees and $z$ in meters above current sea level.
-This model thus accounts for a~weakening of till associated with the presence
-of marine sediments \citep{martin-etal-2011,aschwanden-etal-2013}. Because the
-distribution of marine sediments is assumed constant throughout the run, the
-till friction angle $\phi$ is not affected by changes in bed elevation. Basal
-topography is derived from the ETOPO1 combined topography
-and bathymetry dataset with a~resolution of 1\,arc-min \citep{data:etopo1}.
+\note{Do we need more details (equations) on this new sliding law?}
 
@@
 
-Ice surface accumulation and ablation are computed from monthly mean surface
-air temperature and monthly precipitation by a~temperature-index (positive
+Ice surface accumulation and ablation are computed from monthly mean
+near-surface air temperature, monthly precipitation, and monthly standard
+deviation of near-surface temperature by a~temperature-index (positive
 degree-day) model \citep{hock-2003}. Ice accumulation is equal to precipitation
@@
 variability assuming a~normal distribution along a~central (input) value. The
-temperature standard deviation is part of the forcing climatology and was
+temperature standard deviation is part of the forcing climatology. It was
 computed from daily temperature values from the North American Regional
-Reanalysis \citep{data:narr} and excludes variability associated with the annual
-temperature cycle itself \note{improvement} \citet{seguinot-rogozhina-2014}. The
+Reanalysis \citep{data:narr}, excluding variability associated with the
+seasonal cycle itself \citet{seguinot-rogozhina-2014}. The
 ablation model incorporates degree-day factors of
@@
 precipitation rate fields of the North American Regional Reanalysis (NARR) over
-the period xxxx--xxxx (Fig.~\ref{fig:atm}). Modern climate of the North
+the period 1979--2000 (Fig.~\ref{fig:atm}). Modern climate of the North
 American Cordillera is characterised by strong geographic variations of
@@
 records were calculated using a quadratic equation \aref:
-
-\todo{add equation here,}
-
+\begin{equation}
+    \Delta T = -11.88(\chem{\delta^{18}O}-\chem{\delta^{18}O}_0)
+               -0.1925(\chem{\delta^{18}O}^2-\chem{\delta^{18}O}_0^2)
+\end{equation}
 while the temperature reconstructions from the Antarctic and oceanic cores were

### 2014-05-27 cordillera-cycle.tex
Add equations to describe atmospheric forcing.

@@
 Atmospheric forcing of the model consists of a present-day monthly climatology
-modified by a lapse-rate correction and time-dependent temperature offsets. The
-present-day climatology was computed from the near-surface air temperature and
+modified by time-dependent temperature offsets and a lapse-rate correction.
+\begin{align}
+    T_m(t, x, y) &= T_{m0}(x, y) + {\Delta}T_{TS}(t) + {\Delta}T_{LR}(t, x, y) \\
+    P_m(t, x, y) &= P_{m0}(x, y)
+\end{align}
+
+The present-day climatology $\{T_m, P_m\}$ was computed from the near-surface air temperature and
 precipitation rate fields of the North American Regional Reanalysis (NARR) over
@@
 
-Temperature offset time-series are derived from proxy records from
+Temperature offset time-series ${\Delta}T_{TS}$ are derived from proxy records from
 the Greenland Ice Core Project (GRIP), the North Greenland Ice Core Project
@@
 \begin{equation}
-    \Delta T = -11.88(\chem{\delta^{18}O}-\chem{\delta^{18}O}_0)
-               -0.1925(\chem{\delta^{18}O}^2-\chem{\delta^{18}O}_0^2)
+    {\Delta}T_{TS}(t) = -11.88[\chem{\delta^{18}O(t)}-\chem{\delta^{18}O}(0)]
+                        -0.1925[\chem{\delta^{18}O(t)}^2-\chem{\delta^{18}O}(0)^2]
 \end{equation}
@@
 Prior to surface mass balance computation, the model dynamically applies
-a~lapse-rate correction to surface air temperature. This correction accounts
+a~lapse-rate correction to surface air temperature,
+\begin{align}
+    {\Delta}T_{LR}(t, x, y) &= -\gamma [z_{s}(t, x, y)-z_{ref}] \\
+                            &= -\gamma [h(t, x, y)+z_{b}(t, x, y)-z_{ref}]
+\end{align}
+
+This correction accounts
 for the evolution of ice thickness on the one hand, and differences between the
 climate forcing reference topography and the ice flow model basal topography on
-the other hand.
-An annual lapse rate of 6\,\unit{{\degree}C\,km^{-1}} is used in all
-simulations.
+the other hand. An annual lapse rate of
+$\gamma = 6\,\unit{{\degree}C\,km^{-1}}$
+is used in all simulations.
 

### 2014-05-27 cordillera-cycle.tex
Add LGM figure into manuscript.

@@
 
+The modelled last glacial maximum ice sheet possess a central divide located at
+about 3\,500\,\unit{m} elevation along the spine of the Rocky Mountains. The
+distribution of surface velocites shows a clear assymetry between the maritime,
+western flank, and the continental, eastern flank of the ice sheet
+(Fig.~\ref{fig:lgm}).
+
 % ----------------------------------------------------------------------
@@
 
+% fig:lgm
+\begin{figure}
+  \includegraphics{lgm}
+  \caption{Modelled ice surface topography (100\,\unit{m} contours) and
+           velocity (\unit{m\,yr^{-1}}) corresponding to the last glacial
+           volume maximum.}
+  \label{fig:lgm}
+\end{figure}
+
 % fig:duration

### 2014-05-27 cordillera-cycle.tex
Use 6.1 degC offset for NGRIP.

@@
     GRIP     & \chem{\delta^{18}O} & ?\% & -5.8{\degree}C & \aref \\
-    NGRIP    & \chem{\delta^{18}O} & ?\% & -6.0{\degree}C & \aref \\
+    NGRIP    & \chem{\delta^{18}O} & ?\% & -6.1{\degree}C & \aref \\
     EPICA    & \chem{\delta^{18}O} & ?\% & -5.6{\degree}C & \aref \\

### 2014-05-27 figures/paperglobals.py

@@ -54,7 +54,7 @@ records = ['grip', 'ngrip', 'epica', 'vostok', 'odp1012', 'odp1020']
 labels = ['GRIP', 'NGRIP', 'EPICA', 'Vostok', 'ODP 1012', 'ODP 1020']
 colors = [darkblue, lightblue, darkred, lightred, darkgreen, lightgreen]
 markers = ['s', 'D', 'o', 'h', 'v', '^']
-offsets = [5.8, 6.0, 5.6, 5.6, 5.8, 5.8]
+offsets = [5.8, 6.1, 5.6, 5.6, 5.8, 5.8]
 
 
 # analysis functions

### 2014-05-27 cordillera-cycle.tex
Small corrections regarding snapshots.

@@
   \includegraphics{snapshots}
-  \caption{Snapshots of modelled ice surface topography (1\,\unit{km} contours)
-           and velocity (\unit{m\,yr^{-1}}) at modelled times of glacial
+  \caption{Snapshots of modelled ice surface topography (500\,\unit{m}
+           contours) at modelled times of glacial
            extrema, corresponding to MIS~4, 3, and 2.}

### 2014-05-27 figures/snapshots.py

@@ -52,7 +52,7 @@ for i, rec in enumerate(records):
         iplt.icemargincontour(nc, t)
         cs = iplt.surftopocontour(nc, t, levels=levs, cmap='Blues_r',
                                   norm=BoundaryNorm(levs, 256),
-                                  lw=0.25, alpha=0.75)
+                                  linewidths=0.25, alpha=0.75)
         annotate(ax, '%s kyr' % (mis_times[j]))
 
     # close extra file

### 2014-05-30 cordillera-cycle.tex
Add \aref where a reference is needed.

@@
 contiguous blanket of ice extended from Denali Peak in the north to the Puget
-Sound in the south. This ice mass is known as the former Cordilleran ice
+Sound in the south \aref. This ice mass is known as the former Cordilleran ice
 sheet (Fig.~\ref{fig:locmap}).
@@
 For more than a century, exploration and geological investigation of the
-Cordillera have led to a large collection of evidence of the former ice cover. This
-evidence consists of mapped boundaries of the former ice extent, directions of
-past sliding velocities and locations of former melt-water streams, as well as
-the timing of glacier retreat identified from radiocarbon and cosmogenic dating.
+Cordillera have led to a large collection of evidence of the former ice cover \aref. This
+evidence consists of mapped boundaries of the former ice extent \aref, directions of
+past sliding velocities \aref and locations of former melt-water streams \aref, as well as
+the timing of glacier retreat identified from radiocarbon and cosmogenic dating \aref.
 
@@
 the maximal extent attained by the Cordilleran ice sheet during the last glacial
-cycle. However, former ice thickness and the ice sheet's contribution to the
+cycle \aref. However, former ice thickness and the ice sheet's contribution to the
 last glacial maximum sea-level low-stand remain poorly constrained. Moreover,
@@
 deglaciation phase and on a regional scale, while little is known about the ice
-sheet evolution prior to the last glacial maximum extent. Although
+sheet evolution prior to the last glacial maximum extent \aref. Although
 time-evolving, whole ice-sheet reconstructions of glacial advance and retreat
-patterns are available for the Laurentide and Eurasian ice sheets, such is not
+patterns are available for the Laurentide and Eurasian ice sheets \aref, such is not
 the case for the Cordilleran ice sheet, where complex arrangements of ice flow
-directions emanating from multiple glaciation centres have been identified.
+directions emanating from multiple glaciation centres have been identified \aref.
 
-The present study uses a numerical ice sheet model (PISM) calibrated against
+The present study uses a numerical ice sheet model (PISM, \aref) calibrated against
 field-based evidence to perform a quantitative reconstruction of the
@@
 application, which has only been overcome by recent development of numerical ice
-sheet models and underlying scientific computing tools. In addition to the topographic
+sheet models and underlying scientific computing tools \aref. In addition to the topographic
 complexity, strong climatic contrasts characteristic of the North American
 Cordillera require the use of a high-resolution temperature and precipitation
-forcing as input data to the ice sheet model.
+forcing as input data to the ice sheet model \aref.
 
@@
 and precipitation fields derived from a combination of a present-day
-atmospheric reanalysis (NARR), lapse-rate corrections and temperature offset
+atmospheric reanalysis (NARR, \aref), lapse-rate corrections and temperature offset
 time series. The latter are
@@
 proxy records around the globe, including two \chem{\delta^{18}O} records from
-Greenland ice cores, two \chem{\delta^{18}O} records from Antarctic ice cores,
+Greenland ice cores \aref, two \chem{\delta^{18}O} records from Antarctic ice cores \aref,
 and two alkenone unsaturation index records from Northwest Pacific oceanic
-sediment cores. Model output is compared to geomorphological evidence in terms
+sediment cores \aref. Model output is compared to geomorphological evidence in terms
 of timing and extent of glaciation and patterns of deglaciation.
@@
 development version~11b0a7f), an open-source, finite-difference, shallow ice
-sheet model. The model takes basal topography, sea level, geothermal heat flux
+sheet model \aref. The model takes basal topography, sea level, geothermal heat flux
 and climate forcing, as input fields and computes the evolution of ice extent
@@
 while the temperature reconstructions from the Antarctic and oceanic cores were
-provided by their authors. All records were scaled linearly in
+provided by their authors \aref. All records were scaled linearly in
 order to simulate comparable ice volumes and realistic ice extents at the last

### 2014-05-31 cordillera-cycle.tex
Begin to work on the discussion.

@@
 about 3\,500\,\unit{m} elevation along the spine of the Rocky Mountains. The
-distribution of surface velocites shows a clear assymetry between the maritime,
-western flank, and the continental, eastern flank of the ice sheet
+distribution of surface velocities shows a clear asymmetry between the
+maritime, western flank, and the continental, eastern flank of the ice sheet
 (Fig.~\ref{fig:lgm}).
@@
 
-\note{At this stage, this section is just a collection of thoughts that need to
-be developed into paragraphs. Please comment, criticize, add your own ideas.}
+\subsection{Ice-sheet scale glaciation dynamics}
 
-\subsection{Timing of glaciation}
+During the last glacial cycle, geomorphological evidence has allowed to
+identify two major advances of the Cordilleran ice sheet. These correspond to
+the last glacial maximum \aref, and the older xxx glaciation, found in the
+northernmost parts of the ice sheet \aref. It should be noted that all
+simulations presented here reproduce these two major events, which constitutes
+a first-order agreement between model results and geomorphological evidence.
 
-Timing of glaciation is better reproduced by Greenland and Antarctic records than
-nearby oceanic records. This is because the oceanic cores are affected by
-variations in strength of a local current (the California current) associated
-with the presence of the ice sheet. The timing of the LGM is better reproduced
-by Antarctic than Greenland records.
-
-\subsection{Extent of glaciation}
+However, the timing of these events depends highly on the choice of temperature
+time-series used to force the model. For instance, the timing of the last
+glacial maximum is best reproduced by the Antarctic records, while simulations
+driven by the NGRIP and oceanic records pre-date the last glacial maximum by
+several thousands of years. Regarding the oceanic records, the early volume
+maxima can be explain by an early warming in the palaeo-temperature record.
+This early warming has been previously documented, and explained as a
+weakening of the California current, precisely associated with the presence of
+the Cordilleran ice sheet \aref. This precocious glacial maximum illustrates
+the complexity of climate-ice-sheet feedbacks on a regional scale, and shows
+that, although closely located to the modelling domain, the ODP~1012 and
+ODP~1020 proxy records records can't be reliably used as a temperature forcing
+to model the last Cordilleran ice sheet.
 

### 2014-06-02 cordillera-cycle.tex
Add conclusions first shot.

@@
 
-\note{Nothing here at the moment.}
+Numerical simulations of the Cordilleran ice sheet through the last glacial
+cycle presented in this study consistently produce two glaciation maxima during
+MIS~4 (61.9--55.4\,\unit{kyr}) and MIS~2 (29.5--16.9\,\unit{kyr}), independently
+of the choice of palaeo-temperature record used to force the model. This result
+can be seen as a first-order agreement between the model and geomorphological
+evidence, which has documented palaeo-glaciation corresponding to these two
+stages. However, the timing of glaciation peaks highly depends on which record
+is used to drive the model. The timing of the last glacial maximum is best
+reproduced by the Antarctic record, and occurs too early in all simulations
+that were driven by other records. The mismatch is greatest when forcing the
+model with oceanic records from the Pacific Northwest, which are affected by a
+weakening of the California current during the last glacial maximum.
+
+In all simulations presented here, ice cover is limited to disjoint mountain ice
+caps during most of the glacial cycle. This agrees with earlier interpretations
+of the geomorphological evidence preceding last glacial maximum. However, our
+simulations produce a persistent ice cover over the Skeena mountains during
+the entire glacial cycle. During time periods when the Cordilleran ice sheet did
+not exist, the Skeena ice cap appear to be fed by an eastwards precipitation
+intrusion through a topographic window in the Coast Mountains. The ice cap act
+as a nucleation centre at the onset of the last glacial maximum re-advance, and
+may have contributed to the spectacular glacial erosional imprint of the Skeena
+Mountains landscape.
+
+Concerning the deglaciation phase, when most geological evidence is available,
+none of the simulations presented here produce an ideal picture. The timing of
+timing of the last glacial maximum and early deglaciation is generally best
+reproduced by forcing the model with a palaeo-temperature record from
+Antarctica. However, a Younger Dryas re-advance is best reproduced at places
+where it has been documented using temperature forcing from the GRIP record.
+Nonetheless, the patterns of deglaciation are consistent between different
+simulations. They show a rapid deglaciation of the southern half of the ice
+sheet, followed by unzipping of the Liard Lowland, and centripetal retract of
+the ice margin towards the last palaeo-ice cover in the Skeena Mountains
+during the early Holocene (10.9--9.5\,\unit{kyr}).
+
+One must keep in mind, however, that these results are only accurate for our
+choices of ice-sheet model (PISM), surface mass balance model (PDD), study
+object (the Cordilleran ice sheet). Most importantly, our simplistic
+palaeo-climate forcing does not include precipitation corrections in response
+to the presence of the ice sheet, likely leading to overestimates of glacial
+extent and volume. Nevertheless, our results identify the largely understudied
+Skeena Mountains as a key area to understanding glacial dynamics of the
+Cordilleran ice sheet, highlighting the need for further geological
+investigation of this region.
 

### 2014-06-02 cordillera-cycle.tex
Small corrections in preliminary conclusions.

--- 2014-05-31 cordillera-cycle.tex
@@
 
-\note{Nothing here at the moment.}
+Numerical simulations of the Cordilleran ice sheet through the last glacial
+cycle presented in this study consistently produce two glaciation maxima during
+MIS~4 (61.9--55.4\,\unit{kyr}) and MIS~2 (29.5--16.9\,\unit{kyr}), two periods
+corresponding to documented palaeo-glaciation. This result is independent
+of the choice of palaeo-temperature record used for climate forcing, and
+can be seen as a first-order agreement between the model and geomorphological
+evidence. However, the timing of the two glaciation peaks highly depends on
+which record
+is used to drive the model. The timing of the last glacial maximum is best
+reproduced by the Antarctic record, and occurs too early in all simulations
+that were driven by other records. The mismatch is greatest when forcing the
+model with oceanic records from the Pacific Northwest, which are affected by a
+weakening of the California current during the last glacial maximum.
+
+In all simulations presented here, ice cover is limited to disjoint mountain ice
+caps during most of the glacial cycle. This agrees with earlier interpretations
+of the geological evidence preceding the last glacial maximum. However, our
+simulations produce persistent ice cover over the Skeena mountains during
+the entire glacial cycle. At time when the Cordilleran ice sheet is absent
+the Skeena ice cap appears to be fed by an eastwards precipitation
+intrusion through a topographic window in the Coast Mountains. The ice cap acts
+as a nucleation centre at the onset of the last glacial maximum re-advance, and
+may have contributed to the spectacular glacial imprint of the Skeena Mountains
+erosional landscape.
+
+Concerning the deglaciation phase, when most geological evidence is available,
+none of the palaeo-temperature records used produce an ideal agreement between
+model results and geological evidence. The
+timing of the last glacial maximum and early deglaciation is generally best
+reproduced by forcing the model with an Antarctic
+record. However, the Younger Dryas re-advance is best reproduced at places
+where it has been documented using temperature forcing from the GRIP record.
+Nonetheless, the patterns of deglaciation are consistent between different
+simulations. They show a rapid deglaciation of the southern half of the ice
+sheet, followed by unzipping of the Liard Lowland, and concentric retreat of
+the ice margin towards the last palaeo-ice cover in the Skeena Mountains
+during the early Holocene (10.9--9.5\,\unit{kyr}).
+
+One must keep in mind, however, that these results are only accurate for our
+choices of ice-sheet model (PISM), surface mass balance model (PDD), study
+object (the Cordilleran ice sheet). Most importantly, our simplistic
+palaeo-climate forcing does not include precipitation corrections in response
+to the presence of the ice sheet, likely leading to overestimates of glacial
+extent and volume. Nevertheless, our results identify the largely understudied
+Skeena Mountains as a key area to understanding glacial dynamics of the
+Cordilleran ice sheet, highlighting the need for further geological
+investigation in this region.
 

### 2014-06-02 cordillera-cycle.tex
Update discussion outline.
Corrections from Irina in intro and setup.

@@
 During the last glacial cycle, glaciers and ice caps of the North American
-Cordillera were more extensive than today. During the last glacial maximum, a
+Cordillera were more extensive than today. At the last glacial maximum, a
 contiguous blanket of ice extended from Denali Peak in the north to the Puget
@@
 sheet (Fig.~\ref{fig:locmap}).
+\note{Make sure these locations are marked on Fig.~\ref{fig:locmap}.}
 
@@
 evidence consists of mapped boundaries of the former ice extent \aref, directions of
-past sliding velocities \aref and locations of former melt-water streams \aref, as well as
+past sliding velocities {\aref} and locations of former melt-water streams \aref, as well as
 the timing of glacier retreat identified from radiocarbon and cosmogenic dating \aref.
@@
 our understanding of the glaciation history is mainly restricted to the
-deglaciation phase and on a regional scale, while little is known about the ice
+deglaciation phase and a regional scale, while little is known about the ice
 sheet evolution prior to the last glacial maximum extent \aref. Although
@@
 and climate forcing, as input fields and computes the evolution of ice extent
-and thickness in time, its thermal and dynamic state, and
+and thickness in time, the thermal and dynamic state of the ice sheet, and
 the associated lithospheric response. Our modelling domain encompasses the
@@
 
-\note{Do we need more details (equations) on this new sliding law?}
-
 Sea level is lowered as a function of time according to the SPECMAP
@@
 elastic lithosphere flexure and viscous mantle deformation in a~semi-infinite
-half-space \citep{lingle-clark-1985,bueler-etal-2007}. It uses a lithosphere
-density of 3300\,\unit{kg\,m^{-3}}, a flexural rigidity of $5 \times
-10^{24}$\,\unit{N\,m} and a mantle viscosity of $1 \times
-10^{21}$\,\unit{Pa\,s}. Due to the high mantle viscosity, there exists a time
-lag between ice sheet growth and isostatic bedrock response.
+half-space \citep{lingle-clark-1985,bueler-etal-2007}.
 
@@
 Atmospheric forcing of the model consists of a present-day monthly climatology
-modified by time-dependent temperature offsets and a lapse-rate correction.
+$\{T_m0, P_m0\}$, modified by temperature-offset time series ${\Delta}T_{TS}$
+and a lapse-rate correction ${\Delta}T_{LR}$.
 \begin{align}
@@
 
-The present-day climatology $\{T_m, P_m\}$ was computed from the near-surface air temperature and
+The present-day climatology $\{T_m0, P_m0\}$ was computed from the near-surface air temperature and
 precipitation rate fields of the North American Regional Reanalysis (NARR) over
@@
 
-\subsection{Ice-sheet scale glaciation dynamics}
+\subsection{Timing of the glacial maxima}
 
@@
 
+\subsection{Glaciated areas through the last glacial cycle}
+
 During most of the glacial cycle, ice cover is restricted to high mountain
@@
 
-\subsection{Deglaciation patterns}
+\subsection{History of the last deglaciation}
 

### 2014-06-03 cordillera-cycle.tex
Improve results using Irina's suggestions.

@@
 model output presents consistent features that can be observed across the range
-of forcing data used. In all simulations, ice cover is relatively restricted
+of forcing data used. In all simulations, total ice volume is relatively low
 during most of the glacial cycle, while two major glaciation events occur
 during MIS~4 (61.9--55.4\,\unit{kyr}) and MIS~2 (29.5--16.9\,\unit{kyr},
-Fig.~\ref{fig:timeseries}). A local ice minimum is consistently attained in between these
-two events during MIS~3 (52.2--45.6\,\unit{kyr}).
+Fig.~\ref{fig:timeseries}). A local ice minimum is consistently attained in
+between these two events during MIS~3 (52.2--45.6\,\unit{kyr}).
+However, the magnitude and timing of these events vary significantly
+between our simulations.
 
-However, the magitude and timing of these events vary significantly
-between our simulations. For instance, the use of GRIP and NGRIP forcing
-results in highest variability in modelled ice volume. Simulations driven by
-the Antarctic and GRIP records reach a last glacial maximum in ice volume at
-19.5--16.9\,\unit{kyr}, while simulations driven by oceanic records
-and the NGRIP record attain it prior to 20.0\,\unit{kyr}.
+Simulations forced by the Greenland ice core records (GRIP, NGRIP) produce
+highest variability in modelled ice volume throughout the glaciation history.
+Although simulations forced by the Antarctic (EPICA, Vostok) and GRIP records
+reach the last glacial maximum in ice volume at 19.5--16.9\,\unit{kyr}, those
+driven by oceanic (ODP~1012, ODP~1020) and the NGRIP record attain it several
+thousands of years earlier. The NGRIP record is the only temperature forcing
+that predicts a more extensive glaciation during MIS~4 than at the LGM.
 
-Despite the different timing,
-snapshots of model output show relatively consistent patterns of glaciation
-(Fig.~\ref{fig:snapshots}). In all simulations, a central ice cap persists over
-the Skeena mountains between the two glaciation events. However, the size of
-this ice cap, as well as the magnitude of the MIS~4, depends sensitively on the temperature
-offset time-series used to drive the model.
+Simulations forced by oceanic and Antarctic records generally result in lower
+modelled ice volume during MIS~4 and larger ice volume during MIS~3, thus lower
+ice volume variability throughout the simulation length. The ODP~1012
+record-driven simulation predicts rapid deglaciation prior to 20\,\unit{kyr}
+before present due to a local temperature maximum inconsistent with all other
+temperature forcing time series used. The ODP~1012 record-driven simulation
+predicts an early last glacial volume maximum, followed by slower deglaciation
+than modelled using other records. Finally, simulations forced by Antarctic ice
+core records produce a continuous, rapid deglaciation after the last glacial
+maximum, whereas the simulations driven by the GRIP records produce a rapid
+deglaciation in three steps, including a glacial re-advance around the Younger
+Dryas event.
 
-The modelled last glacial maximum ice sheet possess a central divide located at
-about 3\,500\,\unit{m} elevation along the spine of the Rocky Mountains. The
-distribution of surface velocities shows a clear asymmetry between the
-maritime, western flank, and the continental, eastern flank of the ice sheet
-(Fig.~\ref{fig:lgm}).
+Despite the different timing, snapshots of model output show relatively
+consistent patterns of glaciation (Fig.~\ref{fig:snapshots}). As a result of
+the different scaling factors applied to palaeo-temperature proxy records, the
+ice sheet geometry at the last glacial maximum appears very similar from one
+simulation to the next, including a central divide located at about
+3\,500\,\unit{m} elevation along the spine of the Rocky Mountains. In all
+simulations, a central ice cap persists over the Skeena mountains during MIS~3
+between the two glaciation events. However, the size of this ice cap, as well
+as the magnitude of the MIS~4 glaciation, depends sensitively on the choice of
+palaeo-temperature proxy record used to drive the model.
 

### 2014-06-03 cordillera-cycle.tex
Update notes to co-authors.

@@
 
-\note{Please read through the abstract, introduction and figure captions. All
-      other sections are very preliminary at this stage. I need some feedback
-      on the (currently embryonic) discussion part.}
+\begin{center}
+\begin{tabular}{>{\bfseries}ll}
+    \tophline
+    Introduction & Happy with it except for missing refs.\\
+    Model setup  & Needs minor fixes to avoid repeating the previous paper.\\
+    Results      & Almost happy with it. \\
+    Discussion   & Large gaps still have to be filled.\\
+    Conclusions  & Main contents are here but the text needs to be improved.\\
+    Figures      & I have plans for a couple extra ones.\\
+    \bottomhline
+\end{tabular}
+\end{center}
 
@@
 
-\note{Currently, this section contains parts from the previous paper.}
+\todo{Small adjustments to avoid repetitions with the previous paper.}
 
@@
 
+\todo{Develop the three subsections in order to reach the conclusions.}
+
 \subsection{Timing of the glacial maxima}
@@
 % ----------------------------------------------------------------------
+
+\todo{Improve the style.}
 

### 2014-06-04 cordillera-cycle.tex
Enhance ice cover duration plot.

@@
   \caption{Modelled duration of ice cover during the last 120\,\unit{kyr}.
-           Aside the location of present-day ice caps, note the persistence of
-           a central ice dome over the Skeena mountains during most of the
+           A contiguous ice cover spanning from the Alaska Range (AR) to the
+           southern Coast Mountians (CM) and Rocky Mountains (RM) exists for
+           about 28\,\unit{kyr} in the GRIP simulation and about 29\,\unit{kyr}
+           in the EPICA simulation. Note the irregular colour scale. A central
+           ice cover persists over the Skeena Mountains (SM) during most of the
            simulation. On the other hand, the maximal extent of the ice sheet
-           corresponds to short durations of ice cover along most of the
-           margin.
-           \note{I am trying to improve the readability of low ice cover
-           duration values on this figure.}}
+           generally corresponds to relatively short durations of ice cover.
   \label{fig:duration}

### 2014-06-04 figures/duration.py

@@ -4,7 +4,7 @@
 import numpy as np
 from netCDF4 import Dataset
 from matplotlib import pyplot as plt
-from matplotlib.colors import Normalize
+from matplotlib.colors import BoundaryNorm
 from iceplot import plot as iplt
 from paperglobals import *
 
@@ -12,6 +12,7 @@ from paperglobals import *
 res = '10km'
 records = ['grip', 'epica']
 offsets = [5.8, 5.6]
+cislevs = [28.0, 29.0]
 
 # initialize figure
 fig = iplt.gridfigure((47.5, 95.0), (1, len(records)), axes_pad=2.5*in2mm,
@@ -34,15 +35,24 @@ for i, rec in enumerate(records):
     icecover = (mask[:] == 2).sum(axis=0).T
     icecover *= 120.0/len(nc.variables['time'])
 
-    # plot
-    levs = range(0,121,20)
+    # set contour levels, colors and hatches
+    levs = range(0, 21, 5) + [cislevs[i]] + range(40,121,20)
     levs[0] = 1e-6
-    cf = ax.contourf(x[:], y[:], icecover, levels=levs, cmap='Greens', alpha=0.75)
-    ax.contour(x[:], y[:], icecover, [levs[0]], colors='k')
+    cmap = plt.get_cmap('RdBu')
+    colors = cmap(np.hstack((np.linspace(0.0, 0.5, (len(levs)-1)/2),
+                             np.linspace(0.5, 1.0, (len(levs)-1)/2))))
+    hatches = ['']*5 + ['//'] + ['']*4
+
+    # plot
+    cf = ax.contourf(x[:], y[:], icecover, levels=levs, alpha=0.75,
+                     colors=colors, hatches=hatches)
+    cs = ax.contour(x[:], y[:], icecover, [cislevs[i]], colors='k', linewidths=0.25)
+    cs.clabel(fontsize=4, fmt='%i kyr', manual=[(-1825e3, 1000e3)])
+    ax.contour(x[:], y[:], icecover, [levs[0]], colors='k', linewidths=0.5)
 
     # close extra file
-    annotate(ax, rec.upper())
     nc.close()
+    annotate(ax, rec.upper())
 
 # add colorbar and save
 cb = fig.colorbar(cf, ax.cax)

### 2014-06-05 cordillera-cycle.tex
Minor corrections in model setup.

@@
 development version~11b0a7f), an open-source, finite-difference, shallow ice
-sheet model \aref. The model takes basal topography, sea level, geothermal heat flux
-and climate forcing, as input fields and computes the evolution of ice extent
+sheet model \aref. The model inputs basal topography, sea level, geothermal
+heat flux and climate forcing, and computes the evolution of ice extent
 and thickness in time, the thermal and dynamic state of the ice sheet, and
@@
 provided by their authors \aref. All records were scaled linearly in
-order to simulate comparable ice volumes and realistic ice extents at the last
+order to simulate realistic and comparable ice extents at the last
 glacial maximum (Table~\ref{tab:records}, Fig.~\ref{fig:timeseries}).
@@
 
-This correction accounts
-for the evolution of ice thickness on the one hand, and differences between the
-climate forcing reference topography and the ice flow model basal topography on
-the other hand. An annual lapse rate of
-$\gamma = 6\,\unit{{\degree}C\,km^{-1}}$
-is used in all simulations.
+This correction accounts for the evolution of ice thickness $h$ on the one
+hand, and for differences between the climate forcing reference topography and
+the ice flow model basal topography $z_{b}-z_{ref}$ on the other hand. An
+annual lapse rate of $\gamma = 6\,\unit{{\degree}C\,km^{-1}}$ is used in all
+simulations.
 

### 2014-06-05 cordillera-cycle.tex
Correct small bugs.
Update figures and captions in manuscript.

@@
   \includegraphics{locmap}
-  \caption{Location map. In construction.}
+  \caption{Relief map of northern North America showing a reconstruction of the
+           areas once covered by the Cordilleran (CIS), Laurentide (LIS),
+           Innuitian (IIS) and Greenland (GIS) ice sheets during the last
+           18\,\unit{\chem{^{14}C}\,kyr\,BP} (21.4\,cal\,kyr\,BP)
+           \citep{dyke-2004}. The rectangular box denotes the location of the
+           modelling domain used in this study. Major mountain ranges covered
+           by the ice sheet include the Alaska Range (AR), the Wrangell and
+           St.-Elias mountains (WSE), the Selwyn and MacKenzie mountains (SMK),
+           the Coast Mountains (CM) and the Rocky Mountains (RM). The background
+           map consists of ETOPO1 \citep{data:etopo1} and Natural Earth Data
+           \citep{data:naturalearth}.
+           \todo{Mark palaeo-ice sheets and mountain ranges on the map.}}
   \label{fig:locmap}
@@
            months from the North American Regional Reanalysis (NARR)
-           climatology, used as input fields to the ice sheet model. Note the
+           climatology, used to force the ice sheet model. Note the
            strong contrasts in seasonality, timing of the precipitation peak,
@@
   \includegraphics{timeseries}
-  \caption{\textbf{(top)} temperature offset time-series from ice and sediment
-           core records used as palaeo-climate forcing for the ice sheet model
-           (Table~\ref{tab:records}). \textbf{(bottom)} Modelled ice volume
+  \caption{Temperature offset time-series from ice core and sediment core
+           records (Table~\ref{tab:records}) used as palaeo-climate forcing for
+           the ice sheet model \textbf{(top)}, and modelled ice volume
            through the last 120\,\unit{kyr}, expressed in meters of sea-level
-           equivalent. The grey spans indicate the ranges of modelled times of
-           glacial extrema corresponding to MIS~4 (61.9--55.4\,\unit{kyr}), 3
-           (52.2--45.6\,\unit{kyr}), and 2 (last glacial maximum,
-           29.5--16.9\,\unit{kyr}).
-           \note{The dotted line represents an ongoing 5km-resolution
-           simulation, which may be eventually be used in
-           Figs.~\ref{fig:duration} and \ref{fig:deglac}.}}
+           equivalent \textbf{(bottom)}. Gray spans indicate Marine Isotope
+           Stages (MIS) according to a global compilation of benthic
+           \chem{\delta^{18}O} records \citep{lisiecki-raymo-2005}. Hatched
+           rectangles indicate modelled ice volume extrema corresponding to
+           MIS~4 (61.9--55.4\,\unit{kyr}), MIS~3 (52.2--45.6\,\unit{kyr}), and
+           MIS~2 (last glacial maximum, 29.5--16.9\,\unit{kyr}).}
   \label{fig:timeseries}
@@
   \includegraphics{snapshots}
-  \caption{Snapshots of modelled ice surface topography (500\,\unit{m}
-           contours) at modelled times of glacial
-           extrema, corresponding to MIS~4, 3, and 2.}
+  \caption{Snapshots of modelled surface topography (500\,\unit{m} contours)
+           corresponding to the ice volume extrema indicated on
+           Fig.~\ref{fig:timeseries}. Note the occurence of spatial similarities
+           despite large differences in timing.}
   \label{fig:snapshots}
-\end{figure}
-
-% fig:lgm
-\begin{figure}
-  \includegraphics{lgm}
-  \caption{Modelled ice surface topography (100\,\unit{m} contours) and
-           velocity (\unit{m\,yr^{-1}}) corresponding to the last glacial
-           volume maximum.}
-  \label{fig:lgm}
 \end{figure}
@@
            simulation. On the other hand, the maximal extent of the ice sheet
-           generally corresponds to relatively short durations of ice cover.
+           generally corresponds to relatively short durations of ice cover.}
   \label{fig:duration}
+\end{figure}
+
+% fig:warmbase
+\begin{figure}
+  \includegraphics{warmbase}
+  \caption{Modelled duration of warm-based ice cover during the last
+           120\,\unit{kyr}. Long ice cover durations combined with basal
+           temperatures at the melting point may explain the strong glacial
+           erosional imprint of the Skeena Mountains (SM) landscape. Hatches
+           indicate areas that were covered by cold ice only.
+           \note{indicate location of Skeena Mountains.}}
+  \label{fig:warmbase}
+\end{figure}
+
+% fig:erosion
+\begin{figure}
+  \includegraphics{erosion}
+  \caption{Modelled cumulative basal displacement (integrand of basal velocity)
+           over the last 120\,\unit{kyr}.
+           \note{Even after carefully adjusting the colours, I still find this
+                 figure quite ugly and feel that it does not show much. I may
+                 remove it from this paper and keep the idea for later use.}}
+  \label{fig:erosion}
 \end{figure}
@@
 
+% fig:deglacshots
+\begin{figure}
+  \includegraphics{deglacshots}
+  \caption{Snapshots of modelled surface topography (200\,\unit{m} contours)
+           and surface velocity (colour mapping) from the GRIP simulation,
+           corresponding to the last glacial maximum (-19.5 kyr) and the last
+           deglaciation.}
+  \label{fig:deglacshots}
+\end{figure}
+
 % ----------------------------------------------------------------------

### 2014-06-07 cordillera-cycle.tex
Polish introduction.

@@
 Cordillera were more extensive than today. At the last glacial maximum, a
-contiguous blanket of ice extended from Denali Peak in the north to the Puget
-Sound in the south \aref. This ice mass is known as the former Cordilleran ice
-sheet (Fig.~\ref{fig:locmap}).
-\note{Make sure these locations are marked on Fig.~\ref{fig:locmap}.}
+contiguous blanket of ice extended from the Alaska Range in the north to the
+North Cascades in the south \aref. This ice mass is known as the former
+Cordilleran ice sheet (Fig.~\ref{fig:locmap}).
 
 For more than a century, exploration and geological investigation of the
-Cordillera have led to a large collection of evidence of the former ice cover \aref. This
-evidence consists of mapped boundaries of the former ice extent \aref, directions of
-past sliding velocities {\aref} and locations of former melt-water streams \aref, as well as
-the timing of glacier retreat identified from radiocarbon and cosmogenic dating \aref.
+Cordillera have led to a large collection of evidence of the former ice cover
+\aref. This evidence consists of mapped boundaries of the former ice extent
+\aref, directions of past sliding velocities {\aref} and locations of former
+melt-water streams \aref, as well as the timing of glaciation identified from
+radiocarbon dating \aref, cosmogenic dating \aref, and the offshore sedimental
+record \aref.
 
@@
 the case for the Cordilleran ice sheet, where complex arrangements of ice flow
-directions emanating from multiple glaciation centres have been identified \aref.
+directions emanating from multiple glaciation centres have been identified, but
+remain poorly understood \aref.
 
@@
 and precipitation fields derived from a combination of a present-day
-atmospheric reanalysis (NARR, \aref), lapse-rate corrections and temperature offset
-time series. The latter are
+atmospheric reanalysis (NARR, \aref), lapse-rate corrections,
+and temperature offset time series. The latter are
 obtained by scaling six different palaeo-temperature reconstructions from
@@
            St.-Elias mountains (WSE), the Selwyn and MacKenzie mountains (SMK),
-           the Coast Mountains (CM) and the Rocky Mountains (RM). The background
+           the Skeena Mountains, the Coast Mountains (CM), the Rocky Mountains
+           (RM) and the North Cascades (NC). The background
            map consists of ETOPO1 \citep{data:etopo1} and Natural Earth Data

### 2014-06-07 cordillera-cycle.tex
Add last flow direction map in manuscript.
Polish model setup.

@@
 related to the modelled amount of water in the till, using a formula derived
-from laboratory experiments with till extracted from an Antarctic ice stream.
-Basal topography is derived from the ETOPO1 combined topography
+from laboratory experiments with till extracted from an Antarctic ice stream
+\aref. Basal topography is derived from the ETOPO1 combined topography
 and bathymetry dataset with a~resolution of 1\,arc-min \citep{data:etopo1}.
-
 Sea level is lowered as a function of time according to the SPECMAP
@@
 computed from daily temperature values from the North American Regional
-Reanalysis \citep{data:narr}, excluding variability associated with the
+Reanalysis (NARR) \citep{data:narr}, after excluding variability associated
+with the
 seasonal cycle itself \citet{seguinot-rogozhina-2014}. The
@@
 Atmospheric forcing of the model consists of a present-day monthly climatology
-$\{T_m0, P_m0\}$, modified by temperature-offset time series ${\Delta}T_{TS}$
-and a lapse-rate correction ${\Delta}T_{LR}$.
+$\{T_{m0}, P_{m0}\}$, modified by temperature-offset time series
+${\Delta}T_{TS}$ and a lapse-rate correction ${\Delta}T_{LR}$.
 \begin{align}
@@
 
-The present-day climatology $\{T_m0, P_m0\}$ was computed from the near-surface air temperature and
-precipitation rate fields of the North American Regional Reanalysis (NARR) over
+The present-day climatology $\{T_{m0}, P_{m0}\}$ was computed from the
+near-surface air temperature and precipitation rate fields of the NARR over
 the period 1979--2000 (Fig.~\ref{fig:atm}). Modern climate of the North
@@
 high-resolution precipitation forcing, as outlined in a previous sensitivity
-study.
+study \aref.
 
@@
 Vostok ice core, and Ocean Drilling Program (ODP) sites 1012 and 1020, both
-located off the Californian shore. Palaeo-temperatures from the GRIP and NGRIP
-records were calculated using a quadratic equation \aref:
+located off the Californian shore \aref. Palaeo-temperatures from the GRIP and
+NGRIP records were calculated using a quadratic equation \aref:
 \begin{equation}
@@
 This correction accounts for the evolution of ice thickness $h$ on the one
-hand, and for differences between the climate forcing reference topography and
-the ice flow model basal topography $z_{b}-z_{ref}$ on the other hand. An
-annual lapse rate of $\gamma = 6\,\unit{{\degree}C\,km^{-1}}$ is used in all
-simulations.
+hand, and for differences between the the ice flow model basal topography
+$z_{b}$ and the climate forcing reference topography $z_{ref}$ on the other
+hand. All simulations use an annual temperature lapse rate of
+$\gamma = 6\,\unit{{\degree}C\,km^{-1}}$.
 
@@
 
+% fig:lastflow
+\begin{figure}
+  \includegraphics{lastflow}
+  \caption{Modelled directions of the last basal ice velocities. Hatches
+           indicate areas that remain non-sliding throughout the simulation.
+           Sliding grid cells were distinguished from non-sliding grid cells
+           using a velocity threshold of 1\,\unit{m\,yr^{-1}}.}
+  \label{fig:lastflow}
+\end{figure}
+
 % ----------------------------------------------------------------------

### 2014-06-08 cordillera-cycle.tex
Polish results section.

@@
 during most of the glacial cycle, while two major glaciation events occur
-during MIS~4 (61.9--55.4\,\unit{kyr}) and MIS~2 (29.5--16.9\,\unit{kyr},
-Fig.~\ref{fig:timeseries}). A local ice minimum is consistently attained in
-between these two events during MIS~3 (52.2--45.6\,\unit{kyr}).
+during Marine Isotope Stage (MIS)~4 (61.9--55.4\,\unit{kyr}) and MIS~2
+(29.5--16.9\,\unit{kyr}, Fig.~\ref{fig:timeseries}). A local ice minimum is
+consistently attained between these two events, during MIS~3
+(52.2--45.6\,\unit{kyr}).
 However, the magnitude and timing of these events vary significantly
@@
 thousands of years earlier. The NGRIP record is the only temperature forcing
-that predicts a more extensive glaciation during MIS~4 than at the LGM.
+that predicts a more extensive glaciation during MIS~4 than during the LGM.
 
@@
 ice volume variability throughout the simulation length. The ODP~1012
-record-driven simulation predicts rapid deglaciation prior to 20\,\unit{kyr}
+simulation predicts rapid deglaciation prior to 20\,\unit{kyr}
 before present due to a local temperature maximum inconsistent with all other
-temperature forcing time series used. The ODP~1012 record-driven simulation
-predicts an early last glacial volume maximum, followed by slower deglaciation
+temperature forcing time series used. The ODP~1020 simulation predicts an early last glacial volume maximum (29.5\,\unit{kyr}), followed by slower deglaciation
 than modelled using other records. Finally, simulations forced by Antarctic ice
 core records produce a continuous, rapid deglaciation after the last glacial
-maximum, whereas the simulations driven by the GRIP records produce a rapid
-deglaciation in three steps, including a glacial re-advance around the Younger
-Dryas event.
+maximum, whereas the simulation driven by the GRIP record produce a rapid
+deglaciation in three steps, including stand-stills and readvances of the ice
+margin.
 
@@
 ice sheet geometry at the last glacial maximum appears very similar from one
-simulation to the next, including a central divide located at about
+simulation to the next. It includes a central divide located at about
 3\,500\,\unit{m} elevation along the spine of the Rocky Mountains. In all

### 2014-06-09 cordillera-cycle.tex
Work in progress on the discussion.

@@
 before present due to a local temperature maximum inconsistent with all other
-temperature forcing time series used. The ODP~1020 simulation predicts an early last glacial volume maximum (29.5\,\unit{kyr}), followed by slower deglaciation
+temperature forcing time series used. The ODP~1020 simulation predicts an early
+last glacial volume maximum (29.5\,\unit{kyr}), followed by slower deglaciation
 than modelled using other records. Finally, simulations forced by Antarctic ice
@@
 
-During the last glacial cycle, geomorphological evidence has allowed to
-identify two major advances of the Cordilleran ice sheet. These correspond to
-the last glacial maximum \aref, and the older xxx glaciation, found in the
-northernmost parts of the ice sheet \aref. It should be noted that all
-simulations presented here reproduce these two major events, which constitutes
-a first-order agreement between model results and geomorphological evidence.
+Independently of the choice of palaeo-temperature record used as forcing
+time-series for the ice sheet model, our simulations consistently produce two
+glaciation maxima during MIS~4 (61.9--55.4\,\unit{kyr}) and MIS~2
+(29.5--16.9\,\unit{kyr}). These events broadly correspond in timing to the
+Gladstone (MIS~4) and McConell (MIS~2) glaciation documented by geological
+evidence in the northern parts of the model domain, and to the Fraser (MIS~2)
+glaciation in the southern part.
 
-However, the timing of these events depends highly on the choice of temperature
-time-series used to force the model. For instance, the timing of the last
-glacial maximum is best reproduced by the Antarctic records, while simulations
-driven by the NGRIP and oceanic records pre-date the last glacial maximum by
-several thousands of years. Regarding the oceanic records, the early volume
-maxima can be explain by an early warming in the palaeo-temperature record.
-This early warming has been previously documented, and explained as a
-weakening of the California current, precisely associated with the presence of
-the Cordilleran ice sheet \aref. This precocious glacial maximum illustrates
+However, the timing of the modelled glacial maxima strongly depends on the
+choice of palaeo-temperature record used, and a more precise agreement with
+timing reconstructed from geological evidence is only attained by some of the
+simulations. According to radiocarbon dating, the last glacial maximum advance
+of the southern Cordilleran ice sheet margin has been constrained between
+00.0\,\unit{kyr} and 00.0\,\unit{kyr}. These dates have been confirmed by the
+offshore sediment record, which shows an increase of xxx
+betweem 00.0\,\unit{kyr} and 00.0\,\unit{kyr}.
+
+Among simulations presented here, only those forced by the GRIP, EPICA and
+Vostok records place the last glacial maximum within this range
+(Fig.~\ref{fig:timeseries}, lower panel). Simulations driven by the NGRIP,
+ODP~1012 and ODP~1020 records, on the opposite, produce a last glacial maximum
+that pre-dates reconstructed ages by several thousand years. Concerning
+simulations driven by oceanic (ODP~1012, ODP~1020) records, this early
+deglaciation is caused by an early warming in the palaeo-temperature records
+(Fig.~\ref{fig:timeseries}, upper panel). However, it has been shown that this
+early warming is a local effect, corresponding to a weakening of the California
+current \aref. The California currents is a southwards current driving cold
+waters along the western coast of USA. It is thought to have weakened in the
+presence of the Cordilleran ice sheet, explaining paradoxically warmer
+sea-surface temperatures at the location of the ODP~1012 and ODP~1020 sites
+during the last glacial maximum \aref.
+
+This precocious glacial maximum illustrates
 the complexity of climate-ice-sheet feedbacks on a regional scale, and shows

### 2014-06-10 cordillera-cycle.tex
Develop first part of the discussion.

@@
 Gladstone (MIS~4) and McConell (MIS~2) glaciation documented by geological
-evidence in the northern parts of the model domain, and to the Fraser (MIS~2)
+evidence in the northern part of the model domain, and to the Fraser (MIS~2)
 glaciation in the southern part.
@@
 choice of palaeo-temperature record used, and a more precise agreement with
-timing reconstructed from geological evidence is only attained by some of the
-simulations. According to radiocarbon dating, the last glacial maximum advance
-of the southern Cordilleran ice sheet margin has been constrained between
-00.0\,\unit{kyr} and 00.0\,\unit{kyr}. These dates have been confirmed by the
-offshore sediment record, which shows an increase of xxx
-betweem 00.0\,\unit{kyr} and 00.0\,\unit{kyr}.
+timing reconstructed from geological evidence is only attained in some of our
+simulations. In the Puget lowland, the last glacial maximum advance of the
+southern Cordilleran ice sheet margin has been constrained between by
+radiocarbon dating between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,kyr\,BP}.
+These dates have been confirmed by the offshore sediment record, which shows an
+increase of glaciomarine sedimentation between 19.5 and
+16.2\,\unit{\chem{^{14}C}\,cal\,kyr\,BP}.
 
@@
 early warming is a local effect, corresponding to a weakening of the California
-current \aref. The California currents is a southwards current driving cold
-waters along the western coast of USA. It is thought to have weakened in the
-presence of the Cordilleran ice sheet, explaining paradoxically warmer
+current \aref. The California current, driving cold waters southwards
+along the western coast of North America, is thought to have weakened in the
+presence of the Cordilleran ice sheet, resulting in paradoxically warmer
 sea-surface temperatures at the location of the ODP~1012 and ODP~1020 sites
@@
 
-This precocious glacial maximum illustrates
-the complexity of climate-ice-sheet feedbacks on a regional scale, and shows
-that, although closely located to the modelling domain, the ODP~1012 and
-ODP~1020 proxy records records can't be reliably used as a temperature forcing
-to model the last Cordilleran ice sheet.
+Because most of the Cordilleran ice sheet marine front lies in a different
+oceanic circulation gyre than that holding the California current, it is
+unlikely that it was affected by this early warming. However, the above paradox
+illustrates the complexity of ice-sheet feedbacks on regional climate, and
+demonstrates that, although located in the neighbourhood of the modelling
+domain, the ODP~1012 and ODP~1020 palaeo-temperature proxy records can not be
+used as a realistic forcing time-series to model the Cordilleran ice sheet
+through the last glacial cycle.
+
+These results makes us believe that a palaeo-temperature proxy record located
+in the direct vicinity of the former Cordilleran ice sheet would be of great
+help to understand its glacial history. Because, to our knowledge, no such
+record yet exists, we focus the rest of our analysis on simulations forced by
+temperature records from one Greenland (GRIP) and one Antarctic (EPICA) ice
+cores.
 

### 2014-06-10 cordillera-cycle.tex
Add discussion of glaciated areas.

@@
 record yet exists, we focus the rest of our analysis on simulations forced by
-temperature records from one Greenland (GRIP) and one Antarctic (EPICA) ice
-cores.
+temperature records from a Greenland (GRIP) and an Antarctic (EPICA) ice cores.
 
@@
 
-During most of the glacial cycle, ice cover is restricted to high mountain
-areas. This is particularly true when using the Greenland-based forcing
-(Fig~\ref{fig:duration}). Most of the glacial maxima margin has a transient
-character, showing that the ice sheet is far out of balance with climate. There
-persists a central ice dome over the Skeena mountains over the entire glacial
-period. This ice dome serves as a nucleation centre at the initiation of the
-advance phase preceding the last glacial maximum. It is here because of more
-widespread precipitation over the Skeena mountains than across other East-West
-transects towards south or north of this region. We interpret this as an
-orographic precipitation effect resulting from a "window" through the sharp
-range of the Coast Mountains at this latitude (Fig?).
+Although snapshots from the model output (Fig.~\ref{fig:duration}) give a
+crisp picture of modelled ice dynamics at specific time frames, they often
+can not be directly compared to geomorphological evidence, which is by nature
+time-transgressive. In this section, we use the 6-km horizontal-resolution
+model output from simulations driven by the GRIP and EPICA records to produce
+a series of time-integrated plots containing information aggregated over the
+whole simulation period.
+
+During most of the glacial cycle, modelled ice cover is restricted to disjoint
+ice caps centred on major mountain ranges of the North American Cordillera
+(Fig.~\ref{fig:duration}). A~contiguous ice cover spanning from the Alaska
+Range to the Rocky Mountains exists for about 28\,\unit{kyr} in the GRIP
+simulation and for about 29\,\unit{kyr} in the EPICA simulation. These values
+represent cumulative ice cover over the two glaciations during MIS~2 and MIS~4,
+while a dynamically consistent Cordilleran ice sheet forms for even shorter
+time spans.
+
+With exception of the oceanic front and the northern foothills of the Alaska
+Range, most of the maximal extent of the ice sheet is associated with short
+durations of ice cover, generally a few thousand years, distributed over one
+(GRIP) or two (EPICA) glaciations. This shows that, in our simulations, the
+most extensive stages of the Cordilleran ice sheet are both short-lived and
+out of balance with contemporary climate states, a result which confirms
+inferences from earlier geological studies \aref.
+
+However, our simulations depict, independently of the choice of record used for
+palaeo-temperature forcing, a central ice cap persisting throughout the entire
+last glacial cycle on the Skeena mountains (Fig.~\ref{fig:snapshots} and
+\ref{fig:duration}). In particular, this ice cap persists between the MIS~4
+and MIS~2 glaciations, providing a nucleation centre for the onset of glacial
+readvance towards the last glacial maximum.
+
+During time periods when a fully-grown Cordilleran ice sheet is absent from our
+simulation, the presence of the Skeena ice cap can be explained by more
+widespread winter precipitation there than in other parts of the modelling
+domain (Figs.~\ref{fig:atm}). Along most of the north-western North American
+shore, coastal mountain ranges provide a sharp orographic obstacle to westerly
+winds, capturing atmospheric moisture in the form of precipitation, and leaving
+interior lowlands arid. However, near the centre of our modelling domain, the
+topographic barrier formed by the Coast Mountains is less pronounced than
+elsewhere, allowing westerly winds to carry moisture further inland, until it
+is progressively captured by the broad and mildly elevated group of the Skeena
+Mountains.
+
+In addition, the erosional landscape of the Skeena mountain bears one of the
+strongest glacial imprint (?) found within the area formerly covered by the
+Cordilleran ice sheet \aref. Here we suggest that persistent ice cover
+associated with basal temperatures at the ice melting point
+(Fig.~\ref{fig:warmbase}) may partly explain this spectacular landscape.
 

### 2014-06-10 cordillera-cycle.tex
Update discussion subsection titles.
Introduce 6-km resolution in text.

@@
 Shallow Ice Approximation (SIA) \citep{bueler-brown-2009,winkelmann-etal-2011}.
-SIA and SSA velocities are computed by finite difference methods on a~10\,km
-resolution horizontal grid of 300 by 150 points (the modelling domain). Ice
-softness depends on temperature and water content through an enthalpy
-formulation \citep{aschwanden-blatter-2009,aschwanden-etal-2012}. Enthalpy is
-computed in three dimensions in up to 51 layers within the
-ice, and temperature is additionally computed in 31 layers in bedrock to
-a~depth of 3\,km. Surface air temperature from the atmospheric forcing
-provides the upper boundary condition to the ice enthalpy model, and
-a~geothermal heat flux of 70\,\unit{mW\,m^{-2}} provides the lower boundary
-condition to the bedrock thermal model. Although this uniform value does not
+Ice softness depends on temperature and water content through an enthalpy
+formulation \citep{aschwanden-blatter-2009,aschwanden-etal-2012}. Surface air
+temperature from the atmospheric forcing provides the upper boundary condition
+to the ice enthalpy model. Temperature is computed subglacially to a~depth of
+3\,km, where it is conditioned by a~geothermal heat flux of
+70\,\unit{mW\,m^{-2}}. Although this uniform value does not
 account for the high spatial geothermal variability in the region, it is on
@@
 
+For all records, we firstly run a low-resolution simulation, using
+a~10\,\unit{km}, 300 by 150~points horizontal grid for SSA and SIA velocities,
+up to 51~enthalpy layers in the ice, and 31~temperature layers in the bedrock.
+For the GRIP and EPICA records, we then run a second simulation using a higher
+horizontal resolution of 6\,\unit{km}, up to 101~ice layers, and 61~bedrock
+layers.
+
 % ----------------------------------------------------------------------
@@
 
-\subsection{Timing of the glacial maxima}
+\subsection{Timing of the glaciation peaks}
 
@@
 
-\subsection{Glaciated areas through the last glacial cycle}
+\subsection{Spatial imprint of ice cover}
 
@@
 
-\subsection{History of the last deglaciation}
+\subsection{Deglaciation history}
 

### 2014-06-10 cordillera-cycle.tex
Restructure atmosphere forcing for readability.

@@
 Atmospheric forcing of the model consists of a present-day monthly climatology
-$\{T_{m0}, P_{m0}\}$, modified by temperature-offset time series
-${\Delta}T_{TS}$ and a lapse-rate correction ${\Delta}T_{LR}$.
+$\{T_{m0}, P_{m0}\}$, modified by a lapse-rate correction ${\Delta}T_{LR}$ and
+temperature-offset time series ${\Delta}T_{TS}$.
 \begin{align}
-    T_m(t, x, y) &= T_{m0}(x, y) + {\Delta}T_{TS}(t) + {\Delta}T_{LR}(t, x, y) \\
+    T_m(t, x, y) &= T_{m0}(x, y) + {\Delta}T_{LR}(t) + {\Delta}T_{TS}(t, x, y) \\
     P_m(t, x, y) &= P_{m0}(x, y)
@@
 
-The present-day climatology $\{T_{m0}, P_{m0}\}$ was computed from the
+The present-day climatology $\{T_{m0}, P_{m0}\}$ was computed from
 near-surface air temperature and precipitation rate fields of the NARR over
@@
 
+The lapse-rate correction ${\Delta}T_{LR}$ is computed using the NARR surface
+geopotential height a reference topography $z_{ref}$,
+\begin{align}
+    {\Delta}T_{LR}(t, x, y) &= -\gamma [z_{s}(t, x, y)-z_{ref}] \\
+                            &= -\gamma [h(t, x, y)+z_{b}(t, x, y)-z_{ref}],
+\end{align}
+
+thus accounting for the evolution of ice thickness $h$ on the one hand, and
+for differences between the the ice flow model basal topography $z_{b}$ and the
+NARR topography $z_{ref}$ on the other hand. All simulations use an annual temperature lapse rate of $\gamma = 6\,\unit{{\degree}C\,km^{-1}}$.
+
 Temperature offset time-series ${\Delta}T_{TS}$ are derived from proxy records from
@@
 located off the Californian shore \aref. Palaeo-temperatures from the GRIP and
-NGRIP records were calculated using a quadratic equation \aref:
+NGRIP records were calculated using a quadratic equation \aref,
 \begin{equation}
     {\Delta}T_{TS}(t) = -11.88[\chem{\delta^{18}O(t)}-\chem{\delta^{18}O}(0)]
-                        -0.1925[\chem{\delta^{18}O(t)}^2-\chem{\delta^{18}O}(0)^2]
+                        -0.1925[\chem{\delta^{18}O(t)}^2-\chem{\delta^{18}O}(0)^2],
 \end{equation}
@@
 
-Prior to surface mass balance computation, the model dynamically applies
-a~lapse-rate correction to surface air temperature,
-\begin{align}
-    {\Delta}T_{LR}(t, x, y) &= -\gamma [z_{s}(t, x, y)-z_{ref}] \\
-                            &= -\gamma [h(t, x, y)+z_{b}(t, x, y)-z_{ref}]
-\end{align}
-
-This correction accounts for the evolution of ice thickness $h$ on the one
-hand, and for differences between the the ice flow model basal topography
-$z_{b}$ and the climate forcing reference topography $z_{ref}$ on the other
-hand. All simulations use an annual temperature lapse rate of
-$\gamma = 6\,\unit{{\degree}C\,km^{-1}}$.
-
-For all records, we firstly run a low-resolution simulation, using
+For all proxy records, we firstly run a low-resolution simulation, using
 a~10\,\unit{km}, 300 by 150~points horizontal grid for SSA and SIA velocities,

### 2014-06-10 cordillera-cycle.tex
Add 6-km resolution time-series.

@@
            \chem{\delta^{18}O} records \citep{lisiecki-raymo-2005}. Hatched
-           rectangles indicate modelled ice volume extrema corresponding to
+           rectangles highlight modelled ice volume extrema corresponding to
            MIS~4 (61.9--55.4\,\unit{kyr}), MIS~3 (52.2--45.6\,\unit{kyr}), and
-           MIS~2 (last glacial maximum, 29.5--16.9\,\unit{kyr}).}
+           MIS~2 (last glacial maximum, 29.5--16.9\,\unit{kyr}). Dotted lines
+           correspond to the GRIP and EPICA 6\,\unit{km}-resolution runs.}
   \label{fig:timeseries}

### 2014-06-10 figures/timeseries.py

@@ -44,14 +44,14 @@ for i, rec in enumerate(records):
              color=colors[i], marker=markers[i], label=labels[i])
 
     # look for a high-resolution run
-    #try:
-    #    nc = Dataset(run_path % ('5km', rec, dt*100) + '-ts.nc')
-    #    ts_time = nc.variables['time'][:]*s2ka
-    #    ts_ivol = nc.variables['slvol'][:]
-    #    nc.close()
-    #    ax2.plot(ts_time, ts_ivol, color=colors[i], dashes=(1, 1))
-    #except RuntimeError:
-    #    pass
+    try:
+        nc = Dataset(run_path % ('6km', rec, dt*100) + '-ts.nc')
+        ts_time = nc.variables['time'][:]*s2ka
+        ts_ivol = nc.variables['slvol'][:]
+        nc.close()
+        ax2.plot(ts_time, ts_ivol, color=colors[i], dashes=(1, 1))
+    except RuntimeError:
+        pass
 
 # mark true MIS stages
 # source: http://www.lorraine-lisiecki.com/LR04_MISboundaries.txt

### 2014-06-11 cordillera-cycle.tex
Add deglaciation profiles in manuscript.
Add deglaciation time-series in manuscript.

@@
 
+% fig:deglacseries
+\begin{figure}
+  \includegraphics{deglacseries}
+  \caption{Temperature offset time-series from the GRIP and EPICA ice core
+           records (Table~\ref{tab:records}) \textbf{(top)}, and modelled ice
+           volume during the deglaciation, expressed in meters of sea-level
+           equivalent \textbf{(bottom)}.}
+  \label{fig:deglacseries}
+\end{figure}
+
 % fig:deglac
@@
 
+% fig:profiles
+\begin{figure}
+  \includegraphics{profiles}
+  \caption{Modelled bedrock (black) and ice surface (blue) topography profiles
+           during deglaciation (22.0--8.0\,\unit{kyr}) in the GRIP 6\,\unit{km}
+           simulation, corresponding to the four transects indicated in
+           Fig.~\ref{fig:deglac}.}
+  \label{fig:profiles}
+\end{figure}
+
 % fig:lastflow

### 2014-06-11 cordillera-cycle.tex
Deglaciation history first shot.

@@
 
-Deglaciation happens first in the southern half and then in the northern half
-of the ice-covered area (Fig.~\ref{fig:deglac}). Retreat patterns are different
-than reconstructed over central BC, however they match Martin's reconstruction
-of Liard lowland "unzipping" pretty well. In the GRIP-driven simulation, the
-patterns of de-glaciation and Younger Dryas re-advance match with recent
-cosmogenic exposure dates.
+In the North American Cordillera alike other glaciated regions, the large
+majority of the glacial geologic record relates to the last few millennia of
+glacier cover, most of the older evidence having been overridden by ice retreat
+retreat during the deglaciation.
+
+In our simulations, the timing of the last glacial maximum, and in turn that of
+the deglaciation, depends on the palaeo-temperature time-series used to force
+the model (Figs.~\ref{fig:timeseries} and ~\ref{fig:deglacseries}). The
+6\,\unit{km}-resolution simulation forced by the EPICA record reaches a maximal
+ice volume at 17.3\,\unit{kyr}, followed by a continuous deglaciation between
+about 17 and 11\,\unit{kyr}. That driven by the GRIP record reaches maximal
+volume at 19.1\,\unit{kyr}, and produce a deglaciation in three steps,
+including a first stand-still between 16.6 and 14.5\,\unit{kyr}, and a late
+glacial re-advance between 12.6 and 11.6\,\unit{kyr}.
+
+Although the timing differs, the patterns of glacier retreat are relatively
+consistent between the two simulations (Figs.~\ref{fig:deglac}). In both
+simulations, the southern half of the modelling domain, including the Puget
+Lowland, the Coast and Rocky Mountains and the Interior Plateau of British
+Columbia, are completely deglaciated at a time where there remain significant
+ice cover on the Skeena, Selwyn, MacKenzie, Wrangell and Saint-Elias Mountains
+in the northern half of the domain. At this stage, a late glacial re-advance
+occurs in the GRIP simulation. Both simulations then show an unzipping of the
+Liard Lowland, followed by ice retreat towards the Selwyn Mountains and the
+Skeena Mountains, which host the last remain of the palaeo-ice sheet.
+
+On the Interior Plateau of British Columbia, modelled deglaciation is rapid
+and northwards, resulting in the presence of a slow-flowing ice lobe on the
+plateau, while the surrounding mountain ranges are already deglaciated
+(Fig.~\ref{fig:profiles}). Although this supports the top-down (?)
+deglaciation hypothesis for central British Columbia, modelled, southwards,
+last flow directions do not agree with the local lineation patterns
+(Fig.~\ref{fig:lastflow}). This could indicate that the lineations are older,
+or that the modelled last glacial maximum ice sheet is too extensive to the
+east, resulting in modelled deglaciation flow patterns different from those
+observed in the geomorphology.
 

### 2014-06-11 cordillera-cycle.tex
Update notes and todos.

@@
 % ----------------------------------------------------------------------
-
-\todo{Small adjustments to avoid repetitions with the previous paper.}
 
@@
 % ----------------------------------------------------------------------
-
-\todo{Develop the three subsections in order to reach the conclusions.}
 
@@
            MIS~2 (last glacial maximum, 29.5--16.9\,\unit{kyr}). Dotted lines
-           correspond to the GRIP and EPICA 6\,\unit{km}-resolution runs.}
+           correspond to the GRIP and EPICA 6\,\unit{km}-resolution runs.
+           \todo{add grid lines to improve readability}}
   \label{fig:timeseries}
@@
            over the last 120\,\unit{kyr}.
-           \note{Even after carefully adjusting the colours, I still find this
-                 figure quite ugly and feel that it does not show much. I may
+           \note{I am still unsure about the use of this figure. I may
                  remove it from this paper and keep the idea for later use.}}
@@
            volume during the deglaciation, expressed in meters of sea-level
-           equivalent \textbf{(bottom)}.}
+           equivalent \textbf{(bottom)}.
+           \todo{add grid lines to improve readability}}
   \label{fig:deglacseries}
@@
            corresponding to the last glacial maximum (-19.5 kyr) and the last
-           deglaciation.}
+           deglaciation.
+           \note{should we have a similar figure using EPICA?}}
   \label{fig:deglacshots}
@@
            simulation, corresponding to the four transects indicated in
-           Fig.~\ref{fig:deglac}.}
+           Fig.~\ref{fig:deglac}.
+           \note{should we have a similar figure using EPICA?}}
   \label{fig:profiles}

### 2014-06-11 cordillera-cycle.tex
Update contiguous ice sheet duration.

@@
 ice caps centred on major mountain ranges of the North American Cordillera
-(Fig.~\ref{fig:duration}). A~contiguous ice cover spanning from the Alaska
-Range to the Rocky Mountains exists for about 28\,\unit{kyr} in the GRIP
-simulation and for about 29\,\unit{kyr} in the EPICA simulation. These values
-represent cumulative ice cover over the two glaciations during MIS~2 and MIS~4,
-while a dynamically consistent Cordilleran ice sheet forms for even shorter
-time spans.
+(Fig.~\ref{fig:duration}). In both the GRIP and the EPICA simulation,
+a~contiguous ice cover spanning from the Alaska Range to the Rocky Mountains
+exists for about 29\,\unit{kyr}. This value represent cumulative ice cover over
+the two glaciations during MIS~2 and MIS~4, while a dynamically consistent
+Cordilleran ice sheet forms for even shorter time spans.
 
@@
   \caption{Modelled duration of ice cover during the last 120\,\unit{kyr}.
-           A contiguous ice cover spanning from the Alaska Range (AR) to the
-           southern Coast Mountians (CM) and Rocky Mountains (RM) exists for
-           about 28\,\unit{kyr} in the GRIP simulation and about 29\,\unit{kyr}
-           in the EPICA simulation. Note the irregular colour scale. A central
+           Note the irregular colour scale. A contiguous ice cover spanning
+           from the Alaska Range (AR) to the southern Coast Mountains (CM) and
+           Rocky Mountains (RM) exists for about 29\,\unit{kyr} in both
+           simulations. A central
            ice cover persists over the Skeena Mountains (SM) during most of the

### 2014-06-11 cordillera-cycle.tex
Update state of the manuscript.

@@
     Introduction & Happy with it except for missing refs.\\
-    Model setup  & Needs minor fixes to avoid repeating the previous paper.\\
-    Results      & Almost happy with it. \\
-    Discussion   & Large gaps still have to be filled.\\
-    Conclusions  & Main contents are here but the text needs to be improved.\\
-    Figures      & I have plans for a couple extra ones.\\
+    Model setup  & Happy with it except for missing refs.\\
+    Results      & Happy with it except for missing refs.\\
+    Discussion   & Deglaciation parts needs improvement.\\
+    Conclusions  & Can be compacted and the style improved.\\
     \bottomhline
@@
 
-\todo{Improve the style.}
-
 Numerical simulations of the Cordilleran ice sheet through the last glacial

### 2014-06-12 cordillera-cycle.tex
Rework results for structure and avoid repetition.

@@
 horizontal resolution of 6\,\unit{km}, up to 101~ice layers, and 61~bedrock
-layers.
+layers. In the rest of this paper, we refer to different model runs by the
+proxy record used for palaeo-temperature forcing.
 
@@
 
-Despite large differences in the input temperature offset time-series, the
-model output presents consistent features that can be observed across the range
-of forcing data used. In all simulations, total ice volume is relatively low
+Despite large differences in input temperature offset time-series, the model
+output presents consistent features that can be observed across the range of
+forcing data used. In all simulations, total ice volume is relatively low
 during most of the glacial cycle, while two major glaciation events occur
@@
 (29.5--16.9\,\unit{kyr}, Fig.~\ref{fig:timeseries}). A local ice minimum is
-consistently attained between these two events, during MIS~3
-(52.2--45.6\,\unit{kyr}).
-However, the magnitude and timing of these events vary significantly
-between our simulations.
+consistently attained during MIS~3 (52.2--45.6\,\unit{kyr}). However, the
+magnitude and timing of these events depend significantly on the choice of
+proxy record used for palaeo-temperature forcing.
 
@@
 highest variability in modelled ice volume throughout the glaciation history.
-Although simulations forced by the Antarctic (EPICA, Vostok) and GRIP records
-reach the last glacial maximum in ice volume at 19.5--16.9\,\unit{kyr}, those
-driven by oceanic (ODP~1012, ODP~1020) and the NGRIP record attain it several
-thousands of years earlier. The NGRIP record is the only temperature forcing
-that predicts a more extensive glaciation during MIS~4 than during the LGM.
+Model runs using oceanic and Antarctic records generally result in lower
+modelled ice volume during MIS~4 and larger ice volume during MIS~3, thus
+suggesting lower ice volume variability throughout the simulation length. The
+NGRIP palaeo-temperature forcing is the only one that results in larger
+glaciation during MIS~4 than during the last glacial maximum.
 
-Simulations forced by oceanic and Antarctic records generally result in lower
-modelled ice volume during MIS~4 and larger ice volume during MIS~3, thus lower
-ice volume variability throughout the simulation length. The ODP~1012
-simulation predicts rapid deglaciation prior to 20\,\unit{kyr}
-before present due to a local temperature maximum inconsistent with all other
-temperature forcing time series used. The ODP~1020 simulation predicts an early
-last glacial volume maximum (29.5\,\unit{kyr}), followed by slower deglaciation
-than modelled using other records. Finally, simulations forced by Antarctic ice
-core records produce a continuous, rapid deglaciation after the last glacial
-maximum, whereas the simulation driven by the GRIP record produce a rapid
-deglaciation in three steps, including stand-stills and readvances of the ice
-margin.
+While simulations driven by the Antarctic (EPICA, Vostok) and GRIP records
+reach a last glacial maximum in ice volume between 19.5 and 16.9\,\unit{kyr},
+those steered by oceanic (ODP~1012, ODP~1020) and the NGRIP record attain it
+several thousands of years earlier. Moreover, the ODP~1012 run yields rapid
+deglaciation prior to 20\,\unit{kyr} due to a local temperature maximum that is
+inconsistent with all other records used. The ODP~1020 simulation predicts an
+early last glacial maximum in ice volume at 29.5\,\unit{kyr}, followed by
+slower deglaciation than modelled using other records. Finally, model runs
+informed by Antarctic ice core records produce a continuous, rapid deglaciation
+after the last glacial maximum, whereas the simulations driven by the GRIP
+record produce a rapid deglaciation in three steps, including stand-stills and
+readvances of the ice margin.
 
@@
 consistent patterns of glaciation (Fig.~\ref{fig:snapshots}). As a result of
-the different scaling factors applied to palaeo-temperature proxy records, the
+different scaling factors applied to palaeo-temperature proxy records, the
 ice sheet geometry at the last glacial maximum appears very similar from one

### 2014-06-12 cordillera-cycle.tex
Add corrections from Irina.

@@
   approach reveals large variations in model response to evolving atmospheric
-  forcing, all simulations produce two major glaciation events during MIS~4
-  (61.9--55.4\,\unit{kyr}) and MIS~2 (29.5--16.9\,\unit{kyr}). The timing of
-  glaciation is
+  forcing, all simulations produce two major glaciation events during
+  marine isotope stages~4 (61.9--55.4\,\unit{kyr}) and~2
+  (29.5--16.9\,\unit{kyr}). The timing of glaciation is
   better reproduced using temperature reconstructions from Greenland and
@@
     Results      & Happy with it except for missing refs.\\
-    Discussion   & Deglaciation parts needs improvement.\\
+    Discussion   & Deglaciation part needs improvement.\\
     Conclusions  & Can be compacted and the style improved.\\
@@
 (29.5--16.9\,\unit{kyr}). These events broadly correspond in timing to the
-Gladstone (MIS~4) and McConell (MIS~2) glaciation documented by geological
-evidence in the northern part of the model domain, and to the Fraser (MIS~2)
-glaciation in the southern part.
+Gladstone (MIS~4) and McConell (MIS~2) glaciations documented by geological
+evidence in the northern part of the model domain \aref, and to the Fraser
+(MIS~2) glaciation in the southern part \aref.
 
@@
 simulations. In the Puget lowland, the last glacial maximum advance of the
-southern Cordilleran ice sheet margin has been constrained between by
-radiocarbon dating between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,kyr\,BP}.
+southern Cordilleran ice sheet margin has been constrained by radiocarbon
+dating between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,kyr\,BP} \aref.
 These dates have been confirmed by the offshore sediment record, which shows an
 increase of glaciomarine sedimentation between 19.5 and
-16.2\,\unit{\chem{^{14}C}\,cal\,kyr\,BP}.
+16.2\,\unit{\chem{^{14}C}\,cal\,kyr\,BP} \aref.
 
@@
 (Fig.~\ref{fig:timeseries}, lower panel). Simulations driven by the NGRIP,
-ODP~1012 and ODP~1020 records, on the opposite, produce a last glacial maximum
-that pre-dates reconstructed ages by several thousand years. Concerning
+ODP~1012 and ODP~1020 records, on the contrary, produce a last glacial maximum
+that pre-dates reconstructed ages by several thousand years. Concerning the
 simulations driven by oceanic (ODP~1012, ODP~1020) records, this early
-deglaciation is caused by an early warming in the palaeo-temperature records
+deglaciation is caused by an early warming sealed in the palaeo-temperature records
 (Fig.~\ref{fig:timeseries}, upper panel). However, it has been shown that this
@@
 presence of the Cordilleran ice sheet, resulting in paradoxically warmer
-sea-surface temperatures at the location of the ODP~1012 and ODP~1020 sites
+sea-surface temperatures at the locations of the ODP~1012 and ODP~1020 sites
 during the last glacial maximum \aref.
@@
 
-These results makes us believe that a palaeo-temperature proxy record located
+These results make us believe that a palaeo-temperature proxy record located
 in the direct vicinity of the former Cordilleran ice sheet would be of great
@@
 record yet exists, we focus the rest of our analysis on simulations forced by
-temperature records from a Greenland (GRIP) and an Antarctic (EPICA) ice cores.
+temperature records from Greenland (GRIP) and Antarctic (EPICA) ice cores.
 
@@
 
-Although snapshots from the model output (Fig.~\ref{fig:duration}) give a
+Although snapshots from the model output (Fig.~\ref{fig:snapshots}) give a
 crisp picture of modelled ice dynamics at specific time frames, they often
@@
 ice caps centred on major mountain ranges of the North American Cordillera
-(Fig.~\ref{fig:duration}). In both the GRIP and the EPICA simulation,
-a~contiguous ice cover spanning from the Alaska Range to the Rocky Mountains
-exists for about 29\,\unit{kyr}. This value represent cumulative ice cover over
-the two glaciations during MIS~2 and MIS~4, while a dynamically consistent
-Cordilleran ice sheet forms for even shorter time spans.
+(Fig.~\ref{fig:duration}). In both simulations driven by the GRIP and EPICA
+records, a~contiguous ice cover spanning from the Alaska Range to the Rocky
+Mountains exists for about 29\,\unit{kyr} in total over MIS~2 and~4.
 
-With exception of the oceanic front and the northern foothills of the Alaska
+With the exception of the oceanic front and the northern foothills of the Alaska
 Range, most of the maximal extent of the ice sheet is associated with short
@@
 most extensive stages of the Cordilleran ice sheet are both short-lived and
-out of balance with contemporary climate states, a result which confirms
+out of balance with contemporary climate states, which confirms
 inferences from earlier geological studies \aref.
@@
 However, our simulations depict, independently of the choice of record used for
-palaeo-temperature forcing, a central ice cap persisting throughout the entire
-last glacial cycle on the Skeena mountains (Fig.~\ref{fig:snapshots} and
+palaeo-temperature forcing, a central Skeena mountain ice cap persisting
+throughout the entire last glacial cycle (Fig.~\ref{fig:snapshots} and
 \ref{fig:duration}). In particular, this ice cap persists between the MIS~4
@@
 During time periods when a fully-grown Cordilleran ice sheet is absent from our
-simulation, the presence of the Skeena ice cap can be explained by more
+simulation, the presence of the Skeena ice cap can be explained by a more
 widespread winter precipitation there than in other parts of the modelling
-domain (Figs.~\ref{fig:atm}). Along most of the north-western North American
-shore, coastal mountain ranges provide a sharp orographic obstacle to westerly
+domain (Fig.~\ref{fig:atm}). Along most of the north-western North American
+shore, coastal mountain ranges provide a sharp orographic barrier to westerly
 winds, capturing atmospheric moisture in the form of precipitation, and leaving
-interior lowlands arid. However, near the centre of our modelling domain, the
-topographic barrier formed by the Coast Mountains is less pronounced than
+interior lowlands arid. However, near the centre of our modelling domain, this
+barrier is less pronounced than
 elsewhere, allowing westerly winds to carry moisture further inland, until it
@@
 Cordilleran ice sheet \aref. Here we suggest that persistent ice cover
-associated with basal temperatures at the ice melting point
+associated with basal ice temperatures at the pressure-melting point
 (Fig.~\ref{fig:warmbase}) may partly explain this spectacular landscape.
@@
 about 17 and 11\,\unit{kyr}. That driven by the GRIP record reaches maximal
-volume at 19.1\,\unit{kyr}, and produce a deglaciation in three steps,
-including a first stand-still between 16.6 and 14.5\,\unit{kyr}, and a late
+volume at 19.1\,\unit{kyr}, and produces a deglaciation in three steps,
+including the first stand-still between 16.6 and 14.5\,\unit{kyr}, and a late
 glacial re-advance between 12.6 and 11.6\,\unit{kyr}.
 
-Although the timing differs, the patterns of glacier retreat are relatively
-consistent between the two simulations (Figs.~\ref{fig:deglac}). In both
-simulations, the southern half of the modelling domain, including the Puget
+\todo{discuss the work that supports YD re-advance.}
+
+Although the predicted timing of glacier retreat differs, its patterns are
+relatively consistent between the two simulations (Fig.~\ref{fig:deglac}). In
+both runs, the southern half of the modelling domain, including the Puget
 Lowland, the Coast and Rocky Mountains and the Interior Plateau of British
-Columbia, are completely deglaciated at a time where there remain significant
-ice cover on the Skeena, Selwyn, MacKenzie, Wrangell and Saint-Elias Mountains
+Columbia, are completely deglaciated by 12\,\unit{kyr}, whereas significant
+ice cover remains on the Skeena, Selwyn, MacKenzie, Wrangell and Saint-Elias Mountains
 in the northern half of the domain. At this stage, a late glacial re-advance
@@
 Liard Lowland, followed by ice retreat towards the Selwyn Mountains and the
-Skeena Mountains, which host the last remain of the palaeo-ice sheet.
+Skeena Mountains, which host the last remains of the palaeo-ice sheet.
 
-On the Interior Plateau of British Columbia, modelled deglaciation is rapid
-and northwards, resulting in the presence of a slow-flowing ice lobe on the
+\todo{discuss existing evidence (Martin).}
+
+On the Interior Plateau of British Columbia, modelled deglaciation consists of
+a rapid northwards retreat, leaving a slow-flowing ice lobe on the
 plateau, while the surrounding mountain ranges are already deglaciated
@@
 MIS~4 (61.9--55.4\,\unit{kyr}) and MIS~2 (29.5--16.9\,\unit{kyr}), two periods
-corresponding to documented palaeo-glaciation. This result is independent
-of the choice of palaeo-temperature record used for climate forcing, and
+corresponding to documented extensive palaeo-glaciations. This result is
+independent of the choice of palaeo-temperature record used to approximate the
+past evolution of climate forcing, and
 can be seen as a first-order agreement between the model and geomorphological
-evidence. However, the timing of the two glaciation peaks highly depends on
-which record
+evidence. However, the timing of the two glaciation peaks depends sensitively
+on which record
 is used to drive the model. The timing of the last glacial maximum is best
 reproduced by the Antarctic record, and occurs too early in all simulations
-that were driven by other records. The mismatch is greatest when forcing the
-model with oceanic records from the Pacific Northwest, which are affected by a
+that are driven by other records. The mismatch is greatest when using
+oceanic records from the Pacific Northwest, which are affected by the
 weakening of the California current during the last glacial maximum.
@@
 simulations produce persistent ice cover over the Skeena mountains during
-the entire glacial cycle. At time when the Cordilleran ice sheet is absent
-the Skeena ice cap appears to be fed by an eastwards precipitation
+the entire glacial cycle. At the time when a full-size Cordilleran ice sheet is
+absent, the Skeena ice cap appears to be fed by the eastwards precipitation
 intrusion through a topographic window in the Coast Mountains. The ice cap acts
@@
 Concerning the deglaciation phase, when most geological evidence is available,
-none of the palaeo-temperature records used produce an ideal agreement between
-model results and geological evidence. The
-timing of the last glacial maximum and early deglaciation is generally best
-reproduced by forcing the model with an Antarctic
-record. However, the Younger Dryas re-advance is best reproduced at places
-where it has been documented using temperature forcing from the GRIP record.
-Nonetheless, the patterns of deglaciation are consistent between different
-simulations. They show a rapid deglaciation of the southern half of the ice
+none of the palaeo-temperature records used produces an ideal agreement between
+model results and geological evidence. The timing of the last glacial maximum
+and early deglaciation is generally best reproduced by forcing the model with
+Antarctic records. However, the Younger Dryas re-advance is only reproduced
+using temperature forcing from the GRIP record at places where it has been
+documented. Nonetheless, the patterns of deglaciation are consistent between
+both simulations driven by the GRIP and EPICA record. They show a rapid
+deglaciation of the southern half of the ice
 sheet, followed by unzipping of the Liard Lowland, and concentric retreat of
@@
 One must keep in mind, however, that these results are only accurate for our
-choices of ice-sheet model (PISM), surface mass balance model (PDD), study
-object (the Cordilleran ice sheet). Most importantly, our simplistic
+choices of ice-sheet model (PISM), surface mass balance model (PDD) and climate
+forcing. Most importantly, our simplistic
 palaeo-climate forcing does not include precipitation corrections in response
-to the presence of the ice sheet, likely leading to overestimates of glacial
-extent and volume. Nevertheless, our results identify the largely understudied
-Skeena Mountains as a key area to understanding glacial dynamics of the
-Cordilleran ice sheet, highlighting the need for further geological
-investigation in this region.
+to the presence of an ice cover, likely leading to overestimated glacial
+extent and volume. Nevertheless, our results identify that the largely
+understudied Skeena Mountains are a key area to understanding glacial dynamics
+of the Cordilleran ice sheet, highlighting the need for further geological
+investigation of this region.
 
@@
            120\,\unit{kyr}. Long ice cover durations combined with basal
-           temperatures at the melting point may explain the strong glacial
-           erosional imprint of the Skeena Mountains (SM) landscape. Hatches
-           indicate areas that were covered by cold ice only.
+           temperatures at the pressure-melting point may explain the strong
+           glacial erosional imprint of the Skeena Mountains (SM) landscape.
+           Hatches indicate areas that were covered by cold ice only.
            \note{indicate location of Skeena Mountains.}}
@@
            advance exceeded the last glacial maximum advanced are marked in
-           green. Hatches denote the Younger Dryas re-advance, which is most
-           significant in the GRIP-driven simulation.}
+           green. Hatches denote the Younger Dryas re-advance, which is more
+           pronounced in the GRIP-driven simulation.}
   \label{fig:deglac}

### 2014-06-13 cordillera-cycle.tex
Update to new Copernicus class.
Remove unit command where useless.

@@
 % Copernicus manuscript
-\documentclass[tc, ms]{copernicus}
+\documentclass[tc, manuscript]{copernicus}
 
@@
 % Copernicus discussion paper
-%\documentclass[tcd, hvmath]{copernicus_discussions}
+%\documentclass[tcd, hvmath, online]{copernicus}
 
@@
 % Coloured hyperlinks
-\usepackage[colorlinks]{hyperref}
-
-% Encoding
-\usepackage[T1]{fontenc}
+\hypersetup{colorlinks}
 
@@
 % ----------------------------------------------------------------------
-\begin{document}\hack{\sloppy}
+\begin{document}
+\linenumbers
 % ----------------------------------------------------------------------
@@
 % Authors
-\author[1,2]{J.~Seguinot}
-\author[3]{M.~Margold}
-\author[2]{I.~Rogozhina}
-\author[1]{A.~Stroeven}
+\Author[1,2]{J.}{Seguinot}
+\Author[3]{M.}{Margold}
+\Author[2]{I.}{Rogozhina}
+\Author[1]{A.~P.}{Stroeven}
 \runningauthor{J.~Seguinot et~al.}
@@
 \received{}
+\pubdiscuss{}
+\revised{}
 \accepted{}
@@
   forcing, all simulations produce two major glaciation events during
-  marine isotope stages~4 (61.9--55.4\,\unit{kyr}) and~2
-  (29.5--16.9\,\unit{kyr}). The timing of glaciation is
+  marine isotope stages~4 (61.9--55.4\,kyr) and~2
+  (29.5--16.9\,kyr). The timing of glaciation is
   better reproduced using temperature reconstructions from Greenland and
@@
   and hosts the last remains of Cordilleran ice during the
-  early Holocene (10.9--9.5\,\unit{kyr}).
+  early Holocene (10.9--9.5\,kyr).
 
@@
 For all proxy records, we firstly run a low-resolution simulation, using
-a~10\,\unit{km}, 300 by 150~points horizontal grid for SSA and SIA velocities,
+a~10\,km, 300 by 150~points horizontal grid for SSA and SIA velocities,
 up to 51~enthalpy layers in the ice, and 31~temperature layers in the bedrock.
 For the GRIP and EPICA records, we then run a second simulation using a higher
-horizontal resolution of 6\,\unit{km}, up to 101~ice layers, and 61~bedrock
+horizontal resolution of 6\,km, up to 101~ice layers, and 61~bedrock
 layers. In the rest of this paper, we refer to different model runs by the
@@
 during most of the glacial cycle, while two major glaciation events occur
-during Marine Isotope Stage (MIS)~4 (61.9--55.4\,\unit{kyr}) and MIS~2
-(29.5--16.9\,\unit{kyr}, Fig.~\ref{fig:timeseries}). A local ice minimum is
-consistently attained during MIS~3 (52.2--45.6\,\unit{kyr}). However, the
+during Marine Isotope Stage (MIS)~4 (61.9--55.4\,kyr) and MIS~2
+(29.5--16.9\,kyr, Fig.~\ref{fig:timeseries}). A local ice minimum is
+consistently attained during MIS~3 (52.2--45.6\,kyr). However, the
 magnitude and timing of these events depend significantly on the choice of
@@
 While simulations driven by the Antarctic (EPICA, Vostok) and GRIP records
-reach a last glacial maximum in ice volume between 19.5 and 16.9\,\unit{kyr},
+reach a last glacial maximum in ice volume between 19.5 and 16.9\,kyr,
 those steered by oceanic (ODP~1012, ODP~1020) and the NGRIP record attain it
 several thousands of years earlier. Moreover, the ODP~1012 run yields rapid
-deglaciation prior to 20\,\unit{kyr} due to a local temperature maximum that is
+deglaciation prior to 20\,kyr due to a local temperature maximum that is
 inconsistent with all other records used. The ODP~1020 simulation predicts an
-early last glacial maximum in ice volume at 29.5\,\unit{kyr}, followed by
+early last glacial maximum in ice volume at 29.5\,kyr, followed by
 slower deglaciation than modelled using other records. Finally, model runs
@@
 simulation to the next. It includes a central divide located at about
-3\,500\,\unit{m} elevation along the spine of the Rocky Mountains. In all
+3\,500\,m elevation along the spine of the Rocky Mountains. In all
 simulations, a central ice cap persists over the Skeena mountains during MIS~3
@@
 time-series for the ice sheet model, our simulations consistently produce two
-glaciation maxima during MIS~4 (61.9--55.4\,\unit{kyr}) and MIS~2
-(29.5--16.9\,\unit{kyr}). These events broadly correspond in timing to the
+glaciation maxima during MIS~4 (61.9--55.4\,kyr) and MIS~2
+(29.5--16.9\,kyr). These events broadly correspond in timing to the
 Gladstone (MIS~4) and McConell (MIS~2) glaciations documented by geological
@@
 records, a~contiguous ice cover spanning from the Alaska Range to the Rocky
-Mountains exists for about 29\,\unit{kyr} in total over MIS~2 and~4.
+Mountains exists for about 29\,kyr in total over MIS~2 and~4.
 
@@
 the model (Figs.~\ref{fig:timeseries} and ~\ref{fig:deglacseries}). The
-6\,\unit{km}-resolution simulation forced by the EPICA record reaches a maximal
-ice volume at 17.3\,\unit{kyr}, followed by a continuous deglaciation between
-about 17 and 11\,\unit{kyr}. That driven by the GRIP record reaches maximal
-volume at 19.1\,\unit{kyr}, and produces a deglaciation in three steps,
-including the first stand-still between 16.6 and 14.5\,\unit{kyr}, and a late
-glacial re-advance between 12.6 and 11.6\,\unit{kyr}.
+6\,km-resolution simulation forced by the EPICA record reaches a maximal
+ice volume at 17.3\,kyr, followed by a continuous deglaciation between
+about 17 and 11\,kyr. That driven by the GRIP record reaches maximal
+volume at 19.1\,kyr, and produces a deglaciation in three steps,
+including the first stand-still between 16.6 and 14.5\,kyr, and a late
+glacial re-advance between 12.6 and 11.6\,kyr.
 
@@
 Lowland, the Coast and Rocky Mountains and the Interior Plateau of British
-Columbia, are completely deglaciated by 12\,\unit{kyr}, whereas significant
+Columbia, are completely deglaciated by 12\,kyr, whereas significant
 ice cover remains on the Skeena, Selwyn, MacKenzie, Wrangell and Saint-Elias Mountains
@@
 cycle presented in this study consistently produce two glaciation maxima during
-MIS~4 (61.9--55.4\,\unit{kyr}) and MIS~2 (29.5--16.9\,\unit{kyr}), two periods
+MIS~4 (61.9--55.4\,kyr) and MIS~2 (29.5--16.9\,kyr), two periods
 corresponding to documented extensive palaeo-glaciations. This result is
@@
 the ice margin towards the last palaeo-ice cover in the Skeena Mountains
-during the early Holocene (10.9--9.5\,\unit{kyr}).
+during the early Holocene (10.9--9.5\,kyr).
 
@@
 % References
-%\input{references.bbl}
+%\bibliographystyle{copernicus}
+%\bibliography{references.bib}
 \newpage
@@
            prepare temperature offset time-series used to force the ice sheet
-           model through the last 120\,\unit{kyr}. $T_{[32;22]}$ refers to the
-           mean temperature anomaly during the period -32 to~-22~\unit{kyr} after
+           model through the last 120\,kyr. $T_{[32;22]}$ refers to the
+           mean temperature anomaly during the period -32 to~-22~kyr after
            scaling.}
@@
 % fig:timeseries
-\begin{figure}
+\begin{figure*}
   \includegraphics{timeseries}
@@
            the ice sheet model \textbf{(top)}, and modelled ice volume
-           through the last 120\,\unit{kyr}, expressed in meters of sea-level
+           through the last 120\,kyr, expressed in meters of sea-level
            equivalent \textbf{(bottom)}. Gray spans indicate Marine Isotope
@@
            rectangles highlight modelled ice volume extrema corresponding to
-           MIS~4 (61.9--55.4\,\unit{kyr}), MIS~3 (52.2--45.6\,\unit{kyr}), and
-           MIS~2 (last glacial maximum, 29.5--16.9\,\unit{kyr}). Dotted lines
-           correspond to the GRIP and EPICA 6\,\unit{km}-resolution runs.
+           MIS~4 (61.9--55.4\,kyr), MIS~3 (52.2--45.6\,kyr), and
+           MIS~2 (last glacial maximum, 29.5--16.9\,kyr). Dotted lines
+           correspond to the GRIP and EPICA 6\,km-resolution runs.
            \todo{add grid lines to improve readability}}
   \label{fig:timeseries}
-\end{figure}
+\end{figure*}
 
 % fig:snapshots
-\begin{figure}
+\begin{figure*}
   \includegraphics{snapshots}
-  \caption{Snapshots of modelled surface topography (500\,\unit{m} contours)
+  \caption{Snapshots of modelled surface topography (500\,m contours)
            corresponding to the ice volume extrema indicated on
@@
   \label{fig:snapshots}
-\end{figure}
+\end{figure*}
 
 % fig:duration
-\begin{figure}
+\begin{figure*}
   \includegraphics{duration}
-  \caption{Modelled duration of ice cover during the last 120\,\unit{kyr}.
+  \caption{Modelled duration of ice cover during the last 120\,kyr.
            Note the irregular colour scale. A contiguous ice cover spanning
            from the Alaska Range (AR) to the southern Coast Mountains (CM) and
-           Rocky Mountains (RM) exists for about 29\,\unit{kyr} in both
+           Rocky Mountains (RM) exists for about 29\,kyr in both
            simulations. A central
@@
   \label{fig:duration}
-\end{figure}
+\end{figure*}
 
 % fig:warmbase
-\begin{figure}
+\begin{figure*}
   \includegraphics{warmbase}
   \caption{Modelled duration of warm-based ice cover during the last
-           120\,\unit{kyr}. Long ice cover durations combined with basal
+           120\,kyr. Long ice cover durations combined with basal
            temperatures at the pressure-melting point may explain the strong
@@
   \label{fig:warmbase}
-\end{figure}
+\end{figure*}
 
 % fig:erosion
-\begin{figure}
+\begin{figure*}
   \includegraphics{erosion}
   \caption{Modelled cumulative basal displacement (integrand of basal velocity)
-           over the last 120\,\unit{kyr}.
+           over the last 120\,kyr.
            \note{I am still unsure about the use of this figure. I may
@@
   \label{fig:erosion}
-\end{figure}
+\end{figure*}
 
@@
 % fig:deglac
-\begin{figure}
+\begin{figure*}
   \includegraphics{deglac}
@@
   \label{fig:deglac}
-\end{figure}
+\end{figure*}
 
 % fig:deglacshots
-\begin{figure}
+\begin{figure*}
   \includegraphics{deglacshots}
-  \caption{Snapshots of modelled surface topography (200\,\unit{m} contours)
+  \caption{Snapshots of modelled surface topography (200\,m contours)
            and surface velocity (colour mapping) from the GRIP simulation,
@@
   \label{fig:deglacshots}
-\end{figure}
+\end{figure*}
 
@@
   \caption{Modelled bedrock (black) and ice surface (blue) topography profiles
-           during deglaciation (22.0--8.0\,\unit{kyr}) in the GRIP 6\,\unit{km}
+           during deglaciation (22.0--8.0\,kyr) in the GRIP 6\,km
            simulation, corresponding to the four transects indicated in
@@
 % fig:lastflow
-\begin{figure}
+\begin{figure*}
   \includegraphics{lastflow}
@@
   \label{fig:lastflow}
-\end{figure}
+\end{figure*}
 

### 2014-06-13 cordillera-cycle.tex
Begin to insert references.

@@
 \newcommand{\aref}[0]{\textbf{[ref.]}}
-\renewcommand{\citep}[1]{\aref}
-\renewcommand{\citet}[1]{\aref}
+%\renewcommand{\citep}[1]{\aref}
+%\renewcommand{\citet}[1]{\aref}
 
@@
 The Shallow Shelf Approximation (SSA) is used as a ``sliding law'' for the
-Shallow Ice Approximation (SIA) \citep{bueler-brown-2009,winkelmann-etal-2011}.
+Shallow Ice Approximation (SIA) \citep{Bueler.Brown.2009,Winkelmann.etal.2011}.
 Ice softness depends on temperature and water content through an enthalpy
-formulation \citep{aschwanden-blatter-2009,aschwanden-etal-2012}. Surface air
+formulation \citep{Aschwanden.Blatter.2009,Aschwanden.etal.2012}. Surface air
 temperature from the atmospheric forcing provides the upper boundary condition
@@
 average representative of available heat flow measurements
-\citep{artemieva-mooney-2001,blackwell-richards-2004}.
+\citep{Artemieva.Mooney.2001,Blackwell.Richards.2004}.
 
-A~pseudo-plastic sliding law \citep{aschwanden-etal-2013} relates the
+A~pseudo-plastic sliding law \citep{Aschwanden.etal.2013} relates the
 bed-parallel shear stress and the sliding velocity. The yield stress is
@@
 for a~weakening of till associated with the presence of marine sediments
-\citep{martin-etal-2011,aschwanden-etal-2013}. The till effective pressure is
+\citep{Martin.etal.2011,Aschwanden.etal.2013}. The till effective pressure is
 related to the modelled amount of water in the till, using a formula derived
@@
 \aref. Basal topography is derived from the ETOPO1 combined topography
-and bathymetry dataset with a~resolution of 1\,arc-min \citep{data:etopo1}.
+and bathymetry dataset with a~resolution of 1\,arc-min \citep{Amante.Eakins.2009}.
 Sea level is lowered as a function of time according to the SPECMAP
@@
 elastic lithosphere flexure and viscous mantle deformation in a~semi-infinite
-half-space \citep{lingle-clark-1985,bueler-etal-2007}.
+half-space \citep{Lingle.Clark.1985,Bueler.etal.2007}.
 
@@
 deviation of near-surface temperature by a~temperature-index (positive
-degree-day) model \citep{hock-2003}. Ice accumulation is equal to precipitation
+degree-day) model \citep{Hock.2003}. Ice accumulation is equal to precipitation
 when temperature is below 0\,\unit{{\degree}C}, and decreases to zero linearly
@@
 
-The positive degree-day integral \citep{calov-greve-2005} is numerically
+The positive degree-day integral \citep{Calov.Greve.2005} is numerically
 approximated using week-long sub-intervals. It accounts for temperature
@@
 computed from daily temperature values from the North American Regional
-Reanalysis (NARR) \citep{data:narr}, after excluding variability associated
+Reanalysis (NARR) \citep{Mesinger.etal.2006}, after excluding variability associated
 with the
-seasonal cycle itself \citet{seguinot-rogozhina-2014}. The
+seasonal cycle itself \citet{Seguinot.Rogozhina.2014}. The
 ablation model incorporates degree-day factors of
@@
 mass-balance measurements on contemporary glaciers from the Coast Mountains and
-Rocky Mountains in British Columbia \citep{shea-etal-2009}.
+Rocky Mountains in British Columbia \citep{Shea.etal.2009}.
 
@@
 % References
-%\bibliographystyle{copernicus}
-%\bibliography{references.bib}
+\bibliographystyle{copernicus}
+\bibliography{refs/references.bib}
 \newpage
@@
            18\,\unit{\chem{^{14}C}\,kyr\,BP} (21.4\,cal\,kyr\,BP)
-           \citep{dyke-2004}. The rectangular box denotes the location of the
+           \citep{Dyke.2004}. The rectangular box denotes the location of the
            modelling domain used in this study. Major mountain ranges covered
@@
            (RM) and the North Cascades (NC). The background
-           map consists of ETOPO1 \citep{data:etopo1} and Natural Earth Data
-           \citep{data:naturalearth}.
+           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth Data
+           \citep{web:naturalearth}.
            \todo{Mark palaeo-ice sheets and mountain ranges on the map.}}
@@
            Stages (MIS) according to a global compilation of benthic
-           \chem{\delta^{18}O} records \citep{lisiecki-raymo-2005}. Hatched
+           \chem{\delta^{18}O} records \citep{Lisiecki.Raymo.2005}. Hatched
            rectangles highlight modelled ice volume extrema corresponding to

### 2014-06-16 cordillera-cycle.tex
Update references.
Begin to reference the introduction.

@@
 % Coloured hyperlinks
-\hypersetup{colorlinks}
+\hypersetup{colorlinks, citecolor=blue}
 
@@
 \newcommand{\aref}[0]{\textbf{[ref.]}}
-%\renewcommand{\citep}[1]{\aref}
-%\renewcommand{\citet}[1]{\aref}
 
@@
 contiguous blanket of ice extended from the Alaska Range in the north to the
-North Cascades in the south \aref. This ice mass is known as the former
-Cordilleran ice sheet (Fig.~\ref{fig:locmap}).
+North Cascades in the south (Fig.~\ref{fig:locmap}). This ice mass is known as
+the former Cordilleran ice sheet \citep{Dawson.1888}.
 
@@
 Cordillera have led to a large collection of evidence of the former ice cover
-\aref. This evidence consists of mapped boundaries of the former ice extent
-\aref, directions of past sliding velocities {\aref} and locations of former
-melt-water streams \aref, as well as the timing of glaciation identified from
-radiocarbon dating \aref, cosmogenic dating \aref, and the offshore sedimental
-record \aref.
+\citep{Jackson.Clague.1991}. This evidence consists of mapped boundaries of the
+former ice extent \citep{Clague.1989, Dyke.2004}, directions of past sliding
+velocities \citep{Prest.etal.1968, Kleman.etal.2010}, and locations of former
+melt-water streams \citep{Margold.etal.2011, Margold.etal.2013}, as well as the
+timing of glaciation identified from radiocarbon dating \citep[e.g.,][]
+{Clague.etal.1980, Clague.1981, Porter.Swanson.1998}, cosmogenic dating
+\citep[e.g.,][]{Ward.etal.2007, Stroeven.etal.2010, Stroeven.etal.2013}, and
+the offshore sedimental record \citep{Cosma.etal.2008, Davies.etal.2011}.
 
 Field-based evidence allowed to reconstruct, with a high degree of confidence,
-the maximal extent attained by the Cordilleran ice sheet during the last glacial
-cycle \aref. However, former ice thickness and the ice sheet's contribution to the
-last glacial maximum sea-level low-stand remain poorly constrained. Moreover,
-our understanding of the glaciation history is mainly restricted to the
-deglaciation phase and a regional scale, while little is known about the ice
-sheet evolution prior to the last glacial maximum extent \aref. Although
-time-evolving, whole ice-sheet reconstructions of glacial advance and retreat
-patterns are available for the Laurentide and Eurasian ice sheets \aref, such is not
-the case for the Cordilleran ice sheet, where complex arrangements of ice flow
-directions emanating from multiple glaciation centres have been identified, but
-remain poorly understood \aref.
+the maximal extent attained by the Cordilleran ice sheet during the last
+glacial cycle \citep{Clague.1989}. However, former ice thickness and the ice
+sheet's contribution to the last glacial maximum sea-level low-stand remain
+poorly constrained. Moreover, our understanding of the glaciation history is
+mainly restricted to the deglaciation phase and a regional scale, while little
+is known about the ice sheet evolution prior to the last glacial maximum extent
+\citep{Kleman.etal.2010}. Although time-evolving, whole ice-sheet
+reconstructions of glacial advance and retreat patterns are available for the
+Eurasian and Laurentide and ice sheets \citep{Kleman.etal.1997,
+Kleman.etal.2010}, such is not the case for the Cordilleran ice sheet, where
+complex arrangements of ice flow directions emanating from multiple glaciation
+centres have been identified \citep{Prest.etal.1968, Clague.1989}, but remain
+poorly understood \citep{Kleman.etal.2010}.
 
-The present study uses a numerical ice sheet model (PISM, \aref) calibrated against
+The present study uses a numerical ice sheet model \citep{PISM-authors.2014},
+calibrated against
 field-based evidence to perform a quantitative reconstruction of the
@@
 numerical modelling has been established as a useful tool to improve our
-understanding
-of the Cordilleran ice sheet more than twenty years ago \aref, the ubiquitously
+understanding of the Cordilleran ice sheet more than twenty years ago
+\citep{Jackson.Clague.1991,Robert.1991}, the ubiquitously
 mountainous topography of the region has presented a major challenge to its
-application, which has only been overcome by recent development of numerical ice
-sheet models and underlying scientific computing tools \aref. In addition to the topographic
+application, which can only be overcome by recent development of numerical ice
+sheet models and underlying scientific computing tools \aref.
+In addition to the topographic
 complexity, strong climatic contrasts characteristic of the North American
 Cordillera require the use of a high-resolution temperature and precipitation
-forcing as input data to the ice sheet model \aref.
+forcing as input data to the ice sheet model \citep{Seguinot.etal.2013}.
 
@@
 and precipitation fields derived from a combination of a present-day
-atmospheric reanalysis (NARR, \aref), lapse-rate corrections,
+atmospheric reanalysis \citep{Mesinger.etal.2006}, lapse-rate corrections,
 and temperature offset time series. The latter are
@@
            (RM) and the North Cascades (NC). The background
-           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth Data
-           \citep{web:naturalearth}.
+           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
+           Data \citep{Patterson.Kelso.2014}.
            \todo{Mark palaeo-ice sheets and mountain ranges on the map.}}

### 2014-06-16 cordillera-cycle.tex
Finish to reference the introduction.

@@
 contiguous blanket of ice extended from the Alaska Range in the north to the
-North Cascades in the south (Fig.~\ref{fig:locmap}). This ice mass is known as
-the former Cordilleran ice sheet \citep{Dawson.1888}.
+North Cascades in the south \citep[Fig.~\ref{fig:locmap};][Plate XIV]
+{Chamberlin.1894}. This ice mass is known as the former Cordilleran ice sheet
+\citep{Dawson.1888}.
 
@@
 the maximal extent attained by the Cordilleran ice sheet during the last
-glacial cycle \citep{Clague.1989}. However, former ice thickness and the ice
-sheet's contribution to the last glacial maximum sea-level low-stand remain
-poorly constrained. Moreover, our understanding of the glaciation history is
-mainly restricted to the deglaciation phase and a regional scale, while little
-is known about the ice sheet evolution prior to the last glacial maximum extent
-\citep{Kleman.etal.2010}. Although time-evolving, whole ice-sheet
-reconstructions of glacial advance and retreat patterns are available for the
-Eurasian and Laurentide and ice sheets \citep{Kleman.etal.1997,
-Kleman.etal.2010}, such is not the case for the Cordilleran ice sheet, where
-complex arrangements of ice flow directions emanating from multiple glaciation
-centres have been identified \citep{Prest.etal.1968, Clague.1989}, but remain
-poorly understood \citep{Kleman.etal.2010}.
+glacial cycle \citep{Clague.1989, Dyke.2004}. However, former ice thickness
+and the ice sheet's contribution to the last glacial maximum sea-level
+low-stand remain poorly constrained. Moreover, our understanding of the
+glaciation history is mainly restricted to the deglaciation phase and a
+regional scale, while little is known about the ice sheet evolution prior to
+the last glacial maximum extent \citep{Kleman.etal.2010}. Although
+time-evolving, whole ice-sheet reconstructions of glacial advance and retreat
+patterns are available for the Eurasian and Laurentide and ice sheets
+\citep{Kleman.etal.1997, Kleman.etal.2010}, such is not the case for the
+Cordilleran ice sheet, where complex arrangements of ice flow directions
+emanating from multiple glaciation centres have been identified
+\citep{Prest.etal.1968, Clague.1989}, but remain poorly understood
+\citep{Kleman.etal.2010}.
 
 The present study uses a numerical ice sheet model \citep{PISM-authors.2014},
-calibrated against
-field-based evidence to perform a quantitative reconstruction of the
-Cordilleran ice sheet history through the last glacial cycle. Although
-numerical modelling has been established as a useful tool to improve our
-understanding of the Cordilleran ice sheet more than twenty years ago
-\citep{Jackson.Clague.1991,Robert.1991}, the ubiquitously
-mountainous topography of the region has presented a major challenge to its
-application, which can only be overcome by recent development of numerical ice
-sheet models and underlying scientific computing tools \aref.
-In addition to the topographic
-complexity, strong climatic contrasts characteristic of the North American
-Cordillera require the use of a high-resolution temperature and precipitation
-forcing as input data to the ice sheet model \citep{Seguinot.etal.2013}.
+calibrated against field-based evidence to perform a quantitative
+reconstruction of the Cordilleran ice sheet history through the last glacial
+cycle. Although numerical modelling has been established as a useful tool to
+improve our understanding of the Cordilleran ice sheet more than twenty years
+ago \citep{Jackson.Clague.1991,Robert.1991}, the ubiquitously mountainous
+topography of the region has presented a major challenge to its application.
+In fact, only recent development of numerical ice sheet models and underlying
+scientific computing tools \citep[e.g.,][]{Bueler.Brown.2009} allows
+high-resolution numerical modelling of palaeo-glacier flow on mountainous
+terrain over millenial time scales \citep{Golledge.etal.2012}. In the North
+American Cordillera, the topographic complexity induces strong gradients in
+seasonality and distinct patterns of precipition, thus requiring the use of
+high-resolution atmospheric forcing fields as input to the ice sheet model
+\citep{Seguinot.etal.2013}.
 
@@
 atmospheric reanalysis \citep{Mesinger.etal.2006}, lapse-rate corrections,
-and temperature offset time series. The latter are
-obtained by scaling six different palaeo-temperature reconstructions from
-proxy records around the globe, including two \chem{\delta^{18}O} records from
-Greenland ice cores \aref, two \chem{\delta^{18}O} records from Antarctic ice cores \aref,
+and temperature offset time series. The latter are obtained by scaling six
+different palaeo-temperature reconstructions from proxy records around the
+globe, including two \chem{\delta^{18}O} records from Greenland ice cores
+\citep{Dansgaard.etal.1993, Andersen.etal.2004}, two \chem{\delta^{18}O}
+records from Antarctic ice cores \citep{Petit.etal.1999,Jouzel.etal.2007},
 and two alkenone unsaturation index records from Northwest Pacific oceanic
-sediment cores \aref. Model output is compared to geomorphological evidence in terms
-of timing and extent of glaciation and patterns of deglaciation.
+sediment cores \citep{Herbert.2001}. Model output is compared to
+geomorphological evidence in terms of timing and extent of glaciation and
+patterns of deglaciation.
 

### 2014-06-16 cordillera-cycle.tex
Add missing references in model setup.

@@
 development version~11b0a7f), an open-source, finite-difference, shallow ice
-sheet model \aref. The model inputs basal topography, sea level, geothermal
+sheet model \citep{PISM-authors.2014}. The model inputs basal topography,
+sea level, geothermal
 heat flux and climate forcing, and computes the evolution of ice extent
@@
 average representative of available heat flow measurements
-\citep{Artemieva.Mooney.2001,Blackwell.Richards.2004}.
+\citep{Blackwell.Richards.2004}.
 
@@
 elevation, with lowest values occurring at low elevation, thus accounting
-for a~weakening of till associated with the presence of marine sediments
-\citep{Martin.etal.2011,Aschwanden.etal.2013}. The till effective pressure is
-related to the modelled amount of water in the till, using a formula derived
-from laboratory experiments with till extracted from an Antarctic ice stream
-\aref. Basal topography is derived from the ETOPO1 combined topography
-and bathymetry dataset with a~resolution of 1\,arc-min \citep{Amante.Eakins.2009}.
+for a~weakening of till associated with the presence of marine sediments.
+The till effective pressure is related to the modelled amount of water in the
+till, using a formula derived from laboratory experiments with till extracted
+from an Antarctic ice stream \citep{Tulaczyk.etal.2000}. Basal topography is
+derived from the ETOPO1 combined topography and bathymetry dataset with
+a~resolution of 1\,arc-min \citep{Amante.Eakins.2009}.
 Sea level is lowered as a function of time according to the SPECMAP
@@
 computed from daily temperature values from the North American Regional
-Reanalysis (NARR) \citep{Mesinger.etal.2006}, after excluding variability associated
-with the
-seasonal cycle itself \citet{Seguinot.Rogozhina.2014}. The
+Reanalysis (NARR) \citep{Mesinger.etal.2006}, after excluding variability
+associated with the seasonal cycle itself \citep{Seguinot.Rogozhina.2014}. The
 ablation model incorporates degree-day factors of
@@
 high-resolution precipitation forcing, as outlined in a previous sensitivity
-study \aref.
+study \citep{Seguinot.etal.2013}.
 
@@
 Temperature offset time-series ${\Delta}T_{TS}$ are derived from proxy records from
-the Greenland Ice Core Project (GRIP), the North Greenland Ice Core Project
-(NGRIP), the European Project for Ice Coring in Antarctica (EPICA), the Lake
-Vostok ice core, and Ocean Drilling Program (ODP) sites 1012 and 1020, both
-located off the Californian shore \aref. Palaeo-temperatures from the GRIP and
-NGRIP records were calculated using a quadratic equation \aref,
+the Greenland Ice Core Project \citep[GRIP,][]{Dansgaard.etal.1993}, the
+North Greenland Ice Core Project \citep[NGRIP,][]{Andersen.etal.2004},
+the European Project for Ice Coring in Antarctica \citep[EPICA,][]
+{Jouzel.etal.2007}, the Lake Vostok ice core \citep{Petit.etal.1999}, and Ocean
+Drilling Program (ODP) sites 1012 and 1020, both located off the Californian
+shore \citep{Herbert.2001}. Palaeo-temperatures from the GRIP and NGRIP records
+were calculated using a quadratic equation \citep{Johnsen.etal.1995},
 \begin{equation}
@@
 while the temperature reconstructions from the Antarctic and oceanic cores were
-provided by their authors \aref. All records were scaled linearly in
+provided as such. All records were scaled linearly in
 order to simulate realistic and comparable ice extents at the last
@@
     \tophline
-    Record & Proxy & Scaling factor & $T_{[32;22]}$ & Source\\
+    Record & Proxy & Scaling factor & $T_{[32;22]}$ & Original reference\\
     \middlehline
-    GRIP     & \chem{\delta^{18}O} & ?\% & -5.8{\degree}C & \aref \\
-    NGRIP    & \chem{\delta^{18}O} & ?\% & -6.1{\degree}C & \aref \\
-    EPICA    & \chem{\delta^{18}O} & ?\% & -5.6{\degree}C & \aref \\
-    Vostok   & \chem{\delta^{18}O} & ?\% & -5.6{\degree}C & \aref \\
-    ODP~1012 & \chem{U^{K'}_{37}}  & ?\% & -5.8{\degree}C & \aref \\
-    ODP~1020 & \chem{U^{K'}_{37}}  & ?\% & -5.8{\degree}C & \aref \\
+    GRIP     & \chem{\delta^{18}O} & ?\% & -5.8{\degree}C
+        & \citet{Dansgaard.etal.1993} \\
+    NGRIP    & \chem{\delta^{18}O} & ?\% & -6.1{\degree}C
+        & \citet{Andersen.etal.2004} \\
+    EPICA    & \chem{\delta^{18}O} & ?\% & -5.6{\degree}C
+        & \citet{Jouzel.etal.2007} \\
+    Vostok   & \chem{\delta^{18}O} & ?\% & -5.6{\degree}C
+        & \citet{Petit.etal.1999} \\
+    ODP~1012 & \chem{U^{K'}_{37}}  & ?\% & -5.8{\degree}C
+        & \citet{Herbert.2001} \\
+    ODP~1020 & \chem{U^{K'}_{37}}  & ?\% & -5.8{\degree}C
+        & \citet{Herbert.2001} \\
     \bottomhline

### 2014-06-16 cordillera-cycle.tex
Add some references in the discussion part.

@@
 southern Cordilleran ice sheet margin has been constrained by radiocarbon
-dating between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,kyr\,BP} \aref.
+dating between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,kyr\,BP}
+\citep{Porter.Swanson.1998}.
 These dates have been confirmed by the offshore sediment record, which shows an
 increase of glaciomarine sedimentation between 19.5 and
-16.2\,\unit{\chem{^{14}C}\,cal\,kyr\,BP} \aref.
+16.2\,\unit{\chem{^{14}C}\,cal\,kyr\,BP} \citep{Cosma.etal.2008}.
 
@@
 early warming is a local effect, corresponding to a weakening of the California
-current \aref. The California current, driving cold waters southwards
+current \citep{Herbert.2001}. The California current, driving cold waters southwards
 along the western coast of North America, is thought to have weakened in the
@@
 sea-surface temperatures at the locations of the ODP~1012 and ODP~1020 sites
-during the last glacial maximum \aref.
+during the last glacial maximum \citep{Herbert.2001}.
 
@@
 out of balance with contemporary climate states, which confirms
-inferences from earlier geological studies \aref.
+previous interpretations of the geological record \citep{Clague.etal.1980,
+Stroeven.etal.2010}.
 

### 2014-06-17 cordillera-cycle.tex
Add figure or page number in citations.

@@
 \citep{Jackson.Clague.1991}. This evidence consists of mapped boundaries of the
-former ice extent \citep{Clague.1989, Dyke.2004}, directions of past sliding
-velocities \citep{Prest.etal.1968, Kleman.etal.2010}, and locations of former
+former ice extent (\citealp[Fig.~1.12]{Clague.1989}; \citealp{Duk-Rodkin.1999};
+\citealp{Dyke.2004}), directions of past sliding velocities
+(\citealp{Prest.etal.1968}; \citealp[Fig.~2]{Kleman.etal.2010}), and locations
+of former
 melt-water streams \citep{Margold.etal.2011, Margold.etal.2013}, as well as the
 timing of glaciation identified from radiocarbon dating \citep[e.g.,][]
-{Clague.etal.1980, Clague.1981, Porter.Swanson.1998}, cosmogenic dating
-\citep[e.g.,][]{Ward.etal.2007, Stroeven.etal.2010, Stroeven.etal.2013}, and
-the offshore sedimental record \citep{Cosma.etal.2008, Davies.etal.2011}.
+{Clague.1981, Porter.Swanson.1998}, cosmogenic dating \citep[e.g.,][]
+{Ward.etal.2007, Menounos.etal.2009, Stroeven.etal.2010, Stroeven.etal.2013},
+and the offshore sedimental record \citep{Cosma.etal.2008, Davies.etal.2011}.
 
@@
 the maximal extent attained by the Cordilleran ice sheet during the last
-glacial cycle \citep{Clague.1989, Dyke.2004}. However, former ice thickness
+glacial cycle (\citealp[Fig. 1.12]{Clague.1989}; \citealp{Dyke.2004}). However,
+former ice thickness
 and the ice sheet's contribution to the last glacial maximum sea-level
@@
 regional scale, while little is known about the ice sheet evolution prior to
-the last glacial maximum extent \citep{Kleman.etal.2010}. Although
+the last glacial maximum extent \citep[Fig.~6]{Kleman.etal.2010}. Although
 time-evolving, whole ice-sheet reconstructions of glacial advance and retreat
@@
 emanating from multiple glaciation centres have been identified
-\citep{Prest.etal.1968, Clague.1989}, but remain poorly understood
-\citep{Kleman.etal.2010}.
+\citep[Fig.~1.12]{Prest.etal.1968, Clague.1989}, but remain poorly understood
+\citep[p.~2049]{Kleman.etal.2010}.
 
@@
 improve our understanding of the Cordilleran ice sheet more than twenty years
-ago \citep{Jackson.Clague.1991,Robert.1991}, the ubiquitously mountainous
+ago (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991}), the
+ubiquitously mountainous
 topography of the region has presented a major challenge to its application.
@@
 high-resolution numerical modelling of palaeo-glacier flow on mountainous
-terrain over millenial time scales \citep{Golledge.etal.2012}. In the North
+terrain over millenial time scales \citep[e.g.,][]{Golledge.etal.2012}. In the
+North
 American Cordillera, the topographic complexity induces strong gradients in
@@
 and two alkenone unsaturation index records from Northwest Pacific oceanic
-sediment cores \citep{Herbert.2001}. Model output is compared to
+sediment cores \citep{Herbert.etal.2001}. Model output is compared to
 geomorphological evidence in terms of timing and extent of glaciation and
@@
 The Shallow Shelf Approximation (SSA) is used as a ``sliding law'' for the
-Shallow Ice Approximation (SIA) \citep{Bueler.Brown.2009,Winkelmann.etal.2011}.
+Shallow Ice Approximation (SIA) \citep[Eqn.~15]{Winkelmann.etal.2011}.
 Ice softness depends on temperature and water content through an enthalpy
-formulation \citep{Aschwanden.Blatter.2009,Aschwanden.etal.2012}. Surface air
+formulation \citep{Aschwanden.etal.2012}. Surface air
 temperature from the atmospheric forcing provides the upper boundary condition
@@
 
-A~pseudo-plastic sliding law \citep{Aschwanden.etal.2013} relates the
+A~pseudo-plastic sliding law \citep{PISM-authors.2014} relates the
 bed-parallel shear stress and the sliding velocity. The yield stress is
@@
 till, using a formula derived from laboratory experiments with till extracted
-from an Antarctic ice stream \citep{Tulaczyk.etal.2000}. Basal topography is
+from an Antarctic ice stream \citep[Eqn.~2]{Tulaczyk.etal.2000}. Basal
+topography is
 derived from the ETOPO1 combined topography and bathymetry dataset with
@@
 deviation of near-surface temperature by a~temperature-index (positive
-degree-day) model \citep{Hock.2003}. Ice accumulation is equal to precipitation
+degree-day) model \citep[e.g.,]{Hock.2003}. Ice accumulation is equal to
+precipitation
 when temperature is below 0\,\unit{{\degree}C}, and decreases to zero linearly
@@
 
-The positive degree-day integral \citep{Calov.Greve.2005} is numerically
+The positive degree-day integral \citep[Eqn.~6]{Calov.Greve.2005} is
+numerically
 approximated using week-long sub-intervals. It accounts for temperature
@@
 computed from daily temperature values from the North American Regional
-Reanalysis (NARR) \citep{Mesinger.etal.2006}, after excluding variability
-associated with the seasonal cycle itself \citep{Seguinot.Rogozhina.2014}. The
+Reanalysis \citep[NARR,][]{Mesinger.etal.2006}, after excluding variability
+associated with the seasonal cycle itself \citep[cf.][]
+{Seguinot.Rogozhina.2014}. The
 ablation model incorporates degree-day factors of
@@
 The present-day climatology $\{T_{m0}, P_{m0}\}$ was computed from
-near-surface air temperature and precipitation rate fields of the NARR over
+near-surface air temperature and precipitation rate fields of the NARR
+\citep{Mesinger.etal.2006} over
 the period 1979--2000 (Fig.~\ref{fig:atm}). Modern climate of the North
@@
 Drilling Program (ODP) sites 1012 and 1020, both located off the Californian
-shore \citep{Herbert.2001}. Palaeo-temperatures from the GRIP and NGRIP records
-were calculated using a quadratic equation \citep{Johnsen.etal.1995},
+shore \citep{Herbert.etal.2001}. Palaeo-temperatures from the GRIP and NGRIP
+records were calculated using a quadratic equation \citep{Johnsen.etal.1995},
 \begin{equation}
@@
 dating between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,kyr\,BP}
-\citep{Porter.Swanson.1998}.
+\citep[Fig.~4]{Porter.Swanson.1998}.
 These dates have been confirmed by the offshore sediment record, which shows an
@@
 early warming is a local effect, corresponding to a weakening of the California
-current \citep{Herbert.2001}. The California current, driving cold waters southwards
+current \citep[Fig.~3]{Herbert.etal.2001}. The California current, driving cold
+waters southwards
 along the western coast of North America, is thought to have weakened in the
@@
 sea-surface temperatures at the locations of the ODP~1012 and ODP~1020 sites
-during the last glacial maximum \citep{Herbert.2001}.
+during the last glacial maximum \citep{Herbert.etal.2001}.
 
@@
     ODP~1012 & \chem{U^{K'}_{37}}  & ?\% & -5.8{\degree}C
-        & \citet{Herbert.2001} \\
+        & \citet{Herbert.etal.2001} \\
     ODP~1020 & \chem{U^{K'}_{37}}  & ?\% & -5.8{\degree}C
-        & \citet{Herbert.2001} \\
+        & \citet{Herbert.etal.2001} \\
     \bottomhline

### 2014-06-17 cordillera-cycle.tex
Add EPICA run deglaciation profiles.

@@
 plateau, while the surrounding mountain ranges are already deglaciated
-(Fig.~\ref{fig:profiles}). Although this supports the top-down (?)
+(Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}). Although this
+supports the top-down (?)
 deglaciation hypothesis for central British Columbia, modelled, southwards,
@@
 
-% fig:profiles
+% fig:profiles-grip
 \begin{figure}
-  \includegraphics{profiles}
+  \includegraphics{profiles-grip}
   \caption{Modelled bedrock (black) and ice surface (blue) topography profiles
@@
            simulation, corresponding to the four transects indicated in
-           Fig.~\ref{fig:deglac}.
-           \note{should we have a similar figure using EPICA?}}
-  \label{fig:profiles}
+           Fig.~\ref{fig:deglac}.}
+  \label{fig:profiles-grip}
+\end{figure}
+
+% fig:profiles-epica
+\begin{figure}
+  \includegraphics{profiles-epica}
+  \caption{Modelled bedrock (black) and ice surface (blue) topography profiles
+           during deglaciation (22.0--8.0\,kyr) in the EPICA 6\,km
+           simulation, corresponding to the four transects indicated in
+           Fig.~\ref{fig:deglac}.}
+  \label{fig:profiles-epica}
 \end{figure}

### 2014-06-17 figures/deglac.py

@@ -62,10 +62,10 @@ for i, rec in enumerate(records):
     # annotate
     annotate(ax, rec.upper())
 
-# add profile lines
-for yp in [1.7e6, 1.4e6, 1.1e6, 0.8e6]:
-    fig.grid[0].plot([-2.4e6, -1.25e6], [yp, yp], 'k|',
-                     lw=0.25, ls='--', dashes=(2, 2))
+    # add profile lines
+    for yp in [1.7e6, 1.4e6, 1.1e6, 0.8e6]:
+        ax.plot([-2.4e6, -1.25e6], [yp, yp], 'k|',
+                         lw=0.25, ls='--', dashes=(2, 2))
 
 # add colorbar and save
 print 'saving deglac...'

### 2014-06-17 figures/profiles-epica.py

@@ -1,0 +1,18 @@
+#!/usr/bin/env python2
+# coding: utf-8
+
+from paperglobals import darkred
+from profiles import profiles
+
+# parameters
+res = '6km'
+rec = 'epica'
+dt = 5.6
+color = darkred
+
+# plot
+fig = profiles(res, rec, dt, color)
+
+# save
+print('saving...')
+fig.savefig('profiles-epica')

### 2014-06-17 figures/profiles-grip.py

@@ -1,0 +1,18 @@
+#!/usr/bin/env python2
+# coding: utf-8
+
+from paperglobals import darkblue
+from profiles import profiles
+
+# parameters
+res = '6km'
+rec = 'grip'
+dt = 5.8
+color = darkblue
+
+# plot
+fig = profiles(res, rec, dt, color)
+
+# save
+print('saving...')
+fig.savefig('profiles-grip')

### 2014-06-17 figures/profiles.py

@@ -4,56 +4,54 @@
 import numpy as np
 from netCDF4 import Dataset
 from matplotlib import pyplot as plt
-from matplotlib.colors import BoundaryNorm, Normalize
-from iceplot import plot as iplt
-from paperglobals import *
+from paperglobals import in2mm, run_path, s2ka
 
 # parameters
 tmin, tmax = -22.0, -8.0
 yplist = [1.7e6, 1.4e6, 1.1e6, 0.8e6]
 
-# initialize figure
-figw, figh = 85.0, 100.0
-fig, grid = plt.subplots(len(yplist), figsize=(figw*in2mm, figh*in2mm),
-                         sharex=True, sharey=True)
-fig.subplots_adjust(left=10.0/figw, bottom=10/figh,
-                    right=1-2.5/figw, top=1-2.5/figh,
-                    hspace=1/((1+figh/2.5)/4-1))
+def profiles(res, rec, dt, color):
+    # initialize figure
+    figw, figh = 85.0, 100.0
+    fig, grid = plt.subplots(len(yplist), figsize=(figw*in2mm, figh*in2mm),
+                             sharex=True, sharey=True)
+    fig.subplots_adjust(left=10.0/figw, bottom=10/figh,
+                        right=1-2.5/figw, top=1-2.5/figh,
+                        hspace=1/((1+figh/2.5)/4-1))
 
-# read extra output
-print 'reading %s extra output...' % rec
-nc = Dataset(run_path % ('6km', 'grip', 580) + '-extra.nc')
-x = nc.variables['x']
-y = nc.variables['y']
-time = nc.variables['time']
-mask = nc.variables['mask']
-topg = nc.variables['topg']
-usurf = nc.variables['usurf']
+    # read extra output
+    print 'reading %s extra output...' % rec
+    nc = Dataset(run_path % (res, rec, 100*dt) + '-extra.nc')
+    x = nc.variables['x']
+    y = nc.variables['y']
+    time = nc.variables['time']
+    mask = nc.variables['mask']
+    topg = nc.variables['topg']
+    usurf = nc.variables['usurf']
 
-# plot
-kmin, kmax = [np.argmin(np.abs(time[:]*s2ka-t)) for t in (tmin, tmax)]
[…]

### 2014-06-17 cordillera-cycle.tex
Add EPICA run deglaciation snapshots.

@@
 
-% fig:deglacshots
+% fig:deglacshots-grip
 \begin{figure*}
-  \includegraphics{deglacshots}
+  \includegraphics{deglacshots-grip}
   \caption{Snapshots of modelled surface topography (200\,m contours)
            and surface velocity (colour mapping) from the GRIP simulation,
-           corresponding to the last glacial maximum (-19.5 kyr) and the last
-           deglaciation.
-           \note{should we have a similar figure using EPICA?}}
-  \label{fig:deglacshots}
+           corresponding to the last glacial ice volume maximum (-19.1 kyr) and
+           the last deglaciation.}
+  \label{fig:deglacshots-grip}
+\end{figure*}
+
+% fig:deglacshots-epica
+\begin{figure*}
+  \includegraphics{deglacshots-epica}
+  \caption{Snapshots of modelled surface topography (200\,m contours)
+           and surface velocity (colour mapping) from the EPICA simulation,
+           corresponding to the last glacial ice volume maximum (-17.3 kyr) and
+           the last deglaciation.}
+  \label{fig:deglacshots-epica}
 \end{figure*}

### 2014-06-17 figures/deglacshots-epica.py

@@ -1,0 +1,66 @@
+#!/usr/bin/env python2
+# coding: utf-8
+
+import numpy as np
+from netCDF4 import Dataset
+from matplotlib import pyplot as plt
+from matplotlib.colors import BoundaryNorm, LogNorm
+from mpl_toolkits.axes_grid1.axes_grid import ImageGrid
+from iceplot import plot as iplt
+from paperglobals import *
+
+# parameters
+res = '6km'
+rec = 'epica'
+dt = 5.6
+
+# initialize figure
+figw, figh = 120.0, 100.0
+fig = plt.figure(0, (figw*in2mm, figh*in2mm))
+rect1 = [2.5/figw, 2.5/figh, 47.5/figw, 95.0/figh]
+rect2 = [52.5/figw, 2.5/figh, 56.25/figw, 95.0/figh]
+grid = [None]*5
+grid[0] = fig.add_axes(rect1)
+grid[1:] = ImageGrid(fig, rect2, (2, 2), axes_pad=2.5*in2mm,
+                 cbar_mode='single', cbar_location='right',
+                 cbar_pad=2.5*in2mm, cbar_size=5.0*in2mm)
+remove_ticks(grid)
+
+# get LGM time
+this_run_path = run_path % (res, rec, dt*100)
+lgm_time = get_mis_times(this_run_path + '-ts.nc')[-1][-1]
+plot_times = [lgm_time, -16, -14, -12, -10]
+
+# load extra output
+nc = Dataset(this_run_path + '-extra.nc')
+mask = nc.variables['mask']
+time = nc.variables['time'][:]*s2ka
+topg = nc.variables['topg']
+csurf = nc.variables['velsurf_mag']
+thk = nc.variables['thk']
+
+# round pltting times to nearest slice
+plot_idces = [(np.abs(time[:]-t)).argmin() for t in plot_times]
+
+# plot
+for i, t in enumerate(plot_idces):
+    print 'plotting %s kyr snapshot...' % time[t]
+    ax = grid[i]
+    plt.sca(ax)
+    iplt.bedtopoimage(nc, t, cmap=topo_cmap, norm=topo_norm)
+    iplt.icemargincontour(nc, t, linewidths=0.5)
+    iplt.surftopocontour(nc, t, levels=range(200, 5000, 200), linewidths=0.1)
+    iplt.surftopocontour(nc, t, levels=range(1000, 5000, 1000), linewidths=0.25)
[…]

### 2014-06-17 figures/deglacshots-grip.py

@@ -63,4 +63,4 @@ cb = fig.colorbar(im, ax.cax, extend='both', format='%i',
                   ticks=np.logspace(1, 3.5, 6))
 cb.set_label(r'surface velocity ($m\,yr^{-1}$)', labelpad=-2.0)
 print 'saving...'
-fig.savefig('deglacshots')
+fig.savefig('deglacshots-grip')

### 2014-06-17 figures/warmfrac.py
Add warm-base cover fraction map.

@@ -39,10 +39,10 @@ for i, rec in enumerate(records):
 
     # compute duration of warm-based coved
     warm = np.ma.array((temp[:] > -1e-9), mask=(mask[:] != 2))
-    warm = warm.sum(axis=0).T/10.0
+    warm = warm.mean(axis=0).T
 
     # set levels, colors and hatches
-    levs = [-1] + range(0, 121, 20)
+    levs = [-1, 0.0, 0.5, 0.9, 0.99, 1.0]
     cmap = plt.get_cmap('Reds')
     colors = cmap(np.linspace(0.0, 1.0, len(levs)-1))
     hatches = ['//'] + ['']*(len(levs)-2)
@@ -59,5 +59,5 @@ for i, rec in enumerate(records):
 
 # add colorbar and save
 cb = fig.colorbar(cs, ax.cax, ticks=levs[1:])
-cb.set_label('Duration of warm-based ice cover (kyr)')
-fig.savefig('warmbase.png')
+cb.set_label('Fraction of warm-based ice cover')
+fig.savefig('warmfrac.png')

### 2014-06-17 cordillera-cycle.tex
Add warm-base cover fraction map.
Re-read intro to half-discussion and add notes.

@@
   better reproduced using temperature reconstructions from Greenland and
-  Antarctic ice cores than from regional ocean sediment cores. During most of
+  Antarctic ice cores than from regional oceanic sediment cores. During most of
   the last glacial cycle, the modelled ice cover is discontinuous and
@@
 contiguous blanket of ice extended from the Alaska Range in the north to the
-North Cascades in the south \citep[Fig.~\ref{fig:locmap};][Plate XIV]
-{Chamberlin.1894}. This ice mass is known as the former Cordilleran ice sheet
-\citep{Dawson.1888}.
+North Cascades in the south (Fig.~\ref{fig:locmap}). This ice mass is known as
+the former Cordilleran ice sheet \citep{Dawson.1888}.
 
@@
 {Ward.etal.2007, Menounos.etal.2009, Stroeven.etal.2010, Stroeven.etal.2013},
-and the offshore sedimental record \citep{Cosma.etal.2008, Davies.etal.2011}.
+and the offshore sedimentary record \citep{Cosma.etal.2008, Davies.etal.2011}.
+
+\note{Is the set of references in the above paragraph representative of the
+      current state of the field? Am I missing something important here?}
 
@@
 In fact, only recent development of numerical ice sheet models and underlying
-scientific computing tools \citep[e.g.,][]{Bueler.Brown.2009} allows
-high-resolution numerical modelling of palaeo-glacier flow on mountainous
-terrain over millenial time scales \citep[e.g.,][]{Golledge.etal.2012}. In the
-North
-American Cordillera, the topographic complexity induces strong gradients in
+scientific computing tools \citep{Bueler.Brown.2009, Balay.etal.2014} allows
+high-resolution numerical modelling of palaeo-glaciers on mountainous terrain
+over millenial time scales \citep{Golledge.etal.2012}. In the North
+American Cordillera, this complex topography induces strong gradients in
 seasonality and distinct patterns of precipition, thus requiring the use of
@@
 The simulations presented here were run using Parallel Ice Sheet Model (PISM,
-development version~11b0a7f), an open-source, finite-difference, shallow ice
-sheet model \citep{PISM-authors.2014}. The model inputs basal topography,
-sea level, geothermal
+development version~11b0a7f and stable version~0.6.1), an open-source,
+finite-difference, shallow ice sheet model \citep{PISM-authors.2014}.
+The model inputs basal topography, sea level, geothermal
 heat flux and climate forcing, and computes the evolution of ice extent
@@
 conditions at -120\,kyr, and runs to present. These computations were
-performed on 16 to 128 cores at the Swedish National Supercomputing
-Center.
+performed on 16 to 128 cores at the Swedish National Supercomputing Centre.
 
@@
 The Shallow Shelf Approximation (SSA) is used as a ``sliding law'' for the
-Shallow Ice Approximation (SIA) \citep[Eqn.~15]{Winkelmann.etal.2011}.
+Shallow Ice Approximation (SIA) by adding both velocity solutions
+\citep[Eqn.~15]{Winkelmann.etal.2011}.
 Ice softness depends on temperature and water content through an enthalpy
@@
 
-A~pseudo-plastic sliding law \citep{PISM-authors.2014} relates the
+A~pseudo-plastic sliding law \citep{PISM-authors.2014} \note{ask Ed Bueler if
+this will be published somewhere soon} relates the
 bed-parallel shear stress and the sliding velocity. The yield stress is
@@
 derived from the ETOPO1 combined topography and bathymetry dataset with
-a~resolution of 1\,arc-min \citep{Amante.Eakins.2009}.
-Sea level is lowered as a function of time according to the SPECMAP
-reconstruction, and basal topography responds to ice load
+a~resolution of 1\,arc-min \citep{Amante.Eakins.2009}. Sea level is lowered as
+a function of time based on the Spectral Mapping Project (SPECMAP)
+\citep[SPECMAP,][]{Imbrie.etal.1989} time scale.
+Basal topography responds to ice load
 following a bedrock deformation model that includes point-wise isostasy,
@@
 near-surface air temperature, monthly precipitation, and monthly standard
-deviation of near-surface temperature by a~temperature-index (positive
-degree-day) model \citep[e.g.,]{Hock.2003}. Ice accumulation is equal to
-precipitation
+deviation of near-surface temperature by a~temperature-index model
+\citep[e.g.,][]{Hock.2003}. Ice accumulation is equal to precipitation
 when temperature is below 0\,\unit{{\degree}C}, and decreases to zero linearly
@@
 computed from daily temperature values from the North American Regional
-Reanalysis \citep[NARR,][]{Mesinger.etal.2006}, after excluding variability
+Reanalysis \citep[NARR,][]{Mesinger.etal.2006}, in order to account for the
+effects of spatial and seasonal variations of temperature variability
+\citep[Fig.~\ref{fig:atm};][]{Seguinot.2013}, after excluding variability
 associated with the seasonal cycle itself \citep[cf.][]
-{Seguinot.Rogozhina.2014}. The
-ablation model incorporates degree-day factors of
-3.04\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for snow and
+{Seguinot.Rogozhina.2014}. The ablation model incorporates degree-day factors
+of 3.04\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for snow and
 4.59\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for ice, as derived from
@@
 near-surface air temperature and precipitation rate fields of the NARR
-\citep{Mesinger.etal.2006} over
-the period 1979--2000 (Fig.~\ref{fig:atm}). Modern climate of the North
-American Cordillera is characterised by strong geographic variations of
-seasonality, timing of the precipitation peak and daily temperature variability.
+\citep{Mesinger.etal.2006} over the period 1979--2000. Modern climate of the
+North American Cordillera is characterised by strong geographic variations of
+seasonality, timing of the precipitation peak and daily temperature variability
+(Fig.~\ref{fig:atm}).
 Our choice of data from the NARR is motivated by the need for an accurate,
-high-resolution precipitation forcing, as outlined in a previous sensitivity
+high-resolution precipitation forcing, as identified in a previous sensitivity
 study \citep{Seguinot.etal.2013}.
@@
 The lapse-rate correction ${\Delta}T_{LR}$ is computed using the NARR surface
-geopotential height a reference topography $z_{ref}$,
+geopotential height invariant field as a reference topography $z_{ref}$,
 \begin{align}
@@
 horizontal resolution of 6\,km, up to 101~ice layers, and 61~bedrock
-layers. In the rest of this paper, we refer to different model runs by the
-proxy record used for palaeo-temperature forcing.
+layers. In the rest of this paper, we occasionally refer to different model
+through by the proxy record used for palaeo-temperature forcing.
 
@@
 consistently attained during MIS~3 (52.2--45.6\,kyr). However, the
-magnitude and timing of these events depend significantly on the choice of
-proxy record used for palaeo-temperature forcing.
+magnitude and timing of these three events depend significantly on the choice
+of proxy record used for palaeo-temperature forcing.
 
@@
 highest variability in modelled ice volume throughout the glaciation history.
-Model runs using oceanic and Antarctic records generally result in lower
+Model runs using oceanic (ODP~1012, ODP~1020) and Antarctic (EPICA, Vostok)
+records generally result in lower
 modelled ice volume during MIS~4 and larger ice volume during MIS~3, thus
@@
 NGRIP palaeo-temperature forcing is the only one that results in larger
-glaciation during MIS~4 than during the last glacial maximum.
+glaciation during MIS~4 than during the last glacial maximum (MIS~2).
 
-While simulations driven by the Antarctic (EPICA, Vostok) and GRIP records
+While simulations driven by the Antarctic and GRIP records
 reach a last glacial maximum in ice volume between 19.5 and 16.9\,kyr,
-those steered by oceanic (ODP~1012, ODP~1020) and the NGRIP record attain it
+those steered by oceanic and the NGRIP records attain it
 several thousands of years earlier. Moreover, the ODP~1012 run yields rapid
@@
 consistent patterns of glaciation (Fig.~\ref{fig:snapshots}). As a result of
-different scaling factors applied to palaeo-temperature proxy records, the
-ice sheet geometry at the last glacial maximum appears very similar from one
-simulation to the next. It includes a central divide located at about
+different scaling factors applied to palaeo-temperature proxy records,
+modelled ice sheet geometries at the last glacial maximum appear very similar
+from one simulation to the next. They include a central divide located at about
 3\,500\,m elevation along the spine of the Rocky Mountains. In all
@@
 between the two glaciation events. However, the size of this ice cap, as well
-as the magnitude of the MIS~4 glaciation, depends sensitively on the choice of
+as the magnitude of the MIS~4 glaciation, depend sensitively on the choice of
 palaeo-temperature proxy record used to drive the model.
@@
 
+\note{Could you help me to fill the missing refs in this last sentence?}
+
 However, the timing of the modelled glacial maxima strongly depends on the
@@
 that pre-dates reconstructed ages by several thousand years. Concerning the
-simulations driven by oceanic (ODP~1012, ODP~1020) records, this early
-deglaciation is caused by an early warming sealed in the palaeo-temperature records
+simulations driven by oceanic records, this early deglaciation is caused by an
+early warming sealed in the palaeo-temperature reconstructions
 (Fig.~\ref{fig:timeseries}, upper panel). However, it has been shown that this
@@
 in the direct vicinity of the former Cordilleran ice sheet would be of great
-help to understand its glacial history. Because, to our knowledge, no such
-record yet exists, we focus the rest of our analysis on simulations forced by
-temperature records from Greenland (GRIP) and Antarctic (EPICA) ice cores.
+help to further understand its glacial history. Because, to our knowledge, no
+such record yet exists, we focus the rest of our analysis on simulations forced
+by temperature records from Greenland (GRIP) and Antarctic (EPICA) ice cores
+(Fig.~\ref{fig:timeseries}, lower panel, dotted lines).
 
@@
 Mountains exists for about 29\,kyr in total over MIS~2 and~4.
-
 With the exception of the oceanic front and the northern foothills of the Alaska
@@
 durations of ice cover, generally a few thousand years, distributed over one
-(GRIP) or two (EPICA) glaciations. This shows that, in our simulations, the
-most extensive stages of the Cordilleran ice sheet are both short-lived and
-out of balance with contemporary climate states, which confirms
-previous interpretations of the geological record \citep{Clague.etal.1980,
-Stroeven.etal.2010}.
+(GRIP) or two (EPICA) glaciations (Fig.~\ref{fig:duration}). This illustrates
+that the most extensive stages of the modelled ice sheet are both short-lived
+and out of balance with contemporary climate states, which confirms previous
+inferences of a short-lived Cordilleran ice sheet based on the geological
+record \citep{Clague.etal.1980, Stroeven.etal.2010}.
 
@@
 throughout the entire last glacial cycle (Fig.~\ref{fig:snapshots} and
-\ref{fig:duration}). In particular, this ice cap persists between the MIS~4
-and MIS~2 glaciations, providing a nucleation centre for the onset of glacial
-readvance towards the last glacial maximum.
+\ref{fig:duration}). In particular, this ice cap persists during MIS~3,
+providing a nucleation centre for the onset of glacial
+readvance towards the last glacial maximum (MIS~2).
 
@@
 associated with basal ice temperatures at the pressure-melting point
-(Fig.~\ref{fig:warmbase}) may partly explain this spectacular landscape.
+(Fig.~\ref{fig:warmbase} and ~\ref{fig:warmbase}) may partly explain this
+spectacular landscape.
+
+\note{Can we use a reference here to support the point of strong glacial
+      imprint in the Skeena Mountains?}
 
@@
   \includegraphics{warmbase}
+  \caption{Modelled fraction of warm-based ice cover during the ice-covered
+           period. Note the dominance of warm-based conditions on the
+           continental shelf and major glacial troughs of the coastal ranges.
+           Hatches indicate areas that were covered by cold ice only.}
+  \label{fig:warmbase}
+\end{figure*}
+
+% fig:warmfrac
+\begin{figure*}
+  \includegraphics{warmfrac}
   \caption{Modelled duration of warm-based ice cover during the last
@@
            \note{indicate location of Skeena Mountains.}}
-  \label{fig:warmbase}
+  \label{fig:warmfrac}
 \end{figure*}

### 2014-06-17 cordillera-cycle.tex
Reformat and improve notes and todos.

@@
 % My commands
-\newcommand{\idea}[1]{\textbf{[IDEA: #1]}}
 \newcommand{\note}[1]{\textbf{[NOTE: #1]}}
-\newcommand{\todo}[1]{\textbf{[TODO: #1]}}
+\newcommand{\todo}[1]{\emph{[\textbf{Todo:} #1]}}
 \newcommand{\aref}[0]{\textbf{[ref.]}}
@@
 
-\begin{center}
-\begin{tabular}{>{\bfseries}ll}
-    \tophline
-    Introduction & Happy with it except for missing refs.\\
-    Model setup  & Happy with it except for missing refs.\\
-    Results      & Happy with it except for missing refs.\\
-    Discussion   & Deglaciation part needs improvement.\\
-    Conclusions  & Can be compacted and the style improved.\\
-    \bottomhline
-\end{tabular}
-\end{center}
-
 % ----------------------------------------------------------------------
@@
 {Ward.etal.2007, Menounos.etal.2009, Stroeven.etal.2010, Stroeven.etal.2013},
-and the offshore sedimentary record \citep{Cosma.etal.2008, Davies.etal.2011}.
-
-\note{Is the set of references in the above paragraph representative of the
-      current state of the field? Am I missing something important here?}
+and the offshore sedimentary record \citep{Cosma.etal.2008, Davies.etal.2011}%
+\footnote{\textbf{Arjen, Martin,}, please comment on my choice of references
+    in this paragraph. Is it a fair sample of the current state of the field?
+    Am I missing some other important works?}.
 
@@
 
-A~pseudo-plastic sliding law \citep{PISM-authors.2014} \note{ask Ed Bueler if
-this will be published somewhere soon} relates the
+A~pseudo-plastic sliding law \citep{PISM-authors.2014}\footnote{I need to ask
+    Ed Bueler if the sliding will be published somewhere soon.} relates the
 bed-parallel shear stress and the sliding velocity. The yield stress is
@@
 evidence in the northern part of the model domain \aref, and to the Fraser
-(MIS~2) glaciation in the southern part \aref.
-
-\note{Could you help me to fill the missing refs in this last sentence?}
+(MIS~2) glaciation in the southern part \aref
+\footnote{\textbf{Arjen, Martin, Johan,} can you check that I use these names
+    correctly, and help me to fill the missing refs in this sentence? Is is
+    possible to trace back when these names where first applied?}.
 
@@
 palaeo-temperature forcing, a central Skeena mountain ice cap persisting
-throughout the entire last glacial cycle (Fig.~\ref{fig:snapshots} and
-\ref{fig:duration}). In particular, this ice cap persists during MIS~3,
+throughout the entire last glacial cycle (Figs.~\ref{fig:snapshots}
+and~\ref{fig:duration}). In particular, this ice cap persists during MIS~3,
 providing a nucleation centre for the onset of glacial
@@
 In addition, the erosional landscape of the Skeena mountain bears one of the
-strongest glacial imprint (?) found within the area formerly covered by the
-Cordilleran ice sheet \aref. Here we suggest that persistent ice cover
+strongest glacial imprint found within the area formerly covered by the
+Cordilleran ice sheet \aref%
+\footnote{\textbf{Arjen, Martin, Johan,} can we use a reference here to support
+    this point (observations, mapping, shaded relief)? Alternatively, we could
+    maybe use a photograph, if any of you has one. It would add a bit of
+    variation into the figure list. \textbf{Johan}, you seemed to support this
+    this point during our discussion, any idea?}.
+Here we suggest that persistent ice cover (Fig.~\ref{fig:duration})
 associated with basal ice temperatures at the pressure-melting point
-(Fig.~\ref{fig:warmbase} and ~\ref{fig:warmbase}) may partly explain this
-spectacular landscape.
-
-\note{Can we use a reference here to support the point of strong glacial
-      imprint in the Skeena Mountains?}
+(Fig.~\ref{fig:warmfrac} and ~\ref{fig:warmbase}%
+\footnote{Fig.~\ref{fig:erosion} is tricky because it may be significantly
+    affected by the constant-geothermal heat flux assumption, which is not
+    valid for this region. Arjen, Johan and I decided to leave it for my kappa
+    for now. Regarding Figs.~\ref{fig:warmfrac} and ~\ref{fig:warmbase}, I
+    still wonder if it is useful to show both. Basically we have
+    Fig.~\ref{fig:warmbase}\,=\,Fig.~\ref{fig:duration}\,*\,Fig.~\ref{fig:warmfrac}.
+    Maybe the reader can perform this multiplication visually without need of
+    Fig.~\ref{fig:warmbase}? Tell me what you think of it.})
+may partly explain this spectacular landscape.
 
@@
 
+% fig:warmfrac
+\begin{figure*}
+  \includegraphics{warmfrac}
+  \caption{Modelled fraction of warm-based ice cover during the ice-covered
+           period. Note the dominance of warm-based conditions on the
+           continental shelf and major glacial troughs of the coastal ranges.
+           Hatches indicate areas that were covered by cold ice only.
+           \todo{indicate location of the Skeena Mountains.}}
+  \label{fig:warmfrac}
+\end{figure*}
+
 % fig:warmbase
@@
   \includegraphics{warmbase}
-  \caption{Modelled fraction of warm-based ice cover during the ice-covered
-           period. Note the dominance of warm-based conditions on the
-           continental shelf and major glacial troughs of the coastal ranges.
-           Hatches indicate areas that were covered by cold ice only.}
-  \label{fig:warmbase}
-\end{figure*}
-
-% fig:warmfrac
-\begin{figure*}
-  \includegraphics{warmfrac}
   \caption{Modelled duration of warm-based ice cover during the last
@@
            Hatches indicate areas that were covered by cold ice only.
-           \note{indicate location of Skeena Mountains.}}
-  \label{fig:warmfrac}
+           \todo{indicate location of the Skeena Mountains.}}
+  \label{fig:warmbase}
 \end{figure*}
@@
            over the last 120\,kyr.
-           \note{I am still unsure about the use of this figure. I may
-                 remove it from this paper and keep the idea for later use.}}
+           \todo{remove this figure (see footnote in text).}}
   \label{fig:erosion}

### 2014-06-17 cordillera-cycle.tex
Develop discussion of deglaciation history.

@@
 of former
-melt-water streams \citep{Margold.etal.2011, Margold.etal.2013}, as well as the
+melt-water streams \citep{Margold.etal.2011, Margold.etal.2013a}, as well as the
 timing of glaciation identified from radiocarbon dating \citep[e.g.,][]
 {Clague.1981, Porter.Swanson.1998}, cosmogenic dating \citep[e.g.,][]
-{Ward.etal.2007, Menounos.etal.2009, Stroeven.etal.2010, Stroeven.etal.2013},
+{Ward.etal.2007, Menounos.etal.2008, Stroeven.etal.2010, Stroeven.etal.2013},
 and the offshore sedimentary record \citep{Cosma.etal.2008, Davies.etal.2011}%
@@
 glacier cover, most of the older evidence having been overridden by ice retreat
-retreat during the deglaciation.
+retreat during the deglaciation \footnote{\textbf{Johan,} can we refer to one
+of your papers here?}.
 
@@
 the deglaciation, depends on the palaeo-temperature time-series used to force
-the model (Figs.~\ref{fig:timeseries} and ~\ref{fig:deglacseries}). The
+the model (Fig.~\ref{fig:timeseries}). The
 6\,km-resolution simulation forced by the EPICA record reaches a maximal
 ice volume at 17.3\,kyr, followed by a continuous deglaciation between
-about 17 and 11\,kyr. That driven by the GRIP record reaches maximal
+about 17 and 11\,kyr (Fig.~\ref{fig:deglacseries}, red dotted curve).
+That driven by the GRIP record reaches maximal
 volume at 19.1\,kyr, and produces a deglaciation in three steps,
 including the first stand-still between 16.6 and 14.5\,kyr, and a late
-glacial re-advance between 12.6 and 11.6\,kyr.
+glacial re-advance between 12.6 and 11.6\,kyr (Fig.~\ref{fig:deglacseries},
+blue dotted curve).
 
-\todo{discuss the work that supports YD re-advance.}
+The possibly of a widespread late-glacial re-advance of decaying glaciers of
+the North American Cordilleran has been debated for long. On the southern tip
+of the Coast Mountains, radiocarbon-dated frontal moraines indicate multiple
+stages of glacial re-advance in the Fraser and Squamish valleys, of which
+one corresponds to the Younger Dryas chronozone \citep{Clague.etal.1997,
+Friele.Clague.2002, Friele.Clague.2002a, Kovanen.2002,
+Kovanen.Easterbrook.2002}. Additional evidence for a late-glacial re-advance
+was found in the Finlay River area, northern Rocky Mountains, where
+sharp-crested moraines show an abundant remobilization of late-glacial
+sediments by alpine glaciers in interaction with the main body of the decaying
+ice sheet \citep{Lakeman.etal.2008}. Although further work is needed to
+constrain the timing of the North American Cordillera late-glacial re-advance,
+to seize its extent and identify potential climatic triggers
+\citep{Menounos.etal.2008}, it is interesting to note that the simulation
+driven by the GRIP record produces a late-glacial re-advance in the Coast
+Mountains, Rocky Mountains and Finlay River area, corresponding to where it
+has been identified in the geologic record (Fig.~\ref{fig:deglac}, left panel),
+whereas the EPICA-driven simulation produces a nearly-continuous deglaciation
+with very restrained glacial re-advance (Fig.~\ref{fig:deglac}, right panel).
 
-Although the predicted timing of glacier retreat differs, its patterns are
-relatively consistent between the two simulations (Fig.~\ref{fig:deglac}). In
+Although the predicted timing of glacier retreat differs between the two
+simulations, its patterns are relatively consistent (Fig.~\ref{fig:deglac}). In
 both runs, the southern half of the modelling domain, including the Puget
 Lowland, the Coast and Rocky Mountains and the Interior Plateau of British
-Columbia, are completely deglaciated by 12\,kyr, whereas significant
-ice cover remains on the Skeena, Selwyn, MacKenzie, Wrangell and Saint-Elias Mountains
-in the northern half of the domain. At this stage, a late glacial re-advance
-occurs in the GRIP simulation. Both simulations then show an unzipping of the
-Liard Lowland, followed by ice retreat towards the Selwyn Mountains and the
-Skeena Mountains, which host the last remains of the palaeo-ice sheet.
+Columbia, are completely deglaciated by 12\,kyr, whereas significant ice cover
+remains on the Skeena, Selwyn, MacKenzie, Wrangell and Saint-Elias Mountains
+in the northern half of the domain. After 12\,kyr, deglaciation proceeds
+through an opening of the Liard lowland with radial margin retreat towards the
+surrounding mountain ranges and inward flow towards the depression, in full
+compatibility with the regional melt water record \citep{Margold.etal.2013}.
+Remaining ice continues its decay by retreating towards the Selwyn Mountains
+and, then, the Skeena Mountains. The last remains of the palaeo-ice sheet
+vanish in the Skeena Mountains at around 10.6\,kyr in the EPICA simulation
+versus 9.4\,kyr in the GRIP simulation.
 
-\todo{discuss existing evidence (Martin).}
+Once more, snapshots of modelled deglacial ice sheet configurations
+(Figs.~\ref{fig:deglacshots-grip} and~\ref{fig:deglacshots-epica}) provide a
+poor medium for direct comparison with geomorphological evidence of the ice
+retreat. Here, we compute a map of de-glacial, basal flow direction immediately
+preceding exposure of ice-free land or cessation of sliding upon the approach
+of the retreating margin (Fig.~\ref{fig:lastflow}). On the northern and
+southern front of the ice sheet and in the Liard Lowland, the resulting map
+show similarities with patterns of glacial lineations
+(\citealp{Prest.etal.1968}; \citealp[Fig.~1.12]{Clague.1989};
+\citealp[Fig.~2]{Kleman.etal.2010}; \citep[Fig.~2]{Margold.etal.2013}).
+However, such is not the case on the Interior Plateau of British Columbia,
+where the model produce no deglacial flow directions (Fig.~\ref{fig:lastflow}),
+whereas glacial lineation maps indicate eastward flow.
 
-On the Interior Plateau of British Columbia, modelled deglaciation consists of
-a rapid northwards retreat, leaving a slow-flowing ice lobe on the
-plateau, while the surrounding mountain ranges are already deglaciated
-(Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}). Although this
-supports the top-down (?)
-deglaciation hypothesis for central British Columbia, modelled, southwards,
-last flow directions do not agree with the local lineation patterns
-(Fig.~\ref{fig:lastflow}). This could indicate that the lineations are older,
-or that the modelled last glacial maximum ice sheet is too extensive to the
-east, resulting in modelled deglaciation flow patterns different from those
-observed in the geomorphology.
+On the one hand, this could indicate that the model performs poorly in this
+region, perhaps because the modelled last glacial maximum ice sheet is too
+thick, or too extensive to the west. Because feedback mechanisms between ice
+sheet topography and regional climate such as wind redistribution, orographic
+precipitation and latent warming of moisture-depleted air are absent in our
+model, we regard this possibility as fully probable \citep{Seguinot.etal.2013}.
+
+On the other hand, in both simulations driven by the GRIP and EPICA records,
+the deglaciation of central British Columbia consists of a rapid northwards
+retreat associated with southwards-flowing, non-sliding ice lobes on the
+Interior Plateau, while the surrounding mountain ranges are already ice-free
+(Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}). Thus the model
+suggests that glacial lineations on the Interior Plateau of British Columbia
+may be of older age than the last glacial maximum, and may have remained intact
+throughout the deglaciation.
 
@@
            advance exceeded the last glacial maximum advanced are marked in
-           green. Hatches denote the Younger Dryas re-advance, which is more
+           green. Hatches denote re-advance of mountain-centred ice caps and
+           and the decaying ice sheet between 14 and 10\,kyr., which is more
            pronounced in the GRIP-driven simulation.}

### 2014-06-18 cordillera-cycle.tex
Polish deglaciation history.

@@
 glacier cover, most of the older evidence having been overridden by ice retreat
-retreat during the deglaciation \footnote{\textbf{Johan,} can we refer to one
+retreat during the deglaciation\footnote{\textbf{Johan,} can we refer to one
 of your papers here?}.
@@
 
-The possibly of a widespread late-glacial re-advance of decaying glaciers of
+The possibly of a widespread late-glacial in
 the North American Cordilleran has been debated for long. On the southern tip
@@
 Friele.Clague.2002, Friele.Clague.2002a, Kovanen.2002,
-Kovanen.Easterbrook.2002}. Additional evidence for a late-glacial re-advance
-was found in the Finlay River area, northern Rocky Mountains, where
+Kovanen.Easterbrook.2002}. In the Finlay River area, northern Rocky Mountains,
 sharp-crested moraines show an abundant remobilization of late-glacial
 sediments by alpine glaciers in interaction with the main body of the decaying
-ice sheet \citep{Lakeman.etal.2008}. Although further work is needed to
-constrain the timing of the North American Cordillera late-glacial re-advance,
-to seize its extent and identify potential climatic triggers
+ice sheet \citep{Lakeman.etal.2008}. Additional clues for a late-glacial
+re-advance include alpine glacier moraines in the eastern Coast Mountains,
+Rocky Mountains and Columbia Mountains \citep{Menounos.etal.2008}. Although
+further work is needed to constrain the timing of the late-glacial re-advance,
+to seize its extent and to identify potential climatic triggers
 \citep{Menounos.etal.2008}, it is interesting to note that the simulation
@@
 through an opening of the Liard lowland with radial margin retreat towards the
-surrounding mountain ranges and inward flow towards the depression, in full
+surrounding mountain ranges and inward flow towards the depression, in
 compatibility with the regional melt water record \citep{Margold.etal.2013}.
@@
 Once more, snapshots of modelled deglacial ice sheet configurations
-(Figs.~\ref{fig:deglacshots-grip} and~\ref{fig:deglacshots-epica}) provide a
+(Figs.~\ref{fig:deglacshots-grip} and~\ref{fig:deglacshots-epica}%
+\footnote{I produced one figure for each run, but as you could expect from
+    Fig.~\ref{fig:deglac}, Figs.~\ref{fig:deglacshots-grip}
+    and~\ref{fig:deglacshots-epica} look very similar. One of them would
+    probably be enough, or?}) provide a
 poor medium for direct comparison with geomorphological evidence of the ice
-retreat. Here, we compute a map of de-glacial, basal flow direction immediately
+retreat. Here, we compute a map of deglacial, basal flow directions immediately
 preceding exposure of ice-free land or cessation of sliding upon the approach
@@
 (\citealp{Prest.etal.1968}; \citealp[Fig.~1.12]{Clague.1989};
-\citealp[Fig.~2]{Kleman.etal.2010}; \citep[Fig.~2]{Margold.etal.2013}).
+\citealp[Fig.~2]{Kleman.etal.2010}; \citealp[Fig.~2]{Margold.etal.2013}).
 However, such is not the case on the Interior Plateau of British Columbia,
-where the model produce no deglacial flow directions (Fig.~\ref{fig:lastflow}),
-whereas glacial lineation maps indicate eastward flow.
+where both simulations predicts neglictible basal sliding during deglaciation
+(Fig.~\ref{fig:lastflow}), whereas glacial lineation maps indicate substantial
+eastward flow.
 
@@
 model, we regard this possibility as fully probable \citep{Seguinot.etal.2013}.
-
 On the other hand, in both simulations driven by the GRIP and EPICA records,
@@
 Interior Plateau, while the surrounding mountain ranges are already ice-free
-(Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}). Thus the model
-suggests that glacial lineations on the Interior Plateau of British Columbia
-may be of older age than the last glacial maximum, and may have remained intact
-throughout the deglaciation.
+(Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}). This result
+appears compatible with the prevailing conceptual model of deglaciation of
+central British Columbia, in which mountain ranges emerge from the ice before
+the plateau \citep[Fig.~7]{Fulton.1991}. If it is valid, it may suggest that
+glacial lineations on the Interior Plateau of British Columbia may be of older
+age than the last glacial maximum, and may have remained intact throughout the
+deglaciation.
 

### 2014-06-18 cordillera-cycle.tex
Polish conclusions.

@@
 whereas the EPICA-driven simulation produces a nearly-continuous deglaciation
-with very restrained glacial re-advance (Fig.~\ref{fig:deglac}, right panel).
+with very restrained glacial re-advance (Fig.~\ref{fig:deglac}, right panel)%
+\footnote{Should we mention our work in progress with Brian and/or the Tuya
+    Lake moraine?}.
 
@@
 MIS~4 (61.9--55.4\,kyr) and MIS~2 (29.5--16.9\,kyr), two periods
-corresponding to documented extensive palaeo-glaciations. This result is
-independent of the choice of palaeo-temperature record used to approximate the
-past evolution of climate forcing, and
-can be seen as a first-order agreement between the model and geomorphological
+corresponding to documented extensive glaciations. This result is
+independent of the choice of palaeo-temperature record used to approximate past
+climate evolution, and thus
+can be seen as a first-order agreement between the model and geological
 evidence. However, the timing of the two glaciation peaks depends sensitively
@@
 In all simulations presented here, ice cover is limited to disjoint mountain ice
-caps during most of the glacial cycle. This agrees with earlier interpretations
-of the geological evidence preceding the last glacial maximum. However, our
-simulations produce persistent ice cover over the Skeena mountains during
+caps during most of the glacial cycle, confirming previous inferences from
+geological evidence preceding the last glacial maximum. However, our
+simulations produce persistent ice cover on the Skeena mountains during
 the entire glacial cycle. At the time when a full-size Cordilleran ice sheet is
@@
 
-Concerning the deglaciation phase, when most geological evidence is available,
-none of the palaeo-temperature records used produces an ideal agreement between
-model results and geological evidence. The timing of the last glacial maximum
-and early deglaciation is generally best reproduced by forcing the model with
-Antarctic records. However, the Younger Dryas re-advance is only reproduced
-using temperature forcing from the GRIP record at places where it has been
-documented. Nonetheless, the patterns of deglaciation are consistent between
-both simulations driven by the GRIP and EPICA record. They show a rapid
-deglaciation of the southern half of the ice
-sheet, followed by unzipping of the Liard Lowland, and concentric retreat of
-the ice margin towards the last palaeo-ice cover in the Skeena Mountains
+During deglaciation, none of the palaeo-temperature records used produces an
+ideal agreement between model results and geological evidence. Although
+the EPICA record yields to more realistic timing of the last glacial maximum
+and early deglaciation, only the GRIP record produces a late-glacial re-advance
+at places where it has been documented in the literature. Nonetheless, the
+general patterns of deglaciation are consistent between both simulations driven
+by the GRIP and EPICA record, and show a rapid deglaciation of the southern
+half of the ice including rapid northwards retreat through the Interior
+Plateau of central British Comumbia, followed by an opening of the Liard
+Lowland, and final ice cap retreat of the ice margin towards the Selwyn and,
+then, the Skeena Mountains, which host the last remnant of the ice sheet
 during the early Holocene (10.9--9.5\,kyr).
 
-One must keep in mind, however, that these results are only accurate for our
-choices of ice-sheet model (PISM), surface mass balance model (PDD) and climate
-forcing. Most importantly, our simplistic
+These results strongly depend on our choices of ice-sheet model (PISM), surface
+mass balance model (PDD) and climate forcing. Most importantly, our simplistic
 palaeo-climate forcing does not include precipitation corrections in response
-to the presence of an ice cover, likely leading to overestimated glacial
-extent and volume. Nevertheless, our results identify that the largely
-understudied Skeena Mountains are a key area to understanding glacial dynamics
-of the Cordilleran ice sheet, highlighting the need for further geological
-investigation of this region.
+to the presence of an ice cover, potentially leading to overestimated glacial
+extent and volume in continental regions. Nevertheless, our results identify
+the glacially-scrubbed Skeena Mountains as a central, key area to understanding
+glacial dynamics of the Cordilleran ice sheet, highlighting the need for
+further geological investigation of this region.
 

### 2014-06-18 cordillera-cycle.tex
Stroeven et al 2014 has been published.

@@
 {Clague.1981, Porter.Swanson.1998}, cosmogenic dating \citep[e.g.,][]
-{Ward.etal.2007, Menounos.etal.2008, Stroeven.etal.2010, Stroeven.etal.2013},
+{Ward.etal.2007, Menounos.etal.2008, Stroeven.etal.2010, Stroeven.etal.2014},
 and the offshore sedimentary record \citep{Cosma.etal.2008, Davies.etal.2011}%

### 2014-06-18 cordillera-cycle.tex
Add acknowledgements.

@@
 
+% Author contributions
+\section*{Author contributions}
+\dots
+
 % Acknowledgements
-%\begin{acknowledgements}
-  % Author contributions
-  %\hack{\noindent}\textit{Author contributions.}
-%\end{acknowledgements}
+\begin{acknowledgements}
+We are very thankful to Constantine Khroulev, Ed Bueler and Andy Aschwanden for
+providing constant help and development on PISM. This work was supported by the
+Swedish Research Council~(VR) grant no. 2008-3449 to A.~P.~Stroeven, by the
+German Academic Exchange Service~(DAAD) grant no.~50015537 and a Knut and Alice
+Wallenberg Foundation grant to J.~Seguinot, and by NASA grant no.~NNX13AM16G
+to E.~Bueler. Computer resources were provided by the Swedish National
+Infrastructure for Computing (SNIC) allocation no. 2013/1-159 to A.~P.~Stroeven
+at the National Supercomputing Center (NSC).
+\end{acknowledgements}
 

### 2014-06-18 cordillera-cycle.tex
Add grid lines on timeseries plots.

@@
            MIS~2 (last glacial maximum, 29.5--16.9\,kyr). Dotted lines
-           correspond to the GRIP and EPICA 6\,km-resolution runs.
-           \todo{add grid lines to improve readability}}
+           correspond to the GRIP and EPICA 6\,km-resolution runs.}
   \label{fig:timeseries}
@@
            volume during the deglaciation, expressed in meters of sea-level
-           equivalent \textbf{(bottom)}.
-           \todo{add grid lines to improve readability}}
+           equivalent \textbf{(bottom)}.}
   \label{fig:deglacseries}

### 2014-06-18 figures/timeseries.py

@@ -86,5 +86,7 @@ ax2.set_ylabel('ice volume (m s.-l. eq.)')
 ax1.yaxis.set_label_coords(-0.05, 0.5)
 ax2.yaxis.set_label_coords(-0.05, 0.5)
 ax2.set_xlabel('model time (kyr)')
+ax1.grid(axis='y', c='0.5', ls='-', lw=0.1)
+ax2.grid(axis='y', c='0.5', ls='-', lw=0.1)
 ax2.legend(loc='upper left', ncol=2)
 fig.savefig('timeseries')

### 2014-06-18 cordillera-cycle.tex
Correct caption of deglacial flow directions fig.

@@
   \includegraphics{lastflow}
-  \caption{Modelled directions of the last basal ice velocities. Hatches
-           indicate areas that remain non-sliding throughout the simulation.
+  \caption{Modelled directions of the deglacial basal flow velocities. Hatches
+           indicate areas that remain non-sliding throughout deglaciation
+           (22.0--8.0\,kyr), notably including the Interior Plateau of central
+           British Columbia.
            Sliding grid cells were distinguished from non-sliding grid cells

### 2014-06-18 cordillera-cycle.tex
Add record scaling factors.

@@
     \middlehline
-    GRIP     & \chem{\delta^{18}O} & ?\% & -5.8{\degree}C
+    GRIP     & \chem{\delta^{18}O} & 0.35 & -5.8{\degree}C  % -16.4126
         & \citet{Dansgaard.etal.1993} \\
-    NGRIP    & \chem{\delta^{18}O} & ?\% & -6.1{\degree}C
+    NGRIP    & \chem{\delta^{18}O} & 0.22 & -6.1{\degree}C  % -26.7098
         & \citet{Andersen.etal.2004} \\
-    EPICA    & \chem{\delta^{18}O} & ?\% & -5.6{\degree}C
+    EPICA    & \chem{\delta^{18}O} & 0.60 & -5.6{\degree}C  % -9.2055
         & \citet{Jouzel.etal.2007} \\
-    Vostok   & \chem{\delta^{18}O} & ?\% & -5.6{\degree}C
+    Vostok   & \chem{\delta^{18}O} & 0.70 & -5.6{\degree}C  % -7.9550
         & \citet{Petit.etal.1999} \\
-    ODP~1012 & \chem{U^{K'}_{37}}  & ?\% & -5.8{\degree}C
+    ODP~1012 & \chem{U^{K'}_{37}}  & 1.53 & -5.8{\degree}C  % -3.7889
         & \citet{Herbert.etal.2001} \\
-    ODP~1020 & \chem{U^{K'}_{37}}  & ?\% & -5.8{\degree}C
+    ODP~1020 & \chem{U^{K'}_{37}}  & 1.16 & -5.8{\degree}C  % -5.0000
         & \citet{Herbert.etal.2001} \\

### 2014-06-18 cordillera-cycle.tex
Add geographic location of proxy records.

--- 2014-06-18 cordillera-cycle.tex
@@
            prepare temperature offset time-series used to force the ice sheet
-           model through the last 120\,kyr. $T_{[32;22]}$ refers to the
-           mean temperature anomaly during the period -32 to~-22~kyr after
-           scaling.}
+           model through the last 120\,kyr. $f$ corresponds to the scaling
+           factor used and $T_{[32;22]}$ refers to the mean temperature anomaly
+           during the period -32 to~-22~kyr after scaling.}
   \label{tab:records}
-  {\begin{tabular}{lcccc}
+  {\begin{tabular}{lccccccl}
     \tophline
-    Record & Proxy & Scaling factor & $T_{[32;22]}$ & Original reference\\
+    Record & Lat. & Lon. & Elev. & Proxy & $f$ & $T_{[32;22]}$
+           & Original reference\\
     \middlehline
-    GRIP     & \chem{\delta^{18}O} & ?\% & -5.8{\degree}C
+    GRIP     &  72{\degree} 35' N  % 72.58 (decimal)
+             &  37{\degree} 38' W  % 37.64 (decimal)
+             & 3238\,m
+             & \chem{\delta^{18}O}
+             & 0.35 & -5.8{\degree}C  % -16.4126 (before scaling)
              & \citet{Dansgaard.etal.1993} \\
-    NGRIP    & \chem{\delta^{18}O} & ?\% & -6.1{\degree}C
+
+    NGRIP    &  75{\degree} 06' N  % 75.10
+             &  42{\degree} 19' W  % 42.32
+             & 2917\,m
+             & \chem{\delta^{18}O}
+             & 0.22 & -6.1{\degree}C  % -26.7098
              & \citet{Andersen.etal.2004} \\
-    EPICA    & \chem{\delta^{18}O} & ?\% & -5.6{\degree}C
+
+    EPICA    &  75{\degree} 06' S  % 75.1
+             & 123{\degree} 21' E  % 123.35
+             & 3233\,m
+             & \chem{\delta^{18}O}
+             & 0.60 & -5.6{\degree}C  % -9.2055
              & \citet{Jouzel.etal.2007} \\
-    Vostok   & \chem{\delta^{18}O} & ?\% & -5.6{\degree}C
+
+    Vostok   &  78{\degree} 28' S  % 78.8
+             & 106{\degree} 50' E  % 106.8
+             & 3488\,m
+             & \chem{\delta^{18}O}
+             & 0.70 & -5.6{\degree}C  % -7.9550
              & \citet{Petit.etal.1999} \\
-    ODP~1012 & \chem{U^{K'}_{37}}  & ?\% & -5.8{\degree}C
+
+    ODP~1012 &  32{\degree} 17' N
+             & 118{\degree} 23' W
+             & -1772\,m
+             & \chem{U^{K'}_{37}}
+             & 1.53 & -5.8{\degree}C  % -3.7889
              & \citet{Herbert.etal.2001} \\
-    ODP~1020 & \chem{U^{K'}_{37}}  & ?\% & -5.8{\degree}C
+
+    ODP~1020 &  41{\degree} 00' N
+             & 126{\degree} 26' W
+             & -3038\,m
+             & \chem{U^{K'}_{37}}
+             & 1.16 & -5.8{\degree}C  % -5.0000
              & \citet{Herbert.etal.2001} \\

### 2014-06-19 cordillera-cycle.tex
Express all ages as positive numbers.

@@
 palaeo-temperature proxy records. Each simulation starts from assumed ice-free
-conditions at -120\,kyr, and runs to present. These computations were
+conditions at 120\,kyr, and runs to present. These computations were
 performed on 16 to 128 cores at the Swedish National Supercomputing Centre.
@@
            and surface velocity (colour mapping) from the GRIP simulation,
-           corresponding to the last glacial ice volume maximum (-19.1 kyr) and
+           corresponding to the last glacial ice volume maximum (19.1\,kyr) and
            the last deglaciation.}
@@
            and surface velocity (colour mapping) from the EPICA simulation,
-           corresponding to the last glacial ice volume maximum (-17.3 kyr) and
+           corresponding to the last glacial ice volume maximum (17.3\,kyr) and
            the last deglaciation.}

### 2014-06-19 cordillera-cycle.tex
Small language improvements.

@@
   restricted to high mountain areas. However, widespread precipitation over the
-  Skeena mountains favours the persistence of a central ice dome throughout the
+  Skeena Mountains favours the persistence of a central ice dome throughout the
   glacial cycle. It acts as a nucleation centre before the last glacial maximum
@@
 scientific computing tools \citep{Bueler.Brown.2009, Balay.etal.2014} allows
-high-resolution numerical modelling of palaeo-glaciers on mountainous terrain
-over millenial time scales \citep{Golledge.etal.2012}. In the North
-American Cordillera, this complex topography induces strong gradients in
+high-resolution modelling of palaeo-glaciers on mountainous terrain
+over millenial time scales \citep{Golledge.etal.2012}. In addition, complex
+topography of the North American Cordillera induces strong gradients in
 seasonality and distinct patterns of precipition, thus requiring the use of

### 2014-06-19 cordillera-cycle.tex
Update author list.

@@
 \Author[1]{A.~P.}{Stroeven}
+\Author[1]{J.}{Kleman}
 \runningauthor{J.~Seguinot et~al.}
@@
 However, such is not the case on the Interior Plateau of British Columbia,
-where both simulations predicts neglictible basal sliding during deglaciation
+where both simulations predicts negligible basal sliding during deglaciation
 (Fig.~\ref{fig:lastflow}), whereas glacial lineation maps indicate substantial

### 2014-06-23 kappa.tex
Add tentative outline.

% kappa.tex
% ----------------------------------------------------------------------

% Base class and packages
\documentclass{article}
\usepackage{lineno}
\usepackage[onehalfspacing]{setspace}
\usepackage{physics}
%\usepackage[colorlinks]{hyperref}
%\usepackage{natbib}
%\usepackage{graphicx}

% My commands
%\newcommand{\note}[1]{\textbf{[NOTE: #1]}}
%\newcommand{\todo}[1]{\emph{[\textbf{Todo:} #1]}}
%\newcommand{\aref}[0]{\textbf{[ref.]}}

% Document properties
\title{Numerical simulation of the Cordilleran ice sheet \\
       through the last glacial cycle}
\author{Julien Seguinot}

% ----------------------------------------------------------------------
\begin{document}
% ----------------------------------------------------------------------

\maketitle
\linenumbers

% ----------------------------------------------------------------------
\section{Introduction}
\label{sec:intro}
% ----------------------------------------------------------------------
% Ice as viscous matter
% Glaciers and ice sheets
% The last glacial cycle

% Fig. crevasses
% Fig. glacier flowing
% Fig. recent climate evolution
% Fig. present and palaeo-ice sheets

% ----------------------------------------------------------------------
\section{Field area description}
% ----------------------------------------------------------------------
% Geographic setting
% Climatic setting
% Palaeo-glaciology

% Fig. map of North America
% Fig. map of the Cordillera?
% Fig. photos of the Cordillera
% Fig. photos of fieldwork
% Fig. Johan lineation map?

% ----------------------------------------------------------------------
\section{Numerical ice flow model}
% ----------------------------------------------------------------------
% Overview
% Field equations
% Bedrock response
% Surface mass balance
% Atmospheric forcing
% Numerical implementation

% Fig. PISM = ice + bedrock

% ----------------------------------------------------------------------
\section{Software tools and contributions}
% ----------------------------------------------------------------------
% Computational workflow
% Contributions to PISM
% PyPDD
% pismplotlib
% r.in.worldclim
% r.out.pism

% Fig. flow chart
% Fig. pypdd example
% Fig. pismplotlib example?

% ----------------------------------------------------------------------
\section{Results summary}
% ----------------------------------------------------------------------
% Constant-climate runs
% Effect of daily temperature variability
% Transient-climate runs

% Fig. 3km LGM quiver
% Fig. sigma distribution map
% Fig. sigma effect on Cordillera runs
% Fig. deglaciation streamplot
% Fig. cumulative ice displacement map

% ----------------------------------------------------------------------
\section{Future perpectives}
% ----------------------------------------------------------------------
% Model improvements
% Fieldwork
% Geomorphological-oriented plots

% ----------------------------------------------------------------------
\end{document}
% ----------------------------------------------------------------------

### 2014-06-24 kappa.tex
Throw down some of the main equations.
Document field equations.

@@
 \documentclass{article}
+\usepackage{amsmath}
+\usepackage{bm}
 \usepackage{lineno}
+\usepackage{natbib}
 \usepackage[onehalfspacing]{setspace}
 \usepackage{physics}
-%\usepackage[colorlinks]{hyperref}
-%\usepackage{natbib}
+\usepackage[colorlinks, citecolor=blue]{hyperref}
 %\usepackage{graphicx}
@@
 %\newcommand{\aref}[0]{\textbf{[ref.]}}
+\renewcommand{\div}[1]{\mathrm{div}\,#1}
+\renewcommand{\grad}[1]{\vec{\mathrm{grad}}\,#1}
+\newcommand{\tensdiv}[1]{\vec{\mathrm{div}}\,#1}
+\newcommand{\tensgrad}[1]{\bm{\mathrm{grad}}\,#1}
+
+% Internal references
+\newcommand{\eqn}[1]{Eqn.~\ref{eqn:#1}}
+\newcommand{\fig}[1]{Fig.~\ref{fig:#1}}
+\newcommand{\tab}[1]{Table~\ref{tab:#1}}
 
@@
 % ----------------------------------------------------------------------
-\section{Numerical ice flow model}
+\section{Numerical ice sheet model}
 % ----------------------------------------------------------------------
-% Overview
-% Field equations
-% Bedrock response
-% Surface mass balance
-% Atmospheric forcing
-% Numerical implementation
+
+\subsection{Overview}
+
+The simulations presented in this thesis use Parallel Ice Sheet Model (PISM),
+an open-source, finite-difference, shallow ice-sheet model
+\citep{PISM-authors.2014}. The model reads bedrock topography, sea level,
+geothermal heat flux and climate forcing as inputs, and computes the thermal
+and dynamic state of the ice sheet, and the associated thermal and
+deformational response of an idealised lithosphere. The atmospheric and oceanic
+envelopes are not part of the model. However, PISM includes boundary models
+that provide simple parametrisations of their effect on the ice model at
+supposedly well-defined interfaces (Fig.~??).
 
 % Fig. PISM = ice + bedrock
+
+This section does not attempt to provide an extensive description of PISM,
+which can be found in the on-line documentation \citep{PISM-authors.2014}.
+Instead, it aims to highlight modelling choices made in the context of this
+thesis, and give an overview of the physics involved in these choices.
+Notations used and corresponding parameter choices are summarized in Table~??.
+
+\subsection{Field equations}
+
+The thermodynamic core of the ice sheet model relies on four elementary field
+equations. Firstly, ice flow is assumed \emph{incompressible}. Hence, the
+conservation of mass implies a conservation of volume, which can be expressed
+in terms of the ice velocity vector, $\vec{v}$, by
+
+\begin{equation}
+    \label{eqn:incompressibility}
+    \div{\vec{v}} = 0.
+\end{equation}
+
+Secondly, the \emph{balance of stresses} is expressed by the Stokes equation,
+thus assuming creeping flow. In other words, ice is considered as a slow-moving
+fluid whose deformation is entirely controlled by internal friction, here
+expressed by the Cauchy stress tensor, $\bm\sigma$, and gravity, $\vec{g}$,
+
+\begin{equation}
+    \label{eqn:stressbalance}
+    \tensdiv{\bm\sigma} + \rho\,\vec{g} = \vec{0},
+\end{equation}
+
+where $\rho$ corresponds to ice density. Thirdly, after defining the
+strain-rate tensor
+${\bm{\dot\epsilon}~=~\frac{1}{2}(\tensgrad{\vec{v}} +
+                                  \tensgrad{\vec{v}}^{\mathrm{T}})}$,
+and the deviatoric stress tensor, ${\bm{\tau} = \bm{\sigma} - p\,\bm{\delta}}$,
+where $p=\frac{1}{3}\tr(\bm{\sigma})$ is the hydrostatic pressure and
+$\bm{\delta}$ is the identity tensor, a \emph{constitutive law} for ice based
+on laboratory experiments is used to relate these two quantities,
+
+\begin{equation}
+    \label{eqn:glenslaw}
+    \bm{\dot\epsilon} = A\,\tau_e^{n-1}\,\bm{\tau}.
+\end{equation}
+
+The equivalent stress, $\tau_e$, is defined by
+${\tau_e}^2 = -\mathrm{II}_{\bm{\tau}} = \frac{1}{2} \tr(\bm{\tau}^2)$,
+where $\mathrm{II}_{\bm{\tau}}$ is the second invariant of the stress tensor.
+The ice softness coefficient, $A$, depends on the deviation of temperature $T$
+from the pressure-melting point $T_m$ through
+
+\begin{equation}
+    A = A_0\,e^\frac{-Q}{R(T-T_m)}.
+\end{equation}
+
+Letting ${\dot\epsilon_e}^2 = \frac{1}{2} \tr(\bm{\dot\epsilon}^2)$ and
+$B=A^{-1/n}$, the constitutive law (\eqn{glenslaw}) can be rewritten using a
+sometimes more convenient, inverted formulation,
+
+\begin{equation}
+    \bm{\tau} = B\,\dot{\epsilon}_e^{1/n-1}\,\bm{\dot\epsilon}.
+\end{equation}
+
+By analogy to Newtonian flow, an apparent viscosity, $\nu$, can then be defined
+by $\bm{\tau} = \nu\,\bm{\dot\epsilon}$, thus yielding
+
+\begin{equation}
+    \label{eqn:viscosity}
+    \nu = \frac{B}{2}\,\dot{\epsilon_e}^{1/n-1}.
+\end{equation}
+
+Finally, the evolution of temperature within the ice is governed by the
+\emph{conservation of energy}. In PISM, the conservation of energy uses an
+enthalpy formulation in order to account for thermodynamic effects associated
+with internal phase changes in the glacier. This enthalpy ($H$) formulation
+reads
+
+\begin{equation}
+    \label{eqn:enthalpy}
+    \rho \left(\pdv{H}{t} + \vec{v} \cdot \grad{H} \right)
+        = -\div{\vec{q}} + \frac{\nu\dot{\epsilon_e}^2}{4},
+\end{equation}
+
+where $\vec{q}$ represents the heat flow and
+${\frac{\nu\dot{\epsilon_e}^2}{4} = \tr(\bm{\tau}\bm{\dot\epsilon})}$ is a
+source term corresponding to strain heat release. In the case where ice
+temperature, $T$, is below the pressure-melting point, the enthalpy can be
+expressed as $H=c\,T$, while the heat flow is given by Fourier's law,
+$\vec{q} = k\,\grad{T}$. Consequently, the enthalpy equation (\eqn{enthalpy})
+can be rewritten using a more familiar, cold-ice, temperature formulation,
+
+\begin{equation}
+    \label{eqn:temperature}
+    \frac{\partial T}{\partial t}
+        + \vec{v} \cdot \vec{\mathrm{grad}}\,T
+        = \frac{k}{\rho c} \Delta T
+        + \frac{\nu\dot{\epsilon_e}^2}{4\rho c}.
+\end{equation}
+
+where $c$ denotes the specific heat capacity of the ice and $k$ is its thermal
+conductivity. Equations~\ref{eqn:incompressibility}, \ref{eqn:stressbalance},
+\ref{eqn:glenslaw}, and \ref{eqn:enthalpy} constitute the thermodynamic basis
+of our model. However, their resolution in full form implies a computational
+demand too high for multi-millennial, continental-scale applications, such as
+the numerical modelling of the Cordilleran ice sheet through the last glacial
+cycle, thus requiring shallow approximations.
+
+\subsection{Shallow approximations}
+
+Complete stress balance
+
+\begin{equation}
+    \left\{\begin{aligned}
+        \pdv{\tau_{xx}}{x} + \pdv{\tau_{xy}}{y} + \pdv{\tau_{xz}}{z}
+            &= \pdv{p}{x} \\
+        \pdv{\tau_{yx}}{x} + \pdv{\tau_{yy}}{y} + \pdv{\tau_{yz}}{z}
+            &= \pdv{p}{y} \\
+        \pdv{\tau_{zx}}{x} + \pdv{\tau_{zy}}{y} + \pdv{\tau_{zz}}{z}
+            &= \pdv{p}{z} - \rho g
+    \end{aligned}\right.
+\end{equation}
+
+Shallow ice approximation
+
+\begin{equation}
+    \left\{\begin{aligned}
+        \pdv{\tau_{xz}}{z} &= \pdv{p}{x} \\
+        \pdv{\tau_{yz}}{z} &= \pdv{p}{y} \\
+        0 &= \pdv{p}{z} - \rho g
+    \end{aligned}\right.
+\end{equation}
+
+\begin{equation}
+    \pdv{\vec{v}}{z} = 2A\,(\rho g)^n\,(s-z)^n\,|\grad{s}|^{n-1}\,\grad{s}
+\end{equation}
+
+Shallow shelf approximation
+
+\begin{equation}
+    \left\{\begin{aligned}
+        \pdv{\tau_{xx}}{x} + \pdv{\tau_{xy}}{y} + \pdv{\tau_{xz}}{z}
+            &= \pdv{p}{x} \\
+        \pdv{\tau_{yx}}{x} + \pdv{\tau_{yy}}{y} + \pdv{\tau_{yz}}{z}
+            &= \pdv{p}{y} \\
+        \pdv{\tau_{zz}}{z} &= \pdv{p}{z} - \rho g
+    \end{aligned}\right.
+\end{equation}
+
+\begin{equation}
+    \left\{\begin{aligned}
+        \pdv{x} \left[2\bar{\nu}H
+                      \left(2\pdv{v_x}{x} + \pdv{v_y}{y}\right) \right]
+            + \pdv{y} \left[\bar{\nu}H
+                            \left(\pdv{v_x}{y} + \pdv{v_y}{x}\right) \right]
+            + \tau_{bx} &= \rho_i gH \pdv{h}{x} \\
+        \pdv{x} \left[\bar{\nu}H
+                      \left(\pdv{v_x}{y} + \pdv{v_y}{x}\right) \right]
+            + \pdv{y} \left[2\bar{\nu}H
+                            \left(\pdv{v_x}{x} + 2\pdv{v_y}{y}\right) \right]
+            + \tau_{bx} &= \rho_i gH \pdv{h}{x} \\
+    \end{aligned}\right.
+\end{equation}
+
+\begin{equation}
+    \bar{\nu} = \frac{\bar{B}}{2}\,\dot{\epsilon}^{\frac{1-n}{n}}
+              = \frac{1}{2 \bar{A}^{\frac{1}{n}}}\,
+                \dot{\epsilon}^{\frac{1-n}{n}}
+\end{equation}
+
+Basal sliding
+
+\begin{equation}
+    \bm{\tau}_b = -\tau_c \frac{\vec{v}_b}
+                               {{u_{th}}^q\,|\vec{v}_b|^{1-q}}
+\end{equation}
+
+\begin{equation}
+   \tau_c = c_{0} + \tan{\phi}\cdot N_{til}
+\end{equation}
+
+\newcommand{\phimin}{\phi_{\mathrm{min}}}
+\newcommand{\phimax}{\phi_{\mathrm{max}}}
+\newcommand{\bmin}{b_{\mathrm{min}}}
+\newcommand{\bmax}{b_{\mathrm{max}}}
+
+\begin{equation}
+  \phi(x,y) =
+  \begin{cases}
+    \phimin, & b(x,y) \le \bmin, \\
+    \phimin + (b(x,y) - \bmin) \,M, & \bmin < b(x,y) < \bmax, \\
+    \phimax, & \bmax \le b(x,y).
+  \end{cases}
+  \label{eq:phipiecewise}
+\end{equation}
+
+\begin{equation}
+N_{til} = \delta P_o \, 10^{(e_0/C_c)
+          \left(1 - (W_{til}/W_{til}^{max})\right)}
+\end{equation}
+
+\subsection{Bedrock response}
+\subsection{Surface mass balance}
+\subsection{Atmospheric forcing}
+\subsection{Numerical implementation}
 
@@
 
+% References
+\newcommand{\urlprefix}[0]{}  % remove default "URL" prefix
+\bibliographystyle{copernicus}
+\bibliography{refs/references.bib}
+
 % ----------------------------------------------------------------------

### 2014-06-24 kappa.tex
Begin to document SIA and SSA (todos left).

@@
 %\newcommand{\note}[1]{\textbf{[NOTE: #1]}}
-%\newcommand{\todo}[1]{\emph{[\textbf{Todo:} #1]}}
+\newcommand{\todo}[1]{\emph{[\textbf{Todo:} #1]}}
 %\newcommand{\aref}[0]{\textbf{[ref.]}}
@@
 
-Complete stress balance
+PISM, the ice sheet model that we use, employ two well-documented
+approximations of ice flow equations: the shallow ice approximation and the
+the shallow shelf approximation. Although both approximations can be derived
+from a rigorous scaling approach, we don't expand these lengthy derivations
+here, assuming instead the simplifications and focusing on their effect.
+
+% Fig. SIA + SSA
+
+Distinguishing the hydrostatic and deviatoric components of the stress tensor,
+the balance of stresses (\eqn{stressbalance}) can be expanded into
 
 \begin{equation}
+    \label{eqn:stressbalance}
+    \tensdiv{\bm\tau} - \grad{p} + \rho\,\vec{g} = \vec{0}.
+\end{equation}
+
+Projecting along the three Cartesian coordinate axis, this reads
+
+\begin{equation}
+    \label{eqn:fullstokes}
     \left\{\begin{aligned}
@@
 
-Shallow ice approximation
+The shallow ice approximation (SIA) assumes that basal friction is high enough
+for the vertical shear stresses $\{\tau_{xz}, \tau_{yz}\}$ to predominate over
+all other components of the deviatoric stress tensor. In this framework, the
+projected stress-balance can be simplified to
 
@@
         0 &= \pdv{p}{z} - \rho g
-    \end{aligned}\right.
+    \end{aligned}\right..
 \end{equation}
 
+This yields
+
+\todo{develop intermediary steps}
+
 \begin{equation}
-    \pdv{\vec{v}}{z} = 2A\,(\rho g)^n\,(s-z)^n\,|\grad{s}|^{n-1}\,\grad{s}
+    \pdv{\vec{v}_{SIA}}{z} = 2A\,(\rho g)^n\,(s-z)^n\,
+                             |\grad{s}|^{n-1}\,\grad{s}.
 \end{equation}
 
-Shallow shelf approximation
+The shallow shelf approximation (SSA), on the opposite of the SIA, assumes that
+basal friction is low enough for ice to deform predominantly by horizontal
+expansion within the entire ice column. In this context, the stress-balance can
+be simplified to
 
@@
         \pdv{\tau_{zz}}{z} &= \pdv{p}{z} - \rho g
-    \end{aligned}\right.
+    \end{aligned}\right.,
 \end{equation}
+
+and horizontal velocities can be assumed independent of depth, so that
+
+\begin{equation}
+    \left\{\begin{aligned}
+        \pdv{v_x}{z} &= 0 \\
+        \pdv{v_y}{z} &= 0 \\
+    \end{aligned}\right..
+\end{equation}
+
+This yields
+
+\todo{develop intermediary steps}
 
@@
             + \tau_{bx} &= \rho_i gH \pdv{h}{x} \\
-    \end{aligned}\right.
+    \end{aligned}\right.,
 \end{equation}
 
+where $\bar{\nu}$ is the depth-integrated apparent viscosity defined by
+
 \begin{equation}
-    \bar{\nu} = \frac{\bar{B}}{2}\,\dot{\epsilon}^{\frac{1-n}{n}}
-              = \frac{1}{2 \bar{A}^{\frac{1}{n}}}\,
-                \dot{\epsilon}^{\frac{1-n}{n}}
+    \bar{\nu} = \frac{1}{H}\int_b^s\nu.
 \end{equation}
 
-Basal sliding
+In PISM, the SIA and SSA velocities are added,
+
+\begin{equation}
+    \label{eqn:siassa}
+    \vec{v} = \vec{v}_{SIA} + \vec{v}_{SSA}.
+\end{equation}
+
+Because $\vec{v}_{SSA}$ is small where $\vec{v}_{SIA}$ are large and
+reciprocally, we can expect the model to perform well in regions where
+conditions for applicability of either the SIA or the SSA are met. However,
+equation~\ref{eqn:siassa} is a heuristic, and its validity in the zone of
+transition between shear flow and longitudinal flow has not yet been
+mathematically proven.
+
+\subsection{Basal sliding}
 

### 2014-06-24 kappa.tex
Document basal sliding.

@@
 
+Although the SIA is valid only under non-sliding conditions, the SSA requires
+a sliding law as basal boundary condition, which in fact constitutes the main
+control on SSA velocities and in turn locations of ice streams.
+
+The simulations presented in this thesis use a pseudo-plastic sliding law,
+
 \begin{equation}
-    \bm{\tau}_b = -\tau_c \frac{\vec{v}_b}
-                               {{u_{th}}^q\,|\vec{v}_b|^{1-q}}
+    \label{eqn:pseudoplastic}
+    \vec{\tau}_b = -\tau_c \frac{\vec{v}_b}{{v_{th}}^q\,|\vec{v}_b|^{1-q}},
 \end{equation}
 
+where $\vec{\tau}_b$ is the basal traction force, $\vec{v}_b$ is the basal
+velocity, and $v_{th}$ is an arbitrary velocity threshold. In the case of
+$q=0$, equation~\ref{eqn:pseudoplastic} corresponds to a purely plastic law,
+
 \begin{equation}
-   \tau_c = c_{0} + \tan{\phi}\cdot N_{til}
+    \label{eqn:plastic}
+    \vec{\tau}_b = -\tau_c \frac{\vec{v}_b}{|\vec{v}_b|}.
 \end{equation}
 
-\newcommand{\phimin}{\phi_{\mathrm{min}}}
-\newcommand{\phimax}{\phi_{\mathrm{max}}}
-\newcommand{\bmin}{b_{\mathrm{min}}}
-\newcommand{\bmax}{b_{\mathrm{max}}}
+However a non-zero exponent is chosen here in order to improve convergence. The
+yield stress $\tau_c$ is related to till properties by the Mohr-Coulomb
+criterion,
+
+\begin{equation}
+   \tau_c = c_0 + N\,\tan{\phi}.
+\end{equation}
+
+The till cohesion $c_0$ is assumed to be zero. The effective pressure on the
+till, $N$, is determined by the modelled amount of water at the bed. However,
+two different parametrisations are used within this thesis. In Paper~I, using
+PISM~0.5, the effective pressure is linearly related to water content by a
+simple parametrisation,
+
+\begin{equation}
+    N = \rho gH (1 - \alpha \frac{W}{W_{max}}).
+\end{equation}
+
+In Paper~IV, using PISM~0.6, the effective pressure is determined by an
+empirical parametrisation based on laboratory experiments on Antarctic till,
+
+\begin{equation}
+    N = \delta \rho gH \, 10^{(e_0/C_c) (1 - (W/W_{max}))}
+\end{equation}
+
+Drainage of water at the base of the ice sheet, although implemented in PISM,
+is not included in simulations presented here for the sake of simplicity.
+In turn, the amount of water at the bed, $W$, corresponds to the local
+accumulation of basal melt-water. In varies from zero to $W_{max}$, a threshold
+above which further melt exits the system.
+
+Finally, the till friction angle $\phi$ is taken a~function of modern bed
+elevation, with lowest values occurring at low elevations, thus accounting for
+a weakening of till associated with the presence of marine sediments,
 
@@
   \begin{cases}
-    \phimin, & b(x,y) \le \bmin, \\
-    \phimin + (b(x,y) - \bmin) \,M, & \bmin < b(x,y) < \bmax, \\
-    \phimax, & \bmax \le b(x,y).
+    \phi_0, & \mathrm{for}\ b \le b_0, \\
+    \phi_0 + (\phi_1-\phi_0) \frac{b - b_0}{b_1-b_0},
+        & \mathrm{for}\ b_0 < b < b_1, \\
+    \phi_1, & \mathrm{for}\ b_1 \le b.
   \end{cases}
@@
 
-\begin{equation}
-N_{til} = \delta P_o \, 10^{(e_0/C_c)
-          \left(1 - (W_{til}/W_{til}^{max})\right)}
-\end{equation}
+where $b_0=0$\,m is modern sea-level and $b_1=200$\,m is a rough average of
+highest shoreline elevations in the model domain. The values of $\phi_1$ and
+$\phi_2$ are chosen differently in Paper~I (10 and 30$^\circ$) and Paper~IV
+(15 and 45$^\circ$), but these choices are roughly equivalent in the case of a
+saturated till ($W=W_{max}$).
 

### 2014-06-24 kappa.tex
Describe the bedrock response model.

@@
 \subsection{Bedrock response}
+
+The modelled ice sheet interacts with the underlying bedrock in two ways.
+Firstly, bedrock temperatures are computed to a given depth by appending the
+ice enthalpy model \eqn{enthalpy} with an underlying bedrock thermal model.
+The only process accounted for is heat conduction, hence this model consists of
+a simple diffusion equation.
+
+\begin{equation}
+    \pdv{T}{t} = \frac{k_b}{\rho_b c_b} \Delta T,
+\end{equation}
+
+where $k_b$, $\rho_b$ and $c_b$ are the thermal conductivity, density and
+specific heat capacity of the bedrock. Bedrock temperature is conditioned at
+depth by a fixed upwards geothermal heat flux $q_G$. The bedrock thermal model
+is necessary for modelling the Cordilleran ice sheet on multi-millennial
+time-scales, because temperature fluctuations caused by climate change and
+isolating effects of the ice sheet penetrate several hundred meters into the
+rock, and eventually affect basal ice temperatures.
+
+Secondly, bedrock elevation evolve in response to the ice load. The bedrock
+deformation model used in this thesis describes the flexure of an elastic
+lithosphere on top of an infinite half-space viscous mantle
+\citep{Lingle.Clark.1985}. It can be described by a single differential
+equation of the bed elevation $b$,
+
+\begin{equation}
+    2\nu_m\,|\grad|\,\pdv{b}{t} + \rho_l g b + D\,\Delta^2 b = \sigma_{zzb},
+\end{equation}
+
+where $\nu_m$ is mantle viscosity, $\rho_l$ is lithosphere density, $D$ is the
+lithosphere's flexural rigidity, and $\sigma_{zzb}$ corresponds to the ice load
+\citep{Bueler.etal.2007}. Here, $\Delta^2$ designs the biharmonic operator of
+linear elastic theory, while $|\grad|$ is a pseudo-differential operator
+defined through the Fourier transform by \citet[Eqn.~6]{Bueler.etal.2007}. In
+the left-hand part of this equation, the first term accounts for mantle
+relaxation, the second for point-wise isostasy, and the third for elastic
+flexure. Due to high mantle viscosity $\nu_m$, there exists a time lag between
+the ice sheet growth and the isostatic bedrock response.
+%$\nu=1\times10^{21}$\,Pa\,s
+%$D=5\times10^{24}$\,N\,m
+%$\rho_r = 3300$\,kg\,m$^{-3}$
+
 \subsection{Surface mass balance}

### 2014-06-24 kappa.tex
Improve header visibility in LaTeX source.
Add macros to formalise notations.

@@
 % kappa.tex
-% ----------------------------------------------------------------------
+% ======================================================================
 
@@
 %\newcommand{\aref}[0]{\textbf{[ref.]}}
-\renewcommand{\div}[1]{\mathrm{div}\,#1}
-\renewcommand{\grad}[1]{\vec{\mathrm{grad}}\,#1}
-\newcommand{\tensdiv}[1]{\vec{\mathrm{div}}\,#1}
-\newcommand{\tensgrad}[1]{\bm{\mathrm{grad}}\,#1}
+
+% Vectors and tensors
+\newcommand{\vect}[1]{\va*{#1}} % bold arrow vectors
+\newcommand{\tens}[1]{\vb*{#1}} % bold tensors
+
+% Differential operators
+\renewcommand{\div}[1]{\mathrm{div}\,#1}            % divergence
+\renewcommand{\grad}[1]{\vect{\mathrm{grad}}\,#1}   % gradient
+\newcommand{\tdiv}[1]{\vect{\mathrm{div}}\,#1}      % tensor divergence
+\newcommand{\tgrad}[1]{\tens{\mathrm{grad}}\,#1}    % tensor gradient
+\newcommand{\matdv}[1]{\pdv{#1}{t}+\vect{v}\cdot\grad{}\,#1}  % material dv.
+
+% Common notations
+\newcommand{\doteps}[0]{\dot{\epsilon}} % epsilon dot
+\newcommand{\IDT}[0]{\tens{\delta}}     % Identity tensor
+\newcommand{\CST}[0]{\tens{\sigma}}     % Cauchy stress tensor
+\newcommand{\DST}[0]{\tens{\tau}}       % deviatoric stress tensor
+\newcommand{\SRT}[0]{\tens{\doteps}}    % strain-rate tensor
+\newcommand{\vv}[0]{\vect{v}}           % velocity vector
+\newcommand{\vsia}[0]{\vv_{\mathrm{SIA}}}   % SIA velocity
+\newcommand{\vssa}[0]{\vv_{\mathrm{SSA}}}   % SSA velocity
 
@@
 
-% ----------------------------------------------------------------------
+% ======================================================================
 \begin{document}
-% ----------------------------------------------------------------------
+% ======================================================================
 
@@
 
-% ----------------------------------------------------------------------
+% ======================================================================
 \section{Introduction}
-\label{sec:intro}
-% ----------------------------------------------------------------------
+% ======================================================================
 % Ice as viscous matter
@@
 
-% ----------------------------------------------------------------------
+% ======================================================================
 \section{Field area description}
-% ----------------------------------------------------------------------
+% ======================================================================
 % Geographic setting
@@
 
+% ======================================================================
+\section{Numerical ice sheet model}
+% ======================================================================
+
 % ----------------------------------------------------------------------
-\section{Numerical ice sheet model}
+\subsection{Overview}
 % ----------------------------------------------------------------------
-
-\subsection{Overview}
 
@@
 
+% ----------------------------------------------------------------------
 \subsection{Field equations}
+% ----------------------------------------------------------------------
 
@@
 conservation of mass implies a conservation of volume, which can be expressed
-in terms of the ice velocity vector, $\vec{v}$, by
+in terms of the ice velocity vector, $\vv$, by
 
@@
     \label{eqn:incompressibility}
-    \div{\vec{v}} = 0.
+    \div{\vv} = 0.
 \end{equation}
@@
 fluid whose deformation is entirely controlled by internal friction, here
-expressed by the Cauchy stress tensor, $\bm\sigma$, and gravity, $\vec{g}$,
+expressed by the Cauchy stress tensor, $\CST$, and gravity, $\vect{g}$,
 
@@
     \label{eqn:stressbalance}
-    \tensdiv{\bm\sigma} + \rho\,\vec{g} = \vec{0},
+    \tdiv{\CST} + \rho\,\vect{g} = \vect{0},
 \end{equation}
@@
 strain-rate tensor
-${\bm{\dot\epsilon}~=~\frac{1}{2}(\tensgrad{\vec{v}} +
-                                  \tensgrad{\vec{v}}^{\mathrm{T}})}$,
-and the deviatoric stress tensor, ${\bm{\tau} = \bm{\sigma} - p\,\bm{\delta}}$,
-where $p=\frac{1}{3}\tr(\bm{\sigma})$ is the hydrostatic pressure and
-$\bm{\delta}$ is the identity tensor, a \emph{constitutive law} for ice based
+${\SRT = \frac{1}{2}(\tgrad{\vv} + \tgrad{\vv}^{\mathrm{T}})}$,
+and the deviatoric stress tensor, ${\DST = \CST - p\,\IDT}$,
+where $p=\frac{1}{3}\tr(\CST)$ is the hydrostatic pressure and
+$\tens{\delta}$ is the identity tensor, a \emph{constitutive law} for ice based
 on laboratory experiments is used to relate these two quantities,
@@
     \label{eqn:glenslaw}
-    \bm{\dot\epsilon} = A\,\tau_e^{n-1}\,\bm{\tau}.
+    \SRT = A\,\tau_e^{n-1}\,\DST.
 \end{equation}
@@
 The equivalent stress, $\tau_e$, is defined by
-${\tau_e}^2 = -\mathrm{II}_{\bm{\tau}} = \frac{1}{2} \tr(\bm{\tau}^2)$,
-where $\mathrm{II}_{\bm{\tau}}$ is the second invariant of the stress tensor.
+${\tau_e}^2 = -\mathrm{II}_{\DST} = \frac{1}{2} \tr(\DST^2)$,
+where $\mathrm{II}_{\DST}$ is the second invariant of the stress tensor.
 The ice softness coefficient, $A$, depends on the deviation of temperature $T$
@@
 
-Letting ${\dot\epsilon_e}^2 = \frac{1}{2} \tr(\bm{\dot\epsilon}^2)$ and
+Letting ${\doteps_e}^2 = \frac{1}{2} \tr(\SRT^2)$ and
 $B=A^{-1/n}$, the constitutive law (\eqn{glenslaw}) can be rewritten using a
@@
 \begin{equation}
-    \bm{\tau} = B\,\dot{\epsilon}_e^{1/n-1}\,\bm{\dot\epsilon}.
+    \DST = B\,\doteps_e^{1/n-1}\,\SRT.
 \end{equation}
@@
 By analogy to Newtonian flow, an apparent viscosity, $\nu$, can then be defined
-by $\bm{\tau} = \nu\,\bm{\dot\epsilon}$, thus yielding
+by $\DST = \nu\,\SRT$, thus yielding
 
@@
     \label{eqn:viscosity}
-    \nu = \frac{B}{2}\,\dot{\epsilon_e}^{1/n-1}.
+    \nu = \frac{B}{2}\,\doteps_e^{1/n-1}.
 \end{equation}
@@
     \label{eqn:enthalpy}
-    \rho \left(\pdv{H}{t} + \vec{v} \cdot \grad{H} \right)
-        = -\div{\vec{q}} + \frac{\nu\dot{\epsilon_e}^2}{4},
+    \rho \left(\matdv{H}\right)
+        = -\div{\vect{q}} + \frac{\nu\doteps_e^2}{4},
 \end{equation}
 
-where $\vec{q}$ represents the heat flow and
-${\frac{\nu\dot{\epsilon_e}^2}{4} = \tr(\bm{\tau}\bm{\dot\epsilon})}$ is a
+where $\vect{q}$ represents the heat flow and
+${\frac{\nu\dot{\epsilon_e}^2}{4} = \tr(\DST\SRT)}$ is a
 source term corresponding to strain heat release. In the case where ice
@@
 expressed as $H=c\,T$, while the heat flow is given by Fourier's law,
-$\vec{q} = k\,\grad{T}$. Consequently, the enthalpy equation (\eqn{enthalpy})
+$\vect{q} = k\,\grad{T}$. Consequently, the enthalpy equation (\eqn{enthalpy})
 can be rewritten using a more familiar, cold-ice, temperature formulation,
@@
     \label{eqn:temperature}
-    \frac{\partial T}{\partial t}
-        + \vec{v} \cdot \vec{\mathrm{grad}}\,T
-        = \frac{k}{\rho c} \Delta T
-        + \frac{\nu\dot{\epsilon_e}^2}{4\rho c}.
+    \matdv{T} = \frac{k}{\rho c} \Delta T
+                + \frac{\nu\doteps_e^2}{4\rho c}.
 \end{equation}
@@
 
+% ----------------------------------------------------------------------
 \subsection{Shallow approximations}
+% ----------------------------------------------------------------------
 
@@
     \label{eqn:stressbalance}
-    \tensdiv{\bm\tau} - \grad{p} + \rho\,\vec{g} = \vec{0}.
+    \tdiv{\DST} - \grad{p} + \rho\,\vect{g} = \vect{0}.
 \end{equation}
@@
 \begin{equation}
-    \pdv{\vec{v}_{SIA}}{z} = 2A\,(\rho g)^n\,(s-z)^n\,
-                             |\grad{s}|^{n-1}\,\grad{s}.
+    \pdv{\vsia}{z} = 2A\,(\rho g)^n\,(s-z)^n\,|\grad{s}|^{n-1}\,\grad{s}.
 \end{equation}
@@
     \label{eqn:siassa}
-    \vec{v} = \vec{v}_{SIA} + \vec{v}_{SSA}.
+    \vv = \vsia + \vssa.
 \end{equation}
 
-Because $\vec{v}_{SSA}$ is small where $\vec{v}_{SIA}$ are large and
+Because $\vsia$ is small where $\vssa$ is large and
 reciprocally, we can expect the model to perform well in regions where
@@
 
+% ----------------------------------------------------------------------
 \subsection{Basal sliding}
+% ----------------------------------------------------------------------
 
@@
     \label{eqn:pseudoplastic}
-    \vec{\tau}_b = -\tau_c \frac{\vec{v}_b}{{v_{th}}^q\,|\vec{v}_b|^{1-q}},
+    \vect{\tau}_b = -\tau_c \frac{\vv_b}{{v_{th}}^q\,|\vv_b|^{1-q}},
 \end{equation}
 
-where $\vec{\tau}_b$ is the basal traction force, $\vec{v}_b$ is the basal
+where $\vect{\tau}_b$ is the basal traction force, $\vv_b$ is the basal
 velocity, and $v_{th}$ is an arbitrary velocity threshold. In the case of
@@
     \label{eqn:plastic}
-    \vec{\tau}_b = -\tau_c \frac{\vec{v}_b}{|\vec{v}_b|}.
+    \vect{\tau}_b = -\tau_c \frac{\vv_b}{|\vv_b|}.
 \end{equation}
@@
 
+% ----------------------------------------------------------------------
 \subsection{Bedrock response}
+% ----------------------------------------------------------------------
 
@@
 lithosphere's flexural rigidity, and $\sigma_{zzb}$ corresponds to the ice load
-\citep{Bueler.etal.2007}. Here, $\Delta^2$ designs the biharmonic operator of
+\citep{Bueler.etal.2007}. Here, $\Delta^2$ designs the biharmonic (Laplacian
+square) operator of
 linear elastic theory, while $|\grad|$ is a pseudo-differential operator
@@
 
+% ----------------------------------------------------------------------
 \subsection{Surface mass balance}
-\subsection{Atmospheric forcing}
-\subsection{Numerical implementation}
+% ----------------------------------------------------------------------
 
 % ----------------------------------------------------------------------
+\subsection{Atmospheric forcing}
+% ----------------------------------------------------------------------
+
+% ----------------------------------------------------------------------
+\subsection{Numerical implementation}
+% ----------------------------------------------------------------------
+
+% ======================================================================
 \section{Software tools and contributions}
-% ----------------------------------------------------------------------
+% ======================================================================
 % Computational workflow
@@
 
-% ----------------------------------------------------------------------
+% ======================================================================
 \section{Results summary}
-% ----------------------------------------------------------------------
+% ======================================================================
 % Constant-climate runs
@@
 
-% ----------------------------------------------------------------------
+% ======================================================================
 \section{Future perpectives}
-% ----------------------------------------------------------------------
+% ======================================================================
 % Model improvements

### 2014-06-24 kappa.tex
Use cleveref package for homogeneized int. refs.

@@
 \usepackage[colorlinks, citecolor=blue]{hyperref}
+\usepackage[sort]{cleveref}  % needs to be loaded after hyperref
 %\usepackage{graphicx}
+
+% Copernicus-style internal references
+\crefname{chapter}{Chap.}{Chaps.}
+\crefname{equation}{Eq.}{Eqs.}
+\crefname{figure}{Fig.}{Figs.}
+\crefname{section}{Sect.}{Sects.}
+\crefname{table}{Table}{Tables}
 
@@
 \newcommand{\vssa}[0]{\vv_{\mathrm{SSA}}}   % SSA velocity
-
-% Internal references
-\newcommand{\eqn}[1]{Eqn.~\ref{eqn:#1}}
-\newcommand{\fig}[1]{Fig.~\ref{fig:#1}}
-\newcommand{\tab}[1]{Table~\ref{tab:#1}}
 
@@
 
-Letting ${\doteps_e}^2 = \frac{1}{2} \tr(\SRT^2)$ and
-$B=A^{-1/n}$, the constitutive law (\eqn{glenslaw}) can be rewritten using a
-sometimes more convenient, inverted formulation,
+Letting ${\doteps_e}^2 = \frac{1}{2} \tr(\SRT^2)$ and $B=A^{-1/n}$, the
+constitutive law (\ref{eqn:glenslaw}) can be rewritten using a sometimes more
+convenient, inverted formulation,
 
@@
 expressed as $H=c\,T$, while the heat flow is given by Fourier's law,
-$\vect{q} = k\,\grad{T}$. Consequently, the enthalpy equation (\eqn{enthalpy})
-can be rewritten using a more familiar, cold-ice, temperature formulation,
+$\vect{q} = k\,\grad{T}$. Consequently, the enthalpy equation
+(\ref{eqn:enthalpy}) can be rewritten using a more familiar, cold-ice,
+temperature formulation,
 
@@
 where $c$ denotes the specific heat capacity of the ice and $k$ is its thermal
-conductivity. Equations~\ref{eqn:incompressibility}, \ref{eqn:stressbalance},
-\ref{eqn:glenslaw}, and \ref{eqn:enthalpy} constitute the thermodynamic basis
+conductivity. \Cref{eqn:incompressibility,eqn:stressbalance,eqn:glenslaw,%
+eqn:enthalpy} constitute the thermodynamic basis
 of our model. However, their resolution in full form implies a computational
@@
 Distinguishing the hydrostatic and deviatoric components of the stress tensor,
-the balance of stresses (\eqn{stressbalance}) can be expanded into
+the balance of stresses (\ref{eqn:stressbalance}) can be expanded into
 
 \begin{equation}
-    \label{eqn:stressbalance}
     \tdiv{\DST} - \grad{p} + \rho\,\vect{g} = \vect{0}.
@@
 conditions for applicability of either the SIA or the SSA are met. However,
-equation~\ref{eqn:siassa} is a heuristic, and its validity in the zone of
+\cref{eqn:siassa} is a heuristic, and its validity in the zone of
 transition between shear flow and longitudinal flow has not yet been
@@
 velocity, and $v_{th}$ is an arbitrary velocity threshold. In the case of
-$q=0$, equation~\ref{eqn:pseudoplastic} corresponds to a purely plastic law,
+$q=0$, \cref{eqn:pseudoplastic} corresponds to a purely plastic law,
 
@@
 Firstly, bedrock temperatures are computed to a given depth by appending the
-ice enthalpy model \eqn{enthalpy} with an underlying bedrock thermal model.
-The only process accounted for is heat conduction, hence this model consists of
-a simple diffusion equation.
+ice enthalpy model (\ref{eqn:enthalpy}) with an underlying bedrock thermal
+model. The only process accounted for is heat conduction, hence this model
+consists of a simple diffusion equation.
 
@@
 linear elastic theory, while $|\grad|$ is a pseudo-differential operator
-defined through the Fourier transform by \citet[Eqn.~6]{Bueler.etal.2007}. In
+defined through the Fourier transform by \citet[Eq.~6]{Bueler.etal.2007}. In
 the left-hand part of this equation, the first term accounts for mantle

### 2014-06-24 kappa.tex
Add surface mass balance methods.

@@
 \newcommand{\vssa}[0]{\vv_{\mathrm{SSA}}}   % SSA velocity
+
+% Common units
+\newcommand{\unit}[1]{\ensuremath{\mathrm{#1}}}
+\newcommand{\degree}[0]{\ensuremath{^{\circ}}}
+\newcommand{\degC}[0]{\unit{{\degree}C}}
 
@@
     \end{cases}
-  \label{eq:phipiecewise}
 \end{equation}
@@
 relaxation, the second for point-wise isostasy, and the third for elastic
-flexure. Due to high mantle viscosity $\nu_m$, there exists a time lag between
+flexure. Due to high mantle viscosity, $\nu_m$, there exists a time lag between
 the ice sheet growth and the isostatic bedrock response.
@@
 % ----------------------------------------------------------------------
+
+\newcommand{\PDD}[0]{\mathrm{PDD}}
+\newcommand{\sPDD}[0]{\sigma_{\mathrm{PDD}}}
+
+At the interface between the ice sheet and the atmosphere (Fig.~??),
+surfice mass fluxes are computed from monthly mean surface air temperature,
+$T_m$, and monthly precipitation, $P_m$, by a temperature-index model
+\citep[e.g.,][]{Hock.2003}. Surface accumulation, $a_s$, equals precipitation
+when temperature is below ${T_0=0\,\degC}$, and decreases to zero linearly
+with temperature between $T_0$ and ${T_1=2\,\degC}$,
+
+\begin{equation}
+    a_s =
+    \begin{cases}
+        0,      & \mathrm{for}\ T_m \le T_0, \\
+        P_m \frac{T_m-T_0}{T_1-T_0},
+                & \mathrm{for}\ T_0 < b < T_1, \\
+        P_m,    & \mathrm{for}\ T_1 \le T_m,
+    \end{cases}
+\end{equation}
+
+Surface melt, $m_s$, is assumed proportional to the number of positive degree
+days (PDD), defined over an arbitrary time interval, $[t_1, t_2]$, as the
+integral of positive Celcius temperature $T-T_0$,
+
+\begin{equation}
+    \mathrm{PDD} = \int_{0}^{A}\max(T(t)-T_0,0)\,dt.
+\end{equation}
+
+For multimillenial simulations of palaeo-ice sheet evolution such as the ones
+presented in this thesis, daily or hourly temperature data is not available,
+forcing the PDD computation to rely on an idealised representation of the
+annual temperature cycle $T_{ac}$. Sub-annual temperature variability around
+the freezing point, however, significantly affects surface melt on a multi-year
+scale \citep{Arnold.Mackay.1964}. It is then typically included in the models
+under an assumption of normal temperature distribution, using a standard
+deviation parameter $\sPDD$ in the PDD computation \citep{Braithwaite.1984},
+PDDs can then be computed using a double-integral formulation
+\citep{Reeh.1991},
+
+\begin{equation}
+    \PDD = \frac{1}{\sPDD\sqrt{2\pi}}
+        \int_{t_1}^{t_2} \mathrm{d}t
+        \int_{0}^{\infty} \mathrm{d}T \,
+        T \exp\left({-\frac{(T-T_{ac})^2}{2\sPDD^2}}\right)
+\end{equation}
+
+or more efficiently using an error function formulation
+\citep{Calov.Greve.2005},
+
+\begin{equation}
+    \label{eqn:calovgreve}
+    \PDD = \int_{t_1}^{t_2} \mathrm{d}t
+        \left[\frac{\sPDD}{\sqrt{2\pi}}
+                \exp\left({-\frac{T_{ac}^2}{2\sPDD^2}}\right)
+              + \frac{T_{ac}}{2} \, \mathrm{erfc}
+                \left(-\frac{T_{ac}}{\sqrt{2}\sPDD}\right)\right],
+\end{equation}
+
+In the simulations presented here, \cref{eqn:calovgreve} is numerically
+approximated using week-long sub-intervals. For each sub-interval, surface melt
+is computed from PDDs and the available snow and ice depths using an algorithm
+documented in the on-line documentation of PISM \citep{PISM-authors.2014}. This
+algorythms employ different melt factors for snow ($F_s$) and for ice ($F_i$),
+as derived from mass-balance measurements on contemporary glaciers in the
+Coast and Rocky Mountains of British Columbia \citep{Shea.etal.2009}. In
+Paper~I, the temperature standard deviation $\sPDD$ is a constant model
+parameter. In Paper~IV, it is computed from daily temperature values from the
+North American Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006}, using
+a method initially developed in Paper~II, and further improved in Paper~III.
+%3.04\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for snow and
+%4.59\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for ice
 

### 2014-06-24 kappa.tex
Add atmospheric forcing subsection.

@@
 
+Atmospheric forcing of the model consists of a present-day monthly climatology,
+$\{T_{m0}, P_{m0}\}$, modified by a lapse-rate correction, ${\Delta}T_{LR}$,
+and a temperature-offset correction, ${\Delta}T_{TS}$,
+\begin{align}
+    T_m(t, x, y) &= T_{m0}(x, y) +{\Delta}T_{LR}(t) +{\Delta}T_{TS}(t, x, y),\\
+    P_m(t, x, y) &= P_{m0}(x, y).
+\end{align}
+
+The present-day climatology, $\{T_{m0}, P_{m0}\}$, is computed from
+near-surface air temperature and precipitation rate fields extracted from
+observational data and atmospheric reanalyses.
+The lapse-rate correction, ${\Delta}T_{LR}$, is computed in relation to the
+climate input reference bedrock topography, $b_{ref}$,
+\begin{align}
+    {\Delta}T_{LR}(t, x, y) &= -\gamma [s(t, x, y)-b_{ref}(x, y)] \\
+                            &= -\gamma [h(t, x, y)+b(t, x, y)-b_{ref}(x, y)],
+\end{align}
+
+thus accounting for the evolution of ice thickness, $h$, on the one hand, and
+for differences between the the ice flow model basal topography, $b$, and the
+climate input topography, $b_{ref}$, on the other hand. All simulations use an
+annual temperature lapse rate of $\gamma = 6\,\unit{\degC\,km^{-1}}$.
+
+Paper~I tests the sensitivity of the model to different input present-day
+climatologies, $\{T_{m0}, P_{m0}\}$, while the temperature-offset corrections,
+${\Delta}T_{TS}$, uses constant values ranging from -10 to 0\,\degC. Paper~IV,
+on the opposite, tests model sensitivity to time-dependent temperature-offset
+corrections derived from different palaeo-temperature proxy records.
+
 % ----------------------------------------------------------------------

### 2014-06-25 kappa.tex
Develop SIA derivation.

@@
 \begin{equation}
+    \label{eqn:sia}
     \left\{\begin{aligned}
@@
 
-This yields
+The $z$-component of \cref{eqn:sia} gives the pressure (neglecting atmospheric
+pressure),
 
-\todo{develop intermediary steps}
+\begin{equation}
+    p = \rho g (s-z),
+\end{equation}
+
+which can be re-introduced into the $x$ and $y$-components, thus yielding the
+expression of the horizontal shear stresses,
+
+\begin{equation}
+    \left\{\begin{aligned}
+        \tau_{xz} &= \rho g (s-z) \pdv{s}{x} \\
+        \tau_{yz} &= \rho g (s-z) \pdv{s}{x} \\
+    \end{aligned}\right.,
+\end{equation}
+
+which are sometimes referred to as driving stresses. Using the constitutive
+law (\ref{eqn:glenslaw}), vertical derivatives of horizontal velocities can
+be related to the horizontal shear stresses,
+
+\begin{equation}
+    \left\{\begin{aligned}
+        \pdv{v_x}{z} &= 2A (\tau_{xz}^2 + \tau_{yz}^2)^{n-1} \tau_{xz}\\
+        \pdv{v_y}{z} &= 2A (\tau_{xz}^2 + \tau_{yz}^2)^{n-1} \tau_{yz}\\
+    \end{aligned}\right..
+\end{equation}
+
+Thus, within the framework of the SIA, horizontal velocities $\vv_{SIA}$ can be
+directly expressed as a function of the slope gradient,
+
+\begin{equation}
+    \left\{\begin{aligned}
+        \pdv{v_x}{z} &= 2A (\rho g)^n (s-z)^n
+                           (\pdv{s}{x}+\pdv{s}{y})^{n-1} \pdv{s}{x} \\
+        \pdv{v_y}{z} &= 2A (\rho g)^n (s-z)^n
+                           (\pdv{s}{x}+\pdv{s}{y})^{n-1} \pdv{s}{y} \\
+    \end{aligned}\right.,
+\end{equation}
+
+or, in vectorial notation,
 
@@
 \begin{equation}
+    \label{eqn:ssa}
     \left\{\begin{aligned}

### 2014-06-25 kappa.tex
Lighten typesetting of eqn systems: no braces.
Improve punctuation in math throughout.

--- 2014-06-24 kappa.tex
@@
 The thermodynamic core of the ice sheet model relies on four elementary field
-equations. Firstly, ice flow is assumed \emph{incompressible}. Hence, the
+equations. Firstly, ice flow is assumed \emph{incompressible}. Hence, the ice
+density, $\rho$, is constant, and the
 conservation of mass implies a conservation of volume, which can be expressed
 in terms of the ice velocity vector, $\vv$, by
-
 \begin{equation}
     \label{eqn:incompressibility}
-    \div{\vv} = 0.
+    \div{\vv} = 0 \,.
 \end{equation}
@@
 fluid whose deformation is entirely controlled by internal friction, here
-expressed by the Cauchy stress tensor, $\CST$, and gravity, $\vect{g}$,
-
+expressed by the Cauchy stress tensor, $\CST$, and gravity, $\vect{g}$:
 \begin{equation}
     \label{eqn:stressbalance}
-    \tdiv{\CST} + \rho\,\vect{g} = \vect{0},
+    \tdiv{\CST} + \rho\,\vect{g} = \vect{0} \,.
 \end{equation}
 
-where $\rho$ corresponds to ice density. Thirdly, after defining the
-strain-rate tensor
+Thirdly, after defining the strain-rate tensor
 ${\SRT = \frac{1}{2}(\tgrad{\vv} + \tgrad{\vv}^{\mathrm{T}})}$,
@@
 $\tens{\delta}$ is the identity tensor, a \emph{constitutive law} for ice based
-on laboratory experiments is used to relate these two quantities,
-
+on laboratory experiments is used to relate these two quantities. It reads:
 \begin{equation}
     \label{eqn:glenslaw}
-    \SRT = A\,\tau_e^{n-1}\,\DST.
+    \SRT = A\,\tau_e^{n-1}\,\DST \,.
 \end{equation}
-
 The equivalent stress, $\tau_e$, is defined by
@@
 from the pressure-melting point $T_m$ through
-
 \begin{equation}
-    A = A_0\,e^\frac{-Q}{R(T-T_m)}.
+    A = A_0\,e^\frac{-Q}{R(T-T_m)} \,.
 \end{equation}
@@
 convenient, inverted formulation,
-
 \begin{equation}
-    \DST = B\,\doteps_e^{1/n-1}\,\SRT.
+    \DST = B\,\doteps_e^{1/n-1}\,\SRT \,.
 \end{equation}
-
 By analogy to Newtonian flow, an apparent viscosity, $\nu$, can then be defined
@@
     \label{eqn:viscosity}
-    \nu = \frac{B}{2}\,\doteps_e^{1/n-1}.
+    \nu = \frac{B}{2}\,\doteps_e^{1/n-1} \,.
 \end{equation}
@@
 reads
-
 \begin{equation}
@@
     \rho \left(\matdv{H}\right)
-        = -\div{\vect{q}} + \frac{\nu\doteps_e^2}{4},
+        = -\div{\vect{q}} + \frac{\nu\doteps_e^2}{4} \,,
 \end{equation}
-
 where $\vect{q}$ represents the heat flow and
@@
 temperature formulation,
-
 \begin{equation}
@@
     \matdv{T} = \frac{k}{\rho c} \Delta T
-                + \frac{\nu\doteps_e^2}{4\rho c}.
+                + \frac{\nu\doteps_e^2}{4\rho c} \,,
 \end{equation}
+where $c$ denotes the specific heat capacity of the ice and $k$ is its thermal
+conductivity.
 
-where $c$ denotes the specific heat capacity of the ice and $k$ is its thermal
-conductivity. \Cref{eqn:incompressibility,eqn:stressbalance,eqn:glenslaw,%
-eqn:enthalpy} constitute the thermodynamic basis
+\Cref{eqn:incompressibility,eqn:stressbalance,eqn:glenslaw,eqn:enthalpy}
+constitute the thermodynamic basis
 of our model. However, their resolution in full form implies a computational
@@
 the balance of stresses (\ref{eqn:stressbalance}) can be expanded into
-
 \begin{equation}
-    \tdiv{\DST} - \grad{p} + \rho\,\vect{g} = \vect{0}.
+    \tdiv{\DST} - \grad{p} + \rho\,\vect{g} = \vect{0} \,.
 \end{equation}
@@
 Projecting along the three Cartesian coordinate axis, this reads
-
-\begin{equation}
+\begin{subequations}
 \label{eqn:fullstokes}
-    \left\{\begin{aligned}
+\begin{align}
     \pdv{\tau_{xx}}{x} + \pdv{\tau_{xy}}{y} + \pdv{\tau_{xz}}{z}
-            &= \pdv{p}{x} \\
+        &= \pdv{p}{x} \,, \\
     \pdv{\tau_{yx}}{x} + \pdv{\tau_{yy}}{y} + \pdv{\tau_{yz}}{z}
-            &= \pdv{p}{y} \\
+        &= \pdv{p}{y} \,, \\
     \pdv{\tau_{zx}}{x} + \pdv{\tau_{zy}}{y} + \pdv{\tau_{zz}}{z}
-            &= \pdv{p}{z} - \rho g
-    \end{aligned}\right.
-\end{equation}
+        &= \pdv{p}{z} - \rho g \,.
+\end{align}
+\end{subequations}
 
@@
 projected stress-balance can be simplified to
+\begin{subequations}
+\label{eqn:sia}
+\begin{align}
+    \pdv{\tau_{xz}}{z} &= \pdv{p}{x} \,, \\
+    \pdv{\tau_{yz}}{z} &= \pdv{p}{y} \,, \\
+    0 &= \pdv{p}{z} - \rho g \,.
+\end{align}
+\end{subequations}
 
+The $z$-component of \cref{eqn:sia} gives the pressure (neglecting atmospheric
+pressure),
 \begin{equation}
-    \left\{\begin{aligned}
-        \pdv{\tau_{xz}}{z} &= \pdv{p}{x} \\
-        \pdv{\tau_{yz}}{z} &= \pdv{p}{y} \\
-        0 &= \pdv{p}{z} - \rho g
-    \end{aligned}\right..
+    p = \rho g (s-z) \,,
 \end{equation}
+which can be re-introduced into the $x$ and $y$-components, thus yielding the
+expression of the horizontal shear stresses, sometimes referred to as driving
+stresses,
+\begin{subequations}
+\begin{align}
+    \tau_{xz} &= \rho g (s-z) \pdv{s}{x} \,, \\
+    \tau_{yz} &= \rho g (s-z) \pdv{s}{x} \,.
+\end{align}
+\end{subequations}
 
-This yields
+Using the constitutive law (\ref{eqn:glenslaw}), vertical derivatives of
+horizontal velocities can be related to the horizontal shear stresses:
+\begin{subequations}
+\begin{align}
+    \pdv{v_x}{z} &= 2A (\tau_{xz}^2 + \tau_{yz}^2)^{n-1} \tau_{xz} \,, \\
+    \pdv{v_y}{z} &= 2A (\tau_{xz}^2 + \tau_{yz}^2)^{n-1} \tau_{yz} \,.
+\end{align}
+\end{subequations}
 
-\todo{develop intermediary steps}
-
+Thus, within the framework of the SIA, horizontal velocities $\vv_{SIA}$ can be
+directly expressed as a function of the slope gradient,
+\begin{subequations}
+\begin{align}
+    \pdv{v_x}{z} &= 2A (\rho g)^n (s-z)^n
+                    \left(\pdv{s}{x}+\pdv{s}{y}\right)^{n-1} \pdv{s}{x} \,, \\
+    \pdv{v_y}{z} &= 2A (\rho g)^n (s-z)^n
+                    \left(\pdv{s}{x}+\pdv{s}{y}\right)^{n-1} \pdv{s}{y} \,,
+\end{align}
+\end{subequations}
+which yields, in vectorial notation,
 \begin{equation}
-    \pdv{\vsia}{z} = 2A\,(\rho g)^n\,(s-z)^n\,|\grad{s}|^{n-1}\,\grad{s}.
+    \pdv{\vsia}{z} = 2A\,(\rho g)^n\,(s-z)^n\,|\grad{s}|^{n-1}\,\grad{s} \,.
 \end{equation}
@@
 be simplified to
-
-\begin{equation}
-    \left\{\begin{aligned}
+\begin{subequations}
+\label{eqn:ssa}
+\begin{align}
     \pdv{\tau_{xx}}{x} + \pdv{\tau_{xy}}{y} + \pdv{\tau_{xz}}{z}
-            &= \pdv{p}{x} \\
+        &= \pdv{p}{x} \,, \\
     \pdv{\tau_{yx}}{x} + \pdv{\tau_{yy}}{y} + \pdv{\tau_{yz}}{z}
-            &= \pdv{p}{y} \\
-        \pdv{\tau_{zz}}{z} &= \pdv{p}{z} - \rho g
-    \end{aligned}\right.,
-\end{equation}
-
+        &= \pdv{p}{y} \,, \\
+    \pdv{\tau_{zz}}{z} &= \pdv{p}{z} - \rho g \,,
+\end{align}
+\end{subequations}
 and horizontal velocities can be assumed independent of depth, so that
-
-\begin{equation}
-    \left\{\begin{aligned}
-        \pdv{v_x}{z} &= 0 \\
-        \pdv{v_y}{z} &= 0 \\
-    \end{aligned}\right..
-\end{equation}
+\begin{subequations}
+\begin{align}
+    \pdv{v_x}{z} &= 0 \,, \\
+    \pdv{v_y}{z} &= 0 \,.
+\end{align}
+\end{subequations}
 
 This yields
-
 \todo{develop intermediary steps}
-
-\begin{equation}
-    \left\{\begin{aligned}
+\begin{subequations}
+\begin{align}
     \pdv{x} \left[2\bar{\nu}H
@@
                         \left(\pdv{v_x}{y} + \pdv{v_y}{x}\right) \right]
-            + \tau_{bx} &= \rho_i gH \pdv{h}{x} \\
+        + \tau_{bx} &= \rho_i gH \pdv{h}{x} \,, \\
     \pdv{x} \left[\bar{\nu}H
@@
                         \left(\pdv{v_x}{x} + 2\pdv{v_y}{y}\right) \right]
-            + \tau_{bx} &= \rho_i gH \pdv{h}{x} \\
-    \end{aligned}\right.,
+        + \tau_{bx} &= \rho_i gH \pdv{h}{x} \,,
+\end{align}
+\end{subequations}
+where $\bar{\nu}$ is the depth-integrated apparent viscosity defined by
+\begin{equation}
+    \bar{\nu} = \frac{1}{H}\int_b^s\nu \,.
 \end{equation}
 
-where $\bar{\nu}$ is the depth-integrated apparent viscosity defined by
-
-\begin{equation}
-    \bar{\nu} = \frac{1}{H}\int_b^s\nu.
-\end{equation}
-
-In PISM, the SIA and SSA velocities are added,
-
+In PISM, the SIA and SSA velocities are eventually added,
 \begin{equation}
     \label{eqn:siassa}
-    \vv = \vsia + \vssa.
+    \vv = \vsia + \vssa \,.
 \end{equation}
-
 Because $\vsia$ is small where $\vssa$ is large and
@@
 The simulations presented in this thesis use a pseudo-plastic sliding law,
-
 \begin{equation}
     \label{eqn:pseudoplastic}
-    \vect{\tau}_b = -\tau_c \frac{\vv_b}{{v_{th}}^q\,|\vv_b|^{1-q}},
+    \vect{\tau}_b = -\tau_c \frac{\vv_b}{{v_{th}}^q\,|\vv_b|^{1-q}} \,,
 \end{equation}
-
 where $\vect{\tau}_b$ is the basal traction force, $\vv_b$ is the basal
@@
 $q=0$, \cref{eqn:pseudoplastic} corresponds to a purely plastic law,
-
 \begin{equation}
     \label{eqn:plastic}
-    \vect{\tau}_b = -\tau_c \frac{\vv_b}{|\vv_b|}.
+    \vect{\tau}_b = -\tau_c \frac{\vv_b}{|\vv_b|} \,.
 \end{equation}
@@
 criterion,
-
 \begin{equation}
-   \tau_c = c_0 + N\,\tan{\phi}.
+   \tau_c = c_0 + N\,\tan{\phi} \,.
 \end{equation}
@@
 simple parametrisation,
-
 \begin{equation}
-    N = \rho gH (1 - \alpha \frac{W}{W_{max}}).
+    N = \rho gH (1 - \alpha \frac{W}{W_{max}}) \,.
 \end{equation}
@@
 empirical parametrisation based on laboratory experiments on Antarctic till,
-
 \begin{equation}
-    N = \delta \rho gH \, 10^{(e_0/C_c) (1 - (W/W_{max}))}
+    N = \delta \rho gH \, 10^{(e_0/C_c) (1 - (W/W_{max}))} \,.
 \end{equation}
@@
 elevation, with lowest values occurring at low elevations, thus accounting for
-a weakening of till associated with the presence of marine sediments,
-
+a weakening of till associated with the presence of marine sediments:
 \begin{equation}
@@
     \begin{cases}
-        \phi_0, & \mathrm{for}\ b \le b_0, \\
-        \phi_0 + (\phi_1-\phi_0) \frac{b - b_0}{b_1-b_0},
-            & \mathrm{for}\ b_0 < b < b_1, \\
-        \phi_1, & \mathrm{for}\ b_1 \le b.
+        \phi_0 & \text{if}\ b \le b_0 \,, \\
+        \phi_0 + (\phi_1-\phi_0) \frac{b - b_0}{b_1-b_0}
+                & \text{if}\ b_0 < b < b_1 \,, \\
+        \phi_1 & \text{if}\ b_1 \le b. \,,
     \end{cases}
 \end{equation}
-
 where $b_0=0$\,m is modern sea-level and $b_1=200$\,m is a rough average of
@@
 model. The only process accounted for is heat conduction, hence this model
-consists of a simple diffusion equation.
-
+consists of a simple diffusion equation,
 \begin{equation}
-    \pdv{T}{t} = \frac{k_b}{\rho_b c_b} \Delta T,
+    \pdv{T}{t} = \frac{k_b}{\rho_b c_b} \Delta T \,,
 \end{equation}
-
 where $k_b$, $\rho_b$ and $c_b$ are the thermal conductivity, density and
@@
 \citep{Lingle.Clark.1985}. It can be described by a single differential
-equation of the bed elevation $b$,
-
+equation of the bed elevation, $b$,
 \begin{equation}
-    2\nu_m\,|\grad|\,\pdv{b}{t} + \rho_l g b + D\,\Delta^2 b = \sigma_{zzb},
+    2\nu_m\,|\grad|\,\pdv{b}{t} + \rho_l g b + D\,\Delta^2 b = \sigma_{zzb} \,,
 \end{equation}
-
 where $\nu_m$ is mantle viscosity, $\rho_l$ is lithosphere density, $D$ is the
@@
 with temperature between $T_0$ and ${T_1=2\,\degC}$,
-
 \begin{equation}
@@
     \begin{cases}
-        0,      & \mathrm{for}\ T_m \le T_0, \\
-        P_m \frac{T_m-T_0}{T_1-T_0},
-                & \mathrm{for}\ T_0 < b < T_1, \\
-        P_m,    & \mathrm{for}\ T_1 \le T_m,
+        0       & \text{if}\ T_m \le T_0 \,, \\
+        P_m \frac{T_m-T_0}{T_1-T_0}
+                & \text{if}\ T_0 < b < T_1 \,, \\
+        P_m     & \text{if}\ T_1 \le T_m \,,
     \end{cases}
@@
 integral of positive Celcius temperature $T-T_0$,
-
 \begin{equation}
-    \mathrm{PDD} = \int_{0}^{A}\max(T(t)-T_0,0)\,dt.
+    \mathrm{PDD} = \int_{0}^{A}\max(T(t)-T_0,0)\,dt \,.
 \end{equation}
@@
 \citep{Reeh.1991},
-
 \begin{equation}
@@
         \int_{0}^{\infty} \mathrm{d}T \,
-        T \exp\left({-\frac{(T-T_{ac})^2}{2\sPDD^2}}\right)
+        T \exp\left({-\frac{(T-T_{ac})^2}{2\sPDD^2}}\right) \,,
 \end{equation}
-
 or more efficiently using an error function formulation
 \citep{Calov.Greve.2005},
-
 \begin{equation}
@@
               + \frac{T_{ac}}{2} \, \mathrm{erfc}
-                \left(-\frac{T_{ac}}{\sqrt{2}\sPDD}\right)\right],
+                \left(-\frac{T_{ac}}{\sqrt{2}\sPDD}\right)\right] \,.
 \end{equation}
@@
 and a temperature-offset correction, ${\Delta}T_{TS}$,
+\begin{subequations}
 \begin{align}
-    T_m(t, x, y) &= T_{m0}(x, y) +{\Delta}T_{LR}(t) +{\Delta}T_{TS}(t, x, y),\\
-    P_m(t, x, y) &= P_{m0}(x, y).
+    T_m(t, x, y) &= T_{m0}(x, y) + {\Delta}T_{LR}(t)
+                                 + {\Delta}T_{TS}(t, x, y) \,, \\
+    P_m(t, x, y) &= P_{m0}(x, y) \,.
 \end{align}
+\end{subequations}
 
@@
     {\Delta}T_{LR}(t, x, y) &= -\gamma [s(t, x, y)-b_{ref}(x, y)] \\
-                            &= -\gamma [h(t, x, y)+b(t, x, y)-b_{ref}(x, y)],
+                            &= -\gamma [h(t, x, y)+b(t, x, y)-b_{ref}(x, y)]\,,
 \end{align}
-
 thus accounting for the evolution of ice thickness, $h$, on the one hand, and

### 2014-06-25 kappa.tex
Develop SSA derivation.

@@
 \begin{equation}
+    \label{eqn:invglenslaw}
     \DST = B\,\doteps_e^{1/n-1}\,\SRT \,.
@@
 \begin{align}
+    \label{eqn:siax}
     \pdv{\tau_{xz}}{z} &= \pdv{p}{x} \,, \\
+    \label{eqn:siay}
     \pdv{\tau_{yz}}{z} &= \pdv{p}{y} \,, \\
+    \label{eqn:siaz}
     0 &= \pdv{p}{z} - \rho g \,.
@@
 
-The $z$-component of \cref{eqn:sia} gives the pressure (neglecting atmospheric
-pressure),
+Neglecting atmospheric pressure, \Cref{eqn:siaz} yield
 \begin{equation}
@@
 \end{equation}
-which can be re-introduced into the $x$ and $y$-components, thus yielding the
+which can be re-introduced into \cref{eqn:siax,eqn:siay}, thus yielding the
 expression of the horizontal shear stresses, sometimes referred to as driving
@@
 \begin{align}
-    \tau_{xz} &= \rho g (s-z) \pdv{s}{x} \,, \\
-    \tau_{yz} &= \rho g (s-z) \pdv{s}{x} \,.
+    \tau_{xz} &= -\rho g (s-z) \pdv{s}{x} \,, \\
+    \tau_{yz} &= -\rho g (s-z) \pdv{s}{x} \,.
 \end{align}
@@
 \begin{align}
-    \pdv{v_x}{z} &= 2A (\rho g)^n (s-z)^n
+    \pdv{v_x}{z} &= -2A (\rho g)^n (s-z)^n
                     \left(\pdv{s}{x}+\pdv{s}{y}\right)^{n-1} \pdv{s}{x} \,, \\
-    \pdv{v_y}{z} &= 2A (\rho g)^n (s-z)^n
+    \pdv{v_y}{z} &= -2A (\rho g)^n (s-z)^n
                     \left(\pdv{s}{x}+\pdv{s}{y}\right)^{n-1} \pdv{s}{y} \,,
@@
 \begin{equation}
-    \pdv{\vsia}{z} = 2A\,(\rho g)^n\,(s-z)^n\,|\grad{s}|^{n-1}\,\grad{s} \,.
+    \label{eqn:vsia}
+    \pdv{\vsia}{z} = -2A\,(\rho g)^n\,(s-z)^n\,|\grad{s}|^{n-1}\,\grad{s} \,.
 \end{equation}
@@
 \begin{align}
+    \label{eqn:ssax}
     \pdv{\tau_{xx}}{x} + \pdv{\tau_{xy}}{y} + \pdv{\tau_{xz}}{z}
         &= \pdv{p}{x} \,, \\
+    \label{eqn:ssay}
     \pdv{\tau_{yx}}{x} + \pdv{\tau_{yy}}{y} + \pdv{\tau_{yz}}{z}
         &= \pdv{p}{y} \,, \\
+    \label{eqn:ssaz}
     \pdv{\tau_{zz}}{z} &= \pdv{p}{z} - \rho g \,,
@@
 and horizontal velocities can be assumed independent of depth, so that
-\begin{subequations}
 \begin{align}
@@
 \end{align}
+
+Once again neglecting atmospheric pressure, and remembering that $\tr(\DST)=0$,
+\Cref{eqn:ssaz} yields
+\begin{align}
+    p &= \rho g (s-z) - \tau_{zz} \,, \\
+      &= \rho g (s-z) + \tau_{xx} + \tau_{yy} \,,
+\end{align}
+which can be re-introduced into \cref{eqn:ssax,eqn:ssay}:
+\begin{subequations}
+\begin{align}
+    2\pdv{\tau_{xx}}{x} - \pdv{\tau_{yy}}{x} + \pdv{\tau_{xy}}{y}
+        + \pdv{\tau_{xz}}{z} &= \rho g \pdv{s}{x} \,, \\
+    \pdv{\tau_{yx}}{x} + 2\pdv{\tau_{yy}}{y} - \pdv{\tau_{xx}}{y}
+        + \pdv{\tau_{yz}}{z} &= \rho g \pdv{s}{y} \,.
+\end{align}
 \end{subequations}
 
-This yields
-\todo{develop intermediary steps}
+Using the inverse formulation of the constitutive law \eqref{eqn:invglenslaw},
+components of the deviatoric stress tensor can be replaced by corresponding
+velocity components, yielding
 \begin{subequations}
 \begin{align}
-    \pdv{x} \left[2\bar{\nu}H
+    \pdv{x} \left[2\bar{\nu}
                   \left(2\pdv{v_x}{x} + \pdv{v_y}{y}\right) \right]
-        + \pdv{y} \left[\bar{\nu}H
+        + \pdv{y} \left[\bar{\nu}
                         \left(\pdv{v_x}{y} + \pdv{v_y}{x}\right) \right]
-        + \tau_{bx} &= \rho_i gH \pdv{h}{x} \,, \\
-    \pdv{x} \left[\bar{\nu}H
+        + \pdv{\tau_{xz}}{z} &= \rho g \pdv{s}{x} \,, \\
+    \pdv{x} \left[\bar{\nu}
                   \left(\pdv{v_x}{y} + \pdv{v_y}{x}\right) \right]
-        + \pdv{y} \left[2\bar{\nu}H
+        + \pdv{y} \left[2\bar{\nu}
                         \left(\pdv{v_x}{x} + 2\pdv{v_y}{y}\right) \right]
-        + \tau_{bx} &= \rho_i gH \pdv{h}{x} \,,
+         + \pdv{\tau_{yz}}{z} &= \rho g \pdv{s}{y} \,,
 \end{align}
@@
 \end{equation}
+
+A final integration over the entire ice column yields
+\begin{subequations}
+\begin{align}
+    \label{eqn:vssa}
+    \pdv{x} \left[2\bar{\nu}h
+                  \left(2\pdv{v_x}{x} + \pdv{v_y}{y}\right) \right]
+        + \pdv{y} \left[\bar{\nu}h
+                        \left(\pdv{v_x}{y} + \pdv{v_y}{x}\right) \right]
+        + \tau_{bx} &= \rho gh \pdv{s}{x} \,, \\
+    \pdv{x} \left[\bar{\nu}H
+                  \left(\pdv{v_x}{y} + \pdv{v_y}{x}\right) \right]
+        + \pdv{y} \left[2\bar{\nu}h
+                        \left(\pdv{v_x}{x} + 2\pdv{v_y}{y}\right) \right]
+        + \tau_{by} &= \rho gh \pdv{s}{y} \,,
+\end{align}
+\end{subequations}
+where $\tau_{bx}$ and $\tau_{by}$ correspond to $x$ and $y$-components of the
+basal traction, and will be determined by a sliding law.
 
@@
                 & \text{if}\ b_0 < b < b_1 \,, \\
-        \phi_1 & \text{if}\ b_1 \le b. \,,
+        \phi_1 & \text{if}\ b_1 \le b \,,
     \end{cases}

### 2014-06-25 kappa.tex
Add paper-n commands as the order is unsure.

@@
 \newcommand{\degC}[0]{\unit{{\degree}C}}
+
+% Papers of the thesis
+\newcommand{\CCLI}[0]{Paper~I}      % Cordillera climate
+\newcommand{\PSDV}[0]{Paper~II}     % PDD SD variability
+\newcommand{\PSDP}[0]{Paper~III}    % PDD SD parametrisation
+\newcommand{\CCYC}[0]{Paper~IV}     % Cordillera cycle
 
@@
 till, $N$, is determined by the modelled amount of water at the bed. However,
-two different parametrisations are used within this thesis. In Paper~I, using
+two different parametrisations are used within this thesis. In {\CCLI}, using
 PISM~0.5, the effective pressure is linearly related to water content by a
@@
 
-In Paper~IV, using PISM~0.6, the effective pressure is determined by an
+In {\CCYC}, using PISM~0.6, the effective pressure is determined by an
 empirical parametrisation based on laboratory experiments on Antarctic till,
@@
 highest shoreline elevations in the model domain. The values of $\phi_1$ and
-$\phi_2$ are chosen differently in Paper~I (10 and 30$^\circ$) and Paper~IV
+$\phi_2$ are chosen differently in \CCLI (10 and 30$^\circ$) and \CCYC
 (15 and 45$^\circ$), but these choices are roughly equivalent in the case of a
@@
 Coast and Rocky Mountains of British Columbia \citep{Shea.etal.2009}. In
-Paper~I, the temperature standard deviation $\sPDD$ is a constant model
-parameter. In Paper~IV, it is computed from daily temperature values from the
+{\CCLI}, the temperature standard deviation $\sPDD$ is a constant model
+parameter. In {\CCYC}, it is computed from daily temperature values from the
 North American Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006}, using
-a method initially developed in Paper~II, and further improved in Paper~III.
+a method initially developed in {\PSDV}, and further improved in {\PSDP}.
 %3.04\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for snow and
@@
 
-Paper~I tests the sensitivity of the model to different input present-day
+\CCLI tests the sensitivity of the model to different input present-day
 climatologies, $\{T_{m0}, P_{m0}\}$, while the temperature-offset corrections,
-${\Delta}T_{TS}$, uses constant values ranging from -10 to 0\,\degC. Paper~IV,
+${\Delta}T_{TS}$, uses constant values ranging from -10 to 0\,\degC. \CCYC,
 on the opposite, tests model sensitivity to time-dependent temperature-offset

### 2014-06-26 kappa.tex
Release full outline with planned figs and tables.
Correct the ice rheology formula.

@@
 \linenumbers
+\tableofcontents
 
@@
 % ======================================================================
-% Ice as viscous matter
-% Glaciers and ice sheets
-% The last glacial cycle
 
-% Fig. crevasses
-% Fig. glacier flowing
-% Fig. recent climate evolution
-% Fig. present and palaeo-ice sheets
+\todo{Figure. Glacier crevasse photo}\\
+\todo{Figure. Glacier flowing photo}\\
+\todo{Figure. Glacier moraine photo}\\
+\todo{Figure. Benthic oxygen isotopes stack}\\
+\todo{Figure. Modern and former ice sheets}
 
@@
 % ======================================================================
-% Geographic setting
-% Climatic setting
-% Palaeo-glaciology
 
-% Fig. map of North America
-% Fig. map of the Cordillera?
-% Fig. photos of the Cordillera
-% Fig. photos of fieldwork
-% Fig. Johan lineation map?
+% ----------------------------------------------------------------------
+\subsection{Geographic setting}
+% ----------------------------------------------------------------------
+
+\todo{Figure. Location map of North America}\\
+\todo{Figure. Topographic map of the Cordillera}\\
+\todo{Table. LCC projection parameters?}
+
+% ----------------------------------------------------------------------
+\subsection{Climatic setting}
+% ----------------------------------------------------------------------
+
+\todo{Figure. Cordillera wet climate photo}\\
+\todo{Figure. Cordillera dry climate photo}\\
+\todo{Figure. Climatic patters}
+
+% ----------------------------------------------------------------------
+\subsection{Palaeo-glaciology}
+% ----------------------------------------------------------------------
+
+\todo{Figure. Glacierized region photo}\\
+\todo{Figure. Tuya Lake moraine photo}\\
+\todo{Figure. Meltwater channels photo}\\
+\todo{Figure. Lineation map}
 
@@
 
-% Fig. PISM = ice + bedrock
+\todo{Figure. PISM interfaces}
 
@@
 
+\todo{Table. model variables and parameters}
+
 % ----------------------------------------------------------------------
@@
 % ----------------------------------------------------------------------
+
+\todo{Figure. Ice flow problem sketch}
 
@@
 where $\mathrm{II}_{\DST}$ is the second invariant of the stress tensor.
-The ice softness coefficient, $A$, depends on the deviation of temperature $T$
-from the pressure-melting point $T_m$ through
+The ice softness coefficient, $A$, depends on ice temperature, pressure, and
+water, $\omega$ through an Arrhenius-type law,
 \begin{equation}
-    A = A_0\,e^\frac{-Q}{R(T-T_m)} \,.
+    A =
+    \begin{cases}
+        A_c (1+f\omega)\,e^\frac{-Q_c}{RT_{pa}}
+            & \text{if}\ T < T_{crit} \,, \\
+        A_w (1+f\omega)\,e^\frac{-Q_w}{RT_{pa}}
+            & \text{if}\ T \ge T_{crit} \,,
+    \end{cases}
 \end{equation}
+where $T_{pa}$ is the pressure-adjusted temperature calculated using the
+Clapeyron relation, ${T_{pa} = T - \beta p}$.
 
@@
 
-% Fig. SIA + SSA
+\todo{Figure. SIA and SSA}
 
@@
 \begin{equation}
-    N = \rho gH (1 - \alpha \frac{W}{W_{max}}) \,.
+    N = \rho gh (1 - \alpha \frac{W}{W_{max}}) \,.
 \end{equation}
@@
 \begin{equation}
-    N = \delta \rho gH \, 10^{(e_0/C_c) (1 - (W/W_{max}))} \,.
+    N = \delta \rho gh \, 10^{(e_0/C_c) (1 - (W/W_{max}))} \,.
 \end{equation}
@@
 
+\todo{Figure. SMB model algorithm}
+
 % ----------------------------------------------------------------------
@@
 \end{subequations}
+
+\todo{Figure. Boundary models data flow chart}
 
@@
 
+\todo{Table. Data sources?}
+
 % ----------------------------------------------------------------------
@@
 % ======================================================================
-% Computational workflow
-% Contributions to PISM
-% PyPDD
-% pismplotlib
-% r.in.worldclim
-% r.out.pism
 
-% Fig. flow chart
-% Fig. pypdd example
-% Fig. pismplotlib example?
+% ----------------------------------------------------------------------
+\subsection{Computational workflow}
+% ----------------------------------------------------------------------
+\todo{Figure. computational workflow}
+
+% ----------------------------------------------------------------------
+\subsection{A Python PDD model}
+% ----------------------------------------------------------------------
+\todo{Figure. PyPDD demo example}
+
+% ----------------------------------------------------------------------
+\subsection{Pre-processing scripts for GRASS GIS}
+% ----------------------------------------------------------------------
+
+% ----------------------------------------------------------------------
+\subsection{Plotting tools for PISM}
+% ----------------------------------------------------------------------
+\todo{Figure. Plotting tools example}
+
+% ----------------------------------------------------------------------
+\subsection{Contributions to PISM}
+% ----------------------------------------------------------------------
+\todo{Table. PISM ``issues'' identified}
 
@@
 % ======================================================================
-% Constant-climate runs
-% Effect of daily temperature variability
-% Transient-climate runs
 
-% Fig. 3km LGM quiver
-% Fig. sigma distribution map
-% Fig. sigma effect on Cordillera runs
-% Fig. deglaciation streamplot
-% Fig. cumulative ice displacement map
+% ----------------------------------------------------------------------
+\subsection{Constant-climate runs}
+% ----------------------------------------------------------------------
+\todo{Figure. 3km last glacial maximum poster run?}
+
+% ----------------------------------------------------------------------
+\subsection{Daily temperature variability}
+% ----------------------------------------------------------------------
+\todo{Figure. Sigma distribution map}\\
+\todo{Figure. Sigma effect on Cordillera runs}
+
+% ----------------------------------------------------------------------
+\subsection{Transient-climate runs}
+% ----------------------------------------------------------------------
+\todo{Figure. Deglaciation streamplot}\\
+\todo{Figure. Cumulative ice displacement map}
 

### 2014-06-26 kappa.tex
Add parameter tables.

@@
 \newcommand{\todo}[1]{\emph{[\textbf{Todo:} #1]}}
-%\newcommand{\aref}[0]{\textbf{[ref.]}}
+\newcommand{\mref}[0]{\textbf{[ref.]}}
 
@@
 \newcommand{\vssa}[0]{\vv_{\mathrm{SSA}}}   % SSA velocity
+\newcommand{\PDD}[0]{\mathrm{PDD}}
+\newcommand{\sPDD}[0]{\sigma_{\mathrm{PDD}}}
 
 % Common units
+\newcommand{\e}[1]{\ensuremath{\times 10^{#1}}}
 \newcommand{\unit}[1]{\ensuremath{\mathrm{#1}}}
@@
 
-\todo{Table. model variables and parameters}
-
 % ----------------------------------------------------------------------
@@
 The ice softness coefficient, $A$, depends on ice temperature, pressure, and
-water, $\omega$ through an Arrhenius-type law,
+water content through an Arrhenius-type law,
 \begin{equation}
@@
         A_c (1+f\omega)\,e^\frac{-Q_c}{RT_{pa}}
-            & \text{if}\ T < T_{crit} \,, \\
+            & \text{if}\ T < T_c \,, \\
         A_w (1+f\omega)\,e^\frac{-Q_w}{RT_{pa}}
-            & \text{if}\ T \ge T_{crit} \,,
+            & \text{if}\ T \ge T_c \,,
     \end{cases}
@@
 where $T_{pa}$ is the pressure-adjusted temperature calculated using the
-Clapeyron relation, ${T_{pa} = T - \beta p}$.
+Clapeyron relation, ${T_{pa} = T - \beta p}$. The water fraction, $\omega$, is
+capped at a threshold of 0.01, above which no measurements are available.
 
@@
 cycle, thus requiring shallow approximations.
+
+\begin{table}
+  \centering
+  \caption{Constant parameters of the ice flow model.
+           \todo{find a better reference than \citet{Aschwanden.etal.2012}
+                 for all the physical constants. Double-check
+                 \citet{Paterson.Budd.1982} and add missing references.}}
+  \label{tab:iceflowparams}
+  \makebox[0pt]{\begin{tabular*}{165mm}{@{\extracolsep{\fill}}rlrll}
+    \hline
+    Not.    & Name & Value & Unit & Source \\
+    \hline
+
+    $\rho$  & Ice density
+            & 910
+            & \unit{kg\,m^{-3}}
+            & \citet{Aschwanden.etal.2012} \\
+
+    $g$     & Standard gravity
+            & 9.81
+            & \unit{m\,s^{-2}}
+            & \citet{Aschwanden.etal.2012} \\
+
+    $n$     & Glen exponent
+            & 3
+            & -
+            & \mref \\
+
+    $A_c$   & Ice hardness coefficient cold
+            & 3.61\e{-13}
+            & \unit{Pa^{-3}\,s^{-1}}
+            & \citet{Paterson.Budd.1982} \\
+
+    $A_w$   & Ice hardness coefficient warm
+            & 1.73\e3
+            & \unit{Pa^{-3}\,s^{-1}}
+            & \citet{Paterson.Budd.1982} \\
+
+    $Q_c$   & Flow law activation energy cold
+            & 6.0\e4
+            & \unit{J\,mol^{-1}}
+            & \citet{Paterson.Budd.1982} \\
+
+    $Q_w$   & Flow law activation energy warm
+            & 13.9\e4
+            & \unit{J\,mol^{-1}}
+            & \citet{Paterson.Budd.1982} \\
+
+    $R$     & Ideal gas constant
+            & 8.31441
+            & \unit{J\,mol^{-1}\,K^{-1}}
+            & \mref \\
+
+    $T_c$   & Flow law critical temperature
+            & 263.15
+            & \unit{K}
+            & \citet{Paterson.Budd.1982} \\
+
+    $f$     & Flow law water fraction coeff.
+            & 181.25
+            & -
+            & \citet{Lliboutry.Duval.1985} \\
+
+    $\beta$ & Clapeyron constant
+            & 7.9\e{-8}
+            & \unit{K\,Pa^{-1}}
+            & \citet{Luthi.etal.2002} \\
+
+    $c_i$   & Ice specific heat capacity
+            & 2009
+            & \unit{J\,kg^{-1}\,K^{-1}}
+            & \citet{Aschwanden.etal.2012} \\
+
+    $c_w$   & Water specific heat capacity
+            & 4170
+            & \unit{J\,kg^{-1}\,K^{-1}}
+            & \citet{Aschwanden.etal.2012} \\
+
+    $k$     & Ice thermal conductivity
+            & 2.10
+            & \unit{J\,m^{-1}\,K^{-1}\,s^{-1}}
+            & \citet{Aschwanden.etal.2012} \\
+
+    $L$     & Water latent heat of fusion
+            & 3.34\e5
+            & \unit{J\,kg^{-1}\,K^{-1}}
+            & \citet{Aschwanden.etal.2012} \\
+
+    \hline
+  \end{tabular*}}
+\end{table}
 
@@
 
+\begin{table}
+  \centering
+  \caption{Constant parameters of the sliding model. Missing sources
+           correspond to uncalibrated values.}
+  \label{tab:slidingparams}
+  \makebox[0pt]{\begin{tabular*}{165mm}{@{\extracolsep{\fill}}rlrll}
+    \hline
+    Not.    & Name & Value & Unit & Source \\
+    \hline
+
+    $q$     & Pseudo-plastic sliding exponent
+            & 0.25
+            & -
+            & - \\
+
+    $v_{th}$& Pseudo-plastic threshold velocity
+            & 100.0
+            & \unit{m\,yr^{-1}}
+            & - \\
+
+    $c_0$   & Till cohesion
+            & 0.0
+            & Pa
+            & - \\
+
+    $\alpha$& \todo{check value}
+            & 0.95
+            & -
+            & - \\
+
+    $\delta$& \todo{find a good name}
+            & 0.02
+            & -
+            & - \\
+
+    $e_0$   & Till reference void ratio
+            & 0.69
+            & -
+            & \citet{Tulaczyk.etal.2000} \\
+
+    $C_c$   & Till compressibility coefficient
+            & 0.12
+            & -
+            & \citet{Tulaczyk.etal.2000} \\
+
+    $W_{max}$ & Maximal till water thickness
+            & 2.0
+            & m
+            & - \\
+
+    $b_0$   & Altitude of max. friction angle
+            & 0
+            & m
+            & - \\
+
+    $b_1$   & Altitude of min. friction angle
+            & 200
+            & m
+            & - \\
+
+    $\phi_0$& Minimum friction angle
+            & 10/15
+            & \degree
+            & - \\
+
+    $\phi_1$& Maximum friction angle
+            & 30/45
+            & \degree
+            & - \\
+
+    \hline
+  \end{tabular*}}
+\end{table}
+
 % ----------------------------------------------------------------------
@@
 
+\begin{table}
+  \centering
+  \caption{Constant parameters of the bedrock model.
+           \todo{Find out what the bedrock density corresponds to. Track down
+                 original sources in \citet{Ritz.1997}}}
+  \label{tab:bedrockparams}
+  \makebox[0pt]{\begin{tabular*}{165mm}{@{\extracolsep{\fill}}rlrll}
+    \hline
+    Not.    & Name & Value & Unit & Source \\
+    \hline
+
+    $\rho_b$& Bedrock density
+            & 3300
+            & \unit{kg\,m^{-3}}
+            & \mref \\
+
+    $c_b$   & Bedrock specific heat capacity
+            & 1000
+            & \unit{J\,kg^{-1}\,K^{-1}}
+            & \citet{Ritz.1997} \\
+
+    $k_b$   & Bedrock thermal conductivity
+            & 3.0
+            & \unit{J\,m^{-1}\,K^{-1}\,s^{-1}}
+            & \citet{Ritz.1997} \\
+
+    $\nu_m$ & Mantle viscosity
+            & 1\e21
+            & \unit{Pa\,s}
+            & \citet{Lingle.Clark.1985} \\
+
+    $\rho_l$& Lithosphere density
+            & 3300
+            & \unit{kg\,m^{-3}}
+            & \citet{Lingle.Clark.1985} \\
+
+    $D$     & Lithosphere flexural rigidity
+            & 5.0\e24
+            & \unit{N}
+            & \citet{Lingle.Clark.1985} \\
+
+    \hline
+%	pism overrides:bootstrapping geothermal flux value no var = 0.070 ;
+  \end{tabular*}}
+\end{table}
+
 % ----------------------------------------------------------------------
@@
 % ----------------------------------------------------------------------
-
-\newcommand{\PDD}[0]{\mathrm{PDD}}
-\newcommand{\sPDD}[0]{\sigma_{\mathrm{PDD}}}
 
@@
     \begin{cases}
-        0       & \text{if}\ T_m \le T_0 \,, \\
-        P_m \frac{T_m-T_0}{T_1-T_0}
-                & \text{if}\ T_0 < b < T_1 \,, \\
-        P_m     & \text{if}\ T_1 \le T_m \,,
+        0       & \text{if}\ T_m \le T_s \,, \\
+        P_m \frac{T_m-T_s}{T_r-T_s}
+                & \text{if}\ T_s < T_m < T_r \,, \\
+        P_m     & \text{if}\ T_r \le T_m \,,
     \end{cases}
@@
 \todo{Figure. SMB model algorithm}
+
+\begin{table}
+  \centering
+  \caption{Constant parameters of the surface and atmosphere models.
+           \todo{Find a reference for the snow and rain temperature thresholds,
+                 maybe \citet{Hock.2003}.}}
+  \label{tab:surfaceparams}
+  \makebox[0pt]{\begin{tabular*}{165mm}{@{\extracolsep{\fill}}rlrll}
+    \hline
+    Not.    & Name & Value & Unit & Source \\
+    \hline
+
+    $T_s$   & Temperature of snow precipitation
+            & 273.15
+            & \unit{K}
+            & \mref \\
+
+    $T_r$   & Temperature of rain precipitation
+            & 275.15
+            & \unit{K}
+            & \mref \\
+
+    $F_s$   & Degree-day factor for ice
+            & 3.04\e{-3}
+            & \unit{m\,K^{-1}\,day^{-1}}
+            & \citet{Shea.etal.2009} \\
+
+    $F_i$   & Degree-day factor for ice
+            & 4.59\e{-3}
+            & \unit{m\,K^{-1}\,day^{-1}}
+            & \citet{Shea.etal.2009} \\
+
+    $\gamma$& Air temperature lapse-rate
+            & 6\e{-3}
+            & \unit{K\,m{-1}}
+            & - \\
+
+    \hline
+  \end{tabular*}}
+\end{table}
 

### 2014-06-26 kappa.tex
Add forcing data sources table.

@@
 \newcommand{\e}[1]{\ensuremath{\times 10^{#1}}}
+\newcommand{\chem}[1]{\ensuremath{\mathrm{#1}}}
 \newcommand{\unit}[1]{\ensuremath{\mathrm{#1}}}
@@
 
-\todo{Table. Data sources?}
+\begin{table}
+  \centering
+  \caption{Sources of forcing data used to inform the ice sheet model.}
+  \label{tab:datasources}
+  \makebox[0pt]{\begin{tabular*}{160mm}{@{\extracolsep{\fill}}llll}
+    \hline
+    Name        & Type & Original reference & Employed in \\
+    \hline
+
+    ETOPO1      & Bedrock topography
+                & \citet{Amante.Eakins.2009}
+                & \CCLI, \CCYC \\
+
+    WorldClim   & Gridded observations
+                & \citet{Hijmans.etal.2005}
+                & \CCLI \\
+
+    NCEP/NCAR   & Atmospheric reanalysis
+                & \citet{Kalnay.etal.1996}
+                & \CCLI \\
+
+    ERA-40      & \qquad '' \qquad
+                & \citet{Uppala.etal.2005}
+                & \PSDP \\
+
+    ERA-Interim & \qquad '' \qquad
+                & \citet{Dee.etal.2011}
+                & \CCLI, \PSDV \\
+
+    CFSR        & \qquad '' \qquad
+                & \citet{Saha.etal.2010}
+                & \CCLI \\
+
+    NARR        & \qquad '' \qquad
+                & \citet{Mesinger.etal.2006}
+                & \CCLI, \CCYC \\
+
+    GRIP        & Ice-core \chem{\delta^{18}O} record
+                & \citet{Dansgaard.etal.1993}
+                & \CCYC \\
+
+    NGRIP       & \qquad '' \qquad
+                & \citet{Andersen.etal.2004}
+                & \CCYC \\
+
+    EPICA       & \qquad '' \qquad
+                & \citet{Jouzel.etal.2007}
+                & \CCYC \\
+
+    Vostok      & \qquad '' \qquad
+                & \citet{Petit.etal.1999}
+                & \CCYC \\
+
+    ODP~1012    & Sea-floor \chem{U^{K'}_{37}} record
+                & \citet{Herbert.etal.2001}
+                & \CCYC \\
+
+    ODP~1020    & \qquad '' \qquad
+                & \citet{Herbert.etal.2001}
+                & \CCYC \\
+
+    \hline
+  \end{tabular*}}
+\end{table}
 

### 2014-06-26 kappa.tex
Add projection parameters table.

@@
 \todo{Figure. Topographic map of the Cordillera}\\
-\todo{Table. LCC projection parameters?}
+
+\begin{table}
+  \centering
+  \caption{Properties of the Canadian Atlas Lambert conformal conic projection
+           (EPSG code~3978) and the GRS~1980 ellipsoid, and corresponding
+           PROJ.4 parameter names, as used in all model runs, ice volume and
+           area calculations, and most figures.}
+  \label{tab:projparams}
+  \makebox[0pt]{\begin{tabular*}{80mm}{@{\extracolsep{\fill}}lcc}
+    \hline
+    Parameter               & Name              & Value \\
+    \hline
+    Projection name         & \texttt{+proj}    & \texttt{lcc} \\
+    Central meridian        & \texttt{+lon\_0}  & 95\degree W \\
+    Latitude of origin      & \texttt{+lat\_0}  & 49\degree N \\
+    First secant parallel   & \texttt{+lat\_1}  & 49\degree N \\
+    Second secant parallel  & \texttt{+lat\_2}  & 77\degree N \\
+    Semimajor radius        & \texttt{+a}       & 6378137\,m \\
+    Reciprocal flattening   & \texttt{+rf}      & 298.257222101 \\
+    \hline
+  \end{tabular*}}
+\end{table}
 

### 2014-06-27 figures/model-pdd.tex
Add PDD model algorythm from PISM docs.

\documentclass[border=5mm,tikz]{standalone}
\usepackage{amsmath}
\usetikzlibrary{chains}
\usetikzlibrary{arrows}
\usetikzlibrary{shapes}

\newcommand{\PDD}{\ensuremath{\text{PDD}}}
\newcommand{\Hs}{\ensuremath{H_{\text{snow}}}}
\newcommand{\Ms}{\ensuremath{\text{Melt}_{\text{snow}}}}
\newcommand{\Mi}{\ensuremath{\text{Melt}_{\text{ice}}}}

\begin{document}
  \begin{tikzpicture}[ auto, thick, font=\small, node distance=3mm,
    block/.style={rectangle, rounded corners, draw, fill=gray!5, text width=2in, text centered,
      minimum width=2in, minimum height=1cm},
    big block/.style={block, text width=3in},
    decision/.style={diamond, draw, fill=gray!5},
    line/.style={draw, -latex'},
    start chain=1 going below, every node/.style={on chain},
    every join/.style={-latex'},
    ]

    \node [big block] {Compute snow accumulation \mbox{``Accumulation''}, discarding rain};
    \node [big block, join] {Compute the number of \mbox{positive} degree days (\PDD) };
    \node [big block, join] {Update snow depth \mbox{$\Hs \gets \Hs + \text{Accumulation}$}};
    \node [big block, join] {Compute maximum possible snow melt \mbox{$\Ms^{\text{max}} := \PDD\times F_{s}$}};
    \node (melt all snow) [decision, join] {$\Ms^{\text{max}} > \Hs?$};

    \begin{scope}[start branch=yes, node distance=4em]
      \node (ice melt case) [block, on chain=going right] {\mbox{Melt all available snow:} $\Ms := \Hs$};
    \end{scope}
    \begin{scope}[start branch=no, node distance=4em]
      \node (no ice melt case) [block, on chain=going left] {Melt some of the snow: \mbox{$\Ms := \Ms^{\text{max}}$}};
    \end{scope}

    \node (excess pdds) [big block] {$\PDD_{\text{excess}} := \PDD - \Ms / F_{s}$};
    \node [big block, join] {Melt some ice: \mbox{$\Mi := \PDD_{\text{excess}} \times F_{i}$}};
    \node (refreeze ice) [decision, join] {Refreeze melted ice?};

    \begin{scope}[start branch=yes, node distance=4em]
      \node (refreeze both) [block, on chain=going right] {Refreeze is \mbox{$(\Ms + \Mi) \times \theta_{\text{refreeze}}$}};
    \end{scope}

    \begin{scope}[start branch=no, node distance=4em]
      \node (refreeze snow) [block, on chain=going left] {Refreeze is $\Ms \times \theta_{\text{refreeze}}$};
    \end{scope}

    \node (runoff) [big block] {$\text{Runoff} := \Ms + \Mi - \text{Refreeze}$};
    \node [big block, join] {Compute surface mass balance: $\text{SMB} := \text{Accumulation} - \text{Runoff}$};
    \node [big block, join] {Update snow depth: $\Hs \gets \Hs - \Ms$};
    % special edges:
    \path[line] (melt all snow) -- node {Yes} (ice melt case);
    \path[line] (melt all snow) -- node {No}  (no ice melt case);
    \path[line] (ice melt case) |- (excess pdds);
    \path[line] (no ice melt case) |- (excess pdds);
    \path[line] (refreeze ice) -- node {Yes} (refreeze both);
    \path[line] (refreeze ice) -- node {No}  (refreeze snow);
    \path[line] (refreeze both) |- (runoff);
    \path[line] (refreeze snow) |- (runoff);
  \end{tikzpicture}

\end{document}

### 2014-06-29 kappa.tex
Add figures in the introduction.

@@
 \usepackage[sort]{cleveref}  % needs to be loaded after hyperref
-%\usepackage{graphicx}
+\usepackage{graphicx}
+\graphicspath{{figures/}}
 
@@
 
+% Text label over graphics
+% from http://tex.stackexchange.com/questions/128844/put-subfigure-labels-
+% inside-figures-using-subfig-package
+\newcommand{\subgraphics}[3][,]{%
+  \setbox1=\hbox{\includegraphics[#1]{#3}}% Store image in box
+  \leavevmode\rlap{\usebox1}% Print image
+  \rlap{\hspace*{0.25em}
+        \raisebox{\dimexpr\ht1-3ex}{\textbf{(#2)}}}% Print label
+  \phantom{\usebox1}% Insert appropriate spacing
+  \hspace{1cm}% Add horizontal separator
+}
+
+% Finally to not include graphics
+%\renewcommand{\includegraphics}[2][,]{}
+%\renewcommand{\subgraphics}[3][,]{}
+%\usepackage{endfloat}
+%\renewcommand{\efloatseparator}{\mbox{}}
+
 % Document properties
@@
 
-\todo{Figure. Glacier crevasse photo}\\
-\todo{Figure. Glacier flowing photo}\\
-\todo{Figure. Glacier moraine photo}\\
-\todo{Figure. Benthic oxygen isotopes stack}\\
-\todo{Figure. Modern and former ice sheets}
+\begin{figure}
+  \centering
+  \makebox[0pt]{
+    \subgraphics{a}{photo-glacier-crevasses}
+    \subgraphics{b}{photo-glacier-fold}
+  }
+  \caption{Different mechanical behaviours of glacier ice.
+           \textbf{(a)} Brittle behaviour. Ice flowing over a hidden bedrock
+           knob, resulting in fracturing at the glacier surface.
+           Glacier Blanc, French Alps.
+           \textbf{(b)} Ductile behaviour. Mountain glacier calving into a
+           small lake, revealing a folded structure formed by deformed rock
+           layers in the ice.}
+  \label{fig:glacier-mechanics}
+\end{figure}
+
+% Alternative code using subcaption package
+%\usepackage{subcaption}
+%\begin{figure}
+%  \centering
+%  \makebox[0pt]{
+%    \begin{subfigure}[t]{80mm}
+%        \includegraphics{photo-glacier-crevasses}
+%        \caption{Subcaption a.}
+%        \label{fig:1a}
+%    \end{subfigure}
+%    \hspace{10mm}
+%    \begin{subfigure}[t]{80mm}
+%        \includegraphics{photo-glacier-fold}
+%        \caption{Subcaption b.}
+%        \label{fig:1b}
+%    \end{subfigure}
+%  }
+%  \caption{Global caption.}
+%  \label{fig:1}
+%\end{figure}
+
+\begin{figure}
+  \centering
+  \makebox[0pt]{
+    \subgraphics[width=50mm]{a}{photo-glacial-valley}
+    \subgraphics[width=50mm]{b}{photo-erratic-boulder}
+    \subgraphics[width=50mm]{c}{photo-melt-channels}
+  }
+  \caption{Different indicators of past glaciation.
+           \textbf{(a)} Glacial valley, exhibiting a broad, U-shaped profile
+           characteristic of erosion by glaciers.
+           Cirque du Fer-\`{a}-Cheval, French Alps.
+           \textbf{(b)} Erratic boulders, deposited in an improbable
+           arrangement as the glacier melted underneath.
+           Finse area, Norwegian Scandes.
+           \textbf{(c)} Lateral meltwater channels, formed on a gentle slope
+           as the ice margin retreated towards the valley bottom. Tuya Lake
+           area, Canadian Cordillera.}
+  \label{fig:glaciation-indicators}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=80mm]{paleo-timeseries}
+  \caption{The oxygen isotope record from oceanic sediments is a proxy for ice
+           volume on land during Earth's hitory. Over the last million years,
+           it shows predominance of a 100\,ka cycle of glacial and
+           interglacial periods. This thesis focuses on the last glacial cycle.
+           \todo{This figure is directly from \citet{Lisiecki.Raymo.2005}.
+           Redraw it using the data they provide.}}
+  \label{fig:palaeo-timeseries}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=80mm]{paleo-glaciation}
+  \caption{Spatial extent of the last glacial maximum northern hemisphere ice
+           sheets, after \citep{Ehlers.Gibbard.2007}.
+           \todo{Rotate by 90\degree, re-centre, and compare with present-day
+           ice cover. Maybe label the major ice sheets: CIS, LIS, GIS, FIS...}}
+  \label{fig:palaeo-glaciation}
+\end{figure}
 

### 2014-06-29 kappa.tex
Add paleo-timeseries plot from original data.

@@
   \phantom{\usebox1}% Insert appropriate spacing
-  \hspace{1cm}% Add horizontal separator
 }
@@
     \subgraphics{a}{photo-glacier-crevasses}
+    \hspace{1cm}
     \subgraphics{b}{photo-glacier-fold}
@@
     \subgraphics[width=50mm]{a}{photo-glacial-valley}
+    \hspace{1cm}
     \subgraphics[width=50mm]{b}{photo-erratic-boulder}
+    \hspace{1cm}
     \subgraphics[width=50mm]{c}{photo-melt-channels}
@@
   \includegraphics[width=80mm]{paleo-timeseries}
-  \caption{The oxygen isotope record from oceanic sediments is a proxy for ice
-           volume on land during Earth's hitory. Over the last million years,
-           it shows predominance of a 100\,ka cycle of glacial and
-           interglacial periods. This thesis focuses on the last glacial cycle.
-           \todo{This figure is directly from \citet{Lisiecki.Raymo.2005}.
-           Redraw it using the data they provide.}}
+  \caption{Indicators of climatic fluctuations during the last 800\,ka of
+           Earth history.
+           \textbf{(a)} Benthic oxygen isotope \chem{\delta^{18}O} record from
+           a compilation of globally-distributed oceanic sediment cores
+           \citep{Lisiecki.Raymo.2005}. Higher \chem{\delta^{18}O} values
+           correspond to periods of lower temperatures of the deep ocean, or
+           of lower global sea level, and thus higher ice volume stored in
+           glaciers and ice sheets on land.
+           \textbf{(b)} Temperature anomalies reconstructed from a central
+           Antarctic ice core \citep[EPICA,][]{Jouzel.etal.2007}.
+           The grey boxes indicate the approximate time span of the last
+           glacial period of 110 to 10\,ka, which corresponds to our period of
+           interest in the present thesis.}
   \label{fig:palaeo-timeseries}

### 2014-06-29 figures/paleo-timeseries.py

@@ -1,0 +1,69 @@
+#!/usr/bin/env python
+# coding: utf-8
+
+import os
+import urllib
+import numpy as np
+from matplotlib import pyplot as plt
+
+# data sources
+urlbase = 'ftp://ftp.ncdc.noaa.gov/pub/data/paleo/'
+urldict = {
+    'epica': urlbase + 'icecore/antarctica/epica_domec/edc3deuttemp2007.txt',
+    'lr04':  urlbase + 'contributions_by_author/lisiecki2005/lisiecki2005.txt'}
+
+# additional sea-level data sources
+# 'contributions_by_author/lea2002/lea2002.txt'
+# 'contributions_by_author/siddall2003/siddall2003.txt'
+
+# fetch data
+for rec, url in urldict.iteritems():
+    filename = 'data/%s.txt' % rec
+    print filename
+    if not os.path.isfile(filename):
+        urllib.urlretrieve(url, filename)
+
+
+# Initialize figure
+figw, figh = 80.0, 60.0
+fig, grid = plt.subplots(2, figsize=(figw/25.4, figh/25.4), sharex=True)
+fig.subplots_adjust(left=10.0/figw, bottom=7.5/figh,
+                    right=1-2.5/figw, top=1-2.5/figh,
+                    hspace=1/((1+figh/2.5)/2-1))
+
+
+# plot LR04 data (pick only the last 800 ka)
+ax = grid[0]
+age, d18o = np.genfromtxt('data/lr04.txt', skip_header=89, skip_footer=9387,
+                          unpack=True, usecols=(0, 1))
+ax.plot(age, d18o, 'b')
+
+# set LR04 axes properties
+ax.set_ylim(5.5, 2.5)
+ax.set_yticks([3, 4, 5])
+ax.set_ylabel(u'$\delta^{18}O$ (\u2030)')
+ax.text(0.04, 0.85, '(a)', fontweight='bold', transform=ax.transAxes)
+
+# plot EPICA data
+ax = grid[1]
+age, temp = np.genfromtxt('data/epica.txt', delimiter=(4, 13, 17, 13, 13),
+                          skip_header=104, skip_footer=1,
+                          unpack=True, usecols=(2, 4))
+ax.plot(age/1000.0, temp, 'r')
+
+# set EPICA axes properties
+ax.set_yticks(range(-12, 6, 4))
+ax.set_ylim(-12, 6)
[…]

### 2014-06-30 kappa.tex
Improve paleo-glaciation plot.

@@
   \includegraphics[width=80mm]{paleo-glaciation}
-  \caption{Spatial extent of the last glacial maximum northern hemisphere ice
-           sheets, after \citep{Ehlers.Gibbard.2007}.
-           \todo{Rotate by 90\degree, re-centre, and compare with present-day
-           ice cover. Maybe label the major ice sheets: CIS, LIS, GIS, FIS...}}
+  \caption{Spatial extent of ice sheet and glacier cover in the northern
+           hemisphere \textbf{(a)} at present, and \textbf{(b)} during the last
+           glacial maximum \citep{Ehlers.Gibbard.2007}. Coastlines and present
+           glaciers are made with Natural Earth \citep{Patterson.Kelso.2014}.}
   \label{fig:palaeo-glaciation}

### 2014-06-30 figures/data/get_quatglac.sh

@@ -3,7 +3,7 @@
 url="http://static.us.elsevierhealth.com/ehlers_digital_maps/\
 digital_maps_02_all_other_files.zip"
 archive=$(basename $url)
-input=lgm_global.shp
+input=lgm.shp
 output=etopo1.nc
 
 # download archive
@@ -11,11 +11,8 @@ output=etopo1.nc
 
 # inflate archive
 # (we need to escape the wildcards to not match already inflated files)
-[ -f $input ] || unzip -u $archive ${input%.shp}.\* LGM_global.dbf
-
-# correct naming bug
-[ -f ${input%.shp}.dbf ] || ln -s LGM_global.dbf ${input%.shp}.dbf
+[ -f $input ] || unzip -u $archive ${input%.shp}.\*
 
 # remove small polygons and simplify others
 ogr2ogr ${input%.shp}_simple.shp $input -overwrite \
-    -where "OGR_GEOM_AREA > 1e-3" -simplify 0.1    
+    -where "OGR_GEOM_AREA > 1e-3" -simplify 0.1

### 2014-06-30 figures/paleo-glaciation.py

@@ -12,23 +12,30 @@ bwu = 0.5  # base width unit
 scale = '50m'
 
 # Projection
-proj = ccrs.NorthPolarStereo()
+proj = ccrs.Orthographic(central_longitude=-90.0, central_latitude=75.0)
 
-# Initialize figure
-fig = plt.figure(0, (80/25.4, 80/25.4))
-ax = fig.add_axes([0, 0, 1, 1], projection=proj)
-ax.set_xlim(-6e6, 6e6)
-ax.set_ylim(-6e6, 6e6)
+# Initialize figure in inches
+figw, figh = 80.0/25.4, 40.0/25.4
+fig = plt.figure(0, (figw, figh))
+fig.subplots_adjust(left=0.5/72/figw, bottom=0.5/72/figh,
+                    right=1-0.5/72/figw, top=1-0.5/72/figh,
+                    wspace=1/((1+figw/(1./72))/2-1))
+grid = [fig.add_subplot(121+i, projection=proj) for i in range(2)]
 
-# add NaturalEarth coastlines
-ax.add_feature(cfeature.NaturalEarthFeature(
-    category='physical', name='coastline', scale=scale,
-    edgecolor='k', facecolor='none', lw=0.5*bwu))
+# draw common parts
+for ax in grid:
+    ax.gridlines(color='0.5', linestyle='-', linewidth=0.1)
+    ax.coastlines(edgecolor='k', lw=0.25)
+
+# add present glaciers
+grid[0].add_feature(cfeature.NaturalEarthFeature(
+    category='physical', name='glaciated_areas', scale=scale,
+    edgecolor='none', facecolor='#0978ab', alpha=0.5,))
 
 # add Ehlers & Gibbard 2003 LGM outline
 # try to identify duplicates using centroids
 centroids = []
-shp = shpreader.Reader('data/lgm_global_simple.shp')
+shp = shpreader.Reader('data/lgm_simple.shp')
 for i, geom in enumerate(shp.geometries()):
     x, y = geom.centroid.xy
     x, y = x[0], y[0]
@@ -41,13 +48,8 @@ for i, geom in enumerate(shp.geometries()):
     else:
         print 'adding record %i with area %f ...' % (i, geom.area)
         centroids.append((x, y))
-        ax.add_geometries(geom, ccrs.PlateCarree(), alpha=0.5,
-                  edgecolor='none', facecolor='#0978ab', lw=bwu)
-
-# add graticules
[…]

### 2014-06-30 kappa.tex
Commit work in progress on introduction.

@@
 % ======================================================================
+
+In our daily lives, most of us experience ice as a solid that crackles at the
+contact with warmer liquids. However, at geologic time scales, ice behaves in a
+widely different way. Under the action of precipitation, ice accumulates in the
+mountains and cold regions of the globe. Once a critical mass is attained, it
+flows down by gravity, thus behaving as a fluid (\cref{fig:glacier-mechanics}).
+
+Glaciers, these bodies of moving ice, shape the landscape, thus leaving traces
+of their presence to the eye of the geologist \cref{fig:glaciation-indicators}.
+However it was not until XXXX, that these traces were associated with
+climatic fluctuations at the scale of the planet. There have been extensive studies investigating the formation and the influence of glaciers on morphology and changes of the earth surface.
 

### 2014-06-30 kappa.tex
Add introduction text.

--- 2014-06-30 kappa.tex
@@
 % ======================================================================
+
+In our daily lives, we most commonly experience ice as the small, solid cubes
+that crackle on contact with refreshing liquids. Yet, at geological
+time-scales, and under high stresses, ice behaves in a much different fashion.
+In the cold and mountainous regions of the globe, ice accumulates under the
+action of cumulative snowfall, until it reaches a critical thickness, and
+deforms under its own weight to flow downhill. Glaciers, these bodies of moving
+ice, flow by a combination of water-induced sliding at the base, and
+visco-plastic deformation within the ice body (\cref{fig:glacier-mechanics},
+\mref).
+
+As ice flows and glide on its bed, it transports rock debris and erode the
+landscape, thence leaving geomorphological traces of its former presence
+(\cref{fig:glaciation-indicators}). In many parts of the world, montane people
+and early explorers have since long learned to read these traces, and to
+understand that glaciers had once been more extensive than today. With the
+advent of the glacial theory, birthed the idea that, under colder temperatures
+than today, expansive ice sheets had once covered much of Europe and North
+America \mref.
+
+However, the glacial theory did not gain much belief until the exploration and
+discovery of the two present-day ice sheets on Earth, the Greenland and
+Antarctic ice sheets, provided a modern analogue for the proposed European and
+North American ice sheets. The first traverse of Greenland was performed by
+Nansen in 1888, while Amundsen reached the south pole in 1911. Meanwhile, in
+the European Alps, it was then discovered from observations of the glacial
+landscape, that there had not been only one, but at least four, glacial periods
+\mref.
+
+Today, palaeo-climate records extracted from ice-core and oceanic sediment
+records provide a much more detailed picture of the Earth environmental
+history. Over the last few million years, temperature and global ice volume
+have been constantly fluctuating, thus indicating that neither one, nor four,
+but probably tens of glacial cycles have occured \mref. During the last million
+years, glacial and interglacial cycles have succeeded each other with a 100-kyr
+periodicity \mref. In the present thesis, we focus on the most recent of these
+cycles: the last glacial cycle (\cref{fig:paleo-timeseries}).
+
+Together, the Greenland and Antarctic ice sheets currently host more than two
+thirds of the terrestrial fresh water, corresponding to 65~meters of potential
+sea-level equivalent ice volume \mref. At the apogee of the last glacial cycle,
+the last glacial maximum, sea-level was 120 to 135\,m lower than today \mref,
+implying an additional, double amount of ice stored in ice sheets and glaciers
+on land. Systematic investigation of glacial landforms on all continents
+permitted reconstructions of this former ice cover, indicating that ice volume
+was repartitioned between more extensive Greenland and Antarctic ice sheets,
+the Scandinavian and Barents-Kara ice sheets in the Eurasia arctic, and the
+Innuitian, Laurentide and Cordilleran ice sheets over North America
+(\cref{fig:paleo-glaciation}).
+
+Nevertheless, the landscape record is timely sparse and spatially incomplete,
+and geological reconstructions of former glacier extent demand a high amount of
+interpretation. Palaeo-ice sheets did not leave a systematic record of their
+passage, and much of the evidence has been overridden by multiple glacier
+re-advance and other geomorphological processes. This is where numerical
+models, incorporating the processes of glacier sliding and ice deformation
+observed on modern glaciers, can help. Although these models alone contain too
+many uncertainties to reproduce accurate picture of past glacial cycles, they
+can be calibrated against field-based evidence where it exists, while allowing
+a physics-based, quantitative reconstruction where it does not.
 
@@
            interest in the present thesis.}
-  \label{fig:palaeo-timeseries}
+  \label{fig:paleo-timeseries}
 \end{figure}
@@
            glaciers are made with Natural Earth \citep{Patterson.Kelso.2014}.}
-  \label{fig:palaeo-glaciation}
+  \label{fig:paleo-glaciation}
 \end{figure}

### 2014-07-01 kappa.tex
Add references in the introduction.

--- 2014-06-30 kappa.tex
@@
 % ======================================================================
+
+In our daily lives, we most commonly experience ice as the small, solid cubes
+that crackle on contact with refreshing liquids. Yet, at geological
+time-scales, and under high stresses, ice behaves in a much different fashion.
+In the cold and mountainous regions of the globe, ice accumulates under the
+action of cumulative snowfall, until it reaches a critical thickness, and
+deforms under its own weight to flow downhill (\cref{fig:glacier-mechanics}).
+Glaciers, these bodies of moving ice, flow by a combination of water-induced
+sliding at the base \citep{Saussure.1796}, and visco-plastic deformation within
+the ice body \citep{Forbes.1846b}.
+
+As ice flows and glide on its bed, it transports rock debris and erode the
+landscape, thence leaving geomorphological traces of its former presence
+(\cref{fig:glaciation-indicators}). In many parts of the world, montane people
+and early explorers have since long learned to read these traces, and to
+understand that glaciers had once been more extensive than today. With the
+advent of the glacial theory, birthed the idea that, under colder temperatures
+than today, expansive ice sheets had once covered much of Europe and North
+America \citep{Agassiz.1840}.
+
+However, the glacial theory did not gain much belief until the exploration and
+discovery of the two present-day ice sheets on Earth, the Greenland and
+Antarctic ice sheets, provided a modern analogue for the proposed European and
+North American ice sheets. The first traverse of the Greenland was performed by
+Nansen in 1888, while Amundsen reached the south pole in 1911. Meanwhile, in
+the European Alps, it was then discovered from observations of the glacial
+landscape, that there had not been only one, but at least four, glacial periods
+\citep{Penck.Bruckner.1909}.
+
+More recently, palaeo-climate records extracted from deep-sea sediments have
+provided a much more detailed picture of the Earth environmental history
+\citep[e.g.,][]{Emiliani.1955, Shackleton.Opdyke.1973}, thus indicating neither
+one, nor four, but rather tens of glacial cycles. During the last million
+years, glacial and interglacial cycles have succeeded each other with a
+periodicity 100\,kyr \citep{Hays.etal.1976}. In the present thesis, we focus on
+the most recent of these cycles: the last glacial cycle
+(\cref{fig:paleo-timeseries})
+
+% ice volume 26.07 + 2.93 = 29.00
+% meters sea-level 58.3 + 0.70 = 59.0
+Together, the Greenland and Antarctic ice sheets currently host most glacial
+ice on the planet, corresponding to 59~meters of potential sea-level rise
+\citep{Bamber.etal.2001, Fretwell.etal.2013}. At the apogee of the last glacial
+cycle, the last glacial maximum, sea-level was 120 to 135\,m lower than today
+\citep{Clark.Mix.2002}, implying an additional, double amount of ice stored in
+ice sheets and glaciers on land. Systematic mapping and compilation of glacial
+landforms from all continents permitted reconstructions of this former ice
+cover \citep{Ehlers.Gibbard.2007}, indicating that ice volume
+was repartitioned between more extensive Greenland and Antarctic ice sheets,
+the Scandinavian and Barents-Kara ice sheets in the Eurasia arctic, and the
+Innuitian, Laurentide and Cordilleran ice sheets over North America
+(\cref{fig:paleo-glaciation}).
+
+Nevertheless, the landscape record is timely sparse and spatially incomplete,
+and geological reconstructions of former glacier extent demand a high amount of
+interpretation. Palaeo-ice sheets did not leave a systematic record of their
+passage, and much of the evidence has been overridden by multiple glacier
+re-advance and other geomorphological processes. This is where numerical
+models, incorporating the processes of glacier sliding and ice deformation
+observed on modern glaciers, can help. Although these models alone contain too
+many uncertainties to reproduce accurate picture of past glacial cycles, they
+can be calibrated against field-based evidence where it exists, while allowing
+a physics-based, quantitative reconstruction where it does not.
 
@@
            \textbf{(a)} Benthic oxygen isotope \chem{\delta^{18}O} record from
-           a compilation of globally-distributed oceanic sediment cores
+           a compilation of globally-distributed deep sea sediment cores
            \citep{Lisiecki.Raymo.2005}. Higher \chem{\delta^{18}O} values
-           correspond to periods of lower temperatures of the deep ocean, or
-           of lower global sea level, and thus higher ice volume stored in
-           glaciers and ice sheets on land.
+           correspond to periods of lower temperatures \citep{Emiliani.1955}
+           or lower global sea level, and thus higher ice volume stored in
+           glaciers and ice sheets on land \citep{Shackleton.1967}.
            \textbf{(b)} Temperature anomalies reconstructed from a central

### 2014-07-02 kappa.tex
Add more references in the introduction.

@@
 was repartitioned between more extensive Greenland and Antarctic ice sheets,
-the Scandinavian and Barents-Kara ice sheets in the Eurasia arctic, and the
+the Scandinavian and Barents-Kara ice sheets in the Eurasian arctic, and the
 Innuitian, Laurentide and Cordilleran ice sheets over North America
@@
 
-Nevertheless, the landscape record is timely sparse and spatially incomplete,
-and geological reconstructions of former glacier extent demand a high amount of
-interpretation. Palaeo-ice sheets did not leave a systematic record of their
-passage, and much of the evidence has been overridden by multiple glacier
-re-advance and other geomorphological processes. This is where numerical
-models, incorporating the processes of glacier sliding and ice deformation
-observed on modern glaciers, can help. Although these models alone contain too
-many uncertainties to reproduce accurate picture of past glacial cycles, they
-can be calibrated against field-based evidence where it exists, while allowing
-a physics-based, quantitative reconstruction where it does not.
+Nevertheless, the landscape record is sparse in time and most often spatially
+incomplete. Palaeo-ice sheets did not leave a systematic record of their
+passage, and much of the evidence has been overridden by subsequent glacier
+re-advances and other geomorphological processes \citep{Kleman.1994}.
+Therefore, geomorphological reconstructions of former glacier extent require a
+substantial amount of interpretation. This is where numerical models,
+incorporating the processes of glacier sliding \citep[e.g.,][]{Weertman.1957}
+and deformation \citep[e.g.,][]{Nye.1953} observed on modern glaciers, can help.
+Although these models alone contain too many uncertainties to reproduce
+accurate picture of past glacial cycles \citep[e.g.,][]{Hebeler.etal.2008},
+they can be calibrated against geological data where they exist, while
+allowing a physics-based, quantitative reconstruction where they do not.
 

### 2014-07-02 kappa.tex
Activate endfloat.
Add figures in field area description.

@@
 
+% Move floats to the end
+\usepackage[nolists]{endfloat}
+\renewcommand{\efloatseparator}{\mbox{}}
+
 % Copernicus-style internal references
@@
 %\renewcommand{\subgraphics}[3][,]{}
-%\usepackage{endfloat}
-%\renewcommand{\efloatseparator}{\mbox{}}
 
@@
 
-\todo{Figure. Location map of North America}\\
-\todo{Figure. Topographic map of the Cordillera}\\
+% fig:locmap
+\begin{figure}
+  \centering
+  \makebox[0pt]{\includegraphics{map-northamerica}}
+  \caption{Relief map of northern North America showing a reconstruction of the
+           areas covered by the Cordilleran (CIS), Laurentide (LIS), Innuitian
+           (IIS) and Greenland (GIS) ice sheets during the last glacial maximum
+           \citep[21.4 to 16.8\,cal\,\chem{^{14}C}\,kyr\,BP][]{Dyke.2004}.
+           The rectangular box delimits the model domain used throughout this
+           thesis and on which most figures were drawn. The background
+           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
+           Data \citep{Patterson.Kelso.2014}.
+           \todo{Mark palaeo-ice sheets on the map.}}
+  \label{fig:locmap}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \makebox[0pt]{\includegraphics{map-cordillera}}
+  \caption{North-oriented topographic map of the northern American Cordillera.
+           Cumulative last glacial maximum ice cover from \citet{Dyke.2004}.
+           The background map consists of ETOPO1 \citep{Amante.Eakins.2009}
+           and Natural Earth Data \citep{Patterson.Kelso.2014}.
+           \todo{Mark the main reliefs. Add model domain?}}
+  \label{fig:locmap}
+\end{figure}
 
@@
 
-\todo{Figure. Cordillera wet climate photo}\\
-\todo{Figure. Cordillera dry climate photo}\\
-\todo{Figure. Climatic patters}
+\begin{figure}
+  \centering
+  \includegraphics{photo-fraser-valley}
+  \caption{This dry landscape of the Fraser Valley, central British Columbia,
+           exhibits climatic contrasts at play in the Cordillera.}
+  \label{fig:photo-fraser-valley}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \makebox[0pt]{\includegraphics{plot-atm}}
+  \caption{Some characteristics of the Cordilleran climate according to the
+           North American Regional Reanalysis
+           \citep[NARR,][]{Mesinger.etal.2006}}
+  \label{fig:plot-atm}
+\end{figure}
 
@@
 
-\todo{Figure. Glacierized region photo}\\
-\todo{Figure. Tuya Lake moraine photo}\\
-\todo{Figure. Meltwater channels photo}\\
-\todo{Figure. Lineation map}
+\begin{figure}
+  \centering
+  \makebox[0pt]{\includegraphics{photo-alaska-range}}
+  \caption{The Alaska Range hosts several of the highest peaks in the North
+           American Cordillera. Despite this fact and its northerly location,
+           its North slopes hosts relatively small glaciers only, which had not
+           been much more extensive during the Last Glacial Maximum
+           \citep{Kaufman.Manley.2004}, indicating dry climate conditions.
+           The Alaska Range is rapidly eroding, as shown by landslide-covered
+           valley sides on this photograph.}
+  \label{fig:photo-alaska-range}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \makebox[0pt]{\includegraphics{photo-tuya-moraine}}
+  \caption{The Tuya Lake Moraine in northern British Columbia.}
+  \label{fig:photo-tuya-moraine}
+\end{figure}
 

### 2014-07-03 kappa.tex
Add figures in model set-up.

@@
 that provide simple parametrisations of their effect on the ice model at
-supposedly well-defined interfaces (Fig.~??).
-
-\todo{Figure. PISM interfaces}
+supposedly well-defined interfaces (\cref{fig:model-interfaces}).
 
@@
 thesis, and give an overview of the physics involved in these choices.
-Notations used and corresponding parameter choices are summarized in Table~??.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=80mm]{model-interfaces}
+  \caption{Model interfaces in PISM. From \citet{PISM-authors.2014}.}
+  \label{fig:model-interfaces}
+\end{figure}
 
@@
 
-\todo{Figure. Ice flow problem sketch}
+\begin{figure}
+  \centering
+  \includegraphics{model-variables}
+  \caption{Model variables. \todo{This is an old figure which does not match
+           with notations in use, just to give an idea. Redraw.}}
+  \label{fig:model-variables}
+\end{figure}
 
@@
 
-\todo{Figure. SIA and SSA}
+\begin{figure}
+  \centering
+  \includegraphics[width=80mm]{model-siassa}
+  \caption{The SIA and SSA. From \citet{Winkelmann.etal.2011}. \todo{Redraw.}}
+  \label{fig:model-siassa}
+\end{figure}
 
@@
 
-\todo{Figure. SMB model algorithm}
+\begin{figure}
+  \centering
+  \includegraphics[width=80mm]{model-pdd}
+  \caption{PDD model. From \citet{PISM-authors.2014} \todo{Redraw.}}
+  \label{fig:model-pdd}
+\end{figure}
 

### 2014-07-05 cordillera-cycle.tex
Update author list again.
Completely rework the intro.

@@
 \Author[1,2]{J.}{Seguinot}
+\Author[2]{I.}{Rogozhina}
 \Author[3]{M.}{Margold}
-\Author[2]{I.}{Rogozhina}
+\Author[1]{J.}{Kleman}
 \Author[1]{A.~P.}{Stroeven}
-\Author[1]{J.}{Kleman}
 \runningauthor{J.~Seguinot et~al.}
@@
 
-During the last glacial cycle, glaciers and ice caps of the North American
-Cordillera were more extensive than today. At the last glacial maximum, a
-contiguous blanket of ice extended from the Alaska Range in the north to the
-North Cascades in the south (Fig.~\ref{fig:locmap}). This ice mass is known as
-the former Cordilleran ice sheet \citep{Dawson.1888}.
+During most of the the last glacial cycle, glaciers and ice caps of the North
+American Cordillera were more extensive than today. At the Last Glacial
+Maximum (LGM), a continuous blanket of ice, the Cordilleran ice sheet
+\citep{Dawson.1888}, extended from the Alaska Range in the north to the
+North Cascades in the south. It calved into the Pacific Ocean in the west, and
+merged with the Laurentide ice sheet in the east (Fig.~\ref{fig:locmap}).
 
-For more than a century, exploration and geological investigation of the
-Cordillera have led to a large collection of evidence of the former ice cover
-\citep{Jackson.Clague.1991}. This evidence consists of mapped boundaries of the
-former ice extent (\citealp[Fig.~1.12]{Clague.1989}; \citealp{Duk-Rodkin.1999};
-\citealp{Dyke.2004}), directions of past sliding velocities
-(\citealp{Prest.etal.1968}; \citealp[Fig.~2]{Kleman.etal.2010}), and locations
-of former
-melt-water streams \citep{Margold.etal.2011, Margold.etal.2013a}, as well as the
-timing of glaciation identified from radiocarbon dating \citep[e.g.,][]
-{Clague.1981, Porter.Swanson.1998}, cosmogenic dating \citep[e.g.,][]
-{Ward.etal.2007, Menounos.etal.2008, Stroeven.etal.2010, Stroeven.etal.2014},
-and the offshore sedimentary record \citep{Cosma.etal.2008, Davies.etal.2011}%
-\footnote{\textbf{Arjen, Martin,}, please comment on my choice of references
-    in this paragraph. Is it a fair sample of the current state of the field?
-    Am I missing some other important works?}.
+For more than a century, exploration and geological investigations of the
+Cordilleran mountains have led to many observations in support of the former
+ice sheet
+    \citep{Jackson.Clague.1991}.
+Despite the lack of documented end moraines offshore, in the zone of confluence
+with the Laurentide ice sheet, and in areas swept by the Missoula floods
+    \citep{Carrara.etal.1996},
+the south-western and north-eastern margins provide key constraints that
+allow reasonable reconstructions of the maximum ice extent
+    (\citealp{Prest.etal.1968}; \citealp[Fig. 1.12]{Clague.1989};
+     \citealp{Duk-Rodkin.1999};
+     \citealp{Booth.etal.2003}; \citealp{Dyke.2004}).
+The timing of the last glaciation, reconstructed from radiocarbon dating
+    \citep{Clague.etal.1980, Clague.1985, Clague.1986, Porter.Swanson.1998,
+           Menounos.etal.2008},
+cosmogenic exposure dating
+    \citep{Stroeven.etal.2010, Stroeven.etal.2014, Margold.etal.inpress},
+bedrock deformation in response to the ice load
+    \citep{Clague.James.2002, Clague.etal.2005},
+and the offshore sedimentary record
+    \citep{Cosma.etal.2008, Davies.etal.2011}
+indicate that the LGM ice-sheet was short-lived. However, former ice thickness
+and, therefore, the ice sheet's contribution to the last glacial maximum sea
+level low stand
+    \citep{Carlson.Clark.2012, Clark.Mix.2002}
+remain uncertain.
 
-Field-based evidence allowed to reconstruct, with a high degree of confidence,
-the maximal extent attained by the Cordilleran ice sheet during the last
-glacial cycle (\citealp[Fig. 1.12]{Clague.1989}; \citealp{Dyke.2004}). However,
-former ice thickness
-and the ice sheet's contribution to the last glacial maximum sea-level
-low-stand remain poorly constrained. Moreover, our understanding of the
-glaciation history is mainly restricted to the deglaciation phase and a
-regional scale, while little is known about the ice sheet evolution prior to
-the last glacial maximum extent \citep[Fig.~6]{Kleman.etal.2010}. Although
-time-evolving, whole ice-sheet reconstructions of glacial advance and retreat
-patterns are available for the Eurasian and Laurentide and ice sheets
-\citep{Kleman.etal.1997, Kleman.etal.2010}, such is not the case for the
-Cordilleran ice sheet, where complex arrangements of ice flow directions
-emanating from multiple glaciation centres have been identified
-\citep[Fig.~1.12]{Prest.etal.1968, Clague.1989}, but remain poorly understood
-\citep[p.~2049]{Kleman.etal.2010}.
+Our understanding of the Cordilleran glaciation history prior to the LGM is
+even more fragmentary.
+In parts of the Yukon Territory and Alaska, dated glacial erratics indicate
+an extensive Marine Isotope Stage (MIS)~4 glaciation
+    \citep{Ward.etal.2007, Ward.etal.2008, Briner.Kaufman.2008,
+           Stroeven.etal.2010, Stroeven.etal.2014},
+yet it is not known whether other regions were affected. Lineation patterns in
+the ice sheet interior include flow directions likely older than the LGM
+    \citep[Fig.~2]{Kleman.etal.2010},
+but their absolute age is uncertain.
 
-The present study uses a numerical ice sheet model \citep{PISM-authors.2014},
-calibrated against field-based evidence to perform a quantitative
-reconstruction of the Cordilleran ice sheet history through the last glacial
-cycle. Although numerical modelling has been established as a useful tool to
+In contrast, our understanding of the deglaciation history of the Cordilleran
+ice sheet since the LGM is considerable, albeit mostly at a regional scale.
+Geomorphological evidence from south-central British Columbia suggests a rapid
+deglaciation, including early emergence of the high-elevated areas and
+stagnation of thin ice cover in the lowlands
+    \citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a},
+yet this model may not apply in all areas of the Cordilleran ice sheet
+    \citep{Margold.etal.2013}.
+Finally, disparate clues for a late-glacial re-advance have been found in the
+Coast, Columbia and Rocky Mountains
+    \citep{Clague.etal.1997, Friele.Clague.2002, Friele.Clague.2002a,
+           Kovanen.2002, Kovanen.Easterbrook.2002, Lakeman.etal.2008,
+           Menounos.etal.2008},
+but the evidence is much more sparse than that of regions surrounding the North
+Atlantic \aref, thus leaving space for debate around the possibility of a
+regional late-glacial cold reversal.
+
+In general, the topographic complexity of the Cordillera has not allowed for
+ice-sheet wide reconstructions of glacial advance and retreat patterns such as
+those available for the Fennoscandian and Laurentide and ice sheets
+     \citep{Boulton.etal.2001, Dyke.Prest.1987, Dyke.etal.2003,
+            Kleman.etal.1997, Kleman.etal.2010}.
+Here, we use a numerical ice sheet model \citep{PISM-authors.2014},
+calibrated against field-based evidence, to perform a quantitative
+reconstruction of the ice sheet history through the last glacial cycle, and
+analyse some of the long-standing questions related to its evolution:
+
+\begin{itemize}
+  \item How much ice was locked in the Cordilleran ice sheet during the
+    last glacial maximum?
+  \item Which were the primary dispersal centres? Do they reflect a stable or
+    ephemeral configuration?
+  \item What was the scale of glaciation prior to the last glacial maximum?
+  \item How rapid was the deglaciation? Did it include stand-stills or
+    re-advances?
+\end{itemize}
+
+Although numerical ice sheet modelling has been established as a useful tool to
 improve our understanding of the Cordilleran ice sheet more than twenty years
-ago (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991}), the
-ubiquitously mountainous
-topography of the region has presented a major challenge to its application.
-In fact, only recent development of numerical ice sheet models and underlying
+ago (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991},
+     \citealp{Marshall.etal.2000})
+the ubiquitously mountainous
+topography of the region has presented two major challenges to its application.
+First, only recent development of numerical ice sheet models and underlying
 scientific computing tools \citep{Bueler.Brown.2009, Balay.etal.2014} allows
-high-resolution modelling of palaeo-glaciers on mountainous terrain
-over millenial time scales \citep{Golledge.etal.2012}. In addition, complex
-topography of the North American Cordillera induces strong gradients in
+high-resolution numerical modelling of palaeo-glaciers on mountainous terrain
+over millenial time scales \citep{Golledge.etal.2012}. Second, the complex
+topography of the North American Cordillera also induces strong gradients in
 seasonality and distinct patterns of precipition, thus requiring the use of
 high-resolution atmospheric forcing fields as input to the ice sheet model
-\citep{Seguinot.etal.2013}.
+\citep{Seguinot.etal.2014}.
 
-Because past climate conditions are subject to considerable uncertainty, our
-palaeoclimate forcing is a simplistic approximation of time-evolved temperature
-and precipitation fields derived from a combination of a present-day
-atmospheric reanalysis \citep{Mesinger.etal.2006}, lapse-rate corrections,
+Because climate conditions over the last glacial cycle are subject to
+considerable uncertainty, our palaeoclimate forcing is a simplistic approach
+including time-evolved temperature and precipitation fields derived from a
+combination of a present-day atmospheric reanalysis \citep{Mesinger.etal.2006},
+lapse-rate corrections,
 and temperature offset time series. The latter are obtained by scaling six
 different palaeo-temperature reconstructions from proxy records around the
-globe, including two \chem{\delta^{18}O} records from Greenland ice cores
-\citep{Dansgaard.etal.1993, Andersen.etal.2004}, two \chem{\delta^{18}O}
+globe, including two oxygen isotopes records from Greenland ice cores
+\citep{Dansgaard.etal.1993, Andersen.etal.2004}, two oxygen isotopes
 records from Antarctic ice cores \citep{Petit.etal.1999,Jouzel.etal.2007},
@@
 high-resolution precipitation forcing, as identified in a previous sensitivity
-study \citep{Seguinot.etal.2013}.
+study \citep{Seguinot.etal.2014}.
 
@@
 precipitation and latent warming of moisture-depleted air are absent in our
-model, we regard this possibility as fully probable \citep{Seguinot.etal.2013}.
+model, we regard this possibility as fully probable \citep{Seguinot.etal.2014}.
 On the other hand, in both simulations driven by the GRIP and EPICA records,

### 2014-07-06 kappa.tex
Rescale all intro photos to 500x300 px.
Add Susitna glacier photo by Austin Post.

@@
     \hspace{1cm}
-    \subgraphics{b}{photo-glacier-fold}
+    \subgraphics{b}{photo-glacier-susitna}
+    \hspace{1cm}
+    \subgraphics{c}{photo-glacier-fold}
   }
-  \caption{Different mechanical behaviours of glacier ice.
-           \textbf{(a)} Brittle behaviour. Ice flowing over a hidden bedrock
-           knob, resulting in fracturing at the glacier surface.
-           Glacier Blanc, French Alps.
-           \textbf{(b)} Ductile behaviour. Mountain glacier calving into a
-           small lake, revealing a folded structure formed by deformed rock
-           layers in the ice.}
+  \caption{Illustration of various glacier ice behaviours.
+           \textbf{(a)} Surface crevasses formed by ice flow over a hidden
+           bedrock knob. Glacier Blanc, French Alps, 2004.
+           \textbf{(b)} Medial moraines illustrating downhill ice flow on
+           Susitna Glacier, Alaska Range, by Austin Post, 1970
+           \citep{NSIDC.2009}.
+           \textbf{(c)} Mountain glacier calving into a small lake, revealing a
+           folded structure formed by deformed debris layers in the ice.
+           Glacier d'Arsine, French Alps, 2004.}
   \label{fig:glacier-mechanics}
@@
   \makebox[0pt]{
-    \subgraphics[width=50mm]{a}{photo-glacial-valley}
+    \subgraphics{a}{photo-glacial-valley}
     \hspace{1cm}
-    \subgraphics[width=50mm]{b}{photo-erratic-boulder}
+    \subgraphics{b}{photo-erratic-boulder}
     \hspace{1cm}
-    \subgraphics[width=50mm]{c}{photo-melt-channels}
+    \subgraphics{c}{photo-melt-channels}
   }
@@
            characteristic of erosion by glaciers.
-           Cirque du Fer-\`{a}-Cheval, French Alps.
+           Cirque du Fer-\`{a}-Cheval, French Alps, 2013.
            \textbf{(b)} Erratic boulders, deposited in an improbable
            arrangement as the glacier melted underneath.
-           Finse area, Norwegian Scandes.
+           Finse area, Norwegian Scandes, 2011.
            \textbf{(c)} Lateral meltwater channels, formed on a gentle slope
            as the ice margin retreated towards the valley bottom. Tuya Lake
-           area, Canadian Cordillera.}
+           area, Canadian Cordillera, 2012.}
   \label{fig:glaciation-indicators}

### 2014-07-06 kappa.tex
Add field area description first shot.

@@
 
-% ----------------------------------------------------------------------
-\subsection{Geographic setting}
-% ----------------------------------------------------------------------
-
-% fig:locmap
+During the last glacial cycle, much of North America has been covered by ice.
+At the last glacial maximum, apex of glaciation, three major ice sheets had
+collided to form a continuous blanket of ice that spanned from Alaska to
+Newfoundland, connecting with the Greenland ice sheet in the Canadian Arctic,
+and extending as far south as Chicago and the Great Lakes. The Laurentide ice
+sheet, largest of the three ice sheets, was centred on the modern Hudson Bay,
+and fully covered the Canadian Prairies, Eastern Canada, and the Great Lakes
+Basin. The Innuitian ice sheet, much smaller, covered the northernmost Canadian
+archipelago. Finally, the Cordilleran ice sheet capped the mountain ranges of
+the Western Cordillera, our region of focus in the present thesis.
+
 \begin{figure}
@@
 
+% ----------------------------------------------------------------------
+\subsection{Topographic setting}
+% ----------------------------------------------------------------------
+
+In North American, the Western Cordillera forms a broad mountain system with an
+average east-west width of a about 1000\,km. Its topography is a complex
+arrangement of mountain ranges and intramontane basins. It was formed during
+multiple accretion phases associated with the closure of several back-arc
+oceanic trenches over a 200\,Ma-long geologic history.
+
+Most of the major reliefs are found along two major bands that run parallel to
+each other along the general orientation of the Cordillera. The western band
+consists of the Pacific Coast orogenic belt. Major ice-covered reliefs include,
+from North to South, the Alaska Range, the Wranglell and Saint-Elias Mountains,
+the Coast Mountains and the North Cascades. The eastern band consists of the
+Nevadan and Laramide belts, which are separated by the deep and strikingly
+rectilinear Rocky Mountain / Tintina Trench. It includes the Selwyn and Mac
+Kenzie Mountains to the north, and the Rocky Mountains to the South. The brooks
+range, in northern Alaska, has been glaciated but its glaciers were not
+connected to the main body of the Cordilleran ice sheet.
+
+The Cordillera comports several intermontane bansins, which, as we will see
+further, host particular climate properties and significantly impacted patterns
+of ice sheet flow. The two major depressions are the Interior Plateau of
+central British Columbia, and the Liard Lowland. Between these two basins is
+found a mildly elevated, but extensive and deeply incised mountain group known
+as the Skeena Mountains.
+
+The ubiquitously mountainous topography of the region presents a serious
+obstacle to reconstructing past glacier dynamics in the Cordillera. Because of
+the irregular arrangement of valleys and ridges that forms the ice sheet’s
+floor, ice moved along tortuous and intertwined pathways, sometimes following
+the major topographical troughs, and sometimes flowing across them. In terms of
+numerical glacier modelling, it implies complex patterns of flow and stress
+within the ice, and requires high-resolution, and thus, large, computational
+grids.
+
 \begin{figure}
@@
 \end{figure}
-
-\begin{table}
-  \centering
-  \caption{Properties of the Canadian Atlas Lambert conformal conic projection
-           (EPSG code~3978) and the GRS~1980 ellipsoid, and corresponding
-           PROJ.4 parameter names, as used in all model runs, ice volume and
-           area calculations, and most figures.}
-  \label{tab:projparams}
-  \makebox[0pt]{\begin{tabular*}{80mm}{@{\extracolsep{\fill}}lcc}
-    \hline
-    Parameter               & Name              & Value \\
-    \hline
-    Projection name         & \texttt{+proj}    & \texttt{lcc} \\
-    Central meridian        & \texttt{+lon\_0}  & 95\degree W \\
-    Latitude of origin      & \texttt{+lat\_0}  & 49\degree N \\
-    First secant parallel   & \texttt{+lat\_1}  & 49\degree N \\
-    Second secant parallel  & \texttt{+lat\_2}  & 77\degree N \\
-    Semimajor radius        & \texttt{+a}       & 6378137\,m \\
-    Reciprocal flattening   & \texttt{+rf}      & 298.257222101 \\
-    \hline
-  \end{tabular*}}
-\end{table}
-
-% ----------------------------------------------------------------------
-\subsection{Climatic setting}
-% ----------------------------------------------------------------------
-
-\begin{figure}
-  \centering
-  \includegraphics{photo-fraser-valley}
-  \caption{This dry landscape of the Fraser Valley, central British Columbia,
-           exhibits climatic contrasts at play in the Cordillera.}
-  \label{fig:photo-fraser-valley}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \makebox[0pt]{\includegraphics{plot-atm}}
-  \caption{Some characteristics of the Cordilleran climate according to the
-           North American Regional Reanalysis
-           \citep[NARR,][]{Mesinger.etal.2006}}
-  \label{fig:plot-atm}
-\end{figure}
-
-% ----------------------------------------------------------------------
-\subsection{Palaeo-glaciology}
-% ----------------------------------------------------------------------
 
@@
 
+% ----------------------------------------------------------------------
+\subsection{Climatic setting}
+% ----------------------------------------------------------------------
+
+The topographic complexity of the North American Cordillera is associated with
+distinctive patterns of temperature and precipitation over the model domain.
+Due to the large latitudinal extent of the region, mean annual temperatures
+vary between -12 and 12\,\degC. About half the model domain has a mean annual
+temperature below freezing. However, seasonality, here computed as the
+temperature difference between the warmest and the coldest months, varies
+between around 10\,\degC along most of the Pacific Coast to above 40\,\degC in
+the polar interior. This concerns most of the area above 60\degree~N, and
+high-elevated mountain ranges below this latitude. There exist a partial
+correlation between cold regions and those that have a large seasonality, so
+that many regions where mean annual temperatures are well below freezing point
+do experience warm summers.
+
+Annual precipitation varies between less than half-a-meter in some of the
+interior areas to several meters in some parts of maritime Alaska.
+Precipitation rates are highest on the western slopes of the Pacific Coast
+ranges and decline quickly on the other side of the mountains. In terms of ice
+sheet mass balance, this contrast is even reinforced by the difference in
+timing of the precipitation peak through the year. Coastal regions experience
+most precipitation during winter and fall, thus as snow. On the contrary, many
+inland regions experience dry winters and most of the annual precipitation
+consequently falls as rain during the ablation season. The climatic setting of
+the North American Cordillera is described further in \CCLI, where its effects
+on the ice sheet model have been studied in detail.
+
 \begin{figure}
   \centering
-  \makebox[0pt]{\includegraphics{photo-tuya-moraine}}
-  \caption{The Tuya Lake Moraine in northern British Columbia.}
-  \label{fig:photo-tuya-moraine}
+  \includegraphics{photo-fraser-valley}
+  \caption{This dry landscape of the Fraser Valley, central British Columbia,
+           exhibits climatic contrasts at play in the Cordillera.}
+  \label{fig:photo-fraser-valley}
 \end{figure}
+
+\begin{figure}
+  \centering
+  \makebox[0pt]{\includegraphics{plot-atm}}
+  \caption{Some characteristics of the Cordilleran climate according to the
+           North American Regional Reanalysis
+           \citep[NARR,][]{Mesinger.etal.2006}}
+  \label{fig:plot-atm}
+\end{figure}
+
+% ----------------------------------------------------------------------
+\subsection{Palaeo-glaciology}
+% ----------------------------------------------------------------------
+
+The Cordilleran ice sheet covered most of the Canadian Cordillera, as well
+ajacent some of the adjacent land in north-western contiguous USA and Alaska.
+The southern margin is the one that has been most studied. It consisted of
+different ice lobes, that have formed at the exutory of the main valleys. The
+Puget lobe was the westernmost and largest of these lobes. It covered the Puget
+Lowland, a region that includes the current location of the cities of Vancouver
+and Seattle. To the East of the Cascade Range were found the Okanagan and
+Purcell lobes. The latter once dammed the Clark Fork River to form Glacial Lake
+Missoula. As the ice sheet retreeted, the lake repeatedly and catastrophically
+drained to the Pacific Ocean, forming on curse of the flood a lanscape now
+known as the Channeled Scablands.
+
+The western margin of the ice sheet covered part of the continental shelf. An
+ice cap covering Vancouver Island was well-connected to main body of ice on the
+continent. A large portion of the ice flow thus has engulfed between Vancouver
+Island and the Olympic Peninsula, where it could find an exutory to the Pacific
+Ocean, at the Juan the Fuca Lobe. Further to the North, the ice margin reached
+Queen Charlotte Islands, yet the archipelago may not have been covered by ice
+flowing from the land. Submarine troughs likely idicate the erosional action of
+former ice streams, yet it is not clear wether they have formed during the last
+glacial cycle or earlier.
+
+The northern margin of the ice sheet is in several ways more complex than the
+southern margin. Firstly geomorphological evidence indicate multiple glacier
+advances. During the last glacial cycle, at least two glaciations occured, and
+in some places, the first glaciation was more extensive than the second.
+Secondly, geomorphological evidence indicate that the extent of glaciers in the
+North was relatively restricted, despite very cold temperatures at present.
+
+Finally, the eastern margin of the ice sheet is the most unclear. Although it
+is known that the Cordilleran ice sheet collided with its Laurentide neighbour,
+the precise location of this junction is unknown, and little information exists
+on its timing. More importantly, the nature of the interaction between the two
+ice sheets is uncertain. It is not known for instance, if the junction lasted
+long enough for the ice sheets to significantly alter their respective flow
+dynamics. Because the present thesis focuses specifically on the Cordilleran
+ice sheet, certainly our largest assumption consists in neglecting butressing
+effects from the Laurentide ice sheet.
+
+Although the last glacial maximum extent of the Cordilleran ice sheet, apart of
+its eastern margin, is relatively well known, the history of deglaciation is
+heavily debated, while little information exists on earlier dynamics of ice
+sheet.
 
@@
 
+\begin{table}
+  \centering
+  \caption{Properties of the Canadian Atlas Lambert conformal conic projection
+           (EPSG code~3978) and the GRS~1980 ellipsoid, and corresponding
+           PROJ.4 parameter names, as used in all model runs, ice volume and
+           area calculations, and most figures.}
+  \label{tab:projparams}
+  \makebox[0pt]{\begin{tabular*}{80mm}{@{\extracolsep{\fill}}lcc}
+    \hline
+    Parameter               & Name              & Value \\
+    \hline
+    Projection name         & \texttt{+proj}    & \texttt{lcc} \\
+    Central meridian        & \texttt{+lon\_0}  & 95\degree W \\
+    Latitude of origin      & \texttt{+lat\_0}  & 49\degree N \\
+    First secant parallel   & \texttt{+lat\_1}  & 49\degree N \\
+    Second secant parallel  & \texttt{+lat\_2}  & 77\degree N \\
+    Semimajor radius        & \texttt{+a}       & 6378137\,m \\
+    Reciprocal flattening   & \texttt{+rf}      & 298.257222101 \\
+    \hline
+  \end{tabular*}}
+\end{table}
+
 % ======================================================================
@@
 % Geomorphological-oriented plots
+
+\begin{figure}
+  \centering
+  \makebox[0pt]{\includegraphics{photo-tuya-moraine}}
+  \caption{The Tuya Lake Moraine in northern British Columbia.}
+  \label{fig:photo-tuya-moraine}
+\end{figure}
 

### 2014-07-07 kappa.tex
Small corrections in field area description.

@@
 Newfoundland, connecting with the Greenland ice sheet in the Canadian Arctic,
-and extending as far south as Chicago and the Great Lakes. The Laurentide ice
-sheet, largest of the three ice sheets, was centred on the modern Hudson Bay,
+and extending as far south as Chicago and the Great Lakes
+\cref{fig:map-northamerica}. The Laurentide ice
+sheet, largest of the three, was centred on the modern Hudson Bay,
 and fully covered the Canadian Prairies, Eastern Canada, and the Great Lakes
@@
 archipelago. Finally, the Cordilleran ice sheet capped the mountain ranges of
-the Western Cordillera, our region of focus in the present thesis.
+the Western Cordillera, which forms our region of focus in the present thesis.
 
@@
 In North American, the Western Cordillera forms a broad mountain system with an
-average east-west width of a about 1000\,km. Its topography is a complex
-arrangement of mountain ranges and intramontane basins. It was formed during
+average east-west width of about 1000\,km. Its topography is a complex
+arrangement of mountain ranges and interior basins. It was formed during
 multiple accretion phases associated with the closure of several back-arc
-oceanic trenches over a 200\,Ma-long geologic history.
-
-Most of the major reliefs are found along two major bands that run parallel to
-each other along the general orientation of the Cordillera. The western band
-consists of the Pacific Coast orogenic belt. Major ice-covered reliefs include,
-from North to South, the Alaska Range, the Wranglell and Saint-Elias Mountains,
-the Coast Mountains and the North Cascades. The eastern band consists of the
+oceanic trenches over a 200\,Ma-long geological history \mref.
+
+Most of the dominant reliefs are found along two major strips that run parallel
+to each other along the general orientation of the Cordillera
+\cref{fig:map-cordillera}. The western strip consists of the Pacific Coast
+orogenic belt \mref. Major ice-covered reliefs include,
+from North to South, the Alaska Range, the Wrangell and Saint-Elias Mountains,
+the Coast Mountains and the North Cascades. The eastern strip consists of the
 Nevadan and Laramide belts, which are separated by the deep and strikingly
-rectilinear Rocky Mountain / Tintina Trench. It includes the Selwyn and Mac
-Kenzie Mountains to the north, and the Rocky Mountains to the South. The brooks
-range, in northern Alaska, has been glaciated but its glaciers were not
-connected to the main body of the Cordilleran ice sheet.
-
-The Cordillera comports several intermontane bansins, which, as we will see
-further, host particular climate properties and significantly impacted patterns
+linear Rocky Mountain / Tintina Trench \mref. It includes the Selwyn and
+MacKenzie Mountains in the north, and the Rocky Mountains in the South. The
+Brooks range, in northern Alaska, has been covered by ice, but its glaciers
+were not connected to the main body of the Cordilleran ice sheet.
+
+The Cordillera comports several intermontane basins that impacted patterns
 of ice sheet flow. The two major depressions are the Interior Plateau of
@@
 The ubiquitously mountainous topography of the region presents a serious
-obstacle to reconstructing past glacier dynamics in the Cordillera. Because of
-the irregular arrangement of valleys and ridges that forms the ice sheet’s
+obstacle to reconstructing past dynamics of the Cordilleran ice sheet. Because
+of the irregular arrangement of valleys and ridges that formed the ice sheet's
 floor, ice moved along tortuous and intertwined pathways, sometimes following
@@
 numerical glacier modelling, it implies complex patterns of flow and stress
-within the ice, and requires high-resolution, and thus, large, computational
+within the ice, and requires high-resolution, and thus, large computational
 grids.
@@
 temperature difference between the warmest and the coldest months, varies
-between around 10\,\degC along most of the Pacific Coast to above 40\,\degC in
-the polar interior. This concerns most of the area above 60\degree~N, and
+between around 10\,\degC\ along most of the Pacific Coast to above 40\,\degC\
+in the polar interior. This concerns most of the area above 60\degree~N, and
 high-elevated mountain ranges below this latitude. There exist a partial

### 2014-07-07 kappa.tex
Various style corrections.

--- 2014-07-06 kappa.tex
@@
 
-As ice flows and glide on its bed, it transports rock debris and erode the
+As ice flows and glides on its bed, it transports rock debris and erode the
 landscape, thence leaving geomorphological traces of its former presence
@@
 advent of the glacial theory, birthed the idea that, under colder temperatures
-than today, expansive ice sheets had once covered much of Europe and North
-America \citep{Agassiz.1840}.
+than today, expansive \emph{ice sheets} had once covered much of Europe and
+North America \citep{Agassiz.1840}.
 
@@
 years, glacial and interglacial cycles have succeeded each other with a
-periodicity 100\,kyr \citep{Hays.etal.1976}. In the present thesis, we focus on
-the most recent of these cycles: the last glacial cycle
+100\,kyr periodicity \citep{Hays.etal.1976}. In the present thesis, we focus on
+the most recent of these cycles: \emph{the last glacial cycle}
 (\cref{fig:paleo-timeseries})
@@
 the Scandinavian and Barents-Kara ice sheets in the Eurasian arctic, and the
-Innuitian, Laurentide and Cordilleran ice sheets over North America
+Innuitian, Laurentide and \emph{Cordilleran} ice sheets over North America
 (\cref{fig:paleo-glaciation}).
@@
 
-% ----------------------------------------------------------------------
-\subsection{Geographic setting}
-% ----------------------------------------------------------------------
-
-% fig:locmap
+During the last glacial cycle, much of North America has been covered by ice.
+At the last glacial maximum, apex of glaciation, three major ice sheets had
+collided to form a continuous blanket of ice that spanned from Alaska to
+Newfoundland, connecting with the Greenland ice sheet in the Canadian Arctic,
+and extending as far south as Chicago and the Great Lakes
+\cref{fig:map-northamerica}. The Laurentide ice
+sheet, largest of the three, was centred on the modern Hudson Bay,
+and fully covered the Canadian Prairies, Eastern Canada, and the Great Lakes
+Basin. The Innuitian ice sheet, much smaller, covered the northernmost Canadian
+archipelago. Finally, the Cordilleran ice sheet capped the mountain ranges of
+the Western Cordillera, which forms our region of focus in the present thesis.
+
 \begin{figure}
@@
            \todo{Mark palaeo-ice sheets on the map.}}
-  \label{fig:locmap}
+  \label{fig:map-northamerica}
 \end{figure}
+
+% ----------------------------------------------------------------------
+\subsection{Topographic setting}
+% ----------------------------------------------------------------------
+
+In North American, the Western Cordillera forms a broad mountain system with an
+average east-west width of about 1000\,km. Its topography is a complex
+arrangement of mountain ranges and interior basins. It was formed during
+multiple accretion phases associated with the closure of several back-arc
+oceanic trenches over a 200\,Ma-long geological history \mref.
+
+Most of the dominant reliefs are found along two major strips that run parallel
+to each other along the general orientation of the Cordillera
+\cref{fig:map-cordillera}. The western strip consists of the Pacific Coast
+orogenic belt. Major ice-covered reliefs include,
+from North to South, the Alaska Range, the Wrangell and Saint-Elias Mountains,
+the Coast Mountains and the North Cascades. The eastern strip consists of the
+Nevadan and Laramide belts, which are separated by the deep and strikingly
+linear Rocky Mountain / Tintina Trench. It includes the Selwyn and
+MacKenzie Mountains in the north, and the Rocky Mountains in the South. The
+Brooks range, in northern Alaska, has been covered by ice, but its glaciers
+were not connected to the main body of the Cordilleran ice sheet.
+
+The Cordillera comports several intermontane basins that impacted patterns
+of ice sheet flow. The two major depressions are the Interior Plateau of
+central British Columbia, and the Liard Lowland. Between these two basins is
+found a mildly elevated, but extensive and deeply incised mountain group, known
+as the Skeena Mountains.
+
+The ubiquitously mountainous topography of the region presents a serious
+obstacle to reconstructing past dynamics of the Cordilleran ice sheet. Because
+of the irregular arrangement of valleys and ridges that formed the ice sheet's
+floor, ice moved along tortuous and intertwined pathways, sometimes following
+the major topographical troughs, and sometimes flowing across them \mref. In
+terms of
+numerical glacier modelling, it implies complex patterns of flow and stress
+within the ice, and requires high-resolution, and thus, large computational
+grids.
 
@@
            \todo{Mark the main reliefs. Add model domain?}}
-  \label{fig:locmap}
+  \label{fig:map-cordillera}
 \end{figure}
-
-\begin{table}
-  \centering
-  \caption{Properties of the Canadian Atlas Lambert conformal conic projection
-           (EPSG code~3978) and the GRS~1980 ellipsoid, and corresponding
-           PROJ.4 parameter names, as used in all model runs, ice volume and
-           area calculations, and most figures.}
-  \label{tab:projparams}
-  \makebox[0pt]{\begin{tabular*}{80mm}{@{\extracolsep{\fill}}lcc}
-    \hline
-    Parameter               & Name              & Value \\
-    \hline
-    Projection name         & \texttt{+proj}    & \texttt{lcc} \\
-    Central meridian        & \texttt{+lon\_0}  & 95\degree W \\
-    Latitude of origin      & \texttt{+lat\_0}  & 49\degree N \\
-    First secant parallel   & \texttt{+lat\_1}  & 49\degree N \\
-    Second secant parallel  & \texttt{+lat\_2}  & 77\degree N \\
-    Semimajor radius        & \texttt{+a}       & 6378137\,m \\
-    Reciprocal flattening   & \texttt{+rf}      & 298.257222101 \\
-    \hline
-  \end{tabular*}}
-\end{table}
-
-% ----------------------------------------------------------------------
-\subsection{Climatic setting}
-% ----------------------------------------------------------------------
-
-\begin{figure}
-  \centering
-  \includegraphics{photo-fraser-valley}
-  \caption{This dry landscape of the Fraser Valley, central British Columbia,
-           exhibits climatic contrasts at play in the Cordillera.}
-  \label{fig:photo-fraser-valley}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \makebox[0pt]{\includegraphics{plot-atm}}
-  \caption{Some characteristics of the Cordilleran climate according to the
-           North American Regional Reanalysis
-           \citep[NARR,][]{Mesinger.etal.2006}}
-  \label{fig:plot-atm}
-\end{figure}
-
-% ----------------------------------------------------------------------
-\subsection{Palaeo-glaciology}
-% ----------------------------------------------------------------------
 
@@
 
+% ----------------------------------------------------------------------
+\subsection{Climatic setting}
+% ----------------------------------------------------------------------
+
+The topographic complexity of the North American Cordillera is associated with
+distinctive patterns of temperature and precipitation, whose effect is often
+clearly visible in the landscape (\cref{fig:photo-fraser-valley}). Due to the
+large latitudinal extent of the region, mean annual temperatures vary
+significantly from north to south (\cref{fig:plot-atm}a). About half of the
+model domain has mean annual temperature below freezing. However, seasonality,
+here computed as the temperature difference between the warmest and the coldest
+months, varies between around 10\,\degC\ along most of the Pacific Coast to
+above 40\,\degC\ in the polar interior (\cref{fig:plot-atm}b). There exist a
+partial correlation between cold regions and those that have a large
+seasonality, so that many regions where mean annual temperature is well below
+freezing point do experience warm summers.
+
+Precipitation rates are highest on the western slopes of the Pacific Coast
+ranges and decline quickly on the other side of the mountains
+(\cref{fig:plot-atm}c). In terms of ice sheet mass balance, this contrast is
+even reinforced by the difference in timing of the precipitation peak through
+the year (\cref{fig:plot-atm}d). Coastal regions experience most precipitation
+during winter and fall, thus as snow. On the contrary, many inland regions
+experience dry winters, and most of the annual precipitation consequently falls
+as rain during the ablation season. The climatic setting of the North American
+Cordillera is described further in \CCLI, where its effects on the ice sheet
+model response have been studied in detail.
+
 \begin{figure}
   \centering
-  \makebox[0pt]{\includegraphics{photo-tuya-moraine}}
-  \caption{The Tuya Lake Moraine in northern British Columbia.}
-  \label{fig:photo-tuya-moraine}
+  \includegraphics{photo-fraser-valley}
+  \caption{This dry landscape of the Fraser Valley, central British Columbia,
+           exhibits climatic contrasts at play in the Cordillera.}
+  \label{fig:photo-fraser-valley}
 \end{figure}
+
+\begin{figure}
+  \centering
+  \makebox[0pt]{\includegraphics{plot-atm}}
+  \caption{Some characteristics of the Cordilleran climate according to the
+           North American Regional Reanalysis
+           \citep[NARR,][]{Mesinger.etal.2006}}
+  \label{fig:plot-atm}
+\end{figure}
+
+% ----------------------------------------------------------------------
+\subsection{Palaeo-glaciology}
+% ----------------------------------------------------------------------
+
+During the last glacial maximum, the Cordilleran ice sheet covered most of the
+Canadian Cordillera, and some of the adjacent land in north-western contiguous
+USA and Alaska (\cref{map-cordillera}). The southern margin is the most
+extensively studied. It consists of different piedmont lobes that have formed
+at the outlet of the main valleys. The Puget lobe was the westernmost and
+largest lobe. It covered the Puget Lowland, a region that currently includes
+the cities of Vancouver and Seattle. East of the Cascade Range, were found the
+Okanagan and Purcell lobes. The latter once dammed the Clark Fork River to form
+Glacial Lake Missoula \mref. As the ice sheet retreated, the lake repeatedly
+and catastrophically drained to the Pacific Ocean \mref. These floods
+contributed to shape a distinctive landscape known as the Channeled Scablands.
+
+The western margin of the ice sheet covered part of the continental shelf. An
+ice cap covering Vancouver Island was well-connected to main body of ice on the
+continent. A significant part of the ice sheet thus drained into the Georgia
+Strait and engulfed between Vancouver Island and the Olympic Peninsula before
+it formed the Juan the Fuca marine ice lobe, calving into the Pacific Ocean.
+Further to the North, the ice margin reached Queen Charlotte Islands, yet the
+archipelago may not have been fully covered by ice
+flowing from the land. Submarine troughs likely idicate the erosional action of
+former ice streams, yet it is not clear whether they have formed during the
+last glacial cycle or earlier.
+
+The northern margin of the ice sheet is in several ways more complex than the
+southern margin. Firstly, geomorphological evidence indicate multiple glacier
+advances. During the last glacial cycle, at least two glaciations occured, and
+in some places, the first glaciation was more extensive than the second.
+Secondly, geomorphological evidence indicate that the extent of glaciers in the
+North was relatively restricted, despite very cold temperatures at present.
+
+Finally, the eastern margin of the ice sheet is the most unclear. Although it
+is known that the Cordilleran ice sheet collided with its Laurentide neighbour,
+the precise location of this junction is unknown, and little information exists
+on its timing. More importantly, the nature of the interaction between the two
+ice sheets is uncertain. It is not known for instance, if the junction lasted
+long enough for the ice sheets to significantly alter their respective flow
+dynamics. Because the present thesis focuses specifically on the Cordilleran
+ice sheet, our largest assumption certainly consists of neglecting buttressing
+effects from the Laurentide ice sheet.
+
+Although the last glacial maximum extent of the Cordilleran ice sheet, apart of
+its eastern margin, is relatively well known, the history of deglaciation is
+heavily debated, while little information exists on earlier dynamics of ice
+sheet.
 

### 2014-07-07 figures/mindmap.tex
Add minimal mindmap.

\documentclass[border=5mm,tikz]{standalone}

\usepackage{amsmath}
\usetikzlibrary{mindmap}

\tikzstyle{level 1 concept}+=[font=\bf, sibling angle=120]

\begin{document}
\begin{tikzpicture}
  \path[mindmap, grow cyclic, every node/.style=concept, concept color=black,text=white]
    node {This thesis}
    [clockwise from=150]
    child[concept color=green!50!black] { node {Cordilleran ice sheet}
      [clockwise from=210]
      child { node {Geomorphology} }
      child { node {Geology} }
    }  
    child[concept color=blue] { node {ice sheet modelling} }
    child[concept color=orange!75!black] { node {numerical}
      [clockwise from=-60]
      child { node {discretization} }
      child { node {implementation} }
    }
;
\end{tikzpicture}
\end{document}

### 2014-07-07 kappa.tex
Update figure captions.

@@
   \includegraphics[width=80mm]{paleo-timeseries}
-  \caption{Indicators of climatic fluctuations during the last 800\,ka of
-           Earth history.
-           \textbf{(a)} Benthic oxygen isotope \chem{\delta^{18}O} record from
-           a compilation of globally-distributed deep sea sediment cores
+  \caption{Palaeo-climate records of the last 800\,ka.
+           \textbf{(a)} Benthic oxygen isotope (\chem{\delta^{18}O}) record
+           from a compilation of globally-distributed deep sea sediment cores
            \citep{Lisiecki.Raymo.2005}. Higher \chem{\delta^{18}O} values
@@
            glaciers and ice sheets on land \citep{Shackleton.1967}.
-           \textbf{(b)} Temperature anomalies reconstructed from a central
-           Antarctic ice core \citep[EPICA,][]{Jouzel.etal.2007}.
-           The grey boxes indicate the approximate time span of the last
-           glacial period of 110 to 10\,ka, which corresponds to our period of
-           interest in the present thesis.}
+           \textbf{(b)} Temperature anomalies ($\Delta T$) reconstructed from a
+           central Antarctic ice core \citep[EPICA Dome~C,][]{Jouzel.etal.2007}.
+           The grey bands indicate the approximate time span of the last
+           glacial period of 110 to 10\,ka.}
   \label{fig:paleo-timeseries}
@@
   \caption{Spatial extent of ice sheet and glacier cover in the northern
-           hemisphere \textbf{(a)} at present, and \textbf{(b)} during the last
-           glacial maximum \citep{Ehlers.Gibbard.2007}. Coastlines and present
-           glaciers are made with Natural Earth \citep{Patterson.Kelso.2014}.}
+           hemisphere \textbf{(a)} at present \citep{Patterson.Kelso.2014}, and
+           \textbf{(b)} during the last glacial maximum
+           \citep{Ehlers.Gibbard.2007}.}
   \label{fig:paleo-glaciation}
@@
            (IIS) and Greenland (GIS) ice sheets during the last glacial maximum
-           \citep[21.4 to 16.8\,cal\,\chem{^{14}C}\,kyr\,BP][]{Dyke.2004}.
-           The rectangular box delimits the model domain used throughout this
-           thesis and on which most figures were drawn. The background
+           \citep[21.4 to 16.8\,cal\,\chem{^{14}C}\,kyr\,BP,][]{Dyke.2004}.
+           The rectangular box delimits the model domain used in all simulation
+           and most figures of this thesis. The background
            map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
-           Data \citep{Patterson.Kelso.2014}.
-           \todo{Mark palaeo-ice sheets on the map.}}
+           Data \citep{Patterson.Kelso.2014}.}
   \label{fig:map-northamerica}
@@
   \makebox[0pt]{\includegraphics{map-cordillera}}
-  \caption{North-oriented topographic map of the northern American Cordillera.
-           Cumulative last glacial maximum ice cover from \citet{Dyke.2004}.
+  \caption{Topographic map of the northern American Cordillera, oriented north.
+           Cumulative last glacial maximum ice cover \citep{Dyke.2004} is
+           marked in red, and the model domain in black.
            The background map consists of ETOPO1 \citep{Amante.Eakins.2009}
-           and Natural Earth Data \citep{Patterson.Kelso.2014}.
-           \todo{Mark the main reliefs. Add model domain?}}
+           and Natural Earth Data \citep{Patterson.Kelso.2014}.}
   \label{fig:map-cordillera}
@@
            American Cordillera. Despite this fact and its northerly location,
-           its North slopes hosts relatively small glaciers only, which had not
+           its north slope hosts relatively small glaciers only, which had not
            been much more extensive during the Last Glacial Maximum
-           \citep{Kaufman.Manley.2004}, indicating dry climate conditions.
+           \citep{Kaufman.Manley.2004}, indicating dry climatic conditions.
            The Alaska Range is rapidly eroding, as shown by landslide-covered
@@
   \includegraphics[width=80mm]{model-interfaces}
-  \caption{Model interfaces in PISM. From \citet{PISM-authors.2014}.}
+  \caption{Model interfaces in PISM. From \citet{PISM-authors.2014}.
+           \todo{Redraw.}}
   \label{fig:model-interfaces}
@@
   \includegraphics{model-variables}
-  \caption{Model variables. \todo{This is an old figure which does not match
-           with notations in use, just to give an idea. Redraw.}}
+  \caption{Sketch of notations used for the main variables of the ice sheet
+           model.}
   \label{fig:model-variables}
@@
   \includegraphics[width=80mm]{model-siassa}
-  \caption{The SIA and SSA. From \citet{Winkelmann.etal.2011}. \todo{Redraw.}}
+  \caption{Shallow Ice Approximation (SIA) and Shallow Shelf Approximation
+           (SSA) velocities in PISM. From \citet{Winkelmann.etal.2011}.
+           \todo{Redraw.}}
   \label{fig:model-siassa}

### 2014-07-08 kappa.tex
Add references.

@@
 and extending as far south as Chicago and the Great Lakes
-\cref{fig:map-northamerica}. The Laurentide ice
+(\cref{fig:map-northamerica}). The Laurentide ice
 sheet, largest of the three, was centred on the modern Hudson Bay,
@@
 
-In North American, the Western Cordillera forms a broad mountain system with an
+The North American Cordillera forms a broad mountain system with an
 average east-west width of about 1000\,km. Its topography is a complex
-arrangement of mountain ranges and interior basins. It was formed during
-multiple accretion phases associated with the closure of several back-arc
-oceanic trenches over a 200\,Ma-long geological history \mref.
-
-Most of the dominant reliefs are found along two major strips that run parallel
-to each other along the general orientation of the Cordillera
-\cref{fig:map-cordillera}. The western strip consists of the Pacific Coast
-orogenic belt. Major ice-covered reliefs include,
+jigsaw of mountain ranges and interior basins (\cref{fig:map-cordillera}).
+It was formed during multiple accretion phases associated with the closure of
+several back-arc oceanic trenches over a 200\,Ma-long geological history
+\citep{Sigloch.Mihalynuk.2013}.
+
+Most of the dominant reliefs are found along two major ribbons that run
+parallel to each other along the general orientation of the Cordillera
+(\cref{fig:map-cordillera}). The western ribbon is formed by the Pacific Coast
+orogenic belt. Major glaciated reliefs include,
 from North to South, the Alaska Range, the Wrangell and Saint-Elias Mountains,
-the Coast Mountains and the North Cascades. The eastern strip consists of the
+the Coast Mountains and the North Cascades. The eastern ribbon consists of the
 Nevadan and Laramide belts, which are separated by the deep and strikingly
-linear Rocky Mountain / Tintina Trench. It includes the Selwyn and
+linear Rocky Mountain / Tintina Trench \mref. It includes the Selwyn and
 MacKenzie Mountains in the north, and the Rocky Mountains in the South. The
-Brooks range, in northern Alaska, has been covered by ice, but its glaciers
-were not connected to the main body of the Cordilleran ice sheet.
-
-The Cordillera comports several intermontane basins that impacted patterns
-of ice sheet flow. The two major depressions are the Interior Plateau of
-central British Columbia, and the Liard Lowland. Between these two basins is
-found a mildly elevated, but extensive and deeply incised mountain group, known
-as the Skeena Mountains.
+Brooks Range, in northern Alaska, has been covered by ice, but its glaciers
+were not connected to the main body of the Cordilleran ice sheet
+\citep{Kaufman.Manley.2004}.
+
+\todo{Go to geolibrary and check \emph{Geology of Canada} books and such for a
+      reference about the three orogenic belts.}
+
+The Cordillera comports several intermontane basins, which influenced the
+patterns of ice flow. The two major depressions are the Fraser Plateau in
+south-central British Columbia, and the Liard Lowland in northern British
+Columbia. Between these two basins is found a mildly elevated, but extensive
+and deeply incised mountain group, known as the Skeena Mountains.
 
@@
 floor, ice moved along tortuous and intertwined pathways, sometimes following
-the major topographical troughs, and sometimes flowing across them \mref. In
-terms of
-numerical glacier modelling, it implies complex patterns of flow and stress
+the major topographical troughs, and sometimes flowing across them
+\citep{Davis.Mathews.1944, Kleman.etal.2010}. In terms of numerical glacier
+modelling, this topography implies complex patterns of flow and stress
 within the ice, and requires high-resolution, and thus, large computational
@@
 months, varies between around 10\,\degC\ along most of the Pacific Coast to
-above 40\,\degC\ in the polar interior (\cref{fig:plot-atm}b). There exist a
+above 40\,\degC\ in the polar interior (\cref{fig:plot-atm}b). There exists a
 partial correlation between cold regions and those that have a large
@@
 even reinforced by the difference in timing of the precipitation peak through
-the year (\cref{fig:plot-atm}d). Coastal regions experience most precipitation
+the year (\cref{fig:plot-atm}d). Coastal regions receive most precipitation
 during winter and fall, thus as snow. On the contrary, many inland regions
-experience dry winters, and most of the annual precipitation consequently falls
+experience dry winters, and most of the annual precipitation, there,
+consequently falls
 as rain during the ablation season. The climatic setting of the North American
@@
 Canadian Cordillera, and some of the adjacent land in north-western contiguous
-USA and Alaska (\cref{map-cordillera}). The southern margin is the most
+USA and Alaska (\cref{fig:map-cordillera}). The southern margin is the most
 extensively studied. It consists of different piedmont lobes that have formed
@@
 Okanagan and Purcell lobes. The latter once dammed the Clark Fork River to form
-Glacial Lake Missoula \mref. As the ice sheet retreated, the lake repeatedly
-and catastrophically drained to the Pacific Ocean \mref. These floods
-contributed to shape a distinctive landscape known as the Channeled Scablands.
+Glacial Lake Missoula \citep{Pardee.1910}. As the ice sheet retreated, the lake
+repeatedly and catastrophically drained to the Pacific Ocean
+\citep{Bretz.1923,Waitt.1980}. These floods
+contributed to shape a distinctive landscape known as the Channeled Scabland.
 

### 2014-07-08 kappa.tex
Develop glacial history.

@@
 % ----------------------------------------------------------------------
-\subsection{Palaeo-glaciology}
-% ----------------------------------------------------------------------
-
+\subsection{Glacial history}
+% ----------------------------------------------------------------------
+
+Earliest glaciofluvial deposits associated with the presence of an ice sheet in
+the Cordilleran mountains were dated at $2.64^{+0.20}/_{-0.18}$\,Ma
+\citep{Hidy.etal.2013}. This suggests that the North American Cordillera has
+been glaciated many times, with the first glaciations occuring even before the
+onset of the 100\,ka glacial cycle (\cref{fig:paleo-timeseries}). However, only
+the last few of these episodes can be reconstructed from the geologic record,
+whereas much of the older evidence has been overriden.
+
+In the northern sector of the ice sheet, at least four glaciations have been
+identified \mref. The two older episodes, often referred to as Reid and
+pre-Reid, are older than the last glacial cycle \mref, and thus beyond the
+scope of this study. The two younger glaciations, though, occurred both during
+the last glacial cycle \mref. Indeed, the Cordilleran ice sheet did not keep
+its maximum size during the entire glacial cycle. Instead, it is thought that,
+for most of the last glacial cycle, ice cover was restricted to some of the
+major mountain ranges of the Cordillera, while the maximum, ice-sheet stage,
+corresponded to one or more short-lived expansion cycles.
+
+The northern margin of the ice sheet suggests two such glacial episodes. The
+first expansion cycle, the Gladstone glaciation, occurred at about 55-60\,ka,
+during a local maximum of global ice volume on Earth known as the Marine
+Isotope Stage (MIS)~4 \mref. However, evidence for the Gladstone glaciation has
+not been found the southern part of the ice-covered region, making it difficult
+to seize its magnitude. The second expansion cycle, the McConnell glaciation,
+was dated at 18\,ka, which corresponds to the last global ice volume maximum
+(MIS~2). In opposite to the Gladstone glaciation, the McConnell glaciation has
+been correlated to a contemporaneous glacier advance of the southern margin,
+known as the Fraser glaciation, and dated at 17\,ka. Although a slight
+difference in timing may exists between the timing of maximum extent of the
+northern and southern ice margins, the The McConnell and Fraser glaciation
+correspond to the same expansion cycle, which resulted in the Last Glacial
+Maximum (LGM) extent of the Cordilleran ice sheet. This is the period where
+most geological observations point at.
+
+It is generally thought that the Cordilleran ice sheet built from the
+coalescence of several mountain-centred ice caps \citep{Davis.Mathews.1944}.
 During the last glacial maximum, the Cordilleran ice sheet covered most of the
@@
 
-Although the last glacial maximum extent of the Cordilleran ice sheet, apart of
-its eastern margin, is relatively well known, the history of deglaciation is
-heavily debated, while little information exists on earlier dynamics of ice
-sheet.
+The ice sheet is believed to have retreated quickly. Most of the evidence comes
+from the Fraser Plateau, where it indicates that some of the highest relief
+emerged from the ice prior to the lowland, possibly resulting in thin, stagnant
+ice remnants in the interior
+    \citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a}.
+Much less is known about ice sheet decay in the northern parts of the
+ice-covered areas, but glacial landforms from the Liard Lowland suggest a
+different mode of deglaciation, with active retreat of the ice lobe towards the
+surrounding mountain ranges \citep{Margold.etal.2013}. Patterns of
+deglaciation, as well as the possibility for a late-glacial re-advance, are
+discussed in further detail with consideration of modelling results in \CCYC.
 

### 2014-07-08 kappa.tex
Add references on glacial history.

@@
 In the northern sector of the ice sheet, at least four glaciations have been
-identified \mref. The two older episodes, often referred to as Reid and
-pre-Reid, are older than the last glacial cycle \mref, and thus beyond the
+identified
+    \citep{Duk-Rodkin.1999, Ward.etal.2007, Ward.etal.2008,
+           Briner.Kaufman.2008, Demuro.etal.2012,
+           Stroeven.etal.2010, Stroeven.etal.2014}.
+The two older episodes, often referred to as Reid
+and pre-Reid, are older than the last glacial cycle, and thus beyond the
 scope of this study. The two younger glaciations, though, occurred both during
-the last glacial cycle \mref. Indeed, the Cordilleran ice sheet did not keep
+the last glacial cycle. Indeed, the Cordilleran ice sheet did not keep
 its maximum size during the entire glacial cycle. Instead, it is thought that,
@@
 The northern margin of the ice sheet suggests two such glacial episodes. The
-first expansion cycle, the Gladstone glaciation, occurred at about 55-60\,ka,
+first expansion cycle, the Gladstone glaciation, reached its maximum between
+$50.4\pm3.0$ and $54.3\pm2.0$\,\unit{^{10}Be\,ka} \citep{Ward.etal.2007},
 during a local maximum of global ice volume on Earth known as the Marine
-Isotope Stage (MIS)~4 \mref. However, evidence for the Gladstone glaciation has
+Isotope Stage (MIS)~4. However, evidence for the Gladstone glaciation has
 not been found the southern part of the ice-covered region, making it difficult
 to seize its magnitude. The second expansion cycle, the McConnell glaciation,
-was dated at 18\,ka, which corresponds to the last global ice volume maximum
+built end moraines with oldest minimum apparent exposure ages of
+$15.7\pm1.5$ and $17.7\pm1.6$\,\unit{^{10}Be\,ka} \citep{Stroeven.etal.2014},
+which corresponds to the last global ice volume maximum
 (MIS~2). In opposite to the Gladstone glaciation, the McConnell glaciation has
 been correlated to a contemporaneous glacier advance of the southern margin,
-known as the Fraser glaciation, and dated at 17\,ka. Although a slight
-difference in timing may exists between the timing of maximum extent of the
-northern and southern ice margins, the The McConnell and Fraser glaciation
-correspond to the same expansion cycle, which resulted in the Last Glacial
-Maximum (LGM) extent of the Cordilleran ice sheet. This is the period where
-most geological observations point at.
+known as the Fraser glaciation, with a maximum extent framed between
+$17.4$ and $16.4$\,\unit{^{14}C\,cal\,ka} \citep[Fig.~4]{Porter.Swanson.1998}.
+Within errors inherent to the dating methods \citep{Heyman.etal.2011}, the
+McConnell and Fraser glaciation correspond to the same glacial episode, which
+resulted in the Last Glacial Maximum (LGM) extent of the Cordilleran ice sheet.
+This is the period where most geological observations point at.
 

### 2014-07-08 kappa.tex
Improve glacial history and add more refs.

--- 2014-07-08 kappa.tex
@@
 % ----------------------------------------------------------------------
-\subsection{Palaeo-glaciology}
-% ----------------------------------------------------------------------
-
-During the last glacial maximum, the Cordilleran ice sheet covered most of the
+\subsection{Glacial history}
+% ----------------------------------------------------------------------
+
+The first presence of an ice sheet in the Cordilleran mountains is indicated by
+glaciofluvial deposits dated at $2.64^{+0.20}/_{-0.18}$\,Ma
+\citep{Hidy.etal.2013}. This suggests that the North American Cordillera has
+been glaciated many times, with the first glaciations occurring even before the
+onset of the 100-ka glacial cycle (\cref{fig:paleo-timeseries}). However, only
+the last few glaciations can be reconstructed from the geologic record, whereas
+much of the older evidence has been overridden by subsequent glacial
+re-advances and other geomorphological processes.
+
+In the northern sector of the ice sheet, at least four glaciations have been
+identified
+    \citep{Duk-Rodkin.1999, Ward.etal.2007, Ward.etal.2008,
+           Briner.Kaufman.2008, Demuro.etal.2012,
+           Stroeven.etal.2010, Stroeven.etal.2014}.
+The two older ones, often collectively referred to as the Reid glaciations,
+pre-date the last glacial cycle, and thus are beyond the scope of this study.
+The two younger glaciations, though, occurred both during the last glacial
+cycle. First, the Gladstone glaciation reached its maximum between
+$50.4\pm3.0$ and $54.3\pm2.0$\,\unit{^{10}Be\,ka} \citep{Ward.etal.2007},
+during a period of high global ice volume on Earth, known as the Marine Isotope
+Stage (MIS)~4. However, evidence from the Gladstone glaciation has not been
+found in the southern part of the ice-covered region, making it difficult to
+seize its magnitude. Second, the McConnell glaciation is associated with end
+moraines of oldest minimum apparent exposure ages of
+$15.7\pm1.5$ and $17.7\pm1.6$\,\unit{^{10}Be\,ka} \citep{Stroeven.etal.2014},
+which corresponds to the last global ice volume maximum (MIS~2). In contrary to
+the Gladstone glaciation, the McConnell glaciation has been correlated with a
+contemporaneous advance of the southern ice sheet margin, known as the Fraser
+glaciation, with a maximum extent framed between
+$17.4$ and $16.4$\,\unit{^{14}C\,cal\,ka} \citep[Fig.~4]{Porter.Swanson.1998}.
+Within errors inherent to the dating methods \citep{Heyman.etal.2011}, the
+McConnell and Fraser glaciations correspond to the same glacial episode, which
+resulted in the Last Glacial Maximum (LGM) extent of the Cordilleran ice sheet.
+This is the period where most geological observations point at.
+
+It is generally thought that the Cordilleran ice sheet was short-lived
+    \citep[e.g.]{Clague.etal.1980, Clague.1985, Cosma.etal.2008}.
+The ice sheet originated from the
+coalescence of several mountain-centred ice caps \citep{Davis.Mathews.1944}.
+During the last glacial maximum, ice covered most of the
 Canadian Cordillera, and some of the adjacent land in north-western contiguous
-USA and Alaska (\cref{fig:map-cordillera}). The southern margin is the most
-extensively studied. It consists of different piedmont lobes that have formed
-at the outlet of the main valleys. The Puget lobe was the westernmost and
+USA and Alaska (\cref{fig:map-cordillera}).
+
+The southern margin of the ice sheet has been studied extensively, and is in
+several respects better understood than its northern counterpart
+\citep{Booth.etal.2003}. It consists of different piedmont
+lobes that have formed at the outlet of the main valleys. The Puget lobe
+\citep[e.g.]{Thorson.1980, Porter.Swanson.1998} was the westernmost and
 largest lobe. It covered the Puget Lowland, a region that currently includes
@@
 \citep{Bretz.1923,Waitt.1980}. These floods
-contributed to shape a distinctive landscape known as the Channeled Scabland.
+contributed to shape a distinctive landscape, known as the Channeled Scabland.
 
@@
 ice cap covering Vancouver Island was well-connected to main body of ice on the
-continent. A significant part of the ice sheet thus drained into the Georgia
-Strait and engulfed between Vancouver Island and the Olympic Peninsula before
-it formed the Juan the Fuca marine ice lobe, calving into the Pacific Ocean.
+continent \mref. A significant part of the ice sheet thus drained into the Georgia
+Strait and engulfed between Vancouver Island and the Olympic Peninsula where
+it formed the Juan the Fuca marine ice lobe, calving into the Pacific Ocean \mref.
 Further to the North, the ice margin reached Queen Charlotte Islands, yet the
 archipelago may not have been fully covered by ice
-flowing from the land. Submarine troughs likely idicate the erosional action of
+flowing from the land. Submarine troughs likely indicate the erosional action of
 former ice streams, yet it is not clear whether they have formed during the
-last glacial cycle or earlier.
-
-The northern margin of the ice sheet is in several ways more complex than the
-southern margin. Firstly, geomorphological evidence indicate multiple glacier
-advances. During the last glacial cycle, at least two glaciations occured, and
-in some places, the first glaciation was more extensive than the second.
-Secondly, geomorphological evidence indicate that the extent of glaciers in the
-North was relatively restricted, despite very cold temperatures at present.
+last glacial cycle or earlier \mref.
 
@@
 is known that the Cordilleran ice sheet collided with its Laurentide neighbour,
-the precise location of this junction is unknown, and little information exists
-on its timing. More importantly, the nature of the interaction between the two
-ice sheets is uncertain. It is not known for instance, if the junction lasted
-long enough for the ice sheets to significantly alter their respective flow
-dynamics. Because the present thesis focuses specifically on the Cordilleran
-ice sheet, our largest assumption certainly consists of neglecting buttressing
-effects from the Laurentide ice sheet.
-
-Although the last glacial maximum extent of the Cordilleran ice sheet, apart of
-its eastern margin, is relatively well known, the history of deglaciation is
-heavily debated, while little information exists on earlier dynamics of ice
-sheet.
+the nature of the interaction between the two ice sheets and the precise
+location of this junction is unclear, while little information exists on its
+timing
+    \citep[e.g.][]{Jackson.etal.1997, Bednarski.Smith.2007}.
+Because the
+present thesis focuses specifically on the Cordilleran ice sheet, our largest
+assumption certainly consists of neglecting buttressing effects from the
+Laurentide ice sheet.
+
+The ice sheet is believed to have retreated quickly. Most of the evidence comes
+from the Fraser Plateau, where it indicates that some of the highest relief
+emerged from the ice prior to the lowlands, possibly resulting in thin,
+stagnant ice remnants in the interior
+    \citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a}.
+Much less is known about ice sheet decay in the northern parts of the
+ice-covered areas, but glacial landforms from the Liard Lowland suggest a
+different mode of deglaciation, with active retreat of the ice lobe towards the
+surrounding mountain ranges \citep{Margold.etal.2013}. Patterns of
+deglaciation, as well as the possibility for a late-glacial re-advance, are
+discussed in further detail and compared with modelling results in \CCYC.
 

### 2014-07-08 kappa.tex
Add Vancouver Island wet climate photo.
Add fig, table and biblio refs in model set-up.

@@
 \usepackage{bm}
+\usepackage[T1]{fontenc}
 \usepackage{lineno}
@@
 \end{figure}
-
-% Alternative code using subcaption package
-%\usepackage{subcaption}
-%\begin{figure}
-%  \centering
-%  \makebox[0pt]{
-%    \begin{subfigure}[t]{80mm}
-%        \includegraphics{photo-glacier-crevasses}
-%        \caption{Subcaption a.}
-%        \label{fig:1a}
-%    \end{subfigure}
-%    \hspace{10mm}
-%    \begin{subfigure}[t]{80mm}
-%        \includegraphics{photo-glacier-fold}
-%        \caption{Subcaption b.}
-%        \label{fig:1b}
-%    \end{subfigure}
-%  }
-%  \caption{Global caption.}
-%  \label{fig:1}
-%\end{figure}
 
@@
   \centering
-  \includegraphics{photo-fraser-valley}
-  \caption{This dry landscape of the Fraser Valley, central British Columbia,
-           exhibits climatic contrasts at play in the Cordillera.}
+  \makebox[0pt]{
+    \subgraphics{a}{photo-vancouver-island}
+    \hspace{1cm}
+    \subgraphics{b}{photo-fraser-valley}
+  }
+  \caption{Contrasted climates of the North American Cordillera.
+           \textbf{(a)} Pacific Coast at Vancouver Island.
+           \textbf{(b)} Fraser Valley, south-central British Columbia.}
   \label{fig:photo-fraser-valley}
@@
 conservation of mass implies a conservation of volume, which can be expressed
-in terms of the ice velocity vector, $\vv$, by
+in terms of the ice velocity vector, $\vv$ \cref{fig:model-variables}, by
 \begin{equation}
@@
 where $p=\frac{1}{3}\tr(\CST)$ is the hydrostatic pressure and
-$\tens{\delta}$ is the identity tensor, a \emph{constitutive law} for ice based
-on laboratory experiments is used to relate these two quantities. It reads:
+$\tens{\delta}$ is the identity tensor, a \emph{constitutive law} for ice
+\citep{Nye.1953} based on laboratory experiments \citep{Glen.1952} is used to
+relate these two quantities. It reads:
 \begin{equation}
@@
 The ice softness coefficient, $A$, depends on ice temperature, pressure, and
-water content through an Arrhenius-type law,
+water content through an Arrhenius-type law
+\citep[Eqs.~63--65]{Aschwanden.etal.2012},
 \begin{equation}
@@
 Clapeyron relation, ${T_{pa} = T - \beta p}$. The water fraction, $\omega$, is
-capped at a threshold of 0.01, above which no measurements are available.
+capped at a threshold of 0.01, above which no measurements are available
+\citep{Lliboutry.Duval.1985}.
+
+\todo{I am trusting PISM's manual on that. Check other papers by Duval, or
+      the book by Greve (library).}
 
@@
 By analogy to Newtonian flow, an apparent viscosity, $\nu$, can then be defined
-by $\DST = \nu\,\SRT$, thus yielding
+by $\DST = 2 \nu\,\SRT$, thus yielding
 
@@
 \emph{conservation of energy}. In PISM, the conservation of energy uses an
-enthalpy formulation in order to account for thermodynamic effects associated
-with internal phase changes in the glacier. This enthalpy ($H$) formulation
-reads
+enthalpy formulation \citep[Eqs.~20--21]{Aschwanden.etal.2012} in order to
+account for thermodynamic effects associated with internal phase changes in the
+glacier. This enthalpy ($H$) formulation reads
 \begin{equation}
@@
 where $c$ denotes the specific heat capacity of the ice and $k$ is its thermal
-conductivity.
+conductivity. Constant parameters of the ice flow model use in this thesis are
+given in \cref{tab:iceflowparams}.
 
@@
            \todo{find a better reference than \citet{Aschwanden.etal.2012}
-                 for all the physical constants. Double-check
+                 (library) for all the physical constants. Double-check
                  \citet{Paterson.Budd.1982} and add missing references.}}
@@
 
-PISM, the ice sheet model that we use, employ two well-documented
-approximations of ice flow equations: the shallow ice approximation and the
-the shallow shelf approximation. Although both approximations can be derived
+PISM, the ice sheet model that we use, employs two well-documented
+approximations of ice flow equations: the shallow ice approximation
+    \citep[SIA,][]{Morland.Johnson.1980, Hutter.1983}
+and the the shallow shelf approximation
+    \citep[SSA,][]{Morland.1987, Weis.etal.1999}.
+Although both approximations can be derived
 from a rigorous scaling approach, we don't expand these lengthy derivations
@@
 
-Projecting along the three Cartesian coordinate axis, this reads
+Projecting along the three Cartesian axes, this reads
 \begin{subequations}
@@
 
-The shallow ice approximation (SIA) assumes that basal friction is high enough
+The \emph{shallow ice approximation} (SIA) assumes that basal friction is high
+enough
 for the vertical shear stresses $\{\tau_{xz}, \tau_{yz}\}$ to predominate over
@@
 
+\todo{\citet{Morland.Johnson.1980} derived the SIA in 2D, using a stream
+      function. Their equations look very different from here. Go to
+      geolibrary and check the book by \citet{Hutter.1983}, if it is there.}
+
 Neglecting atmospheric pressure, \Cref{eqn:siaz} yield
@@
 
-The shallow shelf approximation (SSA), on the opposite of the SIA, assumes that
+\todo{Add a few words about the fact that $\vsia$ can be directly obtained
+      after one vertical integration. Show that vertical velocities can be
+      obtained from the compressibility equation.}
+
+The \emph{shallow shelf approximation} (SSA), on the opposite of the SIA,
+assumes that
 basal friction is low enough for ice to deform predominantly by horizontal
 expansion within the entire ice column. In this context, the stress-balance can
-be simplified to
+be simplified to \citep[Eqs.~4.10--4.12]{Weis.etal.1999}
 \begin{subequations}
@@
 \end{subequations}
-and horizontal velocities can be assumed independent of depth, so that
+and horizontal velocities can be assumed independent of depth
+\citep[Eq.~4.22]{Weis.etal.1999}, so that
 \begin{align}
@@
 A final integration over the entire ice column yields
+\citep[Eqs.~7.7--7.8]{Weis.etal.1999}
 \begin{subequations}
@@
 
-In PISM, the SIA and SSA velocities are eventually added,
+In PISM, the SIA and SSA velocities are eventually added
+\citep[Eq.~15]{Winkelmann.etal.2011},
 \begin{equation}
@@
 reciprocally, we can expect the model to perform well in regions where
-conditions for applicability of either the SIA or the SSA are met. However,
+conditions for applicability of either the SIA or the SSA are met
+\cref{fig:model-siassa}. However,
 \cref{eqn:siassa} is a heuristic, and its validity in the zone of
 transition between shear flow and longitudinal flow has not yet been
-mathematically proven.
+mathematically proven
 
@@
 a sliding law as basal boundary condition, which in fact constitutes the main
-control on SSA velocities and in turn locations of ice streams.
+control on SSA velocities and, in turn, the locations of ice streams.
 
@@
 velocity, and $v_{th}$ is an arbitrary velocity threshold. In the case of
-$q=0$, \cref{eqn:pseudoplastic} corresponds to a purely plastic law,
+$q=0$, \cref{eqn:pseudoplastic} corresponds to a purely plastic law
+\citep[Eq.~11]{Winkelmann.etal.2011},
 \begin{equation}
@@
 
-However a non-zero exponent is chosen here in order to improve convergence. The
-yield stress $\tau_c$ is related to till properties by the Mohr-Coulomb
+\todo{\citep[supplement]{Aschwanden.etal.2013} refer to a paper by Schoof and
+      Hindmarsh (2010, doi:10.1093/qjmam/hbp025). Check that.}
+
+However a non-zero exponent $q$ is chosen here in order to improve convergence.
+The yield stress $\tau_c$ is related to till properties by the Mohr-Coulomb
 criterion,
@@
 PISM~0.5, the effective pressure is linearly related to water content by a
-simple parametrisation,
+simple parametrisation \citep[Eq.~13]{Winkelmann.etal.2011},
 \begin{equation}
@@
 In {\CCYC}, using PISM~0.6, the effective pressure is determined by an
-empirical parametrisation based on laboratory experiments on Antarctic till,
+empirical parametrisation based on laboratory experiments on Antarctic till
+\citep{Tulaczyk.etal.2000, Bueler.Pelt.2014},
 \begin{equation}
@@
 accumulation of basal melt-water. In varies from zero to $W_{max}$, a threshold
-above which further melt exits the system.
+above which further melt simply exits the system.
 
@@
 elevation, with lowest values occurring at low elevations, thus accounting for
-a weakening of till associated with the presence of marine sediments:
+a weakening of till associated with the presence of marine sediments
+\citep[cf.][Eq.~10]{Martin.etal.2011},
 \begin{equation}
@@
 where $b_0=0$\,m is modern sea-level and $b_1=200$\,m is a rough average of
-highest shoreline elevations in the model domain. The values of $\phi_1$ and
-$\phi_2$ are chosen differently in \CCLI (10 and 30$^\circ$) and \CCYC
-(15 and 45$^\circ$), but these choices are roughly equivalent in the case of a
-saturated till ($W=W_{max}$).
+highest shoreline elevations in the model domain. Parameter values for the
+sliding model are given in \cref{tab:slidingparams}.
 
@@
 flexure. Due to high mantle viscosity, $\nu_m$, there exists a time lag between
-the ice sheet growth and the isostatic bedrock response.
-%$\nu=1\times10^{21}$\,Pa\,s
-%$D=5\times10^{24}$\,N\,m
-%$\rho_r = 3300$\,kg\,m$^{-3}$
+the ice sheet growth and the isostatic bedrock response. Parameter values for
+bedrock response model are given in \cref{tab:bedrockparams}.
 
@@
   \centering
-  \caption{Constant parameters of the bedrock model.
+  \caption{Constant parameters of the bedrock thermal and deformation models.
            \todo{Find out what the bedrock density corresponds to. Track down
@@
 
-At the interface between the ice sheet and the atmosphere (Fig.~??),
+At the interface between the ice sheet and the atmosphere
+(\cref{fig:model-interfaces}),
 surfice mass fluxes are computed from monthly mean surface air temperature,
@@
 In the simulations presented here, \cref{eqn:calovgreve} is numerically
-approximated using week-long sub-intervals. For each sub-interval, surface melt
-is computed from PDDs and the available snow and ice depths using an algorithm
-documented in the on-line documentation of PISM \citep{PISM-authors.2014}. This
-algorythms employ different melt factors for snow ($F_s$) and for ice ($F_i$),
+approximated using week-long sub-intervals. For each sub-interval, PDDs, which
+serve as a proxy for the available energy for melt, define the amount of snow
+melt through a proportionality factor $F_s$. If all snow is melted, remaining
+PDDs are use to melt ice, using a proportionality factor $F_i$
+(\cref{fig:model-pdd}). The melt factors $F_s$ and $F_i$
+(\cref{tab:surfaceparams})
 as derived from mass-balance measurements on contemporary glaciers in the
@@
 a method initially developed in {\PSDV}, and further improved in {\PSDP}.
-%3.04\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for snow and
-%4.59\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for ice
 
@@
 
-\CCLI tests the sensitivity of the model to different input present-day
+\CCLI\ tests the sensitivity of the model to different input present-day
 climatologies, $\{T_{m0}, P_{m0}\}$, while the temperature-offset corrections,
@@
 on the opposite, tests model sensitivity to time-dependent temperature-offset
-corrections derived from different palaeo-temperature proxy records.
+corrections derived from different palaeo-temperature proxy records. All data
+climatic and topographic data sources used are summarised in
+\cref{tab:datasources}.
 
@@
 % ----------------------------------------------------------------------
+
+\todo{Add here a short section about discretization of the SIA and SSA, and
+      their numerical implementation. The main subtlety, I think, lies in the
+      fact that some computation need to use a ``staggered'' grid in order to
+      ensure that we add SIA and SSA values corresponding to the same point in
+      space. I need to check that. This section is also the place where to
+      introduce geographic transformation parameters that I used throughout
+      the thesis -- except in \cref{fig:map-cordillera}!}
 

### 2014-07-09 kappa.tex
Re-update figure captions.

@@
   }
-  \caption{Illustration of various glacier ice behaviours.
-           \textbf{(a)} Surface crevasses formed by ice flow over a hidden
-           bedrock knob. Glacier Blanc, French Alps, 2004.
-           \textbf{(b)} Medial moraines illustrating downhill ice flow on
+  \caption{Illustration of various glacier ice behaviours:
+           \textbf{(a)} surface crevasses formed by ice flow over a hidden
+           bedrock knob, Glacier Blanc, French Alps, 2004;
+           \textbf{(b)} medial moraines illustrating downhill ice flow on
            Susitna Glacier, Alaska Range, by Austin Post, 1970
-           \citep{NSIDC.2009}.
-           \textbf{(c)} Mountain glacier calving into a small lake, revealing a
-           folded structure formed by deformed debris layers in the ice.
+           \citep{NSIDC.2009};
+           \textbf{(c)} mountain glacier calving into a small lake, revealing a
+           folded structure formed by deformed debris layers in the ice,
            Glacier d'Arsine, French Alps, 2004.}
@@
   }
-  \caption{Different indicators of past glaciation.
-           \textbf{(a)} Glacial valley, exhibiting a broad, U-shaped profile
-           characteristic of erosion by glaciers.
-           Cirque du Fer-\`{a}-Cheval, French Alps, 2013.
-           \textbf{(b)} Erratic boulders, deposited in an improbable
-           arrangement as the glacier melted underneath.
-           Finse area, Norwegian Scandes, 2011.
-           \textbf{(c)} Lateral meltwater channels, formed on a gentle slope
-           as the ice margin retreated towards the valley bottom. Tuya Lake
+  \caption{Different indicators of past glaciation:
+           \textbf{(a)} glacial valley, exhibiting a broad, U-shaped profile
+           characteristic of erosion by glaciers,
+           Cirque du Fer-\`{a}-Cheval, French Alps, 2013;
+           \textbf{(b)} erratic boulders, deposited in an improbable
+           configuration as the glacier melted underneath,
+           Finse area, Norwegian Scandes, 2011;
+           \textbf{(c)} lateral meltwater channels, formed on a gentle slope
+           as the ice margin retreated towards the valley bottom, Tuya Lake
            area, Canadian Cordillera, 2012.}
@@
            \textbf{(b)} during the last glacial maximum
-           \citep{Ehlers.Gibbard.2007}.}
+           \citep{Ehlers.Gibbard.2007}.
+           \todo{Make these globes look a little bigger. Add (a) and (b).}}
   \label{fig:paleo-glaciation}
@@
            \citep{Kaufman.Manley.2004}, indicating dry climatic conditions.
-           The Alaska Range is rapidly eroding, as shown by landslide-covered
-           valley sides on this photograph.}
+           The Alaska Range is tectonically active and rapidly eroding, as
+           shown by landslide-covered valley sides on this photograph.}
   \label{fig:photo-alaska-range}
@@
   }
-  \caption{Contrasted climates of the North American Cordillera.
-           \textbf{(a)} Pacific Coast at Vancouver Island.
-           \textbf{(b)} Fraser Valley, south-central British Columbia.}
+  \caption{Contrasted climates of the North American Cordillera:
+           \textbf{(a)} Pacific Coast at Vancouver Island, western British
+           Columbia, by Martin Margold, 2010.
+           \textbf{(b)} Fraser Valley, south-central British Columbia, 2010.}
   \label{fig:photo-fraser-valley}
@@
            North American Regional Reanalysis
-           \citep[NARR,][]{Mesinger.etal.2006}}
+           \citep[NARR,][]{Mesinger.etal.2006}.}
   \label{fig:plot-atm}
@@
   \caption{Model interfaces in PISM. From \citet{PISM-authors.2014}.
-           \todo{Redraw.}}
+           \todo{This is directly from the PISM manual. I may redraw it here
+                 if I have the time.}}
   \label{fig:model-interfaces}
@@
   \caption{Sketch of notations used for the main variables of the ice sheet
-           model.}
+           model. A particle of ice is moving at velocity $\vv$, under the
+           balance of stresses $\CST$ and gravity $\vect{g}$.}
   \label{fig:model-variables}
@@
            (SSA) velocities in PISM. From \citet{Winkelmann.etal.2011}.
-           \todo{Redraw.}}
+           \todo{Redraw in LaTeX in the same style as
+                 \cref{fig:model-variables}.}}
   \label{fig:model-siassa}
@@
   \centering
-  \includegraphics[width=80mm]{model-pdd}
-  \caption{PDD model. From \citet{PISM-authors.2014} \todo{Redraw.}}
+  \includegraphics[width=\linewidth]{model-pdd}
+  \caption{PDD model. From \citet{PISM-authors.2014}
+           \todo{This is again directly from the PISM manual. I can simplify
+                 it by removing the unused refreezing branch, I will match
+                 notations with the text and refer to equations in use.}}
   \label{fig:model-pdd}
@@
 
-\begin{figure}
-  \centering
-  \makebox[0pt]{\includegraphics{photo-tuya-moraine}}
-  \caption{The Tuya Lake Moraine in northern British Columbia.}
-  \label{fig:photo-tuya-moraine}
-\end{figure}
+%\begin{figure}
+%  \centering
+%  \makebox[0pt]{\includegraphics{photo-tuya-moraine}}
+%  \caption{The Tuya Lake Moraine in northern British Columbia.}
+%  \label{fig:photo-tuya-moraine}
+%\end{figure}
 

### 2014-07-09 kappa.tex
Final worhty check before first milestone.

@@
 
-As ice flows and glides on its bed, it transports rock debris and erode the
+As ice flows and glides on its bed, it transports rock debris and erodes the
 landscape, thence leaving geomorphological traces of its former presence
@@
 However, the glacial theory did not gain much belief until the exploration and
-discovery of the two present-day ice sheets on Earth, the Greenland and
+discovery of the two present-day ice sheets on Earth: the Greenland and
 Antarctic ice sheets, provided a modern analogue for the proposed European and
-North American ice sheets. The first traverse of the Greenland was performed by
+North American ice sheets. The first traverse of Greenland was performed by
 Nansen in 1888, while Amundsen reached the south pole in 1911. Meanwhile, in
@@
 100\,kyr periodicity \citep{Hays.etal.1976}. In the present thesis, we focus on
-the most recent of these cycles: \emph{the last glacial cycle}
+the most recent of these periods: \emph{the last glacial cycle}
 (\cref{fig:paleo-timeseries})
@@
 \citep{Bamber.etal.2001, Fretwell.etal.2013}. At the apogee of the last glacial
-cycle, the last glacial maximum, sea-level was 120 to 135\,m lower than today
+cycle, the \emph{Last Glacial Maximum} (LGM), sea-level was 120 to 135\,m lower
+than today
 \citep{Clark.Mix.2002}, implying an additional, double amount of ice stored in
@@
 the Scandinavian and Barents-Kara ice sheets in the Eurasian arctic, and the
-Innuitian, Laurentide and \emph{Cordilleran} ice sheets over North America
-(\cref{fig:paleo-glaciation}).
+Innuitian and Laurentide ice sheets and, finally, the \emph{Cordilleran ice
+sheet}, over North America (\cref{fig:paleo-glaciation}).
 
@@
 they can be calibrated against geological data where they exist, while
-allowing a physics-based, quantitative reconstruction where they do not.
+allowing a physics-based, quantitative reconstruction where they do not
+\citep[e.g.,][]{Marshall.etal.2002, Tarasov.Peltier.2004}.
 
@@
 Basin. The Innuitian ice sheet, much smaller, covered the northernmost Canadian
-archipelago. Finally, the Cordilleran ice sheet capped the mountain ranges of
-the Western Cordillera, which forms our region of focus in the present thesis.
+archipelago. Finally, \emph{the Cordilleran ice sheet} capped the mountain
+ranges of the Western Cordillera, which forms our region of focus in the
+present thesis.
 
@@
 It was formed during multiple accretion phases associated with the closure of
-several back-arc oceanic trenches over a 200\,Ma-long geological history
+several back-arc oceanic trenches, over a 200\,Ma-long geological history
 \citep{Sigloch.Mihalynuk.2013}.
@@
 (\cref{fig:map-cordillera}). The western ribbon is formed by the Pacific Coast
-orogenic belt. Major glaciated reliefs include,
+orogenic belt \mref. Major glaciated reliefs include,
 from North to South, the Alaska Range, the Wrangell and Saint-Elias Mountains,
@@
 modelling, this topography implies complex patterns of flow and stress
-within the ice, and requires high-resolution, and thus, large computational
+within the ice and requires high-resolution, and thus, large computational
 grids.
@@
 clearly visible in the landscape (\cref{fig:photo-fraser-valley}). Due to the
-large latitudinal extent of the region, mean annual temperatures vary
+large latitudinal extent of the region, \emph{mean annual temperatures} vary
 significantly from north to south (\cref{fig:plot-atm}a). About half of the
-model domain has mean annual temperature below freezing. However, seasonality,
+model domain has mean annual temperature below freezing. However,
+\emph{seasonality},
 here computed as the temperature difference between the warmest and the coldest
@@
 
-Precipitation rates are highest on the western slopes of the Pacific Coast
-ranges and decline quickly on the other side of the mountains
+Rates of \emph{precipitation} are highest on the western slopes of the Pacific
+Coast
+ranges and decline quickly on the eastern side of the mountains
 (\cref{fig:plot-atm}c). In terms of ice sheet mass balance, this contrast is
-even reinforced by the difference in timing of the precipitation peak through
+even reinforced by the difference in timing of the \emph{precipitation peak}
+through
 the year (\cref{fig:plot-atm}d). Coastal regions receive most precipitation
 during winter and fall, thus as snow. On the contrary, many inland regions
-experience dry winters, and most of the annual precipitation, there,
-consequently falls
+experience dry winters, and most of the annual precipitation,
+consequently falls there
 as rain during the ablation season. The climatic setting of the North American
@@
 
-In the northern sector of the ice sheet, at least four glaciations have been
-identified
+In the \emph{northern sector} of the ice sheet, at least four glaciations have
+been identified
     \citep{Duk-Rodkin.1999, Ward.etal.2007, Ward.etal.2008,
@@
 
-The southern margin of the ice sheet has been studied extensively, and is in
-several respects better understood than its northern counterpart
+The \emph{southern margin} of the ice sheet has been studied extensively, and
+is in several respects better understood than its northern counterpart
 \citep{Booth.etal.2003}. It consists of different piedmont
@@
 
-The western margin of the ice sheet covered part of the continental shelf. An
+The \emph{western margin} of the ice sheet covered part of the continental
+shelf. An
 ice cap covering Vancouver Island was well-connected to main body of ice on the
@@
 
-Finally, the eastern margin of the ice sheet is the most unclear. Although it
+Finally, the \emph{eastern margin} of the ice sheet is the most unclear.
+Although it
 is known that the Cordilleran ice sheet collided with its Laurentide neighbour,
@@
 ice-covered areas, but glacial landforms from the Liard Lowland suggest a
-different mode of deglaciation, with active retreat of the ice lobe towards the
-surrounding mountain ranges \citep{Margold.etal.2013}. Patterns of
+different mode of \emph{deglaciation}, with active retreat of the ice lobe
+towards the surrounding mountain ranges \citep{Margold.etal.2013}. Patterns of
 deglaciation, as well as the possibility for a late-glacial re-advance, are
@@
 conservation of mass implies a conservation of volume, which can be expressed
-in terms of the ice velocity vector, $\vv$ \cref{fig:model-variables}, by
+in terms of the ice velocity vector, $\vv$ (\cref{fig:model-variables}), by
 \begin{equation}
@@
 By analogy to Newtonian flow, an apparent viscosity, $\nu$, can then be defined
-by $\DST = 2 \nu\,\SRT$, thus yielding
+by $\DST = 2 \nu\SRT$, thus yielding
 
@@
 where $c$ denotes the specific heat capacity of the ice and $k$ is its thermal
-conductivity. Constant parameters of the ice flow model use in this thesis are
+conductivity. Parameter values for the ice flow model used in this thesis are
 given in \cref{tab:iceflowparams}.
@@
 PISM, the ice sheet model that we use, employs two well-documented
-approximations of ice flow equations: the shallow ice approximation
-    \citep[SIA,][]{Morland.Johnson.1980, Hutter.1983}
-and the the shallow shelf approximation
-    \citep[SSA,][]{Morland.1987, Weis.etal.1999}.
-Although both approximations can be derived
-from a rigorous scaling approach, we don't expand these lengthy derivations
+approximations of ice flow equations: the shallow ice approximation (SIA)
+and the the shallow shelf approximation (SSA). Although both approximations can
+be derived from a rigorous scaling approach
+    \citep[SIA,][]{Morland.Johnson.1980, Hutter.1983,
+                   Morland.1987, Weis.etal.1999},
+we don't expand these lengthy derivations
 here, assuming instead the simplifications and focusing on their effect.
@@
 
-\todo{Add a few words about the fact that $\vsia$ can be directly obtained
+\todo{Add a few words to emphasize that $\vsia$ can be directly obtained
       after one vertical integration. Show that vertical velocities can be
@@
 
+\todo{Add a few words to emphasize that $\vssa$ can not be obtained in a
+      direct manner, requiring to solve a linear system.}
+
 In PISM, the SIA and SSA velocities are eventually added
@@
 conditions for applicability of either the SIA or the SSA are met
-\cref{fig:model-siassa}. However,
+(\cref{fig:model-siassa}). However,
 \cref{eqn:siassa} is a heuristic, and its validity in the zone of
-transition between shear flow and longitudinal flow has not yet been
-mathematically proven
+transition between shear flow and longitudinal flow has not been
+mathematically proven.
 
@@
 
-The simulations presented in this thesis use a pseudo-plastic sliding law,
+The simulations presented in this thesis use a \emph{pseudo-plastic sliding
+law},
 \begin{equation}
@@
 The modelled ice sheet interacts with the underlying bedrock in two ways.
-Firstly, bedrock temperatures are computed to a given depth by appending the
+Firstly, \emph{bedrock temperatures} are computed to a given depth by
+appending the
 ice enthalpy model (\ref{eqn:enthalpy}) with an underlying bedrock thermal
@@
 
-Secondly, bedrock elevation evolve in response to the ice load. The bedrock
-deformation model used in this thesis describes the flexure of an elastic
+Secondly, bedrock elevation evolve in response to the ice load. The
+\emph{bedrock deformation} model used in this thesis describes the flexure of
+an elastic
 lithosphere on top of an infinite half-space viscous mantle
-\citep{Lingle.Clark.1985}. It can be described by a single differential
+\citep{Lingle.Clark.1985}. It can be described by a single pseudo-differential
 equation of the bed elevation, $b$,
 \begin{equation}
+    \label{eqn:beddef}
     2\nu_m\,|\grad|\,\pdv{b}{t} + \rho_l g b + D\,\Delta^2 b = \sigma_{zzb} \,,
@@
 lithosphere's flexural rigidity, and $\sigma_{zzb}$ corresponds to the ice load
-\citep{Bueler.etal.2007}. Here, $\Delta^2$ designs the biharmonic (Laplacian
-square) operator of
-linear elastic theory, while $|\grad|$ is a pseudo-differential operator
-defined through the Fourier transform by \citet[Eq.~6]{Bueler.etal.2007}. In
-the left-hand part of this equation, the first term accounts for mantle
+\citep{Bueler.etal.2007}. Here, ${\Delta^2 = (\div\grad{})^2}$ designs the
+biharmonic (Laplacian square) operator of
+linear elastic theory, while $|\grad{}|$ is a pseudo-differential operator
+defined through its Fourier transform by \citet[Eq.~6]{Bueler.etal.2007}. On
+the left-hand side of \cref{eqn:beddef}, the first term accounts for mantle
 relaxation, the second for point-wise isostasy, and the third for elastic
@@
 (\cref{fig:model-interfaces}),
-surfice mass fluxes are computed from monthly mean surface air temperature,
+surface mass fluxes are computed from monthly mean surface air temperature,
 $T_m$, and monthly precipitation, $P_m$, by a temperature-index model
@@
 melt through a proportionality factor $F_s$. If all snow is melted, remaining
-PDDs are use to melt ice, using a proportionality factor $F_i$
+PDDs are used to melt ice, using a proportionality factor $F_i$
 (\cref{fig:model-pdd}). The melt factors $F_s$ and $F_i$
@@
 \end{subequations}
-
-\todo{Figure. Boundary models data flow chart}
 
@@
 on the opposite, tests model sensitivity to time-dependent temperature-offset
-corrections derived from different palaeo-temperature proxy records. All data
+corrections derived from different palaeo-temperature proxy records. All
 climatic and topographic data sources used are summarised in

### 2014-07-09 cordillera-cycle.tex
Corrections from Arjen in introduction.

@@
 % Title
-\title{Numerical simulation of the Cordilleran ice sheet
+\title{Numerical simulations of the Cordilleran ice sheet
        through the last glacial cycle}
@@
 
-During most of the the last glacial cycle, glaciers and ice caps of the North
-American Cordillera were more extensive than today. At the Last Glacial
+During the last glacial cycle, glaciers and ice caps of the North American
+Cordillera have been more extensive than today. At the Last Glacial
 Maximum (LGM), a continuous blanket of ice, the Cordilleran ice sheet
 \citep{Dawson.1888}, extended from the Alaska Range in the north to the
-North Cascades in the south. It calved into the Pacific Ocean in the west, and
-merged with the Laurentide ice sheet in the east (Fig.~\ref{fig:locmap}).
+North Cascades in the south. In addition, it extended offshore, where it calved
+into the Pacific Ocean, and merged with the western margin of its much larger
+neighbour, the Laurentide ice sheet, east of the Rocky Mountains
+(Fig.~\ref{fig:locmap}).
 
-For more than a century, exploration and geological investigations of the
+More than a century of exploration and geological investigations of the
 Cordilleran mountains have led to many observations in support of the former
@@
     \citep{Carrara.etal.1996},
-the south-western and north-eastern margins provide key constraints that
-allow reasonable reconstructions of the maximum ice extent
+moraines that demarcate the south-western and north-eastern margins provide key
+constraints that allow reasonable reconstructions of maximum ice sheet extents
     (\citealp{Prest.etal.1968}; \citealp[Fig. 1.12]{Clague.1989};
@@
      \citealp{Booth.etal.2003}; \citealp{Dyke.2004}).
-The timing of the last glaciation, reconstructed from radiocarbon dating
+The LGM Cordilleran ice sheet maximum extent was short-lived, as indicated by
+field evidence from radiocarbon dating
     \citep{Clague.etal.1980, Clague.1985, Clague.1986, Porter.Swanson.1998,
@@
     \citep{Stroeven.etal.2010, Stroeven.etal.2014, Margold.etal.inpress},
-bedrock deformation in response to the ice load
+bedrock deformation in response to former ice loads
     \citep{Clague.James.2002, Clague.etal.2005},
-and the offshore sedimentary record
-    \citep{Cosma.etal.2008, Davies.etal.2011}
-indicate that the LGM ice-sheet was short-lived. However, former ice thickness
-and, therefore, the ice sheet's contribution to the last glacial maximum sea
-level low stand
+and offshore sedimentary records
+    \citep{Cosma.etal.2008, Davies.etal.2011}.
+However, former ice thicknesses and, therefore, the ice sheet's contribution to
+the LGM sea level low stand
     \citep{Carlson.Clark.2012, Clark.Mix.2002}
@@
 Our understanding of the Cordilleran glaciation history prior to the LGM is
-even more fragmentary.
-In parts of the Yukon Territory and Alaska, dated glacial erratics indicate
-an extensive Marine Isotope Stage (MIS)~4 glaciation
+even more fragmentary
+    \citep{Barendregt.Irving.1998, Kleman.etal.2010, Rutter.etal.2012},
+although it is clear that maximum glaciation of the Cordilleran ice sheet
+predates the last glacial cycle
+    \citep{Hidy.etal.2013}.
+In parts of the Yukon Territory and Alaska, the distribution of tills
+    \citep{Turner.etal.2013}
+and dated glacial erratics indicate an extensive Marine Oxygen Isotope Stage
+(MIS)~4 glaciation
     \citep{Ward.etal.2007, Ward.etal.2008, Briner.Kaufman.2008,
            Stroeven.etal.2010, Stroeven.etal.2014},
-yet it is not known whether other regions were affected. Lineation patterns in
-the ice sheet interior include flow directions likely older than the LGM
+yet it is not known whether other regions in the study area were affected.
+Landforms in the interior regions include flow sets that are likely
+older than the LGM
     \citep[Fig.~2]{Kleman.etal.2010},
-but their absolute age is uncertain.
+but their absolute age remains uncertain.
 
-In contrast, our understanding of the deglaciation history of the Cordilleran
+In contrast, evidence for the deglaciation history of the Cordilleran
 ice sheet since the LGM is considerable, albeit mostly at a regional scale.
-Geomorphological evidence from south-central British Columbia suggests a rapid
-deglaciation, including early emergence of the high-elevated areas and
-stagnation of thin ice cover in the lowlands
-    \citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a},
-yet this model may not apply in all areas of the Cordilleran ice sheet
+Geomorphological evidence from south-central British Columbia indicates a rapid
+deglaciation, including an early emergence of elevated areas while thin
+stagnant ice still covered surrounding lowlands
+    \citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a}.
+This model, although credible, may not apply in all areas of the Cordilleran
+ice sheet
     \citep{Margold.etal.2013}.
-Finally, disparate clues for a late-glacial re-advance have been found in the
-Coast, Columbia and Rocky Mountains
+Although solid evidence for late-glacial glacier re-advances have been found in
+the Coast, Columbia and Rocky mountains
     \citep{Clague.etal.1997, Friele.Clague.2002, Friele.Clague.2002a,
@@
            Menounos.etal.2008},
-but the evidence is much more sparse than that of regions surrounding the North
-Atlantic \aref, thus leaving space for debate around the possibility of a
-regional late-glacial cold reversal.
+it appears to be more sparse than for formerly glaciated regions surrounding
+the North Atlantic \aref, thus allowing for considerable uncertainty concerning
+the possibility of a regional late glacial cold reversal.
 
-In general, the topographic complexity of the Cordillera has not allowed for
-ice-sheet wide reconstructions of glacial advance and retreat patterns such as
-those available for the Fennoscandian and Laurentide and ice sheets
+In general, the topographic complexity of the North American Cordillera and its
+effect on glacial history have inhibited the construction of ice sheet-wide
+glacial advance and retreat patterns such as
+those available for the Fennoscandian and Laurentide ice sheets
      \citep{Boulton.etal.2001, Dyke.Prest.1987, Dyke.etal.2003,
@@
 calibrated against field-based evidence, to perform a quantitative
-reconstruction of the ice sheet history through the last glacial cycle, and
+reconstruction of the Cordilleran ice sheet history through the last glacial
+cycle, and
 analyse some of the long-standing questions related to its evolution:
@@
     last glacial maximum?
-  \item Which were the primary dispersal centres? Do they reflect a stable or
-    ephemeral configuration?
+  \item Which were its primary dispersal centres? Do they reflect stable or
+    ephemeral configurations?
   \item What was the scale of glaciation prior to the last glacial maximum?
-  \item How rapid was the deglaciation? Did it include stand-stills or
-    re-advances?
+  \item How rapid was the last deglaciation? Did it include late glacial
+    stand-stills or re-advances?
 \end{itemize}
@@
 Although numerical ice sheet modelling has been established as a useful tool to
-improve our understanding of the Cordilleran ice sheet more than twenty years
-ago (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991},
+improve our understanding of the Cordilleran ice sheet
+    (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991},
      \citealp{Marshall.etal.2000})
@@
 high-resolution numerical modelling of palaeo-glaciers on mountainous terrain
-over millenial time scales \citep{Golledge.etal.2012}. Second, the complex
+over millenial time scales \citep[e.g.,][]{Golledge.etal.2012}. Second, the
+complex
 topography of the North American Cordillera also induces strong gradients in
@@
 considerable uncertainty, our palaeoclimate forcing is a simplistic approach
-including time-evolved temperature and precipitation fields derived from a
-combination of a present-day atmospheric reanalysis \citep{Mesinger.etal.2006},
-lapse-rate corrections,
+including temperature and precipitation fields derived from a
+present-day atmospheric reanalysis \citep{Mesinger.etal.2006,
+Seguinot.etal.2014} supplemented by a lapse-rate correction
 and temperature offset time series. The latter are obtained by scaling six
@@
 records from Antarctic ice cores \citep{Petit.etal.1999,Jouzel.etal.2007},
-and two alkenone unsaturation index records from Northwest Pacific oceanic
+and two alkenone unsaturation index records from Northwest Pacific ocean
 sediment cores \citep{Herbert.etal.2001}. Model output is compared to
-geomorphological evidence in terms of timing and extent of glaciation and
+geomorphological evidence in terms of the timing and extent of glaciation and
 patterns of deglaciation.

### 2014-07-11 kappa.tex
Remove lineno, correct biblio and add DOI links.
Add figures in results summary.

@@
 \usepackage[T1]{fontenc}
-\usepackage{lineno}
 \usepackage{natbib}
@@
 \usepackage[sort]{cleveref}  % needs to be loaded after hyperref
+\usepackage{doi}  % needs to be loaded after hyperref
 \usepackage{graphicx}
@@
 \maketitle
-\linenumbers
 \tableofcontents
@@
 % ----------------------------------------------------------------------
-\todo{Figure. 3km last glacial maximum poster run?}
+
+\begin{figure*}
+  \centering
+  \vspace{-16mm}
+  \makebox[0pt]{
+  \begin{minipage}[t]{112.5mm}\vspace{0pt}
+    \includegraphics{plot-lgm}
+  \end{minipage}
+  \hspace{1cm}
+  \begin{minipage}[t]{47.5mm}\vspace{0pt}
+    \caption{Constant-climate simulation of the LGM Cordilleran ice sheet using
+             an horizontal resolution of 3\,km. Modelled basal topography,
+             cold-based areas (hatches), surface topography (100\,m contours)
+             and surface velocities. The simulation is forced by atmospheric
+             fields from the NARR \citep{Mesinger.etal.2006} and a temperature
+             depression of 7\,\degC, using the set-up described in \CCLI. There
+             are several issues with this run. First, all east-facing margins
+             are too extensive as compared to available reconstructions.
+             Second, marine advance on the Pacific Shelf is too restrictive,
+             most notably in the south-western corner where the Puget and Juan
+             de Fuca lobes are missing. Third, en-glacial temperatures are
+             anomalously low due to brutal application of cold climate
+             condition to ice-free bedrock at the inception of the run.
+             \todo{Add black frame, and maybe the LGM outline from Dyke?}}
+  \end{minipage}
+  }
+  \vspace{-16mm}
+  \label{fig:plot-lgm}
+\end{figure*}
 
@@
 % ----------------------------------------------------------------------
-\todo{Figure. Sigma distribution map}\\
-\todo{Figure. Sigma effect on Cordillera runs}
+
+\begin{figure}
+  \centering
+  \makebox[0pt]{\includegraphics{plot-sdmap}}
+  \caption{July, daily temperature variability in the Arctic, as shown by
+           monthly standard deviation of near-surface daily mean air
+           temperature computed from ERA-40 reanalysis
+           \citep{Uppala.etal.2005}. Note the strong differences between
+           continental and maritime regions, notably within the model domain
+           (black rectangle). Although daily variability is lowest in summer,
+           this is when it most importantly contributes to surface melt.}
+  \label{fig:plot-sdmap}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \makebox[0pt]{\includegraphics{plot-sdeffect}}
+  \caption{Effect of temperature variability on Cordilleran ice sheet
+           simulations. Modelled basal topography and surface topography
+           (250\,m contours). The simulations use atmospheric forcing from the
+           NARR \citep{Mesinger.etal.2006}, a temperature depression of
+           7\,\degC, and different treatments of temperature variability
+           including
+           \textbf{(a)} no daily variability, $\sigma=0$;
+           \textbf{(b)} a constant, mean summer (JJA), model domain-averaged 
+           standard deviation value, $\sigma=3.09$\,K;
+           \textbf{(c)} monthly standard deviation grids derived from the NARR
+           and
+           \textbf{(d)} monthly grids for ice-free terrain and the
+           Greenland ice sheet, ERA-40-based \citep{Uppala.etal.2005}
+           parametrisation in \PSDP, ${\sigma=-0.15\cdot{T_m}+1.66}$.
+           Differences in surface elevation are displayed relative to case (c).
+           Note the different colour scales.}
+  \label{fig:plot-sdeffect}
+\end{figure}
 
@@
 % ----------------------------------------------------------------------
-\todo{Figure. Deglaciation streamplot}\\
-\todo{Figure. Cumulative ice displacement map}
+
+\begin{figure}
+  \centering
+  \makebox[0pt]{\includegraphics{plot-snapshots}}
+  \caption{\textbf{(a)} Modelled growth and decay of the last Cordilleran ice
+           sheet. Basal topography surface topography (1\,km contours) and
+           surface velocity (m/a).
+           \textbf{(b)} Evolution of total ice volume during the entire last
+           glacial cycle, expressed as meters of sea-level equivalent. The
+           simulation uses a horizontal resolution of 6\,km and is forced by
+           atmospheric fields from the NARR \citep{Mesinger.etal.2006} and
+           temperature offset time-series derived from the GRIP
+           \citep{Dansgaard.etal.1993} ice-core record as described in \CCYC.
+           \todo{X-tick snapshot times on time series plot and thicken the
+                 vertical lines. Add (a) and (b) subfigure labels. Link lower
+                 and upper panels; see recipe at
+                 \url{http://stackoverflow.com/questions/17543359/}.}}
+  \label{fig:plot-snapshots}
+\end{figure}
 

### 2014-07-12 figures/plot-snapshots.py
Enhance snapshots plot.

@@ -9,6 +9,9 @@ import numpy as np
 from netCDF4 import Dataset
 from matplotlib import pyplot as plt
 from matplotlib.colors import BoundaryNorm, LogNorm, Normalize
+from matplotlib.gridspec import GridSpec
+from matplotlib.transforms import blended_transform_factory
+from matplotlib.patches import Polygon
 from mpl_toolkits.axes_grid1.axes_grid import ImageGrid
 from iceplot import plot as iplt
 
@@ -26,11 +29,6 @@ def bounded_argmax(a, coord, bmin, bmax):
 run_path = '/home/julien/pism/output/cordillera-narr-6km-bl/' \
            'grip3222cool580+ccyc+till1545/y0120000'
 
-# initialize figure
-figw, figh = 170.0, 70.0
-fig = plt.figure(0, (figw/25.4, figh/25.4))
-ax = fig.add_axes([2.5/figw, 5.0/figh, 155.0/figw, 30.0/figh])
-
 # read output time series
 print 'reading time series...'
 nc = Dataset(run_path + '-ts.nc')
@@ -39,18 +37,35 @@ ivol = nc.variables['slvol'][:]
 nc.close()
 
 # locate snapshot times
-tkalist = np.array([
+idxlist = [
     bounded_argmin(ivol, time, -45, -20),
     bounded_argmin(ivol, time, -35, -30),
     bounded_argmin(ivol, time, -30, -25),
     bounded_argmin(ivol, time, -26, -20),
     bounded_argmax(ivol, time, -20, -18),  # LGM
     bounded_argmax(ivol, time, -16, -14),
-    10549,  # 10 ka
+    10699,  # 13 ka
     bounded_argmax(ivol, time, -12, -10),  # YD
     10999,  # 10 ka
-])
-tkalist = time[tkalist]
+]
+tkalist = time[idxlist]
+ivolist = ivol[idxlist]
+ntka = len(tkalist)
+
+# initialize figure
+figw, figh = 170.0, 80.0
+fig = plt.figure(0, (figw/25.4, figh/25.4))
+gs = GridSpec(2, ntka)
+ax = plt.subplot(gs[0, :])
+ax = tsax = plt.subplot(gs[1, :])
+fig.subplots_adjust(left=2.5/figw, right=1-12.5/figw,
+                    bottom=5.0/figh, top=1-2.5/figh,
+                    wspace=1/((1+figw/2.5)/ntka-1),
[…]

### 2014-07-13 kappa.tex
Enhance snapshots plot.
Add list of papers.

@@
 \documentclass{article}
+%\makeatletter\newcommand{\@ptsize}{0}\makeatother  % needed by setspace
 \usepackage{amsmath}
+\usepackage{bibentry}
 \usepackage{bm}
@@
 \usepackage{natbib}
-\usepackage[onehalfspacing]{setspace}
+\usepackage{setspace}
 \usepackage{physics}
@@
 \usepackage{graphicx}
+\onehalfspacing
 \graphicspath{{figures/}}
+\nobibliography*
 
@@
        through the last glacial cycle}
+%\runningtitle{Numerical modelling of the Cordilleran ice sheet}
 \author{Julien Seguinot}
@@
 
+\section*{List of papers}
+
+This doctoral dissertation consists of a cover summary and the following
+papers, which are referred to by their Roman numerals in the text.
+
+\begin{description}
+  \item[\CCLI] \bibentry{Seguinot.etal.2014}
+  \item[\PSDV] \bibentry{Seguinot.2013}
+  \item[\PSDP] \bibentry{Seguinot.Rogozhina.2014}
+  \item[\CCYC] \bibentry{Seguinot.etal.inpreparation}
+\end{description}
+
+\vfill
+\noindent\textbf{Author contributions}
+
+\small{\begin{enumerate}
+  \item[I] I set-up the ice sheet model and supercomputing workflow, ran the
+simulations and wrote the manuscript; C.~Khroulev implemented necessary code
+changes in PISM; I.~Rogozhina guided experiment design; A.~P.~Stroeven
+initiated the Cordilleran ice sheet modelling project; and Q.~Zhang prepared
+some of the forcing data. All authors contributed to improve the text.
+  \item[II] I analysed the weather data, wrote the positive degree day model,
+ran computations and wrote the manuscript. I.~Rogozhina helped to assess the
+reviews.
+  \item[III] I analysed the weather data; I.~Rogozhina and I wrote the manuscript.
+  \item[IV] I implemented necessary code changes in PISM, ran the simulations
+and wrote the manuscript; I.~Rogozhina guided experiment design; M.~Margold,
+J.~Kleman and A.~P.~Stroeven took part in the interpretation. All authors
+contributed to improve the text.
+\end{enumerate}}
+
+\newpage
 \maketitle
 \tableofcontents
+\newpage
 
@@
            temperature offset time-series derived from the GRIP
-           \citep{Dansgaard.etal.1993} ice-core record as described in \CCYC.
-           \todo{X-tick snapshot times on time series plot and thicken the
-                 vertical lines. Add (a) and (b) subfigure labels. Link lower
-                 and upper panels; see recipe at
-                 \url{http://stackoverflow.com/questions/17543359/}.}}
+           \citep{Dansgaard.etal.1993} ice-core record as described in \CCYC.}
   \label{fig:plot-snapshots}

### 2014-07-13 kappa.tex
Add results.

@@
 
+% Github links
+\newcommand{\issue}[1]{\href{https://github.com/pism/pism/issues/#1}{\##1}}
+
 % Text label over graphics
@@
 reviews.
-  \item[III] I analysed the weather data; I.~Rogozhina and I wrote the manuscript.
+  \item[III] I.~Rogozhina provided the original idea; I analysed the weather
+data; and we wrote the manuscript together.
   \item[IV] I implemented necessary code changes in PISM, ran the simulations
@@
 % ----------------------------------------------------------------------
-\subsection{Constant-climate runs}
-% ----------------------------------------------------------------------
+\subsection{Constant-climate simulations}
+% ----------------------------------------------------------------------
+
+\noindent\textbf{\CCLI} --- \bibentry{Seguinot.etal.2014}
+\bigskip
+
+As a first step towards simulations of the Cordilleran ice sheet through the
+last glacial cycle, I have chosen to run the ice sheet model under a constant
+climate, targeting the LGM as the period for which ice sheet limits are best
+known. These simulations, presented in \CCLI, use constant temperature offsets
+from the present climate state. They start from ice-free conditions and run for
+10\,ka, a period chosen as representative of the ice-sheet build-up from
+assumed, negligible ice cover. Although this assumption is disproved in \CCYC,
+the simplistic set-up used in \CCLI\ allowed to test model sensitivity to the
+choice of present-day climate forcing data, without involving too many
+additional parameters.
+
+Discrepancies between datasets used (\cref{tab:datasources}) cause much
+variability in modelled ice sheet geometries (\CCLI, Figs.~5--7). In
+particular, continental parts of the ice-sheet are strongly dependent on input
+precipitation patterns, which are often poorly resolved by atmospheric
+reanalysis fields used to force the model (\CCLI, Figs.~10--14). The closest
+match between model output and the reconstructed ice margin \citep{Dyke.2004}
+was obtained using atmospheric forcing from the North American Regional
+Reanalysis \citep[NARR,][]{Mesinger.etal.2006} and a temperature depression of
+7\,\degC\ (\CCLI, Fig.~15). The relevant simulation is reproduced here, using a
+higher horizontal resolution of 3\,km (\cref{fig:plot-lgm}). The NARR is the
+dataset that I have retained to force simulations of the Cordilleran ice sheet
+presented here (e.g., \cref{fig:plot-lgm}) and in \CCYC.
 
@@
              \todo{Add black frame, and maybe the LGM outline from Dyke?}}
+    \label{fig:plot-lgm}
   \end{minipage}
@@
   \vspace{-16mm}
-  \label{fig:plot-lgm}
 \end{figure*}
@@
 % ----------------------------------------------------------------------
+
+\noindent\textbf{\PSDV} --- \bibentry{Seguinot.2013}
+\bigskip
+
+\noindent\textbf{\PSDP} --- \bibentry{Seguinot.Rogozhina.2014}
+\bigskip
+
+Next to the atmospheric forcing, another critical source of uncertainty in
+palaeo-ice sheet modelling is the choice of parameters for the surface mass
+balance model \citep[e.g.,][]{Hebeler.etal.2008}. In our case, these parameters
+are the degree-day factors for snow, $F_s$, and ice, $F_i$, and the standard
+deviation of near-surface air temperature, $\sPDD$. Although the first two can
+only be deduced from laborious surface mass-balance measurements on present-day
+glaciers \citep[e.g.,][]{Shea.etal.2009}, $\sPDD$ can be estimated from
+observed, modelled, or reanalysed temperature time-series
+\citep[e.g.,][]{Fausto.etal.2011}.
+
+Using temperature time-series from the ERA-Interim reanalysis
+\citep{Dee.etal.2011}, this is what is done of \PSDV, where I show, that,
+although often assumed constant, $\sPDD$ greatly varies through the year and
+around the globe, most particularly in the polar regions
+(\cref{fig:plot-sdmap}; \PSDV, Fig.~1). Using a simple surface mass balance
+model, I then demonstrate the effect of these variations on the computation of
+positive degree-days (\PSDV, Fig.~2), and surface mass-balance (\PSDV, Fig.~3).
+
+The analysis is taken further in \PSDP, using re-analysed air temperature from
+ERA-40 \citep{Uppala.etal.2005}. Although of coarser spatial resolution, the
+ERA-40 time series have a longer time span than that of the ERA-Interim, which
+is necessary to compute standard deviation values that are representative of
+thelong-term daily variability \citep{Rogozhina.Rau.2014}. In \PSDP, we
+demonstrated that, over the Greenland ice sheet, there exist an inverse
+correlation between $\sPDD$ values and the monthly temperature means, $T_m$
+(\PSDP, Fig.~1). This correlation could potentially be used as a
+parametrisation, which can be expressed as (\PSDP, Eq.~5)
+
+\begin{equation}
+    \label{eqn:sigma}
+    \sPDD = -0.15 \cdot T_{m} + 1.66\,.
+\end{equation}
+
+The effect of the PDD standard deviation, $\sPDD$, on simulations of the
+Cordilleran ice sheet is not shown in \PSDV\ and \PSDP. However, following
+these two studies, a natural step further was to implement spatially and
+seasonally variable $\sPDD$ (PISM issue \issue{179}), and the parametrisation
+expressed by \cref{eqn:sigma} over ice-filled grid cells (\issue{265}), in
+PISM. The effect of various temperature variability modelling approaches, and
+associated errors, is demonstrated here using a constant-climate set-up similar
+to that of \CCLI\ (\cref{fig:plot-sdeffect}).
 
@@
            including
-           \textbf{(a)} no daily variability, $\sigma=0$;
+           \textbf{(a)} no daily variability, $\sPDD=0$;
            \textbf{(b)} a constant, mean summer (JJA), model domain-averaged
-           standard deviation value, $\sigma=3.09$\,K;
+           standard deviation value, $\sPDD=3.09$\,K;
            \textbf{(c)} monthly standard deviation grids derived from the NARR
@@
            Greenland ice sheet, ERA-40-based \citep{Uppala.etal.2005}
-           parametrisation in \PSDP, ${\sigma=-0.15\cdot{T_m}+1.66}$.
+           parametrisation in \PSDP, ${\sPDD=-0.15\cdot{T_m}+1.66}$.
            Differences in surface elevation are displayed relative to case (c).
@@
 % ----------------------------------------------------------------------
-\subsection{Transient-climate runs}
-% ----------------------------------------------------------------------
+\subsection{Glacial cycle simulations}
+% ----------------------------------------------------------------------
+
+\noindent\textbf{\CCYC} --- \bibentry{Seguinot.etal.inpreparation}
+\bigskip
+
+From the start, the primary goal of this thesis has been to perform simulations of the Cordilleran ice sheet through the last glacial cycle.
+This is the topic of \CCYC, where monthly air temperature, precipitation, and PDD standard deviation computed from the NARR are amended by palaeo-temperature time-series from six proxy records to produce a time-dependent climate forcing (\CCYC, Fig.~2).
+The resulting simulations all produce two major glaciations between MISs~4 and 2.
+However the differences in timing and magnitude of these events is considerable (\CCYC, Fig.~3--4).
+
+Geological evidence regarding the timing of glaciation allowed us to select two
+simulations, driven by the GRIP \citep{Dansgaard.etal.1993} and EPICA \citep{Jouzel.etal.2007} ice-core records, as producing a more realistic reconstruction of the last glacial cycle.
+I have re-run these two simulations at a higher resolution of 6\,km, to allow for a more detailed study of the primary locations of transient ice cap cover, stability of major ice-dispersal centres, and patterns of deglaciation (\CCYC, Fig.~5--15).
+Modelled ice volume evolution and snapshots of ice flow patterns from one of these two simulations are reproduced here (\cref{fig:plot-snapshots}).
+
+\todo{Verify that figure numbers remain correct as \CCYC\ evolves.}
 
@@
 % References
-\newcommand{\urlprefix}[0]{}  % remove default "URL" prefix
+\renewcommand{\urlprefix}[0]{}  % remove default "URL" prefix
 \bibliographystyle{copernicus}

### 2014-07-13 kappa.tex
Add preliminary conclusions.

@@
 % ======================================================================
-\section{Future perpectives}
+\section{Conclusion and perpectives}
 % ======================================================================
-% Model improvements
-% Fieldwork
-% Geomorphological-oriented plots
-
-%\begin{figure}
-%  \centering
-%  \makebox[0pt]{\includegraphics{photo-tuya-moraine}}
-%  \caption{The Tuya Lake Moraine in northern British Columbia.}
-%  \label{fig:photo-tuya-moraine}
-%\end{figure}
+
+In this thesis, I present numerical reconstructions of the former Cordilleran
+ice sheet to a level of detail that has, to my knowledge, never been attained
+before. The coupled SIA-SSA approximation is crude, yet it is today the only
+approach that allows, combined with massively parallel computing, this level of
+detail over such time-scales as involved in the growth and decay of ice sheets.
+
+Through sensitivity studies, I have identified the predominant sources of
+uncertainty, which I have assessed using a combination of detailed
+meteorological datasets and available geological evidence. The simulations
+supporting this thesis allowed estimates of the LGM Cordilleran ice sheet
+extent and volume, identification of the main ice-dispersal centres, and
+partial understanding of the deglaciation patterns. However, there remain much
+room for improvement of the model set-up and further analysis of the simulation
+output.
+
+% ----------------------------------------------------------------------
+\subsection{Precipitation change}
+% ----------------------------------------------------------------------
+
+Throughout the thesis, precipitation is prescribed as monthly, spatial grids,
+but it is assumed equal to the present values. Although there is no doubt that
+the distribution of precipitation has been affected by the evolution of global
+climate and the intermittent presence of ice sheets and ice caps, such changes
+are difficult to assess. In the North American Cordillera, much of the interior
+was likely dryer than today due to the presence of ice caps reinforcing the
+topographic barrier formed by the Pacific Coast ranges. However, some regions
+may have become wetter as a side effect of redirected winds.
+
+In meteorology, these effects are typically modelled using atmospheric
+circulation models. However, I have shown in \CCLI that atmospheric reanalyses,
+circulation models using data assimilation techniques, fail to agree on the
+present-day pattern of precipitation in the Cordillera due to the highly
+mountainous topography, with dramatic effects on the ice sheet model response.
+Considering additional difficulties inherent to palaeo-climate modelling, it
+can only be expected, that palaeo-climate models without data assimilation will
+produce even much higher discrepancies in the ice sheet model response.
+
+% ----------------------------------------------------------------------
+\subsection{Geothermal heat flux}
+% ----------------------------------------------------------------------
+
+Geothermal heat flux, controlling en-glacial temperature distribution and the
+location of basal melt, was assumed constant. However the North America is an
+active subduction zone with complex geothermal patterns. Including this
+variability in the ice sheet model will offer new opportunities to study its
+output.
+
+Most notably, modelled basal flow could be compared with glacial lineations,
+one of the main sources of evidence used by geomorphologists to reconstruct
+former ice sheets. For instance, in our simulations, large portions of the
+Cordilleran ice sheet have remain cold-based throughout the simulation. As
+cold-based glaciers typically do not form lineations, this could indicate that
+some of the lineation patterns we observe in the North American Cordillera
+could predate deglaciation, and instead have been formed during an earlier
+phase of advance or retreat of the ice sheet.
+
+% ----------------------------------------------------------------------
+\subsection{The Cassiar ice cap}
+% ----------------------------------------------------------------------
+
+In the summer of 2012, I have co-organised and participated in fieldwork with
+Martin Margold and Ping Fu to collect samples of glacial deposits in the
+southern Cassiar Mountains, northern British Columbia. The primary target of
+the fieldwork was a triple-crested, sharp, ca. 50\,m-high moraine located in
+the vicinity of Tuya Lake the western slope of the Cassiar Mountains
+(\cref{fig:photo-tuya-moraine}). Although this is a unique feature in the area,
+the size of the moraine, and its location at the foothills of the mountain
+range indicate that it was likely formed by the tributary from an ice cap,
+rather than a stand-alone valley glacier. Because the relatively low-elevated
+Cassiar Mountains do not appear as a major nucleation centre in our simulations
+(\CCYC, Fig.~5), it is tempting to think that such a large feature may result
+from late-glacial re-advance of a remnant of the decaying ice sheet.
+High-resolution modelling of the deglaciation phase, perhaps at a regional
+scale, may help to assess this hypothesis.
+
+\begin{figure}
+  \centering
+  \makebox[0pt]{\includegraphics{photo-tuya-moraine}}
+  \caption{The Tuya Lake Moraine in northern British Columbia.}
+  \label{fig:photo-tuya-moraine}
+\end{figure}
 

### 2014-07-13 kappa.tex
Rework results and conclusions with Em Yeu.

--- 2014-07-13 kappa.tex
@@
 This doctoral dissertation consists of a cover summary and the following
-papers, which are referred to by their Roman numerals in the text.
+journal articles, which are referred to by their Roman numerals in the text.
 
@@
 % ----------------------------------------------------------------------
-\subsection{Constant-climate runs}
-% ----------------------------------------------------------------------
+\subsection{Constant-climate simulations}
+% ----------------------------------------------------------------------
+
+\noindent\textbf{\CCLI} --- \bibentry{Seguinot.etal.2014}
+\bigskip
+
+As a first step towards simulations of the Cordilleran ice sheet through the
+last glacial cycle, I have chosen to run the ice sheet model under a constant
+climate, targeting the LGM as the period for which ice sheet limits are best
+known. These simulations, presented in \CCLI, use constant temperature offsets
+from the present climate state. They start from ice-free conditions and run for
+10\,ka, a period chosen as representative of the ice-sheet build-up from
+assumed, negligible ice cover. Although this assumption is disproved in \CCYC,
+the simplistic set-up used in \CCLI\ allowed to test model sensitivity to the
+choice of present-day climate forcing data, without involving too many
+additional parameters.
+
+Discrepancies between input data used (\cref{tab:datasources}) caused much
+variability in modelled ice sheet geometries (\CCLI, Figs.~5--7). In
+particular, continental parts of the ice-sheet are strongly dependent on input
+precipitation patterns, which are often poorly resolved by atmospheric
+reanalysis fields used to force the model (\CCLI, Figs.~10--14). The closest
+match between model output and the reconstructed ice margin \citep{Dyke.2004}
+was obtained using atmospheric forcing from the North American Regional
+Reanalysis \citep[NARR,][]{Mesinger.etal.2006} and a temperature depression of
+7\,\degC\ (\CCLI, Fig.~15). The relevant simulation is reproduced here
+(\cref{fig:plot-lgm}), using a higher horizontal resolution of 3\,km, instead
+of 10\,km as in the paper. Following this sensitivity study, I have retained
+the NARR to force subsequent simulations of the Cordilleran ice sheet
+presented here and in \CCYC.
 
@@
 % ----------------------------------------------------------------------
+
+\noindent\textbf{\PSDV} --- \bibentry{Seguinot.2013}
+\bigskip
+
+\noindent\textbf{\PSDP} --- \bibentry{Seguinot.Rogozhina.2014}
+\bigskip
+
+Next to the atmospheric forcing, another critical source of uncertainty in
+palaeo-ice sheet modelling is the choice of parameters used by the surface mass
+balance model \citep[e.g.,][]{Hebeler.etal.2008}. In our case, these parameters
+are the degree-day factors for snow, $F_s$, and ice, $F_i$, and the standard
+deviation of near-surface air temperature, $\sPDD$. Although the first two can
+only be deduced from laborious surface mass-balance measurements on present-day
+glaciers \citep[e.g.,][]{Shea.etal.2009}, $\sPDD$ can be estimated from
+temperature time-series \citep[e.g.,][]{Fausto.etal.2011}.
+
+In \PSDV, I used near-surface air temperatures from the ERA-Interim reanalysis
+\citep{Dee.etal.2011} to compute a global, monthly distribution of $\sPDD$.
+The resulting grids show
+that, although often assumed constant, $\sPDD$ varies greatly by location and
+through the year, most particularly in the polar regions
+(\cref{fig:plot-sdmap}; \PSDV, Fig.~1). Using a simple surface mass balance
+model, I then demonstrated the effect of these variations on the computation of
+positive degree-days (\PSDV, Fig.~2), and surface mass-balance (\PSDV, Fig.~3).
+
+The analysis is taken further in \PSDP, using re-analysed air temperatures from
+ERA-40 \citep{Uppala.etal.2005}. Although of coarser spatial resolution, the
+ERA-40 time series have a longer time span than the ERA-Interim ones, which
+is necessary to compute standard deviation values that are representative of
+the long-term daily variability \citep{Rogozhina.Rau.2014}. In \PSDP, we
+demonstrated that, over the Greenland ice sheet, there exist an inverse
+correlation between $\sPDD$ and monthly mean temperatures, $T_m$
+(\PSDP, Fig.~1). This correlation could potentially be used as a
+parametrisation, and can be expressed as (\PSDP, Eq.~5)
+
+\begin{equation}
+    \label{eqn:sigma}
+    \sPDD = -0.15 \cdot T_{m} + 1.66\,.
+\end{equation}
+
+The effect of the PDD standard deviation, $\sPDD$, on simulations of the
+Cordilleran ice sheet is not shown in any of the papers. However, following
+these two studies, a natural step forward was to implement, in PISM, spatially
+and seasonally variable $\sPDD$ input (PISM issue \issue{179}), and the
+parametrisation expressed by \cref{eqn:sigma} over ice-filled grid cells
+(\issue{265}). Their effect on the modelled Cordilleran ice sheet surface is
+demonstrated here (\cref{fig:plot-sdeffect}) using a constant-climate set-up
+similar to the one used in \CCLI.
 
@@
 % ----------------------------------------------------------------------
-\subsection{Transient-climate runs}
-% ----------------------------------------------------------------------
+\subsection{Glacial cycle simulations}
+% ----------------------------------------------------------------------
+
+\noindent\textbf{\CCYC} --- \bibentry{Seguinot.etal.inpreparation}
+\bigskip
+
+From the beginning, the primary goal of this thesis has been to perform
+simulations of the Cordilleran ice sheet through the last glacial cycle. This
+is the topic of \CCYC, where monthly air temperature, precipitation, and PDD
+standard deviation climatologies from the NARR are amended by
+palaeo-temperature time-series from six proxy records to produce a
+time-dependent climate forcing (\CCYC, Fig.~2). The resulting simulations all
+produce two major glaciations between MISs~4 and 2. However the differences in
+timing and magnitude of these events is considerable (\CCYC, Fig.~3--4).
+
+Geological evidence regarding the timing of glaciation allowed us to select two
+simulations, driven by the GRIP \citep{Dansgaard.etal.1993} and EPICA
+\citep{Jouzel.etal.2007} ice-core records, as producing more realistic
+reconstructions of the last glacial cycle. I have re-run these two simulations
+at a higher resolution of 6\,km, to allow for a more detailed study of
+transient ice cap cover before and after the LGM, stability of major
+ice-dispersal centres, and patterns of deglaciation (\CCYC, Fig.~5--15).
+Modelled ice volume evolution and snapshots of ice flow patterns from one of
+these two simulations are reproduced here (\cref{fig:plot-snapshots}).
+
+\todo{Verify that figure numbers remain correct as \CCYC\ evolves.}
 
@@
 % ======================================================================
-\section{Future perpectives}
+\section{Conclusion and perpectives}
 % ======================================================================
-% Model improvements
-% Fieldwork
-% Geomorphological-oriented plots
-
-%\begin{figure}
-%  \centering
-%  \makebox[0pt]{\includegraphics{photo-tuya-moraine}}
-%  \caption{The Tuya Lake Moraine in northern British Columbia.}
-%  \label{fig:photo-tuya-moraine}
-%\end{figure}
+
+In this thesis, I present numerical reconstructions of the former Cordilleran
+ice sheet to a level of detail that has, to my knowledge, never been attained
+before. The coupled SIA-SSA approximation is crude, yet it is today the only
+approach that allows, combined with massively parallel computing, high-spatial
+resolution over the multi-millenial time-scales of ice sheet growth and decay.
+
+Through sensitivity studies, I have identified the predominant sources of
+uncertainty, which I have assessed using a combination of detailed
+meteorological datasets and available geological evidence. The simulations
+supporting this thesis allowed estimates of the LGM Cordilleran ice sheet
+extent and volume, identification of the main ice-dispersal centres, and
+partial understanding of the deglaciation patterns. However, there remains much
+room for model improvements and further analysis of the simulation output.
+
+% ----------------------------------------------------------------------
+\subsection{Precipitation change}
+% ----------------------------------------------------------------------
+
+Throughout the thesis, precipitation rates are prescribed as monthly, spatial
+grids, but assumed equal to their present values. Although the distribution of
+precipitation has been undoubtedly affected by past changes in global climate
+and the intermittent presence of ice sheets and ice caps in the Northern
+Hemisphere, its changes are difficult to assess. In the North American
+Cordillera, much of the continental interior was likely dryer than today due to
+the presence of ice caps reinforcing the topographic barrier formed by the
+Pacific Coast ranges. However, some regions might have become rainier as a side
+effect of redirected winds.
+
+In meteorology, these effects are typically analysed using atmospheric
+circulation models. However, I have shown in \CCLI\ that atmospheric reanalyses
+which are circulation models that use data assimilation techniques, fail to
+agree on the present-day pattern of precipitation in the Cordillera. The
+discrepancies, due to highly mountainous terrain, have dramatic effects on the
+ice-sheet-model response. Considering additional difficulties inherent to
+modelling past climates, and the lack of data for assimilation, we can only
+expect that palaeo-climate reconstructions will produce even higher
+discrepancies in the ice-sheet-model response.
+
+% ----------------------------------------------------------------------
+\subsection{Geothermal heat flux}
+% ----------------------------------------------------------------------
+
+Geothermal heat flux, controlling en-glacial temperature distribution and the
+location of basal melt areas, was assumed constant. However, the North America
+is an active subduction zone with complex geothermal patterns
+\citep{Blackwell.Richards.2004}. Including this variability in the ice-sheet
+model will offer new opportunities to study its output.
+
+Most notably, modelled basal flow could be compared with glacial lineations,
+one of the main sources of geomorphological evidence used to reconstruct former
+ice sheets. For instance, in our simulations, large portions of the Cordilleran
+ice sheet remain cold-based throughout deglaciation (\CCYC, Fig.~15). As
+cold-based glaciers typically do not deform their beds, this could indicate
+that some of the lineation patterns we observe in the North American Cordillera
+could pre-date deglaciation. Instead, they may have been formed during an
+earlier phase of advance or retreat of the ice sheet.
+
+% ----------------------------------------------------------------------
+\subsection{The Cassiar ice cap}
+% ----------------------------------------------------------------------
+
+In the summer of 2012, I have co-organised and participated in a fieldwork with
+Martin Margold and Ping Fu to collect samples of glacial deposits in the
+southern Cassiar Mountains of northern British Columbia
+(\cref{fig:map-cordillera}). The primary target of the fieldwork was a
+triple-crested, sharp, ca. 50\,m-high moraine located in the vicinity of Tuya
+Lake on the western slope of the Cassiar Mountains
+(\cref{fig:photo-tuya-moraine}). Although this is a unique feature in the area,
+the size of the moraine and its location at the foothill of the mountain range
+indicate that it was likely formed by a tributary glacier from an ice cap,
+rather than a stand-alone valley glacier. The relatively low-elevated Cassiar
+Mountains do not appear as a major nucleation centre in our simulations (\CCYC,
+Fig.~5). Therefore, it is tempting to think that such a large feature may
+result from the late-glacial re-advance of a remnant of the decaying ice sheet.
+High-resolution modelling of the deglaciation phase, perhaps at a regional
+scale, may help to study this hypothesis.
+
+\begin{figure}
+  \centering
+  \makebox[0pt]{\includegraphics{photo-tuya-moraine}}
+  \caption{The Tuya Lake Moraine in northern British Columbia.}
+  \label{fig:photo-tuya-moraine}
+\end{figure}
 
 % References
-\newcommand{\urlprefix}[0]{}  % remove default "URL" prefix
+\renewcommand{\urlprefix}[0]{}  % remove default "URL" prefix
 \bibliographystyle{copernicus}

### 2014-07-14 kappa.tex
Re-think software implementation outline.
Begin to implement Arjen's comments.

@@
 % My commands
-%\newcommand{\note}[1]{\textbf{[NOTE: #1]}}
+\newcommand{\renote}[1]{\footnote{\textbf{Comment reply}: #1}}
 \newcommand{\todo}[1]{\emph{[\textbf{Todo:} #1]}}
@@
 % Document properties
-\title{Numerical simulation of the Cordilleran ice sheet \\
-       through the last glacial cycle}
+\title{Numerical modelling of the Cordilleran ice sheet}
 %\runningtitle{Numerical modelling of the Cordilleran ice sheet}
@@
 \newpage
-\maketitle
+%\maketitle
 \tableofcontents
@@
 
-In our daily lives, we most commonly experience ice as the small, solid cubes
-that crackle on contact with refreshing liquids. Yet, at geological
-time-scales, and under high stresses, ice behaves in a much different fashion.
-In the cold and mountainous regions of the globe, ice accumulates under the
-action of cumulative snowfall, until it reaches a critical thickness, and
+In our daily life, we most commonly experience ice as small, solid cubes
+that crackle on contact with refreshing liquids. Yet, on geological
+time scales, and under high pressure, ice behaves quite differently.
+In the cold and mountainous regions of the globe, ice accumulates through
+cumulative snowfall until it reaches a critical thickness. Beyond this
+threshold, usually in excess of 50\,m, but dependent on temperature, ice
 deforms under its own weight to flow downhill (\cref{fig:glacier-mechanics}).
-Glaciers, these bodies of moving ice, flow by a combination of water-induced
-sliding at the base \citep{Saussure.1796}, and visco-plastic deformation within
-the ice body \citep{Forbes.1846b}.
-
-As ice flows and glides on its bed, it transports rock debris and erodes the
-landscape, thence leaving geomorphological traces of its former presence
-(\cref{fig:glaciation-indicators}). In many parts of the world, montane people
-and early explorers have since long learned to read these traces, and to
-understand that glaciers had once been more extensive than today. With the
-advent of the glacial theory, birthed the idea that, under colder temperatures
+Glaciers, these bodies of moving ice, move by a combination of
+meltwater-induced sliding at the base \citep[\S~532]{Saussure.1786}\renote{
+    I mean volume 1, which as a couple of beautiful paragraphs (would be worth
+    translating if I had time) about how the "heat from the Earth's interior"
+    melts basal ice to lubricate glacier flow. This is precisely those
+    paragraphs that \citet{Forbes.1846} refers to sixty years later when
+    advancing his ``viscous theory of glacier notion''. According to the cover
+    page, volume 1 was published in ``MDCCLXXXVI'', which I just translated to
+    1786 (so 1796 was wrong).},
+and viscous deformation within the
+ice body \citep{Forbes.1846b}.
+
+As glaciers flow and slide across their bed, they transport\renote{
+    I would like to keep this statement general, surface transport is also
+    important. Erosion is not only by basal rock debris but also, indirectly,
+    by glacial meltwater.}
+rock debris and erode the landscape, thereby leaving geomorphological traces of
+their former presence (\cref{fig:glaciation-indicators}). In several parts of
+the world, montane people\renote{
+    I insist on keeping this: people of the Alps, for istance, had understood
+    that glacial erratics originated from glaciers before it has ever been
+    documented by explorers and scientists. It is stated in the old literature.
+    Check the new reference, reprinted in \citet{Mathews.1898} (not quite sure
+    how to cite this material).}
+and early explorers learned since long to read these traces, and to
+understand that glaciers had once been more extensive than today
+\citep[e.g.,][p.~21]{Windham.Martel.1744}. Wtih the emergence of the glacial
+theory came the idea that, under colder temperatures
 than today, expansive \emph{ice sheets} had once covered much of Europe and
-North America \citep{Agassiz.1840}.
-
-However, the glacial theory did not gain much belief until the exploration and
-discovery of the two present-day ice sheets on Earth: the Greenland and
+North America \citep{Agassiz.1840}\renote{
+    I am a bit reluctant to write ``\citep{Agassiz.1840} eventually posed the
+    glacial theory''. Although he published these ideas, do they really
+    originate from him?}.
+
+However, the glacial theory did not gain general acceptance until the
+discovery and exploration
+of the two present-day ice sheets on Earth: the Greenland and
 Antarctic ice sheets, provided a modern analogue for the proposed European and
@@
 
-More recently, palaeo-climate records extracted from deep-sea sediments have
-provided a much more detailed picture of the Earth environmental history
-\citep[e.g.,][]{Emiliani.1955, Shackleton.Opdyke.1973}, thus indicating neither
-one, nor four, but rather tens of glacial cycles. During the last million
-years, glacial and interglacial cycles have succeeded each other with a
-100\,kyr periodicity \citep{Hays.etal.1976}. In the present thesis, we focus on
-the most recent of these periods: \emph{the last glacial cycle}
-(\cref{fig:paleo-timeseries})
+More recently, palaeoclimate records extracted from deep sea sediments and ice
+cores have provided a much more detailed picture of the Earth environmental
+history \citep[e.g.,][]{Emiliani.1955, Shackleton.Opdyke.1973,
+Dansgaard.etal.1993}, indicating neither one, nor four, but rather tens of
+glacial cycles. During the last 800\,ka years, glacial and interglacial cycles
+have succeeded each other with a 100\,ka periodicity \citep{Hays.etal.1976,
+Augustin.etal.2004}. In this thesis, I focus on the most recent of these
+periods: \emph{the last glacial cycle} (\cref{fig:paleo-timeseries})
 
 % ice volume 26.07 + 2.93 = 29.00
-% meters sea-level 58.3 + 0.70 = 59.0
-Together, the Greenland and Antarctic ice sheets currently host most glacial
-ice on the planet, corresponding to 59~meters of potential sea-level rise
-\citep{Bamber.etal.2001, Fretwell.etal.2013}. At the apogee of the last glacial
-cycle, the \emph{Last Glacial Maximum} (LGM), sea-level was 120 to 135\,m lower
-than today
+% meters sea-level 58.3 + 7.36 = 65.6
+Together, the Greenland and Antarctic ice sheets currently host almost all
+glacial ice on the planet, corresponding to 65.6\,m of potential sea level
+rise, were these ice sheets to melt away entirely \citep{Bamber.etal.2013,
+Fretwell.etal.2013}. At the apogee of the last glacial period, global sea level
+was 120 to 135\,m lower than today
 \citep{Clark.Mix.2002}, implying an additional, double amount of ice stored in
 ice sheets and glaciers on land. Systematic mapping and compilation of glacial
-landforms from all continents permitted reconstructions of this former ice
-cover \citep{Ehlers.Gibbard.2007}, indicating that ice volume
+landforms from all continents \citep[e.g.,][]{Prest.etal.1968,
+Boulton.Clark.1990, Kleman.etal.1997, Hattestrand.1998} permitted
+reconstructions of these former ice covers \citep{Ehlers.Gibbard.2007},
+indicating that the excess ice volume
 was repartitioned between more extensive Greenland and Antarctic ice sheets,
-the Scandinavian and Barents-Kara ice sheets in the Eurasian arctic, and the
-Innuitian and Laurentide ice sheets and, finally, the \emph{Cordilleran ice
-sheet}, over North America (\cref{fig:paleo-glaciation}).
-
-Nevertheless, the landscape record is sparse in time and most often spatially
-incomplete. Palaeo-ice sheets did not leave a systematic record of their
-passage, and much of the evidence has been overridden by subsequent glacier
-re-advances and other geomorphological processes \citep{Kleman.1994}.
+the Fennoscandian and Barents-Kara ice sheets in the Eurasian Arctic, and the
+Innuitian, Laurentide, and \emph{Cordilleran} ice
+sheets over North America (\cref{fig:paleo-glaciation}).
+
+Nevertheless, the landform record is sparse in time and, most often, spatially
+incomplete. Palaeo-ice sheets did not leave a continuous imprint on the
+lanscape, and much of this evidence has been overridden by subsequent glacier
+re-advances and other geomorphological processes \citep{Kleman.1994,
+Kleman.etal.2006, Kleman.etal.2010}.
 Therefore, geomorphological reconstructions of former glacier extent require a
-substantial amount of interpretation. This is where numerical models,
+substantial amount of interpretation. This is where numerical ice sheet models,
 incorporating the processes of glacier sliding \citep[e.g.,][]{Weertman.1957}
-and deformation \citep[e.g.,][]{Nye.1953} observed on modern glaciers, can help.
-Although these models alone contain too many uncertainties to reproduce
-accurate picture of past glacial cycles \citep[e.g.,][]{Hebeler.etal.2008},
-they can be calibrated against geological data where they exist, while
-allowing a physics-based, quantitative reconstruction where they do not
+and deformation \citep[e.g.,][]{Nye.1953} observed on modern glaciers, can
+help. Although these models alone contain too many uncertainties to reproduce
+accurate ice sheet reconstructions during former glacial cycles
+\citep[e.g.,][]{Hebeler.etal.2008}, they can be calibrated against geological
+data where they exist \citep{Napieralski.etal.2007}, while allowing
+physics-based, quantitative reconstructions of ice sheet sectors where they do
+not
 \citep[e.g.,][]{Marshall.etal.2002, Tarasov.Peltier.2004}.
+In addition, numerical ice sheet models provide quantitative insights on the
+four-dimensional distribution of stress and deformation in palaeo-ice sheets
+and allow for reconstructions their past thickness and volume evolution, which
+are difficult to infer from geological evidence
+\renote{
+    Is this the place where to cite \citep{Denton.Hughes.1981}? \todo{I need to
+    check the contents of this book (library).}
+}.
+
+In this thesis, I focus on numerical modelling of the last Cordilleran ice
+sheet, which once covered much of the mountainous terrain of western North
+America \citep{Dawson.1888}, a region where topographic complexity has
+presented a particular challenge to reconstructing glacial history based on the
+geomorphological record \citep[e.g.,][]{Kleman.etal.2010} or numerical ice
+sheet modelling \citep{Robert.1991}.
 
@@
 % ======================================================================
-\section{Field area description}
+\section{Physical setting}
 % ======================================================================
 
-During the last glacial cycle, much of North America has been covered by ice.
-At the last glacial maximum, apex of glaciation, three major ice sheets had
+During the last glacial cycle, much of North America has been covered by ice
+(\cref{fig:map-northamerica}).
+During the last glacial maximum, apex of glaciation, three major ice sheets had
 collided to form a continuous blanket of ice that spanned from Alaska to
@@
 % ----------------------------------------------------------------------
-\subsection{Topographic setting}
+\subsection{Topography}
 % ----------------------------------------------------------------------
@@
 % ----------------------------------------------------------------------
-\subsection{Climatic setting}
+\subsection{Climate}
 % ----------------------------------------------------------------------
@@
 % ======================================================================
-\section{Software tools and contributions}
+\section{Software implementation}
 % ======================================================================
@@
 % ----------------------------------------------------------------------
-\subsection{A Python PDD model}
-% ----------------------------------------------------------------------
-\todo{Figure. PyPDD demo example}
-
-% ----------------------------------------------------------------------
-\subsection{Pre-processing scripts for GRASS GIS}
-% ----------------------------------------------------------------------
-
-% ----------------------------------------------------------------------
-\subsection{Plotting tools for PISM}
-% ----------------------------------------------------------------------
-\todo{Figure. Plotting tools example}
-
-% ----------------------------------------------------------------------
 \subsection{Contributions to PISM}
@@
 \todo{Table. PISM ``issues'' identified}
+
+% ----------------------------------------------------------------------
+\subsection{Peripheral tools}
+% ----------------------------------------------------------------------
+A Python PDD model\\
+Pre-processing scripts for GRASS GIS\\
+Plotting tools for PISM\\
 

### 2014-07-15 cordillera-cycle.tex
Implement Arjen's comment on intro to results.

@@
 % My commands
-\newcommand{\note}[1]{\textbf{[NOTE: #1]}}
+\newcommand{\renote}[1]{\footnote{\textbf{Comment reply}: #1}}
 \newcommand{\todo}[1]{\emph{[\textbf{Todo:} #1]}}
@@
 Cordillera have been more extensive than today. At the Last Glacial
-Maximum (LGM), a continuous blanket of ice, the Cordilleran ice sheet
+Maximum (LGM), a continuous blanket of ice, the Cordilleran ice sheet\renote{
+    In \citep{Seguinot.etal.2014}, The Cryosphere changed all my occurences of
+    Cordilleran (Greenland etc) \textbf{I}ce \textbf{S}heet to lowercase
+    \textbf{i}ce \textbf{s}heet. Let's keep it so for now for consistency
+    within the thesis (kappa + papers).}
 \citep{Dawson.1888}, extended from the Alaska Range in the north to the
@@
     last glacial maximum?
+  \item What was the scale of glaciation prior to the last glacial maximum?
   \item Which were its primary dispersal centres? Do they reflect stable or
     ephemeral configurations?
-  \item What was the scale of glaciation prior to the last glacial maximum?
   \item How rapid was the last deglaciation? Did it include late glacial
@@
 
-The simulations presented here were run using Parallel Ice Sheet Model (PISM,
-development version~11b0a7f and stable version~0.6.1), an open-source,
-finite-difference, shallow ice sheet model \citep{PISM-authors.2014}.
-The model inputs basal topography, sea level, geothermal
-heat flux and climate forcing, and computes the evolution of ice extent
-and thickness in time, the thermal and dynamic state of the ice sheet, and
-the associated lithospheric response. Our modelling domain encompasses the
-entire area covered by the Cordilleran ice sheet at the last glacial maximum
-(Fig.~\ref{fig:locmap}).
-
-To reconstruct the successive phases of growth and decay of the last Cordilleran
-ice sheet, palaeo-climatic conditions of the last glacial cycle are mimicked
-by applying time-dependent temperature offsets derived from multiple
-palaeo-temperature proxy records. Each simulation starts from assumed ice-free
-conditions at 120\,kyr, and runs to present. These computations were
-performed on 16 to 128 cores at the Swedish National Supercomputing Centre.
+%\subsection{Forcings}
+
+\todo{Move parameter values into a separate table.}
+
+\todo{Do we need a new subsection title here?}
+
+The simulations presented here were run using the Parallel Ice Sheet Model
+(PISM, development version~11b0a7f and stable version~0.6.1), an open source,
+finite difference, shallow ice sheet model \citep{PISM-authors.2014}. The model
+requires input on basal topography, sea level, geothermal heat flux and
+atmospheric\renote{
+    Changed here from ``climate'' to ``atmospheric''. Although I have used
+    ``climate forcing'' in the previous paper, I now think that ``atmospheric
+    forcing'' is more appropriate: ``climate'' may also encompass the oceanic
+    component, not really assessed here.}
+forcing, and computes the evolution of ice extent
+and thickness over time, the thermal and dynamic\renote{
+    Thermal = temperature, dynamic = force + movement. If I meant movement
+    alone, I would use ``cinematic''. Here we could use ``ice temperatures,
+    stress and deformation (i.e. velocities)'', but that becomes a bit clumsy,
+    I feel.}
+states of the ice sheet, and the associated lithospheric response.
+
+Basal topography is derived from the ETOPO1 combined topography and bathymetry
+dataset with a~resolution of 1\,arc-min \citep{Amante.Eakins.2009}. Sea level
+is lowered as a function of time based on the Spectral Mapping Project\renote{
+    I had no time to check for later products, but if you have suggestions for
+    something more up-to-date, I can try to include it in later simulations
+    (everything will have to be re-run anyway).}
+\citep[SPECMAP;][]{Imbrie.etal.1989} time scale. Geothermal heat flux
+is applied as a constant value of 70\,\unit{mW\,m^{-2}} at 3\,km depth
+(Sect.~\ref{sec:icedyn}). Atmospheric forcing is provided by a monthly
+climatology from the North American Regional Reanalysis
+\citep[NARR;][]{Mesinger.etal.2006} perturbated by time-dependent and
+lapse-rate temperature corrections (Sect.~\ref{sec:atm}).
+
+Each simulation starts from assumed ice-free conditions at 120\,000 years ago
+(120\,ka), and runs to present. Our modelling domain of 1\,500 by 3\,000\,km
+encompasses the entire area covered by the Cordilleran ice sheet at the LGM
+(Fig.~\ref{fig:locmap}). The simulations were run on two distinct grids, using
+a lower horizontal resolution of 10\,km, and a higher horizontal resolution of
+6\,km. These computations were performed on 16 to 128 computing cores at the
+Swedish National Supercomputing Centre.
 
 \subsection{Ice thermodynamics}
-
-PISM is a~shallow model, which implies that the balance of stresses is
-approximated based on their predominant components.
-The Shallow Shelf Approximation (SSA) is used as a ``sliding law'' for the
-Shallow Ice Approximation (SIA) by adding both velocity solutions
-\citep[Eqn.~15]{Winkelmann.etal.2011}.
-Ice softness depends on temperature and water content through an enthalpy
-formulation \citep{Aschwanden.etal.2012}. Surface air
-temperature from the atmospheric forcing provides the upper boundary condition
-to the ice enthalpy model. Temperature is computed subglacially to a~depth of
-3\,km, where it is conditioned by a~geothermal heat flux of
-70\,\unit{mW\,m^{-2}}. Although this uniform value does not
-account for the high spatial geothermal variability in the region, it is on
-average representative of available heat flow measurements
-\citep{Blackwell.Richards.2004}.
-
-A~pseudo-plastic sliding law \citep{PISM-authors.2014}\footnote{I need to ask
-    Ed Bueler if the sliding will be published somewhere soon.} relates the
-bed-parallel shear stress and the sliding velocity. The yield stress is
-modelled using the Mohr--Coulomb criterion. The till friction angle $\phi$
-varies from 15 to 45{\degree}. It is taken as a~function of modern bed
-elevation, with lowest values occurring at low elevation, thus accounting
-for a~weakening of till associated with the presence of marine sediments.
-The till effective pressure is related to the modelled amount of water in the
-till, using a formula derived from laboratory experiments with till extracted
-from an Antarctic ice stream \citep[Eqn.~2]{Tulaczyk.etal.2000}. Basal
-topography is
-derived from the ETOPO1 combined topography and bathymetry dataset with
-a~resolution of 1\,arc-min \citep{Amante.Eakins.2009}. Sea level is lowered as
-a function of time based on the Spectral Mapping Project (SPECMAP)
-\citep[SPECMAP,][]{Imbrie.etal.1989} time scale.
-Basal topography responds to ice load
+\label{sec:icedyn}
+
+Ice sheet dynamics are typically modelled using a combination of internal
+deformation and basal sliding. PISM is a~shallow ice sheet model, which implies
+that the balance of stresses is approximated based on their predominant
+components. The Shallow Shelf Approximation (SSA) is used as a ``sliding law''
+for the Shallow Ice Approximation (SIA) by adding both velocity solutions
+\citep[Eqns.~7--9 and 15]{Bueler.Brown.2009, Winkelmann.etal.2011}\renote{
+    Arjen, I don't agree with your comment. Reference to equation number(s)
+    show the reader where to look. \citet{Winkelmann.etal.2011} is a rather
+    long paper that also treats many other aspects of the model. Their SIA+SSA
+    approach builds on \citep{Bueler.Brown.2009}, but it is slightly different.
+    It is the one in use in PISM, and in my simulations.}.
+Ice rheology depends on temperature and water content through an enthalpy
+formulation \citep{Aschwanden.etal.2012}. Surface air temperature derived from
+the atmospheric forcing (Sect.~\ref{sec:atm}) provides the upper boundary
+condition to the ice enthalpy model. Temperature is computed subglacially to
+a~depth of 3\,km\renote{
+    Yes, I changed this from 1 to 3\,km as we now model longer period of times
+    than in the previous paper. The penetration depth of temperature anomalies
+    (i.e. skin depth) depends on the period of temperature oscillations at the
+    surface. The longer the period (in our case 100\,kyr), the deeper the
+    temperature change. A good way to estimate what this parameter should be is
+    to solve the 1-D temperature diffusion equation in the case of periodic
+    surface forcing, using rock heat capacity and rock thermal conductivity
+    values similar as in the numerical model. This basically forms the basis
+    for my choice of the 3\,km value.},
+where it is conditioned by a lower boundary geothermal heat flux of
+70\,\unit{mW\,m^{-2}}. Although this uniform value does not account for the
+high spatial geothermal variability in the region
+\citep{Blackwell.Richards.2004}, it is, on average, representative of available
+heat flow measurements. In the low-resolution simulations, the vertical grid
+consists of to 51~enthalpy layers in the ice sheet and 31~temperature layers in
+the bedrock. In the high-resolution simulations, up to 101~ice layers, and
+61~bedrock layers are used.
+
+A~pseudo-plastic sliding law \citep{Bueler.Pelt.2014} relates the
+bed-parallel shear stress to the sliding velocity. The yield stress, $\tau_c$,
+is modelled using the Mohr--Coulomb criterion,
+\begin{equation}
+   \tau_c = c_0 + N\,\tan{\phi} \,,
+\end{equation}
+where cohesion, $c_0$, is assumed to be zero. The friction angle, $\phi$,
+varies from 15 to 45{\degree}\renote{
+    Different from the previous paper, but fairly similar in effect (see
+    $\alpha$ and $\delta$ parameter values and corresponding equations in the
+    kappa, basal sliding section). These values are is based on a quick
+    sensitivity test on Greenland.}.
+It is taken to be a~function of modern bed elevation, with lower values at
+lower elevations (<200\,m above sea level), thus accounting for a~weakening of
+till associated with the presence of marine sediments. Effective pressure, $N$,
+is related to the ice overburden stress, $\rho gh$, and the modelled amount of
+subglacial water, using a formula derived from laboratory experiments with till
+extracted from an Antarctic ice stream \citep{Tulaczyk.etal.2000,
+Bueler.Pelt.2014},
+\begin{equation}
+    N = \delta \rho gh \, 10^{(e_0/C_c) (1 - (W/W_{max}))} \,,
+\end{equation}
+where $delta$ is choosed as 0.95, $e_0$ is a measured reference void ratio and
+$C_c$ a measured compressibility coefficient. The amount of water at the base,
+$W$, varies from zero to $W_{max}=2$\,m, a threshold above which instantaneous
+drainage is assumed\renote{
+    Regarding a parameter table, it is in the kappa, I am skipping it for now
+    but will try to include it soon.}.
+Finally, the bedrock topography responds to ice load\renote{
+    I disagree with writing ``the isostatic response''. The main point in
+    using this model is that computes a non-isostatic response.}
 following a bedrock deformation model that includes point-wise isostasy,
-elastic lithosphere flexure and viscous mantle deformation in a~semi-infinite
-half-space \citep{Lingle.Clark.1985,Bueler.etal.2007}.
-
-\subsection{Surface mass-balance}
-
+elastic lithosphere flexure and viscous mantle deformation in the lower
+half-space\renote{
+    I had put an emphasis on semi-infinite at this is the key assumption of
+    that model, allowing effective solution in Fourier domain. This is the
+    whole point of Ed's paper and probably the reason why he picked it for PISM
+    But yes, I guess that ``lower half-space'' is fine, too...}
+\citep{Lingle.Clark.1985,Bueler.etal.2007}.
+
+\subsection{Surface mass balance}
+
+\renote{
+    Surface mass balance is the balance of mass fluxes at the surface,
+    consisting of ice accumulation and ice ablation, which I introduce straight
+    away. I don't know what kind of ``opening sentence'' can be added here.}
 Ice surface accumulation and ablation are computed from monthly mean
-near-surface air temperature, monthly precipitation, and monthly standard
-deviation of near-surface temperature by a~temperature-index model
-\citep[e.g.,][]{Hock.2003}. Ice accumulation is equal to precipitation
-when temperature is below 0\,\unit{{\degree}C}, and decreases to zero linearly
-with temperature between 0 and 2\,\unit{{\degree}C}. Ice ablation is computed
-from the number of positive degree-days, defined as the integral of
-temperatures above 0\,\unit{{\degree}C} in one year.
-
-The positive degree-day integral \citep[Eqn.~6]{Calov.Greve.2005} is
-numerically
-approximated using week-long sub-intervals. It accounts for temperature
-variability assuming a~normal distribution along a~central (input) value. The
-temperature standard deviation is part of the forcing climatology. It was
-computed from daily temperature values from the North American Regional
-Reanalysis \citep[NARR,][]{Mesinger.etal.2006}, in order to account for the
-effects of spatial and seasonal variations of temperature variability
-\citep[Fig.~\ref{fig:atm};][]{Seguinot.2013}, after excluding variability
-associated with the seasonal cycle itself \citep[cf.][]
-{Seguinot.Rogozhina.2014}. The ablation model incorporates degree-day factors
-of 3.04\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for snow and
+near-surface air temperature $T_m$, monthly standard deviation of near-surface
+air temperature $\sigma$, and monthly precipitation $P_m$, using
+a~temperature-index model \citep[e.g.,][]{Hock.2003}. Accumulation is equal to
+precipitation when air temperatures are below 0\,\unit{{\degree}C}, and
+decreases to zero linearly with temperatures between 0 and
+2\,\unit{{\degree}C}. Ablation is computed from the number of positive
+degree-days (PDD), defined as the integral of temperatures above
+0\,\unit{{\degree}C} in one year.
+
+The PDD computation accounts for stochastic temperature variations by assuming
+a normal temperature distribution of standard deviation $\sigma$ aroung the
+expected value $T_m$. It is expressed by an error-function formulation
+\citep{Calov.Greve.2005},
+\begin{equation}
+    \label{eqn:calovgreve}
+    \mathrm{PDD} = \int_{t_1}^{t_2} \mathrm{d}t
+        \left[\frac{\sigma}{\sqrt{2\pi}}
+                \exp\left({-\frac{T_{m}^2}{2\sigma^2}}\right)
+              + \frac{T_{m}}{2} \, \mathrm{erfc}
+                \left(-\frac{T_{m}}{\sqrt{2}\sigma}\right)\right] \,,
+\end{equation}
+which is numerically approximated using week-long sub-intervals. In order to
+account for the effects of spatial and seasonal variations of temperature
+variability \citep{Seguinot.2013}, $\sigma$ is computed from daily temperature
+values from the North American Regional Reanalysis
+\citep[NARR,][]{Mesinger.etal.2006}, after excluding variability associated
+with the seasonal cycle itself \citep[cf.][]{Seguinot.Rogozhina.2014}\renote{
+    I may add a figure in the kappa if I find time, but I don't think that we
+    should do it here. The subject is too marginal for the present study.}.
+The ablation model incorporates degree-day factors of
+3.04\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for snow and
 4.59\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for ice, as derived from
@@
 \subsection{Atmospheric forcing}
-
-Atmospheric forcing of the model consists of a present-day monthly climatology
-$\{T_{m0}, P_{m0}\}$, modified by a lapse-rate correction ${\Delta}T_{LR}$ and
-temperature-offset time series ${\Delta}T_{TS}$.
+\label{sec:atm}
+
+Atmospheric forcing of the model consists of a present-day monthly climatology,
+$\{T_{m0}, P_{m0}\}$, where temperatures are modified by offset time series,
+${\Delta}T_{TS}$, and lapse-rate corrections, ${\Delta}T_{LR}$:
 \begin{align}
-    T_m(t, x, y) &= T_{m0}(x, y) + {\Delta}T_{LR}(t) + {\Delta}T_{TS}(t, x, y) \\
-    P_m(t, x, y) &= P_{m0}(x, y)
+    T_m(t, x, y) &= T_{m0}(x, y) + {\Delta}T_{TS}(t)
+                    + {\Delta}T_{LR}(t, x, y) \,, \\
+    P_m(t, x, y) &= P_{m0}(x, y) \,.
 \end{align}
 
-The present-day climatology $\{T_{m0}, P_{m0}\}$ was computed from
+The present-day monthly climatology, $\{T_{m0}, P_{m0}\}$, was computed from
 near-surface air temperature and precipitation rate fields of the NARR
-\citep{Mesinger.etal.2006} over the period 1979--2000. Modern climate of the
-North American Cordillera is characterised by strong geographic variations of
-seasonality, timing of the precipitation peak and daily temperature variability
-(Fig.~\ref{fig:atm}).
+\citep{Mesinger.etal.2006}, averaged between 1979 and 2000. Modern climate of the
+North American Cordillera is characterised by strong geographic variations in
+temperature seasonality, timing of the maximum annual precipitation, and
+daily temperature variability (Fig.~\ref{fig:atm}).
 Our choice of data from the NARR is motivated by the need for an accurate,
@@
 
-The lapse-rate correction ${\Delta}T_{LR}$ is computed using the NARR surface
-geopotential height invariant field as a reference topography $z_{ref}$,
-\begin{align}
-    {\Delta}T_{LR}(t, x, y) &= -\gamma [z_{s}(t, x, y)-z_{ref}] \\
-                            &= -\gamma [h(t, x, y)+z_{b}(t, x, y)-z_{ref}],
-\end{align}
-
-thus accounting for the evolution of ice thickness $h$ on the one hand, and
-for differences between the the ice flow model basal topography $z_{b}$ and the
-NARR topography $z_{ref}$ on the other hand. All simulations use an annual temperature lapse rate of $\gamma = 6\,\unit{{\degree}C\,km^{-1}}$.
-
-Temperature offset time-series ${\Delta}T_{TS}$ are derived from proxy records from
+Temperature offset time-series, ${\Delta}T_{TS}$, are derived from
+palaeo-temperature proxy records from
 the Greenland Ice Core Project \citep[GRIP,][]{Dansgaard.etal.1993}, the
@@
 the European Project for Ice Coring in Antarctica \citep[EPICA,][]
-{Jouzel.etal.2007}, the Lake Vostok ice core \citep{Petit.etal.1999}, and Ocean
+{Jouzel.etal.2007}, the Vostok ice core \citep{Petit.etal.1999}, and Ocean
 Drilling Program (ODP) sites 1012 and 1020, both located off the Californian
-shore \citep{Herbert.etal.2001}. Palaeo-temperatures from the GRIP and NGRIP
-records were calculated using a quadratic equation \citep{Johnsen.etal.1995},
+coast \citep{Herbert.etal.2001}. Palaeo-temperatures from the GRIP and NGRIP
+records were calculated from oxygen isotope (\chem{\delta^{18}O}) measurements
+using a quadratic equation \citep{Johnsen.etal.1995},
 \begin{equation}
-    {\Delta}T_{TS}(t) = -11.88[\chem{\delta^{18}O(t)}-\chem{\delta^{18}O}(0)]
-                        -0.1925[\chem{\delta^{18}O(t)}^2-\chem{\delta^{18}O}(0)^2],
+    {\Delta}T_{TS}(t) = -11.88 [\chem{\delta^{18}O}(t)
+                               -\chem{\delta^{18}O}(0)]
+                        -0.1925[\chem{\delta^{18}O}(t)^2
+                                -\chem{\delta^{18}O}(0)^2] \,,
 \end{equation}
-while the temperature reconstructions from the Antarctic and oceanic cores were
+while temperature reconstructions from Antarctic and oceanic cores were
 provided as such. All records were scaled linearly in
-order to simulate realistic and comparable ice extents at the last
+order to simulate realistic and comparable ice limits at the last
 glacial maximum (Table~\ref{tab:records}, Fig.~\ref{fig:timeseries}).
-
-For all proxy records, we firstly run a low-resolution simulation, using
-a~10\,km, 300 by 150~points horizontal grid for SSA and SIA velocities,
-up to 51~enthalpy layers in the ice, and 31~temperature layers in the bedrock.
-For the GRIP and EPICA records, we then run a second simulation using a higher
-horizontal resolution of 6\,km, up to 101~ice layers, and 61~bedrock
-layers. In the rest of this paper, we occasionally refer to different model
-through by the proxy record used for palaeo-temperature forcing.
+In the rest of this paper, we occasionally refer to different model
+runs by the name of the proxy record used for the palaeo-temperature forcing.
+
+Finally, lapse-rate corrections, ${\Delta}T_{LR}$, are computed as a function
+of ice surface elevation, $s$, using the NARR surface geopotential height
+invariant field as a reference topography, $b_{ref}$, by
+\begin{align}
+    {\Delta}T_{LR}(t, x, y) &= -\gamma [s(t, x, y)-b_{ref}] \\
+                            &= -\gamma [h(t, x, y)+b(t, x, y)-b_{ref}],
+\end{align}
+
+thus accounting for the evolution of ice thickness, ${h=s-b}$\renote{
+    Small change in notation to match with the kappa},
+on the one hand, and
+for differences between the the ice flow model basal topography, $b$, and the
+NARR reference topography, $b_{ref}$, on the other hand. All simulations use an
+annual temperature lapse rate of $\gamma = 6\,\unit{{\degree}C\,km^{-1}}$.
 
@@
 
-Despite large differences in input temperature offset time-series, the model
-output presents consistent features that can be observed across the range of
-forcing data used. In all simulations, total ice volume is relatively low
-during most of the glacial cycle, while two major glaciation events occur
-during Marine Isotope Stage (MIS)~4 (61.9--55.4\,kyr) and MIS~2
-(29.5--16.9\,kyr, Fig.~\ref{fig:timeseries}). A local ice minimum is
-consistently attained during MIS~3 (52.2--45.6\,kyr). However, the
-magnitude and timing of these three events depend significantly on the choice
-of proxy record used for palaeo-temperature forcing.
-
-Simulations forced by the Greenland ice core records (GRIP, NGRIP) produce
-highest variability in modelled ice volume throughout the glaciation history.
-Model runs using oceanic (ODP~1012, ODP~1020) and Antarctic (EPICA, Vostok)
-records generally result in lower
-modelled ice volume during MIS~4 and larger ice volume during MIS~3, thus
-suggesting lower ice volume variability throughout the simulation length. The
-NGRIP palaeo-temperature forcing is the only one that results in larger
-glaciation during MIS~4 than during the last glacial maximum (MIS~2).
-
-While simulations driven by the Antarctic and GRIP records
-reach a last glacial maximum in ice volume between 19.5 and 16.9\,kyr,
-those steered by oceanic and the NGRIP records attain it
-several thousands of years earlier. Moreover, the ODP~1012 run yields rapid
-deglaciation prior to 20\,kyr due to a local temperature maximum that is
-inconsistent with all other records used. The ODP~1020 simulation predicts an
-early last glacial maximum in ice volume at 29.5\,kyr, followed by
-slower deglaciation than modelled using other records. Finally, model runs
-informed by Antarctic ice core records produce a continuous, rapid deglaciation
-after the last glacial maximum, whereas the simulations driven by the GRIP
-record produce a rapid deglaciation in three steps, including stand-stills and
-readvances of the ice margin.
-
-Despite the different timing, snapshots of model output show relatively
-consistent patterns of glaciation (Fig.~\ref{fig:snapshots}). As a result of
-different scaling factors applied to palaeo-temperature proxy records,
-modelled ice sheet geometries at the last glacial maximum appear very similar
-from one simulation to the next. They include a central divide located at about
-3\,500\,m elevation along the spine of the Rocky Mountains. In all
-simulations, a central ice cap persists over the Skeena mountains during MIS~3
-between the two glaciation events. However, the size of this ice cap, as well
-as the magnitude of the MIS~4 glaciation, depend sensitively on the choice of
-palaeo-temperature proxy record used to drive the model.
+\todo{Add table with MIS~2, 3, and 4 volume and area table.}
+
+Despite large differences in the input atmospheric forcing (temperature offset
+time-series; Fig.~\ref{fig:timeseries}), model output presents consistent
+features that can be observed across the range of forcing data used. In all
+simulations, modelled ice volumes remain relatively low during most of the
+glacial cycle, except during two major glacial events which occur between 61.9
+and 55.4\,ka during Marine Oxygen Isotope Stage (MIS)~4, and between 29.5 and
+16.9\,ka during MIS~2 (Fig.~\ref{fig:timeseries}). An ice volume minimum is
+consistently attained between 52.2 and 45.6\,ka during MIS~3. However, the
+magnitude and precise timing of these three events depend significantly on the
+choice of proxy record used for time-dependent atmospheric forcing.
+
+Simulations forced by the Greenland ice core palaeo-temperature\renote{
+    Throughout this section, I prepended ``palaeo-temperature'' to
+    ``record(s)'' in place of ``climate'' as you originally suggested. If we
+    are to precise what type ``records'' we use, let's be clear about it.}
+records (GRIP, NGRIP) produce the highest variability in modelled ice volume
+throughout the last glacial cycle. In contrast, simulations driven by oceanic
+(ODP~1012, ODP~1020) and Antarctic (EPICA, Vostok) palaeo-temperature records
+generally result in smaller\renote{
+    Until we run out of other ajectives, I don't think that ``modest'' and
+    ``generous'' are adequate terms for describing ice volumes.}
+modelled ice volumes during MIS~4 and larger ice volumes during MIS~3, thus
+indicating. More broadly speaking, they produce lower ice volume variability
+throughout the simulation length. The NGRIP atmospheric forcing is the only one
+that results in a larger ice volume during MIS~4 than during the last glacial
+maximum (MIS~2).
+
+While simulations driven by GRIP and the two Antarctic palaeo-temperature
+records attain a last ice volume maximum between 19.5 and 16.9\,ka, those
+informed by NGRIP and the two oceanic palaeo-temperature records attain their
+maximum ice volume thousands of years earlier. Moreover, the ODP~1012 run
+yields a rapid deglaciation of the modelled area prior to 20\,ka. The ODP~1020
+simulation predicts an early maximum in ice volume at 29.5\,ka, followed by
+slower deglaciation than modelled using other palaeo-temperature records.
+Finally, model runs forced by Antarctic palaeo-temperature records produce a
+rapid an uninterrupted deglaciation after the last glacial maximum, whereas the
+simulation driven by the GRIP palaeo-temperature record also yields a rapid
+deglaciation but in three steps, interrupted by two stand-stills and readvances
+of the ice margin.
+
+Despite such differences in the timing of attained volume maxima and mimima,
+snapshots of model output show relatively consistent patterns of glaciation
+(Fig.~\ref{fig:snapshots}). As a result of scaling factors applied to the
+different palaeo-temperature proxy records (Table~\ref{tab:records}), modelled
+ice sheet geometries during the last glacial maximum (MIS~2;
+Fig.~\ref{fig:snapshots}, lower panels) appear highly similar from one
+simulation to the next. They include a ca. 1\,500\,km-long central divide
+located about 3\,500\,m\ above sea level (a.s.l.) along the spine of the Rocky
+Mountains. A central ice cap persists in all simulations over the Skeena
+Mountains during the MIS~3 ice volume minima (Fig.~\ref{fig:snapshots}, middle
+panels). However, the size of the Skeena Mountains ice cap, and the dimensions
+of the Cordilleran ice sheet during the MIS~4 glaciation, depend sensitively on
+the choice of palaeo-temperature record used to drive the model.
 
@@
 \subsection{Spatial imprint of ice cover}
+
+\todo{Add some point, refer back to Fig.~\ref{fig:atm}.}
 
@@
 \begin{table*}[t]
-  \caption{Palaeo-temperature proxy records and scaling parameters used to
-           prepare temperature offset time-series used to force the ice sheet
-           model through the last 120\,kyr. $f$ corresponds to the scaling
-           factor used and $T_{[32;22]}$ refers to the mean temperature anomaly
-           during the period -32 to~-22~kyr after scaling.}
+  \caption{Palaeo-temperature proxy records and scaling parameters yielding
+           temperature offset time-series used to force the ice sheet model
+           through the last glacial cycle (Fig.~\ref{fig:timeseries}). $f$
+           corresponds to the scaling factor adopted to yield last glacial
+           maximum ice limits in the vicinity of mapped end moraines, and
+           $T_{[32, 22]}$ refers to the resulting mean temperature anomaly
+           during the period -32 to~-22~ka after scaling.}
   \label{tab:records}
@@
     \tophline
-    Record & Lat. & Lon. & Elev. & Proxy & $f$ & $T_{[32;22]}$
-           & Original reference\\
+    Record & Latitude & Longitude & Elev. & Proxy & $f$ & $T_{[32, 22]}$
+           & Reference\\
+    & & & (m\,a.s.l) & & & (K) & \\
     \middlehline
@@
            St.-Elias mountains (WSE), the Selwyn and MacKenzie mountains (SMK),
-           the Skeena Mountains, the Coast Mountains (CM), the Rocky Mountains
-           (RM) and the North Cascades (NC). The background
+           the Skeena Mountains (SM), the Coast Mountains (CM), the Rocky
+           Mountains (RM) and the North Cascades (NC). The background
            map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
@@
            and temperature variability over the model domain, notably between
-           the maritime and continental regions.}
+           the maritime and continental regions.
+           \todo{Dash boundaries between ice sheets?}}
   \label{fig:atm}
@@
            Fig.~\ref{fig:timeseries}. Note the occurence of spatial similarities
-           despite large differences in timing.}
+           despite large differences in timing.
+           \todo{Perhaps indicate the location of the Skeena Mountains.}}
   \label{fig:snapshots}

### 2014-07-15 cordillera-cycle.tex
Add table with modelled ice volumes and extents.

@@
 
-\todo{Add table with MIS~2, 3, and 4 volume and area table.}
-
 Despite large differences in the input atmospheric forcing (temperature offset
@@
 magnitude and precise timing of these three events depend significantly on the
-choice of proxy record used for time-dependent atmospheric forcing.
+choice of proxy record used for time-dependent atmospheric forcing
+(Table~\ref{tab:records})\renote{
+    Hopefully this new table will assess several of your comments at once.}.
 
@@
     ``generous'' are adequate terms for describing ice volumes.}
-modelled ice volumes during MIS~4 and larger ice volumes during MIS~3, thus
-indicating. More broadly speaking, they produce lower ice volume variability
+modelled ice volumes during MIS~4 and larger ice volumes during MIS~3.
+More broadly speaking, they produce lower ice volume variability
 throughout the simulation length. The NGRIP atmospheric forcing is the only one
@@
 located about 3\,500\,m\ above sea level (a.s.l.) along the spine of the Rocky
-Mountains. A central ice cap persists in all simulations over the Skeena
+Mountains. The volumes of these modelled last glacial maximum ice sheets are
+contained within a close range from 8.24 to 8.66\,m of sea-level equivalent
+(s.l.e; Table~\ref{tab:records}).
+A central ice cap persists in all simulations over the Skeena
 Mountains during the MIS~3 ice volume minima (Fig.~\ref{fig:snapshots}, middle
@@
   \label{tab:records}
-  {\begin{tabular}{lccccccl}
+  {\begin{tabular}{l|ccc|ccc|l}
     \tophline
@@
            & Reference\\
-    & & & (m\,a.s.l) & & & (K) & \\
+    & & & (m~a.s.l) & & & (K) & \\
     \middlehline
@@
              & \citet{Herbert.etal.2001} \\
+    \bottomhline
+  \end{tabular}}
+  \belowtable{}
+\end{table*}
+
+% tab:records
+\begin{table*}[t]
+  \caption{Extrema of ice volume and extent corresponding to MIS~4, 3 and 2 for
+           each of the low-resolution simulations (Fig.~\ref{fig:timeseries}).}
+  \label{tab:records}
+  {\begin{tabular}{l*{3}{|ccc}}
+    \tophline
+             & \multicolumn{3}{c}{Age (ka)}
+             & \multicolumn{3}{c}{Ice extent (\unit{10^6\,km^2})}
+             & \multicolumn{3}{c}{Ice volume (m~s.l.e.)} \\
+    Record   & MIS.~4 & MIS.~3 & MIS.~2
+             & MIS.~4 & MIS.~3 & MIS.~2
+             & MIS.~4 & MIS.~3 & MIS.~2 \\
+    \middlehline
+    GRIP     & -57.58 & -49.24 & -19.52
+             &   1.98 &   0.46 &   2.13
+             &   7.52 &   0.89 &   8.52 \\
+    NGRIP    & -60.26 & -50.16 & -22.85
+             &   2.16 &   0.50 &   2.09
+             &   8.69 &   0.93 &   8.24 \\
+    EPICA    & -61.87 & -45.57 & -17.10
+             &   1.57 &   0.95 &   2.08
+             &   5.20 &   2.44 &   8.35 \\
+    Vostok   & -60.87 & -49.68 & -16.87
+             &   1.55 &   0.86 &   2.14
+             &   5.10 &   2.01 &   8.66 \\
+    ODP 1012 & -55.41 & -47.08 & -23.21
+             &   1.44 &   0.85 &   2.13
+             &   4.50 &   2.06 &   8.46 \\
+    ODP 1020 & -60.16 & -52.24 & -29.46
+             &   1.32 &   0.70 &   2.08
+             &   3.88 &   1.52 &   8.32 \\
+    \middlehline
+    Minimum  & -61.87 & -52.24 & -29.46
+             &   1.32 &   0.46 &   2.08
+             &   3.88 &   0.89 &   8.24 \\
+    Maximum  & -55.41 & -45.57 & -16.87
+             &   2.16 &   0.95 &   2.14
+             &   8.69 &   2.44 &   8.66 \\
     \bottomhline

### 2014-07-15 figures/timeseries.py

@@ -15,10 +15,11 @@ ax2 = fig.add_axes([10/figw, 10/figh, 107.5/figw, 32.5/figh])
 mis_idces = np.zeros((len(records), 3), dtype=int)
 mis_times = np.zeros((len(records), 3), dtype=float)
 mis_ivols = np.zeros((len(records), 3), dtype=float)
+mis_iareas = np.zeros((len(records), 3), dtype=float)
 
 # loop on records[i]
+tabline = ' '*4 + '%-8s '+ '& %6.2f '*3
 for i, rec in enumerate(records):
-    print 'plotting %s time series...' % rec
     dt = offsets[i]
 
     # get MIS times
@@ -33,9 +34,17 @@ for i, rec in enumerate(records):
     # load output time series
     nc = Dataset(run_path % ('10km', rec, dt*100) + '-ts.nc')
     ts_time = nc.variables['time'][:]*s2ka
+    ts_iarea = nc.variables['iarea'][:]/1e12
     ts_ivol = nc.variables['slvol'][:]
     nc.close()
+    mis_iareas[i] = ts_iarea[mis_idces[i]]
     mis_ivols[i] = ts_ivol[mis_idces[i]]
+
+    # print info in table style
+    tabline = ' '*4 + '%-8s '+ '& %6.2f '*3
+    print tabline % ( (labels[i],) + tuple(mis_times[i]) )
+    print tabline % ( ('',) + tuple(mis_iareas[i]) )
+    print tabline % ( ('',) + tuple(mis_ivols[i]) ) + '\\\\'
 
     # plot time series
     ax1.plot(dt_time, dt_temp, color=colors[i])
@@ -53,6 +62,7 @@ for i, rec in enumerate(records):
     except RuntimeError:
         pass
 
+
 # mark true MIS stages
 # source: http://www.lorraine-lisiecki.com/LR04_MISboundaries.txt
 ax2.axvspan(-71, -57, fc='0.95', lw=0.25)
@@ -68,13 +78,21 @@ mistmin = mis_times.min(axis=0)
 mistmax = mis_times.max(axis=0)
 misvmin = mis_ivols.min(axis=0)
 misvmax = mis_ivols.max(axis=0)
-print misvmin, misvmax
+misamin = mis_iareas.min(axis=0)
+misamax = mis_iareas.max(axis=0)
 for i in range(3):
[…]

### 2014-07-16 kappa.tex
Add corrections from Arjen.

@@
 initiated the Cordilleran ice sheet modelling project; and Q.~Zhang prepared
-some of the forcing data. All authors contributed to improve the text.
+some of the forcing data. All co-authors contributed to improve the text.
   \item[II] I analysed the weather data, wrote the positive degree day model,
@@
 
-\noindent\textbf{\CCLI} --- \bibentry{Seguinot.etal.2014}
+\noindent\textbf{\CCLI} --- \bibentry{Seguinot.etal.2014}.
 \bigskip
 
-As a first step towards simulations of the Cordilleran ice sheet through the
-last glacial cycle, I have chosen to run the ice sheet model under a constant
-climate, targeting the LGM as the period for which ice sheet limits are best
-known. These simulations, presented in \CCLI, use constant temperature offsets
-from the present climate state. They start from ice-free conditions and run for
-10\,ka, a period chosen as representative of the ice-sheet build-up from
-assumed, negligible ice cover. Although this assumption is disproved in \CCYC,
-the simplistic set-up used in \CCLI\ allowed to test model sensitivity to the
-choice of present-day climate forcing data, without involving too many
-additional parameters.
-
-Discrepancies between input data used (\cref{tab:datasources}) caused much
-variability in modelled ice sheet geometries (\CCLI, Figs.~5--7). In
-particular, continental parts of the ice-sheet are strongly dependent on input
-precipitation patterns, which are often poorly resolved by atmospheric
-reanalysis fields used to force the model (\CCLI, Figs.~10--14). The closest
-match between model output and the reconstructed ice margin \citep{Dyke.2004}
-was obtained using atmospheric forcing from the North American Regional
-Reanalysis \citep[NARR,][]{Mesinger.etal.2006} and a temperature depression of
-7\,\degC\ (\CCLI, Fig.~15). The relevant simulation is reproduced here
-(\cref{fig:plot-lgm}), using a higher horizontal resolution of 3\,km, instead
-of 10\,km as in the paper. Following this sensitivity study, I have retained
-the NARR to force subsequent simulations of the Cordilleran ice sheet
-presented here and in \CCYC.
+As a first step towards a simulation of the Cordilleran ice sheet through the
+last glacial cycle, I have opted to run the ice sheet model under a range of
+constant temperature depressions, targeting the LGM as the period for which ice
+sheet limits are best known. These simulations, presented in \CCLI and run at
+10\,km grid resolution, use constant temperature depressions from the present
+climate state (using five different input climatologies). All model runs start
+from ice free conditions and run for 10\,ka, a period chosen as representative
+of Cordilleran ice sheet build-up from assumed, negligible ice cover. Although
+this assumption appears unjustified from results in \CCYC, the simplistic
+set-up used in \CCLI\ allowed an assessment of model sensitivity to the choice
+of input climatologies, without involving too many additional parameters.
+
+Discrepancies between the five input data sets used (\cref{tab:datasources})
+caused considerable variability in modelled ice sheet geometries (\CCLI,
+Figs.~5--7). In particular, ice growth in continental sectors of the modelling
+domain is strongly dependent on input precipitation patterns, which are often
+poorly resolved by atmospheric reanalysis fields used to force the model
+(\CCLI, Figs.~10--14). The closest match between modelled ice sheet extent and
+the reconstructed LGM ice margin \citep{Dyke.2004} was obtained using the North
+American Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006} climatologies
+and a temperature depression of 7\,K (\CCLI, Fig.~15). The relevant simulation
+is reproduced here, using a higher horizontal resolution of 3\,km
+(\cref{fig:plot-lgm}). Following this sensitivity study, I have retained the
+NARR climatology as input to simulations of the Cordilleran ice sheet presented
+here and in \CCYC.
 
@@
 % ----------------------------------------------------------------------
-\subsection{Daily temperature variability}
-% ----------------------------------------------------------------------
-
-\noindent\textbf{\PSDV} --- \bibentry{Seguinot.2013}
+\subsection{Importance of daily temperature variability for ice sheet modelling}
+% ----------------------------------------------------------------------
+
+\noindent\textbf{\PSDV} --- \bibentry{Seguinot.2013}.
 \bigskip
 
-\noindent\textbf{\PSDP} --- \bibentry{Seguinot.Rogozhina.2014}
+\noindent\textbf{\PSDP} --- \bibentry{Seguinot.Rogozhina.2014}.
 \bigskip
@@
 balance model \citep[e.g.,][]{Hebeler.etal.2008}. In our case, these parameters
-are the degree-day factors for snow, $F_s$, and ice, $F_i$, and the standard
-deviation of near-surface air temperature, $\sPDD$. Although the first two can
-only be deduced from laborious surface mass-balance measurements on present-day
-glaciers \citep[e.g.,][]{Shea.etal.2009}, $\sPDD$ can be estimated from
-temperature time-series \citep[e.g.,][]{Fausto.etal.2011}.
+are the degree-day factors for snow, $F_s$, and ice, $F_i$
+(\cref{fig:model-pdd, tab:surfaceparams}), and the standard deviation of
+near-surface air temperature, $\sPDD$ (\cref{eq:calovgreve,
+tab:surfaceparams}). Although the first two can only be deduced from surface
+mass-balance measurements on present-day glaciers
+\citep[e.g.,][]{Shea.etal.2009} and assumptions that these values are also
+valid for palaeo-simulations, $\sPDD$ can be estimated from temperature time
+series \citep[e.g.,][]{Fausto.etal.2011}.
 
 In \PSDV, I used near-surface air temperatures from the ERA-Interim reanalysis
-\citep{Dee.etal.2011} to compute a global, monthly distribution of $\sPDD$.
-The resulting grids show
-that, although often assumed constant, $\sPDD$ varies greatly by location and
-through the year, most particularly in the polar regions
-(\cref{fig:plot-sdmap}; \PSDV, Fig.~1). Using a simple surface mass balance
-model, I then demonstrated the effect of these variations on the computation of
-positive degree-days (\PSDV, Fig.~2), and surface mass-balance (\PSDV, Fig.~3).
-
-The analysis is taken further in \PSDP, using re-analysed air temperatures from
-ERA-40 \citep{Uppala.etal.2005}. Although of coarser spatial resolution, the
-ERA-40 time series have a longer time span than the ERA-Interim ones, which
-is necessary to compute standard deviation values that are representative of
-the long-term daily variability \citep{Rogozhina.Rau.2014}. In \PSDP, we
-demonstrated that, over the Greenland ice sheet, there exist an inverse
-correlation between $\sPDD$ and monthly mean temperatures, $T_m$
-(\PSDP, Fig.~1). This correlation could potentially be used as a
-parametrisation, and can be expressed as (\PSDP, Eq.~5)
+\citep{Dee.etal.2011} to compute a global, monthly distribution of $\sPDD$. The
+resulting grids show that, although often assumed constant in ice sheet
+modelling, $\sPDD$ exhibits significant spacial and temporal variation, most
+particularly in the polar regions (\cref{fig:plot-sdmap}; \PSDV, Fig.~1). Using
+a simplified surface mass balance model, I then demonstrated the effect of this
+variation on the computation of positive degree-days (\PSDV, Fig.~2), and
+surface mass-balance (\PSDV, Fig.~3). Neglecting this variability in ice sheet
+modelling introduces significant errors in modelled PDD and surface mass
+balance calculations. The transient climate simulations presented in \CCYC\
+therefore now include this effect.
+
+The analysis is taken further extended in \PSDP, using re-analysed air
+temperatures from ERA-40 \citep{Uppala.etal.2005}. Although of coarser spatial
+resolution than the ERA-Interim reanalysis used in \PSDV, the ERA-40 time
+series span over longer time, which is necessary to compute standard deviation
+values that are representative of the long-term daily variability
+\citep{Rogozhina.Rau.2014}. In \PSDP, we demonstrated an inverse correlation
+between $\sPDD$ and monthly mean temperatures, $T_m$, over the Greenland ice
+sheet (\PSDP, Fig.~1). This correlation could potentially be used as a
+parametrisation, and can be expressed as (\PSDP, Eq.~3)
 
@@
 these two studies, a natural step forward was to implement, in PISM, spatially
-and seasonally variable $\sPDD$ input (PISM issue \issue{179}), and the
-parametrisation expressed by \cref{eqn:sigma} over ice-filled grid cells
-(\issue{265}). Their effect on the modelled Cordilleran ice sheet surface is
-demonstrated here (\cref{fig:plot-sdeffect}) using a constant-climate set-up
-similar to the one used in \CCLI.
+and seasonally variable $\sPDD$ input (PISM issue~\issue{179}), and the
+parameterisation expressed by \cref{eqn:sigma} over ice-covered grid cells
+(PISM issue~\issue{265}). Their effect on modelled Cordilleran ice sheet
+surface elevations is demonstrated here using a constant climate set-up similar
+to the one used in \CCLI\ (\cref{fig:plot-sdeffect}).
+
+\todo{You need to insert a couple of sentences of what you learn from Figure
+      16!}
 
@@
            \citep{Uppala.etal.2005}. Note the strong differences between
-           continental and maritime regions, notably within the model domain
+           continental and maritime regions, notably within the modelling domain
            (black rectangle). Although daily variability is lowest in summer,
-           this is when it most importantly contributes to surface melt.}
+           this is when it most importantly contributes to surface melt and,
+           therefore, net mass balance.}
   \label{fig:plot-sdmap}
@@
 
-\noindent\textbf{\CCYC} --- \bibentry{Seguinot.etal.inpreparation}
+\noindent\textbf{\CCYC} --- \bibentry{Seguinot.etal.inpreparation}.
 \bigskip
 
-From the beginning, the primary goal of this thesis has been to perform
-simulations of the Cordilleran ice sheet through the last glacial cycle. This
-is the topic of \CCYC, where monthly air temperature, precipitation, and PDD
-standard deviation climatologies from the NARR are amended by
-palaeo-temperature time-series from six proxy records to produce a
-time-dependent climate forcing (\CCYC, Fig.~2). The resulting simulations all
-produce two major glaciations between MISs~4 and 2. However the differences in
-timing and magnitude of these events is considerable (\CCYC, Fig.~3--4).
-
-Geological evidence regarding the timing of glaciation allowed us to select two
-simulations, driven by the GRIP \citep{Dansgaard.etal.1993} and EPICA
-\citep{Jouzel.etal.2007} ice-core records, as producing more realistic
-reconstructions of the last glacial cycle. I have re-run these two simulations
-at a higher resolution of 6\,km, to allow for a more detailed study of
-transient ice cap cover before and after the LGM, stability of major
-ice-dispersal centres, and patterns of deglaciation (\CCYC, Fig.~5--15).
-Modelled ice volume evolution and snapshots of ice flow patterns from one of
-these two simulations are reproduced here (\cref{fig:plot-snapshots}).
+The primary goal of this thesis has been to perform simulations of the
+Cordilleran ice sheet through the last glacial cycle. This is the topic of
+\CCYC, where monthly air temperature, precipitation, and PDD standard deviation
+climatologies from the NARR are combined with palaeo-temperature time-series
+from six proxy records to produce a time-dependent climate forcing (\CCYC,
+Fig.~2). The resulting simulations, using horizontal resolutions of 6 and
+10\,km, all produced two major glaciations during the last glacial cycle, at
+MIS~4 and MIS~2. However, the differences in timing and magnitude of these
+glaciations are considerable (\CCYC, Figs.~3--4).
+
+Geological evidence for the timing and extent of maximum glaciation durng the
+LGM allowed me to select two simulations, driven by the GRIP
+\citep{Dansgaard.etal.1993} and EPICA \citep{Jouzel.etal.2007} ice core
+records, as producing the most realistic reconstructions of the last glacial
+cycle. I re-ran these two simulations at a higher resolution of 6\,km, in order
+to allow for a more detailed study of transient ice caps before and after the
+LGM, of the stability of major ice-dispersal centres, and of patterns and rates
+of deglaciation (\CCYC, Figs.~5--15). Modelled ice volume evolution (expressed
+as meters of sea level equivalent) and snapshots of ice extent and ice flow
+patterns for the GRIP simulation (\cref{fig:plot-snapshots}) shows...
+\todo{complete}.
 
@@
   \caption{\textbf{(a)} Modelled growth and decay of the last Cordilleran ice
-           sheet. Basal topography surface topography (1\,km contours) and
-           surface velocity (m/a).
-           \textbf{(b)} Evolution of total ice volume during the entire last
-           glacial cycle, expressed as meters of sea-level equivalent. The
-           simulation uses a horizontal resolution of 6\,km and is forced by
+           sheet. Basal topography, surface topography (1\,km contours) and
+           surface velocity (m/a) are model results.
+           \textbf{(b)} Evolution of total ice volume during the last glacial
+           cycle, expressed as metres of sea-level equivalent (m~s.l.e.). The
+           simulation is performed on a 6\,km horizontal grid and adopts
            atmospheric fields from the NARR \citep{Mesinger.etal.2006} and
-           temperature offset time-series derived from the GRIP
-           \citep{Dansgaard.etal.1993} ice-core record as described in \CCYC.}
+           temperature offset time-series derived from the GRIP ice-core record
+           \citep{Dansgaard.etal.1993} as described in \CCYC. Note that, for
+           this run, the Cordilleran ice sheet was almost as voluminous during
+           MIS~4 as it was during the LGM.}
   \label{fig:plot-snapshots}
@@
 In this thesis, I present numerical reconstructions of the former Cordilleran
-ice sheet to a level of detail that has, to my knowledge, never been attained
-before. The coupled SIA-SSA approximation is crude, yet it is today the only
-approach that allows, combined with massively parallel computing, high-spatial
-resolution over the multi-millenial time-scales of ice sheet growth and decay.
+ice sheet to a level of detail that has, to my knowledge, not been attained
+before. The coupled SIA-SSA approximation is crude, yet it is currently the
+only approach that allows, combined with massively parallel computing, high
+spatial resolution over the multi-millenial time scales of ice sheet growth and
+decay.
 
@@
 uncertainty, which I have assessed using a combination of detailed
-meteorological datasets and available geological evidence. The simulations
-supporting this thesis allowed estimates of the LGM Cordilleran ice sheet
-extent and volume, identification of the main ice-dispersal centres, and
-partial understanding of the deglaciation patterns. However, there remains much
-room for model improvements and further analysis of the simulation output.
+meteorological datasets and available geological evidence.\todo{insert a
+statement which these sources of uncertainties are} Simulations supporting this
+thesis yielded estimates of the LGM Cordilleran ice sheet extent and volume,
+identification of the main ice-dispersal centres througout the last glacial
+cycle, and understanding of final deglaciation patterns. However, there remains
+considerable room for model improvements and further analysis of the simulation
+output \todo{introduce your perspectives here}.
 
@@
 
-Throughout the thesis, precipitation rates are prescribed as monthly, spatial
-grids, but assumed equal to their present values. Although the distribution of
-precipitation has been undoubtedly affected by past changes in global climate
-and the intermittent presence of ice sheets and ice caps in the Northern
-Hemisphere, its changes are difficult to assess. In the North American
-Cordillera, much of the continental interior was likely dryer than today due to
-the presence of ice caps reinforcing the topographic barrier formed by the
-Pacific Coast ranges. However, some regions might have become rainier as a side
-effect of redirected winds.
+One of the predominant uncertainties, for which Cordilleran ice sheet modelling
+is highly sensitive, is the treatment of precipitation. Throughout the
+modelling, we base precipitation patterns on current patterns derived from
+reanalysis data. Although the distribution of precipitation has undoubtedly
+been affected by past changes in global climate and the intermittent presence
+of ice sheets and ice caps in the Northern Hemisphere, its changes are
+difficult to assess. In the North American Cordillera, much of the continental
+interior was likely dryer than today due to the presence of ice caps on the
+Pacific Coast ranges, reinforcing the current topographic barrier. However,
+some regions were probably rainier as a side effect of modified wind patterns.
 
 In meteorology, these effects are typically analysed using atmospheric
-circulation models. However, I have shown in \CCLI\ that atmospheric reanalyses
-which are circulation models that use data assimilation techniques, fail to
-agree on the present-day pattern of precipitation in the Cordillera. The
-discrepancies, due to highly mountainous terrain, have dramatic effects on the
-ice-sheet-model response. Considering additional difficulties inherent to
-modelling past climates, and the lack of data for assimilation, we can only
-expect that palaeo-climate reconstructions will produce even higher
-discrepancies in the ice-sheet-model response.
+circulation models. However, I have shown in \CCLI\ that different atmospheric
+reanalyses which are circulation models that use data assimilation techniques,
+fail to agree regarding the present-day pattern of precipitation over the
+modelling domain. These discrepancies, which arise as a consequence of the
+highly mountainous terrain, have dramatic effects on the ice sheet model
+response. Considering additional difficulties inherent to modelling past
+climates, and the lack of data for assimilation, we must regard derived ice
+sheet histories as a highly generalized model output [only expect that
+palaeo-climate reconstructions will produce even higher discrepancies in the
+ice-sheet-model response].
+
+\todo{think tomorrow. can one say something about the robustness of some
+      features (ice caps that form, spreading centers which persist)which occur
+      irrespective of which reanalysis is used?}.
 
@@
 
-Geothermal heat flux, controlling en-glacial temperature distribution and the
-location of basal melt areas, was assumed constant. However, the North America
-is an active subduction zone with complex geothermal patterns
-\citep{Blackwell.Richards.2004}. Including this variability in the ice-sheet
-model will offer new opportunities to study its output.
-
-Most notably, modelled basal flow could be compared with glacial lineations,
-one of the main sources of geomorphological evidence used to reconstruct former
-ice sheets. For instance, in our simulations, large portions of the Cordilleran
-ice sheet remain cold-based throughout deglaciation (\CCYC, Fig.~15). As
-cold-based glaciers typically do not deform their beds, this could indicate
-that some of the lineation patterns we observe in the North American Cordillera
-could pre-date deglaciation. Instead, they may have been formed during an
-earlier phase of advance or retreat of the ice sheet.
+Geothermal heat flux, contributing to the en-glacial temperature distribution
+and the location of basal melt areas, was assumed constant over the modelling
+grid. However, the geology of the study area consists of an amalgamation of
+former subduction zone, resulting in complex geothermal patterns
+\citep{Blackwell.Richards.2004}. I am now at the stage to start including this
+variability as a driver to the ice sheet model, which will offer new
+opportunities to study its output in terms of ice flow patterns, consistent
+basal freezing and basal melting patterns, regions prone for glacial erosion,
+and regions characterised by glacial deposition.
+
+I expect that inclusion of geothermal heat patterns allow modelled basal flow
+directions and distances to be compared with glacial lineations, which is one
+of the main sources of geomorphological evidence used to reconstruct former ice
+sheets \citep{Boulton.Clark.1990, Kleman.etal.1997, Kleman.etal.2010}. For
+instance, in our simulations, large portions of the Cordilleran ice sheet
+remain cold-based throughout deglaciation (\CCYC, Fig.~15). As cold-based
+glaciers typically do not impact their beds significantly, this could indicate
+that some of the lineation patterns we observe within these zones of consistent
+cold-based deglaciation could be older features, formed during an earlier phase
+of advance or retreat of the ice sheet. However, without a better control on
+the geothermal heat flux, such analysis are perhaps too premature.
 

### 2014-07-16 figures/plot-snapshots.py

@@ -72,7 +72,7 @@ figinvtrans = fig.transFigure.inverted()
 print 'plotting time series...'
 ax.plot(time, ivol, color='#0978ab')
 ax.set_xlabel('model time (kyr)')
-ax.set_ylabel('ice volume (m s.-l. eq.)')
+ax.set_ylabel('ice volume (m s.l.e.)')
 ax.yaxis.set_ticks_position('right')
 ax.yaxis.set_label_position('right')
 ax.grid(axis='y', c='0.5', ls='-', lw=0.1)

### 2014-07-16 cordillera-cycle.tex
Change result and discussion headers.
Implement minor comments on section 4.1.

@@
 
-\todo{Do we need a new subsection title here?}
+\noindent\todo{Do we need a new 2.1 subsection title here?}
 
@@
 % ----------------------------------------------------------------------
-\section{Results}
+\section{Sensitivity to atmospheric forcing time-series}
 \label{sec:results}
@@
 choice of proxy record used for time-dependent atmospheric forcing
-(Table~\ref{tab:records})\renote{
+(Table~\ref{tab:extrema})\renote{
     Hopefully this new table will assess several of your comments at once.}.
@@
 Simulations forced by the Greenland ice core palaeo-temperature\renote{
-    Throughout this section, I prepended ``palaeo-temperature'' to
-    ``record(s)'' in place of ``climate'' as you originally suggested. If we
-    are to precise what type ``records'' we use, let's be clear about it.}
+    From here on, I prepend ``palaeo-temperature'' to ``record(s)'' in place of
+    ``climate'' as you (Arjen) originally suggested. If we are to precise what
+    type of ``records'' we use, let's be clear about it.}
 records (GRIP, NGRIP) produce the highest variability in modelled ice volume
@@
 contained within a close range from 8.24 to 8.66\,m of sea-level equivalent
-(s.l.e; Table~\ref{tab:records}).
+(s.l.e; Table~\ref{tab:extrema}).
 A central ice cap persists in all simulations over the Skeena
@@
 % ----------------------------------------------------------------------
-\section{Discussion}
+\section[Comparison to the geologic record]
+        {Comparison to the geologic record\renote{
+    Arjen, you have proposed 2.1 Evaluation of model runs agains field
+    evidence, 2.2 Palaeoglaciology of the Cordilleran ice sheet, and 2.3 The
+    last deglaciation; but I don't think these headers work. In my vision of
+    the paper, this entire ``discussion'' part is all about evaluation of model
+    runs agains field evidence and palaeoglaciology. Because of this, and
+    because we also present new ``results'', or rather different views of the
+    same results, I propose to rename it to a more explicit ``Comparison to the
+    geologic record'' (or similar). I have also renamed the ``results'' to
+    ``Sensitivity to atmospheric forcing time-series'', which is actually what
+    that section is about: raw model output, no discussion against the geologic
+    record.}}
 \label{sec:discussion}
@@
 
-\subsection{Timing of the glaciation peaks}
-
-Independently of the choice of palaeo-temperature record used as forcing
-time-series for the ice sheet model, our simulations consistently produce two
-glaciation maxima during MIS~4 (61.9--55.4\,kyr) and MIS~2
-(29.5--16.9\,kyr). These events broadly correspond in timing to the
-Gladstone (MIS~4) and McConell (MIS~2) glaciations documented by geological
-evidence in the northern part of the model domain \aref, and to the Fraser
-(MIS~2) glaciation in the southern part \aref
-\footnote{\textbf{Arjen, Martin, Johan,} can you check that I use these names
-    correctly, and help me to fill the missing refs in this sentence? Is is
-    possible to trace back when these names where first applied?}.
-
-However, the timing of the modelled glacial maxima strongly depends on the
-choice of palaeo-temperature record used, and a more precise agreement with
-timing reconstructed from geological evidence is only attained in some of our
-simulations. In the Puget lowland, the last glacial maximum advance of the
-southern Cordilleran ice sheet margin has been constrained by radiocarbon
-dating between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,kyr\,BP}
-\citep[Fig.~4]{Porter.Swanson.1998}.
+\subsection{Timing of glaciation}
+
+Independently of the choice of palaeo-temperature record\renote{
+    In my view, ``atmospheric/climate forcing'' would be far too broad here. We
+    could possibly apply a much wilder choice of atmospheric forcing
+    techniques, leading just as wild differences in model output
+    \citep[e.g.,][but possibly much more than that]{Seguinot.etal.2014}.}
+used to force the ice sheet model, our simulations consistently produce two
+glacial maxima during the last glacial cycle. The first maximum configuration
+is obtained during MIS~4 (61.9--55.4\,ka) and the second during MIS~2
+(29.5--16.9\,ka). These events broadly correspond in timing to the Gladstone
+(MIS~4) and McConnell (MIS~2) glaciations documented by geological evidence for
+the northern sector of the Cordilleran ice sheet
+    \citep{Duk-Rodkin.etal.1996, Ward.etal.2007,
+           Stroeven.etal.2010, Stroeven.etal.2014},
+and to the Fraser Glaciation (MIS~2) documented for its southern sector
+    \citep{Porter.Swanson.1998, Margold.etal.inpress}.
+There is patchy stratigraphical evidence for glaciations older than the Fraser
+Glaciation \citep{Clague.Ward.2011} in British Columbia, and their extent and
+timing are therefore still highly conjectural
+    \citep[perhaps MIS~4 or early MIS~3; e.g.,][]{Cosma.etal.2008}.
+
+However, the exact timing of modelled MIS~2 maxima depend strongly on the
+choice of applied palaeo-temperature record, which allows for a more in-depth
+comparison with geological evidence for the timing of maximum Cordilleran ice
+sheet extent. In the Puget lowland (Fig.~\ref{fig:locmap}), the LGM advance of
+the southern Cordilleran ice sheet margin has been constrained by radiocarbon
+dating on wood between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,kyr\,BP}
+\citep{Porter.Swanson.1998}.
 These dates have been confirmed by the offshore sediment record, which shows an
 increase of glaciomarine sedimentation between 19.5 and
-16.2\,\unit{\chem{^{14}C}\,cal\,kyr\,BP} \citep{Cosma.etal.2008}.
-
-Among simulations presented here, only those forced by the GRIP, EPICA and
-Vostok records place the last glacial maximum within this range
-(Fig.~\ref{fig:timeseries}, lower panel). Simulations driven by the NGRIP,
-ODP~1012 and ODP~1020 records, on the contrary, produce a last glacial maximum
-that pre-dates reconstructed ages by several thousand years. Concerning the
-simulations driven by oceanic records, this early deglaciation is caused by an
-early warming sealed in the palaeo-temperature reconstructions
-(Fig.~\ref{fig:timeseries}, upper panel). However, it has been shown that this
+16.2\,\unit{\chem{^{14}C}\,cal\,kyr\,BP} \citep{Cosma.etal.2008}. Radiocarbon
+dating of the northern Cordilleran ice sheet margin is much less constrained
+but straddles presented constraints from the southern margin. However,
+cosmogenic exposure dating by \citep{Stroeven.etal.2010, Stroeven.etal.2014}
+places the timing of maximum CIS extent during the McConnell glaciation close
+to 17\,ka.
+
+Among the simulations presented here, only those forced with the GRIP, EPICA
+and Vostok palaeo-temperature records yield Cordilleran ice sheet maximum
+extent that may be compatible with these field constraints
+(Fig.~\ref{fig:timeseries}, lower panel; Fig.~\ref{fig:snapshots}, lowest
+panels). Simulations driven by the NGRIP, ODP~1012 and ODP~1020
+palaeo-temperature records, on the contrary, yield MIS~2 maximum Cordilleran
+ice sheet extents that pre-date field-based constraints by several thousands of
+years. Concerning the simulations driven by oceanic records, this early
+deglaciation is caused by an early warming present in the alkenone
+palaeo-temperature reconstructions (Fig.~\ref{fig:timeseries}, upper panel;
+\citealp[Fig.3]{Herbert.etal.2001}). However, this
 early warming is a local effect, corresponding to a weakening of the California
 current \citep[Fig.~3]{Herbert.etal.2001}. The California current, driving cold
-waters southwards
-along the western coast of North America, is thought to have weakened in the
-presence of the Cordilleran ice sheet, resulting in paradoxically warmer
-sea-surface temperatures at the locations of the ODP~1012 and ODP~1020 sites
-during the last glacial maximum \citep{Herbert.etal.2001}.
-
-Because most of the Cordilleran ice sheet marine front lies in a different
-oceanic circulation gyre than that holding the California current, it is
-unlikely that it was affected by this early warming. However, the above paradox
+waters southwards along the south-western coast of North America\renote{
+    Not only USA but also Baja California in Mexico.},
+has been shown to have weakened during each peak global glaciation (in SPECMAP)
+durint the past 550\,ka, resulting in paradoxically warmer sea-surface
+temperatures at the locations of the ODP~1012 and ODP~1020 sites also during
+the LGM \citep{Herbert.etal.2001}.
+
+Because most of the marine margine of the Cordilleran ice sheet terminated in a
+sector of the Pacific Ocean unaffected by variations in the California current,
+it probably remained unaffected by this early warming. However, the above paradox
 illustrates the complexity of ice-sheet feedbacks on regional climate, and
 demonstrates that, although located in the neighbourhood of the modelling
-domain, the ODP~1012 and ODP~1020 palaeo-temperature proxy records can not be
-used as a realistic forcing time-series to model the Cordilleran ice sheet
+domain, the ODP~1012 and ODP~1020 palaeo-temperature records cannot be
+used as a realistic forcing to model the Cordilleran ice sheet
 through the last glacial cycle.
 
-These results make us believe that a palaeo-temperature proxy record located
-in the direct vicinity of the former Cordilleran ice sheet would be of great
-help to further understand its glacial history. Because, to our knowledge, no
+\todo{say something about NGRIP.}
+
+... Hence, we focus the rest of our analysis on simulations forced by palaeo-temperature records from GRIP and EPICA ice cores (Fig.~\ref{fig:timeseries}, lower panel, dotted lines).
+
+\todo{avoid repetition with the next paragraph.}
+
+These results indicate that a palaeo-temperature proxy record located in the
+direct vicinity of the former Cordilleran ice sheet would potentially lead to
+improved understanding of the history of advance and retreat of the western ice
+sheet margin. Because, to our knowledge, no
 such record yet exists, we focus the rest of our analysis on simulations forced
-by temperature records from Greenland (GRIP) and Antarctic (EPICA) ice cores
+by palaeo-temperature records from GRIP and EPICA ice cores
 (Fig.~\ref{fig:timeseries}, lower panel, dotted lines).
 
-\subsection{Spatial imprint of ice cover}
+\subsection{Location and longevity of the major ice-despersal centres}
 
@@
 
-\subsection{Deglaciation history}
+\subsection{Pace and patterns of deglaciation}
 
@@
            each of the low-resolution simulations (Fig.~\ref{fig:timeseries}).}
-  \label{tab:records}
+  \label{tab:extrema}
   {\begin{tabular}{l*{3}{|ccc}}
@@
              & \multicolumn{3}{c}{Ice volume (m~s.l.e.)} \\
-    Record   & MIS.~4 & MIS.~3 & MIS.~2
-             & MIS.~4 & MIS.~3 & MIS.~2
-             & MIS.~4 & MIS.~3 & MIS.~2 \\
+    Record   &  MIS~4 &  MIS~3 &  MIS~2
+             &  MIS~4 &  MIS~3 &  MIS~2
+             &  MIS~4 &  MIS~3 &  MIS~2 \\
     \middlehline
@@
            Data \citep{Patterson.Kelso.2014}.
-           \todo{Mark palaeo-ice sheets and mountain ranges on the map.}}
+           \todo{Mark palaeo-ice sheets and mountain ranges on the map.
+                 Mark location of the Puget Lowland.}}
   \label{fig:locmap}

### 2014-07-16 cordillera-cycle.tex
Work in progress on section 4.2.

@@
 
+\todo{Add general statements on the scale of glaciation (but where?).}
+
+\todo{Put more equal weight to the northern sector}
+
 \todo{Add some point, refer back to Fig.~\ref{fig:atm}.}
 
-Although snapshots from the model output (Fig.~\ref{fig:snapshots}) give a
-crisp picture of modelled ice dynamics at specific time frames, they often
-can not be directly compared to geomorphological evidence, which is by nature
-time-transgressive. In this section, we use the 6-km horizontal-resolution
-model output from simulations driven by the GRIP and EPICA records to produce
-a series of time-integrated plots containing information aggregated over the
-whole simulation period.
+\todo{Mention the major nucleation centres. Growth of the CIS from the
+      highlands.}
+
+Although snapshots from the model output (Fig.~\ref{fig:snapshots}) yield a
+clear spatial picture of modelled glaciation patterns at specific time frames,
+they often can not be directly compared to geomorphological evidence, which is
+by nature time-transgressive. In this section, we use high-resolution
+simulations driven by the GRIP and EPICA records to locate the major nucleation
+centres of the Cordilleran ice sheet and assess their longevity during the last
+glacial cycle.
+
+\todo{Rewrite this opening paragraph. Fig.~\ref{fig:snapshots} can actually be
+      compared to geological evidence.}
 
@@
 ice caps centred on major mountain ranges of the North American Cordillera
-(Fig.~\ref{fig:duration}). In both simulations driven by the GRIP and EPICA
-records, a~contiguous ice cover spanning from the Alaska Range to the Rocky
-Mountains exists for about 29\,kyr in total over MIS~2 and~4.
-With the exception of the oceanic front and the northern foothills of the Alaska
-Range, most of the maximal extent of the ice sheet is associated with short
-durations of ice cover, generally a few thousand years, distributed over one
-(GRIP) or two (EPICA) glaciations (Fig.~\ref{fig:duration}). This illustrates
-that the most extensive stages of the modelled ice sheet are both short-lived
-and out of balance with contemporary climate states, which confirms previous
-inferences of a short-lived Cordilleran ice sheet based on the geological
-record \citep{Clague.etal.1980, Stroeven.etal.2010}.
-
-However, our simulations depict, independently of the choice of record used for
-palaeo-temperature forcing, a central Skeena mountain ice cap persisting
-throughout the entire last glacial cycle (Figs.~\ref{fig:snapshots}
-and~\ref{fig:duration}). In particular, this ice cap persists during MIS~3,
-providing a nucleation centre for the onset of glacial
-readvance towards the last glacial maximum (MIS~2).
-
-During time periods when a fully-grown Cordilleran ice sheet is absent from our
-simulation, the presence of the Skeena ice cap can be explained by a more
-widespread winter precipitation there than in other parts of the modelling
-domain (Fig.~\ref{fig:atm}). Along most of the north-western North American
-shore, coastal mountain ranges provide a sharp orographic barrier to westerly
-winds, capturing atmospheric moisture in the form of precipitation, and leaving
-interior lowlands arid. However, near the centre of our modelling domain, this
-barrier is less pronounced than
-elsewhere, allowing westerly winds to carry moisture further inland, until it
-is progressively captured by the broad and mildly elevated group of the Skeena
-Mountains.
-
-In addition, the erosional landscape of the Skeena mountain bears one of the
+(Fig.~\ref{fig:duration}). A 2\,500\,km-long continuous ice cover spanning from
+the Alaska Range in the northeast to the Rocky Mountains in the southwest only
+exists\renote{
+    I chosed to write about model output in the present form to discern in from
+    the reality.}
+for at most 29\,kyr in total over the entire last glacial cycle. However,
+except for its marine margin and the northern foothills of the Alaska Range,
+the maximum extent of the ice sheet is attained for an even more brief period
+of a few thousand years (Figs.~\ref{fig:timeseries}--{fig:duration}). This
+result illustrates that the maximum extents of the modelled ice sheet during
+MIS~4 and 2 were both short-lived and therefore out of balance with
+contemporary climate, which confirms previous inferences of a short-lived
+Cordilleran ice sheet based on the geological record \citep{Clague.etal.1980,
+Stroeven.etal.2010}.
+
+\todo{Include two paragraphs from Martin, but not here. Make sure that figures
+      come in the right order in these paragraphs}.
+
+Pehaps one of the most persistent landscape feature is the Skeena Mountains ice
+cap, which persisted throughout the entire last glacial cycle
+(Figs.~\ref{fig:snapshots} and~\ref{fig:duration}). Regardless of the applied
+forcing (Fig.~\ref{fig:snapshots}, middle panels), this ice cap appears to
+survive MIS~3, and serves as a nucleation centre at the onset of the glacial
+readvance towards the LGM (MIS~2). The importance of residual ice for Noeth
+American glacial history leading up to the LGM has been illustrated by the
+MIS~3 residual ice bodies in northern and eastern Canada as nucleation centres
+for a much more extensive MIS~2 configuration \citep{Kleman.etal.2010}.
+
+The presence of a Skeena Mountains ice cap during most of the last glacial
+cycle can be explained by a more widespread\renote{
+    I mean widespread, precipitation there is lower than elsewhere.},
+current winter precipitation for that region there than for other parts of the
+modelling domain (Fig.~\ref{fig:atm}). Along most of the north-western coast of
+North America, coastal mountain ranges form a pronounced topographic barrier
+for westerly winds, capturing atmospheric moisture in the form of orographic
+precipitation, and resulting in arid interior lowlands. However, near the
+centre of our modelling domain, this barrier is less pronounced than elsewhere,
+allowing westerly winds to carry moisture\renote{
+    I do not agree with ``carry precipitation''. When reading this I get the
+    picture of water drops falling diagonaly under action of the wind. I do not
+    deny that transport of precipitation happens and
+    is an important process for understanding orographic
+    precipitation (think of why it is raining so much in the Swedish mountains
+    even when the wind comes from the west), but it is not what I am trying to
+    describe here. I wrote ``moisture'' and this is what I mean.}
+further inland, until it is progressively captured by the extensive, but mildly
+elevated group of the Skeena Mountains in north-central British Columbia.
+
+\todo{Explain why it would not be good with general statements on thermal
+      evolution of ice cover.}
+
+[In addition, the erosional landscape of the Skeena mountain bears one of the
 strongest glacial imprint found within the area formerly covered by the
@@
     Fig.~\ref{fig:warmbase}? Tell me what you think of it.})
-may partly explain this spectacular landscape.
+may partly explain this spectacular landscape.]
+
+A correlation is observed between the modelled duration of warm based ice cover
+(Fig.~\ref{fig:warmbase}) and the degree of glacial modification of the
+landscape (mainly in terms of the development of deep glacial valleys and
+troughs). We find evidence for this on both the southwestern and northeastern
+slopes of the southern Coast Mountains, on the western slopes of the Columbia
+Mountains, on the western and eastern slopes of the northern Coast Mountains
+and the Saint-Elias Mountains, and radiating off the Skeena Mountains
+(Figs.~\ref{fig:duration}, \ref{fig:warmbase} and~X\renote{
+    Yes it could be interesting to compare basal flow patterns with Fig.~2 of
+    \citet{Kleman.etal.2010, but I don't think that I have the time for this
+    right now.}}).
+The Skeena Mountains, for example, indeed bear a strong glacial imprint that
+indicates ice drainage in a system of distinct glacial troughs emanating in a
+radial pattern from the centre of the mountain range (Fig. X). We suggest that
+persistent ice cover (Fig.~\ref{fig:duration}) associated with basal ice
+temperatures at the pressure-melting point (Figs.~\ref{fig:warmfrac}
+and~\ref{fig:warmbase}) explains the large-scale glacial erosional imprint on
+the landscape. In contrast, a well-developed network of glacial valleys west of
+the MacKenzie Mountains \citep[Fig.~X;][]{Kleman.etal.2010,
+Stroeven.etal.2010}, which is modelled to have hosted warm-based ice
+(Fig.~\ref{fig:warmfrac}) has only been glaciated for a short duration
+(Fig.~\ref{fig:duration}) during the last glacial cycle according to our
+results (Figs.~\ref{fig:duration} and~\ref{fig:warmbase}). We therefore infer
+that either our model does not perform reliably in this part of the domain or
+the observed landscape pattern originates from older glacial cycles and
+indicates an increased relative importance of this ice dispersal centre prior
+to the Late Pleistocene \citep[cf.][]{Ward.etal.2008, Demuro.etal.2012}.
+
+\todo{Paragraph on model uncertainties related to warm-based ice.}
 

### 2014-07-17 cordillera-cycle.tex
Work in progress on (mainly) section 4.3.

@@
 cosmogenic exposure dating
-    \citep{Stroeven.etal.2010, Stroeven.etal.2014, Margold.etal.inpress},
+    \citep{Stroeven.etal.2010, Stroeven.etal.2014, Margold.etal.2014},
 bedrock deformation in response to former ice loads
@@
 and to the Fraser Glaciation (MIS~2) documented for its southern sector
-    \citep{Porter.Swanson.1998, Margold.etal.inpress}.
+    \citep{Porter.Swanson.1998, Margold.etal.2014}.
 There is patchy stratigraphical evidence for glaciations older than the Fraser
@@
     Yes it could be interesting to compare basal flow patterns with Fig.~2 of
-    \citet{Kleman.etal.2010, but I don't think that I have the time for this
-    right now.}}).
+    \citet{Kleman.etal.2010}, but I don't think that I have the time for this
+    right now.}).
 The Skeena Mountains, for example, indeed bear a strong glacial imprint that
@@
 
-In our simulations, the timing of the last glacial maximum, and in turn that of
-the deglaciation, depends on the palaeo-temperature time-series used to force
-the model (Fig.~\ref{fig:timeseries}). The
-6\,km-resolution simulation forced by the EPICA record reaches a maximal
-ice volume at 17.3\,kyr, followed by a continuous deglaciation between
-about 17 and 11\,kyr (Fig.~\ref{fig:deglacseries}, red dotted curve).
-That driven by the GRIP record reaches maximal
-volume at 19.1\,kyr, and produces a deglaciation in three steps,
-including the first stand-still between 16.6 and 14.5\,kyr, and a late
-glacial re-advance between 12.6 and 11.6\,kyr (Fig.~\ref{fig:deglacseries},
-blue dotted curve).
-
-The possibly of a widespread late-glacial in
-the North American Cordilleran has been debated for long. On the southern tip
-of the Coast Mountains, radiocarbon-dated frontal moraines indicate multiple
-stages of glacial re-advance in the Fraser and Squamish valleys, of which
-one corresponds to the Younger Dryas chronozone \citep{Clague.etal.1997,
+\todo{add 1 sentence to connect to the next paragraph: that is easy:D}
+
+The timing of peak ice volume during the last glacial maximum and the pacing of
+deglaciation depend cirticially on the choice of atmospheric forcing
+(Figs.~\ref{fig:timeseries} and~\ref{fig:deglacseries}). Adopting the EPICA
+atmospheric forcing yields peak ice volume at 17.3\,ka and an uninterupted
+deglaciation until 10.6\,ka (Fig.~\ref{fig:deglacseries}, lower panel, red
+curves). The simulation driven by the GRIP palaeo-temperature record yields
+peak ice volume at 19.1\,ka and a deglaciation interupted by two standstills or
+readvances until 9.4\,ka. The first interuption occurs between 16.6 and
+14.5\,ka, and the second between 12.6 and 11.6\,ka
+(Fig.~\ref{fig:deglacseries}, lower panel, blue curve). Hence, the two model
+runs, while similar in overall timing compared to runs with other climate
+drivers, differ in detail in that the EPICA depicts peak glaciation almost
+2\,ka later and shows a faster uninterrupted deglaciation which yields ice-free
+conditions more than 1\,ka earlier.
+
+\todo{From the timing of max glaciation and the duration of the deglaciation,
+is there any evidence to prefer one model output above the other? If so, state
+that, if not, state that too. The behavior of both models is perhaps reasonable
+within the uncertainty envelope that the field evidence provides...?}
+
+The possibility of late glacial readvances in the North American Cordillera has
+been debated for some time \citep{Osborn.Gerloff.1997}, and locally these have
+been reconstructed and dated. Radiocarbon-dated end moraines in the Fraser and
+Squamish valleys off the southern tip of the Coast Mountains, indicates
+consecutive glacier maximal (or stillstands while in overall retreat), one of
+which one corresponds to the Younger Dryas chronozone \citep{Clague.etal.1997,
 Friele.Clague.2002, Friele.Clague.2002a, Kovanen.2002,
-Kovanen.Easterbrook.2002}. In the Finlay River area, northern Rocky Mountains,
-sharp-crested moraines show an abundant remobilization of late-glacial
-sediments by alpine glaciers in interaction with the main body of the decaying
-ice sheet \citep{Lakeman.etal.2008}. Additional clues for a late-glacial
-re-advance include alpine glacier moraines in the eastern Coast Mountains,
-Rocky Mountains and Columbia Mountains \citep{Menounos.etal.2008}. Although
-further work is needed to constrain the timing of the late-glacial re-advance,
-to seize its extent and to identify potential climatic triggers
+Kovanen.Easterbrook.2002}. In the Finlay River area of northern Rocky
+Mountains, sharp-crested moraines indicate a late-glacial readvance of alpine
+glaciers and their interaction with the main body of the decaying Cordilleran
+ice sheet \citep{Lakeman.etal.2008}. Additional evidence for late-glacial
+alpine glacier readvances includes moraines in the eastern Coast Mountains,
+Rocky Mountains and the Columbia Mountains \citep{Osborn.Gerloff.1997,
+Menounos.etal.2008}. Although further work is needed to constrain the timing of
+the late-glacial readvance, to assess its extents and geographical
+distribution, and to identify the potential climatic triggers
 \citep{Menounos.etal.2008}, it is interesting to note that the simulation
-driven by the GRIP record produces a late-glacial re-advance in the Coast
-Mountains, Rocky Mountains and Finlay River area, corresponding to where it
-has been identified in the geologic record (Fig.~\ref{fig:deglac}, left panel),
-whereas the EPICA-driven simulation produces a nearly-continuous deglaciation
-with very restrained glacial re-advance (Fig.~\ref{fig:deglac}, right panel)%
-\footnote{Should we mention our work in progress with Brian and/or the Tuya
-    Lake moraine?}.
-
-Although the predicted timing of glacier retreat differs between the two
-simulations, its patterns are relatively consistent (Fig.~\ref{fig:deglac}). In
-both runs, the southern half of the modelling domain, including the Puget
-Lowland, the Coast and Rocky Mountains and the Interior Plateau of British
-Columbia, are completely deglaciated by 12\,kyr, whereas significant ice cover
-remains on the Skeena, Selwyn, MacKenzie, Wrangell and Saint-Elias Mountains
-in the northern half of the domain. After 12\,kyr, deglaciation proceeds
-through an opening of the Liard lowland with radial margin retreat towards the
-surrounding mountain ranges and inward flow towards the depression, in
-compatibility with the regional melt water record \citep{Margold.etal.2013}.
-Remaining ice continues its decay by retreating towards the Selwyn Mountains
-and, then, the Skeena Mountains. The last remains of the palaeo-ice sheet
-vanish in the Skeena Mountains at around 10.6\,kyr in the EPICA simulation
-versus 9.4\,kyr in the GRIP simulation.
-
-Once more, snapshots of modelled deglacial ice sheet configurations
-(Figs.~\ref{fig:deglacshots-grip} and~\ref{fig:deglacshots-epica}%
-\footnote{I produced one figure for each run, but as you could expect from
-    Fig.~\ref{fig:deglac}, Figs.~\ref{fig:deglacshots-grip}
-    and~\ref{fig:deglacshots-epica} look very similar. One of them would
-    probably be enough, or?}) provide a
-poor medium for direct comparison with geomorphological evidence of the ice
-retreat. Here, we compute a map of deglacial, basal flow directions immediately
-preceding exposure of ice-free land or cessation of sliding upon the approach
-of the retreating margin (Fig.~\ref{fig:lastflow}). On the northern and
-southern front of the ice sheet and in the Liard Lowland, the resulting map
-show similarities with patterns of glacial lineations
-(\citealp{Prest.etal.1968}; \citealp[Fig.~1.12]{Clague.1989};
-\citealp[Fig.~2]{Kleman.etal.2010}; \citealp[Fig.~2]{Margold.etal.2013}).
-However, such is not the case on the Interior Plateau of British Columbia,
-where both simulations predicts negligible basal sliding during deglaciation
-(Fig.~\ref{fig:lastflow}), whereas glacial lineation maps indicate substantial
-eastward flow.
-
-On the one hand, this could indicate that the model performs poorly in this
-region, perhaps because the modelled last glacial maximum ice sheet is too
-thick, or too extensive to the west. Because feedback mechanisms between ice
-sheet topography and regional climate such as wind redistribution, orographic
-precipitation and latent warming of moisture-depleted air are absent in our
-model, we regard this possibility as fully probable \citep{Seguinot.etal.2014}.
-On the other hand, in both simulations driven by the GRIP and EPICA records,
-the deglaciation of central British Columbia consists of a rapid northwards
-retreat associated with southwards-flowing, non-sliding ice lobes on the
-Interior Plateau, while the surrounding mountain ranges are already ice-free
+driven by the GRIP record produces a late-glacial readvance in the Coast
+Mountains, Rocky Mountains and the Finlay River area, corresponding to where it
+has been identified in the geological record (Fig.~\ref{fig:deglac}, left
+panel). In contrast the EPICA-driven simulation produces a nearly-continuous
+deglaciation with only a highly restricted glacial readvance
+(Fig.~\ref{fig:deglac}, right panel).
+
+\todo{Note that most of the evidence for re-advance is for small glaciers,
+      whereas the model output indicates that there was still a big body of ice
+      at that time, as in \citet{Lakeman.etal.2008}.}
+
+Predicted patterns of ice sheet retreat are relatively consistent between the
+two simulations (Fig.~\ref{fig:deglac}). The southern sector of the modelling
+domain, including the Puget Lowland, the Coast and the Rocky mountains, and the
+Interior Plateau of British Columbia, becomes completely deglaciated by 12\,ka,
+whereas a significant ice cover remains over the Skeena, Selwyn, MacKenzie,
+Wrangell and Saint-Elias mountains in the northern sector of the modelling
+domain. After 12\,ka, deglaciation continues to proceed across the Liard
+Lowland with a radial ice margin retreat towards the surrounding mountain
+ranges, and inward flow towards the depression, consistent with the regional
+melt water record of the last deglaciation \citep{Margold.etal.2013}
+\todo{rephrase with ``unzip''?}. Remaining ice continues to decay by retreating
+towards the Selwyn and Skeena mountains. The last remnants of the Cordilleran
+ice sheet finally disappears from the Skeena mountains around 10.6\,ka (EPICA)
+and 9.4\,ka (GRIP).
+
+Because a general tenant in glacial geomorphology is that the majority of
+landforms (lineations and eskers) are part of the deglacial envelop
+\citep[terminology from]{Kleman.etal.2006}, that is that these were formed
+close inside the retreating margin of ice sheets \citep{Boulton.Clark.1990,
+Kleman.etal.1997, Kleman.etal.2010}, we present a map of basal flow directions
+immediately preceding deglaciation or at the time of cessation of sliding
+inside the cold-base retreating margin (Fig.~\ref{fig:lastflow})
+\todo{description of the general trends that the map shows}. Patterns of
+glacial lineation formed in the northern and southern sectors of the
+Cordilleran ice sheet and in the Liard Lowland (\citealp{Prest.etal.1968};
+\citealp[Fig.~1.12]{Clague.1989}; \citealp[Fig.~2]{Kleman.etal.2010};
+\citealp[Fig.~2]{Margold.etal.2013}), similarities with the patterns of
+deglacial ice flow from modelling (Fig.~\ref{fig:lastflow}). However, this is
+not the case for the Interior Plateau of British Columbia, where both
+simulations predict negligible basal sliding during deglaciation
+(Fig.~\ref{fig:lastflow}), but where an impressive set of glacial delineations
+indicate a substantial eastward flow component of the Cordilleran ice sheet
+\citep{Prest.etal.1968, Kleman.etal.2010}.
+
+The Interior Plateau lineation swarm could thus present a smoking gun for the
+reliability of the presented model results. One explanation for the incongruent
+results could therefore be that the modelled LGM ice sheet is too thick, or
+that the ice divide is positioned too far to the east \todo{descript the
+problem for the incongruent results of these two conditions}. Because feedback
+mechanisms between ice sheet topography and regional climate are absent in our
+model, including wind redirection, orographic precipitation effects, and latent
+warming of moisture-depleted air, we regard misfeeds in ice thickness and ice
+devide location as plausible \citep{Seguinot.etal.2014}. Another explanation
+for the incongruent results could be that the Interior Plateau lineation swarm
+predates deglaciation and that deglacitation landforms are largely absent. The
+modelled deglaciation of the Interior Plateau, central British Columbia,
+consists of a rapid northwards retreat (Fig.~\ref{fig:deglac}) of
+southwards-flowing (Fig.~\ref{fig:lastflow}) non-sliding ice lobes positioned
+in-between deglaciated (ice-free) mountain ranges
 (Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}). This result
@@
 glacial lineations on the Interior Plateau of British Columbia may be of older
-age than the last glacial maximum, and may have remained intact throughout the
-deglaciation.
+age than the LGM, and may have remained intact throughout the deglaciation.
 
@@
 Numerical simulations of the Cordilleran ice sheet through the last glacial
-cycle presented in this study consistently produce two glaciation maxima during
-MIS~4 (61.9--55.4\,kyr) and MIS~2 (29.5--16.9\,kyr), two periods
+cycle presented in this study consistently produce two glacial maxima during
+MIS~4 (61.9--55.4\,ka) and MIS~2 (29.5--16.9\,ka), two periods
 corresponding to documented extensive glaciations. This result is
-independent of the choice of palaeo-temperature record used to approximate past
-climate evolution, and thus
-can be seen as a first-order agreement between the model and geological
+independent of the choice of the palaeo-temperature record used to approximate
+the past climate evolution, and thus
+can be seen as a first-order agreement between the model and the geological
 evidence. However, the timing of the two glaciation peaks depends sensitively
 on which record
-is used to drive the model. The timing of the last glacial maximum is best
+is used to drive the model. The timing of the LGM is best
 reproduced by the Antarctic record, and occurs too early in all simulations
-that are driven by other records. The mismatch is greatest when using
+that are driven by the other records. The mismatch is greatest when using
 oceanic records from the Pacific Northwest, which are affected by the
-weakening of the California current during the last glacial maximum.
+weakening of the California current during the LGM.
 
 In all simulations presented here, ice cover is limited to disjoint mountain ice
-caps during most of the glacial cycle, confirming previous inferences from
-geological evidence preceding the last glacial maximum. However, our
-simulations produce persistent ice cover on the Skeena mountains during
+caps during most of the glacial cycle, confirming previous inferences from the
+geological evidence preceding the LGM. However, our
+simulations produce persistent ice cover on the Skeena Mountains during
 the entire glacial cycle. At the time when a full-size Cordilleran ice sheet is
@@
 intrusion through a topographic window in the Coast Mountains. The ice cap acts
-as a nucleation centre at the onset of the last glacial maximum re-advance, and
-may have contributed to the spectacular glacial imprint of the Skeena Mountains
-erosional landscape.
-
-During deglaciation, none of the palaeo-temperature records used produces an
-ideal agreement between model results and geological evidence. Although
-the EPICA record yields to more realistic timing of the last glacial maximum
-and early deglaciation, only the GRIP record produces a late-glacial re-advance
-at places where it has been documented in the literature. Nonetheless, the
+as a nucleation centre at the onset of the LGM readvance, and appears
+consistent with the distinct glacial imprint of the Skeena Mountains landscape.
+
+During deglaciation, none of the palaeo-temperature records used produces a
+close agreement between the model results and the geological evidence. Although
+the EPICA record yields a more realistic timing of the LGM and early
+deglaciation, only the GRIP record produces a late-glacial readvance in areas
+where it has been documented in the literature. Nonetheless, the
 general patterns of deglaciation are consistent between both simulations driven
-by the GRIP and EPICA record, and show a rapid deglaciation of the southern
-half of the ice including rapid northwards retreat through the Interior
-Plateau of central British Comumbia, followed by an opening of the Liard
-Lowland, and final ice cap retreat of the ice margin towards the Selwyn and,
-then, the Skeena Mountains, which host the last remnant of the ice sheet
+by the GRIP and EPICA records, and show a rapid deglaciation of the southern
+half of the ice sheet, including a rapid northwards retreat across the Interior
+Plateau of central British Columbia. This is followed by an opening of the ice
+margin in the Liard Lowland, and a final retreat of the margin of the remaining
+ice caps towards the Selwyn and, later, the Skeena mountains, which host the
+last remnant of the ice sheet
 during the early Holocene (10.9--9.5\,kyr).
 
-These results strongly depend on our choices of ice-sheet model (PISM), surface
+These results are strongly dependent on selected ice-sheet model (PISM),
+surface
 mass balance model (PDD) and climate forcing. Most importantly, our simplistic
@@
 extent and volume in continental regions. Nevertheless, our results identify
-the glacially-scrubbed Skeena Mountains as a central, key area to understanding
+the Skeena Mountains as a key area to understanding
 glacial dynamics of the Cordilleran ice sheet, highlighting the need for
@@
 German Academic Exchange Service~(DAAD) grant no.~50015537 and a Knut and Alice
-Wallenberg Foundation grant to J.~Seguinot, and by NASA grant no.~NNX13AM16G
-to E.~Bueler. Computer resources were provided by the Swedish National
+Wallenberg Foundation grant to J.~Seguinot.
+Computer resources were provided by the Swedish National
 Infrastructure for Computing (SNIC) allocation no. 2013/1-159 to A.~P.~Stroeven

### 2014-07-17 cordillera-cycle.tex
Highlight headers in LaTeX source.
Add new outline.

@@
 % cordillera-climate.tex
-% ----------------------------------------------------------------------
+% =============================================================================
 
@@
 
-% ----------------------------------------------------------------------
+% =============================================================================
 \begin{document}
 \linenumbers
-% ----------------------------------------------------------------------
+% =============================================================================
 
@@
 
-% ----------------------------------------------------------------------
+% =============================================================================
 \introduction
 \label{sec:intro}
-% ----------------------------------------------------------------------
+% =============================================================================
 
@@
 
-% ----------------------------------------------------------------------
+% =============================================================================
 \section{Model setup}
 \label{sec:model}
-% ----------------------------------------------------------------------
-
-%\subsection{Forcings}
+% =============================================================================
+
+% -----------------------------------------------------------------------------
+\subsection{Overview}
+% -----------------------------------------------------------------------------
 
 \todo{Move parameter values into a separate table.}
-
-\noindent\todo{Do we need a new 2.1 subsection title here?}
 
@@
 
+% -----------------------------------------------------------------------------
 \subsection{Ice thermodynamics}
 \label{sec:icedyn}
+% -----------------------------------------------------------------------------
 
@@
 
+% -----------------------------------------------------------------------------
 \subsection{Surface mass balance}
+% -----------------------------------------------------------------------------
 
@@
 
+% -----------------------------------------------------------------------------
 \subsection{Atmospheric forcing}
 \label{sec:atm}
+% -----------------------------------------------------------------------------
 
@@
 
-% ----------------------------------------------------------------------
+% =============================================================================
 \section{Sensitivity to atmospheric forcing time-series}
 \label{sec:results}
-% ----------------------------------------------------------------------
+% =============================================================================
+
+% -----------------------------------------------------------------------------
+\subsection{Evolution of ice volume}
+% -----------------------------------------------------------------------------
 
@@
 of the ice margin.
+
+% -----------------------------------------------------------------------------
+\subsection{Extreme configurations}
+% -----------------------------------------------------------------------------
 
@@
 
-% ----------------------------------------------------------------------
+\todo{Add general statements on the scale of glaciation.}
+
+% =============================================================================
 \section[Comparison to the geologic record]
@@
 \label{sec:discussion}
-% ----------------------------------------------------------------------
-
-\subsection{Timing of glaciation}
+% =============================================================================
+
+% -----------------------------------------------------------------------------
+\subsection{Glacial maxima}
+% -----------------------------------------------------------------------------
+
+\subsubsection{Timing of glaciation}
 
@@
 
-\subsection{Location and longevity of the major ice-despersal centres}
-
-\todo{Add general statements on the scale of glaciation (but where?).}
+\subsubsection{Ice configuration during MIS~2}
+
+\todo{Add MIS~2 ice topo and velocity map. Include two paragraphs from Martin.}
+
+\subsubsection{Ice configuration during MIS~4}
+
+\todo{Add MIS~4 ice topo and velocity map. Discuss that some of the margins
+      are more extensive during MIS~4 than during MIS~2, in the GRIP simu.}
+
+% -----------------------------------------------------------------------------
+\subsection{Nucleation centres}
+% -----------------------------------------------------------------------------
+
+\subsubsection{Transient ice sheet states}
 
@@
 
-\todo{Include two paragraphs from Martin, but not here. Make sure that figures
-      come in the right order in these paragraphs}.
+\subsubsection{Major ice-dispersal centres}
 
@@
 elevated group of the Skeena Mountains in north-central British Columbia.
+
+\subsubsection{Potential imprint on the landscape}
 
@@
 
-\subsection{Pace and patterns of deglaciation}
+% -----------------------------------------------------------------------------
+\subsection{The last deglaciation}
+% -----------------------------------------------------------------------------
+
+\subsubsection{Pace and patterns of deglaciation}
 
@@
 within the uncertainty envelope that the field evidence provides...?}
+
+\subsubsection{Possibility for a late-glacial readvance}
 
@@
 
+\subsubsection{Deglacial flow directions}
+
 Because a general tenant in glacial geomorphology is that the majority of
@@
 
-% ----------------------------------------------------------------------
+% =============================================================================
 \conclusions
 \label{sec:concl}
-% ----------------------------------------------------------------------
+% =============================================================================
 
@@
 
-% ----------------------------------------------------------------------
+% =============================================================================
 % Floats
-% ----------------------------------------------------------------------
+% =============================================================================
 
@@
 
-% ----------------------------------------------------------------------
+% =============================================================================
 \end{document}
 \endinput
-% ----------------------------------------------------------------------
+% =============================================================================

### 2014-07-17 kappa.tex
Update author contrib, and correct broken refs.

@@
 \begin{description}
-  \item[\CCLI] \bibentry{Seguinot.etal.2014}
-  \item[\PSDV] \bibentry{Seguinot.2013}
-  \item[\PSDP] \bibentry{Seguinot.Rogozhina.2014}
-  \item[\CCYC] \bibentry{Seguinot.etal.inpreparation}
+  \item[\CCLI] \bibentry{Seguinot.etal.2014}.
+  \item[\PSDV] \bibentry{Seguinot.2013}.
+  \item[\PSDP] \bibentry{Seguinot.Rogozhina.2014}.
+  \item[\CCYC] \bibentry{Seguinot.etal.manuscript}.\renote{
+    Arjen, you had several comments regarding the style of references
+    throughout the kappa (name formatting, italic text). I am using the
+    Copernicus style here, but is there another recommendation to follow?
+    Maybe Peter knows?}
 \end{description}
@@
 
-\small{\begin{enumerate}
-  \item[I] I set-up the ice sheet model and supercomputing workflow, ran the
-simulations and wrote the manuscript; C.~Khroulev implemented necessary code
-changes in PISM; I.~Rogozhina guided experiment design; A.~P.~Stroeven
-initiated the Cordilleran ice sheet modelling project; and Q.~Zhang prepared
-some of the forcing data. All co-authors contributed to improve the text.
-  \item[II] I analysed the weather data, wrote the positive degree day model,
-ran computations and wrote the manuscript. I.~Rogozhina helped to assess the
-reviews.
+\footnotesize{\begin{enumerate}
+  \item[I] A.~P.~Stroeven initiated the Cordilleran ice sheet modelling
+project; I set-up the ice sheet model (PISM) and supercomputing workflow, ran
+simulations and wrote the first draft; C.~Khroulev implemented necessary code
+changes in the model; I.~Rogozhina contributed to experiment design; A.~P.~Stroeven
+helped with interpretation of the results; and Q.~Zhang prepared some of the
+forcing data. All co-authors contributed to improve the text.
+  \item[II] I analysed the weather data, wrote the positive degree day model
+(PyPDD), ran computations and wrote the manuscript. A.~P.~Stroeven contributed
+with editorial comments at the pre-submission stage and I.~Rogozhina helped to
+assess the reviews.
   \item[III] I.~Rogozhina provided the original idea; I analysed the weather
-data; and we wrote the manuscript together.
-  \item[IV] I implemented necessary code changes in PISM, ran the simulations
-and wrote the manuscript; I.~Rogozhina guided experiment design; M.~Margold,
-J.~Kleman and A.~P.~Stroeven took part in the interpretation. All authors
-contributed to improve the text.
+data and developed a new parameterization of daily temperature standard
+deviation. I.~Rogozhina and I wrote the manuscript. A.~P.~Stroeven contributed
+with editorial comments at the pre-submission stage.
+  \item[IV] I integrated the new parameterization of standard deviation into
+PISM, ran the simulations and wrote the first draft; I.~Rogozhina guided
+experiment design; A.~P.~Stroeven, M.~Margold and J.~Kleman took part in the
+interpretation and comparison of model results against geological evidence. All
+co-authors contributed to improve the text.
 \end{enumerate}}
@@
 (\cref{fig:map-northamerica}).
-During the last glacial maximum, apex of glaciation, three major ice sheets had
-collided to form a continuous blanket of ice that spanned from Alaska to
+During the Last Glacial Maximum (LGM), three major ice sheets had collided to
+form a continuous blanket of ice that spanned from southern Alaska to
 Newfoundland, connecting with the Greenland ice sheet in the Canadian Arctic,
-and extending as far south as Chicago and the Great Lakes
+and extending as far south as Ohio
 (\cref{fig:map-northamerica}). The Laurentide ice
-sheet, largest of the three, was centred on the modern Hudson Bay,
+sheet, largest of the three, was centred on the Hudson Bay,
 and fully covered the Canadian Prairies, Eastern Canada, and the Great Lakes
 Basin. The Innuitian ice sheet, much smaller, covered the northernmost Canadian
-archipelago. Finally, \emph{the Cordilleran ice sheet} capped the mountain
-ranges of the Western Cordillera, which forms our region of focus in the
-present thesis.
+archipelago. Finally, \emph{the Cordilleran ice sheet}, intermediate in size,
+capped the mountain
+ranges of the Western Cordillera.
 
@@
 
-The North American Cordillera forms a broad mountain system with an
-average east-west width of about 1000\,km. Its topography is a complex
-jigsaw of mountain ranges and interior basins (\cref{fig:map-cordillera}).
-It was formed during multiple accretion phases associated with the closure of
-several back-arc oceanic trenches, over a 200\,Ma-long geological history
+The North American Cordillera forms an elongated but broad mountain system with
+an average east-west width of about 1000\,km (\cref{fig:map-cordillera}). Its
+topography is a complex jigsaw of mountain ranges and interior basins. It was
+formed during multiple accretion phases associated with the closure of several
+back-arc oceanic trenches, over a 200 million year (Ma)-long geological history
 \citep{Sigloch.Mihalynuk.2013}.
 
-Most of the dominant reliefs are found along two major ribbons that run
-parallel to each other along the general orientation of the Cordillera
-(\cref{fig:map-cordillera}). The western ribbon is formed by the Pacific Coast
-orogenic belt \mref. Major glaciated reliefs include,
+The dominant reliefs occur along two major ribbons that run
+parallel to each other along the general orientation of the North American Cordillera
+(\cref{photo-alaska-range}). The western ribbon is formed by the Pacific Coast
+orogenic belt \mref\ and includes,
 from North to South, the Alaska Range, the Wrangell and Saint-Elias Mountains,
@@
 Nevadan and Laramide belts, which are separated by the deep and strikingly
-linear Rocky Mountain / Tintina Trench \mref. It includes the Selwyn and
-MacKenzie Mountains in the north, and the Rocky Mountains in the South. The
-Brooks Range, in northern Alaska, has been covered by ice, but its glaciers
-were not connected to the main body of the Cordilleran ice sheet
+linear Rocky Mountain / Tintina Trench \mref, depression and includes the Selwyn and
+MacKenzie Mountains in the north, and the Rocky mountains in the South. The
+Brooks Range of northern Alaska has been covered by ice, but its glaciers
+remained disjunct from the main body of the Cordilleran ice sheet
 \citep{Kaufman.Manley.2004}.
@@
 
-The Cordillera comports several intermontane basins, which influenced the
-patterns of ice flow. The two major depressions are the Fraser Plateau in
-south-central British Columbia, and the Liard Lowland in northern British
-Columbia. Between these two basins is found a mildly elevated, but extensive
-and deeply incised mountain group, known as the Skeena Mountains.
+The North American Cordillera hosts several intermontane basins, the presence
+of which influenced former patterns of Cordilleran ice sheet flow. The two
+major depressions are the Fraser Plateau in south-central British Columbia, and
+the Liard Lowland in northern British Columbia (\cref{fig:map-cordillera}).
+These two basins are separated by the Skeena Mountains, a mildly elevated, but
+extensive and deeply incised mountain group. This is the most dominant upland
+area in-between the dominant topographic ribbons that form the North American
+Cordillera.
 
 The ubiquitously mountainous topography of the region presents a serious
-obstacle to reconstructing past dynamics of the Cordilleran ice sheet. Because
-of the irregular arrangement of valleys and ridges that formed the ice sheet's
-floor, ice moved along tortuous and intertwined pathways, sometimes following
-the major topographical troughs, and sometimes flowing across them
-\citep{Davis.Mathews.1944, Kleman.etal.2010}. In terms of numerical glacier
-modelling, this topography implies complex patterns of flow and stress
-within the ice and requires high-resolution, and thus, large computational
-grids.
+obstacle to reconstructing the palaeoglaciology of the Cordilleran ice sheet.
+Because of the irregular arrangement of valleys and ridges that formed the ice
+sheet's subglacial landscape, ice moved along tortuous and intertwined
+pathways, sometimes following the major topographical troughs, and sometimes
+flowing across them \citep{Davis.Mathews.1944, Kleman.etal.2010}. In terms of
+numerical ice sheet modelling, the presence of steep and irregular subglacial
+topography yields complex flow patterns and stress distributions within the ice
+sheet and requires high-resolution and thus, large computational grids.
 
@@
 The topographic complexity of the North American Cordillera is associated with
-distinctive patterns of temperature and precipitation, whose effect is often
-clearly visible in the landscape (\cref{fig:photo-fraser-valley}). Due to the
-large latitudinal extent of the region, \emph{mean annual temperatures} vary
-significantly from north to south (\cref{fig:plot-atm}a). About half of the
-model domain has mean annual temperature below freezing. However,
-\emph{seasonality},
-here computed as the temperature difference between the warmest and the coldest
-months, varies between around 10\,\degC\ along most of the Pacific Coast to
-above 40\,\degC\ in the polar interior (\cref{fig:plot-atm}b). There exists a
+distinctive patterns of temperature and precipitation, whose effects on
+vegetation are clearly expressed in the landscape
+(\cref{fig:photo-fraser-valley}). Due to the large latitudinal extent of the
+region, \emph{mean annual temperatures} vary significantly from north to south
+(\cref{fig:plot-atm}a). About half of the model domain has mean annual
+temperatures below freezing. However, \emph{temperature seasonality}\renote{
+    This is a common term in the meteorological literature, I think. At least
+    this is what Qiong told me.},
+here computed as the difference between the warmest and the coldest months,
+varies between around 10\,\degC\ along most of the Pacific Coast to above
+40\,\degC\ for the interior Arctic sectors \renote{
+    I think the word ``polar'' or ``Arctic'' is needed here, see
+    \cref{fig:plot-atm}b.}
+(\cref{fig:plot-atm}b). There exists a
 partial correlation between cold regions and those that have a large
-seasonality, so that many regions where mean annual temperature is well below
-freezing point do experience warm summers.
-
-Rates of \emph{precipitation} are highest on the western slopes of the Pacific
-Coast
-ranges and decline quickly on the eastern side of the mountains
-(\cref{fig:plot-atm}c). In terms of ice sheet mass balance, this contrast is
-even reinforced by the difference in timing of the \emph{precipitation peak}
-through
-the year (\cref{fig:plot-atm}d). Coastal regions receive most precipitation
-during winter and fall, thus as snow. On the contrary, many inland regions
-experience dry winters, and most of the annual precipitation,
-consequently falls there
-as rain during the ablation season. The climatic setting of the North American
-Cordillera is described further in \CCLI, where its effects on the ice sheet
-model response have been studied in detail.
+seasonality, so that many regions where mean annual temperatures are well below
+freezing point experience warm summers.
+
+\emph{Annual precipitation rates} are highest on the western slopes of the
+Pacific Coast ranges and decline quickly on the eastern side of the mountains
+(\cref{fig:photo-fraser-valley}, \cref{fig:plot-atm}c). In terms of ice sheet
+mass balance, this contrast is even reinforced through the difference in timing
+of the \emph{precipitation peaks} through the year (\cref{fig:plot-atm}d).
+Coastal regions receive most precipitation during winter and fall, which is
+therefore largely delivered as snow. In contrast, many inland regions
+experience dry winters, and most of the annual precipitation, consequently
+falls there as rain during summer, further enhancing ice melt. The climatic
+setting of the North American Cordillera is further described in \CCLI, where
+its effects on the ice sheet model response have been studied in detail.
 
@@
 
-The first presence of an ice sheet in the Cordilleran mountains is indicated by
-glaciofluvial deposits dated at $2.64^{+0.20}/_{-0.18}$\,Ma
-\citep{Hidy.etal.2013}. This suggests that the North American Cordillera has
-been glaciated many times, with the first glaciations occurring even before the
-onset of the 100-ka glacial cycle (\cref{fig:paleo-timeseries}). However, only
-the last few glaciations can be reconstructed from the geologic record, whereas
-much of the older evidence has been overridden by subsequent glacial
-re-advances and other geomorphological processes.
-
-In the \emph{northern sector} of the ice sheet, at least four glaciations have
+Evidence from glaciofluvial deposited dated using cosmogenic nuclides at $2.64^{+0.20}/_{-0.18}$\,Ma indicates the first presence of an ice sheet in the Cordilleran mountains in the early Pleistocene
+\citep{Hidy.etal.2013}.
+
+This indicates that the North American Cordillera has
+been glaciated many times since the late Pliocene, with the first of these glaciations occurring even before the mid-Pleistocene
+onset of 100 ka glacial cycles (\cref{fig:paleo-timeseries}). However, only fragments of
+the last few glaciations can be reconstructed from the geologic record, implying that
+much of the older evidence has been erased by subsequent glacial
+re-advances and perhaps even by other, non-glacial geomorphological processes.
+
+In the \emph{northern sector} of the Cordilleran ice sheet domain, at least four glaciations have
 been identified
@@
            Stroeven.etal.2010, Stroeven.etal.2014}.
-The two older ones, often collectively referred to as the Reid glaciations,
-pre-date the last glacial cycle, and thus are beyond the scope of this study.
-The two younger glaciations, though, occurred both during the last glacial
-cycle. First, the Gladstone glaciation reached its maximum between
+The two older ones, often collectively referred to as the Reid  and pre-Reid glaciations,
+pre-date the last glacial cycle, and are therefore beyond the scope of this study.
+The two younger glaciations, however, occurred both during the last glacial
+cycle. First, the older of the two, the Gladstone Glaciation, reached its maximum extent between
 $50.4\pm3.0$ and $54.3\pm2.0$\,\unit{^{10}Be\,ka} \citep{Ward.etal.2007},
-during a period of high global ice volume on Earth, known as the Marine Isotope
-Stage (MIS)~4. However, evidence from the Gladstone glaciation has not been
-found in the southern part of the ice-covered region, making it difficult to
-seize its magnitude. Second, the McConnell glaciation is associated with end
+during Marine Oxygen Isotope Stage (MIS) 4, a period of relatively large global ice volume. However, evidence from the Gladstone Glaciation has not been
+documented along the southern ice sheet margin, making it difficult to
+seize its spacial extent. The younger of the two last glacial cycle glaciations, the McConnell Glaciation, is associated with end
 moraines of oldest minimum apparent exposure ages of
 $15.7\pm1.5$ and $17.7\pm1.6$\,\unit{^{10}Be\,ka} \citep{Stroeven.etal.2014},
-which corresponds to the last global ice volume maximum (MIS~2). In contrary to
-the Gladstone glaciation, the McConnell glaciation has been correlated with a
-contemporaneous advance of the southern ice sheet margin, known as the Fraser
-glaciation, with a maximum extent framed between
-$17.4$ and $16.4$\,\unit{^{14}C\,cal\,ka} \citep[Fig.~4]{Porter.Swanson.1998}.
-Within errors inherent to the dating methods \citep{Heyman.etal.2011}, the
+which appears to correspond to the last global ice volume maximum (MIS~2). In contrast to
+the Gladstone Glaciation, the McConnell Glaciation appears contemporaneous with glacial advance of the southern ice sheet margin during the Fraser
+glaciation, with a maximum extent attained between
+$17.4$ and $16.4$\,\unit{^{14}C\,cal\,ka} \citep{Porter.Swanson.1998}.
+Within errors inherent to the cosmogenic nuclide exposure dating methods \citep{Heyman.etal.2011}, the
 McConnell and Fraser glaciations correspond to the same glacial episode, which
 resulted in the Last Glacial Maximum (LGM) extent of the Cordilleran ice sheet.
-This is the period where most geological observations point at.
-
-It is generally thought that the Cordilleran ice sheet was short-lived
-    \citep[e.g.]{Clague.etal.1980, Clague.1985, Cosma.etal.2008}.
-The ice sheet originated from the
-coalescence of several mountain-centred ice caps \citep{Davis.Mathews.1944}.
-During the last glacial maximum, ice covered most of the
-Canadian Cordillera, and some of the adjacent land in north-western contiguous
-USA and Alaska (\cref{fig:map-cordillera}).
+Apart from geological evidence of the final demise of the ice sheet, most field evidence relate to the LGM maximum extent.
+
+Ice has covered most of the
+Canadian Cordillera, and some of the adjacent regions in north-western contiguous
+USA and Alaska (\cref{fig:map-cordillera}). The ice sheet originated from the
+coalescence of several mountain-centred ice caps \citep{Davis.Mathews.1944}. It is generally thought that the LGM configuration of the Cordilleran ice sheet was short-lived
+    \citep{Clague.etal.1980, Clague.1985, Cosma.etal.2008}, a hypothesis which can be studied through numerical modelling.
 
@@
 is in several respects better understood than its northern counterpart
-\citep{Booth.etal.2003}. It consists of different piedmont
-lobes that have formed at the outlet of the main valleys. The Puget lobe
-\citep[e.g.]{Thorson.1980, Porter.Swanson.1998} was the westernmost and
-largest lobe. It covered the Puget Lowland, a region that currently includes
-the cities of Vancouver and Seattle. East of the Cascade Range, were found the
-Okanagan and Purcell lobes. The latter once dammed the Clark Fork River to form
-Glacial Lake Missoula \citep{Pardee.1910}. As the ice sheet retreated, the lake
+\citep{Booth.etal.2003}. It consists of coalescent piedmont
+lobes, each of which formed at the outlet of the main valleys. The Puget Lobe
+\citep{Thorson.1980, Porter.Swanson.1998} was the westernmost and
+largest Lobes. It covered the Puget Lowland, a region that comprises
+the cities of Vancouver and Seattle. Further east, beyond the Cascade Range, the
+Okanagan and Purcell lobes formed the southern Cordilleran ice sheet margin. The latter once dammed the Clark Fork River, inhibiting runoff and forming
+Glacial Lake Missoula \citep{Pardee.1910}. As the ice sheet margin retreated, the lake
 repeatedly and catastrophically drained to the Pacific Ocean
@@
 
-The \emph{western margin} of the ice sheet covered part of the continental
-shelf. An
-ice cap covering Vancouver Island was well-connected to main body of ice on the
-continent \mref. A significant part of the ice sheet thus drained into the Georgia
-Strait and engulfed between Vancouver Island and the Olympic Peninsula where
-it formed the Juan the Fuca marine ice lobe, calving into the Pacific Ocean \mref.
-Further to the North, the ice margin reached Queen Charlotte Islands, yet the
-archipelago may not have been fully covered by ice
-flowing from the land. Submarine troughs likely indicate the erosional action of
-former ice streams, yet it is not clear whether they have formed during the
-last glacial cycle or earlier \mref.
-
-Finally, the \emph{eastern margin} of the ice sheet is the most unclear.
+The \emph{western margin} of the Cordilleran ice sheet covered part of the Pacific continental
+shelf. An independent
+ice cap covering Vancouver Island diverted flow from the main body of ice to drain into the Strait of Georgia between Vancouver Island and the Olympic Peninsula to form the Juan the Fuca marine ice lobe, calving into the Pacific Ocean \mref.
+Further to the North, the ice margin encroached the Queen Charlotte Islands. However, the
+archipelago may not have been fully ice-covered, even during maximum of glaciation. Submarine troughs traversing several portions of the continental shelf indicate the erosional action of
+additional ice streams, yet it is not clear whether they have formed during the
+last glacial cycle or during earlier glaciations \mref.
+
+Finally, the conditions at the \emph{eastern margin} of the Cordilleran ice sheet are the least understood.
 Although it
-is known that the Cordilleran ice sheet collided with its Laurentide neighbour,
+is known that the Cordilleran ice sheet collided with the Laurentide ice sheet at times during the last glacial cycle,
 the nature of the interaction between the two ice sheets and the precise
-location of this junction is unclear, while little information exists on its
+location of this junction remains unclear \citep{Gowan.2013}, while limited information exists on its
 timing
@@
 Because the
-present thesis focuses specifically on the Cordilleran ice sheet, our largest
-assumption certainly consists of neglecting buttressing effects from the
+present thesis focuses specifically on the Cordilleran ice sheet, I, therefore, neglect buttressing effect of the
 Laurentide ice sheet.
 
-The ice sheet is believed to have retreated quickly. Most of the evidence comes
-from the Fraser Plateau, where it indicates that some of the highest relief
-emerged from the ice prior to the lowlands, possibly resulting in thin,
-stagnant ice remnants in the interior
+The retreat of the ice sheet during the last deglaciation was likely rapid. Most of the evidence for the style and pace of deglaciation comes
+from the Fraser Plateau. Studies here have indicated that some of the highest relief
+emerged from the ice prior to the surrounding lowlands, possibly resulting in thin,
+stagnant ice remnants covering parts of the plateau
     \citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a}.
-Much less is known about ice sheet decay in the northern parts of the
-ice-covered areas, but glacial landforms from the Liard Lowland suggest a
-different mode of \emph{deglaciation}, with active retreat of the ice lobe
+Less is known about the deglaciation of the northern sector of the Cordilleran ice sheet, but abundant glacial landforms from the Liard Lowland indicate a
+different mode of \emph{deglaciation}, with active retreat of the ice margin
 towards the surrounding mountain ranges \citep{Margold.etal.2013}. Patterns of
@@
 
+\todo{To make this more complete I suggest two additions: 1.
+a few sentences regarding the timing of deglaciation (to complement your introduction where you have timing of glaciation) - look in my 2010 and Martin's papers for information.
+2.
+A few more references older than Margold et al.
+2013.
+You might check what Ward has written fordeglaciation in the Yukon for example?
+Didn't Dyke write about the deglaciation of the Liard Lowland?}
+
 % ======================================================================
@@
 
-The simulations presented in this thesis use Parallel Ice Sheet Model (PISM),
-an open-source, finite-difference, shallow ice-sheet model
-\citep{PISM-authors.2014}. The model reads bedrock topography, sea level,
+The simulations presented in this thesis employ the Parallel Ice Sheet Model (PISM),
+an open-source, finite-difference, shallow ice sheet model
+\citep{PISM-authors.2014}. The model requires bedrock topography, sea level,
 geothermal heat flux and climate forcing as inputs, and computes the thermal
 and dynamic state of the ice sheet, and the associated thermal and
-deformational response of an idealised lithosphere. The atmospheric and oceanic
-envelopes are not part of the model. However, PISM includes boundary models
-that provide simple parametrisations of their effect on the ice model at
-supposedly well-defined interfaces (\cref{fig:model-interfaces}).
+deformational responses of an idealised lithosphere. However, the atmosphere and ocean
+are not part of the model. Instead, PISM includes boundary models
+that provide simple parametrisations of their interaction with the ice sheet across interfaces that are assumed to be well defined (\cref{fig:model-interfaces}).
 
 This section does not attempt to provide an extensive description of PISM,
-which can be found in the on-line documentation \citep{PISM-authors.2014}.
+which can be found in an on-line documentation \citep{PISM-authors.2014}.
 Instead, it aims to highlight modelling choices made in the context of this
-thesis, and give an overview of the physics involved in these choices.
+thesis, and presents an overview of the physics involved in these choices.
 
@@
 equations. Firstly, ice flow is assumed \emph{incompressible}. Hence, the ice
-density, $\rho$, is constant, and the
-conservation of mass implies a conservation of volume, which can be expressed
+density, $\rho$, is constant, and
+conservation of mass therefore implies a conservation of volume, which can be expressed
 in terms of the ice velocity vector, $\vv$ (\cref{fig:model-variables}), by
@@
 Thirdly, after defining the strain-rate tensor
-${\SRT = \frac{1}{2}(\tgrad{\vv} + \tgrad{\vv}^{\mathrm{T}})}$,
-and the deviatoric stress tensor, ${\DST = \CST - p\,\IDT}$,
-where $p=\frac{1}{3}\tr(\CST)$ is the hydrostatic pressure and
+${\SRT = \frac{1}{2}(\tgrad{\vv} + \tgrad{\vv}^{\mathrm{T}})}$\renote{
+  $\grad$, $\div$, $\tgrad$, $\tdiv$, are the classic gradient and divergence operators, applied on vector and tensor fields. $\tr$ is the trace of a matrix. I don't think these need to be defined.},
+and the deviatoric stress tensor, ${\DST = \CST + p\,\IDT}$,
+where $p=-\frac{1}{3}\tr(\CST)$ is the hydrostatic pressure and
 $\tens{\delta}$ is the identity tensor, a \emph{constitutive law} for ice
@@
 where $\mathrm{II}_{\DST}$ is the second invariant of the stress tensor.
-The ice softness coefficient, $A$, depends on ice temperature, pressure, and
-water content through an Arrhenius-type law
+The ice softness coefficient, $A$, depends on ice temperature, $T$, pressure, $p$, and
+water content, $\omega$, through an Arrhenius-type law
 \citep[Eqs.~63--65]{Aschwanden.etal.2012},
@@
 \todo{I am trusting PISM's manual on that. Check other papers by Duval, or
-      the book by Greve (library).}
+      the book by Greve (library). Explain params (Arjen + Irina) or refer to
+      table, maybe.}
 
@@
 \emph{conservation of energy}. In PISM, the conservation of energy uses an
-enthalpy formulation \citep[Eqs.~20--21]{Aschwanden.etal.2012} in order to
+enthalpy formulation in order to
 account for thermodynamic effects associated with internal phase changes in the
-glacier. This enthalpy ($H$) formulation reads
+glacier \citep[Eqs.~20--21]{Aschwanden.etal.2012}. This enthalpy ($H$) formulation reads
 \begin{equation}
@@
 source term corresponding to strain heat release. In the case where ice
-temperature, $T$, is below the pressure-melting point, the enthalpy can be
-expressed as $H=c\,T$, while the heat flow is given by Fourier's law,
-$\vect{q} = k\,\grad{T}$. Consequently, the enthalpy equation
+temperature, $T$, is below the pressure-melting point, $T_c$, the enthalpy can be
+expressed as $H=c\,T$, where, $c$, denotes the specific heat capacity of the ice, while the heat flow is given by Fourier's law,
+$\vect{q} = k\,\grad{T}$, where $k$ is its thermal conductivity of ice. Consequently, the enthalpy equation
 (\ref{eqn:enthalpy}) can be rewritten using a more familiar, cold-ice,
@@
     \matdv{T} = \frac{k}{\rho c} \Delta T
-                + \frac{\nu\doteps_e^2}{4\rho c} \,,
-\end{equation}
-where $c$ denotes the specific heat capacity of the ice and $k$ is its thermal
-conductivity. Parameter values for the ice flow model used in this thesis are
+                + \frac{\nu\doteps_e^2}{4\rho c} \,.
+\end{equation}
+Parameter values for the ice flow model used in this thesis are
 given in \cref{tab:iceflowparams}.
+
+\todo{Explain $H=c\,T$. Enthalpy here is equivalent to internal energy.}
 
@@
 constitute the thermodynamic basis
-of our model. However, their resolution in full form implies a computational
+of the model. However, their resolution in full form implies a computational
 demand too high for multi-millennial, continental-scale applications, such as
-the numerical modelling of the Cordilleran ice sheet through the last glacial
+the numerical modelling of the Cordilleran ice sheet through the 120\,ka last glacial
 cycle, thus requiring shallow approximations.
@@
 
-PISM, the ice sheet model that we use, employs two well-documented
+PISM employs two well-documented
 approximations of ice flow equations: the shallow ice approximation (SIA)
@@
 be derived from a rigorous scaling approach
-    \citep[SIA,][]{Morland.Johnson.1980, Hutter.1983,
+    \citep{Morland.Johnson.1980, Hutter.1983,
            Morland.1987, Weis.etal.1999},
-we don't expand these lengthy derivations
+I won't expand these lengthy derivations
 here, assuming instead the simplifications and focusing on their effect.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=80mm]{model-siassa}
-  \caption{Shallow Ice Approximation (SIA) and Shallow Shelf Approximation
-           (SSA) velocities in PISM. From \citet{Winkelmann.etal.2011}.
-           \todo{Redraw in LaTeX in the same style as
-                 \cref{fig:model-variables}.}}
-  \label{fig:model-siassa}
-\end{figure}
 
@@
 
-Neglecting atmospheric pressure, \Cref{eqn:siaz} yield
+Neglecting atmospheric pressure, \Cref{eqn:siaz} yields \todo{detail}
 \begin{equation}
@@
     \tau_{xz} &= -\rho g (s-z) \pdv{s}{x} \,, \\
-    \tau_{yz} &= -\rho g (s-z) \pdv{s}{x} \,.
+    \tau_{yz} &= -\rho g (s-z) \pdv{s}{y} \,.
 \end{align}
@@
 
-The \emph{shallow shelf approximation} (SSA), on the opposite of the SIA,
+The \emph{shallow shelf approximation} (SSA), in contrast to SIA,
 assumes that
-basal friction is low enough for ice to deform predominantly by horizontal
-expansion within the entire ice column. In this context, the stress-balance can
+basal friction is low enough for ice to deform uniformly within the entire ice column. In this context, the stress-balance can
 be simplified to \citep[Eqs.~4.10--4.12]{Weis.etal.1999}
@@
 \begin{align}
-    p &= \rho g (s-z) - \tau_{zz} \,, \\
-      &= \rho g (s-z) + \tau_{xx} + \tau_{yy} \,,
+    p &= \rho g (s-z) + \tau_{zz} \,, \\
+      &= \rho g (s-z) - \tau_{xx} - \tau_{yy} \,,
 \end{align}
@@
 \begin{align}
-    2\pdv{\tau_{xx}}{x} - \pdv{\tau_{yy}}{x} + \pdv{\tau_{xy}}{y}
+    2\pdv{\tau_{xx}}{x} + \pdv{\tau_{yy}}{x} + \pdv{\tau_{xy}}{y}
         + \pdv{\tau_{xz}}{z} &= \rho g \pdv{s}{x} \,, \\
-    \pdv{\tau_{yx}}{x} + 2\pdv{\tau_{yy}}{y} - \pdv{\tau_{xx}}{y}
+    \pdv{\tau_{yx}}{x} + 2\pdv{\tau_{yy}}{y} + \pdv{\tau_{xx}}{y}
         + \pdv{\tau_{yz}}{z} &= \rho g \pdv{s}{y} \,.
@@
 \cref{eqn:siassa} is a heuristic, and its validity in the zone of
-transition between shear flow and longitudinal flow has not been
+transition between shear flow (SIA) and longitudinal flow (SSA) has not been
 mathematically proven.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=80mm]{model-siassa}
+  \caption{Shallow Ice Approximation (SIA) and Shallow Shelf Approximation
+           (SSA) velocities in PISM. From \citet{Winkelmann.etal.2011}.
+           \todo{Redraw in LaTeX in the same style as
+                 \cref{fig:model-variables}.}}
+  \label{fig:model-siassa}
+\end{figure}
 
@@
 
-The till cohesion $c_0$ is assumed to be zero. The effective pressure on the
+The till cohesion $c_0$ is assumed to be zero \renote{
+  Yes, we should probably defend this choice, but I am not sure how (except
+  that it just diminishes the number of parameters).}.
+The effective pressure on the
 till, $N$, is determined by the modelled amount of water at the bed. However,
@@
 \begin{equation}
-    N = \rho gh (1 - \alpha \frac{W}{W_{max}}) \,.
-\end{equation}
+    N = \rho gh (1 - \alpha \frac{W}{W_{max}}) \,,
+\end{equation}
+where $W$ is the amount of water at the bed, and varies from zero to $W_{max}$,
+a threshold above which further melt simply exits the system.
 
@@
 
+\todo{explain parameters}
+
 Drainage of water at the base of the ice sheet, although implemented in PISM,
 is not included in simulations presented here for the sake of simplicity.
-In turn, the amount of water at the bed, $W$, corresponds to the local
-accumulation of basal melt-water. In varies from zero to $W_{max}$, a threshold
-above which further melt simply exits the system.
 
@@
 ice enthalpy model (\ref{eqn:enthalpy}) with an underlying bedrock thermal
-model. The only process accounted for is heat conduction, hence this model
+model. The only process accounted for is vertical heat conduction, hence this model
 consists of a simple diffusion equation,
@@
 time-scales, because temperature fluctuations caused by climate change and
-isolating effects of the ice sheet penetrate several hundred meters into the
+varying isolation from the fluctuating ice sheet penetrate several kilometers
+into the
 rock, and eventually affect basal ice temperatures.
 
-Secondly, bedrock elevation evolve in response to the ice load. The
+Secondly, bedrock elevation evolve in response to the ice weight. The
 \emph{bedrock deformation} model used in this thesis describes the flexure of
@@
 the ice sheet growth and the isostatic bedrock response. Parameter values for
-bedrock response model are given in \cref{tab:bedrockparams}.
+the bedrock response model are given in \cref{tab:bedrockparams}.
 
@@
 surface mass fluxes are computed from monthly mean surface air temperature,
-$T_m$, and monthly precipitation, $P_m$, by a temperature-index model
+$T_m$, and monthly precipitation, $P_m$, using a temperature-index model
 \citep[e.g.,][]{Hock.2003}. Surface accumulation, $a_s$, equals precipitation
 when temperature is below ${T_0=0\,\degC}$, and decreases to zero linearly
-with temperature between $T_0$ and ${T_1=2\,\degC}$,
+with temperature between $T_s$ and ${T_r=2\,\degC}$,
 \begin{equation}
@@
     \begin{cases}
-        0       & \text{if}\ T_m \le T_s \,, \\
-        P_m \frac{T_m-T_s}{T_r-T_s}
+        P_m     & \text{if}\ T_m \le T_s \,, \\
+        P_m \frac{T_r-T_m}{T_r-T_s}
                 & \text{if}\ T_s < T_m < T_r \,, \\
-        P_m     & \text{if}\ T_r \le T_m \,,
+        0       & \text{if}\ T_r \le T_m \,,
     \end{cases}
@@
 
-For multimillenial simulations of palaeo-ice sheet evolution such as the ones
+For multimillenial simulations of palaeo-ice sheet evolution such as
 presented in this thesis, daily or hourly temperature data is not available,
@@
 under an assumption of normal temperature distribution, using a standard
-deviation parameter $\sPDD$ in the PDD computation \citep{Braithwaite.1984},
+deviation parameter $\sPDD$ in the PDD computation \citep{Braithwaite.1984}.
 PDDs can then be computed using a double-integral formulation
@@
 PDDs are used to melt ice, using a proportionality factor $F_i$
-(\cref{fig:model-pdd}). The melt factors $F_s$ and $F_i$
+(\cref{fig:model-pdd}). The degree-day factors for snow, $F_s$, and ice, $F_i$,
 (\cref{tab:surfaceparams})
-as derived from mass-balance measurements on contemporary glaciers in the
+are derived from mass-balance measurements on contemporary glaciers in the
 Coast and Rocky Mountains of British Columbia \citep{Shea.etal.2009}. In
@@
 North American Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006}, using
-a method initially developed in {\PSDV}, and further improved in {\PSDP}.
+a method initially developed in {\PSDV}, and further improved upon in {\PSDP}.
+
+\todo{explain why it is possible to linearly relate melt and PDDs.}
 
@@
 
-The present-day climatology, $\{T_{m0}, P_{m0}\}$, is computed from
+The present-day monthly climatology is computed from
 near-surface air temperature and precipitation rate fields extracted from
@@
 climate input topography, $b_{ref}$, on the other hand. All simulations use an
-annual temperature lapse rate of $\gamma = 6\,\unit{\degC\,km^{-1}}$.
-
-\CCLI\ tests the sensitivity of the model to different input present-day
-climatologies, $\{T_{m0}, P_{m0}\}$, while the temperature-offset corrections,
-${\Delta}T_{TS}$, uses constant values ranging from -10 to 0\,\degC. \CCYC,
-on the opposite, tests model sensitivity to time-dependent temperature-offset
-corrections derived from different palaeo-temperature proxy records. All
+annual temperature lapse rate of $\gamma = 6\,\unit{K\,km^{-1}}$
+\cref{tab:surfaceparams}.
+
+In \CCLI, we test the sensitivity of the ice sheet model to five different input present-day
+climatologies, $\{T_{m0}, P_{m0}\}$ \cref{tab:datasources}, while adopting a range of constant temperature-offset corrections,
+${\Delta}T_{TS}$, between -10 and 0\,K. In \CCYC,
+on the opposite, we test ice sheet model sensitivity to time dependent temperature-offset
+corrections derived from six different palaeo-temperature proxy records. All
 climatic and topographic data sources used are summarised in
@@
 are the degree-day factors for snow, $F_s$, and ice, $F_i$
-(\cref{fig:model-pdd, tab:surfaceparams}), and the standard deviation of
-near-surface air temperature, $\sPDD$ (\cref{eq:calovgreve,
-tab:surfaceparams}). Although the first two can only be deduced from surface
-mass-balance measurements on present-day glaciers
+(\cref{fig:model-pdd,tab:surfaceparams}), and the standard deviation of
+near-surface air temperature,$\sPDD$
+(\cref{eqn:calovgreve,tab:surfaceparams}). Although the first two can only be
+deduced from surface mass-balance measurements on present-day glaciers
 \citep[e.g.,][]{Shea.etal.2009} and assumptions that these values are also
@@
 
-\noindent\textbf{\CCYC} --- \bibentry{Seguinot.etal.inpreparation}.
+\noindent\textbf{\CCYC} --- \bibentry{Seguinot.etal.manuscript}.
 \bigskip

### 2014-07-17 cordillera-cycle.tex
Add parameter table.

@@
 
+% tab:params
+\begin{table*}
+  \centering
+  \caption{Main parameters of the ice sheet model.
+           \todo{At the moment, this is pretty extensive (does not fit on one
+                 page using the standard font size). There are probably a
+                 couple of things that are not needed here, like for instance
+                 the very standard things (gravity, gas constant, etc). Many
+                 parameters are not defined in the text. Do we need to
+                 introduce them, or should we skip them? I wonder if the last
+                 column is needed, in which case it needs to be completed.}}
+  \label{tab:params}
+  \footnotesize
+  {\begin{tabular}{llrll}
+    \tophline
+    Not.    & Name & Value & Unit & Source \\
+    \middlehline
+
+    \multicolumn{2}{l}{\emph{Ice rheology}} \\
+
+    $\rho$  & Ice density
+            & 910
+            & \unit{kg\,m^{-3}}
+            & \citet{Aschwanden.etal.2012} \\
+
+    $g$     & Standard gravity
+            & 9.81
+            & \unit{m\,s^{-2}}
+            & \citet{Aschwanden.etal.2012} \\
+
+    $n$     & Glen exponent
+            & 3
+            & -
+            & \aref \\
+
+    $A_c$   & Ice hardness coefficient cold
+            & $3.61\times10^{-13}$
+            & \unit{Pa^{-3}\,s^{-1}}
+            & \citet{Paterson.Budd.1982} \\
+
+    $A_w$   & Ice hardness coefficient warm
+            & $1.73\times10^3$
+            & \unit{Pa^{-3}\,s^{-1}}
+            & \citet{Paterson.Budd.1982} \\
+
+    $Q_c$   & Flow law activation energy cold
+            & $6.0\times10^4$
+            & \unit{J\,mol^{-1}}
+            & \citet{Paterson.Budd.1982} \\
+
+    $Q_w$   & Flow law activation energy warm
+            & $13.9\times10^4$
+            & \unit{J\,mol^{-1}}
+            & \citet{Paterson.Budd.1982} \\
+
+    $R$     & Ideal gas constant
+            & 8.31441
+            & \unit{J\,mol^{-1}\,K^{-1}}
+            & \aref \\
+
+    $T_c$   & Flow law critical temperature
+            & 263.15
+            & \unit{K}
+            & \citet{Paterson.Budd.1982} \\
+
+    $f$     & Flow law water fraction coeff.
+            & 181.25
+            & -
+            & \citet{Lliboutry.Duval.1985} \\
+
+    $\beta$ & Clapeyron constant
+            & $7.9\times10^{-8}$
+            & \unit{K\,Pa^{-1}}
+            & \citet{Luthi.etal.2002} \\
+
+    $c_i$   & Ice specific heat capacity
+            & 2009
+            & \unit{J\,kg^{-1}\,K^{-1}}
+            & \citet{Aschwanden.etal.2012} \\
+
+    $c_w$   & Water specific heat capacity
+            & 4170
+            & \unit{J\,kg^{-1}\,K^{-1}}
+            & \citet{Aschwanden.etal.2012} \\
+
+    $k$     & Ice thermal conductivity
+            & 2.10
+            & \unit{J\,m^{-1}\,K^{-1}\,s^{-1}}
+            & \citet{Aschwanden.etal.2012} \\
+
+    $L$     & Water latent heat of fusion
+            & $3.34\times10^5$
+            & \unit{J\,kg^{-1}\,K^{-1}}
+            & \citet{Aschwanden.etal.2012} \\
+
+    \multicolumn{2}{l}{\emph{Basal sliding}} \\
+
+    $q$     & Pseudo-plastic sliding exponent
+            & 0.25
+            & -
+            & - \\
+
+    $v_{th}$& Pseudo-plastic threshold velocity
+            & 100.0
+            & \unit{m\,yr^{-1}}
+            & - \\
+
+    $c_0$   & Till cohesion
+            & 0.0
+            & Pa
+            & - \\
+
+    $\delta$& Effective pressure coefficient
+            & 0.02
+            & -
+            & - \\
+
+    $e_0$   & Till reference void ratio
+            & 0.69
+            & -
+            & \citet{Tulaczyk.etal.2000} \\
+
+    $C_c$   & Till compressibility coefficient
+            & 0.12
+            & -
+            & \citet{Tulaczyk.etal.2000} \\
+
+    $W_{max}$ & Maximal till water thickness
+            & 2.0
+            & m
+            & - \\
+
+    \multicolumn{2}{l}{\emph{Bedrock and lithosphere}} \\
+
+    $\rho_b$& Bedrock density
+            & 3300
+            & \unit{kg\,m^{-3}}
+            & \aref \\
+
+    $c_b$   & Bedrock specific heat capacity
+            & 1000
+            & \unit{J\,kg^{-1}\,K^{-1}}
+            & \citet{Ritz.1997} \\
+
+    $k_b$   & Bedrock thermal conductivity
+            & 3.0
+            & \unit{J\,m^{-1}\,K^{-1}\,s^{-1}}
+            & \citet{Ritz.1997} \\
+
+    $\nu_m$ & Mantle viscosity
+            & $1\times10^21$
+            & \unit{Pa\,s}
+            & \citet{Lingle.Clark.1985} \\
+
+    $\rho_l$& Lithosphere density
+            & 3300
+            & \unit{kg\,m^{-3}}
+            & \citet{Lingle.Clark.1985} \\
+
+    $D$     & Lithosphere flexural rigidity
+            & $5.0\times10^24$
+            & \unit{N}
+            & \citet{Lingle.Clark.1985} \\
+
+    \multicolumn{2}{l}{\emph{Surface and atmosphere}} \\
+
+    $T_s$   & Temperature of snow precipitation
+            & 273.15
+            & \unit{K}
+            & \aref \\
+
+    $T_r$   & Temperature of rain precipitation
+            & 275.15
+            & \unit{K}
+            & \aref \\
+
+    $F_s$   & Degree-day factor for ice
+            & $3.04\times10^{-3}$
+            & \unit{m\,K^{-1}\,day^{-1}}
+            & \citet{Shea.etal.2009} \\
+
+    $F_i$   & Degree-day factor for ice
+            & $4.59\times10^{-3}$
+            & \unit{m\,K^{-1}\,day^{-1}}
+            & \citet{Shea.etal.2009} \\
+
+    $\gamma$& Air temperature lapse-rate
+            & $6\times10^{-3}$
+            & \unit{K\,m{-1}}
+            & - \\
+
+    \bottomhline
+  \end{tabular}}
+  \belowtable{}
+\end{table*}
+
 % tab:records
@@
 
-% tab:records
+% tab:extrema
 \begin{table*}[t]

### 2014-07-17 cordillera-cycle.tex
Add minor corrections from Martin in Sect. 1--3.

@@
 First, only recent development of numerical ice sheet models and underlying
-scientific computing tools \citep{Bueler.Brown.2009, Balay.etal.2014} allows
-high-resolution numerical modelling of palaeo-glaciers on mountainous terrain
+scientific computing tools \citep{Bueler.Brown.2009, Balay.etal.2014} has
+allowed for high-resolution numerical modelling of glaciers and ice sheets on
+mountainous terrain
 over millenial time scales \citep[e.g.,][]{Golledge.etal.2012}. Second, the
@@
 seasonality and distinct patterns of precipition, thus requiring the use of
-high-resolution atmospheric forcing fields as input to the ice sheet model
+high-resolution atmospheric forcing fields as an input to the ice sheet model
 \citep{Seguinot.etal.2014}.
@@
 and two alkenone unsaturation index records from Northwest Pacific ocean
-sediment cores \citep{Herbert.etal.2001}. Model output is compared to
-geomorphological evidence in terms of the timing and extent of glaciation and
-patterns of deglaciation.
+sediment cores \citep{Herbert.etal.2001}. We then proceed to compare the model
+output to geological evidence and discuss the timing and extent of glaciation
+and the patterns of deglaciation.
 
@@
 Each simulation starts from assumed ice-free conditions at 120\,000 years ago
-(120\,ka), and runs to present. Our modelling domain of 1\,500 by 3\,000\,km
+(120\,ka), and runs to the present. Our modelling domain of 1\,500 by
+3\,000\,km
 encompasses the entire area covered by the Cordilleran ice sheet at the LGM
@@
 
-Atmospheric forcing of the model consists of a present-day monthly climatology,
+Atmospheric forcing of the model consists of a present-day monthly climatology
+\renote{
+    ``Monthly climatology'' is a common term in meteorology to design a set
+    of variables aggregated over several decades, and assumed to represent
+    some climate state.},
 $\{T_{m0}, P_{m0}\}$, where temperatures are modified by offset time series,
@@
 {Jouzel.etal.2007}, the Vostok ice core \citep{Petit.etal.1999}, and Ocean
-Drilling Program (ODP) sites 1012 and 1020, both located off the Californian
-coast \citep{Herbert.etal.2001}. Palaeo-temperatures from the GRIP and NGRIP
+Drilling Program (ODP) sites 1012 and 1020, both located off the coast of
+California \citep{Herbert.etal.2001}. Palaeo-temperatures from the GRIP and
+NGRIP
 records were calculated from oxygen isotope (\chem{\delta^{18}O}) measurements
@@
 glacial maximum (Table~\ref{tab:records}, Fig.~\ref{fig:timeseries}).
-In the rest of this paper, we occasionally refer to different model
-runs by the name of the proxy record used for the palaeo-temperature forcing.
 
@@
     Small change in notation to match with the kappa},
-on the one hand, and
-for differences between the the ice flow model basal topography, $b$, and the
+on the one hand, and for differences between the basal topography of the ice
+flow model, $b$, and the
 NARR reference topography, $b_{ref}$, on the other hand. All simulations use an
 annual temperature lapse rate of $\gamma = 6\,\unit{{\degree}C\,km^{-1}}$.
+In the rest of this paper, we refer to different model runs by the name of the
+proxy record used for the palaeo-temperature forcing.
 
@@
 16.9\,ka during MIS~2 (Fig.~\ref{fig:timeseries}). An ice volume minimum is
-consistently attained between 52.2 and 45.6\,ka during MIS~3. However, the
+consistently reached between 52.2 and 45.6\,ka during MIS~3. However, the
 magnitude and precise timing of these three events depend significantly on the
@@
 throughout the simulation length. The NGRIP atmospheric forcing is the only one
-that results in a larger ice volume during MIS~4 than during the last glacial
-maximum (MIS~2).
+that results in a larger ice volume during MIS~4 than during the LGM (MIS~2).
 
@@
 simulation predicts an early maximum in ice volume at 29.5\,ka, followed by
-slower deglaciation than modelled using other palaeo-temperature records.
-Finally, model runs forced by Antarctic palaeo-temperature records produce a
+slower deglaciation than modelled using the other palaeo-temperature records.
+Finally, model runs forced by Antarctic palaeo-temperature records results in a
 rapid an uninterrupted deglaciation after the last glacial maximum, whereas the
-simulation driven by the GRIP palaeo-temperature record also yields a rapid
+simulation driven by the GRIP palaeo-temperature record also result in a rapid
 deglaciation but in three steps, interrupted by two stand-stills and readvances
@@
 Despite such differences in the timing of attained volume maxima and mimima,
-snapshots of model output show relatively consistent patterns of glaciation
+all the model runs show relatively consistent patterns of glaciation
 (Fig.~\ref{fig:snapshots}). As a result of scaling factors applied to the

### 2014-07-17 cordillera-cycle.tex
Remove thousands separator (just checked in CCLI).

@@
 
-Each simulation starts from assumed ice-free conditions at 120\,000 years ago
-(120\,ka), and runs to the present. Our modelling domain of 1\,500 by
-3\,000\,km
+Each simulation starts from assumed ice-free conditions at 120000 years ago
+(120\,ka), and runs to the present. Our modelling domain of 1500 by 3000\,km
 encompasses the entire area covered by the Cordilleran ice sheet at the LGM
@@
 Fig.~\ref{fig:snapshots}, lower panels) appear highly similar from one
-simulation to the next. They include a ca. 1\,500\,km-long central divide
-located about 3\,500\,m\ above sea level (a.s.l.) along the spine of the Rocky
+simulation to the next. They include a ca. 1500\,km-long central divide
+located about 3500\,m\ above sea level (a.s.l.) along the spine of the Rocky
 Mountains. The volumes of these modelled last glacial maximum ice sheets are
@@
 ice caps centred on major mountain ranges of the North American Cordillera
-(Fig.~\ref{fig:duration}). A 2\,500\,km-long continuous ice cover spanning from
+(Fig.~\ref{fig:duration}). A 2500\,km-long continuous ice cover spanning from
 the Alaska Range in the northeast to the Rocky Mountains in the southwest only

### 2014-07-17 cordillera-cycle.tex
Polish intro and model set-up.

@@
 Cordillera have been more extensive than today. At the Last Glacial
-Maximum (LGM), a continuous blanket of ice, the Cordilleran ice sheet\renote{
-    In \citep{Seguinot.etal.2014}, The Cryosphere changed all my occurences of
+Maximum (LGM\renote{
+    I decided to adopt ``LGM'', which is very common, but not ``CIS'', for the
+    sake of readability. This corresponds to abbreviations used in
+    \citet{Seguinot.etal.2014} (and now, in the kappa).}),
+a continuous blanket of ice, the Cordilleran ice sheet\renote{
+    In \citet{Seguinot.etal.2014}, The Cryosphere changed all my occurences of
     Cordilleran (Greenland etc) \textbf{I}ce \textbf{S}heet to lowercase
@@
     \citep{Turner.etal.2013}
-and dated glacial erratics indicate an extensive Marine Oxygen Isotope Stage
-(MIS)~4 glaciation
+and dated glacial erratics
     \citep{Ward.etal.2007, Ward.etal.2008, Briner.Kaufman.2008,
-           Stroeven.etal.2010, Stroeven.etal.2014},
+           Stroeven.etal.2010, Stroeven.etal.2014}
+indicate an extensive Marine Oxygen Isotope Stage (MIS)~4 glaciation,
 yet it is not known whether other regions in the study area were affected.
@@
 Geomorphological evidence from south-central British Columbia indicates a rapid
-deglaciation, including an early emergence of elevated areas while thin
-stagnant ice still covered surrounding lowlands
+deglaciation, including an early emergence of the elevated areas while thin,
+stagnant ice still covered the surrounding lowlands
     \citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a}.
@@
 it appears to be more sparse than for formerly glaciated regions surrounding
-the North Atlantic \aref, thus allowing for considerable uncertainty concerning
+the North Atlantic \aref\footnote{
+    Do we need to add a couple of refs. from Britain, Scandinavia, the Alps,
+    Eastern America here? Or a review paper? Could yuo help me with this?},
+thus allowing for considerable uncertainty concerning
 the possibility of a regional late glacial cold reversal.
@@
   \item What was the scale of glaciation prior to the last glacial maximum?
-  \item Which were its primary dispersal centres? Do they reflect stable or
+  \item Which were the primary dispersal centres? Do they reflect stable or
     ephemeral configurations?
   \item How rapid was the last deglaciation? Did it include late glacial
-    stand-stills or re-advances?
+    standstills or readvances?
 \end{itemize}
@@
     (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991},
-     \citealp{Marshall.etal.2000})
+     \citealp{Marshall.etal.2000}),
 the ubiquitously mountainous
@@
 Because climate conditions over the last glacial cycle are subject to
-considerable uncertainty, our palaeoclimate forcing is a simplistic approach
+considerable uncertainty, our palaeo-climate forcing is a simplistic approach
 including temperature and precipitation fields derived from a
 present-day atmospheric reanalysis \citep{Mesinger.etal.2006,
-Seguinot.etal.2014} supplemented by a lapse-rate correction
+Seguinot.etal.2014} supplemented by lapse-rate corrections
 and temperature offset time series. The latter are obtained by scaling six
 different palaeo-temperature reconstructions from proxy records around the
-globe, including two oxygen isotopes records from Greenland ice cores
-\citep{Dansgaard.etal.1993, Andersen.etal.2004}, two oxygen isotopes
+globe, including two oxygen isotope records from Greenland ice cores
+\citep{Dansgaard.etal.1993, Andersen.etal.2004}, two oxygen isotope
 records from Antarctic ice cores \citep{Petit.etal.1999,Jouzel.etal.2007},
@@
 % -----------------------------------------------------------------------------
-
-\todo{Move parameter values into a separate table.}
 
@@
     forcing'' is more appropriate: ``climate'' may also encompass the oceanic
-    component, not really assessed here.}
-forcing, and computes the evolution of ice extent
+    component, which is not assessed here.}
+forcing. It computes the evolution of ice extent
 and thickness over time, the thermal and dynamic\renote{
@@
 is applied as a constant value of 70\,\unit{mW\,m^{-2}} at 3\,km depth
-(Sect.~\ref{sec:icedyn}). Atmospheric forcing is provided by a monthly
-climatology from the North American Regional Reanalysis
+(Sect.~\ref{sec:icedyn}). Surface mass balance is computed using a positive
+degree-day (PDD) model (Sect.~\ref{sec:surface}). Atmospheric forcing is
+provided by a monthly climatology from the North American Regional Reanalysis
 \citep[NARR;][]{Mesinger.etal.2006} perturbated by time-dependent and
@@
 components. The Shallow Shelf Approximation (SSA) is used as a ``sliding law''
-for the Shallow Ice Approximation (SIA) by adding both velocity solutions
+for the Shallow Ice Approximation (SIA) by adding velocity solutions of the
+two approximation
 \citep[Eqns.~7--9 and 15]{Bueler.Brown.2009, Winkelmann.etal.2011}\renote{
@@
 heat flow measurements. In the low-resolution simulations, the vertical grid
-consists of to 51~enthalpy layers in the ice sheet and 31~temperature layers in
-the bedrock. In the high-resolution simulations, up to 101~ice layers, and
+consists of up to 51~enthalpy layers in the ice sheet and 31~temperature layers
+in the bedrock. In the high-resolution simulations, up to 101~ice layers, and
 61~bedrock layers are used.
@@
 A~pseudo-plastic sliding law \citep{Bueler.Pelt.2014} relates the
-bed-parallel shear stress to the sliding velocity. The yield stress, $\tau_c$,
+bed-parallel shear stresses to the sliding velocity\footnote{
+    Should I pull in this equation, too?}.
+The yield stress, $\tau_c$,
 is modelled using the Mohr--Coulomb criterion,
@@
     sensitivity test on Greenland.}.
-It is taken to be a~function of modern bed elevation, with lower values at
-lower elevations (<200\,m above sea level), thus accounting for a~weakening of
+It is taken to be a piecewise-linear function of modern bed elevation, with
+the lowest value below modern sea level (0\,m above sea level, m~a.s.l) and the
+highest value above the approximate elevation of highest shorelines
+(200\,m~a.s.l), thus accounting for a~weakening of
 till associated with the presence of marine sediments. Effective pressure, $N$,
@@
 \end{equation}
-where $delta$ is choosed as 0.95, $e_0$ is a measured reference void ratio and
-$C_c$ a measured compressibility coefficient. The amount of water at the base,
-$W$, varies from zero to $W_{max}=2$\,m, a threshold above which instantaneous
-drainage is assumed\renote{
-    Regarding a parameter table, it is in the kappa, I am skipping it for now
-    but will try to include it soon.}.
+where $delta$ is chosen as 0.02, $e_0$ is a measured reference void ratio and
+$C_c$ a measured compressibility coefficient (Table~\ref{tab:params}). The
+amount of water at the base, $W$, varies from zero to $W_{max}=2$\,m, a
+threshold above which instantaneous drainage is assumed.
 Finally, the bedrock topography responds to ice load\renote{
@@
 half-space\renote{
-    I had put an emphasis on semi-infinite at this is the key assumption of
-    that model, allowing effective solution in Fourier domain. This is the
-    whole point of Ed's paper and probably the reason why he picked it for PISM
-    But yes, I guess that ``lower half-space'' is fine, too...}
-\citep{Lingle.Clark.1985,Bueler.etal.2007}.
+    I had put an emphasis on semi-infinite as this is, from a mathematical
+    point of view, the key assumption of that model, allowing effective
+    solution in Fourier domain. This is the whole point of Ed's paper and
+    probably the reason why he picked it for PISM. But yes, I guess that
+    ``lower half-space'' is fine, too...}
+\citep[Table~\ref{tab:params};][]{Lingle.Clark.1985,Bueler.etal.2007}.
 
@@
 \subsection{Surface mass balance}
+\label{sec:surface}
 % -----------------------------------------------------------------------------
@@
     consisting of ice accumulation and ice ablation, which I introduce straight
-    away. I don't know what kind of ``opening sentence'' can be added here.}
+    away. I don't see what kind of ``opening sentence'' can be added here.}
 Ice surface accumulation and ablation are computed from monthly mean
-near-surface air temperature $T_m$, monthly standard deviation of near-surface
-air temperature $\sigma$, and monthly precipitation $P_m$, using
+near-surface air temperature, $T_m$, monthly standard deviation of near-surface
+air temperature, $\sigma$, and monthly precipitation, $P_m$, using
 a~temperature-index model \citep[e.g.,][]{Hock.2003}. Accumulation is equal to
@@
     should do it here. The subject is too marginal for the present study.}.
-The ablation model incorporates degree-day factors of
-3.04\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for snow and
-4.59\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for ice, as derived from
+Degree-day factors for snow and ice melt are derived from
 mass-balance measurements on contemporary glaciers from the Coast Mountains and
-Rocky Mountains in British Columbia \citep{Shea.etal.2009}.
+Rocky Mountains in British Columbia
+\citep[Table~\ref{tab:params};][]{Shea.etal.2009}.
 
@@
 
-Atmospheric forcing of the model consists of a present-day monthly climatology
+Atmospheric forcing of the model consists of a present-day monthly climatology%
 \renote{
-    ``Monthly climatology'' is a common term in meteorology to design a set
+    ``Climatology'' is a common term in meteorology, used to design a set
     of variables aggregated over several decades, and assumed to represent
@@
 The present-day monthly climatology, $\{T_{m0}, P_{m0}\}$, was computed from
-near-surface air temperature and precipitation rate fields of the NARR
+near-surface air temperature and precipitation rate fields from the NARR
 \citep{Mesinger.etal.2006}, averaged between 1979 and 2000. Modern climate of the
@@
 while temperature reconstructions from Antarctic and oceanic cores were
-provided as such. All records were scaled linearly in
-order to simulate realistic and comparable ice limits at the last
-glacial maximum (Table~\ref{tab:records}, Fig.~\ref{fig:timeseries}).
+provided as such. All records were scaled linearly (Table~\ref{tab:records}) in
+order to simulate realistic and comparable ice extents at the LGM
+(Fig.~\ref{fig:timeseries}).
 
@@
 simulation driven by the GRIP palaeo-temperature record also result in a rapid
-deglaciation but in three steps, interrupted by two stand-stills and readvances
+deglaciation but in three steps, interrupted by two standstills and readvances
 of the ice margin.
@@
 simulation to the next. They include a ca. 1500\,km-long central divide
-located about 3500\,m\ above sea level (a.s.l.) along the spine of the Rocky
+located about 3500\,m~a.s.l. along the spine of the Rocky
 Mountains. The volumes of these modelled last glacial maximum ice sheets are
@@
 Squamish valleys off the southern tip of the Coast Mountains, indicates
-consecutive glacier maximal (or stillstands while in overall retreat), one of
+consecutive glacier maximal (or standstills while in overall retreat), one of
 which one corresponds to the Younger Dryas chronozone \citep{Clague.etal.1997,

### 2014-07-17 cordillera-cycle.tex
Rework sensitivity (results) part.

@@
 
-Despite large differences in the input atmospheric forcing (temperature offset
-time-series; Fig.~\ref{fig:timeseries}), model output presents consistent
+Despite large differences in the input atmospheric forcing
+(Fig.~\ref{fig:timeseries}, upper panel), model output presents consistent
 features that can be observed across the range of forcing data used. In all
@@
 glacial cycle, except during two major glacial events which occur between 61.9
-and 55.4\,ka during Marine Oxygen Isotope Stage (MIS)~4, and between 29.5 and
-16.9\,ka during MIS~2 (Fig.~\ref{fig:timeseries}). An ice volume minimum is
+and 55.4\,ka during MIS~4, and between 29.5 and 16.9\,ka during MIS~2
+(Fig.~\ref{fig:timeseries}, lower panel). An ice volume minimum is
 consistently reached between 52.2 and 45.6\,ka during MIS~3. However, the
@@
 (Table~\ref{tab:extrema})\renote{
-    Hopefully this new table will assess several of your comments at once.}.
+    This new table contains modelled ages, volumes and extents at MIS~4--2}.
 
@@
     Until we run out of other ajectives, I don't think that ``modest'' and
-    ``generous'' are adequate terms for describing ice volumes.}
+    ``generous'' are necessary terms for describing ice volumes.}
 modelled ice volumes during MIS~4 and larger ice volumes during MIS~3.
@@
 throughout the simulation length. The NGRIP atmospheric forcing is the only one
-that results in a larger ice volume during MIS~4 than during the LGM (MIS~2).
-
-While simulations driven by GRIP and the two Antarctic palaeo-temperature
+that results in a larger ice volume during MIS~4 than during the MIS~2.
+
+While simulations driven by the GRIP and the two Antarctic palaeo-temperature
 records attain a last ice volume maximum between 19.5 and 16.9\,ka, those
-informed by NGRIP and the two oceanic palaeo-temperature records attain their
+informed by the NGRIP and the two oceanic palaeo-temperature records attain their
 maximum ice volume thousands of years earlier. Moreover, the ODP~1012 run
@@
 slower deglaciation than modelled using the other palaeo-temperature records.
-Finally, model runs forced by Antarctic palaeo-temperature records results in a
-rapid an uninterrupted deglaciation after the last glacial maximum, whereas the
-simulation driven by the GRIP palaeo-temperature record also result in a rapid
-deglaciation but in three steps, interrupted by two standstills and readvances
-of the ice margin.
+Finally, model runs forced by Antarctic palaeo-temperature records result in a
+rapid and uninterrupted deglaciation after the last glacial maximum, whereas the
+simulation driven by the GRIP palaeo-temperature record also results in a rapid
+deglaciation but in three steps, separated by two ice volume standstills.
 
@@
 
-Despite such differences in the timing of attained volume maxima and mimima,
-all the model runs show relatively consistent patterns of glaciation
-(Fig.~\ref{fig:snapshots}). As a result of scaling factors applied to the
-different palaeo-temperature proxy records (Table~\ref{tab:records}), modelled
-ice sheet geometries during the last glacial maximum (MIS~2;
-Fig.~\ref{fig:snapshots}, lower panels) appear highly similar from one
-simulation to the next. They include a ca. 1500\,km-long central divide
-located about 3500\,m~a.s.l. along the spine of the Rocky
-Mountains. The volumes of these modelled last glacial maximum ice sheets are
-contained within a close range from 8.24 to 8.66\,m of sea-level equivalent
-(s.l.e; Table~\ref{tab:extrema}).
-A central ice cap persists in all simulations over the Skeena
-Mountains during the MIS~3 ice volume minima (Fig.~\ref{fig:snapshots}, middle
-panels). However, the size of the Skeena Mountains ice cap, and the dimensions
-of the Cordilleran ice sheet during the MIS~4 glaciation, depend sensitively on
-the choice of palaeo-temperature record used to drive the model.
-
-\todo{Add general statements on the scale of glaciation.}
+Despite such differences in the timing of attained volume maxima and mimima
+(Table~\ref{tab:extrema}), all the model runs show relatively consistent
+patterns of glaciation (Fig.~\ref{fig:snapshots}, upper panels). During MIS~4,
+all simulation produce an extensive ice sheet, covering an area of at least
+half that attained during MIS~4. However, the corresponding maximum ice volumes
+vary differ significantly between model runs, and cary between 3.88 and~8.69\,m
+sea-level equivalent (m.~s.l.e; Table~\ref{tab:extrema}).
+
+During the MIS~3 ice volume minima, a central ice cap persists in all
+simulations over the Skeena Mountains (Fig.~\ref{fig:snapshots}, middle
+panels). Although this ice cap is present in all simulations, its dimensions
+depend sensitively on the choice of palaeo-temperature record used to drive the
+model. Modelled ice volume minima vary between 0.89 and 2.44\,m~s.l.e.
+(Table~\ref{tab:extrema}).
+
+The modelled ice sheet geometries during the last glacial maximum (MIS~2;
+Fig.~\ref{fig:snapshots}, lower panels) include a ca. 1500\,km-long central
+divide located about 3500\,m~a.s.l. along the spine of the Rocky Mountains and
+appear highly similar from one simulation to the next. Although the similatity
+in modelled ice extent is a direct result from the choice of scaling factors
+applied to the different palaeo-temperature proxy records
+(Table~\ref{tab:records}), it is interesting to note that modelled maximum ice
+volumes are also contained within a close range of 8.24 to 8.66\,m~s.l.e.
+(Table~\ref{tab:extrema}).
 

### 2014-07-17 cordillera-cycle.tex
Polish 4.1 (parts missing).

@@
 \subsection{Overview}
+\label{sec:overview}
 % -----------------------------------------------------------------------------
@@
         {Comparison to the geologic record\renote{
-    Arjen, you have proposed 2.1 Evaluation of model runs agains field
-    evidence, 2.2 Palaeoglaciology of the Cordilleran ice sheet, and 2.3 The
-    last deglaciation; but I don't think these headers work. In my vision of
-    the paper, this entire ``discussion'' part is all about evaluation of model
-    runs agains field evidence and palaeoglaciology. Because of this, and
-    because we also present new ``results'', or rather different views of the
-    same results, I propose to rename it to a more explicit ``Comparison to the
-    geologic record'' (or similar). I have also renamed the ``results'' to
-    ``Sensitivity to atmospheric forcing time-series'', which is actually what
-    that section is about: raw model output, no discussion against the geologic
-    record.}}
+    Arjen has proposed: 2.1 Evaluation of model runs agains field evidence, 2.2
+    Palaeoglaciology of the Cordilleran ice sheet, and 2.3 The last
+    deglaciation; but I don't think that these headers work. In my view, this
+    entire ``discussion'' part is all about evaluation of model runs agains
+    field evidence and palaeoglaciology. It is necessary that this comparison
+    starts here, in order to ``kick out'' the records that perform badly (ODP*,
+    NGRIP) and justify that the rest of the analysis is performed on two
+    (high-resolution) simulations only. Such figures as
+    Figs.~\ref{fig:duration}--\ref{fig:erosion} and Fig.~\ref{fig:deglac},
+    which I consider as different views of the results rather than new results
+    (apart the fact that they use higher resolution) are drawn for the purpose
+    of comparing model results against geological evidence. I propose to rename
+    this section to a more explicit ``Comparison to the geologic record'' (or
+    similar). I have also renamed the ``results'' to ``Sensitivity to
+    atmospheric forcing time-series'', which is actually what that section is
+    about: raw model output, no discussion against the geologic record. This
+    new discussion is divided in three parts: 4.1 Glacial maxima, where we
+    study the maximum stages, relatively well-documented by the geology; 4.2
+    Nucleation centres, where we study ``intermediate'' states, for which less
+    is known, but the model can say something; and 4.3 The last deglaciation,
+    which deserves a separate section because many people are interested in
+    this period, and much geological evidence is available.}}
 \label{sec:discussion}
@@
 is obtained during MIS~4 (61.9--55.4\,ka) and the second during MIS~2
-(29.5--16.9\,ka). These events broadly correspond in timing to the Gladstone
+(29.5--16.9\,ka; Figs.~\ref{fig:timeseries} and~\ref{fig:snapshots};
+Table~\ref{tab:extrema}). These events broadly correspond in timing to the
+Gladstone
 (MIS~4) and McConnell (MIS~2) glaciations documented by geological evidence for
@@
 
-However, the exact timing of modelled MIS~2 maxima depend strongly on the
+However, the exact timing of modelled MIS~2 maxima depends strongly on the
 choice of applied palaeo-temperature record, which allows for a more in-depth
@@
 the southern Cordilleran ice sheet margin has been constrained by radiocarbon
-dating on wood between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,kyr\,BP}
+dating on wood between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,ka\,BP}
 \citep{Porter.Swanson.1998}.
-These dates have been confirmed by the offshore sediment record, which shows an
-increase of glaciomarine sedimentation between 19.5 and
-16.2\,\unit{\chem{^{14}C}\,cal\,kyr\,BP} \citep{Cosma.etal.2008}. Radiocarbon
+These dates have been confirmed by the offshore sedimentary record, which shows
+an increase of glaciomarine sedimentation between 19.5 and
+16.2\,\unit{\chem{^{14}C}\,cal\,ka\,BP} \citep{Cosma.etal.2008}. Radiocarbon
 dating of the northern Cordilleran ice sheet margin is much less constrained
 but straddles presented constraints from the southern margin. However,
-cosmogenic exposure dating by \citep{Stroeven.etal.2010, Stroeven.etal.2014}
+cosmogenic exposure dating
 places the timing of maximum CIS extent during the McConnell glaciation close
-to 17\,ka.
+to 17\,ka \citep{Stroeven.etal.2010, Stroeven.etal.2014}.
 
@@
 extent that may be compatible with these field constraints
-(Fig.~\ref{fig:timeseries}, lower panel; Fig.~\ref{fig:snapshots}, lowest
-panels). Simulations driven by the NGRIP, ODP~1012 and ODP~1020
+(Fig.~\ref{fig:timeseries}, lower panel; Table~\ref{tab:extrema}).
+Simulations driven by the NGRIP, ODP~1012 and ODP~1020
 palaeo-temperature records, on the contrary, yield MIS~2 maximum Cordilleran
@@
 palaeo-temperature reconstructions (Fig.~\ref{fig:timeseries}, upper panel;
-\citealp[Fig.3]{Herbert.etal.2001}). However, this
+\citealp[Fig.~3]{Herbert.etal.2001}). However, this
 early warming is a local effect, corresponding to a weakening of the California
-current \citep[Fig.~3]{Herbert.etal.2001}. The California current, driving cold
+current \citep{Herbert.etal.2001}. The California current, driving cold
 waters southwards along the south-western coast of North America\renote{
@@
 has been shown to have weakened during each peak global glaciation (in SPECMAP)
-durint the past 550\,ka, resulting in paradoxically warmer sea-surface
-temperatures at the locations of the ODP~1012 and ODP~1020 sites also during
-the LGM \citep{Herbert.etal.2001}.
+durint the past 550\,ka, including the LGM, resulting in paradoxically warmer
+sea-surface temperatures at the locations of the ODP~1012 and ODP~1020 sites
+\citep{Herbert.etal.2001}.
 
@@
 
-\todo{say something about NGRIP.}
-
-... Hence, we focus the rest of our analysis on simulations forced by palaeo-temperature records from GRIP and EPICA ice cores (Fig.~\ref{fig:timeseries}, lower panel, dotted lines).
-
-\todo{avoid repetition with the next paragraph.}
-
-These results indicate that a palaeo-temperature proxy record located in the
-direct vicinity of the former Cordilleran ice sheet would potentially lead to
-improved understanding of the history of advance and retreat of the western ice
-sheet margin. Because, to our knowledge, no
-such record yet exists, we focus the rest of our analysis on simulations forced
-by palaeo-temperature records from GRIP and EPICA ice cores
-(Fig.~\ref{fig:timeseries}, lower panel, dotted lines).
+\todo{Write something about NGRIP.}
+
+Hence, we focus the rest of our analysis on simulations forced by
+palaeo-temperature records from GRIP and EPICA ice cores. To allow for a more
+detailed comparison against the geologic record, these two simulations were
+re-run on the high-resolution grid (Fig.~\ref{fig:timeseries}, lower panel,
+dotted lines).
 
@@
 
-\todo{Add MIS~2 ice topo and velocity map. Include two paragraphs from Martin.}
+\todo{Add MIS~2 ice topo and velocity map. Include two paragraphs from Martin.
+      Describe model limitations referring to \citet{Seguinot.etal.2014}}
 
@@
 \todo{Add MIS~4 ice topo and velocity map. Discuss that some of the margins
-      are more extensive during MIS~4 than during MIS~2, in the GRIP simu.}
+      are more extensive during MIS~4 than during MIS~2 in the GRIP simulation}
 

### 2014-07-17 cordillera-cycle.tex
Polish section 4.2.1.

@@
 
-\todo{Put more equal weight to the northern sector}
-
-\todo{Add some point, refer back to Fig.~\ref{fig:atm}.}
-
-\todo{Mention the major nucleation centres. Growth of the CIS from the
-      highlands.}
-
-Although snapshots from the model output (Fig.~\ref{fig:snapshots}) yield a
-clear spatial picture of modelled glaciation patterns at specific time frames,
-they often can not be directly compared to geomorphological evidence, which is
-by nature time-transgressive. In this section, we use high-resolution
-simulations driven by the GRIP and EPICA records to locate the major nucleation
-centres of the Cordilleran ice sheet and assess their longevity during the last
-glacial cycle.
-
-\todo{Rewrite this opening paragraph. Fig.~\ref{fig:snapshots} can actually be
-      compared to geological evidence.}
-
-During most of the glacial cycle, modelled ice cover is restricted to disjoint
-ice caps centred on major mountain ranges of the North American Cordillera
-(Fig.~\ref{fig:duration}). A 2500\,km-long continuous ice cover spanning from
-the Alaska Range in the northeast to the Rocky Mountains in the southwest only
-exists\renote{
+The accuracy of palaeo-glaciological reconstructions is generally biased
+towards the stages of maximum extent, for which the geological evidence is
+generally more detailed and more complete than for other periods. However,
+these maximum stages are, by nature, extreme configurations, which do not
+necessarily represent the dominant patterns of glaciation throughout the period
+of ice cover.
+
+For the Cordilleran ice sheet, geological evidence from radiocarbon dating
+    \citep{Clague.etal.1980, Clague.1985, Clague.1986, Porter.Swanson.1998,
+           Menounos.etal.2008},
+cosmogenic exposure dating
+    \citep{Stroeven.etal.2010, Stroeven.etal.2014, Margold.etal.2014},
+bedrock deformation in response to former ice loads
+    \citep{Clague.James.2002, Clague.etal.2005},
+and offshore sedimentary records
+    \citep{Cosma.etal.2008, Davies.etal.2011}
+indicate that the LGM maximum extent was short-lived. To compare this finding
+to our simulations, we use numerical modelling output covering the whole
+simulation period to compute durations of ice cover throughout the last glacial
+cycle (Fig.~\ref{fig:duration}).
+
+The resulting maps show that, during most of the glacial cycle, modelled ice
+cover is restricted to disjoint ice caps centred on major mountain ranges of
+the North American Cordillera (Fig.~\ref{fig:duration}, blue areas). A
+2500\,km-long continuous ice cover spanning from the Alaska Range in the
+northeast to the Rocky Mountains in the southwest only exists\renote{
     I chosed to write about model output in the present form to discern in from
     the reality.}
-for at most 29\,kyr in total over the entire last glacial cycle. However,
+for at most 29\,ka in total over the entire last glacial cycle
+(Fig.~\ref{fig:duration}, hatched areas). However,
 except for its marine margin and the northern foothills of the Alaska Range,
 the maximum extent of the ice sheet is attained for an even more brief period
-of a few thousand years (Figs.~\ref{fig:timeseries}--{fig:duration}). This
+of a few thousand years (Figs.~\ref{fig:duration}, red areas). This
 result illustrates that the maximum extents of the modelled ice sheet during
 MIS~4 and 2 were both short-lived and therefore out of balance with
-contemporary climate, which confirms previous inferences of a short-lived
-Cordilleran ice sheet based on the geological record \citep{Clague.etal.1980,
-Stroeven.etal.2010}.
+contemporary climate.
+
+A notable exception to the transient character of the LGM Cordilleran ice sheet
+is the northern slope of the Alaska Range, where modelled glaciers are confined
+to the foothills during the entire simulation period (Fig.~\ref{fig:duration},
+AR). This apparent insensitivity of modelled glacial extent to temperature
+fluctuations results from a combination of low precipitations, high summer
+temperatures and high temperature standard deviations in this sector of the
+modelling domain (Fig.~\ref{fig:atm}). This result could potentially explain
+the local distribution of glacial deposits, which indicates that these glaciers
+have remained small throughout the Pleistocene \citep{Kaufman.Manley.2004}.
 

### 2014-07-17 cordillera-cycle.tex
Polish section 4.2.2.

@@
 
-Pehaps one of the most persistent landscape feature is the Skeena Mountains ice
-cap, which persisted throughout the entire last glacial cycle
+It is generally believed that the Cordilleran ice sheet formed by the
+coalescence of of several mountain-centred ice caps \citep{Davis.Mathews.1944}.
+In our simulations, major ice-dispersal centres, visible on the modelled ice
+cover duration maps (Fig.~\ref{fig:duration}), are located over the Coast
+Mountains (CM), the Columbia and Rocky mountains (CR), the Skeena Mountains
+(SM), and the Selwyn and MacKenzie Mountains (SMK). Although the Coast,
+Columbia, Rocky and Skeena Mountains (CM, CR, SM) are covered by mountain
+glaciers for most of the model time, providing durable nucleation centre for
+the initiation of ice sheet growth, this is not the case for the Selwyn and
+MacKenzie mountains (SMK), where ice cover on the highest peaks is limited to a
+fraction of the last glacial cycle. In other words, the Selwyn and MacKenzie
+mountains appear as a secondary ice-dispersal centre which ``activates'' only
+during the coldest periods of the last glacial cycle.
+
+Perhaps the most striking feature displayed by the distributions of modelled
+ice cover is the persistence of the Skeena Mountains ice cap throughout the
+entire last glacial cycle and its predominance over other ice-dispersal centres
 (Figs.~\ref{fig:snapshots} and~\ref{fig:duration}). Regardless of the applied
-forcing (Fig.~\ref{fig:snapshots}, middle panels), this ice cap appears to
-survive MIS~3, and serves as a nucleation centre at the onset of the glacial
-readvance towards the LGM (MIS~2). The importance of residual ice for Noeth
+forcing, this ice cap appears to survive MIS~3 (Fig.~\ref{fig:snapshots},
+middle panels), and serves as a nucleation centre at the onset of the glacial
+readvance towards the LGM (MIS~2). The importance of residual ice for North
 American glacial history leading up to the LGM has been illustrated by the
 MIS~3 residual ice bodies in northern and eastern Canada as nucleation centres
-for a much more extensive MIS~2 configuration \citep{Kleman.etal.2010}.
+for a much more extensive MIS~2 configuration \citep{Kleman.etal.2010}\renote{
+    Does this sentence really fit here? I feel that this sudden mention of the
+    Laurentide ice sheet perturbates a bit the flow. Moreover, knowing that
+    the Laurentide ice sheet is probably the main driver for the 100\,ka
+    sea-level cycle, it would have been surprising if it had completely
+    vanished at MIS~3. But for the Cordilleran ice sheet, usually thought of
+    as fast-responding and intermittent, I feel that the presence of residual
+    ice is less trivial.}.
 
 The presence of a Skeena Mountains ice cap during most of the last glacial
-cycle can be explained by a more widespread\renote{
-    I mean widespread, precipitation there is lower than elsewhere.},
-current winter precipitation for that region there than for other parts of the
+cycle can be explained by a currently more widespread\renote{
+    I mean widespread, precipitation there is lower than elsewhere.}
+winter precipitation for that region there than for other parts of the
 modelling domain (Fig.~\ref{fig:atm}). Along most of the north-western coast of

### 2014-07-17 cordillera-cycle.tex
Polish section 4.2.3.

@@
     (high-resolution) simulations only. Such figures as
-    Figs.~\ref{fig:duration}--\ref{fig:erosion} and Fig.~\ref{fig:deglac},
+    Figs.~\ref{fig:duration}--\ref{fig:warmfrac} and Fig.~\ref{fig:deglac},
     which I consider as different views of the results rather than new results
@@
 mountains appear as a secondary ice-dispersal centre which ``activates'' only
-during the coldest periods of the last glacial cycle.
+during the coldest periods of the last glacial cycle. The Northern Rocky
+Mountains (Fig.~\ref{fig:duration}, NR), do not act as a nucleation centre,
+but rather as a pinning point for the ice margin coming from the west.
 
@@
 
-\todo{Explain why it would not be good with general statements on thermal
-      evolution of ice cover.}
-
-[In addition, the erosional landscape of the Skeena mountain bears one of the
-strongest glacial imprint found within the area formerly covered by the
-Cordilleran ice sheet \aref%
-\footnote{\textbf{Arjen, Martin, Johan,} can we use a reference here to support
-    this point (observations, mapping, shaded relief)? Alternatively, we could
-    maybe use a photograph, if any of you has one. It would add a bit of
-    variation into the figure list. \textbf{Johan}, you seemed to support this
-    this point during our discussion, any idea?}.
-Here we suggest that persistent ice cover (Fig.~\ref{fig:duration})
-associated with basal ice temperatures at the pressure-melting point
-(Fig.~\ref{fig:warmfrac} and ~\ref{fig:warmbase}%
-\footnote{Fig.~\ref{fig:erosion} is tricky because it may be significantly
-    affected by the constant-geothermal heat flux assumption, which is not
-    valid for this region. Arjen, Johan and I decided to leave it for my kappa
-    for now. Regarding Figs.~\ref{fig:warmfrac} and ~\ref{fig:warmbase}, I
-    still wonder if it is useful to show both. Basically we have
-    Fig.~\ref{fig:warmbase}\,=\,Fig.~\ref{fig:duration}\,*\,Fig.~\ref{fig:warmfrac}.
-    Maybe the reader can perform this multiplication visually without need of
-    Fig.~\ref{fig:warmbase}? Tell me what you think of it.})
-may partly explain this spectacular landscape.]
-
 A correlation is observed between the modelled duration of warm based ice cover
@@
 and the Saint-Elias Mountains, and radiating off the Skeena Mountains
-(Figs.~\ref{fig:duration}, \ref{fig:warmbase} and~X\renote{
-    Yes it could be interesting to compare basal flow patterns with Fig.~2 of
-    \citet{Kleman.etal.2010}, but I don't think that I have the time for this
-    right now.}).
+(Figs.~\ref{fig:duration} and \ref{fig:warmbase};
+\citealp[Fig.~2]{Kleman.etal.2010})\renote{
+    Yes, I agree that it would be good to eventually reproduce here the basal
+    flow patterns from Johan's map \citep[Fig.~2]{Kleman.etal.2010}, but I
+    don't think that I will have the time for this before thesis submission}.
 The Skeena Mountains, for example, indeed bear a strong glacial imprint that
 indicates ice drainage in a system of distinct glacial troughs emanating in a
-radial pattern from the centre of the mountain range (Fig. X). We suggest that
+radial pattern from the centre of the mountain range
+\citep[Fig.~2]{Kleman.etal.2010}. We suggest that
 persistent ice cover (Fig.~\ref{fig:duration}) associated with basal ice
@@
 and~\ref{fig:warmbase}) explains the large-scale glacial erosional imprint on
-the landscape. In contrast, a well-developed network of glacial valleys west of
-the MacKenzie Mountains \citep[Fig.~X;][]{Kleman.etal.2010,
-Stroeven.etal.2010}, which is modelled to have hosted warm-based ice
-(Fig.~\ref{fig:warmfrac}) has only been glaciated for a short duration
-(Fig.~\ref{fig:duration}) during the last glacial cycle according to our
-results (Figs.~\ref{fig:duration} and~\ref{fig:warmbase}). We therefore infer
-that either our model does not perform reliably in this part of the domain or
+the landscape. A well-developed network of glacial valleys west of
+the MacKenzie Mountains (\citealp[Fig.~2]{Kleman.etal.2010}; \citealp[Fig.~8]
+{Stroeven.etal.2010}) is modelled to have hosted warm-based ice
+(Fig.~\ref{fig:warmfrac}) but it has only been glaciated for a short fraction
+of the last glacial cycle (Fig.~\ref{fig:duration}) according to our
+results. This perhaps indicate that\renote{
+    I don't think that we have to infer that ``our model does not perform
+    reliably'' here. I am actually quite happy to discover that the model
+    puts warm-based ice where there is (another) region of strong glacial
+    erosion. I would rather say that computing the ``duration of
+    warm-based ice cover'' as I did for is a very simple approach and that we
+    can expect it to give a qualitative proxy for glacial erosion but not a
+    quantitative one.}
 the observed landscape pattern originates from older glacial cycles and
-indicates an increased relative importance of this ice dispersal centre prior
+witnesses an increased relative importance of this ice dispersal centre prior
 to the Late Pleistocene \citep[cf.][]{Ward.etal.2008, Demuro.etal.2012}.
 
-\todo{Paragraph on model uncertainties related to warm-based ice.}
+The modelled distribution of warm-based ice cover (Figs.~\ref{fig:warmbase}
+and~\ref{fig:warmfrac}) is inevitably affected by our assumption of a constant,
+70\,\unit{mW\,m^{-2}} geothermal heat flux at 3\,km depth
+(Sect.~\ref{sec:overview}). However, both the Skeena Mountain area and the area
+west of the MacKenzie Mountains are areas of higher-than-average geothermal
+heat flux \citep{Blackwell.Richards.2004}. We can therefore expect a higher
+frequency of warm-based ice cover in this areas if we were to include
+spatially-variable geothermal forcing in our simulations.
 
@@
   \includegraphics{duration}
-  \caption{Modelled duration of ice cover during the last 120\,kyr.
+  \caption{Modelled duration of ice cover during the last 120\,ka.
            Note the irregular colour scale. A contiguous ice cover spanning
            from the Alaska Range (AR) to the southern Coast Mountains (CM) and
-           Rocky Mountains (RM) exists for about 29\,kyr in both
+           Rocky Mountains (RM) exists for about 29\,ka in both
            simulations. A central
@@
            simulation. On the other hand, the maximal extent of the ice sheet
-           generally corresponds to relatively short durations of ice cover.}
+           generally corresponds to relatively short durations of ice cover.
+           \todo{Add WSE, SMK, NR, NC mountain ranges, and changes kyr to ka}}
   \label{fig:duration}
-\end{figure*}
-
-% fig:warmfrac
-\begin{figure*}
-  \includegraphics{warmfrac}
-  \caption{Modelled fraction of warm-based ice cover during the ice-covered
-           period. Note the dominance of warm-based conditions on the
-           continental shelf and major glacial troughs of the coastal ranges.
-           Hatches indicate areas that were covered by cold ice only.
-           \todo{indicate location of the Skeena Mountains.}}
-  \label{fig:warmfrac}
 \end{figure*}
@@
 
-% fig:erosion
+% fig:warmfrac
 \begin{figure*}
-  \includegraphics{erosion}
-  \caption{Modelled cumulative basal displacement (integrand of basal velocity)
-           over the last 120\,kyr.
-           \todo{remove this figure (see footnote in text).}}
-  \label{fig:erosion}
+  \includegraphics{warmfrac}
+  \caption{Modelled fraction of warm-based ice cover during the ice-covered
+           period. Note the dominance of warm-based conditions on the
+           continental shelf and major glacial troughs of the coastal ranges.
+           Hatches indicate areas that were covered by cold ice only.
+           \todo{indicate location of the Skeena Mountains.}}
+  \label{fig:warmfrac}
 \end{figure*}

### 2014-07-17 cordillera-cycle.tex
Add MIS 2 and MIS 4 ice configuration maps.

@@
 
+% avoid "too many unprocessed floats" errors
+\clearpage
+
 % fig:locmap
@@
 
+% fig:mis2
+\begin{figure*}
+  \includegraphics{icemaps-mis2}
+  \caption{Modelled surface topography (200\,m contours) and surface velocity
+           (colour mapping) corresponding to the maximum ice volume during
+           MIS~2 in the GRIP and EPICA high-resolution simulations.}
+  \label{fig:mis2}
+\end{figure*}
+
+% fig:mis4
+\begin{figure*}
+  \includegraphics{icemaps-mis4}
+  \caption{Modelled surface topography (200\,m contours) and surface velocity
+           (colour mapping) corresponding to the maximum ice volume during
+           MIS~4 in the GRIP and EPICA high-resolution simulations.}
+  \label{fig:mis4}
+\end{figure*}
+
 % fig:duration

### 2014-07-17 figures/icemaps-mis2.py

@@ -1,0 +1,12 @@
+#!/usr/bin/env python2
+# coding: utf-8
+
+from paperglobals import darkred
+from icemaps import icemaps
+
+# plot
+fig = icemaps(mis=2)
+
+# save
+print('saving...')
+fig.savefig('icemaps-mis2')

### 2014-07-17 figures/icemaps-mis4.py

@@ -1,0 +1,12 @@
+#!/usr/bin/env python2
+# coding: utf-8
+
+from paperglobals import darkred
+from icemaps import icemaps
+
+# plot
+fig = icemaps(mis=4)
+
+# save
+print('saving...')
+fig.savefig('icemaps-mis4')

### 2014-07-17 figures/icemaps.py

@@ -1,0 +1,59 @@
+#!/usr/bin/env python2
+# coding: utf-8
+
+import numpy as np
+from netCDF4 import Dataset
+from matplotlib import pyplot as plt
+from matplotlib.colors import BoundaryNorm, LogNorm
+from mpl_toolkits.axes_grid1.axes_grid import ImageGrid
+from paperglobals import *
+
+# parameters
+res = '6km'
+records = ['grip', 'epica']
+offsets = [5.8, 5.6]
+
+def icemaps(mis):
+    # initialize figure
+    fig = iplt.gridfigure((47.5, 95.0), (1, len(records)), axes_pad=2.5*in2mm,
+                          cbar_mode='single', cbar_pad=2.5*in2mm, cbar_size=5*in2mm)
+
+    # loop on records
+    for i, rec in enumerate(records):
+        ax = fig.grid[i]
+        this_run_path = run_path % (res, rec, offsets[i]*100)
+
+        # get ice volume maximum
+        t = get_mis_times(this_run_path + '-ts.nc')[-1][1-mis]
+
+        # load extra output
+        nc = Dataset(this_run_path + '-extra.nc')
+        time = nc.variables['time'][:]*s2ka
+
+        # round maximum time to nearest slice
+        idx = (np.abs(time[:]-t)).argmin()
+
+        print 'plotting %s at %s ka...' % (rec, t)
+        ax = fig.grid[i]
+        iplt.imshow(nc, 'topg', idx, ax,
+                    cmap=topo_cmap, norm=topo_norm)
+        iplt.icemargin(nc, idx, ax,
+                       linewidths=0.5)
+        iplt.contour(nc, 'usurf', idx, ax,
+                     levels=range(200, 5000, 200),
+                     cmap=None, colors='k', linewidths=0.1)
+        iplt.contour(nc, 'usurf', idx, ax,
+                     levels=range(1000, 5000, 1000),
+                     cmap=None, colors='k', linewidths=0.25)
+        im = iplt.imshow(nc, 'velsurf_mag', idx, ax,
+                         cmap=vel_cmap, norm=vel_norm, alpha=0.75)
+        annotate(ax, '%s, %s kyr' % (rec.upper(), t))
+
+    # close extra file
+    nc.close()
+
+    # add colorbar and return figure
+    cb = fig.colorbar(im, ax.cax, extend='both', format='%i',
+                  ticks=np.logspace(1, 3.5, 6))
[…]

### 2014-07-17 cordillera-cycle.tex
Regroup GRIP and EPICA deglaciation snapshots.

@@
 
+% fig:deglacshots
+\begin{figure*}
+  \includegraphics{deglacshots}
+  \caption{Snapshots of modelled surface topography (200\,m contours)
+           and surface velocity (colour mapping) during the last deglaciation
+           from the GRIP (top panels) and EPICA (bottom panels) simulations.}
+  \label{fig:deglacshots-grip}
+\end{figure*}
+
 % fig:deglac
@@
   \label{fig:deglac}
-\end{figure*}
-
-% fig:deglacshots-grip
-\begin{figure*}
-  \includegraphics{deglacshots-grip}
-  \caption{Snapshots of modelled surface topography (200\,m contours)
-           and surface velocity (colour mapping) from the GRIP simulation,
-           corresponding to the last glacial ice volume maximum (19.1\,kyr) and
-           the last deglaciation.}
-  \label{fig:deglacshots-grip}
-\end{figure*}
-
-% fig:deglacshots-epica
-\begin{figure*}
-  \includegraphics{deglacshots-epica}
-  \caption{Snapshots of modelled surface topography (200\,m contours)
-           and surface velocity (colour mapping) from the EPICA simulation,
-           corresponding to the last glacial ice volume maximum (17.3\,kyr) and
-           the last deglaciation.}
-  \label{fig:deglacshots-epica}
 \end{figure*}

### 2014-07-17 figures/deglacshots-epica.py

@@ -1,66 +1,0 @@
-#!/usr/bin/env python2
-# coding: utf-8
-
-import numpy as np
-from netCDF4 import Dataset
-from matplotlib import pyplot as plt
-from matplotlib.colors import BoundaryNorm, LogNorm
-from mpl_toolkits.axes_grid1.axes_grid import ImageGrid
-from paperglobals import *
-
-# parameters
-res = '6km'
-rec = 'epica'
-dt = 5.6
-
-# initialize figure
-figw, figh = 120.0, 100.0
-fig = plt.figure(0, (figw*in2mm, figh*in2mm))
-rect1 = [2.5/figw, 2.5/figh, 47.5/figw, 95.0/figh]
-rect2 = [52.5/figw, 2.5/figh, 56.25/figw, 95.0/figh]
-grid = [None]*5
-grid[0] = fig.add_axes(rect1)
-grid[1:] = ImageGrid(fig, rect2, (2, 2), axes_pad=2.5*in2mm,
-                 cbar_mode='single', cbar_location='right',
-                 cbar_pad=2.5*in2mm, cbar_size=5.0*in2mm)
-remove_ticks(grid)
-
-# get LGM time
-this_run_path = run_path % (res, rec, dt*100)
-lgm_time = get_mis_times(this_run_path + '-ts.nc')[-1][-1]
-plot_times = [lgm_time, -16, -14, -12, -10]
-
-# load extra output
-nc = Dataset(this_run_path + '-extra.nc')
-time = nc.variables['time'][:]*s2ka
-
-# round pltting times to nearest slice
-plot_idces = [(np.abs(time[:]-t)).argmin() for t in plot_times]
-
-# plot
-for i, t in enumerate(plot_idces):
-    print 'plotting %s kyr snapshot...' % time[t]
-    ax = grid[i]
-    iplt.imshow(nc, 'topg', t, ax=ax,
-                cmap=topo_cmap, norm=topo_norm)
-    iplt.icemargin(nc, t, ax=grid[i],
-                   linewidths=0.5)
-    iplt.contour(nc, 'usurf', t, ax=ax,
-                 levels=range(200, 5000, 200),
-                 cmap=None, colors='k', linewidths=0.1)
-    iplt.contour(nc, 'usurf', t, ax=ax,
-                 levels=range(1000, 5000, 1000),
-                 cmap=None, colors='k', linewidths=0.25)
[…]

### 2014-07-17 figures/deglacshots-grip.py

@@ -1,66 +1,0 @@
-#!/usr/bin/env python2
-# coding: utf-8
-
-import numpy as np
-from netCDF4 import Dataset
-from matplotlib import pyplot as plt
-from matplotlib.colors import BoundaryNorm, LogNorm
-from mpl_toolkits.axes_grid1.axes_grid import ImageGrid
-from paperglobals import *
-
-# parameters
-res = '6km'
-rec = 'grip'
-dt = 5.8
-
-# initialize figure
-figw, figh = 120.0, 100.0
-fig = plt.figure(0, (figw*in2mm, figh*in2mm))
-rect1 = [2.5/figw, 2.5/figh, 47.5/figw, 95.0/figh]
-rect2 = [52.5/figw, 2.5/figh, 56.25/figw, 95.0/figh]
-grid = [None]*5
-grid[0] = fig.add_axes(rect1)
-grid[1:] = ImageGrid(fig, rect2, (2, 2), axes_pad=2.5*in2mm,
-                 cbar_mode='single', cbar_location='right',
-                 cbar_pad=2.5*in2mm, cbar_size=5.0*in2mm)
-remove_ticks(grid)
-
-# get LGM time
-this_run_path = run_path % (res, rec, dt*100)
-lgm_time = get_mis_times(this_run_path + '-ts.nc')[-1][-1]
-plot_times = [lgm_time, -16, -14, -12, -10]
-
-# load extra output
-nc = Dataset(this_run_path + '-extra.nc')
-time = nc.variables['time'][:]*s2ka
-
-# round pltting times to nearest slice
-plot_idces = [(np.abs(time[:]-t)).argmin() for t in plot_times]
-
-# plot
-for i, t in enumerate(plot_idces):
-    print 'plotting %s kyr snapshot...' % time[t]
-    ax = grid[i]
-    iplt.imshow(nc, 'topg', t, ax=ax,
-                cmap=topo_cmap, norm=topo_norm)
-    iplt.icemargin(nc, t, ax=grid[i],
-                   linewidths=0.5)
-    iplt.contour(nc, 'usurf', t, ax=ax,
-                 levels=range(200, 5000, 200),
-                 cmap=None, colors='k', linewidths=0.1)
-    iplt.contour(nc, 'usurf', t, ax=ax,
-                 levels=range(1000, 5000, 1000),
-                 cmap=None, colors='k', linewidths=0.25)
[…]

### 2014-07-17 figures/deglacshots.py

@@ -1,0 +1,61 @@
+#!/usr/bin/env python2
+# coding: utf-8
+
+import numpy as np
+from netCDF4 import Dataset
+from matplotlib import pyplot as plt
+from matplotlib.colors import BoundaryNorm, LogNorm
+from mpl_toolkits.axes_grid1.axes_grid import ImageGrid
+from paperglobals import *
+
+# parameters
+res = '6km'
+records = ['grip', 'epica']
+offsets = [5.8, 5.6]
+tkalist = [-16, -14, -12, -10]
+
+# initialize figure
+fig = iplt.gridfigure((22.5, 45.0), (len(records), 4), axes_pad=2.5*in2mm,
+                      cbar_mode='single', cbar_pad=2.5*in2mm, cbar_size=5*in2mm)
+
+# loop on records
+for i, rec in enumerate(records):
+    this_run_path = run_path % (res, rec, offsets[i]*100)
+
+    # load extra output
+    nc = Dataset(this_run_path + '-extra.nc')
+    time = nc.variables['time'][:]*s2ka
+
+    # round plotting times to nearest slices
+    idxlist = [(np.abs(time[:]-t)).argmin() for t in tkalist]
+
+    # plot
+    for j, t in enumerate(idxlist):
+        print 'plotting %s at %s ka...' % (rec, time[t])
+        ax = fig.grid[4*i+j]
+        iplt.imshow(nc, 'topg', t, ax,
+                    cmap=topo_cmap, norm=topo_norm)
+        iplt.icemargin(nc, t, ax,
+                       linewidths=0.5)
+        iplt.contour(nc, 'usurf', t, ax,
+                     levels=range(200, 5000, 200),
+                     cmap=None, colors='k', linewidths=0.1)
+        iplt.contour(nc, 'usurf', t, ax,
+                     levels=range(1000, 5000, 1000),
+                     cmap=None, colors='k', linewidths=0.25)
+        im = iplt.imshow(nc, 'velsurf_mag', t, ax,
+                         cmap=vel_cmap, norm=vel_norm, alpha=0.75)
+        annotate(ax, '%s kyr' % time[t])
+
+    # add record label
+    annotate(ax, rec.upper(), bottom=True)
+
+    # close extra file
+    nc.close()
+
+# add colorbar and save
+cb = fig.colorbar(im, ax.cax, extend='both', format='%i',
+                  ticks=np.logspace(1, 3.5, 6))
[…]

### 2014-07-17 cordillera-cycle.tex
Rework section 4.3 (missing parts).

@@
 \subsubsection{Timing of glaciation}
+\label{sec:timing}
 
@@
 \subsubsection{Ice configuration during MIS~2}
+\label{sec:mis2}
 
@@
 \subsubsection{Ice configuration during MIS~4}
+\label{sec:mis4}
 
@@
 majority of the glacial geologic record relates to the last few millennia of
-glacier cover, most of the older evidence having been overridden by ice retreat
-retreat during the deglaciation\footnote{\textbf{Johan,} can we refer to one
-of your papers here?}.
-
-\todo{add 1 sentence to connect to the next paragraph: that is easy:D}
-
-The timing of peak ice volume during the last glacial maximum and the pacing of
-deglaciation depend cirticially on the choice of atmospheric forcing
-(Figs.~\ref{fig:timeseries} and~\ref{fig:deglacseries}). Adopting the EPICA
+glaciation, most of the older evidence having been overridden by ice retreat
+during the deglaciation \citep{Kleman.1994, Kleman.etal.2010}. From a
+numerical modelling perspective, phases of glacier retreat are more challenging
+than phases of growth, because they involve more rapid fluctuations of the ice
+margin, increased flow velocities and longitudinal stress gradients, and poorly
+understood hydrological processes, typically included in the models through
+simple parametrisations \citep[e.g.][]{Clason.etal.2012, Clason.etal.2014,
+Bueler.Pelt.2014}, if included at all. However, after mapped moraines
+indicative of maximal extents (Sect.~\ref{sec:mis2} and~\ref{sec:mis4}),
+geomorphologically-reconstructed patterns of the last deglaciation provide the
+second best source of evidence for validation of our simulations.
+
+In the North American Cordillera, the presence of lateral melt-water channels
+at high elevation \citep{Margold.etal.2011, Margold.etal.2013a,
+Margold.etal.2014}, and abundant esker systems at low elevation
+\citep{Burke.etal.2012, Burke.etal.2012a, Perkins.etal.2013,
+Margold.etal.2013}\footnote{
+    Is this a good choice of references?}
+indicate a rapid deglaciation. The southern and northern margins of the
+Cordilleran ice sheet reached their last glacial maximum extent around
+17\,ka \citep[Sect.~\ref{sec:timing};][]{Porter.Swanson.1998, Cosma.etal.2008,
+Stroeven.etal.2010, Stroeven.etal.2014}, putting an older bound to the
+onset of ice retreat. However, he timing of the late deglaciation is less well
+constrained, but recent cosmogenic dates from north-central British Columbia
+indicate that a seizable ice cap emanating from the central Coast Mountains or
+the Skeena Mountains persisted into the Younger Dryas chronozone, at least
+until 12.4\,ka \citep{Margold.etal.2014}.
+
+In our simulations, the timing of peak ice volume during the last glacial
+maximum and the pacing of deglaciation depend cirticially on the choice of
+atmospheric forcing (Table~\ref{tab:extrema}, Figs.~\ref{fig:timeseries}
+and~\ref{fig:deglacseries}). Adopting the EPICA
 atmospheric forcing yields peak ice volume at 17.3\,ka and an uninterupted
@@
 curves). The simulation driven by the GRIP palaeo-temperature record yields
-peak ice volume at 19.1\,ka and a deglaciation interupted by two standstills or
-readvances until 9.4\,ka. The first interuption occurs between 16.6 and
+peak ice volume at 19.1\,ka and a deglaciation interrupted by two ice volume
+standstills until 9.4\,ka. The first interuption occurs between 16.6 and
 14.5\,ka, and the second between 12.6 and 11.6\,ka
@@
 drivers, differ in detail in that the EPICA depicts peak glaciation almost
-2\,ka later and shows a faster uninterrupted deglaciation which yields ice-free
-conditions more than 1\,ka earlier.
-
-\todo{From the timing of max glaciation and the duration of the deglaciation,
-is there any evidence to prefer one model output above the other? If so, state
-that, if not, state that too. The behavior of both models is perhaps reasonable
-within the uncertainty envelope that the field evidence provides...?}
-
-\subsubsection{Possibility for a late-glacial readvance}
+2\,ka later than the GRIP, in closer agreement to dated maximum extents, and
+shows a faster, uninterrupted deglaciation which yields ice-free conditions
+more than 1\,ka earlier.
+
+Modelled patterns of ice sheet retreat are relatively consistent between the
+two simulations (Figs.~\ref{fig:deglacshots} and.~\ref{fig:deglac}). The
+southern sector of the modelling domain, including the Puget Lowland, the Coast
+and Rocky mountains, and the Interior Plateau of British Columbia, becomes
+completely deglaciated by 12\,ka, whereas a significant ice cover remains over
+the Skeena, Selwyn, MacKenzie, Wrangell and Saint-Elias mountains in the
+northern sector of the modelling domain. After 12\,ka, deglaciation continues
+to proceed across the Liard Lowland with a radial ice margin retreat towards
+the surrounding mountain ranges, consistent with the regional melt water record
+of the last deglaciation \citep{Margold.etal.2013}. Remaining ice continues to
+decay by retreating towards the Selwyn and Skeena mountains. The last remnants
+of the Cordilleran ice sheet finally disappear from the Skeena mountains around
+10.6\,ka (EPICA) and 9.4\,ka (GRIP).
+
+\subsubsection{Late-glacial readvance}
 
@@
 been reconstructed and dated. Radiocarbon-dated end moraines in the Fraser and
-Squamish valleys off the southern tip of the Coast Mountains, indicates
-consecutive glacier maximal (or standstills while in overall retreat), one of
-which one corresponds to the Younger Dryas chronozone \citep{Clague.etal.1997,
+Squamish valleys, off the southern tip of the Coast Mountains, indicate
+consecutive glacier maxima, or standstills while in overall retreat, one of
+which corresponding to the Younger Dryas chronozone \citep{Clague.etal.1997,
 Friele.Clague.2002, Friele.Clague.2002a, Kovanen.2002,
-Kovanen.Easterbrook.2002}. In the Finlay River area of northern Rocky
-Mountains, sharp-crested moraines indicate a late-glacial readvance of alpine
-glaciers and their interaction with the main body of the decaying Cordilleran
-ice sheet \citep{Lakeman.etal.2008}. Additional evidence for late-glacial
+Kovanen.Easterbrook.2002}. Although these moraines characterise independent
+valley glaciers, not necessarily connected to an ice sheet, the Finlay River
+area of the northern Rocky Mountains presents a different kind of evidence.
+There, sharp-crested moraines indicate a late-glacial readvance of local alpine
+glaciers and, more importantly, their interaction with larger, lingering
+remnants of the main body of the Cordilleran ice sheet in the valleys
+\citep{Lakeman.etal.2008}. Additional evidence for late-glacial
 alpine glacier readvances includes moraines in the eastern Coast Mountains,
 Rocky Mountains and the Columbia Mountains \citep{Osborn.Gerloff.1997,
-Menounos.etal.2008}. Although further work is needed to constrain the timing of
-the late-glacial readvance, to assess its extents and geographical
-distribution, and to identify the potential climatic triggers
-\citep{Menounos.etal.2008}, it is interesting to note that the simulation
-driven by the GRIP record produces a late-glacial readvance in the Coast
-Mountains, Rocky Mountains and the Finlay River area, corresponding to where it
-has been identified in the geological record (Fig.~\ref{fig:deglac}, left
-panel). In contrast the EPICA-driven simulation produces a nearly-continuous
-deglaciation with only a highly restricted glacial readvance
+Menounos.etal.2008}.
+
+Although further work is needed to constrain the timing of the late-glacial
+readvance, to assess its extents and geographical distribution, and to identify
+the potential climatic triggers \citep{Menounos.etal.2008}, it is interesting
+to note that the simulation driven by the GRIP record produces a late-glacial
+readvance in the Coast Mountains, Rocky Mountains and the Finlay River area,
+corresponding to where it has been identified in the geological record
+(Fig.~\ref{fig:deglac}, left panel). In addition to matching the location of
+local readvances, the GRIP-driven simulation shows that larger ice bodies
+emanating from the Skeena and MacKenzie mountains may still have exist at the
+time of this late-glacial readvance\renote{
+    Thanks Arjen for your suggestion to reformulate. Good point.}.
+In contrast, the EPICA-driven simulation produces a nearly-continuous
+deglaciation with only a tightly restricted glacial readvance
 (Fig.~\ref{fig:deglac}, right panel).
-
-\todo{Note that most of the evidence for re-advance is for small glaciers,
-      whereas the model output indicates that there was still a big body of ice
-      at that time, as in \citet{Lakeman.etal.2008}.}
-
-Predicted patterns of ice sheet retreat are relatively consistent between the
-two simulations (Fig.~\ref{fig:deglac}). The southern sector of the modelling
-domain, including the Puget Lowland, the Coast and the Rocky mountains, and the
-Interior Plateau of British Columbia, becomes completely deglaciated by 12\,ka,
-whereas a significant ice cover remains over the Skeena, Selwyn, MacKenzie,
-Wrangell and Saint-Elias mountains in the northern sector of the modelling
-domain. After 12\,ka, deglaciation continues to proceed across the Liard
-Lowland with a radial ice margin retreat towards the surrounding mountain
-ranges, and inward flow towards the depression, consistent with the regional
-melt water record of the last deglaciation \citep{Margold.etal.2013}
-\todo{rephrase with ``unzip''?}. Remaining ice continues to decay by retreating
-towards the Selwyn and Skeena mountains. The last remnants of the Cordilleran
-ice sheet finally disappears from the Skeena mountains around 10.6\,ka (EPICA)
-and 9.4\,ka (GRIP).
 
@@
            records (Table~\ref{tab:records}) used as palaeo-climate forcing for
-           the ice sheet model \textbf{(top)}, and modelled ice volume
+           the ice sheet model (top panel), and modelled ice volume
            through the last 120\,kyr, expressed in meters of sea-level
-           equivalent \textbf{(bottom)}. Gray spans indicate Marine Isotope
+           equivalent (bottom panel). Gray spans indicate Marine Isotope
            Stages (MIS) according to a global compilation of benthic
@@
   \caption{Temperature offset time-series from the GRIP and EPICA ice core
-           records (Table~\ref{tab:records}) \textbf{(top)}, and modelled ice
+           records (Table~\ref{tab:records}) (top panel), and modelled ice
            volume during the deglaciation, expressed in meters of sea-level
-           equivalent \textbf{(bottom)}.}
+           equivalent (bottom panel).}
   \label{fig:deglacseries}
@@
            from the GRIP (top panels) and EPICA (bottom panels) simulations.}
-  \label{fig:deglacshots-grip}
+  \label{fig:deglacshots}
 \end{figure*}

### 2014-07-18 cordillera-cycle.tex
Add section 4.1.2.

@@
 
-\todo{Add MIS~2 ice topo and velocity map. Include two paragraphs from Martin.
-      Describe model limitations referring to \citet{Seguinot.etal.2014}}
+During maximum glaciation, the main meridional ice divide is positioned over
+the western flank of the Rocky Mountains (Fig.~\ref{fig:snapshots}, lower
+panels; Fig.~\ref{fig:mis2}). This result appears to contrast with
+palaeoglaciological reconstructions for central and southern British Columbia
+that inferred the ice divide to have been situated in a more westerly position,
+over the western margin of the Interior Plateau \citep{Ryder.etal.1991,
+Stumpf.etal.2000, Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}. A
+latitudinal saddle has been inferred to have connected ice dispersal centres in
+the Columbia Mountains with the main ice divide \citep{Ryder.etal.1991,
+Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}. A latitudinal saddle
+does indeed features in our modelling results, however, in an inverse
+configuration between the main ice divide over the Columbia Mountains and a
+secondary divide over the southern Coast Mountains (Fig.~\ref{fig:mis2}).
+
+On the one hand, this could reflect the fact that our model does not include
+feedback mechanisms between ice sheet topography and the regional climate.
+Firstly, during the build-up phase preceding the LGM, rapid accumulation over
+the Coast Mountains enhanced the topographic barrier formed by these mountain
+ranges, which likely resulted in a decrease of precipitation and, therefore,
+accumulation in the interior. Secondly, latent warming of the moisture-depleted
+air parcels flowing over this enhanced topography could have resulted in an
+inflow of warmer air over the eastern flank of the ice sheet, increasing melt
+along the advancing margin \citep[cf.][]{Langen.etal.2012}. Because these two
+processes, both inhibiting of ice-sheet growth, are absent from our model, the
+eastern margin of the ice sheet and the position of the main meridional ice
+divide are certainly biased towards the east in our simulations
+\citep{Seguinot.etal.2014}.
+
+On the other hand, field-based palaeoglaciological reconstructions have
+struggled to reconcile the more westerly-centred ice divide in south-central
+British Columbia with evidence in the Rocky Mountains and beyond, that the
+Cordilleran ice sheet invaded the western Interior Plains, where it merged with
+the southwestern margin of the Laurentide ice sheet and was deflected to the
+south \citep{Jackson.etal.1997, Bednarski.Smith.2007, Kleman.etal.2010,
+Margold.etal.2013, Margold.etal.2013a}. Ice geometries from our model runs do
+not have this problem, because the position and elevation of the ice divide
+results in significant ice drainage across the Rocky Mountains at the LGM
+(Fig.~\ref{fig:mis2}).
 
@@
       are more extensive during MIS~4 than during MIS~2 in the GRIP simulation}
+
+(Fig.~\ref{fig:snapshots}, upper panels; Fig.~\ref{fig:mis4})
+(Fig.~\ref{fig:mis4})
 

### 2014-07-18 cordillera-cycle.tex
Add section 4.1.3.

@@
 
-\todo{Add MIS~4 ice topo and velocity map. Discuss that some of the margins
-      are more extensive during MIS~4 than during MIS~2 in the GRIP simulation}
-
-(Fig.~\ref{fig:snapshots}, upper panels; Fig.~\ref{fig:mis4})
-(Fig.~\ref{fig:mis4})
+The modelled ice sheet configurations corresponding to ice volume maxima during
+MIS~4 are more variable (Fig.~\ref{fig:snapshots}, upper panels;
+Fig.~\ref{fig:mis4}). The GRIP simulation (Fig.~\ref{fig:mis4}, left panel)
+results in a modelled maximum ice sheet that closely resemble that obtained
+during MIS~2, with the only major differences being a slightly less developed
+northern and eastern sectors. However, the EPICA simulations produce a reduced
+ice volume maximum (Fig.~\ref{fig:timeseries}), which translates in the
+modelled ice sheet geometry into a much less developed southern sector, more
+restricted ice cover on the northern and eastern flanks, and generally lower
+ice surface elevations in the interior (Fig.~\ref{fig:mis4}, left panel).
+
+In the GRIP simulation, some sections of the MIS~4 maximum ice margin are
+slightly more extensive than their MIS~2 counterparts. This is notably the case
+in the Puget Lowland, along parts of the marine margin, and parts of the north
+slope of the Alaska Range. However, these differences generally correspond to a
+few model grid cells. Their interpretation would be hazardous considering
+the simplicity of our atmospheric forcing, and the scale of our simulations.
 

### 2014-07-18 cordillera-cycle.tex
Polish and add missing parts.

@@
   forcing, all simulations produce two major glaciation events during
-  marine isotope stages~4 (61.9--55.4\,kyr) and~2
-  (29.5--16.9\,kyr). The timing of glaciation is
+  marine isotope stages~4 (61.9--55.4\,ka) and~2
+  (29.5--16.9\,ka). The timing of glaciation is
   better reproduced using temperature reconstructions from Greenland and
@@
   and hosts the last remains of Cordilleran ice during the
-  early Holocene (10.9--9.5\,kyr).
+  early Holocene (10.9--9.5\,ka).
 
@@
     (i.e. skin depth) depends on the period of temperature oscillations at the
-    surface. The longer the period (in our case 100\,kyr), the deeper the
+    surface. The longer the period (in our case 100\,ka), the deeper the
     temperature change. A good way to estimate what this parameter should be is
@@
 the southern Cordilleran ice sheet margin has been constrained by radiocarbon
-dating on wood between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,ka\,BP}
+dating on wood between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,ka}
 \citep{Porter.Swanson.1998}.
@@
 an increase of glaciomarine sedimentation between 19.5 and
-16.2\,\unit{\chem{^{14}C}\,cal\,ka\,BP} \citep{Cosma.etal.2008}. Radiocarbon
+16.2\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008}. Radiocarbon
 dating of the northern Cordilleran ice sheet margin is much less constrained
@@
 and Vostok palaeo-temperature records yield Cordilleran ice sheet maximum
-extent that may be compatible with these field constraints
+extents that may be compatible with these field constraints
 (Fig.~\ref{fig:timeseries}, lower panel; Table~\ref{tab:extrema}).
@@
 
-\todo{Write something about NGRIP.}
-
+Similarly, atmospheric forcing of the model using the NGRIP palaseo-temperature
+record results in an early onset of deglaciation (Fig.~\ref{fig:timeseries})
+after the -22.9\,ka last glacial volume maximum (Table~\ref{tab:extrema},
+incompatible with dated evidence of the last glacial maximum advance\renote{
+    I actually have no explanation for that. Did not have the time to dig much
+    into the ice-core literature. Any idea?}.
 Hence, we focus the rest of our analysis on simulations forced by
@@
 detailed comparison against the geologic record, these two simulations were
-re-run on the high-resolution grid (Fig.~\ref{fig:timeseries}, lower panel,
+re-run using the high-resolution grid (Fig.~\ref{fig:timeseries}, lower panel,
 dotted lines).
@@
 ice volume maximum (Fig.~\ref{fig:timeseries}), which translates in the
-modelled ice sheet geometry into a much less developed southern sector, more
+modelled ice sheet geometry into a significantly reduced southern sector, more
 restricted ice cover on the northern and eastern flanks, and generally lower
@@
 slope of the Alaska Range. However, these differences generally correspond to a
-few model grid cells. Their interpretation would be hazardous considering
-the simplicity of our atmospheric forcing, and the scale of our simulations.
+few model grid cells. Considering the simplicity of the applied atmospheric
+forcing, their interpretation would be hazardous at the scale of our
+simulations.
 
@@
 It is generally believed that the Cordilleran ice sheet formed by the
-coalescence of of several mountain-centred ice caps \citep{Davis.Mathews.1944}.
+coalescence of several mountain-centred ice caps \citep{Davis.Mathews.1944}.
 In our simulations, major ice-dispersal centres, visible on the modelled ice
@@
 ice cover is the persistence of the Skeena Mountains ice cap throughout the
-entire last glacial cycle and its predominance over other ice-dispersal centres
+entire last glacial cycle and its predominance over the other ice-dispersal
+centres
 (Figs.~\ref{fig:snapshots} and~\ref{fig:duration}). Regardless of the applied
@@
 frequency of warm-based ice cover in this areas if we were to include
-spatially-variable geothermal forcing in our simulations.
+spatially-variable geothermal forcing in our Cordilleran ice sheet simulations.
 
@@
 Stroeven.etal.2010, Stroeven.etal.2014}, putting an older bound to the
-onset of ice retreat. However, he timing of the late deglaciation is less well
+onset of ice retreat. The timing of the late deglaciation is less well
 constrained, but recent cosmogenic dates from north-central British Columbia
@@
 landforms (lineations and eskers) are part of the deglacial envelop
-\citep[terminology from]{Kleman.etal.2006}, that is that these were formed
+\citep[terminology from][]{Kleman.etal.2006}, that is that these were formed
 close inside the retreating margin of ice sheets \citep{Boulton.Clark.1990,
-Kleman.etal.1997, Kleman.etal.2010}, we present a map of basal flow directions
+Kleman.etal.1997, Kleman.etal.2010}, we present maps of basal flow directions
 immediately preceding deglaciation or at the time of cessation of sliding
-inside the cold-base retreating margin (Fig.~\ref{fig:lastflow})
-\todo{description of the general trends that the map shows}. Patterns of
-glacial lineation formed in the northern and southern sectors of the
-Cordilleran ice sheet and in the Liard Lowland (\citealp{Prest.etal.1968};
-\citealp[Fig.~1.12]{Clague.1989}; \citealp[Fig.~2]{Kleman.etal.2010};
-\citealp[Fig.~2]{Margold.etal.2013}), similarities with the patterns of
-deglacial ice flow from modelling (Fig.~\ref{fig:lastflow}). However, this is
-not the case for the Interior Plateau of British Columbia, where both
+inside the cold-base retreating margin (Fig.~\ref{fig:lastflow}). The
+modelled deglacial flow patterns appear consistent between the GRIP and EPICA
+simulations. They depict an active ice sheet retreat in the peripheral areas,
+followed by stagnant ice decay in some of the interior regions. Several parts
+of the modelling domain do not experience any basal sliding throughout the
+deglaciation phase (Fig.~\ref{fig:lastflow}, hatched areas). This notably
+includes parts of the Interior Plateau in central British Columbia, major
+portions of the Alaskan sector of the ice sheet, and a tortuous ribbon running
+from the Northern Rocky Mountains over the Skeena and Selwyn Mountains and into
+the MacKenzie Mountains.
+
+Patterns of glacial lineation formed in the northern and southern sectors of
+the Cordilleran ice sheet (\citealp{Prest.etal.1968};
+\citealp[Fig.~1.12]{Clague.1989}; \citealp[Fig.~2]{Kleman.etal.2010}) show
+similarities with the patterns of deglacial ice flow from numerical modelling
+(Fig.~\ref{fig:lastflow}). In the northern half
+of the modelling domain, modelled deglacial flow direction depict an
+active downhill flow as the last remnants of the ice sheet retreat towards
+elevated mountain ranges. Converging deglacial flow patterns in the Liard
+Lowland, for instance (Fig.~\ref{fig:lastflow}), closely resemble that
+indicated by the glacial lineations \citep[Fig.~2]{Margold.etal.2013}.
+
+However, on the Interior Plateau of British Columbia, both
 simulations predict negligible basal sliding during deglaciation
-(Fig.~\ref{fig:lastflow}), but where an impressive set of glacial delineations
+(Fig.~\ref{fig:lastflow}), but an impressive set of glacial delineations
 indicate a substantial eastward flow component of the Cordilleran ice sheet
 \citep{Prest.etal.1968, Kleman.etal.2010}.
-
-The Interior Plateau lineation swarm could thus present a smoking gun for the
+The Interior Plateau lineation set could thus present a smoking gun for the
 reliability of the presented model results. One explanation for the incongruent
-results could therefore be that the modelled LGM ice sheet is too thick, or
-that the ice divide is positioned too far to the east \todo{descript the
-problem for the incongruent results of these two conditions}. Because feedback
-mechanisms between ice sheet topography and regional climate are absent in our
-model, including wind redirection, orographic precipitation effects, and latent
-warming of moisture-depleted air, we regard misfeeds in ice thickness and ice
-devide location as plausible \citep{Seguinot.etal.2014}. Another explanation
-for the incongruent results could be that the Interior Plateau lineation swarm
-predates deglaciation and that deglacitation landforms are largely absent. The
+results could therefore be that missing feedback mechanisms between ice sheet
+topography and regional climate position the modelled ice divide of the LGM ice
+sheet too far to the east (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
+\citealp{Seguinot.etal.2014}). A more westerly-located LGM ice divide would
+certainly result in a different deglacial flow patterns over the Interior
+Plateau. Another explanation for the incongruent results could be that the
+Interior Plateau lineation system predates deglaciation, and that deglacial
+landforms are largely absent \footnote{
+    \textbf{Martin, Johan}, any southwards running eskers on the plateau?}.
+
+The
 modelled deglaciation of the Interior Plateau, central British Columbia,
@@
 southwards-flowing (Fig.~\ref{fig:lastflow}) non-sliding ice lobes positioned
-in-between deglaciated (ice-free) mountain ranges
+in-between deglaciated mountain ranges
 (Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}). This result
@@
 central British Columbia, in which mountain ranges emerge from the ice before
-the plateau \citep[Fig.~7]{Fulton.1991}. If it is valid, it may suggest that
-glacial lineations on the Interior Plateau of British Columbia may be of older
-age than the LGM, and may have remained intact throughout the deglaciation.
+the plateau \citep[Fig.~7]{Fulton.1991}.
 
@@
 last remnant of the ice sheet
-during the early Holocene (10.9--9.5\,kyr).
+during the early Holocene (10.9--9.5\,ka).
 
@@
            Innuitian (IIS) and Greenland (GIS) ice sheets during the last
-           18\,\unit{\chem{^{14}C}\,kyr\,BP} (21.4\,cal\,kyr\,BP)
+           18\,\unit{\chem{^{14}C}\,ka} (21.4\,cal\,ka)
            \citep{Dyke.2004}. The rectangular box denotes the location of the
@@
            the ice sheet model (top panel), and modelled ice volume
-           through the last 120\,kyr, expressed in meters of sea-level
+           through the last 120\,ka, expressed in meters of sea-level
            equivalent (bottom panel). Gray spans indicate Marine Isotope
@@
            rectangles highlight modelled ice volume extrema corresponding to
-           MIS~4 (61.9--55.4\,kyr), MIS~3 (52.2--45.6\,kyr), and
-           MIS~2 (last glacial maximum, 29.5--16.9\,kyr). Dotted lines
+           MIS~4 (61.9--55.4\,ka), MIS~3 (52.2--45.6\,ka), and
+           MIS~2 (last glacial maximum, 29.5--16.9\,ka). Dotted lines
            correspond to the GRIP and EPICA 6\,km-resolution runs.}
@@
   \caption{Modelled duration of warm-based ice cover during the last
-           120\,kyr. Long ice cover durations combined with basal
+           120\,ka. Long ice cover durations combined with basal
            temperatures at the pressure-melting point may explain the strong
@@
            green. Hatches denote re-advance of mountain-centred ice caps and
-           and the decaying ice sheet between 14 and 10\,kyr., which is more
+           and the decaying ice sheet between 14 and 10\,ka., which is more
            pronounced in the GRIP-driven simulation.}
@@
 \end{figure*}
-
-% fig:profiles-grip
-\begin{figure}
-  \includegraphics{profiles-grip}
-  \caption{Modelled bedrock (black) and ice surface (blue) topography profiles
-           during deglaciation (22.0--8.0\,kyr) in the GRIP 6\,km
-           simulation, corresponding to the four transects indicated in
-           Fig.~\ref{fig:deglac}.}
-  \label{fig:profiles-grip}
-\end{figure}
-
-% fig:profiles-epica
-\begin{figure}
-  \includegraphics{profiles-epica}
-  \caption{Modelled bedrock (black) and ice surface (blue) topography profiles
-           during deglaciation (22.0--8.0\,kyr) in the EPICA 6\,km
-           simulation, corresponding to the four transects indicated in
-           Fig.~\ref{fig:deglac}.}
-  \label{fig:profiles-epica}
-\end{figure}
 
@@
            indicate areas that remain non-sliding throughout deglaciation
-           (22.0--8.0\,kyr), notably including the Interior Plateau of central
+           (22.0--8.0\,ka), notably including the Interior Plateau of central
            British Columbia.
@@
 
+% fig:profiles-grip
+\begin{figure}
+  \includegraphics{profiles-grip}
+  \caption{Modelled bedrock (black) and ice surface (blue) topography profiles
+           during deglaciation (22.0--8.0\,ka) in the GRIP 6\,km
+           simulation, corresponding to the four transects indicated in
+           Fig.~\ref{fig:deglac}.}
+  \label{fig:profiles-grip}
+\end{figure}
+
+% fig:profiles-epica
+\begin{figure}
+  \includegraphics{profiles-epica}
+  \caption{Modelled bedrock (black) and ice surface (blue) topography profiles
+           during deglaciation (22.0--8.0\,ka) in the EPICA 6\,km
+           simulation, corresponding to the four transects indicated in
+           Fig.~\ref{fig:deglac}.}
+  \label{fig:profiles-epica}
+\end{figure}
+
 % =============================================================================

### 2014-07-18 cordillera-cycle.tex
Prepare for future figure improvements.

@@
 that inferred the ice divide to have been situated in a more westerly position,
-over the western margin of the Interior Plateau \citep{Ryder.etal.1991,
+over the western margin of the Fraser Plateau\footnote{
+    Just checked that this name is correct. Feels lighter than the ``Interior
+    Plateau of south-central British Columbia'' (although having a slightly
+    different meaning).}
+\citep{Ryder.etal.1991,
 Stumpf.etal.2000, Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}. A
@@
 southern sector of the modelling domain, including the Puget Lowland, the Coast
-and Rocky mountains, and the Interior Plateau of British Columbia, becomes
+and Rocky mountains, and the Fraser Plateau, becomes
 completely deglaciated by 12\,ka, whereas a significant ice cover remains over
@@
 deglaciation phase (Fig.~\ref{fig:lastflow}, hatched areas). This notably
-includes parts of the Interior Plateau in central British Columbia, major
+includes parts of the Fraser Plateau, major
 portions of the Alaskan sector of the ice sheet, and a tortuous ribbon running
@@
 
-However, on the Interior Plateau of British Columbia, both
+However, on the Fraser Plateau, both
 simulations predict negligible basal sliding during deglaciation
@@
 \citep{Prest.etal.1968, Kleman.etal.2010}.
-The Interior Plateau lineation set could thus present a smoking gun for the
+The Fraser lineation set could thus present a smoking gun for the
 reliability of the presented model results. One explanation for the incongruent
@@
 Plateau. Another explanation for the incongruent results could be that the
-Interior Plateau lineation system predates deglaciation, and that deglacial
+Fraser Plateau lineation system predates deglaciation, and that deglacial
 landforms are largely absent \footnote{
@@
 
-The
-modelled deglaciation of the Interior Plateau, central British Columbia,
+The modelled deglaciation of the Fraser Plateau
 consists of a rapid northwards retreat (Fig.~\ref{fig:deglac}) of
@@
            the Skeena Mountains (SM), the Coast Mountains (CM), the Rocky
-           Mountains (RM) and the North Cascades (NC). The background
+           Mountains (RM) and the North Cascades (NC). Major depressions
+           include the Liard Lowland (LL), the Fraser Plateau (FP) and the
+           Puget Lowland (PL). The background
            map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
            Data \citep{Patterson.Kelso.2014}.
-           \todo{Mark palaeo-ice sheets and mountain ranges on the map.
-                 Mark location of the Puget Lowland.}}
+           \todo{Mark palaeo-ice sheets, mountain ranges and depressions on
+                 the map. Dash boundaries between ice sheets?}}
   \label{fig:locmap}
@@
            and temperature variability over the model domain, notably between
-           the maritime and continental regions.
-           \todo{Dash boundaries between ice sheets?}}
+           the maritime and continental regions.}
   \label{fig:atm}
@@
            generally corresponds to relatively short durations of ice cover.
-           \todo{Add WSE, SMK, NR, NC mountain ranges, and changes kyr to ka}}
+           \todo{Add WSE, SMK, NR, NC mountain ranges, and change kyr to ka}}
   \label{fig:duration}
@@
            Hatches indicate areas that were covered by cold ice only.
-           \todo{indicate location of the Skeena Mountains.}}
+           \todo{indicate location of the Skeena Mountains and the MacKenzie
+                 Mountains.}}
   \label{fig:warmfrac}
@@
            and surface velocity (colour mapping) during the last deglaciation
-           from the GRIP (top panels) and EPICA (bottom panels) simulations.}
+           from the GRIP (top panels) and EPICA (bottom panels) simulations.
+           \todo{Add the location of profiles used in
+                 Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-grip}.}}
   \label{fig:deglacshots}
@@
            indicate areas that remain non-sliding throughout deglaciation
-           (22.0--8.0\,ka), notably including the Interior Plateau of central
-           British Columbia.
+           (22.0--8.0\,ka), notably including the Fraser Plateau.
            Sliding grid cells were distinguished from non-sliding grid cells
-           using a velocity threshold of 1\,\unit{m\,yr^{-1}}.}
+           using a velocity threshold of 1\,\unit{m\,yr^{-1}}.
+           \todo{indicate the Fraser Plateau and the Liard Lowland. Add the
+                 location of profiles used in Figs.~\ref{fig:profiles-grip}
+                 and~\ref{fig:profiles-grip}.}}
   \label{fig:lastflow}
@@
            simulation, corresponding to the four transects indicated in
-           Fig.~\ref{fig:deglac}.}
+           Fig.~\ref{fig:deglac}.
+           \todo{indicate mountain ranges and depressions.}}
   \label{fig:profiles-grip}
@@
            simulation, corresponding to the four transects indicated in
-           Fig.~\ref{fig:deglac}.}
+           Fig.~\ref{fig:deglac}.
+           \todo{indicate mountain ranges and depressions.}}
   \label{fig:profiles-epica}

### 2014-07-18 cordillera-cycle.tex
Latex2rtf compatibility changes.

@@
 
+% Copernicus-like latex2rtf compatible
+\newif\iflatextortf
+\iflatextortf
+\input{copernicus_rtf}
+\else
+
 % Copernicus manuscript
@@
 %\documentclass[tcd, hvmath, online]{copernicus}
-
-% Copernicus-like latex2rtf compatible
-%\input{copernicus_rtf}
+\fi
 
@@
 
-\renote{
-    Surface mass balance is the balance of mass fluxes at the surface,
-    consisting of ice accumulation and ice ablation, which I introduce straight
-    away. I don't see what kind of ``opening sentence'' can be added here.}
 Ice surface accumulation and ablation are computed from monthly mean
@@
 air temperature, $\sigma$, and monthly precipitation, $P_m$, using
-a~temperature-index model \citep[e.g.,][]{Hock.2003}. Accumulation is equal to
+a~temperature-index model \citep[e.g.,][]{Hock.2003}\renote{
+    Surface mass balance is the balance of mass fluxes at the surface,
+    consisting of ice accumulation and ice ablation, which I introduce straight
+    away. I don't see what kind of ``opening sentence'' can be added here.}.
+Accumulation is equal to
 precipitation when air temperatures are below 0\,\unit{{\degree}C}, and
@@
 % =============================================================================
-\section[Comparison to the geologic record]
-        {Comparison to the geologic record\renote{
-    Arjen has proposed: 2.1 Evaluation of model runs agains field evidence, 2.2
-    Palaeoglaciology of the Cordilleran ice sheet, and 2.3 The last
-    deglaciation; but I don't think that these headers work. In my view, this
+\section{Comparison to the geologic record}
+\label{sec:discussion}
+% =============================================================================
+
+\emph{[\textbf{About the new titles}: In my view, this
     entire ``discussion'' part is all about evaluation of model runs agains
@@
     which deserves a separate section because many people are interested in
-    this period, and much geological evidence is available.}}
-\label{sec:discussion}
-% =============================================================================
+    this period, and much geological evidence is available.]}
 

### 2014-07-18 kappa.tex
Polish introduction.

@@
 deforms under its own weight to flow downhill (\cref{fig:glacier-mechanics}).
-Glaciers, these bodies of moving ice, move by a combination of
-meltwater-induced sliding at the base \citep[\S~532]{Saussure.1786}\renote{
-    I mean volume 1, which as a couple of beautiful paragraphs (would be worth
+\emph{Glaciers}, these bodies of moving ice, move by a combination of
+meltwater-induced sliding at the base \citep[\S532]{Saussure.1786}\renote{
+    Actually I mean volume 1, published in ``MDCCLXXXVI'' (1786, right?),
+    which as a couple of beautiful paragraphs (would be worth
     translating if I had time) about how the "heat from the Earth's interior"
@@
     paragraphs that \citet{Forbes.1846} refers to sixty years later when
-    advancing his ``viscous theory of glacier notion''. According to the cover
-    page, volume 1 was published in ``MDCCLXXXVI'', which I just translated to
-    1786 (so 1796 was wrong).},
+    advancing his ``viscous theory of glacier notion''.},
 and viscous deformation within the
@@
 understand that glaciers had once been more extensive than today
-\citep[e.g.,][p.~21]{Windham.Martel.1744}. Wtih the emergence of the glacial
+\citep[e.g.,][p.~21]{Windham.Martel.1744}. With the emergence of the glacial
 theory came the idea that, under colder temperatures
@@
 North America \citep{Agassiz.1840}\renote{
-    I am a bit reluctant to write ``\citep{Agassiz.1840} eventually posed the
+    I am a bit reluctant to write ``\citet{Agassiz.1840} eventually posed the
     glacial theory''. Although he published these ideas, do they really
@@
 discovery and exploration
-of the two present-day ice sheets on Earth: the Greenland and
+of the two present-day ice sheets on Earth, the Greenland and
 Antarctic ice sheets, provided a modern analogue for the proposed European and
@@
 the European Alps, it was then discovered from observations of the glacial
-landscape, that there had not been only one, but at least four, glacial periods
-\citep{Penck.Bruckner.1909}.
+landscape, that there had not been only one, but at least four, \emph{glacial
+periods} \citep{Penck.Bruckner.1909}.
 
@@
 history \citep[e.g.,][]{Emiliani.1955, Shackleton.Opdyke.1973,
-Dansgaard.etal.1993}, indicating neither one, nor four, but rather tens of
-glacial cycles. During the last 800\,ka years, glacial and interglacial cycles
-have succeeded each other with a 100\,ka periodicity \citep{Hays.etal.1976,
-Augustin.etal.2004}. In this thesis, I focus on the most recent of these
-periods: \emph{the last glacial cycle} (\cref{fig:paleo-timeseries})
+Dansgaard.etal.1993, Augustin.etal.2004}, indicating neither one, nor four, but
+rather tens of glacial cycles. During the last 800000 years (800\,ka), glacial
+and interglacial periods have succeeded each other with a 100\,ka periodicity
+\citep[\cref{fig:paleo-timeseries};]{Hays.etal.1976, Augustin.etal.2004}.
+However, much of the traces left by glaciers on the landscape date from the
+most recent of these cycles, the \emph{last glacial cycle}.
 
 % ice volume 26.07 + 2.93 = 29.00
-% meters sea-level 58.3 + 7.36 = 65.6
+% metres sea-level 58.3 + 7.36 = 65.6
 Together, the Greenland and Antarctic ice sheets currently host almost all
@@
 In addition, numerical ice sheet models provide quantitative insights on the
-four-dimensional distribution of stress and deformation in palaeo-ice sheets
+four-dimensional distribution of stresses and deformations in palaeo-ice sheets
 and allow for reconstructions their past thickness and volume evolution, which
-are difficult to infer from geological evidence
-\renote{
+are difficult to infer from geological evidence\renote{
     Is this the place where to cite \citep{Denton.Hughes.1981}? \todo{I need to
@@
            \textbf{(b)} medial moraines illustrating downhill ice flow on
-           Susitna Glacier, Alaska Range, by Austin Post, 1970
-           \citep{NSIDC.2009};
+           Susitna Glacier, Alaska Range, photo by Austin Post, 1970
+           \citep{NSIDC.2009}; and
            \textbf{(c)} mountain glacier calving into a small lake, revealing a
@@
            configuration as the glacier melted underneath,
-           Finse area, Norwegian Scandes, 2011;
+           Finse area, Norwegian Scandes, 2011; and
            \textbf{(c)} lateral meltwater channels, formed on a gentle slope
@@
            The grey bands indicate the approximate time span of the last
-           glacial period of 110 to 10\,ka.}
+           glacial period of 110 to 10\,ka.
+           \todo{Put nicer blue and red colours.}}
   \label{fig:paleo-timeseries}
@@
 varying isolation from the fluctuating ice sheet penetrate several kilometers
-into the
-rock, and eventually affect basal ice temperatures.
+into the rock, and eventually affect basal ice temperatures.
 
@@
 of deglaciation (\CCYC, Figs.~5--15). Modelled ice volume evolution (expressed
-as meters of sea level equivalent) and snapshots of ice extent and ice flow
+as metres of sea level equivalent) and snapshots of ice extent and ice flow
 patterns for the GRIP simulation (\cref{fig:plot-snapshots}) shows...

### 2014-07-18 kappa.tex
Implement comments from Arjen.

--- 2014-07-17 kappa.tex
@@
 \begin{description}
-  \item[\CCLI] \bibentry{Seguinot.etal.2014}.
-  \item[\PSDV] \bibentry{Seguinot.2013}.
-  \item[\PSDP] \bibentry{Seguinot.Rogozhina.2014}.
-  \item[\CCYC] \bibentry{Seguinot.etal.manuscript}.\renote{
-    Arjen, you had several comments regarding the style of references
-    throughout the kappa (name formatting, italic text). I am using the
-    Copernicus style here, but is there another recommendation to follow?
-    Maybe Peter knows?}
+  \item[\CCLI] \bibentry{Seguinot.etal.2014}
+  \item[\PSDV] \bibentry{Seguinot.2013}
+  \item[\PSDP] \bibentry{Seguinot.Rogozhina.2014}
+  \item[\CCYC] \bibentry{Seguinot.etal.inpreparation}
 \end{description}
@@
 
-\footnotesize{\begin{enumerate}
-  \item[I] A.~P.~Stroeven initiated the Cordilleran ice sheet modelling
-project; I set-up the ice sheet model (PISM) and supercomputing workflow, ran
-simulations and wrote the first draft; C.~Khroulev implemented necessary code
-changes in the model; I.~Rogozhina contributed to experiment design; A.~P.~Stroeven
-helped with interpretation of the results; and Q.~Zhang prepared some of the
-forcing data. All co-authors contributed to improve the text.
-  \item[II] I analysed the weather data, wrote the positive degree day model
-(PyPDD), ran computations and wrote the manuscript. A.~P.~Stroeven contributed
-with editorial comments at the pre-submission stage and I.~Rogozhina helped to
-assess the reviews.
+\small{\begin{enumerate}
+  \item[I] I set-up the ice sheet model and supercomputing workflow, ran the
+simulations and wrote the manuscript; C.~Khroulev implemented necessary code
+changes in PISM; I.~Rogozhina guided experiment design; A.~P.~Stroeven
+initiated the Cordilleran ice sheet modelling project; and Q.~Zhang prepared
+some of the forcing data. All co-authors contributed to improve the text.
+  \item[II] I analysed the weather data, wrote the positive degree day model,
+ran computations and wrote the manuscript. I.~Rogozhina helped to assess the
+reviews.
   \item[III] I.~Rogozhina provided the original idea; I analysed the weather
-data and developed a new parameterization of daily temperature standard
-deviation. I.~Rogozhina and I wrote the manuscript. A.~P.~Stroeven contributed
-with editorial comments at the pre-submission stage.
-  \item[IV] I integrated the new parameterization of standard deviation into
-PISM, ran the simulations and wrote the first draft; I.~Rogozhina guided
-experiment design; A.~P.~Stroeven, M.~Margold and J.~Kleman took part in the
-interpretation and comparison of model results against geological evidence. All
-co-authors contributed to improve the text.
+data; and we wrote the manuscript together.
+  \item[IV] I implemented necessary code changes in PISM, ran the simulations
+and wrote the manuscript; I.~Rogozhina guided experiment design; M.~Margold,
+J.~Kleman and A.~P.~Stroeven took part in the interpretation. All authors
+contributed to improve the text.
 \end{enumerate}}
@@
 are the degree-day factors for snow, $F_s$, and ice, $F_i$
-(\cref{fig:model-pdd,tab:surfaceparams}), and the standard deviation of
-near-surface air temperature,$\sPDD$
-(\cref{eqn:calovgreve,tab:surfaceparams}). Although the first two can only be
-deduced from surface mass-balance measurements on present-day glaciers
+(\cref{fig:model-pdd, tab:surfaceparams}), and the standard deviation of
+near-surface air temperature, $\sPDD$ (\cref{eq:calovgreve,
+tab:surfaceparams}). Although the first two can only be deduced from surface
+mass-balance measurements on present-day glaciers
 \citep[e.g.,][]{Shea.etal.2009} and assumptions that these values are also
@@
 
-\noindent\textbf{\CCYC} --- \bibentry{Seguinot.etal.manuscript}.
+\noindent\textbf{\CCYC} --- \bibentry{Seguinot.etal.inpreparation}.
 \bigskip

### 2014-07-18 figures/plot-atm.py

@@ -51,7 +51,7 @@ cs = grid[1].contourf(x[:], y[:], (temp[:].max(axis=0)-temp[:].min(axis=0)).T,
                       levels=levs, norm=BoundaryNorm(levs, 256),
                       cmap='Oranges', alpha=0.75)
 cb = fig.colorbar(cs, grid[1].cax, orientation='horizontal', ticks=levs[:-1])
-cb.set_label(u'Seasonality (°C)')
+cb.set_label(u'Temperature seasonality (°C)')
 
 # plot mean precipitation
 print 'plotting precipitation maps...'

### 2014-07-18 kappa.tex
Polish physical setting.

--- 2014-07-18 kappa.tex
@@
 During the last glacial cycle, much of North America has been covered by ice
-(\cref{fig:map-northamerica}).
-During the Last Glacial Maximum (LGM), three major ice sheets had collided to
+(\cref{fig:map-northamerica}). During the Last Glacial Maximum (LGM), peak of
+glaciation, three major ice sheets had collided to
 form a continuous blanket of ice that spanned from southern Alaska to
 Newfoundland, connecting with the Greenland ice sheet in the Canadian Arctic,
-and extending as far south as Ohio
-(\cref{fig:map-northamerica}). The Laurentide ice
-sheet, largest of the three, was centred on the Hudson Bay,
+and extending as far south as Chicago\renote{
+    Although already a few 100\,km into the ice margin, I think that Chicago
+    is a more illustrative location than Ohio or smaller cities there.}.
+The Laurentide ice sheet, largest of the three, was centred on the Hudson Bay,
 and fully covered the Canadian Prairies, Eastern Canada, and the Great Lakes
 Basin. The Innuitian ice sheet, much smaller, covered the northernmost Canadian
-archipelago. Finally, \emph{the Cordilleran ice sheet}, intermediate in size,
+archipelago. Finally, the \emph{Cordilleran ice sheet}, intermediate in size,
 capped the mountain
@@
 formed during multiple accretion phases associated with the closure of several
-back-arc oceanic trenches, over a 200 million year (Ma)-long geological history
-\citep{Sigloch.Mihalynuk.2013}.
-
-The dominant reliefs occur along two major ribbons that run
-parallel to each other along the general orientation of the North American Cordillera
-(\cref{photo-alaska-range}). The western ribbon is formed by the Pacific Coast
-orogenic belt \mref\ and includes,
-from North to South, the Alaska Range, the Wrangell and Saint-Elias Mountains,
+back-arc oceanic trenches, over a 200 million year (200\,Ma)-long geological
+history \citep{Sigloch.Mihalynuk.2013}.
+
+The \emph{dominant reliefs} occur along two major ribbons that run parallel to
+each other along the general orientation of the North American Cordillera. The
+western ribbon is formed by the Pacific Coast orogenic belt and includes, from
+North to South, the Alaska Range (\cref{fig:photo-alaska-range}), the Wrangell
+and Saint-Elias Mountains,
 the Coast Mountains and the North Cascades. The eastern ribbon consists of the
 Nevadan and Laramide belts, which are separated by the deep and strikingly
-linear Rocky Mountain / Tintina Trench \mref, depression and includes the Selwyn and
+linear Rocky Mountain / Tintina Trench depression, and includes the Selwyn and
 MacKenzie Mountains in the north, and the Rocky mountains in the South. The
@@
 
-The North American Cordillera hosts several intermontane basins, the presence
+The North American Cordillera hosts several \emph{intermontane basins}, the
+presence
 of which influenced former patterns of Cordilleran ice sheet flow. The two
@@
 (\cref{fig:photo-fraser-valley}). Due to the large latitudinal extent of the
-region, \emph{mean annual temperatures} vary significantly from north to south
+region, mean annual \emph{temperatures} vary significantly from north to south
 (\cref{fig:plot-atm}a). About half of the model domain has mean annual
-temperatures below freezing. However, \emph{temperature seasonality}\renote{
-    This is a common term in the meteorological literature, I think. At least
-    this is what Qiong told me.},
+temperatures below freezing. However, temperature \emph{seasonality}\renote{
+    ``Seasonality'' is a common term in the meteorological literature,
+    implicitly referring to temperature. At least this is what I understood
+    from Qiong.},
 here computed as the difference between the warmest and the coldest months,
 varies between around 10\,\degC\ along most of the Pacific Coast to above
-40\,\degC\ for the interior Arctic sectors \renote{
+40\,\degC\ for the interior Arctic sectors\renote{
     I think the word ``polar'' or ``Arctic'' is needed here, see
@@
 
-\emph{Annual precipitation rates} are highest on the western slopes of the
+Annual \emph{precipitation rates} are highest on the western slopes of the
 Pacific Coast ranges and decline quickly on the eastern side of the mountains
-(\cref{fig:photo-fraser-valley}, \cref{fig:plot-atm}c). In terms of ice sheet
-mass balance, this contrast is even reinforced through the difference in timing
-of the \emph{precipitation peaks} through the year (\cref{fig:plot-atm}d).
+(\cref{fig:photo-fraser-valley,fig:plot-atm}c). In terms of ice sheet
+mass balance, this contrast is even reinforced through the differences in
+\emph{timing of maximum precipitation} through the year (\cref{fig:plot-atm}d).
 Coastal regions receive most precipitation during winter and fall, which is
@@
 
-Evidence from glaciofluvial deposited dated using cosmogenic nuclides at $2.64^{+0.20}/_{-0.18}$\,Ma indicates the first presence of an ice sheet in the Cordilleran mountains in the early Pleistocene
-\citep{Hidy.etal.2013}.
-
-This indicates that the North American Cordillera has
-been glaciated many times since the late Pliocene, with the first of these glaciations occurring even before the mid-Pleistocene
-onset of 100 ka glacial cycles (\cref{fig:paleo-timeseries}). However, only fragments of
-the last few glaciations can be reconstructed from the geologic record, implying that
-much of the older evidence has been erased by subsequent glacial
-re-advances and perhaps even by other, non-glacial geomorphological processes.
-
-In the \emph{northern sector} of the Cordilleran ice sheet domain, at least four glaciations have
-been identified
+Evidence from glaciofluvial deposits dated using cosmogenic nuclides at
+$2.64^{+0.20}/_{-0.18}$\,Ma indicates the first presence of an ice sheet in the
+Cordilleran mountains during the early Pleistocene \citep{Hidy.etal.2013}.
+
+This indicates that the North American Cordillera has been glaciated many times
+since the late Pliocene, with the first of these glaciations occurring even
+before the mid-Pleistocene onset of 100 ka glacial cycles
+(\cref{fig:paleo-timeseries}). However, only fragments of the last few
+glaciations can be reconstructed from the geologic record, implying that much
+of the older evidence has been erased by subsequent glacial re-advances and
+perhaps even by other, non-glacial geomorphological processes.
+
+In the \emph{northern sector} of the Cordilleran ice sheet domain, at least
+four glaciations have been identified
     \citep{Duk-Rodkin.1999, Ward.etal.2007, Ward.etal.2008,
@@
            Stroeven.etal.2010, Stroeven.etal.2014}.
-The two older ones, often collectively referred to as the Reid  and pre-Reid glaciations,
-pre-date the last glacial cycle, and are therefore beyond the scope of this study.
-The two younger glaciations, however, occurred both during the last glacial
-cycle. First, the older of the two, the Gladstone Glaciation, reached its maximum extent between
-$50.4\pm3.0$ and $54.3\pm2.0$\,\unit{^{10}Be\,ka} \citep{Ward.etal.2007},
-during Marine Oxygen Isotope Stage (MIS) 4, a period of relatively large global ice volume. However, evidence from the Gladstone Glaciation has not been
-documented along the southern ice sheet margin, making it difficult to
-seize its spacial extent. The younger of the two last glacial cycle glaciations, the McConnell Glaciation, is associated with end
-moraines of oldest minimum apparent exposure ages of
-$15.7\pm1.5$ and $17.7\pm1.6$\,\unit{^{10}Be\,ka} \citep{Stroeven.etal.2014},
-which appears to correspond to the last global ice volume maximum (MIS~2). In contrast to
-the Gladstone Glaciation, the McConnell Glaciation appears contemporaneous with glacial advance of the southern ice sheet margin during the Fraser
-glaciation, with a maximum extent attained between
-$17.4$ and $16.4$\,\unit{^{14}C\,cal\,ka} \citep{Porter.Swanson.1998}.
-Within errors inherent to the cosmogenic nuclide exposure dating methods \citep{Heyman.etal.2011}, the
-McConnell and Fraser glaciations correspond to the same glacial episode, which
-resulted in the Last Glacial Maximum (LGM) extent of the Cordilleran ice sheet.
-Apart from geological evidence of the final demise of the ice sheet, most field evidence relate to the LGM maximum extent.
-
-Ice has covered most of the
-Canadian Cordillera, and some of the adjacent regions in north-western contiguous
-USA and Alaska (\cref{fig:map-cordillera}). The ice sheet originated from the
-coalescence of several mountain-centred ice caps \citep{Davis.Mathews.1944}. It is generally thought that the LGM configuration of the Cordilleran ice sheet was short-lived
-    \citep{Clague.etal.1980, Clague.1985, Cosma.etal.2008}, a hypothesis which can be studied through numerical modelling.
+The two older ones, often collectively referred to as the Reid and pre-Reid
+glaciations, pre-date the last glacial cycle, and are therefore beyond the
+scope of this study. The two younger glaciations, however, occurred both during
+the last glacial cycle. First, the older of the two, the Gladstone Glaciation,
+reached its maximum extent between $50.4\pm3.0$ and $54.3\pm2.0\,^{10}$Be\,ka
+\citep{Ward.etal.2007}, during Marine Oxygen Isotope Stage (MIS) 4, a period of
+relatively large global ice volume. However, evidence from the Gladstone
+Glaciation has not been documented along the southern ice sheet margin, making
+it difficult to seize its spacial extent. The younger of the two last glacial
+cycle glaciations, the McConnell Glaciation, is associated with end moraines of
+oldest minimum apparent exposure ages of $15.7\pm1.5$ and
+$17.7\pm1.6\,^{10}$Be\,ka \citep{Stroeven.etal.2014}, which appears to
+correspond to the last global ice volume maximum (MIS~2). In contrast to the
+Gladstone Glaciation, the McConnell Glaciation appears contemporaneous with
+glacial advance of the southern ice sheet margin during the Fraser glaciation,
+with a maximum extent attained between $17.4$ and
+$16.4$\,\unit{^{14}C\,cal\,ka} \citep{Porter.Swanson.1998}. Within errors
+inherent to the cosmogenic nuclide exposure dating methods
+\citep{Heyman.etal.2011}, the McConnell and Fraser glaciations correspond to
+the same glacial episode, which resulted in the LGM extent of the Cordilleran
+ice sheet.
+
+Apart from geological evidence of the final demise of the ice sheet, most field
+evidence relate to the LGM extent. It indicates that ice covered most of the
+Canadian Cordillera, and some of the adjacent regions in north-western
+contiguous USA and Alaska (\cref{fig:map-cordillera}). The Cordilleran ice
+sheet originated from the coalescence of several mountain-centred ice caps
+\citep{Davis.Mathews.1944}. It is generally thought that this LGM configuration
+was short-lived \citep{Clague.etal.1980, Clague.1985, Cosma.etal.2008}, a
+hypothesis which can be studied through numerical modelling.
 
@@
 is in several respects better understood than its northern counterpart
-\citep{Booth.etal.2003}. It consists of coalescent piedmont
-lobes, each of which formed at the outlet of the main valleys. The Puget Lobe
-\citep{Thorson.1980, Porter.Swanson.1998} was the westernmost and
-largest Lobes. It covered the Puget Lowland, a region that comprises
-the cities of Vancouver and Seattle. Further east, beyond the Cascade Range, the
-Okanagan and Purcell lobes formed the southern Cordilleran ice sheet margin. The latter once dammed the Clark Fork River, inhibiting runoff and forming
-Glacial Lake Missoula \citep{Pardee.1910}. As the ice sheet margin retreated, the lake
-repeatedly and catastrophically drained to the Pacific Ocean
-\citep{Bretz.1923,Waitt.1980}. These floods
-contributed to shape a distinctive landscape, known as the Channeled Scabland.
-
-The \emph{western margin} of the Cordilleran ice sheet covered part of the Pacific continental
-shelf. An independent
-ice cap covering Vancouver Island diverted flow from the main body of ice to drain into the Strait of Georgia between Vancouver Island and the Olympic Peninsula to form the Juan the Fuca marine ice lobe, calving into the Pacific Ocean \mref.
-Further to the North, the ice margin encroached the Queen Charlotte Islands. However, the
-archipelago may not have been fully ice-covered, even during maximum of glaciation. Submarine troughs traversing several portions of the continental shelf indicate the erosional action of
-additional ice streams, yet it is not clear whether they have formed during the
-last glacial cycle or during earlier glaciations \mref.
-
-Finally, the conditions at the \emph{eastern margin} of the Cordilleran ice sheet are the least understood.
-Although it
-is known that the Cordilleran ice sheet collided with the Laurentide ice sheet at times during the last glacial cycle,
-the nature of the interaction between the two ice sheets and the precise
-location of this junction remains unclear \citep{Gowan.2013}, while limited information exists on its
-timing
-    \citep[e.g.][]{Jackson.etal.1997, Bednarski.Smith.2007}.
-Because the
-present thesis focuses specifically on the Cordilleran ice sheet, I, therefore, neglect buttressing effect of the
-Laurentide ice sheet.
-
-The retreat of the ice sheet during the last deglaciation was likely rapid. Most of the evidence for the style and pace of deglaciation comes
-from the Fraser Plateau. Studies here have indicated that some of the highest relief
-emerged from the ice prior to the surrounding lowlands, possibly resulting in thin,
-stagnant ice remnants covering parts of the plateau
-    \citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a}.
-Less is known about the deglaciation of the northern sector of the Cordilleran ice sheet, but abundant glacial landforms from the Liard Lowland indicate a
-different mode of \emph{deglaciation}, with active retreat of the ice margin
-towards the surrounding mountain ranges \citep{Margold.etal.2013}. Patterns of
+\citep{Booth.etal.2003}. It consists of coalescent piedmont lobes, each of
+which formed at the outlet of the main valleys. The Puget Lobe
+\citep{Thorson.1980, Porter.Swanson.1998} was the westernmost and largest one.
+It covered the Puget Lowland, a region that comprises the cities of Vancouver
+and Seattle. Further east, beyond the Cascade Range, the Okanagan and Purcell
+lobes formed the southern Cordilleran ice sheet margin. The latter once dammed
+the Clark Fork River, inhibiting runoff and forming Glacial Lake Missoula
+\citep{Pardee.1910}. As the ice sheet margin retreated, the lake repeatedly and
+catastrophically drained to the Pacific Ocean \citep{Bretz.1923,Waitt.1980}.
+These floods contributed to shape a distinctive landscape, known as the
+Channeled Scabland.
+
+The \emph{western margin} of the Cordilleran ice sheet covered part of the
+Pacific continental shelf. An independent ice cap covering Vancouver Island
+diverted flow from the main body of ice to drain into the Strait of Georgia
+between Vancouver Island and the Olympic Peninsula, where it formed the Juan
+the Fuca marine ice lobe, calving into the Pacific Ocean \mref. Further to the
+North, the ice margin encroached the Queen Charlotte Islands. However, the
+archipelago may not have been fully ice-covered, even during the maximum of
+glaciation. Submarine troughs traversing several portions of the continental
+shelf indicate the erosional action of additional ice streams, yet it is not
+clear whether they have formed during the last glacial cycle or during earlier
+glaciations \mref.
+
+\todo{Find missing refs}
+
+Finally, the conditions at the \emph{eastern margin} of the Cordilleran ice
+sheet are the least understood. Although it is known that the Cordilleran ice
+sheet collided with the Laurentide ice sheet at times during the last glacial
+cycle \citep[e.g.][]{Margold.etal.2013, Margold.etal.2013a}, the nature of the
+interaction between the two ice sheets and the precise location of this
+junction remains unclear \citep{Gowan.2013}, while limited information exists
+on its timing \citep[e.g.][]{Jackson.etal.1997, Bednarski.Smith.2007}. Because
+the present thesis focuses specifically on the Cordilleran ice sheet, I,
+therefore, neglect buttressing effects of the Laurentide ice sheet.
+
+The retreat of the ice sheet during the last \emph{deglaciation} was likely
+rapid. Most of the evidence for the style and pace of deglaciation comes from
+the Fraser Plateau. Studies there have indicated that some of the highest
+reliefs emerged from the ice sheet prior to the surrounding lowlands, possibly
+resulting in thin, stagnant ice remnants covering parts of the plateau
+\citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a}. Less
+is known about the deglaciation of the northern sector of the Cordilleran ice
+sheet, but abundant glacial landforms from the Liard Lowland indicate a
+different mode of deglaciation, with active retreat of the ice margin towards
+the surrounding mountain ranges \citep{Margold.etal.2013}. Patterns of
 deglaciation, as well as the possibility for a late-glacial re-advance, are
-discussed in further detail and compared with modelling results in \CCYC.
-
-\todo{To make this more complete I suggest two additions: 1.
-a few sentences regarding the timing of deglaciation (to complement your introduction where you have timing of glaciation) - look in my 2010 and Martin's papers for information.
-2.
-A few more references older than Margold et al.
-2013.
-You might check what Ward has written fordeglaciation in the Yukon for example?
-Didn't Dyke write about the deglaciation of the Liard Lowland?}
+discussed in further detail and compared with modelling results in
+\CCYC\renote{
+    I have not had time to dig into new literature. The Ward paper on
+    deglacial valley fills in the Pelly River is like Chinese to me, and I
+    could not access the Dyke paper on the Liard Lowland, assuming this is the
+    one cited by Martin in his paper.}.
 

### 2014-07-18 kappa.tex
Polish model description.

@@
 
-The simulations presented in this thesis employ the Parallel Ice Sheet Model (PISM),
-an open-source, finite-difference, shallow ice sheet model
+The simulations presented in this thesis employ the \emph{Parallel Ice Sheet
+Model} (PISM), an open-source, finite-difference, shallow ice sheet model
 \citep{PISM-authors.2014}. The model requires bedrock topography, sea level,
 geothermal heat flux and climate forcing as inputs, and computes the thermal
-and dynamic state of the ice sheet, and the associated thermal and
-deformational responses of an idealised lithosphere. However, the atmosphere and ocean
-are not part of the model. Instead, PISM includes boundary models
-that provide simple parametrisations of their interaction with the ice sheet across interfaces that are assumed to be well defined (\cref{fig:model-interfaces}).
+and dynamic states of the ice sheet, and the associated thermal and
+deformational responses of an idealised lithosphere. However, the atmosphere
+and ocean are not part of the model. Instead, PISM includes boundary models
+that provide simple parametrisations of their interaction with the ice sheet
+across interfaces that are assumed to be well defined
+(\cref{fig:model-interfaces}).
 
@@
 which can be found in an on-line documentation \citep{PISM-authors.2014}.
-Instead, it aims to highlight modelling choices made in the context of this
-thesis, and presents an overview of the physics involved in these choices.
+Instead, it aims to highlight the modelling choices that I have made in this
+thesis, and to present an overview of the physics involved in these choices.
 
@@
 fluid whose deformation is entirely controlled by internal friction, here
-expressed by the Cauchy stress tensor, $\CST$, and gravity, $\vect{g}$:
+expressed by the Cauchy stress tensor, $\CST$, and the gravity vector,
+$\vect{g}$:
 \begin{equation}
@@
 water content, $\omega$, through an Arrhenius-type law
-\citep[Eqs.~63--65]{Aschwanden.etal.2012},
+\citep[Eqs.~63--65]{Paterson.Budd.1982, Aschwanden.etal.2012},
 \begin{equation}
@@
 where $T_{pa}$ is the pressure-adjusted temperature calculated using the
-Clapeyron relation, ${T_{pa} = T - \beta p}$. The water fraction, $\omega$, is
-capped at a threshold of 0.01, above which no measurements are available
-\citep{Lliboutry.Duval.1985}.
-
-\todo{I am trusting PISM's manual on that. Check other papers by Duval, or
-      the book by Greve (library). Explain params (Arjen + Irina) or refer to
-      table, maybe.}
-
-Letting ${\doteps_e}^2 = \frac{1}{2} \tr(\SRT^2)$ and $B=A^{-1/n}$, the
-constitutive law (\ref{eqn:glenslaw}) can be rewritten using a sometimes more
-convenient, inverted formulation,
+Clapeyron relation, ${T_{pa} = T - \beta p}$. $R$ is the ideal gas constant,
+and $A_c$, $A_w$, $Q_c$ and $Q_w$, are constant parameter corresponding to
+values measured above and below a critical temperature threshold
+${T_c=-10}$\,\degC\ \citep[\cref{tab:iceflowparams};][]{Paterson.Budd.1982}.
+The water fraction, $\omega$, is capped at a maximum value of 0.01, above which
+no measurements are available \citep[Eq.~5.7]{Lliboutry.Duval.1985,
+Greve.1997}.
+
+For a convenient derivation of the shallow shelf approximation, developed in
+the next section (\cref{sec:siassa}), the constitutive law can be rewritten
+using an inverted formulation. Assuming
+${{\doteps_e}^2 = \frac{1}{2} \tr(\SRT^2)}$ and ${B=A^{-1/n}}$,
+\cref{eqn:glenslaw} becomes
 \begin{equation}
@@
 \end{equation}
+
 By analogy to Newtonian flow, an apparent viscosity, $\nu$, can then be defined
-by $\DST = 2 \nu\SRT$, thus yielding
-
+by ${\DST = 2 \nu\SRT}$, thus yielding
 \begin{equation}
@@
 account for thermodynamic effects associated with internal phase changes in the
-glacier \citep[Eqs.~20--21]{Aschwanden.etal.2012}. This enthalpy ($H$) formulation reads
+glacier \citep[Eqs.~20--21]{Aschwanden.etal.2012}. This enthalpy ($H$)
+formulation reads
 \begin{equation}
@@
 where $\vect{q}$ represents the heat flow and
-${\frac{\nu\dot{\epsilon_e}^2}{4} = \tr(\DST\SRT)}$ is a
+${\nu\dot{\epsilon_e}^2/4 = \tr(\DST\SRT)}$ is a
 source term corresponding to strain heat release. In the case where ice
-temperature, $T$, is below the pressure-melting point, $T_c$, the enthalpy can be
-expressed as $H=c\,T$, where, $c$, denotes the specific heat capacity of the ice, while the heat flow is given by Fourier's law,
-$\vect{q} = k\,\grad{T}$, where $k$ is its thermal conductivity of ice. Consequently, the enthalpy equation
-(\ref{eqn:enthalpy}) can be rewritten using a more familiar, cold-ice,
-temperature formulation,
+temperature, $T$, is below the pressure-melting point, the heat flow is
+modelled by Fourier's law, $\vect{q} = k\,\grad{T}$, where $k$ is the thermal
+conductivity of ice, while enthalpy can be expressed as $H=c\,T$, where $c$
+denotes the ice specific heat capacity (\cref{tab:iceflowparams}).
+Consequently, the enthalpy equation (\ref{eqn:enthalpy}) can be rewritten using
+a more familiar, cold-ice, temperature formulation,
 \begin{equation}
@@
 \end{equation}
-Parameter values for the ice flow model used in this thesis are
-given in \cref{tab:iceflowparams}.
-
-\todo{Explain $H=c\,T$. Enthalpy here is equivalent to internal energy.}
+However, note that, here, the term ``enthalpy'' is actually an abuse of
+language, commonly used in glaciology as a synonym of ``interval energy''
+\citep{Aschwanden.etal.2012}.
 
@@
 demand too high for multi-millennial, continental-scale applications, such as
-the numerical modelling of the Cordilleran ice sheet through the 120\,ka last glacial
-cycle, thus requiring shallow approximations.
+the numerical modelling of the Cordilleran ice sheet through the last 100\,ka
+glacial cycle, thus requiring shallow approximations.
 
@@
 \subsection{Shallow approximations}
+\label{sec:siassa}
 % ----------------------------------------------------------------------
@@
 PISM employs two well-documented
-approximations of ice flow equations: the shallow ice approximation (SIA)
+approximations of the ice flow equations: the shallow ice approximation (SIA)
 and the the shallow shelf approximation (SSA). Although both approximations can
@@
 
-Neglecting atmospheric pressure, \Cref{eqn:siaz} yields \todo{detail}
+Neglecting atmospheric pressure, \Cref{eqn:siaz} yields the pressure,
 \begin{equation}
@@
 \end{equation}
-which can be re-introduced into \cref{eqn:siax,eqn:siay}, thus yielding the
+varying from zero at the ice surface, $s$, to $\rho gh$ at the base, ${b=s-h}$
+(\cref{fig:model-variables}). This expression
+can be re-introduced into \cref{eqn:siax,eqn:siay}, thus yielding the
 expression of the horizontal shear stresses, sometimes referred to as driving
@@
 
-\todo{Add a few words to emphasize that $\vsia$ can be directly obtained
-      after one vertical integration. Show that vertical velocities can be
+From \cref{eqn:vsia} it can be seen, that the shallow ice velocities, $\vsia$,
+can be directly computed from the geometry of the ice sheet surface, $s$,
+after a single vertical integration over the ice thickness, in order to
+account for variations of ice rheology, $A$, with depth.
+
+\todo{If I have time, show that vertical velocities can be
       obtained from the compressibility equation.}
 
-The \emph{shallow shelf approximation} (SSA), in contrast to SIA,
-assumes that
-basal friction is low enough for ice to deform uniformly within the entire ice column. In this context, the stress-balance can
+The \emph{shallow shelf approximation} (SSA), in contrast to SIA, assumes that
+basal friction is low enough for ice to deform uniformly within the entire ice
+column. In this context, the stress-balance can
 be simplified to \citep[Eqs.~4.10--4.12]{Weis.etal.1999}
@@
 Once again neglecting atmospheric pressure, and remembering that $\tr(\DST)=0$,
-\Cref{eqn:ssaz} yields
+\Cref{eqn:ssaz} yields the pressure
 \begin{align}
@@
 
-\todo{Add a few words to emphasize that $\vssa$ can not be obtained in a
-      direct manner, requiring to solve a linear system.}
+In contrary to the the SIA velocities, $\vsia$, the SSA velocities, $\vssa$,
+can not be computed in a direct manner. Instead, \cref{eqn:vssa} must be
+resolved numerically using iteration methods, in order to obtain
+approximated values for $\vssa$.
 
 In PISM, the SIA and SSA velocities are eventually added
-\citep[Eq.~15]{Winkelmann.etal.2011},
+\citep[\cref{fig:model-siassa};][Eq.~15]{Winkelmann.etal.2011}:
 \begin{equation}
@@
 a sliding law as basal boundary condition, which in fact constitutes the main
-control on SSA velocities and, in turn, the locations of ice streams.
-
-The simulations presented in this thesis use a \emph{pseudo-plastic sliding
-law},
+control on SSA velocities and, in turn, the locations of ice streams in the
+model.
+
+The simulations presented here use a \emph{pseudo-plastic} sliding law,
 \begin{equation}
@@
 where $\vect{\tau}_b$ is the basal traction force, $\vv_b$ is the basal
-velocity, and $v_{th}$ is an arbitrary velocity threshold. In the case of
+velocity, and $v_{th}$ is an arbitrary velocity threshold
+(\cref{tab:slidingparams}). Note that, in the case of
 $q=0$, \cref{eqn:pseudoplastic} corresponds to a purely plastic law
@@
 
-\todo{\citep[supplement]{Aschwanden.etal.2013} refer to a paper by Schoof and
-      Hindmarsh (2010, doi:10.1093/qjmam/hbp025). Check that.}
-
-However a non-zero exponent $q$ is chosen here in order to improve convergence.
 The yield stress $\tau_c$ is related to till properties by the Mohr-Coulomb
@@
 \begin{equation}
+    \label{eqn:coulomb}
     \tau_c = c_0 + N\,\tan{\phi} \,.
@@
 
-The till cohesion $c_0$ is assumed to be zero \renote{
-  Yes, we should probably defend this choice, but I am not sure how (except
-  that it just diminishes the number of parameters).}.
+The till cohesion $c_0$ is assumed to be zero\renote{
+  Yes, we should probably defend this choice, but I am not aware of any
+  observations justifying this.}.
 The effective pressure on the
@@
 \end{equation}
-where $W$ is the amount of water at the bed, and varies from zero to $W_{max}$,
-a threshold above which further melt simply exits the system.
+where $W$ is the amount of water at the bed, and varies from zero to
+$W_{max}=2$\,m, a threshold above which further melt simply exits the system,
+thus assuming instantaneous drainage. Although mass-conserving water drainage
+is now implemented in PISM \citep{Bueler.Pelt.2014}, is not included in the
+simulations presented here for the sake of simplicity.
 
@@
 \begin{equation}
-    N = \delta \rho gh \, 10^{(e_0/C_c) (1 - (W/W_{max}))} \,.
-\end{equation}
-
-\todo{explain parameters}
-
-Drainage of water at the base of the ice sheet, although implemented in PISM,
-is not included in simulations presented here for the sake of simplicity.
-
-Finally, the till friction angle $\phi$ is taken a~function of modern bed
+    N = \delta \rho gh \, 10^{(e_0/C_c) (1 - (W/W_{max}))} \,,
+\end{equation}
+where $e_0$ is a measured reference void ratio, $C_c$ a measured
+compressibility coefficient, and $\delta$ is a non-dimensional parameter
+chosen as 0.02 based on simple sensitivity tests on the Greenland ice sheet
+(Table~\ref{tab:slidingparams}).
+
+Finally, the till friction angle, $\phi$ (\cref{eqn:coulomb}), is taken as a
+function of modern bed
 elevation, with lowest values occurring at low elevations, thus accounting for
@@
 where $b_0=0$\,m is modern sea-level and $b_1=200$\,m is a rough average of
-highest shoreline elevations in the model domain. Parameter values for the
-sliding model are given in \cref{tab:slidingparams}.
+highest observed palaeo-shoreline elevations in the model domain
+\citep[\cref{tab:slidingparams};][]{Clague.James.2002}.
 
@@
 where $k_b$, $\rho_b$ and $c_b$ are the thermal conductivity, density and
-specific heat capacity of the bedrock. Bedrock temperature is conditioned at
-depth by a fixed upwards geothermal heat flux $q_G$. The bedrock thermal model
+specific heat capacity of the bedrock (\cref{tab:bedrockparams}).
+Bedrock temperature is conditioned at depth by a fixed upwards geothermal heat
+flux $q_G$  (\cref{tab:bedrockparams}). The bedrock thermal model
 is necessary for modelling the Cordilleran ice sheet on multi-millennial
 time-scales, because temperature fluctuations caused by climate change and
-varying isolation from the fluctuating ice sheet penetrate several kilometers
+varying isolation from the fluctuating ice sheet penetrate several kilometres
 into the rock, and eventually affect basal ice temperatures.
@@
 where $\nu_m$ is mantle viscosity, $\rho_l$ is lithosphere density, $D$ is the
-lithosphere's flexural rigidity, and $\sigma_{zzb}$ corresponds to the ice load
+lithosphere's flexural rigidity (\cref{tab:bedrockparams}), and $\sigma_{zzb}$
+corresponds to the ice load
 \citep{Bueler.etal.2007}. Here, ${\Delta^2 = (\div\grad{})^2}$ designs the
@@
 flexure. Due to high mantle viscosity, $\nu_m$, there exists a time lag between
-the ice sheet growth and the isostatic bedrock response. Parameter values for
-the bedrock response model are given in \cref{tab:bedrockparams}.
+the ice sheet growth and the isostatic bedrock response.
 
@@
             & 3.0
-            & \unit{J\,m^{-1}\,K^{-1}\,s^{-1}}
+            & \unit{W\,m^{-1}\,K^{-1}}
             & \citet{Ritz.1997} \\
+
+    $q_G$   & Geothermal heat flux
+            & 70.0
+            & \unit{mW\,m^{-2}}
+            &  \\
 
@@
 \citep[e.g.,][]{Hock.2003}. Surface accumulation, $a_s$, equals precipitation
-when temperature is below ${T_0=0\,\degC}$, and decreases to zero linearly
-with temperature between $T_s$ and ${T_r=2\,\degC}$,
+when temperature is below ${T_0=0\,\degC}$, and decreases to zero\renote{
+    Well spotted, thanks!}
+linearly with temperature between $T_0$ and ${T_1=2\,\degC}$,
 \begin{equation}
@@
     \begin{cases}
-        P_m     & \text{if}\ T_m \le T_s \,, \\
-        P_m \frac{T_r-T_m}{T_r-T_s}
-                & \text{if}\ T_s < T_m < T_r \,, \\
-        0       & \text{if}\ T_r \le T_m \,,
+        P_m     & \text{if}\ T_m \le T_0 \,, \\
+        P_m \frac{T_1-T_m}{T_1-T_0}
+                & \text{if}\ T_0 < T_m < T_1 \,, \\
+        0       & \text{if}\ T_1 \le T_m \,,
     \end{cases}
@@
 melt through a proportionality factor $F_s$. If all snow is melted, remaining
-PDDs are used to melt ice, using a proportionality factor $F_i$
-(\cref{fig:model-pdd}). The degree-day factors for snow, $F_s$, and ice, $F_i$,
-(\cref{tab:surfaceparams})
+PDDs are then used to melt ice, using a proportionality factor $F_i$
+(\cref{fig:model-pdd}). The degree-day factors for snow, $F_s$, and ice, $F_i$
+(\cref{tab:surfaceparams}),
 are derived from mass-balance measurements on contemporary glaciers in the
-Coast and Rocky Mountains of British Columbia \citep{Shea.etal.2009}. In
+Coast and Rocky mountains of British Columbia \citep{Shea.etal.2009}. In
 {\CCLI}, the temperature standard deviation $\sPDD$ is a constant model
@@
 a method initially developed in {\PSDV}, and further improved upon in {\PSDP}.
-
-\todo{explain why it is possible to linearly relate melt and PDDs.}
+Although the PDD model is a crude parametrisation of processes happening at the
+surface of a glacier, its applicability has been proven \citep{Hock.2003},
+while other, physically-based methods \citep[e.g.][]{Hock.2005}, require a
+precise knowledge of the physical state of the atmosphere, typically very
+uncertain for past periods such as the last glacial cycle.
 
@@
 
-    $T_s$   & Temperature of snow precipitation
-            & 273.15
-            & \unit{K}
+    $T_0$   & Temperature of snow precipitation
+            & 0.0
+            & \degC
             & \mref \\
 
-    $T_r$   & Temperature of rain precipitation
-            & 275.15
-            & \unit{K}
+    $T_1$   & Temperature of rain precipitation
+            & 2.0
+            & \degC
             & \mref \\
@@
 Atmospheric forcing of the model consists of a present-day monthly climatology,
-$\{T_{m0}, P_{m0}\}$, modified by a lapse-rate correction, ${\Delta}T_{LR}$,
-and a temperature-offset correction, ${\Delta}T_{TS}$,
+$\{T_{m0}, P_{m0}\}$, modified by temperature-offset corrections,
+${\Delta}T_{TS}$, and lapse-rate corrections, ${\Delta}T_{LR}$,
 \begin{subequations}
 \begin{align}
-    T_m(t, x, y) &= T_{m0}(x, y) + {\Delta}T_{LR}(t)
-                                 + {\Delta}T_{TS}(t, x, y) \,, \\
+    T_m(t, x, y) &= T_{m0}(x, y) + {\Delta}T_{TS}(t)
+                                 + {\Delta}T_{LR}(t, x, y) \,, \\
     P_m(t, x, y) &= P_{m0}(x, y) \,.
@@
 near-surface air temperature and precipitation rate fields extracted from
-observational data and atmospheric reanalyses.
-The lapse-rate correction, ${\Delta}T_{LR}$, is computed in relation to the
+observational data and atmospheric reanalyses (\cref{tab:datasources}).
+The lapse-rate correction, ${\Delta}T_{LR}$, is computed from the ice sheet
+surface elevation, $s$, in relation to the
 climate input reference bedrock topography, $b_{ref}$,
@@
 \end{align}
-thus accounting for the evolution of ice thickness, $h$, on the one hand, and
+thus accounting for the evolution of ice thickness, ${h=s-b}$, on the one hand,
+and
 for differences between the the ice flow model basal topography, $b$, and the
@@
 annual temperature lapse rate of $\gamma = 6\,\unit{K\,km^{-1}}$
-\cref{tab:surfaceparams}.
-
-In \CCLI, we test the sensitivity of the ice sheet model to five different input present-day
-climatologies, $\{T_{m0}, P_{m0}\}$ \cref{tab:datasources}, while adopting a range of constant temperature-offset corrections,
-${\Delta}T_{TS}$, between -10 and 0\,K. In \CCYC,
-on the opposite, we test ice sheet model sensitivity to time dependent temperature-offset
-corrections derived from six different palaeo-temperature proxy records. All
-climatic and topographic data sources used are summarised in
-\cref{tab:datasources}.
+(\cref{tab:surfaceparams}).
+
+In \CCLI, we test the sensitivity of the ice sheet model to five different
+input present-day climatologies, $\{T_{m0}, P_{m0}\}$ (\cref{tab:datasources}),
+while adopting a range of constant temperature-offset corrections,
+${\Delta}T_{TS}$, between -10 and 0\,K. In \CCYC, on the opposite, we test ice
+sheet model sensitivity to time dependent temperature-offset corrections
+derived from six different palaeo-temperature proxy records
+(\cref{tab:datasources}).
 

### 2014-07-18 kappa.tex
Polish results.

@@
 % ----------------------------------------------------------------------
-\subsection{Constant-climate simulations}
+\subsection{Simulations of the Last Glacial Maximum}
 % ----------------------------------------------------------------------
@@
 
-As a first step towards a simulation of the Cordilleran ice sheet through the
+As a first step towards a simulations of the Cordilleran ice sheet through the
 last glacial cycle, I have opted to run the ice sheet model under a range of
 constant temperature depressions, targeting the LGM as the period for which ice
-sheet limits are best known. These simulations, presented in \CCLI and run at
-10\,km grid resolution, use constant temperature depressions from the present
-climate state (using five different input climatologies). All model runs start
+sheet limits are best known. These simulations, presented in \CCLI\ and
+performed on a 10\,km-resolution grid, use constant temperature depressions
+from the present climate state, which is represented using five different input
+climatologies. All model runs start
 from ice free conditions and run for 10\,ka, a period chosen as representative
@@
 % ----------------------------------------------------------------------
-\subsection{Importance of daily temperature variability for ice sheet modelling}
+\subsection{Role of daily temperature variability on melt}
 % ----------------------------------------------------------------------
@@
 (\cref{fig:model-pdd,tab:surfaceparams}), and the standard deviation of
-near-surface air temperature,$\sPDD$
+near-surface air temperature, $\sPDD$
 (\cref{eqn:calovgreve,tab:surfaceparams}). Although the first two can only be
@@
 modelling introduces significant errors in modelled PDD and surface mass
-balance calculations. The transient climate simulations presented in \CCYC\
-therefore now include this effect.
-
-The analysis is taken further extended in \PSDP, using re-analysed air
+balance calculations. Therefore, the Cordilleran ice sheet simulations
+presented in \CCYC\ now include this effect.
+
+The analysis is extended in \PSDP, using re-analysed air
 temperatures from ERA-40 \citep{Uppala.etal.2005}. Although of coarser spatial
@@
 and seasonally variable $\sPDD$ input (PISM issue~\issue{179}), and the
-parameterisation expressed by \cref{eqn:sigma} over ice-covered grid cells
+parameterisation expressed by \cref{eqn:sigma} over ice-filled grid cells
 (PISM issue~\issue{265}). Their effect on modelled Cordilleran ice sheet
 surface elevations is demonstrated here using a constant climate set-up similar
-to the one used in \CCLI\ (\cref{fig:plot-sdeffect}).
-
-\todo{You need to insert a couple of sentences of what you learn from Figure
-      16!}
+to the one used in \CCLI\ (\cref{fig:plot-sdeffect}). These simulations show
+that the effect of daily variability is strongest in continental regions, where
+the $\sPDD$ values are highest (cf. \cref{fig:plot-sdmap}). Where comparing the
+results to the mapped LGM ice limits (\cref{fig:map-cordillera}), which show
+a restricted ice extent in the continental areas despite low temperatures at
+present (\cref{fig:plot-atm}), it appears that including daily temperature
+variability in the ice sheet model results in closer agreement with these
+limits (\cref{fig:plot-sdeffect}b and~c). However, applying the parametrisation
+derived for the Greenland ice sheet (\cref{eqn:sigma}) over ice-filled grid
+cells partly reverts this effect (\cref{fig:plot-sdeffect}d), indicating that
+this parametrization may not be applicable to the former Cordilleran ice sheet.
 
@@
 % ----------------------------------------------------------------------
-\subsection{Glacial cycle simulations}
+\subsection{Simulations of the last glacial cycle}
 % ----------------------------------------------------------------------
@@
 from six proxy records to produce a time-dependent climate forcing (\CCYC,
-Fig.~2). The resulting simulations, using horizontal resolutions of 6 and
+Figs.~2--3). The resulting simulations, using horizontal resolutions of 6 and
 10\,km, all produced two major glaciations during the last glacial cycle, at
@@
 
-Geological evidence for the timing and extent of maximum glaciation durng the
+Geological evidence for the timing and extent of maximum glaciation during the
 LGM allowed me to select two simulations, driven by the GRIP
-\citep{Dansgaard.etal.1993} and EPICA \citep{Jouzel.etal.2007} ice core
-records, as producing the most realistic reconstructions of the last glacial
-cycle. I re-ran these two simulations at a higher resolution of 6\,km, in order
-to allow for a more detailed study of transient ice caps before and after the
-LGM, of the stability of major ice-dispersal centres, and of patterns and rates
-of deglaciation (\CCYC, Figs.~5--15). Modelled ice volume evolution (expressed
-as metres of sea level equivalent) and snapshots of ice extent and ice flow
-patterns for the GRIP simulation (\cref{fig:plot-snapshots}) shows...
-\todo{complete}.
-
-\todo{Verify that figure numbers remain correct as \CCYC\ evolves.}
+\citep{Dansgaard.etal.1993} and EPICA \citep{Jouzel.etal.2007} ice core records
+(\CCYC, Figs.~5--6), as producing the most realistic reconstructions of the
+last glacial cycle. Using aggregated model output from these two runs, I
+analysed the location of transient ice caps before and after the LGM, the
+longevity of major ice-dispersal centres, and the rates and patterns of
+deglaciation (\CCYC, Figs.~6--15), in relation to available geological
+evidence. Modelled ice volume evolution, expressed as metres of sea level
+equivalent, and snapshots of ice extent and flow patterns from the GRIP
+simulation (\cref{fig:plot-snapshots}) show the growth and decay of the
+Cordilleran ice sheet over its last expansion cycle. Ice sheet growth begins on
+a residual ice cap located over the Skeena Mountains, which also corresponds
+to the location of the last remnants of the ice sheet towards the end of the
+deglaciation.
 

### 2014-07-19 kappa.tex
Rework 'perspectives' into 'sources of error'.

@@
 % ======================================================================
-\section{Conclusion and perpectives}
+\section{Sources of error}
 % ======================================================================
 
-In this thesis, I present numerical reconstructions of the former Cordilleran
-ice sheet to a level of detail that has, to my knowledge, not been attained
-before. The coupled SIA-SSA approximation is crude, yet it is currently the
-only approach that allows, combined with massively parallel computing, high
-spatial resolution over the multi-millenial time scales of ice sheet growth and
-decay.
-
 Through sensitivity studies, I have identified the predominant sources of
-uncertainty, which I have assessed using a combination of detailed
-meteorological datasets and available geological evidence.\todo{insert a
-statement which these sources of uncertainties are} Simulations supporting this
-thesis yielded estimates of the LGM Cordilleran ice sheet extent and volume,
+uncertainty involved in numerical modelling of Cordilleran ice sheet. Through
+the different papers constituting this thesis, I present an assessment of some
+of these sources, including the choice of present-day atmospheric forcing, the
+effects of daily temperature variability, and the choice of palaeo-temperature
+time series used to drive the model through past climate history.
+
+The resulting simulations, combined with available geological evidence,
+yielded estimates of the LGM Cordilleran ice sheet extent and volume,
 identification of the main ice-dispersal centres througout the last glacial
@@
 considerable room for model improvements and further analysis of the simulation
-output \todo{introduce your perspectives here}.
-
-% ----------------------------------------------------------------------
-\subsection{Precipitation change}
-% ----------------------------------------------------------------------
-
-One of the predominant uncertainties, for which Cordilleran ice sheet modelling
-is highly sensitive, is the treatment of precipitation. Throughout the
-modelling, we base precipitation patterns on current patterns derived from
-reanalysis data. Although the distribution of precipitation has undoubtedly
+output. Sources of error not assessed in this thesis include potential
+buttressing effects of the Laurentide ice sheet, past changes in precipitation,
+and regional variability of the geothermal heat flux.
+
+% ----------------------------------------------------------------------
+\subsection{Interaction with the Laurentide ice sheet}
+% ----------------------------------------------------------------------
+
+Certainly the most important assumption made in this thesis is that of a
+Cordilleran ice sheet largely disconnected from its much larger neighbour, the
+Laurentide ice sheet. Geomorphological evidence from the suture zone indicates
+that, during the LGM, both Cordilleran and Laurentide ice flow has been
+deflected by their junction \citep{Kleman.etal.2010, Margold.etal.2013,
+Margold.etal.2013a}, although it remains unclear for how long this junction has
+lasted \citep{Jackson.etal.1997, Bednarski.Smith.2007, Gowan.2013}.
+
+Nevertheless, during most of the last glacial cycle, the two ice sheets have
+been disconnected \citep{Kleman.etal.2010}. Thus, this assumption only affects
+numerical reconstructions of the Cordilleran ice sheet presented here during
+the last glacial maximum and during the subsequent deglacation. Other numerical
+modelling results \citep{Gregoire.etal.2012} indicate the Laurentide ice sheet
+may have formed a buttress to ice flow from the Western Cordillera, resulting
+in a thicker Cordilleran ice sheet. However modelling alone will not elucidate
+the nature and duration of the interaction between of the Cordilleran and
+Laurentide ice sheets. For this, additional field constraints will be needed.
+
+% ----------------------------------------------------------------------
+\subsection{Past changes in precipitation}
+% ----------------------------------------------------------------------
+
+Another important source of error is the treatment of precipitation, to which
+the numerical ice sheet model is highly sensitive (\CCLI). In all sensitivity
+studies, we base precipitation patterns on current values derived atmospheric
+reanalyses. Although the distribution of precipitation has undoubtedly
 been affected by past changes in global climate and the intermittent presence
@@
 In meteorology, these effects are typically analysed using atmospheric
-circulation models. However, I have shown in \CCLI\ that different atmospheric
-reanalyses which are circulation models that use data assimilation techniques,
-fail to agree regarding the present-day pattern of precipitation over the
-modelling domain. These discrepancies, which arise as a consequence of the
-highly mountainous terrain, have dramatic effects on the ice sheet model
-response. Considering additional difficulties inherent to modelling past
-climates, and the lack of data for assimilation, we must regard derived ice
-sheet histories as a highly generalized model output [only expect that
-palaeo-climate reconstructions will produce even higher discrepancies in the
-ice-sheet-model response].
-
-\todo{think tomorrow. can one say something about the robustness of some
-      features (ice caps that form, spreading centers which persist)which occur
-      irrespective of which reanalysis is used?}.
-
-% ----------------------------------------------------------------------
-\subsection{Geothermal heat flux}
-% ----------------------------------------------------------------------
-
-Geothermal heat flux, contributing to the en-glacial temperature distribution
-and the location of basal melt areas, was assumed constant over the modelling
-grid. However, the geology of the study area consists of an amalgamation of
-former subduction zone, resulting in complex geothermal patterns
-\citep{Blackwell.Richards.2004}. I am now at the stage to start including this
-variability as a driver to the ice sheet model, which will offer new
-opportunities to study its output in terms of ice flow patterns, consistent
-basal freezing and basal melting patterns, regions prone for glacial erosion,
-and regions characterised by glacial deposition.
-
-I expect that inclusion of geothermal heat patterns allow modelled basal flow
-directions and distances to be compared with glacial lineations, which is one
-of the main sources of geomorphological evidence used to reconstruct former ice
-sheets \citep{Boulton.Clark.1990, Kleman.etal.1997, Kleman.etal.2010}. For
+circulation models. However, I have shown that different atmospheric reanalyses
+--- circulation models using data assimilation techniques --- fail to agree
+regarding the present-day pattern of precipitation over the modelling domain.
+These discrepancies, which arise as a consequence of the highly mountainous
+terrain, have dramatic effects on the ice sheet model response. Considering the
+lack of observational data for modelling past climates, the increased
+computational demand, and a decreased familiarity to processes at play, we must
+expect that, for ice sheet simulations forced by different palaeo-climate
+simulations, the uncertainty associated to the choice of atmospheric forcing
+will be even higher. Perhaps alternative methods of coupling between ice sheet
+evolution and precipitation changes need to be developed.
+
+% ----------------------------------------------------------------------
+\subsection{Geothermal heat flux patterns}
+% ----------------------------------------------------------------------
+
+Geothermal heat flux forcing, a major contributor to the modelled distribution
+of en-glacial temperatures, and the location of basal melt areas, was assumed
+constant over the modelling domain. However, the geology of the study area
+consists of an amalgamation of former subduction zones, resulting in complex
+geothermal patterns \citep{Blackwell.Richards.2004}. I am now at the stage to
+start including this variability as a driver to the ice sheet model, which will
+offer new opportunities to study its output in terms of ice flow patterns,
+consistent basal freezing and basal melting patterns, regions prone for glacial
+erosion, and regions characterised by glacial deposition.
+
+I expect that inclusion of geothermal heat patterns will allow modelled basal
+flow directions and distances to be compared with glacial lineations, which is
+one of the main sources of geomorphological evidence used to reconstruct former
+ice sheets \citep{Boulton.Clark.1990, Kleman.etal.1997, Kleman.etal.2010}. For
 instance, in our simulations, large portions of the Cordilleran ice sheet
-remain cold-based throughout deglaciation (\CCYC, Fig.~15). As cold-based
+remain cold-based throughout the deglaciation (\CCYC, Fig.~15). As cold-based
 glaciers typically do not impact their beds significantly, this could indicate
@@
 
-% ----------------------------------------------------------------------
-\subsection{The Cassiar ice cap}
-% ----------------------------------------------------------------------
-
-In the summer of 2012, I have co-organised and participated in a fieldwork with
-Martin Margold and Ping Fu to collect samples of glacial deposits in the
-southern Cassiar Mountains of northern British Columbia
-(\cref{fig:map-cordillera}). The primary target of the fieldwork was a
-triple-crested, sharp, ca. 50\,m-high moraine located in the vicinity of Tuya
-Lake on the western slope of the Cassiar Mountains
-(\cref{fig:photo-tuya-moraine}). Although this is a unique feature in the area,
-the size of the moraine and its location at the foothill of the mountain range
-indicate that it was likely formed by a tributary glacier from an ice cap,
-rather than a stand-alone valley glacier. The relatively low-elevated Cassiar
-Mountains do not appear as a major nucleation centre in our simulations (\CCYC,
-Fig.~5). Therefore, it is tempting to think that such a large feature may
-result from the late-glacial re-advance of a remnant of the decaying ice sheet.
-High-resolution modelling of the deglaciation phase, perhaps at a regional
-scale, may help to study this hypothesis.
-
-\begin{figure}
-  \centering
-  \makebox[0pt]{\includegraphics{photo-tuya-moraine}}
-  \caption{The Tuya Lake Moraine in northern British Columbia.}
-  \label{fig:photo-tuya-moraine}
-\end{figure}
+% ======================================================================
+\section{Conclusions}
+% ======================================================================
+
+In this thesis, I present numerical reconstructions of the former Cordilleran
+ice sheet to a level of detail that has, to my knowledge, not been attained
+before. Although mathematically uncertain, the coupled SIA-SSA approximation,
+combined with massively parallel computing, allowed ice sheet modelling at
+high spatial resolution and over multi-millenial time scales characteristic
+of ice sheet growth and decay.
 

### 2014-07-19 kappa.tex
Add real conclusions.

@@
 Cordilleran ice sheet largely disconnected from its much larger neighbour, the
-Laurentide ice sheet. Geomorphological evidence from the suture zone indicates
-that, during the LGM, both Cordilleran and Laurentide ice flow has been
-deflected by their junction \citep{Kleman.etal.2010, Margold.etal.2013,
+\emph{Laurentide ice sheet}. Geomorphological evidence from the suture zone
+indicates that, during the LGM, both Cordilleran and Laurentide ice flow has
+been deflected by their junction \citep{Kleman.etal.2010, Margold.etal.2013,
 Margold.etal.2013a}, although it remains unclear for how long this junction has
@@
 
-Another important source of error is the treatment of precipitation, to which
+Another important source of error is the treatment of \emph{precipitation}, to
+which
 the numerical ice sheet model is highly sensitive (\CCLI). In all sensitivity
@@
 
-Geothermal heat flux forcing, a major contributor to the modelled distribution
+\emph{Geothermal heat flux} forcing, a major contributor to the modelled
+distribution
 of en-glacial temperatures, and the location of basal melt areas, was assumed
@@
 
+Through sensitivity studies, and comparison of model output to geological
+evidence, I have gained a personal understanding of Cordilleran ice sheet
+evolution through the last glacial cycle. I attempt to summarise my thoughts
+here in a few points, perhaps also applicable to other ice sheets in the
+Northern Hemisphere.
+
+\begin{enumerate}
+
+\item Although spatial patterns of glaciation in the North American Cordillera
+are sometimes paradoxical, they can be largely explained by the present-day
+distribution of temperature and precipitation in the region, provided an
+accurate treatment of seasonality, orographic precipitation, and daily
+variability in the model, and a simple temperature-index accumulation and melt
+model.
+
+\item The LGM Cordilleran ice sheet corresponds to a transient stage, out of
+balance with the contemporaneous atmospheric temperatures. Its maximum extent
+is the result of a 50\,ka-long evolution, captured in the geologic record at a
+moment of transition from colder to warmer temperatures than those needed to
+sustain a constant ice sheet geometry. Therefore, this maximum stage is not
+representative of average conditions of the last glacial cycle.
+
+\item The patterns of growth and decay of the Cordilleran ice sheet are varied
+and differ by region and by epoch. Two-dimensional, conceptual models of growth
+and decay \citep[e.g.][]{Fulton.1991, Margold.etal.2013a} do not encompass this
+diversity. In order to decipher the complexity of geological evidence, the
+Cordilleran ice sheet needs to be thought of as a three-dimensional body of
+moving ice in perpetual transformation towards evolving climate conditions.
+
+\end{enumerate}
+
 % References

### 2014-07-19 kappa.tex
Various figure adjustements.

@@
 
-% Finally to not include graphics
-%\renewcommand{\includegraphics}[2][,]{}
-%\renewcommand{\subgraphics}[3][,]{}
+% Some general low-priority todos
+% -- switch to kappa.cls
+% -- incorporate figs as pdf, rasterize heavy elements
+% -- use latex in matplotlib figures
+% -- homogeneize subfig label fonts
+% -- add software references
+% -- redraw PISM interface
 
@@
            The grey bands indicate the approximate time span of the last
-           glacial period of 110 to 10\,ka.
-           \todo{Put nicer blue and red colours.}}
+           glacial period of 110 to 10\,ka.}
   \label{fig:paleo-timeseries}
@@
            \textbf{(b)} during the last glacial maximum
-           \citep{Ehlers.Gibbard.2007}.
-           \todo{Make these globes look a little bigger. Add (a) and (b).}}
+           \citep{Ehlers.Gibbard.2007}.}
   \label{fig:paleo-glaciation}
@@
   \centering
-  \includegraphics[width=80mm]{model-interfaces}
-  \caption{Model interfaces in PISM. From \citet{PISM-authors.2014}.
-           \todo{This is directly from the PISM manual. I may redraw it here
-                 if I have the time.}}
+  \includegraphics[width=85mm]{model-interfaces}
+  \caption{PISM is a numerical model for ice and bedrock physics. Processes
+           occurring at the interface with the atmosphere and the ocean, that
+           are progressive transitions in nature, are parametrised along
+           well-defined boundaries in the model.
+           From \citet{PISM-authors.2014}.}
   \label{fig:model-interfaces}
@@
   \caption{Sketch of notations used for the main variables of the ice sheet
-           model. A particle of ice is moving at velocity $\vv$, under the
-           balance of stresses $\CST$ and gravity $\vect{g}$.}
+           model. An infinitesimal ice parcel is moving at velocity $\vv$,
+           under the balance of stresses $\CST$ and gravity $\vect{g}$.}
   \label{fig:model-variables}
@@
   \includegraphics[width=80mm]{model-siassa}
-  \caption{Shallow Ice Approximation (SIA) and Shallow Shelf Approximation
-           (SSA) velocities in PISM. From \citet{Winkelmann.etal.2011}.
-           \todo{Redraw in LaTeX in the same style as
-                 \cref{fig:model-variables}.}}
+  \caption{Combination of Shallow Ice Approximation (SIA) and Shallow Shelf
+           Approximation (SSA) velocities in PISM, assuring a smooth transition
+           between shear-dominant and longitudinal stress regimes. After
+           \citet[Fig.~1]{Winkelmann.etal.2011}. Not to scale.}
   \label{fig:model-siassa}
@@
     \caption{Constant-climate simulation of the LGM Cordilleran ice sheet using
-             an horizontal resolution of 3\,km. Modelled basal topography,
-             cold-based areas (hatches), surface topography (100\,m contours)
-             and surface velocities. The simulation is forced by atmospheric
+             an horizontal resolution of 3\,km. The map depicts modelled basal
+             topography, cold-based areas (hatches), surface topography (100\,m
+             contours) and surface velocities, and a reconstruction of the
+             approximate ice sheet margin at 16.8\,cal\,ka from compiled
+             geomorphological evidence \citep[red contour;][]{Dyke.2004}.
+             The simulation is forced by atmospheric
              fields from the NARR \citep{Mesinger.etal.2006} and a temperature
@@
              anomalously low due to brutal application of cold climate
-             condition to ice-free bedrock at the inception of the run.
-             \todo{Add black frame, and maybe the LGM outline from Dyke?}}
+             condition to ice-free bedrock at the inception of the run.}
     \label{fig:plot-lgm}

### 2014-07-19 figures/paleo-glaciation.py

@@ -15,12 +15,18 @@ scale = '50m'
 proj = ccrs.Orthographic(central_longitude=-90.0, central_latitude=75.0)
 
 # Initialize figure in inches
-figw, figh = 80.0/25.4, 40.0/25.4
+figw, figh = 85.0/25.4, 85.0/25.4
 fig = plt.figure(0, (figw, figh))
 fig.subplots_adjust(left=0.5/72/figw, bottom=0.5/72/figh,
                     right=1-0.5/72/figw, top=1-0.5/72/figh,
                     wspace=1/((1+figw/(1./72))/2-1))
-grid = [fig.add_subplot(121+i, projection=proj) for i in range(2)]
+figw, figh = 85.0, 85.0
+#grid = [fig.add_subplot(121+i, projection=proj) for i in range(2)]
+mapw = 50
+grid = [fig.add_axes([2.5/figw, 1-(2.5+mapw)/figh, mapw/figw, mapw/figh],
+                     projection=proj),
+        fig.add_axes([1-(2.5+mapw)/figw, 2.5/figh, mapw/figw, mapw/figh],
+                     projection=proj)]
 
 # draw common parts
 for ax in grid:
@@ -30,7 +36,7 @@ for ax in grid:
 # add present glaciers
 grid[0].add_feature(cfeature.NaturalEarthFeature(
     category='physical', name='glaciated_areas', scale=scale,
-    edgecolor='none', facecolor='#0978ab', alpha=0.5,))
+    edgecolor='none', facecolor='#1f78b4', alpha=0.5,))  # alt. #0978ab
 
 # add Ehlers & Gibbard 2003 LGM outline
 # try to identify duplicates using centroids
@@ -50,6 +56,11 @@ for i, geom in enumerate(shp.geometries()):
         centroids.append((x, y))
         grid[1].add_geometries(geom, ccrs.PlateCarree(),
                   edgecolor='none', facecolor='#0978ab', alpha=0.5)
+
+# add subfigure labels
+for i, ax in enumerate(grid):
+    ax.text(0.9, 0.9, '(%s)' % chr(97+i),
+            fontweight='bold', transform=ax.transAxes)
 
 # save
 fig.savefig('paleo-glaciation')

### 2014-07-19 figures/paleo-timeseries.py

@@ -36,7 +36,7 @@ fig.subplots_adjust(left=10.0/figw, bottom=7.5/figh,
 ax = grid[0]
 age, d18o = np.genfromtxt('data/lr04.txt', skip_header=89, skip_footer=9387,
                           unpack=True, usecols=(0, 1))
-ax.plot(age, d18o, 'b')
+ax.plot(age, d18o, '#1f78b4')  # alt. wiki color #0978ab
 
 # set LR04 axes properties
 ax.set_ylim(5.5, 2.5)
@@ -49,7 +49,7 @@ ax = grid[1]
 age, temp = np.genfromtxt('data/epica.txt', delimiter=(4, 13, 17, 13, 13),
                           skip_header=104, skip_footer=1,
                           unpack=True, usecols=(2, 4))
-ax.plot(age/1000.0, temp, 'r')
+ax.plot(age/1000.0, temp, '#e31a1c')  # alt. wiki color #e0584e
 
 # set EPICA axes properties
 ax.set_yticks(range(-12, 6, 4))

### 2014-07-19 figures/plot-atm.py

@@ -88,5 +88,10 @@ for ax in grid:
 # close
 nc.close()
 
+# add subfigure labels
+for i, ax in enumerate(grid):
+    ax.text(0.04, 0.96, '(%s)' % chr(97+i),
+            fontweight='bold', transform=ax.transAxes)
+
 # save
 fig.savefig('plot-atm')

### 2014-07-19 figures/plot-lgm.py

@@ -10,6 +10,8 @@ from netCDF4 import Dataset
 from matplotlib import pyplot as plt
 from matplotlib.colors import LinearSegmentedColormap
 from iceplot import plot as iplt
+import cartopy.crs as ccrs
+import cartopy.io.shapereader as shpreader
 
 # file path
 filename = '/home/julien/pism/archive/cordillera-climate-3km/' \
@@ -17,11 +19,17 @@ filename = '/home/julien/pism/archive/cordillera-climate-3km/' \
            'y0010000-extra.nc'
 t = -1
 
+# Canadian Atlas Lambert projection
+proj = ccrs.LambertConformal(
+    central_longitude=-95.0, central_latitude=49.0,
+    false_easting=0.0, false_northing=0.0,
+    secant_latitudes=(49.0, 77.0), globe=None, cutoff=0)
+
 # prepare figure
 figw, figh = 112.5, 225.0
 fig = plt.figure(0, (figw/25.4, figh/25.4))
-ax  = fig.add_axes([0, 0, 1, 1])
-ax.axis('off')
+ax  = fig.add_axes([0, 0, 1, 1], projection=proj)
+ax.outline_patch.set_linewidth(2.0)  # only with cartopy
 nc = Dataset(filename)
 
 # bed topography
@@ -45,8 +53,15 @@ cs = iplt.contour(nc, 'usurf', t, levels=range(1000,6000,1000),
 cs.clabel(fontsize=6, fmt='%g', linewidths=0.5)
 
 # ice margin
-cs = iplt.icemargin(nc, t)
+cs = iplt.icemargin(nc, t, linewidths=0.5, zorder=4)
 nc.close()
+
+# Dyke margin
+sf = shpreader.Reader('data/ice14k.shp')
+for rec in sf.records():
+    if rec.attributes['SYMB'] == 'ICE':
+        ax.add_geometries(rec.geometry, proj,
+                          edgecolor='#800000', facecolor='none', zorder=0.5)
 
 # colorbars
 cax = fig.add_axes([5/figw, 10/figh, 2.5/figw, 40/figh])

### 2014-07-19 figures/plot-snapshots.py

@@ -70,7 +70,7 @@ figinvtrans = fig.transFigure.inverted()
 
 # plot time series
 print 'plotting time series...'
-ax.plot(time, ivol, color='#0978ab')
+ax.plot(time, ivol, color='#1f78b4')  # alt. wiki color #0978ab
 ax.set_xlabel('model time (kyr)')
 ax.set_ylabel('ice volume (m s.l.e.)')
 ax.yaxis.set_ticks_position('right')

### 2014-07-19 figures/model-pdd.tex
Rework PDD model sketch.

@@
-\documentclass[border=5mm,tikz]{standalone}
-\usepackage{amsmath}
-\usetikzlibrary{chains}
-\usetikzlibrary{arrows}
+\documentclass[border=0mm,tikz]{standalone}
 \usetikzlibrary{shapes}
+\usetikzlibrary{calc}
 
-\newcommand{\PDD}{\ensuremath{\text{PDD}}}
-\newcommand{\Hs}{\ensuremath{H_{\text{snow}}}}
-\newcommand{\Ms}{\ensuremath{\text{Melt}_{\text{snow}}}}
-\newcommand{\Mi}{\ensuremath{\text{Melt}_{\text{ice}}}}
+% TikZ styles
+\tikzstyle{block} = [draw, rectangle, fill=gray!5, rounded corners,
+                     text centered, minimum height=1cm]
+\tikzstyle{trunk}  = [block, text width=55mm, minimum width=60mm]
+\tikzstyle{branch} = [block, text width=32.5mm, minimum width=35mm]
+\tikzstyle{split} = [draw, diamond, fill=gray!5, minimum size=2.5cm]
+\newcommand{\PDD}{\mathrm{PDD}}
+\newcommand{\SMB}{\mathrm{SMB}}
 
+% begin document
 \begin{document}
-  \begin{tikzpicture}[ auto, thick, font=\small, node distance=3mm,
-    block/.style={rectangle, rounded corners, draw, fill=gray!5, text width=2in, text centered,
-      minimum width=2in, minimum height=1cm},
-    big block/.style={block, text width=3in},
-    decision/.style={diamond, draw, fill=gray!5},
-    line/.style={draw, -latex'},
-    start chain=1 going below, every node/.style={on chain},
-    every join/.style={-latex'},
-    ]
+\begin{tikzpicture}[thick, font=\small, >=latex]
 
-    \node [big block] {Compute snow accumulation \mbox{``Accumulation''}, discarding rain};
-    \node [big block, join] {Compute the number of \mbox{positive} degree days (\PDD) };
-    \node [big block, join] {Update snow depth \mbox{$\Hs \gets \Hs + \text{Accumulation}$}};
-    \node [big block, join] {Compute maximum possible snow melt \mbox{$\Ms^{\text{max}} := \PDD\times F_{s}$}};
-    \node (melt all snow) [decision, join] {$\Ms^{\text{max}} > \Hs?$};
+% background
+\fill[white] (-0.25,-0.25) rectangle +(8.5,14.75);
+%\draw[help lines, lightgray] (0,0) grid (8,14);
 
-    \begin{scope}[start branch=yes, node distance=4em]
-      \node (ice melt case) [block, on chain=going right] {\mbox{Melt all available snow:} $\Ms := \Hs$};
-    \end{scope}
-    \begin{scope}[start branch=no, node distance=4em]
-      \node (no ice melt case) [block, on chain=going left] {Melt some of the snow: \mbox{$\Ms := \Ms^{\text{max}}$}};
-    \end{scope}
+% add nodes
+\path     (4, 13.75) node [trunk] (acc) {Compute snow accumulation, $A_s$}
+  -- ++(0, -1.5) node [trunk] (pdd) {Compute PDDs}
+  -- ++(0, -1.5) node [trunk]  (s1) {Update snow depth \\
+                                      $S \gets S + A_s$}
+  -- ++(0, -1.5) node [trunk] (msp) {Compute potential snow melt \\
+                                      $M_{s,p} := \PDD \times F_{s}$}
+  -- ++(0, -2.25) node [split] (split) {$M_{s,p} > H_s?$};
 
-    \node (excess pdds) [big block] {$\PDD_{\text{excess}} := \PDD - \Ms / F_{s}$};
-    \node [big block, join] {Melt some ice: \mbox{$\Mi := \PDD_{\text{excess}} \times F_{i}$}};
-    \node (refreeze ice) [decision, join] {Refreeze melted ice?};
+\path ($(split)+(-2,-2)$) node [branch] (no)  {Melt some snow \\
+                                               $M_s := M_{s,p}$}
+  -- ++(0, -1.5) node [branch] (nomi) {Melt no ice \\
+                                      $M_i := 0$};
 
-    \begin{scope}[start branch=yes, node distance=4em]
-      \node (refreeze both) [block, on chain=going right] {Refreeze is \mbox{$(\Ms + \Mi) \times \theta_{\text{refreeze}}$}};
-    \end{scope}
+\path ($(split)+(+2,-2)$) node [branch] (yes) {Melt all snow \\
+                                               $M_s := H_s$}
+  -- ++(0, -1.5) node [branch] (yesmi) {Melt some ice \\
+                                        $M_i := (\PDD - \frac{M_s}{F_s}) \times F_i$};
 
-    \begin{scope}[start branch=no, node distance=4em]
-      \node (refreeze snow) [block, on chain=going left] {Refreeze is $\Ms \times \theta_{\text{refreeze}}$};
-    \end{scope}
+\path ($(split)+(0,-5)$) node [trunk](smb) {Compute surf. mass balance \\
+                                      $\SMB := A_s - M_s - M_i$}
+  -- ++(0, -1.5) node [trunk]  (s2) {Update snow depth \\
+                                      $S \gets S - M_s$};
 
-    \node (runoff) [big block] {$\text{Runoff} := \Ms + \Mi - \text{Refreeze}$};
-    \node [big block, join] {Compute surface mass balance: $\text{SMB} := \text{Accumulation} - \text{Runoff}$};
-    \node [big block, join] {Update snow depth: $\Hs \gets \Hs - \Ms$};
-    % special edges:
-    \path[line] (melt all snow) -- node {Yes} (ice melt case);
-    \path[line] (melt all snow) -- node {No}  (no ice melt case);
-    \path[line] (ice melt case) |- (excess pdds);
-    \path[line] (no ice melt case) |- (excess pdds);
-    \path[line] (refreeze ice) -- node {Yes} (refreeze both);
-    \path[line] (refreeze ice) -- node {No}  (refreeze snow);
-    \path[line] (refreeze both) |- (runoff);
-    \path[line] (refreeze snow) |- (runoff);
+% add connections
+\draw[->] (acc) -- (pdd);
+\draw[->] (pdd) -- (s1);
+\draw[->] (s1) -- (msp);
+\draw[->] (msp) -- (split);
+\draw[->] (split) -| (no) node [pos=0.3, above] {No};
+\draw[->] (no) -- (nomi);
+\draw[->] (nomi) -- (smb);
+\draw[->] (split) -| (yes) node [pos=0.3, above] {Yes};
+\draw[->] (yes) -- (yesmi);
+\draw[->] (yesmi) -- (smb);
+\draw[->] (smb) -- (s2);
+
 \end{tikzpicture}
-
 \end{document}

### 2014-07-19 kappa.tex
Rework PDD model sketch.

@@
 $T_m$, and monthly precipitation, $P_m$, using a temperature-index model
-\citep[e.g.,][]{Hock.2003}. Surface accumulation, $a_s$, equals precipitation
+\citep[\cref{fig:model-pdd}; cf.][]{Hock.2003}. Snow accumulation, $A_s$,
+equals precipitation
 when temperature is below ${T_0=0\,\degC}$, and decreases to zero\renote{
@@
 \begin{equation}
-    a_s =
+    A_s =
     \begin{cases}
@@
 
-Surface melt, $m_s$, is assumed proportional to the number of positive degree
+Surface melt, $M$, is assumed proportional to the number of positive degree
 days (PDD), defined over an arbitrary time interval, $[t_1, t_2]$, as the
@@
   \centering
-  \includegraphics[width=\linewidth]{model-pdd}
-  \caption{PDD model. From \citet{PISM-authors.2014}
-           \todo{This is again directly from the PISM manual. I can simplify
-                 it by removing the unused refreezing branch, I will match
-                 notations with the text and refer to equations in use.}}
+  \includegraphics{model-pdd}
+  \caption{Positive degree-day algorithm used to compute accumulation and melt
+           at the ice sheet surface. Adapted from \citet{PISM-authors.2014}.}
   \label{fig:model-pdd}

### 2014-07-20 cordillera-cycle.tex
Add corrections from Arjen and Irina.

@@
   offsets from six proxy records located around the globe. Although this
-  approach reveals large variations in model response to evolving atmospheric
+  approach reveals large variations in model response to evolving climate
   forcing, all simulations produce two major glaciation events during
-  marine isotope stages~4 (61.9--55.4\,ka) and~2
+  marine oxygen isotope stages~4 (61.9--55.4\,ka) and~2
   (29.5--16.9\,ka). The timing of glaciation is
@@
 \citep{Dawson.1888}, extended from the Alaska Range in the north to the
-North Cascades in the south. In addition, it extended offshore, where it calved
+North Cascades in the south (Fig.~\ref{fig:locmap}).
+In addition, it extended offshore, where it calved
 into the Pacific Ocean, and merged with the western margin of its much larger
-neighbour, the Laurentide ice sheet, east of the Rocky Mountains
-(Fig.~\ref{fig:locmap}).
+neighbour, the Laurentide ice sheet, east of the Rocky Mountains.
 
@@
     \citep{Carrara.etal.1996},
-moraines that demarcate the south-western and north-eastern margins provide key
+moraines that demarcate the northern and south-western margins provide key
 constraints that allow reasonable reconstructions of maximum ice sheet extents
@@
 it appears to be more sparse than for formerly glaciated regions surrounding
-the North Atlantic \aref\footnote{
-    Do we need to add a couple of refs. from Britain, Scandinavia, the Alps,
-    Eastern America here? Or a review paper? Could yuo help me with this?},
+the North Atlantic
+    \citep[e.g.,][]{Sissons.1979, Lundqvist.1987,
+                    Ivy-Ochs.etal.1999, Stea.etal.2011},
 thus allowing for considerable uncertainty concerning
@@
 seasonality and distinct patterns of precipition, thus requiring the use of
-high-resolution atmospheric forcing fields as an input to the ice sheet model
+high-resolution climate forcing fields as an input to the ice sheet model
 \citep{Seguinot.etal.2014}.
@@
 requires input on basal topography, sea level, geothermal heat flux and
-atmospheric\renote{
-    Changed here from ``climate'' to ``atmospheric''. Although I have used
-    ``climate forcing'' in the previous paper, I now think that ``atmospheric
-    forcing'' is more appropriate: ``climate'' may also encompass the oceanic
-    component, which is not assessed here.}
-forcing. It computes the evolution of ice extent
+climate forcing. It computes the evolution of ice extent
 and thickness over time, the thermal and dynamic\renote{
@@
 (Sect.~\ref{sec:icedyn}). Surface mass balance is computed using a positive
-degree-day (PDD) model (Sect.~\ref{sec:surface}). Atmospheric forcing is
-provided by a monthly climatology from the North American Regional Reanalysis
-\citep[NARR;][]{Mesinger.etal.2006} perturbated by time-dependent and
+degree-day (PDD) model (Sect.~\ref{sec:surface}). Climate forcing is
+provided by a monthly climatology averaged from 1979 to 2000 from the North
+American Regional Reanalysis \citep[NARR;][]{Mesinger.etal.2006},
+perturbated by time-dependent and
 lapse-rate temperature corrections (Sect.~\ref{sec:atm}).
@@
 formulation \citep{Aschwanden.etal.2012}. Surface air temperature derived from
-the atmospheric forcing (Sect.~\ref{sec:atm}) provides the upper boundary
+the climate forcing (Sect.~\ref{sec:atm}) provides the upper boundary
 condition to the ice enthalpy model. Temperature is computed subglacially to
@@
 
-A~pseudo-plastic sliding law \citep{Bueler.Pelt.2014} relates the
-bed-parallel shear stresses to the sliding velocity\footnote{
-    Should I pull in this equation, too?}.
+A~pseudo-plastic sliding law,
+\begin{equation}
+    \label{eqn:pseudoplastic}
+    \vec{\tau}_b = -\tau_c \frac{\vec{v}_b}{{v_{th}}^q\,|\vec{v}_b|^{1-q}} \,,
+\end{equation}
+relates the bed-parallel shear stresses, $\vec{\tau}_b$, to the sliding
+velocity, $\vec{v}_b$ (Table~\ref{tab:params}).
 The yield stress, $\tau_c$,
@@
 where cohesion, $c_0$, is assumed to be zero. The friction angle, $\phi$,
-varies from 15 to 45{\degree}\renote{
-    Different from the previous paper, but fairly similar in effect (see
-    $\alpha$ and $\delta$ parameter values and corresponding equations in the
-    kappa, basal sliding section). These values are is based on a quick
-    sensitivity test on Greenland.}.
-It is taken to be a piecewise-linear function of modern bed elevation, with
-the lowest value below modern sea level (0\,m above sea level, m~a.s.l) and the
-highest value above the approximate elevation of highest shorelines
-(200\,m~a.s.l), thus accounting for a~weakening of
-till associated with the presence of marine sediments. Effective pressure, $N$,
+varies from 15 to 45\degree as a piecewise-linear function of modern bed
+elevation, with the lowest value occuring below modern sea level (0\,m above
+sea level, m~a.s.l) and the highest value occuring above the generalised
+elevation of the highest shorelines (200\,m~a.s.l), thus accounting for
+a~weakening of till associated with the presence of marine sediments
+(cf. \citealp{Martin.etal.2011}, \citealp[supplement]{Aschwanden.etal.2013},
+\citealp{PISM-authors.2014}). Effective pressure, $N$,
 is related to the ice overburden stress, $\rho gh$, and the modelled amount of
@@
 \end{equation}
-where $delta$ is chosen as 0.02, $e_0$ is a measured reference void ratio and
+where $\delta$ is chosen as 0.02, $e_0$ is a measured reference void ratio and
 $C_c$ a measured compressibility coefficient (Table~\ref{tab:params}). The
@@
     using this model is that computes a non-isostatic response.}
-following a bedrock deformation model that includes point-wise isostasy,
+following a bedrock deformation model that includes local isostasy,
 elastic lithosphere flexure and viscous mantle deformation in the lower
@@
 variability \citep{Seguinot.2013}, $\sigma$ is computed from daily temperature
-values from the North American Regional Reanalysis
-\citep[NARR,][]{Mesinger.etal.2006}, after excluding variability associated
+values from the NARR, after excluding variability associated
 with the seasonal cycle itself \citep[cf.][]{Seguinot.Rogozhina.2014}\renote{
@@
 % -----------------------------------------------------------------------------
-\subsection{Atmospheric forcing}
+\subsection{Climate forcing}
 \label{sec:atm}
@@
 
-Atmospheric forcing of the model consists of a present-day monthly climatology%
+Climate forcing of the model consists of a present-day monthly climatology%
 \renote{
@@
 
-The present-day monthly climatology, $\{T_{m0}, P_{m0}\}$, was computed from
-near-surface air temperature and precipitation rate fields from the NARR
-\citep{Mesinger.etal.2006}, averaged between 1979 and 2000. Modern climate of the
+The present-day monthly climatology was computed from
+near-surface air temperature and precipitation rate fields from the NARR,
+averaged from 1979 to 2000. Modern climate of the
 North American Cordillera is characterised by strong geographic variations in
@@
 daily temperature variability (Fig.~\ref{fig:atm}).
-Our choice of data from the NARR is motivated by the need for an accurate,
+The use of NARR is motivated by the need for an accurate,
 high-resolution precipitation forcing, as identified in a previous sensitivity
@@
 provided as such. All records were scaled linearly (Table~\ref{tab:records}) in
-order to simulate realistic and comparable ice extents at the LGM
-(Fig.~\ref{fig:timeseries}).
+order to simulate comparable ice extents at the LGM  (Table~\ref{tab:extrema})
+and realistic outlines (Fig.~\ref{fig:snapshots}).
 
@@
 of ice surface elevation, $s$, using the NARR surface geopotential height
-invariant field as a reference topography, $b_{ref}$, by
+invariant field as a reference topography, $b_{ref}$:
 \begin{align}
@@
 NARR reference topography, $b_{ref}$, on the other hand. All simulations use an
-annual temperature lapse rate of $\gamma = 6\,\unit{{\degree}C\,km^{-1}}$.
+annual temperature lapse rate of $\gamma = 6\,\unit{K\,km^{-1}}$.
 In the rest of this paper, we refer to different model runs by the name of the
@@
 % =============================================================================
-\section{Sensitivity to atmospheric forcing time-series}
+\section{Sensitivity to climate forcing time-series}
 \label{sec:results}
@@
 
-Despite large differences in the input atmospheric forcing
+I AM HERE.
+
+Despite large differences in the input climate forcing
 (Fig.~\ref{fig:timeseries}, upper panel), model output presents consistent
@@
 magnitude and precise timing of these three events depend significantly on the
-choice of proxy record used for time-dependent atmospheric forcing
+choice of proxy record used for time-dependent climate forcing
 (Table~\ref{tab:extrema})\renote{
@@
 modelled ice volumes during MIS~4 and larger ice volumes during MIS~3.
-More broadly speaking, they produce lower ice volume variability
-throughout the simulation length. The NGRIP atmospheric forcing is the only one
-that results in a larger ice volume during MIS~4 than during the MIS~2.
+Hence, they produce lower ice volume variability
+throughout the simulation length. The NGRIP climate forcing is the only one
+that results in a larger ice volume during MIS~4 than during MIS~2.
 
@@
 informed by the NGRIP and the two oceanic palaeo-temperature records attain their
-maximum ice volume thousands of years earlier. Moreover, the ODP~1012 run
+maximum ice volumes thousands of years earlier. Moreover, the ODP~1012 run
 yields a rapid deglaciation of the modelled area prior to 20\,ka. The ODP~1020
@@
 simulation driven by the GRIP palaeo-temperature record also results in a rapid
-deglaciation but in three steps, separated by two ice volume standstills.
+deglaciation but in three steps, separated by two ice volume standstills
+(Fig.~\ref{fig:timeseries}).
 
@@
 
-Despite such differences in the timing of attained volume maxima and mimima
+Despite such differences in the timing of attained volume extrema
 (Table~\ref{tab:extrema}), all the model runs show relatively consistent
-patterns of glaciation (Fig.~\ref{fig:snapshots}, upper panels). During MIS~4,
-all simulation produce an extensive ice sheet, covering an area of at least
-half that attained during MIS~4. However, the corresponding maximum ice volumes
-vary differ significantly between model runs, and cary between 3.88 and~8.69\,m
-sea-level equivalent (m.~s.l.e; Table~\ref{tab:extrema}).
-
-During the MIS~3 ice volume minima, a central ice cap persists in all
-simulations over the Skeena Mountains (Fig.~\ref{fig:snapshots}, middle
-panels). Although this ice cap is present in all simulations, its dimensions
-depend sensitively on the choice of palaeo-temperature record used to drive the
-model. Modelled ice volume minima vary between 0.89 and 2.44\,m~s.l.e.
-(Table~\ref{tab:extrema}).
-
-The modelled ice sheet geometries during the last glacial maximum (MIS~2;
-Fig.~\ref{fig:snapshots}, lower panels) include a ca. 1500\,km-long central
-divide located about 3500\,m~a.s.l. along the spine of the Rocky Mountains and
-appear highly similar from one simulation to the next. Although the similatity
-in modelled ice extent is a direct result from the choice of scaling factors
-applied to the different palaeo-temperature proxy records
+patterns of glaciation. During MIS~4, all simulations produce an extensive ice
+sheet, covering an area of at least half of that attained during MIS~2
+(Table~\ref{tab:extrema}; Fig.~\ref{fig:snapshots}, upper panels).
+Corresponding maximum ice volumes also differ significantly between model runs,
+and vary between 3.88 and~8.69\,m sea level equivalents (m.~s.l.e;
+Table~\ref{tab:extrema}).
+
+During MIS~3 ice volume minima, a central ice cap persists over the Skeena
+Mountains (Fig.~\ref{fig:snapshots}, middle panels). Although this ice cap is
+present in all simulations, its dimensions depend sensitively on the choice of
+the applied palaeo-temperature record. Modelled ice volume minima spread over a
+wide range between 0.89 and 2.44\,m~s.l.e. (Table~\ref{tab:extrema}).
+
+Modelled ice sheet geometries during the LGM (MIS~2; Fig.~\ref{fig:snapshots},
+lower panels) invariably include a ca. 1500\,km-long central divide located
+about 3500\,m~a.s.l. along the spine of the Rocky Mountains. Although the
+similatity of modelled ice extents is a direct result from the choice of scaling
+factors applied to different palaeo-temperature proxy records
 (Table~\ref{tab:records}), it is interesting to note that modelled maximum ice
-volumes are also contained within a close range of 8.24 to 8.66\,m~s.l.e.
+volumes also fall within a tight range of 8.24 to 8.66\,m~s.l.e.
 (Table~\ref{tab:extrema}).
@@
     similar). I have also renamed the ``results'' to ``Sensitivity to
-    atmospheric forcing time-series'', which is actually what that section is
+    climate forcing time-series'', which is actually what that section is
     about: raw model output, no discussion against the geologic record. This
@@
 
+Large variations in the model response reveal its sensitivity to evolving
+climate forcing. Nevertheless, geological evidence of former glaciations
+provide a basis for validation of our runs, while the results from numerical
+modelling can perhaps help to analyse some of the complexity of this evidence.
+In this section, we compare model output to the geologic record, in terms of
+timing and configuration of the maximum stages, location and durability of the
+major nucleation centres, and patterns of ice retreat during the last
+deglaciation.
+
 % -----------------------------------------------------------------------------
@@
 
-Independently of the choice of palaeo-temperature record\renote{
-    In my view, ``atmospheric/climate forcing'' would be far too broad here. We
-    could possibly apply a much wilder choice of atmospheric forcing
+Independently of the palaeo-temperature records\renote{
+    In my view, ``climate forcing'' would be far too broad here. We
+    could possibly apply a much wilder choice of climate forcing
     techniques, leading just as wild differences in model output
@@
 is obtained during MIS~4 (61.9--55.4\,ka) and the second during MIS~2
-(29.5--16.9\,ka; Figs.~\ref{fig:timeseries} and~\ref{fig:snapshots};
+(29.5--16.9\,ka; Figs.~\ref{fig:timeseries}, \ref{fig:snapshots};
 Table~\ref{tab:extrema}). These events broadly correspond in timing to the
@@
 There is patchy stratigraphical evidence for glaciations older than the Fraser
-Glaciation \citep{Clague.Ward.2011} in British Columbia, and their extent and
-timing are therefore still highly conjectural
+Glaciation \citep{Clague.Ward.2011} in British Columbia, but their extent and
+timing are still highly conjectural
     \citep[perhaps MIS~4 or early MIS~3; e.g.,][]{Cosma.etal.2008}.
 
-However, the exact timing of modelled MIS~2 maxima depends strongly on the
+The exact timing of modelled MIS~2 maximum ice volume depends strongly on the
 choice of applied palaeo-temperature record, which allows for a more in-depth
@@
 dating on wood between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,ka}
-\citep{Porter.Swanson.1998}.
-These dates have been confirmed by the offshore sedimentary record, which shows
+\citep{Porter.Swanson.1998}. These dates are consistent with radiocarbon dates
+from the offshore sedimentary record, which reveals
 an increase of glaciomarine sedimentation between 19.5 and
@@
 has been shown to have weakened during each peak global glaciation (in SPECMAP)
-durint the past 550\,ka, including the LGM, resulting in paradoxically warmer
+during the past 550\,ka, including the LGM, resulting in paradoxically warmer
 sea-surface temperatures at the locations of the ODP~1012 and ODP~1020 sites
@@
 
-Because most of the marine margine of the Cordilleran ice sheet terminated in a
+Because most of the marine margin of the Cordilleran ice sheet terminated in a
 sector of the Pacific Ocean unaffected by variations in the California current,
-it probably remained unaffected by this early warming. However, the above paradox
+it probably remained insensitive to this local phenomenum. However, the above
+paradox
 illustrates the complexity of ice-sheet feedbacks on regional climate, and
@@
 
-Similarly, atmospheric forcing of the model using the NGRIP palaseo-temperature
-record results in an early onset of deglaciation (Fig.~\ref{fig:timeseries})
-after the -22.9\,ka last glacial volume maximum (Table~\ref{tab:extrema},
-incompatible with dated evidence of the last glacial maximum advance\renote{
-    I actually have no explanation for that. Did not have the time to dig much
-    into the ice-core literature. Any idea?}.
-Hence, we focus the rest of our analysis on simulations forced by
-palaeo-temperature records from GRIP and EPICA ice cores. To allow for a more
+Similarly, the simulation using the NGRIP palaeo-temperature record depicts an
+early onset of deglaciation (Fig.~\ref{fig:timeseries}) following its last
+glacial volume maximum attained about 6\,ka earlier (-22.9\,ka,
+Table~\ref{tab:extrema}) than dated evidence of the last glacial maximum
+advance. Hence, we focus the rest of our analysis on simulations forced by
+palaeo-temperature records from the GRIP and EPICA ice cores that appear to
+produce the most realistic reconstructions of regional glaciation history among
+all records tested. To allow for a more
 detailed comparison against the geologic record, these two simulations were
@@
 
-During maximum glaciation, the main meridional ice divide is positioned over
+During maximum glaciation, both simulations position the main meridional ice
+divide over
 the western flank of the Rocky Mountains (Fig.~\ref{fig:snapshots}, lower
@@
 palaeoglaciological reconstructions for central and southern British Columbia
-that inferred the ice divide to have been situated in a more westerly position,
+that inferred the ice divide in a more westerly position,
 over the western margin of the Fraser Plateau\footnote{
@@
 \citep{Ryder.etal.1991,
-Stumpf.etal.2000, Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}. A
-latitudinal saddle has been inferred to have connected ice dispersal centres in
+Stumpf.etal.2000, Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}.
+These idincate that a latitudinal saddle connected ice dispersal centres in
 the Columbia Mountains with the main ice divide \citep{Ryder.etal.1991,
 Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}. A latitudinal saddle
-does indeed features in our modelling results, however, in an inverse
+does indeed feature in our modelling results, however, in an inverse
 configuration between the main ice divide over the Columbia Mountains and a
@@
 
-On the one hand, this could reflect the fact that our model does not include
+This could reflect the fact that our model does not include
 feedback mechanisms between ice sheet topography and the regional climate.
@@
 along the advancing margin \citep[cf.][]{Langen.etal.2012}. Because these two
-processes, both inhibiting of ice-sheet growth, are absent from our model, the
+processes, both with a tendency to limit of ice-sheet growth, are absent from
+our model, the
 eastern margin of the ice sheet and the position of the main meridional ice
@@
 
-On the other hand, field-based palaeoglaciological reconstructions have
+However, field-based palaeoglaciological reconstructions have
 struggled to reconcile the more westerly-centred ice divide in south-central
@@
 The modelled ice sheet configurations corresponding to ice volume maxima during
-MIS~4 are more variable (Fig.~\ref{fig:snapshots}, upper panels;
+MIS~4 are more sensitive to the choice of atmospheric forcing than those
+corresponding to ice volume maxima during MIS~2 (Fig.~\ref{fig:snapshots},
+upper panels;
 Fig.~\ref{fig:mis4}). The GRIP simulation (Fig.~\ref{fig:mis4}, left panel)
-results in a modelled maximum ice sheet that closely resemble that obtained
-during MIS~2, with the only major differences being a slightly less developed
-northern and eastern sectors. However, the EPICA simulations produce a reduced
+results in a modelled maximum ice sheet extent that closely resembles that
+obtained
+during MIS~2, with the only major difference being slightly less extensive
+northern and eastern sectors. In contrast, the EPICA simulation produces a lower
 ice volume maximum (Fig.~\ref{fig:timeseries}), which translates in the
 modelled ice sheet geometry into a significantly reduced southern sector, more
-restricted ice cover on the northern and eastern flanks, and generally lower
+restricted ice cover in northern and eastern sectors, and generally lower
 ice surface elevations in the interior (Fig.~\ref{fig:mis4}, left panel).
@@
 slightly more extensive than their MIS~2 counterparts. This is notably the case
-in the Puget Lowland, along parts of the marine margin, and parts of the north
-slope of the Alaska Range. However, these differences generally correspond to a
-few model grid cells. Considering the simplicity of the applied atmospheric
-forcing, their interpretation would be hazardous at the scale of our
-simulations.
+in the Puget Lowland and along parts of the ice margin in the marine sectors,
+on the north slope of the Alaska Range and in Yukon Territory. However, these
+differences
+are small and generally correspond to a few grid cells only. Although such
+results may be corroborated by field evidence in some places
+\citep[e.g.][]{Ward.etal.2007}, considering the simplicity of the applied
+climate forcing, modelled differences in extent are not demonstrably
+significant.
 
@@
 
-The accuracy of palaeo-glaciological reconstructions is generally biased
-towards the stages of maximum extent, for which the geological evidence is
-generally more detailed and more complete than for other periods. However,
+Palaeo-glaciological reconstructions are generally more robust
+for maximum ice sheet extents and late ice sheet configurations than for
+intermediate or minimum ice sheet extents and older ice sheet configurations
+\citep{Kleman.etal.2010}. However,
 these maximum stages are, by nature, extreme configurations, which do not
 necessarily represent the dominant patterns of glaciation throughout the period
-of ice cover.
+of ice cover \citep{Porter.1989, Kleman.Stroeven.1997}.
 
@@
 indicate that the LGM maximum extent was short-lived. To compare this finding
-to our simulations, we use numerical modelling output covering the whole
-simulation period to compute durations of ice cover throughout the last glacial
+to our simulations, we use numerical modelling output
+to compute durations of ice cover throughout the last glacial
 cycle (Fig.~\ref{fig:duration}).
@@
 the North American Cordillera (Fig.~\ref{fig:duration}, blue areas). A
-2500\,km-long continuous ice cover spanning from the Alaska Range in the
-northeast to the Rocky Mountains in the southwest only exists\renote{
-    I chosed to write about model output in the present form to discern in from
-    the reality.}
-for at most 29\,ka in total over the entire last glacial cycle
+2500\,km-long continuous expanse of ice, extending from the Alaska Range in the
+northeast to the Rocky Mountains in the southwest, is only in operation
+for at most 29\,ka, which is less than 25\,\% of a glacial cycle
 (Fig.~\ref{fig:duration}, hatched areas). However,
 except for its marine margin and the northern foothills of the Alaska Range,
-the maximum extent of the ice sheet is attained for an even more brief period
-of a few thousand years (Figs.~\ref{fig:duration}, red areas). This
+the maximum extent of the ice sheet is attained for an even shorter period
+of time of few thousand years (Figs.~\ref{fig:duration}, red areas). This
 result illustrates that the maximum extents of the modelled ice sheet during
-MIS~4 and 2 were both short-lived and therefore out of balance with
+MIS~4 and MIS~2 were both short-lived and therefore out of balance with
 contemporary climate.
 
-A notable exception to the transient character of the LGM Cordilleran ice sheet
+A notable exception to the transient character of other sectors of the
+Cordilleran ice sheet
 is the northern slope of the Alaska Range, where modelled glaciers are confined
-to the foothills during the entire simulation period (Fig.~\ref{fig:duration},
+to its foothills during the entire simulation period (Fig.~\ref{fig:duration},
 AR). This apparent insensitivity of modelled glacial extent to temperature
-fluctuations results from a combination of low precipitations, high summer
-temperatures and high temperature standard deviations in this sector of the
+fluctuations results from a combination regimes of low precipitation, high summer
+temperature and large temperature standard deviation for this sector of the
 modelling domain (Fig.~\ref{fig:atm}). This result could potentially explain
@@
 cover duration maps (Fig.~\ref{fig:duration}), are located over the Coast
-Mountains (CM), the Columbia and Rocky mountains (CR), the Skeena Mountains
-(SM), and the Selwyn and MacKenzie Mountains (SMK). Although the Coast,
-Columbia, Rocky and Skeena Mountains (CM, CR, SM) are covered by mountain
-glaciers for most of the model time, providing durable nucleation centre for
-the initiation of ice sheet growth, this is not the case for the Selwyn and
-MacKenzie mountains (SMK), where ice cover on the highest peaks is limited to a
+Mountains (CM), the Columbia and Rocky mountains (CRM), the Skeena Mountains
+(SM), and the Selwyn and MacKenzie Mountains (SMKM). Although the Coast,
+Columbia, Rocky and Skeena mountains (CM, CRM, SM) are covered by mountain
+glaciers for most of the last glacial cycle, providing durable nucleation
+centres for an ice sheet, this is not the case for the Selwyn and
+MacKenzie mountains (SMKM), where ice cover on the highest peaks is limited to a
 fraction of the last glacial cycle. In other words, the Selwyn and MacKenzie
@@
 during the coldest periods of the last glacial cycle. The Northern Rocky
-Mountains (Fig.~\ref{fig:duration}, NR), do not act as a nucleation centre,
-but rather as a pinning point for the ice margin coming from the west.
+Mountains (Fig.~\ref{fig:duration}, NRM), do not act as a nucleation centre,
+but rather as a pinning point for the Cordilleran ice sheet margin coming from
+the west.
 
@@
 MIS~3 residual ice bodies in northern and eastern Canada as nucleation centres
-for a much more extensive MIS~2 configuration \citep{Kleman.etal.2010}\renote{
+for a much more extensive MIS~2 configuration of the Laurentide ice sheet
+\citep{Kleman.etal.2010}\renote{
     Does this sentence really fit here? I feel that this sudden mention of the
@@
 The presence of a Skeena Mountains ice cap during most of the last glacial
-cycle can be explained by a currently more widespread\renote{
-    I mean widespread, precipitation there is lower than elsewhere.}
-winter precipitation for that region there than for other parts of the
+cycle can be explained by a currently more eastwards-spread distribution of
+winter precipitation there than along other transects of the
 modelling domain (Fig.~\ref{fig:atm}). Along most of the north-western coast of
@@
     describe here. I wrote ``moisture'' and this is what I mean.}
-further inland, until it is progressively captured by the extensive, but mildly
-elevated group of the Skeena Mountains in north-central British Columbia.
-
-\subsubsection{Potential imprint on the landscape}
+further inland, until it is captured by the extensive Skeena Mountains in
+north-central British Columbia, thus resulting in a more widespread
+distribution of winter precipitation.
+
+\subsubsection{Erosional imprint on the landscape}
 
@@
 {Stroeven.etal.2010}) is modelled to have hosted warm-based ice
-(Fig.~\ref{fig:warmfrac}) but it has only been glaciated for a short fraction
-of the last glacial cycle (Fig.~\ref{fig:duration}) according to our
-results. This perhaps indicate that\renote{
+(Fig.~\ref{fig:warmfrac}). However, because it has only been glaciated for a
+short fraction
+of the last glacial cycle, (Fig.~\ref{fig:duration}) according to our
+results, this perhaps appears indicate that\renote{
     I don't think that we have to infer that ``our model does not perform
@@
     quantitative one.}
-the observed landscape pattern originates from older glacial cycles and
+the observed landscape pattern originates from multiple glacial cycles and
 witnesses an increased relative importance of this ice dispersal centre prior
@@
 70\,\unit{mW\,m^{-2}} geothermal heat flux at 3\,km depth
-(Sect.~\ref{sec:overview}). However, both the Skeena Mountain area and the area
+(Sect.~\ref{sec:icedyn}). However, both the Skeena Mountains and the area
 west of the MacKenzie Mountains are areas of higher-than-average geothermal
-heat flux \citep{Blackwell.Richards.2004}. We can therefore expect a higher
-frequency of warm-based ice cover in this areas if we were to include
-spatially-variable geothermal forcing in our Cordilleran ice sheet simulations.
+heat flux \citep{Blackwell.Richards.2004}. We can therefore expect longer
+durations of warm-based ice cover for these areas if we were to include
+spatially variable geothermal forcing in our Cordilleran ice sheet simulations.
 
@@
 
-In the North American Cordillera alike other glaciated regions, the large
-majority of the glacial geologic record relates to the last few millennia of
-glaciation, most of the older evidence having been overridden by ice retreat
+Similarly to other glaciated regions, most glacial traces in the North American
+Cordillera relate to the last few millennia of glaciation, because most of the
+older evidence has been overprinted by warm-based ice retreat
 during the deglaciation \citep{Kleman.1994, Kleman.etal.2010}. From a
@@
 margin, increased flow velocities and longitudinal stress gradients, and poorly
-understood hydrological processes, typically included in the models through
-simple parametrisations \citep[e.g.][]{Clason.etal.2012, Clason.etal.2014,
-Bueler.Pelt.2014}, if included at all. However, after mapped moraines
-indicative of maximal extents (Sect.~\ref{sec:mis2} and~\ref{sec:mis4}),
-geomorphologically-reconstructed patterns of the last deglaciation provide the
-second best source of evidence for validation of our simulations.
+understood hydrological processes. The latter are typically included in the
+models through simple parametrisations \citep[e.g.][]{Clason.etal.2012,
+Clason.etal.2014, Bueler.Pelt.2014}, if included at all. However, next after a
+matching of modelled ice margin positions with mapped end moraines
+\citep{Napieralski.etal.2007}, indicative of maximum ice sheet extents
+(Sect.~\ref{sec:mis2} and~\ref{sec:mis4}), geomorphologically-based
+reconstructions of patterns of of ice sheet retreat during the last
+deglaciation provide the second best source of evidence for the validation of
+our simulations.
 
@@
     Is this a good choice of references?}
-indicate a rapid deglaciation. The southern and northern margins of the
+indicate that abundant meltwater was produced over the entire ice sheet surface
+during deglaciation. The southern and northern margins of the
 Cordilleran ice sheet reached their last glacial maximum extent around
 17\,ka \citep[Sect.~\ref{sec:timing};][]{Porter.Swanson.1998, Cosma.etal.2008,
-Stroeven.etal.2010, Stroeven.etal.2014}, putting an older bound to the
-onset of ice retreat. The timing of the late deglaciation is less well
+Stroeven.etal.2010, Stroeven.etal.2014}, which we take as a limiting age for
+the onset of ice retreat. The timing of final deglaciation is less well
 constrained, but recent cosmogenic dates from north-central British Columbia
@@
 In our simulations, the timing of peak ice volume during the last glacial
-maximum and the pacing of deglaciation depend cirticially on the choice of
-atmospheric forcing (Table~\ref{tab:extrema}, Figs.~\ref{fig:timeseries}
+maximum and the pacing of deglaciation depend critically on the choice of
+climate forcing (Table~\ref{tab:extrema}, Figs.~\ref{fig:timeseries}
 and~\ref{fig:deglacseries}). Adopting the EPICA
-atmospheric forcing yields peak ice volume at 17.3\,ka and an uninterupted
+climate forcing yields peak ice volume at 17.3\,ka and an uninterrupted
 deglaciation until 10.6\,ka (Fig.~\ref{fig:deglacseries}, lower panel, red
@@
 peak ice volume at 19.1\,ka and a deglaciation interrupted by two ice volume
-standstills until 9.4\,ka. The first interuption occurs between 16.6 and
+standstills until 9.4\,ka. The first interruption occurs between 16.6 and
 14.5\,ka, and the second between 12.6 and 11.6\,ka
@@
 drivers, differ in detail in that the EPICA depicts peak glaciation almost
-2\,ka later than the GRIP, in closer agreement to dated maximum extents, and
+2\,ka later than the GRIP, in closer agreement with dated maximum extents, and
 shows a faster, uninterrupted deglaciation which yields ice-free conditions
@@
 decay by retreating towards the Selwyn and Skeena mountains. The last remnants
-of the Cordilleran ice sheet finally disappear from the Skeena mountains around
+of the Cordilleran ice sheet finally disappear from the Skeena Mountains around
 10.6\,ka (EPICA) and 9.4\,ka (GRIP).
@@
 consecutive glacier maxima, or standstills while in overall retreat, one of
-which corresponding to the Younger Dryas chronozone \citep{Clague.etal.1997,
+which corresponds to the Younger Dryas chronozone \citep{Clague.etal.1997,
 Friele.Clague.2002, Friele.Clague.2002a, Kovanen.2002,
-Kovanen.Easterbrook.2002}. Although these moraines characterise independent
-valley glaciers, not necessarily connected to an ice sheet, the Finlay River
+Kovanen.Easterbrook.2002}. Although most of these moraines characterise
+independent valley glaciers, that were probably disconnected from the waning
+Cordilleran ice sheet, the Finlay River
 area of the northern Rocky Mountains presents a different kind of evidence.
@@
 local readvances, the GRIP-driven simulation shows that larger ice bodies
-emanating from the Skeena and MacKenzie mountains may still have exist at the
+emanating from the Skeena and MacKenzie mountains may still have existed at the
 time of this late-glacial readvance\renote{
@@
 Because a general tenant in glacial geomorphology is that the majority of
-landforms (lineations and eskers) are part of the deglacial envelop
-\citep[terminology from][]{Kleman.etal.2006}, that is that these were formed
+landforms (lineations and eskers) are part of the deglacial envelope
+\citep[terminology from][]{Kleman.etal.2006}, having been formed
 close inside the retreating margin of ice sheets \citep{Boulton.Clark.1990,
@@
 immediately preceding deglaciation or at the time of cessation of sliding
-inside the cold-base retreating margin (Fig.~\ref{fig:lastflow}). The
+inside a cold-based retreating margin (Fig.~\ref{fig:lastflow}). The
 modelled deglacial flow patterns appear consistent between the GRIP and EPICA
@@
 
-Patterns of glacial lineation formed in the northern and southern sectors of
+Patterns of glacial lineations formed in the northern and southern sectors of
 the Cordilleran ice sheet (\citealp{Prest.etal.1968};
@@
 (Fig.~\ref{fig:lastflow}). In the northern half
-of the modelling domain, modelled deglacial flow direction depict an
+of the modelling domain, modelled deglacial flow directions depict an
 active downhill flow as the last remnants of the ice sheet retreat towards
-elevated mountain ranges. Converging deglacial flow patterns in the Liard
-Lowland, for instance (Fig.~\ref{fig:lastflow}), closely resemble that
-indicated by the glacial lineations \citep[Fig.~2]{Margold.etal.2013}.
+mountain ranges. Converging deglacial flow patterns in the Liard
+Lowland, for instance (Fig.~\ref{fig:lastflow}), closely resemble the pattern
+indicated by glacial lineations \citep[Fig.~2]{Margold.etal.2013}.
 
@@
 simulations predict negligible basal sliding during deglaciation
-(Fig.~\ref{fig:lastflow}), but an impressive set of glacial delineations
+(Fig.~\ref{fig:lastflow}), but an impressive set of glacial lineations
 indicate a substantial eastward flow component of the Cordilleran ice sheet
@@
 \citealp{Seguinot.etal.2014}). A more westerly-located LGM ice divide would
-certainly result in a different deglacial flow patterns over the Interior
-Plateau. Another explanation for the incongruent results could be that the
-Fraser Plateau lineation system predates deglaciation, and that deglacial
-landforms are largely absent \footnote{
-    \textbf{Martin, Johan}, any southwards running eskers on the plateau?}.
+certainly result in a different deglacial flow pattern over the Interior
+Plateau. However, a more westerly-positioned LGM ice divide would certainly be
+associated with a thinner ice sheet than that modelled here. Decreased ice
+thickness would not promote warm-based conditions but, on the
+opposite, enlarge the region of negligible basal sliding
+(Fig.~\ref{fig:lastflow}). Thus, another explanation for the incongruent
+results could be that the Fraser Plateau lineation system predates
+deglaciation, and that deglacial landforms are largely absent. Alternatively,
+local geothermal heat associated with volcanic activity on the Fraser Plateau
+could have triggered the basal sliding \citep{Fahnestock.etal.2001}.
 
@@
 consists of a rapid northwards retreat (Fig.~\ref{fig:deglac}) of
-southwards-flowing (Fig.~\ref{fig:lastflow}) non-sliding ice lobes positioned
+southwards-flowing non-sliding ice lobes (Fig.~\ref{fig:lastflow}) positioned
 in-between deglaciated mountain ranges
@@
             & $6\times10^{-3}$
-            & \unit{K\,m{-1}}
+            & \unit{K\,m^{-1}}
             & - \\
@@
 \begin{table*}[t]
-  \caption{Extrema of ice volume and extent corresponding to MIS~4, 3 and 2 for
-           each of the low-resolution simulations (Fig.~\ref{fig:timeseries}).}
+  \caption{Extremes in Cordilleran ice sheet volume and extent corresponding to
+           MIS~4, 3 and 2 for each of the six low-resolution simulations
+           (Fig.~\ref{fig:timeseries}).}
   \label{tab:extrema}
@@
     \middlehline
-    GRIP     & -57.58 & -49.24 & -19.52
+    GRIP     &  57.58 &  49.24 &  19.52
              &   1.98 &   0.46 &   2.13
              &   7.52 &   0.89 &   8.52 \\
-    NGRIP    & -60.26 & -50.16 & -22.85
+    NGRIP    &  60.26 &  50.16 &  22.85
              &   2.16 &   0.50 &   2.09
              &   8.69 &   0.93 &   8.24 \\
-    EPICA    & -61.87 & -45.57 & -17.10
+    EPICA    &  61.87 &  45.57 &  17.10
              &   1.57 &   0.95 &   2.08
              &   5.20 &   2.44 &   8.35 \\
-    Vostok   & -60.87 & -49.68 & -16.87
+    Vostok   &  60.87 &  49.68 &  16.87
              &   1.55 &   0.86 &   2.14
              &   5.10 &   2.01 &   8.66 \\
-    ODP 1012 & -55.41 & -47.08 & -23.21
+    ODP 1012 &  55.41 &  47.08 &  23.21
              &   1.44 &   0.85 &   2.13
              &   4.50 &   2.06 &   8.46 \\
-    ODP 1020 & -60.16 & -52.24 & -29.46
+    ODP 1020 &  60.16 &  52.24 &  29.46
              &   1.32 &   0.70 &   2.08
@@
     \middlehline
-    Minimum  & -61.87 & -52.24 & -29.46
+    Minimum  &  61.87 &  52.24 &  29.46
              &   1.32 &   0.46 &   2.08
              &   3.88 &   0.89 &   8.24 \\
-    Maximum  & -55.41 & -45.57 & -16.87
+    Maximum  &  55.41 &  45.57 &  16.87
              &   2.16 &   0.95 &   2.14
@@
            standard deviation of daily mean temperature for January and July
-           months from the North American Regional Reanalysis (NARR)
-           climatology, used to force the ice sheet model. Note the
+           from the North American Regional Reanalysis
+           \citep[NARR;][]{Mesinger.etal.2006},
+           used to force the ice sheet model. Note the
            strong contrasts in seasonality, timing of the precipitation peak,
            and temperature variability over the model domain, notably between
-           the maritime and continental regions.}
+           coastal and inland regions.}
   \label{fig:atm}
@@
   \includegraphics{timeseries}
-  \caption{Temperature offset time-series from ice core and sediment core
-           records (Table~\ref{tab:records}) used as palaeo-climate forcing for
-           the ice sheet model (top panel), and modelled ice volume
-           through the last 120\,ka, expressed in meters of sea-level
-           equivalent (bottom panel). Gray spans indicate Marine Isotope
-           Stages (MIS) according to a global compilation of benthic
-           \chem{\delta^{18}O} records \citep{Lisiecki.Raymo.2005}. Hatched
-           rectangles highlight modelled ice volume extrema corresponding to
-           MIS~4 (61.9--55.4\,ka), MIS~3 (52.2--45.6\,ka), and
-           MIS~2 (last glacial maximum, 29.5--16.9\,ka). Dotted lines
-           correspond to the GRIP and EPICA 6\,km-resolution runs.}
+  \caption{Temperature offset time-series from ice core and ocean records
+           (Table~\ref{tab:records}) used as palaeo-climate forcing for the ice
+           sheet model (top panel), and modelled ice volume (bottom panel)
+           through the last 120\,ka. Ice volumes are expressed in meters of sea
+           level equivalent (m~s.l.e.). Gray fields indicate Marine Oxygen
+           Isotope Stage (MIS) boundaries for MIS~2 and MIS~4 according to a
+           global compilation of benthic \chem{\delta^{18}O} records
+           \citep{Lisiecki.Raymo.2005}. Hatched rectangles highlight the
+           time-volume space for ice volume extremes corresponding to MIS~4
+           (61.9--55.4\,ka), MIS~3 (52.2--45.6\,ka), and MIS~2 (LGM,
+           29.5--16.9\,ka). Dotted lines correspond to GRIP- and EPICA-driven
+           6\,km-resolution runs.}
   \label{fig:timeseries}
@@
   \caption{Snapshots of modelled surface topography (500\,m contours)
-           corresponding to the ice volume extrema indicated on
+           corresponding to the ice volume extremes indicated on
            Fig.~\ref{fig:timeseries}. Note the occurence of spatial similarities

### 2014-07-20 cordillera-cycle.tex
Rework conclusions.

@@
 local geothermal heat associated with volcanic activity on the Fraser Plateau
-could have triggered the basal sliding \citep{Fahnestock.etal.2001}.
+could have triggered the basal sliding
+\citep[cf. Greenland ice sheet,][]{Fahnestock.etal.2001}.
 
@@
 cycle presented in this study consistently produce two glacial maxima during
-MIS~4 (61.9--55.4\,ka) and MIS~2 (29.5--16.9\,ka), two periods
-corresponding to documented extensive glaciations. This result is
-independent of the choice of the palaeo-temperature record used to approximate
-the past climate evolution, and thus
-can be seen as a first-order agreement between the model and the geological
-evidence. However, the timing of the two glaciation peaks depends sensitively
-on which record
-is used to drive the model. The timing of the LGM is best
-reproduced by the Antarctic record, and occurs too early in all simulations
-that are driven by the other records. The mismatch is greatest when using
-oceanic records from the Pacific Northwest, which are affected by the
-weakening of the California current during the LGM.
-
-In all simulations presented here, ice cover is limited to disjoint mountain ice
-caps during most of the glacial cycle, confirming previous inferences from the
-geological evidence preceding the LGM. However, our
-simulations produce persistent ice cover on the Skeena Mountains during
-the entire glacial cycle. At the time when a full-size Cordilleran ice sheet is
-absent, the Skeena ice cap appears to be fed by the eastwards precipitation
-intrusion through a topographic window in the Coast Mountains. The ice cap acts
-as a nucleation centre at the onset of the LGM readvance, and appears
-consistent with the distinct glacial imprint of the Skeena Mountains landscape.
-
-During deglaciation, none of the palaeo-temperature records used produces a
-close agreement between the model results and the geological evidence. Although
-the EPICA record yields a more realistic timing of the LGM and early
-deglaciation, only the GRIP record produces a late-glacial readvance in areas
-where it has been documented in the literature. Nonetheless, the
-general patterns of deglaciation are consistent between both simulations driven
-by the GRIP and EPICA records, and show a rapid deglaciation of the southern
-half of the ice sheet, including a rapid northwards retreat across the Interior
-Plateau of central British Columbia. This is followed by an opening of the ice
-margin in the Liard Lowland, and a final retreat of the margin of the remaining
-ice caps towards the Selwyn and, later, the Skeena mountains, which host the
-last remnant of the ice sheet
-during the early Holocene (10.9--9.5\,ka).
-
-These results are strongly dependent on selected ice-sheet model (PISM),
-surface
-mass balance model (PDD) and climate forcing. Most importantly, our simplistic
-palaeo-climate forcing does not include precipitation corrections in response
-to the presence of an ice cover, potentially leading to overestimated glacial
-extent and volume in continental regions. Nevertheless, our results identify
-the Skeena Mountains as a key area to understanding
-glacial dynamics of the Cordilleran ice sheet, highlighting the need for
-further geological investigation of this region.
+MIS~4 (61.9--55.4\,ka, 3.8--8.7\,m~s.l.e) and MIS~2 (29.5--16.9\,ka,
+8.24--8.66\,m~s.l.e.), two periods corresponding to documented extensive
+glaciations. This result is independent of the palaeo-temperature record used
+among the six selected for this study, and thus can be seen as a robust model
+output, which broadly matches geological evidence. However, the timing of the
+two glaciation peaks depends sensitively on which climate record is used to
+drive the model. The timing of the LGM is best reproduced by the EPICA and
+Vostok Antarctic ice core records, and occurs too early in all other
+simulations. The mismatch is largest for the two northwest Pacific ODP
+palaeo-temperature records, which are affected by the weakening of the
+California current during the LGM.
+
+In all simulations presented here, ice cover is limited to disjoint mountain
+ice caps during most of the glacial cycle. The most persistent nucleation
+centres are located in the Coast Mountains, the Columbia and Rocky mountains,
+the Selwyn and MacKenzie mountains, and most importantly, in the Skeena
+Mountains. Throughout the modelled last glacial cycle, the Skeena Mountains
+host an ice cap which appears to be fed by moisture intruding inland from the
+west through a topographic low in the Coast Mountains. The Skeena ice cap acts
+as the main nucleation centre at the onset of the LGM readvance. As indicated
+by persistent, warm-based ice in the model, this ice cap is perhaps associated
+with the distinct glacial erosional imprint on the landscape of the Skeena
+Mountains.
+
+During deglaciation, none of the climate records used can be selected as
+producing an optimal agreement between the model results and the geological
+evidence. Although the EPICA-driven simulation yields the most realistic timing
+of the LGM and, therefore, start of deglaciation, only the GRIP-driven
+simulation produces late glacial readvances in areas where these have been
+documented. Nonetheless, the patterns of ice sheet retreat are consistent
+between the two simulations, and show a rapid deglaciation of the southern
+sector of the ice sheet, including a rapid northwards retreat across the Fraser
+Plateau of central British Columbia. The GRIP-driven simulation then produces a
+late-glacial readvance of local ice caps in the Coast, Columbia and Rocky
+mountains, and of the main body of the decaying Cordilleran ice sheet in the
+Skeena and Selwyn mountains. In both simulations, this is followed by an
+opening of the Liard Lowland, and a final retreat of the remaining ice caps
+towards the Selwyn and, finally, the Skeena mountains, which hosts the last
+remnant of the ice sheet during the early Holocene (10.9--9.5\,ka). Our results
+identify the Skeena Mountains as a key area to understanding glacial dynamics
+of the Cordilleran ice sheet, highlighting the need for further geological
+investigation of this region.
 

### 2014-07-20 cordillera-cycle.tex
Simplify parameter table

@@
   \centering
-  \caption{Main parameters of the ice sheet model.
-           \todo{At the moment, this is pretty extensive (does not fit on one
-                 page using the standard font size). There are probably a
-                 couple of things that are not needed here, like for instance
-                 the very standard things (gravity, gas constant, etc). Many
-                 parameters are not defined in the text. Do we need to
-                 introduce them, or should we skip them? I wonder if the last
-                 column is needed, in which case it needs to be completed.}}
+  \caption{Parameter values used in the ice sheet model.}
   \label{tab:params}
-  \footnotesize
-  {\begin{tabular}{llrll}
+  {\begin{tabular}{llrl}
     \tophline
-    Not.    & Name & Value & Unit & Source \\
+    Not.    & Name & Value & Unit \\
     \middlehline
-
-    \multicolumn{2}{l}{\emph{Ice rheology}} \\
 
@@
             & 910
-            & \unit{kg\,m^{-3}}
-            & \citet{Aschwanden.etal.2012} \\
+            & \unit{kg\,m^{-3}} \\
 
@@
             & 9.81
-            & \unit{m\,s^{-2}}
-            & \citet{Aschwanden.etal.2012} \\
-
-    $n$     & Glen exponent
-            & 3
-            & -
-            & \aref \\
-
-    $A_c$   & Ice hardness coefficient cold
-            & $3.61\times10^{-13}$
-            & \unit{Pa^{-3}\,s^{-1}}
-            & \citet{Paterson.Budd.1982} \\
-
-    $A_w$   & Ice hardness coefficient warm
-            & $1.73\times10^3$
-            & \unit{Pa^{-3}\,s^{-1}}
-            & \citet{Paterson.Budd.1982} \\
-
-    $Q_c$   & Flow law activation energy cold
-            & $6.0\times10^4$
-            & \unit{J\,mol^{-1}}
-            & \citet{Paterson.Budd.1982} \\
-
-    $Q_w$   & Flow law activation energy warm
-            & $13.9\times10^4$
-            & \unit{J\,mol^{-1}}
-            & \citet{Paterson.Budd.1982} \\
-
-    $R$     & Ideal gas constant
-            & 8.31441
-            & \unit{J\,mol^{-1}\,K^{-1}}
-            & \aref \\
-
-    $T_c$   & Flow law critical temperature
-            & 263.15
-            & \unit{K}
-            & \citet{Paterson.Budd.1982} \\
-
-    $f$     & Flow law water fraction coeff.
-            & 181.25
-            & -
-            & \citet{Lliboutry.Duval.1985} \\
-
-    $\beta$ & Clapeyron constant
-            & $7.9\times10^{-8}$
-            & \unit{K\,Pa^{-1}}
-            & \citet{Luthi.etal.2002} \\
-
-    $c_i$   & Ice specific heat capacity
-            & 2009
-            & \unit{J\,kg^{-1}\,K^{-1}}
-            & \citet{Aschwanden.etal.2012} \\
-
-    $c_w$   & Water specific heat capacity
-            & 4170
-            & \unit{J\,kg^{-1}\,K^{-1}}
-            & \citet{Aschwanden.etal.2012} \\
-
-    $k$     & Ice thermal conductivity
-            & 2.10
-            & \unit{J\,m^{-1}\,K^{-1}\,s^{-1}}
-            & \citet{Aschwanden.etal.2012} \\
-
-    $L$     & Water latent heat of fusion
-            & $3.34\times10^5$
-            & \unit{J\,kg^{-1}\,K^{-1}}
-            & \citet{Aschwanden.etal.2012} \\
+            & \unit{m\,s^{-2}} \\
 
@@
             & 0.25
-            & -
             & - \\
@@
             & 100.0
-            & \unit{m\,yr^{-1}}
-            & - \\
+            & \unit{m\,yr^{-1}} \\
 
@@
             & 0.0
-            & Pa
-            & - \\
+            & Pa \\
 
@@
             & 0.02
-            & -
             & - \\
@@
             & 0.69
-            & -
-            & \citet{Tulaczyk.etal.2000} \\
+            & - \\
 
@@
             & 0.12
-            & -
-            & \citet{Tulaczyk.etal.2000} \\
+            & - \\
 
@@
             & 2.0
-            & m
-            & - \\
+            & m \\
 
@@
             & 3300
-            & \unit{kg\,m^{-3}}
-            & \aref \\
+            & \unit{kg\,m^{-3}} \\
 
@@
             & 1000
-            & \unit{J\,kg^{-1}\,K^{-1}}
-            & \citet{Ritz.1997} \\
+            & \unit{J\,kg^{-1}\,K^{-1}} \\
 
@@
             & 3.0
-            & \unit{J\,m^{-1}\,K^{-1}\,s^{-1}}
-            & \citet{Ritz.1997} \\
+            & \unit{J\,m^{-1}\,K^{-1}\,s^{-1}} \\
 
@@
             & $1\times10^21$
-            & \unit{Pa\,s}
-            & \citet{Lingle.Clark.1985} \\
+            & \unit{Pa\,s} \\
 
@@
             & 3300
-            & \unit{kg\,m^{-3}}
-            & \citet{Lingle.Clark.1985} \\
+            & \unit{kg\,m^{-3}} \\
 
@@
             & $5.0\times10^24$
-            & \unit{N}
-            & \citet{Lingle.Clark.1985} \\
+            & \unit{N} \\
 
@@
             & 273.15
-            & \unit{K}
-            & \aref \\
+            & \unit{K} \\
 
@@
             & 275.15
-            & \unit{K}
-            & \aref \\
+            & \unit{K} \\
 
@@
             & $3.04\times10^{-3}$
-            & \unit{m\,K^{-1}\,day^{-1}}
-            & \citet{Shea.etal.2009} \\
+            & \unit{m\,K^{-1}\,day^{-1}} \\
 
@@
             & $4.59\times10^{-3}$
-            & \unit{m\,K^{-1}\,day^{-1}}
-            & \citet{Shea.etal.2009} \\
+            & \unit{m\,K^{-1}\,day^{-1}} \\
 
@@
             & $6\times10^{-3}$
-            & \unit{K\,m^{-1}}
-            & - \\
+            & \unit{K\,m^{-1}} \\
 

### 2014-07-20 cordillera-cycle.tex
Various long-due figure improvements.

@@
            by the ice sheet include the Alaska Range (AR), the Wrangell and
-           St.-Elias mountains (WSE), the Selwyn and MacKenzie mountains (SMK),
-           the Skeena Mountains (SM), the Coast Mountains (CM), the Rocky
-           Mountains (RM) and the North Cascades (NC). Major depressions
+           St.-Elias mountains (WSEM), the Selwyn and MacKenzie mountains (SMKM),
+           the Skeena Mountains (SM), the Coast and Rocky
+           Mountains (CRM) and the North Cascades (NC). Major depressions
            include the Liard Lowland (LL), the Fraser Plateau (FP) and the
@@
            map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
-           Data \citep{Patterson.Kelso.2014}.
-           \todo{Mark palaeo-ice sheets, mountain ranges and depressions on
-                 the map. Dash boundaries between ice sheets?}}
+           Data \citep{Patterson.Kelso.2014}.}
   \label{fig:locmap}
@@
            corresponding to the ice volume extremes indicated on
-           Fig.~\ref{fig:timeseries}. Note the occurence of spatial similarities
-           despite large differences in timing.
-           \todo{Perhaps indicate the location of the Skeena Mountains.}}
+           Fig.~\ref{fig:timeseries}. An ice cap persists over the Skeena
+           Mountains (SM) during MIS~3. Note the occurence of spatial
+           similarities despite large differences in timing.}
   \label{fig:snapshots}
@@
            from the Alaska Range (AR) to the southern Coast Mountains (CM) and
-           Rocky Mountains (RM) exists for about 29\,ka in both
-           simulations. A central
-           ice cover persists over the Skeena Mountains (SM) during most of the
-           simulation. On the other hand, the maximal extent of the ice sheet
-           generally corresponds to relatively short durations of ice cover.
-           \todo{Add WSE, SMK, NR, NC mountain ranges, and change kyr to ka}}
+           the Columbia and Rocky mountains (RM) exists for about 29\,ka in
+           both simulations. The maximal extent of the ice sheet generally
+           corresponds to relatively short durations of ice cover, but ice
+           cover persists over the Skeena Mountains (SM) during most of the
+           simulation.}
   \label{fig:duration}
@@
            glacial erosional imprint of the Skeena Mountains (SM) landscape.
-           Hatches indicate areas that were covered by cold ice only.
-           \todo{indicate location of the Skeena Mountains.}}
+           Hatches indicate areas that were covered by cold ice only.}
   \label{fig:warmbase}
@@
            continental shelf and major glacial troughs of the coastal ranges.
-           Hatches indicate areas that were covered by cold ice only.
-           \todo{indicate location of the Skeena Mountains and the MacKenzie
-                 Mountains.}}
+           Hatches indicate areas that were covered by cold ice only.}
   \label{fig:warmfrac}
@@
            from the GRIP (top panels) and EPICA (bottom panels) simulations.
-           \todo{Add the location of profiles used in
-                 Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-grip}.}}
+           Dashed segments indicate the location of profiles used in
+           Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}.}
   \label{fig:deglacshots}
@@
            and the decaying ice sheet between 14 and 10\,ka., which is more
-           pronounced in the GRIP-driven simulation.}
+           pronounced in the GRIP-driven simulation.
+           Dashed segments indicate the location of profiles used in
+           Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}.}
   \label{fig:deglac}
@@
   \includegraphics{lastflow}
-  \caption{Modelled directions of the deglacial basal flow velocities. Hatches
+  \caption{Modelled of the deglacial basal ice velocities. Hatches
            indicate areas that remain non-sliding throughout deglaciation
-           (22.0--8.0\,ka), notably including the Fraser Plateau.
+           (22.0--8.0\,ka), notably including the Fraser Plateau (FP).
+           Note the concentric patterns of deglacial flow in the Liard
+           Lowland (LL).
            Sliding grid cells were distinguished from non-sliding grid cells
            using a velocity threshold of 1\,\unit{m\,yr^{-1}}.
-           \todo{indicate the Fraser Plateau and the Liard Lowland. Add the
-                 location of profiles used in Figs.~\ref{fig:profiles-grip}
-                 and~\ref{fig:profiles-grip}.}}
+           Dashed segments indicate the location of profiles used in
+           Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}.}
   \label{fig:lastflow}
@@
            simulation, corresponding to the four transects indicated in
-           Fig.~\ref{fig:deglac}.
-           \todo{indicate mountain ranges and depressions.}}
+           Figs.~\ref{fig:deglacshots}--\ref{fig:lastflow}.}
   \label{fig:profiles-grip}
@@
            simulation, corresponding to the four transects indicated in
-           Fig.~\ref{fig:deglac}.
-           \todo{indicate mountain ranges and depressions.}}
+           Figs.~\ref{fig:deglacshots}--\ref{fig:lastflow}.}
   \label{fig:profiles-epica}

### 2014-07-20 figures/duration.py

@@ -46,7 +46,7 @@ for i, rec in enumerate(records):
     cf = ax.contourf(x[:], y[:], icecover, levels=levs, alpha=0.75,
                      colors=colors, hatches=hatches)
     cs = ax.contour(x[:], y[:], icecover, [cislevs[i]], colors='k', linewidths=0.25)
-    cs.clabel(fontsize=6, fmt='%i kyr', manual=[(-1825e3, 1000e3)])
+    cs.clabel(fontsize=6, fmt='%i ka', manual=[(-1825e3, 1000e3)])
     ax.contour(x[:], y[:], icecover, [levs[0]], colors='k', linewidths=0.5)
 
     # close extra file
@@ -59,8 +59,12 @@ txtkwa = dict(ha='center', va='center',
               arrowprops=dict(arrowstyle="->"))
 ax.annotate('AR', xy=(-2300e3, 2600e3), xytext=(-2000e3, 2600e3), **txtkwa)
 ax.annotate('SM', xy=(-2000e3, 1450e3), xytext=(-2350e3, 1450e3), **txtkwa)
-ax.annotate('CM', xy=(-2000e3,  650e3), xytext=(-2350e3,  650e3), **txtkwa)
-ax.annotate('RM', xy=(-1500e3,  650e3), xytext=(-1200e3,  650e3), **txtkwa)
+ax.annotate('CM', xy=(-1950e3,  700e3), xytext=(-2350e3,  700e3), **txtkwa)
+ax.annotate('NC', xy=(-1900e3, 250e3), xytext=(-2350e3, 250e3), **txtkwa)
+ax.annotate('WSEM', xy=(-2200e3, 2150e3), xytext=(-1200e3, 2150e3), **txtkwa)
+ax.annotate('SMKM', xy=(-1550e3, 1900e3), xytext=(-1200e3, 1900e3), **txtkwa)
+ax.annotate('NR', xy=(-1600e3, 1450e3), xytext=(-1200e3, 1450e3), **txtkwa)
+ax.annotate('CRM', xy=(-1550e3,  650e3), xytext=(-1200e3,  650e3), **txtkwa)
 
 # add colorbar and save
 cb = fig.colorbar(cf, ax.cax)

### 2014-07-20 figures/lastflow.py

@@ -81,8 +81,20 @@ for i, rec in enumerate(records):
     ax.contour(x[:], y[:], glaciated, levels=[0.5],
                colors='k', linewidths=0.5)
 
+    # add profile lines
+    for yp in [1.7e6, 1.4e6, 1.1e6, 0.8e6]:
+        ax.plot([-2.4e6, -1.25e6], [yp, yp], 'k|',
+                         lw=0.25, ls='--', dashes=(2, 2))
+
     # annotate
     annotate(ax, rec.upper())
+
+# locate Liard Lowland and Fraser Plateau
+txtkwa = dict(ha='center', va='center',
+              bbox=dict(ec='k', fc='w', alpha=1.0),
+              arrowprops=dict(arrowstyle="->"))
+ax.annotate('LL', xy=(-1700e3, 1600e3), xytext=(-1100e3, 1600e3), **txtkwa)
+ax.annotate('FP', xy=(-1850e3, 900e3), xytext=(-1100e3, 900e3), **txtkwa)
 
 # add colorbar and save
 print 'saving...'

### 2014-07-20 figures/snapshots.py

@@ -71,6 +71,12 @@ for j in range(3):
             ha='right', va='center', rotation='vertical',
             transform=ax.transAxes)
 
+# mark location of the skeena mountains
+grid[8].annotate('SM', xy=(-2000e3, 1450e3), xytext=(-1100e3, 1450e3),
+                       ha='center', va='center',
+                       bbox=dict(ec='k', fc='w', alpha=1.0),
+                       arrowprops=dict(arrowstyle="->"))
+
 # add colorbar and save
 cb = fig.colorbar(cs, ax.cax, ticks=levs[::2],
                   format=FuncFormatter(lambda x, pos: '%g' % (x/1000.0)))

### 2014-07-20 figures/warmbase.py

@@ -56,6 +56,13 @@ for i, rec in enumerate(records):
     annotate(ax, rec.upper())
     nc.close()
 
+# locate Skeena Mountains
+txtkwa = dict(ha='center', va='center',
+              bbox=dict(ec='k', fc='w', alpha=1.0),
+              arrowprops=dict(arrowstyle="->"))
+ax.annotate('SM', xy=(-2000e3, 1450e3), xytext=(-2350e3, 1450e3), **txtkwa)
+ax.annotate('MKM', xy=(-1550e3, 2000e3), xytext=(-1200e3, 2000e3), **txtkwa)
+
 # add colorbar and save
 cb = fig.colorbar(cs, ax.cax, ticks=levs[1:])
 cb.set_label('Duration of warm-based ice cover (kyr)')

### 2014-07-20 figures/warmfrac.py

@@ -56,6 +56,13 @@ for i, rec in enumerate(records):
     annotate(ax, rec.upper())
     nc.close()
 
+# locate Skeena Mountains
+txtkwa = dict(ha='center', va='center',
+              bbox=dict(ec='k', fc='w', alpha=1.0),
+              arrowprops=dict(arrowstyle="->"))
+ax.annotate('SM', xy=(-2000e3, 1450e3), xytext=(-2350e3, 1450e3), **txtkwa)
+ax.annotate('MKM', xy=(-1550e3, 2000e3), xytext=(-1200e3, 2000e3), **txtkwa)
+
 # add colorbar and save
 cb = fig.colorbar(cs, ax.cax, ticks=levs[1:])
 cb.set_label('Fraction of warm-based ice cover')

### 2014-07-20 cordillera-cycle.tex
Change author list and add contributions.

@@
 \Author[2]{I.}{Rogozhina}
+\Author[1]{A.~P.}{Stroeven}
 \Author[3]{M.}{Margold}
 \Author[1]{J.}{Kleman}
-\Author[1]{A.~P.}{Stroeven}
 \runningauthor{J.~Seguinot et~al.}
@@
 \section*{Author contributions}
-\dots
+J.~Seguinot ran the simulations and wrote most of the manuscript; I.~Rogozhina
+guided experiment design; A.~P.~Stroeven, M.~Margold and J.~Kleman took part in
+the interpretation and comparison of model results against geological evidence.
+All co-authors contributed to improve the text.
 

### 2014-07-20 cordillera-cycle.tex
Remove comment replies etc.

@@
 % My commands
-\newcommand{\renote}[1]{\footnote{\textbf{Comment reply}: #1}}
-\newcommand{\todo}[1]{\emph{[\textbf{Todo:} #1]}}
-\newcommand{\aref}[0]{\textbf{[ref.]}}
+%\newcommand{\renote}[1]{\footnote{\textbf{Comment reply}: #1}}
+%\newcommand{\todo}[1]{\emph{[\textbf{Todo:} #1]}}
+%\newcommand{\aref}[0]{\textbf{[ref.]}}
 
@@
 Cordillera have been more extensive than today. At the Last Glacial
-Maximum (LGM\renote{
-    I decided to adopt ``LGM'', which is very common, but not ``CIS'', for the
-    sake of readability. This corresponds to abbreviations used in
-    \citet{Seguinot.etal.2014} (and now, in the kappa).}),
-a continuous blanket of ice, the Cordilleran ice sheet\renote{
-    In \citet{Seguinot.etal.2014}, The Cryosphere changed all my occurences of
-    Cordilleran (Greenland etc) \textbf{I}ce \textbf{S}heet to lowercase
-    \textbf{i}ce \textbf{s}heet. Let's keep it so for now for consistency
-    within the thesis (kappa + papers).}
+Maximum (LGM),
+a continuous blanket of ice, the Cordilleran ice sheet
 \citep{Dawson.1888}, extended from the Alaska Range in the north to the
@@
 climate forcing. It computes the evolution of ice extent
-and thickness over time, the thermal and dynamic\renote{
-    Thermal = temperature, dynamic = force + movement. If I meant movement
-    alone, I would use ``cinematic''. Here we could use ``ice temperatures,
-    stress and deformation (i.e. velocities)'', but that becomes a bit clumsy,
-    I feel.}
+and thickness over time, the thermal and dynamic
 states of the ice sheet, and the associated lithospheric response.
@@
 dataset with a~resolution of 1\,arc-min \citep{Amante.Eakins.2009}. Sea level
-is lowered as a function of time based on the Spectral Mapping Project\renote{
-    I had no time to check for later products, but if you have suggestions for
-    something more up-to-date, I can try to include it in later simulations
-    (everything will have to be re-run anyway).}
+is lowered as a function of time based on the Spectral Mapping Project
 \citep[SPECMAP;][]{Imbrie.etal.1989} time scale. Geothermal heat flux
@@
 two approximation
-\citep[Eqns.~7--9 and 15]{Bueler.Brown.2009, Winkelmann.etal.2011}\renote{
-    Arjen, I don't agree with your comment. Reference to equation number(s)
-    show the reader where to look. \citet{Winkelmann.etal.2011} is a rather
-    long paper that also treats many other aspects of the model. Their SIA+SSA
-    approach builds on \citep{Bueler.Brown.2009}, but it is slightly different.
-    It is the one in use in PISM, and in my simulations.}.
+\citep[Eqns.~7--9 and 15]{Bueler.Brown.2009, Winkelmann.etal.2011}.
 Ice rheology depends on temperature and water content through an enthalpy
@@
 condition to the ice enthalpy model. Temperature is computed subglacially to
-a~depth of 3\,km\renote{
-    Yes, I changed this from 1 to 3\,km as we now model longer period of times
-    than in the previous paper. The penetration depth of temperature anomalies
-    (i.e. skin depth) depends on the period of temperature oscillations at the
-    surface. The longer the period (in our case 100\,ka), the deeper the
-    temperature change. A good way to estimate what this parameter should be is
-    to solve the 1-D temperature diffusion equation in the case of periodic
-    surface forcing, using rock heat capacity and rock thermal conductivity
-    values similar as in the numerical model. This basically forms the basis
-    for my choice of the 3\,km value.},
+a~depth of 3\,km,
 where it is conditioned by a lower boundary geothermal heat flux of
@@
 threshold above which instantaneous drainage is assumed.
-Finally, the bedrock topography responds to ice load\renote{
-    I disagree with writing ``the isostatic response''. The main point in
-    using this model is that computes a non-isostatic response.}
+Finally, the bedrock topography responds to ice load
 following a bedrock deformation model that includes local isostasy,
 elastic lithosphere flexure and viscous mantle deformation in the lower
-half-space\renote{
-    I had put an emphasis on semi-infinite as this is, from a mathematical
-    point of view, the key assumption of that model, allowing effective
-    solution in Fourier domain. This is the whole point of Ed's paper and
-    probably the reason why he picked it for PISM. But yes, I guess that
-    ``lower half-space'' is fine, too...}
+half-space
 \citep[Table~\ref{tab:params};][]{Lingle.Clark.1985,Bueler.etal.2007}.
@@
 air temperature, $\sigma$, and monthly precipitation, $P_m$, using
-a~temperature-index model \citep[e.g.,][]{Hock.2003}\renote{
-    Surface mass balance is the balance of mass fluxes at the surface,
-    consisting of ice accumulation and ice ablation, which I introduce straight
-    away. I don't see what kind of ``opening sentence'' can be added here.}.
+a~temperature-index model \citep[e.g.,][]{Hock.2003}.
 Accumulation is equal to
@@
 values from the NARR, after excluding variability associated
-with the seasonal cycle itself \citep[cf.][]{Seguinot.Rogozhina.2014}\renote{
-    I may add a figure in the kappa if I find time, but I don't think that we
-    should do it here. The subject is too marginal for the present study.}.
+with the seasonal cycle itself \citep[cf.][]{Seguinot.Rogozhina.2014}.
 Degree-day factors for snow and ice melt are derived from
@@
 
-Climate forcing of the model consists of a present-day monthly climatology%
-\renote{
-    ``Climatology'' is a common term in meteorology, used to design a set
-    of variables aggregated over several decades, and assumed to represent
-    some climate state.},
+Climate forcing of the model consists of a present-day monthly climatology,
 $\{T_{m0}, P_{m0}\}$, where temperatures are modified by offset time series,
@@
 
-thus accounting for the evolution of ice thickness, ${h=s-b}$\renote{
-    Small change in notation to match with the kappa},
+thus accounting for the evolution of ice thickness, ${h=s-b}$,
 on the one hand, and for differences between the basal topography of the ice
@@
 % -----------------------------------------------------------------------------
-
-I AM HERE.
 
@@
 choice of proxy record used for time-dependent climate forcing
-(Table~\ref{tab:extrema})\renote{
-    This new table contains modelled ages, volumes and extents at MIS~4--2}.
-
-Simulations forced by the Greenland ice core palaeo-temperature\renote{
-    From here on, I prepend ``palaeo-temperature'' to ``record(s)'' in place of
-    ``climate'' as you (Arjen) originally suggested. If we are to precise what
-    type of ``records'' we use, let's be clear about it.}
+(Table~\ref{tab:extrema}).
+
+Simulations forced by the Greenland ice core palaeo-temperature
 records (GRIP, NGRIP) produce the highest variability in modelled ice volume
@@
 (ODP~1012, ODP~1020) and Antarctic (EPICA, Vostok) palaeo-temperature records
-generally result in smaller\renote{
-    Until we run out of other ajectives, I don't think that ``modest'' and
-    ``generous'' are necessary terms for describing ice volumes.}
+generally result in smaller
 modelled ice volumes during MIS~4 and larger ice volumes during MIS~3.
@@
 
-\emph{[\textbf{About the new titles}: In my view, this
-    entire ``discussion'' part is all about evaluation of model runs agains
-    field evidence and palaeoglaciology. It is necessary that this comparison
-    starts here, in order to ``kick out'' the records that perform badly (ODP*,
-    NGRIP) and justify that the rest of the analysis is performed on two
-    (high-resolution) simulations only. Such figures as
-    Figs.~\ref{fig:duration}--\ref{fig:warmfrac} and Fig.~\ref{fig:deglac},
-    which I consider as different views of the results rather than new results
-    (apart the fact that they use higher resolution) are drawn for the purpose
-    of comparing model results against geological evidence. I propose to rename
-    this section to a more explicit ``Comparison to the geologic record'' (or
-    similar). I have also renamed the ``results'' to ``Sensitivity to
-    climate forcing time-series'', which is actually what that section is
-    about: raw model output, no discussion against the geologic record. This
-    new discussion is divided in three parts: 4.1 Glacial maxima, where we
-    study the maximum stages, relatively well-documented by the geology; 4.2
-    Nucleation centres, where we study ``intermediate'' states, for which less
-    is known, but the model can say something; and 4.3 The last deglaciation,
-    which deserves a separate section because many people are interested in
-    this period, and much geological evidence is available.]}
-
 Large variations in the model response reveal its sensitivity to evolving
@@
 
-Independently of the palaeo-temperature records\renote{
-    In my view, ``climate forcing'' would be far too broad here. We
-    could possibly apply a much wilder choice of climate forcing
-    techniques, leading just as wild differences in model output
-    \citep[e.g.,][but possibly much more than that]{Seguinot.etal.2014}.}
+Independently of the palaeo-temperature records
 used to force the ice sheet model, our simulations consistently produce two
@@
 current \citep{Herbert.etal.2001}. The California current, driving cold
-waters southwards along the south-western coast of North America\renote{
-    Not only USA but also Baja California in Mexico.},
+waters southwards along the south-western coast of North America,
 has been shown to have weakened during each peak global glaciation (in SPECMAP)
@@
 that inferred the ice divide in a more westerly position,
-over the western margin of the Fraser Plateau\footnote{
-    Just checked that this name is correct. Feels lighter than the ``Interior
-    Plateau of south-central British Columbia'' (although having a slightly
-    different meaning).}
+over the western margin of the Fraser Plateau
 \citep{Ryder.etal.1991,
@@
 for a much more extensive MIS~2 configuration of the Laurentide ice sheet
-\citep{Kleman.etal.2010}\renote{
-    Does this sentence really fit here? I feel that this sudden mention of the
-    Laurentide ice sheet perturbates a bit the flow. Moreover, knowing that
-    the Laurentide ice sheet is probably the main driver for the 100\,ka
-    sea-level cycle, it would have been surprising if it had completely
-    vanished at MIS~3. But for the Cordilleran ice sheet, usually thought of
-    as fast-responding and intermittent, I feel that the presence of residual
-    ice is less trivial.}.
+\citep{Kleman.etal.2010}.
 
@@
 centre of our modelling domain, this barrier is less pronounced than elsewhere,
-allowing westerly winds to carry moisture\renote{
-    I do not agree with ``carry precipitation''. When reading this I get the
-    picture of water drops falling diagonaly under action of the wind. I do not
-    deny that transport of precipitation happens and
-    is an important process for understanding orographic
-    precipitation (think of why it is raining so much in the Swedish mountains
-    even when the wind comes from the west), but it is not what I am trying to
-    describe here. I wrote ``moisture'' and this is what I mean.}
+allowing westerly winds to carry moisture
 further inland, until it is captured by the extensive Skeena Mountains in
@@
 (Figs.~\ref{fig:duration} and \ref{fig:warmbase};
-\citealp[Fig.~2]{Kleman.etal.2010})\renote{
-    Yes, I agree that it would be good to eventually reproduce here the basal
-    flow patterns from Johan's map \citep[Fig.~2]{Kleman.etal.2010}, but I
-    don't think that I will have the time for this before thesis submission}.
+\citealp[Fig.~2]{Kleman.etal.2010}).
 The Skeena Mountains, for example, indeed bear a strong glacial imprint that
@@
 of the last glacial cycle, (Fig.~\ref{fig:duration}) according to our
-results, this perhaps appears indicate that\renote{
-    I don't think that we have to infer that ``our model does not perform
-    reliably'' here. I am actually quite happy to discover that the model
-    puts warm-based ice where there is (another) region of strong glacial
-    erosion. I would rather say that computing the ``duration of
-    warm-based ice cover'' as I did for is a very simple approach and that we
-    can expect it to give a qualitative proxy for glacial erosion but not a
-    quantitative one.}
+results, this perhaps appears indicate that
 the observed landscape pattern originates from multiple glacial cycles and
@@
 \citep{Burke.etal.2012, Burke.etal.2012a, Perkins.etal.2013,
-Margold.etal.2013}\footnote{
-    Is this a good choice of references?}
+Margold.etal.2013}
 indicate that abundant meltwater was produced over the entire ice sheet surface
@@
 emanating from the Skeena and MacKenzie mountains may still have existed at the
-time of this late-glacial readvance\renote{
-    Thanks Arjen for your suggestion to reformulate. Good point.}.
+time of this late-glacial readvance.
 In contrast, the EPICA-driven simulation produces a nearly-continuous

### 2014-07-20 kappa.tex
Add abstract.

@@
 % ======================================================================
+
+\maketitle
+
+\begin{abstract}
+
+This doctoral thesis presents a study of the late glacial history of
+north-western North America using numerical ice sheet modelling calibrated by
+geological evidence.
+
+The North Americam Cordillera is an immense maze of mountain ranges and
+intermontane basins which once hosted an ice mass of the size of Greenland: the
+Cordilleran ice sheet. Although less extensive than the former Laurentide and
+Fennoscandian ice sheets, the Cordilleran ice sheet perhaps remains the least
+understood Pleistocene ice sheet in terms of its extent, volume, and dynamics.
+Because of the irregular topography of the ice sheet's subglacial landscape,
+ice flowed along tortuous pathways, sometimes following the major topographical
+troughs, and sometimes flowing across them.
+
+This thesis presents simulations performed with PISM, a numerical ice sheet
+model that approximates glacier flow by basal sliding and internal deformation.
+These simulations show that the somewhat paradoxical geometry of the last
+glacial maximum Cordilleran ice sheet can be largely explained by the present
+distribution of temperature and precipitation in the region, provided an
+accurate treatment of seasonality, orographic precipitation, and daily
+temperature variability in the ice sheet model. Nevertheless, at the scale of
+the last glacial cycle, this configuration appears ephemeral. In contrast, the
+average stage of glaciation consists of isolated glaciers and ice caps covering
+the major mountain ranges of the North American Cordillera.
+
+The simulations presented in this thesis demonstrate the potential of combining
+numerical ice sheet modelling to analysis of geological evidence in order to
+improve our understanding former glaciations.
+
+\end{abstract}
+
+\newpage
 

### 2014-07-20 kappa.tex
Add corrections from Arjen.

@@
 
-The North Americam Cordillera is an immense maze of mountain ranges and
+The North American Cordillera is an immense maze of mountain ranges and
 intermontane basins which once hosted an ice mass of the size of Greenland: the
@@
 numerical ice sheet modelling to analysis of geological evidence in order to
-improve our understanding former glaciations.
+improve our understanding of the former glaciations.
 
@@
 incomplete. Palaeo-ice sheets did not leave a continuous imprint on the
-lanscape, and much of this evidence has been overridden by subsequent glacier
+lanscape, and much of this evidence has been overprinted by subsequent glacier
 re-advances and other geomorphological processes \citep{Kleman.1994,
@@
 four-dimensional distribution of stresses and deformations in palaeo-ice sheets
-and allow for reconstructions their past thickness and volume evolution, which
+and allow for reconstructions of their past thickness and volume evolution, which
 are difficult to infer from geological evidence\renote{
@@
 western ribbon is formed by the Pacific Coast orogenic belt and includes, from
-North to South, the Alaska Range (\cref{fig:photo-alaska-range}), the Wrangell
+north to south, the Alaska Range (\cref{fig:photo-alaska-range}), the Wrangell
 and Saint-Elias Mountains,
@@
 linear Rocky Mountain / Tintina Trench depression, and includes the Selwyn and
-MacKenzie Mountains in the north, and the Rocky mountains in the South. The
+MacKenzie Mountains in the north, and the Rocky mountains in the south. The
 Brooks Range of northern Alaska has been covered by ice, but its glaciers
@@
 $2.64^{+0.20}/_{-0.18}$\,Ma indicates the first presence of an ice sheet in the
-Cordilleran mountains during the early Pleistocene \citep{Hidy.etal.2013}.
-
+Cordilleran mountains during the late Pliocene \citep{Hidy.etal.2013}.
 This indicates that the North American Cordillera has been glaciated many times
@@
 
-Apart from geological evidence of the final demise of the ice sheet, most field
-evidence relate to the LGM extent. It indicates that ice covered most of the
+Apart from geological evidence of the final demise of the Cordilleran ice sheet, most field
+evidence relates to its LGM extent. It indicates that ice covered most of the
 Canadian Cordillera, and some of the adjacent regions in north-western
@@
 and Seattle. Further east, beyond the Cascade Range, the Okanagan and Purcell
-lobes formed the southern Cordilleran ice sheet margin. The latter once dammed
+lobes formed the southern Cordilleran ice sheet margin. The latter dammed
 the Clark Fork River, inhibiting runoff and forming Glacial Lake Missoula
@@
 between Vancouver Island and the Olympic Peninsula, where it formed the Juan
-the Fuca marine ice lobe, calving into the Pacific Ocean \mref. Further to the
+de Fuca marine ice lobe, calving into the Pacific Ocean \mref. Further to the
 North, the ice margin encroached the Queen Charlotte Islands. However, the
@@
 the Fraser Plateau. Studies there have indicated that some of the highest
-reliefs emerged from the ice sheet prior to the surrounding lowlands, possibly
+uplands emerged from the ice sheet prior to the surrounding lowlands, possibly
 resulting in thin, stagnant ice remnants covering parts of the plateau
@@
 
+The thermodynamic core of the ice sheet model relies on four elementary field
+equations: the equation of incompressibility, the balance of stresses, the constitutive law for ice, and the conservation of energy. Firstly, ice flow is assumed \emph{incompressible}. Hence, ice
+density, $\rho$, is constant, and
+conservation of mass therefore implies a conservation of volume, which can be expressed
+in terms of the ice velocity vector, $\vv$ (\cref{fig:model-variables}), by
+\begin{equation}
+    \label{eqn:incompressibility}
+    \div{\vv} = 0 \,.
+\end{equation}
+
 \begin{figure}
@@
 \end{figure}
-
-The thermodynamic core of the ice sheet model relies on four elementary field
-equations. Firstly, ice flow is assumed \emph{incompressible}. Hence, the ice
-density, $\rho$, is constant, and
-conservation of mass therefore implies a conservation of volume, which can be expressed
-in terms of the ice velocity vector, $\vv$ (\cref{fig:model-variables}), by
-\begin{equation}
-    \label{eqn:incompressibility}
-    \div{\vv} = 0 \,.
-\end{equation}
 
@@
 and $A_c$, $A_w$, $Q_c$ and $Q_w$, are constant parameter corresponding to
-values measured above and below a critical temperature threshold
+values measured below and above a critical temperature threshold
 ${T_c=-10}$\,\degC\ \citep[\cref{tab:iceflowparams};][]{Paterson.Budd.1982}.
@@
     \label{eqn:invglenslaw}
-    \DST = B\,\doteps_e^{1/n-1}\,\SRT \,.
+    \DST = B\,\doteps_e^{(1/n)-1}\,\SRT \,.
 \end{equation}
@@
     \label{eqn:viscosity}
-    \nu = \frac{B}{2}\,\doteps_e^{1/n-1} \,.
+    \nu = \frac{B}{2}\,\doteps_e^{(1/n)-1} \,.
 \end{equation}
@@
 
-From \cref{eqn:vsia} it can be seen, that the shallow ice velocities, $\vsia$,
+From \cref{eqn:vsia} it can be seen that shallow ice velocities, $\vsia$,
 can be directly computed from the geometry of the ice sheet surface, $s$,
@@
 \end{equation}
-Because $\vsia$ is small where $\vssa$ is large and
+Because $\vsia$ is small where $\vssa$ is large, and
 reciprocally, we can expect the model to perform well in regions where
@@
 thus assuming instantaneous drainage. Although mass-conserving water drainage
-is now implemented in PISM \citep{Bueler.Pelt.2014}, is not included in the
+is now implemented in PISM \citep{Bueler.Pelt.2014}, it is not included in the
 simulations presented here for the sake of simplicity.
@@
 time-scales, because temperature fluctuations caused by climate change and
-varying isolation from the fluctuating ice sheet penetrate several kilometres
+varying degrees of isolation from atmospheric temperatures by ice sheet
+coverage, penetrate several kilometres
 into the rock, and eventually affect basal ice temperatures.
 
-Secondly, bedrock elevation evolve in response to the ice weight. The
+Secondly, bedrock elevation evolves in response to the ice weight. The
 \emph{bedrock deformation} model used in this thesis describes the flexure of
@@
 Although the PDD model is a crude parametrisation of processes happening at the
-surface of a glacier, its applicability has been proven \citep{Hock.2003},
+surface of a glacier, its applicability has been demonstrated
+\citep[e.g.][]{Hock.2003},
 while other, physically-based methods \citep[e.g.][]{Hock.2005}, require a
-precise knowledge of the physical state of the atmosphere, typically very
+precise knowledge of the physical state of the atmosphere, which typically is highly
 uncertain for past periods such as the last glacial cycle.
@@
 while adopting a range of constant temperature-offset corrections,
-${\Delta}T_{TS}$, between -10 and 0\,K. In \CCYC, on the opposite, we test ice
+${\Delta}T_{TS}$, between -10 and 0\,K. In \CCYC, on the contrary, we test ice
 sheet model sensitivity to time dependent temperature-offset corrections

### 2014-07-20 kappa.tex
Comment-out numerical parts and deactivate todos.

@@
 \newcommand{\renote}[1]{\footnote{\textbf{Comment reply}: #1}}
-\newcommand{\todo}[1]{\emph{[\textbf{Todo:} #1]}}
+\newcommand{\todo}[1]{} %\emph{[\textbf{Todo:} #1]}}
 \newcommand{\mref}[0]{\textbf{[ref.]}}
@@
 The North American Cordillera is an immense maze of mountain ranges and
-intermontane basins which once hosted an ice mass of the size of Greenland: the
+intermontane basins which once hosted an ice mass comparable to Greenland: the
 Cordilleran ice sheet. Although less extensive than the former Laurentide and
@@
 glacial maximum Cordilleran ice sheet can be largely explained by the present
-distribution of temperature and precipitation in the region, provided an
+distribution of temperature and precipitation in the region combined with
+paleoclimate reconstructions, provided an
 accurate treatment of seasonality, orographic precipitation, and daily
@@
 The simulations presented in this thesis demonstrate the potential of combining
-numerical ice sheet modelling to analysis of geological evidence in order to
-improve our understanding of the former glaciations.
+numerical ice sheet modelling with analysis of geological evidence in order to
+improve our understanding of former glaciations.
 
@@
 between Vancouver Island and the Olympic Peninsula, where it formed the Juan
-de Fuca marine ice lobe, calving into the Pacific Ocean \mref. Further to the
+de Fuca marine ice lobe, calving into the Pacific Ocean. Further to the
 North, the ice margin encroached the Queen Charlotte Islands. However, the
@@
 clear whether they have formed during the last glacial cycle or during earlier
-glaciations \mref.
-
-\todo{Find missing refs}
+glaciations.
+
+\todo{Find refs for marine extent of the ice sheet}
 
@@
   \centering
-  \caption{Constant parameters of the ice flow model.
-           \todo{find a better reference than \citet{Aschwanden.etal.2012}
-                 (library) for all the physical constants. Double-check
-                 \citet{Paterson.Budd.1982} and add missing references.}}
+  \caption{Constant parameters of the ice flow model.}
   \label{tab:iceflowparams}
@@
             & -
-            & \mref \\
+            & - \\
 
@@
             & \unit{J\,mol^{-1}\,K^{-1}}
-            & \mref \\
+            & - \\
 
@@
 
-    $\alpha$& \todo{check value}
+    $\alpha$& Effective pressure coefficient (\CCLI)
             & 0.95
@@
 
-    $\delta$& \todo{find a good name}
+    $\delta$& Effective pressure coefficient (\CCYC)
             & 0.02
@@
   \centering
-  \caption{Constant parameters of the bedrock thermal and deformation models.
-           \todo{Find out what the bedrock density corresponds to. Track down
-                 original sources in \citet{Ritz.1997}}}
+  \caption{Constant parameters of the bedrock thermal and deformation models.}
   \label{tab:bedrockparams}
@@
             & \unit{kg\,m^{-3}}
-            & \mref \\
+            & - \\
 
@@
             & \unit{J\,kg^{-1}\,K^{-1}}
-            & \citet{Ritz.1997} \\
+            & - \\
 
@@
             & \unit{W\,m^{-1}\,K^{-1}}
-            & \citet{Ritz.1997} \\
+            & - \\
 
@@
             & \unit{mW\,m^{-2}}
-            &  \\
+            & - \\
 
@@
     \hline
-%	pism overrides:bootstrapping geothermal flux value no var = 0.070 ;
   \end{tabular*}}
@@
   \centering
-  \caption{Constant parameters of the surface and atmosphere models.
-           \todo{Find a reference for the snow and rain temperature thresholds,
-                 maybe \citet{Hock.2003}.}}
+  \caption{Constant parameters of the surface and atmosphere models.}
   \label{tab:surfaceparams}
@@
             & \degC
-            & \mref \\
+            & - \\
 
@@
 % ----------------------------------------------------------------------
-\subsection{Numerical implementation}
-% ----------------------------------------------------------------------
-
-\todo{Add here a short section about discretization of the SIA and SSA, and
-      their numerical implementation. The main subtlety, I think, lies in the
-      fact that some computation need to use a ``staggered'' grid in order to
-      ensure that we add SIA and SSA values corresponding to the same point in
-      space. I need to check that. This section is also the place where to
-      introduce geographic transformation parameters that I used throughout
-      the thesis -- except in \cref{fig:map-cordillera}!}
-
-\begin{table}
-  \centering
-  \caption{Properties of the Canadian Atlas Lambert conformal conic projection
-           (EPSG code~3978) and the GRS~1980 ellipsoid, and corresponding
-           PROJ.4 parameter names, as used in all model runs, ice volume and
-           area calculations, and most figures.}
-  \label{tab:projparams}
-  \makebox[0pt]{\begin{tabular*}{80mm}{@{\extracolsep{\fill}}lcc}
-    \hline
-    Parameter               & Name              & Value \\
-    \hline
-    Projection name         & \texttt{+proj}    & \texttt{lcc} \\
-    Central meridian        & \texttt{+lon\_0}  & 95\degree W \\
-    Latitude of origin      & \texttt{+lat\_0}  & 49\degree N \\
-    First secant parallel   & \texttt{+lat\_1}  & 49\degree N \\
-    Second secant parallel  & \texttt{+lat\_2}  & 77\degree N \\
-    Semimajor radius        & \texttt{+a}       & 6378137\,m \\
-    Reciprocal flattening   & \texttt{+rf}      & 298.257222101 \\
-    \hline
-  \end{tabular*}}
-\end{table}
+%\subsection{Numerical implementation}
+% ----------------------------------------------------------------------
+
+%\todo{Add here a short section about discretization of the SIA and SSA, and
+%      their numerical implementation. The main subtlety, I think, lies in the
+%      fact that some computation need to use a ``staggered'' grid in order to
+%      ensure that we add SIA and SSA values corresponding to the same point in
+%      space. I need to check that. This section is also the place where to
+%      introduce geographic transformation parameters that I used throughout
+%      the thesis -- except in \cref{fig:map-cordillera}!}
+
+%\begin{table}
+%  \centering
+%  \caption{Properties of the Canadian Atlas Lambert conformal conic projection
+%           (EPSG code~3978) and the GRS~1980 ellipsoid, and corresponding
+%           PROJ.4 parameter names, as used in all model runs, ice volume and
+%           area calculations, and most figures.}
+%  \label{tab:projparams}
+%  \makebox[0pt]{\begin{tabular*}{80mm}{@{\extracolsep{\fill}}lcc}
+%    \hline
+%    Parameter               & Name              & Value \\
+%    \hline
+%    Projection name         & \texttt{+proj}    & \texttt{lcc} \\
+%    Central meridian        & \texttt{+lon\_0}  & 95\degree W \\
+%    Latitude of origin      & \texttt{+lat\_0}  & 49\degree N \\
+%    First secant parallel   & \texttt{+lat\_1}  & 49\degree N \\
+%    Second secant parallel  & \texttt{+lat\_2}  & 77\degree N \\
+%    Semimajor radius        & \texttt{+a}       & 6378137\,m \\
+%    Reciprocal flattening   & \texttt{+rf}      & 298.257222101 \\
+%    \hline
+%  \end{tabular*}}
+%\end{table}
 
 % ======================================================================
-\section{Software implementation}
+%\section{Software implementation}
 % ======================================================================
@@
 % ----------------------------------------------------------------------
-\subsection{Computational workflow}
-% ----------------------------------------------------------------------
-\todo{Figure. computational workflow}
-
-% ----------------------------------------------------------------------
-\subsection{Contributions to PISM}
-% ----------------------------------------------------------------------
-\todo{Table. PISM ``issues'' identified}
-
-% ----------------------------------------------------------------------
-\subsection{Peripheral tools}
-% ----------------------------------------------------------------------
-A Python PDD model\\
-Pre-processing scripts for GRASS GIS\\
-Plotting tools for PISM\\
+%\subsection{Computational workflow}
+% ----------------------------------------------------------------------
+%\todo{Figure. computational workflow}
+
+% ----------------------------------------------------------------------
+%\subsection{Contributions to PISM}
+% ----------------------------------------------------------------------
+%\todo{Table. PISM ``issues'' identified}
+
+% ----------------------------------------------------------------------
+%\subsection{Peripheral tools}
+% ----------------------------------------------------------------------
+%A Python PDD model\\
+%Pre-processing scripts for GRASS GIS\\
+%Plotting tools for PISM\\
 

### 2014-07-20 kappa.tex
Remove comment replies.

@@
 % My commands
-\newcommand{\renote}[1]{\footnote{\textbf{Comment reply}: #1}}
+%\newcommand{\renote}[1]{\footnote{\textbf{Comment reply}: #1}}
 \newcommand{\todo}[1]{} %\emph{[\textbf{Todo:} #1]}}
-\newcommand{\mref}[0]{\textbf{[ref.]}}
+%\newcommand{\mref}[0]{\textbf{[ref.]}}
 
@@
   \item[\PSDP] \bibentry{Seguinot.Rogozhina.2014}.
-  \item[\CCYC] \bibentry{Seguinot.etal.manuscript}.\renote{
-    Arjen, you had several comments regarding the style of references
-    throughout the kappa (name formatting, italic text). I am using the
-    Copernicus style here, but is there another recommendation to follow?
-    Maybe Peter knows?}
+  \item[\CCYC] \bibentry{Seguinot.etal.manuscript}.
 \end{description}
@@
 \emph{Glaciers}, these bodies of moving ice, move by a combination of
-meltwater-induced sliding at the base \citep[\S532]{Saussure.1786}\renote{
-    Actually I mean volume 1, published in ``MDCCLXXXVI'' (1786, right?),
-    which as a couple of beautiful paragraphs (would be worth
-    translating if I had time) about how the "heat from the Earth's interior"
-    melts basal ice to lubricate glacier flow. This is precisely those
-    paragraphs that \citet{Forbes.1846} refers to sixty years later when
-    advancing his ``viscous theory of glacier notion''.},
+meltwater-induced sliding at the base \citep[\S532]{Saussure.1786},
 and viscous deformation within the
@@
 
-As glaciers flow and slide across their bed, they transport\renote{
-    I would like to keep this statement general, surface transport is also
-    important. Erosion is not only by basal rock debris but also, indirectly,
-    by glacial meltwater.}
+As glaciers flow and slide across their bed, they transport
 rock debris and erode the landscape, thereby leaving geomorphological traces of
 their former presence (\cref{fig:glaciation-indicators}). In several parts of
-the world, montane people\renote{
-    I insist on keeping this: people of the Alps, for istance, had understood
-    that glacial erratics originated from glaciers before it has ever been
-    documented by explorers and scientists. It is stated in the old literature.
-    Check the new reference, reprinted in \citet{Mathews.1898} (not quite sure
-    how to cite this material).}
+the world, montane people
 and early explorers learned since long to read these traces, and to
@@
 than today, expansive \emph{ice sheets} had once covered much of Europe and
-North America \citep{Agassiz.1840}\renote{
-    I am a bit reluctant to write ``\citet{Agassiz.1840} eventually posed the
-    glacial theory''. Although he published these ideas, do they really
-    originate from him?}.
+North America \citep{Agassiz.1840}.
 
@@
 and allow for reconstructions of their past thickness and volume evolution, which
-are difficult to infer from geological evidence\renote{
-    Is this the place where to cite \citep{Denton.Hughes.1981}? \todo{I need to
-    check the contents of this book (library).}
-}.
+are difficult to infer from geological evidence.
 
@@
 Newfoundland, connecting with the Greenland ice sheet in the Canadian Arctic,
-and extending as far south as Chicago\renote{
-    Although already a few 100\,km into the ice margin, I think that Chicago
-    is a more illustrative location than Ohio or smaller cities there.}.
+and extending as far south as Chicago.
 The Laurentide ice sheet, largest of the three, was centred on the Hudson Bay,
@@
 (\cref{fig:plot-atm}a). About half of the model domain has mean annual
-temperatures below freezing. However, temperature \emph{seasonality}\renote{
-    ``Seasonality'' is a common term in the meteorological literature,
-    implicitly referring to temperature. At least this is what I understood
-    from Qiong.},
+temperatures below freezing. However, temperature \emph{seasonality},
 here computed as the difference between the warmest and the coldest months,
 varies between around 10\,\degC\ along most of the Pacific Coast to above
-40\,\degC\ for the interior Arctic sectors\renote{
-    I think the word ``polar'' or ``Arctic'' is needed here, see
-    \cref{fig:plot-atm}b.}
+40\,\degC\ for the interior Arctic sectors
 (\cref{fig:plot-atm}b). There exists a
@@
 discussed in further detail and compared with modelling results in
-\CCYC\renote{
-    I have not had time to dig into new literature. The Ward paper on
-    deglacial valley fills in the Pelly River is like Chinese to me, and I
-    could not access the Dyke paper on the Liard Lowland, assuming this is the
-    one cited by Martin in his paper.}.
+\CCYC.
 
@@
 Thirdly, after defining the strain-rate tensor
-${\SRT = \frac{1}{2}(\tgrad{\vv} + \tgrad{\vv}^{\mathrm{T}})}$\renote{
-  $\grad$, $\div$, $\tgrad$, $\tdiv$, are the classic gradient and divergence operators, applied on vector and tensor fields. $\tr$ is the trace of a matrix. I don't think these need to be defined.},
+${\SRT = \frac{1}{2}(\tgrad{\vv} + \tgrad{\vv}^{\mathrm{T}})}$,
 and the deviatoric stress tensor, ${\DST = \CST + p\,\IDT}$,
@@
 
-The till cohesion $c_0$ is assumed to be zero\renote{
-  Yes, we should probably defend this choice, but I am not aware of any
-  observations justifying this.}.
+The till cohesion $c_0$ is assumed to be zero.
 The effective pressure on the
@@
 equals precipitation
-when temperature is below ${T_0=0\,\degC}$, and decreases to zero\renote{
-    Well spotted, thanks!}
+when temperature is below ${T_0=0\,\degC}$, and decreases to zero
 linearly with temperature between $T_0$ and ${T_1=2\,\degC}$,

### 2014-07-20 kappa.tex
Rework the abstract.

@@
 
-This doctoral thesis presents a study of the late glacial history of
-north-western North America using numerical ice sheet modelling calibrated by
-geological evidence.
-
-The North American Cordillera is an immense maze of mountain ranges and
-intermontane basins which once hosted an ice mass comparable to Greenland: the
-Cordilleran ice sheet. Although less extensive than the former Laurentide and
-Fennoscandian ice sheets, the Cordilleran ice sheet perhaps remains the least
-understood Pleistocene ice sheet in terms of its extent, volume, and dynamics.
-Because of the irregular topography of the ice sheet's subglacial landscape,
-ice flowed along tortuous pathways, sometimes following the major topographical
-troughs, and sometimes flowing across them.
+This doctoral thesis presents a study of the glacial history of northwestern
+North America using numerical ice sheet modelling calibrated against field
+evidence. The area is characterized by the steep topography of several mountain
+ranges separated by large inter-montane depressions that were once covered by a
+large-scale ice mass, the former Cordilleran ice sheet. Due to this complex
+topographic setting, the Cordilleran ice sheet remains the least understood
+Pleistocene ice sheet in terms of its extent, volume, and dynamics. Because of
+the irregular topography on which the ice sheet formed, field studies have
+often only had local or regional relevance. Numerical modelling, on the other
+hand, allows quantitative reconstructions of whole ice sheet evolution based on
+approximated physics of glacier flow.
 
 This thesis presents simulations performed with PISM, a numerical ice sheet
-model that approximates glacier flow by basal sliding and internal deformation.
-These simulations show that the somewhat paradoxical geometry of the last
-glacial maximum Cordilleran ice sheet can be largely explained by the present
-distribution of temperature and precipitation in the region combined with
-paleoclimate reconstructions, provided an
-accurate treatment of seasonality, orographic precipitation, and daily
-temperature variability in the ice sheet model. Nevertheless, at the scale of
-the last glacial cycle, this configuration appears ephemeral. In contrast, the
-average stage of glaciation consists of isolated glaciers and ice caps covering
-the major mountain ranges of the North American Cordillera.
-
-The simulations presented in this thesis demonstrate the potential of combining
-numerical ice sheet modelling with analysis of geological evidence in order to
-improve our understanding of former glaciations.
+model that approximates glacier flow by basal sliding and internal deformation,
+and allows simulations of ice sheet growth and decay over multi-millenial time
+scales. Sensitivity experiments have shown that simulations adopting climate
+forcing from the North American Regional Reanalysis (NARR) yield ice sheet
+extents that best fit the field constraints during the Last Glacial Maximum
+(LGM). These simulations show that the geometry of the LGM Cordilleran ice
+sheet can be largely explained by present-day temperature and precipitation
+fields combined with palaeo-climate reconstructions if an accurate treatment of
+seasonality, orographic precipitation, and daily temperature variability is
+included in the ice sheet model.
+
+In contrast to this short-lived, maximum ice sheet configuration, most of the
+last glacial cycle is characterized by an intermediate state of glaciation
+including isolated glaciers and ice caps covering the major mountain ranges of
+the North American Cordillera. The largest and most persistent of these ice
+sheet nucleation centres is located over the Skeena Mountains. Modelled
+deglaciation conditions broadly match field evidence, and where they diverge,
+such as over the Fraser Plateau, they indicate either a lack of important
+processes in the model, or a more complex glaciation history than previously
+thought. The model yields a readvance of the ice sheet during late glacial
+times, which appears to match field evidence, before final retreat towards the
+Skeena Mountains, which we identify as a key area to understanding glacial
+dynamics of the Cordilleran ice sheet, highlighting the need for further
+geological investigation of this region.
+
+The simulations presented in this thesis demonstrate the potential of numerical
+ice sheet modelling to inform on ice sheet history over glacial cycles,
+provided that ice sheet models can be calibrated against field constraints.
 

### 2014-07-20 kappa.tex
Add more corrections from Arjen.

@@
 constant temperature depressions, targeting the LGM as the period for which ice
-sheet limits are best known. These simulations, presented in \CCLI\ and
-performed on a 10\,km-resolution grid, use constant temperature depressions
+sheet limits are best known. These simulations,
+performed on a 10\,km grid, use constant temperature depressions
 from the present climate state, which is represented using five different input
@@
 (\cref{fig:plot-lgm}). Following this sensitivity study, I have retained the
-NARR climatology as input to simulations of the Cordilleran ice sheet presented
-here and in \CCYC.
+NARR climatology as input to simulations of the Cordilleran ice sheet through
+the last glacial cycle (\CCYC).
 
@@
 that the effect of daily variability is strongest in continental regions, where
-the $\sPDD$ values are highest (cf. \cref{fig:plot-sdmap}). Where comparing the
-results to the mapped LGM ice limits (\cref{fig:map-cordillera}), which show
-a restricted ice extent in the continental areas despite low temperatures at
+the $\sPDD$ values are highest (cf. \cref{fig:plot-sdmap}). When the results
+are compared to the mapped LGM ice limits (\cref{fig:map-cordillera}), which show
+a restricted ice extent in continental areas despite low temperatures at
 present (\cref{fig:plot-atm}), it appears that including daily temperature
-variability in the ice sheet model results in closer agreement with these
+variability in the ice sheet model results in a closer agreement with these
 limits (\cref{fig:plot-sdeffect}b and~c). However, applying the parametrisation
@@
 
-The primary goal of this thesis has been to perform simulations of the
-Cordilleran ice sheet through the last glacial cycle. This is the topic of
-\CCYC, where monthly air temperature, precipitation, and PDD standard deviation
+The primary goal of my thesis has been to perform numerical reconstructions
+of Cordilleran ice sheet history through the last glacial cycle, which forms
+the topic of \CCYC.
+Here, monthly air temperature, precipitation, and PDD standard deviation
 climatologies from the NARR are combined with palaeo-temperature time-series
-from six proxy records to produce a time-dependent climate forcing (\CCYC,
+from six proxy records to produce a time-dependent climate forcing
+spanning over the last glacial cycle (\CCYC,
 Figs.~2--3). The resulting simulations, using horizontal resolutions of 6 and
@@
 simulation (\cref{fig:plot-snapshots}) show the growth and decay of the
-Cordilleran ice sheet over its last expansion cycle. Ice sheet growth begins on
-a residual ice cap located over the Skeena Mountains, which also corresponds
-to the location of the last remnants of the ice sheet towards the end of the
-deglaciation.
+Cordilleran ice sheet over its last expansion cycle. An ice cap located over
+the Skeena Mountains acts as the primary nucleation center for the Cordilleran
+ice sheet. After the LGM the ice sheet shrinks rapidly such that the last
+remnants of the ice sheet towards the end of the deglaciation in the early
+Holocene reside once again in the Skeena Mountains.
 
@@
 
-Through sensitivity studies, I have identified the predominant sources of
-uncertainty involved in numerical modelling of Cordilleran ice sheet. Through
-the different papers constituting this thesis, I present an assessment of some
-of these sources, including the choice of present-day atmospheric forcing, the
-effects of daily temperature variability, and the choice of palaeo-temperature
-time series used to drive the model through past climate history.
-
-The resulting simulations, combined with available geological evidence,
-yielded estimates of the LGM Cordilleran ice sheet extent and volume,
+Through sensitivity studies presented in my thesis, I have identified some of
+the predominant sources of uncertainty involved in numerical modelling of the
+Cordilleran ice sheet, including the choice of present-day atmospheric forcing
+fields, the effects of daily temperature variability, and the choice of
+palaeoclimate time series used simulate ice sheet growth over glacial cycle
+time scales. The resulting simulations, combined with available geological
+evidence, yielded estimates of the LGM Cordilleran ice sheet extent and volume,
 identification of the main ice-dispersal centres througout the last glacial
 cycle, and understanding of final deglaciation patterns. However, there remains
-considerable room for model improvements and further analysis of the simulation
-output. Sources of error not assessed in this thesis include potential
-buttressing effects of the Laurentide ice sheet, past changes in precipitation,
-and regional variability of the geothermal heat flux.
+considerable room for model improvements beyond the ones achieved in this
+thesis. The foremost sources of error that remain unassessed in this thesis
+include potential buttressing effects of the Laurentide ice sheet, changing
+precipitation patterns with evolving topography, and regional variability of
+the geothermal heat flux.
 
@@
 Certainly the most important assumption made in this thesis is that of a
-Cordilleran ice sheet largely disconnected from its much larger neighbour, the
-\emph{Laurentide ice sheet}. Geomorphological evidence from the suture zone
-indicates that, during the LGM, both Cordilleran and Laurentide ice flow has
-been deflected by their junction \citep{Kleman.etal.2010, Margold.etal.2013,
-Margold.etal.2013a}, although it remains unclear for how long this junction has
-lasted \citep{Jackson.etal.1997, Bednarski.Smith.2007, Gowan.2013}.
+Cordilleran ice sheet evolution entirely disconnected from its much larger
+neighbour, the \emph{Laurentide ice sheet}. Geomorphological evidence from the
+suture zone indicates that, during the LGM, both Cordilleran and Laurentide ice
+flow were deflected \citep{Kleman.etal.2010, Margold.etal.2013,
+Margold.etal.2013a}, although it remains unclear for how long this suture
+condition has lasted \citep{Jackson.etal.1997, Bednarski.Smith.2007,
+Gowan.2013}.
 
@@
 been disconnected \citep{Kleman.etal.2010}. Thus, this assumption only affects
-numerical reconstructions of the Cordilleran ice sheet presented here during
-the last glacial maximum and during the subsequent deglacation. Other numerical
-modelling results \citep{Gregoire.etal.2012} indicate the Laurentide ice sheet
-may have formed a buttress to ice flow from the Western Cordillera, resulting
-in a thicker Cordilleran ice sheet. However modelling alone will not elucidate
-the nature and duration of the interaction between of the Cordilleran and
-Laurentide ice sheets. For this, additional field constraints will be needed.
+numerical reconstructions of the Cordilleran ice sheet during the LGM and
+subsequent deglacation. Modelling results by \citet{Gregoire.etal.2012}
+indicate that the Laurentide ice sheet buttress resulted in a thicker
+Cordilleran ice sheet. However, modelling alone will neither elucidate the
+nature nor the duration of the interaction between of the Cordilleran and
+Laurentide ice sheets. For this, additional field constraints will be required.
 
@@
 the numerical ice sheet model is highly sensitive (\CCLI). In all sensitivity
-studies, we base precipitation patterns on current values derived atmospheric
+studies, we base precipitation patterns on current fields derived from
+atmospheric
 reanalyses. Although the distribution of precipitation has undoubtedly
 been affected by past changes in global climate and the intermittent presence
-of ice sheets and ice caps in the Northern Hemisphere, its changes are
+of ice sheets and ice caps in the Northern Hemisphere, such changes are
 difficult to assess. In the North American Cordillera, much of the continental
-interior was likely dryer than today due to the presence of ice caps on the
+interior was presumed dryer than today due to the presence of ice caps on the
 Pacific Coast ranges, reinforcing the current topographic barrier. However,
-some regions were probably rainier as a side effect of modified wind patterns.
-
-In meteorology, these effects are typically analysed using atmospheric
+some regions were probably wetter as a side effect of modified atmospheric
+circulation patterns.
+
+These effects are typically analysed using atmospheric
 circulation models. However, I have shown that different atmospheric reanalyses
@@
 expect that, for ice sheet simulations forced by different palaeo-climate
-simulations, the uncertainty associated to the choice of atmospheric forcing
-will be even higher. Perhaps alternative methods of coupling between ice sheet
+records, the uncertainty associated to the choice of atmospheric forcing
+will be even higher. Alternative methods of coupling between ice sheet
 evolution and precipitation changes need to be developed.
@@
 of advance or retreat of the ice sheet. However, without a better control on
-the geothermal heat flux, such analysis are perhaps too premature.
+the geothermal heat flux, such analyses are perhaps too premature.
 
@@
 before. Although mathematically uncertain, the coupled SIA-SSA approximation,
-combined with massively parallel computing, allowed ice sheet modelling at
-high spatial resolution and over multi-millenial time scales characteristic
-of ice sheet growth and decay.
-
-Through sensitivity studies, and comparison of model output to geological
-evidence, I have gained a personal understanding of Cordilleran ice sheet
-evolution through the last glacial cycle. I attempt to summarise my thoughts
-here in a few points, perhaps also applicable to other ice sheets in the
-Northern Hemisphere.
+combined with massively parallel computing, allowed for ice sheet modelling at
+large spatial scales, high spatial resolution, and long time scales
+characteristic of ice sheet growth and decay.
+
+From the outcome of sensitivity studies and the comparison of model output to
+field constraints, I conclude the following:
 
@@
 \item Although spatial patterns of glaciation in the North American Cordillera
-are sometimes paradoxical, they can be largely explained by the present-day
-distribution of temperature and precipitation in the region, provided an
+are sometimes poorly understood, they can be largely explained by the
+present-day distributions of temperature and precipitation, provided an
 accurate treatment of seasonality, orographic precipitation, and daily
-variability in the model, and a simple temperature-index accumulation and melt
-model.
-
-\item The LGM Cordilleran ice sheet corresponds to a transient stage, out of
-balance with the contemporaneous atmospheric temperatures. Its maximum extent
-is the result of a 50\,ka-long evolution, captured in the geologic record at a
-moment of transition from colder to warmer temperatures than those needed to
-sustain a constant ice sheet geometry. Therefore, this maximum stage is not
-representative of average conditions of the last glacial cycle.
-
-\item The patterns of growth and decay of the Cordilleran ice sheet are varied
-and differ by region and by epoch. Two-dimensional, conceptual models of growth
-and decay \citep[e.g.][]{Fulton.1991, Margold.etal.2013a} do not encompass this
+temperature variability in the model, and a simple temperature-index
+accumulation and melt model (PDD).
+
+\item The LGM Cordilleran ice sheet was a transient feature, out of balance
+with contemporaneous atmospheric temperatures. Its maximum extent is the result
+of a 50\,ka-long evolution. Peak glacial temperatures were lower than those
+needed sustain a constant ice sheet geometry because the ice sheet growth was
+finally inhibited by the deglacial transition to warmer temperatures .
+Therefore, this maximum stage is not representative of average conditions, as
+shown in the modelling by persistent ice caps throughout most of the last
+glacial cycle.
+
+\item Patterns of growth and decay of the Cordilleran ice sheet vary spatially
+and temporally. Two-dimensional, conceptual models of growth and decay
+\citep[e.g.][]{Fulton.1991, Margold.etal.2013a} do not encompass such
 diversity. In order to decipher the complexity of geological evidence, the

### 2014-07-21 cordillera-cycle.tex
Add last corrections from Martin.

@@
 %\newcommand{\aref}[0]{\textbf{[ref.]}}
+
+% Todo
+% change Fraser Plateau to Interior Plateau
 
@@
 In general, the topographic complexity of the North American Cordillera and its
-effect on glacial history have inhibited the construction of ice sheet-wide
+effect on glacial history have inhibited the reconstruction of ice sheet-wide
 glacial advance and retreat patterns such as
@@
 Mountains (CM), the Columbia and Rocky mountains (CRM), the Skeena Mountains
-(SM), and the Selwyn and MacKenzie Mountains (SMKM). Although the Coast,
+(SM), and the Selwyn and Mackenzie Mountains (SMKM). Although the Coast,
 Columbia, Rocky and Skeena mountains (CM, CRM, SM) are covered by mountain
@@
 centres for an ice sheet, this is not the case for the Selwyn and
-MacKenzie mountains (SMKM), where ice cover on the highest peaks is limited to a
-fraction of the last glacial cycle. In other words, the Selwyn and MacKenzie
+Mackenzie mountains (SMKM), where ice cover on the highest peaks is limited to a
+fraction of the last glacial cycle. In other words, the Selwyn and Mackenzie
 mountains appear as a secondary ice-dispersal centre which ``activates'' only
@@
 the landscape. A well-developed network of glacial valleys west of
-the MacKenzie Mountains (\citealp[Fig.~2]{Kleman.etal.2010}; \citealp[Fig.~8]
+the Mackenzie Mountains (\citealp[Fig.~2]{Kleman.etal.2010}; \citealp[Fig.~8]
 {Stroeven.etal.2010}) is modelled to have hosted warm-based ice
@@
 (Sect.~\ref{sec:icedyn}). However, both the Skeena Mountains and the area
-west of the MacKenzie Mountains are areas of higher-than-average geothermal
+west of the Mackenzie Mountains are areas of higher-than-average geothermal
 heat flux \citep{Blackwell.Richards.2004}. We can therefore expect longer
@@
 
-In the North American Cordillera, the presence of lateral melt-water channels
+In the North American Cordillera, the presence of lateral meltwater channels
 at high elevation \citep{Margold.etal.2011, Margold.etal.2013a,
@@
 completely deglaciated by 12\,ka, whereas a significant ice cover remains over
-the Skeena, Selwyn, MacKenzie, Wrangell and Saint-Elias mountains in the
+the Skeena, Selwyn, Mackenzie, Wrangell and Saint-Elias mountains in the
 northern sector of the modelling domain. After 12\,ka, deglaciation continues
@@
 local readvances, the GRIP-driven simulation shows that larger ice bodies
-emanating from the Skeena and MacKenzie mountains may still have existed at the
+emanating from the Skeena and Mackenzie mountains may still have existed at the
 time of this late-glacial readvance.
@@
 from the Northern Rocky Mountains over the Skeena and Selwyn Mountains and into
-the MacKenzie Mountains.
+the Mackenzie Mountains.
 
@@
 results could be that the Fraser Plateau lineation system predates
-deglaciation, and that deglacial landforms are largely absent. Alternatively,
+deglaciation, as perhaps indicated by some eskers that appear incompatible with
+these glacial lineations \citep[Fig.~9]{Margold.etal.2013a}. Alternatively,
 local geothermal heat associated with volcanic activity on the Fraser Plateau
@@
 centres are located in the Coast Mountains, the Columbia and Rocky mountains,
-the Selwyn and MacKenzie mountains, and most importantly, in the Skeena
+the Selwyn and Mackenzie mountains, and most importantly, in the Skeena
 Mountains. Throughout the modelled last glacial cycle, the Skeena Mountains
@@
            by the ice sheet include the Alaska Range (AR), the Wrangell and
-           St.-Elias mountains (WSEM), the Selwyn and MacKenzie mountains (SMKM),
+           St.-Elias mountains (WSEM), the Selwyn and Mackenzie mountains (SMKM),
            the Skeena Mountains (SM), the Coast and Rocky

### 2014-08-04 kappa.tex
Add corrections from Qiong.

@@
 % ======================================================================
-\section{Physical setting}
+\section{North American Cordillera}
 % ======================================================================
@@
 % ----------------------------------------------------------------------
-\subsection{Topography}
+\subsection{Topographic setting}
 % ----------------------------------------------------------------------
@@
 % ----------------------------------------------------------------------
-\subsection{Climate}
+\subsection{Climatic setting}
 % ----------------------------------------------------------------------
@@
 \emph{timing of maximum precipitation} through the year (\cref{fig:plot-atm}d).
-Coastal regions receive most precipitation during winter and fall, which is
+Coastal regions receive most precipitation during fall and winter, which is
 therefore largely delivered as snow. In contrast, many inland regions
 experience dry winters, and most of the annual precipitation, consequently
-falls there as rain during summer, further enhancing ice melt. The climatic
-setting of the North American Cordillera is further described in \CCLI, where
-its effects on the ice sheet model response have been studied in detail.
+falls there as rain during summer, further enhancing ice melt.
+
+These topographically-induced, spatial variations in temperature and
+precipitations tightly constrain the location of potential ice sheet nucleation
+centres that lead to the formation of a Cordilleran ice sheet. The numerical
+ice sheet model is therefore highly sensitive to the choice of input topography
+and climate used (\cref{sec:atm}). This forms the motivation for the study
+presented in \CCLI, where the climatic setting of the North American Cordillera
+is also described in further detail.
 
@@
 Evidence from glaciofluvial deposits dated using cosmogenic nuclides at
-$2.64^{+0.20}/_{-0.18}$\,Ma indicates the first presence of an ice sheet in the
+$2.64^{+0.20}_{-0.18}$\,Ma (uncertainty range from 2.46 to 2.84\,Ma)
+indicates the first presence of an ice sheet in the
 Cordilleran mountains during the late Pliocene \citep{Hidy.etal.2013}.
@@
 scope of this study. The two younger glaciations, however, occurred both during
-the last glacial cycle. First, the older of the two, the Gladstone Glaciation,
+the last glacial cycle. First, the older one of the two, the Gladstone
+Glaciation,
 reached its maximum extent between $50.4\pm3.0$ and $54.3\pm2.0\,^{10}$Be\,ka
@@
 Glaciation has not been documented along the southern ice sheet margin, making
-it difficult to seize its spacial extent. The younger of the two last glacial
+it difficult to seize its spacial extent. The younger one of the two last
+glacial
 cycle glaciations, the McConnell Glaciation, is associated with end moraines of
@@
 This section does not attempt to provide an extensive description of PISM,
-which can be found in an on-line documentation \citep{PISM-authors.2014}.
+which can be found in an on-line documentation
+\citep[\url{http://www.pism-docs.org},][]{PISM-authors.2014}.
 Instead, it aims to highlight the modelling choices that I have made in this
@@
            well-defined boundaries in the model.
-           From \citet{PISM-authors.2014}.}
+           From \citet{PISM-authors.2014}.
+           \todo{replace air by atmosphere.}}
   \label{fig:model-interfaces}
@@
 \subsection{Atmospheric forcing}
+\label{sec:atm}
 % ----------------------------------------------------------------------
@@
 domain is strongly dependent on input precipitation patterns, which are often
-poorly resolved by atmospheric reanalysis fields used to force the model
+poorly resolved by atmospheric reanalysis fields
 (\CCLI, Figs.~10--14). The closest match between modelled ice sheet extent and
@@
 fields, the effects of daily temperature variability, and the choice of
-palaeoclimate time series used simulate ice sheet growth over glacial cycle
+palaeoclimate time series used to simulate ice sheet growth and decay over
+glacial cycle
 time scales. The resulting simulations, combined with available geological

### 2014-08-04 figures/plot-sdeffect.py
Change figure format from PNG to PDF.

@@ -77,4 +77,4 @@ for i, ax in enumerate(grid):
             fontweight='bold', transform=ax.transAxes)
 
 # save
-fig.savefig('plot-sdeffect.png')
+fig.savefig('plot-sdeffect')

### 2014-08-05 kappa.tex
Change figure format from PNG to PDF.
Small change in PISM interfaces diagram.
Switch to two-column kappa class.

@@
 
-% Base class and packages
-\documentclass{article}
-%\makeatletter\newcommand{\@ptsize}{0}\makeatother  % needed by setspace
+% Kappa class and personal tweaks
+\documentclass[a4paper]{kappa}
+\makeatletter
+\setlength{\textwidth}{170mm}
+\setlength{\columnsep}{1cm}  % increase space betweem columns
+\setlength{\@figureindent}{0cm}  % remore figure caption padding
+\makeatother
+
+% load packages
 \usepackage{amsmath}
@@
 \usepackage{natbib}
-\usepackage{setspace}
 \usepackage{physics}
@@
 \usepackage{graphicx}
-\onehalfspacing
 \graphicspath{{figures/}}
@@
 % Move floats to the end
-\usepackage[nolists]{endfloat}
-\renewcommand{\efloatseparator}{\mbox{}}
+%\usepackage[nolists]{endfloat}
+%\renewcommand{\efloatseparator}{\mbox{}}
 
@@
 % Document properties
-\title{Numerical modelling of the Cordilleran ice sheet}
-%\runningtitle{Numerical modelling of the Cordilleran ice sheet}
-\author{Julien Seguinot}
+%\title{Numerical modelling of the Cordilleran ice sheet}
+%\author{Julien Seguinot}
+\thesistitle{Numerical modelling of the Cordilleran ice sheet}
+\runningtitle{Numerical modelling of the Cordilleran ice sheet}
+\candidate{Julien Seguinot}
+\runningauthor{J. Seguinot}
+\pubyear{2014}
+\department{Department of Physical Geography and Quaternary Geology}
+\university{Stockholm University}
+\printed{Universitetsservice US-AB, 2014.}
+%\issn{}
+%\isbn{}
+%\coverpic{Richard Swan. \textsf {www.richardswan.com}. Back cover portrait: Louise Tränk.}
+%\layout{Summary and Papers IV \& V: Alistair Auffret based on  \LaTeX~templates by Rickard Petterson.
+Published papers typeset by respective publishers, reprinted with permission.
+Graphics created using \textsf{Inkscape}, \textsf{R} and \textsf{OpenOffice Draw}.}
 
@@
 
-\maketitle
+\begin{frontmatter}
 
@@
 
-\newpage
-
-\section*{List of papers}
+%\begin{keyword}
+%\end{keyword}
+%\begin{svabstract}
+%\end{svabstract}
+
+\begin{thepapers}
+
+\textbf{List of papers}
 
@@
 
-\begin{description}
-  \item[\CCLI] \bibentry{Seguinot.etal.2014}.
-  \item[\PSDV] \bibentry{Seguinot.2013}.
-  \item[\PSDP] \bibentry{Seguinot.Rogozhina.2014}.
-  \item[\CCYC] \bibentry{Seguinot.etal.manuscript}.
-\end{description}
+\begin{paper}
+  \item \bibentry{Seguinot.etal.2014}.
+  \item \bibentry{Seguinot.2013}.
+  \item \bibentry{Seguinot.Rogozhina.2014}.
+  \item \bibentry{Seguinot.etal.manuscript}.
+\end{paper}
 
 \vfill
+
 \noindent\textbf{Author contributions}
-
-\footnotesize{\begin{enumerate}
+\footnotesize{\begin{paper}
   \item[I] A.~P.~Stroeven initiated the Cordilleran ice sheet modelling
@@
 co-authors contributed to improve the text.
-\end{enumerate}}
-
-\newpage
-%\maketitle
+\end{paper}}
+
+\end{thepapers}
+
+\end{frontmatter}
+
+\renewcommand{\contentsname}{Contents of cover summary}
 \tableofcontents
-\newpage
 
@@
 
-\begin{figure}
+\begin{figure*}
   \centering
@@
   \label{fig:glacier-mechanics}
-\end{figure}
-
-\begin{figure}
+\end{figure*}
+
+\begin{figure*}
   \centering
@@
   \label{fig:glaciation-indicators}
-\end{figure}
+\end{figure*}
 
@@
   \centering
-  \includegraphics[width=80mm]{paleo-timeseries}
+  \includegraphics{paleo-timeseries}
   \caption{Palaeo-climate records of the last 800\,ka.
@@
   \centering
-  \includegraphics[width=80mm]{paleo-glaciation}
+  \includegraphics{paleo-glaciation}
   \caption{Spatial extent of ice sheet and glacier cover in the northern
@@
   \label{fig:map-northamerica}
-\end{figure}
+\end{figure*}
 
@@
   \label{fig:map-cordillera}
-\end{figure}
-
-\begin{figure}
+\end{figure*}
+
+\begin{figure*}
   \centering
@@
   \label{fig:photo-alaska-range}
-\end{figure}
+\end{figure*}
 
@@
   \label{fig:photo-fraser-valley}
-\end{figure}
-
-\begin{figure}
+\end{figure*}
+
+\begin{figure*}
   \centering
@@
   \label{fig:plot-atm}
-\end{figure}
+\end{figure*}
 
@@
            well-defined boundaries in the model.
-           From \citet{PISM-authors.2014}.
-           \todo{replace air by atmosphere.}}
+           From \citet{PISM-authors.2014}.}
   \label{fig:model-interfaces}
@@
 
-\begin{table}
+\begin{table*}
   \centering
@@
   \end{tabular*}}
-\end{table}
+\end{table*}
 
@@
   \centering
-  \vspace{-16mm}
   \makebox[0pt]{
@@
   \hspace{1cm}
-  \begin{minipage}[t]{47.5mm}\vspace{0pt}
+  \begin{minipage}[t]{47.5mm}\vspace{-12pt}  % kappa class add 12 pt here
     \caption{Constant-climate simulation of the LGM Cordilleran ice sheet using
@@
   }
-  \vspace{-16mm}
 \end{figure*}
@@
   \label{fig:plot-sdeffect}
-\end{figure}
+\end{figure*}
 
@@
   \label{fig:plot-snapshots}
-\end{figure}
+\end{figure*}
 

### 2014-08-06 kappa.tex
Begin to improve figure placement.
Split equations not fitting in one column.

@@
 \usepackage{graphicx}
+\usepackage{multicol}
 \graphicspath{{figures/}}
@@
 
+% ======================================================================
+\onecolumn
 \renewcommand{\contentsname}{Contents of cover summary}
 \tableofcontents
+% ======================================================================
+
+\begin{figure*}[b]
+  \centering
+  \makebox[0pt]{
+    \subgraphics{a}{photo-glacier-crevasses}
+    \hspace{1cm}
+    \subgraphics{b}{photo-glacier-susitna}
+    \hspace{1cm}
+    \subgraphics{c}{photo-glacier-fold}
+  }
+  \caption{Illustration of various glacier ice behaviours:
+           \textbf{(a)} surface crevasses formed by ice flow over a hidden
+           bedrock knob, Glacier Blanc, French Alps, 2004;
+           \textbf{(b)} medial moraines illustrating downhill ice flow on
+           Susitna Glacier, Alaska Range, photo by Austin Post, 1970
+           \citep{NSIDC.2009}; and
+           \textbf{(c)} mountain glacier calving into a small lake, revealing a
+           folded structure formed by deformed debris layers in the ice,
+           Glacier d'Arsine, French Alps, 2004.}
+  \label{fig:glacier-mechanics}
+\end{figure*}
+
+\newpage
+
+\begin{figure*}[b]
+  \centering
+  \makebox[0pt]{
+    \subgraphics{a}{photo-glacial-valley}
+    \hspace{1cm}
+    \subgraphics{b}{photo-erratic-boulder}
+    \hspace{1cm}
+    \subgraphics{c}{photo-melt-channels}
+  }
+  \caption{Different indicators of past glaciation:
+           \textbf{(a)} glacial valley, exhibiting a broad, U-shaped profile
+           characteristic of erosion by glaciers,
+           Cirque du Fer-\`{a}-Cheval, French Alps, 2013;
+           \textbf{(b)} erratic boulders, deposited in an improbable
+           configuration as the glacier melted underneath,
+           Finse area, Norwegian Scandes, 2011; and
+           \textbf{(c)} lateral meltwater channels, formed on a gentle slope
+           as the ice margin retreated towards the valley bottom, Tuya Lake
+           area, Canadian Cordillera, 2012.}
+  \label{fig:glaciation-indicators}
+\end{figure*}
+
+\begin{multicols}{2}  % appears necessary for adequate figure positions
 
@@
 sheets over North America (\cref{fig:paleo-glaciation}).
+
+\end{multicols}
+\twocolumn
+
+\begin{figure}[t]
+  \centering
+  \includegraphics{paleo-timeseries}
+  \caption{Palaeo-climate records of the last 800\,ka.
+           \textbf{(a)} Benthic oxygen isotope (\chem{\delta^{18}O}) record
+           from a compilation of globally-distributed deep sea sediment cores
+           \citep{Lisiecki.Raymo.2005}. Higher \chem{\delta^{18}O} values
+           correspond to periods of lower temperatures \citep{Emiliani.1955}
+           or lower global sea level, and thus higher ice volume stored in
+           glaciers and ice sheets on land \citep{Shackleton.1967}.
+           \textbf{(b)} Temperature anomalies ($\Delta T$) reconstructed from a
+           central Antarctic ice core \citep[EPICA Dome~C,][]{Jouzel.etal.2007}.
+           The grey bands indicate the approximate time span of the last
+           glacial period of 110 to 10\,ka.}
+  \label{fig:paleo-timeseries}
+\end{figure}
+
+\begin{figure}[t]
+  \centering
+  \includegraphics{paleo-glaciation}
+  \caption{Spatial extent of ice sheet and glacier cover in the northern
+           hemisphere \textbf{(a)} at present \citep{Patterson.Kelso.2014}, and
+           \textbf{(b)} during the last glacial maximum
+           \citep{Ehlers.Gibbard.2007}.}
+  \label{fig:paleo-glaciation}
+\end{figure}
 
@@
 
-\begin{figure*}
-  \centering
-  \makebox[0pt]{
-    \subgraphics{a}{photo-glacier-crevasses}
-    \hspace{1cm}
-    \subgraphics{b}{photo-glacier-susitna}
-    \hspace{1cm}
-    \subgraphics{c}{photo-glacier-fold}
-  }
-  \caption{Illustration of various glacier ice behaviours:
-           \textbf{(a)} surface crevasses formed by ice flow over a hidden
-           bedrock knob, Glacier Blanc, French Alps, 2004;
-           \textbf{(b)} medial moraines illustrating downhill ice flow on
-           Susitna Glacier, Alaska Range, photo by Austin Post, 1970
-           \citep{NSIDC.2009}; and
-           \textbf{(c)} mountain glacier calving into a small lake, revealing a
-           folded structure formed by deformed debris layers in the ice,
-           Glacier d'Arsine, French Alps, 2004.}
-  \label{fig:glacier-mechanics}
-\end{figure*}
-
-\begin{figure*}
-  \centering
-  \makebox[0pt]{
-    \subgraphics{a}{photo-glacial-valley}
-    \hspace{1cm}
-    \subgraphics{b}{photo-erratic-boulder}
-    \hspace{1cm}
-    \subgraphics{c}{photo-melt-channels}
-  }
-  \caption{Different indicators of past glaciation:
-           \textbf{(a)} glacial valley, exhibiting a broad, U-shaped profile
-           characteristic of erosion by glaciers,
-           Cirque du Fer-\`{a}-Cheval, French Alps, 2013;
-           \textbf{(b)} erratic boulders, deposited in an improbable
-           configuration as the glacier melted underneath,
-           Finse area, Norwegian Scandes, 2011; and
-           \textbf{(c)} lateral meltwater channels, formed on a gentle slope
-           as the ice margin retreated towards the valley bottom, Tuya Lake
-           area, Canadian Cordillera, 2012.}
-  \label{fig:glaciation-indicators}
-\end{figure*}
-
-\begin{figure}
-  \centering
-  \includegraphics{paleo-timeseries}
-  \caption{Palaeo-climate records of the last 800\,ka.
-           \textbf{(a)} Benthic oxygen isotope (\chem{\delta^{18}O}) record
-           from a compilation of globally-distributed deep sea sediment cores
-           \citep{Lisiecki.Raymo.2005}. Higher \chem{\delta^{18}O} values
-           correspond to periods of lower temperatures \citep{Emiliani.1955}
-           or lower global sea level, and thus higher ice volume stored in
-           glaciers and ice sheets on land \citep{Shackleton.1967}.
-           \textbf{(b)} Temperature anomalies ($\Delta T$) reconstructed from a
-           central Antarctic ice core \citep[EPICA Dome~C,][]{Jouzel.etal.2007}.
-           The grey bands indicate the approximate time span of the last
-           glacial period of 110 to 10\,ka.}
-  \label{fig:paleo-timeseries}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics{paleo-glaciation}
-  \caption{Spatial extent of ice sheet and glacier cover in the northern
-           hemisphere \textbf{(a)} at present \citep{Patterson.Kelso.2014}, and
-           \textbf{(b)} during the last glacial maximum
-           \citep{Ehlers.Gibbard.2007}.}
-  \label{fig:paleo-glaciation}
-\end{figure}
-
 % ======================================================================
@@
 
+\begin{figure*}
+  \centering
+  \makebox[0pt]{\includegraphics{map-cordillera}}
+  \caption{Topographic map of the northern American Cordillera, oriented north.
+           Cumulative last glacial maximum ice cover \citep{Dyke.2004} is
+           marked in red, and the model domain in black.
+           The background map consists of ETOPO1 \citep{Amante.Eakins.2009}
+           and Natural Earth Data \citep{Patterson.Kelso.2014}.}
+  \label{fig:map-cordillera}
+\end{figure*}
+
+\begin{figure*}
+  \centering
+  \makebox[0pt]{\includegraphics{photo-alaska-range}}
+  \caption{The Alaska Range hosts several of the highest peaks in the North
+           American Cordillera. Despite this fact and its northerly location,
+           its north slope hosts relatively small glaciers only, which had not
+           been much more extensive during the Last Glacial Maximum
+           \citep{Kaufman.Manley.2004}, indicating dry climatic conditions.
+           The Alaska Range is tectonically active and rapidly eroding, as
+           shown by landslide-covered valley sides on this photograph.}
+  \label{fig:photo-alaska-range}
+\end{figure*}
+
+\begin{figure*}
+  \centering
+  \makebox[0pt]{
+    \subgraphics{a}{photo-vancouver-island}
+    \hspace{1cm}
+    \subgraphics{b}{photo-fraser-valley}
+  }
+  \caption{Contrasted climates of the North American Cordillera:
+           \textbf{(a)} Pacific Coast at Vancouver Island, western British
+           Columbia, by Martin Margold, 2010.
+           \textbf{(b)} Fraser Valley, south-central British Columbia, 2010.}
+  \label{fig:photo-fraser-valley}
+\end{figure*}
+
 % ----------------------------------------------------------------------
@@
 
+% ----------------------------------------------------------------------
+\subsection{Climatic setting}
+% ----------------------------------------------------------------------
+
 \begin{figure*}
   \centering
-  \makebox[0pt]{\includegraphics{map-cordillera}}
-  \caption{Topographic map of the northern American Cordillera, oriented north.
-           Cumulative last glacial maximum ice cover \citep{Dyke.2004} is
-           marked in red, and the model domain in black.
-           The background map consists of ETOPO1 \citep{Amante.Eakins.2009}
-           and Natural Earth Data \citep{Patterson.Kelso.2014}.}
-  \label{fig:map-cordillera}
+  \makebox[0pt]{\includegraphics{plot-atm}}
+  \caption{Some characteristics of the Cordilleran climate according to the
+           North American Regional Reanalysis
+           \citep[NARR,][]{Mesinger.etal.2006}.}
+  \label{fig:plot-atm}
 \end{figure*}
-
-\begin{figure*}
-  \centering
-  \makebox[0pt]{\includegraphics{photo-alaska-range}}
-  \caption{The Alaska Range hosts several of the highest peaks in the North
-           American Cordillera. Despite this fact and its northerly location,
-           its north slope hosts relatively small glaciers only, which had not
-           been much more extensive during the Last Glacial Maximum
-           \citep{Kaufman.Manley.2004}, indicating dry climatic conditions.
-           The Alaska Range is tectonically active and rapidly eroding, as
-           shown by landslide-covered valley sides on this photograph.}
-  \label{fig:photo-alaska-range}
-\end{figure*}
-
-% ----------------------------------------------------------------------
-\subsection{Climatic setting}
-% ----------------------------------------------------------------------
 
@@
 is also described in further detail.
-
-\begin{figure*}
-  \centering
-  \makebox[0pt]{
-    \subgraphics{a}{photo-vancouver-island}
-    \hspace{1cm}
-    \subgraphics{b}{photo-fraser-valley}
-  }
-  \caption{Contrasted climates of the North American Cordillera:
-           \textbf{(a)} Pacific Coast at Vancouver Island, western British
-           Columbia, by Martin Margold, 2010.
-           \textbf{(b)} Fraser Valley, south-central British Columbia, 2010.}
-  \label{fig:photo-fraser-valley}
-\end{figure*}
-
-\begin{figure*}
-  \centering
-  \makebox[0pt]{\includegraphics{plot-atm}}
-  \caption{Some characteristics of the Cordilleran climate according to the
-           North American Regional Reanalysis
-           \citep[NARR,][]{Mesinger.etal.2006}.}
-  \label{fig:plot-atm}
-\end{figure*}
 
@@
 \begin{subequations}
-\begin{align}
+\begin{gather}
+  \begin{split}
     \pdv{x} \left[2\bar{\nu}
                   \left(2\pdv{v_x}{x} + \pdv{v_y}{y}\right) \right]
-        + \pdv{y} \left[\bar{\nu}
-                        \left(\pdv{v_x}{y} + \pdv{v_y}{x}\right) \right]
-        + \pdv{\tau_{xz}}{z} &= \rho g \pdv{s}{x} \,, \\
+        &+ \pdv{y} \left[\bar{\nu}
+                         \left(\pdv{v_x}{y} + \pdv{v_y}{x}\right) \right] \\
+        &+ \pdv{\tau_{xz}}{z} = \rho g \pdv{s}{x} \,,
+  \end{split}\\
+  \begin{split}
     \pdv{x} \left[\bar{\nu}
                   \left(\pdv{v_x}{y} + \pdv{v_y}{x}\right) \right]
-        + \pdv{y} \left[2\bar{\nu}
-                        \left(\pdv{v_x}{x} + 2\pdv{v_y}{y}\right) \right]
-         + \pdv{\tau_{yz}}{z} &= \rho g \pdv{s}{y} \,,
-\end{align}
+        &+ \pdv{y} \left[2\bar{\nu}
+                         \left(\pdv{v_x}{x} + 2\pdv{v_y}{y}\right) \right] \\
+        &+ \pdv{\tau_{yz}}{z} = \rho g \pdv{s}{y} \,,
+  \end{split}
+\end{gather}
 \end{subequations}
@@
 \begin{subequations}
-\begin{align}
+\begin{gather}
   \label{eqn:vssa}
+  \begin{split}
     \pdv{x} \left[2\bar{\nu}h
                   \left(2\pdv{v_x}{x} + \pdv{v_y}{y}\right) \right]
-        + \pdv{y} \left[\bar{\nu}h
-                        \left(\pdv{v_x}{y} + \pdv{v_y}{x}\right) \right]
-        + \tau_{bx} &= \rho gh \pdv{s}{x} \,, \\
+        &+ \pdv{y} \left[\bar{\nu}h
+                         \left(\pdv{v_x}{y} + \pdv{v_y}{x}\right) \right] \\
+        &+ \tau_{bx} = \rho gh \pdv{s}{x} \,,
+  \end{split}\\
+  \begin{split}
     \pdv{x} \left[\bar{\nu}H
                   \left(\pdv{v_x}{y} + \pdv{v_y}{x}\right) \right]
-        + \pdv{y} \left[2\bar{\nu}h
-                        \left(\pdv{v_x}{x} + 2\pdv{v_y}{y}\right) \right]
-        + \tau_{by} &= \rho gh \pdv{s}{y} \,,
-\end{align}
+        &+ \pdv{y} \left[2\bar{\nu}h
+                         \left(\pdv{v_x}{x} + 2\pdv{v_y}{y}\right) \right] \\
+        &+ \tau_{by} = \rho gh \pdv{s}{y} \,,
+  \end{split}
+\end{gather}
 \end{subequations}
@@
 days (PDD), defined over an arbitrary time interval, $[t_1, t_2]$, as the
-integral of positive Celcius temperature $T-T_0$,
-\begin{equation}
-    \mathrm{PDD} = \int_{0}^{A}\max(T(t)-T_0,0)\,dt \,.
+integral of positive Celcius temperature $\theta = T-T_0$,
+\begin{equation}
+    \mathrm{PDD} = \int_{0}^{A}\max(\theta(t),0)\,dt \,.
 \end{equation}
@@
 forcing the PDD computation to rely on an idealised representation of the
-annual temperature cycle $T_{ac}$. Sub-annual temperature variability around
+annual temperature cycle $\theta_{ac}$. Sub-annual temperature variability around
 the freezing point, however, significantly affects surface melt on a multi-year
@@
 \citep{Reeh.1991},
-\begin{equation}
-    \PDD = \frac{1}{\sPDD\sqrt{2\pi}}
+\begin{multline}
+    \PDD = \frac{1}{\sPDD\sqrt{2\pi}} \\
+        \times
         \int_{t_1}^{t_2} \mathrm{d}t
-        \int_{0}^{\infty} \mathrm{d}T \,
-        T \exp\left({-\frac{(T-T_{ac})^2}{2\sPDD^2}}\right) \,,
-\end{equation}
+        \int_{0}^{\infty} \mathrm{d}\theta \,
+        \theta \exp\left({-\frac{(\theta-\theta_{ac})^2}{2\sPDD^2}}\right) \,,
+\end{multline}
 or more efficiently using an error function formulation
 \citep{Calov.Greve.2005},
-\begin{equation}
+\begin{multline}
     \label{eqn:calovgreve}
     \PDD = \int_{t_1}^{t_2} \mathrm{d}t
-        \left[\frac{\sPDD}{\sqrt{2\pi}}
-                \exp\left({-\frac{T_{ac}^2}{2\sPDD^2}}\right)
-              + \frac{T_{ac}}{2} \, \mathrm{erfc}
-                \left(-\frac{T_{ac}}{\sqrt{2}\sPDD}\right)\right] \,.
-\end{equation}
+        \bigg[\frac{\sPDD}{\sqrt{2\pi}}
+                \exp\left({-\frac{\theta_{ac}^2}{2\sPDD^2}}\right) \\
+              + \frac{\theta_{ac}}{2} \, \mathrm{erfc}
+                \left(-\frac{\theta_{ac}}{\sqrt{2}\sPDD}\right)\bigg] \,.
+\end{multline}
 
@@
 \begin{align}
-    {\Delta}T_{LR}(t, x, y) &= -\gamma [s(t, x, y)-b_{ref}(x, y)] \\
-                            &= -\gamma [h(t, x, y)+b(t, x, y)-b_{ref}(x, y)]\,,
+    {\Delta}T_{LR}(t, x, y) = -\gamma\,[&s(t, x, y) - b_{ref}(x, y)] \\
+                            = -\gamma\,[&h(t, x, y) + b(t, x, y) \notag \\
+                                        &- b_{ref}(x, y)] \,,
 \end{align}

### 2014-08-06 kappa.tex
Regroup all parameter tables into one.

@@
 %\newcommand{\mref}[0]{\textbf{[ref.]}}
+
+% Copernicus-like vertical space in tables
+\newcommand\tophline{\hline\noalign{\vspace{1mm}}}
+\newcommand\middlehline{\noalign{\vspace{1mm}}\hline\noalign{\vspace{1mm}}}
+\newcommand\bottomhline{\noalign{\vspace{1mm}}\hline}
 
@@
 thesis, and to present an overview of the physics involved in these choices.
+Parameter values are summarised in \cref{tab:params}.
 
@@
 
+\begin{table*}
+  \centering
+  \caption{Constant parameters of the ice sheet model.}
+  \label{tab:params}
+  \makebox[0pt]{\begin{tabular*}{170mm}{@{\hspace{2em}}l@{\extracolsep{\fill}}lrll}
+    \tophline
+    Not.    & Name & Value & Unit & Source \\
+    \middlehline
+
+    \multicolumn{2}{l}{\emph{Ice material properties}} \\
+
+    $\rho$  & Ice density
+            & 910
+            & \unit{kg\,m^{-3}}
+            & \citet{Aschwanden.etal.2012} \\
+
+    $g$     & Standard gravity
+            & 9.81
+            & \unit{m\,s^{-2}}
+            & \citet{Aschwanden.etal.2012} \\
+
+    $n$     & Glen exponent
+            & 3
+            & -
+            & - \\
+
+    $A_c$   & Ice hardness coefficient cold
+            & 3.61\e{-13}
+            & \unit{Pa^{-3}\,s^{-1}}
+            & \citet{Paterson.Budd.1982} \\
+
+    $A_w$   & Ice hardness coefficient warm
+            & 1.73\e3
+            & \unit{Pa^{-3}\,s^{-1}}
+            & \citet{Paterson.Budd.1982} \\
+
+    $Q_c$   & Flow law activation energy cold
+            & 6.0\e4
+            & \unit{J\,mol^{-1}}
+            & \citet{Paterson.Budd.1982} \\
+
+    $Q_w$   & Flow law activation energy warm
+            & 13.9\e4
+            & \unit{J\,mol^{-1}}
+            & \citet{Paterson.Budd.1982} \\
+
+    $R$     & Ideal gas constant
+            & 8.31441
+            & \unit{J\,mol^{-1}\,K^{-1}}
+            & - \\
+
+    $T_c$   & Flow law critical temperature
+            & 263.15
+            & \unit{K}
+            & \citet{Paterson.Budd.1982} \\
+
+    $f$     & Flow law water fraction coeff.
+            & 181.25
+            & -
+            & \citet{Lliboutry.Duval.1985} \\
+
+    $\beta$ & Clapeyron constant
+            & 7.9\e{-8}
+            & \unit{K\,Pa^{-1}}
+            & \citet{Luthi.etal.2002} \\
+
+    $c_i$   & Ice specific heat capacity
+            & 2009
+            & \unit{J\,kg^{-1}\,K^{-1}}
+            & \citet{Aschwanden.etal.2012} \\
+
+    $c_w$   & Water specific heat capacity
+            & 4170
+            & \unit{J\,kg^{-1}\,K^{-1}}
+            & \citet{Aschwanden.etal.2012} \\
+
+    $k$     & Ice thermal conductivity
+            & 2.10
+            & \unit{J\,m^{-1}\,K^{-1}\,s^{-1}}
+            & \citet{Aschwanden.etal.2012} \\
+
+    $L$     & Water latent heat of fusion
+            & 3.34\e5
+            & \unit{J\,kg^{-1}\,K^{-1}}
+            & \citet{Aschwanden.etal.2012} \\
+
+    \multicolumn{2}{l}{\emph{Basal sliding}} \\
+
+    $q$     & Pseudo-plastic sliding exponent
+            & 0.25
+            & -
+            & - \\
+
+    $v_{th}$& Pseudo-plastic threshold velocity
+            & 100.0
+            & \unit{m\,yr^{-1}}
+            & - \\
+
+    $c_0$   & Till cohesion
+            & 0.0
+            & Pa
+            & - \\
+
+    $\alpha$& Effective pressure coefficient (\CCLI)
+            & 0.95
+            & -
+            & - \\
+
+    $\delta$& Effective pressure coefficient (\CCYC)
+            & 0.02
+            & -
+            & - \\
+
+    $e_0$   & Till reference void ratio
+            & 0.69
+            & -
+            & \citet{Tulaczyk.etal.2000} \\
+
+    $C_c$   & Till compressibility coefficient
+            & 0.12
+            & -
+            & \citet{Tulaczyk.etal.2000} \\
+
+    $W_{max}$ & Maximal till water thickness
+            & 2.0
+            & m
+            & - \\
+
+    $b_0$   & Altitude of max. friction angle
+            & 0
+            & m
+            & - \\
+
+    $b_1$   & Altitude of min. friction angle
+            & 200
+            & m
+            & - \\
+
+    $\phi_0$& Minimum friction angle
+            & 10/15
+            & \degree
+            & - \\
+
+    $\phi_1$& Maximum friction angle
+            & 30/45
+            & \degree
+            & - \\
+
+    \multicolumn{2}{l}{\emph{Bedrock and lithosphere}} \\
+
+    $\rho_b$& Bedrock density
+            & 3300
+            & \unit{kg\,m^{-3}}
+            & - \\
+
+    $c_b$   & Bedrock specific heat capacity
+            & 1000
+            & \unit{J\,kg^{-1}\,K^{-1}}
+            & - \\
+
+    $k_b$   & Bedrock thermal conductivity
+            & 3.0
+            & \unit{W\,m^{-1}\,K^{-1}}
+            & - \\
+
+    $q_G$   & Geothermal heat flux
+            & 70.0
+            & \unit{mW\,m^{-2}}
+            & - \\
+
+    $\nu_m$ & Mantle viscosity
+            & 1\e21
+            & \unit{Pa\,s}
+            & \citet{Lingle.Clark.1985} \\
+
+    $\rho_l$& Lithosphere density
+            & 3300
+            & \unit{kg\,m^{-3}}
+            & \citet{Lingle.Clark.1985} \\
+
+    $D$     & Lithosphere flexural rigidity
+            & 5.0\e24
+            & \unit{N}
+            & \citet{Lingle.Clark.1985} \\
+
+    \multicolumn{2}{l}{\emph{Surface and atmosphere}} \\
+
+    $T_0$   & Temperature of snow precipitation
+            & 0.0
+            & \degC
+            & - \\
+
+    $T_1$   & Temperature of rain precipitation
+            & 2.0
+            & \degC
+            & - \\
+
+    $F_s$   & Degree-day factor for ice
+            & 3.04\e{-3}
+            & \unit{m\,K^{-1}\,day^{-1}}
+            & \citet{Shea.etal.2009} \\
+
+    $F_i$   & Degree-day factor for ice
+            & 4.59\e{-3}
+            & \unit{m\,K^{-1}\,day^{-1}}
+            & \citet{Shea.etal.2009} \\
+
+    $\gamma$& Air temperature lapse-rate
+            & 6\e{-3}
+            & \unit{K\,m{-1}}
+            & - \\
+
+    \bottomhline
+  \end{tabular*}}
+\end{table*}
+
 % ----------------------------------------------------------------------
@@
 values measured below and above a critical temperature threshold
-${T_c=-10}$\,\degC\ \citep[\cref{tab:iceflowparams};][]{Paterson.Budd.1982}.
+${T_c=-10}$\,\degC\ \citep[\cref{tab:params};][]{Paterson.Budd.1982}.
 The water fraction, $\omega$, is capped at a maximum value of 0.01, above which
@@
 conductivity of ice, while enthalpy can be expressed as $H=c\,T$, where $c$
-denotes the ice specific heat capacity (\cref{tab:iceflowparams}).
+denotes the ice specific heat capacity (\cref{tab:params}).
 Consequently, the enthalpy equation (\ref{eqn:enthalpy}) can be rewritten using
@@
 glacial cycle, thus requiring shallow approximations.
-
-\begin{table*}
-  \centering
-  \caption{Constant parameters of the ice flow model.}
-  \label{tab:iceflowparams}
-  \makebox[0pt]{\begin{tabular*}{165mm}{@{\extracolsep{\fill}}rlrll}
-    \hline
-    Not.    & Name & Value & Unit & Source \\
-    \hline
-
-    $\rho$  & Ice density
-            & 910
-            & \unit{kg\,m^{-3}}
-            & \citet{Aschwanden.etal.2012} \\
-
-    $g$     & Standard gravity
-            & 9.81
-            & \unit{m\,s^{-2}}
-            & \citet{Aschwanden.etal.2012} \\
-
-    $n$     & Glen exponent
-            & 3
-            & -
-            & - \\
-
-    $A_c$   & Ice hardness coefficient cold
-            & 3.61\e{-13}
-            & \unit{Pa^{-3}\,s^{-1}}
-            & \citet{Paterson.Budd.1982} \\
-
-    $A_w$   & Ice hardness coefficient warm
-            & 1.73\e3
-            & \unit{Pa^{-3}\,s^{-1}}
-            & \citet{Paterson.Budd.1982} \\
-
-    $Q_c$   & Flow law activation energy cold
-            & 6.0\e4
-            & \unit{J\,mol^{-1}}
-            & \citet{Paterson.Budd.1982} \\
-
-    $Q_w$   & Flow law activation energy warm
-            & 13.9\e4
-            & \unit{J\,mol^{-1}}
-            & \citet{Paterson.Budd.1982} \\
-
-    $R$     & Ideal gas constant
-            & 8.31441
-            & \unit{J\,mol^{-1}\,K^{-1}}
-            & - \\
-
-    $T_c$   & Flow law critical temperature
-            & 263.15
-            & \unit{K}
-            & \citet{Paterson.Budd.1982} \\
-
-    $f$     & Flow law water fraction coeff.
-            & 181.25
-            & -
-            & \citet{Lliboutry.Duval.1985} \\
-
-    $\beta$ & Clapeyron constant
-            & 7.9\e{-8}
-            & \unit{K\,Pa^{-1}}
-            & \citet{Luthi.etal.2002} \\
-
-    $c_i$   & Ice specific heat capacity
-            & 2009
-            & \unit{J\,kg^{-1}\,K^{-1}}
-            & \citet{Aschwanden.etal.2012} \\
-
-    $c_w$   & Water specific heat capacity
-            & 4170
-            & \unit{J\,kg^{-1}\,K^{-1}}
-            & \citet{Aschwanden.etal.2012} \\
-
-    $k$     & Ice thermal conductivity
-            & 2.10
-            & \unit{J\,m^{-1}\,K^{-1}\,s^{-1}}
-            & \citet{Aschwanden.etal.2012} \\
-
-    $L$     & Water latent heat of fusion
-            & 3.34\e5
-            & \unit{J\,kg^{-1}\,K^{-1}}
-            & \citet{Aschwanden.etal.2012} \\
-
-    \hline
-  \end{tabular*}}
-\end{table*}
 
@@
 velocity, and $v_{th}$ is an arbitrary velocity threshold
-(\cref{tab:slidingparams}). Note that, in the case of
+(\cref{tab:params}). Note that, in the case of
 $q=0$, \cref{eqn:pseudoplastic} corresponds to a purely plastic law
@@
 chosen as 0.02 based on simple sensitivity tests on the Greenland ice sheet
-(Table~\ref{tab:slidingparams}).
+(Table~\ref{tab:params}).
 
@@
 highest observed palaeo-shoreline elevations in the model domain
-\citep[\cref{tab:slidingparams};][]{Clague.James.2002}.
-
-\begin{table*}
-  \centering
-  \caption{Constant parameters of the sliding model. Missing sources
-           correspond to uncalibrated values.}
-  \label{tab:slidingparams}
-  \makebox[0pt]{\begin{tabular*}{165mm}{@{\extracolsep{\fill}}rlrll}
-    \hline
-    Not.    & Name & Value & Unit & Source \\
-    \hline
-
-    $q$     & Pseudo-plastic sliding exponent
-            & 0.25
-            & -
-            & - \\
-
-    $v_{th}$& Pseudo-plastic threshold velocity
-            & 100.0
-            & \unit{m\,yr^{-1}}
-            & - \\
-
-    $c_0$   & Till cohesion
-            & 0.0
-            & Pa
-            & - \\
-
-    $\alpha$& Effective pressure coefficient (\CCLI)
-            & 0.95
-            & -
-            & - \\
-
-    $\delta$& Effective pressure coefficient (\CCYC)
-            & 0.02
-            & -
-            & - \\
-
-    $e_0$   & Till reference void ratio
-            & 0.69
-            & -
-            & \citet{Tulaczyk.etal.2000} \\
-
-    $C_c$   & Till compressibility coefficient
-            & 0.12
-            & -
-            & \citet{Tulaczyk.etal.2000} \\
-
-    $W_{max}$ & Maximal till water thickness
-            & 2.0
-            & m
-            & - \\
-
-    $b_0$   & Altitude of max. friction angle
-            & 0
-            & m
-            & - \\
-
-    $b_1$   & Altitude of min. friction angle
-            & 200
-            & m
-            & - \\
-
-    $\phi_0$& Minimum friction angle
-            & 10/15
-            & \degree
-            & - \\
-
-    $\phi_1$& Maximum friction angle
-            & 30/45
-            & \degree
-            & - \\
-
-    \hline
-  \end{tabular*}}
-\end{table*}
+\citep[\cref{tab:params};][]{Clague.James.2002}.
 
@@
 where $k_b$, $\rho_b$ and $c_b$ are the thermal conductivity, density and
-specific heat capacity of the bedrock (\cref{tab:bedrockparams}).
+specific heat capacity of the bedrock (\cref{tab:params}).
 Bedrock temperature is conditioned at depth by a fixed upwards geothermal heat
-flux $q_G$  (\cref{tab:bedrockparams}). The bedrock thermal model
+flux $q_G$  (\cref{tab:params}). The bedrock thermal model
 is necessary for modelling the Cordilleran ice sheet on multi-millennial
@@
 where $\nu_m$ is mantle viscosity, $\rho_l$ is lithosphere density, $D$ is the
-lithosphere's flexural rigidity (\cref{tab:bedrockparams}), and $\sigma_{zzb}$
+lithosphere's flexural rigidity (\cref{tab:params}), and $\sigma_{zzb}$
 corresponds to the ice load
@@
 the ice sheet growth and the isostatic bedrock response.
-
-\begin{table*}
-  \centering
-  \caption{Constant parameters of the bedrock thermal and deformation models.}
-  \label{tab:bedrockparams}
-  \makebox[0pt]{\begin{tabular*}{165mm}{@{\extracolsep{\fill}}rlrll}
-    \hline
-    Not.    & Name & Value & Unit & Source \\
-    \hline
-
-    $\rho_b$& Bedrock density
-            & 3300
-            & \unit{kg\,m^{-3}}
-            & - \\
-
-    $c_b$   & Bedrock specific heat capacity
-            & 1000
-            & \unit{J\,kg^{-1}\,K^{-1}}
-            & - \\
-
-    $k_b$   & Bedrock thermal conductivity
-            & 3.0
-            & \unit{W\,m^{-1}\,K^{-1}}
-            & - \\
-
-    $q_G$   & Geothermal heat flux
-            & 70.0
-            & \unit{mW\,m^{-2}}
-            & - \\
-
-    $\nu_m$ & Mantle viscosity
-            & 1\e21
-            & \unit{Pa\,s}
-            & \citet{Lingle.Clark.1985} \\
-
-    $\rho_l$& Lithosphere density
-            & 3300
-            & \unit{kg\,m^{-3}}
-            & \citet{Lingle.Clark.1985} \\
-
-    $D$     & Lithosphere flexural rigidity
-            & 5.0\e24
-            & \unit{N}
-            & \citet{Lingle.Clark.1985} \\
-
-    \hline
-  \end{tabular*}}
-\end{table*}
 
@@
 (\cref{fig:model-pdd}). The degree-day factors for snow, $F_s$, and ice, $F_i$
-(\cref{tab:surfaceparams}),
+(\cref{tab:params}),
 are derived from mass-balance measurements on contemporary glaciers in the
@@
 
+% ----------------------------------------------------------------------
+\subsection{Atmospheric forcing}
+\label{sec:atm}
+% ----------------------------------------------------------------------
+
 \begin{table*}
   \centering
-  \caption{Constant parameters of the surface and atmosphere models.}
-  \label{tab:surfaceparams}
-  \makebox[0pt]{\begin{tabular*}{165mm}{@{\extracolsep{\fill}}rlrll}
-    \hline
-    Not.    & Name & Value & Unit & Source \\
-    \hline
-
-    $T_0$   & Temperature of snow precipitation
-            & 0.0
-            & \degC
-            & - \\
-
-    $T_1$   & Temperature of rain precipitation
-            & 2.0
-            & \degC
-            & - \\
-
-    $F_s$   & Degree-day factor for ice
-            & 3.04\e{-3}
-            & \unit{m\,K^{-1}\,day^{-1}}
-            & \citet{Shea.etal.2009} \\
-
-    $F_i$   & Degree-day factor for ice
-            & 4.59\e{-3}
-            & \unit{m\,K^{-1}\,day^{-1}}
-            & \citet{Shea.etal.2009} \\
-
-    $\gamma$& Air temperature lapse-rate
-            & 6\e{-3}
-            & \unit{K\,m{-1}}
-            & - \\
-
-    \hline
+  \caption{Sources of forcing data used to inform the ice sheet model.}
+  \label{tab:datasources}
+  \makebox[0pt]{\begin{tabular*}{170mm}{l@{\extracolsep{\fill}}lll}
+    \tophline
+    Name        & Type & Original reference & Employed in \\
+    \middlehline
+
+    ETOPO1      & Bedrock topography
+                & \citet{Amante.Eakins.2009}
+                & \CCLI, \CCYC \\
+
+    WorldClim   & Gridded observations
+                & \citet{Hijmans.etal.2005}
+                & \CCLI \\
+
+    NCEP/NCAR   & Atmospheric reanalysis
+                & \citet{Kalnay.etal.1996}
+                & \CCLI \\
+
+    ERA-40      & \qquad '' \qquad
+                & \citet{Uppala.etal.2005}
+                & \PSDP \\
+
+    ERA-Interim & \qquad '' \qquad
+                & \citet{Dee.etal.2011}
+                & \CCLI, \PSDV \\
+
+    CFSR        & \qquad '' \qquad
+                & \citet{Saha.etal.2010}
+                & \CCLI \\
+
+    NARR        & \qquad '' \qquad
+                & \citet{Mesinger.etal.2006}
+                & \CCLI, \CCYC \\
+
+    GRIP        & Ice-core \chem{\delta^{18}O} record
+                & \citet{Dansgaard.etal.1993}
+                & \CCYC \\
+
+    NGRIP       & \qquad '' \qquad
+                & \citet{Andersen.etal.2004}
+                & \CCYC \\
+
+    EPICA       & \qquad '' \qquad
+                & \citet{Jouzel.etal.2007}
+                & \CCYC \\
+
+    Vostok      & \qquad '' \qquad
+                & \citet{Petit.etal.1999}
+                & \CCYC \\
+
+    ODP~1012    & Sea-floor \chem{U^{K'}_{37}} record
+                & \citet{Herbert.etal.2001}
+                & \CCYC \\
+
+    ODP~1020    & \qquad '' \qquad
+                & \citet{Herbert.etal.2001}
+                & \CCYC \\
+
+    \bottomhline
   \end{tabular*}}
 \end{table*}
-
-% ----------------------------------------------------------------------
-\subsection{Atmospheric forcing}
-\label{sec:atm}
-% ----------------------------------------------------------------------
 
@@
 annual temperature lapse rate of $\gamma = 6\,\unit{K\,km^{-1}}$
-(\cref{tab:surfaceparams}).
+(\cref{tab:params}).
 
@@
 
-\begin{table*}
-  \centering
-  \caption{Sources of forcing data used to inform the ice sheet model.}
-  \label{tab:datasources}
-  \makebox[0pt]{\begin{tabular*}{160mm}{@{\extracolsep{\fill}}llll}
-    \hline
-    Name        & Type & Original reference & Employed in \\
-    \hline
-
-    ETOPO1      & Bedrock topography
-                & \citet{Amante.Eakins.2009}
-                & \CCLI, \CCYC \\
-
-    WorldClim   & Gridded observations
-                & \citet{Hijmans.etal.2005}
-                & \CCLI \\
-
-    NCEP/NCAR   & Atmospheric reanalysis
-                & \citet{Kalnay.etal.1996}
-                & \CCLI \\
-
-    ERA-40      & \qquad '' \qquad
-                & \citet{Uppala.etal.2005}
-                & \PSDP \\
-
-    ERA-Interim & \qquad '' \qquad
-                & \citet{Dee.etal.2011}
-                & \CCLI, \PSDV \\
-
-    CFSR        & \qquad '' \qquad
-                & \citet{Saha.etal.2010}
-                & \CCLI \\
-
-    NARR        & \qquad '' \qquad
-                & \citet{Mesinger.etal.2006}
-                & \CCLI, \CCYC \\
-
-    GRIP        & Ice-core \chem{\delta^{18}O} record
-                & \citet{Dansgaard.etal.1993}
-                & \CCYC \\
-
-    NGRIP       & \qquad '' \qquad
-                & \citet{Andersen.etal.2004}
-                & \CCYC \\
-
-    EPICA       & \qquad '' \qquad
-                & \citet{Jouzel.etal.2007}
-                & \CCYC \\
-
-    Vostok      & \qquad '' \qquad
-                & \citet{Petit.etal.1999}
-                & \CCYC \\
-
-    ODP~1012    & Sea-floor \chem{U^{K'}_{37}} record
-                & \citet{Herbert.etal.2001}
-                & \CCYC \\
-
-    ODP~1020    & \qquad '' \qquad
-                & \citet{Herbert.etal.2001}
-                & \CCYC \\
-
-    \hline
-  \end{tabular*}}
-\end{table*}
-
 % ----------------------------------------------------------------------
@@
 %           area calculations, and most figures.}
-%  \label{tab:projparams}
+%  \label{tab:params}
 %  \makebox[0pt]{\begin{tabular*}{80mm}{@{\extracolsep{\fill}}lcc}
@@
 are the degree-day factors for snow, $F_s$, and ice, $F_i$
-(\cref{fig:model-pdd,tab:surfaceparams}), and the standard deviation of
+(\cref{fig:model-pdd,tab:params}), and the standard deviation of
 near-surface air temperature, $\sPDD$
-(\cref{eqn:calovgreve,tab:surfaceparams}). Although the first two can only be
+(\cref{eqn:calovgreve,tab:params}). Although the first two can only be
 deduced from surface mass-balance measurements on present-day glaciers

### 2014-08-06 kappa.tex
Move TOC to separate page and re-arrange figures.

@@
 
+\onecolumn
+
 % ======================================================================
-\onecolumn
-\renewcommand{\contentsname}{Contents of cover summary}
+\hfill \textbf{\huge Cover summary}
+\vfill
 \tableofcontents
 % ======================================================================
+
+\newpage
 
@@
 \end{figure*}
-
-\newpage
 
@@
 
-\begin{multicols}{2}  % appears necessary for adequate figure positions
+\begin{multicols}{2}  % appears necessary for bottom figures on first page
 
@@
 
+\end{multicols}
+\twocolumn
+
 More recently, palaeoclimate records extracted from deep sea sediments and ice
@@
 most recent of these cycles, the \emph{last glacial cycle}.
+
+\begin{figure}
+  \centering
+  \includegraphics{paleo-timeseries}
+  \caption{Palaeo-climate records of the last 800\,ka.
+           \textbf{(a)} Benthic oxygen isotope (\chem{\delta^{18}O}) record
+           from a compilation of globally-distributed deep sea sediment cores
+           \citep{Lisiecki.Raymo.2005}. Higher \chem{\delta^{18}O} values
+           correspond to periods of lower temperatures \citep{Emiliani.1955}
+           or lower global sea level, and thus higher ice volume stored in
+           glaciers and ice sheets on land \citep{Shackleton.1967}.
+           \textbf{(b)} Temperature anomalies ($\Delta T$) reconstructed from a
+           central Antarctic ice core \citep[EPICA Dome~C,][]{Jouzel.etal.2007}.
+           The grey bands indicate the approximate time span of the last
+           glacial period of 110 to 10\,ka.}
+  \label{fig:paleo-timeseries}
+\end{figure}
 
@@
 
-\end{multicols}
-\twocolumn
-
-\begin{figure}[t]
-  \centering
-  \includegraphics{paleo-timeseries}
-  \caption{Palaeo-climate records of the last 800\,ka.
-           \textbf{(a)} Benthic oxygen isotope (\chem{\delta^{18}O}) record
-           from a compilation of globally-distributed deep sea sediment cores
-           \citep{Lisiecki.Raymo.2005}. Higher \chem{\delta^{18}O} values
-           correspond to periods of lower temperatures \citep{Emiliani.1955}
-           or lower global sea level, and thus higher ice volume stored in
-           glaciers and ice sheets on land \citep{Shackleton.1967}.
-           \textbf{(b)} Temperature anomalies ($\Delta T$) reconstructed from a
-           central Antarctic ice core \citep[EPICA Dome~C,][]{Jouzel.etal.2007}.
-           The grey bands indicate the approximate time span of the last
-           glacial period of 110 to 10\,ka.}
-  \label{fig:paleo-timeseries}
-\end{figure}
-
-\begin{figure}[t]
+\begin{figure}
   \centering
@@
 physics-based, quantitative reconstructions of ice sheet sectors where they do
-not
-\citep[e.g.,][]{Marshall.etal.2002, Tarasov.Peltier.2004}.
-In addition, numerical ice sheet models provide quantitative insights on the
-four-dimensional distribution of stresses and deformations in palaeo-ice sheets
-and allow for reconstructions of their past thickness and volume evolution, which
-are difficult to infer from geological evidence.
-
-In this thesis, I focus on numerical modelling of the last Cordilleran ice
-sheet, which once covered much of the mountainous terrain of western North
-America \citep{Dawson.1888}, a region where topographic complexity has
-presented a particular challenge to reconstructing glacial history based on the
-geomorphological record \citep[e.g.,][]{Kleman.etal.2010} or numerical ice
-sheet modelling \citep{Robert.1991}.
-
-% ======================================================================
-\section{North American Cordillera}
-% ======================================================================
-
-During the last glacial cycle, much of North America has been covered by ice
-(\cref{fig:map-northamerica}). During the Last Glacial Maximum (LGM), peak of
-glaciation, three major ice sheets had collided to
-form a continuous blanket of ice that spanned from southern Alaska to
-Newfoundland, connecting with the Greenland ice sheet in the Canadian Arctic,
-and extending as far south as Chicago.
-The Laurentide ice sheet, largest of the three, was centred on the Hudson Bay,
-and fully covered the Canadian Prairies, Eastern Canada, and the Great Lakes
-Basin. The Innuitian ice sheet, much smaller, covered the northernmost Canadian
-archipelago. Finally, the \emph{Cordilleran ice sheet}, intermediate in size,
-capped the mountain
-ranges of the Western Cordillera.
+not \citep[e.g.,][]{Marshall.etal.2002, Tarasov.Peltier.2004}.
 
@@
 \end{figure*}
+
+In addition, numerical ice sheet models provide quantitative insights on the
+four-dimensional distribution of stresses and deformations in palaeo-ice sheets
+and allow for reconstructions of their past thickness and volume evolution, which
+are difficult to infer from geological evidence.
+
+In this thesis, I focus on numerical modelling of the last Cordilleran ice
+sheet, which once covered much of the mountainous terrain of western North
+America \citep{Dawson.1888}, a region where topographic complexity has
+presented a particular challenge to reconstructing glacial history based on the
+geomorphological record \citep[e.g.,][]{Kleman.etal.2010} or numerical ice
+sheet modelling \citep{Robert.1991}.
+
+% ======================================================================
+\section{North American Cordillera}
+% ======================================================================
+
+During the last glacial cycle, much of North America has been covered by ice
+(\cref{fig:map-northamerica}). During the Last Glacial Maximum (LGM), peak of
+glaciation, three major ice sheets had collided to
+form a continuous blanket of ice that spanned from southern Alaska to
+Newfoundland, connecting with the Greenland ice sheet in the Canadian Arctic,
+and extending as far south as Chicago.
+The Laurentide ice sheet, largest of the three, was centred on the Hudson Bay,
+and fully covered the Canadian Prairies, Eastern Canada, and the Great Lakes
+Basin. The Innuitian ice sheet, much smaller, covered the northernmost Canadian
+archipelago. Finally, the \emph{Cordilleran ice sheet}, intermediate in size,
+capped the mountain
+ranges of the Western Cordillera.
 
@@
 
-Next to the atmospheric forcing, another critical source of uncertainty in
-palaeo-ice sheet modelling is the choice of parameters used by the surface mass
-balance model \citep[e.g.,][]{Hebeler.etal.2008}. In our case, these parameters
-are the degree-day factors for snow, $F_s$, and ice, $F_i$
-(\cref{fig:model-pdd,tab:params}), and the standard deviation of
-near-surface air temperature, $\sPDD$
-(\cref{eqn:calovgreve,tab:params}). Although the first two can only be
-deduced from surface mass-balance measurements on present-day glaciers
-\citep[e.g.,][]{Shea.etal.2009} and assumptions that these values are also
-valid for palaeo-simulations, $\sPDD$ can be estimated from temperature time
-series \citep[e.g.,][]{Fausto.etal.2011}.
-
-In \PSDV, I used near-surface air temperatures from the ERA-Interim reanalysis
-\citep{Dee.etal.2011} to compute a global, monthly distribution of $\sPDD$. The
-resulting grids show that, although often assumed constant in ice sheet
-modelling, $\sPDD$ exhibits significant spacial and temporal variation, most
-particularly in the polar regions (\cref{fig:plot-sdmap}; \PSDV, Fig.~1). Using
-a simplified surface mass balance model, I then demonstrated the effect of this
-variation on the computation of positive degree-days (\PSDV, Fig.~2), and
-surface mass-balance (\PSDV, Fig.~3). Neglecting this variability in ice sheet
-modelling introduces significant errors in modelled PDD and surface mass
-balance calculations. Therefore, the Cordilleran ice sheet simulations
-presented in \CCYC\ now include this effect.
-
-The analysis is extended in \PSDP, using re-analysed air
-temperatures from ERA-40 \citep{Uppala.etal.2005}. Although of coarser spatial
-resolution than the ERA-Interim reanalysis used in \PSDV, the ERA-40 time
-series span over longer time, which is necessary to compute standard deviation
-values that are representative of the long-term daily variability
-\citep{Rogozhina.Rau.2014}. In \PSDP, we demonstrated an inverse correlation
-between $\sPDD$ and monthly mean temperatures, $T_m$, over the Greenland ice
-sheet (\PSDP, Fig.~1). This correlation could potentially be used as a
-parametrisation, and can be expressed as (\PSDP, Eq.~3)
-
-\begin{equation}
-    \label{eqn:sigma}
-    \sPDD = -0.15 \cdot T_{m} + 1.66\,.
-\end{equation}
-
-The effect of the PDD standard deviation, $\sPDD$, on simulations of the
-Cordilleran ice sheet is not shown in any of the papers. However, following
-these two studies, a natural step forward was to implement, in PISM, spatially
-and seasonally variable $\sPDD$ input (PISM issue~\issue{179}), and the
-parameterisation expressed by \cref{eqn:sigma} over ice-filled grid cells
-(PISM issue~\issue{265}). Their effect on modelled Cordilleran ice sheet
-surface elevations is demonstrated here using a constant climate set-up similar
-to the one used in \CCLI\ (\cref{fig:plot-sdeffect}). These simulations show
-that the effect of daily variability is strongest in continental regions, where
-the $\sPDD$ values are highest (cf. \cref{fig:plot-sdmap}). When the results
-are compared to the mapped LGM ice limits (\cref{fig:map-cordillera}), which show
-a restricted ice extent in continental areas despite low temperatures at
-present (\cref{fig:plot-atm}), it appears that including daily temperature
-variability in the ice sheet model results in a closer agreement with these
-limits (\cref{fig:plot-sdeffect}b and~c). However, applying the parametrisation
-derived for the Greenland ice sheet (\cref{eqn:sigma}) over ice-filled grid
-cells partly reverts this effect (\cref{fig:plot-sdeffect}d), indicating that
-this parametrization may not be applicable to the former Cordilleran ice sheet.
-
 \begin{figure}
@@
 
+Next to the atmospheric forcing, another critical source of uncertainty in
+palaeo-ice sheet modelling is the choice of parameters used by the surface mass
+balance model \citep[e.g.,][]{Hebeler.etal.2008}. In our case, these parameters
+are the degree-day factors for snow, $F_s$, and ice, $F_i$
+(\cref{fig:model-pdd,tab:params}), and the standard deviation of
+near-surface air temperature, $\sPDD$
+(\cref{eqn:calovgreve,tab:params}). Although the first two can only be
+deduced from surface mass-balance measurements on present-day glaciers
+\citep[e.g.,][]{Shea.etal.2009} and assumptions that these values are also
+valid for palaeo-simulations, $\sPDD$ can be estimated from temperature time
+series \citep[e.g.,][]{Fausto.etal.2011}.
+
+In \PSDV, I used near-surface air temperatures from the ERA-Interim reanalysis
+\citep{Dee.etal.2011} to compute a global, monthly distribution of $\sPDD$. The
+resulting grids show that, although often assumed constant in ice sheet
+modelling, $\sPDD$ exhibits significant spacial and temporal variation, most
+particularly in the polar regions (\cref{fig:plot-sdmap}; \PSDV, Fig.~1). Using
+a simplified surface mass balance model, I then demonstrated the effect of this
+variation on the computation of positive degree-days (\PSDV, Fig.~2), and
+surface mass-balance (\PSDV, Fig.~3). Neglecting this variability in ice sheet
+modelling introduces significant errors in modelled PDD and surface mass
+balance calculations. Therefore, the Cordilleran ice sheet simulations
+presented in \CCYC\ now include this effect.
+
+The analysis is extended in \PSDP, using re-analysed air
+temperatures from ERA-40 \citep{Uppala.etal.2005}. Although of coarser spatial
+resolution than the ERA-Interim reanalysis used in \PSDV, the ERA-40 time
+series span over longer time, which is necessary to compute standard deviation
+values that are representative of the long-term daily variability
+\citep{Rogozhina.Rau.2014}. In \PSDP, we demonstrated an inverse correlation
+between $\sPDD$ and monthly mean temperatures, $T_m$, over the Greenland ice
+sheet (\PSDP, Fig.~1). This correlation could potentially be used as a
+parametrisation, and can be expressed as (\PSDP, Eq.~3)
+
+\begin{equation}
+    \label{eqn:sigma}
+    \sPDD = -0.15 \cdot T_{m} + 1.66\,.
+\end{equation}
+
+The effect of the PDD standard deviation, $\sPDD$, on simulations of the
+Cordilleran ice sheet is not shown in any of the papers. However, following
+these two studies, a natural step forward was to implement, in PISM, spatially
+and seasonally variable $\sPDD$ input (PISM issue~\issue{179}), and the
+parameterisation expressed by \cref{eqn:sigma} over ice-filled grid cells
+(PISM issue~\issue{265}). Their effect on modelled Cordilleran ice sheet
+surface elevations is demonstrated here using a constant climate set-up similar
+to the one used in \CCLI\ (\cref{fig:plot-sdeffect}). These simulations show
+that the effect of daily variability is strongest in continental regions, where
+the $\sPDD$ values are highest (cf. \cref{fig:plot-sdmap}). When the results
+are compared to the mapped LGM ice limits (\cref{fig:map-cordillera}), which show
+a restricted ice extent in continental areas despite low temperatures at
+present (\cref{fig:plot-atm}), it appears that including daily temperature
+variability in the ice sheet model results in a closer agreement with these
+limits (\cref{fig:plot-sdeffect}b and~c). However, applying the parametrisation
+derived for the Greenland ice sheet (\cref{eqn:sigma}) over ice-filled grid
+cells partly reverts this effect (\cref{fig:plot-sdeffect}d), indicating that
+this parametrization may not be applicable to the former Cordilleran ice sheet.
+
 % ----------------------------------------------------------------------
@@
 \bigskip
+
+\begin{figure*}
+  \centering
+  \makebox[0pt]{\includegraphics{plot-snapshots}}
+  \caption{\textbf{(a)} Modelled growth and decay of the last Cordilleran ice
+           sheet. Basal topography, surface topography (1\,km contours) and
+           surface velocity (m/a) are model results.
+           \textbf{(b)} Evolution of total ice volume during the last glacial
+           cycle, expressed as metres of sea-level equivalent (m~s.l.e.). The
+           simulation is performed on a 6\,km horizontal grid and adopts
+           atmospheric fields from the NARR \citep{Mesinger.etal.2006} and
+           temperature offset time-series derived from the GRIP ice-core record
+           \citep{Dansgaard.etal.1993} as described in \CCYC. Note that, for
+           this run, the Cordilleran ice sheet was almost as voluminous during
+           MIS~4 as it was during the LGM.}
+  \label{fig:plot-snapshots}
+\end{figure*}
 
@@
 
-\begin{figure*}
-  \centering
-  \makebox[0pt]{\includegraphics{plot-snapshots}}
-  \caption{\textbf{(a)} Modelled growth and decay of the last Cordilleran ice
-           sheet. Basal topography, surface topography (1\,km contours) and
-           surface velocity (m/a) are model results.
-           \textbf{(b)} Evolution of total ice volume during the last glacial
-           cycle, expressed as metres of sea-level equivalent (m~s.l.e.). The
-           simulation is performed on a 6\,km horizontal grid and adopts
-           atmospheric fields from the NARR \citep{Mesinger.etal.2006} and
-           temperature offset time-series derived from the GRIP ice-core record
-           \citep{Dansgaard.etal.1993} as described in \CCYC. Note that, for
-           this run, the Cordilleran ice sheet was almost as voluminous during
-           MIS~4 as it was during the LGM.}
-  \label{fig:plot-snapshots}
-\end{figure*}
-
 % ======================================================================

### 2014-08-06 kappa.tex
Remove spurious subfig whitespace and makeboxes.
Assess or remove todos.

@@
 %\newcommand{\renote}[1]{\footnote{\textbf{Comment reply}: #1}}
-\newcommand{\todo}[1]{} %\emph{[\textbf{Todo:} #1]}}
+%\newcommand{\todo}[1]{} %\emph{[\textbf{Todo:} #1]}}
 %\newcommand{\mref}[0]{\textbf{[ref.]}}
@@
 % Some general low-priority todos
-% -- switch to kappa.cls
-% -- incorporate figs as pdf, rasterize heavy elements
 % -- use latex in matplotlib figures
@@
 \begin{figure*}[b]
-  \centering
-  \makebox[0pt]{
-    \subgraphics{a}{photo-glacier-crevasses}
-    \hspace{1cm}
-    \subgraphics{b}{photo-glacier-susitna}
-    \hspace{1cm}
-    \subgraphics{c}{photo-glacier-fold}
-  }
+  \subgraphics{a}{photo-glacier-crevasses}%
+  \hspace{1cm}%
+  \subgraphics{b}{photo-glacier-susitna}%
+  \hspace{1cm}%
+  \subgraphics{c}{photo-glacier-fold}%
   \caption{Illustration of various glacier ice behaviours:
@@
 \begin{figure*}[b]
-  \centering
-  \makebox[0pt]{
-    \subgraphics{a}{photo-glacial-valley}
-    \hspace{1cm}
-    \subgraphics{b}{photo-erratic-boulder}
-    \hspace{1cm}
-    \subgraphics{c}{photo-melt-channels}
-  }
+  \subgraphics{a}{photo-glacial-valley}%
+  \hspace{1cm}%
+  \subgraphics{b}{photo-erratic-boulder}%
+  \hspace{1cm}%
+  \subgraphics{c}{photo-melt-channels}%
   \caption{Different indicators of past glaciation:
@@
 \begin{figure}
-  \centering
   \includegraphics{paleo-timeseries}
@@
 \begin{figure}
-  \centering
   \includegraphics{paleo-glaciation}
@@
 \begin{figure*}
-  \centering
-  \makebox[0pt]{\includegraphics{map-northamerica}}
+  \includegraphics{map-northamerica}
   \caption{Relief map of northern North America showing a reconstruction of the
@@
 \begin{figure*}
-  \centering
-  \makebox[0pt]{\includegraphics{map-cordillera}}
+  \includegraphics{map-cordillera}
   \caption{Topographic map of the northern American Cordillera, oriented north.
@@
 \begin{figure*}
-  \centering
-  \makebox[0pt]{\includegraphics{photo-alaska-range}}
+  \includegraphics{photo-alaska-range}
   \caption{The Alaska Range hosts several of the highest peaks in the North
@@
 \begin{figure*}
-  \centering
-  \makebox[0pt]{
-    \subgraphics{a}{photo-vancouver-island}
-    \hspace{1cm}
-    \subgraphics{b}{photo-fraser-valley}
-  }
+  \subgraphics{a}{photo-vancouver-island}%
+  \hspace{1cm}%
+  \subgraphics{b}{photo-fraser-valley}%
   \caption{Contrasted climates of the North American Cordillera:
@@
 
-\todo{Go to geolibrary and check \emph{Geology of Canada} books and such for a
-      reference about the three orogenic belts.}
-
 The North American Cordillera hosts several \emph{intermontane basins}, the
@@
 \begin{figure*}
-  \centering
-  \makebox[0pt]{\includegraphics{plot-atm}}
+  \includegraphics{plot-atm}
   \caption{Some characteristics of the Cordilleran climate according to the
@@
 between Vancouver Island and the Olympic Peninsula, where it formed the Juan
-de Fuca marine ice lobe, calving into the Pacific Ocean. Further to the
-North, the ice margin encroached the Queen Charlotte Islands. However, the
+de Fuca marine ice lobe, calving into the Pacific Ocean \citep{Herzer.Bornhold.1982}. Further to the
+North, the ice margin encroached the Queen Charlotte Islands \citep{Blaise.etal.1990}. However, the
 archipelago may not have been fully ice-covered, even during the maximum of
 glaciation. Submarine troughs traversing several portions of the continental
-shelf indicate the erosional action of additional ice streams, yet it is not
+shelf ()\cref{fig:map-cordillera}) indicate the erosional action of additional ice streams, yet it is not
 clear whether they have formed during the last glacial cycle or during earlier
 glaciations.
-
-\todo{Find refs for marine extent of the ice sheet}
 
@@
 \begin{figure}
-  \centering
-  \includegraphics[width=85mm]{model-interfaces}
+  \includegraphics[width=80mm]{model-interfaces}
   \caption{PISM is a numerical model for ice and bedrock physics. Processes
@@
 \begin{table*}
-  \centering
   \caption{Constant parameters of the ice sheet model.}
   \label{tab:params}
-  \makebox[0pt]{\begin{tabular*}{170mm}{@{\hspace{2em}}l@{\extracolsep{\fill}}lrll}
+  \begin{tabular*}{170mm}{@{\hspace{2em}}l@{\extracolsep{\fill}}lrll}
     \tophline
@@
     \bottomhline
-  \end{tabular*}}
+  \end{tabular*}
 \end{table*}
@@
 \begin{figure}
-  \centering
   \includegraphics{model-variables}
@@
 
-\todo{\citet{Morland.Johnson.1980} derived the SIA in 2D, using a stream
-      function. Their equations look very different from here. Go to
-      geolibrary and check the book by \citet{Hutter.1983}, if it is there.}
-
 Neglecting atmospheric pressure, \Cref{eqn:siaz} yields the pressure,
@@
 account for variations of ice rheology, $A$, with depth.
-
-\todo{If I have time, show that vertical velocities can be
-      obtained from the compressibility equation.}
 
@@
 \begin{figure}
-  \centering
   \includegraphics[width=80mm]{model-siassa}
@@
 \begin{figure}
-  \centering
   \includegraphics{model-pdd}
@@
 \begin{table*}
-  \centering
   \caption{Sources of forcing data used to inform the ice sheet model.}
   \label{tab:datasources}
-  \makebox[0pt]{\begin{tabular*}{170mm}{l@{\extracolsep{\fill}}lll}
+  \begin{tabular*}{170mm}{l@{\extracolsep{\fill}}lll}
     \tophline
@@
 %\begin{table}
-%  \centering
 %  \caption{Properties of the Canadian Atlas Lambert conformal conic projection
@@
 %  \label{tab:params}
-%  \makebox[0pt]{\begin{tabular*}{80mm}{@{\extracolsep{\fill}}lcc}
+%  \begin{tabular*}{80mm}{@{\extracolsep{\fill}}lcc}
 %    \hline
@@
 %    \hline
-%  \end{tabular*}}
+%  \end{tabular*}
 %\end{table}
@@
 \begin{figure*}
-  \centering
-  \makebox[0pt]{
   \begin{minipage}[t]{112.5mm}\vspace{0pt}
     \includegraphics{plot-lgm}
-  \end{minipage}
-  \hspace{1cm}
+  \end{minipage}%
+  \hspace{1cm}%
   \begin{minipage}[t]{47.5mm}\vspace{-12pt}  % kappa class add 12 pt here
@@
   \end{minipage}
-  }
 \end{figure*}
@@
 \begin{figure}
-  \centering
-  \makebox[0pt]{\includegraphics{plot-sdmap}}
+  \includegraphics{plot-sdmap}
   \caption{July, daily temperature variability in the Arctic, as shown by
@@
 \begin{figure*}
-  \centering
-  \makebox[0pt]{\includegraphics{plot-sdeffect}}
+  \includegraphics{plot-sdeffect}
   \caption{Effect of temperature variability on Cordilleran ice sheet
@@
 \begin{figure*}
-  \centering
-  \makebox[0pt]{\includegraphics{plot-snapshots}}
+  \includegraphics{plot-snapshots}
   \caption{\textbf{(a)} Modelled growth and decay of the last Cordilleran ice

### 2014-08-07 cordillera-cycle.tex
Change Fraser Plateau to Interior Plateau.

@@
 %\newcommand{\aref}[0]{\textbf{[ref.]}}
-
-% Todo
-% change Fraser Plateau to Interior Plateau
 
@@
 that inferred the ice divide in a more westerly position,
-over the western margin of the Fraser Plateau
-\citep{Ryder.etal.1991,
+over the western margin of the Interior Plateau \citep{Ryder.etal.1991,
 Stumpf.etal.2000, Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}.
@@
 southern sector of the modelling domain, including the Puget Lowland, the Coast
-and Rocky mountains, and the Fraser Plateau, becomes
+and Rocky mountains, and the Interior Plateau of British Columbia, becomes
 completely deglaciated by 12\,ka, whereas a significant ice cover remains over
@@
 deglaciation phase (Fig.~\ref{fig:lastflow}, hatched areas). This notably
-includes parts of the Fraser Plateau, major
+includes parts of the Interior Plateau in British Columbia, major
 portions of the Alaskan sector of the ice sheet, and a tortuous ribbon running
@@
 
-However, on the Fraser Plateau, both
+However, on the Interior Plateau of south-central British Columbia, both
 simulations predict negligible basal sliding during deglaciation
@@
 (Fig.~\ref{fig:lastflow}). Thus, another explanation for the incongruent
-results could be that the Fraser Plateau lineation system predates
+results could be that the Interior Plateau lineation system predates
 deglaciation, as perhaps indicated by some eskers that appear incompatible with
 these glacial lineations \citep[Fig.~9]{Margold.etal.2013a}. Alternatively,
-local geothermal heat associated with volcanic activity on the Fraser Plateau
+local geothermal heat associated with volcanic activity on the Interior Plateau
 could have triggered the basal sliding
@@
 
-The modelled deglaciation of the Fraser Plateau
+The modelled deglaciation of the Interior Plateau of British Columbia
 consists of a rapid northwards retreat (Fig.~\ref{fig:deglac}) of
@@
            by the ice sheet include the Alaska Range (AR), the Wrangell and
-           St.-Elias mountains (WSEM), the Selwyn and Mackenzie mountains (SMKM),
-           the Skeena Mountains (SM), the Coast and Rocky
+           St.-Elias mountains (WSEM), the Selwyn and Mackenzie mountains
+           (SMKM), the Skeena Mountains (SM), the Coast and Rocky
            Mountains (CRM) and the North Cascades (NC). Major depressions
-           include the Liard Lowland (LL), the Fraser Plateau (FP) and the
-           Puget Lowland (PL). The background
+           include the Liard Lowland (LL), the Interior Plateau of British
+           Columbia (IP) and the Puget Lowland (PL). The background
            map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
@@
            indicate areas that remain non-sliding throughout deglaciation
-           (22.0--8.0\,ka), notably including the Fraser Plateau (FP).
+           (22.0--8.0\,ka), notably including the Interior Plateau (IP).
            Note the concentric patterns of deglacial flow in the Liard

### 2014-08-07 figures/lastflow.py

@@ -96,7 +96,7 @@ txtkwa = dict(ha='center', va='center',
               arrowprops=dict(arrowstyle="->"))
 ax.set_rasterization_zorder(2.5)
 ax.annotate('LL', xy=(-1700e3, 1600e3), xytext=(-1100e3, 1600e3), **txtkwa)
-ax.annotate('FP', xy=(-1850e3, 900e3), xytext=(-1100e3, 900e3), **txtkwa)
+ax.annotate('IP', xy=(-1850e3, 900e3), xytext=(-1100e3, 900e3), **txtkwa)
 
 # add colorbar and save
 print 'saving...'

### 2014-08-07 figures/locmap.py

@@ -118,7 +118,7 @@ def add_names():
 
     # add names of intermontane plateaus
     ax.text(-127, 60, 'LL', **txtkwa)
-    ax.text(-124, 54, 'FL', **txtkwa)
+    ax.text(-122, 52, 'IP', **txtkwa)
     ax.text(-122.5, 47.5, 'PL', **txtkwa)
 
 # modelling domain

### 2014-08-07 cordillera-cycle.tex
Switch to print layout for thesis.
Pull figures up in the text.

@@
 % Copernicus manuscript
-\documentclass[tc, manuscript]{copernicus}
+%\documentclass[tc, manuscript]{copernicus}
 
 % Copernicus final print
-%\documentclass[tc]{copernicus}
+\documentclass[tc]{copernicus}
 
@@
 \begin{document}
-\linenumbers
 % =============================================================================
@@
 
+\begin{figure}
+  \includegraphics{locmap}
+  \caption{Relief map of northern North America showing a reconstruction of the
+           areas once covered by the Cordilleran (CIS), Laurentide (LIS),
+           Innuitian (IIS) and Greenland (GIS) ice sheets during the last
+           18\,\unit{\chem{^{14}C}\,ka} (21.4\,cal\,ka)
+           \citep{Dyke.2004}. The rectangular box denotes the location of the
+           modelling domain used in this study. Major mountain ranges covered
+           by the ice sheet include the Alaska Range (AR), the Wrangell and
+           St.-Elias mountains (WSEM), the Selwyn and Mackenzie mountains
+           (SMKM), the Skeena Mountains (SM), the Coast and Rocky
+           Mountains (CRM) and the North Cascades (NC). Major depressions
+           include the Liard Lowland (LL), the Interior Plateau of British
+           Columbia (IP) and the Puget Lowland (PL). The background
+           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
+           Data \citep{Patterson.Kelso.2014}.}
+  \label{fig:locmap}
+\end{figure}
+
 In contrast, evidence for the deglaciation history of the Cordilleran
@@
 % -----------------------------------------------------------------------------
+
+\begin{table*}
+  \centering
+  \caption{Parameter values used in the ice sheet model.}
+  \label{tab:params}
+  {\begin{tabular}{llrl}
+    \tophline
+    Not.    & Name & Value & Unit \\
+    \middlehline
+
+    $\rho$  & Ice density
+            & 910
+            & \unit{kg\,m^{-3}} \\
+
+    $g$     & Standard gravity
+            & 9.81
+            & \unit{m\,s^{-2}} \\
+
+    \multicolumn{2}{l}{\emph{Basal sliding}} \\
+
+    $q$     & Pseudo-plastic sliding exponent
+            & 0.25
+            & - \\
+
+    $v_{th}$& Pseudo-plastic threshold velocity
+            & 100.0
+            & \unit{m\,yr^{-1}} \\
+
+    $c_0$   & Till cohesion
+            & 0.0
+            & Pa \\
+
+    $\delta$& Effective pressure coefficient
+            & 0.02
+            & - \\
+
+    $e_0$   & Till reference void ratio
+            & 0.69
+            & - \\
+
+    $C_c$   & Till compressibility coefficient
+            & 0.12
+            & - \\
+
+    $W_{max}$ & Maximal till water thickness
+            & 2.0
+            & m \\
+
+    \multicolumn{2}{l}{\emph{Bedrock and lithosphere}} \\
+
+    $\rho_b$& Bedrock density
+            & 3300
+            & \unit{kg\,m^{-3}} \\
+
+    $c_b$   & Bedrock specific heat capacity
+            & 1000
+            & \unit{J\,kg^{-1}\,K^{-1}} \\
+
+    $k_b$   & Bedrock thermal conductivity
+            & 3.0
+            & \unit{J\,m^{-1}\,K^{-1}\,s^{-1}} \\
+
+    $\nu_m$ & Mantle viscosity
+            & $1\times10^21$
+            & \unit{Pa\,s} \\
+
+    $\rho_l$& Lithosphere density
+            & 3300
+            & \unit{kg\,m^{-3}} \\
+
+    $D$     & Lithosphere flexural rigidity
+            & $5.0\times10^24$
+            & \unit{N} \\
+
+    \multicolumn{2}{l}{\emph{Surface and atmosphere}} \\
+
+    $T_s$   & Temperature of snow precipitation
+            & 273.15
+            & \unit{K} \\
+
+    $T_r$   & Temperature of rain precipitation
+            & 275.15
+            & \unit{K} \\
+
+    $F_s$   & Degree-day factor for ice
+            & $3.04\times10^{-3}$
+            & \unit{m\,K^{-1}\,day^{-1}} \\
+
+    $F_i$   & Degree-day factor for ice
+            & $4.59\times10^{-3}$
+            & \unit{m\,K^{-1}\,day^{-1}} \\
+
+    $\gamma$& Air temperature lapse-rate
+            & $6\times10^{-3}$
+            & \unit{K\,m^{-1}} \\
+
+    \bottomhline
+  \end{tabular}}
+  \belowtable{}
+\end{table*}
 
@@
 
+\begin{figure}
+  \includegraphics{atm}
+  \caption{Monthly mean near-surface air temperature, precipitation and
+           standard deviation of daily mean temperature for January and July
+           from the North American Regional Reanalysis
+           \citep[NARR;][]{Mesinger.etal.2006},
+           used to force the ice sheet model. Note the
+           strong contrasts in seasonality, timing of the precipitation peak,
+           and temperature variability over the model domain, notably between
+           coastal and inland regions.}
+  \label{fig:atm}
+\end{figure}
+
 Ice surface accumulation and ablation are computed from monthly mean
@@
 variability \citep{Seguinot.2013}, $\sigma$ is computed from daily temperature
-values from the NARR, after excluding variability associated
-with the seasonal cycle itself \citep[cf.][]{Seguinot.Rogozhina.2014}.
+values from the NARR, after excluding variability associated with the seasonal
+cycle itself \citep[Fig.~\ref{fig:atm}, cf.][]{Seguinot.Rogozhina.2014}.
 Degree-day factors for snow and ice melt are derived from
@@
 
-Climate forcing of the model consists of a present-day monthly climatology,
-$\{T_{m0}, P_{m0}\}$, where temperatures are modified by offset time series,
-${\Delta}T_{TS}$, and lapse-rate corrections, ${\Delta}T_{LR}$:
-\begin{align}
-    T_m(t, x, y) &= T_{m0}(x, y) + {\Delta}T_{TS}(t)
-                    + {\Delta}T_{LR}(t, x, y) \,, \\
-    P_m(t, x, y) &= P_{m0}(x, y) \,.
-\end{align}
-
-The present-day monthly climatology was computed from
-near-surface air temperature and precipitation rate fields from the NARR,
-averaged from 1979 to 2000. Modern climate of the
-North American Cordillera is characterised by strong geographic variations in
-temperature seasonality, timing of the maximum annual precipitation, and
-daily temperature variability (Fig.~\ref{fig:atm}).
-The use of NARR is motivated by the need for an accurate,
-high-resolution precipitation forcing, as identified in a previous sensitivity
-study \citep{Seguinot.etal.2014}.
-
-Temperature offset time-series, ${\Delta}T_{TS}$, are derived from
-palaeo-temperature proxy records from
-the Greenland Ice Core Project \citep[GRIP,][]{Dansgaard.etal.1993}, the
-North Greenland Ice Core Project \citep[NGRIP,][]{Andersen.etal.2004},
-the European Project for Ice Coring in Antarctica \citep[EPICA,][]
-{Jouzel.etal.2007}, the Vostok ice core \citep{Petit.etal.1999}, and Ocean
-Drilling Program (ODP) sites 1012 and 1020, both located off the coast of
-California \citep{Herbert.etal.2001}. Palaeo-temperatures from the GRIP and
-NGRIP
-records were calculated from oxygen isotope (\chem{\delta^{18}O}) measurements
-using a quadratic equation \citep{Johnsen.etal.1995},
-\begin{equation}
-    {\Delta}T_{TS}(t) = -11.88 [\chem{\delta^{18}O}(t)
-                               -\chem{\delta^{18}O}(0)]
-                        -0.1925[\chem{\delta^{18}O}(t)^2
-                                -\chem{\delta^{18}O}(0)^2] \,,
-\end{equation}
-while temperature reconstructions from Antarctic and oceanic cores were
-provided as such. All records were scaled linearly (Table~\ref{tab:records}) in
-order to simulate comparable ice extents at the LGM  (Table~\ref{tab:extrema})
-and realistic outlines (Fig.~\ref{fig:snapshots}).
-
-Finally, lapse-rate corrections, ${\Delta}T_{LR}$, are computed as a function
-of ice surface elevation, $s$, using the NARR surface geopotential height
-invariant field as a reference topography, $b_{ref}$:
-\begin{align}
-    {\Delta}T_{LR}(t, x, y) &= -\gamma [s(t, x, y)-b_{ref}] \\
-                            &= -\gamma [h(t, x, y)+b(t, x, y)-b_{ref}],
-\end{align}
-
-thus accounting for the evolution of ice thickness, ${h=s-b}$,
-on the one hand, and for differences between the basal topography of the ice
-flow model, $b$, and the
-NARR reference topography, $b_{ref}$, on the other hand. All simulations use an
-annual temperature lapse rate of $\gamma = 6\,\unit{K\,km^{-1}}$.
-In the rest of this paper, we refer to different model runs by the name of the
-proxy record used for the palaeo-temperature forcing.
-
-% =============================================================================
-\section{Sensitivity to climate forcing time-series}
-\label{sec:results}
-% =============================================================================
-
-% -----------------------------------------------------------------------------
-\subsection{Evolution of ice volume}
-% -----------------------------------------------------------------------------
-
-Despite large differences in the input climate forcing
-(Fig.~\ref{fig:timeseries}, upper panel), model output presents consistent
-features that can be observed across the range of forcing data used. In all
-simulations, modelled ice volumes remain relatively low during most of the
-glacial cycle, except during two major glacial events which occur between 61.9
-and 55.4\,ka during MIS~4, and between 29.5 and 16.9\,ka during MIS~2
-(Fig.~\ref{fig:timeseries}, lower panel). An ice volume minimum is
-consistently reached between 52.2 and 45.6\,ka during MIS~3. However, the
-magnitude and precise timing of these three events depend significantly on the
-choice of proxy record used for time-dependent climate forcing
-(Table~\ref{tab:extrema}).
-
-Simulations forced by the Greenland ice core palaeo-temperature
-records (GRIP, NGRIP) produce the highest variability in modelled ice volume
-throughout the last glacial cycle. In contrast, simulations driven by oceanic
-(ODP~1012, ODP~1020) and Antarctic (EPICA, Vostok) palaeo-temperature records
-generally result in smaller
-modelled ice volumes during MIS~4 and larger ice volumes during MIS~3.
-Hence, they produce lower ice volume variability
-throughout the simulation length. The NGRIP climate forcing is the only one
-that results in a larger ice volume during MIS~4 than during MIS~2.
-
-While simulations driven by the GRIP and the two Antarctic palaeo-temperature
-records attain a last ice volume maximum between 19.5 and 16.9\,ka, those
-informed by the NGRIP and the two oceanic palaeo-temperature records attain their
-maximum ice volumes thousands of years earlier. Moreover, the ODP~1012 run
-yields a rapid deglaciation of the modelled area prior to 20\,ka. The ODP~1020
-simulation predicts an early maximum in ice volume at 29.5\,ka, followed by
-slower deglaciation than modelled using the other palaeo-temperature records.
-Finally, model runs forced by Antarctic palaeo-temperature records result in a
-rapid and uninterrupted deglaciation after the last glacial maximum, whereas the
-simulation driven by the GRIP palaeo-temperature record also results in a rapid
-deglaciation but in three steps, separated by two ice volume standstills
-(Fig.~\ref{fig:timeseries}).
-
-% -----------------------------------------------------------------------------
-\subsection{Extreme configurations}
-% -----------------------------------------------------------------------------
-
-Despite such differences in the timing of attained volume extrema
-(Table~\ref{tab:extrema}), all the model runs show relatively consistent
-patterns of glaciation. During MIS~4, all simulations produce an extensive ice
-sheet, covering an area of at least half of that attained during MIS~2
-(Table~\ref{tab:extrema}; Fig.~\ref{fig:snapshots}, upper panels).
-Corresponding maximum ice volumes also differ significantly between model runs,
-and vary between 3.88 and~8.69\,m sea level equivalents (m.~s.l.e;
-Table~\ref{tab:extrema}).
-
-During MIS~3 ice volume minima, a central ice cap persists over the Skeena
-Mountains (Fig.~\ref{fig:snapshots}, middle panels). Although this ice cap is
-present in all simulations, its dimensions depend sensitively on the choice of
-the applied palaeo-temperature record. Modelled ice volume minima spread over a
-wide range between 0.89 and 2.44\,m~s.l.e. (Table~\ref{tab:extrema}).
-
-Modelled ice sheet geometries during the LGM (MIS~2; Fig.~\ref{fig:snapshots},
-lower panels) invariably include a ca. 1500\,km-long central divide located
-about 3500\,m~a.s.l. along the spine of the Rocky Mountains. Although the
-similatity of modelled ice extents is a direct result from the choice of scaling
-factors applied to different palaeo-temperature proxy records
-(Table~\ref{tab:records}), it is interesting to note that modelled maximum ice
-volumes also fall within a tight range of 8.24 to 8.66\,m~s.l.e.
-(Table~\ref{tab:extrema}).
-
-% =============================================================================
-\section{Comparison to the geologic record}
-\label{sec:discussion}
-% =============================================================================
-
-Large variations in the model response reveal its sensitivity to evolving
-climate forcing. Nevertheless, geological evidence of former glaciations
-provide a basis for validation of our runs, while the results from numerical
-modelling can perhaps help to analyse some of the complexity of this evidence.
-In this section, we compare model output to the geologic record, in terms of
-timing and configuration of the maximum stages, location and durability of the
-major nucleation centres, and patterns of ice retreat during the last
-deglaciation.
-
-% -----------------------------------------------------------------------------
-\subsection{Glacial maxima}
-% -----------------------------------------------------------------------------
-
-\subsubsection{Timing of glaciation}
-\label{sec:timing}
-
-Independently of the palaeo-temperature records
-used to force the ice sheet model, our simulations consistently produce two
-glacial maxima during the last glacial cycle. The first maximum configuration
-is obtained during MIS~4 (61.9--55.4\,ka) and the second during MIS~2
-(29.5--16.9\,ka; Figs.~\ref{fig:timeseries}, \ref{fig:snapshots};
-Table~\ref{tab:extrema}). These events broadly correspond in timing to the
-Gladstone
-(MIS~4) and McConnell (MIS~2) glaciations documented by geological evidence for
-the northern sector of the Cordilleran ice sheet
-    \citep{Duk-Rodkin.etal.1996, Ward.etal.2007,
-           Stroeven.etal.2010, Stroeven.etal.2014},
-and to the Fraser Glaciation (MIS~2) documented for its southern sector
-    \citep{Porter.Swanson.1998, Margold.etal.2014}.
-There is patchy stratigraphical evidence for glaciations older than the Fraser
-Glaciation \citep{Clague.Ward.2011} in British Columbia, but their extent and
-timing are still highly conjectural
-    \citep[perhaps MIS~4 or early MIS~3; e.g.,][]{Cosma.etal.2008}.
-
-The exact timing of modelled MIS~2 maximum ice volume depends strongly on the
-choice of applied palaeo-temperature record, which allows for a more in-depth
-comparison with geological evidence for the timing of maximum Cordilleran ice
-sheet extent. In the Puget lowland (Fig.~\ref{fig:locmap}), the LGM advance of
-the southern Cordilleran ice sheet margin has been constrained by radiocarbon
-dating on wood between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,ka}
-\citep{Porter.Swanson.1998}. These dates are consistent with radiocarbon dates
-from the offshore sedimentary record, which reveals
-an increase of glaciomarine sedimentation between 19.5 and
-16.2\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008}. Radiocarbon
-dating of the northern Cordilleran ice sheet margin is much less constrained
-but straddles presented constraints from the southern margin. However,
-cosmogenic exposure dating
-places the timing of maximum CIS extent during the McConnell glaciation close
-to 17\,ka \citep{Stroeven.etal.2010, Stroeven.etal.2014}.
-
-Among the simulations presented here, only those forced with the GRIP, EPICA
-and Vostok palaeo-temperature records yield Cordilleran ice sheet maximum
-extents that may be compatible with these field constraints
-(Fig.~\ref{fig:timeseries}, lower panel; Table~\ref{tab:extrema}).
-Simulations driven by the NGRIP, ODP~1012 and ODP~1020
-palaeo-temperature records, on the contrary, yield MIS~2 maximum Cordilleran
-ice sheet extents that pre-date field-based constraints by several thousands of
-years. Concerning the simulations driven by oceanic records, this early
-deglaciation is caused by an early warming present in the alkenone
-palaeo-temperature reconstructions (Fig.~\ref{fig:timeseries}, upper panel;
-\citealp[Fig.~3]{Herbert.etal.2001}). However, this
-early warming is a local effect, corresponding to a weakening of the California
-current \citep{Herbert.etal.2001}. The California current, driving cold
-waters southwards along the south-western coast of North America,
-has been shown to have weakened during each peak global glaciation (in SPECMAP)
-during the past 550\,ka, including the LGM, resulting in paradoxically warmer
-sea-surface temperatures at the locations of the ODP~1012 and ODP~1020 sites
-\citep{Herbert.etal.2001}.
-
-Because most of the marine margin of the Cordilleran ice sheet terminated in a
-sector of the Pacific Ocean unaffected by variations in the California current,
-it probably remained insensitive to this local phenomenum. However, the above
-paradox
-illustrates the complexity of ice-sheet feedbacks on regional climate, and
-demonstrates that, although located in the neighbourhood of the modelling
-domain, the ODP~1012 and ODP~1020 palaeo-temperature records cannot be
-used as a realistic forcing to model the Cordilleran ice sheet
-through the last glacial cycle.
-
-Similarly, the simulation using the NGRIP palaeo-temperature record depicts an
-early onset of deglaciation (Fig.~\ref{fig:timeseries}) following its last
-glacial volume maximum attained about 6\,ka earlier (-22.9\,ka,
-Table~\ref{tab:extrema}) than dated evidence of the last glacial maximum
-advance. Hence, we focus the rest of our analysis on simulations forced by
-palaeo-temperature records from the GRIP and EPICA ice cores that appear to
-produce the most realistic reconstructions of regional glaciation history among
-all records tested. To allow for a more
-detailed comparison against the geologic record, these two simulations were
-re-run using the high-resolution grid (Fig.~\ref{fig:timeseries}, lower panel,
-dotted lines).
-
-\subsubsection{Ice configuration during MIS~2}
-\label{sec:mis2}
-
-During maximum glaciation, both simulations position the main meridional ice
-divide over
-the western flank of the Rocky Mountains (Fig.~\ref{fig:snapshots}, lower
-panels; Fig.~\ref{fig:mis2}). This result appears to contrast with
-palaeoglaciological reconstructions for central and southern British Columbia
-that inferred the ice divide in a more westerly position,
-over the western margin of the Interior Plateau \citep{Ryder.etal.1991,
-Stumpf.etal.2000, Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}.
-These idincate that a latitudinal saddle connected ice dispersal centres in
-the Columbia Mountains with the main ice divide \citep{Ryder.etal.1991,
-Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}. A latitudinal saddle
-does indeed feature in our modelling results, however, in an inverse
-configuration between the main ice divide over the Columbia Mountains and a
-secondary divide over the southern Coast Mountains (Fig.~\ref{fig:mis2}).
-
-This could reflect the fact that our model does not include
-feedback mechanisms between ice sheet topography and the regional climate.
-Firstly, during the build-up phase preceding the LGM, rapid accumulation over
-the Coast Mountains enhanced the topographic barrier formed by these mountain
-ranges, which likely resulted in a decrease of precipitation and, therefore,
-accumulation in the interior. Secondly, latent warming of the moisture-depleted
-air parcels flowing over this enhanced topography could have resulted in an
-inflow of warmer air over the eastern flank of the ice sheet, increasing melt
-along the advancing margin \citep[cf.][]{Langen.etal.2012}. Because these two
-processes, both with a tendency to limit of ice-sheet growth, are absent from
-our model, the
-eastern margin of the ice sheet and the position of the main meridional ice
-divide are certainly biased towards the east in our simulations
-\citep{Seguinot.etal.2014}.
-
-However, field-based palaeoglaciological reconstructions have
-struggled to reconcile the more westerly-centred ice divide in south-central
-British Columbia with evidence in the Rocky Mountains and beyond, that the
-Cordilleran ice sheet invaded the western Interior Plains, where it merged with
-the southwestern margin of the Laurentide ice sheet and was deflected to the
-south \citep{Jackson.etal.1997, Bednarski.Smith.2007, Kleman.etal.2010,
-Margold.etal.2013, Margold.etal.2013a}. Ice geometries from our model runs do
-not have this problem, because the position and elevation of the ice divide
-results in significant ice drainage across the Rocky Mountains at the LGM
-(Fig.~\ref{fig:mis2}).
-
-\subsubsection{Ice configuration during MIS~4}
-\label{sec:mis4}
-
-The modelled ice sheet configurations corresponding to ice volume maxima during
-MIS~4 are more sensitive to the choice of atmospheric forcing than those
-corresponding to ice volume maxima during MIS~2 (Fig.~\ref{fig:snapshots},
-upper panels;
-Fig.~\ref{fig:mis4}). The GRIP simulation (Fig.~\ref{fig:mis4}, left panel)
-results in a modelled maximum ice sheet extent that closely resembles that
-obtained
-during MIS~2, with the only major difference being slightly less extensive
-northern and eastern sectors. In contrast, the EPICA simulation produces a lower
-ice volume maximum (Fig.~\ref{fig:timeseries}), which translates in the
-modelled ice sheet geometry into a significantly reduced southern sector, more
-restricted ice cover in northern and eastern sectors, and generally lower
-ice surface elevations in the interior (Fig.~\ref{fig:mis4}, left panel).
-
-In the GRIP simulation, some sections of the MIS~4 maximum ice margin are
-slightly more extensive than their MIS~2 counterparts. This is notably the case
-in the Puget Lowland and along parts of the ice margin in the marine sectors,
-on the north slope of the Alaska Range and in Yukon Territory. However, these
-differences
-are small and generally correspond to a few grid cells only. Although such
-results may be corroborated by field evidence in some places
-\citep[e.g.][]{Ward.etal.2007}, considering the simplicity of the applied
-climate forcing, modelled differences in extent are not demonstrably
-significant.
-
-% -----------------------------------------------------------------------------
-\subsection{Nucleation centres}
-% -----------------------------------------------------------------------------
-
-\subsubsection{Transient ice sheet states}
-
-Palaeo-glaciological reconstructions are generally more robust
-for maximum ice sheet extents and late ice sheet configurations than for
-intermediate or minimum ice sheet extents and older ice sheet configurations
-\citep{Kleman.etal.2010}. However,
-these maximum stages are, by nature, extreme configurations, which do not
-necessarily represent the dominant patterns of glaciation throughout the period
-of ice cover \citep{Porter.1989, Kleman.Stroeven.1997}.
-
-For the Cordilleran ice sheet, geological evidence from radiocarbon dating
-    \citep{Clague.etal.1980, Clague.1985, Clague.1986, Porter.Swanson.1998,
-           Menounos.etal.2008},
-cosmogenic exposure dating
-    \citep{Stroeven.etal.2010, Stroeven.etal.2014, Margold.etal.2014},
-bedrock deformation in response to former ice loads
-    \citep{Clague.James.2002, Clague.etal.2005},
-and offshore sedimentary records
-    \citep{Cosma.etal.2008, Davies.etal.2011}
-indicate that the LGM maximum extent was short-lived. To compare this finding
-to our simulations, we use numerical modelling output
-to compute durations of ice cover throughout the last glacial
-cycle (Fig.~\ref{fig:duration}).
-
-The resulting maps show that, during most of the glacial cycle, modelled ice
-cover is restricted to disjoint ice caps centred on major mountain ranges of
-the North American Cordillera (Fig.~\ref{fig:duration}, blue areas). A
-2500\,km-long continuous expanse of ice, extending from the Alaska Range in the
-northeast to the Rocky Mountains in the southwest, is only in operation
-for at most 29\,ka, which is less than 25\,\% of a glacial cycle
-(Fig.~\ref{fig:duration}, hatched areas). However,
-except for its marine margin and the northern foothills of the Alaska Range,
-the maximum extent of the ice sheet is attained for an even shorter period
-of time of few thousand years (Figs.~\ref{fig:duration}, red areas). This
-result illustrates that the maximum extents of the modelled ice sheet during
-MIS~4 and MIS~2 were both short-lived and therefore out of balance with
-contemporary climate.
-
-A notable exception to the transient character of other sectors of the
-Cordilleran ice sheet
-is the northern slope of the Alaska Range, where modelled glaciers are confined
-to its foothills during the entire simulation period (Fig.~\ref{fig:duration},
-AR). This apparent insensitivity of modelled glacial extent to temperature
-fluctuations results from a combination regimes of low precipitation, high summer
-temperature and large temperature standard deviation for this sector of the
-modelling domain (Fig.~\ref{fig:atm}). This result could potentially explain
-the local distribution of glacial deposits, which indicates that these glaciers
-have remained small throughout the Pleistocene \citep{Kaufman.Manley.2004}.
-
-\subsubsection{Major ice-dispersal centres}
-
-It is generally believed that the Cordilleran ice sheet formed by the
-coalescence of several mountain-centred ice caps \citep{Davis.Mathews.1944}.
-In our simulations, major ice-dispersal centres, visible on the modelled ice
-cover duration maps (Fig.~\ref{fig:duration}), are located over the Coast
-Mountains (CM), the Columbia and Rocky mountains (CRM), the Skeena Mountains
-(SM), and the Selwyn and Mackenzie Mountains (SMKM). Although the Coast,
-Columbia, Rocky and Skeena mountains (CM, CRM, SM) are covered by mountain
-glaciers for most of the last glacial cycle, providing durable nucleation
-centres for an ice sheet, this is not the case for the Selwyn and
-Mackenzie mountains (SMKM), where ice cover on the highest peaks is limited to a
-fraction of the last glacial cycle. In other words, the Selwyn and Mackenzie
-mountains appear as a secondary ice-dispersal centre which ``activates'' only
-during the coldest periods of the last glacial cycle. The Northern Rocky
-Mountains (Fig.~\ref{fig:duration}, NRM), do not act as a nucleation centre,
-but rather as a pinning point for the Cordilleran ice sheet margin coming from
-the west.
-
-Perhaps the most striking feature displayed by the distributions of modelled
-ice cover is the persistence of the Skeena Mountains ice cap throughout the
-entire last glacial cycle and its predominance over the other ice-dispersal
-centres
-(Figs.~\ref{fig:snapshots} and~\ref{fig:duration}). Regardless of the applied
-forcing, this ice cap appears to survive MIS~3 (Fig.~\ref{fig:snapshots},
-middle panels), and serves as a nucleation centre at the onset of the glacial
-readvance towards the LGM (MIS~2). The importance of residual ice for North
-American glacial history leading up to the LGM has been illustrated by the
-MIS~3 residual ice bodies in northern and eastern Canada as nucleation centres
-for a much more extensive MIS~2 configuration of the Laurentide ice sheet
-\citep{Kleman.etal.2010}.
-
-The presence of a Skeena Mountains ice cap during most of the last glacial
-cycle can be explained by a currently more eastwards-spread distribution of
-winter precipitation there than along other transects of the
-modelling domain (Fig.~\ref{fig:atm}). Along most of the north-western coast of
-North America, coastal mountain ranges form a pronounced topographic barrier
-for westerly winds, capturing atmospheric moisture in the form of orographic
-precipitation, and resulting in arid interior lowlands. However, near the
-centre of our modelling domain, this barrier is less pronounced than elsewhere,
-allowing westerly winds to carry moisture
-further inland, until it is captured by the extensive Skeena Mountains in
-north-central British Columbia, thus resulting in a more widespread
-distribution of winter precipitation.
-
-\subsubsection{Erosional imprint on the landscape}
-
-A correlation is observed between the modelled duration of warm based ice cover
-(Fig.~\ref{fig:warmbase}) and the degree of glacial modification of the
-landscape (mainly in terms of the development of deep glacial valleys and
-troughs). We find evidence for this on both the southwestern and northeastern
-slopes of the southern Coast Mountains, on the western slopes of the Columbia
-Mountains, on the western and eastern slopes of the northern Coast Mountains
-and the Saint-Elias Mountains, and radiating off the Skeena Mountains
-(Figs.~\ref{fig:duration} and \ref{fig:warmbase};
-\citealp[Fig.~2]{Kleman.etal.2010}).
-The Skeena Mountains, for example, indeed bear a strong glacial imprint that
-indicates ice drainage in a system of distinct glacial troughs emanating in a
-radial pattern from the centre of the mountain range
-\citep[Fig.~2]{Kleman.etal.2010}. We suggest that
-persistent ice cover (Fig.~\ref{fig:duration}) associated with basal ice
-temperatures at the pressure-melting point (Figs.~\ref{fig:warmfrac}
-and~\ref{fig:warmbase}) explains the large-scale glacial erosional imprint on
-the landscape. A well-developed network of glacial valleys west of
-the Mackenzie Mountains (\citealp[Fig.~2]{Kleman.etal.2010}; \citealp[Fig.~8]
-{Stroeven.etal.2010}) is modelled to have hosted warm-based ice
-(Fig.~\ref{fig:warmfrac}). However, because it has only been glaciated for a
-short fraction
-of the last glacial cycle, (Fig.~\ref{fig:duration}) according to our
-results, this perhaps appears indicate that
-the observed landscape pattern originates from multiple glacial cycles and
-witnesses an increased relative importance of this ice dispersal centre prior
-to the Late Pleistocene \citep[cf.][]{Ward.etal.2008, Demuro.etal.2012}.
-
-The modelled distribution of warm-based ice cover (Figs.~\ref{fig:warmbase}
-and~\ref{fig:warmfrac}) is inevitably affected by our assumption of a constant,
-70\,\unit{mW\,m^{-2}} geothermal heat flux at 3\,km depth
-(Sect.~\ref{sec:icedyn}). However, both the Skeena Mountains and the area
-west of the Mackenzie Mountains are areas of higher-than-average geothermal
-heat flux \citep{Blackwell.Richards.2004}. We can therefore expect longer
-durations of warm-based ice cover for these areas if we were to include
-spatially variable geothermal forcing in our Cordilleran ice sheet simulations.
-
-% -----------------------------------------------------------------------------
-\subsection{The last deglaciation}
-% -----------------------------------------------------------------------------
-
-\subsubsection{Pace and patterns of deglaciation}
-
-Similarly to other glaciated regions, most glacial traces in the North American
-Cordillera relate to the last few millennia of glaciation, because most of the
-older evidence has been overprinted by warm-based ice retreat
-during the deglaciation \citep{Kleman.1994, Kleman.etal.2010}. From a
-numerical modelling perspective, phases of glacier retreat are more challenging
-than phases of growth, because they involve more rapid fluctuations of the ice
-margin, increased flow velocities and longitudinal stress gradients, and poorly
-understood hydrological processes. The latter are typically included in the
-models through simple parametrisations \citep[e.g.][]{Clason.etal.2012,
-Clason.etal.2014, Bueler.Pelt.2014}, if included at all. However, next after a
-matching of modelled ice margin positions with mapped end moraines
-\citep{Napieralski.etal.2007}, indicative of maximum ice sheet extents
-(Sect.~\ref{sec:mis2} and~\ref{sec:mis4}), geomorphologically-based
-reconstructions of patterns of of ice sheet retreat during the last
-deglaciation provide the second best source of evidence for the validation of
-our simulations.
-
-In the North American Cordillera, the presence of lateral meltwater channels
-at high elevation \citep{Margold.etal.2011, Margold.etal.2013a,
-Margold.etal.2014}, and abundant esker systems at low elevation
-\citep{Burke.etal.2012, Burke.etal.2012a, Perkins.etal.2013,
-Margold.etal.2013}
-indicate that abundant meltwater was produced over the entire ice sheet surface
-during deglaciation. The southern and northern margins of the
-Cordilleran ice sheet reached their last glacial maximum extent around
-17\,ka \citep[Sect.~\ref{sec:timing};][]{Porter.Swanson.1998, Cosma.etal.2008,
-Stroeven.etal.2010, Stroeven.etal.2014}, which we take as a limiting age for
-the onset of ice retreat. The timing of final deglaciation is less well
-constrained, but recent cosmogenic dates from north-central British Columbia
-indicate that a seizable ice cap emanating from the central Coast Mountains or
-the Skeena Mountains persisted into the Younger Dryas chronozone, at least
-until 12.4\,ka \citep{Margold.etal.2014}.
-
-In our simulations, the timing of peak ice volume during the last glacial
-maximum and the pacing of deglaciation depend critically on the choice of
-climate forcing (Table~\ref{tab:extrema}, Figs.~\ref{fig:timeseries}
-and~\ref{fig:deglacseries}). Adopting the EPICA
-climate forcing yields peak ice volume at 17.3\,ka and an uninterrupted
-deglaciation until 10.6\,ka (Fig.~\ref{fig:deglacseries}, lower panel, red
-curves). The simulation driven by the GRIP palaeo-temperature record yields
-peak ice volume at 19.1\,ka and a deglaciation interrupted by two ice volume
-standstills until 9.4\,ka. The first interruption occurs between 16.6 and
-14.5\,ka, and the second between 12.6 and 11.6\,ka
-(Fig.~\ref{fig:deglacseries}, lower panel, blue curve). Hence, the two model
-runs, while similar in overall timing compared to runs with other climate
-drivers, differ in detail in that the EPICA depicts peak glaciation almost
-2\,ka later than the GRIP, in closer agreement with dated maximum extents, and
-shows a faster, uninterrupted deglaciation which yields ice-free conditions
-more than 1\,ka earlier.
-
-Modelled patterns of ice sheet retreat are relatively consistent between the
-two simulations (Figs.~\ref{fig:deglacshots} and.~\ref{fig:deglac}). The
-southern sector of the modelling domain, including the Puget Lowland, the Coast
-and Rocky mountains, and the Interior Plateau of British Columbia, becomes
-completely deglaciated by 12\,ka, whereas a significant ice cover remains over
-the Skeena, Selwyn, Mackenzie, Wrangell and Saint-Elias mountains in the
-northern sector of the modelling domain. After 12\,ka, deglaciation continues
-to proceed across the Liard Lowland with a radial ice margin retreat towards
-the surrounding mountain ranges, consistent with the regional melt water record
-of the last deglaciation \citep{Margold.etal.2013}. Remaining ice continues to
-decay by retreating towards the Selwyn and Skeena mountains. The last remnants
-of the Cordilleran ice sheet finally disappear from the Skeena Mountains around
-10.6\,ka (EPICA) and 9.4\,ka (GRIP).
-
-\subsubsection{Late-glacial readvance}
-
-The possibility of late glacial readvances in the North American Cordillera has
-been debated for some time \citep{Osborn.Gerloff.1997}, and locally these have
-been reconstructed and dated. Radiocarbon-dated end moraines in the Fraser and
-Squamish valleys, off the southern tip of the Coast Mountains, indicate
-consecutive glacier maxima, or standstills while in overall retreat, one of
-which corresponds to the Younger Dryas chronozone \citep{Clague.etal.1997,
-Friele.Clague.2002, Friele.Clague.2002a, Kovanen.2002,
-Kovanen.Easterbrook.2002}. Although most of these moraines characterise
-independent valley glaciers, that were probably disconnected from the waning
-Cordilleran ice sheet, the Finlay River
-area of the northern Rocky Mountains presents a different kind of evidence.
-There, sharp-crested moraines indicate a late-glacial readvance of local alpine
-glaciers and, more importantly, their interaction with larger, lingering
-remnants of the main body of the Cordilleran ice sheet in the valleys
-\citep{Lakeman.etal.2008}. Additional evidence for late-glacial
-alpine glacier readvances includes moraines in the eastern Coast Mountains,
-Rocky Mountains and the Columbia Mountains \citep{Osborn.Gerloff.1997,
-Menounos.etal.2008}.
-
-Although further work is needed to constrain the timing of the late-glacial
-readvance, to assess its extents and geographical distribution, and to identify
-the potential climatic triggers \citep{Menounos.etal.2008}, it is interesting
-to note that the simulation driven by the GRIP record produces a late-glacial
-readvance in the Coast Mountains, Rocky Mountains and the Finlay River area,
-corresponding to where it has been identified in the geological record
-(Fig.~\ref{fig:deglac}, left panel). In addition to matching the location of
-local readvances, the GRIP-driven simulation shows that larger ice bodies
-emanating from the Skeena and Mackenzie mountains may still have existed at the
-time of this late-glacial readvance.
-In contrast, the EPICA-driven simulation produces a nearly-continuous
-deglaciation with only a tightly restricted glacial readvance
-(Fig.~\ref{fig:deglac}, right panel).
-
-\subsubsection{Deglacial flow directions}
-
-Because a general tenant in glacial geomorphology is that the majority of
-landforms (lineations and eskers) are part of the deglacial envelope
-\citep[terminology from][]{Kleman.etal.2006}, having been formed
-close inside the retreating margin of ice sheets \citep{Boulton.Clark.1990,
-Kleman.etal.1997, Kleman.etal.2010}, we present maps of basal flow directions
-immediately preceding deglaciation or at the time of cessation of sliding
-inside a cold-based retreating margin (Fig.~\ref{fig:lastflow}). The
-modelled deglacial flow patterns appear consistent between the GRIP and EPICA
-simulations. They depict an active ice sheet retreat in the peripheral areas,
-followed by stagnant ice decay in some of the interior regions. Several parts
-of the modelling domain do not experience any basal sliding throughout the
-deglaciation phase (Fig.~\ref{fig:lastflow}, hatched areas). This notably
-includes parts of the Interior Plateau in British Columbia, major
-portions of the Alaskan sector of the ice sheet, and a tortuous ribbon running
-from the Northern Rocky Mountains over the Skeena and Selwyn Mountains and into
-the Mackenzie Mountains.
-
-Patterns of glacial lineations formed in the northern and southern sectors of
-the Cordilleran ice sheet (\citealp{Prest.etal.1968};
-\citealp[Fig.~1.12]{Clague.1989}; \citealp[Fig.~2]{Kleman.etal.2010}) show
-similarities with the patterns of deglacial ice flow from numerical modelling
-(Fig.~\ref{fig:lastflow}). In the northern half
-of the modelling domain, modelled deglacial flow directions depict an
-active downhill flow as the last remnants of the ice sheet retreat towards
-mountain ranges. Converging deglacial flow patterns in the Liard
-Lowland, for instance (Fig.~\ref{fig:lastflow}), closely resemble the pattern
-indicated by glacial lineations \citep[Fig.~2]{Margold.etal.2013}.
-
-However, on the Interior Plateau of south-central British Columbia, both
-simulations predict negligible basal sliding during deglaciation
-(Fig.~\ref{fig:lastflow}), but an impressive set of glacial lineations
-indicate a substantial eastward flow component of the Cordilleran ice sheet
-\citep{Prest.etal.1968, Kleman.etal.2010}.
-The Fraser lineation set could thus present a smoking gun for the
-reliability of the presented model results. One explanation for the incongruent
-results could therefore be that missing feedback mechanisms between ice sheet
-topography and regional climate position the modelled ice divide of the LGM ice
-sheet too far to the east (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
-\citealp{Seguinot.etal.2014}). A more westerly-located LGM ice divide would
-certainly result in a different deglacial flow pattern over the Interior
-Plateau. However, a more westerly-positioned LGM ice divide would certainly be
-associated with a thinner ice sheet than that modelled here. Decreased ice
-thickness would not promote warm-based conditions but, on the
-opposite, enlarge the region of negligible basal sliding
-(Fig.~\ref{fig:lastflow}). Thus, another explanation for the incongruent
-results could be that the Interior Plateau lineation system predates
-deglaciation, as perhaps indicated by some eskers that appear incompatible with
-these glacial lineations \citep[Fig.~9]{Margold.etal.2013a}. Alternatively,
-local geothermal heat associated with volcanic activity on the Interior Plateau
-could have triggered the basal sliding
-\citep[cf. Greenland ice sheet,][]{Fahnestock.etal.2001}.
-
-The modelled deglaciation of the Interior Plateau of British Columbia
-consists of a rapid northwards retreat (Fig.~\ref{fig:deglac}) of
-southwards-flowing non-sliding ice lobes (Fig.~\ref{fig:lastflow}) positioned
-in-between deglaciated mountain ranges
-(Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}). This result
-appears compatible with the prevailing conceptual model of deglaciation of
-central British Columbia, in which mountain ranges emerge from the ice before
-the plateau \citep[Fig.~7]{Fulton.1991}.
-
-% =============================================================================
-\conclusions
-\label{sec:concl}
-% =============================================================================
-
-Numerical simulations of the Cordilleran ice sheet through the last glacial
-cycle presented in this study consistently produce two glacial maxima during
-MIS~4 (61.9--55.4\,ka, 3.8--8.7\,m~s.l.e) and MIS~2 (29.5--16.9\,ka,
-8.24--8.66\,m~s.l.e.), two periods corresponding to documented extensive
-glaciations. This result is independent of the palaeo-temperature record used
-among the six selected for this study, and thus can be seen as a robust model
-output, which broadly matches geological evidence. However, the timing of the
-two glaciation peaks depends sensitively on which climate record is used to
-drive the model. The timing of the LGM is best reproduced by the EPICA and
-Vostok Antarctic ice core records, and occurs too early in all other
-simulations. The mismatch is largest for the two northwest Pacific ODP
-palaeo-temperature records, which are affected by the weakening of the
-California current during the LGM.
-
-In all simulations presented here, ice cover is limited to disjoint mountain
-ice caps during most of the glacial cycle. The most persistent nucleation
-centres are located in the Coast Mountains, the Columbia and Rocky mountains,
-the Selwyn and Mackenzie mountains, and most importantly, in the Skeena
-Mountains. Throughout the modelled last glacial cycle, the Skeena Mountains
-host an ice cap which appears to be fed by moisture intruding inland from the
-west through a topographic low in the Coast Mountains. The Skeena ice cap acts
-as the main nucleation centre at the onset of the LGM readvance. As indicated
-by persistent, warm-based ice in the model, this ice cap is perhaps associated
-with the distinct glacial erosional imprint on the landscape of the Skeena
-Mountains.
-
-During deglaciation, none of the climate records used can be selected as
-producing an optimal agreement between the model results and the geological
-evidence. Although the EPICA-driven simulation yields the most realistic timing
-of the LGM and, therefore, start of deglaciation, only the GRIP-driven
-simulation produces late glacial readvances in areas where these have been
-documented. Nonetheless, the patterns of ice sheet retreat are consistent
-between the two simulations, and show a rapid deglaciation of the southern
-sector of the ice sheet, including a rapid northwards retreat across the Fraser
-Plateau of central British Columbia. The GRIP-driven simulation then produces a
-late-glacial readvance of local ice caps in the Coast, Columbia and Rocky
-mountains, and of the main body of the decaying Cordilleran ice sheet in the
-Skeena and Selwyn mountains. In both simulations, this is followed by an
-opening of the Liard Lowland, and a final retreat of the remaining ice caps
-towards the Selwyn and, finally, the Skeena mountains, which hosts the last
-remnant of the ice sheet during the early Holocene (10.9--9.5\,ka). Our results
-identify the Skeena Mountains as a key area to understanding glacial dynamics
-of the Cordilleran ice sheet, highlighting the need for further geological
-investigation of this region.
-
-% Author contributions
-\section*{Author contributions}
-J.~Seguinot ran the simulations and wrote most of the manuscript; I.~Rogozhina
-guided experiment design; A.~P.~Stroeven, M.~Margold and J.~Kleman took part in
-the interpretation and comparison of model results against geological evidence.
-All co-authors contributed to improve the text.
-
-% Acknowledgements
-\begin{acknowledgements}
-We are very thankful to Constantine Khroulev, Ed Bueler and Andy Aschwanden for
-providing constant help and development on PISM. This work was supported by the
-Swedish Research Council~(VR) grant no. 2008-3449 to A.~P.~Stroeven, by the
-German Academic Exchange Service~(DAAD) grant no.~50015537 and a Knut and Alice
-Wallenberg Foundation grant to J.~Seguinot.
-Computer resources were provided by the Swedish National
-Infrastructure for Computing (SNIC) allocation no. 2013/1-159 to A.~P.~Stroeven
-at the National Supercomputing Center (NSC).
-\end{acknowledgements}
-
-% References
-\bibliographystyle{copernicus}
-\bibliography{refs/references.bib}
-\newpage
-
-% =============================================================================
-% Floats
-% =============================================================================
-
-% tab:params
-\begin{table*}
-  \centering
-  \caption{Parameter values used in the ice sheet model.}
-  \label{tab:params}
-  {\begin{tabular}{llrl}
-    \tophline
-    Not.    & Name & Value & Unit \\
-    \middlehline
-
-    $\rho$  & Ice density
-            & 910
-            & \unit{kg\,m^{-3}} \\
-
-    $g$     & Standard gravity
-            & 9.81
-            & \unit{m\,s^{-2}} \\
-
-    \multicolumn{2}{l}{\emph{Basal sliding}} \\
-
-    $q$     & Pseudo-plastic sliding exponent
-            & 0.25
-            & - \\
-
-    $v_{th}$& Pseudo-plastic threshold velocity
-            & 100.0
-            & \unit{m\,yr^{-1}} \\
-
-    $c_0$   & Till cohesion
-            & 0.0
-            & Pa \\
-
-    $\delta$& Effective pressure coefficient
-            & 0.02
-            & - \\
-
-    $e_0$   & Till reference void ratio
-            & 0.69
-            & - \\
-
-    $C_c$   & Till compressibility coefficient
-            & 0.12
-            & - \\
-
-    $W_{max}$ & Maximal till water thickness
-            & 2.0
-            & m \\
-
-    \multicolumn{2}{l}{\emph{Bedrock and lithosphere}} \\
-
-    $\rho_b$& Bedrock density
-            & 3300
-            & \unit{kg\,m^{-3}} \\
-
-    $c_b$   & Bedrock specific heat capacity
-            & 1000
-            & \unit{J\,kg^{-1}\,K^{-1}} \\
-
-    $k_b$   & Bedrock thermal conductivity
-            & 3.0
-            & \unit{J\,m^{-1}\,K^{-1}\,s^{-1}} \\
-
-    $\nu_m$ & Mantle viscosity
-            & $1\times10^21$
-            & \unit{Pa\,s} \\
-
-    $\rho_l$& Lithosphere density
-            & 3300
-            & \unit{kg\,m^{-3}} \\
-
-    $D$     & Lithosphere flexural rigidity
-            & $5.0\times10^24$
-            & \unit{N} \\
-
-    \multicolumn{2}{l}{\emph{Surface and atmosphere}} \\
-
-    $T_s$   & Temperature of snow precipitation
-            & 273.15
-            & \unit{K} \\
-
-    $T_r$   & Temperature of rain precipitation
-            & 275.15
-            & \unit{K} \\
-
-    $F_s$   & Degree-day factor for ice
-            & $3.04\times10^{-3}$
-            & \unit{m\,K^{-1}\,day^{-1}} \\
-
-    $F_i$   & Degree-day factor for ice
-            & $4.59\times10^{-3}$
-            & \unit{m\,K^{-1}\,day^{-1}} \\
-
-    $\gamma$& Air temperature lapse-rate
-            & $6\times10^{-3}$
-            & \unit{K\,m^{-1}} \\
-
-    \bottomhline
-  \end{tabular}}
-  \belowtable{}
-\end{table*}
-
-% tab:records
 \begin{table*}[t]
@@
 
-% tab:extrema
-\begin{table*}[t]
+Climate forcing of the model consists of a present-day monthly climatology,
+$\{T_{m0}, P_{m0}\}$, where temperatures are modified by offset time series,
+${\Delta}T_{TS}$, and lapse-rate corrections, ${\Delta}T_{LR}$:
+\begin{align}
+    T_m(t, x, y) &= T_{m0}(x, y) + {\Delta}T_{TS}(t)
+                    + {\Delta}T_{LR}(t, x, y) \,, \\
+    P_m(t, x, y) &= P_{m0}(x, y) \,.
+\end{align}
+
+The present-day monthly climatology was computed from
+near-surface air temperature and precipitation rate fields from the NARR,
+averaged from 1979 to 2000. Modern climate of the
+North American Cordillera is characterised by strong geographic variations in
+temperature seasonality, timing of the maximum annual precipitation, and
+daily temperature variability (Fig.~\ref{fig:atm}).
+The use of NARR is motivated by the need for an accurate,
+high-resolution precipitation forcing, as identified in a previous sensitivity
+study \citep{Seguinot.etal.2014}.
+
+Temperature offset time-series, ${\Delta}T_{TS}$, are derived from
+palaeo-temperature proxy records from
+the Greenland Ice Core Project \citep[GRIP,][]{Dansgaard.etal.1993}, the
+North Greenland Ice Core Project \citep[NGRIP,][]{Andersen.etal.2004},
+the European Project for Ice Coring in Antarctica \citep[EPICA,][]
+{Jouzel.etal.2007}, the Vostok ice core \citep{Petit.etal.1999}, and Ocean
+Drilling Program (ODP) sites 1012 and 1020, both located off the coast of
+California \citep{Herbert.etal.2001}. Palaeo-temperatures from the GRIP and
+NGRIP
+records were calculated from oxygen isotope (\chem{\delta^{18}O}) measurements
+using a quadratic equation \citep{Johnsen.etal.1995},
+\begin{align}
+    {\Delta}T_{TS}(t) = &-11.88 [\chem{\delta^{18}O}(t)
+                                -\chem{\delta^{18}O}(0)] \nonumber \\
+                        &-0.1925[\chem{\delta^{18}O}(t)^2
+                                 -\chem{\delta^{18}O}(0)^2] \,,
+\end{align}
+while temperature reconstructions from Antarctic and oceanic cores were
+provided as such. All records were scaled linearly (Table~\ref{tab:records}) in
+order to simulate comparable ice extents at the LGM  (Table~\ref{tab:extrema})
+and realistic outlines (Fig.~\ref{fig:snapshots}).
+
+Finally, lapse-rate corrections, ${\Delta}T_{LR}$, are computed as a function
+of ice surface elevation, $s$, using the NARR surface geopotential height
+invariant field as a reference topography, $b_{ref}$:
+\begin{align}
+    {\Delta}T_{LR}(t, x, y) &= -\gamma [s(t, x, y)-b_{ref}] \\
+                            &= -\gamma [h(t, x, y)+b(t, x, y)-b_{ref}],
+\end{align}
+
+thus accounting for the evolution of ice thickness, ${h=s-b}$,
+on the one hand, and for differences between the basal topography of the ice
+flow model, $b$, and the
+NARR reference topography, $b_{ref}$, on the other hand. All simulations use an
+annual temperature lapse rate of $\gamma = 6\,\unit{K\,km^{-1}}$.
+In the rest of this paper, we refer to different model runs by the name of the
+proxy record used for the palaeo-temperature forcing.
+
+% =============================================================================
+\section{Sensitivity to climate forcing time-series}
+\label{sec:results}
+% =============================================================================
+
+% -----------------------------------------------------------------------------
+\subsection{Evolution of ice volume}
+% -----------------------------------------------------------------------------
+
+\begin{figure*}
+  \includegraphics{timeseries}
+  \caption{Temperature offset time-series from ice core and ocean records
+           (Table~\ref{tab:records}) used as palaeo-climate forcing for the ice
+           sheet model (top panel), and modelled ice volume (bottom panel)
+           through the last 120\,ka. Ice volumes are expressed in meters of sea
+           level equivalent (m~s.l.e.). Gray fields indicate Marine Oxygen
+           Isotope Stage (MIS) boundaries for MIS~2 and MIS~4 according to a
+           global compilation of benthic \chem{\delta^{18}O} records
+           \citep{Lisiecki.Raymo.2005}. Hatched rectangles highlight the
+           time-volume space for ice volume extremes corresponding to MIS~4
+           (61.9--55.4\,ka), MIS~3 (52.2--45.6\,ka), and MIS~2 (LGM,
+           29.5--16.9\,ka). Dotted lines correspond to GRIP- and EPICA-driven
+           6\,km-resolution runs.}
+  \label{fig:timeseries}
+\end{figure*}
+
+\begin{figure*}
+  \includegraphics{snapshots}
+  \caption{Snapshots of modelled surface topography (500\,m contours)
+           corresponding to the ice volume extremes indicated on
+           Fig.~\ref{fig:timeseries}. An ice cap persists over the Skeena
+           Mountains (SM) during MIS~3. Note the occurence of spatial
+           similarities despite large differences in timing.}
+  \label{fig:snapshots}
+\end{figure*}
+
+\begin{table*}
   \caption{Extremes in Cordilleran ice sheet volume and extent corresponding to
@@
 
-% avoid "too many unprocessed floats" errors
-\clearpage
-
-% fig:locmap
-\begin{figure}
-  \includegraphics{locmap}
-  \caption{Relief map of northern North America showing a reconstruction of the
-           areas once covered by the Cordilleran (CIS), Laurentide (LIS),
-           Innuitian (IIS) and Greenland (GIS) ice sheets during the last
-           18\,\unit{\chem{^{14}C}\,ka} (21.4\,cal\,ka)
-           \citep{Dyke.2004}. The rectangular box denotes the location of the
-           modelling domain used in this study. Major mountain ranges covered
-           by the ice sheet include the Alaska Range (AR), the Wrangell and
-           St.-Elias mountains (WSEM), the Selwyn and Mackenzie mountains
-           (SMKM), the Skeena Mountains (SM), the Coast and Rocky
-           Mountains (CRM) and the North Cascades (NC). Major depressions
-           include the Liard Lowland (LL), the Interior Plateau of British
-           Columbia (IP) and the Puget Lowland (PL). The background
-           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
-           Data \citep{Patterson.Kelso.2014}.}
-  \label{fig:locmap}
-\end{figure}
-
-% fig:atm
-\begin{figure}
-  \includegraphics{atm}
-  \caption{Monthly mean near-surface air temperature, precipitation and
-           standard deviation of daily mean temperature for January and July
-           from the North American Regional Reanalysis
-           \citep[NARR;][]{Mesinger.etal.2006},
-           used to force the ice sheet model. Note the
-           strong contrasts in seasonality, timing of the precipitation peak,
-           and temperature variability over the model domain, notably between
-           coastal and inland regions.}
-  \label{fig:atm}
-\end{figure}
-
-% fig:timeseries
-\begin{figure*}
-  \includegraphics{timeseries}
-  \caption{Temperature offset time-series from ice core and ocean records
-           (Table~\ref{tab:records}) used as palaeo-climate forcing for the ice
-           sheet model (top panel), and modelled ice volume (bottom panel)
-           through the last 120\,ka. Ice volumes are expressed in meters of sea
-           level equivalent (m~s.l.e.). Gray fields indicate Marine Oxygen
-           Isotope Stage (MIS) boundaries for MIS~2 and MIS~4 according to a
-           global compilation of benthic \chem{\delta^{18}O} records
-           \citep{Lisiecki.Raymo.2005}. Hatched rectangles highlight the
-           time-volume space for ice volume extremes corresponding to MIS~4
-           (61.9--55.4\,ka), MIS~3 (52.2--45.6\,ka), and MIS~2 (LGM,
-           29.5--16.9\,ka). Dotted lines correspond to GRIP- and EPICA-driven
-           6\,km-resolution runs.}
-  \label{fig:timeseries}
-\end{figure*}
-
-% fig:snapshots
-\begin{figure*}
-  \includegraphics{snapshots}
-  \caption{Snapshots of modelled surface topography (500\,m contours)
-           corresponding to the ice volume extremes indicated on
-           Fig.~\ref{fig:timeseries}. An ice cap persists over the Skeena
-           Mountains (SM) during MIS~3. Note the occurence of spatial
-           similarities despite large differences in timing.}
-  \label{fig:snapshots}
-\end{figure*}
-
-% fig:mis2
+Despite large differences in the input climate forcing
+(Fig.~\ref{fig:timeseries}, upper panel), model output presents consistent
+features that can be observed across the range of forcing data used. In all
+simulations, modelled ice volumes remain relatively low during most of the
+glacial cycle, except during two major glacial events which occur between 61.9
+and 55.4\,ka during MIS~4, and between 29.5 and 16.9\,ka during MIS~2
+(Fig.~\ref{fig:timeseries}, lower panel). An ice volume minimum is
+consistently reached between 52.2 and 45.6\,ka during MIS~3. However, the
+magnitude and precise timing of these three events depend significantly on the
+choice of proxy record used for time-dependent climate forcing
+(Table~\ref{tab:extrema}).
+
+Simulations forced by the Greenland ice core palaeo-temperature
+records (GRIP, NGRIP) produce the highest variability in modelled ice volume
+throughout the last glacial cycle. In contrast, simulations driven by oceanic
+(ODP~1012, ODP~1020) and Antarctic (EPICA, Vostok) palaeo-temperature records
+generally result in smaller
+modelled ice volumes during MIS~4 and larger ice volumes during MIS~3.
+Hence, they produce lower ice volume variability
+throughout the simulation length. The NGRIP climate forcing is the only one
+that results in a larger ice volume during MIS~4 than during MIS~2.
+
+While simulations driven by the GRIP and the two Antarctic palaeo-temperature
+records attain a last ice volume maximum between 19.5 and 16.9\,ka, those
+informed by the NGRIP and the two oceanic palaeo-temperature records attain their
+maximum ice volumes thousands of years earlier. Moreover, the ODP~1012 run
+yields a rapid deglaciation of the modelled area prior to 20\,ka. The ODP~1020
+simulation predicts an early maximum in ice volume at 29.5\,ka, followed by
+slower deglaciation than modelled using the other palaeo-temperature records.
+Finally, model runs forced by Antarctic palaeo-temperature records result in a
+rapid and uninterrupted deglaciation after the last glacial maximum, whereas the
+simulation driven by the GRIP palaeo-temperature record also results in a rapid
+deglaciation but in three steps, separated by two ice volume standstills
+(Fig.~\ref{fig:timeseries}).
+
+% -----------------------------------------------------------------------------
+\subsection{Extreme configurations}
+% -----------------------------------------------------------------------------
+
+Despite such differences in the timing of attained volume extrema
+(Table~\ref{tab:extrema}), all the model runs show relatively consistent
+patterns of glaciation. During MIS~4, all simulations produce an extensive ice
+sheet, covering an area of at least half of that attained during MIS~2
+(Table~\ref{tab:extrema}; Fig.~\ref{fig:snapshots}, upper panels).
+Corresponding maximum ice volumes also differ significantly between model runs,
+and vary between 3.88 and~8.69\,m sea level equivalents (m.~s.l.e;
+Table~\ref{tab:extrema}).
+
+During MIS~3 ice volume minima, a central ice cap persists over the Skeena
+Mountains (Fig.~\ref{fig:snapshots}, middle panels). Although this ice cap is
+present in all simulations, its dimensions depend sensitively on the choice of
+the applied palaeo-temperature record. Modelled ice volume minima spread over a
+wide range between 0.89 and 2.44\,m~s.l.e. (Table~\ref{tab:extrema}).
+
+Modelled ice sheet geometries during the LGM (MIS~2; Fig.~\ref{fig:snapshots},
+lower panels) invariably include a ca. 1500\,km-long central divide located
+about 3500\,m~a.s.l. along the spine of the Rocky Mountains. Although the
+similatity of modelled ice extents is a direct result from the choice of scaling
+factors applied to different palaeo-temperature proxy records
+(Table~\ref{tab:records}), it is interesting to note that modelled maximum ice
+volumes also fall within a tight range of 8.24 to 8.66\,m~s.l.e.
+(Table~\ref{tab:extrema}).
+
+% =============================================================================
+\section{Comparison to the geologic record}
+\label{sec:discussion}
+% =============================================================================
+
+Large variations in the model response reveal its sensitivity to evolving
+climate forcing. Nevertheless, geological evidence of former glaciations
+provide a basis for validation of our runs, while the results from numerical
+modelling can perhaps help to analyse some of the complexity of this evidence.
+In this section, we compare model output to the geologic record, in terms of
+timing and configuration of the maximum stages, location and durability of the
+major nucleation centres, and patterns of ice retreat during the last
+deglaciation.
+
+% -----------------------------------------------------------------------------
+\subsection{Glacial maxima}
+% -----------------------------------------------------------------------------
+
+\subsubsection{Timing of glaciation}
+\label{sec:timing}
+
+Independently of the palaeo-temperature records
+used to force the ice sheet model, our simulations consistently produce two
+glacial maxima during the last glacial cycle. The first maximum configuration
+is obtained during MIS~4 (61.9--55.4\,ka) and the second during MIS~2
+(29.5--16.9\,ka; Figs.~\ref{fig:timeseries}, \ref{fig:snapshots};
+Table~\ref{tab:extrema}). These events broadly correspond in timing to the
+Gladstone
+(MIS~4) and McConnell (MIS~2) glaciations documented by geological evidence for
+the northern sector of the Cordilleran ice sheet
+    \citep{Duk-Rodkin.etal.1996, Ward.etal.2007,
+           Stroeven.etal.2010, Stroeven.etal.2014},
+and to the Fraser Glaciation (MIS~2) documented for its southern sector
+    \citep{Porter.Swanson.1998, Margold.etal.2014}.
+There is patchy stratigraphical evidence for glaciations older than the Fraser
+Glaciation \citep{Clague.Ward.2011} in British Columbia, but their extent and
+timing are still highly conjectural
+    \citep[perhaps MIS~4 or early MIS~3; e.g.,][]{Cosma.etal.2008}.
+
+The exact timing of modelled MIS~2 maximum ice volume depends strongly on the
+choice of applied palaeo-temperature record, which allows for a more in-depth
+comparison with geological evidence for the timing of maximum Cordilleran ice
+sheet extent. In the Puget lowland (Fig.~\ref{fig:locmap}), the LGM advance of
+the southern Cordilleran ice sheet margin has been constrained by radiocarbon
+dating on wood between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,ka}
+\citep{Porter.Swanson.1998}. These dates are consistent with radiocarbon dates
+from the offshore sedimentary record, which reveals
+an increase of glaciomarine sedimentation between 19.5 and
+16.2\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008}. Radiocarbon
+dating of the northern Cordilleran ice sheet margin is much less constrained
+but straddles presented constraints from the southern margin. However,
+cosmogenic exposure dating
+places the timing of maximum CIS extent during the McConnell glaciation close
+to 17\,ka \citep{Stroeven.etal.2010, Stroeven.etal.2014}.
+
+Among the simulations presented here, only those forced with the GRIP, EPICA
+and Vostok palaeo-temperature records yield Cordilleran ice sheet maximum
+extents that may be compatible with these field constraints
+(Fig.~\ref{fig:timeseries}, lower panel; Table~\ref{tab:extrema}).
+Simulations driven by the NGRIP, ODP~1012 and ODP~1020
+palaeo-temperature records, on the contrary, yield MIS~2 maximum Cordilleran
+ice sheet extents that pre-date field-based constraints by several thousands of
+years. Concerning the simulations driven by oceanic records, this early
+deglaciation is caused by an early warming present in the alkenone
+palaeo-temperature reconstructions (Fig.~\ref{fig:timeseries}, upper panel;
+\citealp[Fig.~3]{Herbert.etal.2001}). However, this
+early warming is a local effect, corresponding to a weakening of the California
+current \citep{Herbert.etal.2001}. The California current, driving cold
+waters southwards along the south-western coast of North America,
+has been shown to have weakened during each peak global glaciation (in SPECMAP)
+during the past 550\,ka, including the LGM, resulting in paradoxically warmer
+sea-surface temperatures at the locations of the ODP~1012 and ODP~1020 sites
+\citep{Herbert.etal.2001}.
+
+Because most of the marine margin of the Cordilleran ice sheet terminated in a
+sector of the Pacific Ocean unaffected by variations in the California current,
+it probably remained insensitive to this local phenomenum. However, the above
+paradox
+illustrates the complexity of ice-sheet feedbacks on regional climate, and
+demonstrates that, although located in the neighbourhood of the modelling
+domain, the ODP~1012 and ODP~1020 palaeo-temperature records cannot be
+used as a realistic forcing to model the Cordilleran ice sheet
+through the last glacial cycle.
+
+Similarly, the simulation using the NGRIP palaeo-temperature record depicts an
+early onset of deglaciation (Fig.~\ref{fig:timeseries}) following its last
+glacial volume maximum attained about 6\,ka earlier (-22.9\,ka,
+Table~\ref{tab:extrema}) than dated evidence of the last glacial maximum
+advance. Hence, we focus the rest of our analysis on simulations forced by
+palaeo-temperature records from the GRIP and EPICA ice cores that appear to
+produce the most realistic reconstructions of regional glaciation history among
+all records tested. To allow for a more
+detailed comparison against the geologic record, these two simulations were
+re-run using the high-resolution grid (Fig.~\ref{fig:timeseries}, lower panel,
+dotted lines).
+
+\subsubsection{Ice configuration during MIS~2}
+\label{sec:mis2}
+
 \begin{figure*}
@@
 
-% fig:mis4
 \begin{figure*}
@@
 
-% fig:duration
+During maximum glaciation, both simulations position the main meridional ice
+divide over
+the western flank of the Rocky Mountains (Fig.~\ref{fig:snapshots}, lower
+panels; Fig.~\ref{fig:mis2}). This result appears to contrast with
+palaeoglaciological reconstructions for central and southern British Columbia
+that inferred the ice divide in a more westerly position,
+over the western margin of the Interior Plateau \citep{Ryder.etal.1991,
+Stumpf.etal.2000, Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}.
+These idincate that a latitudinal saddle connected ice dispersal centres in
+the Columbia Mountains with the main ice divide \citep{Ryder.etal.1991,
+Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}. A latitudinal saddle
+does indeed feature in our modelling results, however, in an inverse
+configuration between the main ice divide over the Columbia Mountains and a
+secondary divide over the southern Coast Mountains (Fig.~\ref{fig:mis2}).
+
+This could reflect the fact that our model does not include
+feedback mechanisms between ice sheet topography and the regional climate.
+Firstly, during the build-up phase preceding the LGM, rapid accumulation over
+the Coast Mountains enhanced the topographic barrier formed by these mountain
+ranges, which likely resulted in a decrease of precipitation and, therefore,
+accumulation in the interior. Secondly, latent warming of the moisture-depleted
+air parcels flowing over this enhanced topography could have resulted in an
+inflow of warmer air over the eastern flank of the ice sheet, increasing melt
+along the advancing margin \citep[cf.][]{Langen.etal.2012}. Because these two
+processes, both with a tendency to limit of ice-sheet growth, are absent from
+our model, the
+eastern margin of the ice sheet and the position of the main meridional ice
+divide are certainly biased towards the east in our simulations
+\citep{Seguinot.etal.2014}.
+
+However, field-based palaeoglaciological reconstructions have
+struggled to reconcile the more westerly-centred ice divide in south-central
+British Columbia with evidence in the Rocky Mountains and beyond, that the
+Cordilleran ice sheet invaded the western Interior Plains, where it merged with
+the southwestern margin of the Laurentide ice sheet and was deflected to the
+south \citep{Jackson.etal.1997, Bednarski.Smith.2007, Kleman.etal.2010,
+Margold.etal.2013, Margold.etal.2013a}. Ice geometries from our model runs do
+not have this problem, because the position and elevation of the ice divide
+results in significant ice drainage across the Rocky Mountains at the LGM
+(Fig.~\ref{fig:mis2}).
+
+\subsubsection{Ice configuration during MIS~4}
+\label{sec:mis4}
+
+The modelled ice sheet configurations corresponding to ice volume maxima during
+MIS~4 are more sensitive to the choice of atmospheric forcing than those
+corresponding to ice volume maxima during MIS~2 (Fig.~\ref{fig:snapshots},
+upper panels;
+Fig.~\ref{fig:mis4}). The GRIP simulation (Fig.~\ref{fig:mis4}, left panel)
+results in a modelled maximum ice sheet extent that closely resembles that
+obtained
+during MIS~2, with the only major difference being slightly less extensive
+northern and eastern sectors. In contrast, the EPICA simulation produces a lower
+ice volume maximum (Fig.~\ref{fig:timeseries}), which translates in the
+modelled ice sheet geometry into a significantly reduced southern sector, more
+restricted ice cover in northern and eastern sectors, and generally lower
+ice surface elevations in the interior (Fig.~\ref{fig:mis4}, left panel).
+
+In the GRIP simulation, some sections of the MIS~4 maximum ice margin are
+slightly more extensive than their MIS~2 counterparts. This is notably the case
+in the Puget Lowland and along parts of the ice margin in the marine sectors,
+on the north slope of the Alaska Range and in Yukon Territory. However, these
+differences
+are small and generally correspond to a few grid cells only. Although such
+results may be corroborated by field evidence in some places
+\citep[e.g.][]{Ward.etal.2007}, considering the simplicity of the applied
+climate forcing, modelled differences in extent are not demonstrably
+significant.
+
+% -----------------------------------------------------------------------------
+\subsection{Nucleation centres}
+% -----------------------------------------------------------------------------
+
+\subsubsection{Transient ice sheet states}
+
 \begin{figure*}
@@
 
-% fig:warmbase
+Palaeo-glaciological reconstructions are generally more robust
+for maximum ice sheet extents and late ice sheet configurations than for
+intermediate or minimum ice sheet extents and older ice sheet configurations
+\citep{Kleman.etal.2010}. However,
+these maximum stages are, by nature, extreme configurations, which do not
+necessarily represent the dominant patterns of glaciation throughout the period
+of ice cover \citep{Porter.1989, Kleman.Stroeven.1997}.
+
+For the Cordilleran ice sheet, geological evidence from radiocarbon dating
+    \citep{Clague.etal.1980, Clague.1985, Clague.1986, Porter.Swanson.1998,
+           Menounos.etal.2008},
+cosmogenic exposure dating
+    \citep{Stroeven.etal.2010, Stroeven.etal.2014, Margold.etal.2014},
+bedrock deformation in response to former ice loads
+    \citep{Clague.James.2002, Clague.etal.2005},
+and offshore sedimentary records
+    \citep{Cosma.etal.2008, Davies.etal.2011}
+indicate that the LGM maximum extent was short-lived. To compare this finding
+to our simulations, we use numerical modelling output
+to compute durations of ice cover throughout the last glacial
+cycle (Fig.~\ref{fig:duration}).
+
+The resulting maps show that, during most of the glacial cycle, modelled ice
+cover is restricted to disjoint ice caps centred on major mountain ranges of
+the North American Cordillera (Fig.~\ref{fig:duration}, blue areas). A
+2500\,km-long continuous expanse of ice, extending from the Alaska Range in the
+northeast to the Rocky Mountains in the southwest, is only in operation
+for at most 29\,ka, which is less than 25\,\% of a glacial cycle
+(Fig.~\ref{fig:duration}, hatched areas). However,
+except for its marine margin and the northern foothills of the Alaska Range,
+the maximum extent of the ice sheet is attained for an even shorter period
+of time of few thousand years (Figs.~\ref{fig:duration}, red areas). This
+result illustrates that the maximum extents of the modelled ice sheet during
+MIS~4 and MIS~2 were both short-lived and therefore out of balance with
+contemporary climate.
+
+A notable exception to the transient character of other sectors of the
+Cordilleran ice sheet
+is the northern slope of the Alaska Range, where modelled glaciers are confined
+to its foothills during the entire simulation period (Fig.~\ref{fig:duration},
+AR). This apparent insensitivity of modelled glacial extent to temperature
+fluctuations results from a combination regimes of low precipitation, high summer
+temperature and large temperature standard deviation for this sector of the
+modelling domain (Fig.~\ref{fig:atm}). This result could potentially explain
+the local distribution of glacial deposits, which indicates that these glaciers
+have remained small throughout the Pleistocene \citep{Kaufman.Manley.2004}.
+
+\subsubsection{Major ice-dispersal centres}
+
+It is generally believed that the Cordilleran ice sheet formed by the
+coalescence of several mountain-centred ice caps \citep{Davis.Mathews.1944}.
+In our simulations, major ice-dispersal centres, visible on the modelled ice
+cover duration maps (Fig.~\ref{fig:duration}), are located over the Coast
+Mountains (CM), the Columbia and Rocky mountains (CRM), the Skeena Mountains
+(SM), and the Selwyn and Mackenzie Mountains (SMKM). Although the Coast,
+Columbia, Rocky and Skeena mountains (CM, CRM, SM) are covered by mountain
+glaciers for most of the last glacial cycle, providing durable nucleation
+centres for an ice sheet, this is not the case for the Selwyn and
+Mackenzie mountains (SMKM), where ice cover on the highest peaks is limited to a
+fraction of the last glacial cycle. In other words, the Selwyn and Mackenzie
+mountains appear as a secondary ice-dispersal centre which ``activates'' only
+during the coldest periods of the last glacial cycle. The Northern Rocky
+Mountains (Fig.~\ref{fig:duration}, NRM), do not act as a nucleation centre,
+but rather as a pinning point for the Cordilleran ice sheet margin coming from
+the west.
+
+Perhaps the most striking feature displayed by the distributions of modelled
+ice cover is the persistence of the Skeena Mountains ice cap throughout the
+entire last glacial cycle and its predominance over the other ice-dispersal
+centres
+(Figs.~\ref{fig:snapshots} and~\ref{fig:duration}). Regardless of the applied
+forcing, this ice cap appears to survive MIS~3 (Fig.~\ref{fig:snapshots},
+middle panels), and serves as a nucleation centre at the onset of the glacial
+readvance towards the LGM (MIS~2). The importance of residual ice for North
+American glacial history leading up to the LGM has been illustrated by the
+MIS~3 residual ice bodies in northern and eastern Canada as nucleation centres
+for a much more extensive MIS~2 configuration of the Laurentide ice sheet
+\citep{Kleman.etal.2010}.
+
+The presence of a Skeena Mountains ice cap during most of the last glacial
+cycle can be explained by a currently more eastwards-spread distribution of
+winter precipitation there than along other transects of the
+modelling domain (Fig.~\ref{fig:atm}). Along most of the north-western coast of
+North America, coastal mountain ranges form a pronounced topographic barrier
+for westerly winds, capturing atmospheric moisture in the form of orographic
+precipitation, and resulting in arid interior lowlands. However, near the
+centre of our modelling domain, this barrier is less pronounced than elsewhere,
+allowing westerly winds to carry moisture
+further inland, until it is captured by the extensive Skeena Mountains in
+north-central British Columbia, thus resulting in a more widespread
+distribution of winter precipitation.
+
+\subsubsection{Erosional imprint on the landscape}
+
 \begin{figure*}
@@
 
-% fig:warmfrac
 \begin{figure*}
+  \vspace{-0.5mm}  % enough to allow the two figures to fit on one page
   \includegraphics{warmfrac}
@@
 
-% fig:deglacseries
+A correlation is observed between the modelled duration of warm based ice cover
+(Fig.~\ref{fig:warmbase}) and the degree of glacial modification of the
+landscape (mainly in terms of the development of deep glacial valleys and
+troughs). We find evidence for this on both the southwestern and northeastern
+slopes of the southern Coast Mountains, on the western slopes of the Columbia
+Mountains, on the western and eastern slopes of the northern Coast Mountains
+and the Saint-Elias Mountains, and radiating off the Skeena Mountains
+(Figs.~\ref{fig:duration} and \ref{fig:warmbase};
+\citealp[Fig.~2]{Kleman.etal.2010}).
+The Skeena Mountains, for example, indeed bear a strong glacial imprint that
+indicates ice drainage in a system of distinct glacial troughs emanating in a
+radial pattern from the centre of the mountain range
+\citep[Fig.~2]{Kleman.etal.2010}. We suggest that
+persistent ice cover (Fig.~\ref{fig:duration}) associated with basal ice
+temperatures at the pressure-melting point (Figs.~\ref{fig:warmfrac}
+and~\ref{fig:warmbase}) explains the large-scale glacial erosional imprint on
+the landscape. A well-developed network of glacial valleys west of
+the Mackenzie Mountains (\citealp[Fig.~2]{Kleman.etal.2010}; \citealp[Fig.~8]
+{Stroeven.etal.2010}) is modelled to have hosted warm-based ice
+(Fig.~\ref{fig:warmfrac}). However, because it has only been glaciated for a
+short fraction
+of the last glacial cycle, (Fig.~\ref{fig:duration}) according to our
+results, this perhaps appears indicate that
+the observed landscape pattern originates from multiple glacial cycles and
+witnesses an increased relative importance of this ice dispersal centre prior
+to the Late Pleistocene \citep[cf.][]{Ward.etal.2008, Demuro.etal.2012}.
+
+The modelled distribution of warm-based ice cover (Figs.~\ref{fig:warmbase}
+and~\ref{fig:warmfrac}) is inevitably affected by our assumption of a constant,
+70\,\unit{mW\,m^{-2}} geothermal heat flux at 3\,km depth
+(Sect.~\ref{sec:icedyn}). However, both the Skeena Mountains and the area
+west of the Mackenzie Mountains are areas of higher-than-average geothermal
+heat flux \citep{Blackwell.Richards.2004}. We can therefore expect longer
+durations of warm-based ice cover for these areas if we were to include
+spatially variable geothermal forcing in our Cordilleran ice sheet simulations.
+
+% -----------------------------------------------------------------------------
+\subsection{The last deglaciation}
+% -----------------------------------------------------------------------------
+
+\subsubsection{Pace and patterns of deglaciation}
+
+Similarly to other glaciated regions, most glacial traces in the North American
+Cordillera relate to the last few millennia of glaciation, because most of the
+older evidence has been overprinted by warm-based ice retreat
+during the deglaciation \citep{Kleman.1994, Kleman.etal.2010}. From a
+numerical modelling perspective, phases of glacier retreat are more challenging
+than phases of growth, because they involve more rapid fluctuations of the ice
+margin, increased flow velocities and longitudinal stress gradients, and poorly
+understood hydrological processes. The latter are typically included in the
+models through simple parametrisations \citep[e.g.][]{Clason.etal.2012,
+Clason.etal.2014, Bueler.Pelt.2014}, if included at all. However, next after a
+matching of modelled ice margin positions with mapped end moraines
+\citep{Napieralski.etal.2007}, indicative of maximum ice sheet extents
+(Sect.~\ref{sec:mis2} and~\ref{sec:mis4}), geomorphologically-based
+reconstructions of patterns of of ice sheet retreat during the last
+deglaciation provide the second best source of evidence for the validation of
+our simulations.
+
 \begin{figure}
@@
 
-% fig:deglacshots
 \begin{figure*}
@@
 
-% fig:deglac
 \begin{figure*}
+  \vspace{-1.5mm}  % enough to allow the two figures to fit on one page
   \includegraphics{deglac}
@@
 
-% fig:lastflow
+In the North American Cordillera, the presence of lateral meltwater channels
+at high elevation \citep{Margold.etal.2011, Margold.etal.2013a,
+Margold.etal.2014}, and abundant esker systems at low elevation
+\citep{Burke.etal.2012, Burke.etal.2012a, Perkins.etal.2013,
+Margold.etal.2013}
+indicate that abundant meltwater was produced over the entire ice sheet surface
+during deglaciation. The southern and northern margins of the
+Cordilleran ice sheet reached their last glacial maximum extent around
+17\,ka \citep[Sect.~\ref{sec:timing};][]{Porter.Swanson.1998, Cosma.etal.2008,
+Stroeven.etal.2010, Stroeven.etal.2014}, which we take as a limiting age for
+the onset of ice retreat. The timing of final deglaciation is less well
+constrained, but recent cosmogenic dates from north-central British Columbia
+indicate that a seizable ice cap emanating from the central Coast Mountains or
+the Skeena Mountains persisted into the Younger Dryas chronozone, at least
+until 12.4\,ka \citep{Margold.etal.2014}.
+
+In our simulations, the timing of peak ice volume during the last glacial
+maximum and the pacing of deglaciation depend critically on the choice of
+climate forcing (Table~\ref{tab:extrema}, Figs.~\ref{fig:timeseries}
+and~\ref{fig:deglacseries}). Adopting the EPICA
+climate forcing yields peak ice volume at 17.3\,ka and an uninterrupted
+deglaciation until 10.6\,ka (Fig.~\ref{fig:deglacseries}, lower panel, red
+curves). The simulation driven by the GRIP palaeo-temperature record yields
+peak ice volume at 19.1\,ka and a deglaciation interrupted by two ice volume
+standstills until 9.4\,ka. The first interruption occurs between 16.6 and
+14.5\,ka, and the second between 12.6 and 11.6\,ka
+(Fig.~\ref{fig:deglacseries}, lower panel, blue curve). Hence, the two model
+runs, while similar in overall timing compared to runs with other climate
+drivers, differ in detail in that the EPICA depicts peak glaciation almost
+2\,ka later than the GRIP, in closer agreement with dated maximum extents, and
+shows a faster, uninterrupted deglaciation which yields ice-free conditions
+more than 1\,ka earlier.
+
+Modelled patterns of ice sheet retreat are relatively consistent between the
+two simulations (Figs.~\ref{fig:deglacshots} and.~\ref{fig:deglac}). The
+southern sector of the modelling domain, including the Puget Lowland, the Coast
+and Rocky mountains, and the Interior Plateau of British Columbia, becomes
+completely deglaciated by 12\,ka, whereas a significant ice cover remains over
+the Skeena, Selwyn, Mackenzie, Wrangell and Saint-Elias mountains in the
+northern sector of the modelling domain. After 12\,ka, deglaciation continues
+to proceed across the Liard Lowland with a radial ice margin retreat towards
+the surrounding mountain ranges, consistent with the regional melt water record
+of the last deglaciation \citep{Margold.etal.2013}. Remaining ice continues to
+decay by retreating towards the Selwyn and Skeena mountains. The last remnants
+of the Cordilleran ice sheet finally disappear from the Skeena Mountains around
+10.6\,ka (EPICA) and 9.4\,ka (GRIP).
+
+\subsubsection{Late-glacial readvance}
+
+The possibility of late glacial readvances in the North American Cordillera has
+been debated for some time \citep{Osborn.Gerloff.1997}, and locally these have
+been reconstructed and dated. Radiocarbon-dated end moraines in the Fraser and
+Squamish valleys, off the southern tip of the Coast Mountains, indicate
+consecutive glacier maxima, or standstills while in overall retreat, one of
+which corresponds to the Younger Dryas chronozone \citep{Clague.etal.1997,
+Friele.Clague.2002, Friele.Clague.2002a, Kovanen.2002,
+Kovanen.Easterbrook.2002}. Although most of these moraines characterise
+independent valley glaciers, that were probably disconnected from the waning
+Cordilleran ice sheet, the Finlay River
+area of the northern Rocky Mountains presents a different kind of evidence.
+There, sharp-crested moraines indicate a late-glacial readvance of local alpine
+glaciers and, more importantly, their interaction with larger, lingering
+remnants of the main body of the Cordilleran ice sheet in the valleys
+\citep{Lakeman.etal.2008}. Additional evidence for late-glacial
+alpine glacier readvances includes moraines in the eastern Coast Mountains,
+Rocky Mountains and the Columbia Mountains \citep{Osborn.Gerloff.1997,
+Menounos.etal.2008}.
+
+Although further work is needed to constrain the timing of the late-glacial
+readvance, to assess its extents and geographical distribution, and to identify
+the potential climatic triggers \citep{Menounos.etal.2008}, it is interesting
+to note that the simulation driven by the GRIP record produces a late-glacial
+readvance in the Coast Mountains, Rocky Mountains and the Finlay River area,
+corresponding to where it has been identified in the geological record
+(Fig.~\ref{fig:deglac}, left panel). In addition to matching the location of
+local readvances, the GRIP-driven simulation shows that larger ice bodies
+emanating from the Skeena and Mackenzie mountains may still have existed at the
+time of this late-glacial readvance.
+In contrast, the EPICA-driven simulation produces a nearly-continuous
+deglaciation with only a tightly restricted glacial readvance
+(Fig.~\ref{fig:deglac}, right panel).
+
+\subsubsection{Deglacial flow directions}
+
 \begin{figure*}
@@
 
-% fig:profiles-grip
 \begin{figure}
@@
 
-% fig:profiles-epica
 \begin{figure}
@@
 
+Because a general tenant in glacial geomorphology is that the majority of
+landforms (lineations and eskers) are part of the deglacial envelope
+\citep[terminology from][]{Kleman.etal.2006}, having been formed
+close inside the retreating margin of ice sheets \citep{Boulton.Clark.1990,
+Kleman.etal.1997, Kleman.etal.2010}, we present maps of basal flow directions
+immediately preceding deglaciation or at the time of cessation of sliding
+inside a cold-based retreating margin (Fig.~\ref{fig:lastflow}). The
+modelled deglacial flow patterns appear consistent between the GRIP and EPICA
+simulations. They depict an active ice sheet retreat in the peripheral areas,
+followed by stagnant ice decay in some of the interior regions. Several parts
+of the modelling domain do not experience any basal sliding throughout the
+deglaciation phase (Fig.~\ref{fig:lastflow}, hatched areas). This notably
+includes parts of the Interior Plateau in British Columbia, major
+portions of the Alaskan sector of the ice sheet, and a tortuous ribbon running
+from the Northern Rocky Mountains over the Skeena and Selwyn Mountains and into
+the Mackenzie Mountains.
+
+Patterns of glacial lineations formed in the northern and southern sectors of
+the Cordilleran ice sheet (\citealp{Prest.etal.1968};
+\citealp[Fig.~1.12]{Clague.1989}; \citealp[Fig.~2]{Kleman.etal.2010}) show
+similarities with the patterns of deglacial ice flow from numerical modelling
+(Fig.~\ref{fig:lastflow}). In the northern half
+of the modelling domain, modelled deglacial flow directions depict an
+active downhill flow as the last remnants of the ice sheet retreat towards
+mountain ranges. Converging deglacial flow patterns in the Liard
+Lowland, for instance (Fig.~\ref{fig:lastflow}), closely resemble the pattern
+indicated by glacial lineations \citep[Fig.~2]{Margold.etal.2013}.
+
+However, on the Interior Plateau of south-central British Columbia, both
+simulations predict negligible basal sliding during deglaciation
+(Fig.~\ref{fig:lastflow}), but an impressive set of glacial lineations
+indicate a substantial eastward flow component of the Cordilleran ice sheet
+\citep{Prest.etal.1968, Kleman.etal.2010}.
+The Fraser lineation set could thus present a smoking gun for the
+reliability of the presented model results. One explanation for the incongruent
+results could therefore be that missing feedback mechanisms between ice sheet
+topography and regional climate position the modelled ice divide of the LGM ice
+sheet too far to the east (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
+\citealp{Seguinot.etal.2014}). A more westerly-located LGM ice divide would
+certainly result in a different deglacial flow pattern over the Interior
+Plateau. However, a more westerly-positioned LGM ice divide would certainly be
+associated with a thinner ice sheet than that modelled here. Decreased ice
+thickness would not promote warm-based conditions but, on the
+opposite, enlarge the region of negligible basal sliding
+(Fig.~\ref{fig:lastflow}). Thus, another explanation for the incongruent
+results could be that the Interior Plateau lineation system predates
+deglaciation, as perhaps indicated by some eskers that appear incompatible with
+these glacial lineations \citep[Fig.~9]{Margold.etal.2013a}. Alternatively,
+local geothermal heat associated with volcanic activity on the Interior Plateau
+could have triggered the basal sliding
+\citep[cf. Greenland ice sheet,][]{Fahnestock.etal.2001}.
+
+The modelled deglaciation of the Interior Plateau of British Columbia
+consists of a rapid northwards retreat (Fig.~\ref{fig:deglac}) of
+southwards-flowing non-sliding ice lobes (Fig.~\ref{fig:lastflow}) positioned
+in-between deglaciated mountain ranges
+(Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}). This result
+appears compatible with the prevailing conceptual model of deglaciation of
+central British Columbia, in which mountain ranges emerge from the ice before
+the plateau \citep[Fig.~7]{Fulton.1991}.
+
+% =============================================================================
+\conclusions
+\label{sec:concl}
+% =============================================================================
+
+Numerical simulations of the Cordilleran ice sheet through the last glacial
+cycle presented in this study consistently produce two glacial maxima during
+MIS~4 (61.9--55.4\,ka, 3.8--8.7\,m~s.l.e) and MIS~2 (29.5--16.9\,ka,
+8.24--8.66\,m~s.l.e.), two periods corresponding to documented extensive
+glaciations. This result is independent of the palaeo-temperature record used
+among the six selected for this study, and thus can be seen as a robust model
+output, which broadly matches geological evidence. However, the timing of the
+two glaciation peaks depends sensitively on which climate record is used to
+drive the model. The timing of the LGM is best reproduced by the EPICA and
+Vostok Antarctic ice core records, and occurs too early in all other
+simulations. The mismatch is largest for the two northwest Pacific ODP
+palaeo-temperature records, which are affected by the weakening of the
+California current during the LGM.
+
+In all simulations presented here, ice cover is limited to disjoint mountain
+ice caps during most of the glacial cycle. The most persistent nucleation
+centres are located in the Coast Mountains, the Columbia and Rocky mountains,
+the Selwyn and Mackenzie mountains, and most importantly, in the Skeena
+Mountains. Throughout the modelled last glacial cycle, the Skeena Mountains
+host an ice cap which appears to be fed by moisture intruding inland from the
+west through a topographic low in the Coast Mountains. The Skeena ice cap acts
+as the main nucleation centre at the onset of the LGM readvance. As indicated
+by persistent, warm-based ice in the model, this ice cap is perhaps associated
+with the distinct glacial erosional imprint on the landscape of the Skeena
+Mountains.
+
+During deglaciation, none of the climate records used can be selected as
+producing an optimal agreement between the model results and the geological
+evidence. Although the EPICA-driven simulation yields the most realistic timing
+of the LGM and, therefore, start of deglaciation, only the GRIP-driven
+simulation produces late glacial readvances in areas where these have been
+documented. Nonetheless, the patterns of ice sheet retreat are consistent
+between the two simulations, and show a rapid deglaciation of the southern
+sector of the ice sheet, including a rapid northwards retreat across the Fraser
+Plateau of central British Columbia. The GRIP-driven simulation then produces a
+late-glacial readvance of local ice caps in the Coast, Columbia and Rocky
+mountains, and of the main body of the decaying Cordilleran ice sheet in the
+Skeena and Selwyn mountains. In both simulations, this is followed by an
+opening of the Liard Lowland, and a final retreat of the remaining ice caps
+towards the Selwyn and, finally, the Skeena mountains, which hosts the last
+remnant of the ice sheet during the early Holocene (10.9--9.5\,ka). Our results
+identify the Skeena Mountains as a key area to understanding glacial dynamics
+of the Cordilleran ice sheet, highlighting the need for further geological
+investigation of this region.
+
+% Author contributions
+\section*{Author contributions}
+J.~Seguinot ran the simulations and wrote most of the manuscript; I.~Rogozhina
+guided experiment design; A.~P.~Stroeven, M.~Margold and J.~Kleman took part in
+the interpretation and comparison of model results against geological evidence.
+All co-authors contributed to improve the text.
+
+% Acknowledgements
+\begin{acknowledgements}
+We are very thankful to Constantine Khroulev, Ed Bueler and Andy Aschwanden for
+providing constant help and development on PISM. This work was supported by the
+Swedish Research Council~(VR) grant no. 2008-3449 to A.~P.~Stroeven, by the
+German Academic Exchange Service~(DAAD) grant no.~50015537 and a Knut and Alice
+Wallenberg Foundation grant to J.~Seguinot.
+Computer resources were provided by the Swedish National
+Infrastructure for Computing (SNIC) allocation no. 2013/1-159 to A.~P.~Stroeven
+at the National Supercomputing Center (NSC).
+\end{acknowledgements}
+
+% References
+\bibliographystyle{copernicus}
+\bibliography{refs/references.bib}
+\newpage
+
+% =============================================================================
+% Floats
+% =============================================================================
+
+% tab:params
+
+% tab:extrema
+
 % =============================================================================

### 2014-08-08 kappa.tex
Replace "I" by "J. Seguinot" in author contrib.

@@
   \item[I] A.~P.~Stroeven initiated the Cordilleran ice sheet modelling
-project; I set-up the ice sheet model (PISM) and supercomputing workflow, ran
-simulations and wrote the first draft; C.~Khroulev implemented necessary code
-changes in the model; I.~Rogozhina contributed to experiment design; A.~P.~Stroeven
-helped with interpretation of the results; and Q.~Zhang prepared some of the
-forcing data. All co-authors contributed to improve the text.
-  \item[II] I analysed the weather data, wrote the positive degree day model
-(PyPDD), ran computations and wrote the manuscript. A.~P.~Stroeven contributed
-with editorial comments at the pre-submission stage and I.~Rogozhina helped to
-assess the reviews.
-  \item[III] I.~Rogozhina provided the original idea; I analysed the weather
-data and developed a new parameterization of daily temperature standard
-deviation. I.~Rogozhina and I wrote the manuscript. A.~P.~Stroeven contributed
-with editorial comments at the pre-submission stage.
-  \item[IV] I integrated the new parameterization of standard deviation into
-PISM, ran the simulations and wrote the first draft; I.~Rogozhina guided
-experiment design; A.~P.~Stroeven, M.~Margold and J.~Kleman took part in the
-interpretation and comparison of model results against geological evidence. All
-co-authors contributed to improve the text.
+project; J.~Seguinot set-up the ice sheet model (PISM) and supercomputing
+workflow, ran simulations and wrote the first draft; C.~Khroulev implemented
+necessary code changes in the model; I.~Rogozhina contributed to experiment
+design; A.~P.~Stroeven helped with interpretation of the results; and Q.~Zhang
+prepared some of the forcing data. All co-authors contributed to improve the
+text.
+  \item[II] J.~Seguinot analysed the weather data, wrote the positive degree
+day model (PyPDD), ran computations and wrote the manuscript. A.~P.~Stroeven
+contributed with editorial comments at the pre-submission stage and
+I.~Rogozhina helped to assess the reviews.
+  \item[III] I.~Rogozhina provided the original idea; J.~Seguinot analysed the
+weather data and developed a new parameterization of daily temperature standard
+deviation. I.~Rogozhina and J.~Seguinot wrote the manuscript. A.~P.~Stroeven
+contributed with editorial comments at the pre-submission stage.
+  \item[IV] J.~Seguinot integrated the new parameterization of standard
+deviation into PISM, ran the simulations and wrote the first draft;
+I.~Rogozhina guided experiment design; A.~P.~Stroeven, M.~Margold and J.~Kleman
+took part in the interpretation and comparison of model results against
+geological evidence. All co-authors contributed to improve the text.
 \end{paper}}

### 2014-08-08 figures/model-pdd.tex
Adjust figure widths.

@@
 % background
-\fill[white] (-0.25,-0.25) rectangle +(8.5,14.75);
+\fill[white] (0,0) rectangle +(8.0,14.75);
 %\draw[help lines, lightgray] (0,0) grid (8,14);
@@
 % add nodes
-\path     (4, 13.75) node [trunk] (acc) {Compute snow accumulation, $A_s$}
+\path     (4, 14) node [trunk] (acc) {Compute snow accumulation, $A_s$}
   -- ++(0, -1.5) node [trunk] (pdd) {Compute PDDs}

### 2014-08-08 figures/paleo-glaciation.py

@@ -15,14 +15,9 @@ scale = '50m'
 proj = ccrs.Orthographic(central_longitude=-90.0, central_latitude=75.0)
 
 # Initialize figure in inches
-figw, figh = 85.0/25.4, 85.0/25.4
-fig = plt.figure(0, (figw, figh))
-fig.subplots_adjust(left=0.5/72/figw, bottom=0.5/72/figh,
-                    right=1-0.5/72/figw, top=1-0.5/72/figh,
-                    wspace=1/((1+figw/(1./72))/2-1))
-figw, figh = 85.0, 85.0
-#grid = [fig.add_subplot(121+i, projection=proj) for i in range(2)]
+figw, figh = 80.0, 80.0
 mapw = 50
+fig = plt.figure(0, (figw/25.4, figh/25.4))
 grid = [fig.add_axes([2.5/figw, 1-(2.5+mapw)/figh, mapw/figw, mapw/figh],
                      projection=proj),
         fig.add_axes([1-(2.5+mapw)/figw, 2.5/figh, mapw/figw, mapw/figh],

### 2014-08-09 kappa.tex
Adjust figure widths.
Rework and shorten abstract.

@@
 
-This doctoral thesis presents a study of the glacial history of northwestern
-North America using numerical ice sheet modelling calibrated against field
-evidence. The area is characterized by the steep topography of several mountain
-ranges separated by large inter-montane depressions that were once covered by a
-large-scale ice mass, the former Cordilleran ice sheet. Due to this complex
-topographic setting, the Cordilleran ice sheet remains the least understood
-Pleistocene ice sheet in terms of its extent, volume, and dynamics. Because of
-the irregular topography on which the ice sheet formed, field studies have
-often only had local or regional relevance. Numerical modelling, on the other
-hand, allows quantitative reconstructions of whole ice sheet evolution based on
-approximated physics of glacier flow.
-
-This thesis presents simulations performed with PISM, a numerical ice sheet
-model that approximates glacier flow by basal sliding and internal deformation,
-and allows simulations of ice sheet growth and decay over multi-millenial time
-scales. Sensitivity experiments have shown that simulations adopting climate
-forcing from the North American Regional Reanalysis (NARR) yield ice sheet
-extents that best fit the field constraints during the Last Glacial Maximum
-(LGM). These simulations show that the geometry of the LGM Cordilleran ice
-sheet can be largely explained by present-day temperature and precipitation
-fields combined with palaeo-climate reconstructions if an accurate treatment of
-seasonality, orographic precipitation, and daily temperature variability is
-included in the ice sheet model.
-
-In contrast to this short-lived, maximum ice sheet configuration, most of the
-last glacial cycle is characterized by an intermediate state of glaciation
-including isolated glaciers and ice caps covering the major mountain ranges of
-the North American Cordillera. The largest and most persistent of these ice
-sheet nucleation centres is located over the Skeena Mountains. Modelled
-deglaciation conditions broadly match field evidence, and where they diverge,
-such as over the Fraser Plateau, they indicate either a lack of important
-processes in the model, or a more complex glaciation history than previously
-thought. The model yields a readvance of the ice sheet during late glacial
-times, which appears to match field evidence, before final retreat towards the
-Skeena Mountains, which we identify as a key area to understanding glacial
-dynamics of the Cordilleran ice sheet, highlighting the need for further
-geological investigation of this region.
+This doctoral dissertation presents a study of the glacial history of the North
+American Cordillera using numerical ice sheet modelling calibrated against
+field evidence. This area, characterized by the steep topography of several
+mountain ranges separated by large inter-montane depressions, was once covered
+by a large-scale ice mass: the former Cordilleran ice sheet. Because of the
+irregular topography on which the ice sheet formed, geological studies have
+often only had local or regional relevance, and thus the Cordilleran ice sheet
+remains the least understood Pleistocene ice sheet in terms of its extent,
+volume, and dynamics.
+
+In this thesis, I present numerical simulations performed with the Parallel Ice
+Sheet Model (PISM), that allow quantitative reconstructions of the former ice
+sheet evolution based on approximated physics of glacier flow. These
+simulations show that the geometry of the Last Glacial Maximum Cordilleran ice
+sheet was largely controlled by sharp contrasts in regional temperature,
+precipitation, and daily temperature variability associated with the presence
+of the mountain ranges.
+
+However, this maximum stage appear short-lived and out of balance with
+contemporaneous climate. During most of the simulated last glacial cycle, the
+North American Cordillera is characterized by an intermediate state of
+glaciation including isolated glaciers and ice caps covering the major mountain
+ranges, the largest of which is located over the Skeena Mountains. Numerically
+modelled patterns of ice sheet growth and decay are diverse, and reveal a
+higher complexity than that encapsulated in previous, conceptual models.
 
@@
 \begin{figure}
-  \includegraphics[width=80mm]{model-siassa}
+  \includegraphics{model-siassa}
   \caption{Combination of Shallow Ice Approximation (SIA) and Shallow Shelf
@@
 needed sustain a constant ice sheet geometry because the ice sheet growth was
-finally inhibited by the deglacial transition to warmer temperatures .
+finally inhibited by the deglacial transition to warmer temperatures.
 Therefore, this maximum stage is not representative of average conditions, as

### 2014-08-10 kappa.tex
Implement Irina's corrections on abstract.

@@
 irregular topography on which the ice sheet formed, geological studies have
-often only had local or regional relevance, and thus the Cordilleran ice sheet
-remains the least understood Pleistocene ice sheet in terms of its extent,
+often had only local or regional relevance, thus leaving the Cordilleran ice
+sheet least understood among Pleistocene ice sheets in terms of its extent,
 volume, and dynamics.
 
-In this thesis, I present numerical simulations performed with the Parallel Ice
-Sheet Model (PISM), that allow quantitative reconstructions of the former ice
+Here, I present numerical simulations that allow quantitative reconstructions
+of the former ice
 sheet evolution based on approximated physics of glacier flow. These
@@
 precipitation, and daily temperature variability associated with the presence
-of the mountain ranges.
-
-However, this maximum stage appear short-lived and out of balance with
+of mountain ranges.
+
+However, this maximum stage appears short-lived and out of balance with
 contemporaneous climate. During most of the simulated last glacial cycle, the
 North American Cordillera is characterized by an intermediate state of
-glaciation including isolated glaciers and ice caps covering the major mountain
-ranges, the largest of which is located over the Skeena Mountains. Numerically
-modelled patterns of ice sheet growth and decay are diverse, and reveal a
-higher complexity than that encapsulated in previous, conceptual models.
-
-The simulations presented in this thesis demonstrate the potential of numerical
-ice sheet modelling to inform on ice sheet history over glacial cycles,
-provided that ice sheet models can be calibrated against field constraints.
+glaciation including isolated glaciers and ice caps covering major mountain
+ranges, the largest of which is located over the Skeena Mountains.
+The numerically modelled Cordilleran ice sheet appears in constant imbalance with evolving climate conditions, while the nature of this transient response transcends the complexity of two-dimensional, conceptual models of ice sheet growth and decay.
+
+This thesis demonstrates the potential of numerical ice sheet modelling to
+inform on ice sheet history and former climate conditions over a glacial cycle,
+given that ice sheet models can be calibrated against field constraints.
 

### 2014-08-10 kappa.tex
Move one figure up.
Fix one sentence in the abstract.

@@
 glaciation including isolated glaciers and ice caps covering major mountain
-ranges, the largest of which is located over the Skeena Mountains.
-The numerically modelled Cordilleran ice sheet appears in constant imbalance with evolving climate conditions, while the nature of this transient response transcends the complexity of two-dimensional, conceptual models of ice sheet growth and decay.
+ranges, the largest of which is located over the Skeena Mountains. The
+numerically modelled Cordilleran ice sheet appears in constant imbalance with
+evolving climate conditions, while the complexity of this transient response
+transcends that encapsulated in two-dimensional, conceptual models of ice sheet
+growth and decay.
 
@@
 
-As a first step towards a simulations of the Cordilleran ice sheet through the
-last glacial cycle, I have opted to run the ice sheet model under a range of
-constant temperature depressions, targeting the LGM as the period for which ice
-sheet limits are best known. These simulations,
-performed on a 10\,km grid, use constant temperature depressions
-from the present climate state, which is represented using five different input
-climatologies. All model runs start
-from ice free conditions and run for 10\,ka, a period chosen as representative
-of Cordilleran ice sheet build-up from assumed, negligible ice cover. Although
-this assumption appears unjustified from results in \CCYC, the simplistic
-set-up used in \CCLI\ allowed an assessment of model sensitivity to the choice
-of input climatologies, without involving too many additional parameters.
-
-Discrepancies between the five input data sets used (\cref{tab:datasources})
-caused considerable variability in modelled ice sheet geometries (\CCLI,
-Figs.~5--7). In particular, ice growth in continental sectors of the modelling
-domain is strongly dependent on input precipitation patterns, which are often
-poorly resolved by atmospheric reanalysis fields
-(\CCLI, Figs.~10--14). The closest match between modelled ice sheet extent and
-the reconstructed LGM ice margin \citep{Dyke.2004} was obtained using the North
-American Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006} climatologies
-and a temperature depression of 7\,K (\CCLI, Fig.~15). The relevant simulation
-is reproduced here, using a higher horizontal resolution of 3\,km
-(\cref{fig:plot-lgm}). Following this sensitivity study, I have retained the
-NARR climatology as input to simulations of the Cordilleran ice sheet through
-the last glacial cycle (\CCYC).
-
 \begin{figure*}
@@
 
+As a first step towards a simulations of the Cordilleran ice sheet through the
+last glacial cycle, I have opted to run the ice sheet model under a range of
+constant temperature depressions, targeting the LGM as the period for which ice
+sheet limits are best known. These simulations,
+performed on a 10\,km grid, use constant temperature depressions
+from the present climate state, which is represented using five different input
+climatologies. All model runs start
+from ice free conditions and run for 10\,ka, a period chosen as representative
+of Cordilleran ice sheet build-up from assumed, negligible ice cover. Although
+this assumption appears unjustified from results in \CCYC, the simplistic
+set-up used in \CCLI\ allowed an assessment of model sensitivity to the choice
+of input climatologies, without involving too many additional parameters.
+
+Discrepancies between the five input data sets used (\cref{tab:datasources})
+caused considerable variability in modelled ice sheet geometries (\CCLI,
+Figs.~5--7). In particular, ice growth in continental sectors of the modelling
+domain is strongly dependent on input precipitation patterns, which are often
+poorly resolved by atmospheric reanalysis fields
+(\CCLI, Figs.~10--14). The closest match between modelled ice sheet extent and
+the reconstructed LGM ice margin \citep{Dyke.2004} was obtained using the North
+American Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006} climatologies
+and a temperature depression of 7\,K (\CCLI, Fig.~15). The relevant simulation
+is reproduced here, using a higher horizontal resolution of 3\,km
+(\cref{fig:plot-lgm}). Following this sensitivity study, I have retained the
+NARR climatology as input to simulations of the Cordilleran ice sheet through
+the last glacial cycle (\CCYC).
+
 % ----------------------------------------------------------------------

### 2014-08-23 cordillera-cycle.tex
Correct many small mistakes and add recent refs.

@@
   approach reveals large variations in model response to evolving climate
-  forcing, all simulations produce two major glaciation events during
+  forcing, all simulations produce two major glaciations during
   marine oxygen isotope stages~4 (61.9--55.4\,ka) and~2
@@
 a continuous blanket of ice, the Cordilleran ice sheet
-\citep{Dawson.1888}, extended from the Alaska Range in the north to the
+\citep{Dawson.1888}, stretched from the Alaska Range in the north to the
 North Cascades in the south (Fig.~\ref{fig:locmap}).
@@
 
-More than a century of exploration and geological investigations of the
+More than a century of exploration and geological investigation of the
 Cordilleran mountains have led to many observations in support of the former
@@
      \citealp{Booth.etal.2003}; \citealp{Dyke.2004}).
-The LGM Cordilleran ice sheet maximum extent was short-lived, as indicated by
+The LGM Cordilleran ice sheet extent was short-lived, as indicated by
 field evidence from radiocarbon dating
@@
            Innuitian (IIS) and Greenland (GIS) ice sheets during the last
-           18\,\unit{\chem{^{14}C}\,ka} (21.4\,cal\,ka)
-           \citep{Dyke.2004}. The rectangular box denotes the location of the
+           18\,\unit{\chem{^{14}C}\,ka} \citep[21.4\,cal\,ka,][]{Dyke.2004}.
+           The rectangular box denotes the location of the
            modelling domain used in this study. Major mountain ranges covered
@@
     \citep[e.g.,][]{Sissons.1979, Lundqvist.1987,
-                    Ivy-Ochs.etal.1999, Stea.etal.2011},
-thus allowing for considerable uncertainty concerning
-the possibility of a regional late glacial cold reversal.
+                    Ivy-Ochs.etal.1999, Stea.etal.2011}.
+Nevertheless, recent oxygen isotope measurements from Gulf of Alaska sediments
+reveals a climatic evolution highly correlated to that of Greenland during this
+period, including a distinct late glacial cold reversal between 14.1 and
+11.7\,ka \citep{Praetorius.Mix.2014}.
 
@@
 topography of the North American Cordillera also induces strong gradients in
+temperature
 seasonality and distinct patterns of precipition, thus requiring the use of
@@
     $\nu_m$ & Mantle viscosity
-            & $1\times10^21$
+            & $1\times10^{21}$
             & \unit{Pa\,s} \\
@@
     $D$     & Lithosphere flexural rigidity
-            & $5.0\times10^24$
+            & $5.0\times10^{24}$
             & \unit{N} \\
@@
 
-    $F_s$   & Degree-day factor for ice
+    $F_s$   & Degree-day factor for snow
             & $3.04\times10^{-3}$
@@
 
-    $\gamma$& Air temperature lapse-rate
+    $\gamma$& Air temperature lapse rate
             & $6\times10^{-3}$
@@
 is lowered as a function of time based on the Spectral Mapping Project
-\citep[SPECMAP;][]{Imbrie.etal.1989} time scale. Geothermal heat flux
+\citep[SPECMAP,][]{Imbrie.etal.1989} time scale. Geothermal heat flux
 is applied as a constant value of 70\,\unit{mW\,m^{-2}} at 3\,km depth
@@
 provided by a monthly climatology averaged from 1979 to 2000 from the North
-American Regional Reanalysis \citep[NARR;][]{Mesinger.etal.2006},
+American Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006},
 perturbated by time-dependent and
@@
 sea level, m~a.s.l) and the highest value occuring above the generalised
-elevation of the highest shorelines (200\,m~a.s.l), thus accounting for
+elevation of the highest shorelines
+\citep[200\,m~a.s.l.,][Fig.~5]{Clague.1981}, thus accounting for
 a~weakening of till associated with the presence of marine sediments
-(cf. \citealp{Martin.etal.2011}, \citealp[supplement]{Aschwanden.etal.2013},
+(cf. \citealp{Martin.etal.2011}; \citealp[supplement]{Aschwanden.etal.2013};
 \citealp{PISM-authors.2014}). Effective pressure, $N$,
@@
 following a bedrock deformation model that includes local isostasy,
-elastic lithosphere flexure and viscous mantle deformation in the lower
+elastic lithosphere flexure and viscous mantle deformation in an infinite
 half-space
@@
   \includegraphics{atm}
-  \caption{Monthly mean near-surface air temperature, precipitation and
+  \caption{Monthly mean near-surface air temperature, precipitation, and
            standard deviation of daily mean temperature for January and July
@@
 Drilling Program (ODP) sites 1012 and 1020, both located off the coast of
-California \citep{Herbert.etal.2001}. Palaeo-temperatures from the GRIP and
-NGRIP
+California \citep{Herbert.etal.2001}. Palaeo-temperature anomalies from the
+GRIP and NGRIP
 records were calculated from oxygen isotope (\chem{\delta^{18}O}) measurements
@@
            \citep{Lisiecki.Raymo.2005}. Hatched rectangles highlight the
-           time-volume space for ice volume extremes corresponding to MIS~4
+           time-volume span for ice volume extremes corresponding to MIS~4
            (61.9--55.4\,ka), MIS~3 (52.2--45.6\,ka), and MIS~2 (LGM,
@@
 Despite such differences in the timing of attained volume extrema
-(Table~\ref{tab:extrema}), all the model runs show relatively consistent
+(Table~\ref{tab:extrema}), all model runs show relatively consistent
 patterns of glaciation. During MIS~4, all simulations produce an extensive ice
@@
 an increase of glaciomarine sedimentation between 19.5 and
-16.2\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008}. Radiocarbon
+16.2\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008, Taylor.etal.2014}.
+Radiocarbon
 dating of the northern Cordilleran ice sheet margin is much less constrained
 but straddles presented constraints from the southern margin. However,
-cosmogenic exposure dating
-places the timing of maximum CIS extent during the McConnell glaciation close
-to 17\,ka \citep{Stroeven.etal.2010, Stroeven.etal.2014}.
+cosmogenic exposure dating places the timing of maximum CIS extent during the
+McConnell glaciation close to 17\,\unit{\chem{^{10}Be}\,ka}
+\citep{Stroeven.etal.2010, Stroeven.etal.2014}. A sharp transition in
+the sediment record of the Gulf of Alaska indicates a retreat of regional
+outlet glaciers onto land at 14.8\,\unit{\chem{^{14}C}\,cal\,ka}
+\citep{Davies.etal.2011}.
 
@@
 air parcels flowing over this enhanced topography could have resulted in an
-inflow of warmer air over the eastern flank of the ice sheet, increasing melt
+inflow of potentially warmer air over the eastern flank of the ice sheet,
+increasing melt
 along the advancing margin \citep[cf.][]{Langen.etal.2012}. Because these two
-processes, both with a tendency to limit of ice-sheet growth, are absent from
+processes, both with a tendency to limit ice-sheet growth, are absent from
 our model, the
@@
 ice surface elevations in the interior (Fig.~\ref{fig:mis4}, left panel).
+Thus, only the GRIP simulation can explain the presence of ice-rafted debris
+in the marine sediment record offshore Vancouver Island at
+ca.\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008}.
 
@@
   \caption{Modelled duration of ice cover during the last 120\,ka.
-           Note the irregular colour scale. A contiguous ice cover spanning
+           Note the irregular colour scale. A continuous ice cover spanning
            from the Alaska Range (AR) to the southern Coast Mountains (CM) and
@@
 AR). This apparent insensitivity of modelled glacial extent to temperature
-fluctuations results from a combination regimes of low precipitation, high summer
+fluctuations results from a combination of low precipitation, high summer
 temperature and large temperature standard deviation for this sector of the
@@
 middle panels), and serves as a nucleation centre at the onset of the glacial
-readvance towards the LGM (MIS~2). The importance of residual ice for North
+readvance towards the LGM (MIS~2). For the neighbouring Laurentide ice sheet,
+the importance of residual ice for North
 American glacial history leading up to the LGM has been illustrated by the
 MIS~3 residual ice bodies in northern and eastern Canada as nucleation centres
-for a much more extensive MIS~2 configuration of the Laurentide ice sheet
-\citep{Kleman.etal.2010}.
+for a much more extensive MIS~2 configuration \citep{Kleman.etal.2010}.
 
@@
 north-central British Columbia, thus resulting in a more widespread
-distribution of winter precipitation.
+distribution of winter precipitation (Fig.~\ref{fig:atm}).
 
@@
 {Stroeven.etal.2010}) is modelled to have hosted warm-based ice
-(Fig.~\ref{fig:warmfrac}). However, because it has only been glaciated for a
-short fraction
-of the last glacial cycle, (Fig.~\ref{fig:duration}) according to our
-results, this perhaps appears indicate that
+(Fig.~\ref{fig:warmfrac}). However, because it is only glaciated for a
+short fraction of the last glacial cycle in our simulations
+(Fig.~\ref{fig:duration}), this perhaps appears indicate that
 the observed landscape pattern originates from multiple glacial cycles and
@@
            green. Hatches denote re-advance of mountain-centred ice caps and
-           and the decaying ice sheet between 14 and 10\,ka., which is more
+           the decaying ice sheet between 14 and 10\,ka., which is more
            pronounced in the GRIP-driven simulation.
@@
 Margold.etal.2014}, and abundant esker systems at low elevation
-\citep{Burke.etal.2012, Burke.etal.2012a, Perkins.etal.2013,
-Margold.etal.2013}
-indicate that abundant meltwater was produced over the entire ice sheet surface
-during deglaciation. The southern and northern margins of the
+\citep{Burke.etal.2012, Burke.etal.2012a, Perkins.etal.2013, Margold.etal.2013}
+indicate that meltwater was produced over large portions of the ice sheet
+surface during deglaciation. The southern and northern margins of the
 Cordilleran ice sheet reached their last glacial maximum extent around
@@
 maximum and the pacing of deglaciation depend critically on the choice of
-climate forcing (Table~\ref{tab:extrema}, Figs.~\ref{fig:timeseries}
-and~\ref{fig:deglacseries}). Adopting the EPICA
-climate forcing yields peak ice volume at 17.3\,ka and an uninterrupted
-deglaciation until 10.6\,ka (Fig.~\ref{fig:deglacseries}, lower panel, red
-curves). The simulation driven by the GRIP palaeo-temperature record yields
-peak ice volume at 19.1\,ka and a deglaciation interrupted by two ice volume
-standstills until 9.4\,ka. The first interruption occurs between 16.6 and
-14.5\,ka, and the second between 12.6 and 11.6\,ka
-(Fig.~\ref{fig:deglacseries}, lower panel, blue curve). Hence, the two model
-runs, while similar in overall timing compared to runs with other climate
-drivers, differ in detail in that the EPICA depicts peak glaciation almost
-2\,ka later than the GRIP, in closer agreement with dated maximum extents, and
-shows a faster, uninterrupted deglaciation which yields ice-free conditions
-more than 1\,ka earlier.
+climate forcing (Table~\ref{tab:extrema}; Figs.~\ref{fig:timeseries}
+and~\ref{fig:deglacseries}). Adopting the EPICA climate forcing yields peak ice
+volume at 17.3\,ka and an uninterrupted deglaciation until 10.6\,ka
+(Fig.~\ref{fig:deglacseries}, lower panel, red curves). On the contrary, the
+simulation driven by the GRIP palaeo-temperature record yields peak ice volume
+at 19.1\,ka and a deglaciation interrupted by two ice volume standstills until
+9.4\,ka. The first interruption occurs between 16.6 and 14.5\,ka, and the
+second between 12.6 and 11.6\,ka (Fig.~\ref{fig:deglacseries}, lower panel,
+blue curve).
+
+Hence, the two model runs, while similar in overall timing compared to runs
+with other climate drivers, differ in detail. On the one hand, the EPICA run
+depicts peak glaciation almost 2\,ka later than the GRIP run, in closer
+agreement with dated maximum extents, and shows a faster, uninterrupted
+deglaciation which yields ice-free conditions more than 1\,ka earlier. On the
+other hand, the GRIP run yields a deglaciation in three steps, compatible with
+marine sediment sequences offshore Vancouver Island, where the distribution of
+ice-rafted debris indicate an ice margin retreat from the Georgia Strait in two
+phases that are contemporary with warming oceanic temperatures from 17.2 to
+16.5 and from 15.5 to 14.0\,\unit{\chem{^{14}C}\,cal\,ka}.
 
@@
 In contrast, the EPICA-driven simulation produces a nearly-continuous
-deglaciation with only a tightly restricted glacial readvance
+deglaciation with only a tightly restricted late-glacial readvance
 (Fig.~\ref{fig:deglac}, right panel).
@@
   \includegraphics{profiles-epica}
-  \caption{Modelled bedrock (black) and ice surface (blue) topography profiles
+  \caption{Modelled bedrock (black) and ice surface (red) topography profiles
            during deglaciation (22.0--8.0\,ka) in the EPICA 6\,km
@@
 central British Columbia, in which mountain ranges emerge from the ice before
-the plateau \citep[Fig.~7]{Fulton.1991}.
+the plateau \citep[Fig.~7]{Fulton.1991}. However, due to different topographic
+and climatic conditions, our simulations produce different deglaciation
+patterns in the northern half of the model domain, indicating that this
+conceptual model may not be applied to the entire area formerly covered by the
+Cordilleran ice sheet.
 
@@
 between the two simulations, and show a rapid deglaciation of the southern
-sector of the ice sheet, including a rapid northwards retreat across the Fraser
+sector of the ice sheet, including a rapid northwards retreat across the Interior
 Plateau of central British Columbia. The GRIP-driven simulation then produces a
@@
 % =============================================================================
-% Floats
-% =============================================================================
-
-% tab:params
-
-% tab:extrema
-
-% =============================================================================
 \end{document}

### 2014-08-24 kappa.tex
Change to Nimbus Roman fonts.
Non-break space after (a), (b), (c), etc.

@@
 \nobibliography*
+
+% Nimbus fonts
+\usepackage{times}
+\urlstyle{same}
 
@@
   \caption{Illustration of various glacier ice behaviours:
-           \textbf{(a)} surface crevasses formed by ice flow over a hidden
+           \textbf{(a)}~surface crevasses formed by ice flow over a hidden
            bedrock knob, Glacier Blanc, French Alps, 2004;
-           \textbf{(b)} medial moraines illustrating downhill ice flow on
+           \textbf{(b)}~medial moraines illustrating downhill ice flow on
            Susitna Glacier, Alaska Range, photo by Austin Post, 1970
            \citep{NSIDC.2009}; and
-           \textbf{(c)} mountain glacier calving into a small lake, revealing a
+           \textbf{(c)}~mountain glacier calving into a small lake, revealing a
            folded structure formed by deformed debris layers in the ice,
@@
   \caption{Different indicators of past glaciation:
-           \textbf{(a)} glacial valley, exhibiting a broad, U-shaped profile
+           \textbf{(a)}~glacial valley, exhibiting a broad, U-shaped profile
            characteristic of erosion by glaciers,
            Cirque du Fer-\`{a}-Cheval, French Alps, 2013;
-           \textbf{(b)} erratic boulders, deposited in an improbable
+           \textbf{(b)}~erratic boulders, deposited in an improbable
            configuration as the glacier melted underneath,
            Finse area, Norwegian Scandes, 2011; and
-           \textbf{(c)} lateral meltwater channels, formed on a gentle slope
+           \textbf{(c)}~lateral meltwater channels, formed on a gentle slope
            as the ice margin retreated towards the valley bottom, Tuya Lake
@@
   \caption{Palaeo-climate records of the last 800\,ka.
-           \textbf{(a)} Benthic oxygen isotope (\chem{\delta^{18}O}) record
+           \textbf{(a)}~Benthic oxygen isotope (\chem{\delta^{18}O}) record
            from a compilation of globally-distributed deep sea sediment cores
@@
            glaciers and ice sheets on land \citep{Shackleton.1967}.
-           \textbf{(b)} Temperature anomalies ($\Delta T$) reconstructed from a
+           \textbf{(b)}~Temperature anomalies ($\Delta T$) reconstructed from a
            central Antarctic ice core \citep[EPICA Dome~C,][]{Jouzel.etal.2007}.
@@
   \caption{Spatial extent of ice sheet and glacier cover in the northern
-           hemisphere \textbf{(a)} at present \citep{Patterson.Kelso.2014}, and
-           \textbf{(b)} during the last glacial maximum
+           hemisphere \textbf{(a)}~at present \citep{Patterson.Kelso.2014}, and
+           \textbf{(b)}~during the last glacial maximum
            \citep{Ehlers.Gibbard.2007}.}
@@
   \caption{Contrasted climates of the North American Cordillera:
-           \textbf{(a)} Pacific Coast at Vancouver Island, western British
+           \textbf{(a)}~Pacific Coast at Vancouver Island, western British
            Columbia, by Martin Margold, 2010.
-           \textbf{(b)} Fraser Valley, south-central British Columbia, 2010.}
+           \textbf{(b)}~Fraser Valley, south-central British Columbia, 2010.}
   \label{fig:photo-fraser-valley}
@@
            including
-           \textbf{(a)} no daily variability, $\sPDD=0$;
-           \textbf{(b)} a constant, mean summer (JJA), model domain-averaged
+           \textbf{(a)}~no daily variability, $\sPDD=0$;
+           \textbf{(b)}~a constant, mean summer (JJA), model domain-averaged
            standard deviation value, $\sPDD=3.09$\,K;
-           \textbf{(c)} monthly standard deviation grids derived from the NARR
+           \textbf{(c)}~monthly standard deviation grids derived from the NARR
            and
-           \textbf{(d)} monthly grids for ice-free terrain and the
+           \textbf{(d)}~monthly grids for ice-free terrain and the
            Greenland ice sheet, ERA-40-based \citep{Uppala.etal.2005}
@@
   \includegraphics{plot-snapshots}
-  \caption{\textbf{(a)} Modelled growth and decay of the last Cordilleran ice
+  \caption{\textbf{(a)}~Modelled growth and decay of the last Cordilleran ice
            sheet. Basal topography, surface topography (1\,km contours) and
            surface velocity (m/a) are model results.
-           \textbf{(b)} Evolution of total ice volume during the last glacial
+           \textbf{(b)}~Evolution of total ice volume during the last glacial
            cycle, expressed as metres of sea-level equivalent (m~s.l.e.). The

### 2014-08-24 kappa.tex
Correct many small mistakes and explain notations.

@@
 % -- add software references
-% -- redraw PISM interface
+% -- add Swedish abstract
+% -- add numerical implementation
 
@@
            \textbf{(b)}~medial moraines illustrating downhill ice flow on
-           Susitna Glacier, Alaska Range, photo by Austin Post, 1970
+           Susitna Glacier, Alaska Range, photograph by Austin Post, 1970
            \citep{NSIDC.2009}; and
@@
            \textbf{(c)}~lateral meltwater channels, formed on a gentle slope
-           as the ice margin retreated towards the valley bottom, Tuya Lake
-           area, Canadian Cordillera, 2012.}
+           as the lateral ice margin retreated towards the valley bottom, Tuya
+           Lake area, Canadian Cordillera, 2012.}
   \label{fig:glaciation-indicators}
@@
 cumulative snowfall until it reaches a critical thickness. Beyond this
-threshold, usually in excess of 50\,m, but dependent on temperature, ice
+threshold, usually of several tens of metres, but dependent on temperature, ice
 deforms under its own weight to flow downhill (\cref{fig:glacier-mechanics}).
-\emph{Glaciers}, these bodies of moving ice, move by a combination of
+\emph{Glaciers}, these bodies of moving ice, flow by a combination of
 meltwater-induced sliding at the base \citep[\S532]{Saussure.1786},
@@
 and interglacial periods have succeeded each other with a 100\,ka periodicity
-\citep[\cref{fig:paleo-timeseries};]{Hays.etal.1976, Augustin.etal.2004}.
+\citep[\cref{fig:paleo-timeseries};][]{Hays.etal.1976, Augustin.etal.2004}.
 However, much of the traces left by glaciers on the landscape date from the
@@
 data where they exist \citep{Napieralski.etal.2007}, while allowing
-physics-based, quantitative reconstructions of ice sheet sectors where they do
-not \citep[e.g.,][]{Marshall.etal.2002, Tarasov.Peltier.2004}.
+quantitative reconstructions of ice sheet sectors where they do not, based on
+approximated physics of glacier flow \citep[e.g.,][]{Marshall.etal.2002,
+Tarasov.Peltier.2004}.
 
@@
            \citep[21.4 to 16.8\,cal\,\chem{^{14}C}\,kyr\,BP,][]{Dyke.2004}.
-           The rectangular box delimits the model domain used in all simulation
-           and most figures of this thesis. The background
+           The rectangular box delimits the model domain used in all
+           simulations and most figures of this thesis. The background
            map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
@@
 geomorphological record \citep[e.g.,][]{Kleman.etal.2010} or numerical ice
-sheet modelling \citep{Robert.1991}.
+sheet modelling \citep[e.g.,][]{Robert.1991}.
 
@@
            \textbf{(a)}~Pacific Coast at Vancouver Island, western British
-           Columbia, by Martin Margold, 2010.
+           Columbia, photograph by Martin Margold, 2010.
            \textbf{(b)}~Fraser Valley, south-central British Columbia, 2010.}
@@
 Nevadan and Laramide belts, which are separated by the deep and strikingly
-linear Rocky Mountain / Tintina Trench depression, and includes the Selwyn and
+linear Rocky Mountain~/ Tintina Trench depression, and includes the Selwyn and
 MacKenzie Mountains in the north, and the Rocky mountains in the south. The
@@
 of which influenced former patterns of Cordilleran ice sheet flow. The two
-major depressions are the Fraser Plateau in south-central British Columbia, and
-the Liard Lowland in northern British Columbia (\cref{fig:map-cordillera}).
+major depressions are the Interior Plateau in south-central British Columbia,
+and the Liard Lowland in northern British Columbia (\cref{fig:map-cordillera}).
 These two basins are separated by the Skeena Mountains, a mildly elevated, but
@@
 varies between around 10\,\degC\ along most of the Pacific Coast to above
-40\,\degC\ for the interior Arctic sectors
+40\,\degC\ for the continental Arctic sectors
 (\cref{fig:plot-atm}b). There exists a
@@
 therefore largely delivered as snow. In contrast, many inland regions
-experience dry winters, and most of the annual precipitation, consequently
+experience dry winters, and most of the annual precipitation consequently
 falls there as rain during summer, further enhancing ice melt.
@@
 $2.64^{+0.20}_{-0.18}$\,Ma (uncertainty range from 2.46 to 2.84\,Ma)
-indicates the first presence of an ice sheet in the
+indicates the first presence of an ice sheet on the
 Cordilleran mountains during the late Pliocene \citep{Hidy.etal.2013}.
@@
 since the late Pliocene, with the first of these glaciations occurring even
-before the mid-Pleistocene onset of 100 ka glacial cycles
+before the mid-Pleistocene onset of 100\,ka glacial cycles
 (\cref{fig:paleo-timeseries}). However, only fragments of the last few
@@
 Canadian Cordillera, and some of the adjacent regions in north-western
-contiguous USA and Alaska (\cref{fig:map-cordillera}). The Cordilleran ice
+contiguous USA and Alaska (\cref{fig:map-cordillera}). The LGM Cordilleran ice
 sheet originated from the coalescence of several mountain-centred ice caps
-\citep{Davis.Mathews.1944}. It is generally thought that this LGM configuration
+\citep{Davis.Mathews.1944}. It is generally thought that the ice sheet
+configuration
 was short-lived \citep{Clague.etal.1980, Clague.1985, Cosma.etal.2008}, a
@@
 Pacific continental shelf. An independent ice cap covering Vancouver Island
-diverted flow from the main body of ice to drain into the Strait of Georgia
-between Vancouver Island and the Olympic Peninsula, where it formed the Juan
-de Fuca marine ice lobe, calving into the Pacific Ocean \citep{Herzer.Bornhold.1982}. Further to the
-North, the ice margin encroached the Queen Charlotte Islands \citep{Blaise.etal.1990}. However, the
+diverted flow from the main body of ice to drain into the Strait of Georgia and
+then between Vancouver Island and the Olympic Peninsula, where it formed the
+Juan de Fuca marine ice lobe, calving into the Pacific Ocean
+\citep{Herzer.Bornhold.1982}. Further to the North, the ice margin encroached
+the Queen Charlotte Islands \citep{Blaise.etal.1990}. However, the
 archipelago may not have been fully ice-covered, even during the maximum of
 glaciation. Submarine troughs traversing several portions of the continental
-shelf ()\cref{fig:map-cordillera}) indicate the erosional action of additional ice streams, yet it is not
+shelf (\cref{fig:map-cordillera}) indicate the erosional action of additional
+ice streams, yet it is not
 clear whether they have formed during the last glacial cycle or during earlier
@@
 junction remains unclear \citep{Gowan.2013}, while limited information exists
-on its timing \citep[e.g.][]{Jackson.etal.1997, Bednarski.Smith.2007}. Because
+on its timing \citep[e.g.,][]{Jackson.etal.1997, Bednarski.Smith.2007}. Because
 the present thesis focuses specifically on the Cordilleran ice sheet, I,
@@
 rapid. Most of the evidence for the style and pace of deglaciation comes from
-the Fraser Plateau. Studies there have indicated that some of the highest
+the Interior Plateau of British Columbia. Studies there have indicated that
+some of the highest
 uplands emerged from the ice sheet prior to the surrounding lowlands, possibly
@@
 thesis, and to present an overview of the physics involved in these choices.
+
+Vectors are identified by bold arrow symbols ($\vv$) and second-order tensors by bold symbols ($\CST$), while scalars use a normal font ($s$).
+The divergence of a vector field is written $\div{\vv}$, and $\grad{s}$ denotes the gradient of a scalar field.
+The divergence of a tensor field is characterised by $\tdiv{\CST}$ and the gradient of a vector field by $\tgrad{\vv}$.
+The race of a tensor $\CST$ is written $\tr(\CST)$.
 Parameter values are summarised in \cref{tab:params}.
+
+\newpage  %FIXME temporary fix
 
@@
            well-defined boundaries in the model.
-           From \citet{PISM-authors.2014}.}
+           After \citet{PISM-authors.2014}.}
   \label{fig:model-interfaces}
@@
             & -
-            & - \\
+            & \citet{Aschwanden.etal.2013} \\
 
@@
             & \unit{m\,yr^{-1}}
-            & - \\
+            & \citet{Aschwanden.etal.2013} \\
 
@@
             & Pa
-            & - \\
+            & \citet{Aschwanden.etal.2013} \\
 
@@
             & -
-            & - \\
+            & \citet{Bueler.Pelt.2014} \\
 
@@
             & m
-            & - \\
+            & \citet{Aschwanden.etal.2013} \\
 
@@
     $\nu_m$ & Mantle viscosity
-            & 1\e21
+            & 1\e{21}
             & \unit{Pa\,s}
@@
     $D$     & Lithosphere flexural rigidity
-            & 5.0\e24
+            & 5.0\e{24}
             & \unit{N}
@@
 
-    $F_s$   & Degree-day factor for ice
+    $F_s$   & Degree-day factor for snow
             & 3.04\e{-3}
@@
 
-    $\gamma$& Air temperature lapse-rate
+    $\gamma$& Air temperature lapse rate
             & 6\e{-3}
@@
 The thermodynamic core of the ice sheet model relies on four elementary field
-equations: the equation of incompressibility, the balance of stresses, the constitutive law for ice, and the conservation of energy. Firstly, ice flow is assumed \emph{incompressible}. Hence, ice
-density, $\rho$, is constant, and
-conservation of mass therefore implies a conservation of volume, which can be expressed
+equations: the equation of incompressibility, the balance of stresses, the
+constitutive law for ice, and the conservation of energy. Firstly, ice flow is
+assumed \emph{incompressible}. Hence, ice density, $\rho$, is constant, and
+conservation of mass therefore implies a conservation of volume, which can be
+expressed
 in terms of the ice velocity vector, $\vv$ (\cref{fig:model-variables}), by
@@
 Clapeyron relation, ${T_{pa} = T - \beta p}$. $R$ is the ideal gas constant,
-and $A_c$, $A_w$, $Q_c$ and $Q_w$, are constant parameter corresponding to
+and $A_c$, $A_w$, $Q_c$ and $Q_w$, are constant parameters corresponding to
 values measured below and above a critical temperature threshold
@@
 Consequently, the enthalpy equation (\ref{eqn:enthalpy}) can be rewritten using
-a more familiar, cold-ice, temperature formulation,
+a perhaps more familiar, cold-ice, temperature formulation,
 \begin{equation}
@@
 \end{equation}
-However, note that, here, the term ``enthalpy'' is actually an abuse of
+Note that, here, the term ``enthalpy'' is actually an abuse of
 language, commonly used in glaciology as a synonym of ``interval energy''
@@
 can not be computed in a direct manner. Instead, \cref{eqn:vssa} must be
-resolved numerically using iteration methods, in order to obtain
+resolved numerically using an iteration method, in order to obtain
 approximated values for $\vssa$.
@@
 
-Finally, the till friction angle, $\phi$ (\cref{eqn:coulomb}), is taken as a
+Finally, the till friction angle, $\phi$ (Eq.~\ref{eqn:coulomb}), is taken as a
 function of modern bed
@@
 varying degrees of isolation from atmospheric temperatures by ice sheet
-coverage, penetrate several kilometres
+coverage penetrate several kilometres
 into the rock, and eventually affect basal ice temperatures.
@@
 the left-hand side of \cref{eqn:beddef}, the first term accounts for mantle
-relaxation, the second for point-wise isostasy, and the third for elastic
+relaxation, the second for local isostasy, and the third for elastic
 flexure. Due to high mantle viscosity, $\nu_m$, there exists a time lag between
@@
 forcing the PDD computation to rely on an idealised representation of the
-annual temperature cycle $\theta_{ac}$. Sub-annual temperature variability around
+annual temperature cycle, $\theta_{ac}$. Sub-annual temperature variability around
 the freezing point, however, significantly affects surface melt on a multi-year
@@
 serve as a proxy for the available energy for melt, define the amount of snow
-melt through a proportionality factor $F_s$. If all snow is melted, remaining
-PDDs are then used to melt ice, using a proportionality factor $F_i$
+melt through a proportionality factor, $F_s$. If all snow is melted, remaining
+PDDs are then used to melt ice, using another proportionality factor, $F_i$
 (\cref{fig:model-pdd}). The degree-day factors for snow, $F_s$, and ice, $F_i$
@@
 Coast and Rocky mountains of British Columbia \citep{Shea.etal.2009}. In
-{\CCLI}, the temperature standard deviation $\sPDD$ is a constant model
+{\CCLI}, the temperature standard deviation, $\sPDD$, is a constant model
 parameter. In {\CCYC}, it is computed from daily temperature values from the
@@
              anomalously low due to brutal application of cold climate
-             condition to ice-free bedrock at the inception of the run.}
+             conditions to ice-free bedrock at the inception of the run.}
     \label{fig:plot-lgm}
@@
 NARR climatology as input to simulations of the Cordilleran ice sheet through
-the last glacial cycle (\CCYC).
+the last glacial cycle (\cref{sec:ccyc}; \CCYC).
 
@@
 (\cref{fig:model-pdd,tab:params}), and the standard deviation of
-near-surface air temperature, $\sPDD$
-(\cref{eqn:calovgreve,tab:params}). Although the first two can only be
+near-surface air temperature, $\sPDD$ (Eq.~\ref{eqn:calovgreve} and
+\cref{tab:params}). Although the first two can only be
 deduced from surface mass-balance measurements on present-day glaciers
@@
 \subsection{Simulations of the last glacial cycle}
+\label{sec:ccyc}
 % ----------------------------------------------------------------------
@@
 
-The primary goal of my thesis has been to perform numerical reconstructions
+The primary goal of my doctoral project has been to perform numerical
+reconstructions
 of Cordilleran ice sheet history through the last glacial cycle, which forms
@@
 Cordilleran ice sheet over its last expansion cycle. An ice cap located over
-the Skeena Mountains acts as the primary nucleation center for the Cordilleran
-ice sheet. After the LGM the ice sheet shrinks rapidly such that the last
-remnants of the ice sheet towards the end of the deglaciation in the early
-Holocene reside once again in the Skeena Mountains.
+the Skeena Mountains acts as the primary nucleation centre for the Cordilleran
+ice sheet. After the LGM, the ice sheet shrinks rapidly such that its last
+remnants reside once again in the Skeena Mountains towards the end of the
+deglaciation in the early Holocene.
 
@@
 numerical reconstructions of the Cordilleran ice sheet during the LGM and
-subsequent deglacation. Modelling results by \citet{Gregoire.etal.2012}
-indicate that the Laurentide ice sheet buttress resulted in a thicker
-Cordilleran ice sheet. However, modelling alone will neither elucidate the
+subsequent deglacation. Numerical odelling results by
+\citet{Gregoire.etal.2012} indicate that the Laurentide ice sheet buttress
+resulted in a thicker Cordilleran ice sheet. However, numerical modelling alone
+will neither elucidate the
 nature nor the duration of the interaction between of the Cordilleran and
@@
 interior was presumed dryer than today due to the presence of ice caps on the
-Pacific Coast ranges, reinforcing the current topographic barrier. However,
+Pacific Coast ranges, reinforcing the current topographic barrier to eastward
+moisture transport. However,
 some regions were probably wetter as a side effect of modified atmospheric
@@
 distribution
-of en-glacial temperatures, and the location of basal melt areas, was assumed
+of englacial temperatures, and the location of basal melt areas, was assumed
 constant over the modelling domain. However, the geology of the study area
@@
 of a 50\,ka-long evolution. Peak glacial temperatures were lower than those
-needed sustain a constant ice sheet geometry because the ice sheet growth was
+needed to sustain a constant ice sheet geometry because ice sheet growth was
 finally inhibited by the deglacial transition to warmer temperatures.

### 2014-08-24 kappa.tex
Add space for Swedish abstract.
Reposition floats.
Add missing preamble information.

@@
 % Document properties
-%\title{Numerical modelling of the Cordilleran ice sheet}
-%\author{Julien Seguinot}
 \thesistitle{Numerical modelling of the Cordilleran ice sheet}
@@
 \university{Stockholm University}
-\printed{Universitetsservice US-AB, 2014.}
-%\issn{}
-%\isbn{}
-%\coverpic{Richard Swan. \textsf {www.richardswan.com}. Back cover portrait: Louise Tränk.}
-%\layout{Summary and Papers IV \& V: Alistair Auffret based on  \LaTeX~templates by Rickard Petterson.
-Published papers typeset by respective publishers, reprinted with permission.
-Graphics created using \textsf{Inkscape}, \textsf{R} and \textsf{OpenOffice Draw}.}
+\issn{1653-7211}
+\isbn{978-91-7447-973-7}
+\coverpic{Numerical reconstruction of the Cordilleran ice sheet at the Last
+          Glacial Maximum. Bedrock topography, ice surface topography, ice
+          surface velocity, and frozen bed areas.}
+\layout{Julien Seguinot based on {\LaTeX} templates by Rickard Petterson
+        (summary) and Copernicus Publications (Paper~IV). Published papers
+        typeset by Copernicus Publications (Paper~I) and the International
+        Glaciology Society (Papers~II and~III). Graphics created using
+        matplotlib, cartopy, {\LaTeX} and Inkscape.}
+\printed{Universitetsservice US-AB, Stockholm, Sweden, 2014.}
 
@@
 %\end{keyword}
-%\begin{svabstract}
-%\end{svabstract}
+
+\begin{svabstract}
+
+Hej!
+Här kommer sammanfattning på svenska. Här kommer sammanfattning på svenska.
+Här kommer sammanfattning på svenska. Här kommer sammanfattning på svenska.
+Här kommer sammanfattning på svenska. Här kommer sammanfattning på svenska.
+Här kommer sammanfattning på svenska. Här kommer sammanfattning på svenska.
+Här kommer sammanfattning på svenska. Här kommer sammanfattning på svenska.
+Här kommer sammanfattning på svenska. Här kommer sammanfattning på svenska.
+Här kommer sammanfattning på svenska. Här kommer sammanfattning på svenska.
+Här kommer sammanfattning på svenska. Här kommer sammanfattning på svenska.
+
+\end{svabstract}
+
+%\begin{svkeyword}
+%\end{svkeyword}
 
@@
 
-Vectors are identified by bold arrow symbols ($\vv$) and second-order tensors by bold symbols ($\CST$), while scalars use a normal font ($s$).
-The divergence of a vector field is written $\div{\vv}$, and $\grad{s}$ denotes the gradient of a scalar field.
-The divergence of a tensor field is characterised by $\tdiv{\CST}$ and the gradient of a vector field by $\tgrad{\vv}$.
-The race of a tensor $\CST$ is written $\tr(\CST)$.
-Parameter values are summarised in \cref{tab:params}.
-
-\newpage  %FIXME temporary fix
+Vectors are identified by bold arrow symbols ($\vv$) and second-order tensors
+by bold symbols ($\CST$), while scalars use a normal font ($s$). The divergence
+of a vector field is written $\div{\vv}$, and $\grad{s}$ denotes the gradient
+of a scalar field. The divergence of a tensor field is characterised by
+$\tdiv{\CST}$ and the gradient of a vector field by $\tgrad{\vv}$. The race of
+a tensor $\CST$ is written $\tr(\CST)$. Parameter values are summarised in
+\cref{tab:params}, \cpageref{tab:params}.
 
@@
 
-\begin{table*}
-  \caption{Constant parameters of the ice sheet model.}
-  \label{tab:params}
-  \begin{tabular*}{170mm}{@{\hspace{2em}}l@{\extracolsep{\fill}}lrll}
-    \tophline
-    Not.    & Name & Value & Unit & Source \\
-    \middlehline
-
-    \multicolumn{2}{l}{\emph{Ice material properties}} \\
-
-    $\rho$  & Ice density
-            & 910
-            & \unit{kg\,m^{-3}}
-            & \citet{Aschwanden.etal.2012} \\
-
-    $g$     & Standard gravity
-            & 9.81
-            & \unit{m\,s^{-2}}
-            & \citet{Aschwanden.etal.2012} \\
-
-    $n$     & Glen exponent
-            & 3
-            & -
-            & - \\
-
-    $A_c$   & Ice hardness coefficient cold
-            & 3.61\e{-13}
-            & \unit{Pa^{-3}\,s^{-1}}
-            & \citet{Paterson.Budd.1982} \\
-
-    $A_w$   & Ice hardness coefficient warm
-            & 1.73\e3
-            & \unit{Pa^{-3}\,s^{-1}}
-            & \citet{Paterson.Budd.1982} \\
-
-    $Q_c$   & Flow law activation energy cold
-            & 6.0\e4
-            & \unit{J\,mol^{-1}}
-            & \citet{Paterson.Budd.1982} \\
-
-    $Q_w$   & Flow law activation energy warm
-            & 13.9\e4
-            & \unit{J\,mol^{-1}}
-            & \citet{Paterson.Budd.1982} \\
-
-    $R$     & Ideal gas constant
-            & 8.31441
-            & \unit{J\,mol^{-1}\,K^{-1}}
-            & - \\
-
-    $T_c$   & Flow law critical temperature
-            & 263.15
-            & \unit{K}
-            & \citet{Paterson.Budd.1982} \\
-
-    $f$     & Flow law water fraction coeff.
-            & 181.25
-            & -
-            & \citet{Lliboutry.Duval.1985} \\
-
-    $\beta$ & Clapeyron constant
-            & 7.9\e{-8}
-            & \unit{K\,Pa^{-1}}
-            & \citet{Luthi.etal.2002} \\
-
-    $c_i$   & Ice specific heat capacity
-            & 2009
-            & \unit{J\,kg^{-1}\,K^{-1}}
-            & \citet{Aschwanden.etal.2012} \\
-
-    $c_w$   & Water specific heat capacity
-            & 4170
-            & \unit{J\,kg^{-1}\,K^{-1}}
-            & \citet{Aschwanden.etal.2012} \\
-
-    $k$     & Ice thermal conductivity
-            & 2.10
-            & \unit{J\,m^{-1}\,K^{-1}\,s^{-1}}
-            & \citet{Aschwanden.etal.2012} \\
-
-    $L$     & Water latent heat of fusion
-            & 3.34\e5
-            & \unit{J\,kg^{-1}\,K^{-1}}
-            & \citet{Aschwanden.etal.2012} \\
-
-    \multicolumn{2}{l}{\emph{Basal sliding}} \\
-
-    $q$     & Pseudo-plastic sliding exponent
-            & 0.25
-            & -
-            & \citet{Aschwanden.etal.2013} \\
-
-    $v_{th}$& Pseudo-plastic threshold velocity
-            & 100.0
-            & \unit{m\,yr^{-1}}
-            & \citet{Aschwanden.etal.2013} \\
-
-    $c_0$   & Till cohesion
-            & 0.0
-            & Pa
-            & \citet{Aschwanden.etal.2013} \\
-
-    $\alpha$& Effective pressure coefficient (\CCLI)
-            & 0.95
-            & -
-            & \citet{Aschwanden.etal.2013} \\
-
-    $\delta$& Effective pressure coefficient (\CCYC)
-            & 0.02
-            & -
-            & \citet{Bueler.Pelt.2014} \\
-
-    $e_0$   & Till reference void ratio
-            & 0.69
-            & -
-            & \citet{Tulaczyk.etal.2000} \\
-
-    $C_c$   & Till compressibility coefficient
-            & 0.12
-            & -
-            & \citet{Tulaczyk.etal.2000} \\
-
-    $W_{max}$ & Maximal till water thickness
-            & 2.0
-            & m
-            & \citet{Aschwanden.etal.2013} \\
-
-    $b_0$   & Altitude of max. friction angle
-            & 0
-            & m
-            & - \\
-
-    $b_1$   & Altitude of min. friction angle
-            & 200
-            & m
-            & - \\
-
-    $\phi_0$& Minimum friction angle
-            & 10/15
-            & \degree
-            & - \\
-
-    $\phi_1$& Maximum friction angle
-            & 30/45
-            & \degree
-            & - \\
-
-    \multicolumn{2}{l}{\emph{Bedrock and lithosphere}} \\
-
-    $\rho_b$& Bedrock density
-            & 3300
-            & \unit{kg\,m^{-3}}
-            & - \\
-
-    $c_b$   & Bedrock specific heat capacity
-            & 1000
-            & \unit{J\,kg^{-1}\,K^{-1}}
-            & - \\
-
-    $k_b$   & Bedrock thermal conductivity
-            & 3.0
-            & \unit{W\,m^{-1}\,K^{-1}}
-            & - \\
-
-    $q_G$   & Geothermal heat flux
-            & 70.0
-            & \unit{mW\,m^{-2}}
-            & - \\
-
-    $\nu_m$ & Mantle viscosity
-            & 1\e{21}
-            & \unit{Pa\,s}
-            & \citet{Lingle.Clark.1985} \\
-
-    $\rho_l$& Lithosphere density
-            & 3300
-            & \unit{kg\,m^{-3}}
-            & \citet{Lingle.Clark.1985} \\
-
-    $D$     & Lithosphere flexural rigidity
-            & 5.0\e{24}
-            & \unit{N}
-            & \citet{Lingle.Clark.1985} \\
-
-    \multicolumn{2}{l}{\emph{Surface and atmosphere}} \\
-
-    $T_0$   & Temperature of snow precipitation
-            & 0.0
-            & \degC
-            & - \\
-
-    $T_1$   & Temperature of rain precipitation
-            & 2.0
-            & \degC
-            & - \\
-
-    $F_s$   & Degree-day factor for snow
-            & 3.04\e{-3}
-            & \unit{m\,K^{-1}\,day^{-1}}
-            & \citet{Shea.etal.2009} \\
-
-    $F_i$   & Degree-day factor for ice
-            & 4.59\e{-3}
-            & \unit{m\,K^{-1}\,day^{-1}}
-            & \citet{Shea.etal.2009} \\
-
-    $\gamma$& Air temperature lapse rate
-            & 6\e{-3}
-            & \unit{K\,m{-1}}
-            & - \\
-
-    \bottomhline
-  \end{tabular*}
-\end{table*}
+\begin{figure}
+  \includegraphics{model-variables}
+  \caption{Sketch of notations used for the main variables of the ice sheet
+           model. An infinitesimal ice parcel is moving at velocity $\vv$,
+           under the balance of stresses $\CST$ and gravity $\vect{g}$.}
+  \label{fig:model-variables}
+\end{figure}
 
@@
 \end{equation}
-
-\begin{figure}
-  \includegraphics{model-variables}
-  \caption{Sketch of notations used for the main variables of the ice sheet
-           model. An infinitesimal ice parcel is moving at velocity $\vv$,
-           under the balance of stresses $\CST$ and gravity $\vect{g}$.}
-  \label{fig:model-variables}
-\end{figure}
 
@@
 % ----------------------------------------------------------------------
+
+\begin{table*}
+  \caption{Constant parameters of the ice sheet model.}
+  \label{tab:params}
+  \begin{tabular*}{170mm}{@{\hspace{2em}}l@{\extracolsep{\fill}}lrll}
+    \tophline
+    Not.    & Name & Value & Unit & Source \\
+    \middlehline
+
+    \multicolumn{2}{l}{\emph{Ice material properties}} \\
+
+    $\rho$  & Ice density
+            & 910
+            & \unit{kg\,m^{-3}}
+            & \citet{Aschwanden.etal.2012} \\
+
+    $g$     & Standard gravity
+            & 9.81
+            & \unit{m\,s^{-2}}
+            & \citet{Aschwanden.etal.2012} \\
+
+    $n$     & Glen exponent
+            & 3
+            & -
+            & - \\
+
+    $A_c$   & Ice hardness coefficient cold
+            & 3.61\e{-13}
+            & \unit{Pa^{-3}\,s^{-1}}
+            & \citet{Paterson.Budd.1982} \\
+
+    $A_w$   & Ice hardness coefficient warm
+            & 1.73\e3
+            & \unit{Pa^{-3}\,s^{-1}}
+            & \citet{Paterson.Budd.1982} \\
+
+    $Q_c$   & Flow law activation energy cold
+            & 6.0\e4
+            & \unit{J\,mol^{-1}}
+            & \citet{Paterson.Budd.1982} \\
+
+    $Q_w$   & Flow law activation energy warm
+            & 13.9\e4
+            & \unit{J\,mol^{-1}}
+            & \citet{Paterson.Budd.1982} \\
+
+    $R$     & Ideal gas constant
+            & 8.31441
+            & \unit{J\,mol^{-1}\,K^{-1}}
+            & - \\
+
+    $T_c$   & Flow law critical temperature
+            & 263.15
+            & \unit{K}
+            & \citet{Paterson.Budd.1982} \\
+
+    $f$     & Flow law water fraction coeff.
+            & 181.25
+            & -
+            & \citet{Lliboutry.Duval.1985} \\
+
+    $\beta$ & Clapeyron constant
+            & 7.9\e{-8}
+            & \unit{K\,Pa^{-1}}
+            & \citet{Luthi.etal.2002} \\
+
+    $c_i$   & Ice specific heat capacity
+            & 2009
+            & \unit{J\,kg^{-1}\,K^{-1}}
+            & \citet{Aschwanden.etal.2012} \\
+
+    $c_w$   & Water specific heat capacity
+            & 4170
+            & \unit{J\,kg^{-1}\,K^{-1}}
+            & \citet{Aschwanden.etal.2012} \\
+
+    $k$     & Ice thermal conductivity
+            & 2.10
+            & \unit{J\,m^{-1}\,K^{-1}\,s^{-1}}
+            & \citet{Aschwanden.etal.2012} \\
+
+    $L$     & Water latent heat of fusion
+            & 3.34\e5
+            & \unit{J\,kg^{-1}\,K^{-1}}
+            & \citet{Aschwanden.etal.2012} \\
+
+    \multicolumn{2}{l}{\emph{Basal sliding}} \\
+
+    $q$     & Pseudo-plastic sliding exponent
+            & 0.25
+            & -
+            & \citet{Aschwanden.etal.2013} \\
+
+    $v_{th}$& Pseudo-plastic threshold velocity
+            & 100.0
+            & \unit{m\,yr^{-1}}
+            & \citet{Aschwanden.etal.2013} \\
+
+    $c_0$   & Till cohesion
+            & 0.0
+            & Pa
+            & \citet{Aschwanden.etal.2013} \\
+
+    $\alpha$& Effective pressure coefficient (\CCLI)
+            & 0.95
+            & -
+            & \citet{Aschwanden.etal.2013} \\
+
+    $\delta$& Effective pressure coefficient (\CCYC)
+            & 0.02
+            & -
+            & \citet{Bueler.Pelt.2014} \\
+
+    $e_0$   & Till reference void ratio
+            & 0.69
+            & -
+            & \citet{Tulaczyk.etal.2000} \\
+
+    $C_c$   & Till compressibility coefficient
+            & 0.12
+            & -
+            & \citet{Tulaczyk.etal.2000} \\
+
+    $W_{max}$ & Maximal till water thickness
+            & 2.0
+            & m
+            & \citet{Aschwanden.etal.2013} \\
+
+    $b_0$   & Altitude of max. friction angle
+            & 0
+            & m
+            & - \\
+
+    $b_1$   & Altitude of min. friction angle
+            & 200
+            & m
+            & - \\
+
+    $\phi_0$& Minimum friction angle
+            & 10/15
+            & \degree
+            & - \\
+
+    $\phi_1$& Maximum friction angle
+            & 30/45
+            & \degree
+            & - \\
+
+    \multicolumn{2}{l}{\emph{Bedrock and lithosphere}} \\
+
+    $\rho_b$& Bedrock density
+            & 3300
+            & \unit{kg\,m^{-3}}
+            & - \\
+
+    $c_b$   & Bedrock specific heat capacity
+            & 1000
+            & \unit{J\,kg^{-1}\,K^{-1}}
+            & - \\
+
+    $k_b$   & Bedrock thermal conductivity
+            & 3.0
+            & \unit{W\,m^{-1}\,K^{-1}}
+            & - \\
+
+    $q_G$   & Geothermal heat flux
+            & 70.0
+            & \unit{mW\,m^{-2}}
+            & - \\
+
+    $\nu_m$ & Mantle viscosity
+            & 1\e{21}
+            & \unit{Pa\,s}
+            & \citet{Lingle.Clark.1985} \\
+
+    $\rho_l$& Lithosphere density
+            & 3300
+            & \unit{kg\,m^{-3}}
+            & \citet{Lingle.Clark.1985} \\
+
+    $D$     & Lithosphere flexural rigidity
+            & 5.0\e{24}
+            & \unit{N}
+            & \citet{Lingle.Clark.1985} \\
+
+    \multicolumn{2}{l}{\emph{Surface and atmosphere}} \\
+
+    $T_0$   & Temperature of snow precipitation
+            & 0.0
+            & \degC
+            & - \\
+
+    $T_1$   & Temperature of rain precipitation
+            & 2.0
+            & \degC
+            & - \\
+
+    $F_s$   & Degree-day factor for snow
+            & 3.04\e{-3}
+            & \unit{m\,K^{-1}\,day^{-1}}
+            & \citet{Shea.etal.2009} \\
+
+    $F_i$   & Degree-day factor for ice
+            & 4.59\e{-3}
+            & \unit{m\,K^{-1}\,day^{-1}}
+            & \citet{Shea.etal.2009} \\
+
+    $\gamma$& Air temperature lapse rate
+            & 6\e{-3}
+            & \unit{K\,m{-1}}
+            & - \\
+
+    \bottomhline
+  \end{tabular*}
+\end{table*}
 
@@
 % ======================================================================
+\newpage
 \section{Results summary}
@@
 
+Next to the atmospheric forcing, another critical source of uncertainty in
+palaeo-ice sheet modelling is the choice of parameters used by the surface mass
+balance model \citep[e.g.,][]{Hebeler.etal.2008}. In our case, these parameters
+are the degree-day factors for snow, $F_s$, and ice, $F_i$
+(\cref{fig:model-pdd,tab:params}), and the standard deviation of
+near-surface air temperature, $\sPDD$ (Eq.~\ref{eqn:calovgreve} and
+\cref{tab:params}). Although the first two can only be
+deduced from surface mass-balance measurements on present-day glaciers
+\citep[e.g.,][]{Shea.etal.2009} and assumptions that these values are also
+valid for palaeo-simulations, $\sPDD$ can be estimated from temperature time
+series \citep[e.g.,][]{Fausto.etal.2011}.
+
+In \PSDV, I used near-surface air temperatures from the ERA-Interim reanalysis
+\citep{Dee.etal.2011} to compute a global, monthly distribution of $\sPDD$. The
+resulting grids show that, although often assumed constant in ice sheet
+modelling, $\sPDD$ exhibits significant spacial and temporal variation, most
+particularly in the polar regions (\cref{fig:plot-sdmap}; \PSDV, Fig.~1). Using
+a simplified surface mass balance model, I then demonstrated the effect of this
+variation on the computation of positive degree-days (\PSDV, Fig.~2), and
+surface mass-balance (\PSDV, Fig.~3). Neglecting this variability in ice sheet
+modelling introduces significant errors in modelled PDD and surface mass
+balance calculations. Therefore, the Cordilleran ice sheet simulations
+presented in \CCYC\ now include this effect.
+
 \begin{figure}
@@
 
-Next to the atmospheric forcing, another critical source of uncertainty in
-palaeo-ice sheet modelling is the choice of parameters used by the surface mass
-balance model \citep[e.g.,][]{Hebeler.etal.2008}. In our case, these parameters
-are the degree-day factors for snow, $F_s$, and ice, $F_i$
-(\cref{fig:model-pdd,tab:params}), and the standard deviation of
-near-surface air temperature, $\sPDD$ (Eq.~\ref{eqn:calovgreve} and
-\cref{tab:params}). Although the first two can only be
-deduced from surface mass-balance measurements on present-day glaciers
-\citep[e.g.,][]{Shea.etal.2009} and assumptions that these values are also
-valid for palaeo-simulations, $\sPDD$ can be estimated from temperature time
-series \citep[e.g.,][]{Fausto.etal.2011}.
-
-In \PSDV, I used near-surface air temperatures from the ERA-Interim reanalysis
-\citep{Dee.etal.2011} to compute a global, monthly distribution of $\sPDD$. The
-resulting grids show that, although often assumed constant in ice sheet
-modelling, $\sPDD$ exhibits significant spacial and temporal variation, most
-particularly in the polar regions (\cref{fig:plot-sdmap}; \PSDV, Fig.~1). Using
-a simplified surface mass balance model, I then demonstrated the effect of this
-variation on the computation of positive degree-days (\PSDV, Fig.~2), and
-surface mass-balance (\PSDV, Fig.~3). Neglecting this variability in ice sheet
-modelling introduces significant errors in modelled PDD and surface mass
-balance calculations. Therefore, the Cordilleran ice sheet simulations
-presented in \CCYC\ now include this effect.
-
 The analysis is extended in \PSDP, using re-analysed air

### 2014-08-25 kappa.tex
Add numerical implementation paragraph.

@@
 \begin{subequations}
+\label{eqn:vssa}
 \begin{gather}
-  \label{eqn:vssa}
   \begin{split}
@@
 % ----------------------------------------------------------------------
-%\subsection{Numerical implementation}
-% ----------------------------------------------------------------------
-
-%\todo{Add here a short section about discretization of the SIA and SSA, and
-%      their numerical implementation. The main subtlety, I think, lies in the
-%      fact that some computation need to use a ``staggered'' grid in order to
-%      ensure that we add SIA and SSA values corresponding to the same point in
-%      space. I need to check that. This section is also the place where to
-%      introduce geographic transformation parameters that I used throughout
-%      the thesis -- except in \cref{fig:map-cordillera}!}
-
-%\begin{table}
-%  \caption{Properties of the Canadian Atlas Lambert conformal conic projection
-%           (EPSG code~3978) and the GRS~1980 ellipsoid, and corresponding
-%           PROJ.4 parameter names, as used in all model runs, ice volume and
-%           area calculations, and most figures.}
-%  \label{tab:params}
-%  \begin{tabular*}{80mm}{@{\extracolsep{\fill}}lcc}
-%    \hline
-%    Parameter               & Name              & Value \\
-%    \hline
-%    Projection name         & \texttt{+proj}    & \texttt{lcc} \\
-%    Central meridian        & \texttt{+lon\_0}  & 95\degree W \\
-%    Latitude of origin      & \texttt{+lat\_0}  & 49\degree N \\
-%    First secant parallel   & \texttt{+lat\_1}  & 49\degree N \\
-%    Second secant parallel  & \texttt{+lat\_2}  & 77\degree N \\
-%    Semimajor radius        & \texttt{+a}       & 6378137\,m \\
-%    Reciprocal flattening   & \texttt{+rf}      & 298.257222101 \\
-%    \hline
-%  \end{tabular*}
-%\end{table}
-
-% ======================================================================
-%\section{Software implementation}
-% ======================================================================
-
-% ----------------------------------------------------------------------
-%\subsection{Computational workflow}
-% ----------------------------------------------------------------------
-%\todo{Figure. computational workflow}
-
-% ----------------------------------------------------------------------
-%\subsection{Contributions to PISM}
-% ----------------------------------------------------------------------
-%\todo{Table. PISM ``issues'' identified}
-
-% ----------------------------------------------------------------------
-%\subsection{Peripheral tools}
-% ----------------------------------------------------------------------
-%A Python PDD model\\
-%Pre-processing scripts for GRASS GIS\\
-%Plotting tools for PISM\\
+\subsection{Numerical implementation}
+% ----------------------------------------------------------------------
+
+In PISM, the SIA and the SSA are resolved numerically using centred, explicit
+finite difference methods. Ice velocities, $\vsia$ and $\vssa$, are computed on
+a staggered grid, i.e., a grid offset by half a grid cell in both directions of
+the $x$- and $y$-axes \citep{Bueler.Brown.2009, Winkelmann.etal.2011}. The
+computation of SSA velocities involves the numerical resolution of
+\cref{eqn:vssa}. After discretization, this can be thought of as the inversion
+of a large but sparse linear system of the form $\tens{A}\vssa=\vect{b}$. This
+sparse linear problem is resolved in parallel by numerical iteration methods
+implemented in PETSc \citep{Balay.etal.2014}. The simulations presented here
+use the generalized minimal residual method \citep[GMRES,][]{Saad.2003}. The
+vertically-averaged mass continuity equation is then treated as a
+diffusion-advection equation, where SIA velocities are diffusive and SSA
+velocities advective \citep{Bueler.Brown.2009, Winkelmann.etal.2011}. The
+horizontal grid uses the Canadian Atlas Lambert conformal conic projection
+(\cref{tab:proj}). Time-stepping is adaptive.
+
+\begin{table}
+  \caption{Properties of the Canadian Atlas Lambert conformal conic projection
+           (EPSG code~3978) and the GRS~1980 ellipsoid, and corresponding
+           PROJ.4 parameter names, as used in all model runs, ice volume and
+           area calculations, and most figures.}
+  \label{tab:proj}
+  \begin{tabular*}{80mm}{@{\extracolsep{\fill}}lcc}
+    \tophline
+    Parameter               & Name              & Value \\
+    \middlehline
+    Projection name         & \texttt{+proj}    & \texttt{lcc} \\
+    Central meridian        & \texttt{+lon\_0}  & 95\degree W \\
+    Latitude of origin      & \texttt{+lat\_0}  & 49\degree N \\
+    First secant parallel   & \texttt{+lat\_1}  & 49\degree N \\
+    Second secant parallel  & \texttt{+lat\_2}  & 77\degree N \\
+    Semimajor radius        & \texttt{+a}       & 6378137\,m \\
+    Reciprocal flattening   & \texttt{+rf}      & 298.257222101 \\
+    \bottomhline
+  \end{tabular*}
+\end{table}
 

### 2014-08-25 kappa.tex
Add Swedish abstract.

@@
           surface velocity, and frozen bed areas.}
-\layout{Julien Seguinot based on {\LaTeX} templates by Rickard Petterson
-        (summary) and Copernicus Publications (Paper~IV). Published papers
+\layout{Julien Seguinot based on {\LaTeX} templates by Rickard Petterson (cover
+        summary) and Copernicus Publications (Paper~IV). Published papers
         typeset by Copernicus Publications (Paper~I) and the International
@@
 
+\noindent
 This doctoral dissertation presents a study of the glacial history of the North
@@
 of mountain ranges.
-
 However, this maximum stage appears short-lived and out of balance with
@@
 
-Hej!
-Här kommer sammanfattning på svenska. Här kommer sammanfattning på svenska.
-Här kommer sammanfattning på svenska. Här kommer sammanfattning på svenska.
-Här kommer sammanfattning på svenska. Här kommer sammanfattning på svenska.
-Här kommer sammanfattning på svenska. Här kommer sammanfattning på svenska.
-Här kommer sammanfattning på svenska. Här kommer sammanfattning på svenska.
-Här kommer sammanfattning på svenska. Här kommer sammanfattning på svenska.
-Här kommer sammanfattning på svenska. Här kommer sammanfattning på svenska.
-Här kommer sammanfattning på svenska. Här kommer sammanfattning på svenska.
+\noindent
+I min avhandling presenterar jag den glaciala historien för de Nordamerikanska
+Kordillerabergen baserad på numerisk ismodellering kalibrerad mot fältdata.
+Området, som karaktäriseras av branta bergskedjor och mellanliggande lägre
+platåer, täcktes tidigare av en inlandsis, Kordilleraisen. Den oregelbundna
+topografin över vilken inlandsisen bildades har bidragit till att tidigare
+fältstudier främst har haft ett lokalt eller regionalt fokus och bland de
+Pleistocena istäckena hör därför Kordilleraisen till de sämst utredda när det
+gäller inlandsisens utbredning, volym och dynamik.
+
+Jag presenterar numeriska simuleringar som är baserade på approximationer av
+isflödesfysik och som möjliggör kvantitativa rekonstruktioner av istäckets
+utveckling. Simuleringarna visar att Kordilleraisen under dess senaste glaciala
+maxima till stor del styrdes av markanta regionala kontraster i temperatur- och
+nederbördsförhållanden samt dygnsmässiga temperaturvariationer.
+Inlandsisens största utbredning under den sista glaciala cykeln var kortlivad.
+Under största delen av den simulerade glaciala cykeln karaktäriserades
+nedisningsområdet av isfält och platåisar samt isolerade glaciärer i
+bergskedjorna, och störst isvolym fanns över Skeenafjället. Den modellerade
+Kordilleraisen verkar ha varit i konstant obalans med det ständigt förändrande
+klimatet. Inlandsisens respons på dessa klimatförändringar var därför vidare
+mer komplext än vad som har presenterats i tvådimensionella konceptuella
+modeller av hur inlandsisen växte till och smälte bort.
+
+I min avhandling demonstrerar jag den potential som numerisk ismodellering har
+för att kunna dra meningsfulla slutsatser om nedisningshistoriken och forna
+klimatförhållanden över glaciala cykler, under förutsättning att modelleringen
+kan kalibreras mot fältdata.
 

### 2014-08-25 backcover.tex
Add backcover imformation.

\documentclass{minimal}
\usepackage{times}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

\setlength{\textwidth}{115mm}

\begin{document}

\textbf{Backcover text}

During the last glacial cycle, continental ice sheets similar to those
currently occupying Greenland and Antarctica covered much of northern Eurasia
and North America. This study focuses on one of these former giants, the
Cordilleran ice sheet in north-western North America. Like other palaeo-ice
sheets of the Northern Hemisphere, the Cordilleran ice sheet left a
characteristic imprint on the landscape that formed its bed. For specific time
periods, this landscape record allows partial reconstructions of glacial
conditions, such as the former ice sheet extents or the directions of flow.
This thesis, however, tackles the problem of palaeo-ice sheet reconstruction
from a different angle; that of numerical ice sheet modelling based on
approximated physics of ice deformation, sliding, surface accumulation and
melt. Output from the numerical model is eventually compared with the landscape
record, with an emphasis on input climate conditions in terms of temperature,
precipitation, and temperature variability needed to simulate the former ice
sheet.

\bigskip

\textbf{Image text}

I was born and raised in the beautiful city of Lille in the flatlands of
ch'Northern France, 570\,km from the nearest glacier. I learned physics at
school and discovered programming on pocket calculators, while I gained a
fascination for icy landscapes during numerous summer hikes in the Alps. I then
studied Earth sciences at \emph{École normale supérieure} in Paris and
\emph{Université Joseph Fourier} in Grenoble, and was introduced to glacial geology at
\emph{Norges geologiske undersøkelse} in Trondheim, before moving to Sweden for
my PhD.

\end{document}

### 2014-08-25 kappa.tex
Add space for acknowledgements.
Disable colour links.
Add acknowledgements.

@@
 \usepackage{bm}
-\usepackage[T1]{fontenc}
+\usepackage[utf8]{inputenc}
+\usepackage[T5, T1]{fontenc}
 \usepackage{natbib}
 \usepackage{physics}
-\usepackage[colorlinks, citecolor=blue]{hyperref}
+\usepackage[hidelinks]{hyperref}
 \usepackage[sort]{cleveref}  % needs to be loaded after hyperref
@@
 }
-
-% Some general low-priority todos
-% -- use latex in matplotlib figures
-% -- homogeneize subfig label fonts
-% -- add software references
-% -- add Swedish abstract
-% -- add numerical implementation
 
@@
 \renewcommand{\urlprefix}[0]{}  % remove default "URL" prefix
+\onecolumn
+\begin{multicols}{2}  % ensure levelled end-of-text
 \bibliographystyle{copernicus}
 \bibliography{refs/references.bib}
+\end{multicols}
+
+% ======================================================================
+\clearpage
+\thispagestyle{plain}
+\hfill \textbf{\huge Acknowledgements}
+\vfill
+% ======================================================================
+
+\noindent This work would not have been possible without the help and
+involvement of many friends and colleagues. I am grateful to Arjen Stroeven, my
+main supervisor, for his continuous support through good and bad times. I most
+particularly thank him for his scrutinized work and tough criticism on all my
+texts, which is hopefully reflected in the quality of this thesis. I also would
+like to thank Irina Rogozhina for her fearless jump into my supervising team
+last year and constant support since then. She has been a great mentor, a good
+friend and a fantastic colleague, and I hope that we will continue to work
+together. I thank Qiong Zhang for her positiveness and encouragement, technical
+help with meteorological data and useful advice in my early writing steps, and
+Johan Kleman for few but always inspiring discussions.
+
+I also wish to particularly thank Jakob Heyman and Martin Margold, who have
+been great tutors on glacial geomorphology and many other things during the
+first years of my PhD. Likewise, I am immensely grateful to Ed Bueler and
+Constantine Khroulev for dedicated support on PISM during the last three years
+and for precious advice during my short visit at the University of Alaska
+Fairbanks.
+
+To my office mates; Carl Lilja, Markus Nathansson, Ping Fu and Natacha
+Gribenski; thank you for cultivating a friendly working atmosphere. I also wish
+to thank colleagues from the glaciology group for many entertaining paper
+discussions, empowered by the incorruptible commitment and delicious cookies of
+Patrick Applegate. Thanks to the administrative staff, and most particularly
+Susanna Blåndman, Carina Henricksson and Linus Richert, for having been
+constantly available for help, as well as to Helle Skånes, for guiding me
+through the last steps towards my doctoral defense.
+
+To colleagues from the German Research Centre for Geosciences, and especially
+to Maik Thomas, Jorge Bernales and Tonio; thank you for supporting my visit and
+welcoming me in Potsdam. To Ola Fredin and Lena Rubensdotter; thank you for
+introducing me to glacial geology during my stay in Trondheim in 2008, for
+encouraging me to pursue this path and for unforgettable fieldwork memories.
+
+To my friends from Stockholm and Potsdam; Max, Nathalie, Juri, Lisa, Reto,
+Peng, Ying, Nikki, Thomas, Viola; thank you for making these years enjoyable.
+Finally, I thank my family, and most specially my parents Lydie and Christophe,
+for having given me the chance to pursue a higher education, and for all those
+summers spent mountain-hiking in the Alps.
+{\fontencoding{T5}\selectfont Th\h{u}y}, your determination to push me to the
+finish line had no limits. Thank you for providing this amazing support, and
+for sharing my frustrations and my joys along the way.
+
+\bigskip
+\noindent\textbf{Funding sources}
+
+\noindent My salary was provided by Stockholm University, and research grants
+from the Swedish Research Council (VR) and the German Academic Exchange Service
+(DAAD). Additional funds for conference participations, course attendances,
+collaborative travels and fieldwork were provided by the Carl Mannerfelt Fund,
+the Bolin Centre for Climate Research, the Ahlmanns and Lagrelius Funds for
+Geographic Research, the De Geer Foundation for Quaternary Research and the
+Wallenberg Foundation.
 

### 2015-05-26 cordillera-cycle.tex
Black hyperlinks.
Update numbers in the text to CCYC4 simulations.

@@
 % Coloured hyperlinks
-\hypersetup{colorlinks, citecolor=blue}
+\hypersetup{hidelinks}
 
@@
   forcing, all simulations produce two major glaciations during
-  marine oxygen isotope stages~4 (61.9--55.4\,ka) and~2
-  (29.5--16.9\,ka). The timing of glaciation is
+  marine oxygen isotope stages~4 (61.9--56.5\,ka) and~2
+  (23.2--16.8\,ka). The timing of glaciation is
   better reproduced using temperature reconstructions from Greenland and
@@
   glacial cycle. It acts as a nucleation centre before the last glacial maximum
-  and hosts the last remains of Cordilleran ice during the
-  early Holocene (10.9--9.5\,ka).
+  and hosts the last remain of Cordilleran ice until the middle Holocene
+  (6.6--6.2\,ka).
 
@@
 
-    $\nu_m$ & Mantle viscosity
-            & $1\times10^{21}$
+    $\nu_m$ & Astenosphere viscosity
+            & $1\times10^{19}$
             & \unit{Pa\,s} \\
@@
 The simulations presented here were run using the Parallel Ice Sheet Model
-(PISM, development version~11b0a7f and stable version~0.6.1), an open source,
+(PISM, development version~8ff7cbe), an open source,
 finite difference, shallow ice sheet model \citep{PISM-authors.2014}. The model
@@
 a lower horizontal resolution of 10\,km, and a higher horizontal resolution of
-6\,km. These computations were performed on 16 to 128 computing cores at the
+5\,km. These computations were performed on 16 to 128 computing cores at the
 Swedish National Supercomputing Centre.
@@
 elevation, with the lowest value occuring below modern sea level (0\,m above
-sea level, m~a.s.l) and the highest value occuring above the generalised
+sea level, m~a.s.l.) and the highest value occuring above the generalised
 elevation of the highest shorelines
@@
 following a bedrock deformation model that includes local isostasy,
-elastic lithosphere flexure and viscous mantle deformation in an infinite
+plastic lithosphere flexure and viscous astenosphere deformation in an infinite
 half-space
 \citep[Table~\ref{tab:params};][]{Lingle.Clark.1985,Bueler.etal.2007}.
+A relatively low viscosity value of $\nu_m = 1\times10^{19}$\,\unit{Pa\,s} is
+used for the astenosphere in accordance to the results of regional glacial
+isostatic adjustment modelling at the northern Cascadia subduction zone
+\citep{James.etal.2009}.
 
@@
 variability \citep{Seguinot.2013}, $\sigma$ is computed from daily temperature
-values from the NARR, after excluding variability associated with the seasonal
-cycle itself \citep[Fig.~\ref{fig:atm}, cf.][]{Seguinot.Rogozhina.2014}.
+values from the NARR, including variability associated with the seasonal cycle
+itself \citep[Fig.~\ref{fig:atm}; i.e. not as in][]{Seguinot.Rogozhina.2014}.
 Degree-day factors for snow and ice melt are derived from
@@
            & Reference\\
-    & & & (m~a.s.l) & & & (K) & \\
+    & & & (m~a.s.l.) & & & (K) & \\
     \middlehline
@@
              & \chem{\delta^{18}O}
-             & 0.35 & -5.8{\degree}C  % -16.4126 (before scaling)
+             & 0.37 & -5.8{\degree}C  % -16.4126 (before scaling)
              & \citet{Dansgaard.etal.1993} \\
@@
              & \chem{\delta^{18}O}
-             & 0.22 & -6.1{\degree}C  % -26.7098
+             & 0.24 & -6.5{\degree}C  % -26.7098
              & \citet{Andersen.etal.2004} \\
@@
              & \chem{\delta^{18}O}
-             & 0.60 & -5.6{\degree}C  % -9.2055
+             & 0.64 & -5.9{\degree}C  % -9.2055
              & \citet{Jouzel.etal.2007} \\
@@
              & \chem{\delta^{18}O}
-             & 0.70 & -5.6{\degree}C  % -7.9550
+             & 0.74 & -5.9{\degree}C  % -7.9550
              & \citet{Petit.etal.1999} \\
@@
              & \chem{U^{K'}_{37}}
-             & 1.53 & -5.8{\degree}C  % -3.7889
+             & 1.61 & -6.1{\degree}C  % -3.7889
              & \citet{Herbert.etal.2001} \\
@@
              & \chem{U^{K'}_{37}}
-             & 1.16 & -5.8{\degree}C  % -5.0000
+             & 1.20 & -6.0{\degree}C  % -5.0000
              & \citet{Herbert.etal.2001} \\
@@
            time-volume span for ice volume extremes corresponding to MIS~4
-           (61.9--55.4\,ka), MIS~3 (52.2--45.6\,ka), and MIS~2 (LGM,
-           29.5--16.9\,ka). Dotted lines correspond to GRIP- and EPICA-driven
-           6\,km-resolution runs.}
+           (61.9--56.5\,ka), MIS~3 (53.0--41.3\,ka), and MIS~2 (LGM,
+           23.2--16.8\,ka). Dotted lines correspond to GRIP- and EPICA-driven
+           5\,km-resolution runs.}
   \label{fig:timeseries}
@@
     \middlehline
-    GRIP     &  57.58 &  49.24 &  19.52
-             &   1.98 &   0.46 &   2.13
-             &   7.52 &   0.89 &   8.52 \\
-    NGRIP    &  60.26 &  50.16 &  22.85
-             &   2.16 &   0.50 &   2.09
-             &   8.69 &   0.93 &   8.24 \\
-    EPICA    &  61.87 &  45.57 &  17.10
-             &   1.57 &   0.95 &   2.08
-             &   5.20 &   2.44 &   8.35 \\
-    Vostok   &  60.87 &  49.68 &  16.87
-             &   1.55 &   0.86 &   2.14
-             &   5.10 &   2.01 &   8.66 \\
-    ODP 1012 &  55.41 &  47.08 &  23.21
-             &   1.44 &   0.85 &   2.13
-             &   4.50 &   2.06 &   8.46 \\
-    ODP 1020 &  60.16 &  52.24 &  29.46
-             &   1.32 &   0.70 &   2.08
-             &   3.88 &   1.52 &   8.32 \\
+    GRIP     &  57.45 &  42.93 &  19.14
+             &   1.87 &   0.69 &   2.07
+             &   7.35 &   1.69 &   8.71 \\
+    NGRIP    &  60.25 &  45.86 &  22.85
+             &   2.11 &   0.74 &   2.10
+             &   8.84 &   1.82 &   8.76 \\
+    EPICA    &  61.89 &  45.30 &  17.23
+             &   1.51 &   1.01 &   2.09
+             &   5.11 &   2.74 &   8.91 \\
+    Vostok   &  60.86 &  41.25 &  16.84
+             &   1.51 &   1.03 &   2.01
+             &   5.15 &   2.88 &   8.40 \\
+    ODP 1012 &  56.46 &  47.39 &  23.19
+             &   1.38 &   0.90 &   2.07
+             &   4.39 &   2.35 &   8.62 \\
+    ODP 1020 &  60.23 &  52.97 &  20.56
+             &   1.25 &   0.72 &   2.09
+             &   3.84 &   1.75 &   8.81 \\
     \middlehline
-    Minimum  &  61.87 &  52.24 &  29.46
-             &   1.32 &   0.46 &   2.08
-             &   3.88 &   0.89 &   8.24 \\
-    Maximum  &  55.41 &  45.57 &  16.87
-             &   2.16 &   0.95 &   2.14
-             &   8.69 &   2.44 &   8.66 \\
+    Minimum  &  61.89 &  52.97 &  23.19
+             &   1.25 &   0.69 &   2.01
+             &   3.84 &   1.69 &   8.40 \\
+    Maximum  &  56.46 &  41.25 &  16.84
+             &   2.11 &   1.03 &   2.10
+             &   8.84 &   2.88 &   8.91 \\
     \bottomhline
@@
 glacial cycle, except during two major glacial events which occur between 61.9
-and 55.4\,ka during MIS~4, and between 29.5 and 16.9\,ka during MIS~2
+and 56.5\,ka during MIS~4, and between 23.2 and 16.8\,ka during MIS~2
 (Fig.~\ref{fig:timeseries}, lower panel). An ice volume minimum is
-consistently reached between 52.2 and 45.6\,ka during MIS~3. However, the
+consistently reached between 53.0 and 41.3\,ka during MIS~3. However, the
 magnitude and precise timing of these three events depend significantly on the
@@
 While simulations driven by the GRIP and the two Antarctic palaeo-temperature
-records attain a last ice volume maximum between 19.5 and 16.9\,ka, those
+records attain a last ice volume maximum between 19.1 and 16.8\,ka, those
 informed by the NGRIP and the two oceanic palaeo-temperature records attain their
 maximum ice volumes thousands of years earlier. Moreover, the ODP~1012 run
-yields a rapid deglaciation of the modelled area prior to 20\,ka. The ODP~1020
-simulation predicts an early maximum in ice volume at 29.5\,ka, followed by
+yields a rapid deglaciation of the modelled area prior to 17\,ka. The ODP~1020
+simulation predicts an early maximum in ice volume at 23.2\,ka, followed by
 slower deglaciation than modelled using the other palaeo-temperature records.
@@
 Corresponding maximum ice volumes also differ significantly between model runs,
-and vary between 3.88 and~8.69\,m sea level equivalents (m.~s.l.e;
+and vary between 3.84 and~8.84\,m sea level equivalents (m.~s.l.e.;
 Table~\ref{tab:extrema}).
@@
 the applied palaeo-temperature record. Modelled ice volume minima spread over a
-wide range between 0.89 and 2.44\,m~s.l.e. (Table~\ref{tab:extrema}).
+wide range between 1.69 and 2.88\,m~s.l.e. (Table~\ref{tab:extrema}).
 
@@
 (Table~\ref{tab:records}), it is interesting to note that modelled maximum ice
-volumes also fall within a tight range of 8.24 to 8.66\,m~s.l.e.
+volumes also fall within a tight range of 8.40 to 8.91\,m~s.l.e.
 (Table~\ref{tab:extrema}).
@@
 glacial maxima during the last glacial cycle. The first maximum configuration
-is obtained during MIS~4 (61.9--55.4\,ka) and the second during MIS~2
-(29.5--16.9\,ka; Figs.~\ref{fig:timeseries}, \ref{fig:snapshots};
+is obtained during MIS~4 (61.9--56.5\,ka) and the second during MIS~2
+(23.2--16.8\,ka; Figs.~\ref{fig:timeseries}, \ref{fig:snapshots};
 Table~\ref{tab:extrema}). These events broadly correspond in timing to the
@@
 early onset of deglaciation (Fig.~\ref{fig:timeseries}) following its last
-glacial volume maximum attained about 6\,ka earlier (-22.9\,ka,
+glacial volume maximum attained about 6\,ka earlier (22.9\,ka,
 Table~\ref{tab:extrema}) than dated evidence of the last glacial maximum
@@
            from the Alaska Range (AR) to the southern Coast Mountains (CM) and
-           the Columbia and Rocky mountains (RM) exists for about 29\,ka in
-           both simulations. The maximal extent of the ice sheet generally
+           the Columbia and Rocky mountains (RM) exists for about 32\,ka in
+           the GRIP simulation and 26\,ka in the EPICA simulation.
+           The maximal extent of the ice sheet generally
            corresponds to relatively short durations of ice cover, but ice
@@
 northeast to the Rocky Mountains in the southwest, is only in operation
-for at most 29\,ka, which is less than 25\,\% of a glacial cycle
+for at most 32\,ka, which is about a third of the modelled glacial cycle
 (Fig.~\ref{fig:duration}, hatched areas). However,
@@
            and surface velocity (colour mapping) during the last deglaciation
-           from the GRIP (top panels) and EPICA (bottom panels) simulations.
+           from the GRIP (top panels) and EPICA (bottom panels) 5\,km simulations.
            Dashed segments indicate the location of profiles used in
@@
 and~\ref{fig:deglacseries}). Adopting the EPICA climate forcing yields peak ice
-volume at 17.3\,ka and an uninterrupted deglaciation until 10.6\,ka
+volume at 17.2\,ka and an uninterrupted deglaciation until about 9\,ka
 (Fig.~\ref{fig:deglacseries}, lower panel, red curves). On the contrary, the
 simulation driven by the GRIP palaeo-temperature record yields peak ice volume
-at 19.1\,ka and a deglaciation interrupted by two ice volume standstills until
-9.4\,ka. The first interruption occurs between 16.6 and 14.5\,ka, and the
-second between 12.6 and 11.6\,ka (Fig.~\ref{fig:deglacseries}, lower panel,
+at 19.3\,ka and a deglaciation interrupted by two ice volume standstills until
+about 8\,ka. The first interruption occurs between 16.6 and 14.5\,ka, and the
+second between 13.1 and 11.6\,ka (Fig.~\ref{fig:deglacseries}, lower panel,
 blue curve).
@@
 with other climate drivers, differ in detail. On the one hand, the EPICA run
-depicts peak glaciation almost 2\,ka later than the GRIP run, in closer
+depicts peak glaciation about 2\,ka later than the GRIP run, in closer
 agreement with dated maximum extents, and shows a faster, uninterrupted
-deglaciation which yields ice-free conditions more than 1\,ka earlier. On the
+deglaciation which yields sporadic ice cover more than 1\,ka earlier. On the
 other hand, the GRIP run yields a deglaciation in three steps, compatible with
@@
 and Rocky mountains, and the Interior Plateau of British Columbia, becomes
-completely deglaciated by 12\,ka, whereas a significant ice cover remains over
+completely deglaciated by 10\,ka, whereas a significant ice cover remains over
 the Skeena, Selwyn, Mackenzie, Wrangell and Saint-Elias mountains in the
-northern sector of the modelling domain. After 12\,ka, deglaciation continues
+northern sector of the modelling domain. After 10\,ka, deglaciation continues
 to proceed across the Liard Lowland with a radial ice margin retreat towards
@@
 decay by retreating towards the Selwyn and Skeena mountains. The last remnants
-of the Cordilleran ice sheet finally disappear from the Skeena Mountains around
-10.6\,ka (EPICA) and 9.4\,ka (GRIP).
+of the Cordilleran ice sheet finally disappear from the Skeena Mountains at
+6.5\,ka (GRIP) and 6.1\,ka (EPICA).
 
@@
   \caption{Modelled bedrock (black) and ice surface (blue) topography profiles
-           during deglaciation (22.0--8.0\,ka) in the GRIP 6\,km
+           during deglaciation (22.0--8.0\,ka) in the GRIP 5\,km
            simulation, corresponding to the four transects indicated in
@@
   \caption{Modelled bedrock (black) and ice surface (red) topography profiles
-           during deglaciation (22.0--8.0\,ka) in the EPICA 6\,km
+           during deglaciation (22.0--8.0\,ka) in the EPICA 5\,km
            simulation, corresponding to the four transects indicated in
@@
 cycle presented in this study consistently produce two glacial maxima during
-MIS~4 (61.9--55.4\,ka, 3.8--8.7\,m~s.l.e) and MIS~2 (29.5--16.9\,ka,
-8.24--8.66\,m~s.l.e.), two periods corresponding to documented extensive
+MIS~4 (61.9--56.5\,ka, 3.8--8.8\,m~s.l.e.) and MIS~2 (23.2--16.8\,ka,
+8.40--8.91\,m~s.l.e.), two periods corresponding to documented extensive
 glaciations. This result is independent of the palaeo-temperature record used
@@
 towards the Selwyn and, finally, the Skeena mountains, which hosts the last
-remnant of the ice sheet during the early Holocene (10.9--9.5\,ka). Our results
+remnant of the ice sheet during the middle Holocene (6.6--6.2\,ka). Our results
 identify the Skeena Mountains as a key area to understanding glacial dynamics

### 2015-05-26 cordillera-cycle.tex
Enhance profile labelling using capital letters.

@@
            simulation, corresponding to the four transects indicated in
-           Figs.~\ref{fig:deglacshots}--\ref{fig:lastflow}.}
+           Figs.~\ref{fig:deglacshots} and~\ref{fig:deglac}.}
   \label{fig:profiles-grip}
@@
            simulation, corresponding to the four transects indicated in
-           Figs.~\ref{fig:deglacshots}--\ref{fig:lastflow}.}
+           Figs.~\ref{fig:deglacshots} and~\ref{fig:deglac}.}
   \label{fig:profiles-epica}

### 2015-05-26 figures/deglac.py

@@ -62,9 +62,11 @@ for i, rec in enumerate(records):
     add_corner_tag(ax, rec.upper())
 
     # add profile lines
-    for yp in [1.7e6, 1.4e6, 1.1e6, 0.8e6]:
+    for k, yp in enumerate([1.7e6, 1.4e6, 1.1e6, 0.8e6]):
         ax.plot([-2.4e6, -1.25e6], [yp, yp], 'k|',
                          lw=0.25, ls='--', dashes=(2, 2))
+        ax.text(-1.225e6, yp, chr(65+k), ha='left', va='bottom')
+
 
 # add colorbar and save
 print 'saving deglac...'

### 2015-05-26 figures/deglacshots.py

@@ -47,9 +47,11 @@ for i, rec in enumerate(records):
         add_corner_tag(ax, '%s kyr' % time[t])
 
         # add profile lines
-        for yp in [1.7e6, 1.4e6, 1.1e6, 0.8e6]:
+        for k, yp in enumerate([1.7e6, 1.4e6, 1.1e6, 0.8e6]):
             ax.plot([-2.4e6, -1.25e6], [yp, yp], 'k|',
                     lw=0.25, ls='--', dashes=(2, 2))
+            if j==3:
+                ax.text(-1.2e6, yp, chr(65+k), ha='left', va='bottom')
 
     # add record label
     add_corner_tag(ax, rec.upper(), va='bottom')

### 2015-05-26 figures/lastflow.py

@@ -78,11 +78,6 @@ for i, rec in enumerate(records):
     ax.contour(x[:], y[:], glaciated, levels=[0.5],
                colors='k', linewidths=0.5)
 
-    # add profile lines
-    for yp in [1.7e6, 1.4e6, 1.1e6, 0.8e6]:
-        ax.plot([-2.4e6, -1.25e6], [yp, yp], 'k|',
-                         lw=0.25, ls='--', dashes=(2, 2))
-
     # annotate
     add_corner_tag(ax, rec.upper())
     nc.close()

### 2015-05-26 figures/profiles.py

@@ -32,7 +32,8 @@ def profiles(res, rec, dt, color):
     kmin, kmax = [np.argmin(np.abs(time[:]*s2ka-t)) for t in (tmin, tmax)]
     if kmin == kmax:  # run has not reached tmin yet
         return fig
-    for ax, yp in zip(grid, yplist):
+    for i, yp in enumerate(yplist):
+        ax = grid[i]
         ax.set_rasterization_zorder(2.5)
         j = np.argmin(np.abs(y[:]-yp))
         xpf = x[:]*1e-3
@@ -45,7 +46,7 @@ def profiles(res, rec, dt, color):
         ax.grid(axis='y', c='0.5', ls='-', lw=0.1)
         ax.plot(xpf, surfpf.T, c=color, lw=0.1)
         ax.plot(xpf, topgpf.T, c='k', lw=0.1)
-        ax.text(0.04, 0.8, 'y = %i km' % (yp*1e-3), transform=ax.transAxes)
+        ax.text(0.04, 0.84, chr(65+i), transform=ax.transAxes)
     nc.close()
 
     # set axes properties

### 2015-05-26 cordillera-cycle.tex
Update affiliations and email adress.

@@
 % Authors
-\Author[1,2]{J.}{Seguinot}
-\Author[2]{I.}{Rogozhina}
-\Author[1]{A.~P.}{Stroeven}
-\Author[3]{M.}{Margold}
-\Author[1]{J.}{Kleman}
+\Author[1,2,3]{J.}{Seguinot}
+\Author[3]{I.}{Rogozhina}
+\Author[2]{A.~P.}{Stroeven}
+\Author[4]{M.}{Margold}
+\Author[2]{J.}{Kleman}
 \runningauthor{J.~Seguinot et~al.}
-\correspondence{J.~Seguinot (julien.seguinot@natgeo.su.se)}
+\correspondence{J.~Seguinot (seguinot@vaw.baug.ethz.ch)}
 
@@
 % Affiliations
-\affil[1]{Department of Physical Geography and Quaternary Geology and the
-          Bolin Centre for Climate Research, Stockholm University,
-          Stockholm, Sweden}
-\affil[2]{Helmholtz Centre Potsdam, GFZ German Research Centre for Geosciences,
+\affil[1]{Laboratory of Hydraulics, Hydrology and Glaciology, ETH Zürich,
+          Zürich, Switzerland}
+\affil[2]{Department of Physical Geography and the Bolin Centre for Climate
+          Research, Stockholm University, Stockholm, Sweden}
+\affil[3]{Helmholtz Centre Potsdam, GFZ German Research Centre for Geosciences,
           Potsdam, Germany}
-\affil[3]{Department of Geography, Durham University, UK}
+\affil[4]{Department of Geography, Durham University, UK}
 

### 2015-05-27 cordillera-cycle.tex
Update text and improve phrasing (Arjen).

@@
 However, former ice thicknesses and, therefore, the ice sheet's contribution to
-the LGM sea level low stand
+the LGM sea level lowstand
     \citep{Carlson.Clark.2012, Clark.Mix.2002}
@@
            map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
-           Data \citep{Patterson.Kelso.2014}.}
+           Data \citep{Patterson.Kelso.2015}.}
   \label{fig:locmap}
@@
            Menounos.etal.2008},
-it appears to be more sparse than for formerly glaciated regions surrounding
+it appears to be sparser than for formerly glaciated regions surrounding
 the North Atlantic
@@
      \citep{Boulton.etal.2001, Dyke.Prest.1987, Dyke.etal.2003,
-            Kleman.etal.1997, Kleman.etal.2010}.
-Here, we use a numerical ice sheet model \citep{PISM-authors.2014},
+            Kleman.etal.1997, Kleman.etal.2010, Stroeven.etal.inreview}.
+Here, we use a numerical ice sheet model \citep{PISM-authors.2015},
 calibrated against field-based evidence, to perform a quantitative
@@
 First, only recent development of numerical ice sheet models and underlying
-scientific computing tools \citep{Bueler.Brown.2009, Balay.etal.2014} has
+scientific computing tools \citep{Bueler.Brown.2009, Balay.etal.2015} has
 allowed for high-resolution numerical modelling of glaciers and ice sheets on
 mountainous terrain
-over millenial time scales \citep[e.g.,][]{Golledge.etal.2012}. Second, the
+over millennial time scales \citep[e.g.,][]{Golledge.etal.2012}. Second, the
 complex
@@
 (PISM, development version~8ff7cbe), an open source,
-finite difference, shallow ice sheet model \citep{PISM-authors.2014}. The model
+finite difference, shallow ice sheet model \citep{PISM-authors.2015}. The model
 requires input on basal topography, sea level, geothermal heat flux and
@@
 (cf. \citealp{Martin.etal.2011}; \citealp[supplement]{Aschwanden.etal.2013};
-\citealp{PISM-authors.2014}). Effective pressure, $N$,
+\citealp{PISM-authors.2015}). Effective pressure, $N$,
 is related to the ice overburden stress, $\rho gh$, and the modelled amount of
@@
 amount of water at the base, $W$, varies from zero to $W_{max}=2$\,m, a
-threshold above which instantaneous drainage is assumed.
-Finally, the bedrock topography responds to ice load
+threshold above which any additional melt water is assumed to drain off
+instantaneously. Finally, the bedrock topography responds to ice load
 following a bedrock deformation model that includes local isostasy,
-plastic lithosphere flexure and viscous astenosphere deformation in an infinite
+elastic lithosphere flexure and viscous astenosphere deformation in an infinite
 half-space
@@
            from the North American Regional Reanalysis
-           \citep[NARR;][]{Mesinger.etal.2006},
-           used to force the ice sheet model. Note the
+           \citep[NARR;][]{Mesinger.etal.2006}, used to force the surface mass
+           balance (PDD) component of the ice sheet model. Note the
            strong contrasts in seasonality, timing of the precipitation peak,
@@
 decreases to zero linearly with temperatures between 0 and
-2\,\unit{{\degree}C}. Ablation is computed from the number of positive
-degree-days (PDD), defined as the integral of temperatures above
-0\,\unit{{\degree}C} in one year.
+2\,\unit{{\degree}C}. Ablation is computed from the PDD, defined as the
+integral of temperatures above 0\,\unit{{\degree}C} in one year.
 
@@
 simulation driven by the GRIP palaeo-temperature record also results in a rapid
-deglaciation but in three steps, separated by two ice volume standstills
+deglaciation but in three steps, separated by two periods of ice volume increase
 (Fig.~\ref{fig:timeseries}).
@@
 ice sheet extents that pre-date field-based constraints by several thousands of
-years. Concerning the simulations driven by oceanic records, this early
+years (about 6, 6 and 4\,ka respectively).
+Concerning the simulations driven by oceanic records, this early
 deglaciation is caused by an early warming present in the alkenone
@@
 Table~\ref{tab:extrema}) than dated evidence of the last glacial maximum
-advance. Hence, we focus the rest of our analysis on simulations forced by
+advance. In addition, there exist a fair agreement between the EPICA and Vostok
+alaeo-temperature records, resulting in only little differences between the
+simulations driven by those records. These differences are not analysed
+further; instead we focus the rest of our analysis on simulations forced by
 palaeo-temperature records from the GRIP and EPICA ice cores that appear to
 produce the most realistic reconstructions of regional glaciation history among
-all records tested. To allow for a more
+all records tested, yet bearing significant disparities. To allow for a more
 detailed comparison against the geologic record, these two simulations were
@@
 ranges, which likely resulted in a decrease of precipitation and, therefore,
+a decrease of
 accumulation in the interior. Secondly, latent warming of the moisture-depleted
@@
 in the marine sediment record offshore Vancouver Island at
-ca.\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008}.
+ca.~45\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008}.
 
@@
            the GRIP simulation and 26\,ka in the EPICA simulation.
-           The maximal extent of the ice sheet generally
+           The maximum extent of the ice sheet generally
            corresponds to relatively short durations of ice cover, but ice
            cover persists over the Skeena Mountains (SM) during most of the
-           simulation.}
+           simulation. See Fig.~\ref{locmap} for a list of abbreviations.}
   \label{fig:duration}
@@
 necessarily represent the dominant patterns of glaciation throughout the period
-of ice cover \citep{Porter.1989, Kleman.Stroeven.1997}.
+of ice cover \citep{Porter.1989, Kleman.Stroeven.1997, Kleman.etal.2008,
+Kleman.etal.2010}.
 
@@
 Mountains (CM), the Columbia and Rocky mountains (CRM), the Skeena Mountains
-(SM), and the Selwyn and Mackenzie Mountains (SMKM). Although the Coast,
+(SM), and the Selwyn and Mackenzie mountains (SMKM). Although the Coast,
 Columbia, Rocky and Skeena mountains (CM, CRM, SM) are covered by mountain
@@
 fraction of the last glacial cycle. In other words, the Selwyn and Mackenzie
-mountains appear as a secondary ice-dispersal centre which ``activates'' only
+mountains appear as a secondary ice-dispersal centre which is only active
 during the coldest periods of the last glacial cycle. The Northern Rocky
-Mountains (Fig.~\ref{fig:duration}, NRM), do not act as a nucleation centre,
+Mountains (Fig.~\ref{fig:duration}, NRM) do not act as a nucleation centre,
 but rather as a pinning point for the Cordilleran ice sheet margin coming from
@@
 centres
-(Figs.~\ref{fig:snapshots} and~\ref{fig:duration}). Regardless of the applied
+(Figs.~\ref{fig:snapshots} and~\ref{fig:duration}, SM). Regardless of the applied
 forcing, this ice cap appears to survive MIS~3 (Fig.~\ref{fig:snapshots},
 middle panels), and serves as a nucleation centre at the onset of the glacial
-readvance towards the LGM (MIS~2). For the neighbouring Laurentide ice sheet,
-the importance of residual ice for North
-American glacial history leading up to the LGM has been illustrated by the
+readvance towards the LGM (MIS~2). This situation appear similar to the neighbouring Laurentide ice sheet, for which the importance of residual ice for
+glacial history leading up to the LGM has been illustrated by the
 MIS~3 residual ice bodies in northern and eastern Canada as nucleation centres
@@
 persistent ice cover (Fig.~\ref{fig:duration}) associated with basal ice
-temperatures at the pressure-melting point (Figs.~\ref{fig:warmfrac}
-and~\ref{fig:warmbase}) explains the large-scale glacial erosional imprint on
+temperatures at the pressure-melting point (Figs.~\ref{fig:warmbase}
+and~\ref{fig:warmfrac}) explains the large-scale glacial erosional imprint on
 the landscape. A well-developed network of glacial valleys west of
 the Mackenzie Mountains (\citealp[Fig.~2]{Kleman.etal.2010}; \citealp[Fig.~8]
-{Stroeven.etal.2010}) is modelled to have hosted warm-based ice
+{Stroeven.etal.2010}) is modelled to have hosted mainly warm-based ice
 (Fig.~\ref{fig:warmfrac}). However, because it is only glaciated for a
 short fraction of the last glacial cycle in our simulations
-(Fig.~\ref{fig:duration}), this perhaps appears indicate that
+(Fig.~\ref{fig:duration}), this perhaps indicates that
 the observed landscape pattern originates from multiple glacial cycles and
-witnesses an increased relative importance of this ice dispersal centre prior
+witnesses an increased relative importance of the Selwyn and Mackenzie
+mountains ice dispersal centre (Fig.~\ref{fig:duration}, SMKM), prior
 to the Late Pleistocene \citep[cf.][]{Ward.etal.2008, Demuro.etal.2012}.
@@
            green. Hatches denote re-advance of mountain-centred ice caps and
-           the decaying ice sheet between 14 and 10\,ka., which is more
-           pronounced in the GRIP-driven simulation.
+           the decaying ice sheet between 14 and 10\,ka., which is very
+           restricted in the EPICA-driven simulation.
            Dashed segments indicate the location of profiles used in
@@
 phases that are contemporary with warming oceanic temperatures from 17.2 to
-16.5 and from 15.5 to 14.0\,\unit{\chem{^{14}C}\,cal\,ka}.
+16.5 and from 15.5 to 14.0\,\unit{\chem{^{14}C}\,cal\,ka}
+\citep{Taylor.etal.2014}.
 
@@
 (Fig.~\ref{fig:deglac}, left panel). In addition to matching the location of
-local readvances, the GRIP-driven simulation shows that larger ice bodies
-emanating from the Skeena and Mackenzie mountains may still have existed at the
-time of this late-glacial readvance.
+local readvances, the GRIP-driven simulation shows a large remnant of the
+decaying ice sheet, centred on the Skeena Mountains, may still have existed
+at the time of this late-glacial readvance.
 In contrast, the EPICA-driven simulation produces a nearly-continuous
-deglaciation with only a tightly restricted late-glacial readvance
+deglaciation with only a tightly restricted late-glacial readvance on the
+western slopes of the Saint-Elias and Coast mountains
 (Fig.~\ref{fig:deglac}, right panel).
@@
   \includegraphics{lastflow}
-  \caption{Modelled of the deglacial basal ice velocities. Hatches
+  \caption{Modelled deglacial basal ice velocities. Hatches
            indicate areas that remain non-sliding throughout deglaciation
-           (22.0--8.0\,ka), notably including the Interior Plateau (IP).
+           (22.0--8.0\,ka), notably including parts of the Interior Plateau (IP).
            Note the concentric patterns of deglacial flow in the Liard
@@
 opposite, enlarge the region of negligible basal sliding
-(Fig.~\ref{fig:lastflow}). Thus, another explanation for the incongruent
-results could be that the Interior Plateau lineation system predates
-deglaciation, as perhaps indicated by some eskers that appear incompatible with
-these glacial lineations \citep[Fig.~9]{Margold.etal.2013a}. Alternatively,
+(Fig.~\ref{fig:lastflow}). Thus, a second explanation for the incongruent
+results could be that the Interior Plateau lineation system observed in the
+geomorphology predates the deglaciation, as perhaps indicated by some eskers
+that appear incompatible with these glacial lineations
+\citep[Fig.~9]{Margold.etal.2013a}. Finally, a third explanation could be that
 local geothermal heat associated with volcanic activity on the Interior Plateau
@@
 glaciations. This result is independent of the palaeo-temperature record used
-among the six selected for this study, and thus can be seen as a robust model
+among the six selected for this study, and thus can be regarded as a robust model
 output, which broadly matches geological evidence. However, the timing of the
@@
 drive the model. The timing of the LGM is best reproduced by the EPICA and
-Vostok Antarctic ice core records, and occurs too early in all other
-simulations. The mismatch is largest for the two northwest Pacific ODP
+Vostok Antarctic ice core records. It occurs about 2\,ka too early in the
+simulation forced by the GRIP ice core record, and even earlier in all other
+simulations. The mismatch is largest for the two Northwest Pacific ODP
 palaeo-temperature records, which are affected by the weakening of the
@@
 host an ice cap which appears to be fed by moisture intruding inland from the
-west through a topographic low in the Coast Mountains. The Skeena ice cap acts
-as the main nucleation centre at the onset of the LGM readvance. As indicated
+west through a topographic breach in the Coast Mountains. The Skeena ice cap acts
+as the main nucleation centre at the onset of the pre-LGM readvance. As indicated
 by persistent, warm-based ice in the model, this ice cap is perhaps associated
@@
 Plateau of central British Columbia. The GRIP-driven simulation then produces a
-late-glacial readvance of local ice caps in the Coast, Columbia and Rocky
-mountains, and of the main body of the decaying Cordilleran ice sheet in the
-Skeena and Selwyn mountains. In both simulations, this is followed by an
+late-glacial readvance of local ice caps and of the main body of the decaying
+Cordilleran ice sheet primarily in the Coast, Columbia and Rocky Mountains.
+In both simulations, this is followed by an
 opening of the Liard Lowland, and a final retreat of the remaining ice caps

### 2015-05-27 cordillera-cycle.tex
Add reference to Perkins and Brennand 2014.

@@
 older evidence has been overprinted by warm-based ice retreat
-during the deglaciation \citep{Kleman.1994, Kleman.etal.2010}. From a
+during the last deglaciation \citep{Kleman.1994, Kleman.etal.2010}. From a
 numerical modelling perspective, phases of glacier retreat are more challenging
@@
 inside a cold-based retreating margin (Fig.~\ref{fig:lastflow}). The
-modelled deglacial flow patterns appear consistent between the GRIP and EPICA
+modelled deglacial flow patterns are mostly consistent between the GRIP and EPICA
 simulations. They depict an active ice sheet retreat in the peripheral areas,
@@
 
-However, on the Interior Plateau of south-central British Columbia, both
-simulations predict negligible basal sliding during deglaciation
-(Fig.~\ref{fig:lastflow}), but an impressive set of glacial lineations
-indicate a substantial eastward flow component of the Cordilleran ice sheet
-\citep{Prest.etal.1968, Kleman.etal.2010}.
+On the Interior Plateau of south-central British Columbia, both simulations
+produce a retreat of the ice margin towards the north-east
+(Fig.~\ref{fig:deglac}), a pattern which is validated by the geomorphological
+and stratigraphical record for ancient pro-glacial lakes dammed by the
+retreating ice sheet \citep{Perkins.Brennand.2014}. However, the two
+simulations differ in the mode of retreat. The GRIP simulation yields an active
+retreat with basal sliding towards the ice margin to the south, whereas the
+EPICA simulation produce negligible basal sliding during deglaciation
+(Fig.~\ref{fig:lastflow}). Yet, the Interior Plateau also hosts an impressive
+set of glacial lineations which indicate a substantial eastward flow component
+of the Cordilleran ice sheet \citep{Prest.etal.1968, Kleman.etal.2010}.
 The Fraser lineation set could thus present a smoking gun for the

### 2015-05-27 cordillera-cycle.tex
Add reference to Troost (2014) GSA poster.

@@
            Stroeven.etal.2010, Stroeven.etal.2014}
-indicate an extensive Marine Oxygen Isotope Stage (MIS)~4 glaciation,
-yet it is not known whether other regions in the study area were affected.
+indicate an extensive Marine Oxygen Isotope Stage (MIS)~4 glaciation, of which
+evidence have also been found in the Puget Lowland \citep{Troost.2014}.
 Landforms in the interior regions include flow sets that are likely
@@
     \citep{Porter.Swanson.1998, Margold.etal.2014}.
-There is patchy stratigraphical evidence for glaciations older than the Fraser
-Glaciation \citep{Clague.Ward.2011} in British Columbia, but their extent and
+There is stratigraphical evidence for glaciations older than the Fraser
+Glaciation in British Columbia \citep{Clague.Ward.2011} and in the Puget
+Lowland \citep{Troost.2014}, but their extent and
 timing are still highly conjectural
@@
 ice surface elevations in the interior (Fig.~\ref{fig:mis4}, left panel).
-Thus, only the GRIP simulation can explain the presence of ice-rafted debris
+Thus, only the GRIP simulation can explain the presence of MIS~4 glacial
+deposits in the Puget Lowland \citep{Troost.2014} and that of ice-rafted debris
 in the marine sediment record offshore Vancouver Island at
-ca.~45\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008}.
+ca.~47\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008}.
 

### 2015-05-28 cordillera-cycle.tex
Add calving section.

@@
 \citep[Eqns.~7--9 and 15]{Bueler.Brown.2009, Winkelmann.etal.2011}.
+
 Ice rheology depends on temperature and water content through an enthalpy
@@
 \citep{James.etal.2009}.
+
+Ice-shelve calving is computed by a double criterion. First, a
+physically-realistic calving flux is computed based on eigenvalues of the
+horizontal strain rate tensor \citep{Winkelmann.etal.2011,
+Levermann.etal.2012}. This allows outlet glaciers to advance in confined
+embayments, but prevents extensive ice-shelves to build-up in the open ocean.
+Second, floating ice thinner than 50\,m is systematically calved off. A subgrid
+scheme \citep{Albrecht.etal.2011} allows for a continuous migration of the
+calving frontm. Such treatment of calving was applied to the Antarctic ice
+sheet and shown to produce realistic calving front positons for many of the
+ice-shelves \citep{Martin.etal.2011}.
 

### 2015-05-28 cordillera-cycle.tex
Add statement on min/max ice volumes in 5 km runs.

@@
 
+During MIS~2, total ice volume peaks at 8.01\,m~s.l.e. in the GRIP simulation
+and at 8.77\,m~s.l.e. in the EPICA simulation.
+
 \subsubsection{Ice configuration during MIS~4}
@@
 
+During MIS~4, total ice volume peaks at 7.19\,m~s.l.e. in the GRIP simulation
+and at 5.04\,m~s.l.e. in the EPICA simulation, corresponding to respectively
+90 and 57\,\% of the MIS~2 ice volumes.
+
 % -----------------------------------------------------------------------------
@@
 distribution of winter precipitation (Fig.~\ref{fig:atm}).
+
+During MIS~3, total ice volume corresponding to these persistent ice-dispersal
+centres reaches a minimum of 1.52\,m~s.l.e. in the GRIP simulation and of
+2.68\,m~s.l.e. in the EPICA simulation, corresponding to respectively 19 and
+31\,\% of the MIS~2 ice volumes.
 

### 2015-05-28 cordillera-cycle.tex
Add location tag for the Omineca Mountains.

@@
 Kovanen.Easterbrook.2002}. Although most of these moraines characterise
-independent valley glaciers, that were probably disconnected from the waning
-Cordilleran ice sheet, the Finlay River
-area of the northern Rocky Mountains presents a different kind of evidence.
+independent valley glaciers, that may have been disconnected from the waning
+Cordilleran ice sheet, the Finlay River area in the Omenica Mountains
+(Fig.~\ref{fig:deglac}, OM) presents a different kind of evidence.
 There, sharp-crested moraines indicate a late-glacial readvance of local alpine
@@
 to note that the simulation driven by the GRIP record produces a late-glacial
-readvance in the Coast Mountains, Rocky Mountains and the Finlay River area,
-corresponding to where it has been identified in the geological record
-(Fig.~\ref{fig:deglac}, left panel). In addition to matching the location of
-local readvances, the GRIP-driven simulation shows a large remnant of the
-decaying ice sheet, centred on the Skeena Mountains, may still have existed
-at the time of this late-glacial readvance.
+readvance in the Coast Mountains and Rocky Mountains (Fig.~\ref{fig:deglac},
+left panel). In addition to matching the location of some local readvances, the
+GRIP-driven simulation shows that a large remnant of the decaying ice sheet
+may still have existed at the time of this late-glacial readvance.
 In contrast, the EPICA-driven simulation produces a nearly-continuous

### 2015-05-28 figures/deglac.py

@@ -67,6 +67,8 @@ for i, rec in enumerate(records):
                          lw=0.25, ls='--', dashes=(2, 2))
         ax.text(-1.225e6, yp, chr(65+k), ha='left', va='bottom')
 
+# mark location of the Omineca mountains
+add_pointer_tag(fig.grid[0], 'OM', xy=(-1800e3, 1400e3), xytext=(-1300e3, 1250e3))
 
 # add colorbar and save
 print 'saving deglac...'

### 2015-05-28 cordillera-cycle.tex
Update acknowledgements.

@@
 scheme \citep{Albrecht.etal.2011} allows for a continuous migration of the
-calving frontm. Such treatment of calving was applied to the Antarctic ice
+calving front. Such treatment of calving was applied to the Antarctic ice
 sheet and shown to produce realistic calving front positons for many of the
@@
 \begin{acknowledgements}
+Foremost, we would like to thank Shawn Marshall for providing a detailed,
+constructive analysis of this study during J.~Seguinot's Ph.D. defence
+(September 2014). His comments were used to improve the model set-up.
 We are very thankful to Constantine Khroulev, Ed Bueler and Andy Aschwanden for
 providing constant help and development on PISM. This work was supported by the
-Swedish Research Council~(VR) grant no. 2008-3449 to A.~P.~Stroeven, by the
+Swedish Research Council~(VR) grant no.~2008-3449 to A.~P.~Stroeven and by the
 German Academic Exchange Service~(DAAD) grant no.~50015537 and a Knut and Alice
-Wallenberg Foundation grant to J.~Seguinot.
-Computer resources were provided by the Swedish National
-Infrastructure for Computing (SNIC) allocation no. 2013/1-159 to A.~P.~Stroeven
-at the National Supercomputing Center (NSC).
+Wallenberg Foundation grant to J.~Seguinot. Computer resources were provided
+by the Swedish National Infrastructure for Computing (SNIC) allocation
+no.~2013/1-159 and~2014/1-159 to A.~P.~Stroeven at the National Supercomputing
+Center (NSC).
 \end{acknowledgements}

### 2015-06-18 cordillera-cycle.tex
Update Martin's affiliation.
Add corrections from Irina (part 1).

@@
 \Author[2]{A.~P.}{Stroeven}
-\Author[4]{M.}{Margold}
+\Author[2]{M.}{Margold}
 \Author[2]{J.}{Kleman}
@@
           Potsdam, Germany}
-\affil[4]{Department of Geography, Durham University, UK}
 
@@
      \citealp{Booth.etal.2003}; \citealp{Dyke.2004}).
-The LGM Cordilleran ice sheet extent was short-lived, as indicated by
-field evidence from radiocarbon dating
+As indicated by field evidence from radiocarbon dating
     \citep{Clague.etal.1980, Clague.1985, Clague.1986, Porter.Swanson.1998,
@@
 and offshore sedimentary records
-    \citep{Cosma.etal.2008, Davies.etal.2011}.
+    \citep{Cosma.etal.2008, Davies.etal.2011},
+the LGM Cordilleran ice sheet extent was short-lived.
 However, former ice thicknesses and, therefore, the ice sheet's contribution to
@@
            Stroeven.etal.2010, Stroeven.etal.2014}
-indicate an extensive Marine Oxygen Isotope Stage (MIS)~4 glaciation, of which
-evidence have also been found in the Puget Lowland \citep{Troost.2014}.
+indicate an extensive Marine Oxygen Isotope Stage (MIS)~4 glaciation, the
+evidence of which have also been found in the Puget Lowland \citep{Troost.2014}.
 Landforms in the interior regions include flow sets that are likely
@@
 \begin{itemize}
-  \item How much ice was locked in the Cordilleran ice sheet during the
-    last glacial maximum?
-  \item What was the scale of glaciation prior to the last glacial maximum?
+  \item How much ice was locked in the Cordilleran ice sheet during the LGM?
+  \item What was the scale of glaciation prior to the LGM?
   \item Which were the primary dispersal centres? Do they reflect stable or
@@
 over millennial time scales \citep[e.g.,][]{Golledge.etal.2012}. Second, the
-complex
-topography of the North American Cordillera also induces strong gradients in
-temperature
-seasonality and distinct patterns of precipition, thus requiring the use of
-high-resolution climate forcing fields as an input to the ice sheet model
-\citep{Seguinot.etal.2014}.
-
-Because climate conditions over the last glacial cycle are subject to
-considerable uncertainty, our palaeo-climate forcing is a simplistic approach
-including temperature and precipitation fields derived from a
-present-day atmospheric reanalysis \citep{Mesinger.etal.2006,
-Seguinot.etal.2014} supplemented by lapse-rate corrections
+complex topography of the North American Cordillera also induces strong
+geographic variations in temperature and precipitation, thus requiring the use
+of high-resolution climate forcing fields as an input to an ice sheet model
+\citep{Seguinot.etal.2014}. However, evolving climate conditions over the last
+glacial cycle are subject to considerable uncertainty and still lie beyond the
+computational reach of atmosphere circulation models.
+
+Our palaeo-climate forcing therefore includes spatial temperature and
+precipitation grids derived from a present-day atmospheric reanalysis
+\citep{Mesinger.etal.2006} that includes the steep precipitation gradients
+previously identified as necessary to model the LGM extent of the Cordilleran
+ice sheet in agreement with its geological imprint
+\citep{Seguinot.etal.2014}. To mimic climate evolution through the last glacial
+cycle, these grids are simply supplemented by lapse-rate corrections
 and temperature offset time series. The latter are obtained by scaling six
@@
 output to geological evidence and discuss the timing and extent of glaciation
-and the patterns of deglaciation.
+and the patterns of deglaciation, based on which we use the applicability of
+different records to modelling the history of the Cordilleran ice sheet.
 
@@
 American Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006},
-perturbated by time-dependent and
+perturbated by time-dependent offsets and
 lapse-rate temperature corrections (Sect.~\ref{sec:atm}).
@@
 for the Shallow Ice Approximation (SIA) by adding velocity solutions of the
-two approximation
+two approximations
 \citep[Eqns.~7--9 and 15]{Bueler.Brown.2009, Winkelmann.etal.2011}.
@@
 consists of up to 51~enthalpy layers in the ice sheet and 31~temperature layers
-in the bedrock. In the high-resolution simulations, up to 101~ice layers, and
+in the bedrock. In the high-resolution simulations, up to 101~ice layers and
 61~bedrock layers are used.
@@
 subglacial water, using a formula derived from laboratory experiments with till
-extracted from an Antarctic ice stream \citep{Tulaczyk.etal.2000,
-Bueler.Pelt.2014},
+extracted from the base of Ice Stream B in West Antarctica
+\citep{Tulaczyk.etal.2000, Bueler.Pelt.2014},
 \begin{equation}
@@
 where $\delta$ is chosen as 0.02, $e_0$ is a measured reference void ratio and
-$C_c$ a measured compressibility coefficient (Table~\ref{tab:params}). The
+$C_c$ is a measured compressibility coefficient (Table~\ref{tab:params}). The
 amount of water at the base, $W$, varies from zero to $W_{max}=2$\,m, a
@@
 A relatively low viscosity value of $\nu_m = 1\times10^{19}$\,\unit{Pa\,s} is
-used for the astenosphere in accordance to the results of regional glacial
+used for the astenosphere in accordance to the results from regional glacial
 isostatic adjustment modelling at the northern Cascadia subduction zone
@@
 
-Ice-shelve calving is computed by a double criterion. First, a
+Ice-shelf calving is computed by a double criterion. First, a
 physically-realistic calving flux is computed based on eigenvalues of the
@@
 Levermann.etal.2012}. This allows outlet glaciers to advance in confined
-embayments, but prevents extensive ice-shelves to build-up in the open ocean.
+embayments, but prevents extensive ice-shelves from building up in the open ocean.
 Second, floating ice thinner than 50\,m is systematically calved off. A subgrid
 scheme \citep{Albrecht.etal.2011} allows for a continuous migration of the
-calving front. Such treatment of calving was applied to the Antarctic ice
-sheet and shown to produce realistic calving front positons for many of the
-ice-shelves \citep{Martin.etal.2011}.
+calving front. Such treatment of calving has been applied to the Antarctic ice
+sheet and has shown to produce a realistic calving front positon for many of
+the present-day ice shelves \citep{Martin.etal.2011}.
 
@@
 decreases to zero linearly with temperatures between 0 and
-2\,\unit{{\degree}C}. Ablation is computed from the PDD, defined as the
+2\,\unit{{\degree}C}. Ablation is computed from PDD, defined as an
 integral of temperatures above 0\,\unit{{\degree}C} in one year.
@@
 account for the effects of spatial and seasonal variations of temperature
-variability \citep{Seguinot.2013}, $\sigma$ is computed from daily temperature
-values from the NARR, including variability associated with the seasonal cycle
+variability \citep{Seguinot.2013}, $\sigma$ is computed from NARR daily
+temperature values from 1979 to 2000 \citep{Mesinger.etal.2006},
+including variability associated with the seasonal cycle
 itself \citep[Fig.~\ref{fig:atm}; i.e. not as in][]{Seguinot.Rogozhina.2014}.
@@
 
-Climate forcing of the model consists of a present-day monthly climatology,
+Climate forcing driving ice sheet simulations consists of a present-day monthly
+climatology,
 $\{T_{m0}, P_{m0}\}$, where temperatures are modified by offset time series,
@@
 magnitude and precise timing of these three events depend significantly on the
-choice of proxy record used for time-dependent climate forcing
+choice of proxy record used to derive a time-dependent climate forcing
 (Table~\ref{tab:extrema}).
@@
 (ODP~1012, ODP~1020) and Antarctic (EPICA, Vostok) palaeo-temperature records
-generally result in smaller
-modelled ice volumes during MIS~4 and larger ice volumes during MIS~3.
-Hence, they produce lower ice volume variability
-throughout the simulation length. The NGRIP climate forcing is the only one
+generally result in lower ice volume variability throughout the simulation
+length, resulting in lower modelled ice volumes during MIS~4 and larger ice
+volumes during MIS~3. The NGRIP climate forcing is the only one
 that results in a larger ice volume during MIS~4 than during MIS~2.
@@
 slower deglaciation than modelled using the other palaeo-temperature records.
-Finally, model runs forced by Antarctic palaeo-temperature records result in a
-rapid and uninterrupted deglaciation after the last glacial maximum, whereas the
+Finally, model runs forced by the Antarctic palaeo-temperature records result in a
+rapid and uninterrupted deglaciation after the LGM, whereas the
 simulation driven by the GRIP palaeo-temperature record also results in a rapid
-deglaciation but in three steps, separated by two periods of ice volume increase
+deglaciation but in three steps, separated by two periods of ice sheet regrowth
 (Fig.~\ref{fig:timeseries}).
@@
 
-Despite such differences in the timing of attained volume extrema
+Despite large differences in the timing of attained volume extrema
 (Table~\ref{tab:extrema}), all model runs show relatively consistent
@@
 
-During MIS~3 ice volume minima, a central ice cap persists over the Skeena
+During the MIS~3 ice volume minima, a central ice cap persists over the Skeena
 Mountains (Fig.~\ref{fig:snapshots}, middle panels). Although this ice cap is
@@
 about 3500\,m~a.s.l. along the spine of the Rocky Mountains. Although the
-similatity of modelled ice extents is a direct result from the choice of scaling
+similatity of modelled ice extents is a direct result of the choice of scaling
 factors applied to different palaeo-temperature proxy records

### 2015-06-19 cordillera-cycle.tex
Add corrections from Irina (part 2).

@@
 % Running title
-\runningtitle{Climate forcing for Cordilleran ice sheet simulations}
+\runningtitle{Numerical simulations of the Cordilleran ice sheet
+              through the last glacial cycle}
 
@@
 % =============================================================================
-\section{Comparison to the geologic record}
+\section{Comparison with the geologic record}
 \label{sec:discussion}
@@
 
-Large variations in the model response reveal its sensitivity to evolving
-climate forcing. Nevertheless, geological evidence of former glaciations
+Large variations in the model responses to evolving climate forcing reveal its
+sensitivity to the choice of palaeo-temperature proxy record. To distinguish
+between different records, geological evidence of former glaciations
 provide a basis for validation of our runs, while the results from numerical
 modelling can perhaps help to analyse some of the complexity of this evidence.
-In this section, we compare model output to the geologic record, in terms of
-timing and configuration of the maximum stages, location and durability of the
+In this section, we compare model outputs to the geologic record, in terms of
+timing and configuration of the maximum stages, location and lifetime of
 major nucleation centres, and patterns of ice retreat during the last
@@
 comparison with geological evidence for the timing of maximum Cordilleran ice
-sheet extent. In the Puget lowland (Fig.~\ref{fig:locmap}), the LGM advance of
+sheet extent. In the Puget Lowland (Fig.~\ref{fig:locmap}), the LGM advance of
 the southern Cordilleran ice sheet margin has been constrained by radiocarbon
@@
 palaeo-temperature records, on the contrary, yield MIS~2 maximum Cordilleran
-ice sheet extents that pre-date field-based constraints by several thousands of
+ice sheet volumes that pre-date field-based constraints by several thousands of
 years (about 6, 6 and 4\,ka respectively).
@@
 current \citep{Herbert.etal.2001}. The California current, driving cold
-waters southwards along the south-western coast of North America,
-has been shown to have weakened during each peak global glaciation (in SPECMAP)
+waters southwards along the south-western coast of North America, has been
+shown to have weakened during each peak of global glaciation (in SPECMAP)
 during the past 550\,ka, including the LGM, resulting in paradoxically warmer
@@
 through the last glacial cycle.
-
 Similarly, the simulation using the NGRIP palaeo-temperature record depicts an
 early onset of deglaciation (Fig.~\ref{fig:timeseries}) following its last
-glacial volume maximum attained about 6\,ka earlier (22.9\,ka,
-Table~\ref{tab:extrema}) than dated evidence of the last glacial maximum
-advance. In addition, there exist a fair agreement between the EPICA and Vostok
+glacial volume maximum (22.9\,ka, Table~\ref{tab:extrema}) attained about 6\,ka
+earlier than dated evidence of the LGM advance.
+There exist a fair agreement between the EPICA and Vostok
 alaeo-temperature records, resulting in only little differences between the
@@
 produce the most realistic reconstructions of regional glaciation history among
-all records tested, yet bearing significant disparities. To allow for a more
-detailed comparison against the geologic record, these two simulations were
-re-run using the high-resolution grid (Fig.~\ref{fig:timeseries}, lower panel,
-dotted lines).
+all records tested, yet bearing significant disparities in model output. To
+allow for a more detailed comparison against the geological record, these two
+simulations were re-run using a higher-resolution grid (Sect.~\ref{sec:model};
+Fig.~\ref{fig:timeseries}, lower panel, dotted lines).
 
@@
 
-This could reflect the fact that our model does not include
+Such deviation from the geological inferences could reflect the fact that our
+model does not include
 feedback mechanisms between ice sheet topography and the regional climate.
@@
 not have this problem, because the position and elevation of the ice divide
-results in significant ice drainage across the Rocky Mountains at the LGM
+ensure significant ice drainage across the Rocky Mountains at the LGM
 (Fig.~\ref{fig:mis2}).
 
-During MIS~2, total ice volume peaks at 8.01\,m~s.l.e. in the GRIP simulation
+During MIS~2, the modelled total ice volume peaks at 8.01\,m~s.l.e. in the GRIP
+simulation
 and at 8.77\,m~s.l.e. in the EPICA simulation.
@@
 results in a modelled maximum ice sheet extent that closely resembles that
-obtained
-during MIS~2, with the only major difference being slightly less extensive
+obtained during MIS~2, with the only major difference of being slightly less
+extensive across
 northern and eastern sectors. In contrast, the EPICA simulation produces a lower
@@
 
-In the GRIP simulation, some sections of the MIS~4 maximum ice margin are
-slightly more extensive than their MIS~2 counterparts. This is notably the case
-in the Puget Lowland and along parts of the ice margin in the marine sectors,
-on the north slope of the Alaska Range and in Yukon Territory. However, these
-differences
-are small and generally correspond to a few grid cells only. Although such
-results may be corroborated by field evidence in some places
-\citep[e.g.][]{Ward.etal.2007}, considering the simplicity of the applied
-climate forcing, modelled differences in extent are not demonstrably
-significant.
-
-During MIS~4, total ice volume peaks at 7.19\,m~s.l.e. in the GRIP simulation
+During MIS~4, the modelled total ice volume peaks at 7.19\,m~s.l.e. in the GRIP
+simulation
 and at 5.04\,m~s.l.e. in the EPICA simulation, corresponding to respectively
@@
 
-A notable exception to the transient character of other sectors of the
+A notable exception to the transient character of the maximum extent of
 Cordilleran ice sheet
@@
 fluctuations results from a combination of low precipitation, high summer
-temperature and large temperature standard deviation for this sector of the
-modelling domain (Fig.~\ref{fig:atm}). This result could potentially explain
-the local distribution of glacial deposits, which indicates that these glaciers
+temperature and large temperature standard deviation in the plains of the
+Alaska Interior (Fig.~\ref{fig:atm}) which confines glaciation to the
+foothills of the mountains. This result could potentially explain
+the local distribution of glacial deposits, which indicates that glaciers
+flowing on the northern slope of the Alaska Range
 have remained small throughout the Pleistocene \citep{Kaufman.Manley.2004}.
@@
 Mountains (CM), the Columbia and Rocky mountains (CRM), the Skeena Mountains
-(SM), and the Selwyn and Mackenzie mountains (SMKM). Although the Coast,
+(SM), and the Selwyn and Mackenzie mountains (SMKM). The Wrangell and
+St.-Elias mountains, heavily glacierized at present, host an ice cap for the
+entire length of both simulations, but that ice cap does not appear to be a
+major feed to the Cordilleran ice sheet (Fig.~\ref{fig:duration}, WSEM).
+Although the Coast,
 Columbia, Rocky and Skeena mountains (CM, CRM, SM) are covered by mountain
@@
 Mackenzie mountains (SMKM), where ice cover on the highest peaks is limited to a
-fraction of the last glacial cycle. In other words, the Selwyn and Mackenzie
+small fraction of the last glacial cycle. In other words, the Selwyn and Mackenzie
 mountains appear as a secondary ice-dispersal centre which is only active
@@
 ice cover is the persistence of the Skeena Mountains ice cap throughout the
-entire last glacial cycle and its predominance over the other ice-dispersal
-centres
+entire last glacial period (ca.~100--10\,ka) and its predominance over the
+other ice-dispersal centres
 (Figs.~\ref{fig:snapshots} and~\ref{fig:duration}, SM). Regardless of the applied
@@
 middle panels), and serves as a nucleation centre at the onset of the glacial
-readvance towards the LGM (MIS~2). This situation appear similar to the neighbouring Laurentide ice sheet, for which the importance of residual ice for
+readvance towards the LGM (MIS~2). This situation appear similar to the neighbouring Laurentide ice sheet, for which the importance of residual ice for the
 glacial history leading up to the LGM has been illustrated by the
@@
 The presence of a Skeena Mountains ice cap during most of the last glacial
-cycle can be explained by a currently more eastwards-spread distribution of
+cycle can be explained by a more eastwards-spread distribution of
 winter precipitation there than along other transects of the
@@
 
-During MIS~3, total ice volume corresponding to these persistent ice-dispersal
+During MIS~3, the modelled total ice volume corresponding to these persistent
+ice-dispersal
 centres reaches a minimum of 1.52\,m~s.l.e. in the GRIP simulation and of
@@
 70\,\unit{mW\,m^{-2}} geothermal heat flux at 3\,km depth
-(Sect.~\ref{sec:icedyn}). However, both the Skeena Mountains and the area
-west of the Mackenzie Mountains are areas of higher-than-average geothermal
-heat flux \citep{Blackwell.Richards.2004}. We can therefore expect longer
+(Sect.~\ref{sec:icedyn}). However, the Skeena Mountains and the area west of
+the Mackenzie Mountains experience higher-than-average geothermal heat flux
+with measured values of ca.~80 and ca.~100\,\unit{mW\,m^{-2}}
+\citep{Blackwell.Richards.2004}. We can therefore expect even longer
 durations of warm-based ice cover for these areas if we were to include
@@
 models through simple parametrisations \citep[e.g.][]{Clason.etal.2012,
-Clason.etal.2014, Bueler.Pelt.2014}, if included at all. However, next after a
-matching of modelled ice margin positions with mapped end moraines
-\citep{Napieralski.etal.2007}, indicative of maximum ice sheet extents
+Clason.etal.2014, Bueler.Pelt.2014}, if included at all. However, next after
+the mapping of maximum ice sheet extents during MIS~2 and MIS~4
 (Sect.~\ref{sec:mis2} and~\ref{sec:mis4}), geomorphologically-based
-reconstructions of patterns of of ice sheet retreat during the last
+reconstructions of patterns of ice sheet retreat during the last
 deglaciation provide the second best source of evidence for the validation of
@@
 
-In our simulations, the timing of peak ice volume during the last glacial
-maximum and the pacing of deglaciation depend critically on the choice of
+In our simulations, the timing of peak ice volume during the LGM
+and the pacing of deglaciation depend critically on the choice of
 climate forcing (Table~\ref{tab:extrema}; Figs.~\ref{fig:timeseries}
@@
 simulation driven by the GRIP palaeo-temperature record yields peak ice volume
-at 19.3\,ka and a deglaciation interrupted by two ice volume standstills until
+at 19.3\,ka and a deglaciation interrupted by two phases of regrowth until
 about 8\,ka. The first interruption occurs between 16.6 and 14.5\,ka, and the
@@
 
-Hence, the two model runs, while similar in overall timing compared to runs
+Hence, the two model runs, while similar in overall timing compared to the runs
 with other climate drivers, differ in detail. On the one hand, the EPICA run
@@
 marine sediment sequences offshore Vancouver Island, where the distribution of
-ice-rafted debris indicate an ice margin retreat from the Georgia Strait in two
+ice-rafted debris indicates an ice margin retreat from the Georgia Strait in two
 phases that are contemporary with warming oceanic temperatures from 17.2 to

### 2015-06-19 cordillera-cycle.tex
Add corrections from Irina (part 3).

@@
 retreat with basal sliding towards the ice margin to the south, whereas the
-EPICA simulation produce negligible basal sliding during deglaciation
+EPICA simulation produces negligible basal sliding on the plateau during deglaciation
 (Fig.~\ref{fig:lastflow}). Yet, the Interior Plateau also hosts an impressive
-set of glacial lineations which indicate a substantial eastward flow component
+set of glacial lineations which indicate a substantial eastwards flow component
 of the Cordilleran ice sheet \citep{Prest.etal.1968, Kleman.etal.2010}.
@@
 Mountains. Throughout the modelled last glacial cycle, the Skeena Mountains
-host an ice cap which appears to be fed by moisture intruding inland from the
+host an ice cap which appears to be fed by the moisture intruding inland from the
 west through a topographic breach in the Coast Mountains. The Skeena ice cap acts
 as the main nucleation centre at the onset of the pre-LGM readvance. As indicated
-by persistent, warm-based ice in the model, this ice cap is perhaps associated
-with the distinct glacial erosional imprint on the landscape of the Skeena
+by persistent, warm-based ice in the model, this ice cap perhaps explains
+the distinct glacial erosional imprint observed on the landscape of the Skeena
 Mountains.
@@
 \section*{Author contributions}
-J.~Seguinot ran the simulations and wrote most of the manuscript; I.~Rogozhina
+J.~Seguinot ran the simulations; I.~Rogozhina
 guided experiment design; A.~P.~Stroeven, M.~Margold and J.~Kleman took part in
 the interpretation and comparison of model results against geological evidence.
-All co-authors contributed to improve the text.
+All authors contributed to the text.
 
@@
 We are very thankful to Constantine Khroulev, Ed Bueler and Andy Aschwanden for
-providing constant help and development on PISM. This work was supported by the
+providing constant help and development with PISM. This work was supported by the
 Swedish Research Council~(VR) grant no.~2008-3449 to A.~P.~Stroeven and by the

### 2015-06-20 cordillera-cycle.tex
Add corrections from Arjen and Martin (part 1).

@@
   Skeena Mountains favours the persistence of a central ice dome throughout the
-  glacial cycle. It acts as a nucleation centre before the last glacial maximum
-  and hosts the last remain of Cordilleran ice until the middle Holocene
+  glacial cycle. It acts as a nucleation centre before the Last Glacial Maximum
+  and hosts the last remains of Cordilleran ice until the middle Holocene
   (6.6--6.2\,ka).
@@
     \citep{Barendregt.Irving.1998, Kleman.etal.2010, Rutter.etal.2012},
-although it is clear that maximum glaciation of the Cordilleran ice sheet
-predates the last glacial cycle
+although it is clear that the Pleistocene maximum extent of the Cordilleran
+ice sheet predates the last glacial cycle
     \citep{Hidy.etal.2013}.
-In parts of the Yukon Territory and Alaska, the distribution of tills
-    \citep{Turner.etal.2013}
+In parts of the Yukon Territory and Alaska, and in the Puget Lowland,
+the distribution of tills
+    \citep{Turner.etal.2013, Troost.2014}
 and dated glacial erratics
@@
            Stroeven.etal.2010, Stroeven.etal.2014}
-indicate an extensive Marine Oxygen Isotope Stage (MIS)~4 glaciation, the
-evidence of which have also been found in the Puget Lowland \citep{Troost.2014}.
+indicate an extensive Marine Oxygen Isotope Stage (MIS)~4 glaciation.
 Landforms in the interior regions include flow sets that are likely
@@
            areas once covered by the Cordilleran (CIS), Laurentide (LIS),
-           Innuitian (IIS) and Greenland (GIS) ice sheets during the last
+           Innuitian (IIS), and Greenland (GIS) ice sheets during the last
            18\,\unit{\chem{^{14}C}\,ka} \citep[21.4\,cal\,ka,][]{Dyke.2004}.
@@
            St.-Elias mountains (WSEM), the Selwyn and Mackenzie mountains
-           (SMKM), the Skeena Mountains (SM), the Coast and Rocky
-           Mountains (CRM) and the North Cascades (NC). Major depressions
+           (SMKM), the Skeena Mountains (SM), the Coast Mountains (CM), the
+           Columbia and Rocky
+           Mountains (CRM), and the North Cascades (NC). Major depressions
            include the Liard Lowland (LL), the Interior Plateau of British
-           Columbia (IP) and the Puget Lowland (PL). The background
+           Columbia (IP), and the Puget Lowland (PL). The background
            map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
@@
 Geomorphological evidence from south-central British Columbia indicates a rapid
-deglaciation, including an early emergence of the elevated areas while thin,
+deglaciation, including an early emergence of elevated areas while thin,
 stagnant ice still covered the surrounding lowlands
@@
 Nevertheless, recent oxygen isotope measurements from Gulf of Alaska sediments
-reveals a climatic evolution highly correlated to that of Greenland during this
-period, including a distinct late glacial cold reversal between 14.1 and
+reveal a climatic evolution highly correlated to that of Greenland during this
+period, including a distinct Late Glacial cold reversal between 14.1 and
 11.7\,ka \citep{Praetorius.Mix.2014}.
@@
 calibrated against field-based evidence, to perform a quantitative
-reconstruction of the Cordilleran ice sheet history through the last glacial
+reconstruction of the Cordilleran ice sheet evolution through the last glacial
 cycle, and
@@
     ephemeral configurations?
-  \item How rapid was the last deglaciation? Did it include late glacial
+  \item How rapid was the last deglaciation? Did it include Late Glacial
     standstills or readvances?
@@
 improve our understanding of the Cordilleran ice sheet
-    (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991},
+    (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991};
      \citealp{Marshall.etal.2000}),
@@
 topography of the region has presented two major challenges to its application.
-First, only recent development of numerical ice sheet models and underlying
-scientific computing tools \citep{Bueler.Brown.2009, Balay.etal.2015} has
+First, only recent developments in numerical ice sheet models and underlying
+scientific computing tools \citep{Bueler.Brown.2009, Balay.etal.2015} have
 allowed for high-resolution numerical modelling of glaciers and ice sheets on
@@
 heat flow measurements. In the low-resolution simulations, the vertical grid
-consists of up to 51~enthalpy layers in the ice sheet and 31~temperature layers
-in the bedrock. In the high-resolution simulations, up to 101~ice layers and
-61~bedrock layers are used.
+consists of 31~temperature layers in the bedrock and up to 51~enthalpy layers
+in the ice sheet, corresponding to a vertical resolution of 100\,m. The
+high-resolution simulations 61~bedrock layers and up to 101~ice layers with a
+vertical resolution of 50\,m.
 
@@
 amount of water at the base, $W$, varies from zero to $W_{max}=2$\,m, a
-threshold above which any additional melt water is assumed to drain off
+threshold above which additional melt water is assumed to drain off
 instantaneously. Finally, the bedrock topography responds to ice load
@@
 elastic lithosphere flexure and viscous astenosphere deformation in an infinite
-half-space
-\citep[Table~\ref{tab:params};][]{Lingle.Clark.1985,Bueler.etal.2007}.
+half-space \citep{Lingle.Clark.1985,Bueler.etal.2007}.
 A relatively low viscosity value of $\nu_m = 1\times10^{19}$\,\unit{Pa\,s} is
-used for the astenosphere in accordance to the results from regional glacial
+used for the astenosphere (Table~\ref{tab:params}) in accordance with the
+results from regional glacial
 isostatic adjustment modelling at the northern Cascadia subduction zone
@@
 
-Ice-shelf calving is computed by a double criterion. First, a
+Ice shelf calving is computed using a double criterion. First, a
 physically-realistic calving flux is computed based on eigenvalues of the
 horizontal strain rate tensor \citep{Winkelmann.etal.2011,
-Levermann.etal.2012}. This allows outlet glaciers to advance in confined
-embayments, but prevents extensive ice-shelves from building up in the open ocean.
+Levermann.etal.2012}. This allows floating ice to advance in confined
+embayments, but prevents the formation of extensive ice shelves in the open ocean.
 Second, floating ice thinner than 50\,m is systematically calved off. A subgrid
-scheme \citep{Albrecht.etal.2011} allows for a continuous migration of the
-calving front. Such treatment of calving has been applied to the Antarctic ice
+scheme by \citet{Albrecht.etal.2011} allows for a continuous migration of the
+calving front. This formulation of calving has been applied to the Antarctic ice
 sheet and has shown to produce a realistic calving front positon for many of
@@
   \caption{Monthly mean near-surface air temperature, precipitation, and
-           standard deviation of daily mean temperature for January and July
-           from the North American Regional Reanalysis
+           standard deviation of daily mean temperature (PDD SD) for January
+           and July from the North American Regional Reanalysis
            \citep[NARR;][]{Mesinger.etal.2006}, used to force the surface mass
@@
 temperature values from 1979 to 2000 \citep{Mesinger.etal.2006},
-including variability associated with the seasonal cycle
-itself \citep[Fig.~\ref{fig:atm}; i.e. not as in][]{Seguinot.Rogozhina.2014}.
+including variability associated with the seasonal cycle (Fig.~\ref{fig:atm}).
 Degree-day factors for snow and ice melt are derived from
@@
              &  37{\degree} 38' W  % 37.64 (decimal)
-             & 3238\,m
+             & 3238
              & \chem{\delta^{18}O}
-             & 0.37 & -5.8{\degree}C  % -16.4126 (before scaling)
+             & 0.37 & -5.8  % -16.4126 (before scaling)
              & \citet{Dansgaard.etal.1993} \\
@@
              &  42{\degree} 19' W  % 42.32
-             & 2917\,m
+             & 2917
              & \chem{\delta^{18}O}
-             & 0.24 & -6.5{\degree}C  % -26.7098
+             & 0.24 & -6.5  % -26.7098
              & \citet{Andersen.etal.2004} \\
@@
              & 123{\degree} 21' E  % 123.35
-             & 3233\,m
+             & 3233
              & \chem{\delta^{18}O}
-             & 0.64 & -5.9{\degree}C  % -9.2055
+             & 0.64 & -5.9  % -9.2055
              & \citet{Jouzel.etal.2007} \\
@@
              & 106{\degree} 50' E  % 106.8
-             & 3488\,m
+             & 3488
              & \chem{\delta^{18}O}
-             & 0.74 & -5.9{\degree}C  % -7.9550
+             & 0.74 & -5.9  % -7.9550
              & \citet{Petit.etal.1999} \\
@@
              & 118{\degree} 23' W
-             & -1772\,m
+             & -1772
              & \chem{U^{K'}_{37}}
-             & 1.61 & -6.1{\degree}C  % -3.7889
+             & 1.61 & -6.1  % -3.7889
              & \citet{Herbert.etal.2001} \\
@@
              & 126{\degree} 26' W
-             & -3038\,m
+             & -3038
              & \chem{U^{K'}_{37}}
-             & 1.20 & -6.0{\degree}C  % -5.0000
+             & 1.20 & -6.0 % -5.0000
              & \citet{Herbert.etal.2001} \\
@@
 slower deglaciation than modelled using the other palaeo-temperature records.
-Finally, model runs forced by the Antarctic palaeo-temperature records result in a
-rapid and uninterrupted deglaciation after the LGM, whereas the
+Finally, whereas model runs forced by the Antarctic palaeo-temperature records
+result in a rapid and uninterrupted deglaciation after the LGM, the
 simulation driven by the GRIP palaeo-temperature record also results in a rapid
@@
 
-During the MIS~3 ice volume minima, a central ice cap persists over the Skeena
+In the MIS~3 ice volume minimum reconstructions, a central ice cap persists
+over the Skeena
 Mountains (Fig.~\ref{fig:snapshots}, middle panels). Although this ice cap is
@@
 Modelled ice sheet geometries during the LGM (MIS~2; Fig.~\ref{fig:snapshots},
-lower panels) invariably include a ca. 1500\,km-long central divide located
-about 3500\,m~a.s.l. along the spine of the Rocky Mountains. Although the
-similatity of modelled ice extents is a direct result of the choice of scaling
+lower panels) invariably include a ca. 1500\,km-long central divide above
+3000\,m~a.s.l. located along the spine of the Rocky Mountains. Although the
+similarity of modelled ice extents is a direct result of the choice of scaling
 factors applied to different palaeo-temperature proxy records
@@
     \citep{Porter.Swanson.1998, Margold.etal.2014}.
-There is stratigraphical evidence for glaciations older than the Fraser
-Glaciation in British Columbia \citep{Clague.Ward.2011} and in the Puget
-Lowland \citep{Troost.2014}, but their extent and
-timing are still highly conjectural
+There is stratigraphical evidence for an MIS~4 glaciation in British Columbia
+\citep{Clague.Ward.2011} and in the Puget Lowland \citep{Troost.2014}, but
+their extent and timing are still highly conjectural
     \citep[perhaps MIS~4 or early MIS~3; e.g.,][]{Cosma.etal.2008}.
@@
 earlier than dated evidence of the LGM advance.
-There exist a fair agreement between the EPICA and Vostok
-alaeo-temperature records, resulting in only little differences between the
-simulations driven by those records. These differences are not analysed
-further; instead we focus the rest of our analysis on simulations forced by
+There is a fair agreement between the EPICA and Vostok
+palaeo-temperature records, resulting in only small differences between the
+simulations driven by those records. These differences are not subject to
+further analysis further; instead we focus on simulations forced by
 palaeo-temperature records from the GRIP and EPICA ice cores that appear to
-produce the most realistic reconstructions of regional glaciation history among
-all records tested, yet bearing significant disparities in model output. To
+produce the most realistic reconstructions of regional glaciation history, yet
+bearing significant disparities in model output. To
 allow for a more detailed comparison against the geological record, these two
@@
 palaeoglaciological reconstructions for central and southern British Columbia
-that inferred the ice divide in a more westerly position,
+with ice divides in a more westerly position,
 over the western margin of the Interior Plateau \citep{Ryder.etal.1991,
 Stumpf.etal.2000, Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}.
-These idincate that a latitudinal saddle connected ice dispersal centres in
+These indicate that a latitudinal saddle connected ice dispersal centres in
 the Columbia Mountains with the main ice divide \citep{Ryder.etal.1991,
@@
 
-During MIS~2, the modelled total ice volume peaks at 8.01\,m~s.l.e. in the GRIP
-simulation
-and at 8.77\,m~s.l.e. in the EPICA simulation.
+During MIS~2, the modelled total ice volume peaks at 8.01\,m~s.l.e. (19.1\,ka)
+in the GRIP simulation and at 8.77\,m~s.l.e. (17.2\,ka) in the EPICA
+simulation.
 
@@
 restricted ice cover in northern and eastern sectors, and generally lower
-ice surface elevations in the interior (Fig.~\ref{fig:mis4}, left panel).
+ice surface elevations in the interior (Fig.~\ref{fig:mis4}, right panel).
 Thus, only the GRIP simulation can explain the presence of MIS~4 glacial
@@
 
-During MIS~4, the modelled total ice volume peaks at 7.19\,m~s.l.e. in the GRIP
-simulation
-and at 5.04\,m~s.l.e. in the EPICA simulation, corresponding to respectively
-90 and 57\,\% of the MIS~2 ice volumes.
+During MIS~4, the modelled total ice volume peaks at 7.19\,m~s.l.e. (57.3\,ka)
+in the GRIP simulation and at 5.04\,m~s.l.e. (61.9\,ka) in the EPICA
+simulation, corresponding to respectively 90\,\% and 57\,\% of modelled MIS~2
+ice volumes.
 
@@
   \includegraphics{duration}
-  \caption{Modelled duration of ice cover during the last 120\,ka.
+  \caption{Modelled duration of ice cover during the last 120\,ka using GRIP
+           and EPICA climate forcing.
            Note the irregular colour scale. A continuous ice cover spanning
-           from the Alaska Range (AR) to the southern Coast Mountains (CM) and
-           the Columbia and Rocky mountains (RM) exists for about 32\,ka in
+           from the Alaska Range (AR) to the Coast Mountains (CM) and
+           the Columbia and Rocky mountains (CRM) exists for about 32\,ka in
            the GRIP simulation and 26\,ka in the EPICA simulation.
@@
            cover persists over the Skeena Mountains (SM) during most of the
-           simulation. See Fig.~\ref{locmap} for a list of abbreviations.}
+           simulation. See Fig.~\ref{fig:locmap} for a list of abbreviations.}
   \label{fig:duration}
@@
 northeast to the Rocky Mountains in the southwest, is only in operation
-for at most 32\,ka, which is about a third of the modelled glacial cycle
-(Fig.~\ref{fig:duration}, hatched areas). However,
-except for its marine margin and the northern foothills of the Alaska Range,
-the maximum extent of the ice sheet is attained for an even shorter period
-of time of few thousand years (Figs.~\ref{fig:duration}, red areas). This
+for at most 32\,ka, which is about a third of the timespan of the last glacial cycle
+(Fig.~\ref{fig:duration}, hatched areas). However, except for its margins in
+the Pacific Ocean and in the northern foothills of the Alaska Range,
+the maximum extent of the ice sheet is attained for a much shorter period
+of time of only few thousand years (Figs.~\ref{fig:duration}, red areas). This
 result illustrates that the maximum extents of the modelled ice sheet during
@@
 fluctuations results from a combination of low precipitation, high summer
-temperature and large temperature standard deviation in the plains of the
+temperature and large temperature standard deviation (PDD SD) in the plains of the
 Alaska Interior (Fig.~\ref{fig:atm}) which confines glaciation to the
@@
 major feed to the Cordilleran ice sheet (Fig.~\ref{fig:duration}, WSEM).
-Although the Coast,
-Columbia, Rocky and Skeena mountains (CM, CRM, SM) are covered by mountain
+Although the Coast, Skeena and Columbia and Rocky mountains (CM, SM, CRM)
+are covered by mountain
 glaciers for most of the last glacial cycle, providing durable nucleation
@@
 small fraction of the last glacial cycle. In other words, the Selwyn and Mackenzie
-mountains appear as a secondary ice-dispersal centre which is only active
+mountains only appear as a secondary ice-dispersal centre
 during the coldest periods of the last glacial cycle. The Northern Rocky
@@
 middle panels), and serves as a nucleation centre at the onset of the glacial
-readvance towards the LGM (MIS~2). This situation appear similar to the neighbouring Laurentide ice sheet, for which the importance of residual ice for the
+readvance towards the LGM (MIS~2). This situation appear similar to the
+neighbouring Laurentide ice sheet, for which the importance of residual ice for the
 glacial history leading up to the LGM has been illustrated by the
@@
 The presence of a Skeena Mountains ice cap during most of the last glacial
-cycle can be explained by a more eastwards-spread distribution of
-winter precipitation there than along other transects of the
-modelling domain (Fig.~\ref{fig:atm}). Along most of the north-western coast of
+cycle can be explained by meteorological conditions more favourable for ice
+growth there than elsewhere. In fact, reanalysed atmospheric fields used to
+force the surface mass balance model show that high winter precipitations are
+mainly confined to the western slope of the Coast Mountains, except in the
+centre of the modelling domain where they also occur further inland than along
+other east-west transects (Fig.~\ref{fig:atm}).
+In fact, along most of the north-western coast of
 North America, coastal mountain ranges form a pronounced topographic barrier
@@
 
-During MIS~3, the modelled total ice volume corresponding to these persistent
-ice-dispersal
-centres reaches a minimum of 1.52\,m~s.l.e. in the GRIP simulation and of
-2.68\,m~s.l.e. in the EPICA simulation, corresponding to respectively 19 and
-31\,\% of the MIS~2 ice volumes.
+The modelled total ice volume corresponding to these persistent ice-dispersal
+centres attains a minimum of 1.52\,m~s.l.e. (42.9\,ka) in the GRIP simulation
+and of 2.68\,m~s.l.e. (51.8\,ka) in the EPICA simulation, corresponding to
+respectively 19\,\% and 31\,\% of the MIS~2 ice volumes.
 

### 2015-06-20 cordillera-cycle.tex
Add corrections from Arjen and Martin (part 2).

@@
            by the ice sheet include the Alaska Range (AR), the Wrangell and
-           St.-Elias mountains (WSEM), the Selwyn and Mackenzie mountains
+           Saint Elias mountains (WSEM), the Selwyn and Mackenzie mountains
            (SMKM), the Skeena Mountains (SM), the Coast Mountains (CM), the
@@
 extracted from the base of Ice Stream B in West Antarctica
-\citep{Tulaczyk.etal.2000, Bueler.Pelt.2014},
+\citep{Tulaczyk.etal.2000, Bueler.Pelt.2015},
 \begin{equation}
@@
 (SM), and the Selwyn and Mackenzie mountains (SMKM). The Wrangell and
-St.-Elias mountains, heavily glacierized at present, host an ice cap for the
+Saint Elias mountains, heavily glacierized at present, host an ice cap for the
 entire length of both simulations, but that ice cap does not appear to be a
@@
            120\,ka. Long ice cover durations combined with basal
-           temperatures at the pressure-melting point may explain the strong
+           temperatures at the pressure melting point may explain the strong
            glacial erosional imprint of the Skeena Mountains (SM) landscape.
@@
 landscape (mainly in terms of the development of deep glacial valleys and
-troughs). We find evidence for this on both the southwestern and northeastern
-slopes of the southern Coast Mountains, on the western slopes of the Columbia
-Mountains, on the western and eastern slopes of the northern Coast Mountains
-and the Saint-Elias Mountains, and radiating off the Skeena Mountains
+troughs). We find evidence for this on the slopes of the Coast Mountains, the
+the Columbia and Rocky mountains, the Wrangell and Saint Elias mountains, and
+radiating off the Skeena Mountains
 (Figs.~\ref{fig:duration} and \ref{fig:warmbase};
@@
 persistent ice cover (Fig.~\ref{fig:duration}) associated with basal ice
-temperatures at the pressure-melting point (Figs.~\ref{fig:warmbase}
+temperatures at the pressure melting point (Figs.~\ref{fig:warmbase}
 and~\ref{fig:warmfrac}) explains the large-scale glacial erosional imprint on
-the landscape. A well-developed network of glacial valleys west of
-the Mackenzie Mountains (\citealp[Fig.~2]{Kleman.etal.2010}; \citealp[Fig.~8]
-{Stroeven.etal.2010}) is modelled to have hosted mainly warm-based ice
+the landscape. A well-developed network of glacial valleys running to the
+north-west on the west slope of the Selwyn
+and Mackenzie Mountains (\citealp[Fig.~2]{Kleman.etal.2010}; \citealp[Fig.~8]
+{Stroeven.etal.2010}) is modelled to have hosted predominantly warm-based ice
 (Fig.~\ref{fig:warmfrac}). However, because it is only glaciated for a
@@
 models through simple parametrisations \citep[e.g.][]{Clason.etal.2012,
-Clason.etal.2014, Bueler.Pelt.2014}, if included at all. However, next after
+Clason.etal.2014, Bueler.Pelt.2015}, if included at all. However, next after
 the mapping of maximum ice sheet extents during MIS~2 and MIS~4
@@
            from the GRIP (top panels) and EPICA (bottom panels) 5\,km simulations.
-           Dashed segments indicate the location of profiles used in
+           Dashed segments (A--D) indicate the location of profiles used in
            Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}.}
@@
   \includegraphics{deglac}
-  \caption{Modelled age of the last deglaciation. Areas where the MIS~4 glacial
-           advance exceeded the last glacial maximum advanced are marked in
-           green. Hatches denote re-advance of mountain-centred ice caps and
-           the decaying ice sheet between 14 and 10\,ka., which is very
-           restricted in the EPICA-driven simulation.
-           Dashed segments indicate the location of profiles used in
+  \caption{Modelled age of the last deglaciation. Areas that have been covered
+           only before the last glacial maximum are marked
+           in green. Hatches denote re-advance of mountain-centred ice caps and
+           the decaying ice sheet between 14 and 10\,ka..
+           Dashed segments (A--D) indicate the location of profiles used in
            Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}.}
@@
 southern sector of the modelling domain, including the Puget Lowland, the Coast
+Mountains, the Columbia and
 and Rocky mountains, and the Interior Plateau of British Columbia, becomes
 completely deglaciated by 10\,ka, whereas a significant ice cover remains over
-the Skeena, Selwyn, Mackenzie, Wrangell and Saint-Elias mountains in the
+the Skeena, the Selwyn and Mackenzie, and the Wrangell and Saint Elias mountains in the
 northern sector of the modelling domain. After 10\,ka, deglaciation continues
@@
 of the last deglaciation \citep{Margold.etal.2013}. Remaining ice continues to
-decay by retreating towards the Selwyn and Skeena mountains. The last remnants
+decay by retreating towards the Selwyn and MacKenzie, and Skeena mountains.
+The last remnants
 of the Cordilleran ice sheet finally disappear from the Skeena Mountains at
@@
 alpine glacier readvances includes moraines in the eastern Coast Mountains,
-Rocky Mountains and the Columbia Mountains \citep{Osborn.Gerloff.1997,
+and the Columbia and Rocky mountains \citep{Osborn.Gerloff.1997,
 Menounos.etal.2008}.
@@
 to note that the simulation driven by the GRIP record produces a late-glacial
-readvance in the Coast Mountains and Rocky Mountains (Fig.~\ref{fig:deglac},
+readvance in the Coast Mountains and in the Columbia and Rocky Mountains
+(Fig.~\ref{fig:deglac},
 left panel). In addition to matching the location of some local readvances, the
@@
 deglaciation with only a tightly restricted late-glacial readvance on the
-western slopes of the Saint-Elias and Coast mountains
+western slopes of the Saint Elias and the Coast mountains
 (Fig.~\ref{fig:deglac}, right panel).
@@
            Sliding grid cells were distinguished from non-sliding grid cells
-           using a velocity threshold of 1\,\unit{m\,yr^{-1}}.
-           Dashed segments indicate the location of profiles used in
-           Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}.}
+           using a basal velocity threshold of 1\,\unit{m\,yr^{-1}}.}
   \label{fig:lastflow}
@@
 of the Cordilleran ice sheet \citep{Prest.etal.1968, Kleman.etal.2010}.
-The Fraser lineation set could thus present a smoking gun for the
+This Interior Plateau lineation set could therefore present a smoking gun for the
 reliability of the presented model results. One explanation for the incongruent
-results could therefore be that missing feedback mechanisms between ice sheet
-topography and regional climate position the modelled ice divide of the LGM ice
-sheet too far to the east (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
+results could be that the missing feedback mechanisms between ice sheet
+topography and regional climate resulted in a modelled ice divide of the LGM
+ice sheet beingtoo far to the east (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
 \citealp{Seguinot.etal.2014}). A more westerly-located LGM ice divide would
@@
 thickness would not promote warm-based conditions but, on the
-opposite, enlarge the region of negligible basal sliding
+contrary, enlarge the region of negligible basal sliding
 (Fig.~\ref{fig:lastflow}). Thus, a second explanation for the incongruent
-results could be that the Interior Plateau lineation system observed in the
-geomorphology predates the deglaciation, as perhaps indicated by some eskers
+results could be that the Interior Plateau lineation system predates
+deglaciation ice flow, as perhaps indicated by some eskers
 that appear incompatible with these glacial lineations
@@
 could have triggered the basal sliding
-\citep[cf. Greenland ice sheet,][]{Fahnestock.etal.2001}.
+\citep[cf. Greenland ice sheet;][]{Fahnestock.etal.2001}.
 
@@
 Vostok Antarctic ice core records. It occurs about 2\,ka too early in the
-simulation forced by the GRIP ice core record, and even earlier in all other
+simulation forced by the GRIP ice core record, and occurs even earlier in all other
 simulations. The mismatch is largest for the two Northwest Pacific ODP
@@
 west through a topographic breach in the Coast Mountains. The Skeena ice cap acts
-as the main nucleation centre at the onset of the pre-LGM readvance. As indicated
+as the main nucleation centre for the glacial reavance towards the LGM
+configuration. As indicated
 by persistent, warm-based ice in the model, this ice cap perhaps explains
@@
 late-glacial readvance of local ice caps and of the main body of the decaying
-Cordilleran ice sheet primarily in the Coast, Columbia and Rocky Mountains.
+Cordilleran ice sheet primarily in the Coast and the Columbia and Rocky Mountains.
 In both simulations, this is followed by an
@@
 (September 2014). His comments were used to improve the model set-up.
-We are very thankful to Constantine Khroulev, Ed Bueler and Andy Aschwanden for
+We are very thankful to Constantine Khroulev, Ed Bueler, and Andy Aschwanden for
 providing constant help and development with PISM. This work was supported by the
@@
 \bibliographystyle{copernicus}
-\bibliography{refs/references.bib}
+\bibliography{refs/references}
 \newpage

### 2015-06-21 abstract.tex
Add separate abstract file.

--- 2013-12-05 abstract.tex
@@
 \documentclass[12pt]{article}
+\usepackage[utf8]{inputenc}
 \usepackage[margin=25mm]{geometry}
@@
 
-\textbf{The effect of climate forcing on numerical simulations of the Cordilleran Ice Sheet at the Last Glacial Maximum}
+% Title
+\textbf{Numerical simulations of the Cordilleran ice sheet
+        through the last glacial cycle}
 
-J. Seguinot\textsuperscript{1},
-C. Khroulev\textsuperscript{2},
-I. Rogozhina\textsuperscript{3},
-A. P. Stroeven\textsuperscript{1},
-and Q. Zhang\textsuperscript{1}
+% Authors
+J.~Seguinot\textsuperscript{1,2,3},
+I.~Rogozhina\textsuperscript{3},
+A.~P.~Stroeven\textsuperscript{2},
+M.~Margold\textsuperscript{2}
+and J.~Kleman\textsuperscript{2}
 
-\textsuperscript{1}Department of Physical Geography and Quaternary Geology, Stockholm University, Stockholm, Sweden \\
-\textsuperscript{2}Geophysical Institute, University of Alaska Fairbanks, Fairbanks, AK, USA \\
-\textsuperscript{3}Helmholtz Centre Potsdam, GFZ German Research Centre for Geosciences, Potsdam, Germany
+% Affiliations
+\textsuperscript{1}{Laboratory of Hydraulics, Hydrology and Glaciology, ETH Zürich,
+          Zürich, Switzerland}\\
+\textsuperscript{2}{Department of Physical Geography and the Bolin Centre for Climate
+          Research, Stockholm University, Stockholm, Sweden}\\
+\textsuperscript{3}{Helmholtz Centre Potsdam, GFZ German Research Centre for Geosciences,
+          Potsdam, Germany}\\
 
-\input{sections/abstract.tex}
+\begin{abstract}
+
+  Despite more than a century of geological observations, the Cordilleran ice
+  sheet of North America remains poorly understood in terms of its former
+  extent, volume and dynamics. Although geomorphological evidence is abundant,
+  its complexity is such that whole ice-sheet reconstructions of advance and
+  retreat patterns are lacking. Here we use a numerical ice sheet model
+  calibrated against field-based evidence to attempt a quantitative
+  reconstruction of the Cordilleran ice sheet history through the last glacial
+  cycle. A series of simulations is driven by time-dependent temperature
+  offsets from six proxy records located around the globe. Although this
+  approach reveals large variations in model response to evolving climate
+  forcing, all simulations produce two major glaciations during
+  marine oxygen isotope stages~4 (61.9--56.5\,ka) and~2
+  (23.2--16.8\,ka). The timing of glaciation is
+  better reproduced using temperature reconstructions from Greenland and
+  Antarctic ice cores than from regional oceanic sediment cores. During most of
+  the last glacial cycle, the modelled ice cover is discontinuous and
+  restricted to high mountain areas. However, widespread precipitation over the
+  Skeena Mountains favours the persistence of a central ice dome throughout the
+  glacial cycle. It acts as a nucleation centre before the Last Glacial Maximum
+  and hosts the last remains of Cordilleran ice until the middle Holocene
+  (6.6--6.2\,ka).
+
+\end{abstract}
 

### 2015-11-11 reviews/response-jarosch.tex
Initialize responses to the reviews.
Quote and reformat comments from the referees.

\documentclass[10pt]{article}

\usepackage[T1]{fontenc}
\usepackage[pdftex]{graphicx}
\usepackage[pdftex]{color}

\sloppy
\setlength{\parskip}{1.2ex}
\setlength{\parindent}{0em}
\renewcommand\familydefault{\sfdefault}

\graphicspath{{../figures/}}
\definecolor{journalname}{rgb}{0.34,0.59,0.82}

\begin{document}
\textbf{Authors' response to A.~H.~Jarosch}
\bigskip

% ----------------------------------------------------------------------
% Interactive comment text begins
% ----------------------------------------------------------------------

\newcommand{\sechead}[1]{\bigskip\noindent\textbf{#1}}
\newcommand{\referee}[1]{\bigskip\textcolor{journalname}{\textit{#1}}}
\newcommand{\msquote}[1]{\begin{quote}\textit{#1}\end{quote}}
\newcommand{\todo}[1]{\textcolor{red}{TODO: #1}}

To A.~H.~Jarosch,

% ----------------------------------------------------------------------

\sechead{1 \quad General comments}

\referee{%
    Seguinot et al. present in this well written and structured manuscript a
    numerical modelling study of the Cordilleran ice sheet through the last
    glacial cycle. The model is driven by several temperature reconstructions
    based on proxy data and model output is subsequently compared in detail to
    the existing geological evidence in the region. The study is of significant
    relevance as it focuses on the Cordilleran ice sheet evolution in the past,
    which is still poorly understood.}

\referee{%
    Nevertheless, the manuscript is quite unbalanced in its presentation as it
    focuses strongly on section~4 (Comparison with geological record) and by
    doing so neglects crucial details in section~2 (Model setup). This poses a
    fundamental challenge for understanding the science presented. If it is not
    quite clear what the model does and how it performs to start with, it
    becomes difficult to discuss the results of the modelling study and why
    there are mismatches with geological evidence.}

\referee{%
    An overall sensitivity study of the parameters used in the model is
    completely lacking, thereby making it almost impossible to understand
    different responses of the ice sheet model. After reading the manuscript,
    one is left with the impression that the authors assume the PISM ice sheet
    model to be a black box which just requires one initial ``correct'' setup
    with literature values. This notion is reflected in the current manuscript,
    where almost all mismatches of model output with geological evidence (as
    discussed in section~4) are attributed to climate variations lacking in the
    proxy data, or climate-ice sheet feedback mechanisms not represented in the
    model chain. Similarly in a previous study Seguinot et al. ( 2014 ) have
    focused only on the driving climate sensitivities and have omitted
    influences of the ice sheet model as well as mass balance model pa-
    rameters even though they note in that study that these sensitivities
    require attention as well.}

\referee{%
    What I advocate at this point is not a complete, strict sensitivity study
    of all parameters involved in the model setup (that would be probably a
    work package large enough to fill a science career). However several key
    parameters can be investigated with not too much effort. Contrasting the
    influence of e.g. basal sliding and ice rheology parameters with the
    influence of driving climate on the model results would help to estimate
    the overall sensitivity of the model system as well as help guiding future
    efforts performing such modelling studies. Implicitly the authors assume
    that all other model sensitivities are negligibly small in comparison to
    the driving climate. However it is obvious from an ice sheet model
    perspective that at least chosen basal sliding parameters as well as ice
    rheology parameters will strongly influence the shape and volume of the
    modelled ice sheet. Thus it would be nice to see evidence supporting the
    claim that driving climate is the only input to worry about being presented
    in the current manuscript. Or should it turn out that basal sliding and ice
    rheology play an important role too, as one would expect, then the relative
    importance of each including error estimates on the chosen parameters
    should be presented as well.}

\referee{%
    Generally section~4 appears to be quite long and seems to re-summarize
    known geological evidence for the region. At times the language is quite
    speculative, for example P4162 L1 and 7, P4164 L18, P4171 L7, L9, L17, L19
    and L20 and so forth. I would recommend to shorten that section to focus
    only on the geological evidence which can or can not clearly be reproduced
    by the presented model and avoid extensive speculation on what the reasons
    for mismatch are, especially in the present form of the manuscript, where a
    sensitivity study of the model itself is completely missing. However I
    leave the choice of how much geological evidence is discussed in the
    manuscript entirely up to the authors.}

% ----------------------------------------------------------------------

\sechead{2 \quad Specific Comments}

\referee{%
    I refer to text locations in the discussion paper by page number (P) and
    line number range (L) for the specific comments.}

\referee{%
    \textbf{P4151 L11-16:}
    In this sentence the authors refer back to their previous work ( Seguinot
    et al. , 2014 ) and highlight that the NARR temperature and precipitation
    fields are the most suitable present day climate datasets to be used.
    Especially since the NARR precipitation fields include steep precipitation
    gradients which are required as identified by Seguinot et al. ( 2014 ).
    NARR is delivered on a 32\,km Lambert grid, and thus it is ques- tionable
    how ``steep'' these gradients can be, given the rather smooth
    representation of the existing topography on a 32\,km grid. Seguinot et al.
    ( 2014 ) have partly discussed that however. NARR precipitation and
    temperature fields have been evaluated in detail based on available station
    data for large parts of the study domain dealt with in this manuscript.
    This evaluation ( Jarosch et al. , 2012 ) demonstrated that NARR has
    difficulties simulating orographic processes in the Coast Mountains which
    in turn results in unrealistic atmospheric conditions over the Rocky
    Mountains. Jarosch et al. ( 2012 ) further concluded that physics based
    downscaling is required to adequately drive glacier models in that region.
    The authors should argue in more detail here why they think that NARR
    precipitation fields at 32\,km are adequate to drive their model and
    reflect their arguments with the findings of Jarosch et al. ( 2012 ). A
    solid argument here is of special importance as the authors assume the
    present day precipitation fields to be valid throughout their model time
    period (120\,ky to present) without further corrections (cf. section~2.4
    equation~6).}

\referee{%
    \textbf{P4152 L11:}
    Basal topography is ``derived'' from ETOPO1 data. What does this mean? Do
    the authors just re-sample the DEM data to their 10\,km and 5\,km model
    grids (P4152 L21-22) or is there more processing done? The ETOPO1 data
    contains the present day ice volumes within the study region. Clarke et al.
    ( 2013 ) have estimated the ice volume in parts of that region to be $2530
    \pm 220\,km^2$, with maximum ice thicknesses up to 200\,m. It can be argued
    that the volume is negligible in this study and the authors should do so if
    they think this is appropriate, but I wonder about the ice thicknesses.
    Assuming that the authors did not remove the present day ice cover, basal
    topography could be up to 200\,m higher that it actually is in reality.
    Given their used temperature lapse rate of $6\,K\,km^{-1}$ (P4157 L1),
    parts of the Cordilleran ice sheet growing in those regions with 200\,m too
    high topography would experience a 1.2\,K colder atmosphere than it
    actually should in reality. This favours unrealistic ice growth and thus
    the omission of present day ice cover removal should be clearly argued for
    in the manuscript.}

\referee{%
    \textbf{P4153 L2-3:}
    That the ``shallow shelf approximation'' (SSA) is used as a ``sliding law''
    for the ``shallow ice approximation'' (SIA) is a confusing statement in
    this context. Bueler has coined the term in his 2009 paper as cited in the
    manuscript. However the casual reader will be confused at this point,
    especially since the authors state the pseudoplastic sliding law the model
    actually uses in equation~1. I would recommend to leave out the statement
    on the SSA being the ``sliding law'' for the SIA.}

\referee{%
    \textbf{P4153 L5-6:}
    As stated here, ice rheology within the used ice sheet model is based on
    Aschwanden et al. ( 2012 ). This enthalpy based formulation has proofed
    itself to be very suitable for estimating ice rheology in ice sheet models,
    but it also depends on several parameters to translate enthalpy within the
    ice to ice viscosity (see Aschwanden et al. ( 2012 ), equations 62-65). The
    authors do not mention any of these parameters (e.g. any of the rate
    factors or nonlinear power $n$ ) within the manuscript or in Table~1. I
    have mentioned above in the general comments section that parameters used
    in ice rheology and basal sliding formulations are important model
    parameters which will influence the ice sheet model output and that a basic
    sensitivity study on those parameters is required to understand the model
    results. Here the authors could start with listing the parameters used in
    the ice rheology formulation, than continue with estimating uncertainties
    for those from literature and afterwards perform additional model
    simulations to identify the influence of the chosen parameter sets on the
    ice volume and ice margin position history the model creates. In the end
    the authors will be able to identify the relative importance of
    uncertainties in driving climate as well as model parameters, which will
    strengthen their discussion in section~4.}

\referee{%
    \textbf{P4153 L8:}
    It is not clear where the geothermal heat flux boundary is located. Does
    the ``depth of 3\,km'' refer to a depth measured from the ice surface,
    which would not make much sense for a ice thickness evolving ice sheet
    model or is it measured from the ice-bedrock interface downward. In that
    case the term ``computed subglacially'' is confusing as it refers to the
    ice-bedrock interface. Please be more specific here.}

\referee{%
    \textbf{P4153 L16 - P4154 L0:}
    Here the authors describe the basal sliding setup in their model. However
    they do not explain how they came up with the parameters used in
    equations~1-3 that are listed in Table~1 (part on ``basal sliding''). What
    motivates these parameter choices (references?) and how sensitive is the
    model and its results to these choices? Both question come instantly to
    mind and need to be addressed in detail. Here a basic sensitivity study on
    how basal sliding parameters in the model control the outcome discussed in
    section~4 is in order and I strongly recommend to include one in the
    manuscript. The authors can start by estimating the uncertainties in the
    chosen basal sliding parameters and run two extra simulation runs with
    their preferred climate forcing and the end member values of the
    uncertainties. This would create the most simple sensitivity study with
    respect to basal sliding, but would be extremely helpful for the argument
    made above in my general comments.}

\referee{%
    \textbf{P4156 L3-5:}
    In addition to what I have stated above on the NARR precipitation fields
    and their suitability, it is important to state at this location in the
    manuscript how the 32\,km NARR data is translated to the 10 and 5\,km
    computational grids of the current study. I disagree with the notion that a
    32\,km precipitation field can be called ``high- resolution'' in the
    context of 10 and 5\,km grid based ice sheet modelling. The input data is
    either 3 or 6 times coarser than the numerical grid, thus not at all
    high-resolution. Seguinot et al. ( 2014 ) state in their section~3.3 that
    the NARR data fields have been bilinearly interpolated to 10\,km resolution
    in their work. Did the authors do the same here for their 10 and 5\,km
    working grids? This is crucial information to be included in the
    manuscript. It has been demonstrated by spectral power analysis ( Jarosch
    et al. , 2012 ) that the NARR precipitation fields do not contain any
    significant spacial information below approximately 39\,km resolution and
    that bilinear interpolation does not add any information whatsoever on
    smaller scales, which should come to no surprise. Physics-based downscaling
    techniques however are able to add spatial information to precipitation
    fields down to about 1\,km grid sizes ( Jarosch et al. , 2012 ). Taking
    these findings into the current context of the manuscript at hand, the NARR
    precipitation fields can hardly be called ``high resolution'' with their
    effective precipitation grid size of 39\,km. The authors should argue for
    their choice of not performing any downscaling whatsoever to their
    computational grids of 10 and 5\,km for precipitation and temperature and
    discuss their choice in the light of the findings from Jarosch et al. (
    2012 ). Temperature however is better constrained in NARR ( Jarosch et al.
    , 2012 ) and contains spectral information down to 10\,km resolution, which
    justifies the usage of NARR temperature fields on the 10\,km computational
    grid of this study. The 5\,km grid still needs to be argued for.}

\referee{%
    \textbf{P4157 L1:}
    How is a fixed temperature lapse rate justified for simulations over 120k
    years, when there is ample published evidence that temperature lapse rates
    vary significantly within space and time? I am sure that the choice of
    $\gamma$ in this study has a significant influence on the model outcome and
    I leave it to the authors to explore this possibility.}

% ----------------------------------------------------------------------

\sechead{3 \quad Technical Corrections}

\referee{%
    \textbf{P4161 L10:}
    ``further analysis further;'' maybe change to ``further analysis'' or
    ``further analysis here''.}

\referee{%
    \textbf{P4166 L13:}
    double ``the'' in the sentence.}

\referee{%
    I hope the authors find my comments helpful in revising their manuscript
    and wish them success for their future endeavours.}

% ----------------------------------------------------------------------
% Interactive comment text ends
% ----------------------------------------------------------------------

\end{document}

### 2015-11-11 reviews/response-marshall.tex
Initialize responses to the reviews.
Quote and reformat comments from the referees.

\documentclass[10pt]{article}

\usepackage[T1]{fontenc}
\usepackage[pdftex]{graphicx}
\usepackage[pdftex]{color}

\sloppy
\setlength{\parskip}{1.2ex}
\setlength{\parindent}{0em}
\renewcommand\familydefault{\sfdefault}

\graphicspath{{../figures/}}
\definecolor{journalname}{rgb}{0.341,0.596,0.824}
\definecolor{darkred}{rgb}{0.439,0.157,0.145}  % from Advances in Geoscience

\begin{document}
\textbf{Authors' response to S.~J.~Marshall}
\bigskip

% ----------------------------------------------------------------------
% Interactive comment text begins
% ----------------------------------------------------------------------

\newcommand{\sechead}[1]{\bigskip\noindent\textbf{#1}}
\newcommand{\referee}[1]{\bigskip\textcolor{journalname}{\textit{#1}}}
\newcommand{\msquote}[1]{\begin{quote}\textit{#1}\end{quote}}
\newcommand{\todo}[1]{\textcolor{darkred}{TODO: #1}}

To S.~J.~Marshall,

% ----------------------------------------------------------------------

\sechead{1 \quad Summary comments}

\referee{%
    Seguinot and colleagues provide the first detailed glaciological modelling
    of that I am aware of for the Cordilleran Ice Sheet in western North
    America, making this a novel and long overdue contribution. The authors
    have not only made new advances with this contribution, they have done so
    in an impressive leap forward. This is an excellent and carefully-presented
    study which is likely to rejuvenate interest and debate in Cordilleran Ice
    Sheet reconstructions. The balance between numerical modelling and glacial
    geological/geomorphological considerations is unusually strong, and the
    authors can be commended for this emphasis. This adds tremendous value to
    the results and increases confidence in the modelling, and I also
    appreciate that the authors point out areas where the numerical model is
    not in accord with the geological record.}

\referee{%
    The manuscript is well-written and beautifully illustrated, and I have very
    few substantive comments. The choices made by the authors are logical and
    well-explained, and they reach several well-substantiated conclusions: a
    two-phase Cordilleran glaciation, a reasonably robust estimate of CIS
    volume at LGM, the general model of CIS growth through multiple alpine
    icefields, and the importance of the Skeena Mountain inception centre. One
    can always quibble with specific aspects of the model design and climate
    scenarios, but the authors have explored a reasonable span of `solution
    space' and these aspects of the Cordilleran ice sheet history appear to be
    robust features of the simulations.}

\referee{%
    The modelling strategy and results presented here stand to be widely cited,
    and I expect that it will serve as a springboard for additional studies
    from others in the international community. I recommend this manuscript for
    publication in The Cryosphere without reservations.}

% ----------------------------------------------------------------------

\sechead{1 \quad Specific comments}

\referee{%
    The Cordilleran Ice Sheet is difficult to model due to its complex
    topography and multiple inception centres (and possibly multiple
    domes/divides), strong regional climatic gradients, which require
    relatively high-resolution climate input fields, and a dearth of
    paleoclimate proxies for western North America to inform spatial and
    temporal variations in climate conditions during the glacial period. The
    authors confront these challenges well, with a adequate ice sheet model
    resolution (5 to 10\,km) and ice physics, carefully calibrated `control'
    climatology (published in Seguinot et al., 2014), and a good exploration of
    different paleoclimate time series histories in this contribution.}

\referee{%
    Nevertheless, it is not clear that ice-core based paleoclimate proxies from
    Greenland or Antarctica are appropriate for western North America. This may
    be particularly true of Greenland proxies, where the amplitude of D-O
    (millennial) climate variations is exceptionally strong and is likely to be
    regional. Because these remote ice core records are `scaled' based on only
    one constraint, producing a CIS maximum configuration that resembles the
    geological record, it is difficult to assess the pre-LGM simulations or the
    details of the modelled ice divide structure, ice thickness, etc. The
    robustness of the conclusion that Greenland and Antarctic ice core records
    are good proxies of glacial climate variability in western North America is
    therefore not so clear, but it is admittedly hard to do better at this
    time. I do wonder if there is any hope from more regional climate proxies
    such as the Logan ice cores or the off-shore Vancouver Island sediment
    records that are cited from Cosma et al. This is worth a short discussion.}

\todo{%
    Logan ice core covers only the last 16\,Ka and is not a good proxy for
    temperature. The Vancouver SST record covers only deglaciation. Check new
    papers for updates. Discuss in the manuscript.}

\referee{%
    Similarly it is difficult to know the errors and uncertainties associated
    with the assumption of fixed modern-day spatial patterns for temperature
    and precipitation. I suspect that the sensitivity of this assumption far
    exceeds that associated with the different paleoclimate proxies. Such that,
    for example, one could readily imagine different assumptions, such as a
    maritime effect that gives reduced glacial cooling near the coast vs. in
    the interior, that is a stronger effect than the difference between
    different paleo-climate proxies with respect to the timing of LGM, ice
    divide structure, etc. But this is a very reasonable start, what the
    authors have done -- there is always going to be more parameter space to
    explore in future studies. As above, I would perhaps just suggest a small
    discussion of the authors' opinion on this question, the uncertainty or
    possible influence of this assumption of modern-day climate patterns.}

\todo{Discuss in the manuscript.}

\referee{%
    Several minor points and grammatical corrections are included in the
    attached text. Nothing that will require much thought -- this is a really
    impressive piece of research, overall, and I am hard-pressed to find any
    criticism of it. It is one of the easiest reviews I have ever done.
    Congratulations to the authors and thanks for this fine work.}

\todo{Correct mistakes.}

% ----------------------------------------------------------------------
% Interactive comment text ends
% ----------------------------------------------------------------------

\end{document}

### 2015-11-11 reviews/response-stumpf.tex
Initialize responses to the reviews.
Quote and reformat comments from the referees.

\documentclass[10pt]{article}

\usepackage[T1]{fontenc}
\usepackage[pdftex]{graphicx}
\usepackage[pdftex]{color}

\sloppy
\setlength{\parskip}{1.2ex}
\setlength{\parindent}{0em}
\renewcommand\familydefault{\sfdefault}

\graphicspath{{../figures/}}
\definecolor{journalname}{rgb}{0.34,0.59,0.82}

\begin{document}
\textbf{Authors' response to A.~Stumpf}
\bigskip

% ----------------------------------------------------------------------
% Interactive comment text begins
% ----------------------------------------------------------------------

\newcommand{\sechead}[1]{\bigskip\noindent\textbf{#1}}
\newcommand{\referee}[1]{\bigskip\textcolor{journalname}{\textit{#1}}}
\newcommand{\msquote}[1]{\begin{quote}\textit{#1}\end{quote}}
\newcommand{\todo}[1]{\textcolor{red}{TODO: #1}}

To A.~Stumpf,

\referee{%
    Seguinot et al. present a well designed numerical model for the Cordilleran
    Ice Sheet (CIS) in North America for the last two glaciations occurring
    over the past 120000 years. Although the constraints on such a model are
    not yet all fully understood, their proposed simulations attempt to take
    account all the complexity of the glacial system, and utilizes a variety of
    data sets. The modeling confirms what geologists have observed in the
    field; there was a fast decay of the CIS during MIS~2 and MIS~4 and
    non-glacial conditions existed during MIS (Olympia Nonglacial Interval;
    e.g., Plouffe and Jett\'e, 1997). I applaud them for undertaking of such a
    difficult task, and notifying the reader where input data is sparse or
    inconclusive.}

\referee{%
    I provide the following general comments and observations which may help
    the authors in revising the manuscript for final publication. Many of these
    points are both my personal suggestions and also the recommendations of
    others currently researching the CIS or who have completed studies in the
    past.}

\referee{%
    1) With the large amount of research that has been undertaken to map the
    landforms and deposits of the CIS by federal and provincial scientists,
    academic faculty, and undergraduate and graduate students and determine the
    extent, volume and dynamics ice sheet, the impression left on the reader by
    the opening sentence would be incorrect. These studies have greatly
    advanced our understanding of the CIS, and could be an important dataset to
    test against the modeling. Stumpf et al. (2014) provides a list of some of
    these studies.}

\referee{%
    2) To help the reader better understand the maps presented, I would
    recommend some spatial information be added (e.g., latitude/longitude
    grids; political boundaries; lakes and rivers, place names etc..).}

\referee{%
    3) Was the model tested against regional-scale ground-based data (e.g.,
    Ferbey et al. 2013) to constrain ice divide positions, ice flow direction,
    and ice sheet thickness?}

\referee{%
    4) For the central sector of the CIS, Stumpf et al. (2000) provides some
    insight into the chronology and effectiveness of glacial erosional during
    the MIS. In figure~5, and in the accompanying text, they describe how
    landforms on the surface formed. For example, in lake valleys east of the
    Skeena Mountains, it appeared the major glacial streamlined landforms were
    formed during a longer glacial advance phase, with ice flow paralleling the
    valleys, and later flows, some perpendicular the valley flow, only weakly
    impacted them.}

\referee{%
    5) Stumpf et al. (2000) was the first study to extensively document a
    predominant westerly directed ice flow across high elevations in the Skeena
    and Coast Mountains. This flow direction appeared to continue into the
    late-glacial period. Other subsequent studies also confirm that
    late-glacial readvance eastward out of these mountains and retreat of ice
    margins westward into these mountains was limited.}

% ----------------------------------------------------------------------
% Interactive comment text ends
% ----------------------------------------------------------------------

\end{document}

### 2015-11-12 reviews/response-jarosch.tex
Add references from the reviews.

@@
 \usepackage[pdftex]{color}
+\usepackage{hyperref}
+\usepackage{natbib}
 
@@
 
+\newcommand{\doi}[1]{doi:\allowbreak\href{http://dx.doi.org/#1}{#1}}
 \newcommand{\sechead}[1]{\bigskip\noindent\textbf{#1}}
@@
     proxy data, or climate-ice sheet feedback mechanisms not represented in the
-    model chain. Similarly in a previous study Seguinot et al. ( 2014 ) have
+    model chain. Similarly in a previous study \citet{Seguinot.etal.2014} have
     focused only on the driving climate sensitivities and have omitted
@@
     \textbf{P4151 L11-16:}
-    In this sentence the authors refer back to their previous work ( Seguinot
-    et al. , 2014 ) and highlight that the NARR temperature and precipitation
+    In this sentence the authors refer back to their previous work
+    \citep{Seguinot.etal.2014} and highlight that the NARR temperature and precipitation
     fields are the most suitable present day climate datasets to be used.
     Especially since the NARR precipitation fields include steep precipitation
-    gradients which are required as identified by Seguinot et al. ( 2014 ).
+    gradients which are required as identified by et \citet{Seguinot.etal.2014}.
     NARR is delivered on a 32\,km Lambert grid, and thus it is ques- tionable
     how ``steep'' these gradients can be, given the rather smooth
-    representation of the existing topography on a 32\,km grid. Seguinot et al.
-    ( 2014 ) have partly discussed that however. NARR precipitation and
+    representation of the existing topography on a 32\,km grid. \citet{Seguinot.etal.2014}
+    have partly discussed that however. NARR precipitation and
     temperature fields have been evaluated in detail based on available station
     data for large parts of the study domain dealt with in this manuscript.
-    This evaluation ( Jarosch et al. , 2012 ) demonstrated that NARR has
+    This evaluation \citep{Jarosch.etal.2010} demonstrated that NARR has
     difficulties simulating orographic processes in the Coast Mountains which
     in turn results in unrealistic atmospheric conditions over the Rocky
-    Mountains. Jarosch et al. ( 2012 ) further concluded that physics based
+    Mountains. \citet{Jarosch.etal.2010} further concluded that physics based
     downscaling is required to adequately drive glacier models in that region.
@@
     precipitation fields at 32\,km are adequate to drive their model and
-    reflect their arguments with the findings of Jarosch et al. ( 2012 ). A
+    reflect their arguments with the findings of \citet{Jarosch.etal.2010}. A
     solid argument here is of special importance as the authors assume the
@@
     grids (P4152 L21-22) or is there more processing done? The ETOPO1 data
-    contains the present day ice volumes within the study region. Clarke et al.
-    ( 2013 ) have estimated the ice volume in parts of that region to be $2530
+    contains the present day ice volumes within the study region.
+    \citet{Clarke.etal.2013}
+    have estimated the ice volume in parts of that region to be $2530
     \pm 220\,km^2$, with maximum ice thicknesses up to 200\,m. It can be argued
@@
     As stated here, ice rheology within the used ice sheet model is based on
-    Aschwanden et al. ( 2012 ). This enthalpy based formulation has proofed
+    \citet{Aschwanden.etal.2012}. This enthalpy based formulation has proofed
     itself to be very suitable for estimating ice rheology in ice sheet models,
     but it also depends on several parameters to translate enthalpy within the
-    ice to ice viscosity (see Aschwanden et al. ( 2012 ), equations 62-65). The
+    ice to ice viscosity \citep[see][equations 62-65)]{Aschwanden.etal.2012}. The
     authors do not mention any of these parameters (e.g. any of the rate
@@
     either 3 or 6 times coarser than the numerical grid, thus not at all
-    high-resolution. Seguinot et al. ( 2014 ) state in their section~3.3 that
+    high-resolution. \citet{Seguinot.etal.2014} state in their section~3.3 that
     the NARR data fields have been bilinearly interpolated to 10\,km resolution
@@
     working grids? This is crucial information to be included in the
-    manuscript. It has been demonstrated by spectral power analysis ( Jarosch
-    et al. , 2012 ) that the NARR precipitation fields do not contain any
+    manuscript. It has been demonstrated by spectral power analysis \citep{Jarosch.etal.2010}
+    that the NARR precipitation fields do not contain any
     significant spacial information below approximately 39\,km resolution and
@@
     techniques however are able to add spatial information to precipitation
-    fields down to about 1\,km grid sizes ( Jarosch et al. , 2012 ). Taking
+    fields down to about 1\,km grid sizes \citep{Jarosch.etal.2010}. Taking
     these findings into the current context of the manuscript at hand, the NARR
@@
     computational grids of 10 and 5\,km for precipitation and temperature and
-    discuss their choice in the light of the findings from Jarosch et al. (
-    2012 ). Temperature however is better constrained in NARR ( Jarosch et al.
-    , 2012 ) and contains spectral information down to 10\,km resolution, which
+    discuss their choice in the light of the findings from \citet{Jarosch.etal.2010}.
+    Temperature however is better constrained in NARR \citep{Jarosch.etal.2010}
+    and contains spectral information down to 10\,km resolution, which
     justifies the usage of NARR temperature fields on the 10\,km computational
@@
 % ----------------------------------------------------------------------
+
+\begin{thebibliography}{69}
+
+\bibitem[{Aschwanden et~al.(2012)}]{Aschwanden.etal.2012}
+Aschwanden, A., Bueler, E., Khroulev, C., and Blatter, H.: An enthalpy
+  formulation for glaciers and ice sheets, J. Glaciol., 58, 441--457,
+  \doi{10.3189/2012JoG11J088}, 2012.
+
+\bibitem[{Clarke et~al.(2013)}]{Clarke.etal.2013}
+Clarke, G. K.~C., Anslow, F.~S., Jarosch, A.~H., Radi{\'{c}}, V., Menounos, B.,
+  Bolch, T., and Berthier, E.: Ice Volume and Subglacial Topography for Western
+  Canadian Glaciers from Mass Balance Fields, Thinning Rates, and a Bed Stress
+  Model, J. Climate, 26, 4282--4303, \doi{10.1175/jcli-d-12-00513.1}, 2013.
+
+\bibitem[{Jarosch et~al.(2010)Jarosch, Anslow, and Clarke}]{Jarosch.etal.2010}
+Jarosch, A.~H., Anslow, F.~S., and Clarke, G. K.~C.: High-resolution
+  precipitation and temperature downscaling for glacier models, Clim. Dynam.,
+  38, 391--409, \doi{10.1007/s00382-010-0949-1}, 2010.
+
+\bibitem[{Seguinot et~al.(2014)Seguinot, Khroulev, Rogozhina, Stroeven, and
+  Zhang}]{Seguinot.etal.2014}
+Seguinot, J., Khroulev, C., Rogozhina, I., Stroeven, A.~P., and Zhang, Q.: The
+  effect of climate forcing on numerical simulations of the {C}ordilleran ice
+  sheet at the {L}ast {G}lacial {M}aximum, The Cryosphere, 8, 1087--1103,
+  \doi{10.5194/tc-8-1087-2014}, 2014.
+
+\end{thebibliography}
+
+% ----------------------------------------------------------------------
 % Interactive comment text ends

### 2015-11-12 reviews/response-marshall.tex
Add references from the reviews.

@@
 
+\newpage
 \sechead{1 \quad Specific comments}
@@
     resolution (5 to 10\,km) and ice physics, carefully calibrated `control'
-    climatology (published in Seguinot et al., 2014), and a good exploration of
-    different paleoclimate time series histories in this contribution.}
+    climatology \citep[published in][]{Seguinot.etal.2014}), and a good
+    exploration of different paleoclimate time series histories in this
+    contribution.}
 
@@
 % ----------------------------------------------------------------------
+
+\begin{thebibliography}{69}
+
+\bibitem[{Seguinot et~al.(2014)Seguinot, Khroulev, Rogozhina, Stroeven, and
+  Zhang}]{Seguinot.etal.2014}
+Seguinot, J., Khroulev, C., Rogozhina, I., Stroeven, A.~P., and Zhang, Q.: The
+  effect of climate forcing on numerical simulations of the {C}ordilleran ice
+  sheet at the {L}ast {G}lacial {M}aximum, The Cryosphere, 8, 1087--1103,
+  \doi{10.5194/tc-8-1087-2014}, 2014.
+
+\end{thebibliography}
+
+% ----------------------------------------------------------------------
 % Interactive comment text ends

### 2015-11-12 reviews/response-stumpf.tex
Add references from the reviews.

@@
     non-glacial conditions existed during MIS (Olympia Nonglacial Interval;
-    e.g., Plouffe and Jett\'e, 1997). I applaud them for undertaking of such a
+    \citealp[e.g.,][]{Plouffe.Jette.1997}). I applaud them for undertaking of such a
     difficult task, and notifying the reader where input data is sparse or
@@
     advanced our understanding of the CIS, and could be an important dataset to
-    test against the modeling. Stumpf et al. (2014) provides a list of some of
+    test against the modeling. \citet{Stumpf.etal.2014} provides a list of some of
     these studies.}
@@
 \referee{%
-    3) Was the model tested against regional-scale ground-based data (e.g.,
-    Ferbey et al. 2013) to constrain ice divide positions, ice flow direction,
+    3) Was the model tested against regional-scale ground-based data
+    \citep[e.g.,][]{Ferbey.etal.2013} to constrain ice divide positions, ice flow direction,
     and ice sheet thickness?}
@@
 \referee{%
-    4) For the central sector of the CIS, Stumpf et al. (2000) provides some
+    4) For the central sector of the CIS, \citet{Stumpf.etal.2000} provides some
     insight into the chronology and effectiveness of glacial erosional during
@@
 \referee{%
-    5) Stumpf et al. (2000) was the first study to extensively document a
+    5) \citet{Stumpf.etal.2000} was the first study to extensively document a
     predominant westerly directed ice flow across high elevations in the Skeena
@@
 % ----------------------------------------------------------------------
+
+\begin{thebibliography}{69}
+
+\bibitem[{Ferbey et~al.(2013)}]{Ferbey.etal.2013}
+Ferbey, T., Arnold, H., and Hickin, A.: Ice-flow indicator compilation, British
+  Columbia., British Columbia Geol. Surv., Victoria, BC, Open-File 2013-06,
+  \url{http://www.em.gov.bc.ca/Mining/Geoscience/PublicationsCatalogue/OpenFiles/2013/Pages/2013-06.aspx},
+  2013.
+
+\bibitem[{Seguinot et~al.(2014)}]{Seguinot.etal.2014}
+Seguinot, J., Khroulev, C., Rogozhina, I., Stroeven, A.~P., and Zhang, Q.: The
+  effect of climate forcing on numerical simulations of the {C}ordilleran ice
+  sheet at the {L}ast {G}lacial {M}aximum, The Cryosphere, 8, 1087--1103,
+  \doi{10.5194/tc-8-1087-2014}, 2014.
+
+\bibitem[{Stumpf et~al.(2000)}]{Stumpf.etal.2000}
+Stumpf, A.~J., Broster, B.~E., and Levson, V.~M.: Multiphase flow of the late
+  Wisconsinan Cordilleran ice sheet in western Canada, Geol. Soc. Am. Bull.,
+  112, 1850--1863, \doi{10.1130/0016-7606(2000)112<1850:mfotlw>2.0.co;2}, 2000.
+
+\bibitem[{Stumpf et~al.(2014)}]{Stumpf.etal.2014}
+Stumpf, A.~J., Ferbey, T., Plouffe, A., Clague, J.~J., Ward, B.~C., Paulen,
+  R.~C., and Bush, A.~B.: Discussion: ``Streamlined erosional residuals and
+  drumlins in central British Columbia, Canada'' by J. Donald {McClenagan},
+  (2013) Geomorphology 189, 41--54, Geomorphology, 209, 147--150,
+  \doi{10.1016/j.geomorph.2013.10.019}, 2014.
+
+\bibitem[{Plouffe and Jett{\'{e}}(1997)}]{Plouffe.Jette.1997}
+Plouffe, A. and Jett{\'{e}}, H.: Middle Wisconsinan sediments and paleoecology
+  of central British Columbia: sites at Necoslie and Nautley rivers, Can. J.
+  Earth Sci., 34, 200--208, \doi{10.1139/e17-016}, 1997.
+
+\end{thebibliography}
+
+% ----------------------------------------------------------------------
 % Interactive comment text ends

### 2015-11-12 reviews/response-jarosch.tex
Reformat referee comments (split long lines).

@@
 \referee{%
-    \textbf{P4151 L11-16:}
-    In this sentence the authors refer back to their previous work
-    \citep{Seguinot.etal.2014} and highlight that the NARR temperature and precipitation
-    fields are the most suitable present day climate datasets to be used.
-    Especially since the NARR precipitation fields include steep precipitation
-    gradients which are required as identified by et \citet{Seguinot.etal.2014}.
-    NARR is delivered on a 32\,km Lambert grid, and thus it is ques- tionable
-    how ``steep'' these gradients can be, given the rather smooth
-    representation of the existing topography on a 32\,km grid. \citet{Seguinot.etal.2014}
-    have partly discussed that however. NARR precipitation and
-    temperature fields have been evaluated in detail based on available station
-    data for large parts of the study domain dealt with in this manuscript.
-    This evaluation \citep{Jarosch.etal.2010} demonstrated that NARR has
-    difficulties simulating orographic processes in the Coast Mountains which
-    in turn results in unrealistic atmospheric conditions over the Rocky
-    Mountains. \citet{Jarosch.etal.2010} further concluded that physics based
-    downscaling is required to adequately drive glacier models in that region.
-    The authors should argue in more detail here why they think that NARR
-    precipitation fields at 32\,km are adequate to drive their model and
-    reflect their arguments with the findings of \citet{Jarosch.etal.2010}. A
-    solid argument here is of special importance as the authors assume the
+    \textbf{P4151 L11-16:} In this sentence the authors refer back to their
+    previous work \citep{Seguinot.etal.2014} and highlight that the NARR
+    temperature and precipitation fields are the most suitable present day
+    climate datasets to be used. Especially since the NARR precipitation fields
+    include steep precipitation gradients which are required as identified by
+    et \citet{Seguinot.etal.2014}. NARR is delivered on a 32\,km Lambert grid,
+    and thus it is ques- tionable how ``steep'' these gradients can be, given
+    the rather smooth representation of the existing topography on a 32\,km
+    grid. \citet{Seguinot.etal.2014} have partly discussed that however. NARR
+    precipitation and temperature fields have been evaluated in detail based on
+    available station data for large parts of the study domain dealt with in
+    this manuscript. This evaluation \citep{Jarosch.etal.2010} demonstrated
+    that NARR has difficulties simulating orographic processes in the Coast
+    Mountains which in turn results in unrealistic atmospheric conditions over
+    the Rocky Mountains. \citet{Jarosch.etal.2010} further concluded that
+    physics based downscaling is required to adequately drive glacier models in
+    that region. The authors should argue in more detail here why they think
+    that NARR precipitation fields at 32\,km are adequate to drive their model
+    and reflect their arguments with the findings of \citet{Jarosch.etal.2010}.
+    A solid argument here is of special importance as the authors assume the
     present day precipitation fields to be valid throughout their model time
@@
     but it also depends on several parameters to translate enthalpy within the
-    ice to ice viscosity \citep[see][equations 62-65)]{Aschwanden.etal.2012}. The
-    authors do not mention any of these parameters (e.g. any of the rate
+    ice to ice viscosity \citep[see][equations 62-65)]{Aschwanden.etal.2012}.
+    The authors do not mention any of these parameters (e.g. any of the rate
     factors or nonlinear power $n$ ) within the manuscript or in Table~1. I
@@
     working grids? This is crucial information to be included in the
-    manuscript. It has been demonstrated by spectral power analysis \citep{Jarosch.etal.2010}
-    that the NARR precipitation fields do not contain any
-    significant spacial information below approximately 39\,km resolution and
-    that bilinear interpolation does not add any information whatsoever on
+    manuscript. It has been demonstrated by spectral power analysis
+    \citep{Jarosch.etal.2010} that the NARR precipitation fields do not contain
+    any significant spacial information below approximately 39\,km resolution
+    and that bilinear interpolation does not add any information whatsoever on
     smaller scales, which should come to no surprise. Physics-based downscaling
@@
     computational grids of 10 and 5\,km for precipitation and temperature and
-    discuss their choice in the light of the findings from \citet{Jarosch.etal.2010}.
-    Temperature however is better constrained in NARR \citep{Jarosch.etal.2010}
-    and contains spectral information down to 10\,km resolution, which
-    justifies the usage of NARR temperature fields on the 10\,km computational
-    grid of this study. The 5\,km grid still needs to be argued for.}
+    discuss their choice in the light of the findings from
+    \citet{Jarosch.etal.2010}. Temperature however is better constrained in
+    NARR \citep{Jarosch.etal.2010} and contains spectral information down to
+    10\,km resolution, which justifies the usage of NARR temperature fields on
+    the 10\,km computational grid of this study. The 5\,km grid still needs to
+    be argued for.}
 

### 2015-11-12 reviews/response-jarosch.tex
Begin to add TODO items to assess the reviews.

@@
 \definecolor{journalname}{rgb}{0.34,0.59,0.82}
+\definecolor{adgeored}{rgb}{0.439,0.157,0.145}  % from Advances in Geoscience
 
@@
 \newcommand{\msquote}[1]{\begin{quote}\textit{#1}\end{quote}}
-\newcommand{\todo}[1]{\textcolor{red}{TODO: #1}}
+\newcommand{\todo}[1]{\textcolor{adgeored}{TODO: #1}}
 
@@
     there are mismatches with geological evidence.}
+
+\todo{%
+    Perhaps explain here that this fundamental challenge is not one posed by
+    the presentation of the manuscript, but one linked to the mutual
+    understanding of the glaciological and geomorphological communities.}
 
@@
 
+\todo{%
+    Perform new sensitivity tests on rheological parameters (water content,
+    activation energies, rate factors) and sliding (friction angle). This will
+    take some time. Based on the results, decide if they should be presented in
+    the manuscript (before/after) discussion, in a supplement, or perhaps just
+    here in this response.}
+
 \referee{%
@@
     manuscript entirely up to the authors.}
+
+\todo{%
+    I have no intention to remove half the manuscript. Argue against it here.
+    Geomorphology is no exact science and has to be speculative.}
 
@@
     equation~6).}
+
+\todo{%
+    Decide what to do!}
 

### 2015-11-12 reviews/response-marshall.tex
Begin to add TODO items to assess the reviews.

@@
 \definecolor{journalname}{rgb}{0.341,0.596,0.824}
-\definecolor{darkred}{rgb}{0.439,0.157,0.145}  % from Advances in Geoscience
+\definecolor{adgeored}{rgb}{0.439,0.157,0.145}  % from Advances in Geoscience
 
@@
 \newcommand{\msquote}[1]{\begin{quote}\textit{#1}\end{quote}}
-\newcommand{\todo}[1]{\textcolor{darkred}{TODO: #1}}
+\newcommand{\todo}[1]{\textcolor{adgeored}{TODO: #1}}
 
@@
 
-\todo{Discuss in the manuscript.}
+\todo{%
+    Perhaps we forgot to mention that as a limitation in the manuscript. Find
+    the appropriate place to add this in the discussion, and rephrase the above
+    points into the manuscript.}
 
@@
 
-\todo{Correct mistakes.}
+\todo{%
+    Apply grammar corrections.}
 

### 2015-11-13 reviews/response-jarosch.tex
Continue to add todo items.

@@
 
+\todo{%
+    Mention that ice cover could be overestimated in this region. The main
+    problematic area are the St.-Elias Mountains. However, there is no ice
+    thickness data available as far as I know. Check whether there is more info
+    about that in \citet{Waechter.etal.2015}}
+
 \referee{%
@@
     on the SSA being the ``sliding law'' for the SIA.}
+
+\todo{%
+    Agree. Reformulate.}
 
@@
 
+\todo{%
+    Pull values from parameter table in PhD kappa. Wait for results of
+    sensitivity tests.}
+
 \referee{%
@@
     ice-bedrock interface. Please be more specific here.}
+
+\todo{%
+    From the ice-bedrock interface downward. Clarify.}
 
@@
     made above in my general comments.}
+
+\todo{%
+    Pull parameters and explanation from Kappa. Check for newer PISM refs on
+    modelling Greenland or Antarctica. Wait for results from sensitivity tests}
 
@@
 
+\todo{%
+    NARR data fields have been bilinearly interpolated. Decide what to do!}
+
 \referee{%
@@
     I leave it to the authors to explore this possibility.}
+
+\todo{%
+    I am not sure how to adress this at the moment. Temperature lapse rates
+    vary significantly at present, but I am not sure if there is much knowledge
+    on how they varied in the past.}
 
@@
 
+\todo{%
+    Correct.}
+
 \referee{%
@@
     double ``the'' in the sentence.}
+
+\todo{%
+    Correct.}
 
@@
 
+\bibitem[{Waechter et~al.(2015}]{Waechter.etal.2015}
+Waechter, A., Copland, L., and Herdes, E.: Modern glacier velocities across the
+  Icefield Ranges, St Elias Mountains, and variability at selected glaciers
+  from 1959 to 2012, J. Glaciol., 61, 624--634, \doi{10.3189/2015jog14j147},
+  2015.
+
 \end{thebibliography}

### 2015-11-13 reviews/response-stumpf.tex
Reformat referee comments (split long lines).
Begin to add TODO items to assess the reviews.
Continue to add todo items.

@@
     non-glacial conditions existed during MIS (Olympia Nonglacial Interval;
-    \citealp[e.g.,][]{Plouffe.Jette.1997}). I applaud them for undertaking of such a
-    difficult task, and notifying the reader where input data is sparse or
-    inconclusive.}
+    \citealp[e.g.,][]{Plouffe.Jette.1997}). I applaud them for undertaking of
+    such a difficult task, and notifying the reader where input data is sparse
+    or inconclusive.}
 
@@
     advanced our understanding of the CIS, and could be an important dataset to
-    test against the modeling. \citet{Stumpf.etal.2014} provides a list of some of
-    these studies.}
+    test against the modeling. \citet{Stumpf.etal.2014} provides a list of some
+    of these studies.}
+
+\todo{%
+    Rephrase, and perhaps add a few references from \citet{Stumpf.etal.2014},
+    although I feel I already have many.}
 
@@
 
+\todo{%
+    Add info where it will not disturb the figures. Or perhaps a larger
+    location map like in the Kappa?}
+
 \referee{%
     3) Was the model tested against regional-scale ground-based data
-    \citep[e.g.,][]{Ferbey.etal.2013} to constrain ice divide positions, ice flow direction,
-    and ice sheet thickness?}
+    \citep[e.g.,][]{Ferbey.etal.2013} to constrain ice divide positions, ice
+    flow direction, and ice sheet thickness?}
+
+\todo{%
+    Nice dataset. Not quantitatively tested, if that is what the comment means.
+    In my understanding, there is very little reliable info about ice sheet
+    thickness, ice divide positions are reconstructions that always involve
+    some level of interpretation. Comparing flow direction against
+    \citet{Ferbey.etal.2013} would be very nice but I don't think we are this
+    confident in the basal velocities for such a comparison yet.}
 
 \referee{%
-    4) For the central sector of the CIS, \citet{Stumpf.etal.2000} provides some
-    insight into the chronology and effectiveness of glacial erosional during
-    the MIS. In figure~5, and in the accompanying text, they describe how
-    landforms on the surface formed. For example, in lake valleys east of the
-    Skeena Mountains, it appeared the major glacial streamlined landforms were
-    formed during a longer glacial advance phase, with ice flow paralleling the
-    valleys, and later flows, some perpendicular the valley flow, only weakly
-    impacted them.}
+    4) For the central sector of the CIS, \citet{Stumpf.etal.2000} provides
+    some insight into the chronology and effectiveness of glacial erosional
+    during the MIS. In figure~5, and in the accompanying text, they describe
+    how landforms on the surface formed. For example, in lake valleys east of
+    the Skeena Mountains, it appeared the major glacial streamlined landforms
+    were formed during a longer glacial advance phase, with ice flow
+    paralleling the valleys, and later flows, some perpendicular the valley
+    flow, only weakly impacted them.}
+
+\todo{%
+    Add reference to \citet{Stumpf.etal.2000}...}
 
@@
     margins westward into these mountains was limited.}
+
+\todo{%
+    Add reference to \citet{Stumpf.etal.2000}...}
 

### 2015-12-16 cordillera-cycle.tex
Restore manuscript format.
Rename all figures. Fix Makefile.

@@
 % Copernicus manuscript
-%\documentclass[tc, manuscript]{copernicus}
+\documentclass[tc, manuscript]{copernicus}
 
 % Copernicus final print
-\documentclass[tc]{copernicus}
+%\documentclass[tc]{copernicus}
 
@@
 \hypersetup{hidelinks}
+
+% Push floats to the end
+\usepackage[nolists, nomarkers]{endfloat}
 
@@
            temperature offset time-series used to force the ice sheet model
-           through the last glacial cycle (Fig.~\ref{fig:timeseries}). $f$
+           through the last glacial cycle (Fig.~\ref{fig:lr_ts}). $f$
            corresponds to the scaling factor adopted to yield last glacial
@@
 order to simulate comparable ice extents at the LGM  (Table~\ref{tab:extrema})
-and realistic outlines (Fig.~\ref{fig:snapshots}).
+and realistic outlines (Fig.~\ref{fig:lr_maps}).
 
@@
 \begin{figure*}
-  \includegraphics{timeseries}
+  \includegraphics{lr_ts}
   \caption{Temperature offset time-series from ice core and ocean records
@@
            5\,km-resolution runs.}
-  \label{fig:timeseries}
+  \label{fig:lr_ts}
 \end{figure*}
@@
 \begin{figure*}
-  \includegraphics{snapshots}
+  \includegraphics{lr_maps}
   \caption{Snapshots of modelled surface topography (500\,m contours)
            corresponding to the ice volume extremes indicated on
-           Fig.~\ref{fig:timeseries}. An ice cap persists over the Skeena
+           Fig.~\ref{fig:lr_ts}. An ice cap persists over the Skeena
            Mountains (SM) during MIS~3. Note the occurence of spatial
            similarities despite large differences in timing.}
-  \label{fig:snapshots}
+  \label{fig:lr_maps}
 \end{figure*}
@@
            MIS~4, 3 and 2 for each of the six low-resolution simulations
-           (Fig.~\ref{fig:timeseries}).}
+           (Fig.~\ref{fig:lr_ts}).}
   \label{tab:extrema}
@@
 Despite large differences in the input climate forcing
-(Fig.~\ref{fig:timeseries}, upper panel), model output presents consistent
+(Fig.~\ref{fig:lr_ts}, upper panel), model output presents consistent
 features that can be observed across the range of forcing data used. In all
@@
 and 56.5\,ka during MIS~4, and between 23.2 and 16.8\,ka during MIS~2
-(Fig.~\ref{fig:timeseries}, lower panel). An ice volume minimum is
+(Fig.~\ref{fig:lr_ts}, lower panel). An ice volume minimum is
 consistently reached between 53.0 and 41.3\,ka during MIS~3. However, the
@@
 deglaciation but in three steps, separated by two periods of ice sheet regrowth
-(Fig.~\ref{fig:timeseries}).
+(Fig.~\ref{fig:lr_ts}).
 
@@
 sheet, covering an area of at least half of that attained during MIS~2
-(Table~\ref{tab:extrema}; Fig.~\ref{fig:snapshots}, upper panels).
+(Table~\ref{tab:extrema}; Fig.~\ref{fig:lr_maps}, upper panels).
 Corresponding maximum ice volumes also differ significantly between model runs,
@@
 over the Skeena
-Mountains (Fig.~\ref{fig:snapshots}, middle panels). Although this ice cap is
+Mountains (Fig.~\ref{fig:lr_maps}, middle panels). Although this ice cap is
 present in all simulations, its dimensions depend sensitively on the choice of
@@
 
-Modelled ice sheet geometries during the LGM (MIS~2; Fig.~\ref{fig:snapshots},
+Modelled ice sheet geometries during the LGM (MIS~2; Fig.~\ref{fig:lr_maps},
 lower panels) invariably include a ca. 1500\,km-long central divide above
@@
 is obtained during MIS~4 (61.9--56.5\,ka) and the second during MIS~2
-(23.2--16.8\,ka; Figs.~\ref{fig:timeseries}, \ref{fig:snapshots};
+(23.2--16.8\,ka; Figs.~\ref{fig:lr_ts}, \ref{fig:lr_maps};
 Table~\ref{tab:extrema}). These events broadly correspond in timing to the
@@
 extents that may be compatible with these field constraints
-(Fig.~\ref{fig:timeseries}, lower panel; Table~\ref{tab:extrema}).
+(Fig.~\ref{fig:lr_ts}, lower panel; Table~\ref{tab:extrema}).
 Simulations driven by the NGRIP, ODP~1012 and ODP~1020
@@
 deglaciation is caused by an early warming present in the alkenone
-palaeo-temperature reconstructions (Fig.~\ref{fig:timeseries}, upper panel;
+palaeo-temperature reconstructions (Fig.~\ref{fig:lr_ts}, upper panel;
 \citealp[Fig.~3]{Herbert.etal.2001}). However, this
@@
 Similarly, the simulation using the NGRIP palaeo-temperature record depicts an
-early onset of deglaciation (Fig.~\ref{fig:timeseries}) following its last
+early onset of deglaciation (Fig.~\ref{fig:lr_ts}) following its last
 glacial volume maximum (22.9\,ka, Table~\ref{tab:extrema}) attained about 6\,ka
@@
 simulations were re-run using a higher-resolution grid (Sect.~\ref{sec:model};
-Fig.~\ref{fig:timeseries}, lower panel, dotted lines).
+Fig.~\ref{fig:lr_ts}, lower panel, dotted lines).
 
@@
 \begin{figure*}
-  \includegraphics{icemaps-mis2}
+  \includegraphics{hr_maps_mis2}
   \caption{Modelled surface topography (200\,m contours) and surface velocity
@@
 \begin{figure*}
-  \includegraphics{icemaps-mis4}
+  \includegraphics{hr_maps_mis4}
   \caption{Modelled surface topography (200\,m contours) and surface velocity
@@
 divide over
-the western flank of the Rocky Mountains (Fig.~\ref{fig:snapshots}, lower
+the western flank of the Rocky Mountains (Fig.~\ref{fig:lr_maps}, lower
 panels; Fig.~\ref{fig:mis2}). This result appears to contrast with
@@
 MIS~4 are more sensitive to the choice of atmospheric forcing than those
-corresponding to ice volume maxima during MIS~2 (Fig.~\ref{fig:snapshots},
+corresponding to ice volume maxima during MIS~2 (Fig.~\ref{fig:lr_maps},
 upper panels;
@@
 northern and eastern sectors. In contrast, the EPICA simulation produces a lower
-ice volume maximum (Fig.~\ref{fig:timeseries}), which translates in the
+ice volume maximum (Fig.~\ref{fig:lr_ts}), which translates in the
 modelled ice sheet geometry into a significantly reduced southern sector, more
@@
 \begin{figure*}
-  \includegraphics{duration}
+  \includegraphics{hr_geom_duration}
   \caption{Modelled duration of ice cover during the last 120\,ka using GRIP
@@
            simulation. See Fig.~\ref{fig:locmap} for a list of abbreviations.}
-  \label{fig:duration}
+  \label{fig:hr_geom_duration}
 \end{figure*}
@@
 to compute durations of ice cover throughout the last glacial
-cycle (Fig.~\ref{fig:duration}).
+cycle (Fig.~\ref{fig:hr_geom_duration}).
 
@@
 cover is restricted to disjoint ice caps centred on major mountain ranges of
-the North American Cordillera (Fig.~\ref{fig:duration}, blue areas). A
+the North American Cordillera (Fig.~\ref{fig:hr_geom_duration}, blue areas). A
 2500\,km-long continuous expanse of ice, extending from the Alaska Range in the
@@
 for at most 32\,ka, which is about a third of the timespan of the last glacial cycle
-(Fig.~\ref{fig:duration}, hatched areas). However, except for its margins in
+(Fig.~\ref{fig:hr_geom_duration}, hatched areas). However, except for its margins in
 the Pacific Ocean and in the northern foothills of the Alaska Range,
 the maximum extent of the ice sheet is attained for a much shorter period
-of time of only few thousand years (Figs.~\ref{fig:duration}, red areas). This
+of time of only few thousand years (Figs.~\ref{fig:hr_geom_duration}, red areas). This
 result illustrates that the maximum extents of the modelled ice sheet during
@@
 is the northern slope of the Alaska Range, where modelled glaciers are confined
-to its foothills during the entire simulation period (Fig.~\ref{fig:duration},
+to its foothills during the entire simulation period (Fig.~\ref{fig:hr_geom_duration},
 AR). This apparent insensitivity of modelled glacial extent to temperature
@@
 In our simulations, major ice-dispersal centres, visible on the modelled ice
-cover duration maps (Fig.~\ref{fig:duration}), are located over the Coast
+cover duration maps (Fig.~\ref{fig:hr_geom_duration}), are located over the Coast
 Mountains (CM), the Columbia and Rocky mountains (CRM), the Skeena Mountains
@@
 entire length of both simulations, but that ice cap does not appear to be a
-major feed to the Cordilleran ice sheet (Fig.~\ref{fig:duration}, WSEM).
+major feed to the Cordilleran ice sheet (Fig.~\ref{fig:hr_geom_duration}, WSEM).
 Although the Coast, Skeena and Columbia and Rocky mountains (CM, SM, CRM)
@@
 during the coldest periods of the last glacial cycle. The Northern Rocky
-Mountains (Fig.~\ref{fig:duration}, NRM) do not act as a nucleation centre,
+Mountains (Fig.~\ref{fig:hr_geom_duration}, NRM) do not act as a nucleation centre,
 but rather as a pinning point for the Cordilleran ice sheet margin coming from
@@
 other ice-dispersal centres
-(Figs.~\ref{fig:snapshots} and~\ref{fig:duration}, SM). Regardless of the applied
-forcing, this ice cap appears to survive MIS~3 (Fig.~\ref{fig:snapshots},
+(Figs.~\ref{fig:lr_maps} and~\ref{fig:hr_geom_duration}, SM). Regardless of the applied
+forcing, this ice cap appears to survive MIS~3 (Fig.~\ref{fig:lr_maps},
 middle panels), and serves as a nucleation centre at the onset of the glacial
@@
 \begin{figure*}
-  \includegraphics{warmbase}
+  \includegraphics{hr_geom_warmbase}
   \caption{Modelled duration of warm-based ice cover during the last
@@
            Hatches indicate areas that were covered by cold ice only.}
-  \label{fig:warmbase}
+  \label{fig:hr_geom_warmbase}
 \end{figure*}
@@
   \vspace{-0.5mm}  % enough to allow the two figures to fit on one page
-  \includegraphics{warmfrac}
+  \includegraphics{hr_geom_warmfrac}
   \caption{Modelled fraction of warm-based ice cover during the ice-covered
@@
            Hatches indicate areas that were covered by cold ice only.}
-  \label{fig:warmfrac}
+  \label{fig:hr_geom_warmfrac}
 \end{figure*}
@@
 A correlation is observed between the modelled duration of warm based ice cover
-(Fig.~\ref{fig:warmbase}) and the degree of glacial modification of the
+(Fig.~\ref{fig:hr_geom_warmbase}) and the degree of glacial modification of the
 landscape (mainly in terms of the development of deep glacial valleys and
@@
 radiating off the Skeena Mountains
-(Figs.~\ref{fig:duration} and \ref{fig:warmbase};
+(Figs.~\ref{fig:hr_geom_duration} and \ref{fig:hr_geom_warmbase};
 \citealp[Fig.~2]{Kleman.etal.2010}).
@@
 \citep[Fig.~2]{Kleman.etal.2010}. We suggest that
-persistent ice cover (Fig.~\ref{fig:duration}) associated with basal ice
-temperatures at the pressure melting point (Figs.~\ref{fig:warmbase}
-and~\ref{fig:warmfrac}) explains the large-scale glacial erosional imprint on
+persistent ice cover (Fig.~\ref{fig:hr_geom_duration}) associated with basal ice
+temperatures at the pressure melting point (Figs.~\ref{fig:hr_geom_warmbase}
+and~\ref{fig:hr_geom_warmfrac}) explains the large-scale glacial erosional imprint on
 the landscape. A well-developed network of glacial valleys running to the
@@
 {Stroeven.etal.2010}) is modelled to have hosted predominantly warm-based ice
-(Fig.~\ref{fig:warmfrac}). However, because it is only glaciated for a
+(Fig.~\ref{fig:hr_geom_warmfrac}). However, because it is only glaciated for a
 short fraction of the last glacial cycle in our simulations
-(Fig.~\ref{fig:duration}), this perhaps indicates that
+(Fig.~\ref{fig:hr_geom_duration}), this perhaps indicates that
 the observed landscape pattern originates from multiple glacial cycles and
 witnesses an increased relative importance of the Selwyn and Mackenzie
-mountains ice dispersal centre (Fig.~\ref{fig:duration}, SMKM), prior
+mountains ice dispersal centre (Fig.~\ref{fig:hr_geom_duration}, SMKM), prior
 to the Late Pleistocene \citep[cf.][]{Ward.etal.2008, Demuro.etal.2012}.
 
-The modelled distribution of warm-based ice cover (Figs.~\ref{fig:warmbase}
-and~\ref{fig:warmfrac}) is inevitably affected by our assumption of a constant,
+The modelled distribution of warm-based ice cover (Figs.~\ref{fig:hr_geom_warmbase}
+and~\ref{fig:hr_geom_warmfrac}) is inevitably affected by our assumption of a constant,
 70\,\unit{mW\,m^{-2}} geothermal heat flux at 3\,km depth
@@
 \begin{figure}
-  \includegraphics{deglacseries}
+  \includegraphics{lr_ts_deglac}
   \caption{Temperature offset time-series from the GRIP and EPICA ice core
@@
            equivalent (bottom panel).}
-  \label{fig:deglacseries}
+  \label{fig:lr_ts_deglac}
 \end{figure}
@@
 \begin{figure*}
-  \includegraphics{deglacshots}
+  \includegraphics{hr_maps_deglac}
   \caption{Snapshots of modelled surface topography (200\,m contours)
@@
            Dashed segments (A--D) indicate the location of profiles used in
-           Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}.}
-  \label{fig:deglacshots}
+           Figs.~\ref{fig:hr_pf_grip} and~\ref{fig:hr_pf_epica}.}
+  \label{fig:hr_geom_deglacageshots}
 \end{figure*}
@@
   \vspace{-1.5mm}  % enough to allow the two figures to fit on one page
-  \includegraphics{deglac}
+  \includegraphics{hr_geom_deglacage}
   \caption{Modelled age of the last deglaciation. Areas that have been covered
@@
            Dashed segments (A--D) indicate the location of profiles used in
-           Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}.}
-  \label{fig:deglac}
+           Figs.~\ref{fig:hr_pf_grip} and~\ref{fig:hr_pf_epica}.}
+  \label{fig:hr_geom_deglacage}
 \end{figure*}
@@
 and the pacing of deglaciation depend critically on the choice of
-climate forcing (Table~\ref{tab:extrema}; Figs.~\ref{fig:timeseries}
-and~\ref{fig:deglacseries}). Adopting the EPICA climate forcing yields peak ice
+climate forcing (Table~\ref{tab:extrema}; Figs.~\ref{fig:lr_ts}
+and~\ref{fig:lr_ts_deglac}). Adopting the EPICA climate forcing yields peak ice
 volume at 17.2\,ka and an uninterrupted deglaciation until about 9\,ka
-(Fig.~\ref{fig:deglacseries}, lower panel, red curves). On the contrary, the
+(Fig.~\ref{fig:lr_ts_deglac}, lower panel, red curves). On the contrary, the
 simulation driven by the GRIP palaeo-temperature record yields peak ice volume
@@
 about 8\,ka. The first interruption occurs between 16.6 and 14.5\,ka, and the
-second between 13.1 and 11.6\,ka (Fig.~\ref{fig:deglacseries}, lower panel,
+second between 13.1 and 11.6\,ka (Fig.~\ref{fig:lr_ts_deglac}, lower panel,
 blue curve).
@@
 Modelled patterns of ice sheet retreat are relatively consistent between the
-two simulations (Figs.~\ref{fig:deglacshots} and.~\ref{fig:deglac}). The
+two simulations (Figs.~\ref{fig:hr_geom_deglacageshots} and.~\ref{fig:hr_geom_deglacage}). The
 southern sector of the modelling domain, including the Puget Lowland, the Coast
@@
 Cordilleran ice sheet, the Finlay River area in the Omenica Mountains
-(Fig.~\ref{fig:deglac}, OM) presents a different kind of evidence.
+(Fig.~\ref{fig:hr_geom_deglacage}, OM) presents a different kind of evidence.
 There, sharp-crested moraines indicate a late-glacial readvance of local alpine
@@
 readvance in the Coast Mountains and in the Columbia and Rocky Mountains
-(Fig.~\ref{fig:deglac},
+(Fig.~\ref{fig:hr_geom_deglacage},
 left panel). In addition to matching the location of some local readvances, the
@@
 western slopes of the Saint Elias and the Coast mountains
-(Fig.~\ref{fig:deglac}, right panel).
+(Fig.~\ref{fig:hr_geom_deglacage}, right panel).
 
@@
 \begin{figure*}
-  \includegraphics{lastflow}
+  \includegraphics{hr_geom_lastflow}
   \caption{Modelled deglacial basal ice velocities. Hatches
@@
            using a basal velocity threshold of 1\,\unit{m\,yr^{-1}}.}
-  \label{fig:lastflow}
+  \label{fig:hr_geom_lastflow}
 \end{figure*}
@@
 \begin{figure}
-  \includegraphics{profiles-grip}
+  \includegraphics{hr_pf_grip}
   \caption{Modelled bedrock (black) and ice surface (blue) topography profiles
@@
            simulation, corresponding to the four transects indicated in
-           Figs.~\ref{fig:deglacshots} and~\ref{fig:deglac}.}
-  \label{fig:profiles-grip}
+           Figs.~\ref{fig:hr_geom_deglacageshots} and~\ref{fig:hr_geom_deglacage}.}
+  \label{fig:hr_pf_grip}
 \end{figure}
@@
 \begin{figure}
-  \includegraphics{profiles-epica}
+  \includegraphics{hr_pf_epica}
   \caption{Modelled bedrock (black) and ice surface (red) topography profiles
@@
            simulation, corresponding to the four transects indicated in
-           Figs.~\ref{fig:deglacshots} and~\ref{fig:deglac}.}
-  \label{fig:profiles-epica}
+           Figs.~\ref{fig:hr_geom_deglacageshots} and~\ref{fig:hr_geom_deglacage}.}
+  \label{fig:hr_pf_epica}
 \end{figure}
@@
 immediately preceding deglaciation or at the time of cessation of sliding
-inside a cold-based retreating margin (Fig.~\ref{fig:lastflow}). The
+inside a cold-based retreating margin (Fig.~\ref{fig:hr_geom_lastflow}). The
 modelled deglacial flow patterns are mostly consistent between the GRIP and EPICA
@@
 of the modelling domain do not experience any basal sliding throughout the
-deglaciation phase (Fig.~\ref{fig:lastflow}, hatched areas). This notably
+deglaciation phase (Fig.~\ref{fig:hr_geom_lastflow}, hatched areas). This notably
 includes parts of the Interior Plateau in British Columbia, major
@@
 similarities with the patterns of deglacial ice flow from numerical modelling
-(Fig.~\ref{fig:lastflow}). In the northern half
+(Fig.~\ref{fig:hr_geom_lastflow}). In the northern half
 of the modelling domain, modelled deglacial flow directions depict an
@@
 mountain ranges. Converging deglacial flow patterns in the Liard
-Lowland, for instance (Fig.~\ref{fig:lastflow}), closely resemble the pattern
+Lowland, for instance (Fig.~\ref{fig:hr_geom_lastflow}), closely resemble the pattern
 indicated by glacial lineations \citep[Fig.~2]{Margold.etal.2013}.
@@
 produce a retreat of the ice margin towards the north-east
-(Fig.~\ref{fig:deglac}), a pattern which is validated by the geomorphological
+(Fig.~\ref{fig:hr_geom_deglacage}), a pattern which is validated by the geomorphological
 and stratigraphical record for ancient pro-glacial lakes dammed by the
@@
 EPICA simulation produces negligible basal sliding on the plateau during deglaciation
-(Fig.~\ref{fig:lastflow}). Yet, the Interior Plateau also hosts an impressive
+(Fig.~\ref{fig:hr_geom_lastflow}). Yet, the Interior Plateau also hosts an impressive
 set of glacial lineations which indicate a substantial eastwards flow component
@@
 contrary, enlarge the region of negligible basal sliding
-(Fig.~\ref{fig:lastflow}). Thus, a second explanation for the incongruent
+(Fig.~\ref{fig:hr_geom_lastflow}). Thus, a second explanation for the incongruent
 results could be that the Interior Plateau lineation system predates
@@
 The modelled deglaciation of the Interior Plateau of British Columbia
-consists of a rapid northwards retreat (Fig.~\ref{fig:deglac}) of
-southwards-flowing non-sliding ice lobes (Fig.~\ref{fig:lastflow}) positioned
+consists of a rapid northwards retreat (Fig.~\ref{fig:hr_geom_deglacage}) of
+southwards-flowing non-sliding ice lobes (Fig.~\ref{fig:hr_geom_lastflow}) positioned
 in-between deglaciated mountain ranges
-(Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}). This result
+(Figs.~\ref{fig:hr_pf_grip} and~\ref{fig:hr_pf_epica}). This result
 appears compatible with the prevailing conceptual model of deglaciation of

### 2015-12-16 figures/hr_geom_basedispl.py

@@ -61,4 +61,4 @@ for i, rec in enumerate(records):
 # add colorbar and save
 cb = fig.colorbar(cf, cax, format='%i')
 cb.set_label('Cumulative basal displacement (km)')
-fig.savefig('erosion')
+fig.savefig('hr_geom_basedispl')

### 2015-12-16 figures/hr_geom_deglacage.py

@@ -81,5 +81,5 @@ ut.pl.add_pointer_tag(grid[0], 'OM', xy=(-1800e3, 1400e3), xytext=(-1300e3, 1250
 print 'saving deglac...'
 cb = fig.colorbar(cs, cax, ticks=ages)
 cb.set_label('Deglaciation age (ka)')
-fig.savefig('deglac')
+fig.savefig('hr_geom_deglacage')
 nc.close()

### 2015-12-16 figures/hr_geom_duration.py

@@ -74,4 +74,4 @@ ut.pl.add_pointer_tag(ax, 'CRM', xy=(-1550e3,  650e3), xytext=(-1200e3,  650e3))
 # add colorbar and save
 cb = fig.colorbar(cf, cax)
 cb.set_label('Duration of ice cover (ka)')
-fig.savefig('duration')
+fig.savefig('hr_geom_duration')

### 2015-12-16 figures/hr_geom_lastflow.py

@@ -95,4 +95,4 @@ ut.pl.add_pointer_tag(ax, 'IP', xy=(-1850e3, 900e3), xytext=(-1100e3, 900e3))
 print 'saving...'
 cb = ColorbarBase(cax, cmap=cmap, norm=norm, ticks=range(8, 23, 2))
 cb.set_label('Age of last basal sliding (ka)')
-fig.savefig('lastflow')
+fig.savefig('hr_geom_lastflow')

### 2015-12-16 figures/hr_geom_warmbase.py

@@ -67,4 +67,4 @@ ut.pl.add_pointer_tag(ax, 'MKM', xy=(-1550e3, 2000e3), xytext=(-1200e3, 2000e3))
 # add colorbar and save
 cb = fig.colorbar(cs, cax, ticks=levs[1:])
 cb.set_label('Duration of warm-based ice cover (ka)')
-fig.savefig('warmbase')
+fig.savefig('hr_geom_warmbase')

### 2015-12-16 figures/hr_geom_warmfrac.py

@@ -67,4 +67,4 @@ ut.pl.add_pointer_tag(ax, 'MKM', xy=(-1550e3, 2000e3), xytext=(-1200e3, 2000e3))
 # add colorbar and save
 cb = fig.colorbar(cs, cax, ticks=levs[1:])
 cb.set_label('Fraction of warm-based ice cover')
-fig.savefig('warmfrac')
+fig.savefig('hr_geom_warmfrac')

### 2015-12-16 figures/hr_maps_deglac.py

@@ -58,4 +58,4 @@ cb = fig.colorbar(im, cax, extend='both', format='%i',
                   ticks=np.logspace(1, 3.5, 6))
 cb.set_label(r'surface velocity ($m\,yr^{-1}$)', labelpad=-2.0)
 print 'saving...'
-fig.savefig('deglacshots')
+fig.savefig('hr_maps_deglac')

### 2015-12-16 figures/hr_maps_mis2.py

@@ -1,11 +1,11 @@
 #!/usr/bin/env python2
 # coding: utf-8
 
-from icemaps import icemaps
+from hr_maps_base import icemaps
 
 # plot
 fig = icemaps(mis=2)
 
 # save
 print('saving...')
-fig.savefig('icemaps-mis2')
+fig.savefig('hr_maps_mis2')

### 2015-12-16 figures/hr_maps_mis3.py

@@ -1,11 +1,11 @@
 #!/usr/bin/env python2
 # coding: utf-8
 
-from icemaps import icemaps
+from hr_maps_base import icemaps
 
 # plot
 fig = icemaps(mis=3)
 
 # save
 print('saving...')
-fig.savefig('icemaps-mis3')
+fig.savefig('hr_maps_mis3')

### 2015-12-16 figures/hr_maps_mis4.py

@@ -1,11 +1,11 @@
 #!/usr/bin/env python2
 # coding: utf-8
 
-from icemaps import icemaps
+from hr_maps_base import icemaps
 
 # plot
 fig = icemaps(mis=4)
 
 # save
 print('saving...')
-fig.savefig('icemaps-mis4')
+fig.savefig('hr_maps_mis4')

### 2015-12-16 figures/hr_maps_puget.py

@@ -58,5 +58,5 @@ for i, t in enumerate(times):
 
 # save
 print 'saving puget...'
-fig.savefig('puget')
+fig.savefig('hr_maps_puget')
 nc.close()

### 2015-12-16 figures/hr_pf_epica.py

@@ -2,7 +2,7 @@
 # coding: utf-8
 
 import util as ut
-from profiles import profiles
+from hr_pf_base import profiles
 
 # parameters
 res = '5km'
@@ -15,4 +15,4 @@ fig = profiles(res, rec, dt, color)
 
 # save
 print('saving...')
-fig.savefig('profiles-epica')
+fig.savefig('hr_pf_epica')

### 2015-12-16 figures/hr_pf_grip.py

@@ -2,7 +2,7 @@
 # coding: utf-8
 
 import util as ut
-from profiles import profiles
+from hr_pf_base import profiles
 
 # parameters
 res = '5km'
@@ -15,4 +15,4 @@ fig = profiles(res, rec, dt, color)
 
 # save
 print('saving...')
-fig.savefig('profiles-grip')
+fig.savefig('hr_pf_grip')

### 2015-12-16 figures/lr_maps.py

@@ -71,4 +71,4 @@ cb = fig.colorbar(cs, cax, ticks=levs[::2],
                   format=FuncFormatter(lambda x, pos: '%g' % (x/1000.0)))
 cb.set_label(r'surface elevation (km)')  #, labelpad=-1.5*pt2mm)
 print 'saving snapshots...'
-fig.savefig('snapshots')
+fig.savefig('lr_maps')

### 2015-12-16 figures/lr_ts.py

@@ -116,4 +116,4 @@ ax2.set_xlabel('model age (ka)')
 ax1.grid(axis='y')
 ax2.grid(axis='y')
 ax2.legend(loc='upper left', ncol=2)
-fig.savefig('timeseries')
+fig.savefig('lr_ts')

### 2015-12-16 figures/lr_ts_deglac.py

@@ -53,4 +53,4 @@ ax2.set_xlabel('model age (ka)')
 ax2.legend(loc='upper right', bbox_to_anchor=(1, 1+2.5/32.5))
 ax1.grid(axis='y')
 ax2.grid(axis='y')
-fig.savefig('deglacseries')
+fig.savefig('lr_ts_deglac')

### 2015-12-16 figures/lr_ts_diff.py

@@ -68,4 +68,4 @@ ax.yaxis.set_label_coords(-0.05, 0.5)
 ax.set_xlabel('model age (ka)')
 ax.grid(axis='y', c='0.5', ls='-', lw=0.1)
 ax.legend(loc='upper left', ncol=2)
-fig.savefig('diffseries')
+fig.savefig('lr_ts_diff')

### 2015-12-16 figures/lr_violins.py

@@ -23,4 +23,4 @@ iplt.violinplot(data)
 print 'saving violins...'
 ax.set_ylim(0.0, 10.0)
 ax.set_ylabel('ice volume (m s.-l. eq.)')
-fig.savefig('violins')
+fig.savefig('lr_violins')

### 2015-12-16 figures/sens_ntil.py

@@ -92,4 +92,4 @@ if __name__ == '__main__':
     grid[1].set_yscale('log')
 
     # save
-    fig.savefig('sens-ntil')
+    fig.savefig('sens_ntil')

### 2015-12-16 figures/sens_rheo.py

@@ -81,4 +81,4 @@ if __name__ == '__main__':
                  va='center', rotation=90.0, transform=grid[1].transAxes)
     
     # save
-    fig.savefig('sens-rheo')
+    fig.savefig('sens_rheo')

### 2015-12-29 reviews/response-jarosch.tex
Begin to respond to A. Jarosch.

@@
 
+We apologize for our delayed response and thank you for this detailed review of
+our manuscript. Following your comments, we decide to amend the manuscript with
+a new section that explores the sensitivity of one of our runs to some of the
+parameters governing ice deformation and sliding. Although the discussion of
+model results against geological evidence is indeed extensive and partly
+speculative, we prefer not to shorten it for reasons we will detail below.
+Although you may feel we have not applied all the changes you requested, we
+hope that those we do plan fill much of the voids you pointed at.
+
 % ----------------------------------------------------------------------
@@
 
+Thank you for this positive summary of our work.
+
 \referee{%
@@
 
-\todo{%
-    Perhaps explain here that this fundamental challenge is not one posed by
-    the presentation of the manuscript, but one linked to the mutual
-    understanding of the glaciological and geomorphological communities.}
+In our view, the fundamental challenge you refer to is not one specifically
+posed by the presentation of this manuscript, but instead a more long-standing
+one, related to the mutual understanding of two scientific communities that
+have long remain poorly connected: that of numerical ice sheet modellers on one
+side, and that of glacial geologists on the other.
+
+Because our manuscript is aimed at both communities, it is important for us, as
+a team of co-authors with different backgrounds, that a balance is kept between
+the description of the physics embedded in the numerical model (documented
+elsewhere) and the level of regional detail in the discussion of geological
+evidence (documented elsewhere too, though perhaps in a more fragmented way).
+
+Yet, in doing so, we may have omitted crucial details in the model set-up. We
+hope that our changes described below will correct for that.
 
@@
     focused only on the driving climate sensitivities and have omitted
-    influences of the ice sheet model as well as mass balance model pa-
-    rameters even though they note in that study that these sensitivities
+    influences of the ice sheet model as well as mass balance model
+    parameters even though they note in that study that these sensitivities
     require attention as well.}

### 2015-12-29 reviews/response-jarosch.tex
Resp. Jarosch: announce new outline.

@@
     require attention as well.}
+
+As announced earlier, we have decided to add a new section to the manuscript
+containing a short sensitivity study to some of the most influential rheologic
+and basal sliding parameters. Thus the new outline will be as follow:
+
+\begin{enumerate}
+    \item{Introduction}
+    \item{Model setup}
+    \item{Sensitivity to climate forcing time-series}
+    \item{Sensitivity to ice flow parameters}
+    \item{Comparison with the geologic record}
+\end{enumerate}
+
+In accordance, Sect.~2 (Model setup) will be amended with an description of
+default and alternative parameters for ice rheology and basal sliding with
+illustrations of their role in the model. Sect.~2.2 (Ice thermodynamics) will
+be divided into:
+
+\begin{enumerate}
+    \item[2.2]{Ice rheology}
+    \item[2.3]{Basal sliding}
+    \item[2.4]{Ice shelf calving}
+\end{enumerate}
 

### 2015-12-29 reviews/response-jarosch.tex
Resp. Jarosch: temporarily use bibtex.
I will manually pull in references at the end, but there is a long way to go.

@@
 
-\begin{thebibliography}{69}
-
-\bibitem[{Aschwanden et~al.(2012)}]{Aschwanden.etal.2012}
-Aschwanden, A., Bueler, E., Khroulev, C., and Blatter, H.: An enthalpy
-  formulation for glaciers and ice sheets, J. Glaciol., 58, 441--457,
-  \doi{10.3189/2012JoG11J088}, 2012.
-
-\bibitem[{Clarke et~al.(2013)}]{Clarke.etal.2013}
-Clarke, G. K.~C., Anslow, F.~S., Jarosch, A.~H., Radi{\'{c}}, V., Menounos, B.,
-  Bolch, T., and Berthier, E.: Ice Volume and Subglacial Topography for Western
-  Canadian Glaciers from Mass Balance Fields, Thinning Rates, and a Bed Stress
-  Model, J. Climate, 26, 4282--4303, \doi{10.1175/jcli-d-12-00513.1}, 2013.
-
-\bibitem[{Jarosch et~al.(2010)Jarosch, Anslow, and Clarke}]{Jarosch.etal.2010}
-Jarosch, A.~H., Anslow, F.~S., and Clarke, G. K.~C.: High-resolution
-  precipitation and temperature downscaling for glacier models, Clim. Dynam.,
-  38, 391--409, \doi{10.1007/s00382-010-0949-1}, 2010.
-
-\bibitem[{Seguinot et~al.(2014)Seguinot, Khroulev, Rogozhina, Stroeven, and
-  Zhang}]{Seguinot.etal.2014}
-Seguinot, J., Khroulev, C., Rogozhina, I., Stroeven, A.~P., and Zhang, Q.: The
-  effect of climate forcing on numerical simulations of the {C}ordilleran ice
-  sheet at the {L}ast {G}lacial {M}aximum, The Cryosphere, 8, 1087--1103,
-  \doi{10.5194/tc-8-1087-2014}, 2014.
-
-\bibitem[{Waechter et~al.(2015}]{Waechter.etal.2015}
-Waechter, A., Copland, L., and Herdes, E.: Modern glacier velocities across the
-  Icefield Ranges, St Elias Mountains, and variability at selected glaciers
-  from 1959 to 2012, J. Glaciol., 61, 624--634, \doi{10.3189/2015jog14j147},
-  2015.
-
-\end{thebibliography}
+% FIXME: no bibtex in final response
+\bibliographystyle{copernicus}
+\bibliography{../refs/references}
 

### 2015-12-29 reviews/response-jarosch.tex
Resp. Jarosch: add ice rheology eqns and params.

@@
 
+\usepackage{amsmath}
 \usepackage[T1]{fontenc}
@@
 \renewcommand\familydefault{\sfdefault}
+\renewcommand{\vec}[1]{\ensuremath{\mbox{\boldmath$#1$}}}
 
@@
 
-\todo{%
-    Perform new sensitivity tests on rheological parameters (water content,
-    activation energies, rate factors) and sliding (friction angle). This will
-    take some time. Based on the results, decide if they should be presented in
-    the manuscript (before/after) discussion, in a supplement, or perhaps just
-    here in this response.}
+To keep this sensitivity study as simple as possible, we choose to present only
+four additional runs, two with varying rheologic parameters and two with
+varying sliding parameters, using the simulation driven by the GRIP record as
+a control, at the horizontal resolution of 10\,km.
+
+In all simulations, ice deformation is governed by the constitutive law for ice
+\citep{Glen.1952, Nye.1953},
+
+\begin{equation}
+    \label{eqn:glenslaw}
+    \vec{\dot{\epsilon}} = A\,\tau_e^{n-1}\,\vec{\tau} \,.
+\end{equation}
+
+Where $\vec{\dot{\epsilon}}$ is the the strain-rate tensor, $\vec{\tau}$ the
+deviatoric stress tensor, and $\tau_e$ the equivalent stress defined by
+${\tau_e}^2 = \frac{1}{2} \mathrm{tr}(\vec{\tau}^2)$.
+The ice softness coefficient, $A$, depends on ice temperature, $T$, pressure, $p$, and
+water content, $\omega$, through a piece-wise Arrhenius-type law
+\citep[Eqs.~63--65]{Aschwanden.etal.2012},
+
+\begin{equation}
+    A = E\cdot
+    \begin{cases}
+        A_c \,e^\frac{-Q_c}{RT_{pa}}
+            & \text{if}\ T_{pa} < T_c \,, \\
+        A_w (1+f\omega)\,e^\frac{-Q_w}{RT_{pa}}
+            & \text{if}\ T_{pa} \ge T_c \,,
+    \end{cases}
+\end{equation}
+
+where $T_{pa}$ is the pressure-adjusted ice temperature calculated using the
+Clapeyron relation, ${T_{pa} = T - \beta p}$.
+$R=8.31441$\,J\,mol$^{-1}$\,K$^{-1}$ is the ideal gas constant, and $A_c$,
+$A_w$, $Q_c$ and $Q_w$, are constant parameters corresponding to values
+measured below and above a critical temperature threshold ${T_c=-10}^\circ$C
+\citep[p.~72]{Paterson.Budd.1982,Cuffey.Paterson.2010}. The water fraction,
+$\omega$, is capped at a maximum value of 0.01, above which no measurements
+are available \citep[Eq.~5.7]{Lliboutry.Duval.1985, Greve.1997}. Finally,
+$E$ is a non-dimensional enhancement factor which can take different values,
+$E_{SIA}$, in the Shallow Ice Approximation (SIA) and $E_{SSA}$, in the Shallow
+Shelf Approximation (SSA).
+
+In our short sensitivity study, we set constant the power-law exponent, $n=3$,
+according to \citet[p.~55--57]{Cuffey.Paterson.2010}, the Clapeyron constant,
+$\beta=7.9\times 10^{-8}$\,K\,Pa$^{-1}$, according to \citep{Luthi.etal.2002},
+the water fraction coefficient, $f=181.25$, according to
+\citet{Lliboutry.Duval.1985}, and the SSA enhancement factor, $E_{SSA}=1$,
+according to \citep[p.~77]{Cuffey.Paterson.2010}.
+
+On the other hand, we test different values for the two creep parameters, $A_c$
+and $A_w$, the two activation energies, $Q_c$ and $Q_w$, and the SIA
+enhancement factor, $E_{SIA}$, as follow.
+
+\begin{itemize}
+    \item{Our \emph{default} configuration used in the control run and all other
+          simulations in the manuscript include rheologic parameters, $A_c$,
+          $A_w$, $Q_c$ and $Q_w$, derived from \citet{Paterson.Budd.1982} and
+          given in \citet[Eqn.~5]{Bueler.Brown.2009}, and $E_{SIA}=1$.}
+    \item{Our \emph{hard ice} configuration include rheologic parameters,
+          $A_c$, $A_w$, $Q_c$ and $Q_w$, derived from
+          \citet[p.~72 and 76]{Cuffey.Paterson.2010}, and $E_{SIA}=1$, which
+          correspond to a stiffer rheology than that used in the control run.}
+    \item{Our \emph{soft ice} configuration include rheologic parameters,
+          $A_c$, $A_w$, $Q_c$ and $Q_w$, derived from
+          \citet[p.~72 and 76]{Cuffey.Paterson.2010}, and $E_{SIA}=5$, the
+          recommended value by \citet[p.~77]{Cuffey.Paterson.2010} for ice age
+          polar ice.}
+\end{itemize}
 

### 2015-12-29 reviews/response-jarosch.tex
Resp. Jarosch: add ice rheology plot.

@@
 \end{itemize}
+
+The effect of these three different parametrisations on temperature-dependent
+ice softness, $A$, are illustrated in Fig.~\ref{fig:sens_plot_rheo}.
+
+\begin{figure}
+    \centering
+    \includegraphics{sens_plot_rheo}
+    \caption{Ice softness parameter, $A$, as a function of pressure-adjusted
+             temperature, $T_{pa}$, for the default \citet{Paterson.Budd.1982},
+             hard ice \citet{Cuffey.Paterson.2010}, and soft ice
+             \citet[with $E_{SIA}=5$]{Cuffey.Paterson.2010} rheologies, using
+             a linear scale (top panel) and logarithmic scale (bottom panel).}
+    \label{fig:sens_plot_rheo}
+\end{figure}
 

### 2015-12-29 reviews/response-jarosch.tex
Add common header for responses.
Resp. Jarosch: add rheologic parameters table.

@@
-\documentclass[10pt]{article}
-
-\usepackage{amsmath}
-\usepackage[T1]{fontenc}
-\usepackage[pdftex]{graphicx}
-\usepackage[pdftex]{color}
-\usepackage{hyperref}
-\usepackage{natbib}
-
-\sloppy
-\setlength{\parskip}{1.2ex}
-\setlength{\parindent}{0em}
-\renewcommand\familydefault{\sfdefault}
-\renewcommand{\vec}[1]{\ensuremath{\mbox{\boldmath$#1$}}}
-
-\graphicspath{{../figures/}}
-\definecolor{journalname}{rgb}{0.34,0.59,0.82}
-\definecolor{adgeored}{rgb}{0.439,0.157,0.145}  % from Advances in Geoscience
+% response-jarosch.tex
+% ----------------------------------------------------------------------
+\input{response-header}
 
@@
 
-\newcommand{\doi}[1]{doi:\allowbreak\href{http://dx.doi.org/#1}{#1}}
 \newcommand{\sechead}[1]{\bigskip\noindent\textbf{#1}}
@@
 \newcommand{\msquote}[1]{\begin{quote}\textit{#1}\end{quote}}
-\newcommand{\todo}[1]{\textcolor{adgeored}{TODO: #1}}
 
@@
           \citet[p.~72 and 76]{Cuffey.Paterson.2010}, and $E_{SIA}=5$, the
-          recommended value by \citet[p.~77]{Cuffey.Paterson.2010} for ice age
-          polar ice.}
+          recommended value for ice age polar ice
+          \citep[p.~77]{Cuffey.Paterson.2010}.}
 \end{itemize}
 
-The effect of these three different parametrisations on temperature-dependent
-ice softness, $A$, are illustrated in Fig.~\ref{fig:sens_plot_rheo}.
+Actual parameter values for $A_c$, $A_w$, $Q_c$, $Q_w$ and $E_{SIA}$ used in
+our simulations are given in Table~\ref{tab:rheo_params}, while the effect of
+the three different parametrisations on temperature-dependent ice softness,
+$A$, is illustrated in Fig.~\ref{fig:sens_plot_rheo}.
+
+\begin{table*}
+  \caption{Rheologic parameter values used in the sensitivity test.}
+  \label{tab:rheo_params}
+  \centering
+  {\begin{tabular}{l|ccccc}
+    \tophline
+    Config. & $A_c$ & $A_w$ & $Q_c$ & $Q_w$ & $E_{SIA}$ \\
+            & \multicolumn{2}{c}{(\unit{Pa^{-3}\,s^{-1}})}
+            & \multicolumn{2}{c}{(\unit{J\,mol^{-1}})} & \\
+    \middlehline
+    Default$^1$  & $ 3.61\times 10^{-13}$
+                 & $ 1.73\times 10^3$
+                 & $   60\times 10^3$
+                 & $  139\times 10^3$
+                 & 1 \\
+    Hard ice$^2$ & $2.847\times 10^{-13}$
+                 & $2.356\times 10^{-2}$
+                 & $   60\times 10^3$
+                 & $  115\times 10^3$
+                 & 1 \\
+    Soft ice$^2$ & $2.847\times 10^{-13}$
+                 & $2.356\times 10^{-2}$
+                 & $   60\times 10^3$
+                 & $  115\times 10^3$
+                 & 5 \\
+    \bottomhline
+  \end{tabular}}
+  \belowtable{According to $^1$\citet{Paterson.Budd.1982} and
+              $^2$\citet{Cuffey.Paterson.2010}.}
+\end{table*}
 

### 2015-12-29 reviews/response-jarosch.tex
Resp. Jarosch: already a few changes in rheology.

@@
 and basal sliding parameters. Thus the new outline will be as follow:
-
+%
 \begin{enumerate}
@@
 \end{enumerate}
-
+%
 In accordance, Sect.~2 (Model setup) will be amended with an description of
@@
 be divided into:
-
+%
 \begin{enumerate}
@@
 
-In all simulations, ice deformation is governed by the constitutive law for ice
+In our simulations, ice deformation is governed by the constitutive law for ice
 \citep{Glen.1952, Nye.1953},
-
+%
 \begin{equation}
@@
 \end{equation}
-
-Where $\vec{\dot{\epsilon}}$ is the the strain-rate tensor, $\vec{\tau}$ the
+%
+where $\vec{\dot{\epsilon}}$ is the the strain-rate tensor, $\vec{\tau}$ the
 deviatoric stress tensor, and $\tau_e$ the equivalent stress defined by
@@
 \citep[Eqs.~63--65]{Aschwanden.etal.2012},
-
+%
 \begin{equation}
@@
 \end{equation}
-
+%
 where $T_{pa}$ is the pressure-adjusted ice temperature calculated using the
@@
 \end{itemize}
+
+An additional simulation with rheologic parameters, $A_c$, $A_w$, $Q_c$ and
+$Q_w$, derived from \citet{Cuffey.Paterson.2010}, and $E_{SIA}=2$, the
+recommended value for Holocene polar ice \citep[p.~77]{Cuffey.Paterson.2010}
+was performed, but its results were very similar to that of our default run,
+thus we decided to not present it here.
 

### 2015-12-29 reviews/response-jarosch.tex
Resp. Jarosch: add basal sliding eqns and params.

@@
 
+\clearpage  % moving on to basal sliding
+
+In our simulations, basal sliding is governed by a~pseudo-plastic sliding law,
+already given in the manuscript but recalled here for the sake of completeness,
+%
+\begin{equation}
+    \label{eqn:pseudoplastic}
+    \vec{\tau}_b = -\tau_c \frac{\vec{v}_b}{{v_{th}}^q\,|\vec{v}_b|^{1-q}} \,,
+\end{equation}
+%
+which relates the bed-parallel shear stresses, $\vec{\tau}_b$, to the sliding
+velocity, $\vec{v}_b$. The yield stress, $\tau_c$, is modelled using the
+Mohr--Coulomb criterion,
+%
+\begin{equation}
+   \tau_c = c_0 + N\,\tan{\phi} \,,
+\end{equation}
+%
+where cohesion, $c_0$, is assumed to be zero. Effective pressure, $N$,
+is related to the ice overburden stress, $\rho gh$, and the modelled amount of
+subglacial water, using a formula derived from laboratory experiments with till
+extracted from the base of Ice Stream B in West Antarctica
+\citep[Eqn.~23]{Tulaczyk.etal.2000, Bueler.Pelt.2015},
+\begin{equation}
+    N = \delta \rho gh \, 10^{(e_0/C_c) (1 - (W/W_{max}))} \,,
+\end{equation}
+where $\delta$ sets the minimum ratio between the effective pressure, $N$, and
+the ice overburden stress, $\rho gh$, $e_0$ is a measured reference void ratio
+and $C_c$ is a measured compressibility coefficient. The amount of water at the
+base, $W$, varies from zero to $W_{max}$, a threshold above which additional
+melt water is assumed to drain off instantaneously.
+
+In our sensitivity test, we set constant the pseudo-plastic sliding exponent,
+$q=0.25$, and the threshold velocity, $v_{th}=100$\,\unit{m\,s^{-1}}, according
+to values used by \citet{Aschwanden.etal.2013}, the till cohesion, $c_0=0$,
+whose measured values are consistently negligible
+\citep[p.~268]{Tulaczyk.etal.2000, Cuffey.Paterson.2010}, the till reference
+void ratio, $e_0=0.69$, and the till compressibility coefficient, $C_c=0.12$,
+according to the only measurements available to our knowledge by
+\citep{Tulaczyk.etal.2000}.
+
+We also use a constant spatial distribution of the till friction angle, $\phi$,
+whose values vary from 15 to 45{\degree} as a piecewise-linear function of
+modern bed elevation, with the lowest value occuring below modern sea level
+(0\,m above sea level, m~a.s.l.) and the highest value occuring above the
+generalised elevation of the highest shorelines
+\citep[200\,m~a.s.l.,][Fig.~5]{Clague.1981}. This range of value span over the
+range of measured values for glacial till of 18 to 40{\degree}
+\citep[p.~268]{Cuffey.Paterson.2010}. It accounts for frictional basal
+conditions associated with discontinuous till cover at high elevations, and
+a~weakening of till associated with the presence of marine sediments at low
+elevations (cf. \citealp{Martin.etal.2011};
+\citealp[supplement]{Aschwanden.etal.2013}; \citealp{PISM-authors.2015}).
+
+An additional simulation with a constant till friction angle, $\phi=30\degree$,
+corresponding to the average value in \citet[p.~268]{Cuffey.Paterson.2010}, was
+actually performed, but the induced variability was small as compared to that
+which will be presented here, and therefore we decided to not include this run.
+
+On the other hand, we test different values for the minimum ratio between the
+effective pressure and the overburden stress, $\delta$, and the maximum water
+height in the till, $W_{max}$, as follow.
+
+\begin{itemize}
+    \item{Our \emph{default} configuration used in the control run and all other
+          simulations in the manuscript include $\delta=0.02$ and
+          $W_{max}=2$\,m as in \citet{Bueler.Pelt.2015}.}
+    \item{Our \emph{hard bed} configuration use $\delta=0.01$ and
+          $W_{max}=1$\,m.}
+    \item{Our \emph{soft bed} configuration use $\delta=0.05$ and
+          $W_{max}=5$\,m.}
+\end{itemize}
+
 \referee{%

### 2015-12-29 reviews/response-jarosch.tex
Resp. Jarosch: add effective pressure plot.

@@
           $W_{max}=2$\,m as in \citet{Bueler.Pelt.2015}.}
-    \item{Our \emph{hard bed} configuration use $\delta=0.01$ and
+    \item{Our \emph{hard bed} configuration use $\delta=0.05$ and
+          $W_{max}=5$\,m.}
+    \item{Our \emph{soft bed} configuration use $\delta=0.01$ and
           $W_{max}=1$\,m.}
-    \item{Our \emph{soft bed} configuration use $\delta=0.05$ and
-          $W_{max}=5$\,m.}
 \end{itemize}
+
+These parameter choices are recalled in Table~\ref{tab:ntil_params}, while the
+effect of the three different parametrisations on the effective pressure on
+the till, $N$, in response to water content, $W$, is illustrated in
+Fig.~\ref{fig:sens_plot_ntil}.
+
+\begin{figure}
+    \centering
+    \includegraphics{sens_plot_ntil}
+    \caption{Effective pressure, $N$, as a function of water height at the
+             base, $W$, for the default ($\delta=0.02$, $W_{max}=2$\,m),
+             hard bed ($\delta=0.05$, $W_{max}=5$\,m), and soft bed
+             ($\delta=0.01$, $W_{max}=1$\,m) sliding parametrisations, using a
+             linear scale (top panel) and a logarithmic scale (bottom panel).}
+    \label{fig:sens_plot_ntil}
+\end{figure}
 

### 2015-12-29 reviews/response-jarosch.tex
Resp. Jarosch: add basal sliding parameters table.

@@
 
-These parameter choices are recalled in Table~\ref{tab:ntil_params}, while the
+These parameter choices are recalled in Table~\ref{tab:slip_params}, while the
 effect of the three different parametrisations on the effective pressure on
@@
 Fig.~\ref{fig:sens_plot_ntil}.
+
+\begin{table*}
+  \caption{Basal sliding parameter values used in the sensitivity test.}
+  \label{tab:slip_params}
+  \centering
+  {\begin{tabular}{l|cc}
+    \tophline
+    Config. & $\delta$ & $W_{max}$ \\
+            &          & m \\
+    \middlehline
+    Default$^1$ & 0.02 & 2 \\
+    Hard bed    & 0.05 & 5 \\
+    Soft bed    & 0.01 & 1 \\
+    \bottomhline
+  \end{tabular}}
+  \belowtable{$^1$According to \citet{Bueler.Pelt.2015}.}
+\end{table*}
 
@@
 \end{figure}
+
+\clearpage  % moving on to the results
 

### 2015-12-30 reviews/response-jarosch.tex
Resp. Jarosch: merge the two param tables.

--- 2015-12-29 reviews/response-jarosch.tex
@@
 Actual parameter values for $A_c$, $A_w$, $Q_c$, $Q_w$ and $E_{SIA}$ used in
-our simulations are given in Table~\ref{tab:rheo_params}, while the effect of
+our simulations are given in Table~\ref{tab:sens_params}, while the effect of
 the three different parametrisations on temperature-dependent ice softness,
@@
 \begin{table*}
-  \caption{Rheologic parameter values used in the sensitivity test.}
-  \label{tab:rheo_params}
-  \centering
-  {\begin{tabular}{l|ccccc}
+  \caption{Parameter values used in the sensitivity test.}
+  \label{tab:sens_params}
+  \centering\makebox[\textwidth]
+  {\begin{tabular}{l|ccccc|cc}
     \tophline
-    Config. & $A_c$ & $A_w$ & $Q_c$ & $Q_w$ & $E_{SIA}$ \\
+    Config. & $A_c$ & $A_w$ & $Q_c$ & $Q_w$ & $E_{SIA}$
+            & $\delta$ & $W_{max}$ \\
             & \multicolumn{2}{c}{(\unit{Pa^{-3}\,s^{-1}})}
-            & \multicolumn{2}{c}{(\unit{J\,mol^{-1}})} & \\
+            & \multicolumn{2}{c}{(\unit{J\,mol^{-1}})}
+            & & & (m) \\
     \middlehline
@@
                  & $  139\times 10^3$
-                 & 1 \\
+                 & 1 & 0.02 & 2 \\
+    \middlehline
     Hard ice$^2$ & $2.847\times 10^{-13}$
@@
                  & $  115\times 10^3$
-                 & 1 \\
+                 & 1 & 0.02 & 2 \\
     Soft ice$^2$ & $2.847\times 10^{-13}$
@@
                  & $  115\times 10^3$
-                 & 5 \\
+                 & 5 & 0.02 & 2 \\
+    \middlehline
+    Hard bed     & $ 3.61\times 10^{-13}$
+                 & $ 1.73\times 10^3$
+                 & $   60\times 10^3$
+                 & $  139\times 10^3$
+                 & 1 & 0.05 & 5 \\
+    Soft bed     & $ 3.61\times 10^{-13}$
+                 & $ 1.73\times 10^3$
+                 & $   60\times 10^3$
+                 & $  139\times 10^3$
+                 & 1 & 0.01 & 1 \\
     \bottomhline
   \end{tabular}}
-  \belowtable{According to $^1$\citet{Paterson.Budd.1982} and
-              $^2$\citet{Cuffey.Paterson.2010}.}
+  \belowtable{After $^1$\citet{Paterson.Budd.1982,Bueler.Pelt.2015};
+              and $^2$\citet{Cuffey.Paterson.2010}.}
 \end{table*}
@@
 
-These parameter choices are recalled in Table~\ref{tab:ntil_params}, while the
-effect of the three different parametrisations on the effective pressure on
+The effect of the three different parametrisations on the effective pressure on
 the till, $N$, in response to water content, $W$, is illustrated in
 Fig.~\ref{fig:sens_plot_ntil}.
+All parameter choices are recalled in Table~\ref{tab:sens_params}.
 

### 2015-12-30 reviews/response-jarosch.tex
Resp. Jarosch: add offsets to param table.
Resp. Jarosch: add sensitivity time series plot.

@@
   \centering\makebox[\textwidth]
-  {\begin{tabular}{l|ccccc|cc}
+  {\begin{tabular}{l|ccccc|cc|cc}
     \tophline
+            & \multicolumn{5}{c|}{Rheology}
+            & \multicolumn{2}{c|}{Sliding}
+            & \multicolumn{2}{c}{GRIP scaling} \\
     Config. & $A_c$ & $A_w$ & $Q_c$ & $Q_w$ & $E_{SIA}$
-            & $\delta$ & $W_{max}$ \\
+            & $\delta$ & $W_{max}$ & $f$ & $T_{[32, 22]}$ \\
             & \multicolumn{2}{c}{(\unit{Pa^{-3}\,s^{-1}})}
@@
                  & $  139\times 10^3$
-                 & 1 & 0.02 & 2 \\
+                 & 1 & 0.02 & 2 & 0.38 & 6.2 \\
     \middlehline
@@
                  & $  115\times 10^3$
-                 & 1 & 0.02 & 2 \\
+                 & 1 & 0.02 & 2 & 0.37 & 6.0 \\
     Soft ice$^2$ & $2.847\times 10^{-13}$
@@
                  & $  115\times 10^3$
-                 & 5 & 0.02 & 2 \\
+                 & 5 & 0.02 & 2 & 0.40 & 6.6 \\
     \middlehline
@@
                  & $  139\times 10^3$
-                 & 1 & 0.05 & 5 \\
+                 & 1 & 0.05 & 5 & 0.36 & 5.9 \\
     Soft bed     & $ 3.61\times 10^{-13}$
@@
                  & $  139\times 10^3$
-                 & 1 & 0.01 & 1 \\
+                 & 1 & 0.01 & 1 & 0.40 & 6.5 \\
     \bottomhline
@@
 
+Finally, we adjusted the GRIP linear scaling factor for each run, so that they
+result in a similar glaciated area at the Last Glacial Maximum (LGM) to that
+modelled with the default configuration (Table~\ref{tab:sens_params}).
+
 \begin{figure}
@@
 \clearpage  % moving on to the results
+
+\begin{figure*}
+  \includegraphics{sens_ts}
+  \caption{Modelled sea-level relevant ice volume through the last 120\,ka
+           in the simulation forced by the GRIP paleo-climate record, using
+           default parameters (black curves), different ice rheology parameters,
+           (top panel) and different basal sliding parameters (bottom panel).
+           Gray fields indicate Marine Oxygen Isotope Stage (MIS) boundaries
+           for MIS~2 and MIS~4 according to a global compilation of benthic
+           \chem{\delta^{18}O} records \citep{Lisiecki.Raymo.2005}.}
+  \label{fig:sens_ts}
+\end{figure*}
 

### 2015-12-30 reviews/response-jarosch.tex
Resp. Jarosch: add sensitivity extremes table.

@@
 
+\begin{table*}
+  \caption{Extremes in Cordilleran ice sheet volume and extent corresponding to
+           MIS~4, 3 and 2 using the GRIP paleo-climate forcing with each
+           parameter configuration (Fig.~\ref{fig:sens_ts}). Relative
+           differences (r. diff.) give rough error estimates related to
+           varying selected ice rheology and basal sliding parameters
+           (Table.~\ref{tab:sens_params}).}
+  \label{tab:sens_extrema}
+  \centering\makebox[\textwidth]
+  {\begin{tabular}{l*{3}{|ccc}}
+    \tophline
+             & \multicolumn{3}{c}{Age (ka)}
+             & \multicolumn{3}{c}{Ice extent (\unit{10^6\,km^2})}
+             & \multicolumn{3}{c}{Ice volume (m~s.l.e.)} \\
+    Config.  &  MIS~4 &  MIS~3 &  MIS~2
+             &  MIS~4 &  MIS~3 &  MIS~2
+             &  MIS~4 &  MIS~3 &  MIS~2 \\
+    \middlehline
+    Default  &  57.59 &  42.91 &  19.14
+             &   1.93 &   0.67 &   2.09
+             &   7.43 &   1.54 &   8.62 \\
+    \middlehline
+    Hard ice &  57.32 &  42.90 &  19.14
+             &   1.90 &   0.75 &   2.12
+             &   7.83 &   1.91 &   9.46 \\
+    Soft ice &  58.89 &  49.97 &  21.57
+             &   1.96 &   0.54 &   2.08
+             &   6.58 &   1.03 &   6.88 \\
+    R. diff. &    3\% &   16\% &   13\%
+             &    3\% &   31\% &    2\%
+             &   17\% &   57\% &   30\% \\
+    \middlehline
+    Hard bed &  57.31 &  42.91 &  19.14
+             &   1.93 &   0.96 &   2.13
+             &   7.99 &   2.89 &   9.31 \\
+    Soft bed &  58.90 &  49.21 &  19.53
+             &   1.88 &   0.55 &   2.05
+             &   6.46 &   1.03 &   7.52 \\
+    R. diff. &    3\% &   15\% &    2\%
+             &    3\% &   62\% &    4\%
+             &   21\% &  120\% &   21\% \\
+    \bottomhline
+  \end{tabular}}
+  \belowtable{}
+\end{table*}
+
+\clearpage  % eventually moving on to the next point
+
 \referee{%

### 2015-12-30 reviews/response-jarosch.tex
Resp. Jarosch: reorder sensitivity test labels.
iResp. Jarosch: describe sensitivity test results.

@@
     \middlehline
+    Soft ice$^2$ & $2.847\times 10^{-13}$
+                 & $2.356\times 10^{-2}$
+                 & $   60\times 10^3$
+                 & $  115\times 10^3$
+                 & 5 & 0.02 & 2 & 0.40 & 6.6 \\
     Hard ice$^2$ & $2.847\times 10^{-13}$
@@
                  & 1 & 0.02 & 2 & 0.37 & 6.0 \\
-    Soft ice$^2$ & $2.847\times 10^{-13}$
-                 & $2.356\times 10^{-2}$
+    \middlehline
+    Soft bed     & $ 3.61\times 10^{-13}$
+                 & $ 1.73\times 10^3$
                  & $   60\times 10^3$
-                 & $  115\times 10^3$
-                 & 5 & 0.02 & 2 & 0.40 & 6.6 \\
-    \middlehline
+                 & $  139\times 10^3$
+                 & 1 & 0.01 & 1 & 0.40 & 6.5 \\
     Hard bed     & $ 3.61\times 10^{-13}$
@@
                  & 1 & 0.05 & 5 & 0.36 & 5.9 \\
-    Soft bed     & $ 3.61\times 10^{-13}$
-                 & $ 1.73\times 10^3$
-                 & $   60\times 10^3$
-                 & $  139\times 10^3$
-                 & 1 & 0.01 & 1 & 0.40 & 6.5 \\
     \bottomhline
@@
 whose values vary from 15 to 45{\degree} as a piecewise-linear function of
-modern bed elevation, with the lowest value occuring below modern sea level
-(0\,m above sea level, m~a.s.l.) and the highest value occuring above the
+modern bed elevation, with the lowest value occurring below modern sea level
+(0\,m above sea level, m~a.s.l.) and the highest value occurring above the
 generalised elevation of the highest shorelines
@@
           $W_{max}=2$\,m as in \citet{Bueler.Pelt.2015}.}
+    \item{Our \emph{soft bed} configuration use $\delta=0.01$ and
+          $W_{max}=1$\,m.}
     \item{Our \emph{hard bed} configuration use $\delta=0.05$ and
           $W_{max}=5$\,m.}
-    \item{Our \emph{soft bed} configuration use $\delta=0.01$ and
-          $W_{max}=1$\,m.}
 \end{itemize}
@@
 modelled with the default configuration (Table~\ref{tab:sens_params}).
+In other words, the sensitivity in modelled ice sheet extent at MIS~2 is
+expressed through the scaling factor required to obtain a fixed target area,
+since we consider this scaling factor as a free parameter in our study.
 
@@
 
-\begin{figure*}
-  \includegraphics{sens_ts}
-  \caption{Modelled sea-level relevant ice volume through the last 120\,ka
-           in the simulation forced by the GRIP paleo-climate record, using
-           default parameters (black curves), different ice rheology parameters,
-           (top panel) and different basal sliding parameters (bottom panel).
-           Gray fields indicate Marine Oxygen Isotope Stage (MIS) boundaries
-           for MIS~2 and MIS~4 according to a global compilation of benthic
-           \chem{\delta^{18}O} records \citep{Lisiecki.Raymo.2005}.}
-  \label{fig:sens_ts}
-\end{figure*}
+By analogy to our manuscript's Table~3 (ice volume and extent extrema) and
+Fig.~3 (sea-level relevant ice volume time series), the results of our
+sensitivity study are presented here using a similar layout in
+Table~\ref{tab:sens_extrema} and Fig.~\ref{fig:sens_ts}.
+
+As a result of the different scaling factors applied, the resulting simulations
+show very little difference in the modelled glaciated areas corresponding to
+maximum ice volumes during MIS~2, but also during MIS~4
+(Table~\ref{tab:sens_extrema}). However, this can not be said for the modelled
+glaciated area corresponding to minimum ice volume during MIS~3. In fact, the
+extent of the remnant ice cap which persists over the Skeena Mountains during
+this stage shows a significant sensitivity to ice rheology of 31\%, and
+an even more important sensitivity to basal sliding parameters of 62\%
+(Table~\ref{tab:sens_extrema}).
+
+The modelled sea-level relevant ice volumes show more variability than the
+modelled glaciated areas (Table~\ref{tab:sens_extrema},
+Fig.~\ref{fig:sens_ts}). As could be expected, softer ice and weaker till both
+result in a thinner ice sheet, while harder ice and stronger till result in a
+thicker ice sheet. For instance, peak ice volume during the MIS~2 (LGM) varies
+by 30\% between the two parametrisations of ice rheology used, and by 21\%
+between the two parametrisations of till properties used. The differences in
+sea-level relevant ice volume are greatest during the MIS~3
+(Table~\ref{tab:sens_extrema}, Fig.~\ref{fig:sens_ts}) where both the areal and
+thickness contributions add up.
 
@@
     \middlehline
+    Soft ice &  58.89 &  49.97 &  21.57
+             &   1.96 &   0.54 &   2.08
+             &   6.58 &   1.03 &   6.88 \\
     Hard ice &  57.32 &  42.90 &  19.14
@@
              &   7.83 &   1.91 &   9.46 \\
-    Soft ice &  58.89 &  49.97 &  21.57
-             &   1.96 &   0.54 &   2.08
-             &   6.58 &   1.03 &   6.88 \\
     R. diff. &    3\% &   16\% &   13\%
@@
     \middlehline
+    Soft bed &  58.90 &  49.21 &  19.53
+             &   1.88 &   0.55 &   2.05
+             &   6.46 &   1.03 &   7.52 \\
     Hard bed &  57.31 &  42.91 &  19.14
@@
              &   7.99 &   2.89 &   9.31 \\
-    Soft bed &  58.90 &  49.21 &  19.53
-             &   1.88 &   0.55 &   2.05
-             &   6.46 &   1.03 &   7.52 \\
     R. diff. &    3\% &   15\% &    2\%
@@
 \end{table*}
+
+\begin{figure*}
+  \centering
+  \includegraphics{sens_ts}
+  \caption{Modelled sea-level relevant ice volume through the last 120\,ka
+           in the simulation forced by the GRIP paleo-climate record, using
+           default parameters (black curves), different ice rheology parameters,
+           (top panel) and different basal sliding parameters (bottom panel).
+           Gray fields indicate Marine Oxygen Isotope Stage (MIS) boundaries
+           for MIS~2 and MIS~4 according to a global compilation of benthic
+           \chem{\delta^{18}O} records \citep{Lisiecki.Raymo.2005}.}
+  \label{fig:sens_ts}
+\end{figure*}
 
@@
     et \citet{Seguinot.etal.2014}. NARR is delivered on a 32\,km Lambert grid,
-    and thus it is ques- tionable how ``steep'' these gradients can be, given
+    and thus it is questionable how ``steep'' these gradients can be, given
     the rather smooth representation of the existing topography on a 32\,km

### 2015-12-30 reviews/response-jarosch.tex
Resp. Jarosch: explain ms changes for sens test.

@@
 \begin{equation}
+    \label{eqn:softness}
     A = E\cdot
@@
 
+All the information detailed above, including
+Eqns~\ref{eqn:glenslaw} and~\ref{eqn:softness}, default parameter values,
+Tables~\ref{tab:sens_params} and~\ref{tab:sens_extrema}, and
+Figs.~\ref{fig:sens_plot_rheo}--\ref{fig:sens_ts} will be included in Sect.~2
+(Model setup) or in the new section (Sensitivity to ice flow parameters).
+Relevant discussion points in Sect.~4(Comparison with the geologic record)
+will be updated to account for these new results.
+
 \referee{%

### 2015-12-30 reviews/response-jarosch.tex
Resp. Jarosch: introduce overburden pressure, P_0.

@@
 %
-where cohesion, $c_0$, is assumed to be zero. Effective pressure, $N$,
-is related to the ice overburden stress, $\rho gh$, and the modelled amount of
+where cohesion, $c_0$, is assumed to be zero. Effective pressure, $N$, is
+related to the overburden pressure, $P_0=\rho gh$, and the modelled amount of
 subglacial water, using a formula derived from laboratory experiments with till
@@
 \begin{equation}
-    N = \delta \rho gh \, 10^{(e_0/C_c) (1 - (W/W_{max}))} \,,
+    N = \delta P_0 \, 10^{(e_0/C_c) (1 - (W/W_{max}))} \,,
 \end{equation}
-where $\delta$ sets the minimum ratio between the effective pressure, $N$, and
-the ice overburden stress, $\rho gh$, $e_0$ is a measured reference void ratio
+where $\delta$ sets the minimum ratio between the effective and
+overburden pressures, $e_0$ is a measured reference void ratio
 and $C_c$ is a measured compressibility coefficient. The amount of water at the
@@
 On the other hand, we test different values for the minimum ratio between the
-effective pressure and the overburden stress, $\delta$, and the maximum water
+effective and overburden pressures, $\delta$, and the maximum water
 height in the till, $W_{max}$, as follow.

### 2015-12-30 reviews/response-jarosch.tex
Resp. Jarosch: add cross-refs in fig captions.

@@
              \citet[with $E_{SIA}=5$]{Cuffey.Paterson.2010} rheologies, using
-             a linear scale (top panel) and logarithmic scale (bottom panel).}
+             a linear scale (top panel) and logarithmic scale (bottom panel).
+             Figure made using Eqn.~\ref{eqn:softness} with parameters from
+             Table~\ref{tab:sens_params}.}
     \label{fig:sens_plot_rheo}
@@
 \begin{equation}
+    \label{eqn:ntil}
     N = \delta P_0 \, 10^{(e_0/C_c) (1 - (W/W_{max}))} \,,
@@
              ($\delta=0.01$, $W_{max}=1$\,m) sliding parametrisations, using a
-             linear scale (top panel) and a logarithmic scale (bottom panel).}
+             linear scale (top panel) and a logarithmic scale (bottom panel).
+             Calculations are made for an ice thickness, $h$, of 1000\,m.
+             Figure made using Eqn.~\ref{eqn:ntil} with parameters from
+             Table~\ref{tab:sens_params} after
+             \citet[Fig.~1]{Bueler.Pelt.2015}.}
     \label{fig:sens_plot_ntil}

### 2015-12-30 reviews/response-jarosch.tex
Resp. Jarosch: address second general comment.

@@
 
-\todo{%
-    I have no intention to remove half the manuscript. Argue against it here.
-    Geomorphology is no exact science and has to be speculative.}
+Our choice of language P4162 L1 and 7 (discussion of a the longitudinal shift
+between the modelled and geological reconstructed eastern margin and, in turn,
+the position of the main ice divide), P4164 L18 (discussion of the confined
+modelled ice extent in agreement with the regional geomorphology), and P4171
+L7, L9, L17, L19 and L20 (discussion of the modelled flow patterns and the
+observed lineation pattern on Interior Plateau of British Columbia) is
+intentional. So is our choice to discuss some aspects of the model results that
+are not unequivocal, as a result of uncertain model parameters (as partly
+addressed by our new sensitivity study) or, more often, due to processes not
+included in the model.
+
+Our rationale for including this ``extensive speculation'' takes into account
+the fact that glacial geology is no exact science. In fact, it often has to be
+speculative and to include a large part of interpretation in order to reach
+conclusions that are often uncertain but without which the field could not move
+forward. Because we expect that part of our readership will consist of glacial
+geologists who have sometimes spent an entire scientific career into studying
+some aspects of the Cordilleran ice sheet (cf. comment by A.~Stumpf in this
+discussion), we chose to address some of the long-standing debates on
+Cordilleran ice sheet glacial geology, such as the location of ice divides or
+the enigmatic lineation pattern on the Interior Plateau.
+
+This is not only to prevent over-interpretation of our model results by
+researchers with little or no background in ice sheet modelling, but also to
+hint at potential model improvements that may be needed to approach these
+debates through numerical modelling. Thus we do not plan to shorten section~4.
 

### 2015-12-30 cordillera-cycle.tex
Correct a hidden relict.
Fix typos found by Jarosch.

@@
-% cordillera-climate.tex
+% cordillera-cycle.tex
 % =============================================================================
@@
 simulations driven by those records. These differences are not subject to
-further analysis further; instead we focus on simulations forced by
+further analysis here; instead we focus on simulations forced by
 palaeo-temperature records from the GRIP and EPICA ice cores that appear to
@@
 troughs). We find evidence for this on the slopes of the Coast Mountains, the
-the Columbia and Rocky mountains, the Wrangell and Saint Elias mountains, and
+Columbia and Rocky mountains, the Wrangell and Saint Elias mountains, and
 radiating off the Skeena Mountains
@@
 topography and regional climate resulted in a modelled ice divide of the LGM
-ice sheet beingtoo far to the east (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
+ice sheet being too far to the east (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
 \citealp{Seguinot.etal.2014}). A more westerly-located LGM ice divide would

### 2015-12-30 reviews/response-jarosch.tex
Fix typos found by Jarosch.
Resp. Jarosch: polish general comments response.
I believe this part is good to go.

--- 2015-12-30 reviews/response-jarosch.tex
@@
 
-We apologize for our delayed response and thank you for this detailed review of
-our manuscript. Following your comments, we decide to amend the manuscript with
+We thank you for this detailed review of our manuscript and apologize for our
+delayed response, due to the fact that I moved to a different institution and
+had to set-up the sensitivity study presented hereafter on a different
+supercomputer than the one used for the first version of the manuscript, using
+a new allocation to which I have access only since October.
+
+Following your comments, we decide to amend the manuscript with
 a new section that explores the sensitivity of one of our runs to some of the
@@
 speculative, we prefer not to shorten it for reasons we will detail below.
-Although you may feel we have not applied all the changes you requested, we
-hope that those we do plan fill much of the voids you pointed at.
+We hope that the changes described below address your comments.
 
@@
 evidence (documented elsewhere too, though perhaps in a more fragmented way).
+We would be grateful if you can keep this point in mind when reading on.
 
-Yet, in doing so, we may have omitted crucial details in the model set-up. We
-hope that our changes described below will correct for that.
+Yet, in trying to keep this balance, we admit that we may have omitted crucial
+details in the model set-up. We will try to correct for that in the present
+response and in the new version of the manuscript.
 
@@
 illustrations of their role in the model. Sect.~2.2 (Ice thermodynamics) will
-be divided into:
+be expanded and divided into:
 %
@@
 
-In our short sensitivity study, we set constant the power-law exponent, $n=3$,
+In our sensitivity study, we set constant the power-law exponent, $n=3$,
 according to \citet[p.~55--57]{Cuffey.Paterson.2010}, the Clapeyron constant,
@@
           correspond to a stiffer rheology than that used in the control run.}
-    \item{Our \emph{soft ice} configuration include rheologic parameters,
-          $A_c$, $A_w$, $Q_c$ and $Q_w$, derived from
-          \citet[p.~72 and 76]{Cuffey.Paterson.2010}, and $E_{SIA}=5$, the
+    \item{Our \emph{soft ice} configuration include rheologic parameters
+          from \citet{Cuffey.Paterson.2010}, and $E_{SIA}=5$, the
           recommended value for ice age polar ice
@@
 
-An additional simulation with rheologic parameters, $A_c$, $A_w$, $Q_c$ and
-$Q_w$, derived from \citet{Cuffey.Paterson.2010}, and $E_{SIA}=2$, the
+An additional simulation using the ice rheology from
+\citet{Cuffey.Paterson.2010} and $E_{SIA}=2$, the
 recommended value for Holocene polar ice \citep[p.~77]{Cuffey.Paterson.2010}
@@
              temperature, $T_{pa}$, for the default \citet{Paterson.Budd.1982},
-             hard ice \citet{Cuffey.Paterson.2010}, and soft ice
+             hard ice \citet[with $E_{SIA}=5$]{Cuffey.Paterson.2010}, and soft ice
              \citet[with $E_{SIA}=5$]{Cuffey.Paterson.2010} rheologies, using
@@
 void ratio, $e_0=0.69$, and the till compressibility coefficient, $C_c=0.12$,
-according to the only measurements available to our knowledge by
+according to the only measurements available to our knowledge, published by
 \citep{Tulaczyk.etal.2000}.
@@
 actually performed, but the induced variability was small as compared to that
-which will be presented here, and therefore we decided to not include this run.
+which will be presented here, and therefore we decided to not include this run
+in our sensitivity study.
 
@@
     \includegraphics{sens_plot_ntil}
-    \caption{Effective pressure, $N$, as a function of water height at the
-             base, $W$, for the default ($\delta=0.02$, $W_{max}=2$\,m),
+    \caption{Effective pressure, $N$, as a function of water content in the
+             till, $W$, for the default ($\delta=0.02$, $W_{max}=2$\,m),
              hard bed ($\delta=0.05$, $W_{max}=5$\,m), and soft bed
@@
 maximum ice volumes during MIS~2, but also during MIS~4
-(Table~\ref{tab:sens_extrema}). However, this can not be said for the modelled
+(Table~\ref{tab:sens_extrema}). However, such can not be said of the modelled
 glaciated area corresponding to minimum ice volume during MIS~3. In fact, the
@@
 modelled glaciated areas (Table~\ref{tab:sens_extrema},
-Fig.~\ref{fig:sens_ts}). As could be expected, softer ice and weaker till both
+Fig.~\ref{fig:sens_ts}). As one could expect, softer ice and weaker till both
 result in a thinner ice sheet, while harder ice and stronger till result in a
@@
 by 30\% between the two parametrisations of ice rheology used, and by 21\%
-between the two parametrisations of till properties used. The differences in
+between the two parametrisations of basal sliding used. The differences in
 sea-level relevant ice volume are greatest during the MIS~3
@@
 (Model setup) or in the new section (Sensitivity to ice flow parameters).
-Relevant discussion points in Sect.~4(Comparison with the geologic record)
-will be updated to account for these new results.
+Relevant discussion points in Sect.~4 (Comparison with the geologic record)
+will also be updated to account for these new results.
 
@@
 
-\todo{%
-    I have no intention to remove half the manuscript. Argue against it here.
-    Geomorphology is no exact science and has to be speculative.}
+Our choice of language P4162 L1 and 7 (discussion of a the longitudinal shift
+between the modelled and geological reconstructed eastern margin and, in turn,
+the position of the main ice divide), P4164 L18 (discussion of the confined
+modelled ice extent on the Alaska Range north slope in agreement with the
+regional reconstructions), and P4171
+L7, L9, L17, L19 and L20 (discussion of the modelled flow patterns and the
+observed lineation pattern on the Interior Plateau of British Columbia) is
+intentional. So is our choice to discuss some aspects of the model results that
+are not unequivocal, as a result of uncertain model parameters or, more often,
+due to processes not included in the model.
+
+Our rationale for including this ``extensive speculation'' takes into account
+the fact that glacial geology is no exact science. In fact, it often has to be
+speculative and to include a large part of interpretation in order to reach
+conclusions that are often uncertain but without which the field could not move
+forward. Because we expect that part of our readership will consist of glacial
+geologists who have sometimes spent an entire scientific career into studying
+some aspects of the Cordilleran ice sheet (cf. comment by A.~Stumpf in this
+discussion), we chose to address some of the long-standing debates on
+Cordilleran ice sheet glacial geology, such as the location of ice divides or
+the enigmatic lineation pattern on the Interior Plateau.
+
+This is not only to prevent an over-interpretation of our model results by
+readers with little or no background in ice sheet modelling, but also to
+hint at potential model improvements that may be needed to approach these
+debates through numerical modelling in the future. We believe that such
+discussion, though speculative, is pertinent, and thus do not plan to shorten
+section~4.
 
@@
 
-\todo{%
-    Correct.}
-
 \referee{%
@@
 
-\todo{%
-    Correct.}
+Thank you for spotting these two typos. We have corrected them.
 

### 2015-12-31 reviews/response-jarosch.tex
Resp. Jarosch: add quick downscaling first shot.

@@
 
-\todo{%
-    Decide what to do!}
+We acknowledge that extending the downscaling method presented by
+\citet{Jarosch.etal.2010} to the entire modelling domain and using the
+downscaled temperature and precipitation fields as input to the model would be
+a significant and much needed improvement to our study. However, this would
+also be (for us) an entirely new work package which we can not complete during
+the time frame of this peer review. What we can and will do is acknowledge the
+limitations implied by not including temperature and precipitation downscaling,
+and refer to the findings of \citet{Jarosch.etal.2010}. We will also rework
+parts of the interpretations in the light of your comment, for instance
+regarding the excess ice on an beyond the Columbia and Rocky mountains in our
+runs.
 

### 2015-12-31 reviews/response-jarosch.tex
Resp. Jarosch: discuss ETOPO1 first shot.

@@
 
-\todo{%
-    Mention that ice cover could be overestimated in this region. The main
-    problematic area are the St.-Elias Mountains. However, there is no ice
-    thickness data available as far as I know. Check whether there is more info
-    about that in \citet{Waechter.etal.2015}}
+By ``derived'' we meant that the ETOPO1 data is simply re-sampled with linear
+interpolation from the original to the projected grid. We will clarify this in
+the methods.
+
+Thank you for pointing out the study by \citet{Clarke.etal.2013}.
+Indeed, the ETOPO1 data contains present day ice volumes. The most problematic
+part of the model domain in this respect is by far that of the Wrangell and
+St.-Elias mountains where ice thicknesses up to 1200\,m have recently been
+measured by low-frequency radar \citep{Rignot.etal.2013}. In this area, located
+over the USA Canada border and just north of 60{\degree}N and thus not included
+in the study by \citet{Clarke.etal.2013}, temperature ice, surge dynamics
+and deep subglacial depressions in the icefield interior pose a fundamental
+challenge to reconstructing basal topography for the entire ice cap. Although
+it is clear that our model overestimate ice thickness in this region, we are
+not aware of bed topography data or reconstructions that could be used to force
+the ice sheet model. This said, with the exception of the Wrangell and
+St.-Elias mountains ice field, present-day ice volumes are small relative to
+the ice volumes concerned in our study.
 

### 2015-12-31 reviews/response-jarosch.tex
Resp. Jarosch: adress several minor comments.

@@
 
-\todo{%
-    Agree. Reformulate.}
+We agree and will reformulate. Thank you for spotting this.
 
@@
 
-\todo{%
-    Pull values from parameter table in PhD kappa. Wait for results of
-    sensitivity tests.}
+Our changes described above partly address this comment. To make our parameter
+choices clear, we also plan to include all default parameters values mentioned
+in this response in the manuscript's Table~1 (alternative values used in the
+sensitivity tests being given in a separate table)
 
@@
 
-\todo{%
-    From the ice-bedrock interface downward. Clarify.}
+The ``depth of 3\,km'' is measured the ice-bedrock interface downward. This is
+where we apply geothermal heat flux as the lower boundary condition to the
+bedrock thermal model. By ``computed subglacially'' we mean below the
+ice-bedrock interface, into the bedrock. We will clarify these two points in
+the text.
 
@@
 
-\todo{%
-    Pull parameters and explanation from Kappa. Check for newer PISM refs on
-    modelling Greenland or Antarctica. Wait for results from sensitivity tests}
+Our changes described above partly address this comment. We also plan to add
+add a new column to Table~1 containing references motivating default parameter
+choices.
 

### 2015-12-31 reviews/response-jarosch.tex
Resp. Jarosch: respond to NARR interpolation.

@@
 regarding the excess ice on an beyond the Columbia and Rocky mountains in our
-runs.
+runs. We have previously shown that forcing PISM with bilinearly-interpolated
+climate fields from the NARR without temperature and precipitation downscaling
+resulted in realistic ice margins at first order relative to the geological
+record, with limitations that have been identified \citep{Seguinot.etal.2014}.
 
@@
     computational grids of the current study. I disagree with the notion that a
-    32\,km precipitation field can be called ``high- resolution'' in the
+    32\,km precipitation field can be called ``high-resolution'' in the
     context of 10 and 5\,km grid based ice sheet modelling. The input data is
@@
 
-\todo{%
-    NARR data fields have been bilinearly interpolated. Decide what to do!}
+As in \citet{Seguinot.etal.2014}, the NARR data fields have been bilinearly
+interpolated from the NARR 32\,km Lambert grid to the computational domain
+grids. We apologize for omitting this information and will add it to the
+manuscript.
+
+We called the NARR data ``high resolution'' because its spatial resolution is
+higher than most other atmospheric reanalyses. But we agree that it is still
+too coarse for ice flow modelling and that temperature and precipitation
+downscaling techniques can correct for errors caused by resolution gap. We will
+rework this sentence to make it clear in manuscript.
 

### 2015-12-31 reviews/response-jarosch.tex
Resp. Jarosch: first shot on lapse rate.

@@
 
-\todo{%
-    I am not sure how to adress this at the moment. Temperature lapse rates
-    vary significantly at present, but I am not sure if there is much knowledge
-    on how they varied in the past.}
+A fixed temperature lapse rate, $\gamma=6{\degree}C$, is certainly another
+coarse approximation in our model setup. Although it is clear that $\gamma$
+varies spatially, seasonally, and varied in the past with glacial fluctuations,
+including these variations in the model setup would require to introduce new
+degrees of freedom in the study, which we have been trying to avoid.
 

### 2015-12-31 reviews/response-jarosch.tex
Resp. Jarosch: apply corrections from Irina (1).

@@
 
-We thank you for this detailed review of our manuscript and apologize for our
-delayed response, due to the fact that I moved to a different institution and
-had to set-up the sensitivity study presented hereafter on a different
-supercomputer than the one used for the first version of the manuscript, using
-a new allocation to which I have access only since October.
+We thank you for this detailed review of our manuscript.
 
-Following your comments, we decide to amend the manuscript with
-a new section that explores the sensitivity of one of our runs to some of the
+Following your comments, we have decided to amend the manuscript with a new
+section that explores the sensitivity of our preferred run to some of the
 parameters governing ice deformation and sliding. Although the discussion of
 model results against geological evidence is indeed extensive and partly
-speculative, we prefer not to shorten it for reasons we will detail below.
+speculative, we prefer not to shorten it for the reasons we detail below.
 We hope that the changes described below address your comments.
@@
 
-In our view, the fundamental challenge you refer to is not one specifically
-posed by the presentation of this manuscript, but instead a more long-standing
-one, related to the mutual understanding of two scientific communities that
-have long remain poorly connected: that of numerical ice sheet modellers on one
-side, and that of glacial geologists on the other.
+The aim of this manuscript, as we see it, is to create a bridge between two
+scientific communities, which have long remained largely disconnected, by
+presenting our work in such a way that both communities will understand it. We
+refer here to the community of numerical ice sheet modellers on the one side,
+and that of glacial geologists on the other. Thus from our viewpoint as
+co-authors, the fundamental challenge you refer to is to communicate to each
+community using their own slang and their methods to describe the results.
 
@@
 a team of co-authors with different backgrounds, that a balance is kept between
-the description of the physics embedded in the numerical model (documented
-elsewhere) and the level of regional detail in the discussion of geological
-evidence (documented elsewhere too, though perhaps in a more fragmented way).
-We would be grateful if you can keep this point in mind when reading on.
+the description of the physics embedded in the numerical model, documented in
+detail in many publications by PISM developers and users elsewhere
+\citep[e.g.,][]{Bueler.Brown.2009, Winkelmann.etal.2011, Martin.etal.2011,
+Aschwanden.etal.2013, Seguinot.2014, Bueler.Pelt.2015}, and the level of
+regional detail in the discussion of geological evidence, documented elsewhere
+as well though in a more fragmented way. We would be grateful if this point
+could be taken into account when reading on.
 
-Yet, in trying to keep this balance, we admit that we may have omitted crucial
-details in the model set-up. We will try to correct for that in the present
-response and in the new version of the manuscript.
+Yet, while trying to keep this balance, we may have indeed omitted crucial
+details related to the model set-up. We have made every attempt to correct this
+in the new version of the manuscript.
 
@@
 As announced earlier, we have decided to add a new section to the manuscript
-containing a short sensitivity study to some of the most influential rheologic
-and basal sliding parameters. Thus the new outline will be as follow:
+aiming to assess the sensitivity of the model to some of the most influential
+ice flow parameters. Thus the new outline is now as follows:
 %
@@
 %
-In accordance, Sect.~2 (Model setup) will be amended with an description of
+In accordance, Sect.~2 (Model setup) will be amended with a description of
 default and alternative parameters for ice rheology and basal sliding with
 illustrations of their role in the model. Sect.~2.2 (Ice thermodynamics) will
-be expanded and divided into:
+has been expanded and divided into:
 %
@@
 To keep this sensitivity study as simple as possible, we choose to present only
-four additional runs, two with varying rheologic parameters and two with
+four additional runs, two with varying rheological parameters and two with
 varying sliding parameters, using the simulation driven by the GRIP record as
@@
     \item{Our \emph{default} configuration used in the control run and all other
-          simulations in the manuscript include rheologic parameters, $A_c$,
+          simulations in the manuscript include rheological parameters, $A_c$,
           $A_w$, $Q_c$ and $Q_w$, derived from \citet{Paterson.Budd.1982} and
           given in \citet[Eqn.~5]{Bueler.Brown.2009}, and $E_{SIA}=1$.}
-    \item{Our \emph{hard ice} configuration include rheologic parameters,
+    \item{Our \emph{hard ice} configuration include rheological parameters,
           $A_c$, $A_w$, $Q_c$ and $Q_w$, derived from
@@
           correspond to a stiffer rheology than that used in the control run.}
-    \item{Our \emph{soft ice} configuration include rheologic parameters
+    \item{Our \emph{soft ice} configuration include rheological parameters
           from \citet{Cuffey.Paterson.2010}, and $E_{SIA}=5$, the
@@
 
-Our choice of language P4162 L1 and 7 (discussion of a the longitudinal shift
+Our choice of language on pages P4162 L1 and 7 (discussion of a longitudinal shift
 between the modelled and geological reconstructed eastern margin and, in turn,
@@
 forward. Because we expect that part of our readership will consist of glacial
-geologists who have sometimes spent an entire scientific career into studying
+geologists who have sometimes spent an entire scientific career on studying
 some aspects of the Cordilleran ice sheet (cf. comment by A.~Stumpf in this
 discussion), we chose to address some of the long-standing debates on
-Cordilleran ice sheet glacial geology, such as the location of ice divides or
+Cordilleran glacial geology, such as the location of ice divides or
 the enigmatic lineation pattern on the Interior Plateau.

### 2015-12-31 reviews/response-stumpf.tex
Add common header for responses.
Begin to work on the response to A. Stumpf.

@@
-\documentclass[10pt]{article}
-
-\usepackage[T1]{fontenc}
-\usepackage[pdftex]{graphicx}
-\usepackage[pdftex]{color}
-\usepackage{hyperref}
-\usepackage{natbib}
-
-\sloppy
-\setlength{\parskip}{1.2ex}
-\setlength{\parindent}{0em}
-\renewcommand\familydefault{\sfdefault}
-
-\graphicspath{{../figures/}}
-\definecolor{journalname}{rgb}{0.34,0.59,0.82}
-\definecolor{adgeored}{rgb}{0.439,0.157,0.145}  % from Advances in Geoscience
+% response-stumpf.tex
+% ----------------------------------------------------------------------
+\input{response-header}
 
@@
 \newcommand{\msquote}[1]{\begin{quote}\textit{#1}\end{quote}}
-\newcommand{\todo}[1]{\textcolor{adgeored}{TODO: #1}}
 
 To A.~Stumpf,
+
+Thank you very much for taking the time to read and comment on our manuscript.
+We are very thankful to receive feedback from the glacial geology community,
+more particularly so here on The Cryosphere Discussion which traditionally
+has a stronger presence in quantitative glaciology.
 
@@
     or inconclusive.}
+
+Thank you very much for this positive summary of our work and for your
+appreciative words!
 

### 2015-12-31 reviews/response-stumpf.tex
Resp. Stumpf: rework first sentence.

@@
 
-\todo{%
-    Rephrase, and perhaps add a few references from \citet{Stumpf.etal.2014},
-    although I feel I already have many.}
+By no means our opening sentence was meant to undervalue the large amount of
+geological work performed on the Cordilleran ice sheet, much of which has been
+reviewed and discussed against our model results in later parts of the
+manuscript. Despite all the work done, our impression is that the Cordilleran
+ice sheet, due to its complexity, remains less understood than its Laurentide
+and Eurasian counterparts, especially when it comes to reconstructing the
+dynamics of the different phases of advance and retreat of the ice sheet
+through the last glacial cycle. This is the impression that we tried to convey
+in the first sentence of the abstract. To avoid any further misunderstanding,
+we replaced the opening two sentences with:
+
+\msquote{%
+    After more than a century of geological research, the Cordilleran ice sheet
+    of North America remains the least understood among Pleistocene ice sheets
+    in terms of its former extent, volume and dynamics. Because of the
+    irregular topography on which the ice sheet formed, geological studies have
+    often had only local or regional relevance, and evidenced such a complexity
+    that whole ice-sheet reconstructions of advance and retreat patterns are
+    lacking.}
 

### 2015-12-31 reviews/response-stumpf.tex
Resp. Stumpf: geographic info and Ferbey dataset.

@@
 
-\todo{%
-    Add info where it will not disturb the figures. Or perhaps a larger
-    location map like in the Kappa?}
+Graticules, rivers and lakes were added to all maps, the latter two as a
+background to model result so that they do not interfere with them. Because we
+feel that more geographic information would disturb the visibility of the main
+figure contents, we have also reworked and enlarged Fig.~1 (location map) to
+replace abbreviations by full geographic names, which should also help the
+reader to follow discussions in the text.
 
@@
 
-\todo{%
-    Nice dataset. Not quantitatively tested, if that is what the comment means.
-    In my understanding, there is very little reliable info about ice sheet
-    thickness, ice divide positions are reconstructions that always involve
-    some level of interpretation. Comparing flow direction against
-    \citet{Ferbey.etal.2013} would be very nice but I don't think we are this
-    confident in the basal velocities for such a comparison yet.}
+The model was tested qualitatively against an extensive body of glacial
+geology literature, which is the topic of Sect.~4 (Comparison to the geologic
+record). The model was not tested quantitatively against geologic evidence,
+because we did not know of any publicly available dataset covering a
+significant part of the model domain until reading your comment.
+
+Therefore, we thank you very much for pointing to us the recent map and dataset
+by \citet{Ferbey.etal.2013}. Indeed, it would make much sense to compare this
+database against modelled basal velocities in the future. However, we feel that
+such a quantitative comparison is premature at this stage. In fact, modelled
+basal velocities are sensitive to hydrological properties of the subglacial
+till (cf. our response to A.~Jarosch in this discussion), and even more to
+thermal conditions at the base, which in turn are largely influenced by
+geothermal heat flow, kept constant in this study.
 

### 2016-01-03 reviews/response-stumpf.tex
Resp. Stumpf: address the last two comments.
Resp. Stumpf: apply corrections from Arjen.

@@
 
-Thank you very much for taking the time to read and comment on our manuscript.
+Thank you very much for your constructive review.
 We are very thankful to receive feedback from the glacial geology community,
@@
     After more than a century of geological research, the Cordilleran ice sheet
-    of North America remains the least understood among Pleistocene ice sheets
-    in terms of its former extent, volume and dynamics. Because of the
-    irregular topography on which the ice sheet formed, geological studies have
-    often had only local or regional relevance, and evidenced such a complexity
-    that whole ice-sheet reconstructions of advance and retreat patterns are
-    lacking.}
+    of North America remains among the least understood in terms of its former
+    extent, volume, and dynamics. Because of the mountainous topography on
+    which the ice sheet formed, geological studies have often had only local or
+    regional relevance, and shown such a complexity that ice sheet-wide spatial
+    reconstructions of advance and retreat patterns are lacking.}
 
@@
 
-\todo{%
-    Add reference to \citet{Stumpf.etal.2000}...}
+Thank your for pointing this out. We have added a reference to
+\citet{Stumpf.etal.2000} in the discussion of the erosional imprint on the
+landscape (Sect.~4.2.3).
 
@@
 
-\todo{%
-    Add reference to \citet{Stumpf.etal.2000}...}
+Thank you. We have added a reference to \citet{Stumpf.etal.2000} in the
+discussion of deglacial flow directions (Sect.~4.3.3).
 

### 2016-01-03 reviews/response-jarosch.tex
Resp. Jarosch: apply corrections from Arjen.

@@
 speculative, we prefer not to shorten it for the reasons we detail below.
-We hope that the changes described below address your comments.
+We hope that the changes described below address your concerns.
 
@@
 and that of glacial geologists on the other. Thus from our viewpoint as
-co-authors, the fundamental challenge you refer to is to communicate to each
-community using their own slang and their methods to describe the results.
+authors, the fundamental challenge you refer to is to communicate to each
+community using their own terminology and methods to describe the results.
 
@@
 aiming to assess the sensitivity of the model to some of the most influential
-ice flow parameters. Thus the new outline is now as follows:
+ice flow parameters. Thus the new outline now includes a ``Sensitivity to ice
+flow parameters'' section and becomes:
 %
@@
 %
-In accordance, Sect.~2 (Model setup) will be amended with a description of
-default and alternative parameters for ice rheology and basal sliding with
-illustrations of their role in the model. Sect.~2.2 (Ice thermodynamics) will
-has been expanded and divided into:
+In accordance, Sect.~2 (Model setup) has been amended with a description of
+default and alternative parameters for ice rheology and basal sliding,
+including illustrations of their role in the model. Sect.~2.2 (Ice
+thermodynamics) has therefore been expanded and divided into:
 %
@@
 \citet{Lliboutry.Duval.1985}, and the SSA enhancement factor, $E_{SSA}=1$,
-according to \citep[p.~77]{Cuffey.Paterson.2010}.
+according to \citet[p.~77]{Cuffey.Paterson.2010}.
 
@@
     \caption{Ice softness parameter, $A$, as a function of pressure-adjusted
-             temperature, $T_{pa}$, for the default \citet{Paterson.Budd.1982},
-             hard ice \citet[with $E_{SIA}=5$]{Cuffey.Paterson.2010}, and soft ice
-             \citet[with $E_{SIA}=5$]{Cuffey.Paterson.2010} rheologies, using
+             temperature, $T_{pa}$, for the default \citep{Paterson.Budd.1982},
+             hard ice \citep[with $E_{SIA}=1$]{Cuffey.Paterson.2010}, and soft ice
+             \citep[with $E_{SIA}=5$]{Cuffey.Paterson.2010} rheologies, using
              a linear scale (top panel) and logarithmic scale (bottom panel).
@@
 In our sensitivity test, we set constant the pseudo-plastic sliding exponent,
-$q=0.25$, and the threshold velocity, $v_{th}=100$\,\unit{m\,s^{-1}}, according
+$q=0.25$, and the threshold velocity, $v_{th}=100$\,\unit{m\,a^{-1}}, according
 to values used by \citet{Aschwanden.etal.2013}, the till cohesion, $c_0=0$,
@@
 generalised elevation of the highest shorelines
-\citep[200\,m~a.s.l.,][Fig.~5]{Clague.1981}. This range of value span over the
+\citep[200\,m~a.s.l.,][Fig.~5]{Clague.1981}. This range of values span over the
 range of measured values for glacial till of 18 to 40{\degree}
@@
 Fig.~\ref{fig:sens_plot_ntil}.
-All parameter choices are recalled in Table~\ref{tab:sens_params}.
+All parameter choices are listed in Table~\ref{tab:sens_params}.
 
@@
            parameter configuration (Fig.~\ref{fig:sens_ts}). Relative
-           differences (r. diff.) give rough error estimates related to
+           differences (R. diff.) give rough error estimates related to
            varying selected ice rheology and basal sliding parameters
-           (Table.~\ref{tab:sens_params}).}
+           (Table~\ref{tab:sens_params}).}
   \label{tab:sens_extrema}
@@
            in the simulation forced by the GRIP paleo-climate record, using
-           default parameters (black curves), different ice rheology parameters,
-           (top panel) and different basal sliding parameters (bottom panel).
+           default parameters (black curves), different ice rheology parameters
+           (top panel), and different basal sliding parameters (bottom panel).
            Gray fields indicate Marine Oxygen Isotope Stage (MIS) boundaries
@@
 Tables~\ref{tab:sens_params} and~\ref{tab:sens_extrema}, and
-Figs.~\ref{fig:sens_plot_rheo}--\ref{fig:sens_ts} will be included in Sect.~2
-(Model setup) or in the new section (Sensitivity to ice flow parameters).
-Relevant discussion points in Sect.~4 (Comparison with the geologic record)
-will also be updated to account for these new results.
+Figs.~\ref{fig:sens_plot_rheo}--\ref{fig:sens_ts} have been included in Sect.~2
+(Model setup) and in the new section (Sensitivity to ice flow parameters).
+Relevant discussion points in Sect.~5 (Comparison with the geologic record)
+have been revised to account for these new results.
 
@@
 debates through numerical modelling in the future. We believe that such
-discussion, though speculative, is pertinent, and thus do not plan to shorten
-section~4.
+discussion, though speculative, is pertinent, and, consequently, have not
+shortened this section.
 
@@
 also be (for us) an entirely new work package which we can not complete during
-the time frame of this peer review. What we can and will do is acknowledge the
+the time frame of this peer review. What we have done is to acknowledge the
 limitations implied by not including temperature and precipitation downscaling,
-and refer to the findings of \citet{Jarosch.etal.2010}. We will also rework
+and refer to the findings of \citet{Jarosch.etal.2010}. We have also modified
 parts of the interpretations in the light of your comment, for instance
-regarding the excess ice on an beyond the Columbia and Rocky mountains in our
+regarding the excess ice on and beyond the Columbia and Rocky mountains in our
 runs. We have previously shown that forcing PISM with bilinearly-interpolated
@@
 By ``derived'' we meant that the ETOPO1 data is simply re-sampled with linear
-interpolation from the original to the projected grid. We will clarify this in
+interpolation from the original to the projected grid. We have clarified this in
 the methods.
@@
 over the USA Canada border and just north of 60{\degree}N and thus not included
-in the study by \citet{Clarke.etal.2013}, temperature ice, surge dynamics
+in the study by \citet{Clarke.etal.2013}, temperate ice, surge dynamics
 and deep subglacial depressions in the icefield interior pose a fundamental
 challenge to reconstructing basal topography for the entire ice cap. Although
-it is clear that our model overestimate ice thickness in this region, we are
+it is clear that our model overestimates ice thickness in this region, we are
 not aware of bed topography data or reconstructions that could be used to force
@@
 
-We agree and will reformulate. Thank you for spotting this.
+We agree and have reformulated this statement. Thank you for spotting this.
 
@@
 Our changes described above partly address this comment. To make our parameter
-choices clear, we also plan to include all default parameters values mentioned
+choices clear, we have included all default parameters values mentioned
 in this response in the manuscript's Table~1 (alternative values used in the
@@
 bedrock thermal model. By ``computed subglacially'' we mean below the
-ice-bedrock interface, into the bedrock. We will clarify these two points in
+ice-bedrock interface, into the bedrock. We have clarified these two points in
 the text.
@@
 
-Our changes described above partly address this comment. We also plan to add
-add a new column to Table~1 containing references motivating default parameter
-choices.
+Our changes described above partly address this comment. We have added a new
+column to Table~1 containing references motivating default parameter choices.
 
@@
 interpolated from the NARR 32\,km Lambert grid to the computational domain
-grids. We apologize for omitting this information and will add it to the
+grids. We apologize for omitting this information and have now added it to the
 manuscript.
@@
 too coarse for ice flow modelling and that temperature and precipitation
-downscaling techniques can correct for errors caused by resolution gap. We will
-rework this sentence to make it clear in manuscript.
+downscaling techniques can correct for errors caused by resolution gap. We have
+rephrased this sentence to highlight this.
 

### 2016-01-03 reviews/response-marshall.tex
Add common header for responses.
Resp. Marshall: temporarily use bibtex.

@@
-\documentclass[10pt]{article}
-
-\usepackage[T1]{fontenc}
-\usepackage[pdftex]{graphicx}
-\usepackage[pdftex]{color}
-\usepackage{hyperref}
-\usepackage{natbib}
-
-\sloppy
-\setlength{\parskip}{1.2ex}
-\setlength{\parindent}{0em}
-\renewcommand\familydefault{\sfdefault}
-
-\graphicspath{{../figures/}}
-\definecolor{journalname}{rgb}{0.341,0.596,0.824}
-\definecolor{adgeored}{rgb}{0.439,0.157,0.145}  % from Advances in Geoscience
+% response-marshall.tex
+% ----------------------------------------------------------------------
+\input{response-header}
 
@@
 
+Vancouver Island: \citet{Cosma.etal.2008, Taylor.etal.2014, Taylor.etal.2015}.
+
+Logan ice core: \citet{Fisher.etal.2004, Fisher.etal.2008}.
+
 \todo{%
@@
 
-\begin{thebibliography}{69}
-
-\bibitem[{Seguinot et~al.(2014)Seguinot, Khroulev, Rogozhina, Stroeven, and
-  Zhang}]{Seguinot.etal.2014}
-Seguinot, J., Khroulev, C., Rogozhina, I., Stroeven, A.~P., and Zhang, Q.: The
-  effect of climate forcing on numerical simulations of the {C}ordilleran ice
-  sheet at the {L}ast {G}lacial {M}aximum, The Cryosphere, 8, 1087--1103,
-  \doi{10.5194/tc-8-1087-2014}, 2014.
-
-\end{thebibliography}
+% FIXME: no bibtex in final response
+\bibliographystyle{copernicus}
+\bibliography{../refs/references}
 

### 2016-01-03 reviews/response-jarosch.tex
Resp. Jarosch: correct date in Jarosch.etal.2012.

@@
     available station data for large parts of the study domain dealt with in
-    this manuscript. This evaluation \citep{Jarosch.etal.2010} demonstrated
+    this manuscript. This evaluation \citep{Jarosch.etal.2012} demonstrated
     that NARR has difficulties simulating orographic processes in the Coast
     Mountains which in turn results in unrealistic atmospheric conditions over
-    the Rocky Mountains. \citet{Jarosch.etal.2010} further concluded that
+    the Rocky Mountains. \citet{Jarosch.etal.2012} further concluded that
     physics based downscaling is required to adequately drive glacier models in
@@
     that NARR precipitation fields at 32\,km are adequate to drive their model
-    and reflect their arguments with the findings of \citet{Jarosch.etal.2010}.
+    and reflect their arguments with the findings of \citet{Jarosch.etal.2012}.
     A solid argument here is of special importance as the authors assume the
@@
 We acknowledge that extending the downscaling method presented by
-\citet{Jarosch.etal.2010} to the entire modelling domain and using the
+\citet{Jarosch.etal.2012} to the entire modelling domain and using the
 downscaled temperature and precipitation fields as input to the model would be
@@
 limitations implied by not including temperature and precipitation downscaling,
-and refer to the findings of \citet{Jarosch.etal.2010}. We have also modified
+and refer to the findings of \citet{Jarosch.etal.2012}. We have also modified
 parts of the interpretations in the light of your comment, for instance
@@
     manuscript. It has been demonstrated by spectral power analysis
-    \citep{Jarosch.etal.2010} that the NARR precipitation fields do not contain
+    \citep{Jarosch.etal.2012} that the NARR precipitation fields do not contain
     any significant spacial information below approximately 39\,km resolution
@@
     techniques however are able to add spatial information to precipitation
-    fields down to about 1\,km grid sizes \citep{Jarosch.etal.2010}. Taking
+    fields down to about 1\,km grid sizes \citep{Jarosch.etal.2012}. Taking
     these findings into the current context of the manuscript at hand, the NARR
@@
     discuss their choice in the light of the findings from
-    \citet{Jarosch.etal.2010}. Temperature however is better constrained in
-    NARR \citep{Jarosch.etal.2010} and contains spectral information down to
+    \citet{Jarosch.etal.2012}. Temperature however is better constrained in
+    NARR \citep{Jarosch.etal.2012} and contains spectral information down to
     10\,km resolution, which justifies the usage of NARR temperature fields on

### 2016-01-03 reviews/response-marshall.tex
Resp. Marshall: adress choice of proxy records.

@@
 To S.~J.~Marshall,
+
+We thank you very much for this positive review of our manuscript.
 
@@
 
+Thank you for summarizing so well the most challenging aspects of our study.
+
 \referee{%
@@
 
-Vancouver Island: \citet{Cosma.etal.2008, Taylor.etal.2014, Taylor.etal.2015}.
+Thank you for raising this point. Choosing proxy records to drive the model
+was not trivial and, realising that have not explained this choice in the
+first version of the manuscript, we are happy to do it here.
 
-Logan ice core: \citet{Fisher.etal.2004, Fisher.etal.2008}.
+First, we focus on records that have been recognized as proxies for
+palaeo-temperature. This includes oxygen isotope (\chem{\delta^{18}O}) records
+from the Greenland and Antarctic ice sheet. The Mount Logan ice core
+\chem{\delta^{18}O} record has been interpreted as a proxy for source region
+rather than palaeotemperature and therefore does not enter this category
+\citep{Fisher.etal.2004, Fisher.etal.2008}. Besides ice core records, we use
+alkenone unsaturation index (\chem{U^{K'}_{37}}) series from oceanic sediment
+cores, a known proxy for sea-surface temperatures \citep{Prahl.Wakeham.1987,
+Prahl.etal.1988,Muller.etal.1998}. We do not use $\delta^{18}O$ records from
+benthic foraminifera, which have been evaluated worldwide
+\citep{Lisiecki.Raymo.2005}, including for core~MD02-2496 near Vancouver Island
+\citep{Cosma.etal.2008}, as they are commonly interpreted as proxies for ice
+volume on land \citep{Shackleton.1967} rather than temperatures.
 
-\todo{%
-    Logan ice core covers only the last 16\,Ka and is not a good proxy for
-    temperature. The Vancouver SST record covers only deglaciation. Check new
-    papers for updates. Discuss in the manuscript.}
+Second, we choose to use only records that span over the last 120\,ka in order
+to avoid spinning-up the model using a different record than the one tested.
+This is a second reason for not including the Mount Logan \chem{\delta^{18}O}
+record \citep{Fisher.etal.2004}. Offshore Vancouver Island,
+sea-surface temperatures have been reconstructed for core~JT96-09 from alkenone
+unsaturation indices over the last 16 ka \citep{Kienast.McKay.2001}, and for nearby
+core~MD02-2496 from Mg/Ca ratio in planktonic foraminifera (\emph{N. pachyderma}
+and \emph{G. bulloides}) between 12 and 21\,\unit{\chem{^{14}C}\,cal\,ka}
+\citep{Taylor.etal.2014}. More recently, the planktonic foraminifera record has
+been extended to 10 to ca.~50\,\unit{\chem{^{14}C}\,cal\,ka}
+\citep{Taylor.etal.2015}, which could perhaps be used to model Cordilleran ice
+sheet growth and decay during the Marine Oxygen Isotope Stage (MIS) 2, but
+coult not be directly compared with other runs without spinning-up the model
+using a different proxy record.
+
+A shortened version of the preceding explanation has been added to the climate
+forcing section of the manuscript. We have also clarified in our conclusions
+that we do not consider the GRIP and EPICA ice core records, our preferred
+climate drivers, as adequate proxies of climate variability in western North
+America, and highlight the need of more regional proxy records to understand
+the dynamics of the Cordilleran ice sheet through the last glacial cycle.
 

### 2016-01-03 reviews/response-marshall.tex
Resp. Marshall: discuss modern climate patterns.

@@
 
-\todo{%
-    Perhaps we forgot to mention that as a limitation in the manuscript. Find
-    the appropriate place to add this in the discussion, and rephrase the above
-    points into the manuscript.}
+Indeed, we chose to keep modern-day spatial patterns for temperature and
+precipitation constant in order to limit the number of degrees of freedom in
+our study.
+
+Regarding temperature, it is reasonable to think that the changes were greater
+inland than near the coast. However, our simulation already produce an
+excessive of ice inland. Including such a temperature continentality gradient
+in the model while keeping the precipitation pattern constant would thus cause
+an even greater mismatch between model results and the geologically
+reconstructed Last Glacial Maximum (LGM) ice margins. We have introduced a
+short comment of this effect in Sect.~5.1.2 (previously~4.1.2, Ice
+configuration during MIS~2).
+
+However, the mismatch we observe between the modelled and reconstructed LGM ice
+margins let us think that the assumption of fixed modern-day precipitation
+patterns is more critical than the assumption of fixed modern-day temperature
+patterns. During phases of ice sheet growth, the presence of mountain ice caps
+on the Coast Mountains likely resulted in a decrease of precipitation inland
+(Sect.~5.1.2). Furthermore, the adiabatic warming associated with moisture
+depletion in the interior (Sect.~5.1.2) may have counterbalanced the potential
+continentality gradient discussed above.
 
@@
 
-\todo{%
-    Apply grammar corrections.}
+We have corrected grammatical mistakes given in the attachment. Thank you for
+spotting them. And thank you again for supporting our manuscript with such
+enthusiasm!
 

### 2016-01-03 reviews/response-marshall.tex
Resp. Marshall: rework choice of proxy records.

--- 2016-01-03 reviews/response-marshall.tex
@@
 
-Vancouver Island: \citet{Cosma.etal.2008, Taylor.etal.2014, Taylor.etal.2015}.
+Thank you for raising this point. Before explaining our choice of
+palaeo-climate proxy records, we would like to clarify that we do not consider
+the GRIP and EPICA ice core records, our preferred climate drivers, as
+satisfactory proxies of climate variability in western North America. Instead,
+we think that more regional proxy records are needed in order to better
+understand the dynamics of the Cordilleran ice sheet through the last glacial
+cycle, and have updated the conclusions to highlight this fact in the
+manuscript.
 
-Logan ice core: \citet{Fisher.etal.2004, Fisher.etal.2008}.
+To choose palaeo-climate proxy records to force the model, we first focused on
+records that are recognized as proxies for temperature. This includes oxygen
+isotope (\chem{\delta^{18}O}) records from the Greenland and Antarctic ice
+sheets. Besides ice core records, we use alkenone unsaturation index
+(\chem{U^{K'}_{37}}) series from oceanic sediment cores, a known proxy for
+sea-surface temperatures \citep{Prahl.Wakeham.1987, Prahl.etal.1988,
+Muller.etal.1998}. However, we do not use $\delta^{18}O$ records from benthic
+foraminifera, which have been evaluated worldwide \citep{Lisiecki.Raymo.2005},
+including for core~MD02-2496 near Vancouver Island \citep{Cosma.etal.2008}, but
+are commonly interpreted as proxies for global ice volume on land rather than
+for temperatures \citep{Shackleton.1967}. Second, we chose to use only records
+that span over the last 120\,ka in order to avoid spinning-up the model using a
+different record than the one being tested. The Mount Logan ice core
+\chem{\delta^{18}O} record covers only the last 30\,Ka and has been interpreted
+as a proxy for source region rather than for palaeo-temperature
+\citep{Fisher.etal.2004, Fisher.etal.2008} and thus do not fit our criteria.
 
-\todo{%
-    Logan ice core covers only the last 16\,Ka and is not a good proxy for
-    temperature. The Vancouver SST record covers only deglaciation. Check new
-    papers for updates. Discuss in the manuscript.}
+Concerning sediment cores offshore Vancouver Island, sea-surface temperatures
+have been reconstructed for core~JT96-09 from alkenone unsaturation indices
+over the last 16 ka \citep{Kienast.McKay.2001}, and for the nearby
+core~MD02-2496 from the Mg/Ca ratio in planktonic foraminifera (\emph{N.
+pachyderma} and \emph{G. bulloides}) between 12 and
+21\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Taylor.etal.2014}. Very recently, this
+planktonic foraminifera record has been extended to cover the period from 10 to
+ca.~50\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Taylor.etal.2015}, and thus it
+could perhaps be used to model Cordilleran ice sheet growth and decay during
+the Marine Oxygen Isotope Stage (MIS) 2, but such a simulation could not be
+directly compared with other runs without spinning-up the model using a
+different proxy record.
+
+A shortened version of the preceding explanation has been added to the climate
+forcing section of the manuscript.
 

### 2016-01-03 reviews/response-marshall.tex
Resp. Marshall: make small corrections.

@@
 Regarding temperature, it is reasonable to think that the changes were greater
-inland than near the coast. However, our simulation already produce an
-excessive of ice inland. Including such a temperature continentality gradient
+inland than near the coast. However, our simulations already produce an
+excess of ice inland. Including such a temperature continentality gradient
 in the model while keeping the precipitation pattern constant would thus cause
-an even greater mismatch between model results and the geologically
+an even greater mismatch between the model results and the geologically
 reconstructed Last Glacial Maximum (LGM) ice margins. We have introduced a
@@
 We have corrected grammatical mistakes given in the attachment. Thank you for
-spotting them. And thank you again for supporting our manuscript with such
-enthusiasm!
+spotting them. And thank you again very much for supporting our manuscript with
+such enthusiasm!
 

### 2016-01-04 reviews/response-marshall.tex
Resp. Marshall: add corrections from Irina.

@@
 To choose palaeo-climate proxy records to force the model, we first focused on
-records that are recognized as proxies for temperature. This includes oxygen
+records that are recognized as proxies for temperature. These include oxygen
 isotope (\chem{\delta^{18}O}) records from the Greenland and Antarctic ice
@@
 as a proxy for source region rather than for palaeo-temperature
-\citep{Fisher.etal.2004, Fisher.etal.2008} and thus do not fit our criteria.
+\citep{Fisher.etal.2004, Fisher.etal.2008} and thus does not fit our criteria.
 
@@
 ca.~50\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Taylor.etal.2015}, and thus it
-could perhaps be used to model Cordilleran ice sheet growth and decay during
+could perhaps be used to model the growth and decay of the Cordilleran ice
+sheet during
 the Marine Oxygen Isotope Stage (MIS) 2, but such a simulation could not be

### 2016-01-04 reviews/response-jarosch.tex
Resp. Jarosch: mention changes related to ETOPO1.
Resp. Jarosch: add corrections from Irina.

@@
 
+We have added a description of the limitations of ETOPO1 data in the methods,
+and a comment on overestimated ice thicknesses in the Wrangell and St.-Elias
+mountains in Sect.~5.2.2 (Major ice-dispersal centres).
+
 \referee{%
@@
 Our changes described above partly address this comment. To make our parameter
-choices clear, we have included all default parameters values mentioned
-in this response in the manuscript's Table~1 (alternative values used in the
-sensitivity tests being given in a separate table)
+choices clear, we have included all default parameter values mentioned
+in this response in the manuscript's Table~1 (with the alternative values used
+in the sensitivity tests being given in a separate table).
 
@@
 too coarse for ice flow modelling and that temperature and precipitation
-downscaling techniques can correct for errors caused by resolution gap. We have
-rephrased this sentence to highlight this.
+downscaling techniques could potentially correct for the errors caused by
+the resolution gap. We have rephrased the sentence to highlight this.
 

### 2016-01-04 reviews/response-jarosch.tex
Resp. Jarosch: rework response on downscaling.

@@
 
-We acknowledge that extending the downscaling method presented by
-\citet{Jarosch.etal.2012} to the entire modelling domain and using the
-downscaled temperature and precipitation fields as input to the model would be
-a significant and much needed improvement to our study. However, this would
-also be (for us) an entirely new work package which we can not complete during
-the time frame of this peer review. What we have done is to acknowledge the
-limitations implied by not including temperature and precipitation downscaling,
-and refer to the findings of \citet{Jarosch.etal.2012}. We have also modified
-parts of the interpretations in the light of your comment, for instance
-regarding the excess ice on and beyond the Columbia and Rocky mountains in our
-runs. We have previously shown that forcing PISM with bilinearly-interpolated
-climate fields from the NARR without temperature and precipitation downscaling
-resulted in realistic ice margins at first order relative to the geological
-record, with limitations that have been identified \citep{Seguinot.etal.2014}.
+We admit that the NARR climate forcing used in our simulations has its
+limitations.
+
+In our previous study \citep{Seguinot.etal.2014}, we have evaluated the
+performance of NARR in forcing constant-climate simulations of the Cordilleran
+ice sheet against that of an observation-based data set
+\citep[WorldClim]{Hijmans.etal.2005}. Indeed, the use of NARR in these
+simulations produced slightly different patterns of glaciation relative to
+WorldClim, including more extensive ice cover on the Columbia and Rocky
+mountains \citep[Figs.~6--7]{Seguinot.etal.2014}. Our simulations have shown
+that these differences are mainly caused by disparities in the precipitation
+fields of the two data sets \citep[Figs.13--14]{Seguinot.etal.2014}, which may,
+indeed, be related to the fact that the NARR has difficulties with simulating
+orographic processes in some areas of steep topography as discussed by
+\citet{Jarosch.etal.2012}.
+
+In the light of your comments, we have introduced a discussion of the NARR and
+its limitations in Sect.~2.6 (Climate forcing) and in Sect.~5.1.2 (Ice
+configuration during MIS~2). We acknowledge that the implementation of the
+downscaling method presented by \citet{Jarosch.etal.2012} may address these
+limitations. However, extending this downscaling method to the entire the model
+domain used in our study could be challenging, because the northern part of the
+model domain is characterized by stronger precipitation gradients and includes
+much fewer weather stations than the computational domain of
+\citet{Jarosch.etal.2012}.
 

### 2016-01-04 reviews/response-jarosch.tex
Resp. Jarosch: pull in references.

@@
 
-% FIXME: no bibtex in final response
-\bibliographystyle{copernicus}
-\bibliography{../refs/references}
+\begin{thebibliography}{69}
+
+\bibitem[{Aschwanden et~al.(2012)Aschwanden, Bueler, Khroulev, and
+  Blatter}]{Aschwanden.etal.2012}
+Aschwanden, A., Bueler, E., Khroulev, C., and Blatter, H.: An enthalpy
+  formulation for glaciers and ice sheets, J. Glaciol., 58, 441--457,
+  \doi{10.3189/2012JoG11J088}, 2012.
+
+\bibitem[{Aschwanden et~al.(2013)Aschwanden, A{\dh}algeirsd\'{o}ttir, and
+  Khroulev}]{Aschwanden.etal.2013}
+Aschwanden, A., A{\dh}algeirsd\'{o}ttir, G., and Khroulev, C.: Hindcasting to
+  measure ice sheet model sensitivity to initial states, The Cryosphere, 7,
+  1083--1093, \doi{10.5194/tc-7-1083-2013}, 2013.
+
+\bibitem[{Bueler and Brown(2009)}]{Bueler.Brown.2009}
+Bueler, E. and Brown, J.: Shallow shelf approximation as a ``sliding law'' in
+  a~thermodynamically coupled ice sheet model, J. Geophys. Res., 114, F03\,008,
+  \doi{10.1029/2008JF001179}, 2009.
+
+\bibitem[{Bueler and van Pelt(2015)}]{Bueler.Pelt.2015}
+Bueler, E. and van Pelt, W.: Mass-conserving subglacial hydrology in the
+  Parallel Ice Sheet Model version 0.6, Geosci. Model Dev., 8, 1613--1635,
+  \doi{10.5194/gmd-8-1613-2015}, 2015.
+
+\bibitem[{Clague(1981)}]{Clague.1981}
+Clague, J.~J.: Late Quaternary Geology and Geochronology of British Columbia
+  Part 2: Summary and Discussion of Radiocarbon-Dated Quaternary History, Geol.
+  Surv. of Can., Ottawa, ON, Paper 80-35, \doi{10.4095/119439}, 1981.
+
+\bibitem[{Clarke et~al.(2013)Clarke, Anslow, Jarosch, Radi{\'{c}}, Menounos,
+  Bolch, and Berthier}]{Clarke.etal.2013}
+Clarke, G. K.~C., Anslow, F.~S., Jarosch, A.~H., Radi{\'{c}}, V., Menounos, B.,
+  Bolch, T., and Berthier, E.: Ice Volume and Subglacial Topography for Western
+  Canadian Glaciers from Mass Balance Fields, Thinning Rates, and a Bed Stress
+  Model, J. Climate, 26, 4282--4303, \doi{10.1175/jcli-d-12-00513.1}, 2013.
+
+\bibitem[{Cuffey and Paterson(2010)}]{Cuffey.Paterson.2010}
+Cuffey, K.~M. and Paterson, W. S.~B.: The physics of glaciers, Elsevier,
+  Amsterdam, 2010.
+
+\bibitem[{Glen(1952)}]{Glen.1952}
+Glen, J.: Experiments on the deformation of ice, J. Glaciol., 2, 111--114,
+  1952.
+
+\bibitem[{Greve(1997)}]{Greve.1997}
+Greve, R.: A continuum-mechanical formulation for shallow polythermal ice
+  sheets, Philos. T. R. Soc. A, 355, 921--974, \doi{10.1098/rsta.1997.0050},
+  1997.
+
+\bibitem[{Hijmans et~al.(2005)Hijmans, Cameron, Parra, Jones, and
+  Jarvis}]{Hijmans.etal.2005}
+Hijmans, R.~J., Cameron, S.~E., Parra, J.~L., Jones, P.~G., and Jarvis, A.:
+  Very high resolution interpolated climate surfaces for global land areas,
+  Int. J. Climatol., 25, 1965--1978, \doi{10.1002/joc.1276}, 2005.
+
+\bibitem[{Jarosch et~al.(2012)Jarosch, Anslow, and Clarke}]{Jarosch.etal.2012}
+Jarosch, A.~H., Anslow, F.~S., and Clarke, G. K.~C.: High-resolution
+  precipitation and temperature downscaling for glacier models, Clim. Dynam.,
+  38, 391--409, \doi{10.1007/s00382-010-0949-1}, 2012.
+
+\bibitem[{Lisiecki and Raymo(2005)}]{Lisiecki.Raymo.2005}
+Lisiecki, L.~E. and Raymo, M.~E.: A Pliocene-Pleistocene stack of 57 globally
+  distributed benthic $\delta^{18}$O records, Paleoceanography, 20, PA1003,
+  \doi{10.1029/2004pa001071}, 2005.
+
+\bibitem[{Lliboutry and Duval(1985)}]{Lliboutry.Duval.1985}
+Lliboutry, L.~A. and Duval, P.: Various isotropic and anisotropic ices found in
+  glaciers and polar ice caps and their corresponding rheologies, Ann.
+  Geophys., 3, 207--224, 1985.
+
+\bibitem[{L{\"u}thi et~al.(2002)L{\"u}thi, Funk, Iken, Gogineni, and
+  Truffer}]{Luthi.etal.2002}
+L{\"u}thi, M., Funk, M., Iken, A., Gogineni, S., and Truffer, M.: Mechanisms of
+  fast flow in {J}akobshavns {I}sbr{\ae}, {G}reenland; {P}art {III}:
+  measurements of ice deformation, temperature and cross-borehole conductivity
+  in boreholes to the bedrock, J. Glaciol., 48, 369--385,
+  \doi{10.3189/172756502781831322}, 2002.
+
+\bibitem[{Martin et~al.(2011)Martin, Winkelmann, Haseloff, Albrecht, Bueler,
+  Khroulev, and Levermann}]{Martin.etal.2011}
+Martin, M.~A., Winkelmann, R., Haseloff, M., Albrecht, T., Bueler, E.,
+  Khroulev, C., and Levermann, A.: The Potsdam Parallel Ice Sheet Model
+  (PISM-PIK) -- Part 2: Dynamic equilibrium simulation of the Antarctic ice
+  sheet, The Cryosphere, 5, 727--740, \doi{10.5194/tc-5-727-2011}, 2011.
+
+\bibitem[{Nye(1953)}]{Nye.1953}
+Nye, J.~F.: The Flow Law of Ice from Measurements in Glacier Tunnels,
+  Laboratory Experiments and the Jungfraufirn Borehole Experiment, Proc. R.
+  Soc. London, Ser. A, 219, 477--489, 1953.
+
+\bibitem[{Paterson and Budd(1982)}]{Paterson.Budd.1982}
+Paterson, W. S.~B. and Budd, W.~F.: Flow parameters for ice sheet modeling,
+  Cold Reg. Sci. Technol., 6, 175--177, 1982.
+
+\bibitem[{Rignot et~al.(2013)Rignot, Mouginot, Larsen, Gim, and
+  Kirchner}]{Rignot.etal.2013}
+Rignot, E., Mouginot, J., Larsen, C.~F., Gim, Y., and Kirchner, D.:
+  Low-frequency radar sounding of temperate ice masses in Southern Alaska,
+  Geophys. Res. Lett., 40, 5399--5405, \doi{10.1002/2013gl057452}, 2013.
+
+\bibitem[{Seguinot(2014)}]{Seguinot.2014}
+Seguinot, J.: Numerical modelling of the Cordilleran ice sheet, Ph.D. thesis,
+  Stockholm University,
+  \urlprefix\url{http://urn.kb.se/resolve?urn=urn:nbn:se:su:diva-106815}, 2014.
+
+\bibitem[{Seguinot et~al.(2014)Seguinot, Khroulev, Rogozhina, Stroeven, and
+  Zhang}]{Seguinot.etal.2014}
+Seguinot, J., Khroulev, C., Rogozhina, I., Stroeven, A.~P., and Zhang, Q.: The
+  effect of climate forcing on numerical simulations of the {C}ordilleran ice
+  sheet at the {L}ast {G}lacial {M}aximum, The Cryosphere, 8, 1087--1103,
+  \doi{10.5194/tc-8-1087-2014}, 2014.
+
+\bibitem[{the PISM~authors(2015)}]{PISM-authors.2015}
+the PISM~authors: {PISM}, a~{P}arallel {I}ce {S}heet {M}odel,
+  \urlprefix\url{http://www.pism-docs.org}, 2015.
+
+\bibitem[{Tulaczyk et~al.(2000)Tulaczyk, Kamb, and
+  Engelhardt}]{Tulaczyk.etal.2000}
+Tulaczyk, S., Kamb, W.~B., and Engelhardt, H.~F.: Basal mechanics of Ice Stream
+  B, west Antarctica: 1. Till mechanics, J. Geophys. Res., 105, 463,
+  \doi{10.1029/1999jb900329}, 2000.
+
+\bibitem[{Winkelmann et~al.(2011)Winkelmann, Martin, Haseloff, Albrecht,
+  Bueler, Khroulev, and Levermann}]{Winkelmann.etal.2011}
+Winkelmann, R., Martin, M.~A., Haseloff, M., Albrecht, T., Bueler, E.,
+  Khroulev, C., and Levermann, A.: The {P}otsdam {P}arallel {I}ce {S}heet
+  {M}odel ({PISM-PIK}) -- Part 1: model description, The Cryosphere, 5,
+  715--726, \doi{10.5194/tc-5-715-2011}, 2011.
+
+\end{thebibliography}
 

### 2016-01-04 reviews/response-marshall.tex
Resp. Marshall: pull in references.

@@
 
-% FIXME: no bibtex in final response
-\bibliographystyle{copernicus}
-\bibliography{../refs/references}
+\begin{thebibliography}{69}
 
+\bibitem[{Cosma et~al.(2008)Cosma, Hendy, and Chang}]{Cosma.etal.2008}
+Cosma, T., Hendy, I., and Chang, A.: Chronological constraints on Cordilleran
+  Ice Sheet glaciomarine sedimentation from core MD02-2496 off Vancouver Island
+  (western Canada), Quaternary Sci. Rev., 27, 941--955,
+  \doi{10.1016/j.quascirev.2008.01.013}, 2008.
+
+\bibitem[{Fisher et~al.(2008)Fisher, Osterberg, Dyke, Dahl-Jensen, Demuth,
+  Zdanowicz, Bourgeois, Koerner, Mayewski, Wake, Kreutz, Steig, Zheng, Yalcin,
+  Goto-Azuma, Luckman, and Rupper}]{Fisher.etal.2008}
+Fisher, D., Osterberg, E., Dyke, A., Dahl-Jensen, D., Demuth, M., Zdanowicz,
+  C., Bourgeois, J., Koerner, R.~M., Mayewski, P., Wake, C., Kreutz, K., Steig,
+  E., Zheng, J., Yalcin, K., Goto-Azuma, K., Luckman, B., and Rupper, S.: The
+  Mt Logan Holocene--late Wisconsinan isotope record: tropical Pacific--Yukon
+  connections, The Holocene, 18, 667--677, \doi{10.1177/0959683608092236},
+  2008.
+
+\bibitem[{Fisher et~al.(2004)Fisher, Wake, Kreutz, Yalcin, Steig, Mayewski,
+  Anderson, Zheng, Rupper, Zdanowicz, Demuth, Waszkiewicz, Dahl-Jensen,
+  Goto-Azuma, Bourgeois, Koerner, Sekerka, Osterberg, Abbott, Finney, and
+  Burns}]{Fisher.etal.2004}
+Fisher, D.~A., Wake, C., Kreutz, K., Yalcin, K., Steig, E., Mayewski, P.,
+  Anderson, L., Zheng, J., Rupper, S., Zdanowicz, C., Demuth, M., Waszkiewicz,
+  M., Dahl-Jensen, D., Goto-Azuma, K., Bourgeois, J.~B., Koerner, R.~M.,
+  Sekerka, J., Osterberg, E., Abbott, M.~B., Finney, B.~P., and Burns, S.~J.:
+  Stable Isotope Records from Mount Logan, Eclipse Ice Cores and Nearby
+  Jellybean Lake. Water Cycle of the North Pacific Over 2000 Years and Over
+  Five Vertical Kilometres: Sudden Shifts and Tropical Connections, G\'{e}ogr.
+  phys. Quatern., 58, 337, \doi{10.7202/013147ar}, 2004.
+
+\bibitem[{Kienast and McKay(2001)}]{Kienast.McKay.2001}
+Kienast, S.~S. and McKay, J.~L.: Sea surface temperatures in the subarctic
+  northeast Pacific reflect millennial-scale climate oscillations during the
+  last 16 kyrs, Geophys. Res. Lett., 28, 1563--1566,
+  \doi{10.1029/2000GL012543}, 2001.
+
+\bibitem[{Lisiecki and Raymo(2005)}]{Lisiecki.Raymo.2005}
+Lisiecki, L.~E. and Raymo, M.~E.: A Pliocene-Pleistocene stack of 57 globally
+  distributed benthic $\delta^{18}$O records, Paleoceanography, 20, PA1003,
+  \doi{10.1029/2004pa001071}, 2005.
+
+\bibitem[{M\"{u}ller et~al.(1998)M\"{u}ller, Kirst, Ruhland, von Storch, and
+  Rosell-Mel{\'{e}}}]{Muller.etal.1998}
+M\"{u}ller, P.~J., Kirst, G., Ruhland, G., von Storch, I., and
+  Rosell-Mel{\'{e}}, A.: Calibration of the alkenone paleotemperature index
+  $U^{K'}_{37}$ based on core-tops from the eastern South Atlantic and the
+  global ocean (60$^\circ$N-60$^\circ$S), Geochim. Cosmochim. Ac., 62,
+  1757--1772, \doi{10.1016/s0016-7037(98)00097-0}, 1998.
+
+\bibitem[{Prahl and Wakeham(1987)}]{Prahl.Wakeham.1987}
+Prahl, F.~G. and Wakeham, S.~G.: Calibration of unsaturation patterns in
+  long-chain ketone compositions for palaeotemperature assessment, Nature, 330,
+  367--369, \doi{10.1038/330367a0}, 1987.
+
+\bibitem[{Prahl et~al.(1988)Prahl, Muehlhausen, and Zahnle}]{Prahl.etal.1988}
+Prahl, F.~G., Muehlhausen, L.~A., and Zahnle, D.~L.: Further evaluation of
+  long-chain alkenones as indicators of paleoceanographic conditions, Geochim.
+  Cosmochim. Ac., 52, 2303--2310, \doi{10.1016/0016-7037(88)90132-9}, 1988.
+
+\bibitem[{Seguinot et~al.(2014)Seguinot, Khroulev, Rogozhina, Stroeven, and
+  Zhang}]{Seguinot.etal.2014}
+Seguinot, J., Khroulev, C., Rogozhina, I., Stroeven, A.~P., and Zhang, Q.: The
+  effect of climate forcing on numerical simulations of the {C}ordilleran ice
+  sheet at the {L}ast {G}lacial {M}aximum, The Cryosphere, 8, 1087--1103,
+  \doi{10.5194/tc-8-1087-2014}, 2014.
+
+\bibitem[{Shackleton(1967)}]{Shackleton.1967}
+Shackleton, N.: Oxygen Isotope Analyses and Pleistocene Temperatures
+  Re-assessed, Nature, 215, 15--17, \doi{10.1038/215015a0}, 1967.
+
+\bibitem[{Taylor et~al.(2014)Taylor, Hendy, and Pak}]{Taylor.etal.2014}
+Taylor, M., Hendy, I., and Pak, D.: Deglacial ocean warming and marine margin
+  retreat of the Cordilleran Ice Sheet in the North Pacific Ocean, Earth
+  Planet. Sc. Lett., 403, 89--98, \doi{10.1016/j.epsl.2014.06.026}, 2014.
+
+\bibitem[{Taylor et~al.(2015)Taylor, Hendy, and Pak}]{Taylor.etal.2015}
+Taylor, M.~A., Hendy, I.~L., and Pak, D.~K.: The California Current System as a
+  transmitter of millennial scale climate change on the northeastern Pacific
+  margin from 10 to 50 ka, Paleoceanography, 30, 1168--1182,
+  \doi{10.1002/2014pa002738}, 2015.
+
+\end{thebibliography}
 % ----------------------------------------------------------------------

### 2016-01-04 reviews/response-stumpf.tex
Resp. Stumpf: update references.

@@
 
-\bibitem[{Ferbey et~al.(2013)}]{Ferbey.etal.2013}
+\bibitem[{Ferbey et~al.(2013)Ferbey, Arnold, and Hickin}]{Ferbey.etal.2013}
 Ferbey, T., Arnold, H., and Hickin, A.: Ice-flow indicator compilation, British
   Columbia., British Columbia Geol. Surv., Victoria, BC, Open-File 2013-06,
-  \url{http://www.em.gov.bc.ca/Mining/Geoscience/PublicationsCatalogue/OpenFiles/2013/Pages/2013-06.aspx},
+  \urlprefix\url{http://www.em.gov.bc.ca/Mining/Geoscience/PublicationsCatalogue/OpenFiles/2013/Pages/2013-06.aspx},
   2013.
 
-\bibitem[{Seguinot et~al.(2014)}]{Seguinot.etal.2014}
-Seguinot, J., Khroulev, C., Rogozhina, I., Stroeven, A.~P., and Zhang, Q.: The
-  effect of climate forcing on numerical simulations of the {C}ordilleran ice
-  sheet at the {L}ast {G}lacial {M}aximum, The Cryosphere, 8, 1087--1103,
-  \doi{10.5194/tc-8-1087-2014}, 2014.
+\bibitem[{Plouffe and Jett{\'{e}}(1997)}]{Plouffe.Jette.1997}
+Plouffe, A. and Jett{\'{e}}, H.: Middle Wisconsinan sediments and paleoecology
+  of central British Columbia: sites at Necoslie and Nautley rivers, Can. J.
+  Earth Sci., 34, 200--208, \doi{10.1139/e17-016}, 1997.
 
-\bibitem[{Stumpf et~al.(2000)}]{Stumpf.etal.2000}
+\bibitem[{Stumpf et~al.(2000)Stumpf, Broster, and Levson}]{Stumpf.etal.2000}
 Stumpf, A.~J., Broster, B.~E., and Levson, V.~M.: Multiphase flow of the late
@@
 
-\bibitem[{Stumpf et~al.(2014)}]{Stumpf.etal.2014}
+\bibitem[{Stumpf et~al.(2014)Stumpf, Ferbey, Plouffe, Clague, Ward, Paulen, and
+  Bush}]{Stumpf.etal.2014}
 Stumpf, A.~J., Ferbey, T., Plouffe, A., Clague, J.~J., Ward, B.~C., Paulen,
@@
   \doi{10.1016/j.geomorph.2013.10.019}, 2014.
-
-\bibitem[{Plouffe and Jett{\'{e}}(1997)}]{Plouffe.Jette.1997}
-Plouffe, A. and Jett{\'{e}}, H.: Middle Wisconsinan sediments and paleoecology
-  of central British Columbia: sites at Necoslie and Nautley rivers, Can. J.
-  Earth Sci., 34, 200--208, \doi{10.1139/e17-016}, 1997.
 

### 2016-01-05 reviews/response-jarosch.tex
DOI command need to be defined in responses.
Resp. Jarosch: figures should be out of comment.
Resp. Jarosch: no natbib in referee comments.

@@
 \newcommand{\msquote}[1]{\begin{quote}\textit{#1}\end{quote}}
+\newcommand{\doi}[1]{doi:\allowbreak\href{http://dx.doi.org/#1}{#1}}
 
@@
     proxy data, or climate-ice sheet feedback mechanisms not represented in the
-    model chain. Similarly in a previous study \citet{Seguinot.etal.2014} have
+    model chain. Similarly in a previous study Seguinot et~al. (2014) have
     focused only on the driving climate sensitivities and have omitted
@@
 
-\begin{figure}
-    \centering
-    \includegraphics{sens_plot_rheo}
-    \caption{Ice softness parameter, $A$, as a function of pressure-adjusted
-             temperature, $T_{pa}$, for the default \citep{Paterson.Budd.1982},
-             hard ice \citep[with $E_{SIA}=1$]{Cuffey.Paterson.2010}, and soft ice
-             \citep[with $E_{SIA}=5$]{Cuffey.Paterson.2010} rheologies, using
-             a linear scale (top panel) and logarithmic scale (bottom panel).
-             Figure made using Eqn.~\ref{eqn:softness} with parameters from
-             Table~\ref{tab:sens_params}.}
-    \label{fig:sens_plot_rheo}
-\end{figure}
-
-\clearpage  % moving on to basal sliding
+%\clearpage  % moving on to basal sliding
 
@@
 
-\begin{figure}
-    \centering
-    \includegraphics{sens_plot_ntil}
-    \caption{Effective pressure, $N$, as a function of water content in the
-             till, $W$, for the default ($\delta=0.02$, $W_{max}=2$\,m),
-             hard bed ($\delta=0.05$, $W_{max}=5$\,m), and soft bed
-             ($\delta=0.01$, $W_{max}=1$\,m) sliding parametrisations, using a
-             linear scale (top panel) and a logarithmic scale (bottom panel).
-             Calculations are made for an ice thickness, $h$, of 1000\,m.
-             Figure made using Eqn.~\ref{eqn:ntil} with parameters from
-             Table~\ref{tab:sens_params} after
-             \citet[Fig.~1]{Bueler.Pelt.2015}.}
-    \label{fig:sens_plot_ntil}
-\end{figure}
-
-\clearpage  % moving on to the results
+%\clearpage  % moving on to the results
 
@@
 
-\begin{figure*}
-  \centering
-  \includegraphics{sens_ts}
-  \caption{Modelled sea-level relevant ice volume through the last 120\,ka
-           in the simulation forced by the GRIP paleo-climate record, using
-           default parameters (black curves), different ice rheology parameters
-           (top panel), and different basal sliding parameters (bottom panel).
-           Gray fields indicate Marine Oxygen Isotope Stage (MIS) boundaries
-           for MIS~2 and MIS~4 according to a global compilation of benthic
-           \chem{\delta^{18}O} records \citep{Lisiecki.Raymo.2005}.}
-  \label{fig:sens_ts}
-\end{figure*}
-
-\clearpage  % eventually moving on to the next point
+%\clearpage  % eventually moving on to the next point
 
@@
     \textbf{P4151 L11-16:} In this sentence the authors refer back to their
-    previous work \citep{Seguinot.etal.2014} and highlight that the NARR
+    previous work (Seguinot et~al., 2014) and highlight that the NARR
     temperature and precipitation fields are the most suitable present day
@@
     include steep precipitation gradients which are required as identified by
-    et \citet{Seguinot.etal.2014}. NARR is delivered on a 32\,km Lambert grid,
+    et Seguinot et~al. (2014). NARR is delivered on a 32\,km Lambert grid,
     and thus it is questionable how ``steep'' these gradients can be, given
     the rather smooth representation of the existing topography on a 32\,km
-    grid. \citet{Seguinot.etal.2014} have partly discussed that however. NARR
+    grid. Seguinot et~al. (2014) have partly discussed that however. NARR
     precipitation and temperature fields have been evaluated in detail based on
     available station data for large parts of the study domain dealt with in
-    this manuscript. This evaluation \citep{Jarosch.etal.2012} demonstrated
+    this manuscript. This evaluation (Jarosch et~al., 2012) demonstrated
     that NARR has difficulties simulating orographic processes in the Coast
     Mountains which in turn results in unrealistic atmospheric conditions over
-    the Rocky Mountains. \citet{Jarosch.etal.2012} further concluded that
+    the Rocky Mountains. Jarosch et~al. (2012) further concluded that
     physics based downscaling is required to adequately drive glacier models in
@@
     that NARR precipitation fields at 32\,km are adequate to drive their model
-    and reflect their arguments with the findings of \citet{Jarosch.etal.2012}.
+    and reflect their arguments with the findings of Jarosch et~al. (2012).
     A solid argument here is of special importance as the authors assume the
@@
     contains the present day ice volumes within the study region.
-    \citet{Clarke.etal.2013}
+    Clarke et~al (2013)
     have estimated the ice volume in parts of that region to be $2530
@@
     As stated here, ice rheology within the used ice sheet model is based on
-    \citet{Aschwanden.etal.2012}. This enthalpy based formulation has proofed
+    Aschwanden et~al. (2012). This enthalpy based formulation has proofed
     itself to be very suitable for estimating ice rheology in ice sheet models,
     but it also depends on several parameters to translate enthalpy within the
-    ice to ice viscosity \citep[see][equations 62-65)]{Aschwanden.etal.2012}.
+    ice to ice viscosity (Aschwanden see et~al., 2012, equations 62-65).
     The authors do not mention any of these parameters (e.g. any of the rate
-    factors or nonlinear power $n$ ) within the manuscript or in Table~1. I
+    factors or nonlinear power $n$) within the manuscript or in Table~1. I
     have mentioned above in the general comments section that parameters used
@@
     either 3 or 6 times coarser than the numerical grid, thus not at all
-    high-resolution. \citet{Seguinot.etal.2014} state in their section~3.3 that
+    high-resolution. Seguinot et~al. (2014) state in their section~3.3 that
     the NARR data fields have been bilinearly interpolated to 10\,km resolution
@@
     manuscript. It has been demonstrated by spectral power analysis
-    \citep{Jarosch.etal.2012} that the NARR precipitation fields do not contain
+    (Jarosch et~al., 2012) that the NARR precipitation fields do not contain
     any significant spacial information below approximately 39\,km resolution
@@
     techniques however are able to add spatial information to precipitation
-    fields down to about 1\,km grid sizes \citep{Jarosch.etal.2012}. Taking
+    fields down to about 1\,km grid sizes (Jarosch et~al., 2012). Taking
     these findings into the current context of the manuscript at hand, the NARR
@@
     discuss their choice in the light of the findings from
-    \citet{Jarosch.etal.2012}. Temperature however is better constrained in
-    NARR \citep{Jarosch.etal.2012} and contains spectral information down to
+    Jarosch et~al. (2012). Temperature however is better constrained in
+    NARR (Jarosch et~al., 2012) and contains spectral information down to
     10\,km resolution, which justifies the usage of NARR temperature fields on
@@
 
+\begin{figure}
+    \centering
+    \includegraphics{sens_plot_rheo}
+    \caption{Ice softness parameter, $A$, as a function of pressure-adjusted
+             temperature, $T_{pa}$, for the default \citep{Paterson.Budd.1982},
+             hard ice \citep[with $E_{SIA}=1$]{Cuffey.Paterson.2010}, and soft ice
+             \citep[with $E_{SIA}=5$]{Cuffey.Paterson.2010} rheologies, using
+             a linear scale (top panel) and logarithmic scale (bottom panel).
+             Figure made using Eqn.~\ref{eqn:softness} with parameters from
+             Table~\ref{tab:sens_params}.}
+    \label{fig:sens_plot_rheo}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics{sens_plot_ntil}
+    \caption{Effective pressure, $N$, as a function of water content in the
+             till, $W$, for the default ($\delta=0.02$, $W_{max}=2$\,m),
+             hard bed ($\delta=0.05$, $W_{max}=5$\,m), and soft bed
+             ($\delta=0.01$, $W_{max}=1$\,m) sliding parametrisations, using a
+             linear scale (top panel) and a logarithmic scale (bottom panel).
+             Calculations are made for an ice thickness, $h$, of 1000\,m.
+             Figure made using Eqn.~\ref{eqn:ntil} with parameters from
+             Table~\ref{tab:sens_params} after
+             \citet[Fig.~1]{Bueler.Pelt.2015}.}
+    \label{fig:sens_plot_ntil}
+\end{figure}
+
+\begin{figure*}
+  \centering
+  \includegraphics{sens_ts}
+  \caption{Modelled sea-level relevant ice volume through the last 120\,ka
+           in the simulation forced by the GRIP paleo-climate record, using
+           default parameters (black curves), different ice rheology parameters
+           (top panel), and different basal sliding parameters (bottom panel).
+           Gray fields indicate Marine Oxygen Isotope Stage (MIS) boundaries
+           for MIS~2 and MIS~4 according to a global compilation of benthic
+           \chem{\delta^{18}O} records \citep{Lisiecki.Raymo.2005}.}
+  \label{fig:sens_ts}
+\end{figure*}
+
 \end{document}

### 2016-01-05 reviews/response-jarosch.tex
Resp. Jarosch: introduce short figs captions.
The long ones do not fit in and math is not rendered.
Move tables to the end on a separate page.
Resp. Jarosch: pull table cmds in comment text.
Resp. Jarosch: figure references are not allowed.

@@
 
+% Replacements for Copernicus commands
+\newcommand{\doi}[1]{doi:\allowbreak\href{http://dx.doi.org/#1}{#1}}
+\newcommand{\tophline}{\noalign{\vspace{1ex}}\hline\noalign{\vspace{1mm}}}
+\newcommand{\middlehline}{\noalign{\vspace{1mm}}\hline\noalign{\vspace{1mm}}}
+\newcommand{\bottomhline}{\noalign{\vspace{1mm}}\hline}
+\newcommand{\belowtable}[1]{\par\vspace{1ex}\raggedleft{\scriptsize#1}}
+
+% Personal commands
 \newcommand{\sechead}[1]{\bigskip\noindent\textbf{#1}}
@@
 \newcommand{\msquote}[1]{\begin{quote}\textit{#1}\end{quote}}
-\newcommand{\doi}[1]{doi:\allowbreak\href{http://dx.doi.org/#1}{#1}}
 
@@
 the three different parametrisations on temperature-dependent ice softness,
-$A$, is illustrated in Fig.~\ref{fig:sens_plot_rheo}.
-
-\begin{table*}
-  \caption{Parameter values used in the sensitivity test.}
-  \label{tab:sens_params}
-  \centering\makebox[\textwidth]
-  {\begin{tabular}{l|ccccc|cc|cc}
-    \tophline
-            & \multicolumn{5}{c|}{Rheology}
-            & \multicolumn{2}{c|}{Sliding}
-            & \multicolumn{2}{c}{GRIP scaling} \\
-    Config. & $A_c$ & $A_w$ & $Q_c$ & $Q_w$ & $E_{SIA}$
-            & $\delta$ & $W_{max}$ & $f$ & $T_{[32, 22]}$ \\
-            & \multicolumn{2}{c}{(\unit{Pa^{-3}\,s^{-1}})}
-            & \multicolumn{2}{c}{(\unit{J\,mol^{-1}})}
-            & & & (m) \\
-    \middlehline
-    Default$^1$  & $ 3.61\times 10^{-13}$
-                 & $ 1.73\times 10^3$
-                 & $   60\times 10^3$
-                 & $  139\times 10^3$
-                 & 1 & 0.02 & 2 & 0.38 & 6.2 \\
-    \middlehline
-    Soft ice$^2$ & $2.847\times 10^{-13}$
-                 & $2.356\times 10^{-2}$
-                 & $   60\times 10^3$
-                 & $  115\times 10^3$
-                 & 5 & 0.02 & 2 & 0.40 & 6.6 \\
-    Hard ice$^2$ & $2.847\times 10^{-13}$
-                 & $2.356\times 10^{-2}$
-                 & $   60\times 10^3$
-                 & $  115\times 10^3$
-                 & 1 & 0.02 & 2 & 0.37 & 6.0 \\
-    \middlehline
-    Soft bed     & $ 3.61\times 10^{-13}$
-                 & $ 1.73\times 10^3$
-                 & $   60\times 10^3$
-                 & $  139\times 10^3$
-                 & 1 & 0.01 & 1 & 0.40 & 6.5 \\
-    Hard bed     & $ 3.61\times 10^{-13}$
-                 & $ 1.73\times 10^3$
-                 & $   60\times 10^3$
-                 & $  139\times 10^3$
-                 & 1 & 0.05 & 5 & 0.36 & 5.9 \\
-    \bottomhline
-  \end{tabular}}
-  \belowtable{After $^1$\citet{Paterson.Budd.1982,Bueler.Pelt.2015};
-              and $^2$\citet{Cuffey.Paterson.2010}.}
-\end{table*}
+$A$, is illustrated in Fig.~1.
 
@@
 the till, $N$, in response to water content, $W$, is illustrated in
-Fig.~\ref{fig:sens_plot_ntil}.
+Fig.~2.
 All parameter choices are listed in Table~\ref{tab:sens_params}.
@@
 sensitivity study are presented here using a similar layout in
-Table~\ref{tab:sens_extrema} and Fig.~\ref{fig:sens_ts}.
+Table~\ref{tab:sens_extrema} and Fig.~3.
 
@@
 modelled glaciated areas (Table~\ref{tab:sens_extrema},
-Fig.~\ref{fig:sens_ts}). As one could expect, softer ice and weaker till both
+Fig.~3). As one could expect, softer ice and weaker till both
 result in a thinner ice sheet, while harder ice and stronger till result in a
@@
 sea-level relevant ice volume are greatest during the MIS~3
-(Table~\ref{tab:sens_extrema}, Fig.~\ref{fig:sens_ts}) where both the areal and
+(Table~\ref{tab:sens_extrema}, Fig.~3) where both the areal and
 thickness contributions add up.
-
-\begin{table*}
-  \caption{Extremes in Cordilleran ice sheet volume and extent corresponding to
-           MIS~4, 3 and 2 using the GRIP paleo-climate forcing with each
-           parameter configuration (Fig.~\ref{fig:sens_ts}). Relative
-           differences (R. diff.) give rough error estimates related to
-           varying selected ice rheology and basal sliding parameters
-           (Table~\ref{tab:sens_params}).}
-  \label{tab:sens_extrema}
-  \centering\makebox[\textwidth]
-  {\begin{tabular}{l*{3}{|ccc}}
-    \tophline
-             & \multicolumn{3}{c}{Age (ka)}
-             & \multicolumn{3}{c}{Ice extent (\unit{10^6\,km^2})}
-             & \multicolumn{3}{c}{Ice volume (m~s.l.e.)} \\
-    Config.  &  MIS~4 &  MIS~3 &  MIS~2
-             &  MIS~4 &  MIS~3 &  MIS~2
-             &  MIS~4 &  MIS~3 &  MIS~2 \\
-    \middlehline
-    Default  &  57.59 &  42.91 &  19.14
-             &   1.93 &   0.67 &   2.09
-             &   7.43 &   1.54 &   8.62 \\
-    \middlehline
-    Soft ice &  58.89 &  49.97 &  21.57
-             &   1.96 &   0.54 &   2.08
-             &   6.58 &   1.03 &   6.88 \\
-    Hard ice &  57.32 &  42.90 &  19.14
-             &   1.90 &   0.75 &   2.12
-             &   7.83 &   1.91 &   9.46 \\
-    R. diff. &    3\% &   16\% &   13\%
-             &    3\% &   31\% &    2\%
-             &   17\% &   57\% &   30\% \\
-    \middlehline
-    Soft bed &  58.90 &  49.21 &  19.53
-             &   1.88 &   0.55 &   2.05
-             &   6.46 &   1.03 &   7.52 \\
-    Hard bed &  57.31 &  42.91 &  19.14
-             &   1.93 &   0.96 &   2.13
-             &   7.99 &   2.89 &   9.31 \\
-    R. diff. &    3\% &   15\% &    2\%
-             &    3\% &   62\% &    4\%
-             &   21\% &  120\% &   21\% \\
-    \bottomhline
-  \end{tabular}}
-  \belowtable{}
-\end{table*}
 
@@
 Tables~\ref{tab:sens_params} and~\ref{tab:sens_extrema}, and
-Figs.~\ref{fig:sens_plot_rheo}--\ref{fig:sens_ts} have been included in Sect.~2
+Figs.~1--3 have been included in Sect.~2
 (Model setup) and in the new section (Sensitivity to ice flow parameters).
@@
 % ----------------------------------------------------------------------
+
+\begin{table*}[p]
+  \caption{Parameter values used in the sensitivity test.}
+  \label{tab:sens_params}
+  \centering\makebox[\textwidth]
+  {\begin{tabular}{l|ccccc|cc|cc}
+    \tophline
+            & \multicolumn{5}{c|}{Rheology}
+            & \multicolumn{2}{c|}{Sliding}
+            & \multicolumn{2}{c}{GRIP scaling} \\
+    Config. & $A_c$ & $A_w$ & $Q_c$ & $Q_w$ & $E_{SIA}$
+            & $\delta$ & $W_{max}$ & $f$ & $T_{[32, 22]}$ \\
+            & \multicolumn{2}{c}{(\unit{Pa^{-3}\,s^{-1}})}
+            & \multicolumn{2}{c}{(\unit{J\,mol^{-1}})}
+            & & & (m) \\
+    \middlehline
+    Default$^1$  & $ 3.61\times 10^{-13}$
+                 & $ 1.73\times 10^3$
+                 & $   60\times 10^3$
+                 & $  139\times 10^3$
+                 & 1 & 0.02 & 2 & 0.38 & 6.2 \\
+    \middlehline
+    Soft ice$^2$ & $2.847\times 10^{-13}$
+                 & $2.356\times 10^{-2}$
+                 & $   60\times 10^3$
+                 & $  115\times 10^3$
+                 & 5 & 0.02 & 2 & 0.40 & 6.6 \\
+    Hard ice$^2$ & $2.847\times 10^{-13}$
+                 & $2.356\times 10^{-2}$
+                 & $   60\times 10^3$
+                 & $  115\times 10^3$
+                 & 1 & 0.02 & 2 & 0.37 & 6.0 \\
+    \middlehline
+    Soft bed     & $ 3.61\times 10^{-13}$
+                 & $ 1.73\times 10^3$
+                 & $   60\times 10^3$
+                 & $  139\times 10^3$
+                 & 1 & 0.01 & 1 & 0.40 & 6.5 \\
+    Hard bed     & $ 3.61\times 10^{-13}$
+                 & $ 1.73\times 10^3$
+                 & $   60\times 10^3$
+                 & $  139\times 10^3$
+                 & 1 & 0.05 & 5 & 0.36 & 5.9 \\
+    \bottomhline
+  \end{tabular}}
+  \belowtable{After $^1$\citet{Paterson.Budd.1982,Bueler.Pelt.2015};
+              and $^2$\citet{Cuffey.Paterson.2010}.}
+\end{table*}
+
+\begin{table*}[p]
+  \caption{Extremes in Cordilleran ice sheet volume and extent corresponding to
+           MIS~4, 3 and 2 using the GRIP paleo-climate forcing with each
+           parameter configuration (Fig.~3). Relative
+           differences (R. diff.) give rough error estimates related to
+           varying selected ice rheology and basal sliding parameters
+           (Table~\ref{tab:sens_params}).}
+  \label{tab:sens_extrema}
+  \centering\makebox[\textwidth]
+  {\begin{tabular}{l*{3}{|ccc}}
+    \tophline
+             & \multicolumn{3}{c}{Age (ka)}
+             & \multicolumn{3}{c}{Ice extent (\unit{10^6\,km^2})}
+             & \multicolumn{3}{c}{Ice volume (m~s.l.e.)} \\
+    Config.  &  MIS~4 &  MIS~3 &  MIS~2
+             &  MIS~4 &  MIS~3 &  MIS~2
+             &  MIS~4 &  MIS~3 &  MIS~2 \\
+    \middlehline
+    Default  &  57.59 &  42.91 &  19.14
+             &   1.93 &   0.67 &   2.09
+             &   7.43 &   1.54 &   8.62 \\
+    \middlehline
+    Soft ice &  58.89 &  49.97 &  21.57
+             &   1.96 &   0.54 &   2.08
+             &   6.58 &   1.03 &   6.88 \\
+    Hard ice &  57.32 &  42.90 &  19.14
+             &   1.90 &   0.75 &   2.12
+             &   7.83 &   1.91 &   9.46 \\
+    R. diff. &    3\% &   16\% &   13\%
+             &    3\% &   31\% &    2\%
+             &   17\% &   57\% &   30\% \\
+    \middlehline
+    Soft bed &  58.90 &  49.21 &  19.53
+             &   1.88 &   0.55 &   2.05
+             &   6.46 &   1.03 &   7.52 \\
+    Hard bed &  57.31 &  42.91 &  19.14
+             &   1.93 &   0.96 &   2.13
+             &   7.99 &   2.89 &   9.31 \\
+    R. diff. &    3\% &   15\% &    2\%
+             &    3\% &   62\% &    4\%
+             &   21\% &  120\% &   21\% \\
+    \bottomhline
+  \end{tabular}}
+  \belowtable{}
+\end{table*}
+
+% ----------------------------------------------------------------------
 % Interactive comment text ends
@@
              Table~\ref{tab:sens_params}.}
+    % Short caption:
+    % Ice softness as a function of pressure-adjusted temperature for the
+    % default, hard ice, and soft ice rheologies, using a linear scale (top
+    % panel) and a logarithmic scale (bottom panel).
     \label{fig:sens_plot_rheo}
@@
              \citet[Fig.~1]{Bueler.Pelt.2015}.}
+    % Short caption:
+    % Effective pressure as a function of water content in the till for the
+    % default, hard bed, and soft bed sliding parametrisations, for an ice
+    % thickness of 1000 m.
     \label{fig:sens_plot_ntil}
@@
            \chem{\delta^{18}O} records \citep{Lisiecki.Raymo.2005}.}
+    % Short caption
+    % Modelled sea-level relevant ice volume using default parameters (black
+    % curves), different ice rheology parameters (top panel), and different
+    % basal sliding parameters (bottom panel).
   \label{fig:sens_ts}

### 2016-01-05 reviews/response-jarosch.tex
Resp. Jarosch: amsmath is not available.

@@
     A = E\cdot
-    \begin{cases}
+    \left\{\begin{array}{ll}
         A_c \,e^\frac{-Q_c}{RT_{pa}}
-            & \text{if}\ T_{pa} < T_c \,, \\
+            & \mathsf{if}\ T_{pa} < T_c \,, \\
         A_w (1+f\omega)\,e^\frac{-Q_w}{RT_{pa}}
-            & \text{if}\ T_{pa} \ge T_c \,,
-    \end{cases}
+            & \mathrm{if}\ T_{pa} \ge T_c \,,
+    \end{array}\right.
+    % or using amsmath:
+    %\begin{cases}
+    %    A_c \,e^\frac{-Q_c}{RT_{pa}}
+    %        & \text{if}\ T_{pa} < T_c \,, \\
+    %    A_w (1+f\omega)\,e^\frac{-Q_w}{RT_{pa}}
+    %        & \text{if}\ T_{pa} \ge T_c \,,
+    %\end{cases}
 \end{equation}

### 2016-01-05 reviews/response-editor.tex
Initialize response to the editor.
This is just to save an email for later that we meant to send to the editor this morning if I had not overslept.

% response-marshall.tex
% ----------------------------------------------------------------------
\input{response-header}

\begin{document}
\textbf{Authors' response to the Editor}
\bigskip

Dear Hilmar Gudmundsson,
(cc co-authors and Copernicus publications)

As you probably noticed, Copernicus stopped peer review of our manuscript
"Numerical simulations of the Cordilleran ice sheet through the last glacial
cycle" when we were nearly ready to submit our responses. Since the deadline
coincided with new year celebrations, I could not obtain feedback from my
co-authors on time to finalize the responses.

We greatly apologize for this additional delay and ask for your assistance to
continue with the peer review that has required a large time investment from
our side and from the side of the reviewers. To some extent, the delay has been
caused by our wish to address the reviews as well as possible. In addition to a
change of my working environment, setting up and running the sensitivity study
asked by Alexander Jarosch on a different supercomputer, using a new allocation
that was meant for a different project and became effective only in October,
took us considerable time.

I contacted Svenja Lange prior to the submission deadline to inquire after the
possibility of a short-term extension but later realized that Copernicus
offices were closed for the new year celebration until yesterday. I have tried
to contact Natascha T\"{o}pfer yesterday (as suggested in the following email from
Copernicus) but then realised on the Cryosphere website that she is on a leave.
Is there a chance that the on-line submission system could be re-opened for us
to submit our responses?

Thank you very much in advance for your understanding,

Best regards,

Julien.

\end{document}

### 2016-01-16 typeset.tex
Temporarily add discussion typeset in repo.
I want to see what changes were made.

\documentclass[tcd, online, hvmath]{copernicus}

\begin{document}\hack{\sloppy}

\title{Numerical simulations of the Cordilleran ice sheet
       through the last glacial cycle}

\Author[1,2,3]{J.}{Seguinot}
\Author[3]{I.}{Rogozhina}
\Author[2]{A.~P.}{Stroeven}
\Author[2]{M.}{Margold}
\Author[2]{J.}{Kleman}

\runningauthor{J.~Seguinot et~al.}

\correspondence{J.~Seguinot (seguinot@vaw.baug.ethz.ch)}

\runningtitle{Numerical simulations of the Cordilleran ice sheet
              through the last glacial cycle}

\affil[1]{Laboratory of Hydraulics, Hydrology and Glaciology, ETH Z\"{u}rich, Z\"{u}rich, Switzerland}
\affil[2]{Department of Physical Geography and the Bolin Centre for Climate Research, Stockholm University, Stockholm, Sweden}
\affil[3]{Helmholtz Centre Potsdam, GFZ German Research Centre for Geosciences, Potsdam,~Germany}

\hack{\allowdisplaybreaks}

\received{21~June 2015}
\accepted{25~June 2015}
\published{}

\firstpage{1}

\maketitle

\begin{abstract}
Despite more than a century of geological observations, the
Cordilleran ice sheet of North America remains poorly understood in
terms of its former extent, volume and dynamics. Although
geomorphological evidence is abundant, its complexity is such that
whole ice-sheet reconstructions of advance and retreat patterns are
lacking. Here we use a numerical ice sheet model calibrated against
field-based evidence to attempt a quantitative reconstruction of the
Cordilleran ice sheet history through the last glacial
cycle. A series of simulations is driven by time-dependent
temperature offsets from six proxy records located around the
globe. Although this approach reveals large variations in model
response to evolving climate forcing, all simulations produce two
major glaciations during marine oxygen isotope stages 4
(61.9--56.5\,\unit{ka}) and 2 (23.2--16.8\,\unit{ka}). The timing of
glaciation is better reproduced using temperature reconstructions
from Greenland and Antarctic ice cores than from regional oceanic
sediment cores. During most of the last glacial cycle, the modelled
ice cover is discontinuous and restricted to high mountain
areas. However, widespread precipitation over the Skeena Mountains
favours the persistence of a central ice dome throughout the glacial
cycle. It acts as a nucleation centre before the Last Glacial
Maximum and hosts the last remains of Cordilleran ice until the
middle Holocene (6.6--6.2\,ka).
\end{abstract}

\introduction
\label{sec:intro}

      During the last glacial cycle, glaciers and ice caps of the North
      American Cordillera have been more extensive than today. At the Last
      Glacial Maximum (LGM), a~continuous blanket of ice, the Cordilleran
      ice sheet \citep{Dawson.1888}, stretched from the Alaska Range in the
      north to the North Cascades in the south (Fig.~\ref{fig:locmap}).  In
      addition, it extended offshore, where it calved into the Pacific
      Ocean, and merged with the western margin of its much larger
      neighbour, the Laurentide ice sheet, east of the Rocky Mountains.

      More than a~century of exploration and geological investigation of the
      Cordilleran mountains have led to many observations in support of the
      former ice sheet \citep{Jackson.Clague.1991}.  Despite the lack of
      documented end moraines offshore, in the zone of confluence with the
      Laurentide ice sheet, and in areas swept by the Missoula floods
      \citep{Carrara.etal.1996}, moraines that demarcate the northern and
      south-western margins provide key constraints that allow reasonable
      reconstructions of maximum ice sheet extents
      (\citealp{Prest.etal.1968}; \citealp[Fig.~1.12]{Clague.1989};
      \citealp{Duk-Rodkin.1999}; \citealp{Booth.etal.2003};
      \citealp{Dyke.2004}).  As indicated by field evidence from radiocarbon
      dating \citep{Clague.etal.1980, Clague.1985, Clague.1986,
      Porter.Swanson.1998, Menounos.etal.2008}, cosmogenic exposure dating
      \citep{Stroeven.etal.2010, Stroeven.etal.2014, Margold.etal.2014},
      bedrock deformation in response to former ice loads
      \citep{Clague.James.2002, Clague.etal.2005}, and offshore sedimentary
      records \citep{Cosma.etal.2008, Davies.etal.2011}, the LGM Cordilleran
      ice sheet extent was short-lived.  However, former ice thicknesses
      and, therefore, the ice sheet's contribution to the LGM sea level
      lowstand \citep{Carlson.Clark.2012, Clark.Mix.2002} remain uncertain.

      Our understanding of the Cordilleran glaciation history prior to the
      LGM is even more fragmentary \citep{Barendregt.Irving.1998,
      Kleman.etal.2010, Rutter.etal.2012}, although it is clear that the
      Pleistocene maximum extent of the Cordilleran ice sheet predates the
      last glacial cycle \citep{Hidy.etal.2013}.  In parts of the Yukon
      Territory and Alaska, and in the Puget Lowland, the distribution of
      tills \citep{Turner.etal.2013, Troost.2014} and dated glacial erratics
      \citep{Ward.etal.2007, Ward.etal.2008, Briner.Kaufman.2008,
      Stroeven.etal.2010, Stroeven.etal.2014} indicate an extensive Marine
      Oxygen Isotope Stage (MIS)~4 glaciation.  Landforms in the interior
      regions include flow sets that are likely older than the LGM
      \citep[Fig.~2]{Kleman.etal.2010}, but their absolute age remains
      uncertain.

      In contrast, evidence for the deglaciation history of the Cordilleran
      ice sheet since the LGM is considerable, albeit mostly at a~regional
      scale.  Geomorphological evidence from south-central British Columbia
      indicates a~rapid deglaciation, including an early emergence of
      elevated areas while thin, stagnant ice still covered the surrounding
      lowlands \citep{Fulton.1967, Fulton.1991, Margold.etal.2011,
      Margold.etal.2013a}.  This model, although credible, may not apply in
      all areas of the Cordilleran ice sheet \citep{Margold.etal.2013}.
      Although solid evidence for late-glacial glacier re-advances have been
      found in the Coast, Columbia and Rocky mountains
      \citep{Clague.etal.1997, Friele.Clague.2002, Friele.Clague.2002a,
      Kovanen.2002, Kovanen.Easterbrook.2002, Lakeman.etal.2008,
      Menounos.etal.2008}, it appears to be sparser than for formerly
      glaciated regions surrounding the North Atlantic
      \citep[e.g.,][]{Sissons.1979, Lundqvist.1987, Ivy-Ochs.etal.1999,
      Stea.etal.2011}.  Nevertheless, recent oxygen isotope measurements
      from Gulf of Alaska sediments reveal a~climatic evolution highly
      correlated to that of Greenland during this period, including
      a~distinct Late Glacial cold reversal between 14.1 and 11.7\,\unit{ka}
      \citep{Praetorius.Mix.2014}.

      In general, the topographic complexity of the North American
      Cordillera and its effect on glacial history have inhibited the
      reconstruction of ice sheet-wide glacial advance and retreat patterns
      such as those available for the Fennoscandian and Laurentide ice
      sheets \citep{Boulton.etal.2001, Dyke.Prest.1987, Dyke.etal.2003,
      Kleman.etal.1997, Kleman.etal.2010, Stroeven.etal.inreview}.  Here, we
      use a~numerical ice sheet model \citep{PISM-authors.2015}, calibrated
      against field-based evidence, to perform a~quantitative reconstruction
      of the Cordilleran ice sheet evolution through the last glacial cycle,
      and analyse some of the long-standing questions related to its
      evolution:
\begin{itemize}
  \item How much ice was locked in the Cordilleran ice sheet during the LGM?
  \item What was the scale of glaciation prior to the LGM?
  \item Which were the primary dispersal centres? Do they reflect stable or
    ephemeral configurations?
  \item How rapid was the last deglaciation? Did it include Late Glacial
    standstills or readvances?
\end{itemize}
      Although numerical ice sheet modelling has been established as
      a~useful tool to improve our understanding of the Cordilleran ice
      sheet (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991};
      \citealp{Marshall.etal.2000}), the ubiquitously mountainous topography
      of the region has presented two major challenges to its application.
      First, only recent developments in numerical ice sheet models and
      underlying scientific computing tools \citep{Bueler.Brown.2009,
      Balay.etal.2015} have allowed for high-resolution numerical modelling
      of glaciers and ice sheets on mountainous terrain over millennial time
      scales \citep[e.g.,][]{Golledge.etal.2012}. Second, the complex
      topography of the North American Cordillera also induces strong
      geographic variations in temperature and precipitation, thus requiring
      the use of high-resolution climate forcing fields as an input to an
      ice sheet model \citep{Seguinot.etal.2014}. However, evolving climate
      conditions over the last glacial cycle are subject to considerable
      uncertainty and still lie beyond the computational reach of atmosphere
      circulation models.

      Our palaeo-climate forcing therefore includes spatial temperature and
      precipitation grids derived from a~present-day atmospheric reanalysis
      \citep{Mesinger.etal.2006} that includes the steep precipitation
      gradients previously identified as necessary to model the LGM extent
      of the Cordilleran ice sheet in agreement with its geological imprint
      \citep{Seguinot.etal.2014}. To mimic climate evolution through the
      last glacial cycle, these grids are simply supplemented by lapse-rate
      corrections and temperature offset time series. The latter are
      obtained by scaling six different palaeo-temperature reconstructions
      from proxy records around the globe, including two oxygen isotope
      records from Greenland ice cores \citep{Dansgaard.etal.1993,
      Andersen.etal.2004}, two oxygen isotope records from Antarctic ice
      cores \citep{Petit.etal.1999,Jouzel.etal.2007}, and two alkenone
      unsaturation index records from Northwest Pacific ocean sediment cores
      \citep{Herbert.etal.2001}. We then proceed to compare the model output
      to geological evidence and discuss the timing and extent of glaciation
      and the patterns of deglaciation, based on which we use the
      applicability of different records to modelling the history of the
      Cordilleran ice sheet.

\section{Model setup}
\label{sec:model}

\subsection{Overview}
\label{sec:overview}%

      The simulations presented here were run using the Parallel Ice Sheet
      Model (PISM, development version~8ff7cbe), an open source, finite
      difference, shallow ice sheet model \citep{PISM-authors.2015}. The
      model requires input on basal topography, sea level, geothermal heat
      flux and climate forcing. It computes the evolution of ice extent and
      thickness over time, the thermal and dynamic states of the ice sheet,
      and the associated lithospheric response.

      Basal topography is derived from the ETOPO1 combined topography and
      bathymetry dataset with a~resolution of 1\,arc-min
      \citep{Amante.Eakins.2009}. Sea level is lowered as a~function of time
      based on the Spectral Mapping Project
      \citep[SPECMAP,][]{Imbrie.etal.1989} time scale. Geothermal heat flux
      is applied as a~constant value of 70\,\unit{mW\,m^{-2}} at
      3\,\unit{km} depth (Sect.~\ref{sec:icedyn}). Surface mass balance is
      computed using a~positive degree-day (PDD) model
      (Sect.~\ref{sec:surface}). Climate forcing is provided by a~monthly
      climatology averaged from 1979 to 2000 from the North American
      Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006}, perturbated
      by time-dependent offsets and lapse-rate temperature corrections
      (Sect.~\ref{sec:atm}).

      Each simulation starts from assumed ice-free conditions at
      12\,0000\,\unit{years} ago (120\,\unit{ka}), and runs to the
      present. Our modelling domain of 1500 by 3000\,\unit{km} encompasses
      the entire area covered by the Cordilleran ice sheet at the LGM
      (Fig.~\ref{fig:locmap}). The simulations were run on two distinct
      grids, using a~lower horizontal resolution of 10\,\unit{km}, and
      a~higher horizontal resolution of 5\,\unit{km}. These computations
      were performed on 16 to 128 computing cores at the Swedish National
      Supercomputing Centre.

\subsection{Ice thermodynamics}
\label{sec:icedyn}

      Ice sheet dynamics are typically modelled using a~combination of
      internal deformation and basal sliding. PISM is a~shallow ice sheet
      model, which implies that the balance of stresses is approximated
      based on their predominant components. The Shallow Shelf Approximation
      (SSA) is used as a ``sliding law'' for the Shallow Ice Approximation
      (SIA) by adding velocity solutions of the two approximations
      \citep[Eqs.~7--9 and 15]{Bueler.Brown.2009, Winkelmann.etal.2011}.

      Ice rheology depends on temperature and water content through an
      enthalpy formulation \citep{Aschwanden.etal.2012}. Surface air
      temperature derived from the climate forcing (Sect.~\ref{sec:atm})
      provides the upper boundary condition to the ice enthalpy
      model. Temperature is computed subglacially to a~depth of
      3\,\unit{km}, where it is conditioned by a~lower boundary geothermal
      heat flux of 70\,\unit{mW\,m^{-2}}. Although this uniform value does
      not account for the high spatial geothermal variability in the region
      \citep{Blackwell.Richards.2004}, it is, on average, representative of
      available heat flow measurements. In the low-resolution simulations,
      the vertical grid consists of 31~temperature layers in the bedrock and
      up to 51~enthalpy layers in the ice sheet, corresponding to a~vertical
      resolution of 100\,\unit{m}. The high-resolution simulations
      61~bedrock layers and up to 101~ice layers with a~vertical resolution
      of 50\,\unit{m}.

      A~pseudo-plastic sliding law,
\begin{align}
&\label{eqn:pseudoplastic}
    \vec{\tau}_{\mathrm{b}} = -\tau_{\mathrm{c}} \frac{\vec{v}_{\mathrm{b}}}{{v_{\text{th}}}^q\,|\vec{v}_{\mathrm{b}}|^{1-q}} \,,
\end{align}
      relates the bed-parallel shear stresses, $\vec{\tau}_{\mathrm{b}}$, to
      the sliding velocity, $\vec{v}_{\mathrm{b}}$ (Table~\ref{tab:params}).
      The yield stress, $\tau_{\mathrm{c}}$, is modelled using the
      Mohr--Coulomb criterion,
\begin{align}
&\tau_{\mathrm{c}} = c_0 + N\,\tan{\phi} \,,
\end{align}
      where cohesion, $c_0$, is assumed to be zero. The friction angle,
      $\phi$, varies from 15 to 45\unit{\degree} as a~piecewise-linear
      function of modern bed elevation, with the lowest value occuring below
      modern sea level (0\,\unit{m\,a.s.l.}) and the highest value occuring
      above the generalised elevation of the highest shorelines
      \citep[200\,\unit{m\,a.s.l.},][Fig.~5]{Clague.1981}, thus accounting
      for a~weakening of till associated with the presence of marine
      sediments (cf. \citealp{Martin.etal.2011};
      \citealp[Supplement]{Aschwanden.etal.2013};
      \citealp{PISM-authors.2015}). Effective pressure, $N$, is related to
      the ice overburden stress, $\rho gh$, and the modelled amount of
      subglacial water, using a~formula derived from laboratory experiments
      with till extracted from the base of Ice Stream B in West Antarctica
      \citep{Tulaczyk.etal.2000, Bueler.Pelt.2015},
\begin{align}
&N = \delta \rho gh \, 10^{(e_0/C_{\mathrm{c}}) (1 - (W/W_{\text{max}}))} \,,
\end{align}
      where $\delta$ is chosen as 0.02, $e_0$ is a~measured reference void
      ratio and $C_{\mathrm{c}}$ is a~measured compressibility coefficient
      (Table~\ref{tab:params}). The amount of water at the base, $W$, varies
      from zero to $W_{\text{max}}=2$\,\unit{m}, a~threshold above which
      additional melt water is assumed to drain off
      instantaneously. Finally, the bedrock topography responds to ice load
      following a~bedrock deformation model that includes local isostasy,
      elastic lithosphere flexure and viscous astenosphere deformation in an
      infinite half-space \citep{Lingle.Clark.1985,Bueler.etal.2007}.
      A~relatively low viscosity value of $\nu_{m} =
      1\times10^{19}$\,\unit{Pa\,s} is used for the astenosphere
      (Table~\ref{tab:params}) in accordance with the results from regional
      glacial isostatic adjustment modelling at the northern Cascadia
      subduction zone \citep{James.etal.2009}.

      Ice shelf calving is computed using a~double criterion. First,
      a~physically-realistic calving flux is computed based on eigenvalues
      of the horizontal strain rate tensor \citep{Winkelmann.etal.2011,
      Levermann.etal.2012}. This allows floating ice to advance in confined
      embayments, but prevents the formation of extensive ice shelves in the
      open ocean.  Second, floating ice thinner than 50\,\unit{m} is
      systematically calved off. A~subgrid scheme by
      \citet{Albrecht.etal.2011} allows for a~continuous migration of the
      calving front. This formulation of calving has been applied to the
      Antarctic ice sheet and has shown to produce a~realistic calving front
      positon for many of the present-day ice shelves
      \citep{Martin.etal.2011}.

\subsection{Surface mass balance}
\label{sec:surface}

      Ice surface accumulation and ablation are computed from monthly mean
      near-surface air temperature, $T_{\mathrm{m}}$, monthly standard deviation of
      near-surface air temperature, $\sigma$, and monthly precipitation,
      $P_{\mathrm{m}}$, using a~temperature-index model \citep[e.g.,][]{Hock.2003}.
      Accumulation is equal to precipitation when air temperatures are below
      0\,\unit{{\degree}C}, and decreases to zero linearly with temperatures
      between 0 and 2\,\unit{{\degree}C}. Ablation is computed from PDD,
      defined as an integral of temperatures above 0\,\unit{{\degree}C} in
      one year.

      The PDD computation accounts for stochastic temperature variations by
      assuming a~normal temperature distribution of standard deviation
      $\sigma$ aroung the expected value $T_{\mathrm{m}}$. It is expressed
      by an error-function formulation \citep{Calov.Greve.2005},
\begin{align}
&\label{eqn:calovgreve}
    {\text{PDD}} = \int_{t_1}^{t_2} \mathrm{d}t
        \left[\frac{\sigma}{\sqrt{2\pi}}
                \exp\left({-\frac{T_{\mathrm{m}}^2}{2\sigma^2}}\right)
              + \frac{T_{\mathrm{m}}}{2} \, {\text{erfc}}
                \left(-\frac{T_{\mathrm{m}}}{\sqrt{2}\sigma}\right)\right] \,,
\end{align}
      which is numerically approximated using week-long sub-intervals. In
      order to account for the effects of spatial and seasonal variations of
      temperature variability \citep{Seguinot.2013}, $\sigma$ is computed
      from NARR daily temperature values from 1979 to 2000
      \citep{Mesinger.etal.2006}, including variability associated with the
      seasonal cycle (Fig.~\ref{fig:atm}).  Degree-day factors for snow and
      ice melt are derived from mass-balance measurements on contemporary
      glaciers from the Coast Mountains and Rocky Mountains in British
      Columbia \citep[Table~\ref{tab:params};][]{Shea.etal.2009}.

\subsection{Climate forcing}
\label{sec:atm}%

      Climate forcing driving ice sheet simulations consists of
      a~present-day monthly climatology, $\{T_{\mathrm{m}0}, P_{\mathrm{m}0}\}$, where
      temperatures are modified by offset time series,
      ${\Delta}T_{\text{TS}}$, and lapse-rate corrections,
      ${\Delta}T_{\text{LR}}$:
\begin{align}
&T_{\mathrm{m}}(t, x, y) = T_{\mathrm{m}0}(x, y) + {\Delta}T_{\text{TS}}(t)
                    + {\Delta}T_{\text{LR}}(t, x, y) \,, \\
&    P_{\mathrm{m}}(t, x, y) = P_{\mathrm{m}0}(x, y) \,.
\end{align}
      The present-day monthly climatology was computed from near-surface air
      temperature and precipitation rate fields from the NARR, averaged from
      1979 to 2000. Modern climate of the North American Cordillera is
      characterised by strong geographic variations in temperature
      seasonality, timing of the maximum annual precipitation, and daily
      temperature variability (Fig.~\ref{fig:atm}).  The use of NARR is
      motivated by the need for an accurate, high-resolution precipitation
      forcing, as identified in a~previous sensitivity study
      \citep{Seguinot.etal.2014}.

      Temperature offset time-series, ${\Delta}T_{\text{TS}}$, are derived
      from palaeo-temperature proxy records from the Greenland Ice Core
      Project \citep[GRIP,][]{Dansgaard.etal.1993}, the North Greenland Ice
      Core Project \citep[NGRIP,][]{Andersen.etal.2004}, the European
      Project for Ice Coring in Antarctica \citep[EPICA,][]
      {Jouzel.etal.2007}, the Vostok ice core \citep{Petit.etal.1999}, and
      Ocean Drilling Program (ODP) sites 1012 and 1020, both located off the
      coast of California \citep{Herbert.etal.2001}. Palaeo-temperature
      anomalies from the GRIP and NGRIP records were calculated from oxygen
      isotope (\chem{\delta^{18}O}) measurements using a~quadratic equation
      \citep{Johnsen.etal.1995},
\begin{align}
{\Delta}T_{\text{TS}}(t) ={~}&-11.88 [\chem{\delta^{18}O}(t)
                                -\chem{\delta^{18}O}(0)] \nonumber \\
                        &-0.1925[\chem{\delta^{18}O}(t)^2
                                 -\chem{\delta^{18}O}(0)^2] \,,
\end{align}
      while temperature reconstructions from Antarctic and oceanic cores
      were provided as such. All records were scaled linearly
      (Table~\ref{tab:records}) in order to simulate comparable ice extents
      at the LGM (Table~\ref{tab:extrema}) and realistic outlines
      (Fig.~\ref{fig:snapshots}).

      Finally, lapse-rate corrections, ${\Delta}T_{\text{LR}}$, are computed
      as a~function of ice surface elevation, $s$, using the NARR surface
      geopotential height invariant field as a~reference topography,
      $b_{\text{ref}}$:
\begin{align}
{\Delta}T_{\text{LR}}(t, x, y) &= -\gamma [s(t, x, y)-b_{\text{ref}}] \\
                            &= -\gamma [h(t, x, y)+b(t, x, y)-b_{\text{ref}}],
\end{align}
      thus accounting for the evolution of ice thickness, ${h=s-b}$, on the
      one hand, and for differences between the basal topography of the ice
      flow model, $b$, and the NARR reference topography, $b_{\text{ref}}$,
      on the other hand. All simulations use an annual temperature lapse
      rate of $\gamma = 6\,\unit{K\,km^{-1}}$.  In the rest of this paper,
      we refer to different model runs by the name of the proxy record used
      for the palaeo-temperature forcing.

\section{Sensitivity to climate forcing time-series}
\label{sec:results}

\subsection{Evolution of ice volume}

%%%

      Despite large differences in the input climate forcing
      (Fig.~\ref{fig:timeseries}, upper panel), model output presents
      consistent features that can be observed across the range of forcing
      data used. In all simulations, modelled ice volumes remain relatively
      low during most of the glacial cycle, except during two major glacial
      events which occur between 61.9 and 56.5\,\unit{ka} during MIS~4, and
      between 23.2 and 16.8\,\unit{ka} during MIS~2
      (Fig.~\ref{fig:timeseries}, lower panel). An ice volume minimum is
      consistently reached between 53.0 and 41.3\,\unit{ka} during
      MIS~3. However, the magnitude and precise timing of these three events
      depend significantly on the choice of proxy record used to derive
      a~time-dependent climate forcing (Table~\ref{tab:extrema}).

      Simulations forced by the Greenland ice core palaeo-temperature
      records (GRIP, NGRIP) produce the highest variability in modelled ice
      volume throughout the last glacial cycle. In contrast, simulations
      driven by oceanic (ODP~1012, ODP~1020) and Antarctic (EPICA, Vostok)
      palaeo-temperature records generally result in lower ice volume
      variability throughout the simulation length, resulting in lower
      modelled ice volumes during MIS~4 and larger ice volumes during
      MIS~3. The NGRIP climate forcing is the only one that results in
      a~larger ice volume during MIS~4 than during MIS~2.

      While simulations driven by the GRIP and the two Antarctic
      palaeo-temperature records attain a~last ice volume maximum between
      19.1 and 16.8\,\unit{ka}, those informed by the NGRIP and the two
      oceanic palaeo-temperature records attain their maximum ice volumes
      thousands of years earlier. Moreover, the ODP~1012 run yields a~rapid
      deglaciation of the modelled area prior to 17\,\unit{ka}. The ODP~1020
      simulation predicts an early maximum in ice volume at 23.2\,\unit{ka},
      followed by slower deglaciation than modelled using the other
      palaeo-temperature records.  Finally, whereas model runs forced by the
      Antarctic palaeo-temperature records result in a~rapid and
      uninterrupted deglaciation after the LGM, the simulation driven by the
      GRIP palaeo-temperature record also results in a~rapid deglaciation
      but in three steps, separated by two periods of ice sheet regrowth
      (Fig.~\ref{fig:timeseries}).

   \subsection{Extreme configurations}

      Despite large differences in the timing of attained volume extrema
      (Table~\ref{tab:extrema}), all model runs show relatively consistent
      patterns of glaciation. During MIS~4, all simulations produce an
      extensive ice sheet, covering an area of at least half of that
      attained during MIS~2 (Table~\ref{tab:extrema};
      Fig.~\ref{fig:snapshots}, upper panels).  Corresponding maximum ice
      volumes also differ significantly between model runs, and vary between
      3.84 and~8.84\,\unit{m} sea level equivalents (m\,s.l.e.;
      Table~\ref{tab:extrema}).

      In the MIS~3 ice volume minimum reconstructions, a~central ice cap
      persists over the Skeena Mountains (Fig.~\ref{fig:snapshots}, middle
      panels). Although this ice cap is present in all simulations, its
      dimensions depend sensitively on the choice of the applied
      palaeo-temperature record. Modelled ice volume minima spread over
      a~wide range between 1.69 and
      2.88\,\unit{m}\,s.l.e. (Table~\ref{tab:extrema}).

      Modelled ice sheet geometries during the LGM (MIS~2;
      Fig.~\ref{fig:snapshots}, lower panels) invariably include
      a~ca. 1500\,\unit{km}-long central divide above 3000\,\unit{m\,a.s.l.}
      located along the spine of the Rocky Mountains. Although the
      similarity of modelled ice extents is a~direct result of the choice of
      scaling factors applied to different palaeo-temperature proxy records
      (Table~\ref{tab:records}), it is interesting to note that modelled
      maximum ice volumes also fall within a~tight range of 8.40 to
      8.91\,\unit{m}\,s.l.e.  (Table~\ref{tab:extrema}).

\section{Comparison with the geologic record}
\label{sec:discussion}

      Large variations in the model responses to evolving climate forcing
      reveal its sensitivity to the choice of palaeo-temperature proxy
      record. To distinguish between different records, geological evidence
      of former glaciations provide a~basis for validation of our runs,
      while the results from numerical modelling can perhaps help to analyse
      some of the complexity of this evidence.  In this section, we compare
      model outputs to the geologic record, in terms of timing and
      configuration of the maximum stages, location and lifetime of major
      nucleation centres, and patterns of ice retreat during the last
      deglaciation.

\subsection{Glacial maxima}

\subsubsection{Timing of glaciation}
\label{sec:timing}

      Independently of the palaeo-temperature records used to force the ice
      sheet model, our simulations consistently produce two glacial maxima
      during the last glacial cycle. The first maximum configuration is
      obtained during MIS~4 (61.9--56.5\,\unit{ka}) and the second during
      MIS~2 (23.2--16.8\,\unit{ka}; Figs.~\ref{fig:timeseries},
      \ref{fig:snapshots}; Table~\ref{tab:extrema}). These events broadly
      correspond in timing to the Gladstone (MIS~4) and McConnell (MIS~2)
      glaciations documented by geological evidence for the northern sector
      of the Cordilleran ice sheet \citep{Duk-Rodkin.etal.1996,
      Ward.etal.2007, Stroeven.etal.2010, Stroeven.etal.2014}, and to the
      Fraser Glaciation (MIS~2) documented for its southern sector
      \citep{Porter.Swanson.1998, Margold.etal.2014}.  There is
      stratigraphical evidence for an MIS~4 glaciation in British Columbia
      \citep{Clague.Ward.2011} and in the Puget Lowland \citep{Troost.2014},
      but their extent and timing are still highly conjectural
      \citep[perhaps MIS~4 or early MIS~3; e.g.,][]{Cosma.etal.2008}.

      The exact timing of modelled MIS~2 maximum ice volume depends strongly
      on the choice of applied palaeo-temperature record, which allows for
      a~more in-depth comparison with geological evidence for the timing of
      maximum Cordilleran ice sheet extent. In the Puget Lowland
      (Fig.~\ref{fig:locmap}), the LGM advance of the southern Cordilleran
      ice sheet margin has been constrained by radiocarbon dating on wood
      between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,ka}
      \citep{Porter.Swanson.1998}. These dates are consistent with
      radiocarbon dates from the offshore sedimentary record, which reveals
      an increase of glaciomarine sedimentation between 19.5 and
      16.2\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008,
      Taylor.etal.2014}.  Radiocarbon dating of the northern Cordilleran ice
      sheet margin is much less constrained but straddles presented
      constraints from the southern margin. However, cosmogenic exposure
      dating places the timing of maximum CIS extent during the McConnell
      glaciation close to 17\,\unit{\chem{^{10}Be}\,ka}
      \citep{Stroeven.etal.2010, Stroeven.etal.2014}. A~sharp transition in
      the sediment record of the Gulf of Alaska indicates a~retreat of
      regional outlet glaciers onto land at
      14.8\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Davies.etal.2011}.

      Among the simulations presented here, only those forced with the GRIP,
      EPICA and Vostok palaeo-temperature records yield Cordilleran ice
      sheet maximum extents that may be compatible with these field
      constraints (Fig.~\ref{fig:timeseries}, lower panel;
      Table~\ref{tab:extrema}).  Simulations driven by the NGRIP, ODP~1012
      and ODP~1020 palaeo-temperature records, on the contrary, yield MIS~2
      maximum Cordilleran ice sheet volumes that pre-date field-based
      constraints by several thousands of years (about 6, 6 and 4\,\unit{ka}
      respectively).  Concerning the simulations driven by oceanic records,
      this early deglaciation is caused by an early warming present in the
      alkenone palaeo-temperature reconstructions
      (Fig.~\ref{fig:timeseries}, upper panel;
      \citealp[Fig.~3]{Herbert.etal.2001}). However, this early warming is
      a~local effect, corresponding to a~weakening of the California current
      \citep{Herbert.etal.2001}. The California current, driving cold waters
      southwards along the south-western coast of North America, has been
      shown to have weakened during each peak of global glaciation (in
      SPECMAP) during the past 550\,\unit{ka}, including the LGM, resulting
      in paradoxically warmer sea-surface temperatures at the locations of
      the ODP~1012 and ODP~1020 sites \citep{Herbert.etal.2001}.

      Because most of the marine margin of the Cordilleran ice sheet
      terminated in a~sector of the Pacific Ocean unaffected by variations
      in the California current, it probably remained insensitive to this
      local phenomenum. However, the above paradox illustrates the
      complexity of ice-sheet feedbacks on regional climate, and
      demonstrates that, although located in the neighbourhood of the
      modelling domain, the ODP~1012 and ODP~1020 palaeo-temperature records
      cannot be used as a~realistic forcing to model the Cordilleran ice
      sheet through the last glacial cycle.  Similarly, the simulation using
      the NGRIP palaeo-temperature record depicts an early onset of
      deglaciation (Fig.~\ref{fig:timeseries}) following its last glacial
      volume maximum (22.9\,\unit{ka}, Table~\ref{tab:extrema}) attained
      about 6\,\unit{ka} earlier than dated evidence of the LGM advance.
      There is a~fair agreement between the EPICA and Vostok
      palaeo-temperature records, resulting in only small differences
      between the simulations driven by those records. These differences are
      not subject to further analysis further; instead we focus on
      simulations forced by palaeo-temperature records from the GRIP and
      EPICA ice cores that appear to produce the most realistic
      reconstructions of regional glaciation history, yet bearing
      significant disparities in model output. To allow for a~more detailed
      comparison against the geological record, these two simulations were
      re-run using a~higher-resolution grid (Sect.~\ref{sec:model};
      Fig.~\ref{fig:timeseries}, lower panel, dotted lines).

\subsubsection{Ice configuration during MIS~2}
\label{sec:mis2}

      During maximum glaciation, both simulations position the main
      meridional ice divide over the western flank of the Rocky Mountains
      (Figs.~\ref{fig:snapshots}, lower panels and \ref{fig:mis2}). This
      result appears to contrast with palaeoglaciological reconstructions
      for central and southern British Columbia with ice divides in a~more
      westerly position, over the western margin of the Interior Plateau
      \citep{Ryder.etal.1991, Stumpf.etal.2000, Kleman.etal.2010,
      Clague.Ward.2011, Margold.etal.2013a}.  These indicate that
      a~latitudinal saddle connected ice dispersal centres in the Columbia
      Mountains with the main ice divide \citep{Ryder.etal.1991,
      Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}. A~latitudinal
      saddle does indeed feature in our modelling results, however, in an
      inverse configuration between the main ice divide over the Columbia
      Mountains and a~secondary divide over the southern Coast Mountains
      (Fig.~\ref{fig:mis2}).

      Such deviation from the geological inferences could reflect the fact
      that our model does not include feedback mechanisms between ice sheet
      topography and the regional climate.  Firstly, during the build-up
      phase preceding the LGM, rapid accumulation over the Coast Mountains
      enhanced the topographic barrier formed by these mountain ranges,
      which likely resulted in a~decrease of precipitation and, therefore,
      a~decrease of accumulation in the interior. Secondly, latent warming
      of the moisture-depleted air parcels flowing over this enhanced
      topography could have resulted in an inflow of potentially warmer air
      over the eastern flank of the ice sheet, increasing melt along the
      advancing margin \citep[cf.][]{Langen.etal.2012}. Because these two
      processes, both with a~tendency to limit ice-sheet growth, are absent
      from our model, the eastern margin of the ice sheet and the position
      of the main meridional ice divide are certainly biased towards the
      east in our simulations \citep{Seguinot.etal.2014}.

      However, field-based palaeoglaciological reconstructions have
      struggled to reconcile the more westerly-centred ice divide in
      south-central British Columbia with evidence in the Rocky Mountains
      and beyond, that the Cordilleran ice sheet invaded the western
      Interior Plains, where it merged with the southwestern margin of the
      Laurentide ice sheet and was deflected to the south
      \citep{Jackson.etal.1997, Bednarski.Smith.2007, Kleman.etal.2010,
      Margold.etal.2013, Margold.etal.2013a}. Ice geometries from our model
      runs do not have this problem, because the position and elevation of
      the ice divide ensure significant ice drainage across the Rocky
      Mountains at the LGM (Fig.~\ref{fig:mis2}).

      During MIS~2, the modelled total ice volume peaks at
      8.01\,\unit{m}\,s.l.e. (19.1\,\unit{ka}) in the GRIP simulation and at
      8.77\,\unit{m}\,s.l.e. (17.2\,\unit{ka}) in the EPICA simulation.

\subsubsection{Ice configuration during MIS~4}
\label{sec:mis4}

      The modelled ice sheet configurations corresponding to ice volume
      maxima during MIS~4 are more sensitive to the choice of atmospheric
      forcing than those corresponding to ice volume maxima during MIS~2
      (Figs.~\ref{fig:snapshots}, upper panels and \ref{fig:mis4}). The GRIP
      simulation (Fig.~\ref{fig:mis4}, left panel) results in a~modelled
      maximum ice sheet extent that closely resembles that obtained during
      MIS~2, with the only major difference of being slightly less extensive
      across northern and eastern sectors. In contrast, the EPICA simulation
      produces a~lower ice volume maximum (Fig.~\ref{fig:timeseries}), which
      translates in the modelled ice sheet geometry into a~significantly
      reduced southern sector, more restricted ice cover in northern and
      eastern sectors, and generally lower ice surface elevations in the
      interior (Fig.~\ref{fig:mis4}, right panel).  Thus, only the GRIP
      simulation can explain the presence of MIS~4 glacial deposits in the
      Puget Lowland \citep{Troost.2014} and that of ice-rafted debris in the
      marine sediment record offshore Vancouver Island at
      ca.~47\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008}.

      During MIS~4, the modelled total ice volume peaks at
      7.19\,\unit{m}\,s.l.e. (57.3\,\unit{ka}) in the GRIP simulation and at
      5.04\,\unit{m}\,s.l.e. (61.9\,\unit{ka}) in the EPICA simulation,
      corresponding to respectively 90 and 57\,\unit{\%} of modelled MIS~2
      ice volumes.

\subsection{Nucleation centres}

\subsubsection{Transient ice sheet states}

      Palaeo-glaciological reconstructions are generally more robust for
      maximum ice sheet extents and late ice sheet configurations than for
      intermediate or minimum ice sheet extents and older ice sheet
      configurations \citep{Kleman.etal.2010}. However, these maximum stages
      are, by nature, extreme configurations, which do not necessarily
      represent the dominant patterns of glaciation throughout the period of
      ice cover \citep{Porter.1989, Kleman.Stroeven.1997, Kleman.etal.2008,
      Kleman.etal.2010}.

      For the Cordilleran ice sheet, geological evidence from radiocarbon
      dating \citep{Clague.etal.1980, Clague.1985, Clague.1986,
      Porter.Swanson.1998, Menounos.etal.2008}, cosmogenic exposure dating
      \citep{Stroeven.etal.2010, Stroeven.etal.2014, Margold.etal.2014},
      bedrock deformation in response to former ice loads
      \citep{Clague.James.2002, Clague.etal.2005}, and offshore sedimentary
      records \citep{Cosma.etal.2008, Davies.etal.2011} indicate that the
      LGM maximum extent was short-lived. To compare this finding to our
      simulations, we use numerical modelling output to compute durations of
      ice cover throughout the last glacial cycle (Fig.~\ref{fig:duration}).

      The resulting maps show that, during most of the glacial cycle,
      modelled ice cover is restricted to disjoint ice caps centred on major
      mountain ranges of the North American Cordillera
      (Fig.~\ref{fig:duration}, blue areas). A~2500\,\unit{km}-long
      continuous expanse of ice, extending from the Alaska Range in the
      northeast to the Rocky Mountains in the southwest, is only in
      operation for at most 32\,\unit{ka}, which is about a~third of the
      timespan of the last glacial cycle (Fig.~\ref{fig:duration}, hatched
      areas). However, except for its margins in the Pacific Ocean and in
      the northern foothills of the Alaska Range, the maximum extent of the
      ice sheet is attained for a~much shorter period of time of only few
      thousand years (Fig.~\ref{fig:duration}, red areas). This result
      illustrates that the maximum extents of the modelled ice sheet during
      MIS~4 and MIS~2 were both short-lived and therefore out of balance
      with contemporary climate.

      A~notable exception to the transient character of the maximum extent
      of Cordilleran ice sheet is the northern slope of the Alaska Range,
      where modelled glaciers are confined to its foothills during the
      entire simulation period (Fig.~\ref{fig:duration}, AR). This apparent
      insensitivity of modelled glacial extent to temperature fluctuations
      results from a~combination of low precipitation, high summer
      temperature and large temperature standard deviation (PDD SD) in the
      plains of the Alaska Interior (Fig.~\ref{fig:atm}) which confines
      glaciation to the foothills of the mountains. This result could
      potentially explain the local distribution of glacial deposits, which
      indicates that glaciers flowing on the northern slope of the Alaska
      Range have remained small throughout the Pleistocene
      \citep{Kaufman.Manley.2004}.

\subsubsection{Major ice-dispersal centres}

      It is generally believed that the Cordilleran ice sheet formed by the
      coalescence of several mountain-centred ice caps
      \citep{Davis.Mathews.1944}.  In our simulations, major ice-dispersal
      centres, visible on the modelled ice cover duration maps
      (Fig.~\ref{fig:duration}), are located over the Coast Mountains (CM),
      the Columbia and Rocky mountains (CRM), the Skeena Mountains (SM), and
      the Selwyn and Mackenzie mountains (SMKM). The Wrangell and Saint
      Elias mountains, heavily glacierized at present, host an ice cap for
      the entire length of both simulations, but that ice cap does not
      appear to be a~major feed to the Cordilleran ice sheet
      (Fig.~\ref{fig:duration}, WSEM).  Although the Coast, Skeena and
      Columbia and Rocky mountains (CM, SM, CRM) are covered by mountain
      glaciers for most of the last glacial cycle, providing durable
      nucleation centres for an ice sheet, this is not the case for the
      Selwyn and Mackenzie mountains (SMKM), where ice cover on the highest
      peaks is limited to a~small fraction of the last glacial cycle. In
      other words, the Selwyn and Mackenzie mountains only appear as
      a~secondary ice-dispersal centre during the coldest periods of the
      last glacial cycle. The Northern Rocky Mountains
      (Fig.~\ref{fig:duration}, NRM) do not act as a~nucleation centre, but
      rather as a~pinning point for the Cordilleran ice sheet margin coming
      from the west.

      Perhaps the most striking feature displayed by the distributions of
      modelled ice cover is the persistence of the Skeena Mountains ice cap
      throughout the entire last glacial period (ca.~100--10\,\unit{ka}) and
      its predominance over the other ice-dispersal centres
      (Figs.~\ref{fig:snapshots} and~\ref{fig:duration}, SM). Regardless of
      the applied forcing, this ice cap appears to survive MIS~3
      (Fig.~\ref{fig:snapshots}, middle panels), and serves as a~nucleation
      centre at the onset of the glacial readvance towards the LGM
      (MIS~2). This situation appear similar to the neighbouring Laurentide
      ice sheet, for which the importance of residual ice for the glacial
      history leading up to the LGM has been illustrated by the MIS~3
      residual ice bodies in northern and eastern Canada as nucleation
      centres for a~much more extensive MIS~2 configuration
      \citep{Kleman.etal.2010}.

      The presence of a~Skeena Mountains ice cap during most of the last
      glacial cycle can be explained by meteorological conditions more
      favourable for ice growth there than elsewhere. In fact, reanalysed
      atmospheric fields used to force the surface mass balance model show
      that high winter precipitations are mainly confined to the western
      slope of the Coast Mountains, except in the centre of the modelling
      domain where they also occur further inland than along other east-west
      transects (Fig.~\ref{fig:atm}).  In fact, along most of the
      north-western coast of North America, coastal mountain ranges form
      a~pronounced topographic barrier for westerly winds, capturing
      atmospheric moisture in the form of orographic precipitation, and
      resulting in arid interior lowlands. However, near the centre of our
      modelling domain, this barrier is less pronounced than elsewhere,
      allowing westerly winds to carry moisture further inland, until it is
      captured by the extensive Skeena Mountains in north-central British
      Columbia, thus resulting in a~more widespread distribution of winter
      precipitation (Fig.~\ref{fig:atm}).

      The modelled total ice volume corresponding to these persistent
      ice-dispersal centres attains a~minimum of
      1.52\,\unit{m}\,s.l.e. (42.9\,\unit{ka}) in the GRIP simulation and of
      2.68\,\unit{m}\,s.l.e. (51.8\,\unit{ka}) in the EPICA simulation,
      corresponding to respectively 19 and 31\,\unit{\%} of the MIS~2 ice
      volumes.

\subsubsection{Erosional imprint on the landscape}

      A~correlation is observed between the modelled duration of warm based
      ice cover (Fig.~\ref{fig:warmbase}) and the degree of glacial
      modification of the landscape (mainly in terms of the development of
      deep glacial valleys and troughs). We find evidence for this on the
      slopes of the Coast Mountains, the the Columbia and Rocky mountains,
      the Wrangell and Saint Elias mountains, and radiating off the Skeena
      Mountains (Figs.~\ref{fig:duration} and \ref{fig:warmbase};
      \citealp[Fig.~2]{Kleman.etal.2010}).  The Skeena Mountains, for
      example, indeed bear a~strong glacial imprint that indicates ice
      drainage in a~system of distinct glacial troughs emanating in a~radial
      pattern from the centre of the mountain range
      \citep[Fig.~2]{Kleman.etal.2010}. We suggest that persistent ice cover
      (Fig.~\ref{fig:duration}) associated with basal ice temperatures at
      the pressure melting point (Figs.~\ref{fig:warmbase}
      and~\ref{fig:warmfrac}) explains the large-scale glacial erosional
      imprint on the landscape. A~well-developed network of glacial valleys
      running to the north-west on the west slope of the Selwyn and
      Mackenzie Mountains (\citealp[Fig.~2]{Kleman.etal.2010};
      \citealp[Fig.~8] {Stroeven.etal.2010}) is modelled to have hosted
      predominantly warm-based ice (Fig.~\ref{fig:warmfrac}). However,
      because it is only glaciated for a~short fraction of the last glacial
      cycle in our simulations (Fig.~\ref{fig:duration}), this perhaps
      indicates that the observed landscape pattern originates from multiple
      glacial cycles and witnesses an increased relative importance of the
      Selwyn and Mackenzie mountains ice dispersal centre
      (Fig.~\ref{fig:duration}, SMKM), prior to the Late Pleistocene
      \citep[cf.][]{Ward.etal.2008, Demuro.etal.2012}.

      The modelled distribution of warm-based ice cover
      (Figs.~\ref{fig:warmbase} and~\ref{fig:warmfrac}) is inevitably
      affected by our assumption of a~constant, 70\,\unit{mW\,m^{-2}}
      geothermal heat flux at 3\,\unit{km} depth
      (Sect.~\ref{sec:icedyn}). However, the Skeena Mountains and the area
      west of the Mackenzie Mountains experience higher-than-average
      geothermal heat flux with measured values of ca.~80 and
      ca.~100\,\unit{mW\,m^{-2}} \citep{Blackwell.Richards.2004}. We can
      therefore expect even longer durations of warm-based ice cover for
      these areas if we were to include spatially variable geothermal
      forcing in our Cordilleran ice sheet simulations.

\subsection{The last deglaciation}

   \subsubsection{Pace and patterns of deglaciation}

      Similarly to other glaciated regions, most glacial traces in the North
      American Cordillera relate to the last few millennia of glaciation,
      because most of the older evidence has been overprinted by warm-based
      ice retreat during the last deglaciation \citep{Kleman.1994,
      Kleman.etal.2010}. From a~numerical modelling perspective, phases of
      glacier retreat are more challenging than phases of growth, because
      they involve more rapid fluctuations of the ice margin, increased flow
      velocities and longitudinal stress gradients, and poorly understood
      hydrological processes. The latter are typically included in the
      models through simple parametrisations
      \citep[e.g.][]{Clason.etal.2012, Clason.etal.2014, Bueler.Pelt.2015},
      if included at all. However, next after the mapping of maximum ice
      sheet extents during MIS~2 and MIS~4 (Sects.~\ref{sec:mis2}
      and~\ref{sec:mis4}), geomorphologically-based reconstructions of
      patterns of ice sheet retreat during the last deglaciation provide the
      second best source of evidence for the validation of our simulations.

      In the North American Cordillera, the presence of lateral meltwater
      channels at high elevation \citep{Margold.etal.2011,
      Margold.etal.2013a, Margold.etal.2014}, and abundant esker systems at
      low elevation \citep{Burke.etal.2012, Burke.etal.2012a,
      Perkins.etal.2013, Margold.etal.2013} indicate that meltwater was
      produced over large portions of the ice sheet surface during
      deglaciation. The southern and northern margins of the Cordilleran ice
      sheet reached their last glacial maximum extent around 17\,\unit{ka}
      \citep[Sect.~\ref{sec:timing};][]{Porter.Swanson.1998,
      Cosma.etal.2008, Stroeven.etal.2010, Stroeven.etal.2014}, which we
      take as a~limiting age for the onset of ice retreat. The timing of
      final deglaciation is less well constrained, but recent cosmogenic
      dates from north-central British Columbia indicate that a~seizable ice
      cap emanating from the central Coast Mountains or the Skeena Mountains
      persisted into the Younger Dryas chronozone, at least until
      12.4\,\unit{ka} \citep{Margold.etal.2014}.

      In our simulations, the timing of peak ice volume during the LGM and
      the pacing of deglaciation depend critically on the choice of climate
      forcing (Table~\ref{tab:extrema}; Figs.~\ref{fig:timeseries}
      and~\ref{fig:deglacseries}). Adopting the EPICA climate forcing yields
      peak ice volume at 17.2\,\unit{ka} and an uninterrupted deglaciation
      until about 9\,\unit{ka} (Fig.~\ref{fig:deglacseries}, lower panel,
      red curves). On the contrary, the simulation driven by the GRIP
      palaeo-temperature record yields peak ice volume at 19.3\,\unit{ka}
      and a~deglaciation interrupted by two phases of regrowth until about
      8\,\unit{ka}. The first interruption occurs between 16.6 and
      14.5\,\unit{ka}, and the second between 13.1 and 11.6\,\unit{ka}
      (Fig.~\ref{fig:deglacseries}, lower panel, blue curve).

      Hence, the two model runs, while similar in overall timing compared to
      the runs with other climate drivers, differ in detail. On the one
      hand, the EPICA run depicts peak glaciation about 2\,\unit{ka} later
      than the GRIP run, in closer agreement with dated maximum extents, and
      shows a~faster, uninterrupted deglaciation which yields sporadic ice
      cover more than 1\,\unit{ka} earlier. On the other hand, the GRIP run
      yields a~deglaciation in three steps, compatible with marine sediment
      sequences offshore Vancouver Island, where the distribution of
      ice-rafted debris indicates an ice margin retreat from the Georgia
      Strait in two phases that are contemporary with warming oceanic
      temperatures from 17.2 to 16.5 and from 15.5 to
      14.0\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Taylor.etal.2014}.

      Modelled patterns of ice sheet retreat are relatively consistent
      between the two simulations (Figs.~\ref{fig:deglacshots}
      and~\ref{fig:deglac}). The southern sector of the modelling domain,
      including the Puget Lowland, the Coast Mountains, the Columbia and and
      Rocky mountains, and the Interior Plateau of British Columbia, becomes
      completely deglaciated by 10\,\unit{ka}, whereas a~significant ice
      cover remains over the Skeena, the Selwyn and Mackenzie, and the
      Wrangell and Saint Elias mountains in the northern sector of the
      modelling domain. After 10\,\unit{ka}, deglaciation continues to
      proceed across the Liard Lowland with a~radial ice margin retreat
      towards the surrounding mountain ranges, consistent with the regional
      melt water record of the last deglaciation
      \citep{Margold.etal.2013}. Remaining ice continues to decay by
      retreating towards the Selwyn and MacKenzie, and Skeena mountains.
      The last remnants of the Cordilleran ice sheet finally disappear from
      the Skeena Mountains at 6.5\,\unit{ka} (GRIP) and 6.1\,\unit{ka}
      (EPICA).

   \subsubsection{Late-glacial readvance}

      The possibility of late glacial readvances in the North American
      Cordillera has been debated for some time \citep{Osborn.Gerloff.1997},
      and locally these have been reconstructed and dated. Radiocarbon-dated
      end moraines in the Fraser and Squamish valleys, off the southern tip
      of the Coast Mountains, indicate consecutive glacier maxima, or
      standstills while in overall retreat, one of which corresponds to the
      Younger Dryas chronozone \citep{Clague.etal.1997, Friele.Clague.2002,
      Friele.Clague.2002a, Kovanen.2002, Kovanen.Easterbrook.2002}. Although
      most of these moraines characterise independent valley glaciers, that
      may have been disconnected from the waning Cordilleran ice sheet, the
      Finlay River area in the Omenica Mountains (Fig.~\ref{fig:deglac}, OM)
      presents a~different kind of evidence.  There, sharp-crested moraines
      indicate a~late-glacial readvance of local alpine glaciers and, more
      importantly, their interaction with larger, lingering remnants of the
      main body of the Cordilleran ice sheet in the valleys
      \citep{Lakeman.etal.2008}. Additional evidence for late-glacial alpine
      glacier readvances includes moraines in the eastern Coast Mountains,
      and the Columbia and Rocky mountains \citep{Osborn.Gerloff.1997,
      Menounos.etal.2008}.

      Although further work is needed to constrain the timing of the
      late-glacial readvance, to assess its extents and geographical
      distribution, and to identify the potential climatic triggers
      \citep{Menounos.etal.2008}, it is interesting to note that the
      simulation driven by the GRIP record produces a~late-glacial readvance
      in the Coast Mountains and in the Columbia and Rocky Mountains
      (Fig.~\ref{fig:deglac}, left panel). In addition to matching the
      location of some local readvances, the GRIP-driven simulation shows
      that a~large remnant of the decaying ice sheet may still have existed
      at the time of this late-glacial readvance.  In contrast, the
      EPICA-driven simulation produces a~nearly-continuous deglaciation with
      only a~tightly restricted late-glacial readvance on the western slopes
      of the Saint Elias and the Coast mountains (Fig.~\ref{fig:deglac},
      right panel).

   \subsubsection{Deglacial flow directions}

      Because a~general tenant in glacial geomorphology is that the majority
      of landforms (lineations and eskers) are part of the deglacial
      envelope \citep[terminology from][]{Kleman.etal.2006}, having been
      formed close inside the retreating margin of ice sheets
      \citep{Boulton.Clark.1990, Kleman.etal.1997, Kleman.etal.2010}, we
      present maps of basal flow directions immediately preceding
      deglaciation or at the time of cessation of sliding inside
      a~cold-based retreating margin (Fig.~\ref{fig:lastflow}). The modelled
      deglacial flow patterns are mostly consistent between the GRIP and
      EPICA simulations. They depict an active ice sheet retreat in the
      peripheral areas, followed by stagnant ice decay in some of the
      interior regions. Several parts of the modelling domain do not
      experience any basal sliding throughout the deglaciation phase
      (Fig.~\ref{fig:lastflow}, hatched areas). This notably includes parts
      of the Interior Plateau in British Columbia, major portions of the
      Alaskan sector of the ice sheet, and a~tortuous ribbon running from
      the Northern Rocky Mountains over the Skeena and Selwyn Mountains and
      into the Mackenzie Mountains.

      Patterns of glacial lineations formed in the northern and southern
      sectors of the Cordilleran ice sheet (\citealp{Prest.etal.1968};
      \citealp[Fig.~1.12]{Clague.1989}; \citealp[Fig.~2]{Kleman.etal.2010})
      show similarities with the patterns of deglacial ice flow from
      numerical modelling (Fig.~\ref{fig:lastflow}). In the northern half of
      the modelling domain, modelled deglacial flow directions depict an
      active downhill flow as the last remnants of the ice sheet retreat
      towards mountain ranges. Converging deglacial flow patterns in the
      Liard Lowland, for instance (Fig.~\ref{fig:lastflow}), closely
      resemble the pattern indicated by glacial lineations
      \citep[Fig.~2]{Margold.etal.2013}.

      On the Interior Plateau of south-central British Columbia, both
      simulations produce a~retreat of the ice margin towards the north-east
      (Fig.~\ref{fig:deglac}), a~pattern which is validated by the
      geomorphological and stratigraphical record for ancient pro-glacial
      lakes dammed by the retreating ice sheet
      \citep{Perkins.Brennand.2014}. However, the two simulations differ in
      the mode of retreat. The GRIP simulation yields an active retreat with
      basal sliding towards the ice margin to the south, whereas the EPICA
      simulation produces negligible basal sliding on the plateau during
      deglaciation (Fig.~\ref{fig:lastflow}). Yet, the Interior Plateau also
      hosts an impressive set of glacial lineations which indicate
      a~substantial eastwards flow component of the Cordilleran ice sheet
      \citep{Prest.etal.1968, Kleman.etal.2010}.  This Interior Plateau
      lineation set could therefore present a~smoking gun for the
      reliability of the presented model results. One explanation for the
      incongruent results could be that the missing feedback mechanisms
      between ice sheet topography and regional climate resulted in
      a~modelled ice divide of the LGM ice sheet beingtoo far to the east
      (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
      \citealp{Seguinot.etal.2014}). A~more westerly-located LGM ice divide
      would certainly result in a~different deglacial flow pattern over the
      Interior Plateau. However, a~more westerly-positioned LGM ice divide
      would certainly be associated with a~thinner ice sheet than that
      modelled here. Decreased ice thickness would not promote warm-based
      conditions but, on the contrary, enlarge the region of negligible
      basal sliding (Fig.~\ref{fig:lastflow}). Thus, a~second explanation
      for the incongruent results could be that the Interior Plateau
      lineation system predates deglaciation ice flow, as perhaps indicated
      by some eskers that appear incompatible with these glacial lineations
      \citep[Fig.~9]{Margold.etal.2013a}. Finally, a~third explanation could
      be that local geothermal heat associated with volcanic activity on the
      Interior Plateau could have triggered the basal sliding
      \citep[cf. Greenland ice sheet;][]{Fahnestock.etal.2001}.

      The modelled deglaciation of the Interior Plateau of British Columbia
      consists of a~rapid northwards retreat (Fig.~\ref{fig:deglac}) of
      southwards-flowing non-sliding ice lobes (Fig.~\ref{fig:lastflow})
      positioned in-between deglaciated mountain ranges
      (Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}). This
      result appears compatible with the prevailing conceptual model of
      deglaciation of central British Columbia, in which mountain ranges
      emerge from the ice before the plateau
      \citep[Fig.~7]{Fulton.1991}. However, due to different topographic and
      climatic conditions, our simulations produce different deglaciation
      patterns in the northern half of the model domain, indicating that
      this conceptual model may not be applied to the entire area formerly
      covered by the Cordilleran ice sheet.

\conclusions
\label{sec:concl}

      Numerical simulations of the Cordilleran ice sheet through the last
      glacial cycle presented in this study consistently produce two glacial
      maxima during MIS~4 (61.9--56.5\,\unit{ka}, 3.8--8.8\,\unit{m}\,s.l.e.)
      and MIS~2 (23.2--16.8\,\unit{ka}, 8.40--8.91\,\unit{m}\,s.l.e.), two
      periods corresponding to documented extensive glaciations. This result
      is independent of the palaeo-temperature record used among the six
      selected for this study, and thus can be regarded as a~robust model
      output, which broadly matches geological evidence. However, the timing
      of the two glaciation peaks depends sensitively on which climate
      record is used to drive the model. The timing of the LGM is best
      reproduced by the EPICA and Vostok Antarctic ice core records. It
      occurs about 2\,\unit{ka} too early in the simulation forced by the
      GRIP ice core record, and occurs even earlier in all other
      simulations. The mismatch is largest for the two Northwest Pacific ODP
      palaeo-temperature records, which are affected by the weakening of the
      California current during the LGM.

      In all simulations presented here, ice cover is limited to disjoint
      mountain ice caps during most of the glacial cycle. The most
      persistent nucleation centres are located in the Coast Mountains, the
      Columbia and Rocky mountains, the Selwyn and Mackenzie mountains, and
      most importantly, in the Skeena Mountains. Throughout the modelled
      last glacial cycle, the Skeena Mountains host an ice cap which appears
      to be fed by the moisture intruding inland from the west through
      a~topographic breach in the Coast Mountains. The Skeena ice cap acts
      as the main nucleation centre for the glacial reavance towards the LGM
      configuration. As indicated by persistent, warm-based ice in the
      model, this ice cap perhaps explains the distinct glacial erosional
      imprint observed on the landscape of the Skeena Mountains.

      During deglaciation, none of the climate records used can be selected
      as producing an optimal agreement between the model results and the
      geological evidence. Although the EPICA-driven simulation yields the
      most realistic timing of the LGM and, therefore, start of
      deglaciation, only the GRIP-driven simulation produces late glacial
      readvances in areas where these have been documented. Nonetheless, the
      patterns of ice sheet retreat are consistent between the two
      simulations, and show a~rapid deglaciation of the southern sector of
      the ice sheet, including a~rapid northwards retreat across the
      Interior Plateau of central British Columbia. The GRIP-driven
      simulation then produces a~late-glacial readvance of local ice caps
      and of the main body of the decaying Cordilleran ice sheet primarily
      in the Coast and the Columbia and Rocky Mountains.  In both
      simulations, this is followed by an opening of the Liard Lowland, and
      a~final retreat of the remaining ice caps towards the Selwyn and,
      finally, the Skeena mountains, which hosts the last remnant of the ice
      sheet during the middle Holocene (6.6--6.2\,\unit{ka}). Our results
      identify the Skeena Mountains as a~key area to understanding glacial
      dynamics of the Cordilleran ice sheet, highlighting the need for
      further geological investigation of this region.

\Supplementary{zip}

\authorcontribution{%
      J.~Seguinot ran the simulations; I.~Rogozhina guided experiment
      design; A.~P.~Stroeven, M.~Margold and J.~Kleman took part in the
      interpretation and comparison of model results against geological
      evidence.  All authors contributed to the text.}

\begin{acknowledgements}
  Foremost, we would like to thank Shawn Marshall for providing
  a~detailed, constructive analysis of this study during J.~Seguinot's
  PhD defence (September 2014). His comments were used to improve the
  model set-up.  We are very thankful to Constantine Khroulev, Ed
  Bueler, and Andy Aschwanden for providing constant help and
  development with PISM. This work was supported by the Swedish
  Research Council~(VR) grant no.~2008-3449 to A.~P.~Stroeven and by
  the German Academic Exchange Service~(DAAD) grant no.~50015537 and
  a~Knut and Alice Wallenberg Foundation grant to
  J.~Seguinot. Computer resources were provided by the Swedish
  National Infrastructure for Computing (SNIC) allocation
  no.~2013/1-159 and~2014/1-159 to A.~P.~Stroeven at the National
  Supercomputing Center (NSC).
\end{acknowledgements}

\begin{thebibliography}{114}

\bibitem[{Albrecht et~al.(2011)Albrecht, Martin, Haseloff, Winkelmann, and
  Levermann}]{Albrecht.etal.2011}
 Albrecht,~T., Martin,~M., Haseloff,~M., Winkelmann,~R., and Levermann,~A.: Parameterization for subgrid-scale motion of ice-shelf calving fronts, The Cryosphere, 5, 35--44,
doi:\href{http://dx.doi.org/10.5194/tc-5-35-2011}{10.5194/tc-5-35-2011}, 2011.

\bibitem[{Amante and Eakins(2009)}]{Amante.Eakins.2009}
Amante,~C. and Eakins,~B.~W.: ETOPO1 1 arc-minute global relief model: procedures, data sources and analysis, NOAA technical memorandum NESDIS NGDC-24, Natl. Geophys. Data Center, NOAA, Boulder, CO,
doi:\href{http://dx.doi.org/10.7289/V5C8276M}{10.7289/V5C8276M}, 2009.

\bibitem[{Andersen et~al.(2004)Andersen, Azuma, Barnola, Bigler, Biscaye, Caillon, Chappellaz, Clausen, Dahl-Jensen, Fischer, Fl\"uckiger, Fritzsche, Fujii, Goto-Azuma, Gr{\o}nvold, Gundestrup, Hansson, Huber, Hvidberg, Johnsen, Jonsell, Jouzel, Kipfstuhl, Landais, Leuenberger, Lorrain, Masson-Delmotte, Miller, Motoyama, Narita, Popp, Rasmussen, Raynaud, Rothlisberger, Ruth, Samyn, Schwander, Shoji, Siggard-Andersen, Steffensen, Stocker, Sveinbj\"ornsd\'ottir, Svensson, Takata, Tison, Thorsteinsson, Watanabe, Wilhelms, and White}]{Andersen.etal.2004}
Andersen,~K.~K., Azuma,~N., Barnola,~J.-M., Bigler,~M., Biscaye,~P., Caillon,~N., Chappellaz,~J., Clausen,~H.~B., Dahl-Jensen,~D., Fischer,~H., Fl\"uckiger,~J., Fritzsche,~D., Fujii,~Y., Goto-Azuma,~K., Gr{\o}nvold,~K., Gundestrup,~N.~S., Hansson,~M., Huber,~C., Hvidberg,~C.~S., Johnsen,~S.~J., Jonsell,~U., Jouzel,~J., Kipfstuhl,~S., Landais,~A., Leuenberger,~M., Lorrain,~R., Masson-Delmotte,~V., Miller,~H., Motoyama,~H., Narita,~H., Popp,~T., Rasmussen,~S.~O., Raynaud,~D., Rothlisberger,~R., Ruth,~U., Samyn,~D., Schwander,~J., Shoji,~H., Siggard-Andersen,~M.-L., Steffensen,~J.~P., Stocker,~T., Sveinbj\"ornsd\'ottir,~A.~E., Svensson,~A., Takata,~M., Tison,~J.-L., Thorsteinsson,~T., Watanabe,~O., Wilhelms,~F., and White,~J.~W.~C.: High-resolution record of Northern Hemisphere climate extending into the last interglacial period, Nature, 431, 147--151,
doi:\href{http://dx.doi.org/10.1038/nature02805}{10.1038/nature02805}, data archived at the World Data Center for Paleoclimatology, Boulder, Colorado, USA, 2004.

\bibitem[{Aschwanden et~al.(2012)Aschwanden, Bueler, Khroulev, and Blatter}]{Aschwanden.etal.2012}
Aschwanden,~A., Bueler,~E., Khroulev,~C., and Blatter,~H.: An enthalpy formulation for glaciers and ice sheets,~J. Glaciol., 58, 441--457,
doi:\href{http://dx.doi.org/10.3189/2012JoG11J088}{10.3189/2012JoG11J088}, 2012.

\bibitem[{Aschwanden et~al.(2013)Aschwanden, A{\dh}algeirsd\'{o}ttir, and Khroulev}]{Aschwanden.etal.2013}
 Aschwanden,~A., A{\dh}algeirsd\'{o}ttir,~G., and Khroulev,~C.: Hindcasting to measure ice sheet model sensitivity to   initial states, The Cryosphere, 7, 1083--1093,
doi:\href{http://dx.doi.org/10.5194/tc-7-1083-2013}{10.5194/tc-7-1083-2013}, 2013.

\bibitem[{Balay et~al.(2015)Balay, Abhyankar, Adams, Brown, Brune, Buschelman, Eijkhout, Gropp, Kaushik, Knepley, McInnes, Rupp, Smith, and Zhang}]{Balay.etal.2015}
Balay,~S., Abhyankar,~S., Adams,~M.~F., Brown,~J., Brune,~P., Buschelman,~K.,
Eijkhout,~V., Gropp,~W.~D., Kaushik,~D., Knepley,~M.~G., McInnes,~L.~C.,
Rupp,~K., Smith,~B.~F., and Zhang,~H.: {PETS}c {W}eb page, available at:
\url{http://www.mcs.anl.gov/petsc} (last access: 2015), 2015.

\bibitem[{Barendregt and Irving(1998)}]{Barendregt.Irving.1998}
Barendregt,~R.~W. and Irving,~E.: Changes in the extent of North American ice sheets during the late Cenozoic, Can.~J. Earth Sci., 35, 504--509,
doi:\href{http://dx.doi.org/10.1139/e97-126}{10.1139/e97-126}, 1998.

\bibitem[{Bednarski and Smith(2007)}]{Bednarski.Smith.2007}
Bednarski,~J.~M. and Smith,~I.~R.: Laurentide and montane glaciation along the Rocky Mountain Foothills of northeastern British Columbia, Can.~J. Earth Sci., 44, 445--457,
doi:\href{http://dx.doi.org/10.1139/e06-095}{10.1139/e06-095}, 2007.

\bibitem[{Blackwell and Richards(2004)}]{Blackwell.Richards.2004}
Blackwell,~D.~D. and Richards,~M.: Geothermal Map of North America, Am. Assoc. Petr. Geol., Tulsa, OK, 2004.

\bibitem[{Booth et~al.(2003)Booth, Troost, Clague, and Waitt}]{Booth.etal.2003}
Booth,~D.~B., Troost,~K.~G., Clague,~J.~J., and Waitt,~R.~B.: The Cordilleran ice sheet, in: The Quaternary Period in the United States, edited by: Gillespie,~A., Porter,~S., and Atwater,~B., vol.~1 of Dev. Quaternary Sci., Elsevier, Amsterdam, 17--43,
doi:\href{http://dx.doi.org/10.1016/s1571-0866(03)01002-9}{10.1016/s1571-0866(03)01002-9}, 2003.

\bibitem[{Boulton and Clark(1990)}]{Boulton.Clark.1990}
Boulton,~G.~S. and Clark,~C.~D.: A highly mobile Laurentide ice sheet revealed by satellite images of glacial lineations, Nature, 346, 813--817,
doi:\href{http://dx.doi.org/10.1038/346813a0}{10.1038/346813a0}, 1990.

\bibitem[{Boulton et~al.(2001)Boulton, Dongelmans, Punkari, and Broadgate}]{Boulton.etal.2001}
Boulton,~G.~S., Dongelmans,~P., Punkari,~M., and Broadgate,~M.: Palaeoglaciology of an ice sheet through a glacial cycle: the European ice sheet through the Weichselian, Quaternary Res., 20, 591--625,
doi:\href{http://dx.doi.org/10.1016/s0277-3791(00)00160-8}{10.1016/s0277-3791(00)00160-8}, 2001.

\bibitem[{Briner and Kaufman(2008)}]{Briner.Kaufman.2008}
Briner,~J.~P. and Kaufman,~D.~S.: Late Pleistocene mountain glaciation in Alaska: key chronologies,~J. Quaternary Sci., 23, 659--670,
doi:\href{http://dx.doi.org/10.1002/jqs.1196}{10.1002/jqs.1196}, 2008.

\bibitem[{Bueler and Brown(2009)}]{Bueler.Brown.2009}
Bueler,~E. and Brown,~J.: Shallow shelf approximation as a ``sliding law'' in a~thermodynamically coupled ice sheet model,~J. Geophys. Res., 114, F03008,
doi:\href{http://dx.doi.org/10.1029/2008JF001179}{10.1029/2008JF001179}, 2009.

\bibitem[{Bueler and van Pelt(2015)}]{Bueler.Pelt.2015}
 Bueler,~E. and van~Pelt,~W.: Mass-conserving subglacial hydrology in the Parallel Ice Sheet Model version 0.6, Geosci. Model Dev., 8, 1613--1635,
doi:\href{http://dx.doi.org/10.5194/gmd-8-1613-2015}{10.5194/gmd-8-1613-2015}, 2015.

\bibitem[{Bueler et~al.(2007)Bueler, Lingle, and Brown}]{Bueler.etal.2007}
Bueler,~E., Lingle,~C.~S., and Brown,~J.: Fast computation of a viscoelastic deformable Earth model for ice-sheet simulations, Ann. Glaciol., 46, 97--105,
doi:\href{http://dx.doi.org/10.3189/172756407782871567}{10.3189/172756407782871567}, 2007.

\bibitem[{Burke et~al.(2012{\natexlab{a}})Burke, Brennand, and Perkins}]{Burke.etal.2012}
Burke,~M.~J., Brennand,~T.~A., and Perkins,~A.~J.: Transient subglacial hydrology of a thin ice sheet: insights from the Chasm esker, British Columbia, Canada, Quaternary Res., 58, 30--55,
doi:\href{http://dx.doi.org/10.1016/j.quascirev.2012.09.004}{10.1016/j.quascirev.2012.09.004}, 2012{\natexlab{a}}.

\bibitem[{Burke et~al.(2012{\natexlab{b}})Burke, Brennand, and Perkins}]{Burke.etal.2012a}
Burke,~M.~J., Brennand,~T.~A., and Perkins,~A.~J.: Evolution of the subglacial hydrologic system beneath the rapidly decaying Cordilleran Ice Sheet caused by ice-dammed lake drainage: implications for meltwater-induced ice acceleration, Quaternary Res., 50, 125--140,
doi:\href{http://dx.doi.org/10.1016/j.quascirev.2012.07.005}{10.1016/j.quascirev.2012.07.005}, 2012{\natexlab{b}}.

\bibitem[{Calov and Greve(2005)}]{Calov.Greve.2005}
Calov,~R. and Greve,~R.: A~semi-analytical solution for the positive degree-day model with stochastic temperature variations,~J. Glaciol., 51, 173--175,
doi:\href{http://dx.doi.org/10.3189/172756505781829601}{10.3189/172756505781829601}, 2005.

\bibitem[{Carlson and Clark(2012)}]{Carlson.Clark.2012}
Carlson,~A.~E. and Clark,~P.~U.: Ice sheet sources of sea level rise and
freshwater discharge during the last deglaciation, Rev. Geophys., 50, RG4007,
doi:\href{http://dx.doi.org/10.1029/2011rg000371}{10.1029/2011rg000371},
2012.

\bibitem[{Carrara et~al.(1996)Carrara, Kiver, and Stradling}]{Carrara.etal.1996}
Carrara,~P.~E., Kiver,~E.~P., and Stradling,~D.~F.: The southern limit of Cordilleran ice in the Colville and Pend Oreille valleys of northeastern Washington during the Late Wisconsin glaciation, Can.~J. Earth Sci., 33, 769--778,
doi:\href{http://dx.doi.org/10.1139/e96-059}{10.1139/e96-059}, 1996.

\bibitem[{Clague et~al.(1980)Clague, Armstrong, and Mathews}]{Clague.etal.1980}
Clague,~J., Armstrong,~J., and Mathews,~W.: Advance of the late Wisconsin Cordilleran Ice Sheet in southern British Columbia since 22\,000~yr~BP, Quaternary Res., 13, 322--326,
doi:\href{http://dx.doi.org/10.1016/0033-5894(80)90060-5}{10.1016/0033-5894(80)90060-5}, 1980.

\bibitem[{Clague(1981)}]{Clague.1981}
Clague,~J.~J.: Late Quaternary Geology and Geochronology of British Columbia Part 2: Summary and Discussion of Radiocarbon-Dated Quaternary History, Geol. Surv. of Can., Ottawa, ON, Paper 80-35,
doi:\href{http://dx.doi.org/10.4095/119439}{10.4095/119439}, 1981.

\bibitem[{Clague(1985)}]{Clague.1985}
Clague,~J.~J.: Delaciation of the Prince Rupert -- Kitimat area, British Columbia, Can.~J. Earth Sci., 22, 256--265,
doi:\href{http://dx.doi.org/10.1139/e85-022}{10.1139/e85-022}, 1985.

\bibitem[{Clague(1986)}]{Clague.1986}
Clague,~J.~J.: The Quaternary stratigraphic record of British Columbia --- evidence for episodic sedimentation and erosion controlled by glaciation, Can.~J. Earth Sci., 23, 885--894,
doi:\href{http://dx.doi.org/10.1139/e86-090}{10.1139/e86-090}, 1986.

\bibitem[{Clague(1989)}]{Clague.1989}
Clague,~J.~J.: Character and distribution of Quaternary deposits (Canadian Cordillera), in: Quaternary Geology of Canada and Greenland, edited by: Fulton,~R.~J., vol.~1 of Geology of Canada, Geol. Surv. of Can., Ottawa, ON, 34--48,
doi:\href{http://dx.doi.org/10.4095/127905}{10.4095/127905}, 1989.

\bibitem[{Clague and James(2002)}]{Clague.James.2002}
Clague,~J.~J. and James,~T.~S.: History and isostatic effects of the last ice sheet in southern British Columbia, Quaternary Res., 21, 71--87,
doi:\href{http://dx.doi.org/10.1016/s0277-3791(01)00070-1}{10.1016/s0277-3791(01)00070-1}, 2002.

\bibitem[{Clague and Ward(2011)}]{Clague.Ward.2011}
Clague,~J.~J. and Ward,~B.: Pleistocene Glaciation of British
Columbia, in:   Dev. Quaternary Sci., vol.~15, edited by: Ehlers,~J., Gibbard,~P.~L., and Hughes,~P.~D., Elsevier, Amsterdam, 563--573,
doi:\href{http://dx.doi.org/10.1016/b978-0-444-53447-7.00044-1}{10.1016/b978-0-444-53447-7.00044-1}, 2011.

\bibitem[{Clague et~al.(1997)Clague, Mathewes, Guilbault, Hutchinson, and Ricketts}]{Clague.etal.1997}
Clague,~J.~J., Mathewes,~R.~W., Guilbault,~J.-P., Hutchinson,~I., and Ricketts,~B.~D.: Pre-Younger Dryas resurgence of the southwestern margin of the Cordilleran ice sheet, British Columbia, Canada, Boreas, 26, 261--278,
doi:\href{http://dx.doi.org/10.1111/j.1502-3885.1997.tb00855.x}{10.1111/j.1502-3885.1997.tb00855.x}, 1997.

\bibitem[{Clague et~al.(2005)Clague, Froese, Hutchinson, James, and Simon}]{Clague.etal.2005}
Clague,~J.~J., Froese,~D., Hutchinson,~I., James,~T.~S., and Simon,~K.~M.: Early growth of the last Cordilleran ice sheet deduced from glacio-isostatic depression in southwest British Columbia, Canada, Quaternary Res., 63, 53--59,
doi:\href{http://dx.doi.org/10.1016/j.yqres.2004.09.007}{10.1016/j.yqres.2004.09.007}, 2005.

\bibitem[{Clark and Mix(2002)}]{Clark.Mix.2002}
Clark,~P.~U. and Mix,~A.~C.: Ice sheets and sea level of the Last Glacial Maximum, Quaternary Res., 21, 1--7,
doi:\href{http://dx.doi.org/10.1016/s0277-3791(01)00118-4}{10.1016/s0277-3791(01)00118-4}, 2002.

\bibitem[{Clason et~al.(2012)Clason, Mair, Burgess, and Nienow}]{Clason.etal.2012}
Clason,~C., Mair,~D.~W., Burgess,~D.~O., and Nienow,~P.~W.: Modelling the delivery of supraglacial meltwater to the ice/bed interface: application to southwest Devon Ice Cap, Nunavut, Canada,~J. Glaciol., 58, 361--374,
doi:\href{http://dx.doi.org/10.3189/2012jog11j129}{10.3189/2012jog11j129}, 2012.

\bibitem[{Clason et~al.(2014)Clason, Applegate, and Holmlund}]{Clason.etal.2014}
Clason,~C., Applegate,~P., and Holmlund,~P.: Modelling Late Weichselian evolution of the Eurasian ice sheets forced by surface meltwater-enhanced basal sliding,~J. Glaciol., 60, 29--40,
doi:\href{http://dx.doi.org/10.3189/2014jog13j037}{10.3189/2014jog13j037}, 2014.

\bibitem[{Cosma et~al.(2008)Cosma, Hendy, and Chang}]{Cosma.etal.2008}
Cosma,~T., Hendy,~I., and Chang,~A.: Chronological constraints on Cordilleran Ice Sheet glaciomarine sedimentation from core MD02-2496 off Vancouver Island (western Canada), Quaternary Sci. Rev., 27, 941--955,
doi:\href{http://dx.doi.org/10.1016/j.quascirev.2008.01.013}{10.1016/j.quascirev.2008.01.013}, 2008.

\bibitem[{Dansgaard et~al.(1993)Dansgaard, Johnsen, Clausen, Dahl-Jensen, Gundestrup, Hammer, Hvidberg, Steffensen, Sveinbj\"ornsdottir, Jouzel, and Bond}]{Dansgaard.etal.1993}
Dansgaard,~W., Johnsen,~S.~J., Clausen,~H.~B., Dahl-Jensen,~D., Gundestrup,~N.~S., Hammer,~C.~U., Hvidberg,~C.~S., Steffensen,~J.~P., Sveinbj\"ornsdottir,~A.~E., Jouzel,~J., and Bond,~G.: Evidence for general instability of past climate from a 250-kyr ice-core record, Nature, 364, 218--220,
doi:\href{http://dx.doi.org/10.1038/364218a0}{10.1038/364218a0}, data archived at the World Data Center for Paleoclimatology, Boulder, Colorado, USA., 1993.

\bibitem[{Davies et~al.(2011)Davies, Mix, Stoner, Addison, Jaeger, Finney, and Wiest}]{Davies.etal.2011}
Davies,~M.~H., Mix,~A.~C., Stoner,~J.~S., Addison,~J.~A., Jaeger,~J., Finney,~B., and Wiest,~J.: The deglacial transition on the southeastern Alaska Margin: Meltwater input, sea level rise, marine productivity, and sedimentary anoxia, Paleoceanography, 26, PA2223,
doi:\href{http://dx.doi.org/10.1029/2010pa002051}{10.1029/2010pa002051}, 2011.

\bibitem[{Davis and Mathews(1944)}]{Davis.Mathews.1944}
Davis,~N.~F.~G. and Mathews,~W.~H.: Four phases of glaciation with illustrations from Southwestern British Columbia,~J. Geol., 52, 403--413,
doi:\href{http://dx.doi.org/10.1086/625236}{10.1086/625236}, 1944.

\bibitem[{Dawson(1888)}]{Dawson.1888}
Dawson,~G.~M.: III. -- Recent observations on the glaciation of British Columbia and adjacent regions, Geol. Mag., 5, 347--350,
doi:\href{http://dx.doi.org/10.1017/s0016756800182159}{10.1017/s0016756800182159}, 1888.

\bibitem[{Demuro et~al.(2012)Demuro, Froese, Arnold, and Roberts}]{Demuro.etal.2012}
Demuro,~M., Froese,~D.~G., Arnold,~L.~J., and Roberts,~R.~G.: Single-grain OSL dating of glaciofluvial quartz constrains Reid glaciation in NW Canada to MIS 6, Quaternary Res., 77, 305--316,
doi:\href{http://dx.doi.org/10.1016/j.yqres.2011.11.009}{10.1016/j.yqres.2011.11.009}, 2012.

\bibitem[{Duk-Rodkin(1999)}]{Duk-Rodkin.1999}
Duk-Rodkin,~A.: Glacial limits map of Yukon Territory, Open File 3694, Geol. Surv. of Can., Ottawa, ON,
doi:\href{http://dx.doi.org/10.4095/210739}{10.4095/210739}, 1999.

\bibitem[{Duk-Rodkin et~al.(1996)Duk-Rodkin, Barendregt, Tarnocai, and Phillips}]{Duk-Rodkin.etal.1996}
Duk-Rodkin,~A., Barendregt,~R.~W., Tarnocai,~C., and Phillips,~F.~M.: Late Tertiary to late Quaternary record in the Mackenzie Mountains, Northwest Territories, Canada: stratigraphy, paleosols, paleomagnetism, and chlorine-36, Can.~J. Earth Sci., 33, 875--895,
doi:\href{http://dx.doi.org/10.1139/e96-066}{10.1139/e96-066}, 1996.

\bibitem[{Dyke(2004)}]{Dyke.2004}
Dyke,~A.~S.: An outline of North American deglaciation with emphasis
on central and northern Canada, in: Dev. Quaternary Sci., vol.~2,
edited by: Ehlers,~J. and Gibbard,~P.~L., Elsevier, Amsterdam, 373--424,
doi:\href{http://dx.doi.org/10.1016/S1571-0866(04)80209-4}{10.1016/S1571-0866(04)80209-4}, 2004.

\bibitem[{Dyke and Prest(1987)}]{Dyke.Prest.1987}
Dyke,~A.~S. and Prest,~V.~K.: Late wisconsinan and holocene history of the laurentide ice sheet, G{e}ogr. Phys. Quatern., 41, 237--263,
doi:\href{http://dx.doi.org/10.7202/032681ar}{10.7202/032681ar}, 1987.

\bibitem[{Dyke et~al.(2003)Dyke, Moore, and Robertson}]{Dyke.etal.2003}
Dyke,~A.~S., Moore,~A., and Robertson,~L.: Deglaciation of North America, Open File 1547, Geol. Surv. of Can., Ottawa, ON, 2003.

\bibitem[{Ehlers and Gibbard(2004)}]{Ehlers.Gibbard.2004}
Ehlers,~J. and Gibbard,~P.~L. (Eds.): Dev. Quaternary Sci., vol.~2, Elsevier, Amsterdam, 2004.

\bibitem[{Ehlers et~al.(2011)Ehlers, Gibbard, and Hughes}]{Ehlers.etal.2011}
Ehlers,~J., Gibbard,~P.~L., and Hughes,~P.~D. (Eds.):  Dev. Quaternary
Sci., vol.~15, Elsevier, Amsterdam, 2011.

\bibitem[{Fahnestock et~al.(2001)Fahnestock, Abdalati, Joughin, Brozena, and Gogineni}]{Fahnestock.etal.2001}
Fahnestock,~M., Abdalati,~W., Joughin,~I., Brozena,~J., and Gogineni,~P.: High geothermal heat flow, basal melt, and the origin of rapid ice flow in central greenland, Sience, 294, 2338--2342,
doi:\href{http://dx.doi.org/10.1126/science.1065370}{10.1126/science.1065370}, 2001.

\bibitem[{Friele and Clague(2002{\natexlab{a}})}]{Friele.Clague.2002}
Friele,~P.~A. and Clague,~J.~J.: Readvance of glaciers in the British Columbia Coast Mountains at the end of the last glaciation, Quatern. Int., 87, 45--58,
doi:\href{http://dx.doi.org/10.1016/s1040-6182(01)00061-1}{10.1016/s1040-6182(01)00061-1}, 2002{\natexlab{a}}.

\bibitem[{Friele and Clague(2002{\natexlab{b}})}]{Friele.Clague.2002a}
Friele,~P.~A. and Clague,~J.~J.: Younger Dryas readvance in Squamish river valley, southern Coast mountains, British Columbia, Quaternary Res., 21, 1925--1933,
doi:\href{http://dx.doi.org/10.1016/s0277-3791(02)00081-1}{10.1016/s0277-3791(02)00081-1}, 2002{\natexlab{b}}.

\bibitem[{Fulton(1967)}]{Fulton.1967}
Fulton,~R.~J.: Deglaciation studies in Kamloops region, an area of moderate relief, British Columbia, vol. 154 of Bull., Geol. Surv. of Can., Ottawa, ON,
doi:\href{http://dx.doi.org/10.4095/101467}{10.4095/101467}, 1967.

\bibitem[{Fulton(1991)}]{Fulton.1991}
Fulton,~R.~J.: A Conceptual model for growth and decay of the Cordilleran Ice Sheet, G{e}ogr. Phys. Quatern., 45, 281--286,
doi:\href{http://dx.doi.org/10.7202/032875ar}{10.7202/032875ar}, 1991.

\bibitem[{Golledge et~al.(2012)Golledge, Mackintosh, Anderson, Buckley, Doughty, Barrell, Denton, Vandergoes, Andersen, and Schaefer}]{Golledge.etal.2012}
Golledge,~N.~R., Mackintosh,~A.~N., Anderson,~B.~M., Buckley,~K.~M., Doughty,~A.~M., Barrell,~D.~J., Denton,~G.~H., Vandergoes,~M.~J., Andersen,~B.~G., and Schaefer,~J.~M.: Last Glacial Maximum climate in {N}ew {Z}ealand inferred from a modelled {S}outhern {A}lps icefield, Quaternary Res., 46, 30--45,
doi:\href{http://dx.doi.org/10.1016/j.quascirev.2012.05.004}{10.1016/j.quascirev.2012.05.004}, 2012.

\bibitem[{Herbert et~al.(2001)Herbert, Schuffert, Andreasen, Heusser, Lyle, Mix, Ravelo, Stott, and Herguera}]{Herbert.etal.2001}
Herbert,~T.~D., Schuffert,~J.~D., Andreasen,~D., Heusser,~L., Lyle,~M., Mix,~A., Ravelo,~A.~C., Stott,~L.~D., and Herguera,~J.~C.: Collapse of the California current during glacial maxima linked to climate change on land, Sience, 293, 71--76,
doi:\href{http://dx.doi.org/10.1126/science.1059209}{10.1126/science.1059209}, data archived at the World Data Center for Paleoclimatology, Boulder, Colorado, USA, 2001.

\bibitem[{Hidy et~al.(2013)Hidy, Gosse, Froese, Bond, and Rood}]{Hidy.etal.2013}
Hidy,~A.~J., Gosse,~J.~C., Froese,~D.~G., Bond,~J.~D., and Rood,~D.~H.: A latest Pliocene age for the earliest and most extensive Cordilleran Ice Sheet in northwestern Canada, Quaternary Res., 61, 77--84,
doi:\href{http://dx.doi.org/10.1016/j.quascirev.2012.11.009}{10.1016/j.quascirev.2012.11.009}, 2013.

\bibitem[{Hock(2003)}]{Hock.2003}
Hock,~R.: Temperature index melt modelling in mountain areas,~J. Hydrol., 282, 104--115,
doi:\href{http://dx.doi.org/10.1016/S0022-1694(03)00257-9}{10.1016/S0022-1694(03)00257-9}, 2003.

\bibitem[{Imbrie et~al.(1989)Imbrie, McIntyre, and Mix}]{Imbrie.etal.1989}
Imbrie,~J., McIntyre,~A., and Mix,~A.: Oceanic response to orbital forcing in the late quaternary: observational and experimental strategies, in: Climate and Geo-Sciences, edited by: Berger,~A., Schneider,~S., and Duplessy,~J., vol. 285 of NATO ASI Series C, Kluwer, Norwell, MA, 121--164,
doi:\href{http://dx.doi.org/10.1007/978-94-009-2446-8_7}{10.1007/978-94-009-2446-8\_7}, 1989.

\bibitem[{Ivy-Ochs et~al.(1999)Ivy-Ochs, Schluchter, Kubik, and Denton}]{Ivy-Ochs.etal.1999}
Ivy-Ochs,~S., Schluchter,~C., Kubik,~P.~W., and Denton,~G.~H.: Moraine
exposure dates imply synchronous Younger Dryas Glacier advances in the
European Alps and in the Southern Alps of New Zealand, Geogr. Ann. A, 81,
313--323,
doi:\href{http://dx.doi.org/10.1111/1468-0459.00060}{10.1111/1468-0459.00060},
1999.

\bibitem[{Jackson and Clague(1991)}]{Jackson.Clague.1991}
Jackson,~L.~E. and Clague,~J.~J.: The Cordilleran ice sheet: one hundred and fifty years of exploration and discovery, G{e}ogr. Phys. Quatern., 45, 269--280,
doi:\href{http://dx.doi.org/10.7202/032874ar}{10.7202/032874ar}, 1991.

\bibitem[{Jackson et~al.(1997)Jackson, Phillips, Shimamura, and Little}]{Jackson.etal.1997}
Jackson,~L.~E., Phillips,~F.~M., Shimamura,~K., and Little,~E.~C.: Cosmogenic $^{36}$Cl dating of the Foothills erratics train, Alberta, Canada, Geology, 25, 195,
doi:\href{http://dx.doi.org/10.1130/0091-7613(1997)025<0195:ccdotf>2.3.co;2}{10.1130/0091-7613(1997)025\textless0195:ccdotf\textgreater2.3.co;2}, 1997.

\bibitem[{James et~al.(2009)James, Gowan, Wada, and Wang}]{James.etal.2009}
James,~T.~S., Gowan,~E.~J., Wada,~I., and Wang,~K.: Viscosity of the
asthenosphere from glacial isostatic adjustment and subduction dynamics at
the northern Cascadia subduction zone, British Columbia, Canada,~J. Geophys.
Res., 114, B04405,
doi:\href{http://dx.doi.org/10.1029/2008jb006077}{10.1029/2008jb006077},
2009.

\bibitem[{Johnsen et~al.(1995)Johnsen, Dahl-Jensen, Dansgaard, and Gundestrup}]{Johnsen.etal.1995}
Johnsen,~S.~J., Dahl-Jensen,~D., Dansgaard,~W., and Gundestrup,~N.: Greenland palaeotemperatures derived from GRIP bore hole temperature and ice core isotope profiles, Tellus B, 47, 624--629,
doi:\href{http://dx.doi.org/10.1034/j.1600-0889.47.issue5.9.x}{10.1034/j.1600-0889.47.issue5.9.x}, 1995.

\bibitem[{Jouzel et~al.(2007)Jouzel, Masson-Delmotte, Cattani, Dreyfus, Falourd, Hoffmann, Minster, Nouet, Barnola, Chappellaz, Fischer, Gallet, Johnsen, Leuenberger, Loulergue, Luethi, Oerter, Parrenin, Raisbeck, Raynaud, Schilt, Schwander, Selmo, Souchez, Spahni, Stauffer, Steffensen, Stenni, Stocker, Tison, Werner, and Wolff}]{Jouzel.etal.2007}
Jouzel,~J., Masson-Delmotte,~V., Cattani,~O., Dreyfus,~G., Falourd,~S., Hoffmann,~G., Minster,~B., Nouet,~J., Barnola,~J.~M., Chappellaz,~J., Fischer,~H., Gallet,~J.~C., Johnsen,~S., Leuenberger,~M., Loulergue,~L., Luethi,~D., Oerter,~H., Parrenin,~F., Raisbeck,~G., Raynaud,~D., Schilt,~A., Schwander,~J., Selmo,~E., Souchez,~R., Spahni,~R., Stauffer,~B., Steffensen,~J.~P., Stenni,~B., Stocker,~T.~F., Tison,~J.~L., Werner,~M., and Wolff,~E.~W.: Orbital and millennial antarctic climate variability over the past 800\,000~years, Sience, 317, 793--796,
doi:\href{http://dx.doi.org/10.1126/science.1141038}{10.1126/science.1141038}, data archived at the World Data Center for Paleoclimatology, Boulder, Colorado, USA., 2007.

\bibitem[{Kaufman and Manley(2004)}]{Kaufman.Manley.2004}
Kaufman,~D.~S. and Manley,~W.~F.: Pleistocene maximum and late wisconsinan glacier extents across Alaska, {USA}, in: Dev. Quaternary Sci., vol.~2,
edited by: Ehlers,~J. and Gibbard,~P.~L., Elsevier, Amsterdam, 9--27,
doi:\href{http://dx.doi.org/10.1016/S1571-0866(04)80182-9}{10.1016/S1571-0866(04)80182-9}, 2004.

\bibitem[{Kleman(1994)}]{Kleman.1994}
Kleman,~J.: Preservation of landforms under ice sheets and ice caps, Geomorphology, 9, 19--32,
doi:\href{http://dx.doi.org/10.1016/0169-555x(94)90028-0}{10.1016/0169-555x(94)90028-0}, 1994.

\bibitem[{Kleman and Stroeven(1997)}]{Kleman.Stroeven.1997}
Kleman,~J. and Stroeven,~A.~P.: Preglacial surface remnants and Quaternary glacial regimes in northwestern Sweden, Geomorphology, 19, 35--54,
doi:\href{http://dx.doi.org/10.1016/s0169-555x(96)00046-3}{10.1016/s0169-555x(96)00046-3}, 1997.

\bibitem[{Kleman et~al.(1997)Kleman, H{\"a}ttestrand, Borgstr{\"o}m, and Stroeven}]{Kleman.etal.1997}
Kleman,~J., H{\"a}ttestrand,~C., Borgstr{\"o}m,~I., and Stroeven,~A.: Fennoscandian paleoglaciology reconstructed using a glacial geological inversion model,~J. Glaciol., 43, 283--299, 1997.

\bibitem[{Kleman et~al.(2006)Kleman, H{\"a}ttestrand, Stroeven, Jansson, De~Angelis, and Borgstr{\"o}m}]{Kleman.etal.2006}
Kleman,~J., H{\"a}ttestrand,~C., Stroeven,~A.~P., Jansson,~K.~N., De~Angelis,~H., and Borgstr{\"o}m,~I.: Reconstruction of palaeo-ice sheets -- inversion of their glacial geomorphological record, in: Glacier Science and Environmental Change, edited by: Knight,~P.~G., Blackwell, Malden, MA, 192--198,
doi:\href{http://dx.doi.org/10.1002/9780470750636.ch38}{10.1002/9780470750636.ch38}, 2006.

\bibitem[{Kleman et~al.(2008)Kleman, Stroeven, and Lundqvist}]{Kleman.etal.2008}
Kleman,~J., Stroeven,~A.~P., and Lundqvist,~J.: Patterns of quaternary ice sheet erosion and deposition in Fennoscandia and a theoretical framework for explanation, Geomorphology, 97, 73--90,
doi:\href{http://dx.doi.org/10.1016/j.geomorph.2007.02.049}{10.1016/j.geomorph.2007.02.049}, 2008.

\bibitem[{Kleman et~al.(2010)Kleman, Jansson, De~Angelis, Stroeven, H\"{a}ttestrand, Alm, and Glasser}]{Kleman.etal.2010}
Kleman,~J., Jansson,~K., De~Angelis,~H., Stroeven,~A., H\"{a}ttestrand,~C., Alm,~G., and Glasser,~N.: North American ice sheet build-up during the last glacial cycle, 115--21\,kyr, Quaternary Sci. Rev., 29, 2036--2051,
doi:\href{http://dx.doi.org/10.1016/j.quascirev.2010.04.021}{10.1016/j.quascirev.2010.04.021}, 2010.

\bibitem[{Kovanen(2002)}]{Kovanen.2002}
Kovanen,~D.~J.: Morphologic and stratigraphic evidence for Aller{\o}d and Younger Dryas age glacier fluctuations of the Cordilleran Ice Sheet, British Columbia, Canada and Northwest Washington,~USA, Boreas, 31, 163--184,
doi:\href{http://dx.doi.org/10.1111/j.1502-3885.2002.tb01064.x}{10.1111/j.1502-3885.2002.tb01064.x}, 2002.

\bibitem[{Kovanen and Easterbrook(2002)}]{Kovanen.Easterbrook.2002}
Kovanen,~D.~J. and Easterbrook,~D.~J.: Timing and extent of Aller{\o}d and Younger Dryas age (ca. 12\,500--10\,000 $^{14}$C~yr~BP) oscillations of the Cordilleran Ice Sheet in the Fraser Lowland, Western North America, Quaternary Res., 57, 208--224,
doi:\href{http://dx.doi.org/10.1006/qres.2001.2307}{10.1006/qres.2001.2307}, 2002.

\bibitem[{Lakeman et~al.(2008)Lakeman, Clague, and Menounos}]{Lakeman.etal.2008}
Lakeman,~T.~R., Clague,~J.~J., and Menounos,~B.: Advance of alpine glaciers during final retreat of the Cordilleran ice sheet in the Finlay River area, northern British Columbia, Canada, Quaternary Res., 69, 188--200,
doi:\href{http://dx.doi.org/10.1016/j.yqres.2008.01.002}{10.1016/j.yqres.2008.01.002}, 2008.

\bibitem[{Langen et~al.(2012)Langen, Solgaard, and Hvidberg}]{Langen.etal.2012}
Langen,~P.~L., Solgaard,~A.~M., and Hvidberg,~C.~S.: Self-inhibiting growth of the Greenland ice sheet, Geophys. Res. Lett., 39, L12502,
doi:\href{http://dx.doi.org/10.1029/2012GL051810}{10.1029/2012GL051810}, 2012.

\bibitem[{Levermann et~al.(2012)Levermann, Albrecht, Winkelmann, Martin, Haseloff, and Joughin}]{Levermann.etal.2012}
 Levermann,~A., Albrecht,~T., Winkelmann,~R., Martin,~M.~A., Haseloff,~M., and Joughin,~I.: Kinematic first-order calving law implies potential for  abrupt ice-shelf retreat, The Cryosphere, 6, 273--286,
doi:\href{http://dx.doi.org/10.5194/tc-6-273-2012}{10.5194/tc-6-273-2012}, 2012.

\bibitem[{Lingle and Clark(1985)}]{Lingle.Clark.1985}
Lingle,~C.~S. and Clark,~J.~A.: A~numerical model of interactions between a~marine ice sheet and the Solid Earth: application to a~West Antarctic ice stream,~J. Geophys. Res., 90, 1100--1114,
doi:\href{http://dx.doi.org/10.1029/JC090iC01p01100}{10.1029/JC090iC01p01100}, 1985.

\bibitem[{Lisiecki and Raymo(2005)}]{Lisiecki.Raymo.2005}
Lisiecki,~L.~E. and Raymo,~M.~E.: A Pliocene-Pleistocene stack of 57 globally distributed benthic $\delta^{18}$O records, Paleoceanography, 20, PA1003,
doi:\href{http://dx.doi.org/10.1029/2004pa001071}{10.1029/2004pa001071}, 2005.

\bibitem[{Lundqvist(1987)}]{Lundqvist.1987}
Lundqvist,~J.: Glaciodynamics of the Younger Dryas Marginal zone in Scandinavia: implications of a revised glaciation model, Geogr. Ann. A, 69, 305,
doi:\href{http://dx.doi.org/10.2307/521191}{10.2307/521191}, 1987.

\bibitem[{Margold et~al.(2011)Margold, Jansson, Kleman, and Stroeven}]{Margold.etal.2011}
Margold,~M., Jansson,~K.~N., Kleman,~J., and Stroeven,~A.~P.: Glacial meltwater landforms of central British Columbia,~J. Maps, 7, 486--506,
doi:\href{http://dx.doi.org/10.4113/jom.2011.1205}{10.4113/jom.2011.1205}, 2011.

\bibitem[{Margold et~al.(2013{\natexlab{a}})Margold, Jansson, Kleman, and Stroeven}]{Margold.etal.2013}
Margold,~M., Jansson,~K.~N., Kleman,~J., and Stroeven,~A.~P.: Lateglacial ice dynamics of the Cordilleran Ice Sheet in northern British Columbia and southern Yukon Territory: retreat pattern of the Liard Lobe reconstructed from the glacial landform record,~J. Quaternary Sci., 28, 180--188,
doi:\href{http://dx.doi.org/10.1002/jqs.2604}{10.1002/jqs.2604}, 2013{\natexlab{a}}.

\bibitem[{Margold et~al.(2013{\natexlab{b}})Margold, Jansson, Kleman, Stroeven, and Clague}]{Margold.etal.2013a}
Margold,~M., Jansson,~K.~N., Kleman,~J., Stroeven,~A.~P., and Clague,~J.~J.: Retreat pattern of the Cordilleran Ice Sheet in central British Columbia at the end of the last glaciation reconstructed from glacial meltwater landforms, Boreas, 42, 830--847,
doi:\href{http://dx.doi.org/10.1111/bor.12007}{10.1111/bor.12007}, 2013{\natexlab{b}}.

\bibitem[{Margold et~al.(2014)Margold, Stroeven, Clague, and Heyman}]{Margold.etal.2014}
Margold,~M., Stroeven,~A.~P., Clague,~J.~J., and Heyman,~J.: Timing of terminal Pleistocene deglaciation at high elevations in southern and central British Columbia constrained by $^{10}$Be exposure dating, Quaternary Res., 99, 193--202,
doi:\href{http://dx.doi.org/10.1016/j.quascirev.2014.06.027}{10.1016/j.quascirev.2014.06.027}, 2014.

\bibitem[{Marshall et~al.(2000)Marshall, Tarasov, Clarke, and Peltier}]{Marshall.etal.2000}
Marshall,~S.~J., Tarasov,~L., Clarke,~G.~K., and Peltier,~W.~R.: Glaciological reconstruction of the Laurentide Ice Sheet: physical processes and modelling challenges, Can.~J. Earth Sci., 37, 769--793,
doi:\href{http://dx.doi.org/10.1139/e99-113}{10.1139/e99-113}, 2000.

\bibitem[{Martin et~al.(2011)Martin, Winkelmann, Haseloff, Albrecht, Bueler, Khroulev, and Levermann}]{Martin.etal.2011}
 Martin,~M.~A., Winkelmann,~R., Haseloff,~M., Albrecht,~T., Bueler,~E., Khroulev,~C., and Levermann,~A.: The Potsdam Parallel Ice Sheet Model (PISM-PIK) -- Part 2: Dynamic equilibrium simulation of the Antarctic ice sheet, The Cryosphere, 5, 727--740,
doi:\href{http://dx.doi.org/10.5194/tc-5-727-2011}{10.5194/tc-5-727-2011}, 2011.

\bibitem[{Menounos et~al.(2008)Menounos, Osborn, Clague, and Luckman}]{Menounos.etal.2008}
Menounos,~B., Osborn,~G., Clague,~J.~J., and Luckman,~B.~H.: Latest Pleistocene and Holocene glacier fluctuations in western Canada, Quaternary Sci. Rev., 28, 2049--2074,
doi:\href{http://dx.doi.org/10.1016/j.quascirev.2008.10.018}{10.1016/j.quascirev.2008.10.018}, 2008.

\bibitem[{Mesinger et~al.(2006)Mesinger, DiMego, Kalnay, Mitchell, Shafran, Ebisuzaki, Jovi\'{c}, Woollen, Rogers, Berbery, Ek, Fan, Grumbine, Higgins, Li, Lin, Manikin, Parrish, and Shi}]{Mesinger.etal.2006}
Mesinger,~F., DiMego,~G., Kalnay,~E., Mitchell,~K., Shafran,~P.~C., Ebisuzaki,~W., Jovi\'{c},~D., Woollen,~J., Rogers,~E., Berbery,~E.~H., Ek,~M.~B., Fan,~Y., Grumbine,~R., Higgins,~W., Li,~H., Lin,~Y., Manikin,~G., Parrish,~D., and Shi,~W.: North American regional reanalysis,~B. Am. Meteorol. Soc., 87, 343--360,
doi:\href{http://dx.doi.org/10.1175/BAMS-87-3-343}{10.1175/BAMS-87-3-343}, 2006.

\bibitem[{Osborn and Gerloff(1997)}]{Osborn.Gerloff.1997}
Osborn,~G. and Gerloff,~L.: Latest pleistocene and early Holocene fluctuations of glaciers in the Canadian and northern American Rockies, Quatern. Int., 38--39, 7--19,
doi:\href{http://dx.doi.org/10.1016/s1040-6182(96)00026-2}{10.1016/s1040-6182(96)00026-2}, 1997.

\bibitem[{Patterson and Kelso(2015)}]{Patterson.Kelso.2015}
Patterson,~T. and Kelso,~N.~V.: {N}atural {E}arth, {F}ree vector and raster
map data, available at: \url{http://naturalearthdata.com} (last access:
2015), 2015.

\bibitem[{Perkins and Brennand(2014)}]{Perkins.Brennand.2014}
Perkins,~A.~J. and Brennand,~T.~A.: Refining the pattern and style of Cordilleran Ice Sheet retreat: palaeogeography, evolution and implications of lateglacial ice-dammed lake systems on the southern Fraser Plateau, British Columbia, Canada, Boreas, 44, 319--342,
doi:\href{http://dx.doi.org/10.1111/bor.12100}{10.1111/bor.12100}, 2014.

\bibitem[{Perkins et~al.(2013)Perkins, Brennand, and Burke}]{Perkins.etal.2013}
Perkins,~A.~J., Brennand,~T.~A., and Burke,~M.~J.: Genesis of an esker-like ridge over the southern Fraser Plateau, British Columbia: implications for paleo-ice sheet reconstruction based on geomorphic inversion, Geomorphology, 190, 27--39,
doi:\href{http://dx.doi.org/10.1016/j.geomorph.2013.02.005}{10.1016/j.geomorph.2013.02.005}, 2013.

\bibitem[{Petit et~al.(1999)Petit, Jouzel, Raynaud, Barkov, Barnola, Basile, Bender, Chappellaz, Davis, Delaygue, Delmotte, Kotlyakov, Legrand, Lipenkov, Lorius, P{\'e}pin, Ritz, Saltzman, and Stievenard}]{Petit.etal.1999}
Petit,~J.~R., Jouzel,~J., Raynaud,~D., Barkov,~N.~I., Barnola,~J.-M., Basile,~I., Bender,~M., Chappellaz,~J., Davis,~M., Delaygue,~G., Delmotte,~M., Kotlyakov,~V.~M., Legrand,~M., Lipenkov,~V.~Y., Lorius,~C., P{\'e}pin,~L., Ritz,~C., Saltzman,~E., and Stievenard,~M.: Climate and atmospheric history of the past 420,000 years from the Vostok ice core, Antarctica, Nature, 399, 429--436,
doi:\href{http://dx.doi.org/10.1038/20859}{10.1038/20859}, data archived at the World Data Center for Paleoclimatology, Boulder, Colorado, USA., 1999.

\bibitem[{Porter(1989)}]{Porter.1989}
Porter,~S.~C.: Some geological implications of average Quaternary glacial conditions, Quaternary Res., 32, 245--261,
doi:\href{http://dx.doi.org/10.1016/0033-5894(89)90092-6}{10.1016/0033-5894(89)90092-6}, 1989.

\bibitem[{Porter and Swanson(1998)}]{Porter.Swanson.1998}
Porter,~S.~C. and Swanson,~T.~W.: Radiocarbon age constraints on rates of advance and retreat of the {P}uget {L}obe of the {C}ordilleran {I}ce {S}heet during the last glaciation, Quaternary Res., 50, 205--213,
doi:\href{http://dx.doi.org/10.1006/qres.1998.2004}{10.1006/qres.1998.2004}, 1998.

\bibitem[{Praetorius and Mix(2014)}]{Praetorius.Mix.2014}
Praetorius,~S.~K. and Mix,~A.~C.: Synchronization of North Pacific and Greenland climates preceded abrupt deglacial warming, Sience, 345, 444--448,
doi:\href{http://dx.doi.org/10.1126/science.1252000}{10.1126/science.1252000}, 2014.

\bibitem[{Prest et~al.(1968)Prest, Grant, and Rampton}]{Prest.etal.1968}
Prest,~V.~K., Grant,~D.~R., and Rampton,~V.~N.: Glacial map of Canada, ``A'' Series Map 1253A, Geol. Surv. of Can., Ottawa, ON,
doi:\href{http://dx.doi.org/10.4095/108979}{10.4095/108979}, 1968.

\bibitem[{Robert(1991)}]{Robert.1991}
Robert,~B.~L.: Modeling the Cordilleran ice sheet, G{e}ogr. Phys. Quatern., 45, 287--299,
doi:\href{http://dx.doi.org/10.7202/032876ar}{10.7202/032876ar}, 1991.

\bibitem[{Rutter et~al.(2012)Rutter, Coronato, Helmens, Rabassa, and Z{\'a}rate}]{Rutter.etal.2012}
Rutter,~N., Coronato,~A., Helmens,~K., Rabassa,~J., and Z{\'a}rate,~M.: Glaciations in North and South America from the Miocene to the Last Glacial Maximum: Comparisons, Linkages and Uncertainties, Springer Briefs in Earth System Sciences, Springer, Dordrecht,
doi:\href{http://dx.doi.org/10.1007/978-94-007-4399-1}{10.1007/978-94-007-4399-1}, 2012.

\bibitem[{Ryder et~al.(1991)Ryder, Fulton, and Clague}]{Ryder.etal.1991}
Ryder,~J.~M., Fulton,~R.~J., and Clague,~J.~J.: The Cordilleran ice sheet and the glacial geomorphology of southern and central British Colombia, G{e}ogr. Phys. Quatern., 45, 365--377,
doi:\href{http://dx.doi.org/10.7202/032882ar}{10.7202/032882ar}, 1991.

\bibitem[{Seguinot(2013)}]{Seguinot.2013}
Seguinot,~J.: Spatial and seasonal effects of temperature variability in a positive degree-day glacier surface mass-balance model,~J. Glaciol., 59, 1202--1204,
doi:\href{http://dx.doi.org/10.3189/2013JoG13J081}{10.3189/2013JoG13J081}, 2013.

\bibitem[{Seguinot et~al.(2014)Seguinot, Khroulev, Rogozhina, Stroeven, and Zhang}]{Seguinot.etal.2014}
 Seguinot,~J., Khroulev,~C., Rogozhina,~I., Stroeven,~A.~P., and Zhang,~Q.: The effect of climate forcing on numerical simulations of the Cordilleran ice sheet at the Last Glacial Maximum, The Cryosphere, 8, 1087--1103,
doi:\href{http://dx.doi.org/10.5194/tc-8-1087-2014}{10.5194/tc-8-1087-2014}, 2014.

\bibitem[{Shea et~al.(2009)Shea, Moore, and Stahl}]{Shea.etal.2009}
Shea,~J.~M., Moore,~R.~D., and Stahl,~K.: Derivation of melt factors from glacier mass-balance records in western Canada,~J. Glaciol., 55, 123--130,
doi:\href{http://dx.doi.org/10.3189/002214309788608886}{10.3189/002214309788608886}, 2009.

\bibitem[{Sissons(1979)}]{Sissons.1979}
Sissons,~J.~B.: The Loch Lomond Stadial in the British Isles, Nature, 280, 199--203,
doi:\href{http://dx.doi.org/10.1038/280199a0}{10.1038/280199a0}, 1979.

\bibitem[{Stea et~al.(2011)Stea, Seaman, Pronk, Parkhill, Allard, and Utting}]{Stea.etal.2011}
Stea,~R.~R., Seaman,~A.~A., Pronk,~T., Parkhill,~M.~A., Allard,~S., and Utting,~D.: The Appalachian Glacier Complex in Maritime Canada, in:   Dev. Quaternary Sci., vol.~15, edited by: Ehlers,~J., Gibbard,~P.~L., and Hughes,~P.~D., Elsevier, Amsterdam, 631--659,
doi:\href{http://dx.doi.org/10.1016/b978-0-444-53447-7.00048-9}{10.1016/b978-0-444-53447-7.00048-9}, 2011.

\bibitem[{Stroeven et~al.(2010)Stroeven, Fabel, Codilean, Kleman, Clague, Miguens-Rodriguez, and Xu}]{Stroeven.etal.2010}
Stroeven,~A.~P., Fabel,~D., Codilean,~A.~T., Kleman,~J., Clague,~J.~J., Miguens-Rodriguez,~M., and Xu,~S.: Investigating the glacial history of the northern sector of the Cordilleran ice sheet with cosmogenic $^{10}$Be concentrations in quartz, Quaternary Sci. Rev., 29, 3630--3643,
doi:\href{http://dx.doi.org/10.1016/j.quascirev.2010.07.010}{10.1016/j.quascirev.2010.07.010}, 2010.

\bibitem[{Stroeven et~al.(2014)Stroeven, Fabel, Margold, Clague, and Xu}]{Stroeven.etal.2014}
Stroeven,~A.~P., Fabel,~D., Margold,~M., Clague,~J.~J., and Xu,~S.: Investigating absolute chronologies of glacial advances in the {NW} sector of the Cordilleran ice sheet with terrestrial in situ cosmogenic nuclides, Quaternary Sci. Rev., 92, 429--443,
doi:\href{http://dx.doi.org/10.1016/j.quascirev.2013.09.026}{10.1016/j.quascirev.2013.09.026}, 2014.

\bibitem[{Stroeven et~al.(2015)Stroeven, H{\"a}ttestrand, Kleman, Heyman, Fabel, Fredin, Goodfellow, Harbor, Jansen, Olsen, Caffee, Fink, Lundqvist, Rosqvist, Str\"omberg, and Jansson}]{Stroeven.etal.inreview}
Stroeven,~A.~P., H{\"a}ttestrand,~C., Kleman,~J., Heyman,~J., Fabel,~D.,
Fredin,~O., Goodfellow,~B.~W., Harbor,~J.~M., Jansen,~J.~D., Olsen,~L.,
Caffee,~M.~W., Fink,~D., Lundqvist,~J., Rosqvist,~G.~C., Str\"omberg,~B., and
Jansson,~K.~N.: Deglaciation of Fennoscandia, Quaternary Sci. Rev., in
review, 2015.

\bibitem[{Stumpf et~al.(2000)Stumpf, Broster, and Levson}]{Stumpf.etal.2000}
Stumpf,~A.~J., Broster,~B.~E., and Levson,~V.~M.: Multiphase flow of the late Wisconsinan Cordilleran ice sheet in western Canada, Geol. Soc. Am. Bull., 112, 1850--1863,
doi:\href{http://dx.doi.org/10.1130/0016-7606(2000)112<1850:mfotlw>2.0.co;2}{10.1130/0016-7606(2000)112\textless1850:mfotlw\textgreater2.0.co;2}, 2000.

\bibitem[{Taylor et~al.(2014)Taylor, Hendy, and Pak}]{Taylor.etal.2014}
Taylor,~M., Hendy,~I., and Pak,~D.: Deglacial ocean warming and marine margin retreat of the Cordilleran Ice Sheet in the North Pacific Ocean, Earth Planet. Sc. Lett., 403, 89--98,
doi:\href{http://dx.doi.org/10.1016/j.epsl.2014.06.026}{10.1016/j.epsl.2014.06.026}, 2014.

\bibitem[{the PISM~authors(2015)}]{PISM-authors.2015}
the PISM~authors: {PISM}, a~{P}arallel {I}ce {S}heet {M}odel, available at:
\url{http://www.pism-docs.org} (last access: 2015), 2015.

\bibitem[{Troost(2014)}]{Troost.2014}
Troost,~K.~G.: The penultimate glaciation and mid-to late-pleistocene
stratigraphy in the Central Puget Lowland, Washington, in: 2014 GSA Annual
Meeting, Vancouver, 19--22 October 2014, abstract no. 138-9, 2014.

\bibitem[{Tulaczyk et~al.(2000)Tulaczyk, Kamb, and Engelhardt}]{Tulaczyk.etal.2000}
Tulaczyk,~S., Kamb,~W.~B., and Engelhardt,~H.~F.: Basal mechanics of Ice Stream B, west Antarctica: 1. Till mechanics,~J. Geophys. Res., 105, 463,
doi:\href{http://dx.doi.org/10.1029/1999jb900329}{10.1029/1999jb900329}, 2000.

\bibitem[{Turner et~al.(2013)Turner, Ward, Bond, Jensen, Froese, Telka, Zazula, and Bigelow}]{Turner.etal.2013}
Turner,~D.~G., Ward,~B.~C., Bond,~J.~D., Jensen,~B.~J., Froese,~D.~G., Telka,~A.~M., Zazula,~G.~D., and Bigelow,~N.~H.: Middle to Late Pleistocene ice extents, tephrochronology and paleoenvironments of the White River area, southwest Yukon, Quaternary Res., 75, 59--77,
doi:\href{http://dx.doi.org/10.1016/j.quascirev.2013.05.011}{10.1016/j.quascirev.2013.05.011}, 2013.

\bibitem[{Ward et~al.(2007)Ward, Bond, and Gosse}]{Ward.etal.2007}
Ward,~B.~C., Bond,~J.~D., and Gosse,~J.~C.: Evidence for a 55--50 ka (early Wisconsin) glaciation of the Cordilleran ice sheet, Yukon Territory, Canada, Quaternary Res., 68, 141--150,
doi:\href{http://dx.doi.org/10.1016/j.yqres.2007.04.002}{10.1016/j.yqres.2007.04.002}, 2007.

\bibitem[{Ward et~al.(2008)Ward, Bond, Froese, and Jensen}]{Ward.etal.2008}
Ward,~B.~C., Bond,~J.~D., Froese,~D., and Jensen,~B.: Old Crow tephra ($140\pm10$\,ka) constrains penultimate Reid glaciation in central Yukon Territory, Quaternary Res., 27, 1909--1915,
doi:\href{http://dx.doi.org/10.1016/j.quascirev.2008.07.012}{10.1016/j.quascirev.2008.07.012}, 2008.

\bibitem[{Winkelmann et~al.(2011)Winkelmann, Martin, Haseloff, Albrecht, Bueler, Khroulev, and Levermann}]{Winkelmann.etal.2011}
 Winkelmann,~R., Martin,~M.~A., Haseloff,~M., Albrecht,~T., Bueler,~E., Khroulev,~C., and Levermann,~A.: The Potsdam Parallel Ice Sheet Model (PISM-PIK) -- Part 1: Model description, The Cryosphere, 5, 715--726,
doi:\href{http://dx.doi.org/10.5194/tc-5-715-2011}{10.5194/tc-5-715-2011}, 2011.

\end{thebibliography}

\begin{table}[t]
\caption{Parameter values used in the ice sheet model.}
%\scalebox{.85}[.85]
{\begin{tabular}{llcc}
\tophline
    Not.    & Name & Value & Unit \\
\middlehline
  $\rho$  & Ice density
            & 910
            & \unit{kg\,m^{-3}} \\

    $g$     & Standard gravity
            & 9.81
            & \unit{m\,s^{-2}} \\
\cline{1-4}
    \multicolumn{2}{l}{{Basal sliding}} \\
\cline{1-4}
    $q$     & Pseudo-plastic sliding exponent
            & 0.25
            & -- \\

    $v_{\text{th}}$& Pseudo-plastic threshold velocity
            & 100.0
            & \unit{m\,yr^{-1}} \\

    $c_0$   & Till cohesion
            & 0.0
            & Pa \\

    $\delta$& Effective pressure coefficient
            & 0.02
            & -- \\

    $e_0$   & Till reference void ratio
            & 0.69
            & -- \\

    $C_{\mathrm{c}}$   & Till compressibility coefficient
            & 0.12
            & -- \\

    $W_{\text{max}}$ & Maximal till water thickness
            & 2.0
            & m \\
\cline{1-4}
    \multicolumn{2}{l}{{Bedrock and lithosphere}} \\
\cline{1-4}
    $\rho_{\mathrm{b}}$& Bedrock density
            & 3300
            & \unit{kg\,m^{-3}} \\

    $c_{\mathrm{b}}$   & Bedrock specific heat capacity
            & 1000
            & \unit{J\,kg^{-1}\,K^{-1}} \\

    $k_{\mathrm{b}}$   & Bedrock thermal conductivity
            & 3.0
            & \unit{J\,m^{-1}\,K^{-1}\,s^{-1}} \\

    $\nu_m$ & Astenosphere viscosity
            & $1\times10^{19}$
            & \unit{Pa\,s} \\

    $\rho_l$& Lithosphere density
            & 3300
            & \unit{kg\,m^{-3}} \\

    $D$     & Lithosphere flexural rigidity
            & $5.0\times10^{24}$
            & \unit{N} \\
\cline{1-4}
    \multicolumn{2}{l}{{Surface and atmosphere}} \\
\cline{1-4}
    $T_{\mathrm{s}}$   & Temperature of snow precipitation
            & 273.15
            & \unit{K} \\

    $T_{\mathrm{r}}$   & Temperature of rain precipitation
            & 275.15
            & \unit{K} \\

    $F_{\mathrm{s}}$   & Degree-day factor for snow
            & $3.04\times10^{-3}$
            & \unit{m\,K^{-1}\,day^{-1}} \\

    $F_{\mathrm{i}}$   & Degree-day factor for ice
            & $4.59\times10^{-3}$
            & \unit{m\,K^{-1}\,day^{-1}} \\

    $\gamma$& Air temperature lapse rate
            & $6\times10^{-3}$
            & \unit{K\,m^{-1}} \\
\bottomhline
\end{tabular}}
%\hack{
%\setlength\tabularwidth{0.9\tabularwidth}
%}
%\scalebox{.7}[.7]{
\belowtable{%
%\hack{\vspace*{2mm}}
}\label{tab:params}
%}
\end{table}

\begin{table}[t]
  \caption{%
Palaeo-temperature proxy records and scaling parameters yielding
           temperature offset time-series used to force the ice sheet model
           through the last glacial cycle (Fig.~\ref{fig:timeseries}). $f$
           corresponds to the scaling factor adopted to yield last glacial
           maximum ice limits in the vicinity of mapped end moraines, and
           $T_{[32, 22]}$ refers to the resulting mean temperature anomaly
           during the period $-$32 to~$-$22\,\unit{ka} after scaling.}
\label{tab:records}
  {
\scalebox{.850}[.850]
{\begin{tabular}{lccccccl}
\tophline
    Record & Latitude & Longitude & Elev. & Proxy & $f$ & $T_{[32, 22]}$
           & Reference\\
    & & & (\unit{m\,a.s.l.}) & & & (K) & \\
    \middlehline
    GRIP     &  72{\degree}35$^{\prime}$\,N   % 72.58 (decimal)
             &  37{\degree}38$^{\prime}$\,W   % 37.64 (decimal)
             & 3238
             & \chem{\delta^{18}O}
             & 0.37 & $-$5.8  % $-$16.4126 (before scaling)
             & \citet{Dansgaard.etal.1993} \\

    NGRIP    &  75{\degree}06$^{\prime}$\,N   % 75.10
             &  42{\degree}19$^{\prime}$\,W   % 42.32
             & 2917
             & \chem{\delta^{18}O}
             & 0.24 & $-$6.5  % $-$26.7098
             & \citet{Andersen.etal.2004} \\

    EPICA    &  75{\degree}06$^{\prime}$\,S   % 75.1
             & 123{\degree}21$^{\prime}$\,E   % 123.35
             & 3233
             & \chem{\delta^{18}O}
             & 0.64 & $-$5.9  % $-$9.2055
             & \citet{Jouzel.etal.2007} \\

    Vostok   &  78{\degree}28$^{\prime}$\,S   % 78.8
             & 106{\degree}50$^{\prime}$\,E   % 106.8
             & 3488
             & \chem{\delta^{18}O}
             & 0.74 & $-$5.9  % $-$7.9550
             & \citet{Petit.etal.1999} \\

    ODP~1012 &  32{\degree}17$^{\prime}$\,N
             & 118{\degree}23$^{\prime}$\,W
             & $-$1772
             & \chem{U^{K'}_{37}}
             & 1.61 & $-$6.1  % $-$3.7889
             & \citet{Herbert.etal.2001} \\

    ODP~1020 &  41{\degree}00$^{\prime}$\,N
             & 126{\degree}26$^{\prime}$\,W
             & $-$3038
             & \chem{U^{K'}_{37}}
             & 1.20 & $-$6.0 % $-$5.0000
             & \citet{Herbert.etal.2001} \\
    \bottomhline
\end{tabular}}}
\end{table}

\begin{table}
  \caption{%
Extremes in Cordilleran ice sheet volume and extent corresponding to
           MIS~4, 3 and 2 for each of the six low-resolution simulations
           (Fig.~\ref{fig:timeseries}).}
\label{tab:extrema}
  {
%\scalebox{.50}[.50]
{\begin{tabular}{l|ccc|ccc|ccc}
    \tophline
             & \multicolumn{3}{c|}{Age (ka)}
             & \multicolumn{3}{c|}{Ice extent (\unit{10^6\,km^2})}
             & \multicolumn{3}{c}{Ice volume (m\,s.l.e.)} \\
    Record   &  MIS~4 &  MIS~3 &  MIS~2
             &  MIS~4 &  MIS~3 &  MIS~2
             &  MIS~4 &  MIS~3 &  MIS~2 \\
    \middlehline
    GRIP     &  57.45 &  42.93 &  19.14
             &   1.87 &   0.69 &   2.07
             &   7.35 &   1.69 &   8.71 \\
    NGRIP    &  60.25 &  45.86 &  22.85
             &   2.11 &   0.74 &   2.10
             &   8.84 &   1.82 &   8.76 \\
    EPICA    &  61.89 &  45.30 &  17.23
             &   1.51 &   1.01 &   2.09
             &   5.11 &   2.74 &   8.91 \\
    Vostok   &  60.86 &  41.25 &  16.84
             &   1.51 &   1.03 &   2.01
             &   5.15 &   2.88 &   8.40 \\
    ODP 1012 &  56.46 &  47.39 &  23.19
             &   1.38 &   0.90 &   2.07
             &   4.39 &   2.35 &   8.62 \\
    ODP 1020 &  60.23 &  52.97 &  20.56
             &   1.25 &   0.72 &   2.09
             &   3.84 &   1.75 &   8.81 \\
    \cline{1-10}
    Minimum  &  61.89 &  52.97 &  23.19
             &   1.25 &   0.69 &   2.01
             &   3.84 &   1.69 &   8.40 \\
    Maximum  &  56.46 &  41.25 &  16.84
             &   2.11 &   1.03 &   2.10
             &   8.84 &   2.88 &   8.91 \\
    \bottomhline
\end{tabular}}}
\end{table}

\begin{figure}%
\includegraphics[width=80mm]{tc-2015-114-discussions-f01.pdf}
\caption{%
Relief map of northern North America showing a~reconstruction of the
           areas once covered by the Cordilleran (CIS), Laurentide (LIS),
           Innuitian (IIS), and Greenland (GIS) ice sheets during the last
           18\,\unit{\chem{^{14}C}\,ka} \citep[21.4\,cal\,ka,][]{Dyke.2004}.
           The rectangular box denotes the location of the
           modelling domain used in this study. Major mountain ranges covered
           by the ice sheet include the Alaska Range (AR), the Wrangell and
           Saint Elias mountains (WSEM), the Selwyn and Mackenzie mountains
           (SMKM), the Skeena Mountains (SM), the Coast Mountains (CM), the
           Columbia and Rocky
           Mountains (CRM), and the North Cascades (NC). Major depressions
           include the Liard Lowland (LL), the Interior Plateau of British
           Columbia (IP), and the Puget Lowland (PL). The background
           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
           Data \citep{Patterson.Kelso.2015}.}
\label{fig:locmap}%
\end{figure}%

\begin{figure}%
\includegraphics[height=90mm]{tc-2015-114-discussions-f02.pdf}
\caption{%
Monthly mean near-surface air temperature, precipitation, and
           standard deviation of daily mean temperature (PDD SD) for January
           and July from the North American Regional Reanalysis
           \citep[NARR;][]{Mesinger.etal.2006}, used to force the surface mass
           balance (PDD) component of the ice sheet model. Note the
           strong contrasts in seasonality, timing of the precipitation peak,
           and temperature variability over the model domain, notably between
           coastal and inland regions.}
\label{fig:atm}%
\end{figure}%

\begin{figure}%
\includegraphics[width=120mm]{tc-2015-114-discussions-f03.pdf}
\caption{%
Temperature offset time-series from ice core and ocean records
           (Table~\ref{tab:records}) used as palaeo-climate forcing for the ice
           sheet model (top panel), and modelled ice volume (bottom panel)
           through the last 120\,\unit{ka}. Ice volumes are expressed in meters of sea
           level equivalent (m\,s.l.e.). Gray fields indicate Marine Oxygen
           Isotope Stage (MIS) boundaries for MIS~2 and MIS~4 according to
           a~global compilation of benthic \chem{\delta^{18}O} records
           \citep{Lisiecki.Raymo.2005}. Hatched rectangles highlight the
           time-volume span for ice volume extremes corresponding to MIS~4
           (61.9--56.5\,\unit{ka}), MIS~3 (53.0--41.3\,\unit{ka}), and MIS~2 (LGM,
           23.2--16.8\,\unit{ka}). Dotted lines correspond to GRIP- and EPICA-driven
           5\,\unit{km}-resolution runs.}
\label{fig:timeseries}%
\end{figure}%

\begin{figure}%
\includegraphics[width=120mm]{tc-2015-114-discussions-f04.pdf}
\caption{%
Snapshots of modelled surface topography (500\,\unit{m} contours)
           corresponding to the ice volume extremes indicated on
           Fig.~\ref{fig:timeseries}. An ice cap persists over the Skeena
           Mountains (SM) during MIS~3. Note the occurence of spatial
           similarities despite large differences in timing.}
\label{fig:snapshots}%
\end{figure}%

\begin{figure}%
\includegraphics[width=120mm]{tc-2015-114-discussions-f05.pdf}
\caption{%
Modelled surface topography (200\,\unit{m} contours) and surface velocity
           (colour mapping) corresponding to the maximum ice volume during
           MIS~2 in the GRIP and EPICA high-resolution simulations.}
\label{fig:mis2}%
\end{figure}%

\begin{figure}%
\includegraphics[width=120mm]{tc-2015-114-discussions-f06.pdf}
\caption{%
Modelled surface topography (200\,\unit{m} contours) and surface velocity
           (colour mapping) corresponding to the maximum ice volume during
           MIS~4 in the GRIP and EPICA high-resolution simulations.}
\label{fig:mis4}%
\end{figure}%

\begin{figure}%
\includegraphics[width=120mm]{tc-2015-114-discussions-f07.pdf}
\caption{%
Modelled duration of ice cover during the last 120\,\unit{ka} using GRIP
           and EPICA climate forcing.
           Note the irregular colour scale. A~continuous ice cover spanning
           from the Alaska Range (AR) to the Coast Mountains (CM) and
           the Columbia and Rocky mountains (CRM) exists for about 32\,\unit{ka} in
           the GRIP simulation and 26\,\unit{ka} in the EPICA simulation.
           The maximum extent of the ice sheet generally
           corresponds to relatively short durations of ice cover, but ice
           cover persists over the Skeena Mountains (SM) during most of the
           simulation. See Fig.~\ref{fig:locmap} for a~list of abbreviations.}
\label{fig:duration}%
\end{figure}%

\begin{figure}%
\includegraphics[width=120mm]{tc-2015-114-discussions-f08.pdf}
\caption{%
Modelled duration of warm-based ice cover during the last
           120\,\unit{ka}. Long ice cover durations combined with basal
           temperatures at the pressure melting point may explain the strong
           glacial erosional imprint of the Skeena Mountains (SM) landscape.
           Hatches indicate areas that were covered by cold ice only.}
\label{fig:warmbase}%
\end{figure}%

\begin{figure}%
\includegraphics[width=120mm]{tc-2015-114-discussions-f09.pdf}
\caption{%
Modelled fraction of warm-based ice cover during the ice-covered
           period. Note the dominance of warm-based conditions on the
           continental shelf and major glacial troughs of the coastal ranges.
           Hatches indicate areas that were covered by cold ice only.}
\label{fig:warmfrac}%
\end{figure}%

\begin{figure}%
\includegraphics[width=120mm]{tc-2015-114-discussions-f10.pdf}
\caption{%
Temperature offset time-series from the GRIP and EPICA ice core
           records (Table~\ref{tab:records}) (top panel), and modelled ice
           volume during the deglaciation, expressed in meters of sea-level
           equivalent (bottom panel).}
\label{fig:deglacseries}%
\end{figure}%

\begin{figure}%
\includegraphics[width=120mm]{tc-2015-114-discussions-f11.pdf}
\caption{%
Snapshots of modelled surface topography (200\,\unit{m} contours)
           and surface velocity (colour mapping) during the last deglaciation
           from the GRIP (top panels) and EPICA (bottom panels) 5\,\unit{km} simulations.
           Dashed segments \textbf{(A--D)} indicate the location of profiles used in
           Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}.}
\label{fig:deglacshots}%
\end{figure}%

\begin{figure}%
\includegraphics[width=120mm]{tc-2015-114-discussions-f12.pdf}
\caption{%
Modelled age of the last deglaciation. Areas that have been covered
           only before the last glacial maximum are marked
           in green. Hatches denote re-advance of mountain-centred ice caps and
           the decaying ice sheet between 14 and 10\,\unit{ka}.
           Dashed segments \textbf{(A--D)} indicate the location of profiles used in
           Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}.}
\label{fig:deglac}%
\end{figure}%

\begin{figure}%
\includegraphics[width=120mm]{tc-2015-114-discussions-f13.pdf}
\caption{%
Modelled deglacial basal ice velocities. Hatches
           indicate areas that remain non-sliding throughout deglaciation
           (22.0--8.0\,\unit{ka}), notably including parts of the Interior Plateau (IP).
           Note the concentric patterns of deglacial flow in the Liard
           Lowland (LL).
           Sliding grid cells were distinguished from non-sliding grid cells
           using a~basal velocity threshold of 1\,\unit{m\,yr^{-1}}.}
\label{fig:lastflow}%
\end{figure}%

\begin{figure}%
\includegraphics[height=100mm]{tc-2015-114-discussions-f14.pdf}
\caption{%
Modelled bedrock (black) and ice surface (blue) topography profiles
           during deglaciation (22.0--8.0\,\unit{ka}) in the GRIP 5\,\unit{km}
           simulation, corresponding to the four transects indicated in
           Figs.~\ref{fig:deglacshots} and~\ref{fig:deglac}.}
\label{fig:profiles-grip}%
\end{figure}%

\begin{figure}%
\includegraphics[height=100mm]{tc-2015-114-discussions-f15.pdf}
\caption{%
Modelled bedrock (black) and ice surface (red) topography profiles
           during deglaciation (22.0--8.0\,\unit{ka}) in the EPICA 5\,\unit{km}
           simulation, corresponding to the four transects indicated in
           Figs.~\ref{fig:deglacshots} and~\ref{fig:deglac}.}
\label{fig:profiles-epica}%
\end{figure}

\end{document}
\endinput%%%%%%

### 2016-01-16 cordillera-cycle.tex
Typeset: use 6-space indentation and 76 char line.
To fit with typeset version which I am supposed to use now.

@@
 
-During the last glacial cycle, glaciers and ice caps of the North American
-Cordillera have been more extensive than today. At the Last Glacial
-Maximum (LGM),
-a continuous blanket of ice, the Cordilleran ice sheet
-\citep{Dawson.1888}, stretched from the Alaska Range in the north to the
-North Cascades in the south (Fig.~\ref{fig:locmap}).
-In addition, it extended offshore, where it calved
-into the Pacific Ocean, and merged with the western margin of its much larger
+      During the last glacial cycle, glaciers and ice caps of the North
+      American Cordillera have been more extensive than today. At the Last
+      Glacial Maximum (LGM), a continuous blanket of ice, the Cordilleran
+      ice sheet \citep{Dawson.1888}, stretched from the Alaska Range in the
+      north to the North Cascades in the south (Fig.~\ref{fig:locmap}). In
+      addition, it extended offshore, where it calved into the Pacific
+      Ocean, and merged with the western margin of its much larger
       neighbour, the Laurentide ice sheet, east of the Rocky Mountains.
@@
       More than a century of exploration and geological investigation of the
-Cordilleran mountains have led to many observations in support of the former
-ice sheet
-    \citep{Jackson.Clague.1991}.
-Despite the lack of documented end moraines offshore, in the zone of confluence
-with the Laurentide ice sheet, and in areas swept by the Missoula floods
-    \citep{Carrara.etal.1996},
-moraines that demarcate the northern and south-western margins provide key
-constraints that allow reasonable reconstructions of maximum ice sheet extents
+      Cordilleran mountains have led to many observations in support of the
+      former ice sheet \citep{Jackson.Clague.1991}. Despite the lack of
+      documented end moraines offshore, in the zone of confluence with the
+      Laurentide ice sheet, and in areas swept by the Missoula floods
+      \citep{Carrara.etal.1996}, moraines that demarcate the northern and
+      south-western margins provide key constraints that allow reasonable
+      reconstructions of maximum ice sheet extents
       (\citealp{Prest.etal.1968}; \citealp[Fig. 1.12]{Clague.1989};
-     \citealp{Duk-Rodkin.1999};
-     \citealp{Booth.etal.2003}; \citealp{Dyke.2004}).
-As indicated by field evidence from radiocarbon dating
-    \citep{Clague.etal.1980, Clague.1985, Clague.1986, Porter.Swanson.1998,
-           Menounos.etal.2008},
-cosmogenic exposure dating
+      \citealp{Duk-Rodkin.1999}; \citealp{Booth.etal.2003};
+      \citealp{Dyke.2004}). As indicated by field evidence from radiocarbon
+      dating \citep{Clague.etal.1980, Clague.1985, Clague.1986,
+      Porter.Swanson.1998, Menounos.etal.2008}, cosmogenic exposure dating
       \citep{Stroeven.etal.2010, Stroeven.etal.2014, Margold.etal.2014},
       bedrock deformation in response to former ice loads
-    \citep{Clague.James.2002, Clague.etal.2005},
-and offshore sedimentary records
-    \citep{Cosma.etal.2008, Davies.etal.2011},
-the LGM Cordilleran ice sheet extent was short-lived.
-However, former ice thicknesses and, therefore, the ice sheet's contribution to
-the LGM sea level lowstand
-    \citep{Carlson.Clark.2012, Clark.Mix.2002}
-remain uncertain.
-
-Our understanding of the Cordilleran glaciation history prior to the LGM is
-even more fragmentary
-    \citep{Barendregt.Irving.1998, Kleman.etal.2010, Rutter.etal.2012},
-although it is clear that the Pleistocene maximum extent of the Cordilleran
-ice sheet predates the last glacial cycle
-    \citep{Hidy.etal.2013}.
-In parts of the Yukon Territory and Alaska, and in the Puget Lowland,
-the distribution of tills
-    \citep{Turner.etal.2013, Troost.2014}
-and dated glacial erratics
+      \citep{Clague.James.2002, Clague.etal.2005}, and offshore sedimentary
+      records \citep{Cosma.etal.2008, Davies.etal.2011}, the LGM Cordilleran
+      ice sheet extent was short-lived. However, former ice thicknesses and,
+      therefore, the ice sheet's contribution to the LGM sea level lowstand
+      \citep{Carlson.Clark.2012, Clark.Mix.2002} remain uncertain.
+
+      Our understanding of the Cordilleran glaciation history prior to the
+      LGM is even more fragmentary \citep{Barendregt.Irving.1998,
+      Kleman.etal.2010, Rutter.etal.2012}, although it is clear that the
+      Pleistocene maximum extent of the Cordilleran ice sheet predates the
+      last glacial cycle \citep{Hidy.etal.2013}. In parts of the Yukon
+      Territory and Alaska, and in the Puget Lowland, the distribution of
+      tills \citep{Turner.etal.2013, Troost.2014} and dated glacial erratics
       \citep{Ward.etal.2007, Ward.etal.2008, Briner.Kaufman.2008,
-           Stroeven.etal.2010, Stroeven.etal.2014}
-indicate an extensive Marine Oxygen Isotope Stage (MIS)~4 glaciation.
-Landforms in the interior regions include flow sets that are likely
-older than the LGM
-    \citep[Fig.~2]{Kleman.etal.2010},
-but their absolute age remains uncertain.
+      Stroeven.etal.2010, Stroeven.etal.2014} indicate an extensive Marine
+      Oxygen Isotope Stage (MIS)~4 glaciation. Landforms in the interior
+      regions include flow sets that are likely older than the LGM
+      \citep[Fig.~2]{Kleman.etal.2010}, but their absolute age remains
+      uncertain.
 
@@
       In contrast, evidence for the deglaciation history of the Cordilleran
-ice sheet since the LGM is considerable, albeit mostly at a regional scale.
-Geomorphological evidence from south-central British Columbia indicates a rapid
-deglaciation, including an early emergence of elevated areas while thin,
-stagnant ice still covered the surrounding lowlands
-    \citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a}.
-This model, although credible, may not apply in all areas of the Cordilleran
-ice sheet
-    \citep{Margold.etal.2013}.
-Although solid evidence for late-glacial glacier re-advances have been found in
-the Coast, Columbia and Rocky mountains
+      ice sheet since the LGM is considerable, albeit mostly at a regional
+      scale. Geomorphological evidence from south-central British Columbia
+      indicates a rapid deglaciation, including an early emergence of
+      elevated areas while thin, stagnant ice still covered the surrounding
+      lowlands \citep{Fulton.1967, Fulton.1991, Margold.etal.2011,
+      Margold.etal.2013a}. This model, although credible, may not apply in
+      all areas of the Cordilleran ice sheet \citep{Margold.etal.2013}.
+      Although solid evidence for late-glacial glacier re-advances have been
+      found in the Coast, Columbia and Rocky mountains
       \citep{Clague.etal.1997, Friele.Clague.2002, Friele.Clague.2002a,
       Kovanen.2002, Kovanen.Easterbrook.2002, Lakeman.etal.2008,
-           Menounos.etal.2008},
-it appears to be sparser than for formerly glaciated regions surrounding
-the North Atlantic
-    \citep[e.g.,][]{Sissons.1979, Lundqvist.1987,
-                    Ivy-Ochs.etal.1999, Stea.etal.2011}.
-Nevertheless, recent oxygen isotope measurements from Gulf of Alaska sediments
-reveal a climatic evolution highly correlated to that of Greenland during this
-period, including a distinct Late Glacial cold reversal between 14.1 and
-11.7\,ka \citep{Praetorius.Mix.2014}.
-
-In general, the topographic complexity of the North American Cordillera and its
-effect on glacial history have inhibited the reconstruction of ice sheet-wide
-glacial advance and retreat patterns such as
-those available for the Fennoscandian and Laurentide ice sheets
-     \citep{Boulton.etal.2001, Dyke.Prest.1987, Dyke.etal.2003,
-            Kleman.etal.1997, Kleman.etal.2010, Stroeven.etal.inreview}.
-Here, we use a numerical ice sheet model \citep{PISM-authors.2015},
-calibrated against field-based evidence, to perform a quantitative
-reconstruction of the Cordilleran ice sheet evolution through the last glacial
-cycle, and
-analyse some of the long-standing questions related to its evolution:
+      Menounos.etal.2008}, it appears to be sparser than for formerly
+      glaciated regions surrounding the North Atlantic
+      \citep[e.g.,][]{Sissons.1979, Lundqvist.1987, Ivy-Ochs.etal.1999,
+      Stea.etal.2011}. Nevertheless, recent oxygen isotope measurements from
+      Gulf of Alaska sediments reveal a climatic evolution highly correlated
+      to that of Greenland during this period, including a distinct Late
+      Glacial cold reversal between 14.1 and 11.7\,ka
+      \citep{Praetorius.Mix.2014}.
+
+      In general, the topographic complexity of the North American
+      Cordillera and its effect on glacial history have inhibited the
+      reconstruction of ice sheet-wide glacial advance and retreat patterns
+      such as those available for the Fennoscandian and Laurentide ice
+      sheets \citep{Boulton.etal.2001, Dyke.Prest.1987, Dyke.etal.2003,
+      Kleman.etal.1997, Kleman.etal.2010, Stroeven.etal.inreview}. Here, we
+      use a numerical ice sheet model \citep{PISM-authors.2015}, calibrated
+      against field-based evidence, to perform a quantitative reconstruction
+      of the Cordilleran ice sheet evolution through the last glacial cycle,
+      and analyse some of the long-standing questions related to its
+      evolution:
 
@@
 
-Although numerical ice sheet modelling has been established as a useful tool to
-improve our understanding of the Cordilleran ice sheet
+      Although numerical ice sheet modelling has been established as a
+      useful tool to improve our understanding of the Cordilleran ice sheet
       (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991};
-     \citealp{Marshall.etal.2000}),
-the ubiquitously mountainous
-topography of the region has presented two major challenges to its application.
-First, only recent developments in numerical ice sheet models and underlying
-scientific computing tools \citep{Bueler.Brown.2009, Balay.etal.2015} have
-allowed for high-resolution numerical modelling of glaciers and ice sheets on
-mountainous terrain
-over millennial time scales \citep[e.g.,][]{Golledge.etal.2012}. Second, the
-complex topography of the North American Cordillera also induces strong
-geographic variations in temperature and precipitation, thus requiring the use
-of high-resolution climate forcing fields as an input to an ice sheet model
-\citep{Seguinot.etal.2014}. However, evolving climate conditions over the last
-glacial cycle are subject to considerable uncertainty and still lie beyond the
-computational reach of atmosphere circulation models.
+      \citealp{Marshall.etal.2000}), the ubiquitously mountainous topography
+      of the region has presented two major challenges to its application.
+      First, only recent developments in numerical ice sheet models and
+      underlying scientific computing tools \citep{Bueler.Brown.2009,
+      Balay.etal.2015} have allowed for high-resolution numerical modelling
+      of glaciers and ice sheets on mountainous terrain over millennial time
+      scales \citep[e.g.,][]{Golledge.etal.2012}. Second, the complex
+      topography of the North American Cordillera also induces strong
+      geographic variations in temperature and precipitation, thus requiring
+      the use of high-resolution climate forcing fields as an input to an
+      ice sheet model \citep{Seguinot.etal.2014}. However, evolving climate
+      conditions over the last glacial cycle are subject to considerable
+      uncertainty and still lie beyond the computational reach of atmosphere
+      circulation models.
 
@@
       precipitation grids derived from a present-day atmospheric reanalysis
-\citep{Mesinger.etal.2006} that includes the steep precipitation gradients
-previously identified as necessary to model the LGM extent of the Cordilleran
-ice sheet in agreement with its geological imprint
-\citep{Seguinot.etal.2014}. To mimic climate evolution through the last glacial
-cycle, these grids are simply supplemented by lapse-rate corrections
-and temperature offset time series. The latter are obtained by scaling six
-different palaeo-temperature reconstructions from proxy records around the
-globe, including two oxygen isotope records from Greenland ice cores
-\citep{Dansgaard.etal.1993, Andersen.etal.2004}, two oxygen isotope
-records from Antarctic ice cores \citep{Petit.etal.1999,Jouzel.etal.2007},
-and two alkenone unsaturation index records from Northwest Pacific ocean
-sediment cores \citep{Herbert.etal.2001}. We then proceed to compare the model
-output to geological evidence and discuss the timing and extent of glaciation
-and the patterns of deglaciation, based on which we use the applicability of
-different records to modelling the history of the Cordilleran ice sheet.
+      \citep{Mesinger.etal.2006} that includes the steep precipitation
+      gradients previously identified as necessary to model the LGM extent
+      of the Cordilleran ice sheet in agreement with its geological imprint
+      \citep{Seguinot.etal.2014}. To mimic climate evolution through the
+      last glacial cycle, these grids are simply supplemented by lapse-rate
+      corrections and temperature offset time series. The latter are
+      obtained by scaling six different palaeo-temperature reconstructions
+      from proxy records around the globe, including two oxygen isotope
+      records from Greenland ice cores \citep{Dansgaard.etal.1993,
+      Andersen.etal.2004}, two oxygen isotope records from Antarctic ice
+      cores \citep{Petit.etal.1999,Jouzel.etal.2007}, and two alkenone
+      unsaturation index records from Northwest Pacific ocean sediment cores
+      \citep{Herbert.etal.2001}. We then proceed to compare the model output
+      to geological evidence and discuss the timing and extent of glaciation
+      and the patterns of deglaciation, based on which we use the
+      applicability of different records to modelling the history of the
+      Cordilleran ice sheet.
 
@@
 
-The simulations presented here were run using the Parallel Ice Sheet Model
-(PISM, development version~8ff7cbe), an open source,
-finite difference, shallow ice sheet model \citep{PISM-authors.2015}. The model
-requires input on basal topography, sea level, geothermal heat flux and
-climate forcing. It computes the evolution of ice extent
-and thickness over time, the thermal and dynamic
-states of the ice sheet, and the associated lithospheric response.
-
-Basal topography is derived from the ETOPO1 combined topography and bathymetry
-dataset with a~resolution of 1\,arc-min \citep{Amante.Eakins.2009}. Sea level
-is lowered as a function of time based on the Spectral Mapping Project
+      The simulations presented here were run using the Parallel Ice Sheet
+      Model (PISM, development version~8ff7cbe), an open source, finite
+      difference, shallow ice sheet model \citep{PISM-authors.2015}. The
+      model requires input on basal topography, sea level, geothermal heat
+      flux and climate forcing. It computes the evolution of ice extent and
+      thickness over time, the thermal and dynamic states of the ice sheet,
+      and the associated lithospheric response.
+
+      Basal topography is derived from the ETOPO1 combined topography and
+      bathymetry dataset with a~resolution of 1\,arc-min
+      \citep{Amante.Eakins.2009}. Sea level is lowered as a function of time
+      based on the Spectral Mapping Project
       \citep[SPECMAP,][]{Imbrie.etal.1989} time scale. Geothermal heat flux
       is applied as a constant value of 70\,\unit{mW\,m^{-2}} at 3\,km depth
-(Sect.~\ref{sec:icedyn}). Surface mass balance is computed using a positive
-degree-day (PDD) model (Sect.~\ref{sec:surface}). Climate forcing is
-provided by a monthly climatology averaged from 1979 to 2000 from the North
-American Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006},
-perturbated by time-dependent offsets and
-lapse-rate temperature corrections (Sect.~\ref{sec:atm}).
-
-Each simulation starts from assumed ice-free conditions at 120000 years ago
-(120\,ka), and runs to the present. Our modelling domain of 1500 by 3000\,km
-encompasses the entire area covered by the Cordilleran ice sheet at the LGM
-(Fig.~\ref{fig:locmap}). The simulations were run on two distinct grids, using
-a lower horizontal resolution of 10\,km, and a higher horizontal resolution of
-5\,km. These computations were performed on 16 to 128 computing cores at the
+      (Sect.~\ref{sec:icedyn}). Surface mass balance is computed using a
+      positive degree-day (PDD) model (Sect.~\ref{sec:surface}). Climate
+      forcing is provided by a monthly climatology averaged from 1979 to
+      2000 from the North American Regional Reanalysis
+      \citep[NARR,][]{Mesinger.etal.2006}, perturbated by time-dependent
+      offsets and lapse-rate temperature corrections (Sect.~\ref{sec:atm}).
+
+      Each simulation starts from assumed ice-free conditions at 120000
+      years ago (120\,ka), and runs to the present. Our modelling domain of
+      1500 by 3000\,km encompasses the entire area covered by the
+      Cordilleran ice sheet at the LGM (Fig.~\ref{fig:locmap}). The
+      simulations were run on two distinct grids, using a lower horizontal
+      resolution of 10\,km, and a higher horizontal resolution of 5\,km.
+      These computations were performed on 16 to 128 computing cores at the
       Swedish National Supercomputing Centre.
@@
 
-Ice sheet dynamics are typically modelled using a combination of internal
-deformation and basal sliding. PISM is a~shallow ice sheet model, which implies
-that the balance of stresses is approximated based on their predominant
-components. The Shallow Shelf Approximation (SSA) is used as a ``sliding law''
-for the Shallow Ice Approximation (SIA) by adding velocity solutions of the
-two approximations
+      Ice sheet dynamics are typically modelled using a combination of
+      internal deformation and basal sliding. PISM is a~shallow ice sheet
+      model, which implies that the balance of stresses is approximated
+      based on their predominant components. The Shallow Shelf Approximation
+      (SSA) is used as a ``sliding law'' for the Shallow Ice Approximation
+      (SIA) by adding velocity solutions of the two approximations
       \citep[Eqns.~7--9 and 15]{Bueler.Brown.2009, Winkelmann.etal.2011}.
 
-Ice rheology depends on temperature and water content through an enthalpy
-formulation \citep{Aschwanden.etal.2012}. Surface air temperature derived from
-the climate forcing (Sect.~\ref{sec:atm}) provides the upper boundary
-condition to the ice enthalpy model. Temperature is computed subglacially to
-a~depth of 3\,km,
-where it is conditioned by a lower boundary geothermal heat flux of
-70\,\unit{mW\,m^{-2}}. Although this uniform value does not account for the
-high spatial geothermal variability in the region
-\citep{Blackwell.Richards.2004}, it is, on average, representative of available
-heat flow measurements. In the low-resolution simulations, the vertical grid
-consists of 31~temperature layers in the bedrock and up to 51~enthalpy layers
-in the ice sheet, corresponding to a vertical resolution of 100\,m. The
-high-resolution simulations 61~bedrock layers and up to 101~ice layers with a
-vertical resolution of 50\,m.
+      Ice rheology depends on temperature and water content through an
+      enthalpy formulation \citep{Aschwanden.etal.2012}. Surface air
+      temperature derived from the climate forcing (Sect.~\ref{sec:atm})
+      provides the upper boundary condition to the ice enthalpy model.
+      Temperature is computed subglacially to a~depth of 3\,km, where it is
+      conditioned by a lower boundary geothermal heat flux of
+      70\,\unit{mW\,m^{-2}}. Although this uniform value does not account
+      for the high spatial geothermal variability in the region
+      \citep{Blackwell.Richards.2004}, it is, on average, representative of
+      available heat flow measurements. In the low-resolution simulations,
+      the vertical grid consists of 31~temperature layers in the bedrock and
+      up to 51~enthalpy layers in the ice sheet, corresponding to a vertical
+      resolution of 100\,m. The high-resolution simulations 61~bedrock
+      layers and up to 101~ice layers with a vertical resolution of 50\,m.
 
@@
 \end{equation}
-relates the bed-parallel shear stresses, $\vec{\tau}_b$, to the sliding
-velocity, $\vec{v}_b$ (Table~\ref{tab:params}).
-The yield stress, $\tau_c$,
-is modelled using the Mohr--Coulomb criterion,
+      relates the bed-parallel shear stresses, $\vec{\tau}_b$, to the
+      sliding velocity, $\vec{v}_b$ (Table~\ref{tab:params}). The yield
+      stress, $\tau_c$, is modelled using the Mohr--Coulomb criterion,
 \begin{equation}
@@
 \end{equation}
-where cohesion, $c_0$, is assumed to be zero. The friction angle, $\phi$,
-varies from 15 to 45\degree as a piecewise-linear function of modern bed
-elevation, with the lowest value occuring below modern sea level (0\,m above
-sea level, m~a.s.l.) and the highest value occuring above the generalised
-elevation of the highest shorelines
+      where cohesion, $c_0$, is assumed to be zero. The friction angle,
+      $\phi$, varies from 15 to 45\degree as a piecewise-linear function of
+      modern bed elevation, with the lowest value occuring below modern sea
+      level (0\,m above sea level, m~a.s.l.) and the highest value occuring
+      above the generalised elevation of the highest shorelines
       \citep[200\,m~a.s.l.,][Fig.~5]{Clague.1981}, thus accounting for
       a~weakening of till associated with the presence of marine sediments
-(cf. \citealp{Martin.etal.2011}; \citealp[supplement]{Aschwanden.etal.2013};
-\citealp{PISM-authors.2015}). Effective pressure, $N$,
-is related to the ice overburden stress, $\rho gh$, and the modelled amount of
-subglacial water, using a formula derived from laboratory experiments with till
-extracted from the base of Ice Stream B in West Antarctica
+      (cf. \citealp{Martin.etal.2011};
+      \citealp[supplement]{Aschwanden.etal.2013};
+      \citealp{PISM-authors.2015}). Effective pressure, $N$, is related to
+      the ice overburden stress, $\rho gh$, and the modelled amount of
+      subglacial water, using a formula derived from laboratory experiments
+      with till extracted from the base of Ice Stream B in West Antarctica
       \citep{Tulaczyk.etal.2000, Bueler.Pelt.2015},
@@
 \end{equation}
-where $\delta$ is chosen as 0.02, $e_0$ is a measured reference void ratio and
-$C_c$ is a measured compressibility coefficient (Table~\ref{tab:params}). The
-amount of water at the base, $W$, varies from zero to $W_{max}=2$\,m, a
-threshold above which additional melt water is assumed to drain off
-instantaneously. Finally, the bedrock topography responds to ice load
-following a bedrock deformation model that includes local isostasy,
-elastic lithosphere flexure and viscous astenosphere deformation in an infinite
-half-space \citep{Lingle.Clark.1985,Bueler.etal.2007}.
-A relatively low viscosity value of $\nu_m = 1\times10^{19}$\,\unit{Pa\,s} is
-used for the astenosphere (Table~\ref{tab:params}) in accordance with the
-results from regional glacial
-isostatic adjustment modelling at the northern Cascadia subduction zone
-\citep{James.etal.2009}.
+      where $\delta$ is chosen as 0.02, $e_0$ is a measured reference void
+      ratio and $C_c$ is a measured compressibility coefficient
+      (Table~\ref{tab:params}). The amount of water at the base, $W$, varies
+      from zero to $W_{max}=2$\,m, a threshold above which additional melt
+      water is assumed to drain off instantaneously. Finally, the bedrock
+      topography responds to ice load following a bedrock deformation model
+      that includes local isostasy, elastic lithosphere flexure and viscous
+      astenosphere deformation in an infinite half-space
+      \citep{Lingle.Clark.1985,Bueler.etal.2007}. A relatively low viscosity
+      value of $\nu_m = 1\times10^{19}$\,\unit{Pa\,s} is used for the
+      astenosphere (Table~\ref{tab:params}) in accordance with the results
+      from regional glacial isostatic adjustment modelling at the northern
+      Cascadia subduction zone \citep{James.etal.2009}.
 
       Ice shelf calving is computed using a double criterion. First, a
-physically-realistic calving flux is computed based on eigenvalues of the
-horizontal strain rate tensor \citep{Winkelmann.etal.2011,
+      physically-realistic calving flux is computed based on eigenvalues of
+      the horizontal strain rate tensor \citep{Winkelmann.etal.2011,
       Levermann.etal.2012}. This allows floating ice to advance in confined
-embayments, but prevents the formation of extensive ice shelves in the open ocean.
-Second, floating ice thinner than 50\,m is systematically calved off. A subgrid
-scheme by \citet{Albrecht.etal.2011} allows for a continuous migration of the
-calving front. This formulation of calving has been applied to the Antarctic ice
-sheet and has shown to produce a realistic calving front positon for many of
-the present-day ice shelves \citep{Martin.etal.2011}.
+      embayments, but prevents the formation of extensive ice shelves in the
+      open ocean. Second, floating ice thinner than 50\,m is systematically
+      calved off. A subgrid scheme by \citet{Albrecht.etal.2011} allows for
+      a continuous migration of the calving front. This formulation of
+      calving has been applied to the Antarctic ice sheet and has shown to
+      produce a realistic calving front positon for many of the present-day
+      ice shelves \citep{Martin.etal.2011}.
 
@@
       Ice surface accumulation and ablation are computed from monthly mean
-near-surface air temperature, $T_m$, monthly standard deviation of near-surface
-air temperature, $\sigma$, and monthly precipitation, $P_m$, using
-a~temperature-index model \citep[e.g.,][]{Hock.2003}.
-Accumulation is equal to
-precipitation when air temperatures are below 0\,\unit{{\degree}C}, and
-decreases to zero linearly with temperatures between 0 and
-2\,\unit{{\degree}C}. Ablation is computed from PDD, defined as an
-integral of temperatures above 0\,\unit{{\degree}C} in one year.
-
-The PDD computation accounts for stochastic temperature variations by assuming
-a normal temperature distribution of standard deviation $\sigma$ aroung the
-expected value $T_m$. It is expressed by an error-function formulation
-\citep{Calov.Greve.2005},
+      near-surface air temperature, $T_m$, monthly standard deviation of
+      near-surface air temperature, $\sigma$, and monthly precipitation,
+      $P_m$, using a~temperature-index model \citep[e.g.,][]{Hock.2003}.
+      Accumulation is equal to precipitation when air temperatures are below
+      0\,\unit{{\degree}C}, and decreases to zero linearly with temperatures
+      between 0 and 2\,\unit{{\degree}C}. Ablation is computed from PDD,
+      defined as an integral of temperatures above 0\,\unit{{\degree}C} in
+      one year.
+
+      The PDD computation accounts for stochastic temperature variations by
+      assuming a normal temperature distribution of standard deviation
+      $\sigma$ aroung the expected value $T_m$. It is expressed by an
+      error-function formulation \citep{Calov.Greve.2005},
 \begin{equation}
@@
 \end{equation}
-which is numerically approximated using week-long sub-intervals. In order to
-account for the effects of spatial and seasonal variations of temperature
-variability \citep{Seguinot.2013}, $\sigma$ is computed from NARR daily
-temperature values from 1979 to 2000 \citep{Mesinger.etal.2006},
-including variability associated with the seasonal cycle (Fig.~\ref{fig:atm}).
-Degree-day factors for snow and ice melt are derived from
-mass-balance measurements on contemporary glaciers from the Coast Mountains and
-Rocky Mountains in British Columbia
-\citep[Table~\ref{tab:params};][]{Shea.etal.2009}.
+      which is numerically approximated using week-long sub-intervals. In
+      order to account for the effects of spatial and seasonal variations of
+      temperature variability \citep{Seguinot.2013}, $\sigma$ is computed
+      from NARR daily temperature values from 1979 to 2000
+      \citep{Mesinger.etal.2006}, including variability associated with the
+      seasonal cycle (Fig.~\ref{fig:atm}). Degree-day factors for snow and
+      ice melt are derived from mass-balance measurements on contemporary
+      glaciers from the Coast Mountains and Rocky Mountains in British
+      Columbia \citep[Table~\ref{tab:params};][]{Shea.etal.2009}.
 
@@
 
-Climate forcing driving ice sheet simulations consists of a present-day monthly
-climatology,
+      Climate forcing driving ice sheet simulations consists of a
+      present-day monthly climatology,
 $\{T_{m0}, P_{m0}\}$, where temperatures are modified by offset time series,
@@
 
-The present-day monthly climatology was computed from
-near-surface air temperature and precipitation rate fields from the NARR,
-averaged from 1979 to 2000. Modern climate of the
-North American Cordillera is characterised by strong geographic variations in
-temperature seasonality, timing of the maximum annual precipitation, and
-daily temperature variability (Fig.~\ref{fig:atm}).
-The use of NARR is motivated by the need for an accurate,
-high-resolution precipitation forcing, as identified in a previous sensitivity
-study \citep{Seguinot.etal.2014}.
+      The present-day monthly climatology was computed from near-surface air
+      temperature and precipitation rate fields from the NARR, averaged from
+      1979 to 2000. Modern climate of the North American Cordillera is
+      characterised by strong geographic variations in temperature
+      seasonality, timing of the maximum annual precipitation, and daily
+      temperature variability (Fig.~\ref{fig:atm}). The use of NARR is
+      motivated by the need for an accurate, high-resolution precipitation
+      forcing, as identified in a previous sensitivity study
+      \citep{Seguinot.etal.2014}.
 
       Temperature offset time-series, ${\Delta}T_{TS}$, are derived from
-palaeo-temperature proxy records from
-the Greenland Ice Core Project \citep[GRIP,][]{Dansgaard.etal.1993}, the
-North Greenland Ice Core Project \citep[NGRIP,][]{Andersen.etal.2004},
-the European Project for Ice Coring in Antarctica \citep[EPICA,][]
-{Jouzel.etal.2007}, the Vostok ice core \citep{Petit.etal.1999}, and Ocean
-Drilling Program (ODP) sites 1012 and 1020, both located off the coast of
-California \citep{Herbert.etal.2001}. Palaeo-temperature anomalies from the
-GRIP and NGRIP
-records were calculated from oxygen isotope (\chem{\delta^{18}O}) measurements
-using a quadratic equation \citep{Johnsen.etal.1995},
+      palaeo-temperature proxy records from the Greenland Ice Core Project
+      \citep[GRIP,][]{Dansgaard.etal.1993}, the North Greenland Ice Core
+      Project \citep[NGRIP,][]{Andersen.etal.2004}, the European Project for
+      Ice Coring in Antarctica \citep[EPICA,][] {Jouzel.etal.2007}, the
+      Vostok ice core \citep{Petit.etal.1999}, and Ocean Drilling Program
+      (ODP) sites 1012 and 1020, both located off the coast of California
+      \citep{Herbert.etal.2001}. Palaeo-temperature anomalies from the GRIP
+      and NGRIP records were calculated from oxygen isotope
+      (\chem{\delta^{18}O}) measurements using a quadratic equation
+      \citep{Johnsen.etal.1995},
 \begin{align}
@@
 \end{align}
-while temperature reconstructions from Antarctic and oceanic cores were
-provided as such. All records were scaled linearly (Table~\ref{tab:records}) in
-order to simulate comparable ice extents at the LGM  (Table~\ref{tab:extrema})
-and realistic outlines (Fig.~\ref{fig:lr_maps}).
-
-Finally, lapse-rate corrections, ${\Delta}T_{LR}$, are computed as a function
-of ice surface elevation, $s$, using the NARR surface geopotential height
-invariant field as a reference topography, $b_{ref}$:
+      while temperature reconstructions from Antarctic and oceanic cores
+      were provided as such. All records were scaled linearly
+      (Table~\ref{tab:records}) in order to simulate comparable ice extents
+      at the LGM (Table~\ref{tab:extrema}) and realistic outlines
+      (Fig.~\ref{fig:lr_maps}).
+
+      Finally, lapse-rate corrections, ${\Delta}T_{LR}$, are computed as a
+      function of ice surface elevation, $s$, using the NARR surface
+      geopotential height invariant field as a reference topography,
+      $b_{ref}$:
 \begin{align}
@@
 
-thus accounting for the evolution of ice thickness, ${h=s-b}$,
-on the one hand, and for differences between the basal topography of the ice
-flow model, $b$, and the
-NARR reference topography, $b_{ref}$, on the other hand. All simulations use an
-annual temperature lapse rate of $\gamma = 6\,\unit{K\,km^{-1}}$.
-In the rest of this paper, we refer to different model runs by the name of the
-proxy record used for the palaeo-temperature forcing.
+      thus accounting for the evolution of ice thickness, ${h=s-b}$, on the
+      one hand, and for differences between the basal topography of the ice
+      flow model, $b$, and the NARR reference topography, $b_{ref}$, on the
+      other hand. All simulations use an annual temperature lapse rate of
+      $\gamma = 6\,\unit{K\,km^{-1}}$. In the rest of this paper, we refer
+      to different model runs by the name of the proxy record used for the
+      palaeo-temperature forcing.
 
@@
       (Fig.~\ref{fig:lr_ts}, upper panel), model output presents consistent
-features that can be observed across the range of forcing data used. In all
-simulations, modelled ice volumes remain relatively low during most of the
-glacial cycle, except during two major glacial events which occur between 61.9
-and 56.5\,ka during MIS~4, and between 23.2 and 16.8\,ka during MIS~2
-(Fig.~\ref{fig:lr_ts}, lower panel). An ice volume minimum is
-consistently reached between 53.0 and 41.3\,ka during MIS~3. However, the
-magnitude and precise timing of these three events depend significantly on the
-choice of proxy record used to derive a time-dependent climate forcing
-(Table~\ref{tab:extrema}).
+      features that can be observed across the range of forcing data used.
+      In all simulations, modelled ice volumes remain relatively low during
+      most of the glacial cycle, except during two major glacial events
+      which occur between 61.9 and 56.5\,ka during MIS~4, and between 23.2
+      and 16.8\,ka during MIS~2 (Fig.~\ref{fig:lr_ts}, lower panel). An ice
+      volume minimum is consistently reached between 53.0 and 41.3\,ka
+      during MIS~3. However, the magnitude and precise timing of these three
+      events depend significantly on the choice of proxy record used to
+      derive a time-dependent climate forcing (Table~\ref{tab:extrema}).
 
       Simulations forced by the Greenland ice core palaeo-temperature
-records (GRIP, NGRIP) produce the highest variability in modelled ice volume
-throughout the last glacial cycle. In contrast, simulations driven by oceanic
-(ODP~1012, ODP~1020) and Antarctic (EPICA, Vostok) palaeo-temperature records
-generally result in lower ice volume variability throughout the simulation
-length, resulting in lower modelled ice volumes during MIS~4 and larger ice
-volumes during MIS~3. The NGRIP climate forcing is the only one
-that results in a larger ice volume during MIS~4 than during MIS~2.
-
-While simulations driven by the GRIP and the two Antarctic palaeo-temperature
-records attain a last ice volume maximum between 19.1 and 16.8\,ka, those
-informed by the NGRIP and the two oceanic palaeo-temperature records attain their
-maximum ice volumes thousands of years earlier. Moreover, the ODP~1012 run
-yields a rapid deglaciation of the modelled area prior to 17\,ka. The ODP~1020
-simulation predicts an early maximum in ice volume at 23.2\,ka, followed by
-slower deglaciation than modelled using the other palaeo-temperature records.
-Finally, whereas model runs forced by the Antarctic palaeo-temperature records
-result in a rapid and uninterrupted deglaciation after the LGM, the
-simulation driven by the GRIP palaeo-temperature record also results in a rapid
-deglaciation but in three steps, separated by two periods of ice sheet regrowth
+      records (GRIP, NGRIP) produce the highest variability in modelled ice
+      volume throughout the last glacial cycle. In contrast, simulations
+      driven by oceanic (ODP~1012, ODP~1020) and Antarctic (EPICA, Vostok)
+      palaeo-temperature records generally result in lower ice volume
+      variability throughout the simulation length, resulting in lower
+      modelled ice volumes during MIS~4 and larger ice volumes during MIS~3.
+      The NGRIP climate forcing is the only one that results in a larger ice
+      volume during MIS~4 than during MIS~2.
+
+      While simulations driven by the GRIP and the two Antarctic
+      palaeo-temperature records attain a last ice volume maximum between
+      19.1 and 16.8\,ka, those informed by the NGRIP and the two oceanic
+      palaeo-temperature records attain their maximum ice volumes thousands
+      of years earlier. Moreover, the ODP~1012 run yields a rapid
+      deglaciation of the modelled area prior to 17\,ka. The ODP~1020
+      simulation predicts an early maximum in ice volume at 23.2\,ka,
+      followed by slower deglaciation than modelled using the other
+      palaeo-temperature records. Finally, whereas model runs forced by the
+      Antarctic palaeo-temperature records result in a rapid and
+      uninterrupted deglaciation after the LGM, the simulation driven by the
+      GRIP palaeo-temperature record also results in a rapid deglaciation
+      but in three steps, separated by two periods of ice sheet regrowth
       (Fig.~\ref{fig:lr_ts}).
@@
       (Table~\ref{tab:extrema}), all model runs show relatively consistent
-patterns of glaciation. During MIS~4, all simulations produce an extensive ice
-sheet, covering an area of at least half of that attained during MIS~2
-(Table~\ref{tab:extrema}; Fig.~\ref{fig:lr_maps}, upper panels).
-Corresponding maximum ice volumes also differ significantly between model runs,
-and vary between 3.84 and~8.84\,m sea level equivalents (m.~s.l.e.;
+      patterns of glaciation. During MIS~4, all simulations produce an
+      extensive ice sheet, covering an area of at least half of that
+      attained during MIS~2 (Table~\ref{tab:extrema};
+      Fig.~\ref{fig:lr_maps}, upper panels). Corresponding maximum ice
+      volumes also differ significantly between model runs, and vary between
+      3.84 and~8.84\,m sea level equivalents (m.~s.l.e.;
       Table~\ref{tab:extrema}).
 
-In the MIS~3 ice volume minimum reconstructions, a central ice cap persists
-over the Skeena
-Mountains (Fig.~\ref{fig:lr_maps}, middle panels). Although this ice cap is
-present in all simulations, its dimensions depend sensitively on the choice of
-the applied palaeo-temperature record. Modelled ice volume minima spread over a
+      In the MIS~3 ice volume minimum reconstructions, a central ice cap
+      persists over the Skeena Mountains (Fig.~\ref{fig:lr_maps}, middle
+      panels). Although this ice cap is present in all simulations, its
+      dimensions depend sensitively on the choice of the applied
+      palaeo-temperature record. Modelled ice volume minima spread over a
       wide range between 1.69 and 2.88\,m~s.l.e. (Table~\ref{tab:extrema}).
 
-Modelled ice sheet geometries during the LGM (MIS~2; Fig.~\ref{fig:lr_maps},
-lower panels) invariably include a ca. 1500\,km-long central divide above
-3000\,m~a.s.l. located along the spine of the Rocky Mountains. Although the
-similarity of modelled ice extents is a direct result of the choice of scaling
-factors applied to different palaeo-temperature proxy records
-(Table~\ref{tab:records}), it is interesting to note that modelled maximum ice
-volumes also fall within a tight range of 8.40 to 8.91\,m~s.l.e.
+      Modelled ice sheet geometries during the LGM (MIS~2;
+      Fig.~\ref{fig:lr_maps}, lower panels) invariably include a ca.
+      1500\,km-long central divide above 3000\,m~a.s.l. located along the
+      spine of the Rocky Mountains. Although the similarity of modelled ice
+      extents is a direct result of the choice of scaling factors applied to
+      different palaeo-temperature proxy records (Table~\ref{tab:records}),
+      it is interesting to note that modelled maximum ice volumes also fall
+      within a tight range of 8.40 to 8.91\,m~s.l.e.
       (Table~\ref{tab:extrema}).
@@
 
-Large variations in the model responses to evolving climate forcing reveal its
-sensitivity to the choice of palaeo-temperature proxy record. To distinguish
-between different records, geological evidence of former glaciations
-provide a basis for validation of our runs, while the results from numerical
-modelling can perhaps help to analyse some of the complexity of this evidence.
-In this section, we compare model outputs to the geologic record, in terms of
-timing and configuration of the maximum stages, location and lifetime of
-major nucleation centres, and patterns of ice retreat during the last
+      Large variations in the model responses to evolving climate forcing
+      reveal its sensitivity to the choice of palaeo-temperature proxy
+      record. To distinguish between different records, geological evidence
+      of former glaciations provide a basis for validation of our runs,
+      while the results from numerical modelling can perhaps help to analyse
+      some of the complexity of this evidence. In this section, we compare
+      model outputs to the geologic record, in terms of timing and
+      configuration of the maximum stages, location and lifetime of major
+      nucleation centres, and patterns of ice retreat during the last
       deglaciation.
@@
 
-Independently of the palaeo-temperature records
-used to force the ice sheet model, our simulations consistently produce two
-glacial maxima during the last glacial cycle. The first maximum configuration
-is obtained during MIS~4 (61.9--56.5\,ka) and the second during MIS~2
+      Independently of the palaeo-temperature records used to force the ice
+      sheet model, our simulations consistently produce two glacial maxima
+      during the last glacial cycle. The first maximum configuration is
+      obtained during MIS~4 (61.9--56.5\,ka) and the second during MIS~2
       (23.2--16.8\,ka; Figs.~\ref{fig:lr_ts}, \ref{fig:lr_maps};
-Table~\ref{tab:extrema}). These events broadly correspond in timing to the
-Gladstone
-(MIS~4) and McConnell (MIS~2) glaciations documented by geological evidence for
-the northern sector of the Cordilleran ice sheet
-    \citep{Duk-Rodkin.etal.1996, Ward.etal.2007,
-           Stroeven.etal.2010, Stroeven.etal.2014},
-and to the Fraser Glaciation (MIS~2) documented for its southern sector
-    \citep{Porter.Swanson.1998, Margold.etal.2014}.
-There is stratigraphical evidence for an MIS~4 glaciation in British Columbia
-\citep{Clague.Ward.2011} and in the Puget Lowland \citep{Troost.2014}, but
-their extent and timing are still highly conjectural
-    \citep[perhaps MIS~4 or early MIS~3; e.g.,][]{Cosma.etal.2008}.
-
-The exact timing of modelled MIS~2 maximum ice volume depends strongly on the
-choice of applied palaeo-temperature record, which allows for a more in-depth
-comparison with geological evidence for the timing of maximum Cordilleran ice
-sheet extent. In the Puget Lowland (Fig.~\ref{fig:locmap}), the LGM advance of
-the southern Cordilleran ice sheet margin has been constrained by radiocarbon
-dating on wood between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,ka}
-\citep{Porter.Swanson.1998}. These dates are consistent with radiocarbon dates
-from the offshore sedimentary record, which reveals
+      Table~\ref{tab:extrema}). These events broadly correspond in timing to
+      the Gladstone (MIS~4) and McConnell (MIS~2) glaciations documented by
+      geological evidence for the northern sector of the Cordilleran ice
+      sheet \citep{Duk-Rodkin.etal.1996, Ward.etal.2007, Stroeven.etal.2010,
+      Stroeven.etal.2014}, and to the Fraser Glaciation (MIS~2) documented
+      for its southern sector \citep{Porter.Swanson.1998,
+      Margold.etal.2014}. There is stratigraphical evidence for an MIS~4
+      glaciation in British Columbia \citep{Clague.Ward.2011} and in the
+      Puget Lowland \citep{Troost.2014}, but their extent and timing are
+      still highly conjectural \citep[perhaps MIS~4 or early MIS~3;
+      e.g.,][]{Cosma.etal.2008}.
+
+      The exact timing of modelled MIS~2 maximum ice volume depends strongly
+      on the choice of applied palaeo-temperature record, which allows for a
+      more in-depth comparison with geological evidence for the timing of
+      maximum Cordilleran ice sheet extent. In the Puget Lowland
+      (Fig.~\ref{fig:locmap}), the LGM advance of the southern Cordilleran
+      ice sheet margin has been constrained by radiocarbon dating on wood
+      between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,ka}
+      \citep{Porter.Swanson.1998}. These dates are consistent with
+      radiocarbon dates from the offshore sedimentary record, which reveals
       an increase of glaciomarine sedimentation between 19.5 and
-16.2\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008, Taylor.etal.2014}.
-Radiocarbon
-dating of the northern Cordilleran ice sheet margin is much less constrained
-but straddles presented constraints from the southern margin. However,
-cosmogenic exposure dating places the timing of maximum CIS extent during the
-McConnell glaciation close to 17\,\unit{\chem{^{10}Be}\,ka}
+      16.2\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008,
+      Taylor.etal.2014}. Radiocarbon dating of the northern Cordilleran ice
+      sheet margin is much less constrained but straddles presented
+      constraints from the southern margin. However, cosmogenic exposure
+      dating places the timing of maximum CIS extent during the McConnell
+      glaciation close to 17\,\unit{\chem{^{10}Be}\,ka}
       \citep{Stroeven.etal.2010, Stroeven.etal.2014}. A sharp transition in
-the sediment record of the Gulf of Alaska indicates a retreat of regional
-outlet glaciers onto land at 14.8\,\unit{\chem{^{14}C}\,cal\,ka}
-\citep{Davies.etal.2011}.
-
-Among the simulations presented here, only those forced with the GRIP, EPICA
-and Vostok palaeo-temperature records yield Cordilleran ice sheet maximum
-extents that may be compatible with these field constraints
-(Fig.~\ref{fig:lr_ts}, lower panel; Table~\ref{tab:extrema}).
-Simulations driven by the NGRIP, ODP~1012 and ODP~1020
-palaeo-temperature records, on the contrary, yield MIS~2 maximum Cordilleran
-ice sheet volumes that pre-date field-based constraints by several thousands of
-years (about 6, 6 and 4\,ka respectively).
-Concerning the simulations driven by oceanic records, this early
-deglaciation is caused by an early warming present in the alkenone
-palaeo-temperature reconstructions (Fig.~\ref{fig:lr_ts}, upper panel;
-\citealp[Fig.~3]{Herbert.etal.2001}). However, this
-early warming is a local effect, corresponding to a weakening of the California
-current \citep{Herbert.etal.2001}. The California current, driving cold
-waters southwards along the south-western coast of North America, has been
-shown to have weakened during each peak of global glaciation (in SPECMAP)
-during the past 550\,ka, including the LGM, resulting in paradoxically warmer
-sea-surface temperatures at the locations of the ODP~1012 and ODP~1020 sites
+      the sediment record of the Gulf of Alaska indicates a retreat of
+      regional outlet glaciers onto land at
+      14.8\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Davies.etal.2011}.
+
+      Among the simulations presented here, only those forced with the GRIP,
+      EPICA and Vostok palaeo-temperature records yield Cordilleran ice
+      sheet maximum extents that may be compatible with these field
+      constraints (Fig.~\ref{fig:lr_ts}, lower panel;
+      Table~\ref{tab:extrema}). Simulations driven by the NGRIP, ODP~1012
+      and ODP~1020 palaeo-temperature records, on the contrary, yield MIS~2
+      maximum Cordilleran ice sheet volumes that pre-date field-based
+      constraints by several thousands of years (about 6, 6 and 4\,ka
+      respectively). Concerning the simulations driven by oceanic records,
+      this early deglaciation is caused by an early warming present in the
+      alkenone palaeo-temperature reconstructions (Fig.~\ref{fig:lr_ts},
+      upper panel; \citealp[Fig.~3]{Herbert.etal.2001}). However, this early
+      warming is a local effect, corresponding to a weakening of the
+      California current \citep{Herbert.etal.2001}. The California current,
+      driving cold waters southwards along the south-western coast of North
+      America, has been shown to have weakened during each peak of global
+      glaciation (in SPECMAP) during the past 550\,ka, including the LGM,
+      resulting in paradoxically warmer sea-surface temperatures at the
+      locations of the ODP~1012 and ODP~1020 sites
       \citep{Herbert.etal.2001}.
 
-Because most of the marine margin of the Cordilleran ice sheet terminated in a
-sector of the Pacific Ocean unaffected by variations in the California current,
-it probably remained insensitive to this local phenomenum. However, the above
-paradox
-illustrates the complexity of ice-sheet feedbacks on regional climate, and
-demonstrates that, although located in the neighbourhood of the modelling
-domain, the ODP~1012 and ODP~1020 palaeo-temperature records cannot be
-used as a realistic forcing to model the Cordilleran ice sheet
-through the last glacial cycle.
-Similarly, the simulation using the NGRIP palaeo-temperature record depicts an
-early onset of deglaciation (Fig.~\ref{fig:lr_ts}) following its last
-glacial volume maximum (22.9\,ka, Table~\ref{tab:extrema}) attained about 6\,ka
-earlier than dated evidence of the LGM advance.
-There is a fair agreement between the EPICA and Vostok
-palaeo-temperature records, resulting in only small differences between the
-simulations driven by those records. These differences are not subject to
-further analysis here; instead we focus on simulations forced by
-palaeo-temperature records from the GRIP and EPICA ice cores that appear to
-produce the most realistic reconstructions of regional glaciation history, yet
-bearing significant disparities in model output. To
-allow for a more detailed comparison against the geological record, these two
-simulations were re-run using a higher-resolution grid (Sect.~\ref{sec:model};
-Fig.~\ref{fig:lr_ts}, lower panel, dotted lines).
+      Because most of the marine margin of the Cordilleran ice sheet
+      terminated in a sector of the Pacific Ocean unaffected by variations
+      in the California current, it probably remained insensitive to this
+      local phenomenum. However, the above paradox illustrates the
+      complexity of ice-sheet feedbacks on regional climate, and
+      demonstrates that, although located in the neighbourhood of the
+      modelling domain, the ODP~1012 and ODP~1020 palaeo-temperature records
+      cannot be used as a realistic forcing to model the Cordilleran ice
+      sheet through the last glacial cycle. Similarly, the simulation using
+      the NGRIP palaeo-temperature record depicts an early onset of
+      deglaciation (Fig.~\ref{fig:lr_ts}) following its last glacial volume
+      maximum (22.9\,ka, Table~\ref{tab:extrema}) attained about 6\,ka
+      earlier than dated evidence of the LGM advance. There is a fair
+      agreement between the EPICA and Vostok palaeo-temperature records,
+      resulting in only small differences between the simulations driven by
+      those records. These differences are not subject to further analysis
+      here; instead we focus on simulations forced by palaeo-temperature
+      records from the GRIP and EPICA ice cores that appear to produce the
+      most realistic reconstructions of regional glaciation history, yet
+      bearing significant disparities in model output. To allow for a more
+      detailed comparison against the geological record, these two
+      simulations were re-run using a higher-resolution grid
+      (Sect.~\ref{sec:model}; Fig.~\ref{fig:lr_ts}, lower panel, dotted
+      lines).
 
@@
 
-During maximum glaciation, both simulations position the main meridional ice
-divide over
-the western flank of the Rocky Mountains (Fig.~\ref{fig:lr_maps}, lower
-panels; Fig.~\ref{fig:mis2}). This result appears to contrast with
-palaeoglaciological reconstructions for central and southern British Columbia
-with ice divides in a more westerly position,
-over the western margin of the Interior Plateau \citep{Ryder.etal.1991,
-Stumpf.etal.2000, Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}.
-These indicate that a latitudinal saddle connected ice dispersal centres in
-the Columbia Mountains with the main ice divide \citep{Ryder.etal.1991,
-Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}. A latitudinal saddle
-does indeed feature in our modelling results, however, in an inverse
-configuration between the main ice divide over the Columbia Mountains and a
-secondary divide over the southern Coast Mountains (Fig.~\ref{fig:mis2}).
-
-Such deviation from the geological inferences could reflect the fact that our
-model does not include
-feedback mechanisms between ice sheet topography and the regional climate.
-Firstly, during the build-up phase preceding the LGM, rapid accumulation over
-the Coast Mountains enhanced the topographic barrier formed by these mountain
-ranges, which likely resulted in a decrease of precipitation and, therefore,
-a decrease of
-accumulation in the interior. Secondly, latent warming of the moisture-depleted
-air parcels flowing over this enhanced topography could have resulted in an
-inflow of potentially warmer air over the eastern flank of the ice sheet,
-increasing melt
-along the advancing margin \citep[cf.][]{Langen.etal.2012}. Because these two
-processes, both with a tendency to limit ice-sheet growth, are absent from
-our model, the
-eastern margin of the ice sheet and the position of the main meridional ice
-divide are certainly biased towards the east in our simulations
-\citep{Seguinot.etal.2014}.
+      During maximum glaciation, both simulations position the main
+      meridional ice divide over the western flank of the Rocky Mountains
+      (Fig.~\ref{fig:lr_maps}, lower panels; Fig.~\ref{fig:mis2}). This
+      result appears to contrast with palaeoglaciological reconstructions
+      for central and southern British Columbia with ice divides in a more
+      westerly position, over the western margin of the Interior Plateau
+      \citep{Ryder.etal.1991, Stumpf.etal.2000, Kleman.etal.2010,
+      Clague.Ward.2011, Margold.etal.2013a}. These indicate that a
+      latitudinal saddle connected ice dispersal centres in the Columbia
+      Mountains with the main ice divide \citep{Ryder.etal.1991,
+      Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}. A latitudinal
+      saddle does indeed feature in our modelling results, however, in an
+      inverse configuration between the main ice divide over the Columbia
+      Mountains and a secondary divide over the southern Coast Mountains
+      (Fig.~\ref{fig:mis2}).
+
+      Such deviation from the geological inferences could reflect the fact
+      that our model does not include feedback mechanisms between ice sheet
+      topography and the regional climate. Firstly, during the build-up
+      phase preceding the LGM, rapid accumulation over the Coast Mountains
+      enhanced the topographic barrier formed by these mountain ranges,
+      which likely resulted in a decrease of precipitation and, therefore, a
+      decrease of accumulation in the interior. Secondly, latent warming of
+      the moisture-depleted air parcels flowing over this enhanced
+      topography could have resulted in an inflow of potentially warmer air
+      over the eastern flank of the ice sheet, increasing melt along the
+      advancing margin \citep[cf.][]{Langen.etal.2012}. Because these two
+      processes, both with a tendency to limit ice-sheet growth, are absent
+      from our model, the eastern margin of the ice sheet and the position
+      of the main meridional ice divide are certainly biased towards the
+      east in our simulations \citep{Seguinot.etal.2014}.
 
       However, field-based palaeoglaciological reconstructions have
-struggled to reconcile the more westerly-centred ice divide in south-central
-British Columbia with evidence in the Rocky Mountains and beyond, that the
-Cordilleran ice sheet invaded the western Interior Plains, where it merged with
-the southwestern margin of the Laurentide ice sheet and was deflected to the
-south \citep{Jackson.etal.1997, Bednarski.Smith.2007, Kleman.etal.2010,
-Margold.etal.2013, Margold.etal.2013a}. Ice geometries from our model runs do
-not have this problem, because the position and elevation of the ice divide
-ensure significant ice drainage across the Rocky Mountains at the LGM
-(Fig.~\ref{fig:mis2}).
-
-During MIS~2, the modelled total ice volume peaks at 8.01\,m~s.l.e. (19.1\,ka)
-in the GRIP simulation and at 8.77\,m~s.l.e. (17.2\,ka) in the EPICA
-simulation.
+      struggled to reconcile the more westerly-centred ice divide in
+      south-central British Columbia with evidence in the Rocky Mountains
+      and beyond, that the Cordilleran ice sheet invaded the western
+      Interior Plains, where it merged with the southwestern margin of the
+      Laurentide ice sheet and was deflected to the south
+      \citep{Jackson.etal.1997, Bednarski.Smith.2007, Kleman.etal.2010,
+      Margold.etal.2013, Margold.etal.2013a}. Ice geometries from our model
+      runs do not have this problem, because the position and elevation of
+      the ice divide ensure significant ice drainage across the Rocky
+      Mountains at the LGM (Fig.~\ref{fig:mis2}).
+
+      During MIS~2, the modelled total ice volume peaks at 8.01\,m~s.l.e.
+      (19.1\,ka) in the GRIP simulation and at 8.77\,m~s.l.e. (17.2\,ka) in
+      the EPICA simulation.
 
@@
 
-The modelled ice sheet configurations corresponding to ice volume maxima during
-MIS~4 are more sensitive to the choice of atmospheric forcing than those
-corresponding to ice volume maxima during MIS~2 (Fig.~\ref{fig:lr_maps},
-upper panels;
-Fig.~\ref{fig:mis4}). The GRIP simulation (Fig.~\ref{fig:mis4}, left panel)
-results in a modelled maximum ice sheet extent that closely resembles that
-obtained during MIS~2, with the only major difference of being slightly less
-extensive across
-northern and eastern sectors. In contrast, the EPICA simulation produces a lower
-ice volume maximum (Fig.~\ref{fig:lr_ts}), which translates in the
-modelled ice sheet geometry into a significantly reduced southern sector, more
-restricted ice cover in northern and eastern sectors, and generally lower
-ice surface elevations in the interior (Fig.~\ref{fig:mis4}, right panel).
-Thus, only the GRIP simulation can explain the presence of MIS~4 glacial
-deposits in the Puget Lowland \citep{Troost.2014} and that of ice-rafted debris
-in the marine sediment record offshore Vancouver Island at
+      The modelled ice sheet configurations corresponding to ice volume
+      maxima during MIS~4 are more sensitive to the choice of atmospheric
+      forcing than those corresponding to ice volume maxima during MIS~2
+      (Fig.~\ref{fig:lr_maps}, upper panels; Fig.~\ref{fig:mis4}). The GRIP
+      simulation (Fig.~\ref{fig:mis4}, left panel) results in a modelled
+      maximum ice sheet extent that closely resembles that obtained during
+      MIS~2, with the only major difference of being slightly less extensive
+      across northern and eastern sectors. In contrast, the EPICA simulation
+      produces a lower ice volume maximum (Fig.~\ref{fig:lr_ts}), which
+      translates in the modelled ice sheet geometry into a significantly
+      reduced southern sector, more restricted ice cover in northern and
+      eastern sectors, and generally lower ice surface elevations in the
+      interior (Fig.~\ref{fig:mis4}, right panel). Thus, only the GRIP
+      simulation can explain the presence of MIS~4 glacial deposits in the
+      Puget Lowland \citep{Troost.2014} and that of ice-rafted debris in the
+      marine sediment record offshore Vancouver Island at
       ca.~47\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008}.
 
-During MIS~4, the modelled total ice volume peaks at 7.19\,m~s.l.e. (57.3\,ka)
-in the GRIP simulation and at 5.04\,m~s.l.e. (61.9\,ka) in the EPICA
-simulation, corresponding to respectively 90\,\% and 57\,\% of modelled MIS~2
-ice volumes.
+      During MIS~4, the modelled total ice volume peaks at 7.19\,m~s.l.e.
+      (57.3\,ka) in the GRIP simulation and at 5.04\,m~s.l.e. (61.9\,ka) in
+      the EPICA simulation, corresponding to respectively 90\,\% and 57\,\%
+      of modelled MIS~2 ice volumes.
 
@@
 
-Palaeo-glaciological reconstructions are generally more robust
-for maximum ice sheet extents and late ice sheet configurations than for
-intermediate or minimum ice sheet extents and older ice sheet configurations
-\citep{Kleman.etal.2010}. However,
-these maximum stages are, by nature, extreme configurations, which do not
-necessarily represent the dominant patterns of glaciation throughout the period
-of ice cover \citep{Porter.1989, Kleman.Stroeven.1997, Kleman.etal.2008,
+      Palaeo-glaciological reconstructions are generally more robust for
+      maximum ice sheet extents and late ice sheet configurations than for
+      intermediate or minimum ice sheet extents and older ice sheet
+      configurations \citep{Kleman.etal.2010}. However, these maximum stages
+      are, by nature, extreme configurations, which do not necessarily
+      represent the dominant patterns of glaciation throughout the period of
+      ice cover \citep{Porter.1989, Kleman.Stroeven.1997, Kleman.etal.2008,
       Kleman.etal.2010}.
 
-For the Cordilleran ice sheet, geological evidence from radiocarbon dating
-    \citep{Clague.etal.1980, Clague.1985, Clague.1986, Porter.Swanson.1998,
-           Menounos.etal.2008},
-cosmogenic exposure dating
+      For the Cordilleran ice sheet, geological evidence from radiocarbon
+      dating \citep{Clague.etal.1980, Clague.1985, Clague.1986,
+      Porter.Swanson.1998, Menounos.etal.2008}, cosmogenic exposure dating
       \citep{Stroeven.etal.2010, Stroeven.etal.2014, Margold.etal.2014},
       bedrock deformation in response to former ice loads
-    \citep{Clague.James.2002, Clague.etal.2005},
-and offshore sedimentary records
-    \citep{Cosma.etal.2008, Davies.etal.2011}
-indicate that the LGM maximum extent was short-lived. To compare this finding
-to our simulations, we use numerical modelling output
-to compute durations of ice cover throughout the last glacial
-cycle (Fig.~\ref{fig:hr_geom_duration}).
-
-The resulting maps show that, during most of the glacial cycle, modelled ice
-cover is restricted to disjoint ice caps centred on major mountain ranges of
-the North American Cordillera (Fig.~\ref{fig:hr_geom_duration}, blue areas). A
-2500\,km-long continuous expanse of ice, extending from the Alaska Range in the
-northeast to the Rocky Mountains in the southwest, is only in operation
-for at most 32\,ka, which is about a third of the timespan of the last glacial cycle
-(Fig.~\ref{fig:hr_geom_duration}, hatched areas). However, except for its margins in
-the Pacific Ocean and in the northern foothills of the Alaska Range,
-the maximum extent of the ice sheet is attained for a much shorter period
-of time of only few thousand years (Figs.~\ref{fig:hr_geom_duration}, red areas). This
-result illustrates that the maximum extents of the modelled ice sheet during
-MIS~4 and MIS~2 were both short-lived and therefore out of balance with
-contemporary climate.
-
-A notable exception to the transient character of the maximum extent of
-Cordilleran ice sheet
-is the northern slope of the Alaska Range, where modelled glaciers are confined
-to its foothills during the entire simulation period (Fig.~\ref{fig:hr_geom_duration},
-AR). This apparent insensitivity of modelled glacial extent to temperature
-fluctuations results from a combination of low precipitation, high summer
-temperature and large temperature standard deviation (PDD SD) in the plains of the
-Alaska Interior (Fig.~\ref{fig:atm}) which confines glaciation to the
-foothills of the mountains. This result could potentially explain
-the local distribution of glacial deposits, which indicates that glaciers
-flowing on the northern slope of the Alaska Range
-have remained small throughout the Pleistocene \citep{Kaufman.Manley.2004}.
+      \citep{Clague.James.2002, Clague.etal.2005}, and offshore sedimentary
+      records \citep{Cosma.etal.2008, Davies.etal.2011} indicate that the
+      LGM maximum extent was short-lived. To compare this finding to our
+      simulations, we use numerical modelling output to compute durations of
+      ice cover throughout the last glacial cycle
+      (Fig.~\ref{fig:hr_geom_duration}).
+
+      The resulting maps show that, during most of the glacial cycle,
+      modelled ice cover is restricted to disjoint ice caps centred on major
+      mountain ranges of the North American Cordillera
+      (Fig.~\ref{fig:hr_geom_duration}, blue areas). A 2500\,km-long
+      continuous expanse of ice, extending from the Alaska Range in the
+      northeast to the Rocky Mountains in the southwest, is only in
+      operation for at most 32\,ka, which is about a third of the timespan
+      of the last glacial cycle (Fig.~\ref{fig:hr_geom_duration}, hatched
+      areas). However, except for its margins in the Pacific Ocean and in
+      the northern foothills of the Alaska Range, the maximum extent of the
+      ice sheet is attained for a much shorter period of time of only few
+      thousand years (Figs.~\ref{fig:hr_geom_duration}, red areas). This
+      result illustrates that the maximum extents of the modelled ice sheet
+      during MIS~4 and MIS~2 were both short-lived and therefore out of
+      balance with contemporary climate.
+
+      A notable exception to the transient character of the maximum extent
+      of Cordilleran ice sheet is the northern slope of the Alaska Range,
+      where modelled glaciers are confined to its foothills during the
+      entire simulation period (Fig.~\ref{fig:hr_geom_duration}, AR). This
+      apparent insensitivity of modelled glacial extent to temperature
+      fluctuations results from a combination of low precipitation, high
+      summer temperature and large temperature standard deviation (PDD SD)
+      in the plains of the Alaska Interior (Fig.~\ref{fig:atm}) which
+      confines glaciation to the foothills of the mountains. This result
+      could potentially explain the local distribution of glacial deposits,
+      which indicates that glaciers flowing on the northern slope of the
+      Alaska Range have remained small throughout the Pleistocene
+      \citep{Kaufman.Manley.2004}.
 
@@
       It is generally believed that the Cordilleran ice sheet formed by the
-coalescence of several mountain-centred ice caps \citep{Davis.Mathews.1944}.
-In our simulations, major ice-dispersal centres, visible on the modelled ice
-cover duration maps (Fig.~\ref{fig:hr_geom_duration}), are located over the Coast
-Mountains (CM), the Columbia and Rocky mountains (CRM), the Skeena Mountains
-(SM), and the Selwyn and Mackenzie mountains (SMKM). The Wrangell and
-Saint Elias mountains, heavily glacierized at present, host an ice cap for the
-entire length of both simulations, but that ice cap does not appear to be a
-major feed to the Cordilleran ice sheet (Fig.~\ref{fig:hr_geom_duration}, WSEM).
-Although the Coast, Skeena and Columbia and Rocky mountains (CM, SM, CRM)
-are covered by mountain
-glaciers for most of the last glacial cycle, providing durable nucleation
-centres for an ice sheet, this is not the case for the Selwyn and
-Mackenzie mountains (SMKM), where ice cover on the highest peaks is limited to a
-small fraction of the last glacial cycle. In other words, the Selwyn and Mackenzie
-mountains only appear as a secondary ice-dispersal centre
-during the coldest periods of the last glacial cycle. The Northern Rocky
-Mountains (Fig.~\ref{fig:hr_geom_duration}, NRM) do not act as a nucleation centre,
-but rather as a pinning point for the Cordilleran ice sheet margin coming from
-the west.
-
-Perhaps the most striking feature displayed by the distributions of modelled
-ice cover is the persistence of the Skeena Mountains ice cap throughout the
-entire last glacial period (ca.~100--10\,ka) and its predominance over the
-other ice-dispersal centres
-(Figs.~\ref{fig:lr_maps} and~\ref{fig:hr_geom_duration}, SM). Regardless of the applied
-forcing, this ice cap appears to survive MIS~3 (Fig.~\ref{fig:lr_maps},
-middle panels), and serves as a nucleation centre at the onset of the glacial
-readvance towards the LGM (MIS~2). This situation appear similar to the
-neighbouring Laurentide ice sheet, for which the importance of residual ice for the
+      coalescence of several mountain-centred ice caps
+      \citep{Davis.Mathews.1944}. In our simulations, major ice-dispersal
+      centres, visible on the modelled ice cover duration maps
+      (Fig.~\ref{fig:hr_geom_duration}), are located over the Coast
+      Mountains (CM), the Columbia and Rocky mountains (CRM), the Skeena
+      Mountains (SM), and the Selwyn and Mackenzie mountains (SMKM). The
+      Wrangell and Saint Elias mountains, heavily glacierized at present,
+      host an ice cap for the entire length of both simulations, but that
+      ice cap does not appear to be a major feed to the Cordilleran ice
+      sheet (Fig.~\ref{fig:hr_geom_duration}, WSEM). Although the Coast,
+      Skeena and Columbia and Rocky mountains (CM, SM, CRM) are covered by
+      mountain glaciers for most of the last glacial cycle, providing
+      durable nucleation centres for an ice sheet, this is not the case for
+      the Selwyn and Mackenzie mountains (SMKM), where ice cover on the
+      highest peaks is limited to a small fraction of the last glacial
+      cycle. In other words, the Selwyn and Mackenzie mountains only appear
+      as a secondary ice-dispersal centre during the coldest periods of the
+      last glacial cycle. The Northern Rocky Mountains
+      (Fig.~\ref{fig:hr_geom_duration}, NRM) do not act as a nucleation
+      centre, but rather as a pinning point for the Cordilleran ice sheet
+      margin coming from the west.
+
+      Perhaps the most striking feature displayed by the distributions of
+      modelled ice cover is the persistence of the Skeena Mountains ice cap
+      throughout the entire last glacial period (ca.~100--10\,ka) and its
+      predominance over the other ice-dispersal centres
+      (Figs.~\ref{fig:lr_maps} and~\ref{fig:hr_geom_duration}, SM).
+      Regardless of the applied forcing, this ice cap appears to survive
+      MIS~3 (Fig.~\ref{fig:lr_maps}, middle panels), and serves as a
+      nucleation centre at the onset of the glacial readvance towards the
+      LGM (MIS~2). This situation appear similar to the neighbouring
+      Laurentide ice sheet, for which the importance of residual ice for the
       glacial history leading up to the LGM has been illustrated by the
-MIS~3 residual ice bodies in northern and eastern Canada as nucleation centres
-for a much more extensive MIS~2 configuration \citep{Kleman.etal.2010}.
-
-The presence of a Skeena Mountains ice cap during most of the last glacial
-cycle can be explained by meteorological conditions more favourable for ice
-growth there than elsewhere. In fact, reanalysed atmospheric fields used to
-force the surface mass balance model show that high winter precipitations are
-mainly confined to the western slope of the Coast Mountains, except in the
-centre of the modelling domain where they also occur further inland than along
-other east-west transects (Fig.~\ref{fig:atm}).
-In fact, along most of the north-western coast of
-North America, coastal mountain ranges form a pronounced topographic barrier
-for westerly winds, capturing atmospheric moisture in the form of orographic
-precipitation, and resulting in arid interior lowlands. However, near the
-centre of our modelling domain, this barrier is less pronounced than elsewhere,
-allowing westerly winds to carry moisture
-further inland, until it is captured by the extensive Skeena Mountains in
-north-central British Columbia, thus resulting in a more widespread
-distribution of winter precipitation (Fig.~\ref{fig:atm}).
-
-The modelled total ice volume corresponding to these persistent ice-dispersal
-centres attains a minimum of 1.52\,m~s.l.e. (42.9\,ka) in the GRIP simulation
-and of 2.68\,m~s.l.e. (51.8\,ka) in the EPICA simulation, corresponding to
-respectively 19\,\% and 31\,\% of the MIS~2 ice volumes.
+      MIS~3 residual ice bodies in northern and eastern Canada as nucleation
+      centres for a much more extensive MIS~2 configuration
+      \citep{Kleman.etal.2010}.
+
+      The presence of a Skeena Mountains ice cap during most of the last
+      glacial cycle can be explained by meteorological conditions more
+      favourable for ice growth there than elsewhere. In fact, reanalysed
+      atmospheric fields used to force the surface mass balance model show
+      that high winter precipitations are mainly confined to the western
+      slope of the Coast Mountains, except in the centre of the modelling
+      domain where they also occur further inland than along other east-west
+      transects (Fig.~\ref{fig:atm}). In fact, along most of the
+      north-western coast of North America, coastal mountain ranges form a
+      pronounced topographic barrier for westerly winds, capturing
+      atmospheric moisture in the form of orographic precipitation, and
+      resulting in arid interior lowlands. However, near the centre of our
+      modelling domain, this barrier is less pronounced than elsewhere,
+      allowing westerly winds to carry moisture further inland, until it is
+      captured by the extensive Skeena Mountains in north-central British
+      Columbia, thus resulting in a more widespread distribution of winter
+      precipitation (Fig.~\ref{fig:atm}).
+
+      The modelled total ice volume corresponding to these persistent
+      ice-dispersal centres attains a minimum of 1.52\,m~s.l.e. (42.9\,ka)
+      in the GRIP simulation and of 2.68\,m~s.l.e. (51.8\,ka) in the EPICA
+      simulation, corresponding to respectively 19\,\% and 31\,\% of the
+      MIS~2 ice volumes.
 
@@
 
-A correlation is observed between the modelled duration of warm based ice cover
-(Fig.~\ref{fig:hr_geom_warmbase}) and the degree of glacial modification of the
-landscape (mainly in terms of the development of deep glacial valleys and
-troughs). We find evidence for this on the slopes of the Coast Mountains, the
-Columbia and Rocky mountains, the Wrangell and Saint Elias mountains, and
-radiating off the Skeena Mountains
-(Figs.~\ref{fig:hr_geom_duration} and \ref{fig:hr_geom_warmbase};
-\citealp[Fig.~2]{Kleman.etal.2010}).
-The Skeena Mountains, for example, indeed bear a strong glacial imprint that
-indicates ice drainage in a system of distinct glacial troughs emanating in a
-radial pattern from the centre of the mountain range
-\citep[Fig.~2]{Kleman.etal.2010}. We suggest that
-persistent ice cover (Fig.~\ref{fig:hr_geom_duration}) associated with basal ice
-temperatures at the pressure melting point (Figs.~\ref{fig:hr_geom_warmbase}
-and~\ref{fig:hr_geom_warmfrac}) explains the large-scale glacial erosional imprint on
-the landscape. A well-developed network of glacial valleys running to the
-north-west on the west slope of the Selwyn
-and Mackenzie Mountains (\citealp[Fig.~2]{Kleman.etal.2010}; \citealp[Fig.~8]
-{Stroeven.etal.2010}) is modelled to have hosted predominantly warm-based ice
-(Fig.~\ref{fig:hr_geom_warmfrac}). However, because it is only glaciated for a
-short fraction of the last glacial cycle in our simulations
-(Fig.~\ref{fig:hr_geom_duration}), this perhaps indicates that
-the observed landscape pattern originates from multiple glacial cycles and
-witnesses an increased relative importance of the Selwyn and Mackenzie
-mountains ice dispersal centre (Fig.~\ref{fig:hr_geom_duration}, SMKM), prior
-to the Late Pleistocene \citep[cf.][]{Ward.etal.2008, Demuro.etal.2012}.
-
-The modelled distribution of warm-based ice cover (Figs.~\ref{fig:hr_geom_warmbase}
-and~\ref{fig:hr_geom_warmfrac}) is inevitably affected by our assumption of a constant,
+      A correlation is observed between the modelled duration of warm based
+      ice cover (Fig.~\ref{fig:hr_geom_warmbase}) and the degree of glacial
+      modification of the landscape (mainly in terms of the development of
+      deep glacial valleys and troughs). We find evidence for this on the
+      slopes of the Coast Mountains, the Columbia and Rocky mountains, the
+      Wrangell and Saint Elias mountains, and radiating off the Skeena
+      Mountains (Figs.~\ref{fig:hr_geom_duration} and
+      \ref{fig:hr_geom_warmbase}; \citealp[Fig.~2]{Kleman.etal.2010}). The
+      Skeena Mountains, for example, indeed bear a strong glacial imprint
+      that indicates ice drainage in a system of distinct glacial troughs
+      emanating in a radial pattern from the centre of the mountain range
+      \citep[Fig.~2]{Kleman.etal.2010}. We suggest that persistent ice cover
+      (Fig.~\ref{fig:hr_geom_duration}) associated with basal ice
+      temperatures at the pressure melting point
+      (Figs.~\ref{fig:hr_geom_warmbase} and~\ref{fig:hr_geom_warmfrac})
+      explains the large-scale glacial erosional imprint on the landscape. A
+      well-developed network of glacial valleys running to the north-west on
+      the west slope of the Selwyn and Mackenzie Mountains
+      (\citealp[Fig.~2]{Kleman.etal.2010}; \citealp[Fig.~8]
+      {Stroeven.etal.2010}) is modelled to have hosted predominantly
+      warm-based ice (Fig.~\ref{fig:hr_geom_warmfrac}). However, because it
+      is only glaciated for a short fraction of the last glacial cycle in
+      our simulations (Fig.~\ref{fig:hr_geom_duration}), this perhaps
+      indicates that the observed landscape pattern originates from multiple
+      glacial cycles and witnesses an increased relative importance of the
+      Selwyn and Mackenzie mountains ice dispersal centre
+      (Fig.~\ref{fig:hr_geom_duration}, SMKM), prior to the Late Pleistocene
+      \citep[cf.][]{Ward.etal.2008, Demuro.etal.2012}.
+
+      The modelled distribution of warm-based ice cover
+      (Figs.~\ref{fig:hr_geom_warmbase} and~\ref{fig:hr_geom_warmfrac}) is
+      inevitably affected by our assumption of a constant,
       70\,\unit{mW\,m^{-2}} geothermal heat flux at 3\,km depth
-(Sect.~\ref{sec:icedyn}). However, the Skeena Mountains and the area west of
-the Mackenzie Mountains experience higher-than-average geothermal heat flux
-with measured values of ca.~80 and ca.~100\,\unit{mW\,m^{-2}}
-\citep{Blackwell.Richards.2004}. We can therefore expect even longer
-durations of warm-based ice cover for these areas if we were to include
-spatially variable geothermal forcing in our Cordilleran ice sheet simulations.
+      (Sect.~\ref{sec:icedyn}). However, the Skeena Mountains and the area
+      west of the Mackenzie Mountains experience higher-than-average
+      geothermal heat flux with measured values of ca.~80 and
+      ca.~100\,\unit{mW\,m^{-2}} \citep{Blackwell.Richards.2004}. We can
+      therefore expect even longer durations of warm-based ice cover for
+      these areas if we were to include spatially variable geothermal
+      forcing in our Cordilleran ice sheet simulations.
 
@@
 
-Similarly to other glaciated regions, most glacial traces in the North American
-Cordillera relate to the last few millennia of glaciation, because most of the
-older evidence has been overprinted by warm-based ice retreat
-during the last deglaciation \citep{Kleman.1994, Kleman.etal.2010}. From a
-numerical modelling perspective, phases of glacier retreat are more challenging
-than phases of growth, because they involve more rapid fluctuations of the ice
-margin, increased flow velocities and longitudinal stress gradients, and poorly
-understood hydrological processes. The latter are typically included in the
-models through simple parametrisations \citep[e.g.][]{Clason.etal.2012,
-Clason.etal.2014, Bueler.Pelt.2015}, if included at all. However, next after
-the mapping of maximum ice sheet extents during MIS~2 and MIS~4
-(Sect.~\ref{sec:mis2} and~\ref{sec:mis4}), geomorphologically-based
-reconstructions of patterns of ice sheet retreat during the last
-deglaciation provide the second best source of evidence for the validation of
-our simulations.
+      Similarly to other glaciated regions, most glacial traces in the North
+      American Cordillera relate to the last few millennia of glaciation,
+      because most of the older evidence has been overprinted by warm-based
+      ice retreat during the last deglaciation \citep{Kleman.1994,
+      Kleman.etal.2010}. From a numerical modelling perspective, phases of
+      glacier retreat are more challenging than phases of growth, because
+      they involve more rapid fluctuations of the ice margin, increased flow
+      velocities and longitudinal stress gradients, and poorly understood
+      hydrological processes. The latter are typically included in the
+      models through simple parametrisations
+      \citep[e.g.][]{Clason.etal.2012, Clason.etal.2014, Bueler.Pelt.2015},
+      if included at all. However, next after the mapping of maximum ice
+      sheet extents during MIS~2 and MIS~4 (Sect.~\ref{sec:mis2}
+      and~\ref{sec:mis4}), geomorphologically-based reconstructions of
+      patterns of ice sheet retreat during the last deglaciation provide the
+      second best source of evidence for the validation of our simulations.
 
@@
 
-In the North American Cordillera, the presence of lateral meltwater channels
-at high elevation \citep{Margold.etal.2011, Margold.etal.2013a,
-Margold.etal.2014}, and abundant esker systems at low elevation
-\citep{Burke.etal.2012, Burke.etal.2012a, Perkins.etal.2013, Margold.etal.2013}
-indicate that meltwater was produced over large portions of the ice sheet
-surface during deglaciation. The southern and northern margins of the
-Cordilleran ice sheet reached their last glacial maximum extent around
-17\,ka \citep[Sect.~\ref{sec:timing};][]{Porter.Swanson.1998, Cosma.etal.2008,
-Stroeven.etal.2010, Stroeven.etal.2014}, which we take as a limiting age for
-the onset of ice retreat. The timing of final deglaciation is less well
-constrained, but recent cosmogenic dates from north-central British Columbia
-indicate that a seizable ice cap emanating from the central Coast Mountains or
-the Skeena Mountains persisted into the Younger Dryas chronozone, at least
-until 12.4\,ka \citep{Margold.etal.2014}.
-
-In our simulations, the timing of peak ice volume during the LGM
-and the pacing of deglaciation depend critically on the choice of
-climate forcing (Table~\ref{tab:extrema}; Figs.~\ref{fig:lr_ts}
-and~\ref{fig:lr_ts_deglac}). Adopting the EPICA climate forcing yields peak ice
-volume at 17.2\,ka and an uninterrupted deglaciation until about 9\,ka
-(Fig.~\ref{fig:lr_ts_deglac}, lower panel, red curves). On the contrary, the
-simulation driven by the GRIP palaeo-temperature record yields peak ice volume
-at 19.3\,ka and a deglaciation interrupted by two phases of regrowth until
-about 8\,ka. The first interruption occurs between 16.6 and 14.5\,ka, and the
-second between 13.1 and 11.6\,ka (Fig.~\ref{fig:lr_ts_deglac}, lower panel,
-blue curve).
-
-Hence, the two model runs, while similar in overall timing compared to the runs
-with other climate drivers, differ in detail. On the one hand, the EPICA run
-depicts peak glaciation about 2\,ka later than the GRIP run, in closer
-agreement with dated maximum extents, and shows a faster, uninterrupted
-deglaciation which yields sporadic ice cover more than 1\,ka earlier. On the
-other hand, the GRIP run yields a deglaciation in three steps, compatible with
-marine sediment sequences offshore Vancouver Island, where the distribution of
-ice-rafted debris indicates an ice margin retreat from the Georgia Strait in two
-phases that are contemporary with warming oceanic temperatures from 17.2 to
+      In the North American Cordillera, the presence of lateral meltwater
+      channels at high elevation \citep{Margold.etal.2011,
+      Margold.etal.2013a, Margold.etal.2014}, and abundant esker systems at
+      low elevation \citep{Burke.etal.2012, Burke.etal.2012a,
+      Perkins.etal.2013, Margold.etal.2013} indicate that meltwater was
+      produced over large portions of the ice sheet surface during
+      deglaciation. The southern and northern margins of the Cordilleran ice
+      sheet reached their last glacial maximum extent around 17\,ka
+      \citep[Sect.~\ref{sec:timing};][]{Porter.Swanson.1998,
+      Cosma.etal.2008, Stroeven.etal.2010, Stroeven.etal.2014}, which we
+      take as a limiting age for the onset of ice retreat. The timing of
+      final deglaciation is less well constrained, but recent cosmogenic
+      dates from north-central British Columbia indicate that a seizable ice
+      cap emanating from the central Coast Mountains or the Skeena Mountains
+      persisted into the Younger Dryas chronozone, at least until 12.4\,ka
+      \citep{Margold.etal.2014}.
+
+      In our simulations, the timing of peak ice volume during the LGM and
+      the pacing of deglaciation depend critically on the choice of climate
+      forcing (Table~\ref{tab:extrema}; Figs.~\ref{fig:lr_ts}
+      and~\ref{fig:lr_ts_deglac}). Adopting the EPICA climate forcing yields
+      peak ice volume at 17.2\,ka and an uninterrupted deglaciation until
+      about 9\,ka (Fig.~\ref{fig:lr_ts_deglac}, lower panel, red curves). On
+      the contrary, the simulation driven by the GRIP palaeo-temperature
+      record yields peak ice volume at 19.3\,ka and a deglaciation
+      interrupted by two phases of regrowth until about 8\,ka. The first
+      interruption occurs between 16.6 and 14.5\,ka, and the second between
+      13.1 and 11.6\,ka (Fig.~\ref{fig:lr_ts_deglac}, lower panel, blue
+      curve).
+
+      Hence, the two model runs, while similar in overall timing compared to
+      the runs with other climate drivers, differ in detail. On the one
+      hand, the EPICA run depicts peak glaciation about 2\,ka later than the
+      GRIP run, in closer agreement with dated maximum extents, and shows a
+      faster, uninterrupted deglaciation which yields sporadic ice cover
+      more than 1\,ka earlier. On the other hand, the GRIP run yields a
+      deglaciation in three steps, compatible with marine sediment sequences
+      offshore Vancouver Island, where the distribution of ice-rafted debris
+      indicates an ice margin retreat from the Georgia Strait in two phases
+      that are contemporary with warming oceanic temperatures from 17.2 to
       16.5 and from 15.5 to 14.0\,\unit{\chem{^{14}C}\,cal\,ka}
@@
 
-Modelled patterns of ice sheet retreat are relatively consistent between the
-two simulations (Figs.~\ref{fig:hr_geom_deglacageshots} and.~\ref{fig:hr_geom_deglacage}). The
-southern sector of the modelling domain, including the Puget Lowland, the Coast
-Mountains, the Columbia and
-and Rocky mountains, and the Interior Plateau of British Columbia, becomes
-completely deglaciated by 10\,ka, whereas a significant ice cover remains over
-the Skeena, the Selwyn and Mackenzie, and the Wrangell and Saint Elias mountains in the
-northern sector of the modelling domain. After 10\,ka, deglaciation continues
-to proceed across the Liard Lowland with a radial ice margin retreat towards
-the surrounding mountain ranges, consistent with the regional melt water record
-of the last deglaciation \citep{Margold.etal.2013}. Remaining ice continues to
-decay by retreating towards the Selwyn and MacKenzie, and Skeena mountains.
-The last remnants
-of the Cordilleran ice sheet finally disappear from the Skeena Mountains at
-6.5\,ka (GRIP) and 6.1\,ka (EPICA).
+      Modelled patterns of ice sheet retreat are relatively consistent
+      between the two simulations (Figs.~\ref{fig:hr_geom_deglacageshots}
+      and.~\ref{fig:hr_geom_deglacage}). The southern sector of the
+      modelling domain, including the Puget Lowland, the Coast Mountains,
+      the Columbia and and Rocky mountains, and the Interior Plateau of
+      British Columbia, becomes completely deglaciated by 10\,ka, whereas a
+      significant ice cover remains over the Skeena, the Selwyn and
+      Mackenzie, and the Wrangell and Saint Elias mountains in the northern
+      sector of the modelling domain. After 10\,ka, deglaciation continues
+      to proceed across the Liard Lowland with a radial ice margin retreat
+      towards the surrounding mountain ranges, consistent with the regional
+      melt water record of the last deglaciation \citep{Margold.etal.2013}.
+      Remaining ice continues to decay by retreating towards the Selwyn and
+      MacKenzie, and Skeena mountains. The last remnants of the Cordilleran
+      ice sheet finally disappear from the Skeena Mountains at 6.5\,ka
+      (GRIP) and 6.1\,ka (EPICA).
 
@@
 
-The possibility of late glacial readvances in the North American Cordillera has
-been debated for some time \citep{Osborn.Gerloff.1997}, and locally these have
-been reconstructed and dated. Radiocarbon-dated end moraines in the Fraser and
-Squamish valleys, off the southern tip of the Coast Mountains, indicate
-consecutive glacier maxima, or standstills while in overall retreat, one of
-which corresponds to the Younger Dryas chronozone \citep{Clague.etal.1997,
-Friele.Clague.2002, Friele.Clague.2002a, Kovanen.2002,
-Kovanen.Easterbrook.2002}. Although most of these moraines characterise
-independent valley glaciers, that may have been disconnected from the waning
-Cordilleran ice sheet, the Finlay River area in the Omenica Mountains
-(Fig.~\ref{fig:hr_geom_deglacage}, OM) presents a different kind of evidence.
-There, sharp-crested moraines indicate a late-glacial readvance of local alpine
-glaciers and, more importantly, their interaction with larger, lingering
-remnants of the main body of the Cordilleran ice sheet in the valleys
-\citep{Lakeman.etal.2008}. Additional evidence for late-glacial
-alpine glacier readvances includes moraines in the eastern Coast Mountains,
-and the Columbia and Rocky mountains \citep{Osborn.Gerloff.1997,
-Menounos.etal.2008}.
-
-Although further work is needed to constrain the timing of the late-glacial
-readvance, to assess its extents and geographical distribution, and to identify
-the potential climatic triggers \citep{Menounos.etal.2008}, it is interesting
-to note that the simulation driven by the GRIP record produces a late-glacial
-readvance in the Coast Mountains and in the Columbia and Rocky Mountains
-(Fig.~\ref{fig:hr_geom_deglacage},
-left panel). In addition to matching the location of some local readvances, the
-GRIP-driven simulation shows that a large remnant of the decaying ice sheet
-may still have existed at the time of this late-glacial readvance.
-In contrast, the EPICA-driven simulation produces a nearly-continuous
-deglaciation with only a tightly restricted late-glacial readvance on the
-western slopes of the Saint Elias and the Coast mountains
+      The possibility of late glacial readvances in the North American
+      Cordillera has been debated for some time \citep{Osborn.Gerloff.1997},
+      and locally these have been reconstructed and dated. Radiocarbon-dated
+      end moraines in the Fraser and Squamish valleys, off the southern tip
+      of the Coast Mountains, indicate consecutive glacier maxima, or
+      standstills while in overall retreat, one of which corresponds to the
+      Younger Dryas chronozone \citep{Clague.etal.1997, Friele.Clague.2002,
+      Friele.Clague.2002a, Kovanen.2002, Kovanen.Easterbrook.2002}. Although
+      most of these moraines characterise independent valley glaciers, that
+      may have been disconnected from the waning Cordilleran ice sheet, the
+      Finlay River area in the Omenica Mountains
+      (Fig.~\ref{fig:hr_geom_deglacage}, OM) presents a different kind of
+      evidence. There, sharp-crested moraines indicate a late-glacial
+      readvance of local alpine glaciers and, more importantly, their
+      interaction with larger, lingering remnants of the main body of the
+      Cordilleran ice sheet in the valleys \citep{Lakeman.etal.2008}.
+      Additional evidence for late-glacial alpine glacier readvances
+      includes moraines in the eastern Coast Mountains, and the Columbia and
+      Rocky mountains \citep{Osborn.Gerloff.1997, Menounos.etal.2008}.
+
+      Although further work is needed to constrain the timing of the
+      late-glacial readvance, to assess its extents and geographical
+      distribution, and to identify the potential climatic triggers
+      \citep{Menounos.etal.2008}, it is interesting to note that the
+      simulation driven by the GRIP record produces a late-glacial readvance
+      in the Coast Mountains and in the Columbia and Rocky Mountains
+      (Fig.~\ref{fig:hr_geom_deglacage}, left panel). In addition to
+      matching the location of some local readvances, the GRIP-driven
+      simulation shows that a large remnant of the decaying ice sheet may
+      still have existed at the time of this late-glacial readvance. In
+      contrast, the EPICA-driven simulation produces a nearly-continuous
+      deglaciation with only a tightly restricted late-glacial readvance on
+      the western slopes of the Saint Elias and the Coast mountains
       (Fig.~\ref{fig:hr_geom_deglacage}, right panel).
@@
 
-Because a general tenant in glacial geomorphology is that the majority of
-landforms (lineations and eskers) are part of the deglacial envelope
-\citep[terminology from][]{Kleman.etal.2006}, having been formed
-close inside the retreating margin of ice sheets \citep{Boulton.Clark.1990,
-Kleman.etal.1997, Kleman.etal.2010}, we present maps of basal flow directions
-immediately preceding deglaciation or at the time of cessation of sliding
-inside a cold-based retreating margin (Fig.~\ref{fig:hr_geom_lastflow}). The
-modelled deglacial flow patterns are mostly consistent between the GRIP and EPICA
-simulations. They depict an active ice sheet retreat in the peripheral areas,
-followed by stagnant ice decay in some of the interior regions. Several parts
-of the modelling domain do not experience any basal sliding throughout the
-deglaciation phase (Fig.~\ref{fig:hr_geom_lastflow}, hatched areas). This notably
+      Because a general tenant in glacial geomorphology is that the majority
+      of landforms (lineations and eskers) are part of the deglacial
+      envelope \citep[terminology from][]{Kleman.etal.2006}, having been
+      formed close inside the retreating margin of ice sheets
+      \citep{Boulton.Clark.1990, Kleman.etal.1997, Kleman.etal.2010}, we
+      present maps of basal flow directions immediately preceding
+      deglaciation or at the time of cessation of sliding inside a
+      cold-based retreating margin (Fig.~\ref{fig:hr_geom_lastflow}). The
+      modelled deglacial flow patterns are mostly consistent between the
+      GRIP and EPICA simulations. They depict an active ice sheet retreat in
+      the peripheral areas, followed by stagnant ice decay in some of the
+      interior regions. Several parts of the modelling domain do not
+      experience any basal sliding throughout the deglaciation phase
+      (Fig.~\ref{fig:hr_geom_lastflow}, hatched areas). This notably
       includes parts of the Interior Plateau in British Columbia, major
-portions of the Alaskan sector of the ice sheet, and a tortuous ribbon running
-from the Northern Rocky Mountains over the Skeena and Selwyn Mountains and into
-the Mackenzie Mountains.
-
-Patterns of glacial lineations formed in the northern and southern sectors of
-the Cordilleran ice sheet (\citealp{Prest.etal.1968};
-\citealp[Fig.~1.12]{Clague.1989}; \citealp[Fig.~2]{Kleman.etal.2010}) show
-similarities with the patterns of deglacial ice flow from numerical modelling
-(Fig.~\ref{fig:hr_geom_lastflow}). In the northern half
-of the modelling domain, modelled deglacial flow directions depict an
-active downhill flow as the last remnants of the ice sheet retreat towards
-mountain ranges. Converging deglacial flow patterns in the Liard
-Lowland, for instance (Fig.~\ref{fig:hr_geom_lastflow}), closely resemble the pattern
-indicated by glacial lineations \citep[Fig.~2]{Margold.etal.2013}.
-
-On the Interior Plateau of south-central British Columbia, both simulations
-produce a retreat of the ice margin towards the north-east
-(Fig.~\ref{fig:hr_geom_deglacage}), a pattern which is validated by the geomorphological
-and stratigraphical record for ancient pro-glacial lakes dammed by the
-retreating ice sheet \citep{Perkins.Brennand.2014}. However, the two
-simulations differ in the mode of retreat. The GRIP simulation yields an active
-retreat with basal sliding towards the ice margin to the south, whereas the
-EPICA simulation produces negligible basal sliding on the plateau during deglaciation
-(Fig.~\ref{fig:hr_geom_lastflow}). Yet, the Interior Plateau also hosts an impressive
-set of glacial lineations which indicate a substantial eastwards flow component
-of the Cordilleran ice sheet \citep{Prest.etal.1968, Kleman.etal.2010}.
-This Interior Plateau lineation set could therefore present a smoking gun for the
-reliability of the presented model results. One explanation for the incongruent
-results could be that the missing feedback mechanisms between ice sheet
-topography and regional climate resulted in a modelled ice divide of the LGM
-ice sheet being too far to the east (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
-\citealp{Seguinot.etal.2014}). A more westerly-located LGM ice divide would
-certainly result in a different deglacial flow pattern over the Interior
-Plateau. However, a more westerly-positioned LGM ice divide would certainly be
-associated with a thinner ice sheet than that modelled here. Decreased ice
-thickness would not promote warm-based conditions but, on the
-contrary, enlarge the region of negligible basal sliding
-(Fig.~\ref{fig:hr_geom_lastflow}). Thus, a second explanation for the incongruent
-results could be that the Interior Plateau lineation system predates
-deglaciation ice flow, as perhaps indicated by some eskers
-that appear incompatible with these glacial lineations
-\citep[Fig.~9]{Margold.etal.2013a}. Finally, a third explanation could be that
-local geothermal heat associated with volcanic activity on the Interior Plateau
-could have triggered the basal sliding
-\citep[cf. Greenland ice sheet;][]{Fahnestock.etal.2001}.
+      portions of the Alaskan sector of the ice sheet, and a tortuous ribbon
+      running from the Northern Rocky Mountains over the Skeena and Selwyn
+      Mountains and into the Mackenzie Mountains.
+
+      Patterns of glacial lineations formed in the northern and southern
+      sectors of the Cordilleran ice sheet (\citealp{Prest.etal.1968};
+      \citealp[Fig.~1.12]{Clague.1989}; \citealp[Fig.~2]{Kleman.etal.2010})
+      show similarities with the patterns of deglacial ice flow from
+      numerical modelling (Fig.~\ref{fig:hr_geom_lastflow}). In the northern
+      half of the modelling domain, modelled deglacial flow directions
+      depict an active downhill flow as the last remnants of the ice sheet
+      retreat towards mountain ranges. Converging deglacial flow patterns in
+      the Liard Lowland, for instance (Fig.~\ref{fig:hr_geom_lastflow}),
+      closely resemble the pattern indicated by glacial lineations
+      \citep[Fig.~2]{Margold.etal.2013}.
+
+      On the Interior Plateau of south-central British Columbia, both
+      simulations produce a retreat of the ice margin towards the north-east
+      (Fig.~\ref{fig:hr_geom_deglacage}), a pattern which is validated by
+      the geomorphological and stratigraphical record for ancient
+      pro-glacial lakes dammed by the retreating ice sheet
+      \citep{Perkins.Brennand.2014}. However, the two simulations differ in
+      the mode of retreat. The GRIP simulation yields an active retreat with
+      basal sliding towards the ice margin to the south, whereas the EPICA
+      simulation produces negligible basal sliding on the plateau during
+      deglaciation (Fig.~\ref{fig:hr_geom_lastflow}). Yet, the Interior
+      Plateau also hosts an impressive set of glacial lineations which
+      indicate a substantial eastwards flow component of the Cordilleran ice
+      sheet \citep{Prest.etal.1968, Kleman.etal.2010}. This Interior Plateau
+      lineation set could therefore present a smoking gun for the
+      reliability of the presented model results. One explanation for the
+      incongruent results could be that the missing feedback mechanisms
+      between ice sheet topography and regional climate resulted in a
+      modelled ice divide of the LGM ice sheet being too far to the east
+      (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
+      \citealp{Seguinot.etal.2014}). A more westerly-located LGM ice divide
+      would certainly result in a different deglacial flow pattern over the
+      Interior Plateau. However, a more westerly-positioned LGM ice divide
+      would certainly be associated with a thinner ice sheet than that
+      modelled here. Decreased ice thickness would not promote warm-based
+      conditions but, on the contrary, enlarge the region of negligible
+      basal sliding (Fig.~\ref{fig:hr_geom_lastflow}). Thus, a second
+      explanation for the incongruent results could be that the Interior
+      Plateau lineation system predates deglaciation ice flow, as perhaps
+      indicated by some eskers that appear incompatible with these glacial
+      lineations \citep[Fig.~9]{Margold.etal.2013a}. Finally, a third
+      explanation could be that local geothermal heat associated with
+      volcanic activity on the Interior Plateau could have triggered the
+      basal sliding \citep[cf. Greenland ice
+      sheet;][]{Fahnestock.etal.2001}.
 
       The modelled deglaciation of the Interior Plateau of British Columbia
-consists of a rapid northwards retreat (Fig.~\ref{fig:hr_geom_deglacage}) of
-southwards-flowing non-sliding ice lobes (Fig.~\ref{fig:hr_geom_lastflow}) positioned
-in-between deglaciated mountain ranges
-(Figs.~\ref{fig:hr_pf_grip} and~\ref{fig:hr_pf_epica}). This result
-appears compatible with the prevailing conceptual model of deglaciation of
-central British Columbia, in which mountain ranges emerge from the ice before
-the plateau \citep[Fig.~7]{Fulton.1991}. However, due to different topographic
-and climatic conditions, our simulations produce different deglaciation
-patterns in the northern half of the model domain, indicating that this
-conceptual model may not be applied to the entire area formerly covered by the
-Cordilleran ice sheet.
+      consists of a rapid northwards retreat
+      (Fig.~\ref{fig:hr_geom_deglacage}) of southwards-flowing non-sliding
+      ice lobes (Fig.~\ref{fig:hr_geom_lastflow}) positioned in-between
+      deglaciated mountain ranges (Figs.~\ref{fig:hr_pf_grip}
+      and~\ref{fig:hr_pf_epica}). This result appears compatible with the
+      prevailing conceptual model of deglaciation of central British
+      Columbia, in which mountain ranges emerge from the ice before the
+      plateau \citep[Fig.~7]{Fulton.1991}. However, due to different
+      topographic and climatic conditions, our simulations produce different
+      deglaciation patterns in the northern half of the model domain,
+      indicating that this conceptual model may not be applied to the entire
+      area formerly covered by the Cordilleran ice sheet.
 
@@
 
-Numerical simulations of the Cordilleran ice sheet through the last glacial
-cycle presented in this study consistently produce two glacial maxima during
-MIS~4 (61.9--56.5\,ka, 3.8--8.8\,m~s.l.e.) and MIS~2 (23.2--16.8\,ka,
-8.40--8.91\,m~s.l.e.), two periods corresponding to documented extensive
-glaciations. This result is independent of the palaeo-temperature record used
-among the six selected for this study, and thus can be regarded as a robust model
-output, which broadly matches geological evidence. However, the timing of the
-two glaciation peaks depends sensitively on which climate record is used to
-drive the model. The timing of the LGM is best reproduced by the EPICA and
-Vostok Antarctic ice core records. It occurs about 2\,ka too early in the
-simulation forced by the GRIP ice core record, and occurs even earlier in all other
-simulations. The mismatch is largest for the two Northwest Pacific ODP
-palaeo-temperature records, which are affected by the weakening of the
-California current during the LGM.
-
-In all simulations presented here, ice cover is limited to disjoint mountain
-ice caps during most of the glacial cycle. The most persistent nucleation
-centres are located in the Coast Mountains, the Columbia and Rocky mountains,
-the Selwyn and Mackenzie mountains, and most importantly, in the Skeena
-Mountains. Throughout the modelled last glacial cycle, the Skeena Mountains
-host an ice cap which appears to be fed by the moisture intruding inland from the
-west through a topographic breach in the Coast Mountains. The Skeena ice cap acts
-as the main nucleation centre for the glacial reavance towards the LGM
-configuration. As indicated
-by persistent, warm-based ice in the model, this ice cap perhaps explains
-the distinct glacial erosional imprint observed on the landscape of the Skeena
-Mountains.
-
-During deglaciation, none of the climate records used can be selected as
-producing an optimal agreement between the model results and the geological
-evidence. Although the EPICA-driven simulation yields the most realistic timing
-of the LGM and, therefore, start of deglaciation, only the GRIP-driven
-simulation produces late glacial readvances in areas where these have been
-documented. Nonetheless, the patterns of ice sheet retreat are consistent
-between the two simulations, and show a rapid deglaciation of the southern
-sector of the ice sheet, including a rapid northwards retreat across the Interior
-Plateau of central British Columbia. The GRIP-driven simulation then produces a
-late-glacial readvance of local ice caps and of the main body of the decaying
-Cordilleran ice sheet primarily in the Coast and the Columbia and Rocky Mountains.
-In both simulations, this is followed by an
-opening of the Liard Lowland, and a final retreat of the remaining ice caps
-towards the Selwyn and, finally, the Skeena mountains, which hosts the last
-remnant of the ice sheet during the middle Holocene (6.6--6.2\,ka). Our results
-identify the Skeena Mountains as a key area to understanding glacial dynamics
-of the Cordilleran ice sheet, highlighting the need for further geological
-investigation of this region.
+      Numerical simulations of the Cordilleran ice sheet through the last
+      glacial cycle presented in this study consistently produce two glacial
+      maxima during MIS~4 (61.9--56.5\,ka, 3.8--8.8\,m~s.l.e.) and MIS~2
+      (23.2--16.8\,ka, 8.40--8.91\,m~s.l.e.), two periods corresponding to
+      documented extensive glaciations. This result is independent of the
+      palaeo-temperature record used among the six selected for this study,
+      and thus can be regarded as a robust model output, which broadly
+      matches geological evidence. However, the timing of the two glaciation
+      peaks depends sensitively on which climate record is used to drive the
+      model. The timing of the LGM is best reproduced by the EPICA and
+      Vostok Antarctic ice core records. It occurs about 2\,ka too early in
+      the simulation forced by the GRIP ice core record, and occurs even
+      earlier in all other simulations. The mismatch is largest for the two
+      Northwest Pacific ODP palaeo-temperature records, which are affected
+      by the weakening of the California current during the LGM.
+
+      In all simulations presented here, ice cover is limited to disjoint
+      mountain ice caps during most of the glacial cycle. The most
+      persistent nucleation centres are located in the Coast Mountains, the
+      Columbia and Rocky mountains, the Selwyn and Mackenzie mountains, and
+      most importantly, in the Skeena Mountains. Throughout the modelled
+      last glacial cycle, the Skeena Mountains host an ice cap which appears
+      to be fed by the moisture intruding inland from the west through a
+      topographic breach in the Coast Mountains. The Skeena ice cap acts as
+      the main nucleation centre for the glacial reavance towards the LGM
+      configuration. As indicated by persistent, warm-based ice in the
+      model, this ice cap perhaps explains the distinct glacial erosional
+      imprint observed on the landscape of the Skeena Mountains.
+
+      During deglaciation, none of the climate records used can be selected
+      as producing an optimal agreement between the model results and the
+      geological evidence. Although the EPICA-driven simulation yields the
+      most realistic timing of the LGM and, therefore, start of
+      deglaciation, only the GRIP-driven simulation produces late glacial
+      readvances in areas where these have been documented. Nonetheless, the
+      patterns of ice sheet retreat are consistent between the two
+      simulations, and show a rapid deglaciation of the southern sector of
+      the ice sheet, including a rapid northwards retreat across the
+      Interior Plateau of central British Columbia. The GRIP-driven
+      simulation then produces a late-glacial readvance of local ice caps
+      and of the main body of the decaying Cordilleran ice sheet primarily
+      in the Coast and the Columbia and Rocky Mountains. In both
+      simulations, this is followed by an opening of the Liard Lowland, and
+      a final retreat of the remaining ice caps towards the Selwyn and,
+      finally, the Skeena mountains, which hosts the last remnant of the ice
+      sheet during the middle Holocene (6.6--6.2\,ka). Our results identify
+      the Skeena Mountains as a key area to understanding glacial dynamics
+      of the Cordilleran ice sheet, highlighting the need for further
+      geological investigation of this region.
 
@@
 \section*{Author contributions}
-J.~Seguinot ran the simulations; I.~Rogozhina
-guided experiment design; A.~P.~Stroeven, M.~Margold and J.~Kleman took part in
-the interpretation and comparison of model results against geological evidence.
-All authors contributed to the text.
+      J.~Seguinot ran the simulations; I.~Rogozhina guided experiment
+      design; A.~P.~Stroeven, M.~Margold and J.~Kleman took part in the
+      interpretation and comparison of model results against geological
+      evidence. All authors contributed to the text.
 
@@
   constructive analysis of this study during J.~Seguinot's Ph.D. defence
-(September 2014). His comments were used to improve the model set-up.
-We are very thankful to Constantine Khroulev, Ed Bueler, and Andy Aschwanden for
-providing constant help and development with PISM. This work was supported by the
-Swedish Research Council~(VR) grant no.~2008-3449 to A.~P.~Stroeven and by the
-German Academic Exchange Service~(DAAD) grant no.~50015537 and a Knut and Alice
-Wallenberg Foundation grant to J.~Seguinot. Computer resources were provided
-by the Swedish National Infrastructure for Computing (SNIC) allocation
-no.~2013/1-159 and~2014/1-159 to A.~P.~Stroeven at the National Supercomputing
-Center (NSC).
+  (September 2014). His comments were used to improve the model set-up. We
+  are very thankful to Constantine Khroulev, Ed Bueler, and Andy Aschwanden
+  for providing constant help and development with PISM. This work was
+  supported by the Swedish Research Council~(VR) grant no.~2008-3449 to
+  A.~P.~Stroeven and by the German Academic Exchange Service~(DAAD) grant
+  no.~50015537 and a Knut and Alice Wallenberg Foundation grant to
+  J.~Seguinot. Computer resources were provided by the Swedish National
+  Infrastructure for Computing (SNIC) allocation no.~2013/1-159
+  and~2014/1-159 to A.~P.~Stroeven at the National Supercomputing Center
+  (NSC).
 \end{acknowledgements}

### 2016-01-16 cordillera-cycle.tex
Typeset: simplify header.
Typeset: move tables to the bottom.
Typeset: move figures to the end. Remove endfloat.
Typeset: apply nonsense linebreaks in abstract.

@@
-% cordillera-cycle.tex
-% =============================================================================
-
-% Copernicus-like latex2rtf compatible
-\newif\iflatextortf
-\iflatextortf
-\input{copernicus_rtf}
-\else
-
-% Copernicus manuscript
 \documentclass[tc, manuscript]{copernicus}
 
-% Copernicus final print
-%\documentclass[tc]{copernicus}
-
-% Copernicus discussion paper
-%\documentclass[tcd, hvmath, online]{copernicus}
-\fi
-
-% Coloured hyperlinks
-\hypersetup{hidelinks}
-
-% Push floats to the end
-\usepackage[nolists, nomarkers]{endfloat}
-
-% Figure directory
 \graphicspath{{figures/}}
-
-% My commands
-%\newcommand{\renote}[1]{\footnote{\textbf{Comment reply}: #1}}
-%\newcommand{\todo}[1]{\emph{[\textbf{Todo:} #1]}}
-%\newcommand{\aref}[0]{\textbf{[ref.]}}
 
@@
 
-% Title
 \title{Numerical simulations of the Cordilleran ice sheet
@@
 
-% Authors
 \Author[1,2,3]{J.}{Seguinot}
@@
 
-% Running title
 \runningtitle{Numerical simulations of the Cordilleran ice sheet
@@
 
-% Affiliations
 \affil[1]{Laboratory of Hydraulics, Hydrology and Glaciology, ETH Zürich,
@@
 
-% For Copernicus
 \received{}
@@
 
-% Title
 \firstpage{1}
@@
 
-% Abstract
 \begin{abstract}
-
-  Despite more than a century of geological observations, the Cordilleran ice
-  sheet of North America remains poorly understood in terms of its former
-  extent, volume and dynamics. Although geomorphological evidence is abundant,
-  its complexity is such that whole ice-sheet reconstructions of advance and
-  retreat patterns are lacking. Here we use a numerical ice sheet model
-  calibrated against field-based evidence to attempt a quantitative
-  reconstruction of the Cordilleran ice sheet history through the last glacial
-  cycle. A series of simulations is driven by time-dependent temperature
-  offsets from six proxy records located around the globe. Although this
-  approach reveals large variations in model response to evolving climate
-  forcing, all simulations produce two major glaciations during
-  marine oxygen isotope stages~4 (61.9--56.5\,ka) and~2
-  (23.2--16.8\,ka). The timing of glaciation is
-  better reproduced using temperature reconstructions from Greenland and
-  Antarctic ice cores than from regional oceanic sediment cores. During most of
-  the last glacial cycle, the modelled ice cover is discontinuous and
-  restricted to high mountain areas. However, widespread precipitation over the
-  Skeena Mountains favours the persistence of a central ice dome throughout the
-  glacial cycle. It acts as a nucleation centre before the Last Glacial Maximum
-  and hosts the last remains of Cordilleran ice until the middle Holocene
-  (6.6--6.2\,ka).
-
+Despite more than a century of geological observations, the
+Cordilleran ice sheet of North America remains poorly understood in
+terms of its former extent, volume and dynamics. Although
+geomorphological evidence is abundant, its complexity is such that
+whole ice-sheet reconstructions of advance and retreat patterns are
+lacking. Here we use a numerical ice sheet model calibrated against
+field-based evidence to attempt a quantitative reconstruction of the
+Cordilleran ice sheet history through the last glacial
+cycle. A series of simulations is driven by time-dependent
+temperature offsets from six proxy records located around the
+globe. Although this approach reveals large variations in model
+response to evolving climate forcing, all simulations produce two
+major glaciations during marine oxygen isotope stages~4
+(61.9--56.5\,ka) and~2 (23.2--16.8\,ka). The timing of
+glaciation is better reproduced using temperature reconstructions
+from Greenland and Antarctic ice cores than from regional oceanic
+sediment cores. During most of the last glacial cycle, the modelled
+ice cover is discontinuous and restricted to high mountain
+areas. However, widespread precipitation over the Skeena Mountains
+favours the persistence of a central ice dome throughout the glacial
+cycle. It acts as a nucleation centre before the Last Glacial
+Maximum and hosts the last remains of Cordilleran ice until the
+qmiddle Holocene (6.6--6.2\,ka).
 \end{abstract}
@@
 
-\begin{figure}
-  \includegraphics{locmap}
-  \caption{Relief map of northern North America showing a reconstruction of the
-           areas once covered by the Cordilleran (CIS), Laurentide (LIS),
-           Innuitian (IIS), and Greenland (GIS) ice sheets during the last
-           18\,\unit{\chem{^{14}C}\,ka} \citep[21.4\,cal\,ka,][]{Dyke.2004}.
-           The rectangular box denotes the location of the
-           modelling domain used in this study. Major mountain ranges covered
-           by the ice sheet include the Alaska Range (AR), the Wrangell and
-           Saint Elias mountains (WSEM), the Selwyn and Mackenzie mountains
-           (SMKM), the Skeena Mountains (SM), the Coast Mountains (CM), the
-           Columbia and Rocky
-           Mountains (CRM), and the North Cascades (NC). Major depressions
-           include the Liard Lowland (LL), the Interior Plateau of British
-           Columbia (IP), and the Puget Lowland (PL). The background
-           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
-           Data \citep{Patterson.Kelso.2015}.}
-  \label{fig:locmap}
-\end{figure}
-
       In contrast, evidence for the deglaciation history of the Cordilleran
@@
 % -----------------------------------------------------------------------------
-
-\begin{table*}
-  \centering
-  \caption{Parameter values used in the ice sheet model.}
-  \label{tab:params}
-  {\begin{tabular}{llrl}
-    \tophline
-    Not.    & Name & Value & Unit \\
-    \middlehline
-
-    $\rho$  & Ice density
-            & 910
-            & \unit{kg\,m^{-3}} \\
-
-    $g$     & Standard gravity
-            & 9.81
-            & \unit{m\,s^{-2}} \\
-
-    \multicolumn{2}{l}{\emph{Basal sliding}} \\
-
-    $q$     & Pseudo-plastic sliding exponent
-            & 0.25
-            & - \\
-
-    $v_{th}$& Pseudo-plastic threshold velocity
-            & 100.0
-            & \unit{m\,yr^{-1}} \\
-
-    $c_0$   & Till cohesion
-            & 0.0
-            & Pa \\
-
-    $\delta$& Effective pressure coefficient
-            & 0.02
-            & - \\
-
-    $e_0$   & Till reference void ratio
-            & 0.69
-            & - \\
-
-    $C_c$   & Till compressibility coefficient
-            & 0.12
-            & - \\
-
-    $W_{max}$ & Maximal till water thickness
-            & 2.0
-            & m \\
-
-    \multicolumn{2}{l}{\emph{Bedrock and lithosphere}} \\
-
-    $\rho_b$& Bedrock density
-            & 3300
-            & \unit{kg\,m^{-3}} \\
-
-    $c_b$   & Bedrock specific heat capacity
-            & 1000
-            & \unit{J\,kg^{-1}\,K^{-1}} \\
-
-    $k_b$   & Bedrock thermal conductivity
-            & 3.0
-            & \unit{J\,m^{-1}\,K^{-1}\,s^{-1}} \\
-
-    $\nu_m$ & Astenosphere viscosity
-            & $1\times10^{19}$
-            & \unit{Pa\,s} \\
-
-    $\rho_l$& Lithosphere density
-            & 3300
-            & \unit{kg\,m^{-3}} \\
-
-    $D$     & Lithosphere flexural rigidity
-            & $5.0\times10^{24}$
-            & \unit{N} \\
-
-    \multicolumn{2}{l}{\emph{Surface and atmosphere}} \\
-
-    $T_s$   & Temperature of snow precipitation
-            & 273.15
-            & \unit{K} \\
-
-    $T_r$   & Temperature of rain precipitation
-            & 275.15
-            & \unit{K} \\
-
-    $F_s$   & Degree-day factor for snow
-            & $3.04\times10^{-3}$
-            & \unit{m\,K^{-1}\,day^{-1}} \\
-
-    $F_i$   & Degree-day factor for ice
-            & $4.59\times10^{-3}$
-            & \unit{m\,K^{-1}\,day^{-1}} \\
-
-    $\gamma$& Air temperature lapse rate
-            & $6\times10^{-3}$
-            & \unit{K\,m^{-1}} \\
-
-    \bottomhline
-  \end{tabular}}
-  \belowtable{}
-\end{table*}
 
@@
 
-\begin{figure}
-  \includegraphics{atm}
-  \caption{Monthly mean near-surface air temperature, precipitation, and
-           standard deviation of daily mean temperature (PDD SD) for January
-           and July from the North American Regional Reanalysis
-           \citep[NARR;][]{Mesinger.etal.2006}, used to force the surface mass
-           balance (PDD) component of the ice sheet model. Note the
-           strong contrasts in seasonality, timing of the precipitation peak,
-           and temperature variability over the model domain, notably between
-           coastal and inland regions.}
-  \label{fig:atm}
-\end{figure}
-
       Ice surface accumulation and ablation are computed from monthly mean
@@
 % -----------------------------------------------------------------------------
-
-\begin{table*}[t]
-  \caption{Palaeo-temperature proxy records and scaling parameters yielding
-           temperature offset time-series used to force the ice sheet model
-           through the last glacial cycle (Fig.~\ref{fig:lr_ts}). $f$
-           corresponds to the scaling factor adopted to yield last glacial
-           maximum ice limits in the vicinity of mapped end moraines, and
-           $T_{[32, 22]}$ refers to the resulting mean temperature anomaly
-           during the period -32 to~-22~ka after scaling.}
-  \label{tab:records}
-  {\begin{tabular}{l|ccc|ccc|l}
-    \tophline
-    Record & Latitude & Longitude & Elev. & Proxy & $f$ & $T_{[32, 22]}$
-           & Reference\\
-    & & & (m~a.s.l.) & & & (K) & \\
-    \middlehline
-    GRIP     &  72{\degree} 35' N  % 72.58 (decimal)
-             &  37{\degree} 38' W  % 37.64 (decimal)
-             & 3238
-             & \chem{\delta^{18}O}
-             & 0.37 & -5.8  % -16.4126 (before scaling)
-             & \citet{Dansgaard.etal.1993} \\
-
-    NGRIP    &  75{\degree} 06' N  % 75.10
-             &  42{\degree} 19' W  % 42.32
-             & 2917
-             & \chem{\delta^{18}O}
-             & 0.24 & -6.5  % -26.7098
-             & \citet{Andersen.etal.2004} \\
-
-    EPICA    &  75{\degree} 06' S  % 75.1
-             & 123{\degree} 21' E  % 123.35
-             & 3233
-             & \chem{\delta^{18}O}
-             & 0.64 & -5.9  % -9.2055
-             & \citet{Jouzel.etal.2007} \\
-
-    Vostok   &  78{\degree} 28' S  % 78.8
-             & 106{\degree} 50' E  % 106.8
-             & 3488
-             & \chem{\delta^{18}O}
-             & 0.74 & -5.9  % -7.9550
-             & \citet{Petit.etal.1999} \\
-
-    ODP~1012 &  32{\degree} 17' N
-             & 118{\degree} 23' W
-             & -1772
-             & \chem{U^{K'}_{37}}
-             & 1.61 & -6.1  % -3.7889
-             & \citet{Herbert.etal.2001} \\
-
-    ODP~1020 &  41{\degree} 00' N
-             & 126{\degree} 26' W
-             & -3038
-             & \chem{U^{K'}_{37}}
-             & 1.20 & -6.0 % -5.0000
-             & \citet{Herbert.etal.2001} \\
-    \bottomhline
-  \end{tabular}}
-  \belowtable{}
-\end{table*}
 
@@
 
-\begin{figure*}
-  \includegraphics{lr_ts}
-  \caption{Temperature offset time-series from ice core and ocean records
-           (Table~\ref{tab:records}) used as palaeo-climate forcing for the ice
-           sheet model (top panel), and modelled ice volume (bottom panel)
-           through the last 120\,ka. Ice volumes are expressed in meters of sea
-           level equivalent (m~s.l.e.). Gray fields indicate Marine Oxygen
-           Isotope Stage (MIS) boundaries for MIS~2 and MIS~4 according to a
-           global compilation of benthic \chem{\delta^{18}O} records
-           \citep{Lisiecki.Raymo.2005}. Hatched rectangles highlight the
-           time-volume span for ice volume extremes corresponding to MIS~4
-           (61.9--56.5\,ka), MIS~3 (53.0--41.3\,ka), and MIS~2 (LGM,
-           23.2--16.8\,ka). Dotted lines correspond to GRIP- and EPICA-driven
-           5\,km-resolution runs.}
-  \label{fig:lr_ts}
-\end{figure*}
-
-\begin{figure*}
-  \includegraphics{lr_maps}
-  \caption{Snapshots of modelled surface topography (500\,m contours)
-           corresponding to the ice volume extremes indicated on
-           Fig.~\ref{fig:lr_ts}. An ice cap persists over the Skeena
-           Mountains (SM) during MIS~3. Note the occurence of spatial
-           similarities despite large differences in timing.}
-  \label{fig:lr_maps}
-\end{figure*}
+      Despite large differences in the input climate forcing
+      (Fig.~\ref{fig:lr_ts}, upper panel), model output presents consistent
+      features that can be observed across the range of forcing data used.
+      In all simulations, modelled ice volumes remain relatively low during
+      most of the glacial cycle, except during two major glacial events
+      which occur between 61.9 and 56.5\,ka during MIS~4, and between 23.2
+      and 16.8\,ka during MIS~2 (Fig.~\ref{fig:lr_ts}, lower panel). An ice
+      volume minimum is consistently reached between 53.0 and 41.3\,ka
+      during MIS~3. However, the magnitude and precise timing of these three
+      events depend significantly on the choice of proxy record used to
+      derive a time-dependent climate forcing (Table~\ref{tab:extrema}).
+
+      Simulations forced by the Greenland ice core palaeo-temperature
+      records (GRIP, NGRIP) produce the highest variability in modelled ice
+      volume throughout the last glacial cycle. In contrast, simulations
+      driven by oceanic (ODP~1012, ODP~1020) and Antarctic (EPICA, Vostok)
+      palaeo-temperature records generally result in lower ice volume
+      variability throughout the simulation length, resulting in lower
+      modelled ice volumes during MIS~4 and larger ice volumes during MIS~3.
+      The NGRIP climate forcing is the only one that results in a larger ice
+      volume during MIS~4 than during MIS~2.
+
+      While simulations driven by the GRIP and the two Antarctic
+      palaeo-temperature records attain a last ice volume maximum between
+      19.1 and 16.8\,ka, those informed by the NGRIP and the two oceanic
+      palaeo-temperature records attain their maximum ice volumes thousands
+      of years earlier. Moreover, the ODP~1012 run yields a rapid
+      deglaciation of the modelled area prior to 17\,ka. The ODP~1020
+      simulation predicts an early maximum in ice volume at 23.2\,ka,
+      followed by slower deglaciation than modelled using the other
+      palaeo-temperature records. Finally, whereas model runs forced by the
+      Antarctic palaeo-temperature records result in a rapid and
+      uninterrupted deglaciation after the LGM, the simulation driven by the
+      GRIP palaeo-temperature record also results in a rapid deglaciation
+      but in three steps, separated by two periods of ice sheet regrowth
+      (Fig.~\ref{fig:lr_ts}).
+
+% -----------------------------------------------------------------------------
+\subsection{Extreme configurations}
+% -----------------------------------------------------------------------------
+
+      Despite large differences in the timing of attained volume extrema
+      (Table~\ref{tab:extrema}), all model runs show relatively consistent
+      patterns of glaciation. During MIS~4, all simulations produce an
+      extensive ice sheet, covering an area of at least half of that
+      attained during MIS~2 (Table~\ref{tab:extrema};
+      Fig.~\ref{fig:lr_maps}, upper panels). Corresponding maximum ice
+      volumes also differ significantly between model runs, and vary between
+      3.84 and~8.84\,m sea level equivalents (m.~s.l.e.;
+      Table~\ref{tab:extrema}).
+
+      In the MIS~3 ice volume minimum reconstructions, a central ice cap
+      persists over the Skeena Mountains (Fig.~\ref{fig:lr_maps}, middle
+      panels). Although this ice cap is present in all simulations, its
+      dimensions depend sensitively on the choice of the applied
+      palaeo-temperature record. Modelled ice volume minima spread over a
+      wide range between 1.69 and 2.88\,m~s.l.e. (Table~\ref{tab:extrema}).
+
+      Modelled ice sheet geometries during the LGM (MIS~2;
+      Fig.~\ref{fig:lr_maps}, lower panels) invariably include a ca.
+      1500\,km-long central divide above 3000\,m~a.s.l. located along the
+      spine of the Rocky Mountains. Although the similarity of modelled ice
+      extents is a direct result of the choice of scaling factors applied to
+      different palaeo-temperature proxy records (Table~\ref{tab:records}),
+      it is interesting to note that modelled maximum ice volumes also fall
+      within a tight range of 8.40 to 8.91\,m~s.l.e.
+      (Table~\ref{tab:extrema}).
+
+% =============================================================================
+\section{Comparison with the geologic record}
+\label{sec:discussion}
+% =============================================================================
+
+      Large variations in the model responses to evolving climate forcing
+      reveal its sensitivity to the choice of palaeo-temperature proxy
+      record. To distinguish between different records, geological evidence
+      of former glaciations provide a basis for validation of our runs,
+      while the results from numerical modelling can perhaps help to analyse
+      some of the complexity of this evidence. In this section, we compare
+      model outputs to the geologic record, in terms of timing and
+      configuration of the maximum stages, location and lifetime of major
+      nucleation centres, and patterns of ice retreat during the last
+      deglaciation.
+
+% -----------------------------------------------------------------------------
+\subsection{Glacial maxima}
+% -----------------------------------------------------------------------------
+
+\subsubsection{Timing of glaciation}
+\label{sec:timing}
+
+      Independently of the palaeo-temperature records used to force the ice
+      sheet model, our simulations consistently produce two glacial maxima
+      during the last glacial cycle. The first maximum configuration is
+      obtained during MIS~4 (61.9--56.5\,ka) and the second during MIS~2
+      (23.2--16.8\,ka; Figs.~\ref{fig:lr_ts}, \ref{fig:lr_maps};
+      Table~\ref{tab:extrema}). These events broadly correspond in timing to
+      the Gladstone (MIS~4) and McConnell (MIS~2) glaciations documented by
+      geological evidence for the northern sector of the Cordilleran ice
+      sheet \citep{Duk-Rodkin.etal.1996, Ward.etal.2007, Stroeven.etal.2010,
+      Stroeven.etal.2014}, and to the Fraser Glaciation (MIS~2) documented
+      for its southern sector \citep{Porter.Swanson.1998,
+      Margold.etal.2014}. There is stratigraphical evidence for an MIS~4
+      glaciation in British Columbia \citep{Clague.Ward.2011} and in the
+      Puget Lowland \citep{Troost.2014}, but their extent and timing are
+      still highly conjectural \citep[perhaps MIS~4 or early MIS~3;
+      e.g.,][]{Cosma.etal.2008}.
+
+      The exact timing of modelled MIS~2 maximum ice volume depends strongly
+      on the choice of applied palaeo-temperature record, which allows for a
+      more in-depth comparison with geological evidence for the timing of
+      maximum Cordilleran ice sheet extent. In the Puget Lowland
+      (Fig.~\ref{fig:locmap}), the LGM advance of the southern Cordilleran
+      ice sheet margin has been constrained by radiocarbon dating on wood
+      between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,ka}
+      \citep{Porter.Swanson.1998}. These dates are consistent with
+      radiocarbon dates from the offshore sedimentary record, which reveals
+      an increase of glaciomarine sedimentation between 19.5 and
+      16.2\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008,
+      Taylor.etal.2014}. Radiocarbon dating of the northern Cordilleran ice
+      sheet margin is much less constrained but straddles presented
+      constraints from the southern margin. However, cosmogenic exposure
+      dating places the timing of maximum CIS extent during the McConnell
+      glaciation close to 17\,\unit{\chem{^{10}Be}\,ka}
+      \citep{Stroeven.etal.2010, Stroeven.etal.2014}. A sharp transition in
+      the sediment record of the Gulf of Alaska indicates a retreat of
+      regional outlet glaciers onto land at
+      14.8\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Davies.etal.2011}.
+
+      Among the simulations presented here, only those forced with the GRIP,
+      EPICA and Vostok palaeo-temperature records yield Cordilleran ice
+      sheet maximum extents that may be compatible with these field
+      constraints (Fig.~\ref{fig:lr_ts}, lower panel;
+      Table~\ref{tab:extrema}). Simulations driven by the NGRIP, ODP~1012
+      and ODP~1020 palaeo-temperature records, on the contrary, yield MIS~2
+      maximum Cordilleran ice sheet volumes that pre-date field-based
+      constraints by several thousands of years (about 6, 6 and 4\,ka
+      respectively). Concerning the simulations driven by oceanic records,
+      this early deglaciation is caused by an early warming present in the
+      alkenone palaeo-temperature reconstructions (Fig.~\ref{fig:lr_ts},
+      upper panel; \citealp[Fig.~3]{Herbert.etal.2001}). However, this early
+      warming is a local effect, corresponding to a weakening of the
+      California current \citep{Herbert.etal.2001}. The California current,
+      driving cold waters southwards along the south-western coast of North
+      America, has been shown to have weakened during each peak of global
+      glaciation (in SPECMAP) during the past 550\,ka, including the LGM,
+      resulting in paradoxically warmer sea-surface temperatures at the
+      locations of the ODP~1012 and ODP~1020 sites
+      \citep{Herbert.etal.2001}.
+
+      Because most of the marine margin of the Cordilleran ice sheet
+      terminated in a sector of the Pacific Ocean unaffected by variations
+      in the California current, it probably remained insensitive to this
+      local phenomenum. However, the above paradox illustrates the
+      complexity of ice-sheet feedbacks on regional climate, and
+      demonstrates that, although located in the neighbourhood of the
+      modelling domain, the ODP~1012 and ODP~1020 palaeo-temperature records
+      cannot be used as a realistic forcing to model the Cordilleran ice
+      sheet through the last glacial cycle. Similarly, the simulation using
+      the NGRIP palaeo-temperature record depicts an early onset of
+      deglaciation (Fig.~\ref{fig:lr_ts}) following its last glacial volume
+      maximum (22.9\,ka, Table~\ref{tab:extrema}) attained about 6\,ka
+      earlier than dated evidence of the LGM advance. There is a fair
+      agreement between the EPICA and Vostok palaeo-temperature records,
+      resulting in only small differences between the simulations driven by
+      those records. These differences are not subject to further analysis
+      here; instead we focus on simulations forced by palaeo-temperature
+      records from the GRIP and EPICA ice cores that appear to produce the
+      most realistic reconstructions of regional glaciation history, yet
+      bearing significant disparities in model output. To allow for a more
+      detailed comparison against the geological record, these two
+      simulations were re-run using a higher-resolution grid
+      (Sect.~\ref{sec:model}; Fig.~\ref{fig:lr_ts}, lower panel, dotted
+      lines).
+
+\subsubsection{Ice configuration during MIS~2}
+\label{sec:mis2}
+
+      During maximum glaciation, both simulations position the main
+      meridional ice divide over the western flank of the Rocky Mountains
+      (Fig.~\ref{fig:lr_maps}, lower panels; Fig.~\ref{fig:mis2}). This
+      result appears to contrast with palaeoglaciological reconstructions
+      for central and southern British Columbia with ice divides in a more
+      westerly position, over the western margin of the Interior Plateau
+      \citep{Ryder.etal.1991, Stumpf.etal.2000, Kleman.etal.2010,
+      Clague.Ward.2011, Margold.etal.2013a}. These indicate that a
+      latitudinal saddle connected ice dispersal centres in the Columbia
+      Mountains with the main ice divide \citep{Ryder.etal.1991,
+      Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}. A latitudinal
+      saddle does indeed feature in our modelling results, however, in an
+      inverse configuration between the main ice divide over the Columbia
+      Mountains and a secondary divide over the southern Coast Mountains
+      (Fig.~\ref{fig:mis2}).
+
+      Such deviation from the geological inferences could reflect the fact
+      that our model does not include feedback mechanisms between ice sheet
+      topography and the regional climate. Firstly, during the build-up
+      phase preceding the LGM, rapid accumulation over the Coast Mountains
+      enhanced the topographic barrier formed by these mountain ranges,
+      which likely resulted in a decrease of precipitation and, therefore, a
+      decrease of accumulation in the interior. Secondly, latent warming of
+      the moisture-depleted air parcels flowing over this enhanced
+      topography could have resulted in an inflow of potentially warmer air
+      over the eastern flank of the ice sheet, increasing melt along the
+      advancing margin \citep[cf.][]{Langen.etal.2012}. Because these two
+      processes, both with a tendency to limit ice-sheet growth, are absent
+      from our model, the eastern margin of the ice sheet and the position
+      of the main meridional ice divide are certainly biased towards the
+      east in our simulations \citep{Seguinot.etal.2014}.
+
+      However, field-based palaeoglaciological reconstructions have
+      struggled to reconcile the more westerly-centred ice divide in
+      south-central British Columbia with evidence in the Rocky Mountains
+      and beyond, that the Cordilleran ice sheet invaded the western
+      Interior Plains, where it merged with the southwestern margin of the
+      Laurentide ice sheet and was deflected to the south
+      \citep{Jackson.etal.1997, Bednarski.Smith.2007, Kleman.etal.2010,
+      Margold.etal.2013, Margold.etal.2013a}. Ice geometries from our model
+      runs do not have this problem, because the position and elevation of
+      the ice divide ensure significant ice drainage across the Rocky
+      Mountains at the LGM (Fig.~\ref{fig:mis2}).
+
+      During MIS~2, the modelled total ice volume peaks at 8.01\,m~s.l.e.
+      (19.1\,ka) in the GRIP simulation and at 8.77\,m~s.l.e. (17.2\,ka) in
+      the EPICA simulation.
+
+\subsubsection{Ice configuration during MIS~4}
+\label{sec:mis4}
+
+      The modelled ice sheet configurations corresponding to ice volume
+      maxima during MIS~4 are more sensitive to the choice of atmospheric
+      forcing than those corresponding to ice volume maxima during MIS~2
+      (Fig.~\ref{fig:lr_maps}, upper panels; Fig.~\ref{fig:mis4}). The GRIP
+      simulation (Fig.~\ref{fig:mis4}, left panel) results in a modelled
+      maximum ice sheet extent that closely resembles that obtained during
+      MIS~2, with the only major difference of being slightly less extensive
+      across northern and eastern sectors. In contrast, the EPICA simulation
+      produces a lower ice volume maximum (Fig.~\ref{fig:lr_ts}), which
+      translates in the modelled ice sheet geometry into a significantly
+      reduced southern sector, more restricted ice cover in northern and
+      eastern sectors, and generally lower ice surface elevations in the
+      interior (Fig.~\ref{fig:mis4}, right panel). Thus, only the GRIP
+      simulation can explain the presence of MIS~4 glacial deposits in the
+      Puget Lowland \citep{Troost.2014} and that of ice-rafted debris in the
+      marine sediment record offshore Vancouver Island at
+      ca.~47\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008}.
+
+      During MIS~4, the modelled total ice volume peaks at 7.19\,m~s.l.e.
+      (57.3\,ka) in the GRIP simulation and at 5.04\,m~s.l.e. (61.9\,ka) in
+      the EPICA simulation, corresponding to respectively 90\,\% and 57\,\%
+      of modelled MIS~2 ice volumes.
+
+% -----------------------------------------------------------------------------
+\subsection{Nucleation centres}
+% -----------------------------------------------------------------------------
+
+\subsubsection{Transient ice sheet states}
+
+      Palaeo-glaciological reconstructions are generally more robust for
+      maximum ice sheet extents and late ice sheet configurations than for
+      intermediate or minimum ice sheet extents and older ice sheet
+      configurations \citep{Kleman.etal.2010}. However, these maximum stages
+      are, by nature, extreme configurations, which do not necessarily
+      represent the dominant patterns of glaciation throughout the period of
+      ice cover \citep{Porter.1989, Kleman.Stroeven.1997, Kleman.etal.2008,
+      Kleman.etal.2010}.
+
+      For the Cordilleran ice sheet, geological evidence from radiocarbon
+      dating \citep{Clague.etal.1980, Clague.1985, Clague.1986,
+      Porter.Swanson.1998, Menounos.etal.2008}, cosmogenic exposure dating
+      \citep{Stroeven.etal.2010, Stroeven.etal.2014, Margold.etal.2014},
+      bedrock deformation in response to former ice loads
+      \citep{Clague.James.2002, Clague.etal.2005}, and offshore sedimentary
+      records \citep{Cosma.etal.2008, Davies.etal.2011} indicate that the
+      LGM maximum extent was short-lived. To compare this finding to our
+      simulations, we use numerical modelling output to compute durations of
+      ice cover throughout the last glacial cycle
+      (Fig.~\ref{fig:hr_geom_duration}).
+
+      The resulting maps show that, during most of the glacial cycle,
+      modelled ice cover is restricted to disjoint ice caps centred on major
+      mountain ranges of the North American Cordillera
+      (Fig.~\ref{fig:hr_geom_duration}, blue areas). A 2500\,km-long
+      continuous expanse of ice, extending from the Alaska Range in the
+      northeast to the Rocky Mountains in the southwest, is only in
+      operation for at most 32\,ka, which is about a third of the timespan
+      of the last glacial cycle (Fig.~\ref{fig:hr_geom_duration}, hatched
+      areas). However, except for its margins in the Pacific Ocean and in
+      the northern foothills of the Alaska Range, the maximum extent of the
+      ice sheet is attained for a much shorter period of time of only few
+      thousand years (Figs.~\ref{fig:hr_geom_duration}, red areas). This
+      result illustrates that the maximum extents of the modelled ice sheet
+      during MIS~4 and MIS~2 were both short-lived and therefore out of
+      balance with contemporary climate.
+
+      A notable exception to the transient character of the maximum extent
+      of Cordilleran ice sheet is the northern slope of the Alaska Range,
+      where modelled glaciers are confined to its foothills during the
+      entire simulation period (Fig.~\ref{fig:hr_geom_duration}, AR). This
+      apparent insensitivity of modelled glacial extent to temperature
+      fluctuations results from a combination of low precipitation, high
+      summer temperature and large temperature standard deviation (PDD SD)
+      in the plains of the Alaska Interior (Fig.~\ref{fig:atm}) which
+      confines glaciation to the foothills of the mountains. This result
+      could potentially explain the local distribution of glacial deposits,
+      which indicates that glaciers flowing on the northern slope of the
+      Alaska Range have remained small throughout the Pleistocene
+      \citep{Kaufman.Manley.2004}.
+
+\subsubsection{Major ice-dispersal centres}
+
+      It is generally believed that the Cordilleran ice sheet formed by the
+      coalescence of several mountain-centred ice caps
+      \citep{Davis.Mathews.1944}. In our simulations, major ice-dispersal
+      centres, visible on the modelled ice cover duration maps
+      (Fig.~\ref{fig:hr_geom_duration}), are located over the Coast
+      Mountains (CM), the Columbia and Rocky mountains (CRM), the Skeena
+      Mountains (SM), and the Selwyn and Mackenzie mountains (SMKM). The
+      Wrangell and Saint Elias mountains, heavily glacierized at present,
+      host an ice cap for the entire length of both simulations, but that
+      ice cap does not appear to be a major feed to the Cordilleran ice
+      sheet (Fig.~\ref{fig:hr_geom_duration}, WSEM). Although the Coast,
+      Skeena and Columbia and Rocky mountains (CM, SM, CRM) are covered by
+      mountain glaciers for most of the last glacial cycle, providing
+      durable nucleation centres for an ice sheet, this is not the case for
+      the Selwyn and Mackenzie mountains (SMKM), where ice cover on the
+      highest peaks is limited to a small fraction of the last glacial
+      cycle. In other words, the Selwyn and Mackenzie mountains only appear
+      as a secondary ice-dispersal centre during the coldest periods of the
+      last glacial cycle. The Northern Rocky Mountains
+      (Fig.~\ref{fig:hr_geom_duration}, NRM) do not act as a nucleation
+      centre, but rather as a pinning point for the Cordilleran ice sheet
+      margin coming from the west.
+
+      Perhaps the most striking feature displayed by the distributions of
+      modelled ice cover is the persistence of the Skeena Mountains ice cap
+      throughout the entire last glacial period (ca.~100--10\,ka) and its
+      predominance over the other ice-dispersal centres
+      (Figs.~\ref{fig:lr_maps} and~\ref{fig:hr_geom_duration}, SM).
+      Regardless of the applied forcing, this ice cap appears to survive
+      MIS~3 (Fig.~\ref{fig:lr_maps}, middle panels), and serves as a
+      nucleation centre at the onset of the glacial readvance towards the
+      LGM (MIS~2). This situation appear similar to the neighbouring
+      Laurentide ice sheet, for which the importance of residual ice for the
+      glacial history leading up to the LGM has been illustrated by the
+      MIS~3 residual ice bodies in northern and eastern Canada as nucleation
+      centres for a much more extensive MIS~2 configuration
+      \citep{Kleman.etal.2010}.
+
+      The presence of a Skeena Mountains ice cap during most of the last
+      glacial cycle can be explained by meteorological conditions more
+      favourable for ice growth there than elsewhere. In fact, reanalysed
+      atmospheric fields used to force the surface mass balance model show
+      that high winter precipitations are mainly confined to the western
+      slope of the Coast Mountains, except in the centre of the modelling
+      domain where they also occur further inland than along other east-west
+      transects (Fig.~\ref{fig:atm}). In fact, along most of the
+      north-western coast of North America, coastal mountain ranges form a
+      pronounced topographic barrier for westerly winds, capturing
+      atmospheric moisture in the form of orographic precipitation, and
+      resulting in arid interior lowlands. However, near the centre of our
+      modelling domain, this barrier is less pronounced than elsewhere,
+      allowing westerly winds to carry moisture further inland, until it is
+      captured by the extensive Skeena Mountains in north-central British
+      Columbia, thus resulting in a more widespread distribution of winter
+      precipitation (Fig.~\ref{fig:atm}).
+
+      The modelled total ice volume corresponding to these persistent
+      ice-dispersal centres attains a minimum of 1.52\,m~s.l.e. (42.9\,ka)
+      in the GRIP simulation and of 2.68\,m~s.l.e. (51.8\,ka) in the EPICA
+      simulation, corresponding to respectively 19\,\% and 31\,\% of the
+      MIS~2 ice volumes.
+
+\subsubsection{Erosional imprint on the landscape}
+
+      A correlation is observed between the modelled duration of warm based
+      ice cover (Fig.~\ref{fig:hr_geom_warmbase}) and the degree of glacial
+      modification of the landscape (mainly in terms of the development of
+      deep glacial valleys and troughs). We find evidence for this on the
+      slopes of the Coast Mountains, the Columbia and Rocky mountains, the
+      Wrangell and Saint Elias mountains, and radiating off the Skeena
+      Mountains (Figs.~\ref{fig:hr_geom_duration} and
+      \ref{fig:hr_geom_warmbase}; \citealp[Fig.~2]{Kleman.etal.2010}). The
+      Skeena Mountains, for example, indeed bear a strong glacial imprint
+      that indicates ice drainage in a system of distinct glacial troughs
+      emanating in a radial pattern from the centre of the mountain range
+      \citep[Fig.~2]{Kleman.etal.2010}. We suggest that persistent ice cover
+      (Fig.~\ref{fig:hr_geom_duration}) associated with basal ice
+      temperatures at the pressure melting point
+      (Figs.~\ref{fig:hr_geom_warmbase} and~\ref{fig:hr_geom_warmfrac})
+      explains the large-scale glacial erosional imprint on the landscape. A
+      well-developed network of glacial valleys running to the north-west on
+      the west slope of the Selwyn and Mackenzie Mountains
+      (\citealp[Fig.~2]{Kleman.etal.2010}; \citealp[Fig.~8]
+      {Stroeven.etal.2010}) is modelled to have hosted predominantly
+      warm-based ice (Fig.~\ref{fig:hr_geom_warmfrac}). However, because it
+      is only glaciated for a short fraction of the last glacial cycle in
+      our simulations (Fig.~\ref{fig:hr_geom_duration}), this perhaps
+      indicates that the observed landscape pattern originates from multiple
+      glacial cycles and witnesses an increased relative importance of the
+      Selwyn and Mackenzie mountains ice dispersal centre
+      (Fig.~\ref{fig:hr_geom_duration}, SMKM), prior to the Late Pleistocene
+      \citep[cf.][]{Ward.etal.2008, Demuro.etal.2012}.
+
+      The modelled distribution of warm-based ice cover
+      (Figs.~\ref{fig:hr_geom_warmbase} and~\ref{fig:hr_geom_warmfrac}) is
+      inevitably affected by our assumption of a constant,
+      70\,\unit{mW\,m^{-2}} geothermal heat flux at 3\,km depth
+      (Sect.~\ref{sec:icedyn}). However, the Skeena Mountains and the area
+      west of the Mackenzie Mountains experience higher-than-average
+      geothermal heat flux with measured values of ca.~80 and
+      ca.~100\,\unit{mW\,m^{-2}} \citep{Blackwell.Richards.2004}. We can
+      therefore expect even longer durations of warm-based ice cover for
+      these areas if we were to include spatially variable geothermal
+      forcing in our Cordilleran ice sheet simulations.
+
+% -----------------------------------------------------------------------------
+\subsection{The last deglaciation}
+% -----------------------------------------------------------------------------
+
+\subsubsection{Pace and patterns of deglaciation}
+
+      Similarly to other glaciated regions, most glacial traces in the North
+      American Cordillera relate to the last few millennia of glaciation,
+      because most of the older evidence has been overprinted by warm-based
+      ice retreat during the last deglaciation \citep{Kleman.1994,
+      Kleman.etal.2010}. From a numerical modelling perspective, phases of
+      glacier retreat are more challenging than phases of growth, because
+      they involve more rapid fluctuations of the ice margin, increased flow
+      velocities and longitudinal stress gradients, and poorly understood
+      hydrological processes. The latter are typically included in the
+      models through simple parametrisations
+      \citep[e.g.][]{Clason.etal.2012, Clason.etal.2014, Bueler.Pelt.2015},
+      if included at all. However, next after the mapping of maximum ice
+      sheet extents during MIS~2 and MIS~4 (Sect.~\ref{sec:mis2}
+      and~\ref{sec:mis4}), geomorphologically-based reconstructions of
+      patterns of ice sheet retreat during the last deglaciation provide the
+      second best source of evidence for the validation of our simulations.
+
+      In the North American Cordillera, the presence of lateral meltwater
+      channels at high elevation \citep{Margold.etal.2011,
+      Margold.etal.2013a, Margold.etal.2014}, and abundant esker systems at
+      low elevation \citep{Burke.etal.2012, Burke.etal.2012a,
+      Perkins.etal.2013, Margold.etal.2013} indicate that meltwater was
+      produced over large portions of the ice sheet surface during
+      deglaciation. The southern and northern margins of the Cordilleran ice
+      sheet reached their last glacial maximum extent around 17\,ka
+      \citep[Sect.~\ref{sec:timing};][]{Porter.Swanson.1998,
+      Cosma.etal.2008, Stroeven.etal.2010, Stroeven.etal.2014}, which we
+      take as a limiting age for the onset of ice retreat. The timing of
+      final deglaciation is less well constrained, but recent cosmogenic
+      dates from north-central British Columbia indicate that a seizable ice
+      cap emanating from the central Coast Mountains or the Skeena Mountains
+      persisted into the Younger Dryas chronozone, at least until 12.4\,ka
+      \citep{Margold.etal.2014}.
+
+      In our simulations, the timing of peak ice volume during the LGM and
+      the pacing of deglaciation depend critically on the choice of climate
+      forcing (Table~\ref{tab:extrema}; Figs.~\ref{fig:lr_ts}
+      and~\ref{fig:lr_ts_deglac}). Adopting the EPICA climate forcing yields
+      peak ice volume at 17.2\,ka and an uninterrupted deglaciation until
+      about 9\,ka (Fig.~\ref{fig:lr_ts_deglac}, lower panel, red curves). On
+      the contrary, the simulation driven by the GRIP palaeo-temperature
+      record yields peak ice volume at 19.3\,ka and a deglaciation
+      interrupted by two phases of regrowth until about 8\,ka. The first
+      interruption occurs between 16.6 and 14.5\,ka, and the second between
+      13.1 and 11.6\,ka (Fig.~\ref{fig:lr_ts_deglac}, lower panel, blue
+      curve).
+
+      Hence, the two model runs, while similar in overall timing compared to
+      the runs with other climate drivers, differ in detail. On the one
+      hand, the EPICA run depicts peak glaciation about 2\,ka later than the
+      GRIP run, in closer agreement with dated maximum extents, and shows a
+      faster, uninterrupted deglaciation which yields sporadic ice cover
+      more than 1\,ka earlier. On the other hand, the GRIP run yields a
+      deglaciation in three steps, compatible with marine sediment sequences
+      offshore Vancouver Island, where the distribution of ice-rafted debris
+      indicates an ice margin retreat from the Georgia Strait in two phases
+      that are contemporary with warming oceanic temperatures from 17.2 to
+      16.5 and from 15.5 to 14.0\,\unit{\chem{^{14}C}\,cal\,ka}
+      \citep{Taylor.etal.2014}.
+
+      Modelled patterns of ice sheet retreat are relatively consistent
+      between the two simulations (Figs.~\ref{fig:hr_geom_deglacageshots}
+      and.~\ref{fig:hr_geom_deglacage}). The southern sector of the
+      modelling domain, including the Puget Lowland, the Coast Mountains,
+      the Columbia and and Rocky mountains, and the Interior Plateau of
+      British Columbia, becomes completely deglaciated by 10\,ka, whereas a
+      significant ice cover remains over the Skeena, the Selwyn and
+      Mackenzie, and the Wrangell and Saint Elias mountains in the northern
+      sector of the modelling domain. After 10\,ka, deglaciation continues
+      to proceed across the Liard Lowland with a radial ice margin retreat
+      towards the surrounding mountain ranges, consistent with the regional
+      melt water record of the last deglaciation \citep{Margold.etal.2013}.
+      Remaining ice continues to decay by retreating towards the Selwyn and
+      MacKenzie, and Skeena mountains. The last remnants of the Cordilleran
+      ice sheet finally disappear from the Skeena Mountains at 6.5\,ka
+      (GRIP) and 6.1\,ka (EPICA).
+
+\subsubsection{Late-glacial readvance}
+
+      The possibility of late glacial readvances in the North American
+      Cordillera has been debated for some time \citep{Osborn.Gerloff.1997},
+      and locally these have been reconstructed and dated. Radiocarbon-dated
+      end moraines in the Fraser and Squamish valleys, off the southern tip
+      of the Coast Mountains, indicate consecutive glacier maxima, or
+      standstills while in overall retreat, one of which corresponds to the
+      Younger Dryas chronozone \citep{Clague.etal.1997, Friele.Clague.2002,
+      Friele.Clague.2002a, Kovanen.2002, Kovanen.Easterbrook.2002}. Although
+      most of these moraines characterise independent valley glaciers, that
+      may have been disconnected from the waning Cordilleran ice sheet, the
+      Finlay River area in the Omenica Mountains
+      (Fig.~\ref{fig:hr_geom_deglacage}, OM) presents a different kind of
+      evidence. There, sharp-crested moraines indicate a late-glacial
+      readvance of local alpine glaciers and, more importantly, their
+      interaction with larger, lingering remnants of the main body of the
+      Cordilleran ice sheet in the valleys \citep{Lakeman.etal.2008}.
+      Additional evidence for late-glacial alpine glacier readvances
+      includes moraines in the eastern Coast Mountains, and the Columbia and
+      Rocky mountains \citep{Osborn.Gerloff.1997, Menounos.etal.2008}.
+
+      Although further work is needed to constrain the timing of the
+      late-glacial readvance, to assess its extents and geographical
+      distribution, and to identify the potential climatic triggers
+      \citep{Menounos.etal.2008}, it is interesting to note that the
+      simulation driven by the GRIP record produces a late-glacial readvance
+      in the Coast Mountains and in the Columbia and Rocky Mountains
+      (Fig.~\ref{fig:hr_geom_deglacage}, left panel). In addition to
+      matching the location of some local readvances, the GRIP-driven
+      simulation shows that a large remnant of the decaying ice sheet may
+      still have existed at the time of this late-glacial readvance. In
+      contrast, the EPICA-driven simulation produces a nearly-continuous
+      deglaciation with only a tightly restricted late-glacial readvance on
+      the western slopes of the Saint Elias and the Coast mountains
+      (Fig.~\ref{fig:hr_geom_deglacage}, right panel).
+
+\subsubsection{Deglacial flow directions}
+
+      Because a general tenant in glacial geomorphology is that the majority
+      of landforms (lineations and eskers) are part of the deglacial
+      envelope \citep[terminology from][]{Kleman.etal.2006}, having been
+      formed close inside the retreating margin of ice sheets
+      \citep{Boulton.Clark.1990, Kleman.etal.1997, Kleman.etal.2010}, we
+      present maps of basal flow directions immediately preceding
+      deglaciation or at the time of cessation of sliding inside a
+      cold-based retreating margin (Fig.~\ref{fig:hr_geom_lastflow}). The
+      modelled deglacial flow patterns are mostly consistent between the
+      GRIP and EPICA simulations. They depict an active ice sheet retreat in
+      the peripheral areas, followed by stagnant ice decay in some of the
+      interior regions. Several parts of the modelling domain do not
+      experience any basal sliding throughout the deglaciation phase
+      (Fig.~\ref{fig:hr_geom_lastflow}, hatched areas). This notably
+      includes parts of the Interior Plateau in British Columbia, major
+      portions of the Alaskan sector of the ice sheet, and a tortuous ribbon
+      running from the Northern Rocky Mountains over the Skeena and Selwyn
+      Mountains and into the Mackenzie Mountains.
+
+      Patterns of glacial lineations formed in the northern and southern
+      sectors of the Cordilleran ice sheet (\citealp{Prest.etal.1968};
+      \citealp[Fig.~1.12]{Clague.1989}; \citealp[Fig.~2]{Kleman.etal.2010})
+      show similarities with the patterns of deglacial ice flow from
+      numerical modelling (Fig.~\ref{fig:hr_geom_lastflow}). In the northern
+      half of the modelling domain, modelled deglacial flow directions
+      depict an active downhill flow as the last remnants of the ice sheet
+      retreat towards mountain ranges. Converging deglacial flow patterns in
+      the Liard Lowland, for instance (Fig.~\ref{fig:hr_geom_lastflow}),
+      closely resemble the pattern indicated by glacial lineations
+      \citep[Fig.~2]{Margold.etal.2013}.
+
+      On the Interior Plateau of south-central British Columbia, both
+      simulations produce a retreat of the ice margin towards the north-east
+      (Fig.~\ref{fig:hr_geom_deglacage}), a pattern which is validated by
+      the geomorphological and stratigraphical record for ancient
+      pro-glacial lakes dammed by the retreating ice sheet
+      \citep{Perkins.Brennand.2014}. However, the two simulations differ in
+      the mode of retreat. The GRIP simulation yields an active retreat with
+      basal sliding towards the ice margin to the south, whereas the EPICA
+      simulation produces negligible basal sliding on the plateau during
+      deglaciation (Fig.~\ref{fig:hr_geom_lastflow}). Yet, the Interior
+      Plateau also hosts an impressive set of glacial lineations which
+      indicate a substantial eastwards flow component of the Cordilleran ice
+      sheet \citep{Prest.etal.1968, Kleman.etal.2010}. This Interior Plateau
+      lineation set could therefore present a smoking gun for the
+      reliability of the presented model results. One explanation for the
+      incongruent results could be that the missing feedback mechanisms
+      between ice sheet topography and regional climate resulted in a
+      modelled ice divide of the LGM ice sheet being too far to the east
+      (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
+      \citealp{Seguinot.etal.2014}). A more westerly-located LGM ice divide
+      would certainly result in a different deglacial flow pattern over the
+      Interior Plateau. However, a more westerly-positioned LGM ice divide
+      would certainly be associated with a thinner ice sheet than that
+      modelled here. Decreased ice thickness would not promote warm-based
+      conditions but, on the contrary, enlarge the region of negligible
+      basal sliding (Fig.~\ref{fig:hr_geom_lastflow}). Thus, a second
+      explanation for the incongruent results could be that the Interior
+      Plateau lineation system predates deglaciation ice flow, as perhaps
+      indicated by some eskers that appear incompatible with these glacial
+      lineations \citep[Fig.~9]{Margold.etal.2013a}. Finally, a third
+      explanation could be that local geothermal heat associated with
+      volcanic activity on the Interior Plateau could have triggered the
+      basal sliding \citep[cf. Greenland ice
+      sheet;][]{Fahnestock.etal.2001}.
+
+      The modelled deglaciation of the Interior Plateau of British Columbia
+      consists of a rapid northwards retreat
+      (Fig.~\ref{fig:hr_geom_deglacage}) of southwards-flowing non-sliding
+      ice lobes (Fig.~\ref{fig:hr_geom_lastflow}) positioned in-between
+      deglaciated mountain ranges (Figs.~\ref{fig:hr_pf_grip}
+      and~\ref{fig:hr_pf_epica}). This result appears compatible with the
+      prevailing conceptual model of deglaciation of central British
+      Columbia, in which mountain ranges emerge from the ice before the
+      plateau \citep[Fig.~7]{Fulton.1991}. However, due to different
+      topographic and climatic conditions, our simulations produce different
+      deglaciation patterns in the northern half of the model domain,
+      indicating that this conceptual model may not be applied to the entire
+      area formerly covered by the Cordilleran ice sheet.
+
+% =============================================================================
+\conclusions
+\label{sec:concl}
+% =============================================================================
+
+      Numerical simulations of the Cordilleran ice sheet through the last
+      glacial cycle presented in this study consistently produce two glacial
+      maxima during MIS~4 (61.9--56.5\,ka, 3.8--8.8\,m~s.l.e.) and MIS~2
+      (23.2--16.8\,ka, 8.40--8.91\,m~s.l.e.), two periods corresponding to
+      documented extensive glaciations. This result is independent of the
+      palaeo-temperature record used among the six selected for this study,
+      and thus can be regarded as a robust model output, which broadly
+      matches geological evidence. However, the timing of the two glaciation
+      peaks depends sensitively on which climate record is used to drive the
+      model. The timing of the LGM is best reproduced by the EPICA and
+      Vostok Antarctic ice core records. It occurs about 2\,ka too early in
+      the simulation forced by the GRIP ice core record, and occurs even
+      earlier in all other simulations. The mismatch is largest for the two
+      Northwest Pacific ODP palaeo-temperature records, which are affected
+      by the weakening of the California current during the LGM.
+
+      In all simulations presented here, ice cover is limited to disjoint
+      mountain ice caps during most of the glacial cycle. The most
+      persistent nucleation centres are located in the Coast Mountains, the
+      Columbia and Rocky mountains, the Selwyn and Mackenzie mountains, and
+      most importantly, in the Skeena Mountains. Throughout the modelled
+      last glacial cycle, the Skeena Mountains host an ice cap which appears
+      to be fed by the moisture intruding inland from the west through a
+      topographic breach in the Coast Mountains. The Skeena ice cap acts as
+      the main nucleation centre for the glacial reavance towards the LGM
+      configuration. As indicated by persistent, warm-based ice in the
+      model, this ice cap perhaps explains the distinct glacial erosional
+      imprint observed on the landscape of the Skeena Mountains.
+
+      During deglaciation, none of the climate records used can be selected
+      as producing an optimal agreement between the model results and the
+      geological evidence. Although the EPICA-driven simulation yields the
+      most realistic timing of the LGM and, therefore, start of
+      deglaciation, only the GRIP-driven simulation produces late glacial
+      readvances in areas where these have been documented. Nonetheless, the
+      patterns of ice sheet retreat are consistent between the two
+      simulations, and show a rapid deglaciation of the southern sector of
+      the ice sheet, including a rapid northwards retreat across the
+      Interior Plateau of central British Columbia. The GRIP-driven
+      simulation then produces a late-glacial readvance of local ice caps
+      and of the main body of the decaying Cordilleran ice sheet primarily
+      in the Coast and the Columbia and Rocky Mountains. In both
+      simulations, this is followed by an opening of the Liard Lowland, and
+      a final retreat of the remaining ice caps towards the Selwyn and,
+      finally, the Skeena mountains, which hosts the last remnant of the ice
+      sheet during the middle Holocene (6.6--6.2\,ka). Our results identify
+      the Skeena Mountains as a key area to understanding glacial dynamics
+      of the Cordilleran ice sheet, highlighting the need for further
+      geological investigation of this region.
+
+% Author contributions
+\section*{Author contributions}
+      J.~Seguinot ran the simulations; I.~Rogozhina guided experiment
+      design; A.~P.~Stroeven, M.~Margold and J.~Kleman took part in the
+      interpretation and comparison of model results against geological
+      evidence. All authors contributed to the text.
+
+% Acknowledgements
+\begin{acknowledgements}
+  Foremost, we would like to thank Shawn Marshall for providing a detailed,
+  constructive analysis of this study during J.~Seguinot's Ph.D. defence
+  (September 2014). His comments were used to improve the model set-up. We
+  are very thankful to Constantine Khroulev, Ed Bueler, and Andy Aschwanden
+  for providing constant help and development with PISM. This work was
+  supported by the Swedish Research Council~(VR) grant no.~2008-3449 to
+  A.~P.~Stroeven and by the German Academic Exchange Service~(DAAD) grant
+  no.~50015537 and a Knut and Alice Wallenberg Foundation grant to
+  J.~Seguinot. Computer resources were provided by the Swedish National
+  Infrastructure for Computing (SNIC) allocation no.~2013/1-159
+  and~2014/1-159 to A.~P.~Stroeven at the National Supercomputing Center
+  (NSC).
+\end{acknowledgements}
+
+% References
+\bibliographystyle{copernicus}
+\bibliography{refs/references}
+\newpage
+
+\begin{table*}
+  \centering
+  \caption{Parameter values used in the ice sheet model.}
+  \label{tab:params}
+  {\begin{tabular}{llrl}
+    \tophline
+    Not.    & Name & Value & Unit \\
+    \middlehline
+
+    $\rho$  & Ice density
+            & 910
+            & \unit{kg\,m^{-3}} \\
+
+    $g$     & Standard gravity
+            & 9.81
+            & \unit{m\,s^{-2}} \\
+
+    \multicolumn{2}{l}{\emph{Basal sliding}} \\
+
+    $q$     & Pseudo-plastic sliding exponent
+            & 0.25
+            & - \\
+
+    $v_{th}$& Pseudo-plastic threshold velocity
+            & 100.0
+            & \unit{m\,yr^{-1}} \\
+
+    $c_0$   & Till cohesion
+            & 0.0
+            & Pa \\
+
+    $\delta$& Effective pressure coefficient
+            & 0.02
+            & - \\
+
+    $e_0$   & Till reference void ratio
+            & 0.69
+            & - \\
+
+    $C_c$   & Till compressibility coefficient
+            & 0.12
+            & - \\
+
+    $W_{max}$ & Maximal till water thickness
+            & 2.0
+            & m \\
+
+    \multicolumn{2}{l}{\emph{Bedrock and lithosphere}} \\
+
+    $\rho_b$& Bedrock density
+            & 3300
+            & \unit{kg\,m^{-3}} \\
+
+    $c_b$   & Bedrock specific heat capacity
+            & 1000
+            & \unit{J\,kg^{-1}\,K^{-1}} \\
+
+    $k_b$   & Bedrock thermal conductivity
+            & 3.0
+            & \unit{J\,m^{-1}\,K^{-1}\,s^{-1}} \\
+
+    $\nu_m$ & Astenosphere viscosity
+            & $1\times10^{19}$
+            & \unit{Pa\,s} \\
+
+    $\rho_l$& Lithosphere density
+            & 3300
+            & \unit{kg\,m^{-3}} \\
+
+    $D$     & Lithosphere flexural rigidity
+            & $5.0\times10^{24}$
+            & \unit{N} \\
+
+    \multicolumn{2}{l}{\emph{Surface and atmosphere}} \\
+
+    $T_s$   & Temperature of snow precipitation
+            & 273.15
+            & \unit{K} \\
+
+    $T_r$   & Temperature of rain precipitation
+            & 275.15
+            & \unit{K} \\
+
+    $F_s$   & Degree-day factor for snow
+            & $3.04\times10^{-3}$
+            & \unit{m\,K^{-1}\,day^{-1}} \\
+
+    $F_i$   & Degree-day factor for ice
+            & $4.59\times10^{-3}$
+            & \unit{m\,K^{-1}\,day^{-1}} \\
+
+    $\gamma$& Air temperature lapse rate
+            & $6\times10^{-3}$
+            & \unit{K\,m^{-1}} \\
+
+    \bottomhline
+  \end{tabular}}
+  \belowtable{}
+\end{table*}
+
+\begin{table*}[t]
+  \caption{Palaeo-temperature proxy records and scaling parameters yielding
+           temperature offset time-series used to force the ice sheet model
+           through the last glacial cycle (Fig.~\ref{fig:lr_ts}). $f$
+           corresponds to the scaling factor adopted to yield last glacial
+           maximum ice limits in the vicinity of mapped end moraines, and
+           $T_{[32, 22]}$ refers to the resulting mean temperature anomaly
+           during the period -32 to~-22~ka after scaling.}
+  \label{tab:records}
+  {\begin{tabular}{l|ccc|ccc|l}
+    \tophline
+    Record & Latitude & Longitude & Elev. & Proxy & $f$ & $T_{[32, 22]}$
+           & Reference\\
+    & & & (m~a.s.l.) & & & (K) & \\
+    \middlehline
+    GRIP     &  72{\degree} 35' N  % 72.58 (decimal)
+             &  37{\degree} 38' W  % 37.64 (decimal)
+             & 3238
+             & \chem{\delta^{18}O}
+             & 0.37 & -5.8  % -16.4126 (before scaling)
+             & \citet{Dansgaard.etal.1993} \\
+
+    NGRIP    &  75{\degree} 06' N  % 75.10
+             &  42{\degree} 19' W  % 42.32
+             & 2917
+             & \chem{\delta^{18}O}
+             & 0.24 & -6.5  % -26.7098
+             & \citet{Andersen.etal.2004} \\
+
+    EPICA    &  75{\degree} 06' S  % 75.1
+             & 123{\degree} 21' E  % 123.35
+             & 3233
+             & \chem{\delta^{18}O}
+             & 0.64 & -5.9  % -9.2055
+             & \citet{Jouzel.etal.2007} \\
+
+    Vostok   &  78{\degree} 28' S  % 78.8
+             & 106{\degree} 50' E  % 106.8
+             & 3488
+             & \chem{\delta^{18}O}
+             & 0.74 & -5.9  % -7.9550
+             & \citet{Petit.etal.1999} \\
+
+    ODP~1012 &  32{\degree} 17' N
+             & 118{\degree} 23' W
+             & -1772
+             & \chem{U^{K'}_{37}}
+             & 1.61 & -6.1  % -3.7889
+             & \citet{Herbert.etal.2001} \\
+
+    ODP~1020 &  41{\degree} 00' N
+             & 126{\degree} 26' W
+             & -3038
+             & \chem{U^{K'}_{37}}
+             & 1.20 & -6.0 % -5.0000
+             & \citet{Herbert.etal.2001} \\
+    \bottomhline
+  \end{tabular}}
+  \belowtable{}
+\end{table*}
 
@@
 
-      Despite large differences in the input climate forcing
-      (Fig.~\ref{fig:lr_ts}, upper panel), model output presents consistent
-      features that can be observed across the range of forcing data used.
-      In all simulations, modelled ice volumes remain relatively low during
-      most of the glacial cycle, except during two major glacial events
-      which occur between 61.9 and 56.5\,ka during MIS~4, and between 23.2
-      and 16.8\,ka during MIS~2 (Fig.~\ref{fig:lr_ts}, lower panel). An ice
-      volume minimum is consistently reached between 53.0 and 41.3\,ka
-      during MIS~3. However, the magnitude and precise timing of these three
-      events depend significantly on the choice of proxy record used to
-      derive a time-dependent climate forcing (Table~\ref{tab:extrema}).
-
-      Simulations forced by the Greenland ice core palaeo-temperature
-      records (GRIP, NGRIP) produce the highest variability in modelled ice
-      volume throughout the last glacial cycle. In contrast, simulations
-      driven by oceanic (ODP~1012, ODP~1020) and Antarctic (EPICA, Vostok)
-      palaeo-temperature records generally result in lower ice volume
-      variability throughout the simulation length, resulting in lower
-      modelled ice volumes during MIS~4 and larger ice volumes during MIS~3.
-      The NGRIP climate forcing is the only one that results in a larger ice
-      volume during MIS~4 than during MIS~2.
-
-      While simulations driven by the GRIP and the two Antarctic
-      palaeo-temperature records attain a last ice volume maximum between
-      19.1 and 16.8\,ka, those informed by the NGRIP and the two oceanic
-      palaeo-temperature records attain their maximum ice volumes thousands
-      of years earlier. Moreover, the ODP~1012 run yields a rapid
-      deglaciation of the modelled area prior to 17\,ka. The ODP~1020
-      simulation predicts an early maximum in ice volume at 23.2\,ka,
-      followed by slower deglaciation than modelled using the other
-      palaeo-temperature records. Finally, whereas model runs forced by the
-      Antarctic palaeo-temperature records result in a rapid and
-      uninterrupted deglaciation after the LGM, the simulation driven by the
-      GRIP palaeo-temperature record also results in a rapid deglaciation
-      but in three steps, separated by two periods of ice sheet regrowth
-      (Fig.~\ref{fig:lr_ts}).
-
-% -----------------------------------------------------------------------------
-\subsection{Extreme configurations}
-% -----------------------------------------------------------------------------
-
-      Despite large differences in the timing of attained volume extrema
-      (Table~\ref{tab:extrema}), all model runs show relatively consistent
-      patterns of glaciation. During MIS~4, all simulations produce an
-      extensive ice sheet, covering an area of at least half of that
-      attained during MIS~2 (Table~\ref{tab:extrema};
-      Fig.~\ref{fig:lr_maps}, upper panels). Corresponding maximum ice
-      volumes also differ significantly between model runs, and vary between
-      3.84 and~8.84\,m sea level equivalents (m.~s.l.e.;
-      Table~\ref{tab:extrema}).
-
-      In the MIS~3 ice volume minimum reconstructions, a central ice cap
-      persists over the Skeena Mountains (Fig.~\ref{fig:lr_maps}, middle
-      panels). Although this ice cap is present in all simulations, its
-      dimensions depend sensitively on the choice of the applied
-      palaeo-temperature record. Modelled ice volume minima spread over a
-      wide range between 1.69 and 2.88\,m~s.l.e. (Table~\ref{tab:extrema}).
-
-      Modelled ice sheet geometries during the LGM (MIS~2;
-      Fig.~\ref{fig:lr_maps}, lower panels) invariably include a ca.
-      1500\,km-long central divide above 3000\,m~a.s.l. located along the
-      spine of the Rocky Mountains. Although the similarity of modelled ice
-      extents is a direct result of the choice of scaling factors applied to
-      different palaeo-temperature proxy records (Table~\ref{tab:records}),
-      it is interesting to note that modelled maximum ice volumes also fall
-      within a tight range of 8.40 to 8.91\,m~s.l.e.
-      (Table~\ref{tab:extrema}).
-
-% =============================================================================
-\section{Comparison with the geologic record}
-\label{sec:discussion}
-% =============================================================================
-
-      Large variations in the model responses to evolving climate forcing
-      reveal its sensitivity to the choice of palaeo-temperature proxy
-      record. To distinguish between different records, geological evidence
-      of former glaciations provide a basis for validation of our runs,
-      while the results from numerical modelling can perhaps help to analyse
-      some of the complexity of this evidence. In this section, we compare
-      model outputs to the geologic record, in terms of timing and
-      configuration of the maximum stages, location and lifetime of major
-      nucleation centres, and patterns of ice retreat during the last
-      deglaciation.
-
-% -----------------------------------------------------------------------------
-\subsection{Glacial maxima}
-% -----------------------------------------------------------------------------
-
-\subsubsection{Timing of glaciation}
-\label{sec:timing}
-
-      Independently of the palaeo-temperature records used to force the ice
-      sheet model, our simulations consistently produce two glacial maxima
-      during the last glacial cycle. The first maximum configuration is
-      obtained during MIS~4 (61.9--56.5\,ka) and the second during MIS~2
-      (23.2--16.8\,ka; Figs.~\ref{fig:lr_ts}, \ref{fig:lr_maps};
-      Table~\ref{tab:extrema}). These events broadly correspond in timing to
-      the Gladstone (MIS~4) and McConnell (MIS~2) glaciations documented by
-      geological evidence for the northern sector of the Cordilleran ice
-      sheet \citep{Duk-Rodkin.etal.1996, Ward.etal.2007, Stroeven.etal.2010,
-      Stroeven.etal.2014}, and to the Fraser Glaciation (MIS~2) documented
-      for its southern sector \citep{Porter.Swanson.1998,
-      Margold.etal.2014}. There is stratigraphical evidence for an MIS~4
-      glaciation in British Columbia \citep{Clague.Ward.2011} and in the
-      Puget Lowland \citep{Troost.2014}, but their extent and timing are
-      still highly conjectural \citep[perhaps MIS~4 or early MIS~3;
-      e.g.,][]{Cosma.etal.2008}.
-
-      The exact timing of modelled MIS~2 maximum ice volume depends strongly
-      on the choice of applied palaeo-temperature record, which allows for a
-      more in-depth comparison with geological evidence for the timing of
-      maximum Cordilleran ice sheet extent. In the Puget Lowland
-      (Fig.~\ref{fig:locmap}), the LGM advance of the southern Cordilleran
-      ice sheet margin has been constrained by radiocarbon dating on wood
-      between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,ka}
-      \citep{Porter.Swanson.1998}. These dates are consistent with
-      radiocarbon dates from the offshore sedimentary record, which reveals
-      an increase of glaciomarine sedimentation between 19.5 and
-      16.2\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008,
-      Taylor.etal.2014}. Radiocarbon dating of the northern Cordilleran ice
-      sheet margin is much less constrained but straddles presented
-      constraints from the southern margin. However, cosmogenic exposure
-      dating places the timing of maximum CIS extent during the McConnell
-      glaciation close to 17\,\unit{\chem{^{10}Be}\,ka}
-      \citep{Stroeven.etal.2010, Stroeven.etal.2014}. A sharp transition in
-      the sediment record of the Gulf of Alaska indicates a retreat of
-      regional outlet glaciers onto land at
-      14.8\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Davies.etal.2011}.
-
-      Among the simulations presented here, only those forced with the GRIP,
-      EPICA and Vostok palaeo-temperature records yield Cordilleran ice
-      sheet maximum extents that may be compatible with these field
-      constraints (Fig.~\ref{fig:lr_ts}, lower panel;
-      Table~\ref{tab:extrema}). Simulations driven by the NGRIP, ODP~1012
-      and ODP~1020 palaeo-temperature records, on the contrary, yield MIS~2
-      maximum Cordilleran ice sheet volumes that pre-date field-based
-      constraints by several thousands of years (about 6, 6 and 4\,ka
-      respectively). Concerning the simulations driven by oceanic records,
-      this early deglaciation is caused by an early warming present in the
-      alkenone palaeo-temperature reconstructions (Fig.~\ref{fig:lr_ts},
-      upper panel; \citealp[Fig.~3]{Herbert.etal.2001}). However, this early
-      warming is a local effect, corresponding to a weakening of the
-      California current \citep{Herbert.etal.2001}. The California current,
-      driving cold waters southwards along the south-western coast of North
-      America, has been shown to have weakened during each peak of global
-      glaciation (in SPECMAP) during the past 550\,ka, including the LGM,
-      resulting in paradoxically warmer sea-surface temperatures at the
-      locations of the ODP~1012 and ODP~1020 sites
-      \citep{Herbert.etal.2001}.
-
-      Because most of the marine margin of the Cordilleran ice sheet
-      terminated in a sector of the Pacific Ocean unaffected by variations
-      in the California current, it probably remained insensitive to this
-      local phenomenum. However, the above paradox illustrates the
-      complexity of ice-sheet feedbacks on regional climate, and
-      demonstrates that, although located in the neighbourhood of the
-      modelling domain, the ODP~1012 and ODP~1020 palaeo-temperature records
-      cannot be used as a realistic forcing to model the Cordilleran ice
-      sheet through the last glacial cycle. Similarly, the simulation using
-      the NGRIP palaeo-temperature record depicts an early onset of
-      deglaciation (Fig.~\ref{fig:lr_ts}) following its last glacial volume
-      maximum (22.9\,ka, Table~\ref{tab:extrema}) attained about 6\,ka
-      earlier than dated evidence of the LGM advance. There is a fair
-      agreement between the EPICA and Vostok palaeo-temperature records,
-      resulting in only small differences between the simulations driven by
-      those records. These differences are not subject to further analysis
-      here; instead we focus on simulations forced by palaeo-temperature
-      records from the GRIP and EPICA ice cores that appear to produce the
-      most realistic reconstructions of regional glaciation history, yet
-      bearing significant disparities in model output. To allow for a more
-      detailed comparison against the geological record, these two
-      simulations were re-run using a higher-resolution grid
-      (Sect.~\ref{sec:model}; Fig.~\ref{fig:lr_ts}, lower panel, dotted
-      lines).
-
-\subsubsection{Ice configuration during MIS~2}
-\label{sec:mis2}
+\begin{figure}
+  \includegraphics{locmap}
+  \caption{Relief map of northern North America showing a reconstruction of the
+           areas once covered by the Cordilleran (CIS), Laurentide (LIS),
+           Innuitian (IIS), and Greenland (GIS) ice sheets during the last
+           18\,\unit{\chem{^{14}C}\,ka} \citep[21.4\,cal\,ka,][]{Dyke.2004}.
+           The rectangular box denotes the location of the
+           modelling domain used in this study. Major mountain ranges covered
+           by the ice sheet include the Alaska Range (AR), the Wrangell and
+           Saint Elias mountains (WSEM), the Selwyn and Mackenzie mountains
+           (SMKM), the Skeena Mountains (SM), the Coast Mountains (CM), the
+           Columbia and Rocky
+           Mountains (CRM), and the North Cascades (NC). Major depressions
+           include the Liard Lowland (LL), the Interior Plateau of British
+           Columbia (IP), and the Puget Lowland (PL). The background
+           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
+           Data \citep{Patterson.Kelso.2015}.}
+  \label{fig:locmap}
+\end{figure}
+
+\begin{figure}
+  \includegraphics{atm}
+  \caption{Monthly mean near-surface air temperature, precipitation, and
+           standard deviation of daily mean temperature (PDD SD) for January
+           and July from the North American Regional Reanalysis
+           \citep[NARR;][]{Mesinger.etal.2006}, used to force the surface mass
+           balance (PDD) component of the ice sheet model. Note the
+           strong contrasts in seasonality, timing of the precipitation peak,
+           and temperature variability over the model domain, notably between
+           coastal and inland regions.}
+  \label{fig:atm}
+\end{figure}
+
+\begin{figure*}
+  \includegraphics{lr_ts}
+  \caption{Temperature offset time-series from ice core and ocean records
+           (Table~\ref{tab:records}) used as palaeo-climate forcing for the ice
+           sheet model (top panel), and modelled ice volume (bottom panel)
+           through the last 120\,ka. Ice volumes are expressed in meters of sea
+           level equivalent (m~s.l.e.). Gray fields indicate Marine Oxygen
+           Isotope Stage (MIS) boundaries for MIS~2 and MIS~4 according to a
+           global compilation of benthic \chem{\delta^{18}O} records
+           \citep{Lisiecki.Raymo.2005}. Hatched rectangles highlight the
+           time-volume span for ice volume extremes corresponding to MIS~4
+           (61.9--56.5\,ka), MIS~3 (53.0--41.3\,ka), and MIS~2 (LGM,
+           23.2--16.8\,ka). Dotted lines correspond to GRIP- and EPICA-driven
+           5\,km-resolution runs.}
+  \label{fig:lr_ts}
+\end{figure*}
+
+\begin{figure*}
+  \includegraphics{lr_maps}
+  \caption{Snapshots of modelled surface topography (500\,m contours)
+           corresponding to the ice volume extremes indicated on
+           Fig.~\ref{fig:lr_ts}. An ice cap persists over the Skeena
+           Mountains (SM) during MIS~3. Note the occurence of spatial
+           similarities despite large differences in timing.}
+  \label{fig:lr_maps}
+\end{figure*}
 
@@
 \end{figure*}
-
-      During maximum glaciation, both simulations position the main
-      meridional ice divide over the western flank of the Rocky Mountains
-      (Fig.~\ref{fig:lr_maps}, lower panels; Fig.~\ref{fig:mis2}). This
-      result appears to contrast with palaeoglaciological reconstructions
-      for central and southern British Columbia with ice divides in a more
-      westerly position, over the western margin of the Interior Plateau
-      \citep{Ryder.etal.1991, Stumpf.etal.2000, Kleman.etal.2010,
-      Clague.Ward.2011, Margold.etal.2013a}. These indicate that a
-      latitudinal saddle connected ice dispersal centres in the Columbia
-      Mountains with the main ice divide \citep{Ryder.etal.1991,
-      Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}. A latitudinal
-      saddle does indeed feature in our modelling results, however, in an
-      inverse configuration between the main ice divide over the Columbia
-      Mountains and a secondary divide over the southern Coast Mountains
-      (Fig.~\ref{fig:mis2}).
-
-      Such deviation from the geological inferences could reflect the fact
-      that our model does not include feedback mechanisms between ice sheet
-      topography and the regional climate. Firstly, during the build-up
-      phase preceding the LGM, rapid accumulation over the Coast Mountains
-      enhanced the topographic barrier formed by these mountain ranges,
-      which likely resulted in a decrease of precipitation and, therefore, a
-      decrease of accumulation in the interior. Secondly, latent warming of
-      the moisture-depleted air parcels flowing over this enhanced
-      topography could have resulted in an inflow of potentially warmer air
-      over the eastern flank of the ice sheet, increasing melt along the
-      advancing margin \citep[cf.][]{Langen.etal.2012}. Because these two
-      processes, both with a tendency to limit ice-sheet growth, are absent
-      from our model, the eastern margin of the ice sheet and the position
-      of the main meridional ice divide are certainly biased towards the
-      east in our simulations \citep{Seguinot.etal.2014}.
-
-      However, field-based palaeoglaciological reconstructions have
-      struggled to reconcile the more westerly-centred ice divide in
-      south-central British Columbia with evidence in the Rocky Mountains
-      and beyond, that the Cordilleran ice sheet invaded the western
-      Interior Plains, where it merged with the southwestern margin of the
-      Laurentide ice sheet and was deflected to the south
-      \citep{Jackson.etal.1997, Bednarski.Smith.2007, Kleman.etal.2010,
-      Margold.etal.2013, Margold.etal.2013a}. Ice geometries from our model
-      runs do not have this problem, because the position and elevation of
-      the ice divide ensure significant ice drainage across the Rocky
-      Mountains at the LGM (Fig.~\ref{fig:mis2}).
-
-      During MIS~2, the modelled total ice volume peaks at 8.01\,m~s.l.e.
-      (19.1\,ka) in the GRIP simulation and at 8.77\,m~s.l.e. (17.2\,ka) in
-      the EPICA simulation.
-
-\subsubsection{Ice configuration during MIS~4}
-\label{sec:mis4}
-
-      The modelled ice sheet configurations corresponding to ice volume
-      maxima during MIS~4 are more sensitive to the choice of atmospheric
-      forcing than those corresponding to ice volume maxima during MIS~2
-      (Fig.~\ref{fig:lr_maps}, upper panels; Fig.~\ref{fig:mis4}). The GRIP
-      simulation (Fig.~\ref{fig:mis4}, left panel) results in a modelled
-      maximum ice sheet extent that closely resembles that obtained during
-      MIS~2, with the only major difference of being slightly less extensive
-      across northern and eastern sectors. In contrast, the EPICA simulation
-      produces a lower ice volume maximum (Fig.~\ref{fig:lr_ts}), which
-      translates in the modelled ice sheet geometry into a significantly
-      reduced southern sector, more restricted ice cover in northern and
-      eastern sectors, and generally lower ice surface elevations in the
-      interior (Fig.~\ref{fig:mis4}, right panel). Thus, only the GRIP
-      simulation can explain the presence of MIS~4 glacial deposits in the
-      Puget Lowland \citep{Troost.2014} and that of ice-rafted debris in the
-      marine sediment record offshore Vancouver Island at
-      ca.~47\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008}.
-
-      During MIS~4, the modelled total ice volume peaks at 7.19\,m~s.l.e.
-      (57.3\,ka) in the GRIP simulation and at 5.04\,m~s.l.e. (61.9\,ka) in
-      the EPICA simulation, corresponding to respectively 90\,\% and 57\,\%
-      of modelled MIS~2 ice volumes.
-
-% -----------------------------------------------------------------------------
-\subsection{Nucleation centres}
-% -----------------------------------------------------------------------------
-
-\subsubsection{Transient ice sheet states}
 
@@
 
-      Palaeo-glaciological reconstructions are generally more robust for
-      maximum ice sheet extents and late ice sheet configurations than for
-      intermediate or minimum ice sheet extents and older ice sheet
-      configurations \citep{Kleman.etal.2010}. However, these maximum stages
-      are, by nature, extreme configurations, which do not necessarily
-      represent the dominant patterns of glaciation throughout the period of
-      ice cover \citep{Porter.1989, Kleman.Stroeven.1997, Kleman.etal.2008,
-      Kleman.etal.2010}.
-
-      For the Cordilleran ice sheet, geological evidence from radiocarbon
-      dating \citep{Clague.etal.1980, Clague.1985, Clague.1986,
-      Porter.Swanson.1998, Menounos.etal.2008}, cosmogenic exposure dating
-      \citep{Stroeven.etal.2010, Stroeven.etal.2014, Margold.etal.2014},
-      bedrock deformation in response to former ice loads
-      \citep{Clague.James.2002, Clague.etal.2005}, and offshore sedimentary
-      records \citep{Cosma.etal.2008, Davies.etal.2011} indicate that the
-      LGM maximum extent was short-lived. To compare this finding to our
-      simulations, we use numerical modelling output to compute durations of
-      ice cover throughout the last glacial cycle
-      (Fig.~\ref{fig:hr_geom_duration}).
-
-      The resulting maps show that, during most of the glacial cycle,
-      modelled ice cover is restricted to disjoint ice caps centred on major
-      mountain ranges of the North American Cordillera
-      (Fig.~\ref{fig:hr_geom_duration}, blue areas). A 2500\,km-long
-      continuous expanse of ice, extending from the Alaska Range in the
-      northeast to the Rocky Mountains in the southwest, is only in
-      operation for at most 32\,ka, which is about a third of the timespan
-      of the last glacial cycle (Fig.~\ref{fig:hr_geom_duration}, hatched
-      areas). However, except for its margins in the Pacific Ocean and in
-      the northern foothills of the Alaska Range, the maximum extent of the
-      ice sheet is attained for a much shorter period of time of only few
-      thousand years (Figs.~\ref{fig:hr_geom_duration}, red areas). This
-      result illustrates that the maximum extents of the modelled ice sheet
-      during MIS~4 and MIS~2 were both short-lived and therefore out of
-      balance with contemporary climate.
-
-      A notable exception to the transient character of the maximum extent
-      of Cordilleran ice sheet is the northern slope of the Alaska Range,
-      where modelled glaciers are confined to its foothills during the
-      entire simulation period (Fig.~\ref{fig:hr_geom_duration}, AR). This
-      apparent insensitivity of modelled glacial extent to temperature
-      fluctuations results from a combination of low precipitation, high
-      summer temperature and large temperature standard deviation (PDD SD)
-      in the plains of the Alaska Interior (Fig.~\ref{fig:atm}) which
-      confines glaciation to the foothills of the mountains. This result
-      could potentially explain the local distribution of glacial deposits,
-      which indicates that glaciers flowing on the northern slope of the
-      Alaska Range have remained small throughout the Pleistocene
-      \citep{Kaufman.Manley.2004}.
-
-\subsubsection{Major ice-dispersal centres}
-
-      It is generally believed that the Cordilleran ice sheet formed by the
-      coalescence of several mountain-centred ice caps
-      \citep{Davis.Mathews.1944}. In our simulations, major ice-dispersal
-      centres, visible on the modelled ice cover duration maps
-      (Fig.~\ref{fig:hr_geom_duration}), are located over the Coast
-      Mountains (CM), the Columbia and Rocky mountains (CRM), the Skeena
-      Mountains (SM), and the Selwyn and Mackenzie mountains (SMKM). The
-      Wrangell and Saint Elias mountains, heavily glacierized at present,
-      host an ice cap for the entire length of both simulations, but that
-      ice cap does not appear to be a major feed to the Cordilleran ice
-      sheet (Fig.~\ref{fig:hr_geom_duration}, WSEM). Although the Coast,
-      Skeena and Columbia and Rocky mountains (CM, SM, CRM) are covered by
-      mountain glaciers for most of the last glacial cycle, providing
-      durable nucleation centres for an ice sheet, this is not the case for
-      the Selwyn and Mackenzie mountains (SMKM), where ice cover on the
-      highest peaks is limited to a small fraction of the last glacial
-      cycle. In other words, the Selwyn and Mackenzie mountains only appear
-      as a secondary ice-dispersal centre during the coldest periods of the
-      last glacial cycle. The Northern Rocky Mountains
-      (Fig.~\ref{fig:hr_geom_duration}, NRM) do not act as a nucleation
-      centre, but rather as a pinning point for the Cordilleran ice sheet
-      margin coming from the west.
-
-      Perhaps the most striking feature displayed by the distributions of
-      modelled ice cover is the persistence of the Skeena Mountains ice cap
-      throughout the entire last glacial period (ca.~100--10\,ka) and its
-      predominance over the other ice-dispersal centres
-      (Figs.~\ref{fig:lr_maps} and~\ref{fig:hr_geom_duration}, SM).
-      Regardless of the applied forcing, this ice cap appears to survive
-      MIS~3 (Fig.~\ref{fig:lr_maps}, middle panels), and serves as a
-      nucleation centre at the onset of the glacial readvance towards the
-      LGM (MIS~2). This situation appear similar to the neighbouring
-      Laurentide ice sheet, for which the importance of residual ice for the
-      glacial history leading up to the LGM has been illustrated by the
-      MIS~3 residual ice bodies in northern and eastern Canada as nucleation
-      centres for a much more extensive MIS~2 configuration
-      \citep{Kleman.etal.2010}.
-
-      The presence of a Skeena Mountains ice cap during most of the last
-      glacial cycle can be explained by meteorological conditions more
-      favourable for ice growth there than elsewhere. In fact, reanalysed
-      atmospheric fields used to force the surface mass balance model show
-      that high winter precipitations are mainly confined to the western
-      slope of the Coast Mountains, except in the centre of the modelling
-      domain where they also occur further inland than along other east-west
-      transects (Fig.~\ref{fig:atm}). In fact, along most of the
-      north-western coast of North America, coastal mountain ranges form a
-      pronounced topographic barrier for westerly winds, capturing
-      atmospheric moisture in the form of orographic precipitation, and
-      resulting in arid interior lowlands. However, near the centre of our
-      modelling domain, this barrier is less pronounced than elsewhere,
-      allowing westerly winds to carry moisture further inland, until it is
-      captured by the extensive Skeena Mountains in north-central British
-      Columbia, thus resulting in a more widespread distribution of winter
-      precipitation (Fig.~\ref{fig:atm}).
-
-      The modelled total ice volume corresponding to these persistent
-      ice-dispersal centres attains a minimum of 1.52\,m~s.l.e. (42.9\,ka)
-      in the GRIP simulation and of 2.68\,m~s.l.e. (51.8\,ka) in the EPICA
-      simulation, corresponding to respectively 19\,\% and 31\,\% of the
-      MIS~2 ice volumes.
-
-\subsubsection{Erosional imprint on the landscape}
-
 \begin{figure*}
@@
 \end{figure*}
-
-      A correlation is observed between the modelled duration of warm based
-      ice cover (Fig.~\ref{fig:hr_geom_warmbase}) and the degree of glacial
-      modification of the landscape (mainly in terms of the development of
-      deep glacial valleys and troughs). We find evidence for this on the
-      slopes of the Coast Mountains, the Columbia and Rocky mountains, the
-      Wrangell and Saint Elias mountains, and radiating off the Skeena
-      Mountains (Figs.~\ref{fig:hr_geom_duration} and
-      \ref{fig:hr_geom_warmbase}; \citealp[Fig.~2]{Kleman.etal.2010}). The
-      Skeena Mountains, for example, indeed bear a strong glacial imprint
-      that indicates ice drainage in a system of distinct glacial troughs
-      emanating in a radial pattern from the centre of the mountain range
-      \citep[Fig.~2]{Kleman.etal.2010}. We suggest that persistent ice cover
-      (Fig.~\ref{fig:hr_geom_duration}) associated with basal ice
-      temperatures at the pressure melting point
-      (Figs.~\ref{fig:hr_geom_warmbase} and~\ref{fig:hr_geom_warmfrac})
-      explains the large-scale glacial erosional imprint on the landscape. A
-      well-developed network of glacial valleys running to the north-west on
-      the west slope of the Selwyn and Mackenzie Mountains
-      (\citealp[Fig.~2]{Kleman.etal.2010}; \citealp[Fig.~8]
-      {Stroeven.etal.2010}) is modelled to have hosted predominantly
-      warm-based ice (Fig.~\ref{fig:hr_geom_warmfrac}). However, because it
-      is only glaciated for a short fraction of the last glacial cycle in
-      our simulations (Fig.~\ref{fig:hr_geom_duration}), this perhaps
-      indicates that the observed landscape pattern originates from multiple
-      glacial cycles and witnesses an increased relative importance of the
-      Selwyn and Mackenzie mountains ice dispersal centre
-      (Fig.~\ref{fig:hr_geom_duration}, SMKM), prior to the Late Pleistocene
-      \citep[cf.][]{Ward.etal.2008, Demuro.etal.2012}.
-
-      The modelled distribution of warm-based ice cover
-      (Figs.~\ref{fig:hr_geom_warmbase} and~\ref{fig:hr_geom_warmfrac}) is
-      inevitably affected by our assumption of a constant,
-      70\,\unit{mW\,m^{-2}} geothermal heat flux at 3\,km depth
-      (Sect.~\ref{sec:icedyn}). However, the Skeena Mountains and the area
-      west of the Mackenzie Mountains experience higher-than-average
-      geothermal heat flux with measured values of ca.~80 and
-      ca.~100\,\unit{mW\,m^{-2}} \citep{Blackwell.Richards.2004}. We can
-      therefore expect even longer durations of warm-based ice cover for
-      these areas if we were to include spatially variable geothermal
-      forcing in our Cordilleran ice sheet simulations.
-
-% -----------------------------------------------------------------------------
-\subsection{The last deglaciation}
-% -----------------------------------------------------------------------------
-
-\subsubsection{Pace and patterns of deglaciation}
-
-      Similarly to other glaciated regions, most glacial traces in the North
-      American Cordillera relate to the last few millennia of glaciation,
-      because most of the older evidence has been overprinted by warm-based
-      ice retreat during the last deglaciation \citep{Kleman.1994,
-      Kleman.etal.2010}. From a numerical modelling perspective, phases of
-      glacier retreat are more challenging than phases of growth, because
-      they involve more rapid fluctuations of the ice margin, increased flow
-      velocities and longitudinal stress gradients, and poorly understood
-      hydrological processes. The latter are typically included in the
-      models through simple parametrisations
-      \citep[e.g.][]{Clason.etal.2012, Clason.etal.2014, Bueler.Pelt.2015},
-      if included at all. However, next after the mapping of maximum ice
-      sheet extents during MIS~2 and MIS~4 (Sect.~\ref{sec:mis2}
-      and~\ref{sec:mis4}), geomorphologically-based reconstructions of
-      patterns of ice sheet retreat during the last deglaciation provide the
-      second best source of evidence for the validation of our simulations.
 
@@
 
-      In the North American Cordillera, the presence of lateral meltwater
-      channels at high elevation \citep{Margold.etal.2011,
-      Margold.etal.2013a, Margold.etal.2014}, and abundant esker systems at
-      low elevation \citep{Burke.etal.2012, Burke.etal.2012a,
-      Perkins.etal.2013, Margold.etal.2013} indicate that meltwater was
-      produced over large portions of the ice sheet surface during
-      deglaciation. The southern and northern margins of the Cordilleran ice
-      sheet reached their last glacial maximum extent around 17\,ka
-      \citep[Sect.~\ref{sec:timing};][]{Porter.Swanson.1998,
-      Cosma.etal.2008, Stroeven.etal.2010, Stroeven.etal.2014}, which we
-      take as a limiting age for the onset of ice retreat. The timing of
-      final deglaciation is less well constrained, but recent cosmogenic
-      dates from north-central British Columbia indicate that a seizable ice
-      cap emanating from the central Coast Mountains or the Skeena Mountains
-      persisted into the Younger Dryas chronozone, at least until 12.4\,ka
-      \citep{Margold.etal.2014}.
-
-      In our simulations, the timing of peak ice volume during the LGM and
-      the pacing of deglaciation depend critically on the choice of climate
-      forcing (Table~\ref{tab:extrema}; Figs.~\ref{fig:lr_ts}
-      and~\ref{fig:lr_ts_deglac}). Adopting the EPICA climate forcing yields
-      peak ice volume at 17.2\,ka and an uninterrupted deglaciation until
-      about 9\,ka (Fig.~\ref{fig:lr_ts_deglac}, lower panel, red curves). On
-      the contrary, the simulation driven by the GRIP palaeo-temperature
-      record yields peak ice volume at 19.3\,ka and a deglaciation
-      interrupted by two phases of regrowth until about 8\,ka. The first
-      interruption occurs between 16.6 and 14.5\,ka, and the second between
-      13.1 and 11.6\,ka (Fig.~\ref{fig:lr_ts_deglac}, lower panel, blue
-      curve).
-
-      Hence, the two model runs, while similar in overall timing compared to
-      the runs with other climate drivers, differ in detail. On the one
-      hand, the EPICA run depicts peak glaciation about 2\,ka later than the
-      GRIP run, in closer agreement with dated maximum extents, and shows a
-      faster, uninterrupted deglaciation which yields sporadic ice cover
-      more than 1\,ka earlier. On the other hand, the GRIP run yields a
-      deglaciation in three steps, compatible with marine sediment sequences
-      offshore Vancouver Island, where the distribution of ice-rafted debris
-      indicates an ice margin retreat from the Georgia Strait in two phases
-      that are contemporary with warming oceanic temperatures from 17.2 to
-      16.5 and from 15.5 to 14.0\,\unit{\chem{^{14}C}\,cal\,ka}
-      \citep{Taylor.etal.2014}.
-
-      Modelled patterns of ice sheet retreat are relatively consistent
-      between the two simulations (Figs.~\ref{fig:hr_geom_deglacageshots}
-      and.~\ref{fig:hr_geom_deglacage}). The southern sector of the
-      modelling domain, including the Puget Lowland, the Coast Mountains,
-      the Columbia and and Rocky mountains, and the Interior Plateau of
-      British Columbia, becomes completely deglaciated by 10\,ka, whereas a
-      significant ice cover remains over the Skeena, the Selwyn and
-      Mackenzie, and the Wrangell and Saint Elias mountains in the northern
-      sector of the modelling domain. After 10\,ka, deglaciation continues
-      to proceed across the Liard Lowland with a radial ice margin retreat
-      towards the surrounding mountain ranges, consistent with the regional
-      melt water record of the last deglaciation \citep{Margold.etal.2013}.
-      Remaining ice continues to decay by retreating towards the Selwyn and
-      MacKenzie, and Skeena mountains. The last remnants of the Cordilleran
-      ice sheet finally disappear from the Skeena Mountains at 6.5\,ka
-      (GRIP) and 6.1\,ka (EPICA).
-
-\subsubsection{Late-glacial readvance}
-
-      The possibility of late glacial readvances in the North American
-      Cordillera has been debated for some time \citep{Osborn.Gerloff.1997},
-      and locally these have been reconstructed and dated. Radiocarbon-dated
-      end moraines in the Fraser and Squamish valleys, off the southern tip
-      of the Coast Mountains, indicate consecutive glacier maxima, or
-      standstills while in overall retreat, one of which corresponds to the
-      Younger Dryas chronozone \citep{Clague.etal.1997, Friele.Clague.2002,
-      Friele.Clague.2002a, Kovanen.2002, Kovanen.Easterbrook.2002}. Although
-      most of these moraines characterise independent valley glaciers, that
-      may have been disconnected from the waning Cordilleran ice sheet, the
-      Finlay River area in the Omenica Mountains
-      (Fig.~\ref{fig:hr_geom_deglacage}, OM) presents a different kind of
-      evidence. There, sharp-crested moraines indicate a late-glacial
-      readvance of local alpine glaciers and, more importantly, their
-      interaction with larger, lingering remnants of the main body of the
-      Cordilleran ice sheet in the valleys \citep{Lakeman.etal.2008}.
-      Additional evidence for late-glacial alpine glacier readvances
-      includes moraines in the eastern Coast Mountains, and the Columbia and
-      Rocky mountains \citep{Osborn.Gerloff.1997, Menounos.etal.2008}.
-
-      Although further work is needed to constrain the timing of the
-      late-glacial readvance, to assess its extents and geographical
-      distribution, and to identify the potential climatic triggers
-      \citep{Menounos.etal.2008}, it is interesting to note that the
-      simulation driven by the GRIP record produces a late-glacial readvance
-      in the Coast Mountains and in the Columbia and Rocky Mountains
-      (Fig.~\ref{fig:hr_geom_deglacage}, left panel). In addition to
-      matching the location of some local readvances, the GRIP-driven
-      simulation shows that a large remnant of the decaying ice sheet may
-      still have existed at the time of this late-glacial readvance. In
-      contrast, the EPICA-driven simulation produces a nearly-continuous
-      deglaciation with only a tightly restricted late-glacial readvance on
-      the western slopes of the Saint Elias and the Coast mountains
-      (Fig.~\ref{fig:hr_geom_deglacage}, right panel).
-
-\subsubsection{Deglacial flow directions}
-
 \begin{figure*}
@@
 
-      Because a general tenant in glacial geomorphology is that the majority
-      of landforms (lineations and eskers) are part of the deglacial
-      envelope \citep[terminology from][]{Kleman.etal.2006}, having been
-      formed close inside the retreating margin of ice sheets
-      \citep{Boulton.Clark.1990, Kleman.etal.1997, Kleman.etal.2010}, we
-      present maps of basal flow directions immediately preceding
-      deglaciation or at the time of cessation of sliding inside a
-      cold-based retreating margin (Fig.~\ref{fig:hr_geom_lastflow}). The
-      modelled deglacial flow patterns are mostly consistent between the
-      GRIP and EPICA simulations. They depict an active ice sheet retreat in
-      the peripheral areas, followed by stagnant ice decay in some of the
-      interior regions. Several parts of the modelling domain do not
-      experience any basal sliding throughout the deglaciation phase
-      (Fig.~\ref{fig:hr_geom_lastflow}, hatched areas). This notably
-      includes parts of the Interior Plateau in British Columbia, major
-      portions of the Alaskan sector of the ice sheet, and a tortuous ribbon
-      running from the Northern Rocky Mountains over the Skeena and Selwyn
-      Mountains and into the Mackenzie Mountains.
-
-      Patterns of glacial lineations formed in the northern and southern
-      sectors of the Cordilleran ice sheet (\citealp{Prest.etal.1968};
-      \citealp[Fig.~1.12]{Clague.1989}; \citealp[Fig.~2]{Kleman.etal.2010})
-      show similarities with the patterns of deglacial ice flow from
-      numerical modelling (Fig.~\ref{fig:hr_geom_lastflow}). In the northern
-      half of the modelling domain, modelled deglacial flow directions
-      depict an active downhill flow as the last remnants of the ice sheet
-      retreat towards mountain ranges. Converging deglacial flow patterns in
-      the Liard Lowland, for instance (Fig.~\ref{fig:hr_geom_lastflow}),
-      closely resemble the pattern indicated by glacial lineations
-      \citep[Fig.~2]{Margold.etal.2013}.
-
-      On the Interior Plateau of south-central British Columbia, both
-      simulations produce a retreat of the ice margin towards the north-east
-      (Fig.~\ref{fig:hr_geom_deglacage}), a pattern which is validated by
-      the geomorphological and stratigraphical record for ancient
-      pro-glacial lakes dammed by the retreating ice sheet
-      \citep{Perkins.Brennand.2014}. However, the two simulations differ in
-      the mode of retreat. The GRIP simulation yields an active retreat with
-      basal sliding towards the ice margin to the south, whereas the EPICA
-      simulation produces negligible basal sliding on the plateau during
-      deglaciation (Fig.~\ref{fig:hr_geom_lastflow}). Yet, the Interior
-      Plateau also hosts an impressive set of glacial lineations which
-      indicate a substantial eastwards flow component of the Cordilleran ice
-      sheet \citep{Prest.etal.1968, Kleman.etal.2010}. This Interior Plateau
-      lineation set could therefore present a smoking gun for the
-      reliability of the presented model results. One explanation for the
-      incongruent results could be that the missing feedback mechanisms
-      between ice sheet topography and regional climate resulted in a
-      modelled ice divide of the LGM ice sheet being too far to the east
-      (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
-      \citealp{Seguinot.etal.2014}). A more westerly-located LGM ice divide
-      would certainly result in a different deglacial flow pattern over the
-      Interior Plateau. However, a more westerly-positioned LGM ice divide
-      would certainly be associated with a thinner ice sheet than that
-      modelled here. Decreased ice thickness would not promote warm-based
-      conditions but, on the contrary, enlarge the region of negligible
-      basal sliding (Fig.~\ref{fig:hr_geom_lastflow}). Thus, a second
-      explanation for the incongruent results could be that the Interior
-      Plateau lineation system predates deglaciation ice flow, as perhaps
-      indicated by some eskers that appear incompatible with these glacial
-      lineations \citep[Fig.~9]{Margold.etal.2013a}. Finally, a third
-      explanation could be that local geothermal heat associated with
-      volcanic activity on the Interior Plateau could have triggered the
-      basal sliding \citep[cf. Greenland ice
-      sheet;][]{Fahnestock.etal.2001}.
-
-      The modelled deglaciation of the Interior Plateau of British Columbia
-      consists of a rapid northwards retreat
-      (Fig.~\ref{fig:hr_geom_deglacage}) of southwards-flowing non-sliding
-      ice lobes (Fig.~\ref{fig:hr_geom_lastflow}) positioned in-between
-      deglaciated mountain ranges (Figs.~\ref{fig:hr_pf_grip}
-      and~\ref{fig:hr_pf_epica}). This result appears compatible with the
-      prevailing conceptual model of deglaciation of central British
-      Columbia, in which mountain ranges emerge from the ice before the
-      plateau \citep[Fig.~7]{Fulton.1991}. However, due to different
-      topographic and climatic conditions, our simulations produce different
-      deglaciation patterns in the northern half of the model domain,
-      indicating that this conceptual model may not be applied to the entire
-      area formerly covered by the Cordilleran ice sheet.
-
-% =============================================================================
-\conclusions
-\label{sec:concl}
-% =============================================================================
-
-      Numerical simulations of the Cordilleran ice sheet through the last
-      glacial cycle presented in this study consistently produce two glacial
-      maxima during MIS~4 (61.9--56.5\,ka, 3.8--8.8\,m~s.l.e.) and MIS~2
-      (23.2--16.8\,ka, 8.40--8.91\,m~s.l.e.), two periods corresponding to
-      documented extensive glaciations. This result is independent of the
-      palaeo-temperature record used among the six selected for this study,
-      and thus can be regarded as a robust model output, which broadly
-      matches geological evidence. However, the timing of the two glaciation
-      peaks depends sensitively on which climate record is used to drive the
-      model. The timing of the LGM is best reproduced by the EPICA and
-      Vostok Antarctic ice core records. It occurs about 2\,ka too early in
-      the simulation forced by the GRIP ice core record, and occurs even
-      earlier in all other simulations. The mismatch is largest for the two
-      Northwest Pacific ODP palaeo-temperature records, which are affected
-      by the weakening of the California current during the LGM.
-
-      In all simulations presented here, ice cover is limited to disjoint
-      mountain ice caps during most of the glacial cycle. The most
-      persistent nucleation centres are located in the Coast Mountains, the
-      Columbia and Rocky mountains, the Selwyn and Mackenzie mountains, and
-      most importantly, in the Skeena Mountains. Throughout the modelled
-      last glacial cycle, the Skeena Mountains host an ice cap which appears
-      to be fed by the moisture intruding inland from the west through a
-      topographic breach in the Coast Mountains. The Skeena ice cap acts as
-      the main nucleation centre for the glacial reavance towards the LGM
-      configuration. As indicated by persistent, warm-based ice in the
-      model, this ice cap perhaps explains the distinct glacial erosional
-      imprint observed on the landscape of the Skeena Mountains.
-
-      During deglaciation, none of the climate records used can be selected
-      as producing an optimal agreement between the model results and the
-      geological evidence. Although the EPICA-driven simulation yields the
-      most realistic timing of the LGM and, therefore, start of
-      deglaciation, only the GRIP-driven simulation produces late glacial
-      readvances in areas where these have been documented. Nonetheless, the
-      patterns of ice sheet retreat are consistent between the two
-      simulations, and show a rapid deglaciation of the southern sector of
-      the ice sheet, including a rapid northwards retreat across the
-      Interior Plateau of central British Columbia. The GRIP-driven
-      simulation then produces a late-glacial readvance of local ice caps
-      and of the main body of the decaying Cordilleran ice sheet primarily
-      in the Coast and the Columbia and Rocky Mountains. In both
-      simulations, this is followed by an opening of the Liard Lowland, and
-      a final retreat of the remaining ice caps towards the Selwyn and,
-      finally, the Skeena mountains, which hosts the last remnant of the ice
-      sheet during the middle Holocene (6.6--6.2\,ka). Our results identify
-      the Skeena Mountains as a key area to understanding glacial dynamics
-      of the Cordilleran ice sheet, highlighting the need for further
-      geological investigation of this region.
-
-% Author contributions
-\section*{Author contributions}
-      J.~Seguinot ran the simulations; I.~Rogozhina guided experiment
-      design; A.~P.~Stroeven, M.~Margold and J.~Kleman took part in the
-      interpretation and comparison of model results against geological
-      evidence. All authors contributed to the text.
-
-% Acknowledgements
-\begin{acknowledgements}
-  Foremost, we would like to thank Shawn Marshall for providing a detailed,
-  constructive analysis of this study during J.~Seguinot's Ph.D. defence
-  (September 2014). His comments were used to improve the model set-up. We
-  are very thankful to Constantine Khroulev, Ed Bueler, and Andy Aschwanden
-  for providing constant help and development with PISM. This work was
-  supported by the Swedish Research Council~(VR) grant no.~2008-3449 to
-  A.~P.~Stroeven and by the German Academic Exchange Service~(DAAD) grant
-  no.~50015537 and a Knut and Alice Wallenberg Foundation grant to
-  J.~Seguinot. Computer resources were provided by the Swedish National
-  Infrastructure for Computing (SNIC) allocation no.~2013/1-159
-  and~2014/1-159 to A.~P.~Stroeven at the National Supercomputing Center
-  (NSC).
-\end{acknowledgements}
-
-% References
-\bibliographystyle{copernicus}
-\bibliography{refs/references}
-\newpage
-
 % =============================================================================

### 2016-01-16 cordillera-cycle.tex
Typeset: Fix lon/lat formatting in record table.

@@
     \middlehline
-    GRIP     &  72{\degree} 35' N  % 72.58 (decimal)
-             &  37{\degree} 38' W  % 37.64 (decimal)
+    GRIP     &  72{\degree}35$^{\prime}$\,N   % 72.58 (decimal)
+             &  37{\degree}38$^{\prime}$\,W   % 37.64 (decimal)
              & 3238
@@
 
-    NGRIP    &  75{\degree} 06' N  % 75.10
-             &  42{\degree} 19' W  % 42.32
+    NGRIP    &  75{\degree}06$^{\prime}$\,N   % 75.10
+             &  42{\degree}19$^{\prime}$\,W   % 42.32
              & 2917
@@
 
-    EPICA    &  75{\degree} 06' S  % 75.1
-             & 123{\degree} 21' E  % 123.35
+    EPICA    &  75{\degree}06$^{\prime}$\,S   % 75.1
+             & 123{\degree}21$^{\prime}$\,E   % 123.35
              & 3233
@@
 
-    Vostok   &  78{\degree} 28' S  % 78.8
-             & 106{\degree} 50' E  % 106.8
+    Vostok   &  78{\degree}28$^{\prime}$\,S   % 78.8
+             & 106{\degree}50$^{\prime}$\,E   % 106.8
              & 3488
@@
 
-    ODP~1012 &  32{\degree} 17' N
-             & 118{\degree} 23' W
+    ODP~1012 &  32{\degree}17$^{\prime}$\,N
+             & 118{\degree}23$^{\prime}$\,W
              & -1772
@@
 
-    ODP~1020 &  41{\degree} 00' N
-             & 126{\degree} 26' W
+    ODP~1020 &  41{\degree}00$^{\prime}$\,N
+             & 126{\degree}26$^{\prime}$\,W
              & -3038

### 2016-01-16 cordillera-cycle.tex
Typeset: no-break space after 'A' and 'a'.

@@
       American Cordillera have been more extensive than today. At the Last
-      Glacial Maximum (LGM), a continuous blanket of ice, the Cordilleran
+      Glacial Maximum (LGM), a~continuous blanket of ice, the Cordilleran
       ice sheet \citep{Dawson.1888}, stretched from the Alaska Range in the
@@
 
-      More than a century of exploration and geological investigation of the
+      More than a~century of exploration and geological investigation of the
       Cordilleran mountains have led to many observations in support of the
@@
       In contrast, evidence for the deglaciation history of the Cordilleran
-      ice sheet since the LGM is considerable, albeit mostly at a regional
+      ice sheet since the LGM is considerable, albeit mostly at a~regional
       scale. Geomorphological evidence from south-central British Columbia
-      indicates a rapid deglaciation, including an early emergence of
+      indicates a~rapid deglaciation, including an early emergence of
       elevated areas while thin, stagnant ice still covered the surrounding
@@
       Stea.etal.2011}. Nevertheless, recent oxygen isotope measurements from
-      Gulf of Alaska sediments reveal a climatic evolution highly correlated
-      to that of Greenland during this period, including a distinct Late
+      Gulf of Alaska sediments reveal a~climatic evolution highly correlated
+      to that of Greenland during this period, including a~distinct Late
       Glacial cold reversal between 14.1 and 11.7\,ka
@@
       Kleman.etal.1997, Kleman.etal.2010, Stroeven.etal.inreview}. Here, we
-      use a numerical ice sheet model \citep{PISM-authors.2015}, calibrated
-      against field-based evidence, to perform a quantitative reconstruction
+      use a~numerical ice sheet model \citep{PISM-authors.2015}, calibrated
+      against field-based evidence, to perform a~quantitative reconstruction
       of the Cordilleran ice sheet evolution through the last glacial cycle,
@@
       Our palaeo-climate forcing therefore includes spatial temperature and
-      precipitation grids derived from a present-day atmospheric reanalysis
+      precipitation grids derived from a~present-day atmospheric reanalysis
       \citep{Mesinger.etal.2006} that includes the steep precipitation
@@
       bathymetry dataset with a~resolution of 1\,arc-min
-      \citep{Amante.Eakins.2009}. Sea level is lowered as a function of time
+      \citep{Amante.Eakins.2009}. Sea level is lowered as a~function of time
       based on the Spectral Mapping Project
       \citep[SPECMAP,][]{Imbrie.etal.1989} time scale. Geothermal heat flux
-      is applied as a constant value of 70\,\unit{mW\,m^{-2}} at 3\,km depth
+      is applied as a~constant value of 70\,\unit{mW\,m^{-2}} at 3\,km depth
       (Sect.~\ref{sec:icedyn}). Surface mass balance is computed using a
       positive degree-day (PDD) model (Sect.~\ref{sec:surface}). Climate
-      forcing is provided by a monthly climatology averaged from 1979 to
+      forcing is provided by a~monthly climatology averaged from 1979 to
       2000 from the North American Regional Reanalysis
@@
       Cordilleran ice sheet at the LGM (Fig.~\ref{fig:locmap}). The
-      simulations were run on two distinct grids, using a lower horizontal
-      resolution of 10\,km, and a higher horizontal resolution of 5\,km.
+      simulations were run on two distinct grids, using a~lower horizontal
+      resolution of 10\,km, and a~higher horizontal resolution of 5\,km.
       These computations were performed on 16 to 128 computing cores at the
@@
 
-      Ice sheet dynamics are typically modelled using a combination of
+      Ice sheet dynamics are typically modelled using a~combination of
       internal deformation and basal sliding. PISM is a~shallow ice sheet
@@
       based on their predominant components. The Shallow Shelf Approximation
-      (SSA) is used as a ``sliding law'' for the Shallow Ice Approximation
+      (SSA) is used as a~``sliding law'' for the Shallow Ice Approximation
       (SIA) by adding velocity solutions of the two approximations
@@
       Temperature is computed subglacially to a~depth of 3\,km, where it is
-      conditioned by a lower boundary geothermal heat flux of
+      conditioned by a~lower boundary geothermal heat flux of
       70\,\unit{mW\,m^{-2}}. Although this uniform value does not account
@@
       the vertical grid consists of 31~temperature layers in the bedrock and
-      up to 51~enthalpy layers in the ice sheet, corresponding to a vertical
+      up to 51~enthalpy layers in the ice sheet, corresponding to a~vertical
       resolution of 100\,m. The high-resolution simulations 61~bedrock
-      layers and up to 101~ice layers with a vertical resolution of 50\,m.
+      layers and up to 101~ice layers with a~vertical resolution of 50\,m.
 
@@
       where cohesion, $c_0$, is assumed to be zero. The friction angle,
-      $\phi$, varies from 15 to 45\degree as a piecewise-linear function of
+      $\phi$, varies from 15 to 45\degree as a~piecewise-linear function of
       modern bed elevation, with the lowest value occuring below modern sea
@@
       the ice overburden stress, $\rho gh$, and the modelled amount of
-      subglacial water, using a formula derived from laboratory experiments
+      subglacial water, using a~formula derived from laboratory experiments
       with till extracted from the base of Ice Stream B in West Antarctica
@@
 \end{equation}
-      where $\delta$ is chosen as 0.02, $e_0$ is a measured reference void
-      ratio and $C_c$ is a measured compressibility coefficient
+      where $\delta$ is chosen as 0.02, $e_0$ is a~measured reference void
+      ratio and $C_c$ is a~measured compressibility coefficient
       (Table~\ref{tab:params}). The amount of water at the base, $W$, varies
-      from zero to $W_{max}=2$\,m, a threshold above which additional melt
+      from zero to $W_{max}=2$\,m, a~threshold above which additional melt
       water is assumed to drain off instantaneously. Finally, the bedrock
-      topography responds to ice load following a bedrock deformation model
+      topography responds to ice load following a~bedrock deformation model
       that includes local isostasy, elastic lithosphere flexure and viscous
       astenosphere deformation in an infinite half-space
-      \citep{Lingle.Clark.1985,Bueler.etal.2007}. A relatively low viscosity
+      \citep{Lingle.Clark.1985,Bueler.etal.2007}. A~relatively low viscosity
       value of $\nu_m = 1\times10^{19}$\,\unit{Pa\,s} is used for the
@@
 
-      Ice shelf calving is computed using a double criterion. First, a
+      Ice shelf calving is computed using a~double criterion. First, a
       physically-realistic calving flux is computed based on eigenvalues of
@@
       open ocean. Second, floating ice thinner than 50\,m is systematically
-      calved off. A subgrid scheme by \citet{Albrecht.etal.2011} allows for
-      a continuous migration of the calving front. This formulation of
+      calved off. A~subgrid scheme by \citet{Albrecht.etal.2011} allows for
+      a~continuous migration of the calving front. This formulation of
       calving has been applied to the Antarctic ice sheet and has shown to
-      produce a realistic calving front positon for many of the present-day
+      produce a~realistic calving front positon for many of the present-day
       ice shelves \citep{Martin.etal.2011}.
@@
       The PDD computation accounts for stochastic temperature variations by
-      assuming a normal temperature distribution of standard deviation
+      assuming a~normal temperature distribution of standard deviation
       $\sigma$ aroung the expected value $T_m$. It is expressed by an
@@
       motivated by the need for an accurate, high-resolution precipitation
-      forcing, as identified in a previous sensitivity study
+      forcing, as identified in a~previous sensitivity study
       \citep{Seguinot.etal.2014}.
@@
       and NGRIP records were calculated from oxygen isotope
-      (\chem{\delta^{18}O}) measurements using a quadratic equation
+      (\chem{\delta^{18}O}) measurements using a~quadratic equation
       \citep{Johnsen.etal.1995},
@@
       function of ice surface elevation, $s$, using the NARR surface
-      geopotential height invariant field as a reference topography,
+      geopotential height invariant field as a~reference topography,
       $b_{ref}$:
@@
       events depend significantly on the choice of proxy record used to
-      derive a time-dependent climate forcing (Table~\ref{tab:extrema}).
+      derive a~time-dependent climate forcing (Table~\ref{tab:extrema}).
 
@@
       modelled ice volumes during MIS~4 and larger ice volumes during MIS~3.
-      The NGRIP climate forcing is the only one that results in a larger ice
+      The NGRIP climate forcing is the only one that results in a~larger ice
       volume during MIS~4 than during MIS~2.
@@
       While simulations driven by the GRIP and the two Antarctic
-      palaeo-temperature records attain a last ice volume maximum between
+      palaeo-temperature records attain a~last ice volume maximum between
       19.1 and 16.8\,ka, those informed by the NGRIP and the two oceanic
       palaeo-temperature records attain their maximum ice volumes thousands
-      of years earlier. Moreover, the ODP~1012 run yields a rapid
+      of years earlier. Moreover, the ODP~1012 run yields a~rapid
       deglaciation of the modelled area prior to 17\,ka. The ODP~1020
@@
       palaeo-temperature records. Finally, whereas model runs forced by the
-      Antarctic palaeo-temperature records result in a rapid and
+      Antarctic palaeo-temperature records result in a~rapid and
       uninterrupted deglaciation after the LGM, the simulation driven by the
-      GRIP palaeo-temperature record also results in a rapid deglaciation
+      GRIP palaeo-temperature record also results in a~rapid deglaciation
       but in three steps, separated by two periods of ice sheet regrowth
@@
 
-      In the MIS~3 ice volume minimum reconstructions, a central ice cap
+      In the MIS~3 ice volume minimum reconstructions, a~central ice cap
       persists over the Skeena Mountains (Fig.~\ref{fig:lr_maps}, middle
@@
       Modelled ice sheet geometries during the LGM (MIS~2;
-      Fig.~\ref{fig:lr_maps}, lower panels) invariably include a ca.
+      Fig.~\ref{fig:lr_maps}, lower panels) invariably include a~ca.
       1500\,km-long central divide above 3000\,m~a.s.l. located along the
       spine of the Rocky Mountains. Although the similarity of modelled ice
-      extents is a direct result of the choice of scaling factors applied to
+      extents is a~direct result of the choice of scaling factors applied to
       different palaeo-temperature proxy records (Table~\ref{tab:records}),
       it is interesting to note that modelled maximum ice volumes also fall
-      within a tight range of 8.40 to 8.91\,m~s.l.e.
+      within a~tight range of 8.40 to 8.91\,m~s.l.e.
       (Table~\ref{tab:extrema}).
@@
       record. To distinguish between different records, geological evidence
-      of former glaciations provide a basis for validation of our runs,
+      of former glaciations provide a~basis for validation of our runs,
       while the results from numerical modelling can perhaps help to analyse
@@
       glaciation close to 17\,\unit{\chem{^{10}Be}\,ka}
-      \citep{Stroeven.etal.2010, Stroeven.etal.2014}. A sharp transition in
-      the sediment record of the Gulf of Alaska indicates a retreat of
+      \citep{Stroeven.etal.2010, Stroeven.etal.2014}. A~sharp transition in
+      the sediment record of the Gulf of Alaska indicates a~retreat of
       regional outlet glaciers onto land at
@@
       upper panel; \citealp[Fig.~3]{Herbert.etal.2001}). However, this early
-      warming is a local effect, corresponding to a weakening of the
+      warming is a~local effect, corresponding to a~weakening of the
       California current \citep{Herbert.etal.2001}. The California current,
@@
       Because most of the marine margin of the Cordilleran ice sheet
-      terminated in a sector of the Pacific Ocean unaffected by variations
+      terminated in a~sector of the Pacific Ocean unaffected by variations
       in the California current, it probably remained insensitive to this
@@
       modelling domain, the ODP~1012 and ODP~1020 palaeo-temperature records
-      cannot be used as a realistic forcing to model the Cordilleran ice
+      cannot be used as a~realistic forcing to model the Cordilleran ice
       sheet through the last glacial cycle. Similarly, the simulation using
@@
       maximum (22.9\,ka, Table~\ref{tab:extrema}) attained about 6\,ka
-      earlier than dated evidence of the LGM advance. There is a fair
+      earlier than dated evidence of the LGM advance. There is a~fair
       agreement between the EPICA and Vostok palaeo-temperature records,
@@
       most realistic reconstructions of regional glaciation history, yet
-      bearing significant disparities in model output. To allow for a more
+      bearing significant disparities in model output. To allow for a~more
       detailed comparison against the geological record, these two
-      simulations were re-run using a higher-resolution grid
+      simulations were re-run using a~higher-resolution grid
       (Sect.~\ref{sec:model}; Fig.~\ref{fig:lr_ts}, lower panel, dotted
@@
       result appears to contrast with palaeoglaciological reconstructions
-      for central and southern British Columbia with ice divides in a more
+      for central and southern British Columbia with ice divides in a~more
       westerly position, over the western margin of the Interior Plateau
@@
       Mountains with the main ice divide \citep{Ryder.etal.1991,
-      Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}. A latitudinal
+      Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}. A~latitudinal
       saddle does indeed feature in our modelling results, however, in an
       inverse configuration between the main ice divide over the Columbia
-      Mountains and a secondary divide over the southern Coast Mountains
+      Mountains and a~secondary divide over the southern Coast Mountains
       (Fig.~\ref{fig:mis2}).
@@
       enhanced the topographic barrier formed by these mountain ranges,
-      which likely resulted in a decrease of precipitation and, therefore, a
+      which likely resulted in a~decrease of precipitation and, therefore, a
       decrease of accumulation in the interior. Secondly, latent warming of
@@
       advancing margin \citep[cf.][]{Langen.etal.2012}. Because these two
-      processes, both with a tendency to limit ice-sheet growth, are absent
+      processes, both with a~tendency to limit ice-sheet growth, are absent
       from our model, the eastern margin of the ice sheet and the position
@@
       (Fig.~\ref{fig:lr_maps}, upper panels; Fig.~\ref{fig:mis4}). The GRIP
-      simulation (Fig.~\ref{fig:mis4}, left panel) results in a modelled
+      simulation (Fig.~\ref{fig:mis4}, left panel) results in a~modelled
       maximum ice sheet extent that closely resembles that obtained during
@@
       across northern and eastern sectors. In contrast, the EPICA simulation
-      produces a lower ice volume maximum (Fig.~\ref{fig:lr_ts}), which
-      translates in the modelled ice sheet geometry into a significantly
+      produces a~lower ice volume maximum (Fig.~\ref{fig:lr_ts}), which
+      translates in the modelled ice sheet geometry into a~significantly
       reduced southern sector, more restricted ice cover in northern and
@@
       mountain ranges of the North American Cordillera
-      (Fig.~\ref{fig:hr_geom_duration}, blue areas). A 2500\,km-long
+      (Fig.~\ref{fig:hr_geom_duration}, blue areas). A~2500\,km-long
       continuous expanse of ice, extending from the Alaska Range in the
       northeast to the Rocky Mountains in the southwest, is only in
-      operation for at most 32\,ka, which is about a third of the timespan
+      operation for at most 32\,ka, which is about a~third of the timespan
       of the last glacial cycle (Fig.~\ref{fig:hr_geom_duration}, hatched
@@
       the northern foothills of the Alaska Range, the maximum extent of the
-      ice sheet is attained for a much shorter period of time of only few
+      ice sheet is attained for a~much shorter period of time of only few
       thousand years (Figs.~\ref{fig:hr_geom_duration}, red areas). This
@@
 
-      A notable exception to the transient character of the maximum extent
+      A~notable exception to the transient character of the maximum extent
       of Cordilleran ice sheet is the northern slope of the Alaska Range,
@@
       apparent insensitivity of modelled glacial extent to temperature
-      fluctuations results from a combination of low precipitation, high
+      fluctuations results from a~combination of low precipitation, high
       summer temperature and large temperature standard deviation (PDD SD)
@@
       host an ice cap for the entire length of both simulations, but that
-      ice cap does not appear to be a major feed to the Cordilleran ice
+      ice cap does not appear to be a~major feed to the Cordilleran ice
       sheet (Fig.~\ref{fig:hr_geom_duration}, WSEM). Although the Coast,
@@
       the Selwyn and Mackenzie mountains (SMKM), where ice cover on the
-      highest peaks is limited to a small fraction of the last glacial
+      highest peaks is limited to a~small fraction of the last glacial
       cycle. In other words, the Selwyn and Mackenzie mountains only appear
-      as a secondary ice-dispersal centre during the coldest periods of the
+      as a~secondary ice-dispersal centre during the coldest periods of the
       last glacial cycle. The Northern Rocky Mountains
-      (Fig.~\ref{fig:hr_geom_duration}, NRM) do not act as a nucleation
-      centre, but rather as a pinning point for the Cordilleran ice sheet
+      (Fig.~\ref{fig:hr_geom_duration}, NRM) do not act as a~nucleation
+      centre, but rather as a~pinning point for the Cordilleran ice sheet
       margin coming from the west.
@@
       MIS~3 residual ice bodies in northern and eastern Canada as nucleation
-      centres for a much more extensive MIS~2 configuration
+      centres for a~much more extensive MIS~2 configuration
       \citep{Kleman.etal.2010}.
 
-      The presence of a Skeena Mountains ice cap during most of the last
+      The presence of a~Skeena Mountains ice cap during most of the last
       glacial cycle can be explained by meteorological conditions more
@@
       captured by the extensive Skeena Mountains in north-central British
-      Columbia, thus resulting in a more widespread distribution of winter
+      Columbia, thus resulting in a~more widespread distribution of winter
       precipitation (Fig.~\ref{fig:atm}).
@@
       The modelled total ice volume corresponding to these persistent
-      ice-dispersal centres attains a minimum of 1.52\,m~s.l.e. (42.9\,ka)
+      ice-dispersal centres attains a~minimum of 1.52\,m~s.l.e. (42.9\,ka)
       in the GRIP simulation and of 2.68\,m~s.l.e. (51.8\,ka) in the EPICA
@@
 
-      A correlation is observed between the modelled duration of warm based
+      A~correlation is observed between the modelled duration of warm based
       ice cover (Fig.~\ref{fig:hr_geom_warmbase}) and the degree of glacial
@@
       \ref{fig:hr_geom_warmbase}; \citealp[Fig.~2]{Kleman.etal.2010}). The
-      Skeena Mountains, for example, indeed bear a strong glacial imprint
-      that indicates ice drainage in a system of distinct glacial troughs
-      emanating in a radial pattern from the centre of the mountain range
+      Skeena Mountains, for example, indeed bear a~strong glacial imprint
+      that indicates ice drainage in a~system of distinct glacial troughs
+      emanating in a~radial pattern from the centre of the mountain range
       \citep[Fig.~2]{Kleman.etal.2010}. We suggest that persistent ice cover
@@
       warm-based ice (Fig.~\ref{fig:hr_geom_warmfrac}). However, because it
-      is only glaciated for a short fraction of the last glacial cycle in
+      is only glaciated for a~short fraction of the last glacial cycle in
       our simulations (Fig.~\ref{fig:hr_geom_duration}), this perhaps
@@
       (Figs.~\ref{fig:hr_geom_warmbase} and~\ref{fig:hr_geom_warmfrac}) is
-      inevitably affected by our assumption of a constant,
+      inevitably affected by our assumption of a~constant,
       70\,\unit{mW\,m^{-2}} geothermal heat flux at 3\,km depth
@@
       ice retreat during the last deglaciation \citep{Kleman.1994,
-      Kleman.etal.2010}. From a numerical modelling perspective, phases of
+      Kleman.etal.2010}. From a~numerical modelling perspective, phases of
       glacier retreat are more challenging than phases of growth, because
@@
       Cosma.etal.2008, Stroeven.etal.2010, Stroeven.etal.2014}, which we
-      take as a limiting age for the onset of ice retreat. The timing of
+      take as a~limiting age for the onset of ice retreat. The timing of
       final deglaciation is less well constrained, but recent cosmogenic
-      dates from north-central British Columbia indicate that a seizable ice
+      dates from north-central British Columbia indicate that a~seizable ice
       cap emanating from the central Coast Mountains or the Skeena Mountains
@@
       the contrary, the simulation driven by the GRIP palaeo-temperature
-      record yields peak ice volume at 19.3\,ka and a deglaciation
+      record yields peak ice volume at 19.3\,ka and a~deglaciation
       interrupted by two phases of regrowth until about 8\,ka. The first
@@
       sector of the modelling domain. After 10\,ka, deglaciation continues
-      to proceed across the Liard Lowland with a radial ice margin retreat
+      to proceed across the Liard Lowland with a~radial ice margin retreat
       towards the surrounding mountain ranges, consistent with the regional
@@
       Finlay River area in the Omenica Mountains
-      (Fig.~\ref{fig:hr_geom_deglacage}, OM) presents a different kind of
-      evidence. There, sharp-crested moraines indicate a late-glacial
+      (Fig.~\ref{fig:hr_geom_deglacage}, OM) presents a~different kind of
+      evidence. There, sharp-crested moraines indicate a~late-glacial
       readvance of local alpine glaciers and, more importantly, their
@@
       \citep{Menounos.etal.2008}, it is interesting to note that the
-      simulation driven by the GRIP record produces a late-glacial readvance
+      simulation driven by the GRIP record produces a~late-glacial readvance
       in the Coast Mountains and in the Columbia and Rocky Mountains
@@
       matching the location of some local readvances, the GRIP-driven
-      simulation shows that a large remnant of the decaying ice sheet may
+      simulation shows that a~large remnant of the decaying ice sheet may
       still have existed at the time of this late-glacial readvance. In
-      contrast, the EPICA-driven simulation produces a nearly-continuous
-      deglaciation with only a tightly restricted late-glacial readvance on
+      contrast, the EPICA-driven simulation produces a~nearly-continuous
+      deglaciation with only a~tightly restricted late-glacial readvance on
       the western slopes of the Saint Elias and the Coast mountains
@@
 
-      Because a general tenant in glacial geomorphology is that the majority
+      Because a~general tenant in glacial geomorphology is that the majority
       of landforms (lineations and eskers) are part of the deglacial
@@
       includes parts of the Interior Plateau in British Columbia, major
-      portions of the Alaskan sector of the ice sheet, and a tortuous ribbon
+      portions of the Alaskan sector of the ice sheet, and a~tortuous ribbon
       running from the Northern Rocky Mountains over the Skeena and Selwyn
@@
       On the Interior Plateau of south-central British Columbia, both
-      simulations produce a retreat of the ice margin towards the north-east
-      (Fig.~\ref{fig:hr_geom_deglacage}), a pattern which is validated by
+      simulations produce a~retreat of the ice margin towards the north-east
+      (Fig.~\ref{fig:hr_geom_deglacage}), a~pattern which is validated by
       the geomorphological and stratigraphical record for ancient
@@
       Plateau also hosts an impressive set of glacial lineations which
-      indicate a substantial eastwards flow component of the Cordilleran ice
+      indicate a~substantial eastwards flow component of the Cordilleran ice
       sheet \citep{Prest.etal.1968, Kleman.etal.2010}. This Interior Plateau
-      lineation set could therefore present a smoking gun for the
+      lineation set could therefore present a~smoking gun for the
       reliability of the presented model results. One explanation for the
@@
       (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
-      \citealp{Seguinot.etal.2014}). A more westerly-located LGM ice divide
-      would certainly result in a different deglacial flow pattern over the
-      Interior Plateau. However, a more westerly-positioned LGM ice divide
-      would certainly be associated with a thinner ice sheet than that
+      \citealp{Seguinot.etal.2014}). A~more westerly-located LGM ice divide
+      would certainly result in a~different deglacial flow pattern over the
+      Interior Plateau. However, a~more westerly-positioned LGM ice divide
+      would certainly be associated with a~thinner ice sheet than that
       modelled here. Decreased ice thickness would not promote warm-based
       conditions but, on the contrary, enlarge the region of negligible
-      basal sliding (Fig.~\ref{fig:hr_geom_lastflow}). Thus, a second
+      basal sliding (Fig.~\ref{fig:hr_geom_lastflow}). Thus, a~second
       explanation for the incongruent results could be that the Interior
@@
       indicated by some eskers that appear incompatible with these glacial
-      lineations \citep[Fig.~9]{Margold.etal.2013a}. Finally, a third
+      lineations \citep[Fig.~9]{Margold.etal.2013a}. Finally, a~third
       explanation could be that local geothermal heat associated with
@@
       The modelled deglaciation of the Interior Plateau of British Columbia
-      consists of a rapid northwards retreat
+      consists of a~rapid northwards retreat
       (Fig.~\ref{fig:hr_geom_deglacage}) of southwards-flowing non-sliding
@@
       palaeo-temperature record used among the six selected for this study,
-      and thus can be regarded as a robust model output, which broadly
+      and thus can be regarded as a~robust model output, which broadly
       matches geological evidence. However, the timing of the two glaciation
@@
       patterns of ice sheet retreat are consistent between the two
-      simulations, and show a rapid deglaciation of the southern sector of
-      the ice sheet, including a rapid northwards retreat across the
+      simulations, and show a~rapid deglaciation of the southern sector of
+      the ice sheet, including a~rapid northwards retreat across the
       Interior Plateau of central British Columbia. The GRIP-driven
-      simulation then produces a late-glacial readvance of local ice caps
+      simulation then produces a~late-glacial readvance of local ice caps
       and of the main body of the decaying Cordilleran ice sheet primarily
@@
       simulations, this is followed by an opening of the Liard Lowland, and
-      a final retreat of the remaining ice caps towards the Selwyn and,
+      a~final retreat of the remaining ice caps towards the Selwyn and,
       finally, the Skeena mountains, which hosts the last remnant of the ice
       sheet during the middle Holocene (6.6--6.2\,ka). Our results identify
-      the Skeena Mountains as a key area to understanding glacial dynamics
+      the Skeena Mountains as a~key area to understanding glacial dynamics
       of the Cordilleran ice sheet, highlighting the need for further
@@
 \begin{acknowledgements}
-  Foremost, we would like to thank Shawn Marshall for providing a detailed,
+  Foremost, we would like to thank Shawn Marshall for providing a~detailed,
   constructive analysis of this study during J.~Seguinot's Ph.D. defence
@@
   A.~P.~Stroeven and by the German Academic Exchange Service~(DAAD) grant
-  no.~50015537 and a Knut and Alice Wallenberg Foundation grant to
+  no.~50015537 and a~Knut and Alice Wallenberg Foundation grant to
   J.~Seguinot. Computer resources were provided by the Swedish National
@@
   \includegraphics{locmap}
-  \caption{Relief map of northern North America showing a reconstruction of the
+  \caption{Relief map of northern North America showing a~reconstruction of the
            areas once covered by the Cordilleran (CIS), Laurentide (LIS),
@@
            and EPICA climate forcing.
-           Note the irregular colour scale. A continuous ice cover spanning
+           Note the irregular colour scale. A~continuous ice cover spanning
            from the Alaska Range (AR) to the Coast Mountains (CM) and

### 2016-01-16 cordillera-cycle.tex
Typeset: add missing unit cmds and rewrap paras.

@@
 major glaciations during marine oxygen isotope stages~4
-(61.9--56.5\,ka) and~2 (23.2--16.8\,ka). The timing of
+(61.9--56.5\,\unit{ka}) and~2 (23.2--16.8\,\unit{ka}). The timing of
 glaciation is better reproduced using temperature reconstructions
@@
 Maximum and hosts the last remains of Cordilleran ice until the
-qmiddle Holocene (6.6--6.2\,ka).
+qmiddle Holocene (6.6--6.2\,\unit{ka}).
 \end{abstract}
@@
       to that of Greenland during this period, including a~distinct Late
-      Glacial cold reversal between 14.1 and 11.7\,ka
+      Glacial cold reversal between 14.1 and 11.7\,\unit{ka}
       \citep{Praetorius.Mix.2014}.
@@
       \citep[SPECMAP,][]{Imbrie.etal.1989} time scale. Geothermal heat flux
-      is applied as a~constant value of 70\,\unit{mW\,m^{-2}} at 3\,km depth
-      (Sect.~\ref{sec:icedyn}). Surface mass balance is computed using a
-      positive degree-day (PDD) model (Sect.~\ref{sec:surface}). Climate
-      forcing is provided by a~monthly climatology averaged from 1979 to
-      2000 from the North American Regional Reanalysis
-      \citep[NARR,][]{Mesinger.etal.2006}, perturbated by time-dependent
-      offsets and lapse-rate temperature corrections (Sect.~\ref{sec:atm}).
-
-      Each simulation starts from assumed ice-free conditions at 120000
-      years ago (120\,ka), and runs to the present. Our modelling domain of
-      1500 by 3000\,km encompasses the entire area covered by the
-      Cordilleran ice sheet at the LGM (Fig.~\ref{fig:locmap}). The
-      simulations were run on two distinct grids, using a~lower horizontal
-      resolution of 10\,km, and a~higher horizontal resolution of 5\,km.
-      These computations were performed on 16 to 128 computing cores at the
-      Swedish National Supercomputing Centre.
+      is applied as a~constant value of 70\,\unit{mW\,m^{-2}} at
+      3\,\unit{km} depth (Sect.~\ref{sec:icedyn}). Surface mass balance is
+      computed using a positive degree-day (PDD) model
+      (Sect.~\ref{sec:surface}). Climate forcing is provided by a~monthly
+      climatology averaged from 1979 to 2000 from the North American
+      Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006}, perturbated
+      by time-dependent offsets and lapse-rate temperature corrections
+      (Sect.~\ref{sec:atm}).
+
+      Each simulation starts from assumed ice-free conditions at
+      12\,0000\,\unit{years} ago (120\,\unit{ka}), and runs to the
+      present. Our modelling domain of 1500 by 3000\,\unit{km} encompasses
+      the entire area covered by the Cordilleran ice sheet at the LGM
+      (Fig.~\ref{fig:locmap}). The simulations were run on two distinct
+      grids, using a~lower horizontal resolution of 10\,\unit{km}, and
+      a~higher horizontal resolution of 5\,\unit{km}. These computations
+      were performed on 16 to 128 computing cores at the Swedish National
+      Supercomputing Centre.
 
@@
       temperature derived from the climate forcing (Sect.~\ref{sec:atm})
-      provides the upper boundary condition to the ice enthalpy model.
-      Temperature is computed subglacially to a~depth of 3\,km, where it is
-      conditioned by a~lower boundary geothermal heat flux of
-      70\,\unit{mW\,m^{-2}}. Although this uniform value does not account
-      for the high spatial geothermal variability in the region
+      provides the upper boundary condition to the ice enthalpy
+      model. Temperature is computed subglacially to a~depth of
+      3\,\unit{km}, where it is conditioned by a~lower boundary geothermal
+      heat flux of 70\,\unit{mW\,m^{-2}}. Although this uniform value does
+      not account for the high spatial geothermal variability in the region
       \citep{Blackwell.Richards.2004}, it is, on average, representative of
@@
       up to 51~enthalpy layers in the ice sheet, corresponding to a~vertical
-      resolution of 100\,m. The high-resolution simulations 61~bedrock
-      layers and up to 101~ice layers with a~vertical resolution of 50\,m.
+      resolution of 100\,\unit{m}. The high-resolution simulations
+      61~bedrock layers and up to 101~ice layers with a~vertical resolution
+      of 50\,\unit{m}.
 
@@
       modern bed elevation, with the lowest value occuring below modern sea
-      level (0\,m above sea level, m~a.s.l.) and the highest value occuring
-      above the generalised elevation of the highest shorelines
-      \citep[200\,m~a.s.l.,][Fig.~5]{Clague.1981}, thus accounting for
-      a~weakening of till associated with the presence of marine sediments
-      (cf. \citealp{Martin.etal.2011};
+      level (0\,\unit{m} above sea level, \unit{m\,a.s.l.}) and the highest
+      value occuring above the generalised elevation of the highest
+      shorelines \citep[200\,\unit{m\,a.s.l.},][Fig.~5]{Clague.1981}, thus
+      accounting for a~weakening of till associated with the presence of
+      marine sediments (cf. \citealp{Martin.etal.2011};
       \citealp[supplement]{Aschwanden.etal.2013};
@@
       (Table~\ref{tab:params}). The amount of water at the base, $W$, varies
-      from zero to $W_{max}=2$\,m, a~threshold above which additional melt
-      water is assumed to drain off instantaneously. Finally, the bedrock
-      topography responds to ice load following a~bedrock deformation model
-      that includes local isostasy, elastic lithosphere flexure and viscous
-      astenosphere deformation in an infinite half-space
+      from zero to $W_{max}=2$\,\unit{m}, a~threshold above which additional
+      melt water is assumed to drain off instantaneously. Finally, the
+      bedrock topography responds to ice load following a~bedrock
+      deformation model that includes local isostasy, elastic lithosphere
+      flexure and viscous astenosphere deformation in an infinite half-space
       \citep{Lingle.Clark.1985,Bueler.etal.2007}. A~relatively low viscosity
@@
       embayments, but prevents the formation of extensive ice shelves in the
-      open ocean. Second, floating ice thinner than 50\,m is systematically
-      calved off. A~subgrid scheme by \citet{Albrecht.etal.2011} allows for
-      a~continuous migration of the calving front. This formulation of
-      calving has been applied to the Antarctic ice sheet and has shown to
-      produce a~realistic calving front positon for many of the present-day
-      ice shelves \citep{Martin.etal.2011}.
+      open ocean. Second, floating ice thinner than 50\,\unit{m} is
+      systematically calved off. A~subgrid scheme by
+      \citet{Albrecht.etal.2011} allows for a~continuous migration of the
+      calving front. This formulation of calving has been applied to the
+      Antarctic ice sheet and has shown to produce a~realistic calving front
+      positon for many of the present-day ice shelves
+      \citep{Martin.etal.2011}.
 
@@
       most of the glacial cycle, except during two major glacial events
-      which occur between 61.9 and 56.5\,ka during MIS~4, and between 23.2
-      and 16.8\,ka during MIS~2 (Fig.~\ref{fig:lr_ts}, lower panel). An ice
-      volume minimum is consistently reached between 53.0 and 41.3\,ka
-      during MIS~3. However, the magnitude and precise timing of these three
-      events depend significantly on the choice of proxy record used to
-      derive a~time-dependent climate forcing (Table~\ref{tab:extrema}).
+      which occur between 61.9 and 56.5\,\unit{ka} during MIS~4, and between
+      23.2 and 16.8\,\unit{ka} during MIS~2 (Fig.~\ref{fig:lr_ts}, lower
+      panel). An ice volume minimum is consistently reached between 53.0 and
+      41.3\,\unit{ka} during MIS~3. However, the magnitude and precise
+      timing of these three events depend significantly on the choice of
+      proxy record used to derive a~time-dependent climate forcing
+      (Table~\ref{tab:extrema}).
 
@@
       palaeo-temperature records attain a~last ice volume maximum between
-      19.1 and 16.8\,ka, those informed by the NGRIP and the two oceanic
-      palaeo-temperature records attain their maximum ice volumes thousands
-      of years earlier. Moreover, the ODP~1012 run yields a~rapid
-      deglaciation of the modelled area prior to 17\,ka. The ODP~1020
-      simulation predicts an early maximum in ice volume at 23.2\,ka,
+      19.1 and 16.8\,\unit{ka}, those informed by the NGRIP and the two
+      oceanic palaeo-temperature records attain their maximum ice volumes
+      thousands of years earlier. Moreover, the ODP~1012 run yields a~rapid
+      deglaciation of the modelled area prior to 17\,\unit{ka}. The ODP~1020
+      simulation predicts an early maximum in ice volume at 23.2\,\unit{ka},
       followed by slower deglaciation than modelled using the other
@@
       palaeo-temperature record. Modelled ice volume minima spread over a
-      wide range between 1.69 and 2.88\,m~s.l.e. (Table~\ref{tab:extrema}).
+      wide range between 1.69 and 2.88\,\unit{m}\,s.l.e.
+      (Table~\ref{tab:extrema}).
 
@@
       Fig.~\ref{fig:lr_maps}, lower panels) invariably include a~ca.
-      1500\,km-long central divide above 3000\,m~a.s.l. located along the
-      spine of the Rocky Mountains. Although the similarity of modelled ice
-      extents is a~direct result of the choice of scaling factors applied to
-      different palaeo-temperature proxy records (Table~\ref{tab:records}),
-      it is interesting to note that modelled maximum ice volumes also fall
-      within a~tight range of 8.40 to 8.91\,m~s.l.e.
-      (Table~\ref{tab:extrema}).
+      1500\,\unit{km}-long central divide above 3000\,\unit{m\,a.s.l.}
+      located along the spine of the Rocky Mountains. Although the
+      similarity of modelled ice extents is a~direct result of the choice of
+      scaling factors applied to different palaeo-temperature proxy records
+      (Table~\ref{tab:records}), it is interesting to note that modelled
+      maximum ice volumes also fall within a~tight range of 8.40 to
+      8.91\,\unit{m}\,s.l.e. (Table~\ref{tab:extrema}).
 
@@
       during the last glacial cycle. The first maximum configuration is
-      obtained during MIS~4 (61.9--56.5\,ka) and the second during MIS~2
-      (23.2--16.8\,ka; Figs.~\ref{fig:lr_ts}, \ref{fig:lr_maps};
-      Table~\ref{tab:extrema}). These events broadly correspond in timing to
-      the Gladstone (MIS~4) and McConnell (MIS~2) glaciations documented by
-      geological evidence for the northern sector of the Cordilleran ice
-      sheet \citep{Duk-Rodkin.etal.1996, Ward.etal.2007, Stroeven.etal.2010,
-      Stroeven.etal.2014}, and to the Fraser Glaciation (MIS~2) documented
-      for its southern sector \citep{Porter.Swanson.1998,
-      Margold.etal.2014}. There is stratigraphical evidence for an MIS~4
-      glaciation in British Columbia \citep{Clague.Ward.2011} and in the
-      Puget Lowland \citep{Troost.2014}, but their extent and timing are
-      still highly conjectural \citep[perhaps MIS~4 or early MIS~3;
-      e.g.,][]{Cosma.etal.2008}.
+      obtained during MIS~4 (61.9--56.5\,\unit{ka}) and the second during
+      MIS~2 (23.2--16.8\,\unit{ka}; Figs.~\ref{fig:lr_ts},
+      \ref{fig:lr_maps}; Table~\ref{tab:extrema}). These events broadly
+      correspond in timing to the Gladstone (MIS~4) and McConnell (MIS~2)
+      glaciations documented by geological evidence for the northern sector
+      of the Cordilleran ice sheet \citep{Duk-Rodkin.etal.1996,
+      Ward.etal.2007, Stroeven.etal.2010, Stroeven.etal.2014}, and to the
+      Fraser Glaciation (MIS~2) documented for its southern sector
+      \citep{Porter.Swanson.1998, Margold.etal.2014}. There is
+      stratigraphical evidence for an MIS~4 glaciation in British Columbia
+      \citep{Clague.Ward.2011} and in the Puget Lowland \citep{Troost.2014},
+      but their extent and timing are still highly conjectural
+      \citep[perhaps MIS~4 or early MIS~3; e.g.,][]{Cosma.etal.2008}.
 
@@
       maximum Cordilleran ice sheet volumes that pre-date field-based
-      constraints by several thousands of years (about 6, 6 and 4\,ka
+      constraints by several thousands of years (about 6, 6 and 4\,\unit{ka}
       respectively). Concerning the simulations driven by oceanic records,
@@
       America, has been shown to have weakened during each peak of global
-      glaciation (in SPECMAP) during the past 550\,ka, including the LGM,
-      resulting in paradoxically warmer sea-surface temperatures at the
+      glaciation (in SPECMAP) during the past 550\,\unit{ka}, including the
+      LGM, resulting in paradoxically warmer sea-surface temperatures at the
       locations of the ODP~1012 and ODP~1020 sites
@@
       deglaciation (Fig.~\ref{fig:lr_ts}) following its last glacial volume
-      maximum (22.9\,ka, Table~\ref{tab:extrema}) attained about 6\,ka
-      earlier than dated evidence of the LGM advance. There is a~fair
-      agreement between the EPICA and Vostok palaeo-temperature records,
-      resulting in only small differences between the simulations driven by
-      those records. These differences are not subject to further analysis
-      here; instead we focus on simulations forced by palaeo-temperature
-      records from the GRIP and EPICA ice cores that appear to produce the
-      most realistic reconstructions of regional glaciation history, yet
-      bearing significant disparities in model output. To allow for a~more
-      detailed comparison against the geological record, these two
-      simulations were re-run using a~higher-resolution grid
-      (Sect.~\ref{sec:model}; Fig.~\ref{fig:lr_ts}, lower panel, dotted
+      maximum (22.9\,\unit{ka}, Table~\ref{tab:extrema}) attained about
+      6\,\unit{ka} earlier than dated evidence of the LGM advance. There is
+      a~fair agreement between the EPICA and Vostok palaeo-temperature
+      records, resulting in only small differences between the simulations
+      driven by those records. These differences are not subject to further
+      analysis here; instead we focus on simulations forced by
+      palaeo-temperature records from the GRIP and EPICA ice cores that
+      appear to produce the most realistic reconstructions of regional
+      glaciation history, yet bearing significant disparities in model
+      output. To allow for a~more detailed comparison against the geological
+      record, these two simulations were re-run using a~higher-resolution
+      grid (Sect.~\ref{sec:model}; Fig.~\ref{fig:lr_ts}, lower panel, dotted
       lines).
@@
 
-      During MIS~2, the modelled total ice volume peaks at 8.01\,m~s.l.e.
-      (19.1\,ka) in the GRIP simulation and at 8.77\,m~s.l.e. (17.2\,ka) in
-      the EPICA simulation.
+      During MIS~2, the modelled total ice volume peaks at
+      8.01\,\unit{m}\,s.l.e. (19.1\,\unit{ka}) in the GRIP simulation and at
+      8.77\,\unit{m}\,s.l.e. (17.2\,\unit{ka}) in the EPICA simulation.
 
@@
 
-      During MIS~4, the modelled total ice volume peaks at 7.19\,m~s.l.e.
-      (57.3\,ka) in the GRIP simulation and at 5.04\,m~s.l.e. (61.9\,ka) in
-      the EPICA simulation, corresponding to respectively 90\,\% and 57\,\%
-      of modelled MIS~2 ice volumes.
+      During MIS~4, the modelled total ice volume peaks at
+      7.19\,\unit{m}\,s.l.e. (57.3\,\unit{ka}) in the GRIP simulation and at
+      5.04\,\unit{m}\,s.l.e. (61.9\,\unit{ka}) in the EPICA simulation,
+      corresponding to respectively 90\,\% and 57\,\% of modelled MIS~2 ice
+      volumes.
 
@@
       mountain ranges of the North American Cordillera
-      (Fig.~\ref{fig:hr_geom_duration}, blue areas). A~2500\,km-long
+      (Fig.~\ref{fig:hr_geom_duration}, blue areas). A~2500\,\unit{km}-long
       continuous expanse of ice, extending from the Alaska Range in the
       northeast to the Rocky Mountains in the southwest, is only in
-      operation for at most 32\,ka, which is about a~third of the timespan
-      of the last glacial cycle (Fig.~\ref{fig:hr_geom_duration}, hatched
-      areas). However, except for its margins in the Pacific Ocean and in
-      the northern foothills of the Alaska Range, the maximum extent of the
-      ice sheet is attained for a~much shorter period of time of only few
-      thousand years (Figs.~\ref{fig:hr_geom_duration}, red areas). This
+      operation for at most 32\,\unit{ka}, which is about a~third of the
+      timespan of the last glacial cycle (Fig.~\ref{fig:hr_geom_duration},
+      hatched areas). However, except for its margins in the Pacific Ocean
+      and in the northern foothills of the Alaska Range, the maximum extent
+      of the ice sheet is attained for a~much shorter period of time of only
+      few thousand years (Figs.~\ref{fig:hr_geom_duration}, red areas). This
       result illustrates that the maximum extents of the modelled ice sheet
@@
       modelled ice cover is the persistence of the Skeena Mountains ice cap
-      throughout the entire last glacial period (ca.~100--10\,ka) and its
-      predominance over the other ice-dispersal centres
+      throughout the entire last glacial period (ca.~100--10\,\unit{ka}) and
+      its predominance over the other ice-dispersal centres
       (Figs.~\ref{fig:lr_maps} and~\ref{fig:hr_geom_duration}, SM).
@@
       The modelled total ice volume corresponding to these persistent
-      ice-dispersal centres attains a~minimum of 1.52\,m~s.l.e. (42.9\,ka)
-      in the GRIP simulation and of 2.68\,m~s.l.e. (51.8\,ka) in the EPICA
-      simulation, corresponding to respectively 19\,\% and 31\,\% of the
-      MIS~2 ice volumes.
+      ice-dispersal centres attains a~minimum of 1.52\,\unit{m}\,s.l.e.
+      (42.9\,\unit{ka}) in the GRIP simulation and of 2.68\,\unit{m}\,s.l.e.
+      (51.8\,\unit{ka}) in the EPICA simulation, corresponding to
+      respectively 19\,\% and 31\,\% of the MIS~2 ice volumes.
 
@@
       inevitably affected by our assumption of a~constant,
-      70\,\unit{mW\,m^{-2}} geothermal heat flux at 3\,km depth
+      70\,\unit{mW\,m^{-2}} geothermal heat flux at 3\,\unit{km} depth
       (Sect.~\ref{sec:icedyn}). However, the Skeena Mountains and the area
@@
       deglaciation. The southern and northern margins of the Cordilleran ice
-      sheet reached their last glacial maximum extent around 17\,ka
+      sheet reached their last glacial maximum extent around 17\,\unit{ka}
       \citep[Sect.~\ref{sec:timing};][]{Porter.Swanson.1998,
@@
       cap emanating from the central Coast Mountains or the Skeena Mountains
-      persisted into the Younger Dryas chronozone, at least until 12.4\,ka
-      \citep{Margold.etal.2014}.
+      persisted into the Younger Dryas chronozone, at least until
+      12.4\,\unit{ka} \citep{Margold.etal.2014}.
 
@@
       and~\ref{fig:lr_ts_deglac}). Adopting the EPICA climate forcing yields
-      peak ice volume at 17.2\,ka and an uninterrupted deglaciation until
-      about 9\,ka (Fig.~\ref{fig:lr_ts_deglac}, lower panel, red curves). On
-      the contrary, the simulation driven by the GRIP palaeo-temperature
-      record yields peak ice volume at 19.3\,ka and a~deglaciation
-      interrupted by two phases of regrowth until about 8\,ka. The first
-      interruption occurs between 16.6 and 14.5\,ka, and the second between
-      13.1 and 11.6\,ka (Fig.~\ref{fig:lr_ts_deglac}, lower panel, blue
-      curve).
+      peak ice volume at 17.2\,\unit{ka} and an uninterrupted deglaciation
+      until about 9\,\unit{ka} (Fig.~\ref{fig:lr_ts_deglac}, lower panel,
+      red curves). On the contrary, the simulation driven by the GRIP
+      palaeo-temperature record yields peak ice volume at 19.3\,\unit{ka}
+      and a~deglaciation interrupted by two phases of regrowth until about
+      8\,\unit{ka}. The first interruption occurs between 16.6 and
+      14.5\,\unit{ka}, and the second between 13.1 and 11.6\,\unit{ka}
+      (Fig.~\ref{fig:lr_ts_deglac}, lower panel, blue curve).
 
@@
       the runs with other climate drivers, differ in detail. On the one
-      hand, the EPICA run depicts peak glaciation about 2\,ka later than the
-      GRIP run, in closer agreement with dated maximum extents, and shows a
-      faster, uninterrupted deglaciation which yields sporadic ice cover
-      more than 1\,ka earlier. On the other hand, the GRIP run yields a
-      deglaciation in three steps, compatible with marine sediment sequences
-      offshore Vancouver Island, where the distribution of ice-rafted debris
-      indicates an ice margin retreat from the Georgia Strait in two phases
-      that are contemporary with warming oceanic temperatures from 17.2 to
-      16.5 and from 15.5 to 14.0\,\unit{\chem{^{14}C}\,cal\,ka}
-      \citep{Taylor.etal.2014}.
+      hand, the EPICA run depicts peak glaciation about 2\,\unit{ka} later
+      than the GRIP run, in closer agreement with dated maximum extents, and
+      shows a faster, uninterrupted deglaciation which yields sporadic ice
+      cover more than 1\,\unit{ka} earlier. On the other hand, the GRIP run
+      yields a deglaciation in three steps, compatible with marine sediment
+      sequences offshore Vancouver Island, where the distribution of
+      ice-rafted debris indicates an ice margin retreat from the Georgia
+      Strait in two phases that are contemporary with warming oceanic
+      temperatures from 17.2 to 16.5 and from 15.5 to
+      14.0\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Taylor.etal.2014}.
 
@@
       the Columbia and and Rocky mountains, and the Interior Plateau of
-      British Columbia, becomes completely deglaciated by 10\,ka, whereas a
-      significant ice cover remains over the Skeena, the Selwyn and
-      Mackenzie, and the Wrangell and Saint Elias mountains in the northern
-      sector of the modelling domain. After 10\,ka, deglaciation continues
-      to proceed across the Liard Lowland with a~radial ice margin retreat
-      towards the surrounding mountain ranges, consistent with the regional
-      melt water record of the last deglaciation \citep{Margold.etal.2013}.
-      Remaining ice continues to decay by retreating towards the Selwyn and
-      MacKenzie, and Skeena mountains. The last remnants of the Cordilleran
-      ice sheet finally disappear from the Skeena Mountains at 6.5\,ka
-      (GRIP) and 6.1\,ka (EPICA).
+      British Columbia, becomes completely deglaciated by 10\,\unit{ka},
+      whereas a significant ice cover remains over the Skeena, the Selwyn
+      and Mackenzie, and the Wrangell and Saint Elias mountains in the
+      northern sector of the modelling domain. After 10\,\unit{ka},
+      deglaciation continues to proceed across the Liard Lowland with
+      a~radial ice margin retreat towards the surrounding mountain ranges,
+      consistent with the regional melt water record of the last
+      deglaciation \citep{Margold.etal.2013}. Remaining ice continues to
+      decay by retreating towards the Selwyn and MacKenzie, and Skeena
+      mountains. The last remnants of the Cordilleran ice sheet finally
+      disappear from the Skeena Mountains at 6.5\,\unit{ka} (GRIP) and
+      6.1\,\unit{ka} (EPICA).
 
@@
       glacial cycle presented in this study consistently produce two glacial
-      maxima during MIS~4 (61.9--56.5\,ka, 3.8--8.8\,m~s.l.e.) and MIS~2
-      (23.2--16.8\,ka, 8.40--8.91\,m~s.l.e.), two periods corresponding to
-      documented extensive glaciations. This result is independent of the
+      maxima during MIS~4 (61.9--56.5\,\unit{ka},
+      3.8--8.8\,\unit{m}\,s.l.e.) and MIS~2 (23.2--16.8\,\unit{ka},
+      8.40--8.91\,\unit{m}\,s.l.e.), two periods corresponding to documented
+      extensive glaciations. This result is independent of the
       palaeo-temperature record used among the six selected for this study,
@@
       model. The timing of the LGM is best reproduced by the EPICA and
-      Vostok Antarctic ice core records. It occurs about 2\,ka too early in
-      the simulation forced by the GRIP ice core record, and occurs even
-      earlier in all other simulations. The mismatch is largest for the two
-      Northwest Pacific ODP palaeo-temperature records, which are affected
-      by the weakening of the California current during the LGM.
+      Vostok Antarctic ice core records. It occurs about 2\,\unit{ka} too
+      early in the simulation forced by the GRIP ice core record, and occurs
+      even earlier in all other simulations. The mismatch is largest for the
+      two Northwest Pacific ODP palaeo-temperature records, which are
+      affected by the weakening of the California current during the LGM.
 
@@
       finally, the Skeena mountains, which hosts the last remnant of the ice
-      sheet during the middle Holocene (6.6--6.2\,ka). Our results identify
-      the Skeena Mountains as a~key area to understanding glacial dynamics
-      of the Cordilleran ice sheet, highlighting the need for further
-      geological investigation of this region.
+      sheet during the middle Holocene (6.6--6.2\,\unit{ka}). Our results
+      identify the Skeena Mountains as a~key area to understanding glacial
+      dynamics of the Cordilleran ice sheet, highlighting the need for
+      further geological investigation of this region.
 
@@
            & Reference\\
-    & & & (m~a.s.l.) & & & (K) & \\
+    & & & (\unit{m\,a.s.l.}) & & & (K) & \\
     \middlehline
@@
              & \multicolumn{3}{c}{Ice extent (\unit{10^6\,km^2})}
-             & \multicolumn{3}{c}{Ice volume (m~s.l.e.)} \\
+             & \multicolumn{3}{c}{Ice volume (\unit{m}\,s.l.e.)} \\
     Record   &  MIS~4 &  MIS~3 &  MIS~2
@@
            sheet model (top panel), and modelled ice volume (bottom panel)
-           through the last 120\,ka. Ice volumes are expressed in meters of sea
-           level equivalent (m~s.l.e.). Gray fields indicate Marine Oxygen
+           through the last 120\,\unit{ka}. Ice volumes are expressed in meters of sea
+           level equivalent (\unit{m}\,s.l.e.). Gray fields indicate Marine Oxygen
            Isotope Stage (MIS) boundaries for MIS~2 and MIS~4 according to a
@@
            time-volume span for ice volume extremes corresponding to MIS~4
-           (61.9--56.5\,ka), MIS~3 (53.0--41.3\,ka), and MIS~2 (LGM,
-           23.2--16.8\,ka). Dotted lines correspond to GRIP- and EPICA-driven
-           5\,km-resolution runs.}
+           (61.9--56.5\,\unit{ka}), MIS~3 (53.0--41.3\,\unit{ka}), and MIS~2 (LGM,
+           23.2--16.8\,\unit{ka}). Dotted lines correspond to GRIP- and EPICA-driven
+           5\,\unit{km}-resolution runs.}
   \label{fig:lr_ts}
@@
   \includegraphics{lr_maps}
-  \caption{Snapshots of modelled surface topography (500\,m contours)
+  \caption{Snapshots of modelled surface topography (500\,\unit{m} contours)
            corresponding to the ice volume extremes indicated on
@@
   \includegraphics{hr_maps_mis2}
-  \caption{Modelled surface topography (200\,m contours) and surface velocity
+  \caption{Modelled surface topography (200\,\unit{m} contours) and surface velocity
            (colour mapping) corresponding to the maximum ice volume during
@@
   \includegraphics{hr_maps_mis4}
-  \caption{Modelled surface topography (200\,m contours) and surface velocity
+  \caption{Modelled surface topography (200\,\unit{m} contours) and surface velocity
            (colour mapping) corresponding to the maximum ice volume during
@@
   \includegraphics{hr_geom_duration}
-  \caption{Modelled duration of ice cover during the last 120\,ka using GRIP
+  \caption{Modelled duration of ice cover during the last 120\,\unit{ka} using GRIP
            and EPICA climate forcing.
@@
            from the Alaska Range (AR) to the Coast Mountains (CM) and
-           the Columbia and Rocky mountains (CRM) exists for about 32\,ka in
-           the GRIP simulation and 26\,ka in the EPICA simulation.
+           the Columbia and Rocky mountains (CRM) exists for about 32\,\unit{ka} in
+           the GRIP simulation and 26\,\unit{ka} in the EPICA simulation.
            The maximum extent of the ice sheet generally
@@
   \caption{Modelled duration of warm-based ice cover during the last
-           120\,ka. Long ice cover durations combined with basal
+           120\,\unit{ka}. Long ice cover durations combined with basal
            temperatures at the pressure melting point may explain the strong
@@
   \includegraphics{hr_maps_deglac}
-  \caption{Snapshots of modelled surface topography (200\,m contours)
+  \caption{Snapshots of modelled surface topography (200\,\unit{m} contours)
            and surface velocity (colour mapping) during the last deglaciation
-           from the GRIP (top panels) and EPICA (bottom panels) 5\,km simulations.
+           from the GRIP (top panels) and EPICA (bottom panels) 5\,\unit{km} simulations.
            Dashed segments (A--D) indicate the location of profiles used in
@@
            in green. Hatches denote re-advance of mountain-centred ice caps and
-           the decaying ice sheet between 14 and 10\,ka..
+           the decaying ice sheet between 14 and 10\,\unit{ka}..
            Dashed segments (A--D) indicate the location of profiles used in
@@
            indicate areas that remain non-sliding throughout deglaciation
-           (22.0--8.0\,ka), notably including parts of the Interior Plateau (IP).
+           (22.0--8.0\,\unit{ka}), notably including parts of the Interior Plateau (IP).
            Note the concentric patterns of deglacial flow in the Liard
@@
   \caption{Modelled bedrock (black) and ice surface (blue) topography profiles
-           during deglaciation (22.0--8.0\,ka) in the GRIP 5\,km
+           during deglaciation (22.0--8.0\,\unit{ka}) in the GRIP 5\,\unit{km}
            simulation, corresponding to the four transects indicated in
@@
   \caption{Modelled bedrock (black) and ice surface (red) topography profiles
-           during deglaciation (22.0--8.0\,ka) in the EPICA 5\,km
+           during deglaciation (22.0--8.0\,\unit{ka}) in the EPICA 5\,\unit{km}
            simulation, corresponding to the four transects indicated in

### 2016-01-17 cordillera-cycle.tex
Typeset: remove comment lines.
Typeset: more no-break space after 'a' and 'A'.

@@
 
-% =============================================================================
 \begin{document}
-% =============================================================================
 
@@
 
-% =============================================================================
 \introduction
 \label{sec:intro}
-% =============================================================================
 
@@
 
-      Although numerical ice sheet modelling has been established as a
-      useful tool to improve our understanding of the Cordilleran ice sheet
+      Although numerical ice sheet modelling has been established as
+      a~useful tool to improve our understanding of the Cordilleran ice sheet
       (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991};
@@
 
-% =============================================================================
 \section{Model setup}
 \label{sec:model}
-% =============================================================================
-
-% -----------------------------------------------------------------------------
+
 \subsection{Overview}
 \label{sec:overview}
-% -----------------------------------------------------------------------------
 
@@
       3\,\unit{km} depth (Sect.~\ref{sec:icedyn}). Surface mass balance is
-      computed using a positive degree-day (PDD) model
+      computed using a~positive degree-day (PDD) model
       (Sect.~\ref{sec:surface}). Climate forcing is provided by a~monthly
@@
 
-% -----------------------------------------------------------------------------
 \subsection{Ice thermodynamics}
 \label{sec:icedyn}
-% -----------------------------------------------------------------------------
 
@@
       based on their predominant components. The Shallow Shelf Approximation
-      (SSA) is used as a~``sliding law'' for the Shallow Ice Approximation
+      (SSA) is used as a ``sliding law'' for the Shallow Ice Approximation
       (SIA) by adding velocity solutions of the two approximations
@@
 
-      Ice shelf calving is computed using a~double criterion. First, a
-      physically-realistic calving flux is computed based on eigenvalues of
+      Ice shelf calving is computed using a~double criterion. First,
+      a~physically-realistic calving flux is computed based on eigenvalues of
       the horizontal strain rate tensor \citep{Winkelmann.etal.2011,
@@
 
-% -----------------------------------------------------------------------------
 \subsection{Surface mass balance}
 \label{sec:surface}
-% -----------------------------------------------------------------------------
 
@@
 
-% -----------------------------------------------------------------------------
 \subsection{Climate forcing}
 \label{sec:atm}
-% -----------------------------------------------------------------------------
-
-      Climate forcing driving ice sheet simulations consists of a
-      present-day monthly climatology,
+
+      Climate forcing driving ice sheet simulations consists of
+      a~present-day monthly climatology,
 $\{T_{m0}, P_{m0}\}$, where temperatures are modified by offset time series,
@@
 
-      Finally, lapse-rate corrections, ${\Delta}T_{LR}$, are computed as a
-      function of ice surface elevation, $s$, using the NARR surface
+      Finally, lapse-rate corrections, ${\Delta}T_{LR}$, are computed as
+      a~function of ice surface elevation, $s$, using the NARR surface
       geopotential height invariant field as a~reference topography,
@@
 
-% =============================================================================
 \section{Sensitivity to climate forcing time-series}
 \label{sec:results}
-% =============================================================================
-
-% -----------------------------------------------------------------------------
+
 \subsection{Evolution of ice volume}
-% -----------------------------------------------------------------------------
 
@@
 
-% -----------------------------------------------------------------------------
 \subsection{Extreme configurations}
-% -----------------------------------------------------------------------------
 
@@
       dimensions depend sensitively on the choice of the applied
-      palaeo-temperature record. Modelled ice volume minima spread over a
-      wide range between 1.69 and 2.88\,\unit{m}\,s.l.e.
+      palaeo-temperature record. Modelled ice volume minima spread over
+      a~wide range between 1.69 and 2.88\,\unit{m}\,s.l.e.
       (Table~\ref{tab:extrema}).
@@
 
-% =============================================================================
 \section{Comparison with the geologic record}
 \label{sec:discussion}
-% =============================================================================
 
@@
 
-% -----------------------------------------------------------------------------
 \subsection{Glacial maxima}
-% -----------------------------------------------------------------------------
 
@@
       The exact timing of modelled MIS~2 maximum ice volume depends strongly
-      on the choice of applied palaeo-temperature record, which allows for a
-      more in-depth comparison with geological evidence for the timing of
+      on the choice of applied palaeo-temperature record, which allows for
+      a~more in-depth comparison with geological evidence for the timing of
       maximum Cordilleran ice sheet extent. In the Puget Lowland
@@
       \citep{Ryder.etal.1991, Stumpf.etal.2000, Kleman.etal.2010,
-      Clague.Ward.2011, Margold.etal.2013a}. These indicate that a
-      latitudinal saddle connected ice dispersal centres in the Columbia
+      Clague.Ward.2011, Margold.etal.2013a}. These indicate that
+      a~latitudinal saddle connected ice dispersal centres in the Columbia
       Mountains with the main ice divide \citep{Ryder.etal.1991,
@@
       enhanced the topographic barrier formed by these mountain ranges,
-      which likely resulted in a~decrease of precipitation and, therefore, a
-      decrease of accumulation in the interior. Secondly, latent warming of
+      which likely resulted in a~decrease of precipitation and, therefore,
+      a~decrease of accumulation in the interior. Secondly, latent warming of
       the moisture-depleted air parcels flowing over this enhanced
@@
 
-% -----------------------------------------------------------------------------
 \subsection{Nucleation centres}
-% -----------------------------------------------------------------------------
 
@@
       Regardless of the applied forcing, this ice cap appears to survive
-      MIS~3 (Fig.~\ref{fig:lr_maps}, middle panels), and serves as a
-      nucleation centre at the onset of the glacial readvance towards the
+      MIS~3 (Fig.~\ref{fig:lr_maps}, middle panels), and serves as
+      a~nucleation centre at the onset of the glacial readvance towards the
       LGM (MIS~2). This situation appear similar to the neighbouring
@@
       transects (Fig.~\ref{fig:atm}). In fact, along most of the
-      north-western coast of North America, coastal mountain ranges form a
-      pronounced topographic barrier for westerly winds, capturing
+      north-western coast of North America, coastal mountain ranges form
+      a~pronounced topographic barrier for westerly winds, capturing
       atmospheric moisture in the form of orographic precipitation, and
@@
       (Figs.~\ref{fig:hr_geom_warmbase} and~\ref{fig:hr_geom_warmfrac})
-      explains the large-scale glacial erosional imprint on the landscape. A
-      well-developed network of glacial valleys running to the north-west on
+      explains the large-scale glacial erosional imprint on the landscape.
+      A~well-developed network of glacial valleys running to the north-west on
       the west slope of the Selwyn and Mackenzie Mountains
@@
 
-% -----------------------------------------------------------------------------
 \subsection{The last deglaciation}
-% -----------------------------------------------------------------------------
 
@@
       than the GRIP run, in closer agreement with dated maximum extents, and
-      shows a faster, uninterrupted deglaciation which yields sporadic ice
+      shows a~faster, uninterrupted deglaciation which yields sporadic ice
       cover more than 1\,\unit{ka} earlier. On the other hand, the GRIP run
-      yields a deglaciation in three steps, compatible with marine sediment
+      yields a~deglaciation in three steps, compatible with marine sediment
       sequences offshore Vancouver Island, where the distribution of
@@
       British Columbia, becomes completely deglaciated by 10\,\unit{ka},
-      whereas a significant ice cover remains over the Skeena, the Selwyn
+      whereas a~significant ice cover remains over the Skeena, the Selwyn
       and Mackenzie, and the Wrangell and Saint Elias mountains in the
@@
       present maps of basal flow directions immediately preceding
-      deglaciation or at the time of cessation of sliding inside a
-      cold-based retreating margin (Fig.~\ref{fig:hr_geom_lastflow}). The
+      deglaciation or at the time of cessation of sliding inside
+      a~cold-based retreating margin (Fig.~\ref{fig:hr_geom_lastflow}). The
       modelled deglacial flow patterns are mostly consistent between the
@@
       incongruent results could be that the missing feedback mechanisms
-      between ice sheet topography and regional climate resulted in a
-      modelled ice divide of the LGM ice sheet being too far to the east
+      between ice sheet topography and regional climate resulted in
+      a~modelled ice divide of the LGM ice sheet being too far to the east
       (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
@@
 
-% =============================================================================
 \conclusions
 \label{sec:concl}
-% =============================================================================
 
@@
       last glacial cycle, the Skeena Mountains host an ice cap which appears
-      to be fed by the moisture intruding inland from the west through a
-      topographic breach in the Coast Mountains. The Skeena ice cap acts as
+      to be fed by the moisture intruding inland from the west through
+      a~topographic breach in the Coast Mountains. The Skeena ice cap acts as
       the main nucleation centre for the glacial reavance towards the LGM
@@
 
-% Author contributions
 \section*{Author contributions}
@@
 
-% Acknowledgements
 \begin{acknowledgements}
@@
 
-% References
 \bibliographystyle{copernicus}
@@
            level equivalent (\unit{m}\,s.l.e.). Gray fields indicate Marine Oxygen
-           Isotope Stage (MIS) boundaries for MIS~2 and MIS~4 according to a
-           global compilation of benthic \chem{\delta^{18}O} records
+           Isotope Stage (MIS) boundaries for MIS~2 and MIS~4 according to
+      a~     global compilation of benthic \chem{\delta^{18}O} records
            \citep{Lisiecki.Raymo.2005}. Hatched rectangles highlight the
@@
            cover persists over the Skeena Mountains (SM) during most of the
-           simulation. See Fig.~\ref{fig:locmap} for a list of abbreviations.}
+           simulation. See Fig.~\ref{fig:locmap} for a~list of abbreviations.}
   \label{fig:hr_geom_duration}
@@
            Sliding grid cells were distinguished from non-sliding grid cells
-           using a basal velocity threshold of 1\,\unit{m\,yr^{-1}}.}
+           using a~basal velocity threshold of 1\,\unit{m\,yr^{-1}}.}
   \label{fig:hr_geom_lastflow}
@@
 
-% =============================================================================
 \end{document}
 \endinput
-% =============================================================================

### 2016-01-17 typeset.tex
Typeset: remove useless double spaces in typset.
Typeset: update figure labels in typeset.

@@
       at the LGM (Table~\ref{tab:extrema}) and realistic outlines
-      (Fig.~\ref{fig:snapshots}).
+      (Fig.~\ref{fig:lr_maps}).
 
@@
       Despite large differences in the input climate forcing
-      (Fig.~\ref{fig:timeseries}, upper panel), model output presents
+      (Fig.~\ref{fig:lr_ts}, upper panel), model output presents
       consistent features that can be observed across the range of forcing
@@
       between 23.2 and 16.8\,\unit{ka} during MIS~2
-      (Fig.~\ref{fig:timeseries}, lower panel). An ice volume minimum is
+      (Fig.~\ref{fig:lr_ts}, lower panel). An ice volume minimum is
       consistently reached between 53.0 and 41.3\,\unit{ka} during
@@
       but in three steps, separated by two periods of ice sheet regrowth
-      (Fig.~\ref{fig:timeseries}).
+      (Fig.~\ref{fig:lr_ts}).
 
@@
       attained during MIS~2 (Table~\ref{tab:extrema};
-      Fig.~\ref{fig:snapshots}, upper panels).  Corresponding maximum ice
+      Fig.~\ref{fig:lr_maps}, upper panels). Corresponding maximum ice
       volumes also differ significantly between model runs, and vary between
@@
       In the MIS~3 ice volume minimum reconstructions, a~central ice cap
-      persists over the Skeena Mountains (Fig.~\ref{fig:snapshots}, middle
+      persists over the Skeena Mountains (Fig.~\ref{fig:lr_maps}, middle
       panels). Although this ice cap is present in all simulations, its
@@
       Modelled ice sheet geometries during the LGM (MIS~2;
-      Fig.~\ref{fig:snapshots}, lower panels) invariably include
+      Fig.~\ref{fig:lr_maps}, lower panels) invariably include
       a~ca. 1500\,\unit{km}-long central divide above 3000\,\unit{m\,a.s.l.}
@@
       obtained during MIS~4 (61.9--56.5\,\unit{ka}) and the second during
-      MIS~2 (23.2--16.8\,\unit{ka}; Figs.~\ref{fig:timeseries},
-      \ref{fig:snapshots}; Table~\ref{tab:extrema}). These events broadly
+      MIS~2 (23.2--16.8\,\unit{ka}; Figs.~\ref{fig:lr_ts},
+      \ref{fig:lr_maps}; Table~\ref{tab:extrema}). These events broadly
       correspond in timing to the Gladstone (MIS~4) and McConnell (MIS~2)
@@
       sheet maximum extents that may be compatible with these field
-      constraints (Fig.~\ref{fig:timeseries}, lower panel;
+      constraints (Fig.~\ref{fig:lr_ts}, lower panel;
       Table~\ref{tab:extrema}). Simulations driven by the NGRIP, ODP~1012
@@
       alkenone palaeo-temperature reconstructions
-      (Fig.~\ref{fig:timeseries}, upper panel;
+      (Fig.~\ref{fig:lr_ts}, upper panel;
       \citealp[Fig.~3]{Herbert.etal.2001}). However, this early warming is
@@
       the NGRIP palaeo-temperature record depicts an early onset of
-      deglaciation (Fig.~\ref{fig:timeseries}) following its last glacial
+      deglaciation (Fig.~\ref{fig:lr_ts}) following its last glacial
       volume maximum (22.9\,\unit{ka}, Table~\ref{tab:extrema}) attained
@@
       re-run using a~higher-resolution grid (Sect.~\ref{sec:model};
-      Fig.~\ref{fig:timeseries}, lower panel, dotted lines).
+      Fig.~\ref{fig:lr_ts}, lower panel, dotted lines).
 
@@
       meridional ice divide over the western flank of the Rocky Mountains
-      (Figs.~\ref{fig:snapshots}, lower panels and \ref{fig:mis2}). This
+      (Figs.~\ref{fig:lr_maps}, lower panels and \ref{fig:mis2}). This
       result appears to contrast with palaeoglaciological reconstructions
@@
       forcing than those corresponding to ice volume maxima during MIS~2
-      (Figs.~\ref{fig:snapshots}, upper panels and \ref{fig:mis4}). The GRIP
+      (Figs.~\ref{fig:lr_maps}, upper panels and \ref{fig:mis4}). The GRIP
       simulation (Fig.~\ref{fig:mis4}, left panel) results in a~modelled
@@
       across northern and eastern sectors. In contrast, the EPICA simulation
-      produces a~lower ice volume maximum (Fig.~\ref{fig:timeseries}), which
+      produces a~lower ice volume maximum (Fig.~\ref{fig:lr_ts}), which
       translates in the modelled ice sheet geometry into a~significantly
@@
       simulations, we use numerical modelling output to compute durations of
-      ice cover throughout the last glacial cycle (Fig.~\ref{fig:duration}).
+      ice cover throughout the last glacial cycle (Fig.~\ref{fig:hr_geom_duration}).
 
@@
       mountain ranges of the North American Cordillera
-      (Fig.~\ref{fig:duration}, blue areas). A~2500\,\unit{km}-long
+      (Fig.~\ref{fig:hr_geom_duration}, blue areas). A~2500\,\unit{km}-long
       continuous expanse of ice, extending from the Alaska Range in the
@@
       operation for at most 32\,\unit{ka}, which is about a~third of the
-      timespan of the last glacial cycle (Fig.~\ref{fig:duration}, hatched
+      timespan of the last glacial cycle (Fig.~\ref{fig:hr_geom_duration}, hatched
       areas). However, except for its margins in the Pacific Ocean and in
@@
       ice sheet is attained for a~much shorter period of time of only few
-      thousand years (Fig.~\ref{fig:duration}, red areas). This result
+      thousand years (Fig.~\ref{fig:hr_geom_duration}, red areas). This result
       illustrates that the maximum extents of the modelled ice sheet during
@@
       where modelled glaciers are confined to its foothills during the
-      entire simulation period (Fig.~\ref{fig:duration}, AR). This apparent
+      entire simulation period (Fig.~\ref{fig:hr_geom_duration}, AR). This apparent
       insensitivity of modelled glacial extent to temperature fluctuations
@@
       centres, visible on the modelled ice cover duration maps
-      (Fig.~\ref{fig:duration}), are located over the Coast Mountains (CM),
+      (Fig.~\ref{fig:hr_geom_duration}), are located over the Coast Mountains (CM),
       the Columbia and Rocky mountains (CRM), the Skeena Mountains (SM), and
@@
       appear to be a~major feed to the Cordilleran ice sheet
-      (Fig.~\ref{fig:duration}, WSEM).  Although the Coast, Skeena and
+      (Fig.~\ref{fig:hr_geom_duration}, WSEM). Although the Coast, Skeena and
       Columbia and Rocky mountains (CM, SM, CRM) are covered by mountain
@@
       last glacial cycle. The Northern Rocky Mountains
-      (Fig.~\ref{fig:duration}, NRM) do not act as a~nucleation centre, but
+      (Fig.~\ref{fig:hr_geom_duration}, NRM) do not act as a~nucleation centre, but
       rather as a~pinning point for the Cordilleran ice sheet margin coming
@@
       its predominance over the other ice-dispersal centres
-      (Figs.~\ref{fig:snapshots} and~\ref{fig:duration}, SM). Regardless of
+      (Figs.~\ref{fig:lr_maps} and~\ref{fig:hr_geom_duration}, SM). Regardless of
       the applied forcing, this ice cap appears to survive MIS~3
-      (Fig.~\ref{fig:snapshots}, middle panels), and serves as a~nucleation
+      (Fig.~\ref{fig:lr_maps}, middle panels), and serves as a~nucleation
       centre at the onset of the glacial readvance towards the LGM
@@
       A~correlation is observed between the modelled duration of warm based
-      ice cover (Fig.~\ref{fig:warmbase}) and the degree of glacial
+      ice cover (Fig.~\ref{fig:hr_geom_warmbase}) and the degree of glacial
       modification of the landscape (mainly in terms of the development of
@@
       the Wrangell and Saint Elias mountains, and radiating off the Skeena
-      Mountains (Figs.~\ref{fig:duration} and \ref{fig:warmbase};
+      Mountains (Figs.~\ref{fig:hr_geom_duration} and \ref{fig:hr_geom_warmbase};
       \citealp[Fig.~2]{Kleman.etal.2010}). The Skeena Mountains, for
@@
       \citep[Fig.~2]{Kleman.etal.2010}. We suggest that persistent ice cover
-      (Fig.~\ref{fig:duration}) associated with basal ice temperatures at
-      the pressure melting point (Figs.~\ref{fig:warmbase}
-      and~\ref{fig:warmfrac}) explains the large-scale glacial erosional
+      (Fig.~\ref{fig:hr_geom_duration}) associated with basal ice temperatures at
+      the pressure melting point (Figs.~\ref{fig:hr_geom_warmbase}
+      and~\ref{fig:hr_geom_warmfrac}) explains the large-scale glacial erosional
       imprint on the landscape. A~well-developed network of glacial valleys
@@
       \citealp[Fig.~8] {Stroeven.etal.2010}) is modelled to have hosted
-      predominantly warm-based ice (Fig.~\ref{fig:warmfrac}). However,
+      predominantly warm-based ice (Fig.~\ref{fig:hr_geom_warmfrac}). However,
       because it is only glaciated for a~short fraction of the last glacial
-      cycle in our simulations (Fig.~\ref{fig:duration}), this perhaps
+      cycle in our simulations (Fig.~\ref{fig:hr_geom_duration}), this perhaps
       indicates that the observed landscape pattern originates from multiple
@@
       Selwyn and Mackenzie mountains ice dispersal centre
-      (Fig.~\ref{fig:duration}, SMKM), prior to the Late Pleistocene
+      (Fig.~\ref{fig:hr_geom_duration}, SMKM), prior to the Late Pleistocene
       \citep[cf.][]{Ward.etal.2008, Demuro.etal.2012}.
@@
       The modelled distribution of warm-based ice cover
-      (Figs.~\ref{fig:warmbase} and~\ref{fig:warmfrac}) is inevitably
+      (Figs.~\ref{fig:hr_geom_warmbase} and~\ref{fig:hr_geom_warmfrac}) is inevitably
       affected by our assumption of a~constant, 70\,\unit{mW\,m^{-2}}
@@
       the pacing of deglaciation depend critically on the choice of climate
-      forcing (Table~\ref{tab:extrema}; Figs.~\ref{fig:timeseries}
-      and~\ref{fig:deglacseries}). Adopting the EPICA climate forcing yields
+      forcing (Table~\ref{tab:extrema}; Figs.~\ref{fig:lr_ts}
+      and~\ref{fig:lr_ts_deglac}). Adopting the EPICA climate forcing yields
       peak ice volume at 17.2\,\unit{ka} and an uninterrupted deglaciation
-      until about 9\,\unit{ka} (Fig.~\ref{fig:deglacseries}, lower panel,
+      until about 9\,\unit{ka} (Fig.~\ref{fig:lr_ts_deglac}, lower panel,
       red curves). On the contrary, the simulation driven by the GRIP
@@
       14.5\,\unit{ka}, and the second between 13.1 and 11.6\,\unit{ka}
-      (Fig.~\ref{fig:deglacseries}, lower panel, blue curve).
+      (Fig.~\ref{fig:lr_ts_deglac}, lower panel, blue curve).
 
@@
       Modelled patterns of ice sheet retreat are relatively consistent
-      between the two simulations (Figs.~\ref{fig:deglacshots}
-      and~\ref{fig:deglac}). The southern sector of the modelling domain,
+      between the two simulations (Figs.~\ref{fig:hr_geom_deglacageshots}
+      and~\ref{fig:hr_geom_deglacage}). The southern sector of the modelling domain,
       including the Puget Lowland, the Coast Mountains, the Columbia and and
@@
       may have been disconnected from the waning Cordilleran ice sheet, the
-      Finlay River area in the Omenica Mountains (Fig.~\ref{fig:deglac}, OM)
+      Finlay River area in the Omenica Mountains (Fig.~\ref{fig:hr_geom_deglacage}, OM)
       presents a~different kind of evidence. There, sharp-crested moraines
@@
       in the Coast Mountains and in the Columbia and Rocky Mountains
-      (Fig.~\ref{fig:deglac}, left panel). In addition to matching the
+      (Fig.~\ref{fig:hr_geom_deglacage}, left panel). In addition to matching the
       location of some local readvances, the GRIP-driven simulation shows
@@
       only a~tightly restricted late-glacial readvance on the western slopes
-      of the Saint Elias and the Coast mountains (Fig.~\ref{fig:deglac},
+      of the Saint Elias and the Coast mountains (Fig.~\ref{fig:hr_geom_deglacage},
       right panel).
@@
       deglaciation or at the time of cessation of sliding inside
-      a~cold-based retreating margin (Fig.~\ref{fig:lastflow}). The modelled
+      a~cold-based retreating margin (Fig.~\ref{fig:hr_geom_lastflow}). The modelled
       deglacial flow patterns are mostly consistent between the GRIP and
@@
       experience any basal sliding throughout the deglaciation phase
-      (Fig.~\ref{fig:lastflow}, hatched areas). This notably includes parts
+      (Fig.~\ref{fig:hr_geom_lastflow}, hatched areas). This notably includes parts
       of the Interior Plateau in British Columbia, major portions of the
@@
       show similarities with the patterns of deglacial ice flow from
-      numerical modelling (Fig.~\ref{fig:lastflow}). In the northern half of
+      numerical modelling (Fig.~\ref{fig:hr_geom_lastflow}). In the northern half of
       the modelling domain, modelled deglacial flow directions depict an
@@
       towards mountain ranges. Converging deglacial flow patterns in the
-      Liard Lowland, for instance (Fig.~\ref{fig:lastflow}), closely
+      Liard Lowland, for instance (Fig.~\ref{fig:hr_geom_lastflow}), closely
       resemble the pattern indicated by glacial lineations
@@
       simulations produce a~retreat of the ice margin towards the north-east
-      (Fig.~\ref{fig:deglac}), a~pattern which is validated by the
+      (Fig.~\ref{fig:hr_geom_deglacage}), a~pattern which is validated by the
       geomorphological and stratigraphical record for ancient pro-glacial
@@
       simulation produces negligible basal sliding on the plateau during
-      deglaciation (Fig.~\ref{fig:lastflow}). Yet, the Interior Plateau also
+      deglaciation (Fig.~\ref{fig:hr_geom_lastflow}). Yet, the Interior Plateau also
       hosts an impressive set of glacial lineations which indicate
@@
       conditions but, on the contrary, enlarge the region of negligible
-      basal sliding (Fig.~\ref{fig:lastflow}). Thus, a~second explanation
+      basal sliding (Fig.~\ref{fig:hr_geom_lastflow}). Thus, a~second explanation
       for the incongruent results could be that the Interior Plateau
@@
       The modelled deglaciation of the Interior Plateau of British Columbia
-      consists of a~rapid northwards retreat (Fig.~\ref{fig:deglac}) of
-      southwards-flowing non-sliding ice lobes (Fig.~\ref{fig:lastflow})
+      consists of a~rapid northwards retreat (Fig.~\ref{fig:hr_geom_deglacage}) of
+      southwards-flowing non-sliding ice lobes (Fig.~\ref{fig:hr_geom_lastflow})
       positioned in-between deglaciated mountain ranges
-      (Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}). This
+      (Figs.~\ref{fig:hr_pf_grip} and~\ref{fig:hr_pf_epica}). This
       result appears compatible with the prevailing conceptual model of
@@
            MIS~4, 3 and 2 for each of the six low-resolution simulations
-           (Fig.~\ref{fig:timeseries}).}
+           (Fig.~\ref{fig:lr_ts}).}
 \label{tab:extrema}
@@
            5\,\unit{km}-resolution runs.}
-\label{fig:timeseries}%
+\label{fig:lr_ts}%
 \end{figure}%
@@
            corresponding to the ice volume extremes indicated on
-           Fig.~\ref{fig:timeseries}. An ice cap persists over the Skeena
+           Fig.~\ref{fig:lr_ts}. An ice cap persists over the Skeena
            Mountains (SM) during MIS~3. Note the occurence of spatial
            similarities despite large differences in timing.}
-\label{fig:snapshots}%
+\label{fig:lr_maps}%
 \end{figure}%
@@
            simulation. See Fig.~\ref{fig:locmap} for a~list of abbreviations.}
-\label{fig:duration}%
+\label{fig:hr_geom_duration}%
 \end{figure}%
@@
            Hatches indicate areas that were covered by cold ice only.}
-\label{fig:warmbase}%
+\label{fig:hr_geom_warmbase}%
 \end{figure}%
@@
            Hatches indicate areas that were covered by cold ice only.}
-\label{fig:warmfrac}%
+\label{fig:hr_geom_warmfrac}%
 \end{figure}%
@@
            equivalent (bottom panel).}
-\label{fig:deglacseries}%
+\label{fig:lr_ts_deglac}%
 \end{figure}%
@@
            Dashed segments \textbf{(A--D)} indicate the location of profiles used in
-           Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}.}
-\label{fig:deglacshots}%
+           Figs.~\ref{fig:hr_pf_grip} and~\ref{fig:hr_pf_epica}.}
+\label{fig:hr_geom_deglacageshots}%
 \end{figure}%
@@
            Dashed segments \textbf{(A--D)} indicate the location of profiles used in
-           Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}.}
-\label{fig:deglac}%
+           Figs.~\ref{fig:hr_pf_grip} and~\ref{fig:hr_pf_epica}.}
+\label{fig:hr_geom_deglacage}%
 \end{figure}%
@@
            using a~basal velocity threshold of 1\,\unit{m\,yr^{-1}}.}
-\label{fig:lastflow}%
+\label{fig:hr_geom_lastflow}%
 \end{figure}%
@@
            simulation, corresponding to the four transects indicated in
-           Figs.~\ref{fig:deglacshots} and~\ref{fig:deglac}.}
-\label{fig:profiles-grip}%
+           Figs.~\ref{fig:hr_geom_deglacageshots} and~\ref{fig:hr_geom_deglacage}.}
+\label{fig:hr_pf_grip}%
 \end{figure}%
@@
            simulation, corresponding to the four transects indicated in
-           Figs.~\ref{fig:deglacshots} and~\ref{fig:deglac}.}
-\label{fig:profiles-epica}%
+           Figs.~\ref{fig:hr_geom_deglacageshots} and~\ref{fig:hr_geom_deglacage}.}
+\label{fig:hr_pf_epica}%
 \end{figure}

### 2016-01-17 cordillera-cycle.tex
Typeset: add LaTeX hacks in header.
Typeset: add received and accepted dates.
Typeset: fix affiliations.

@@
 \documentclass[tc, manuscript]{copernicus}
-
 \graphicspath{{figures/}}
 
-\begin{document}
+\begin{document}\hack{\sloppy}
 
@@
 \Author[2]{J.}{Kleman}
+
 \runningauthor{J.~Seguinot et~al.}
+
 \correspondence{J.~Seguinot (seguinot@vaw.baug.ethz.ch)}
@@
 
-\affil[1]{Laboratory of Hydraulics, Hydrology and Glaciology, ETH Zürich,
-          Zürich, Switzerland}
-\affil[2]{Department of Physical Geography and the Bolin Centre for Climate
-          Research, Stockholm University, Stockholm, Sweden}
-\affil[3]{Helmholtz Centre Potsdam, GFZ German Research Centre for Geosciences,
-          Potsdam, Germany}
-
-\received{}
-\pubdiscuss{}
-\revised{}
-\accepted{}
+\affil[1]{Laboratory of Hydraulics, Hydrology and Glaciology, ETH Z\"{u}rich, Z\"{u}rich, Switzerland}
+\affil[2]{Department of Physical Geography and the Bolin Centre for Climate Research, Stockholm University, Stockholm, Sweden}
+\affil[3]{Helmholtz Centre Potsdam, GFZ German Research Centre for Geosciences, Potsdam,~Germany}
+
+\hack{\allowdisplaybreaks}
+
+\received{21~June 2015}
+\accepted{25~June 2015}
 \published{}
@@
 \firstpage{1}
+
 \maketitle

### 2016-01-17 cordillera-cycle.tex
Typeset: remove no-break space from abstract.
It is apparently not allowed.

@@
 response to evolving climate forcing, all simulations produce two
-major glaciations during marine oxygen isotope stages~4
-(61.9--56.5\,\unit{ka}) and~2 (23.2--16.8\,\unit{ka}). The timing of
+major glaciations during marine oxygen isotope stages 4
+(61.9--56.5\,\unit{ka}) and 2 (23.2--16.8\,\unit{ka}). The timing of
 glaciation is better reproduced using temperature reconstructions
@@
 Maximum and hosts the last remains of Cordilleran ice until the
-qmiddle Holocene (6.6--6.2\,\unit{ka}).
+middle Holocene (6.6--6.2\,\unit{ka}).
 \end{abstract}

### 2016-01-17 cordillera-cycle.tex
Typeset: change fonts in math subscripts.

@@
     \label{eqn:pseudoplastic}
-    \vec{\tau}_b = -\tau_c \frac{\vec{v}_b}{{v_{th}}^q\,|\vec{v}_b|^{1-q}} \,,
+    \vec{\tau}_{\mathrm{b}} = -\tau_{\mathrm{c}} \frac{\vec{v}_{\mathrm{b}}}{{v_{\text{th}}}^q\,|\vec{v}_{\mathrm{b}}|^{1-q}} \,,
 \end{equation}
-      relates the bed-parallel shear stresses, $\vec{\tau}_b$, to the
-      sliding velocity, $\vec{v}_b$ (Table~\ref{tab:params}). The yield
-      stress, $\tau_c$, is modelled using the Mohr--Coulomb criterion,
+      relates the bed-parallel shear stresses, $\vec{\tau}_{\mathrm{b}}$, to the
+      sliding velocity, $\vec{v}_{\mathrm{b}}$ (Table~\ref{tab:params}). The yield
+      stress, $\tau_{\mathrm{c}}$, is modelled using the Mohr--Coulomb criterion,
 \begin{equation}
-   \tau_c = c_0 + N\,\tan{\phi} \,,
+   \tau_{\mathrm{c}} = c_0 + N\,\tan{\phi} \,,
 \end{equation}
@@
 \begin{equation}
-    N = \delta \rho gh \, 10^{(e_0/C_c) (1 - (W/W_{max}))} \,,
+    N = \delta \rho gh \, 10^{(e_0/C_{\mathrm{c}}) (1 - (W/W_{\text{max}}))} \,,
 \end{equation}
       where $\delta$ is chosen as 0.02, $e_0$ is a~measured reference void
-      ratio and $C_c$ is a~measured compressibility coefficient
+      ratio and $C_{\mathrm{c}}$ is a~measured compressibility coefficient
       (Table~\ref{tab:params}). The amount of water at the base, $W$, varies
-      from zero to $W_{max}=2$\,\unit{m}, a~threshold above which additional
+      from zero to $W_{\text{max}}=2$\,\unit{m}, a~threshold above which additional
       melt water is assumed to drain off instantaneously. Finally, the
@@
       \citep{Lingle.Clark.1985,Bueler.etal.2007}. A~relatively low viscosity
-      value of $\nu_m = 1\times10^{19}$\,\unit{Pa\,s} is used for the
+      value of $\nu_{m} = 1\times10^{19}$\,\unit{Pa\,s} is used for the
       astenosphere (Table~\ref{tab:params}) in accordance with the results
@@
       Ice surface accumulation and ablation are computed from monthly mean
-      near-surface air temperature, $T_m$, monthly standard deviation of
+      near-surface air temperature, $T_{\mathrm{m}}$, monthly standard deviation of
       near-surface air temperature, $\sigma$, and monthly precipitation,
-      $P_m$, using a~temperature-index model \citep[e.g.,][]{Hock.2003}.
+      $P_{\mathrm{m}}$, using a~temperature-index model \citep[e.g.,][]{Hock.2003}.
       Accumulation is equal to precipitation when air temperatures are below
@@
       assuming a~normal temperature distribution of standard deviation
-      $\sigma$ aroung the expected value $T_m$. It is expressed by an
+      $\sigma$ aroung the expected value $T_{\mathrm{m}}$. It is expressed by an
       error-function formulation \citep{Calov.Greve.2005},
@@
     \label{eqn:calovgreve}
-    \mathrm{PDD} = \int_{t_1}^{t_2} \mathrm{d}t
+    {\text{PDD}} = \int_{t_1}^{t_2} \mathrm{d}t
         \left[\frac{\sigma}{\sqrt{2\pi}}
-                \exp\left({-\frac{T_{m}^2}{2\sigma^2}}\right)
-              + \frac{T_{m}}{2} \, \mathrm{erfc}
-                \left(-\frac{T_{m}}{\sqrt{2}\sigma}\right)\right] \,,
+                \exp\left({-\frac{T_{\mathrm{m}}^2}{2\sigma^2}}\right)
+              + \frac{T_{\mathrm{m}}}{2} \, {\text{erfc}}
+                \left(-\frac{T_{\mathrm{m}}}{\sqrt{2}\sigma}\right)\right] \,,
 \end{equation}
@@
       a~present-day monthly climatology,
-$\{T_{m0}, P_{m0}\}$, where temperatures are modified by offset time series,
-${\Delta}T_{TS}$, and lapse-rate corrections, ${\Delta}T_{LR}$:
+$\{T_{\mathrm{m}0}, P_{\mathrm{m}0}\}$, where temperatures are modified by offset time series,
+${\Delta}T_{\text{TS}}$, and lapse-rate corrections, ${\Delta}T_{\text{LR}}$:
 \begin{align}
-    T_m(t, x, y) &= T_{m0}(x, y) + {\Delta}T_{TS}(t)
-                    + {\Delta}T_{LR}(t, x, y) \,, \\
-    P_m(t, x, y) &= P_{m0}(x, y) \,.
+    T_{\mathrm{m}}(t, x, y) &= T_{\mathrm{m}0}(x, y) + {\Delta}T_{\text{TS}}(t)
+                    + {\Delta}T_{\text{LR}}(t, x, y) \,, \\
+    P_{\mathrm{m}}(t, x, y) &= P_{\mathrm{m}0}(x, y) \,.
 \end{align}
@@
 
-      Temperature offset time-series, ${\Delta}T_{TS}$, are derived from
+      Temperature offset time-series, ${\Delta}T_{\text{TS}}$, are derived from
       palaeo-temperature proxy records from the Greenland Ice Core Project
@@
 \begin{align}
-    {\Delta}T_{TS}(t) = &-11.88 [\chem{\delta^{18}O}(t)
+    {\Delta}T_{\text{TS}}(t) = &-11.88 [\chem{\delta^{18}O}(t)
                                 -\chem{\delta^{18}O}(0)] \nonumber \\
@@
 
-      Finally, lapse-rate corrections, ${\Delta}T_{LR}$, are computed as
+      Finally, lapse-rate corrections, ${\Delta}T_{\text{LR}}$, are computed as
       a~function of ice surface elevation, $s$, using the NARR surface
       geopotential height invariant field as a~reference topography,
-      $b_{ref}$:
+      $b_{\text{ref}}$:
 \begin{align}
-    {\Delta}T_{LR}(t, x, y) &= -\gamma [s(t, x, y)-b_{ref}] \\
-                            &= -\gamma [h(t, x, y)+b(t, x, y)-b_{ref}],
+    {\Delta}T_{\text{LR}}(t, x, y) &= -\gamma [s(t, x, y)-b_{\text{ref}}] \\
+                            &= -\gamma [h(t, x, y)+b(t, x, y)-b_{\text{ref}}],
 \end{align}
@@
       one hand, and for differences between the basal topography of the ice
-      flow model, $b$, and the NARR reference topography, $b_{ref}$, on the
+      flow model, $b$, and the NARR reference topography, $b_{\text{ref}}$, on the
       other hand. All simulations use an annual temperature lapse rate of
@@
 
-    $v_{th}$& Pseudo-plastic threshold velocity
+    $v_{\text{th}}$& Pseudo-plastic threshold velocity
             & 100.0
@@
 
-    $C_c$   & Till compressibility coefficient
+    $C_{\mathrm{c}}$   & Till compressibility coefficient
             & 0.12
@@
 
-    $W_{max}$ & Maximal till water thickness
+    $W_{\text{max}}$ & Maximal till water thickness
             & 2.0
@@
 
-    $\rho_b$& Bedrock density
+    $\rho_{\mathrm{b}}$& Bedrock density
             & 3300
@@
 
-    $c_b$   & Bedrock specific heat capacity
+    $c_{\mathrm{b}}$   & Bedrock specific heat capacity
             & 1000
@@
 
-    $k_b$   & Bedrock thermal conductivity
+    $k_{\mathrm{b}}$   & Bedrock thermal conductivity
             & 3.0
@@
 
-    $T_s$   & Temperature of snow precipitation
+    $T_{\mathrm{s}}$   & Temperature of snow precipitation
             & 273.15
@@
 
-    $T_r$   & Temperature of rain precipitation
+    $T_{\mathrm{r}}$   & Temperature of rain precipitation
             & 275.15
@@
 
-    $F_s$   & Degree-day factor for snow
+    $F_{\mathrm{s}}$   & Degree-day factor for snow
             & $3.04\times10^{-3}$
@@
 
-    $F_i$   & Degree-day factor for ice
+    $F_{\mathrm{i}}$   & Degree-day factor for ice
             & $4.59\times10^{-3}$
@@
            from the GRIP (top panels) and EPICA (bottom panels) 5\,\unit{km} simulations.
-           Dashed segments (A--D) indicate the location of profiles used in
+           Dashed segments \textbf{(A--D)} indicate the location of profiles used in
            Figs.~\ref{fig:hr_pf_grip} and~\ref{fig:hr_pf_epica}.}
@@
            the decaying ice sheet between 14 and 10\,\unit{ka}..
-           Dashed segments (A--D) indicate the location of profiles used in
+           Dashed segments \textbf{(A--D)} indicate the location of profiles used in
            Figs.~\ref{fig:hr_pf_grip} and~\ref{fig:hr_pf_epica}.}

### 2016-01-17 cordillera-cycle.tex
Typeset: replace equation environments by align.

@@
       A~pseudo-plastic sliding law,
-\begin{equation}
-    \label{eqn:pseudoplastic}
+\begin{align}
+&\label{eqn:pseudoplastic}
     \vec{\tau}_{\mathrm{b}} = -\tau_{\mathrm{c}} \frac{\vec{v}_{\mathrm{b}}}{{v_{\text{th}}}^q\,|\vec{v}_{\mathrm{b}}|^{1-q}} \,,
-\end{equation}
+\end{align}
       relates the bed-parallel shear stresses, $\vec{\tau}_{\mathrm{b}}$, to the
@@
       stress, $\tau_{\mathrm{c}}$, is modelled using the Mohr--Coulomb criterion,
-\begin{equation}
-   \tau_{\mathrm{c}} = c_0 + N\,\tan{\phi} \,,
-\end{equation}
+\begin{align}
+&\tau_{\mathrm{c}} = c_0 + N\,\tan{\phi} \,,
+\end{align}
       where cohesion, $c_0$, is assumed to be zero. The friction angle,
@@
       \citep{Tulaczyk.etal.2000, Bueler.Pelt.2015},
-\begin{equation}
-    N = \delta \rho gh \, 10^{(e_0/C_{\mathrm{c}}) (1 - (W/W_{\text{max}}))} \,,
-\end{equation}
+\begin{align}
+&N = \delta \rho gh \, 10^{(e_0/C_{\mathrm{c}}) (1 - (W/W_{\text{max}}))} \,,
+\end{align}
       where $\delta$ is chosen as 0.02, $e_0$ is a~measured reference void
@@
       error-function formulation \citep{Calov.Greve.2005},
-\begin{equation}
-    \label{eqn:calovgreve}
+\begin{align}
+&\label{eqn:calovgreve}
     {\text{PDD}} = \int_{t_1}^{t_2} \mathrm{d}t
@@
                 \left(-\frac{T_{\mathrm{m}}}{\sqrt{2}\sigma}\right)\right] \,,
-\end{equation}
+\end{align}
       which is numerically approximated using week-long sub-intervals. In
@@
 \begin{align}
-    T_{\mathrm{m}}(t, x, y) &= T_{\mathrm{m}0}(x, y) + {\Delta}T_{\text{TS}}(t)
+&T_{\mathrm{m}}(t, x, y) = T_{\mathrm{m}0}(x, y) + {\Delta}T_{\text{TS}}(t)
                     + {\Delta}T_{\text{LR}}(t, x, y) \,, \\
-    P_{\mathrm{m}}(t, x, y) &= P_{\mathrm{m}0}(x, y) \,.
+&    P_{\mathrm{m}}(t, x, y) = P_{\mathrm{m}0}(x, y) \,.
 \end{align}
@@
 \begin{align}
-    {\Delta}T_{\text{TS}}(t) = &-11.88 [\chem{\delta^{18}O}(t)
+{\Delta}T_{\text{TS}}(t) ={~}&-11.88 [\chem{\delta^{18}O}(t)
                                 -\chem{\delta^{18}O}(0)] \nonumber \\

### 2016-01-17 cordillera-cycle.tex
Typeset: format internal and external references.

@@
       reconstructions of maximum ice sheet extents
-      (\citealp{Prest.etal.1968}; \citealp[Fig. 1.12]{Clague.1989};
+      (\citealp{Prest.etal.1968}; \citealp[Fig.~1.12]{Clague.1989};
       \citealp{Duk-Rodkin.1999}; \citealp{Booth.etal.2003};
@@
       (SIA) by adding velocity solutions of the two approximations
-      \citep[Eqns.~7--9 and 15]{Bueler.Brown.2009, Winkelmann.etal.2011}.
+      \citep[Eqs.~7--9 and 15]{Bueler.Brown.2009, Winkelmann.etal.2011}.
 
@@
       marine sediments (cf. \citealp{Martin.etal.2011};
-      \citealp[supplement]{Aschwanden.etal.2013};
+      \citealp[Supplement]{Aschwanden.etal.2013};
       \citealp{PISM-authors.2015}). Effective pressure, $N$, is related to
@@
       meridional ice divide over the western flank of the Rocky Mountains
-      (Fig.~\ref{fig:lr_maps}, lower panels; Fig.~\ref{fig:mis2}). This
+      (Figs.~\ref{fig:lr_maps}, lower panels and \ref{fig:mis2}). This
       result appears to contrast with palaeoglaciological reconstructions
@@
       forcing than those corresponding to ice volume maxima during MIS~2
-      (Fig.~\ref{fig:lr_maps}, upper panels; Fig.~\ref{fig:mis4}). The GRIP
+      (Figs.~\ref{fig:lr_maps}, upper panels and \ref{fig:mis4}). The GRIP
       simulation (Fig.~\ref{fig:mis4}, left panel) results in a~modelled
@@
       of the ice sheet is attained for a~much shorter period of time of only
-      few thousand years (Figs.~\ref{fig:hr_geom_duration}, red areas). This
+      few thousand years (Fig.~\ref{fig:hr_geom_duration}, red areas). This
       result illustrates that the maximum extents of the modelled ice sheet
@@
       if included at all. However, next after the mapping of maximum ice
-      sheet extents during MIS~2 and MIS~4 (Sect.~\ref{sec:mis2}
+      sheet extents during MIS~2 and MIS~4 (Sects.~\ref{sec:mis2}
       and~\ref{sec:mis4}), geomorphologically-based reconstructions of
@@
       between the two simulations (Figs.~\ref{fig:hr_geom_deglacageshots}
-      and.~\ref{fig:hr_geom_deglacage}). The southern sector of the
+      and~\ref{fig:hr_geom_deglacage}). The southern sector of the
       modelling domain, including the Puget Lowland, the Coast Mountains,

### 2016-01-17 cordillera-cycle.tex
Typeset: more unit formatting.

@@
       where cohesion, $c_0$, is assumed to be zero. The friction angle,
-      $\phi$, varies from 15 to 45\degree as a~piecewise-linear function of
+      $\phi$, varies from 15 to 45\unit{\degree} as a~piecewise-linear function of
       modern bed elevation, with the lowest value occuring below modern sea
@@
       volumes also differ significantly between model runs, and vary between
-      3.84 and~8.84\,m sea level equivalents (m.~s.l.e.;
+      3.84 and~8.84\,\unit{m} sea level equivalents (m\,s.l.e.;
       Table~\ref{tab:extrema}).
@@
       5.04\,\unit{m}\,s.l.e. (61.9\,\unit{ka}) in the EPICA simulation,
-      corresponding to respectively 90\,\% and 57\,\% of modelled MIS~2 ice
+      corresponding to respectively 90 and 57\,\unit{\%} of modelled MIS~2 ice
       volumes.
@@
       (51.8\,\unit{ka}) in the EPICA simulation, corresponding to
-      respectively 19\,\% and 31\,\% of the MIS~2 ice volumes.
+      respectively 19 and 31\,\unit{\%} of the MIS~2 ice volumes.
 
@@
            $T_{[32, 22]}$ refers to the resulting mean temperature anomaly
-           during the period -32 to~-22~ka after scaling.}
+           during the period -32 to~-22\,\unit{ka} after scaling.}
   \label{tab:records}
@@
              & \multicolumn{3}{c}{Ice extent (\unit{10^6\,km^2})}
-             & \multicolumn{3}{c}{Ice volume (\unit{m}\,s.l.e.)} \\
+             & \multicolumn{3}{c}{Ice volume (m\,s.l.e.)} \\
     Record   &  MIS~4 &  MIS~3 &  MIS~2
@@
            through the last 120\,\unit{ka}. Ice volumes are expressed in meters of sea
-           level equivalent (\unit{m}\,s.l.e.). Gray fields indicate Marine Oxygen
+           level equivalent (\unit{m\,s.l.e.}). Gray fields indicate Marine Oxygen
            Isotope Stage (MIS) boundaries for MIS~2 and MIS~4 according to
-      a~     global compilation of benthic \chem{\delta^{18}O} records
+           a~global compilation of benthic \chem{\delta^{18}O} records
            \citep{Lisiecki.Raymo.2005}. Hatched rectangles highlight the
@@
            in green. Hatches denote re-advance of mountain-centred ice caps and
-           the decaying ice sheet between 14 and 10\,\unit{ka}..
+           the decaying ice sheet between 14 and 10\,\unit{ka}.
            Dashed segments \textbf{(A--D)} indicate the location of profiles used in

### 2016-01-17 cordillera-cycle.tex
Typeset: use new author contrib and suppl cmds.

@@
 
-\section*{Author contributions}
+\Supplementary{zip}
+
+\authorcontribution{%
       J.~Seguinot ran the simulations; I.~Rogozhina guided experiment
@@
       interpretation and comparison of model results against geological
-      evidence. All authors contributed to the text.
+      evidence. All authors contributed to the text.}
 

### 2016-01-17 typeset.tex
Typeset: use manuscript format in typeset.
I will use this to produce the final diff for the editor.
Typeset: whitespace changes in typeset.

@@
-\documentclass[tcd, online, hvmath]{copernicus}
+\documentclass[tc, manuscript]{copernicus}
 
@@
       records \citep{Cosma.etal.2008, Davies.etal.2011}, the LGM Cordilleran
-      ice sheet extent was short-lived. However, former ice thicknesses
-      and, therefore, the ice sheet's contribution to the LGM sea level
-      lowstand \citep{Carlson.Clark.2012, Clark.Mix.2002} remain uncertain.
+      ice sheet extent was short-lived. However, former ice thicknesses and,
+      therefore, the ice sheet's contribution to the LGM sea level lowstand
+      \citep{Carlson.Clark.2012, Clark.Mix.2002} remain uncertain.
 
@@
       \citep[e.g.,][]{Sissons.1979, Lundqvist.1987, Ivy-Ochs.etal.1999,
-      Stea.etal.2011}. Nevertheless, recent oxygen isotope measurements
-      from Gulf of Alaska sediments reveal a~climatic evolution highly
-      correlated to that of Greenland during this period, including
-      a~distinct Late Glacial cold reversal between 14.1 and 11.7\,\unit{ka}
+      Stea.etal.2011}. Nevertheless, recent oxygen isotope measurements from
+      Gulf of Alaska sediments reveal a~climatic evolution highly correlated to
+      that of Greenland during this period, including a~distinct Late Glacial
+      cold reversal between 14.1 and 11.7\,\unit{ka}
       \citep{Praetorius.Mix.2014}.
@@
       Although numerical ice sheet modelling has been established as
-      a~useful tool to improve our understanding of the Cordilleran ice
-      sheet (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991};
+      a~useful tool to improve our understanding of the Cordilleran ice sheet
+      (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991};
       \citealp{Marshall.etal.2000}), the ubiquitously mountainous topography
@@
       where cohesion, $c_0$, is assumed to be zero. The friction angle,
-      $\phi$, varies from 15 to 45\unit{\degree} as a~piecewise-linear
-      function of modern bed elevation, with the lowest value occuring below
-      modern sea level (0\,\unit{m\,a.s.l.}) and the highest value occuring
-      above the generalised elevation of the highest shorelines
-      \citep[200\,\unit{m\,a.s.l.},][Fig.~5]{Clague.1981}, thus accounting
-      for a~weakening of till associated with the presence of marine
-      sediments (cf. \citealp{Martin.etal.2011};
+      $\phi$, varies from 15 to 45\unit{\degree} as a~piecewise-linear function
+      of modern bed elevation, with the lowest value occuring below modern sea
+      level (0\,\unit{m\,a.s.l.}) and the highest
+      value occuring above the generalised elevation of the highest shorelines
+      \citep[200\,\unit{m\,a.s.l.},][Fig.~5]{Clague.1981}, thus accounting for
+      a~weakening of till associated with the presence of marine sediments (cf.
+      \citealp{Martin.etal.2011};
       \citealp[Supplement]{Aschwanden.etal.2013};
@@
       from zero to $W_{\text{max}}=2$\,\unit{m}, a~threshold above which
-      additional melt water is assumed to drain off
-      instantaneously. Finally, the bedrock topography responds to ice load
-      following a~bedrock deformation model that includes local isostasy,
-      elastic lithosphere flexure and viscous astenosphere deformation in an
-      infinite half-space \citep{Lingle.Clark.1985,Bueler.etal.2007}.
-      A~relatively low viscosity value of $\nu_{m} =
-      1\times10^{19}$\,\unit{Pa\,s} is used for the astenosphere
-      (Table~\ref{tab:params}) in accordance with the results from regional
-      glacial isostatic adjustment modelling at the northern Cascadia
+      additional melt water is assumed to drain off instantaneously. Finally,
+      the bedrock topography responds to ice load following a~bedrock
+      deformation model that includes local isostasy, elastic lithosphere
+      flexure and viscous astenosphere deformation in an infinite half-space
+      \citep{Lingle.Clark.1985,Bueler.etal.2007}. A~relatively low viscosity
+      value of $\nu_{m} = 1\times10^{19}$\,\unit{Pa\,s} is used for the
+      astenosphere (Table~\ref{tab:params}) in accordance with the results from
+      regional glacial isostatic adjustment modelling at the northern Cascadia
       subduction zone \citep{James.etal.2009}.
@@
       Ice shelf calving is computed using a~double criterion. First,
-      a~physically-realistic calving flux is computed based on eigenvalues
-      of the horizontal strain rate tensor \citep{Winkelmann.etal.2011,
+      a~physically-realistic calving flux is computed based on eigenvalues of
+      the horizontal strain rate tensor \citep{Winkelmann.etal.2011,
       Levermann.etal.2012}. This allows floating ice to advance in confined
@@
       assuming a~normal temperature distribution of standard deviation
-      $\sigma$ aroung the expected value $T_{\mathrm{m}}$. It is expressed
-      by an error-function formulation \citep{Calov.Greve.2005},
+      $\sigma$ aroung the expected value $T_{\mathrm{m}}$. It is expressed by
+      an error-function formulation \citep{Calov.Greve.2005},
 \begin{align}
@@
       Climate forcing driving ice sheet simulations consists of
-      a~present-day monthly climatology, $\{T_{\mathrm{m}0}, P_{\mathrm{m}0}\}$, where
-      temperatures are modified by offset time series,
-      ${\Delta}T_{\text{TS}}$, and lapse-rate corrections,
+      a~present-day monthly climatology, $\{T_{\mathrm{m}0},
+      P_{\mathrm{m}0}\}$, where temperatures are modified by offset time
+      series, ${\Delta}T_{\text{TS}}$, and lapse-rate corrections,
       ${\Delta}T_{\text{LR}}$:
@@
 
-      Temperature offset time-series, ${\Delta}T_{\text{TS}}$, are derived
-      from palaeo-temperature proxy records from the Greenland Ice Core
-      Project \citep[GRIP,][]{Dansgaard.etal.1993}, the North Greenland Ice
-      Core Project \citep[NGRIP,][]{Andersen.etal.2004}, the European
-      Project for Ice Coring in Antarctica \citep[EPICA,][]
-      {Jouzel.etal.2007}, the Vostok ice core \citep{Petit.etal.1999}, and
-      Ocean Drilling Program (ODP) sites 1012 and 1020, both located off the
-      coast of California \citep{Herbert.etal.2001}. Palaeo-temperature
-      anomalies from the GRIP and NGRIP records were calculated from oxygen
-      isotope (\chem{\delta^{18}O}) measurements using a~quadratic equation
-      \citep{Johnsen.etal.1995},
+      Temperature offset time-series, ${\Delta}T_{\text{TS}}$, are derived from
+      palaeo-temperature proxy records from the Greenland Ice Core Project
+      \citep[GRIP,][]{Dansgaard.etal.1993}, the North Greenland Ice Core
+      Project \citep[NGRIP,][]{Andersen.etal.2004}, the European Project for
+      Ice Coring in Antarctica \citep[EPICA,][] {Jouzel.etal.2007}, the Vostok
+      ice core \citep{Petit.etal.1999}, and Ocean Drilling Program (ODP) sites
+      1012 and 1020, both located off the coast of California
+      \citep{Herbert.etal.2001}. Palaeo-temperature anomalies from the GRIP and
+      NGRIP records were calculated from oxygen isotope (\chem{\delta^{18}O})
+      measurements using a~quadratic equation \citep{Johnsen.etal.1995},
 \begin{align}
@@
 
-      Finally, lapse-rate corrections, ${\Delta}T_{\text{LR}}$, are computed
-      as a~function of ice surface elevation, $s$, using the NARR surface
+      Finally, lapse-rate corrections, ${\Delta}T_{\text{LR}}$, are computed as
+      a~function of ice surface elevation, $s$, using the NARR surface
       geopotential height invariant field as a~reference topography,
@@
       one hand, and for differences between the basal topography of the ice
-      flow model, $b$, and the NARR reference topography, $b_{\text{ref}}$,
-      on the other hand. All simulations use an annual temperature lapse
-      rate of $\gamma = 6\,\unit{K\,km^{-1}}$. In the rest of this paper,
-      we refer to different model runs by the name of the proxy record used
-      for the palaeo-temperature forcing.
+      flow model, $b$, and the NARR reference topography, $b_{\text{ref}}$, on
+      the other hand. All simulations use an annual temperature lapse rate of
+      $\gamma = 6\,\unit{K\,km^{-1}}$. In the rest of this paper, we refer to
+      different model runs by the name of the proxy record used for the
+      palaeo-temperature forcing.
 
@@
       Despite large differences in the input climate forcing
-      (Fig.~\ref{fig:lr_ts}, upper panel), model output presents
-      consistent features that can be observed across the range of forcing
-      data used. In all simulations, modelled ice volumes remain relatively
-      low during most of the glacial cycle, except during two major glacial
-      events which occur between 61.9 and 56.5\,\unit{ka} during MIS~4, and
-      between 23.2 and 16.8\,\unit{ka} during MIS~2
-      (Fig.~\ref{fig:lr_ts}, lower panel). An ice volume minimum is
-      consistently reached between 53.0 and 41.3\,\unit{ka} during
-      MIS~3. However, the magnitude and precise timing of these three events
-      depend significantly on the choice of proxy record used to derive
+      (Fig.~\ref{fig:lr_ts}, upper panel), model output presents consistent
+      features that can be observed across the range of forcing data used. In
+      all simulations, modelled ice volumes remain relatively low during most
+      of the glacial cycle, except during two major glacial events which occur
+      between 61.9 and 56.5\,\unit{ka} during MIS~4, and between 23.2 and
+      16.8\,\unit{ka} during MIS~2 (Fig.~\ref{fig:lr_ts}, lower panel). An ice
+      volume minimum is consistently reached between 53.0 and 41.3\,\unit{ka}
+      during MIS~3. However, the magnitude and precise timing of these three
+      events depend significantly on the choice of proxy record used to derive
       a~time-dependent climate forcing (Table~\ref{tab:extrema}).
@@
       variability throughout the simulation length, resulting in lower
-      modelled ice volumes during MIS~4 and larger ice volumes during
-      MIS~3. The NGRIP climate forcing is the only one that results in
-      a~larger ice volume during MIS~4 than during MIS~2.
+      modelled ice volumes during MIS~4 and larger ice volumes during MIS~3.
+      The NGRIP climate forcing is the only one that results in a~larger ice
+      volume during MIS~4 than during MIS~2.
 
@@
       palaeo-temperature record. Modelled ice volume minima spread over
-      a~wide range between 1.69 and
-      2.88\,\unit{m}\,s.l.e. (Table~\ref{tab:extrema}).
+      a~wide range between 1.69 and 2.88\,\unit{m}\,s.l.e.
+      (Table~\ref{tab:extrema}).
 
       Modelled ice sheet geometries during the LGM (MIS~2;
-      Fig.~\ref{fig:lr_maps}, lower panels) invariably include
-      a~ca. 1500\,\unit{km}-long central divide above 3000\,\unit{m\,a.s.l.}
+      Fig.~\ref{fig:lr_maps}, lower panels) invariably include a~ca.
+      1500\,\unit{km}-long central divide above 3000\,\unit{m\,a.s.l.}
       located along the spine of the Rocky Mountains. Although the
@@
       this early deglaciation is caused by an early warming present in the
-      alkenone palaeo-temperature reconstructions
-      (Fig.~\ref{fig:lr_ts}, upper panel;
-      \citealp[Fig.~3]{Herbert.etal.2001}). However, this early warming is
-      a~local effect, corresponding to a~weakening of the California current
+      alkenone palaeo-temperature reconstructions (Fig.~\ref{fig:lr_ts}, upper
+      panel; \citealp[Fig.~3]{Herbert.etal.2001}). However, this early warming
+      is a~local effect, corresponding to a~weakening of the California current
       \citep{Herbert.etal.2001}. The California current, driving cold waters
-      southwards along the south-western coast of North America, has been
-      shown to have weakened during each peak of global glaciation (in
-      SPECMAP) during the past 550\,\unit{ka}, including the LGM, resulting
-      in paradoxically warmer sea-surface temperatures at the locations of
-      the ODP~1012 and ODP~1020 sites \citep{Herbert.etal.2001}.
+      southwards along the south-western coast of North America, has been shown
+      to have weakened during each peak of global glaciation (in SPECMAP)
+      during the past 550\,\unit{ka}, including the LGM, resulting in
+      paradoxically warmer sea-surface temperatures at the locations of the
+      ODP~1012 and ODP~1020 sites \citep{Herbert.etal.2001}.
 
@@
       the NGRIP palaeo-temperature record depicts an early onset of
-      deglaciation (Fig.~\ref{fig:lr_ts}) following its last glacial
-      volume maximum (22.9\,\unit{ka}, Table~\ref{tab:extrema}) attained
-      about 6\,\unit{ka} earlier than dated evidence of the LGM advance.
-      There is a~fair agreement between the EPICA and Vostok
-      palaeo-temperature records, resulting in only small differences
-      between the simulations driven by those records. These differences are
-      not subject to further analysis further; instead we focus on
-      simulations forced by palaeo-temperature records from the GRIP and
-      EPICA ice cores that appear to produce the most realistic
-      reconstructions of regional glaciation history, yet bearing
+      deglaciation (Fig.~\ref{fig:lr_ts}) following its last glacial volume
+      maximum (22.9\,\unit{ka}, Table~\ref{tab:extrema}) attained about
+      6\,\unit{ka} earlier than dated evidence of the LGM advance. There is
+      a~fair agreement between the EPICA and Vostok palaeo-temperature records,
+      resulting in only small differences between the simulations driven by
+      those records. These differences are not subject to further analysis
+      further; instead we focus on simulations forced by palaeo-temperature
+      records from the GRIP and EPICA ice cores that appear to produce the most
+      realistic reconstructions of regional glaciation history, yet bearing
       significant disparities in model output. To allow for a~more detailed
@@
       which likely resulted in a~decrease of precipitation and, therefore,
-      a~decrease of accumulation in the interior. Secondly, latent warming
-      of the moisture-depleted air parcels flowing over this enhanced
+      a~decrease of accumulation in the interior. Secondly, latent warming of
+      the moisture-depleted air parcels flowing over this enhanced
       topography could have resulted in an inflow of potentially warmer air
@@
       5.04\,\unit{m}\,s.l.e. (61.9\,\unit{ka}) in the EPICA simulation,
-      corresponding to respectively 90 and 57\,\unit{\%} of modelled MIS~2
-      ice volumes.
+      corresponding to respectively 90 and 57\,\unit{\%} of modelled MIS~2 ice
+      volumes.
 
@@
       simulations, we use numerical modelling output to compute durations of
-      ice cover throughout the last glacial cycle (Fig.~\ref{fig:hr_geom_duration}).
+      ice cover throughout the last glacial cycle
+      (Fig.~\ref{fig:hr_geom_duration}).
 
@@
       operation for at most 32\,\unit{ka}, which is about a~third of the
-      timespan of the last glacial cycle (Fig.~\ref{fig:hr_geom_duration}, hatched
-      areas). However, except for its margins in the Pacific Ocean and in
-      the northern foothills of the Alaska Range, the maximum extent of the
+      timespan of the last glacial cycle (Fig.~\ref{fig:hr_geom_duration},
+      hatched areas). However, except for its margins in the Pacific Ocean and
+      in the northern foothills of the Alaska Range, the maximum extent of the
       ice sheet is attained for a~much shorter period of time of only few
@@
       illustrates that the maximum extents of the modelled ice sheet during
-      MIS~4 and MIS~2 were both short-lived and therefore out of balance
-      with contemporary climate.
+      MIS~4 and MIS~2 were both short-lived and therefore out of balance with
+      contemporary climate.
 
@@
       where modelled glaciers are confined to its foothills during the
-      entire simulation period (Fig.~\ref{fig:hr_geom_duration}, AR). This apparent
-      insensitivity of modelled glacial extent to temperature fluctuations
-      results from a~combination of low precipitation, high summer
+      entire simulation period (Fig.~\ref{fig:hr_geom_duration}, AR). This
+      apparent insensitivity of modelled glacial extent to temperature
+      fluctuations results from a~combination of low precipitation, high summer
       temperature and large temperature standard deviation (PDD SD) in the
@@
       potentially explain the local distribution of glacial deposits, which
-      indicates that glaciers flowing on the northern slope of the Alaska
-      Range have remained small throughout the Pleistocene
+      indicates that glaciers flowing on the northern slope of the Alaska Range
+      have remained small throughout the Pleistocene
       \citep{Kaufman.Manley.2004}.
@@
       centres, visible on the modelled ice cover duration maps
-      (Fig.~\ref{fig:hr_geom_duration}), are located over the Coast Mountains (CM),
-      the Columbia and Rocky mountains (CRM), the Skeena Mountains (SM), and
-      the Selwyn and Mackenzie mountains (SMKM). The Wrangell and Saint
-      Elias mountains, heavily glacierized at present, host an ice cap for
-      the entire length of both simulations, but that ice cap does not
-      appear to be a~major feed to the Cordilleran ice sheet
+      (Fig.~\ref{fig:hr_geom_duration}), are located over the Coast Mountains
+      (CM), the Columbia and Rocky mountains (CRM), the Skeena Mountains (SM),
+      and the Selwyn and Mackenzie mountains (SMKM). The Wrangell and Saint
+      Elias mountains, heavily glacierized at present, host an ice cap for the
+      entire length of both simulations, but that ice cap does not appear to be
+      a~major feed to the Cordilleran ice sheet
       (Fig.~\ref{fig:hr_geom_duration}, WSEM). Although the Coast, Skeena and
       Columbia and Rocky mountains (CM, SM, CRM) are covered by mountain
-      glaciers for most of the last glacial cycle, providing durable
-      nucleation centres for an ice sheet, this is not the case for the
-      Selwyn and Mackenzie mountains (SMKM), where ice cover on the highest
-      peaks is limited to a~small fraction of the last glacial cycle. In
-      other words, the Selwyn and Mackenzie mountains only appear as
-      a~secondary ice-dispersal centre during the coldest periods of the
-      last glacial cycle. The Northern Rocky Mountains
-      (Fig.~\ref{fig:hr_geom_duration}, NRM) do not act as a~nucleation centre, but
-      rather as a~pinning point for the Cordilleran ice sheet margin coming
-      from the west.
+      glaciers for most of the last glacial cycle, providing durable nucleation
+      centres for an ice sheet, this is not the case for the Selwyn and
+      Mackenzie mountains (SMKM), where ice cover on the highest peaks is
+      limited to a~small fraction of the last glacial cycle. In other words,
+      the Selwyn and Mackenzie mountains only appear as a~secondary
+      ice-dispersal centre during the coldest periods of the last glacial
+      cycle. The Northern Rocky Mountains (Fig.~\ref{fig:hr_geom_duration},
+      NRM) do not act as a~nucleation centre, but rather as a~pinning point for
+      the Cordilleran ice sheet margin coming from the west.
 
@@
       its predominance over the other ice-dispersal centres
-      (Figs.~\ref{fig:lr_maps} and~\ref{fig:hr_geom_duration}, SM). Regardless of
-      the applied forcing, this ice cap appears to survive MIS~3
+      (Figs.~\ref{fig:lr_maps} and~\ref{fig:hr_geom_duration}, SM). Regardless
+      of the applied forcing, this ice cap appears to survive MIS~3
       (Fig.~\ref{fig:lr_maps}, middle panels), and serves as a~nucleation
-      centre at the onset of the glacial readvance towards the LGM
-      (MIS~2). This situation appear similar to the neighbouring Laurentide
-      ice sheet, for which the importance of residual ice for the glacial
-      history leading up to the LGM has been illustrated by the MIS~3
-      residual ice bodies in northern and eastern Canada as nucleation
-      centres for a~much more extensive MIS~2 configuration
-      \citep{Kleman.etal.2010}.
+      centre at the onset of the glacial readvance towards the LGM (MIS~2).
+      This situation appear similar to the neighbouring Laurentide ice sheet,
+      for which the importance of residual ice for the glacial history leading
+      up to the LGM has been illustrated by the MIS~3 residual ice bodies in
+      northern and eastern Canada as nucleation centres for a~much more
+      extensive MIS~2 configuration \citep{Kleman.etal.2010}.
 
@@
       The modelled total ice volume corresponding to these persistent
-      ice-dispersal centres attains a~minimum of
-      1.52\,\unit{m}\,s.l.e. (42.9\,\unit{ka}) in the GRIP simulation and of
-      2.68\,\unit{m}\,s.l.e. (51.8\,\unit{ka}) in the EPICA simulation,
-      corresponding to respectively 19 and 31\,\unit{\%} of the MIS~2 ice
-      volumes.
+      ice-dispersal centres attains a~minimum of 1.52\,\unit{m}\,s.l.e.
+      (42.9\,\unit{ka}) in the GRIP simulation and of 2.68\,\unit{m}\,s.l.e.
+      (51.8\,\unit{ka}) in the EPICA simulation, corresponding to respectively
+      19 and 31\,\unit{\%} of the MIS~2 ice volumes.
 
@@
 
-      A~correlation is observed between the modelled duration of warm based
-      ice cover (Fig.~\ref{fig:hr_geom_warmbase}) and the degree of glacial
-      modification of the landscape (mainly in terms of the development of
-      deep glacial valleys and troughs). We find evidence for this on the
-      slopes of the Coast Mountains, the the Columbia and Rocky mountains,
-      the Wrangell and Saint Elias mountains, and radiating off the Skeena
-      Mountains (Figs.~\ref{fig:hr_geom_duration} and \ref{fig:hr_geom_warmbase};
-      \citealp[Fig.~2]{Kleman.etal.2010}). The Skeena Mountains, for
-      example, indeed bear a~strong glacial imprint that indicates ice
-      drainage in a~system of distinct glacial troughs emanating in a~radial
-      pattern from the centre of the mountain range
-      \citep[Fig.~2]{Kleman.etal.2010}. We suggest that persistent ice cover
-      (Fig.~\ref{fig:hr_geom_duration}) associated with basal ice temperatures at
-      the pressure melting point (Figs.~\ref{fig:hr_geom_warmbase}
-      and~\ref{fig:hr_geom_warmfrac}) explains the large-scale glacial erosional
-      imprint on the landscape. A~well-developed network of glacial valleys
-      running to the north-west on the west slope of the Selwyn and
-      Mackenzie Mountains (\citealp[Fig.~2]{Kleman.etal.2010};
-      \citealp[Fig.~8] {Stroeven.etal.2010}) is modelled to have hosted
-      predominantly warm-based ice (Fig.~\ref{fig:hr_geom_warmfrac}). However,
-      because it is only glaciated for a~short fraction of the last glacial
-      cycle in our simulations (Fig.~\ref{fig:hr_geom_duration}), this perhaps
-      indicates that the observed landscape pattern originates from multiple
-      glacial cycles and witnesses an increased relative importance of the
-      Selwyn and Mackenzie mountains ice dispersal centre
+      A~correlation is observed between the modelled duration of warm based ice
+      cover (Fig.~\ref{fig:hr_geom_warmbase}) and the degree of glacial
+      modification of the landscape (mainly in terms of the development of deep
+      glacial valleys and troughs). We find evidence for this on the slopes of
+      the Coast Mountains, the the Columbia and Rocky mountains, the Wrangell
+      and Saint Elias mountains, and radiating off the Skeena Mountains
+      (Figs.~\ref{fig:hr_geom_duration} and \ref{fig:hr_geom_warmbase};
+      \citealp[Fig.~2]{Kleman.etal.2010}). The Skeena Mountains, for example,
+      indeed bear a~strong glacial imprint that indicates ice drainage in
+      a~system of distinct glacial troughs emanating in a~radial pattern from
+      the centre of the mountain range \citep[Fig.~2]{Kleman.etal.2010}. We
+      suggest that persistent ice cover (Fig.~\ref{fig:hr_geom_duration})
+      associated with basal ice temperatures at the pressure melting point
+      (Figs.~\ref{fig:hr_geom_warmbase} and~\ref{fig:hr_geom_warmfrac})
+      explains the large-scale glacial erosional imprint on the landscape.
+      A~well-developed network of glacial valleys running to the north-west on
+      the west slope of the Selwyn and Mackenzie Mountains
+      (\citealp[Fig.~2]{Kleman.etal.2010}; \citealp[Fig.~8]
+      {Stroeven.etal.2010}) is modelled to have hosted predominantly warm-based
+      ice (Fig.~\ref{fig:hr_geom_warmfrac}). However, because it is only
+      glaciated for a~short fraction of the last glacial cycle in our
+      simulations (Fig.~\ref{fig:hr_geom_duration}), this perhaps indicates
+      that the observed landscape pattern originates from multiple glacial
+      cycles and witnesses an increased relative importance of the Selwyn and
+      Mackenzie mountains ice dispersal centre
       (Fig.~\ref{fig:hr_geom_duration}, SMKM), prior to the Late Pleistocene
@@
       The modelled distribution of warm-based ice cover
-      (Figs.~\ref{fig:hr_geom_warmbase} and~\ref{fig:hr_geom_warmfrac}) is inevitably
-      affected by our assumption of a~constant, 70\,\unit{mW\,m^{-2}}
-      geothermal heat flux at 3\,\unit{km} depth
-      (Sect.~\ref{sec:icedyn}). However, the Skeena Mountains and the area
-      west of the Mackenzie Mountains experience higher-than-average
-      geothermal heat flux with measured values of ca.~80 and
-      ca.~100\,\unit{mW\,m^{-2}} \citep{Blackwell.Richards.2004}. We can
-      therefore expect even longer durations of warm-based ice cover for
-      these areas if we were to include spatially variable geothermal
-      forcing in our Cordilleran ice sheet simulations.
+      (Figs.~\ref{fig:hr_geom_warmbase} and~\ref{fig:hr_geom_warmfrac}) is
+      inevitably affected by our assumption of a~constant,
+      70\,\unit{mW\,m^{-2}} geothermal heat flux at 3\,\unit{km} depth
+      (Sect.~\ref{sec:icedyn}). However, the Skeena Mountains and the area west
+      of the Mackenzie Mountains experience higher-than-average geothermal heat
+      flux with measured values of ca.~80 and ca.~100\,\unit{mW\,m^{-2}}
+      \citep{Blackwell.Richards.2004}. We can therefore expect even longer
+      durations of warm-based ice cover for these areas if we were to include
+      spatially variable geothermal forcing in our Cordilleran ice sheet
+      simulations.
 
@@
 
-      Modelled patterns of ice sheet retreat are relatively consistent
-      between the two simulations (Figs.~\ref{fig:hr_geom_deglacageshots}
-      and~\ref{fig:hr_geom_deglacage}). The southern sector of the modelling domain,
-      including the Puget Lowland, the Coast Mountains, the Columbia and and
-      Rocky mountains, and the Interior Plateau of British Columbia, becomes
-      completely deglaciated by 10\,\unit{ka}, whereas a~significant ice
-      cover remains over the Skeena, the Selwyn and Mackenzie, and the
+      Modelled patterns of ice sheet retreat are relatively consistent between
+      the two simulations (Figs.~\ref{fig:hr_geom_deglacageshots}
+      and~\ref{fig:hr_geom_deglacage}). The southern sector of the modelling
+      domain, including the Puget Lowland, the Coast Mountains, the Columbia
+      and and Rocky mountains, and the Interior Plateau of British Columbia,
+      becomes completely deglaciated by 10\,\unit{ka}, whereas a~significant
+      ice cover remains over the Skeena, the Selwyn and Mackenzie, and the
       Wrangell and Saint Elias mountains in the northern sector of the
-      modelling domain. After 10\,\unit{ka}, deglaciation continues to
-      proceed across the Liard Lowland with a~radial ice margin retreat
-      towards the surrounding mountain ranges, consistent with the regional
-      melt water record of the last deglaciation
-      \citep{Margold.etal.2013}. Remaining ice continues to decay by
-      retreating towards the Selwyn and MacKenzie, and Skeena mountains.
-      The last remnants of the Cordilleran ice sheet finally disappear from
-      the Skeena Mountains at 6.5\,\unit{ka} (GRIP) and 6.1\,\unit{ka}
-      (EPICA).
+      modelling domain. After 10\,\unit{ka}, deglaciation continues to proceed
+      across the Liard Lowland with a~radial ice margin retreat towards the
+      surrounding mountain ranges, consistent with the regional melt water
+      record of the last deglaciation \citep{Margold.etal.2013}. Remaining ice
+      continues to decay by retreating towards the Selwyn and MacKenzie, and
+      Skeena mountains. The last remnants of the Cordilleran ice sheet finally
+      disappear from the Skeena Mountains at 6.5\,\unit{ka} (GRIP) and
+      6.1\,\unit{ka} (EPICA).
 
@@
       may have been disconnected from the waning Cordilleran ice sheet, the
-      Finlay River area in the Omenica Mountains (Fig.~\ref{fig:hr_geom_deglacage}, OM)
-      presents a~different kind of evidence. There, sharp-crested moraines
-      indicate a~late-glacial readvance of local alpine glaciers and, more
-      importantly, their interaction with larger, lingering remnants of the
-      main body of the Cordilleran ice sheet in the valleys
-      \citep{Lakeman.etal.2008}. Additional evidence for late-glacial alpine
-      glacier readvances includes moraines in the eastern Coast Mountains,
-      and the Columbia and Rocky mountains \citep{Osborn.Gerloff.1997,
-      Menounos.etal.2008}.
+      Finlay River area in the Omenica Mountains
+      (Fig.~\ref{fig:hr_geom_deglacage}, OM) presents a~different kind of
+      evidence. There, sharp-crested moraines indicate a~late-glacial readvance
+      of local alpine glaciers and, more importantly, their interaction with
+      larger, lingering remnants of the main body of the Cordilleran ice sheet
+      in the valleys \citep{Lakeman.etal.2008}. Additional evidence for
+      late-glacial alpine glacier readvances includes moraines in the eastern
+      Coast Mountains, and the Columbia and Rocky mountains
+      \citep{Osborn.Gerloff.1997, Menounos.etal.2008}.
 
@@
       distribution, and to identify the potential climatic triggers
-      \citep{Menounos.etal.2008}, it is interesting to note that the
-      simulation driven by the GRIP record produces a~late-glacial readvance
-      in the Coast Mountains and in the Columbia and Rocky Mountains
-      (Fig.~\ref{fig:hr_geom_deglacage}, left panel). In addition to matching the
-      location of some local readvances, the GRIP-driven simulation shows
-      that a~large remnant of the decaying ice sheet may still have existed
-      at the time of this late-glacial readvance. In contrast, the
-      EPICA-driven simulation produces a~nearly-continuous deglaciation with
-      only a~tightly restricted late-glacial readvance on the western slopes
-      of the Saint Elias and the Coast mountains (Fig.~\ref{fig:hr_geom_deglacage},
-      right panel).
+      \citep{Menounos.etal.2008}, it is interesting to note that the simulation
+      driven by the GRIP record produces a~late-glacial readvance in the Coast
+      Mountains and in the Columbia and Rocky Mountains
+      (Fig.~\ref{fig:hr_geom_deglacage}, left panel). In addition to matching
+      the location of some local readvances, the GRIP-driven simulation shows
+      that a~large remnant of the decaying ice sheet may still have existed at
+      the time of this late-glacial readvance. In contrast, the EPICA-driven
+      simulation produces a~nearly-continuous deglaciation with only a~tightly
+      restricted late-glacial readvance on the western slopes of the Saint
+      Elias and the Coast mountains (Fig.~\ref{fig:hr_geom_deglacage}, right
+      panel).
 
@@
 
-      Because a~general tenant in glacial geomorphology is that the majority
-      of landforms (lineations and eskers) are part of the deglacial
-      envelope \citep[terminology from][]{Kleman.etal.2006}, having been
-      formed close inside the retreating margin of ice sheets
-      \citep{Boulton.Clark.1990, Kleman.etal.1997, Kleman.etal.2010}, we
-      present maps of basal flow directions immediately preceding
-      deglaciation or at the time of cessation of sliding inside
-      a~cold-based retreating margin (Fig.~\ref{fig:hr_geom_lastflow}). The modelled
-      deglacial flow patterns are mostly consistent between the GRIP and
-      EPICA simulations. They depict an active ice sheet retreat in the
-      peripheral areas, followed by stagnant ice decay in some of the
-      interior regions. Several parts of the modelling domain do not
-      experience any basal sliding throughout the deglaciation phase
-      (Fig.~\ref{fig:hr_geom_lastflow}, hatched areas). This notably includes parts
-      of the Interior Plateau in British Columbia, major portions of the
-      Alaskan sector of the ice sheet, and a~tortuous ribbon running from
-      the Northern Rocky Mountains over the Skeena and Selwyn Mountains and
-      into the Mackenzie Mountains.
+      Because a~general tenant in glacial geomorphology is that the majority of
+      landforms (lineations and eskers) are part of the deglacial envelope
+      \citep[terminology from][]{Kleman.etal.2006}, having been formed close
+      inside the retreating margin of ice sheets \citep{Boulton.Clark.1990,
+      Kleman.etal.1997, Kleman.etal.2010}, we present maps of basal flow
+      directions immediately preceding deglaciation or at the time of cessation
+      of sliding inside a~cold-based retreating margin
+      (Fig.~\ref{fig:hr_geom_lastflow}). The modelled deglacial flow patterns
+      are mostly consistent between the GRIP and EPICA simulations. They depict
+      an active ice sheet retreat in the peripheral areas, followed by stagnant
+      ice decay in some of the interior regions. Several parts of the modelling
+      domain do not experience any basal sliding throughout the deglaciation
+      phase (Fig.~\ref{fig:hr_geom_lastflow}, hatched areas). This notably
+      includes parts of the Interior Plateau in British Columbia, major
+      portions of the Alaskan sector of the ice sheet, and a~tortuous ribbon
+      running from the Northern Rocky Mountains over the Skeena and Selwyn
+      Mountains and into the Mackenzie Mountains.
 
@@
       \citealp[Fig.~1.12]{Clague.1989}; \citealp[Fig.~2]{Kleman.etal.2010})
-      show similarities with the patterns of deglacial ice flow from
-      numerical modelling (Fig.~\ref{fig:hr_geom_lastflow}). In the northern half of
-      the modelling domain, modelled deglacial flow directions depict an
-      active downhill flow as the last remnants of the ice sheet retreat
-      towards mountain ranges. Converging deglacial flow patterns in the
-      Liard Lowland, for instance (Fig.~\ref{fig:hr_geom_lastflow}), closely
-      resemble the pattern indicated by glacial lineations
+      show similarities with the patterns of deglacial ice flow from numerical
+      modelling (Fig.~\ref{fig:hr_geom_lastflow}). In the northern half of the
+      modelling domain, modelled deglacial flow directions depict an active
+      downhill flow as the last remnants of the ice sheet retreat towards
+      mountain ranges. Converging deglacial flow patterns in the Liard Lowland,
+      for instance (Fig.~\ref{fig:hr_geom_lastflow}), closely resemble the
+      pattern indicated by glacial lineations
       \citep[Fig.~2]{Margold.etal.2013}.
@@
       (Fig.~\ref{fig:hr_geom_deglacage}), a~pattern which is validated by the
-      geomorphological and stratigraphical record for ancient pro-glacial
-      lakes dammed by the retreating ice sheet
-      \citep{Perkins.Brennand.2014}. However, the two simulations differ in
-      the mode of retreat. The GRIP simulation yields an active retreat with
-      basal sliding towards the ice margin to the south, whereas the EPICA
-      simulation produces negligible basal sliding on the plateau during
-      deglaciation (Fig.~\ref{fig:hr_geom_lastflow}). Yet, the Interior Plateau also
-      hosts an impressive set of glacial lineations which indicate
-      a~substantial eastwards flow component of the Cordilleran ice sheet
+      geomorphological and stratigraphical record for ancient pro-glacial lakes
+      dammed by the retreating ice sheet \citep{Perkins.Brennand.2014}.
+      However, the two simulations differ in the mode of retreat. The GRIP
+      simulation yields an active retreat with basal sliding towards the ice
+      margin to the south, whereas the EPICA simulation produces negligible
+      basal sliding on the plateau during deglaciation
+      (Fig.~\ref{fig:hr_geom_lastflow}). Yet, the Interior Plateau also hosts
+      an impressive set of glacial lineations which indicate a~substantial
+      eastwards flow component of the Cordilleran ice sheet
       \citep{Prest.etal.1968, Kleman.etal.2010}. This Interior Plateau
-      lineation set could therefore present a~smoking gun for the
-      reliability of the presented model results. One explanation for the
-      incongruent results could be that the missing feedback mechanisms
-      between ice sheet topography and regional climate resulted in
-      a~modelled ice divide of the LGM ice sheet beingtoo far to the east
-      (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
-      \citealp{Seguinot.etal.2014}). A~more westerly-located LGM ice divide
-      would certainly result in a~different deglacial flow pattern over the
-      Interior Plateau. However, a~more westerly-positioned LGM ice divide
-      would certainly be associated with a~thinner ice sheet than that
-      modelled here. Decreased ice thickness would not promote warm-based
-      conditions but, on the contrary, enlarge the region of negligible
-      basal sliding (Fig.~\ref{fig:hr_geom_lastflow}). Thus, a~second explanation
-      for the incongruent results could be that the Interior Plateau
-      lineation system predates deglaciation ice flow, as perhaps indicated
-      by some eskers that appear incompatible with these glacial lineations
-      \citep[Fig.~9]{Margold.etal.2013a}. Finally, a~third explanation could
-      be that local geothermal heat associated with volcanic activity on the
-      Interior Plateau could have triggered the basal sliding
+      lineation set could therefore present a~smoking gun for the reliability
+      of the presented model results. One explanation for the incongruent
+      results could be that the missing feedback mechanisms between ice sheet
+      topography and regional climate resulted in a~modelled ice divide of the
+      LGM ice sheet beingtoo far to the east (Sect.~\ref{sec:mis2};
+      Fig.~\ref{fig:mis2}; \citealp{Seguinot.etal.2014}). A~more
+      westerly-located LGM ice divide would certainly result in a~different
+      deglacial flow pattern over the Interior Plateau. However, a~more
+      westerly-positioned LGM ice divide would certainly be associated with
+      a~thinner ice sheet than that modelled here. Decreased ice thickness
+      would not promote warm-based conditions but, on the contrary, enlarge the
+      region of negligible basal sliding (Fig.~\ref{fig:hr_geom_lastflow}).
+      Thus, a~second explanation for the incongruent results could be that the
+      Interior Plateau lineation system predates deglaciation ice flow, as
+      perhaps indicated by some eskers that appear incompatible with these
+      glacial lineations \citep[Fig.~9]{Margold.etal.2013a}. Finally, a~third
+      explanation could be that local geothermal heat associated with volcanic
+      activity on the Interior Plateau could have triggered the basal sliding
       \citep[cf. Greenland ice sheet;][]{Fahnestock.etal.2001}.
@@
       The modelled deglaciation of the Interior Plateau of British Columbia
-      consists of a~rapid northwards retreat (Fig.~\ref{fig:hr_geom_deglacage}) of
-      southwards-flowing non-sliding ice lobes (Fig.~\ref{fig:hr_geom_lastflow})
-      positioned in-between deglaciated mountain ranges
-      (Figs.~\ref{fig:hr_pf_grip} and~\ref{fig:hr_pf_epica}). This
-      result appears compatible with the prevailing conceptual model of
-      deglaciation of central British Columbia, in which mountain ranges
-      emerge from the ice before the plateau
-      \citep[Fig.~7]{Fulton.1991}. However, due to different topographic and
-      climatic conditions, our simulations produce different deglaciation
-      patterns in the northern half of the model domain, indicating that
-      this conceptual model may not be applied to the entire area formerly
-      covered by the Cordilleran ice sheet.
+      consists of a~rapid northwards retreat (Fig.~\ref{fig:hr_geom_deglacage})
+      of southwards-flowing non-sliding ice lobes
+      (Fig.~\ref{fig:hr_geom_lastflow}) positioned in-between deglaciated
+      mountain ranges (Figs.~\ref{fig:hr_pf_grip} and~\ref{fig:hr_pf_epica}).
+      This result appears compatible with the prevailing conceptual model of
+      deglaciation of central British Columbia, in which mountain ranges emerge
+      from the ice before the plateau \citep[Fig.~7]{Fulton.1991}. However, due
+      to different topographic and climatic conditions, our simulations produce
+      different deglaciation patterns in the northern half of the model domain,
+      indicating that this conceptual model may not be applied to the entire
+      area formerly covered by the Cordilleran ice sheet.
 
@@
       and MIS~2 (23.2--16.8\,\unit{ka}, 8.40--8.91\,\unit{m}\,s.l.e.), two
-      periods corresponding to documented extensive glaciations. This result
-      is independent of the palaeo-temperature record used among the six
-      selected for this study, and thus can be regarded as a~robust model
-      output, which broadly matches geological evidence. However, the timing
-      of the two glaciation peaks depends sensitively on which climate
-      record is used to drive the model. The timing of the LGM is best
-      reproduced by the EPICA and Vostok Antarctic ice core records. It
-      occurs about 2\,\unit{ka} too early in the simulation forced by the
-      GRIP ice core record, and occurs even earlier in all other
-      simulations. The mismatch is largest for the two Northwest Pacific ODP
-      palaeo-temperature records, which are affected by the weakening of the
-      California current during the LGM.
+      periods corresponding to documented extensive glaciations. This result is
+      independent of the palaeo-temperature record used among the six selected
+      for this study, and thus can be regarded as a~robust model output, which
+      broadly matches geological evidence. However, the timing of the two
+      glaciation peaks depends sensitively on which climate record is used to
+      drive the model. The timing of the LGM is best reproduced by the EPICA
+      and Vostok Antarctic ice core records. It occurs about 2\,\unit{ka} too
+      early in the simulation forced by the GRIP ice core record, and occurs
+      even earlier in all other simulations. The mismatch is largest for the
+      two Northwest Pacific ODP palaeo-temperature records, which are affected
+      by the weakening of the California current during the LGM.
 
@@
       to be fed by the moisture intruding inland from the west through
-      a~topographic breach in the Coast Mountains. The Skeena ice cap acts
-      as the main nucleation centre for the glacial reavance towards the LGM
+      a~topographic breach in the Coast Mountains. The Skeena ice cap acts as
+      the main nucleation centre for the glacial reavance towards the LGM
       configuration. As indicated by persistent, warm-based ice in the
@@
 \begin{acknowledgements}
-  Foremost, we would like to thank Shawn Marshall for providing
-  a~detailed, constructive analysis of this study during J.~Seguinot's
-  PhD defence (September 2014). His comments were used to improve the
-  model set-up. We are very thankful to Constantine Khroulev, Ed
-  Bueler, and Andy Aschwanden for providing constant help and
-  development with PISM. This work was supported by the Swedish
-  Research Council~(VR) grant no.~2008-3449 to A.~P.~Stroeven and by
-  the German Academic Exchange Service~(DAAD) grant no.~50015537 and
-  a~Knut and Alice Wallenberg Foundation grant to
-  J.~Seguinot. Computer resources were provided by the Swedish
-  National Infrastructure for Computing (SNIC) allocation
-  no.~2013/1-159 and~2014/1-159 to A.~P.~Stroeven at the National
-  Supercomputing Center (NSC).
+      Foremost, we would like to thank Shawn Marshall for providing a~detailed,
+      constructive analysis of this study during J.~Seguinot's PhD defence
+      (September 2014). His comments were used to improve the model set-up. We
+      are very thankful to Constantine Khroulev, Ed Bueler, and Andy Aschwanden
+      for providing constant help and development with PISM. This work was
+      supported by the Swedish Research Council~(VR) grant no.~2008-3449 to
+      A.~P.~Stroeven and by the German Academic Exchange Service~(DAAD) grant
+      no.~50015537 and a~Knut and Alice Wallenberg Foundation grant to
+      J.~Seguinot. Computer resources were provided by the Swedish National
+      Infrastructure for Computing (SNIC) allocation no.~2013/1-159
+      and~2014/1-159 to A.~P.~Stroeven at the National Supercomputing Center
+      (NSC).
 \end{acknowledgements}

### 2016-01-17 cordillera-cycle.tex
Typeset: whitespace changes in manuscript.
Also one found one typo: aroung -> around.

@@
       Stea.etal.2011}. Nevertheless, recent oxygen isotope measurements from
-      Gulf of Alaska sediments reveal a~climatic evolution highly correlated
-      to that of Greenland during this period, including a~distinct Late
-      Glacial cold reversal between 14.1 and 11.7\,\unit{ka}
+      Gulf of Alaska sediments reveal a~climatic evolution highly correlated to
+      that of Greenland during this period, including a~distinct Late Glacial
+      cold reversal between 14.1 and 11.7\,\unit{ka}
       \citep{Praetorius.Mix.2014}.
@@
       evolution:
-
 \begin{itemize}
@@
 \end{itemize}
-
       Although numerical ice sheet modelling has been established as
@@
 \end{align}
-      relates the bed-parallel shear stresses, $\vec{\tau}_{\mathrm{b}}$, to the
-      sliding velocity, $\vec{v}_{\mathrm{b}}$ (Table~\ref{tab:params}). The yield
-      stress, $\tau_{\mathrm{c}}$, is modelled using the Mohr--Coulomb criterion,
+      relates the bed-parallel shear stresses, $\vec{\tau}_{\mathrm{b}}$, to
+      the sliding velocity, $\vec{v}_{\mathrm{b}}$ (Table~\ref{tab:params}).
+      The yield stress, $\tau_{\mathrm{c}}$, is modelled using the
+      Mohr--Coulomb criterion,
 \begin{align}
@@
       where cohesion, $c_0$, is assumed to be zero. The friction angle,
-      $\phi$, varies from 15 to 45\unit{\degree} as a~piecewise-linear function of
-      modern bed elevation, with the lowest value occuring below modern sea
+      $\phi$, varies from 15 to 45\unit{\degree} as a~piecewise-linear function
+      of modern bed elevation, with the lowest value occuring below modern sea
       level (0\,\unit{m} above sea level, \unit{m\,a.s.l.}) and the highest
-      value occuring above the generalised elevation of the highest
-      shorelines \citep[200\,\unit{m\,a.s.l.},][Fig.~5]{Clague.1981}, thus
-      accounting for a~weakening of till associated with the presence of
-      marine sediments (cf. \citealp{Martin.etal.2011};
+      value occuring above the generalised elevation of the highest shorelines
+      \citep[200\,\unit{m\,a.s.l.},][Fig.~5]{Clague.1981}, thus accounting for
+      a~weakening of till associated with the presence of marine sediments (cf.
+      \citealp{Martin.etal.2011};
       \citealp[Supplement]{Aschwanden.etal.2013};
@@
       (Table~\ref{tab:params}). The amount of water at the base, $W$, varies
-      from zero to $W_{\text{max}}=2$\,\unit{m}, a~threshold above which additional
-      melt water is assumed to drain off instantaneously. Finally, the
-      bedrock topography responds to ice load following a~bedrock
+      from zero to $W_{\text{max}}=2$\,\unit{m}, a~threshold above which
+      additional melt water is assumed to drain off instantaneously. Finally,
+      the bedrock topography responds to ice load following a~bedrock
       deformation model that includes local isostasy, elastic lithosphere
@@
       value of $\nu_{m} = 1\times10^{19}$\,\unit{Pa\,s} is used for the
-      astenosphere (Table~\ref{tab:params}) in accordance with the results
-      from regional glacial isostatic adjustment modelling at the northern
-      Cascadia subduction zone \citep{James.etal.2009}.
+      astenosphere (Table~\ref{tab:params}) in accordance with the results from
+      regional glacial isostatic adjustment modelling at the northern Cascadia
+      subduction zone \citep{James.etal.2009}.
 
@@
       assuming a~normal temperature distribution of standard deviation
-      $\sigma$ aroung the expected value $T_{\mathrm{m}}$. It is expressed by an
-      error-function formulation \citep{Calov.Greve.2005},
+      $\sigma$ around the expected value $T_{\mathrm{m}}$. It is expressed by
+      an error-function formulation \citep{Calov.Greve.2005},
 \begin{align}
@@
       Climate forcing driving ice sheet simulations consists of
-      a~present-day monthly climatology,
-$\{T_{\mathrm{m}0}, P_{\mathrm{m}0}\}$, where temperatures are modified by offset time series,
-${\Delta}T_{\text{TS}}$, and lapse-rate corrections, ${\Delta}T_{\text{LR}}$:
+      a~present-day monthly climatology, $\{T_{\mathrm{m}0},
+      P_{\mathrm{m}0}\}$, where temperatures are modified by offset time
+      series, ${\Delta}T_{\text{TS}}$, and lapse-rate corrections,
+      ${\Delta}T_{\text{LR}}$:
 \begin{align}
@@
 \end{align}
-
       The present-day monthly climatology was computed from near-surface air
@@
       Project \citep[NGRIP,][]{Andersen.etal.2004}, the European Project for
-      Ice Coring in Antarctica \citep[EPICA,][] {Jouzel.etal.2007}, the
-      Vostok ice core \citep{Petit.etal.1999}, and Ocean Drilling Program
-      (ODP) sites 1012 and 1020, both located off the coast of California
-      \citep{Herbert.etal.2001}. Palaeo-temperature anomalies from the GRIP
-      and NGRIP records were calculated from oxygen isotope
-      (\chem{\delta^{18}O}) measurements using a~quadratic equation
-      \citep{Johnsen.etal.1995},
+      Ice Coring in Antarctica \citep[EPICA,][] {Jouzel.etal.2007}, the Vostok
+      ice core \citep{Petit.etal.1999}, and Ocean Drilling Program (ODP) sites
+      1012 and 1020, both located off the coast of California
+      \citep{Herbert.etal.2001}. Palaeo-temperature anomalies from the GRIP and
+      NGRIP records were calculated from oxygen isotope (\chem{\delta^{18}O})
+      measurements using a~quadratic equation \citep{Johnsen.etal.1995},
 \begin{align}
@@
 \end{align}
-
       thus accounting for the evolution of ice thickness, ${h=s-b}$, on the
       one hand, and for differences between the basal topography of the ice
-      flow model, $b$, and the NARR reference topography, $b_{\text{ref}}$, on the
-      other hand. All simulations use an annual temperature lapse rate of
-      $\gamma = 6\,\unit{K\,km^{-1}}$. In the rest of this paper, we refer
-      to different model runs by the name of the proxy record used for the
+      flow model, $b$, and the NARR reference topography, $b_{\text{ref}}$, on
+      the other hand. All simulations use an annual temperature lapse rate of
+      $\gamma = 6\,\unit{K\,km^{-1}}$. In the rest of this paper, we refer to
+      different model runs by the name of the proxy record used for the
       palaeo-temperature forcing.
@@
       (Fig.~\ref{fig:lr_ts}, upper panel), model output presents consistent
-      features that can be observed across the range of forcing data used.
-      In all simulations, modelled ice volumes remain relatively low during
-      most of the glacial cycle, except during two major glacial events
-      which occur between 61.9 and 56.5\,\unit{ka} during MIS~4, and between
-      23.2 and 16.8\,\unit{ka} during MIS~2 (Fig.~\ref{fig:lr_ts}, lower
-      panel). An ice volume minimum is consistently reached between 53.0 and
-      41.3\,\unit{ka} during MIS~3. However, the magnitude and precise
-      timing of these three events depend significantly on the choice of
-      proxy record used to derive a~time-dependent climate forcing
-      (Table~\ref{tab:extrema}).
+      features that can be observed across the range of forcing data used. In
+      all simulations, modelled ice volumes remain relatively low during most
+      of the glacial cycle, except during two major glacial events which occur
+      between 61.9 and 56.5\,\unit{ka} during MIS~4, and between 23.2 and
+      16.8\,\unit{ka} during MIS~2 (Fig.~\ref{fig:lr_ts}, lower panel). An ice
+      volume minimum is consistently reached between 53.0 and 41.3\,\unit{ka}
+      during MIS~3. However, the magnitude and precise timing of these three
+      events depend significantly on the choice of proxy record used to derive
+      a~time-dependent climate forcing (Table~\ref{tab:extrema}).
 
@@
       this early deglaciation is caused by an early warming present in the
-      alkenone palaeo-temperature reconstructions (Fig.~\ref{fig:lr_ts},
-      upper panel; \citealp[Fig.~3]{Herbert.etal.2001}). However, this early
-      warming is a~local effect, corresponding to a~weakening of the
-      California current \citep{Herbert.etal.2001}. The California current,
-      driving cold waters southwards along the south-western coast of North
-      America, has been shown to have weakened during each peak of global
-      glaciation (in SPECMAP) during the past 550\,\unit{ka}, including the
-      LGM, resulting in paradoxically warmer sea-surface temperatures at the
-      locations of the ODP~1012 and ODP~1020 sites
-      \citep{Herbert.etal.2001}.
+      alkenone palaeo-temperature reconstructions (Fig.~\ref{fig:lr_ts}, upper
+      panel; \citealp[Fig.~3]{Herbert.etal.2001}). However, this early warming
+      is a~local effect, corresponding to a~weakening of the California current
+      \citep{Herbert.etal.2001}. The California current, driving cold waters
+      southwards along the south-western coast of North America, has been shown
+      to have weakened during each peak of global glaciation (in SPECMAP)
+      during the past 550\,\unit{ka}, including the LGM, resulting in
+      paradoxically warmer sea-surface temperatures at the locations of the
+      ODP~1012 and ODP~1020 sites \citep{Herbert.etal.2001}.
 
@@
       6\,\unit{ka} earlier than dated evidence of the LGM advance. There is
-      a~fair agreement between the EPICA and Vostok palaeo-temperature
-      records, resulting in only small differences between the simulations
-      driven by those records. These differences are not subject to further
-      analysis here; instead we focus on simulations forced by
-      palaeo-temperature records from the GRIP and EPICA ice cores that
-      appear to produce the most realistic reconstructions of regional
-      glaciation history, yet bearing significant disparities in model
-      output. To allow for a~more detailed comparison against the geological
-      record, these two simulations were re-run using a~higher-resolution
-      grid (Sect.~\ref{sec:model}; Fig.~\ref{fig:lr_ts}, lower panel, dotted
-      lines).
+      a~fair agreement between the EPICA and Vostok palaeo-temperature records,
+      resulting in only small differences between the simulations driven by
+      those records. These differences are not subject to further analysis
+      here; instead we focus on simulations forced by palaeo-temperature
+      records from the GRIP and EPICA ice cores that appear to produce the most
+      realistic reconstructions of regional glaciation history, yet bearing
+      significant disparities in model output. To allow for a~more detailed
+      comparison against the geological record, these two simulations were
+      re-run using a~higher-resolution grid (Sect.~\ref{sec:model};
+      Fig.~\ref{fig:lr_ts}, lower panel, dotted lines).
 
@@
       timespan of the last glacial cycle (Fig.~\ref{fig:hr_geom_duration},
-      hatched areas). However, except for its margins in the Pacific Ocean
-      and in the northern foothills of the Alaska Range, the maximum extent
-      of the ice sheet is attained for a~much shorter period of time of only
-      few thousand years (Fig.~\ref{fig:hr_geom_duration}, red areas). This
-      result illustrates that the maximum extents of the modelled ice sheet
-      during MIS~4 and MIS~2 were both short-lived and therefore out of
-      balance with contemporary climate.
+      hatched areas). However, except for its margins in the Pacific Ocean and
+      in the northern foothills of the Alaska Range, the maximum extent of the
+      ice sheet is attained for a~much shorter period of time of only few
+      thousand years (Fig.~\ref{fig:hr_geom_duration}, red areas). This result
+      illustrates that the maximum extents of the modelled ice sheet during
+      MIS~4 and MIS~2 were both short-lived and therefore out of balance with
+      contemporary climate.
 
@@
       apparent insensitivity of modelled glacial extent to temperature
-      fluctuations results from a~combination of low precipitation, high
-      summer temperature and large temperature standard deviation (PDD SD)
-      in the plains of the Alaska Interior (Fig.~\ref{fig:atm}) which
-      confines glaciation to the foothills of the mountains. This result
-      could potentially explain the local distribution of glacial deposits,
-      which indicates that glaciers flowing on the northern slope of the
-      Alaska Range have remained small throughout the Pleistocene
+      fluctuations results from a~combination of low precipitation, high summer
+      temperature and large temperature standard deviation (PDD SD) in the
+      plains of the Alaska Interior (Fig.~\ref{fig:atm}) which confines
+      glaciation to the foothills of the mountains. This result could
+      potentially explain the local distribution of glacial deposits, which
+      indicates that glaciers flowing on the northern slope of the Alaska Range
+      have remained small throughout the Pleistocene
       \citep{Kaufman.Manley.2004}.
@@
       centres, visible on the modelled ice cover duration maps
-      (Fig.~\ref{fig:hr_geom_duration}), are located over the Coast
-      Mountains (CM), the Columbia and Rocky mountains (CRM), the Skeena
-      Mountains (SM), and the Selwyn and Mackenzie mountains (SMKM). The
-      Wrangell and Saint Elias mountains, heavily glacierized at present,
-      host an ice cap for the entire length of both simulations, but that
-      ice cap does not appear to be a~major feed to the Cordilleran ice
-      sheet (Fig.~\ref{fig:hr_geom_duration}, WSEM). Although the Coast,
-      Skeena and Columbia and Rocky mountains (CM, SM, CRM) are covered by
-      mountain glaciers for most of the last glacial cycle, providing
-      durable nucleation centres for an ice sheet, this is not the case for
-      the Selwyn and Mackenzie mountains (SMKM), where ice cover on the
-      highest peaks is limited to a~small fraction of the last glacial
-      cycle. In other words, the Selwyn and Mackenzie mountains only appear
-      as a~secondary ice-dispersal centre during the coldest periods of the
-      last glacial cycle. The Northern Rocky Mountains
-      (Fig.~\ref{fig:hr_geom_duration}, NRM) do not act as a~nucleation
-      centre, but rather as a~pinning point for the Cordilleran ice sheet
-      margin coming from the west.
+      (Fig.~\ref{fig:hr_geom_duration}), are located over the Coast Mountains
+      (CM), the Columbia and Rocky mountains (CRM), the Skeena Mountains (SM),
+      and the Selwyn and Mackenzie mountains (SMKM). The Wrangell and Saint
+      Elias mountains, heavily glacierized at present, host an ice cap for the
+      entire length of both simulations, but that ice cap does not appear to be
+      a~major feed to the Cordilleran ice sheet
+      (Fig.~\ref{fig:hr_geom_duration}, WSEM). Although the Coast, Skeena and
+      Columbia and Rocky mountains (CM, SM, CRM) are covered by mountain
+      glaciers for most of the last glacial cycle, providing durable nucleation
+      centres for an ice sheet, this is not the case for the Selwyn and
+      Mackenzie mountains (SMKM), where ice cover on the highest peaks is
+      limited to a~small fraction of the last glacial cycle. In other words,
+      the Selwyn and Mackenzie mountains only appear as a~secondary
+      ice-dispersal centre during the coldest periods of the last glacial
+      cycle. The Northern Rocky Mountains (Fig.~\ref{fig:hr_geom_duration},
+      NRM) do not act as a~nucleation centre, but rather as a~pinning point for
+      the Cordilleran ice sheet margin coming from the west.
 
@@
       its predominance over the other ice-dispersal centres
-      (Figs.~\ref{fig:lr_maps} and~\ref{fig:hr_geom_duration}, SM).
-      Regardless of the applied forcing, this ice cap appears to survive
-      MIS~3 (Fig.~\ref{fig:lr_maps}, middle panels), and serves as
-      a~nucleation centre at the onset of the glacial readvance towards the
-      LGM (MIS~2). This situation appear similar to the neighbouring
-      Laurentide ice sheet, for which the importance of residual ice for the
-      glacial history leading up to the LGM has been illustrated by the
-      MIS~3 residual ice bodies in northern and eastern Canada as nucleation
-      centres for a~much more extensive MIS~2 configuration
-      \citep{Kleman.etal.2010}.
+      (Figs.~\ref{fig:lr_maps} and~\ref{fig:hr_geom_duration}, SM). Regardless
+      of the applied forcing, this ice cap appears to survive MIS~3
+      (Fig.~\ref{fig:lr_maps}, middle panels), and serves as a~nucleation
+      centre at the onset of the glacial readvance towards the LGM (MIS~2).
+      This situation appear similar to the neighbouring Laurentide ice sheet,
+      for which the importance of residual ice for the glacial history leading
+      up to the LGM has been illustrated by the MIS~3 residual ice bodies in
+      northern and eastern Canada as nucleation centres for a~much more
+      extensive MIS~2 configuration \citep{Kleman.etal.2010}.
 
@@
       (42.9\,\unit{ka}) in the GRIP simulation and of 2.68\,\unit{m}\,s.l.e.
-      (51.8\,\unit{ka}) in the EPICA simulation, corresponding to
-      respectively 19 and 31\,\unit{\%} of the MIS~2 ice volumes.
+      (51.8\,\unit{ka}) in the EPICA simulation, corresponding to respectively
+      19 and 31\,\unit{\%} of the MIS~2 ice volumes.
 
@@
 
-      A~correlation is observed between the modelled duration of warm based
-      ice cover (Fig.~\ref{fig:hr_geom_warmbase}) and the degree of glacial
-      modification of the landscape (mainly in terms of the development of
-      deep glacial valleys and troughs). We find evidence for this on the
-      slopes of the Coast Mountains, the Columbia and Rocky mountains, the
-      Wrangell and Saint Elias mountains, and radiating off the Skeena
-      Mountains (Figs.~\ref{fig:hr_geom_duration} and
-      \ref{fig:hr_geom_warmbase}; \citealp[Fig.~2]{Kleman.etal.2010}). The
-      Skeena Mountains, for example, indeed bear a~strong glacial imprint
-      that indicates ice drainage in a~system of distinct glacial troughs
-      emanating in a~radial pattern from the centre of the mountain range
-      \citep[Fig.~2]{Kleman.etal.2010}. We suggest that persistent ice cover
-      (Fig.~\ref{fig:hr_geom_duration}) associated with basal ice
-      temperatures at the pressure melting point
+      A~correlation is observed between the modelled duration of warm based ice
+      cover (Fig.~\ref{fig:hr_geom_warmbase}) and the degree of glacial
+      modification of the landscape (mainly in terms of the development of deep
+      glacial valleys and troughs). We find evidence for this on the slopes of
+      the Coast Mountains, the Columbia and Rocky mountains, the Wrangell
+      and Saint Elias mountains, and radiating off the Skeena Mountains
+      (Figs.~\ref{fig:hr_geom_duration} and \ref{fig:hr_geom_warmbase};
+      \citealp[Fig.~2]{Kleman.etal.2010}). The Skeena Mountains, for example,
+      indeed bear a~strong glacial imprint that indicates ice drainage in
+      a~system of distinct glacial troughs emanating in a~radial pattern from
+      the centre of the mountain range \citep[Fig.~2]{Kleman.etal.2010}. We
+      suggest that persistent ice cover (Fig.~\ref{fig:hr_geom_duration})
+      associated with basal ice temperatures at the pressure melting point
       (Figs.~\ref{fig:hr_geom_warmbase} and~\ref{fig:hr_geom_warmfrac})
@@
       (\citealp[Fig.~2]{Kleman.etal.2010}; \citealp[Fig.~8]
-      {Stroeven.etal.2010}) is modelled to have hosted predominantly
-      warm-based ice (Fig.~\ref{fig:hr_geom_warmfrac}). However, because it
-      is only glaciated for a~short fraction of the last glacial cycle in
-      our simulations (Fig.~\ref{fig:hr_geom_duration}), this perhaps
-      indicates that the observed landscape pattern originates from multiple
-      glacial cycles and witnesses an increased relative importance of the
-      Selwyn and Mackenzie mountains ice dispersal centre
+      {Stroeven.etal.2010}) is modelled to have hosted predominantly warm-based
+      ice (Fig.~\ref{fig:hr_geom_warmfrac}). However, because it is only
+      glaciated for a~short fraction of the last glacial cycle in our
+      simulations (Fig.~\ref{fig:hr_geom_duration}), this perhaps indicates
+      that the observed landscape pattern originates from multiple glacial
+      cycles and witnesses an increased relative importance of the Selwyn and
+      Mackenzie mountains ice dispersal centre
       (Fig.~\ref{fig:hr_geom_duration}, SMKM), prior to the Late Pleistocene
@@
       70\,\unit{mW\,m^{-2}} geothermal heat flux at 3\,\unit{km} depth
-      (Sect.~\ref{sec:icedyn}). However, the Skeena Mountains and the area
-      west of the Mackenzie Mountains experience higher-than-average
-      geothermal heat flux with measured values of ca.~80 and
-      ca.~100\,\unit{mW\,m^{-2}} \citep{Blackwell.Richards.2004}. We can
-      therefore expect even longer durations of warm-based ice cover for
-      these areas if we were to include spatially variable geothermal
-      forcing in our Cordilleran ice sheet simulations.
+      (Sect.~\ref{sec:icedyn}). However, the Skeena Mountains and the area west
+      of the Mackenzie Mountains experience higher-than-average geothermal heat
+      flux with measured values of ca.~80 and ca.~100\,\unit{mW\,m^{-2}}
+      \citep{Blackwell.Richards.2004}. We can therefore expect even longer
+      durations of warm-based ice cover for these areas if we were to include
+      spatially variable geothermal forcing in our Cordilleran ice sheet
+      simulations.
 
@@
 
-      Modelled patterns of ice sheet retreat are relatively consistent
-      between the two simulations (Figs.~\ref{fig:hr_geom_deglacageshots}
-      and~\ref{fig:hr_geom_deglacage}). The southern sector of the
-      modelling domain, including the Puget Lowland, the Coast Mountains,
-      the Columbia and and Rocky mountains, and the Interior Plateau of
-      British Columbia, becomes completely deglaciated by 10\,\unit{ka},
-      whereas a~significant ice cover remains over the Skeena, the Selwyn
-      and Mackenzie, and the Wrangell and Saint Elias mountains in the
-      northern sector of the modelling domain. After 10\,\unit{ka},
-      deglaciation continues to proceed across the Liard Lowland with
-      a~radial ice margin retreat towards the surrounding mountain ranges,
-      consistent with the regional melt water record of the last
-      deglaciation \citep{Margold.etal.2013}. Remaining ice continues to
-      decay by retreating towards the Selwyn and MacKenzie, and Skeena
-      mountains. The last remnants of the Cordilleran ice sheet finally
+      Modelled patterns of ice sheet retreat are relatively consistent between
+      the two simulations (Figs.~\ref{fig:hr_geom_deglacageshots}
+      and~\ref{fig:hr_geom_deglacage}). The southern sector of the modelling
+      domain, including the Puget Lowland, the Coast Mountains, the Columbia
+      and and Rocky mountains, and the Interior Plateau of British Columbia,
+      becomes completely deglaciated by 10\,\unit{ka}, whereas a~significant
+      ice cover remains over the Skeena, the Selwyn and Mackenzie, and the
+      Wrangell and Saint Elias mountains in the northern sector of the
+      modelling domain. After 10\,\unit{ka}, deglaciation continues to proceed
+      across the Liard Lowland with a~radial ice margin retreat towards the
+      surrounding mountain ranges, consistent with the regional melt water
+      record of the last deglaciation \citep{Margold.etal.2013}. Remaining ice
+      continues to decay by retreating towards the Selwyn and MacKenzie, and
+      Skeena mountains. The last remnants of the Cordilleran ice sheet finally
       disappear from the Skeena Mountains at 6.5\,\unit{ka} (GRIP) and
@@
       (Fig.~\ref{fig:hr_geom_deglacage}, OM) presents a~different kind of
-      evidence. There, sharp-crested moraines indicate a~late-glacial
-      readvance of local alpine glaciers and, more importantly, their
-      interaction with larger, lingering remnants of the main body of the
-      Cordilleran ice sheet in the valleys \citep{Lakeman.etal.2008}.
-      Additional evidence for late-glacial alpine glacier readvances
-      includes moraines in the eastern Coast Mountains, and the Columbia and
-      Rocky mountains \citep{Osborn.Gerloff.1997, Menounos.etal.2008}.
+      evidence. There, sharp-crested moraines indicate a~late-glacial readvance
+      of local alpine glaciers and, more importantly, their interaction with
+      larger, lingering remnants of the main body of the Cordilleran ice sheet
+      in the valleys \citep{Lakeman.etal.2008}. Additional evidence for
+      late-glacial alpine glacier readvances includes moraines in the eastern
+      Coast Mountains, and the Columbia and Rocky mountains
+      \citep{Osborn.Gerloff.1997, Menounos.etal.2008}.
 
@@
       distribution, and to identify the potential climatic triggers
-      \citep{Menounos.etal.2008}, it is interesting to note that the
-      simulation driven by the GRIP record produces a~late-glacial readvance
-      in the Coast Mountains and in the Columbia and Rocky Mountains
-      (Fig.~\ref{fig:hr_geom_deglacage}, left panel). In addition to
-      matching the location of some local readvances, the GRIP-driven
-      simulation shows that a~large remnant of the decaying ice sheet may
-      still have existed at the time of this late-glacial readvance. In
-      contrast, the EPICA-driven simulation produces a~nearly-continuous
-      deglaciation with only a~tightly restricted late-glacial readvance on
-      the western slopes of the Saint Elias and the Coast mountains
-      (Fig.~\ref{fig:hr_geom_deglacage}, right panel).
+      \citep{Menounos.etal.2008}, it is interesting to note that the simulation
+      driven by the GRIP record produces a~late-glacial readvance in the Coast
+      Mountains and in the Columbia and Rocky Mountains
+      (Fig.~\ref{fig:hr_geom_deglacage}, left panel). In addition to matching
+      the location of some local readvances, the GRIP-driven simulation shows
+      that a~large remnant of the decaying ice sheet may still have existed at
+      the time of this late-glacial readvance. In contrast, the EPICA-driven
+      simulation produces a~nearly-continuous deglaciation with only a~tightly
+      restricted late-glacial readvance on the western slopes of the Saint
+      Elias and the Coast mountains (Fig.~\ref{fig:hr_geom_deglacage}, right
+      panel).
 
@@
 
-      Because a~general tenant in glacial geomorphology is that the majority
-      of landforms (lineations and eskers) are part of the deglacial
-      envelope \citep[terminology from][]{Kleman.etal.2006}, having been
-      formed close inside the retreating margin of ice sheets
-      \citep{Boulton.Clark.1990, Kleman.etal.1997, Kleman.etal.2010}, we
-      present maps of basal flow directions immediately preceding
-      deglaciation or at the time of cessation of sliding inside
-      a~cold-based retreating margin (Fig.~\ref{fig:hr_geom_lastflow}). The
-      modelled deglacial flow patterns are mostly consistent between the
-      GRIP and EPICA simulations. They depict an active ice sheet retreat in
-      the peripheral areas, followed by stagnant ice decay in some of the
-      interior regions. Several parts of the modelling domain do not
-      experience any basal sliding throughout the deglaciation phase
-      (Fig.~\ref{fig:hr_geom_lastflow}, hatched areas). This notably
+      Because a~general tenant in glacial geomorphology is that the majority of
+      landforms (lineations and eskers) are part of the deglacial envelope
+      \citep[terminology from][]{Kleman.etal.2006}, having been formed close
+      inside the retreating margin of ice sheets \citep{Boulton.Clark.1990,
+      Kleman.etal.1997, Kleman.etal.2010}, we present maps of basal flow
+      directions immediately preceding deglaciation or at the time of cessation
+      of sliding inside a~cold-based retreating margin
+      (Fig.~\ref{fig:hr_geom_lastflow}). The modelled deglacial flow patterns
+      are mostly consistent between the GRIP and EPICA simulations. They depict
+      an active ice sheet retreat in the peripheral areas, followed by stagnant
+      ice decay in some of the interior regions. Several parts of the modelling
+      domain do not experience any basal sliding throughout the deglaciation
+      phase (Fig.~\ref{fig:hr_geom_lastflow}, hatched areas). This notably
       includes parts of the Interior Plateau in British Columbia, major
@@
       \citealp[Fig.~1.12]{Clague.1989}; \citealp[Fig.~2]{Kleman.etal.2010})
-      show similarities with the patterns of deglacial ice flow from
-      numerical modelling (Fig.~\ref{fig:hr_geom_lastflow}). In the northern
-      half of the modelling domain, modelled deglacial flow directions
-      depict an active downhill flow as the last remnants of the ice sheet
-      retreat towards mountain ranges. Converging deglacial flow patterns in
-      the Liard Lowland, for instance (Fig.~\ref{fig:hr_geom_lastflow}),
-      closely resemble the pattern indicated by glacial lineations
+      show similarities with the patterns of deglacial ice flow from numerical
+      modelling (Fig.~\ref{fig:hr_geom_lastflow}). In the northern half of the
+      modelling domain, modelled deglacial flow directions depict an active
+      downhill flow as the last remnants of the ice sheet retreat towards
+      mountain ranges. Converging deglacial flow patterns in the Liard Lowland,
+      for instance (Fig.~\ref{fig:hr_geom_lastflow}), closely resemble the
+      pattern indicated by glacial lineations
       \citep[Fig.~2]{Margold.etal.2013}.
@@
       simulations produce a~retreat of the ice margin towards the north-east
-      (Fig.~\ref{fig:hr_geom_deglacage}), a~pattern which is validated by
-      the geomorphological and stratigraphical record for ancient
-      pro-glacial lakes dammed by the retreating ice sheet
-      \citep{Perkins.Brennand.2014}. However, the two simulations differ in
-      the mode of retreat. The GRIP simulation yields an active retreat with
-      basal sliding towards the ice margin to the south, whereas the EPICA
-      simulation produces negligible basal sliding on the plateau during
-      deglaciation (Fig.~\ref{fig:hr_geom_lastflow}). Yet, the Interior
-      Plateau also hosts an impressive set of glacial lineations which
-      indicate a~substantial eastwards flow component of the Cordilleran ice
-      sheet \citep{Prest.etal.1968, Kleman.etal.2010}. This Interior Plateau
-      lineation set could therefore present a~smoking gun for the
-      reliability of the presented model results. One explanation for the
-      incongruent results could be that the missing feedback mechanisms
-      between ice sheet topography and regional climate resulted in
-      a~modelled ice divide of the LGM ice sheet being too far to the east
-      (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
-      \citealp{Seguinot.etal.2014}). A~more westerly-located LGM ice divide
-      would certainly result in a~different deglacial flow pattern over the
-      Interior Plateau. However, a~more westerly-positioned LGM ice divide
-      would certainly be associated with a~thinner ice sheet than that
-      modelled here. Decreased ice thickness would not promote warm-based
-      conditions but, on the contrary, enlarge the region of negligible
-      basal sliding (Fig.~\ref{fig:hr_geom_lastflow}). Thus, a~second
-      explanation for the incongruent results could be that the Interior
-      Plateau lineation system predates deglaciation ice flow, as perhaps
-      indicated by some eskers that appear incompatible with these glacial
-      lineations \citep[Fig.~9]{Margold.etal.2013a}. Finally, a~third
-      explanation could be that local geothermal heat associated with
-      volcanic activity on the Interior Plateau could have triggered the
-      basal sliding \citep[cf. Greenland ice
-      sheet;][]{Fahnestock.etal.2001}.
+      (Fig.~\ref{fig:hr_geom_deglacage}), a~pattern which is validated by the
+      geomorphological and stratigraphical record for ancient pro-glacial lakes
+      dammed by the retreating ice sheet \citep{Perkins.Brennand.2014}.
+      However, the two simulations differ in the mode of retreat. The GRIP
+      simulation yields an active retreat with basal sliding towards the ice
+      margin to the south, whereas the EPICA simulation produces negligible
+      basal sliding on the plateau during deglaciation
+      (Fig.~\ref{fig:hr_geom_lastflow}). Yet, the Interior Plateau also hosts
+      an impressive set of glacial lineations which indicate a~substantial
+      eastwards flow component of the Cordilleran ice sheet
+      \citep{Prest.etal.1968, Kleman.etal.2010}. This Interior Plateau
+      lineation set could therefore present a~smoking gun for the reliability
+      of the presented model results. One explanation for the incongruent
+      results could be that the missing feedback mechanisms between ice sheet
+      topography and regional climate resulted in a~modelled ice divide of the
+      LGM ice sheet being too far to the east (Sect.~\ref{sec:mis2};
+      Fig.~\ref{fig:mis2}; \citealp{Seguinot.etal.2014}). A~more
+      westerly-located LGM ice divide would certainly result in a~different
+      deglacial flow pattern over the Interior Plateau. However, a~more
+      westerly-positioned LGM ice divide would certainly be associated with
+      a~thinner ice sheet than that modelled here. Decreased ice thickness
+      would not promote warm-based conditions but, on the contrary, enlarge the
+      region of negligible basal sliding (Fig.~\ref{fig:hr_geom_lastflow}).
+      Thus, a~second explanation for the incongruent results could be that the
+      Interior Plateau lineation system predates deglaciation ice flow, as
+      perhaps indicated by some eskers that appear incompatible with these
+      glacial lineations \citep[Fig.~9]{Margold.etal.2013a}. Finally, a~third
+      explanation could be that local geothermal heat associated with volcanic
+      activity on the Interior Plateau could have triggered the basal sliding
+      \citep[cf. Greenland ice sheet;][]{Fahnestock.etal.2001}.
 
       The modelled deglaciation of the Interior Plateau of British Columbia
-      consists of a~rapid northwards retreat
-      (Fig.~\ref{fig:hr_geom_deglacage}) of southwards-flowing non-sliding
-      ice lobes (Fig.~\ref{fig:hr_geom_lastflow}) positioned in-between
-      deglaciated mountain ranges (Figs.~\ref{fig:hr_pf_grip}
-      and~\ref{fig:hr_pf_epica}). This result appears compatible with the
-      prevailing conceptual model of deglaciation of central British
-      Columbia, in which mountain ranges emerge from the ice before the
-      plateau \citep[Fig.~7]{Fulton.1991}. However, due to different
-      topographic and climatic conditions, our simulations produce different
-      deglaciation patterns in the northern half of the model domain,
+      consists of a~rapid northwards retreat (Fig.~\ref{fig:hr_geom_deglacage})
+      of southwards-flowing non-sliding ice lobes
+      (Fig.~\ref{fig:hr_geom_lastflow}) positioned in-between deglaciated
+      mountain ranges (Figs.~\ref{fig:hr_pf_grip} and~\ref{fig:hr_pf_epica}).
+      This result appears compatible with the prevailing conceptual model of
+      deglaciation of central British Columbia, in which mountain ranges emerge
+      from the ice before the plateau \citep[Fig.~7]{Fulton.1991}. However, due
+      to different topographic and climatic conditions, our simulations produce
+      different deglaciation patterns in the northern half of the model domain,
       indicating that this conceptual model may not be applied to the entire
@@
       glacial cycle presented in this study consistently produce two glacial
-      maxima during MIS~4 (61.9--56.5\,\unit{ka},
-      3.8--8.8\,\unit{m}\,s.l.e.) and MIS~2 (23.2--16.8\,\unit{ka},
-      8.40--8.91\,\unit{m}\,s.l.e.), two periods corresponding to documented
-      extensive glaciations. This result is independent of the
-      palaeo-temperature record used among the six selected for this study,
-      and thus can be regarded as a~robust model output, which broadly
-      matches geological evidence. However, the timing of the two glaciation
-      peaks depends sensitively on which climate record is used to drive the
-      model. The timing of the LGM is best reproduced by the EPICA and
-      Vostok Antarctic ice core records. It occurs about 2\,\unit{ka} too
+      maxima during MIS~4 (61.9--56.5\,\unit{ka}, 3.8--8.8\,\unit{m}\,s.l.e.)
+      and MIS~2 (23.2--16.8\,\unit{ka}, 8.40--8.91\,\unit{m}\,s.l.e.), two
+      periods corresponding to documented extensive glaciations. This result is
+      independent of the palaeo-temperature record used among the six selected
+      for this study, and thus can be regarded as a~robust model output, which
+      broadly matches geological evidence. However, the timing of the two
+      glaciation peaks depends sensitively on which climate record is used to
+      drive the model. The timing of the LGM is best reproduced by the EPICA
+      and Vostok Antarctic ice core records. It occurs about 2\,\unit{ka} too
       early in the simulation forced by the GRIP ice core record, and occurs
       even earlier in all other simulations. The mismatch is largest for the
-      two Northwest Pacific ODP palaeo-temperature records, which are
-      affected by the weakening of the California current during the LGM.
+      two Northwest Pacific ODP palaeo-temperature records, which are affected
+      by the weakening of the California current during the LGM.
 

### 2016-01-18 cordillera-cycle.tex
Typeset: add trailing percent signs.

@@
 \subsection{Overview}
-\label{sec:overview}
+\label{sec:overview}%
 
@@
 \subsection{Climate forcing}
-\label{sec:atm}
+\label{sec:atm}%
 
@@
 
-\begin{figure}
+\begin{figure}%
 \includegraphics{locmap}
-  \caption{Relief map of northern North America showing a~reconstruction of the
+\caption{%
+Relief map of northern North America showing a~reconstruction of the
            areas once covered by the Cordilleran (CIS), Laurentide (LIS),
@@
            Data \citep{Patterson.Kelso.2015}.}
-  \label{fig:locmap}
-\end{figure}
-
-\begin{figure}
+\label{fig:locmap}%
+\end{figure}%
+
+\begin{figure}%
 \includegraphics{atm}
-  \caption{Monthly mean near-surface air temperature, precipitation, and
+\caption{%
+Monthly mean near-surface air temperature, precipitation, and
            standard deviation of daily mean temperature (PDD SD) for January
@@
            coastal and inland regions.}
-  \label{fig:atm}
-\end{figure}
-
-\begin{figure*}
+\label{fig:atm}%
+\end{figure}%
+
+\begin{figure*}%
 \includegraphics{lr_ts}
-  \caption{Temperature offset time-series from ice core and ocean records
+\caption{%
+Temperature offset time-series from ice core and ocean records
            (Table~\ref{tab:records}) used as palaeo-climate forcing for the ice
@@
            5\,\unit{km}-resolution runs.}
-  \label{fig:lr_ts}
-\end{figure*}
-
-\begin{figure*}
+\label{fig:lr_ts}%
+\end{figure*}%
+
+\begin{figure*}%
 \includegraphics{lr_maps}
-  \caption{Snapshots of modelled surface topography (500\,\unit{m} contours)
+\caption{%
+Snapshots of modelled surface topography (500\,\unit{m} contours)
            corresponding to the ice volume extremes indicated on
@@
            similarities despite large differences in timing.}
-  \label{fig:lr_maps}
-\end{figure*}
-
-\begin{figure*}
+\label{fig:lr_maps}%
+\end{figure*}%
+
+\begin{figure*}%
 \includegraphics{hr_maps_mis2}
-  \caption{Modelled surface topography (200\,\unit{m} contours) and surface velocity
+\caption{%
+Modelled surface topography (200\,\unit{m} contours) and surface velocity
            (colour mapping) corresponding to the maximum ice volume during
            MIS~2 in the GRIP and EPICA high-resolution simulations.}
-  \label{fig:mis2}
-\end{figure*}
-
-\begin{figure*}
+\label{fig:mis2}%
+\end{figure*}%
+
+\begin{figure*}%
 \includegraphics{hr_maps_mis4}
-  \caption{Modelled surface topography (200\,\unit{m} contours) and surface velocity
+\caption{%
+Modelled surface topography (200\,\unit{m} contours) and surface velocity
            (colour mapping) corresponding to the maximum ice volume during
            MIS~4 in the GRIP and EPICA high-resolution simulations.}
-  \label{fig:mis4}
-\end{figure*}
-
-\begin{figure*}
+\label{fig:mis4}%
+\end{figure*}%
+
+\begin{figure*}%
 \includegraphics{hr_geom_duration}
-  \caption{Modelled duration of ice cover during the last 120\,\unit{ka} using GRIP
+\caption{%
+Modelled duration of ice cover during the last 120\,\unit{ka} using GRIP
            and EPICA climate forcing.
@@
            simulation. See Fig.~\ref{fig:locmap} for a~list of abbreviations.}
-  \label{fig:hr_geom_duration}
-\end{figure*}
-
-\begin{figure*}
+\label{fig:hr_geom_duration}%
+\end{figure*}%
+
+\begin{figure*}%
 \includegraphics{hr_geom_warmbase}
-  \caption{Modelled duration of warm-based ice cover during the last
+\caption{%
+Modelled duration of warm-based ice cover during the last
            120\,\unit{ka}. Long ice cover durations combined with basal
@@
            Hatches indicate areas that were covered by cold ice only.}
-  \label{fig:hr_geom_warmbase}
-\end{figure*}
-
-\begin{figure*}
-  \vspace{-0.5mm}  % enough to allow the two figures to fit on one page
+\label{fig:hr_geom_warmbase}%
+\end{figure*}%
+
+\begin{figure*}%
 \includegraphics{hr_geom_warmfrac}
-  \caption{Modelled fraction of warm-based ice cover during the ice-covered
+\caption{%
+Modelled fraction of warm-based ice cover during the ice-covered
            period. Note the dominance of warm-based conditions on the
@@
            Hatches indicate areas that were covered by cold ice only.}
-  \label{fig:hr_geom_warmfrac}
-\end{figure*}
-
-\begin{figure}
+\label{fig:hr_geom_warmfrac}%
+\end{figure*}%
+
+\begin{figure}%
   \includegraphics{lr_ts_deglac}
-  \caption{Temperature offset time-series from the GRIP and EPICA ice core
+\caption{%
+Temperature offset time-series from the GRIP and EPICA ice core
            records (Table~\ref{tab:records}) (top panel), and modelled ice
@@
            equivalent (bottom panel).}
-  \label{fig:lr_ts_deglac}
-\end{figure}
-
-\begin{figure*}
+\label{fig:lr_ts_deglac}%
+\end{figure}%
+
+\begin{figure*}%
 \includegraphics{hr_maps_deglac}
-  \caption{Snapshots of modelled surface topography (200\,\unit{m} contours)
+\caption{%
+Snapshots of modelled surface topography (200\,\unit{m} contours)
            and surface velocity (colour mapping) during the last deglaciation
@@
            Figs.~\ref{fig:hr_pf_grip} and~\ref{fig:hr_pf_epica}.}
-  \label{fig:hr_geom_deglacageshots}
-\end{figure*}
-
-\begin{figure*}
-  \vspace{-1.5mm}  % enough to allow the two figures to fit on one page
+\label{fig:hr_geom_deglacageshots}%
+\end{figure*}%
+
+\begin{figure*}%
 \includegraphics{hr_geom_deglacage}
-  \caption{Modelled age of the last deglaciation. Areas that have been covered
+\caption{%
+Modelled age of the last deglaciation. Areas that have been covered
            only before the last glacial maximum are marked
@@
            Figs.~\ref{fig:hr_pf_grip} and~\ref{fig:hr_pf_epica}.}
-  \label{fig:hr_geom_deglacage}
-\end{figure*}
-
-\begin{figure*}
+\label{fig:hr_geom_deglacage}%
+\end{figure*}%
+
+\begin{figure*}%
 \includegraphics{hr_geom_lastflow}
-  \caption{Modelled deglacial basal ice velocities. Hatches
+\caption{%
+Modelled deglacial basal ice velocities. Hatches
            indicate areas that remain non-sliding throughout deglaciation
@@
            using a~basal velocity threshold of 1\,\unit{m\,yr^{-1}}.}
-  \label{fig:hr_geom_lastflow}
-\end{figure*}
-
-\begin{figure}
+\label{fig:hr_geom_lastflow}%
+\end{figure*}%
+
+\begin{figure}%
 \includegraphics{hr_pf_grip}
-  \caption{Modelled bedrock (black) and ice surface (blue) topography profiles
+\caption{%
+Modelled bedrock (black) and ice surface (blue) topography profiles
            during deglaciation (22.0--8.0\,\unit{ka}) in the GRIP 5\,\unit{km}
@@
            Figs.~\ref{fig:hr_geom_deglacageshots} and~\ref{fig:hr_geom_deglacage}.}
-  \label{fig:hr_pf_grip}
-\end{figure}
-
-\begin{figure}
+\label{fig:hr_pf_grip}%
+\end{figure}%
+
+\begin{figure}%
 \includegraphics{hr_pf_epica}
-  \caption{Modelled bedrock (black) and ice surface (red) topography profiles
+\caption{%
+Modelled bedrock (black) and ice surface (red) topography profiles
            during deglaciation (22.0--8.0\,\unit{ka}) in the EPICA 5\,\unit{km}
@@
            Figs.~\ref{fig:hr_geom_deglacageshots} and~\ref{fig:hr_geom_deglacage}.}
-  \label{fig:hr_pf_epica}
+\label{fig:hr_pf_epica}%
 \end{figure}
@@
 \end{document}
-\endinput
+\endinput%%%%%%

### 2016-01-18 cordillera-cycle.tex
Typeset: fix figure paths in typset.
To allow for diffs.
Typeset: some table formatting.
I purposely did not apply all changes from the typesetter.

@@
 \begin{table*}
-  \centering
 \caption{Parameter values used in the ice sheet model.}
 \label{tab:params}
+%\scalebox{.85}[.85]
 {\begin{tabular}{llrl}
@@
     \middlehline
-
     $\rho$  & Ice density
@@
             & \unit{m\,s^{-2}} \\
-
-    \multicolumn{2}{l}{\emph{Basal sliding}} \\
-
+    \cline{1-4}
+    \multicolumn{2}{l}{{Basal sliding}} \\
+    \cline{1-4}
     $q$     & Pseudo-plastic sliding exponent
             & 0.25
-            & - \\
+            & -- \\
 
@@
             & 0.02
-            & - \\
+            & -- \\
 
@@
             & 0.69
-            & - \\
+            & -- \\
 
@@
             & 0.12
-            & - \\
+            & -- \\
 
@@
             & m \\
-
-    \multicolumn{2}{l}{\emph{Bedrock and lithosphere}} \\
-
+    \cline{1-4}
+    \multicolumn{2}{l}{{Bedrock and lithosphere}} \\
+    \cline{1-4}
     $\rho_{\mathrm{b}}$& Bedrock density
@@
             & \unit{N} \\
-
-    \multicolumn{2}{l}{\emph{Surface and atmosphere}} \\
-
+    \cline{1-4}
+    \multicolumn{2}{l}{{Surface and atmosphere}} \\
+    \cline{1-4}
     $T_{\mathrm{s}}$   & Temperature of snow precipitation
@@
             & \unit{K\,m^{-1}} \\
-
     \bottomhline
 \end{tabular}}
-  \belowtable{}
 \end{table*}
 
-\begin{table*}[t]
-  \caption{Palaeo-temperature proxy records and scaling parameters yielding
+\begin{table*}
+  \caption{%
+Palaeo-temperature proxy records and scaling parameters yielding
            temperature offset time-series used to force the ice sheet model
@@
 \label{tab:records}
+%\scalebox{.850}[.850]
 {\begin{tabular}{l|ccc|ccc|l}
@@
 \end{tabular}}
-  \belowtable{}
 \end{table*}
@@
 \begin{table*}
-  \caption{Extremes in Cordilleran ice sheet volume and extent corresponding to
+\caption{%
+Extremes in Cordilleran ice sheet volume and extent corresponding to
            MIS~4, 3 and 2 for each of the six low-resolution simulations
@@
 \label{tab:extrema}
-  {\begin{tabular}{l*{3}{|ccc}}
+%\scalebox{.50}[.50]
+{\begin{tabular}{l|ccc|ccc|ccc}
     \tophline
-             & \multicolumn{3}{c}{Age (ka)}
-             & \multicolumn{3}{c}{Ice extent (\unit{10^6\,km^2})}
+             & \multicolumn{3}{c|}{Age (ka)}
+             & \multicolumn{3}{c|}{Ice extent (\unit{10^6\,km^2})}
              & \multicolumn{3}{c}{Ice volume (m\,s.l.e.)} \\
@@
              &   3.84 &   1.75 &   8.81 \\
-    \middlehline
+    \cline{1-10}
     Minimum  &  61.89 &  52.97 &  23.19

### 2016-01-18 cordillera-cycle.tex
Typeset: rewrap text to 76 lines.
I made some errors yesterday.

@@
       Stea.etal.2011}. Nevertheless, recent oxygen isotope measurements from
-      Gulf of Alaska sediments reveal a~climatic evolution highly correlated to
-      that of Greenland during this period, including a~distinct Late Glacial
-      cold reversal between 14.1 and 11.7\,\unit{ka}
+      Gulf of Alaska sediments reveal a~climatic evolution highly correlated
+      to that of Greenland during this period, including a~distinct Late
+      Glacial cold reversal between 14.1 and 11.7\,\unit{ka}
       \citep{Praetorius.Mix.2014}.
@@
       Although numerical ice sheet modelling has been established as
-      a~useful tool to improve our understanding of the Cordilleran ice sheet
-      (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991};
+      a~useful tool to improve our understanding of the Cordilleran ice
+      sheet (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991};
       \citealp{Marshall.etal.2000}), the ubiquitously mountainous topography
@@
       where cohesion, $c_0$, is assumed to be zero. The friction angle,
-      $\phi$, varies from 15 to 45\unit{\degree} as a~piecewise-linear function
-      of modern bed elevation, with the lowest value occuring below modern sea
-      level (0\,\unit{m} above sea level, \unit{m\,a.s.l.}) and the highest
-      value occuring above the generalised elevation of the highest shorelines
-      \citep[200\,\unit{m\,a.s.l.},][Fig.~5]{Clague.1981}, thus accounting for
-      a~weakening of till associated with the presence of marine sediments (cf.
-      \citealp{Martin.etal.2011};
+      $\phi$, varies from 15 to 45\unit{\degree} as a~piecewise-linear
+      function of modern bed elevation, with the lowest value occuring below
+      modern sea level (0\,\unit{m} above sea level, \unit{m\,a.s.l.}) and
+      the highest value occuring above the generalised elevation of the
+      highest shorelines
+      \citep[200\,\unit{m\,a.s.l.},][Fig.~5]{Clague.1981}, thus accounting
+      for a~weakening of till associated with the presence of marine
+      sediments (cf. \citealp{Martin.etal.2011};
       \citealp[Supplement]{Aschwanden.etal.2013};
@@
       from zero to $W_{\text{max}}=2$\,\unit{m}, a~threshold above which
-      additional melt water is assumed to drain off instantaneously. Finally,
-      the bedrock topography responds to ice load following a~bedrock
-      deformation model that includes local isostasy, elastic lithosphere
-      flexure and viscous astenosphere deformation in an infinite half-space
-      \citep{Lingle.Clark.1985,Bueler.etal.2007}. A~relatively low viscosity
-      value of $\nu_{m} = 1\times10^{19}$\,\unit{Pa\,s} is used for the
-      astenosphere (Table~\ref{tab:params}) in accordance with the results from
-      regional glacial isostatic adjustment modelling at the northern Cascadia
+      additional melt water is assumed to drain off instantaneously.
+      Finally, the bedrock topography responds to ice load following
+      a~bedrock deformation model that includes local isostasy, elastic
+      lithosphere flexure and viscous astenosphere deformation in an
+      infinite half-space \citep{Lingle.Clark.1985,Bueler.etal.2007}.
+      A~relatively low viscosity value of $\nu_{m} =
+      1\times10^{19}$\,\unit{Pa\,s} is used for the astenosphere
+      (Table~\ref{tab:params}) in accordance with the results from regional
+      glacial isostatic adjustment modelling at the northern Cascadia
       subduction zone \citep{James.etal.2009}.
@@
       Ice shelf calving is computed using a~double criterion. First,
-      a~physically-realistic calving flux is computed based on eigenvalues of
-      the horizontal strain rate tensor \citep{Winkelmann.etal.2011,
+      a~physically-realistic calving flux is computed based on eigenvalues
+      of the horizontal strain rate tensor \citep{Winkelmann.etal.2011,
       Levermann.etal.2012}. This allows floating ice to advance in confined
@@
       Ice surface accumulation and ablation are computed from monthly mean
-      near-surface air temperature, $T_{\mathrm{m}}$, monthly standard deviation of
-      near-surface air temperature, $\sigma$, and monthly precipitation,
-      $P_{\mathrm{m}}$, using a~temperature-index model \citep[e.g.,][]{Hock.2003}.
-      Accumulation is equal to precipitation when air temperatures are below
-      0\,\unit{{\degree}C}, and decreases to zero linearly with temperatures
-      between 0 and 2\,\unit{{\degree}C}. Ablation is computed from PDD,
-      defined as an integral of temperatures above 0\,\unit{{\degree}C} in
-      one year.
+      near-surface air temperature, $T_{\mathrm{m}}$, monthly standard
+      deviation of near-surface air temperature, $\sigma$, and monthly
+      precipitation, $P_{\mathrm{m}}$, using a~temperature-index model
+      \citep[e.g.,][]{Hock.2003}. Accumulation is equal to precipitation
+      when air temperatures are below 0\,\unit{{\degree}C}, and decreases to
+      zero linearly with temperatures between 0 and 2\,\unit{{\degree}C}.
+      Ablation is computed from PDD, defined as an integral of temperatures
+      above 0\,\unit{{\degree}C} in one year.
 
@@
       assuming a~normal temperature distribution of standard deviation
-      $\sigma$ around the expected value $T_{\mathrm{m}}$. It is expressed by
-      an error-function formulation \citep{Calov.Greve.2005},
+      $\sigma$ around the expected value $T_{\mathrm{m}}$. It is expressed
+      by an error-function formulation \citep{Calov.Greve.2005},
 \begin{align}
@@
 
-      Temperature offset time-series, ${\Delta}T_{\text{TS}}$, are derived from
-      palaeo-temperature proxy records from the Greenland Ice Core Project
-      \citep[GRIP,][]{Dansgaard.etal.1993}, the North Greenland Ice Core
-      Project \citep[NGRIP,][]{Andersen.etal.2004}, the European Project for
-      Ice Coring in Antarctica \citep[EPICA,][] {Jouzel.etal.2007}, the Vostok
-      ice core \citep{Petit.etal.1999}, and Ocean Drilling Program (ODP) sites
-      1012 and 1020, both located off the coast of California
-      \citep{Herbert.etal.2001}. Palaeo-temperature anomalies from the GRIP and
-      NGRIP records were calculated from oxygen isotope (\chem{\delta^{18}O})
-      measurements using a~quadratic equation \citep{Johnsen.etal.1995},
+      Temperature offset time-series, ${\Delta}T_{\text{TS}}$, are derived
+      from palaeo-temperature proxy records from the Greenland Ice Core
+      Project \citep[GRIP,][]{Dansgaard.etal.1993}, the North Greenland Ice
+      Core Project \citep[NGRIP,][]{Andersen.etal.2004}, the European
+      Project for Ice Coring in Antarctica \citep[EPICA,][]
+      {Jouzel.etal.2007}, the Vostok ice core \citep{Petit.etal.1999}, and
+      Ocean Drilling Program (ODP) sites 1012 and 1020, both located off the
+      coast of California \citep{Herbert.etal.2001}. Palaeo-temperature
+      anomalies from the GRIP and NGRIP records were calculated from oxygen
+      isotope (\chem{\delta^{18}O}) measurements using a~quadratic equation
+      \citep{Johnsen.etal.1995},
 \begin{align}
@@
 
-      Finally, lapse-rate corrections, ${\Delta}T_{\text{LR}}$, are computed as
-      a~function of ice surface elevation, $s$, using the NARR surface
+      Finally, lapse-rate corrections, ${\Delta}T_{\text{LR}}$, are computed
+      as a~function of ice surface elevation, $s$, using the NARR surface
       geopotential height invariant field as a~reference topography,
@@
       one hand, and for differences between the basal topography of the ice
-      flow model, $b$, and the NARR reference topography, $b_{\text{ref}}$, on
-      the other hand. All simulations use an annual temperature lapse rate of
-      $\gamma = 6\,\unit{K\,km^{-1}}$. In the rest of this paper, we refer to
-      different model runs by the name of the proxy record used for the
-      palaeo-temperature forcing.
+      flow model, $b$, and the NARR reference topography, $b_{\text{ref}}$,
+      on the other hand. All simulations use an annual temperature lapse
+      rate of $\gamma = 6\,\unit{K\,km^{-1}}$. In the rest of this paper, we
+      refer to different model runs by the name of the proxy record used for
+      the palaeo-temperature forcing.
 
@@
       (Fig.~\ref{fig:lr_ts}, upper panel), model output presents consistent
-      features that can be observed across the range of forcing data used. In
-      all simulations, modelled ice volumes remain relatively low during most
-      of the glacial cycle, except during two major glacial events which occur
-      between 61.9 and 56.5\,\unit{ka} during MIS~4, and between 23.2 and
-      16.8\,\unit{ka} during MIS~2 (Fig.~\ref{fig:lr_ts}, lower panel). An ice
-      volume minimum is consistently reached between 53.0 and 41.3\,\unit{ka}
-      during MIS~3. However, the magnitude and precise timing of these three
-      events depend significantly on the choice of proxy record used to derive
-      a~time-dependent climate forcing (Table~\ref{tab:extrema}).
+      features that can be observed across the range of forcing data used.
+      In all simulations, modelled ice volumes remain relatively low during
+      most of the glacial cycle, except during two major glacial events
+      which occur between 61.9 and 56.5\,\unit{ka} during MIS~4, and between
+      23.2 and 16.8\,\unit{ka} during MIS~2 (Fig.~\ref{fig:lr_ts}, lower
+      panel). An ice volume minimum is consistently reached between 53.0 and
+      41.3\,\unit{ka} during MIS~3. However, the magnitude and precise
+      timing of these three events depend significantly on the choice of
+      proxy record used to derive a~time-dependent climate forcing
+      (Table~\ref{tab:extrema}).
 
@@
       this early deglaciation is caused by an early warming present in the
-      alkenone palaeo-temperature reconstructions (Fig.~\ref{fig:lr_ts}, upper
-      panel; \citealp[Fig.~3]{Herbert.etal.2001}). However, this early warming
-      is a~local effect, corresponding to a~weakening of the California current
-      \citep{Herbert.etal.2001}. The California current, driving cold waters
-      southwards along the south-western coast of North America, has been shown
-      to have weakened during each peak of global glaciation (in SPECMAP)
-      during the past 550\,\unit{ka}, including the LGM, resulting in
-      paradoxically warmer sea-surface temperatures at the locations of the
-      ODP~1012 and ODP~1020 sites \citep{Herbert.etal.2001}.
+      alkenone palaeo-temperature reconstructions (Fig.~\ref{fig:lr_ts},
+      upper panel; \citealp[Fig.~3]{Herbert.etal.2001}). However, this early
+      warming is a~local effect, corresponding to a~weakening of the
+      California current \citep{Herbert.etal.2001}. The California current,
+      driving cold waters southwards along the south-western coast of North
+      America, has been shown to have weakened during each peak of global
+      glaciation (in SPECMAP) during the past 550\,\unit{ka}, including the
+      LGM, resulting in paradoxically warmer sea-surface temperatures at the
+      locations of the ODP~1012 and ODP~1020 sites
+      \citep{Herbert.etal.2001}.
 
@@
       6\,\unit{ka} earlier than dated evidence of the LGM advance. There is
-      a~fair agreement between the EPICA and Vostok palaeo-temperature records,
-      resulting in only small differences between the simulations driven by
-      those records. These differences are not subject to further analysis
-      here; instead we focus on simulations forced by palaeo-temperature
-      records from the GRIP and EPICA ice cores that appear to produce the most
-      realistic reconstructions of regional glaciation history, yet bearing
-      significant disparities in model output. To allow for a~more detailed
-      comparison against the geological record, these two simulations were
-      re-run using a~higher-resolution grid (Sect.~\ref{sec:model};
-      Fig.~\ref{fig:lr_ts}, lower panel, dotted lines).
+      a~fair agreement between the EPICA and Vostok palaeo-temperature
+      records, resulting in only small differences between the simulations
+      driven by those records. These differences are not subject to further
+      analysis here; instead we focus on simulations forced by
+      palaeo-temperature records from the GRIP and EPICA ice cores that
+      appear to produce the most realistic reconstructions of regional
+      glaciation history, yet bearing significant disparities in model
+      output. To allow for a~more detailed comparison against the geological
+      record, these two simulations were re-run using a~higher-resolution
+      grid (Sect.~\ref{sec:model}; Fig.~\ref{fig:lr_ts}, lower panel, dotted
+      lines).
 
@@
       which likely resulted in a~decrease of precipitation and, therefore,
-      a~decrease of accumulation in the interior. Secondly, latent warming of
-      the moisture-depleted air parcels flowing over this enhanced
+      a~decrease of accumulation in the interior. Secondly, latent warming
+      of the moisture-depleted air parcels flowing over this enhanced
       topography could have resulted in an inflow of potentially warmer air
@@
       5.04\,\unit{m}\,s.l.e. (61.9\,\unit{ka}) in the EPICA simulation,
-      corresponding to respectively 90 and 57\,\unit{\%} of modelled MIS~2 ice
-      volumes.
+      corresponding to respectively 90 and 57\,\unit{\%} of modelled MIS~2
+      ice volumes.
 
@@
       timespan of the last glacial cycle (Fig.~\ref{fig:hr_geom_duration},
-      hatched areas). However, except for its margins in the Pacific Ocean and
-      in the northern foothills of the Alaska Range, the maximum extent of the
-      ice sheet is attained for a~much shorter period of time of only few
-      thousand years (Fig.~\ref{fig:hr_geom_duration}, red areas). This result
-      illustrates that the maximum extents of the modelled ice sheet during
-      MIS~4 and MIS~2 were both short-lived and therefore out of balance with
-      contemporary climate.
+      hatched areas). However, except for its margins in the Pacific Ocean
+      and in the northern foothills of the Alaska Range, the maximum extent
+      of the ice sheet is attained for a~much shorter period of time of only
+      few thousand years (Fig.~\ref{fig:hr_geom_duration}, red areas). This
+      result illustrates that the maximum extents of the modelled ice sheet
+      during MIS~4 and MIS~2 were both short-lived and therefore out of
+      balance with contemporary climate.
 
@@
       apparent insensitivity of modelled glacial extent to temperature
-      fluctuations results from a~combination of low precipitation, high summer
-      temperature and large temperature standard deviation (PDD SD) in the
-      plains of the Alaska Interior (Fig.~\ref{fig:atm}) which confines
-      glaciation to the foothills of the mountains. This result could
-      potentially explain the local distribution of glacial deposits, which
-      indicates that glaciers flowing on the northern slope of the Alaska Range
-      have remained small throughout the Pleistocene
+      fluctuations results from a~combination of low precipitation, high
+      summer temperature and large temperature standard deviation (PDD SD)
+      in the plains of the Alaska Interior (Fig.~\ref{fig:atm}) which
+      confines glaciation to the foothills of the mountains. This result
+      could potentially explain the local distribution of glacial deposits,
+      which indicates that glaciers flowing on the northern slope of the
+      Alaska Range have remained small throughout the Pleistocene
       \citep{Kaufman.Manley.2004}.
@@
       centres, visible on the modelled ice cover duration maps
-      (Fig.~\ref{fig:hr_geom_duration}), are located over the Coast Mountains
-      (CM), the Columbia and Rocky mountains (CRM), the Skeena Mountains (SM),
-      and the Selwyn and Mackenzie mountains (SMKM). The Wrangell and Saint
-      Elias mountains, heavily glacierized at present, host an ice cap for the
-      entire length of both simulations, but that ice cap does not appear to be
-      a~major feed to the Cordilleran ice sheet
-      (Fig.~\ref{fig:hr_geom_duration}, WSEM). Although the Coast, Skeena and
-      Columbia and Rocky mountains (CM, SM, CRM) are covered by mountain
-      glaciers for most of the last glacial cycle, providing durable nucleation
-      centres for an ice sheet, this is not the case for the Selwyn and
-      Mackenzie mountains (SMKM), where ice cover on the highest peaks is
-      limited to a~small fraction of the last glacial cycle. In other words,
-      the Selwyn and Mackenzie mountains only appear as a~secondary
-      ice-dispersal centre during the coldest periods of the last glacial
-      cycle. The Northern Rocky Mountains (Fig.~\ref{fig:hr_geom_duration},
-      NRM) do not act as a~nucleation centre, but rather as a~pinning point for
-      the Cordilleran ice sheet margin coming from the west.
+      (Fig.~\ref{fig:hr_geom_duration}), are located over the Coast
+      Mountains (CM), the Columbia and Rocky mountains (CRM), the Skeena
+      Mountains (SM), and the Selwyn and Mackenzie mountains (SMKM). The
+      Wrangell and Saint Elias mountains, heavily glacierized at present,
+      host an ice cap for the entire length of both simulations, but that
+      ice cap does not appear to be a~major feed to the Cordilleran ice
+      sheet (Fig.~\ref{fig:hr_geom_duration}, WSEM). Although the Coast,
+      Skeena and Columbia and Rocky mountains (CM, SM, CRM) are covered by
+      mountain glaciers for most of the last glacial cycle, providing
+      durable nucleation centres for an ice sheet, this is not the case for
+      the Selwyn and Mackenzie mountains (SMKM), where ice cover on the
+      highest peaks is limited to a~small fraction of the last glacial
+      cycle. In other words, the Selwyn and Mackenzie mountains only appear
+      as a~secondary ice-dispersal centre during the coldest periods of the
+      last glacial cycle. The Northern Rocky Mountains
+      (Fig.~\ref{fig:hr_geom_duration}, NRM) do not act as a~nucleation
+      centre, but rather as a~pinning point for the Cordilleran ice sheet
+      margin coming from the west.
 
@@
       its predominance over the other ice-dispersal centres
-      (Figs.~\ref{fig:lr_maps} and~\ref{fig:hr_geom_duration}, SM). Regardless
-      of the applied forcing, this ice cap appears to survive MIS~3
-      (Fig.~\ref{fig:lr_maps}, middle panels), and serves as a~nucleation
-      centre at the onset of the glacial readvance towards the LGM (MIS~2).
-      This situation appear similar to the neighbouring Laurentide ice sheet,
-      for which the importance of residual ice for the glacial history leading
-      up to the LGM has been illustrated by the MIS~3 residual ice bodies in
-      northern and eastern Canada as nucleation centres for a~much more
-      extensive MIS~2 configuration \citep{Kleman.etal.2010}.
+      (Figs.~\ref{fig:lr_maps} and~\ref{fig:hr_geom_duration}, SM).
+      Regardless of the applied forcing, this ice cap appears to survive
+      MIS~3 (Fig.~\ref{fig:lr_maps}, middle panels), and serves as
+      a~nucleation centre at the onset of the glacial readvance towards the
+      LGM (MIS~2). This situation appear similar to the neighbouring
+      Laurentide ice sheet, for which the importance of residual ice for the
+      glacial history leading up to the LGM has been illustrated by the
+      MIS~3 residual ice bodies in northern and eastern Canada as nucleation
+      centres for a~much more extensive MIS~2 configuration
+      \citep{Kleman.etal.2010}.
 
@@
       (42.9\,\unit{ka}) in the GRIP simulation and of 2.68\,\unit{m}\,s.l.e.
-      (51.8\,\unit{ka}) in the EPICA simulation, corresponding to respectively
-      19 and 31\,\unit{\%} of the MIS~2 ice volumes.
+      (51.8\,\unit{ka}) in the EPICA simulation, corresponding to
+      respectively 19 and 31\,\unit{\%} of the MIS~2 ice volumes.
 
@@
 
-      A~correlation is observed between the modelled duration of warm based ice
-      cover (Fig.~\ref{fig:hr_geom_warmbase}) and the degree of glacial
-      modification of the landscape (mainly in terms of the development of deep
-      glacial valleys and troughs). We find evidence for this on the slopes of
-      the Coast Mountains, the Columbia and Rocky mountains, the Wrangell
-      and Saint Elias mountains, and radiating off the Skeena Mountains
-      (Figs.~\ref{fig:hr_geom_duration} and \ref{fig:hr_geom_warmbase};
-      \citealp[Fig.~2]{Kleman.etal.2010}). The Skeena Mountains, for example,
-      indeed bear a~strong glacial imprint that indicates ice drainage in
-      a~system of distinct glacial troughs emanating in a~radial pattern from
-      the centre of the mountain range \citep[Fig.~2]{Kleman.etal.2010}. We
-      suggest that persistent ice cover (Fig.~\ref{fig:hr_geom_duration})
-      associated with basal ice temperatures at the pressure melting point
+      A~correlation is observed between the modelled duration of warm based
+      ice cover (Fig.~\ref{fig:hr_geom_warmbase}) and the degree of glacial
+      modification of the landscape (mainly in terms of the development of
+      deep glacial valleys and troughs). We find evidence for this on the
+      slopes of the Coast Mountains, the Columbia and Rocky mountains,
+      the Wrangell and Saint Elias mountains, and radiating off the Skeena
+      Mountains (Figs.~\ref{fig:hr_geom_duration} and
+      \ref{fig:hr_geom_warmbase}; \citealp[Fig.~2]{Kleman.etal.2010}). The
+      Skeena Mountains, for example, indeed bear a~strong glacial imprint
+      that indicates ice drainage in a~system of distinct glacial troughs
+      emanating in a~radial pattern from the centre of the mountain range
+      \citep[Fig.~2]{Kleman.etal.2010}. We suggest that persistent ice cover
+      (Fig.~\ref{fig:hr_geom_duration}) associated with basal ice
+      temperatures at the pressure melting point
       (Figs.~\ref{fig:hr_geom_warmbase} and~\ref{fig:hr_geom_warmfrac})
       explains the large-scale glacial erosional imprint on the landscape.
-      A~well-developed network of glacial valleys running to the north-west on
-      the west slope of the Selwyn and Mackenzie Mountains
+      A~well-developed network of glacial valleys running to the north-west
+      on the west slope of the Selwyn and Mackenzie Mountains
       (\citealp[Fig.~2]{Kleman.etal.2010}; \citealp[Fig.~8]
-      {Stroeven.etal.2010}) is modelled to have hosted predominantly warm-based
-      ice (Fig.~\ref{fig:hr_geom_warmfrac}). However, because it is only
-      glaciated for a~short fraction of the last glacial cycle in our
-      simulations (Fig.~\ref{fig:hr_geom_duration}), this perhaps indicates
-      that the observed landscape pattern originates from multiple glacial
-      cycles and witnesses an increased relative importance of the Selwyn and
-      Mackenzie mountains ice dispersal centre
+      {Stroeven.etal.2010}) is modelled to have hosted predominantly
+      warm-based ice (Fig.~\ref{fig:hr_geom_warmfrac}). However, because it
+      is only glaciated for a~short fraction of the last glacial cycle in
+      our simulations (Fig.~\ref{fig:hr_geom_duration}), this perhaps
+      indicates that the observed landscape pattern originates from multiple
+      glacial cycles and witnesses an increased relative importance of the
+      Selwyn and Mackenzie mountains ice dispersal centre
       (Fig.~\ref{fig:hr_geom_duration}, SMKM), prior to the Late Pleistocene
@@
       70\,\unit{mW\,m^{-2}} geothermal heat flux at 3\,\unit{km} depth
-      (Sect.~\ref{sec:icedyn}). However, the Skeena Mountains and the area west
-      of the Mackenzie Mountains experience higher-than-average geothermal heat
-      flux with measured values of ca.~80 and ca.~100\,\unit{mW\,m^{-2}}
-      \citep{Blackwell.Richards.2004}. We can therefore expect even longer
-      durations of warm-based ice cover for these areas if we were to include
-      spatially variable geothermal forcing in our Cordilleran ice sheet
-      simulations.
+      (Sect.~\ref{sec:icedyn}). However, the Skeena Mountains and the area
+      west of the Mackenzie Mountains experience higher-than-average
+      geothermal heat flux with measured values of ca.~80 and
+      ca.~100\,\unit{mW\,m^{-2}} \citep{Blackwell.Richards.2004}. We can
+      therefore expect even longer durations of warm-based ice cover for
+      these areas if we were to include spatially variable geothermal
+      forcing in our Cordilleran ice sheet simulations.
 
@@
 
-      Modelled patterns of ice sheet retreat are relatively consistent between
-      the two simulations (Figs.~\ref{fig:hr_geom_deglacageshots}
+      Modelled patterns of ice sheet retreat are relatively consistent
+      between the two simulations (Figs.~\ref{fig:hr_geom_deglacageshots}
       and~\ref{fig:hr_geom_deglacage}). The southern sector of the modelling
@@
       Wrangell and Saint Elias mountains in the northern sector of the
-      modelling domain. After 10\,\unit{ka}, deglaciation continues to proceed
-      across the Liard Lowland with a~radial ice margin retreat towards the
-      surrounding mountain ranges, consistent with the regional melt water
-      record of the last deglaciation \citep{Margold.etal.2013}. Remaining ice
-      continues to decay by retreating towards the Selwyn and MacKenzie, and
-      Skeena mountains. The last remnants of the Cordilleran ice sheet finally
-      disappear from the Skeena Mountains at 6.5\,\unit{ka} (GRIP) and
-      6.1\,\unit{ka} (EPICA).
+      modelling domain. After 10\,\unit{ka}, deglaciation continues to
+      proceed across the Liard Lowland with a~radial ice margin retreat
+      towards the surrounding mountain ranges, consistent with the regional
+      melt water record of the last deglaciation \citep{Margold.etal.2013}.
+      Remaining ice continues to decay by retreating towards the Selwyn and
+      MacKenzie, and Skeena mountains. The last remnants of the Cordilleran
+      ice sheet finally disappear from the Skeena Mountains at
+      6.5\,\unit{ka} (GRIP) and 6.1\,\unit{ka} (EPICA).
 
@@
       (Fig.~\ref{fig:hr_geom_deglacage}, OM) presents a~different kind of
-      evidence. There, sharp-crested moraines indicate a~late-glacial readvance
-      of local alpine glaciers and, more importantly, their interaction with
-      larger, lingering remnants of the main body of the Cordilleran ice sheet
-      in the valleys \citep{Lakeman.etal.2008}. Additional evidence for
-      late-glacial alpine glacier readvances includes moraines in the eastern
-      Coast Mountains, and the Columbia and Rocky mountains
-      \citep{Osborn.Gerloff.1997, Menounos.etal.2008}.
+      evidence. There, sharp-crested moraines indicate a~late-glacial
+      readvance of local alpine glaciers and, more importantly, their
+      interaction with larger, lingering remnants of the main body of the
+      Cordilleran ice sheet in the valleys \citep{Lakeman.etal.2008}.
+      Additional evidence for late-glacial alpine glacier readvances
+      includes moraines in the eastern Coast Mountains, and the Columbia and
+      Rocky mountains \citep{Osborn.Gerloff.1997, Menounos.etal.2008}.
 
@@
       distribution, and to identify the potential climatic triggers
-      \citep{Menounos.etal.2008}, it is interesting to note that the simulation
-      driven by the GRIP record produces a~late-glacial readvance in the Coast
-      Mountains and in the Columbia and Rocky Mountains
-      (Fig.~\ref{fig:hr_geom_deglacage}, left panel). In addition to matching
-      the location of some local readvances, the GRIP-driven simulation shows
-      that a~large remnant of the decaying ice sheet may still have existed at
-      the time of this late-glacial readvance. In contrast, the EPICA-driven
-      simulation produces a~nearly-continuous deglaciation with only a~tightly
-      restricted late-glacial readvance on the western slopes of the Saint
-      Elias and the Coast mountains (Fig.~\ref{fig:hr_geom_deglacage}, right
-      panel).
+      \citep{Menounos.etal.2008}, it is interesting to note that the
+      simulation driven by the GRIP record produces a~late-glacial readvance
+      in the Coast Mountains and in the Columbia and Rocky Mountains
+      (Fig.~\ref{fig:hr_geom_deglacage}, left panel). In addition to
+      matching the location of some local readvances, the GRIP-driven
+      simulation shows that a~large remnant of the decaying ice sheet may
+      still have existed at the time of this late-glacial readvance. In
+      contrast, the EPICA-driven simulation produces a~nearly-continuous
+      deglaciation with only a~tightly restricted late-glacial readvance on
+      the western slopes of the Saint Elias and the Coast mountains
+      (Fig.~\ref{fig:hr_geom_deglacage}, right panel).
 
@@
 
-      Because a~general tenant in glacial geomorphology is that the majority of
-      landforms (lineations and eskers) are part of the deglacial envelope
-      \citep[terminology from][]{Kleman.etal.2006}, having been formed close
-      inside the retreating margin of ice sheets \citep{Boulton.Clark.1990,
-      Kleman.etal.1997, Kleman.etal.2010}, we present maps of basal flow
-      directions immediately preceding deglaciation or at the time of cessation
-      of sliding inside a~cold-based retreating margin
-      (Fig.~\ref{fig:hr_geom_lastflow}). The modelled deglacial flow patterns
-      are mostly consistent between the GRIP and EPICA simulations. They depict
-      an active ice sheet retreat in the peripheral areas, followed by stagnant
-      ice decay in some of the interior regions. Several parts of the modelling
-      domain do not experience any basal sliding throughout the deglaciation
-      phase (Fig.~\ref{fig:hr_geom_lastflow}, hatched areas). This notably
+      Because a~general tenant in glacial geomorphology is that the majority
+      of landforms (lineations and eskers) are part of the deglacial
+      envelope \citep[terminology from][]{Kleman.etal.2006}, having been
+      formed close inside the retreating margin of ice sheets
+      \citep{Boulton.Clark.1990, Kleman.etal.1997, Kleman.etal.2010}, we
+      present maps of basal flow directions immediately preceding
+      deglaciation or at the time of cessation of sliding inside
+      a~cold-based retreating margin (Fig.~\ref{fig:hr_geom_lastflow}). The
+      modelled deglacial flow patterns are mostly consistent between the
+      GRIP and EPICA simulations. They depict an active ice sheet retreat in
+      the peripheral areas, followed by stagnant ice decay in some of the
+      interior regions. Several parts of the modelling domain do not
+      experience any basal sliding throughout the deglaciation phase
+      (Fig.~\ref{fig:hr_geom_lastflow}, hatched areas). This notably
       includes parts of the Interior Plateau in British Columbia, major
@@
       \citealp[Fig.~1.12]{Clague.1989}; \citealp[Fig.~2]{Kleman.etal.2010})
-      show similarities with the patterns of deglacial ice flow from numerical
-      modelling (Fig.~\ref{fig:hr_geom_lastflow}). In the northern half of the
-      modelling domain, modelled deglacial flow directions depict an active
-      downhill flow as the last remnants of the ice sheet retreat towards
-      mountain ranges. Converging deglacial flow patterns in the Liard Lowland,
-      for instance (Fig.~\ref{fig:hr_geom_lastflow}), closely resemble the
-      pattern indicated by glacial lineations
+      show similarities with the patterns of deglacial ice flow from
+      numerical modelling (Fig.~\ref{fig:hr_geom_lastflow}). In the northern
+      half of the modelling domain, modelled deglacial flow directions
+      depict an active downhill flow as the last remnants of the ice sheet
+      retreat towards mountain ranges. Converging deglacial flow patterns in
+      the Liard Lowland, for instance (Fig.~\ref{fig:hr_geom_lastflow}),
+      closely resemble the pattern indicated by glacial lineations
       \citep[Fig.~2]{Margold.etal.2013}.
@@
       simulations produce a~retreat of the ice margin towards the north-east
-      (Fig.~\ref{fig:hr_geom_deglacage}), a~pattern which is validated by the
-      geomorphological and stratigraphical record for ancient pro-glacial lakes
-      dammed by the retreating ice sheet \citep{Perkins.Brennand.2014}.
-      However, the two simulations differ in the mode of retreat. The GRIP
-      simulation yields an active retreat with basal sliding towards the ice
-      margin to the south, whereas the EPICA simulation produces negligible
-      basal sliding on the plateau during deglaciation
-      (Fig.~\ref{fig:hr_geom_lastflow}). Yet, the Interior Plateau also hosts
-      an impressive set of glacial lineations which indicate a~substantial
-      eastwards flow component of the Cordilleran ice sheet
-      \citep{Prest.etal.1968, Kleman.etal.2010}. This Interior Plateau
-      lineation set could therefore present a~smoking gun for the reliability
-      of the presented model results. One explanation for the incongruent
-      results could be that the missing feedback mechanisms between ice sheet
-      topography and regional climate resulted in a~modelled ice divide of the
-      LGM ice sheet being too far to the east (Sect.~\ref{sec:mis2};
-      Fig.~\ref{fig:mis2}; \citealp{Seguinot.etal.2014}). A~more
-      westerly-located LGM ice divide would certainly result in a~different
-      deglacial flow pattern over the Interior Plateau. However, a~more
-      westerly-positioned LGM ice divide would certainly be associated with
-      a~thinner ice sheet than that modelled here. Decreased ice thickness
-      would not promote warm-based conditions but, on the contrary, enlarge the
-      region of negligible basal sliding (Fig.~\ref{fig:hr_geom_lastflow}).
-      Thus, a~second explanation for the incongruent results could be that the
-      Interior Plateau lineation system predates deglaciation ice flow, as
-      perhaps indicated by some eskers that appear incompatible with these
-      glacial lineations \citep[Fig.~9]{Margold.etal.2013a}. Finally, a~third
-      explanation could be that local geothermal heat associated with volcanic
-      activity on the Interior Plateau could have triggered the basal sliding
-      \citep[cf. Greenland ice sheet;][]{Fahnestock.etal.2001}.
+      (Fig.~\ref{fig:hr_geom_deglacage}), a~pattern which is validated by
+      the geomorphological and stratigraphical record for ancient
+      pro-glacial lakes dammed by the retreating ice sheet
+      \citep{Perkins.Brennand.2014}. However, the two simulations differ in
+      the mode of retreat. The GRIP simulation yields an active retreat with
+      basal sliding towards the ice margin to the south, whereas the EPICA
+      simulation produces negligible basal sliding on the plateau during
+      deglaciation (Fig.~\ref{fig:hr_geom_lastflow}). Yet, the Interior
+      Plateau also hosts an impressive set of glacial lineations which
+      indicate a~substantial eastwards flow component of the Cordilleran ice
+      sheet \citep{Prest.etal.1968, Kleman.etal.2010}. This Interior Plateau
+      lineation set could therefore present a~smoking gun for the
+      reliability of the presented model results. One explanation for the
+      incongruent results could be that the missing feedback mechanisms
+      between ice sheet topography and regional climate resulted in
+      a~modelled ice divide of the LGM ice sheet being too far to the east
+      (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
+      \citealp{Seguinot.etal.2014}). A~more westerly-located LGM ice divide
+      would certainly result in a~different deglacial flow pattern over the
+      Interior Plateau. However, a~more westerly-positioned LGM ice divide
+      would certainly be associated with a~thinner ice sheet than that
+      modelled here. Decreased ice thickness would not promote warm-based
+      conditions but, on the contrary, enlarge the region of negligible
+      basal sliding (Fig.~\ref{fig:hr_geom_lastflow}). Thus, a~second
+      explanation for the incongruent results could be that the Interior
+      Plateau lineation system predates deglaciation ice flow, as perhaps
+      indicated by some eskers that appear incompatible with these glacial
+      lineations \citep[Fig.~9]{Margold.etal.2013a}. Finally, a~third
+      explanation could be that local geothermal heat associated with
+      volcanic activity on the Interior Plateau could have triggered the
+      basal sliding \citep[cf. Greenland ice
+      sheet;][]{Fahnestock.etal.2001}.
 
       The modelled deglaciation of the Interior Plateau of British Columbia
-      consists of a~rapid northwards retreat (Fig.~\ref{fig:hr_geom_deglacage})
-      of southwards-flowing non-sliding ice lobes
-      (Fig.~\ref{fig:hr_geom_lastflow}) positioned in-between deglaciated
-      mountain ranges (Figs.~\ref{fig:hr_pf_grip} and~\ref{fig:hr_pf_epica}).
-      This result appears compatible with the prevailing conceptual model of
-      deglaciation of central British Columbia, in which mountain ranges emerge
-      from the ice before the plateau \citep[Fig.~7]{Fulton.1991}. However, due
-      to different topographic and climatic conditions, our simulations produce
-      different deglaciation patterns in the northern half of the model domain,
+      consists of a~rapid northwards retreat
+      (Fig.~\ref{fig:hr_geom_deglacage}) of southwards-flowing non-sliding
+      ice lobes (Fig.~\ref{fig:hr_geom_lastflow}) positioned in-between
+      deglaciated mountain ranges (Figs.~\ref{fig:hr_pf_grip}
+      and~\ref{fig:hr_pf_epica}). This result appears compatible with the
+      prevailing conceptual model of deglaciation of central British
+      Columbia, in which mountain ranges emerge from the ice before the
+      plateau \citep[Fig.~7]{Fulton.1991}. However, due to different
+      topographic and climatic conditions, our simulations produce different
+      deglaciation patterns in the northern half of the model domain,
       indicating that this conceptual model may not be applied to the entire
@@
       glacial cycle presented in this study consistently produce two glacial
-      maxima during MIS~4 (61.9--56.5\,\unit{ka}, 3.8--8.8\,\unit{m}\,s.l.e.)
-      and MIS~2 (23.2--16.8\,\unit{ka}, 8.40--8.91\,\unit{m}\,s.l.e.), two
-      periods corresponding to documented extensive glaciations. This result is
-      independent of the palaeo-temperature record used among the six selected
-      for this study, and thus can be regarded as a~robust model output, which
-      broadly matches geological evidence. However, the timing of the two
-      glaciation peaks depends sensitively on which climate record is used to
-      drive the model. The timing of the LGM is best reproduced by the EPICA
-      and Vostok Antarctic ice core records. It occurs about 2\,\unit{ka} too
+      maxima during MIS~4 (61.9--56.5\,\unit{ka},
+      3.8--8.8\,\unit{m}\,s.l.e.) and MIS~2 (23.2--16.8\,\unit{ka},
+      8.40--8.91\,\unit{m}\,s.l.e.), two periods corresponding to documented
+      extensive glaciations. This result is independent of the
+      palaeo-temperature record used among the six selected for this study,
+      and thus can be regarded as a~robust model output, which broadly
+      matches geological evidence. However, the timing of the two glaciation
+      peaks depends sensitively on which climate record is used to drive the
+      model. The timing of the LGM is best reproduced by the EPICA and
+      Vostok Antarctic ice core records. It occurs about 2\,\unit{ka} too
       early in the simulation forced by the GRIP ice core record, and occurs
       even earlier in all other simulations. The mismatch is largest for the
-      two Northwest Pacific ODP palaeo-temperature records, which are affected
-      by the weakening of the California current during the LGM.
+      two Northwest Pacific ODP palaeo-temperature records, which are
+      affected by the weakening of the California current during the LGM.
 
@@
       to be fed by the moisture intruding inland from the west through
-      a~topographic breach in the Coast Mountains. The Skeena ice cap acts as
-      the main nucleation centre for the glacial reavance towards the LGM
+      a~topographic breach in the Coast Mountains. The Skeena ice cap acts
+      as the main nucleation centre for the glacial reavance towards the LGM
       configuration. As indicated by persistent, warm-based ice in the
@@
 \begin{acknowledgements}
-      Foremost, we would like to thank Shawn Marshall for providing a~detailed,
-      constructive analysis of this study during J.~Seguinot's Ph.D. defence
-      (September 2014). His comments were used to improve the model set-up. We
-      are very thankful to Constantine Khroulev, Ed Bueler, and Andy Aschwanden
-      for providing constant help and development with PISM. This work was
-      supported by the Swedish Research Council~(VR) grant no.~2008-3449 to
-      A.~P.~Stroeven and by the German Academic Exchange Service~(DAAD) grant
-      no.~50015537 and a~Knut and Alice Wallenberg Foundation grant to
-      J.~Seguinot. Computer resources were provided by the Swedish National
-      Infrastructure for Computing (SNIC) allocation no.~2013/1-159
-      and~2014/1-159 to A.~P.~Stroeven at the National Supercomputing Center
-      (NSC).
+      Foremost, we would like to thank Shawn Marshall for providing
+      a~detailed, constructive analysis of this study during J.~Seguinot's
+      Ph.D. defence (September 2014). His comments were used to improve the
+      model set-up. We are very thankful to Constantine Khroulev, Ed Bueler,
+      and Andy Aschwanden for providing constant help and development with
+      PISM. This work was supported by the Swedish Research Council~(VR)
+      grant no.~2008-3449 to A.~P.~Stroeven and by the German Academic
+      Exchange Service~(DAAD) grant no.~50015537 and a~Knut and Alice
+      Wallenberg Foundation grant to J.~Seguinot. Computer resources were
+      provided by the Swedish National Infrastructure for Computing (SNIC)
+      allocation no.~2013/1-159 and~2014/1-159 to A.~P.~Stroeven at the
+      National Supercomputing Center (NSC).
 \end{acknowledgements}

### 2016-01-18 typeset.tex
Typeset: add trailing percent signs.
Typeset: fix figure paths in typset.
To allow for diffs.
Typeset: some table formatting.
I purposely did not apply all changes from the typesetter.
Typeset: rewrap text to 76 lines.
I made some errors yesterday.

--- 2016-01-17 typeset.tex
@@
-\documentclass[tcd, online, hvmath]{copernicus}
+\documentclass[tc, manuscript]{copernicus}
+\graphicspath{{figures/}}
 
@@
       \citep[e.g.,][]{Sissons.1979, Lundqvist.1987, Ivy-Ochs.etal.1999,
-      Stea.etal.2011}. Nevertheless, recent oxygen isotope measurements
-      from Gulf of Alaska sediments reveal a~climatic evolution highly
-      correlated to that of Greenland during this period, including
-      a~distinct Late Glacial cold reversal between 14.1 and 11.7\,\unit{ka}
+      Stea.etal.2011}. Nevertheless, recent oxygen isotope measurements from
+      Gulf of Alaska sediments reveal a~climatic evolution highly correlated
+      to that of Greenland during this period, including a~distinct Late
+      Glacial cold reversal between 14.1 and 11.7\,\unit{ka}
       \citep{Praetorius.Mix.2014}.
@@
       function of modern bed elevation, with the lowest value occuring below
-      modern sea level (0\,\unit{m\,a.s.l.}) and the highest value occuring
-      above the generalised elevation of the highest shorelines
+      modern sea level (0\,\unit{m\,a.s.l.}) and
+      the highest value occuring above the generalised elevation of the
+      highest shorelines
       \citep[200\,\unit{m\,a.s.l.},][Fig.~5]{Clague.1981}, thus accounting
@@
       from zero to $W_{\text{max}}=2$\,\unit{m}, a~threshold above which
-      additional melt water is assumed to drain off
-      instantaneously. Finally, the bedrock topography responds to ice load
-      following a~bedrock deformation model that includes local isostasy,
-      elastic lithosphere flexure and viscous astenosphere deformation in an
+      additional melt water is assumed to drain off instantaneously.
+      Finally, the bedrock topography responds to ice load following
+      a~bedrock deformation model that includes local isostasy, elastic
+      lithosphere flexure and viscous astenosphere deformation in an
       infinite half-space \citep{Lingle.Clark.1985,Bueler.etal.2007}.
@@
       on the other hand. All simulations use an annual temperature lapse
-      rate of $\gamma = 6\,\unit{K\,km^{-1}}$. In the rest of this paper,
-      we refer to different model runs by the name of the proxy record used
-      for the palaeo-temperature forcing.
+      rate of $\gamma = 6\,\unit{K\,km^{-1}}$. In the rest of this paper, we
+      refer to different model runs by the name of the proxy record used for
+      the palaeo-temperature forcing.
 
@@
       Despite large differences in the input climate forcing
-      (Fig.~\ref{fig:lr_ts}, upper panel), model output presents
-      consistent features that can be observed across the range of forcing
-      data used. In all simulations, modelled ice volumes remain relatively
-      low during most of the glacial cycle, except during two major glacial
-      events which occur between 61.9 and 56.5\,\unit{ka} during MIS~4, and
-      between 23.2 and 16.8\,\unit{ka} during MIS~2
-      (Fig.~\ref{fig:lr_ts}, lower panel). An ice volume minimum is
-      consistently reached between 53.0 and 41.3\,\unit{ka} during
-      MIS~3. However, the magnitude and precise timing of these three events
-      depend significantly on the choice of proxy record used to derive
-      a~time-dependent climate forcing (Table~\ref{tab:extrema}).
+      (Fig.~\ref{fig:lr_ts}, upper panel), model output presents consistent
+      features that can be observed across the range of forcing data used.
+      In all simulations, modelled ice volumes remain relatively low during
+      most of the glacial cycle, except during two major glacial events
+      which occur between 61.9 and 56.5\,\unit{ka} during MIS~4, and between
+      23.2 and 16.8\,\unit{ka} during MIS~2 (Fig.~\ref{fig:lr_ts}, lower
+      panel). An ice volume minimum is consistently reached between 53.0 and
+      41.3\,\unit{ka} during MIS~3. However, the magnitude and precise
+      timing of these three events depend significantly on the choice of
+      proxy record used to derive a~time-dependent climate forcing
+      (Table~\ref{tab:extrema}).
 
@@
       this early deglaciation is caused by an early warming present in the
-      alkenone palaeo-temperature reconstructions
-      (Fig.~\ref{fig:lr_ts}, upper panel;
-      \citealp[Fig.~3]{Herbert.etal.2001}). However, this early warming is
-      a~local effect, corresponding to a~weakening of the California current
-      \citep{Herbert.etal.2001}. The California current, driving cold waters
-      southwards along the south-western coast of North America, has been
-      shown to have weakened during each peak of global glaciation (in
-      SPECMAP) during the past 550\,\unit{ka}, including the LGM, resulting
-      in paradoxically warmer sea-surface temperatures at the locations of
-      the ODP~1012 and ODP~1020 sites \citep{Herbert.etal.2001}.
+      alkenone palaeo-temperature reconstructions (Fig.~\ref{fig:lr_ts},
+      upper panel; \citealp[Fig.~3]{Herbert.etal.2001}). However, this early
+      warming is a~local effect, corresponding to a~weakening of the
+      California current \citep{Herbert.etal.2001}. The California current,
+      driving cold waters southwards along the south-western coast of North
+      America, has been shown to have weakened during each peak of global
+      glaciation (in SPECMAP) during the past 550\,\unit{ka}, including the
+      LGM, resulting in paradoxically warmer sea-surface temperatures at the
+      locations of the ODP~1012 and ODP~1020 sites
+      \citep{Herbert.etal.2001}.
 
@@
       the NGRIP palaeo-temperature record depicts an early onset of
-      deglaciation (Fig.~\ref{fig:lr_ts}) following its last glacial
-      volume maximum (22.9\,\unit{ka}, Table~\ref{tab:extrema}) attained
-      about 6\,\unit{ka} earlier than dated evidence of the LGM advance.
-      There is a~fair agreement between the EPICA and Vostok
-      palaeo-temperature records, resulting in only small differences
-      between the simulations driven by those records. These differences are
-      not subject to further analysis further; instead we focus on
-      simulations forced by palaeo-temperature records from the GRIP and
-      EPICA ice cores that appear to produce the most realistic
-      reconstructions of regional glaciation history, yet bearing
-      significant disparities in model output. To allow for a~more detailed
-      comparison against the geological record, these two simulations were
-      re-run using a~higher-resolution grid (Sect.~\ref{sec:model};
-      Fig.~\ref{fig:lr_ts}, lower panel, dotted lines).
+      deglaciation (Fig.~\ref{fig:lr_ts}) following its last glacial volume
+      maximum (22.9\,\unit{ka}, Table~\ref{tab:extrema}) attained about
+      6\,\unit{ka} earlier than dated evidence of the LGM advance. There is
+      a~fair agreement between the EPICA and Vostok palaeo-temperature
+      records, resulting in only small differences between the simulations
+      driven by those records. These differences are not subject to further
+      analysis further; instead we focus on simulations forced by
+      palaeo-temperature records from the GRIP and EPICA ice cores that
+      appear to produce the most realistic reconstructions of regional
+      glaciation history, yet bearing significant disparities in model
+      output. To allow for a~more detailed comparison against the geological
+      record, these two simulations were re-run using a~higher-resolution
+      grid (Sect.~\ref{sec:model}; Fig.~\ref{fig:lr_ts}, lower panel, dotted
+      lines).
 
@@
       operation for at most 32\,\unit{ka}, which is about a~third of the
-      timespan of the last glacial cycle (Fig.~\ref{fig:hr_geom_duration}, hatched
-      areas). However, except for its margins in the Pacific Ocean and in
-      the northern foothills of the Alaska Range, the maximum extent of the
-      ice sheet is attained for a~much shorter period of time of only few
-      thousand years (Fig.~\ref{fig:hr_geom_duration}, red areas). This result
-      illustrates that the maximum extents of the modelled ice sheet during
-      MIS~4 and MIS~2 were both short-lived and therefore out of balance
-      with contemporary climate.
+      timespan of the last glacial cycle (Fig.~\ref{fig:hr_geom_duration},
+      hatched areas). However, except for its margins in the Pacific Ocean
+      and in the northern foothills of the Alaska Range, the maximum extent
+      of the ice sheet is attained for a~much shorter period of time of only
+      few thousand years (Fig.~\ref{fig:hr_geom_duration}, red areas). This
+      result illustrates that the maximum extents of the modelled ice sheet
+      during MIS~4 and MIS~2 were both short-lived and therefore out of
+      balance with contemporary climate.
 
@@
       where modelled glaciers are confined to its foothills during the
-      entire simulation period (Fig.~\ref{fig:hr_geom_duration}, AR). This apparent
-      insensitivity of modelled glacial extent to temperature fluctuations
-      results from a~combination of low precipitation, high summer
-      temperature and large temperature standard deviation (PDD SD) in the
-      plains of the Alaska Interior (Fig.~\ref{fig:atm}) which confines
-      glaciation to the foothills of the mountains. This result could
-      potentially explain the local distribution of glacial deposits, which
-      indicates that glaciers flowing on the northern slope of the Alaska
-      Range have remained small throughout the Pleistocene
+      entire simulation period (Fig.~\ref{fig:hr_geom_duration}, AR). This
+      apparent insensitivity of modelled glacial extent to temperature
+      fluctuations results from a~combination of low precipitation, high
+      summer temperature and large temperature standard deviation (PDD SD)
+      in the plains of the Alaska Interior (Fig.~\ref{fig:atm}) which
+      confines glaciation to the foothills of the mountains. This result
+      could potentially explain the local distribution of glacial deposits,
+      which indicates that glaciers flowing on the northern slope of the
+      Alaska Range have remained small throughout the Pleistocene
       \citep{Kaufman.Manley.2004}.
@@
       centres, visible on the modelled ice cover duration maps
-      (Fig.~\ref{fig:hr_geom_duration}), are located over the Coast Mountains (CM),
-      the Columbia and Rocky mountains (CRM), the Skeena Mountains (SM), and
-      the Selwyn and Mackenzie mountains (SMKM). The Wrangell and Saint
-      Elias mountains, heavily glacierized at present, host an ice cap for
-      the entire length of both simulations, but that ice cap does not
-      appear to be a~major feed to the Cordilleran ice sheet
-      (Fig.~\ref{fig:hr_geom_duration}, WSEM). Although the Coast, Skeena and
-      Columbia and Rocky mountains (CM, SM, CRM) are covered by mountain
-      glaciers for most of the last glacial cycle, providing durable
-      nucleation centres for an ice sheet, this is not the case for the
-      Selwyn and Mackenzie mountains (SMKM), where ice cover on the highest
-      peaks is limited to a~small fraction of the last glacial cycle. In
-      other words, the Selwyn and Mackenzie mountains only appear as
-      a~secondary ice-dispersal centre during the coldest periods of the
+      (Fig.~\ref{fig:hr_geom_duration}), are located over the Coast
+      Mountains (CM), the Columbia and Rocky mountains (CRM), the Skeena
+      Mountains (SM), and the Selwyn and Mackenzie mountains (SMKM). The
+      Wrangell and Saint Elias mountains, heavily glacierized at present,
+      host an ice cap for the entire length of both simulations, but that
+      ice cap does not appear to be a~major feed to the Cordilleran ice
+      sheet (Fig.~\ref{fig:hr_geom_duration}, WSEM). Although the Coast,
+      Skeena and Columbia and Rocky mountains (CM, SM, CRM) are covered by
+      mountain glaciers for most of the last glacial cycle, providing
+      durable nucleation centres for an ice sheet, this is not the case for
+      the Selwyn and Mackenzie mountains (SMKM), where ice cover on the
+      highest peaks is limited to a~small fraction of the last glacial
+      cycle. In other words, the Selwyn and Mackenzie mountains only appear
+      as a~secondary ice-dispersal centre during the coldest periods of the
       last glacial cycle. The Northern Rocky Mountains
-      (Fig.~\ref{fig:hr_geom_duration}, NRM) do not act as a~nucleation centre, but
-      rather as a~pinning point for the Cordilleran ice sheet margin coming
-      from the west.
+      (Fig.~\ref{fig:hr_geom_duration}, NRM) do not act as a~nucleation
+      centre, but rather as a~pinning point for the Cordilleran ice sheet
+      margin coming from the west.
 
@@
       its predominance over the other ice-dispersal centres
-      (Figs.~\ref{fig:lr_maps} and~\ref{fig:hr_geom_duration}, SM). Regardless of
-      the applied forcing, this ice cap appears to survive MIS~3
-      (Fig.~\ref{fig:lr_maps}, middle panels), and serves as a~nucleation
-      centre at the onset of the glacial readvance towards the LGM
-      (MIS~2). This situation appear similar to the neighbouring Laurentide
-      ice sheet, for which the importance of residual ice for the glacial
-      history leading up to the LGM has been illustrated by the MIS~3
-      residual ice bodies in northern and eastern Canada as nucleation
+      (Figs.~\ref{fig:lr_maps} and~\ref{fig:hr_geom_duration}, SM).
+      Regardless of the applied forcing, this ice cap appears to survive
+      MIS~3 (Fig.~\ref{fig:lr_maps}, middle panels), and serves as
+      a~nucleation centre at the onset of the glacial readvance towards the
+      LGM (MIS~2). This situation appear similar to the neighbouring
+      Laurentide ice sheet, for which the importance of residual ice for the
+      glacial history leading up to the LGM has been illustrated by the
+      MIS~3 residual ice bodies in northern and eastern Canada as nucleation
       centres for a~much more extensive MIS~2 configuration
@@
       The modelled total ice volume corresponding to these persistent
-      ice-dispersal centres attains a~minimum of
-      1.52\,\unit{m}\,s.l.e. (42.9\,\unit{ka}) in the GRIP simulation and of
-      2.68\,\unit{m}\,s.l.e. (51.8\,\unit{ka}) in the EPICA simulation,
-      corresponding to respectively 19 and 31\,\unit{\%} of the MIS~2 ice
-      volumes.
+      ice-dispersal centres attains a~minimum of 1.52\,\unit{m}\,s.l.e.
+      (42.9\,\unit{ka}) in the GRIP simulation and of 2.68\,\unit{m}\,s.l.e.
+      (51.8\,\unit{ka}) in the EPICA simulation, corresponding to
+      respectively 19 and 31\,\unit{\%} of the MIS~2 ice volumes.
 
@@
       the Wrangell and Saint Elias mountains, and radiating off the Skeena
-      Mountains (Figs.~\ref{fig:hr_geom_duration} and \ref{fig:hr_geom_warmbase};
-      \citealp[Fig.~2]{Kleman.etal.2010}). The Skeena Mountains, for
-      example, indeed bear a~strong glacial imprint that indicates ice
-      drainage in a~system of distinct glacial troughs emanating in a~radial
-      pattern from the centre of the mountain range
+      Mountains (Figs.~\ref{fig:hr_geom_duration} and
+      \ref{fig:hr_geom_warmbase}; \citealp[Fig.~2]{Kleman.etal.2010}). The
+      Skeena Mountains, for example, indeed bear a~strong glacial imprint
+      that indicates ice drainage in a~system of distinct glacial troughs
+      emanating in a~radial pattern from the centre of the mountain range
       \citep[Fig.~2]{Kleman.etal.2010}. We suggest that persistent ice cover
-      (Fig.~\ref{fig:hr_geom_duration}) associated with basal ice temperatures at
-      the pressure melting point (Figs.~\ref{fig:hr_geom_warmbase}
-      and~\ref{fig:hr_geom_warmfrac}) explains the large-scale glacial erosional
-      imprint on the landscape. A~well-developed network of glacial valleys
-      running to the north-west on the west slope of the Selwyn and
-      Mackenzie Mountains (\citealp[Fig.~2]{Kleman.etal.2010};
-      \citealp[Fig.~8] {Stroeven.etal.2010}) is modelled to have hosted
-      predominantly warm-based ice (Fig.~\ref{fig:hr_geom_warmfrac}). However,
-      because it is only glaciated for a~short fraction of the last glacial
-      cycle in our simulations (Fig.~\ref{fig:hr_geom_duration}), this perhaps
+      (Fig.~\ref{fig:hr_geom_duration}) associated with basal ice
+      temperatures at the pressure melting point
+      (Figs.~\ref{fig:hr_geom_warmbase} and~\ref{fig:hr_geom_warmfrac})
+      explains the large-scale glacial erosional imprint on the landscape.
+      A~well-developed network of glacial valleys running to the north-west
+      on the west slope of the Selwyn and Mackenzie Mountains
+      (\citealp[Fig.~2]{Kleman.etal.2010}; \citealp[Fig.~8]
+      {Stroeven.etal.2010}) is modelled to have hosted predominantly
+      warm-based ice (Fig.~\ref{fig:hr_geom_warmfrac}). However, because it
+      is only glaciated for a~short fraction of the last glacial cycle in
+      our simulations (Fig.~\ref{fig:hr_geom_duration}), this perhaps
       indicates that the observed landscape pattern originates from multiple
@@
       The modelled distribution of warm-based ice cover
-      (Figs.~\ref{fig:hr_geom_warmbase} and~\ref{fig:hr_geom_warmfrac}) is inevitably
-      affected by our assumption of a~constant, 70\,\unit{mW\,m^{-2}}
-      geothermal heat flux at 3\,\unit{km} depth
+      (Figs.~\ref{fig:hr_geom_warmbase} and~\ref{fig:hr_geom_warmfrac}) is
+      inevitably affected by our assumption of a~constant,
+      70\,\unit{mW\,m^{-2}} geothermal heat flux at 3\,\unit{km} depth
       (Sect.~\ref{sec:icedyn}). However, the Skeena Mountains and the area
@@
       between the two simulations (Figs.~\ref{fig:hr_geom_deglacageshots}
-      and~\ref{fig:hr_geom_deglacage}). The southern sector of the modelling domain,
-      including the Puget Lowland, the Coast Mountains, the Columbia and and
-      Rocky mountains, and the Interior Plateau of British Columbia, becomes
-      completely deglaciated by 10\,\unit{ka}, whereas a~significant ice
-      cover remains over the Skeena, the Selwyn and Mackenzie, and the
+      and~\ref{fig:hr_geom_deglacage}). The southern sector of the modelling
+      domain, including the Puget Lowland, the Coast Mountains, the Columbia
+      and and Rocky mountains, and the Interior Plateau of British Columbia,
+      becomes completely deglaciated by 10\,\unit{ka}, whereas a~significant
+      ice cover remains over the Skeena, the Selwyn and Mackenzie, and the
       Wrangell and Saint Elias mountains in the northern sector of the
@@
       towards the surrounding mountain ranges, consistent with the regional
-      melt water record of the last deglaciation
-      \citep{Margold.etal.2013}. Remaining ice continues to decay by
-      retreating towards the Selwyn and MacKenzie, and Skeena mountains.
-      The last remnants of the Cordilleran ice sheet finally disappear from
-      the Skeena Mountains at 6.5\,\unit{ka} (GRIP) and 6.1\,\unit{ka}
-      (EPICA).
+      melt water record of the last deglaciation \citep{Margold.etal.2013}.
+      Remaining ice continues to decay by retreating towards the Selwyn and
+      MacKenzie, and Skeena mountains. The last remnants of the Cordilleran
+      ice sheet finally disappear from the Skeena Mountains at
+      6.5\,\unit{ka} (GRIP) and 6.1\,\unit{ka} (EPICA).
 
@@
       may have been disconnected from the waning Cordilleran ice sheet, the
-      Finlay River area in the Omenica Mountains (Fig.~\ref{fig:hr_geom_deglacage}, OM)
-      presents a~different kind of evidence. There, sharp-crested moraines
-      indicate a~late-glacial readvance of local alpine glaciers and, more
-      importantly, their interaction with larger, lingering remnants of the
-      main body of the Cordilleran ice sheet in the valleys
-      \citep{Lakeman.etal.2008}. Additional evidence for late-glacial alpine
-      glacier readvances includes moraines in the eastern Coast Mountains,
-      and the Columbia and Rocky mountains \citep{Osborn.Gerloff.1997,
-      Menounos.etal.2008}.
+      Finlay River area in the Omenica Mountains
+      (Fig.~\ref{fig:hr_geom_deglacage}, OM) presents a~different kind of
+      evidence. There, sharp-crested moraines indicate a~late-glacial
+      readvance of local alpine glaciers and, more importantly, their
+      interaction with larger, lingering remnants of the main body of the
+      Cordilleran ice sheet in the valleys \citep{Lakeman.etal.2008}.
+      Additional evidence for late-glacial alpine glacier readvances
+      includes moraines in the eastern Coast Mountains, and the Columbia and
+      Rocky mountains \citep{Osborn.Gerloff.1997, Menounos.etal.2008}.
 
@@
       in the Coast Mountains and in the Columbia and Rocky Mountains
-      (Fig.~\ref{fig:hr_geom_deglacage}, left panel). In addition to matching the
-      location of some local readvances, the GRIP-driven simulation shows
-      that a~large remnant of the decaying ice sheet may still have existed
-      at the time of this late-glacial readvance. In contrast, the
-      EPICA-driven simulation produces a~nearly-continuous deglaciation with
-      only a~tightly restricted late-glacial readvance on the western slopes
-      of the Saint Elias and the Coast mountains (Fig.~\ref{fig:hr_geom_deglacage},
-      right panel).
+      (Fig.~\ref{fig:hr_geom_deglacage}, left panel). In addition to
+      matching the location of some local readvances, the GRIP-driven
+      simulation shows that a~large remnant of the decaying ice sheet may
+      still have existed at the time of this late-glacial readvance. In
+      contrast, the EPICA-driven simulation produces a~nearly-continuous
+      deglaciation with only a~tightly restricted late-glacial readvance on
+      the western slopes of the Saint Elias and the Coast mountains
+      (Fig.~\ref{fig:hr_geom_deglacage}, right panel).
 
@@
       deglaciation or at the time of cessation of sliding inside
-      a~cold-based retreating margin (Fig.~\ref{fig:hr_geom_lastflow}). The modelled
-      deglacial flow patterns are mostly consistent between the GRIP and
-      EPICA simulations. They depict an active ice sheet retreat in the
-      peripheral areas, followed by stagnant ice decay in some of the
+      a~cold-based retreating margin (Fig.~\ref{fig:hr_geom_lastflow}). The
+      modelled deglacial flow patterns are mostly consistent between the
+      GRIP and EPICA simulations. They depict an active ice sheet retreat in
+      the peripheral areas, followed by stagnant ice decay in some of the
       interior regions. Several parts of the modelling domain do not
       experience any basal sliding throughout the deglaciation phase
-      (Fig.~\ref{fig:hr_geom_lastflow}, hatched areas). This notably includes parts
-      of the Interior Plateau in British Columbia, major portions of the
-      Alaskan sector of the ice sheet, and a~tortuous ribbon running from
-      the Northern Rocky Mountains over the Skeena and Selwyn Mountains and
-      into the Mackenzie Mountains.
+      (Fig.~\ref{fig:hr_geom_lastflow}, hatched areas). This notably
+      includes parts of the Interior Plateau in British Columbia, major
+      portions of the Alaskan sector of the ice sheet, and a~tortuous ribbon
+      running from the Northern Rocky Mountains over the Skeena and Selwyn
+      Mountains and into the Mackenzie Mountains.
 
@@
       show similarities with the patterns of deglacial ice flow from
-      numerical modelling (Fig.~\ref{fig:hr_geom_lastflow}). In the northern half of
-      the modelling domain, modelled deglacial flow directions depict an
-      active downhill flow as the last remnants of the ice sheet retreat
-      towards mountain ranges. Converging deglacial flow patterns in the
-      Liard Lowland, for instance (Fig.~\ref{fig:hr_geom_lastflow}), closely
-      resemble the pattern indicated by glacial lineations
+      numerical modelling (Fig.~\ref{fig:hr_geom_lastflow}). In the northern
+      half of the modelling domain, modelled deglacial flow directions
+      depict an active downhill flow as the last remnants of the ice sheet
+      retreat towards mountain ranges. Converging deglacial flow patterns in
+      the Liard Lowland, for instance (Fig.~\ref{fig:hr_geom_lastflow}),
+      closely resemble the pattern indicated by glacial lineations
       \citep[Fig.~2]{Margold.etal.2013}.
@@
       simulations produce a~retreat of the ice margin towards the north-east
-      (Fig.~\ref{fig:hr_geom_deglacage}), a~pattern which is validated by the
-      geomorphological and stratigraphical record for ancient pro-glacial
-      lakes dammed by the retreating ice sheet
+      (Fig.~\ref{fig:hr_geom_deglacage}), a~pattern which is validated by
+      the geomorphological and stratigraphical record for ancient
+      pro-glacial lakes dammed by the retreating ice sheet
       \citep{Perkins.Brennand.2014}. However, the two simulations differ in
@@
       simulation produces negligible basal sliding on the plateau during
-      deglaciation (Fig.~\ref{fig:hr_geom_lastflow}). Yet, the Interior Plateau also
-      hosts an impressive set of glacial lineations which indicate
-      a~substantial eastwards flow component of the Cordilleran ice sheet
-      \citep{Prest.etal.1968, Kleman.etal.2010}. This Interior Plateau
+      deglaciation (Fig.~\ref{fig:hr_geom_lastflow}). Yet, the Interior
+      Plateau also hosts an impressive set of glacial lineations which
+      indicate a~substantial eastwards flow component of the Cordilleran ice
+      sheet \citep{Prest.etal.1968, Kleman.etal.2010}. This Interior Plateau
       lineation set could therefore present a~smoking gun for the
@@
       conditions but, on the contrary, enlarge the region of negligible
-      basal sliding (Fig.~\ref{fig:hr_geom_lastflow}). Thus, a~second explanation
-      for the incongruent results could be that the Interior Plateau
-      lineation system predates deglaciation ice flow, as perhaps indicated
-      by some eskers that appear incompatible with these glacial lineations
-      \citep[Fig.~9]{Margold.etal.2013a}. Finally, a~third explanation could
-      be that local geothermal heat associated with volcanic activity on the
-      Interior Plateau could have triggered the basal sliding
-      \citep[cf. Greenland ice sheet;][]{Fahnestock.etal.2001}.
+      basal sliding (Fig.~\ref{fig:hr_geom_lastflow}). Thus, a~second
+      explanation for the incongruent results could be that the Interior
+      Plateau lineation system predates deglaciation ice flow, as perhaps
+      indicated by some eskers that appear incompatible with these glacial
+      lineations \citep[Fig.~9]{Margold.etal.2013a}. Finally, a~third
+      explanation could be that local geothermal heat associated with
+      volcanic activity on the Interior Plateau could have triggered the
+      basal sliding \citep[cf. Greenland ice
+      sheet;][]{Fahnestock.etal.2001}.
 
       The modelled deglaciation of the Interior Plateau of British Columbia
-      consists of a~rapid northwards retreat (Fig.~\ref{fig:hr_geom_deglacage}) of
-      southwards-flowing non-sliding ice lobes (Fig.~\ref{fig:hr_geom_lastflow})
-      positioned in-between deglaciated mountain ranges
-      (Figs.~\ref{fig:hr_pf_grip} and~\ref{fig:hr_pf_epica}). This
-      result appears compatible with the prevailing conceptual model of
-      deglaciation of central British Columbia, in which mountain ranges
-      emerge from the ice before the plateau
-      \citep[Fig.~7]{Fulton.1991}. However, due to different topographic and
-      climatic conditions, our simulations produce different deglaciation
-      patterns in the northern half of the model domain, indicating that
-      this conceptual model may not be applied to the entire area formerly
-      covered by the Cordilleran ice sheet.
+      consists of a~rapid northwards retreat
+      (Fig.~\ref{fig:hr_geom_deglacage}) of southwards-flowing non-sliding
+      ice lobes (Fig.~\ref{fig:hr_geom_lastflow}) positioned in-between
+      deglaciated mountain ranges (Figs.~\ref{fig:hr_pf_grip}
+      and~\ref{fig:hr_pf_epica}). This result appears compatible with the
+      prevailing conceptual model of deglaciation of central British
+      Columbia, in which mountain ranges emerge from the ice before the
+      plateau \citep[Fig.~7]{Fulton.1991}. However, due to different
+      topographic and climatic conditions, our simulations produce different
+      deglaciation patterns in the northern half of the model domain,
+      indicating that this conceptual model may not be applied to the entire
+      area formerly covered by the Cordilleran ice sheet.
 
@@
       glacial cycle presented in this study consistently produce two glacial
-      maxima during MIS~4 (61.9--56.5\,\unit{ka}, 3.8--8.8\,\unit{m}\,s.l.e.)
-      and MIS~2 (23.2--16.8\,\unit{ka}, 8.40--8.91\,\unit{m}\,s.l.e.), two
-      periods corresponding to documented extensive glaciations. This result
-      is independent of the palaeo-temperature record used among the six
-      selected for this study, and thus can be regarded as a~robust model
-      output, which broadly matches geological evidence. However, the timing
-      of the two glaciation peaks depends sensitively on which climate
-      record is used to drive the model. The timing of the LGM is best
-      reproduced by the EPICA and Vostok Antarctic ice core records. It
-      occurs about 2\,\unit{ka} too early in the simulation forced by the
-      GRIP ice core record, and occurs even earlier in all other
-      simulations. The mismatch is largest for the two Northwest Pacific ODP
-      palaeo-temperature records, which are affected by the weakening of the
-      California current during the LGM.
+      maxima during MIS~4 (61.9--56.5\,\unit{ka},
+      3.8--8.8\,\unit{m}\,s.l.e.) and MIS~2 (23.2--16.8\,\unit{ka},
+      8.40--8.91\,\unit{m}\,s.l.e.), two periods corresponding to documented
+      extensive glaciations. This result is independent of the
+      palaeo-temperature record used among the six selected for this study,
+      and thus can be regarded as a~robust model output, which broadly
+      matches geological evidence. However, the timing of the two glaciation
+      peaks depends sensitively on which climate record is used to drive the
+      model. The timing of the LGM is best reproduced by the EPICA and
+      Vostok Antarctic ice core records. It occurs about 2\,\unit{ka} too
+      early in the simulation forced by the GRIP ice core record, and occurs
+      even earlier in all other simulations. The mismatch is largest for the
+      two Northwest Pacific ODP palaeo-temperature records, which are
+      affected by the weakening of the California current during the LGM.
 
@@
       PhD defence (September 2014). His comments were used to improve the
-  model set-up. We are very thankful to Constantine Khroulev, Ed
-  Bueler, and Andy Aschwanden for providing constant help and
-  development with PISM. This work was supported by the Swedish
-  Research Council~(VR) grant no.~2008-3449 to A.~P.~Stroeven and by
-  the German Academic Exchange Service~(DAAD) grant no.~50015537 and
-  a~Knut and Alice Wallenberg Foundation grant to
-  J.~Seguinot. Computer resources were provided by the Swedish
-  National Infrastructure for Computing (SNIC) allocation
-  no.~2013/1-159 and~2014/1-159 to A.~P.~Stroeven at the National
-  Supercomputing Center (NSC).
+      model set-up. We are very thankful to Constantine Khroulev, Ed Bueler,
+      and Andy Aschwanden for providing constant help and development with
+      PISM. This work was supported by the Swedish Research Council~(VR)
+      grant no.~2008-3449 to A.~P.~Stroeven and by the German Academic
+      Exchange Service~(DAAD) grant no.~50015537 and a~Knut and Alice
+      Wallenberg Foundation grant to J.~Seguinot. Computer resources were
+      provided by the Swedish National Infrastructure for Computing (SNIC)
+      allocation no.~2013/1-159 and~2014/1-159 to A.~P.~Stroeven at the
+      National Supercomputing Center (NSC).
 \end{acknowledgements}
@@
 \begin{figure}%
-\includegraphics[width=80mm]{tc-2015-114-discussions-f01.pdf}
+\includegraphics{locmap}
 \caption{%
@@
 \begin{figure}%
-\includegraphics[height=90mm]{tc-2015-114-discussions-f02.pdf}
+\includegraphics{atm}
 \caption{%
@@
 \begin{figure}%
-\includegraphics[width=120mm]{tc-2015-114-discussions-f03.pdf}
+\includegraphics{lr_ts}
 \caption{%
@@
 \begin{figure}%
-\includegraphics[width=120mm]{tc-2015-114-discussions-f04.pdf}
+\includegraphics{lr_maps}
 \caption{%
@@
 \begin{figure}%
-\includegraphics[width=120mm]{tc-2015-114-discussions-f05.pdf}
+\includegraphics{hr_maps_mis2}
 \caption{%
@@
 \begin{figure}%
-\includegraphics[width=120mm]{tc-2015-114-discussions-f06.pdf}
+\includegraphics{hr_maps_mis4}
 \caption{%
@@
 \begin{figure}%
-\includegraphics[width=120mm]{tc-2015-114-discussions-f07.pdf}
+\includegraphics{hr_geom_duration}
 \caption{%
@@
 \begin{figure}%
-\includegraphics[width=120mm]{tc-2015-114-discussions-f08.pdf}
+\includegraphics{hr_geom_warmbase}
 \caption{%
@@
 \begin{figure}%
-\includegraphics[width=120mm]{tc-2015-114-discussions-f09.pdf}
+\includegraphics{hr_geom_warmfrac}
 \caption{%
@@
 \begin{figure}%
-\includegraphics[width=120mm]{tc-2015-114-discussions-f10.pdf}
+\includegraphics{lr_ts_deglac}
 \caption{%
@@
 \begin{figure}%
-\includegraphics[width=120mm]{tc-2015-114-discussions-f11.pdf}
+\includegraphics{hr_maps_deglac}
 \caption{%
@@
 \begin{figure}%
-\includegraphics[width=120mm]{tc-2015-114-discussions-f12.pdf}
+\includegraphics{hr_geom_deglacage}
 \caption{%
@@
 \begin{figure}%
-\includegraphics[width=120mm]{tc-2015-114-discussions-f13.pdf}
+\includegraphics{hr_geom_lastflow}
 \caption{%
@@
 \begin{figure}%
-\includegraphics[height=100mm]{tc-2015-114-discussions-f14.pdf}
+\includegraphics{hr_pf_grip}
 \caption{%
@@
 \begin{figure}%
-\includegraphics[height=100mm]{tc-2015-114-discussions-f15.pdf}
+\includegraphics{hr_pf_epica}
 \caption{%

### 2016-01-18 cordillera-cycle.tex
Typeset: rewrap figure and table captions.
Typeset: negative numbers and Ph.D. to PhD.

@@
       a~detailed, constructive analysis of this study during J.~Seguinot's
-      Ph.D. defence (September 2014). His comments were used to improve the
+      PhD defence (September 2014). His comments were used to improve the
       model set-up. We are very thankful to Constantine Khroulev, Ed Bueler,
@@
 \begin{table*}
-\caption{Parameter values used in the ice sheet model.}
+\caption{%
+      Parameter values used in the ice sheet model.}
 \label{tab:params}
@@
       temperature offset time-series used to force the ice sheet model
-           through the last glacial cycle (Fig.~\ref{fig:lr_ts}). $f$
-           corresponds to the scaling factor adopted to yield last glacial
-           maximum ice limits in the vicinity of mapped end moraines, and
-           $T_{[32, 22]}$ refers to the resulting mean temperature anomaly
-           during the period -32 to~-22\,\unit{ka} after scaling.}
+      through the last glacial cycle (Fig.~\ref{fig:lr_ts}). $f$ corresponds
+      to the scaling factor adopted to yield last glacial maximum ice limits
+      in the vicinity of mapped end moraines, and $T_{[32, 22]}$ refers to
+      the resulting mean temperature anomaly during the period $-$32
+      to~$-$22\,\unit{ka} after scaling.}
 \label{tab:records}
@@
              & \chem{\delta^{18}O}
-             & 0.37 & -5.8  % -16.4126 (before scaling)
+             & 0.37 & $-$5.8  % -16.4126 (before scaling)
              & \citet{Dansgaard.etal.1993} \\
@@
              & \chem{\delta^{18}O}
-             & 0.24 & -6.5  % -26.7098
+             & 0.24 & $-$6.5  % -26.7098
              & \citet{Andersen.etal.2004} \\
@@
              & \chem{\delta^{18}O}
-             & 0.64 & -5.9  % -9.2055
+             & 0.64 & $-$5.9  % -9.2055
              & \citet{Jouzel.etal.2007} \\
@@
              & \chem{\delta^{18}O}
-             & 0.74 & -5.9  % -7.9550
+             & 0.74 & $-$5.9  % -7.9550
              & \citet{Petit.etal.1999} \\
@@
              & 118{\degree}23$^{\prime}$\,W
-             & -1772
+             & $-$1772
              & \chem{U^{K'}_{37}}
-             & 1.61 & -6.1  % -3.7889
+             & 1.61 & $-$6.1  % -3.7889
              & \citet{Herbert.etal.2001} \\
@@
              & 126{\degree}26$^{\prime}$\,W
-             & -3038
+             & $-$3038
              & \chem{U^{K'}_{37}}
-             & 1.20 & -6.0 % -5.0000
+             & 1.20 & $-$6.0 % -5.0000
              & \citet{Herbert.etal.2001} \\
@@
       Innuitian (IIS), and Greenland (GIS) ice sheets during the last
-           18\,\unit{\chem{^{14}C}\,ka} \citep[21.4\,cal\,ka,][]{Dyke.2004}.
-           The rectangular box denotes the location of the
-           modelling domain used in this study. Major mountain ranges covered
-           by the ice sheet include the Alaska Range (AR), the Wrangell and
-           Saint Elias mountains (WSEM), the Selwyn and Mackenzie mountains
-           (SMKM), the Skeena Mountains (SM), the Coast Mountains (CM), the
-           Columbia and Rocky
-           Mountains (CRM), and the North Cascades (NC). Major depressions
-           include the Liard Lowland (LL), the Interior Plateau of British
-           Columbia (IP), and the Puget Lowland (PL). The background
-           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
-           Data \citep{Patterson.Kelso.2015}.}
+      18\,\unit{\chem{^{14}C}\,ka} \citep[21.4\,cal\,ka,][]{Dyke.2004}. The
+      rectangular box denotes the location of the modelling domain used in
+      this study. Major mountain ranges covered by the ice sheet include the
+      Alaska Range (AR), the Wrangell and Saint Elias mountains (WSEM), the
+      Selwyn and Mackenzie mountains (SMKM), the Skeena Mountains (SM), the
+      Coast Mountains (CM), the Columbia and Rocky Mountains (CRM), and the
+      North Cascades (NC). Major depressions include the Liard Lowland (LL),
+      the Interior Plateau of British Columbia (IP), and the Puget Lowland
+      (PL). The background map consists of ETOPO1 \citep{Amante.Eakins.2009}
+      and Natural Earth Data \citep{Patterson.Kelso.2015}.}
 \label{fig:locmap}%
@@
 \caption{%
-Monthly mean near-surface air temperature, precipitation, and
-           standard deviation of daily mean temperature (PDD SD) for January
-           and July from the North American Regional Reanalysis
+      Monthly mean near-surface air temperature, precipitation, and standard
+      deviation of daily mean temperature (PDD SD) for January and July from
+      the North American Regional Reanalysis
       \citep[NARR;][]{Mesinger.etal.2006}, used to force the surface mass
-           balance (PDD) component of the ice sheet model. Note the
-           strong contrasts in seasonality, timing of the precipitation peak,
-           and temperature variability over the model domain, notably between
-           coastal and inland regions.}
+      balance (PDD) component of the ice sheet model. Note the strong
+      contrasts in seasonality, timing of the precipitation peak, and
+      temperature variability over the model domain, notably between coastal
+      and inland regions.}
 \label{fig:atm}%
@@
       sheet model (top panel), and modelled ice volume (bottom panel)
-           through the last 120\,\unit{ka}. Ice volumes are expressed in meters of sea
-           level equivalent (\unit{m\,s.l.e.}). Gray fields indicate Marine Oxygen
-           Isotope Stage (MIS) boundaries for MIS~2 and MIS~4 according to
+      through the last 120\,\unit{ka}. Ice volumes are expressed in meters
+      of sea level equivalent (\unit{m\,s.l.e.}). Gray fields indicate Marine
+      Oxygen Isotope Stage (MIS) boundaries for MIS~2 and MIS~4 according to
       a~global compilation of benthic \chem{\delta^{18}O} records
@@
       time-volume span for ice volume extremes corresponding to MIS~4
-           (61.9--56.5\,\unit{ka}), MIS~3 (53.0--41.3\,\unit{ka}), and MIS~2 (LGM,
-           23.2--16.8\,\unit{ka}). Dotted lines correspond to GRIP- and EPICA-driven
-           5\,\unit{km}-resolution runs.}
+      (61.9--56.5\,\unit{ka}), MIS~3 (53.0--41.3\,\unit{ka}), and MIS~2
+      (LGM, 23.2--16.8\,\unit{ka}). Dotted lines correspond to GRIP- and
+      EPICA-driven 5\,\unit{km}-resolution runs.}
 \label{fig:lr_ts}%
@@
       corresponding to the ice volume extremes indicated on
-           Fig.~\ref{fig:lr_ts}. An ice cap persists over the Skeena
-           Mountains (SM) during MIS~3. Note the occurence of spatial
-           similarities despite large differences in timing.}
+      Fig.~\ref{fig:lr_ts}. An ice cap persists over the Skeena Mountains
+      (SM) during MIS~3. Note the occurence of spatial similarities despite
+      large differences in timing.}
 \label{fig:lr_maps}%
@@
 \caption{%
-Modelled surface topography (200\,\unit{m} contours) and surface velocity
-           (colour mapping) corresponding to the maximum ice volume during
-           MIS~2 in the GRIP and EPICA high-resolution simulations.}
+      Modelled surface topography (200\,\unit{m} contours) and surface
+      velocity (colour mapping) corresponding to the maximum ice volume
+      during MIS~2 in the GRIP and EPICA high-resolution simulations.}
 \label{fig:mis2}%
@@
 \caption{%
-Modelled surface topography (200\,\unit{m} contours) and surface velocity
-           (colour mapping) corresponding to the maximum ice volume during
-           MIS~4 in the GRIP and EPICA high-resolution simulations.}
+      Modelled surface topography (200\,\unit{m} contours) and surface
+      velocity (colour mapping) corresponding to the maximum ice volume
+      during MIS~4 in the GRIP and EPICA high-resolution simulations.}
 \label{fig:mis4}%
@@
 \caption{%
-Modelled duration of ice cover during the last 120\,\unit{ka} using GRIP
-           and EPICA climate forcing.
-           Note the irregular colour scale. A~continuous ice cover spanning
-           from the Alaska Range (AR) to the Coast Mountains (CM) and
-           the Columbia and Rocky mountains (CRM) exists for about 32\,\unit{ka} in
-           the GRIP simulation and 26\,\unit{ka} in the EPICA simulation.
-           The maximum extent of the ice sheet generally
-           corresponds to relatively short durations of ice cover, but ice
-           cover persists over the Skeena Mountains (SM) during most of the
-           simulation. See Fig.~\ref{fig:locmap} for a~list of abbreviations.}
+      Modelled duration of ice cover during the last 120\,\unit{ka} using
+      GRIP and EPICA climate forcing. Note the irregular colour scale.
+      A~continuous ice cover spanning from the Alaska Range (AR) to the
+      Coast Mountains (CM) and the Columbia and Rocky mountains (CRM) exists
+      for about 32\,\unit{ka} in the GRIP simulation and 26\,\unit{ka} in
+      the EPICA simulation. The maximum extent of the ice sheet generally
+      corresponds to relatively short durations of ice cover, but ice cover
+      persists over the Skeena Mountains (SM) during most of the simulation.
+      See Fig.~\ref{fig:locmap} for a~list of abbreviations.}
 \label{fig:hr_geom_duration}%
@@
       Modelled fraction of warm-based ice cover during the ice-covered
-           period. Note the dominance of warm-based conditions on the
-           continental shelf and major glacial troughs of the coastal ranges.
-           Hatches indicate areas that were covered by cold ice only.}
+      period. Note the dominance of warm-based conditions on the continental
+      shelf and major glacial troughs of the coastal ranges. Hatches
+      indicate areas that were covered by cold ice only.}
 \label{fig:hr_geom_warmfrac}%
@@
       Temperature offset time-series from the GRIP and EPICA ice core
-           records (Table~\ref{tab:records}) (top panel), and modelled ice
-           volume during the deglaciation, expressed in meters of sea-level
-           equivalent (bottom panel).}
+      records (Table~\ref{tab:records}) (top panel), and modelled ice volume
+      during the deglaciation, expressed in meters of sea-level equivalent
+      (bottom panel).}
 \label{fig:lr_ts_deglac}%
@@
 \caption{%
-Snapshots of modelled surface topography (200\,\unit{m} contours)
-           and surface velocity (colour mapping) during the last deglaciation
-           from the GRIP (top panels) and EPICA (bottom panels) 5\,\unit{km} simulations.
-           Dashed segments \textbf{(A--D)} indicate the location of profiles used in
-           Figs.~\ref{fig:hr_pf_grip} and~\ref{fig:hr_pf_epica}.}
+      Snapshots of modelled surface topography (200\,\unit{m} contours) and
+      surface velocity (colour mapping) during the last deglaciation from
+      the GRIP (top panels) and EPICA (bottom panels) 5\,\unit{km}
+      simulations. Dashed segments \textbf{(A--D)} indicate the location of
+      profiles used in Figs.~\ref{fig:hr_pf_grip}
+      and~\ref{fig:hr_pf_epica}.}
 \label{fig:hr_geom_deglacageshots}%
@@
       Modelled age of the last deglaciation. Areas that have been covered
-           only before the last glacial maximum are marked
-           in green. Hatches denote re-advance of mountain-centred ice caps and
-           the decaying ice sheet between 14 and 10\,\unit{ka}.
-           Dashed segments \textbf{(A--D)} indicate the location of profiles used in
-           Figs.~\ref{fig:hr_pf_grip} and~\ref{fig:hr_pf_epica}.}
+      only before the last glacial maximum are marked in green. Hatches
+      denote re-advance of mountain-centred ice caps and the decaying ice
+      sheet between 14 and 10\,\unit{ka}. Dashed segments \textbf{(A--D)}
+      indicate the location of profiles used in Figs.~\ref{fig:hr_pf_grip}
+      and~\ref{fig:hr_pf_epica}.}
 \label{fig:hr_geom_deglacage}%
@@
 \caption{%
-Modelled deglacial basal ice velocities. Hatches
-           indicate areas that remain non-sliding throughout deglaciation
-           (22.0--8.0\,\unit{ka}), notably including parts of the Interior Plateau (IP).
-           Note the concentric patterns of deglacial flow in the Liard
-           Lowland (LL).
+      Modelled deglacial basal ice velocities. Hatches indicate areas that
+      remain non-sliding throughout deglaciation (22.0--8.0\,\unit{ka}),
+      notably including parts of the Interior Plateau (IP). Note the
+      concentric patterns of deglacial flow in the Liard Lowland (LL).
       Sliding grid cells were distinguished from non-sliding grid cells
@@
       simulation, corresponding to the four transects indicated in
-           Figs.~\ref{fig:hr_geom_deglacageshots} and~\ref{fig:hr_geom_deglacage}.}
+      Figs.~\ref{fig:hr_geom_deglacageshots}
+      and~\ref{fig:hr_geom_deglacage}.}
 \label{fig:hr_pf_grip}%
@@
       simulation, corresponding to the four transects indicated in
-           Figs.~\ref{fig:hr_geom_deglacageshots} and~\ref{fig:hr_geom_deglacage}.}
+      Figs.~\ref{fig:hr_geom_deglacageshots}
+      and~\ref{fig:hr_geom_deglacage}.}
 \label{fig:hr_pf_epica}%

### 2016-01-18 cordillera-cycle.tex
Typeset: pull typesetted references.

@@
 
-\bibliographystyle{copernicus}
-\bibliography{refs/references}
-\newpage
+\begin{thebibliography}{114}
+
+\bibitem[{Albrecht et~al.(2011)Albrecht, Martin, Haseloff, Winkelmann, and
+  Levermann}]{Albrecht.etal.2011}
+ Albrecht,~T., Martin,~M., Haseloff,~M., Winkelmann,~R., and Levermann,~A.: Parameterization for subgrid-scale motion of ice-shelf calving fronts, The Cryosphere, 5, 35--44,
+doi:\href{http://dx.doi.org/10.5194/tc-5-35-2011}{10.5194/tc-5-35-2011}, 2011.
+
+\bibitem[{Amante and Eakins(2009)}]{Amante.Eakins.2009}
+Amante,~C. and Eakins,~B.~W.: ETOPO1 1 arc-minute global relief model: procedures, data sources and analysis, NOAA technical memorandum NESDIS NGDC-24, Natl. Geophys. Data Center, NOAA, Boulder, CO,
+doi:\href{http://dx.doi.org/10.7289/V5C8276M}{10.7289/V5C8276M}, 2009.
+
+\bibitem[{Andersen et~al.(2004)Andersen, Azuma, Barnola, Bigler, Biscaye, Caillon, Chappellaz, Clausen, Dahl-Jensen, Fischer, Fl\"uckiger, Fritzsche, Fujii, Goto-Azuma, Gr{\o}nvold, Gundestrup, Hansson, Huber, Hvidberg, Johnsen, Jonsell, Jouzel, Kipfstuhl, Landais, Leuenberger, Lorrain, Masson-Delmotte, Miller, Motoyama, Narita, Popp, Rasmussen, Raynaud, Rothlisberger, Ruth, Samyn, Schwander, Shoji, Siggard-Andersen, Steffensen, Stocker, Sveinbj\"ornsd\'ottir, Svensson, Takata, Tison, Thorsteinsson, Watanabe, Wilhelms, and White}]{Andersen.etal.2004}
+Andersen,~K.~K., Azuma,~N., Barnola,~J.-M., Bigler,~M., Biscaye,~P., Caillon,~N., Chappellaz,~J., Clausen,~H.~B., Dahl-Jensen,~D., Fischer,~H., Fl\"uckiger,~J., Fritzsche,~D., Fujii,~Y., Goto-Azuma,~K., Gr{\o}nvold,~K., Gundestrup,~N.~S., Hansson,~M., Huber,~C., Hvidberg,~C.~S., Johnsen,~S.~J., Jonsell,~U., Jouzel,~J., Kipfstuhl,~S., Landais,~A., Leuenberger,~M., Lorrain,~R., Masson-Delmotte,~V., Miller,~H., Motoyama,~H., Narita,~H., Popp,~T., Rasmussen,~S.~O., Raynaud,~D., Rothlisberger,~R., Ruth,~U., Samyn,~D., Schwander,~J., Shoji,~H., Siggard-Andersen,~M.-L., Steffensen,~J.~P., Stocker,~T., Sveinbj\"ornsd\'ottir,~A.~E., Svensson,~A., Takata,~M., Tison,~J.-L., Thorsteinsson,~T., Watanabe,~O., Wilhelms,~F., and White,~J.~W.~C.: High-resolution record of Northern Hemisphere climate extending into the last interglacial period, Nature, 431, 147--151,
+doi:\href{http://dx.doi.org/10.1038/nature02805}{10.1038/nature02805}, data archived at the World Data Center for Paleoclimatology, Boulder, Colorado, USA, 2004.
+
+\bibitem[{Aschwanden et~al.(2012)Aschwanden, Bueler, Khroulev, and Blatter}]{Aschwanden.etal.2012}
+Aschwanden,~A., Bueler,~E., Khroulev,~C., and Blatter,~H.: An enthalpy formulation for glaciers and ice sheets,~J. Glaciol., 58, 441--457,
+doi:\href{http://dx.doi.org/10.3189/2012JoG11J088}{10.3189/2012JoG11J088}, 2012.
+
+\bibitem[{Aschwanden et~al.(2013)Aschwanden, A{\dh}algeirsd\'{o}ttir, and Khroulev}]{Aschwanden.etal.2013}
+ Aschwanden,~A., A{\dh}algeirsd\'{o}ttir,~G., and Khroulev,~C.: Hindcasting to measure ice sheet model sensitivity to   initial states, The Cryosphere, 7, 1083--1093,
+doi:\href{http://dx.doi.org/10.5194/tc-7-1083-2013}{10.5194/tc-7-1083-2013}, 2013.
+
+\bibitem[{Balay et~al.(2015)Balay, Abhyankar, Adams, Brown, Brune, Buschelman, Eijkhout, Gropp, Kaushik, Knepley, McInnes, Rupp, Smith, and Zhang}]{Balay.etal.2015}
+Balay,~S., Abhyankar,~S., Adams,~M.~F., Brown,~J., Brune,~P., Buschelman,~K.,
+Eijkhout,~V., Gropp,~W.~D., Kaushik,~D., Knepley,~M.~G., McInnes,~L.~C.,
+Rupp,~K., Smith,~B.~F., and Zhang,~H.: {PETS}c {W}eb page, available at:
+\url{http://www.mcs.anl.gov/petsc} (last access: 2015), 2015.
+
+\bibitem[{Barendregt and Irving(1998)}]{Barendregt.Irving.1998}
+Barendregt,~R.~W. and Irving,~E.: Changes in the extent of North American ice sheets during the late Cenozoic, Can.~J. Earth Sci., 35, 504--509,
+doi:\href{http://dx.doi.org/10.1139/e97-126}{10.1139/e97-126}, 1998.
+
+\bibitem[{Bednarski and Smith(2007)}]{Bednarski.Smith.2007}
+Bednarski,~J.~M. and Smith,~I.~R.: Laurentide and montane glaciation along the Rocky Mountain Foothills of northeastern British Columbia, Can.~J. Earth Sci., 44, 445--457,
+doi:\href{http://dx.doi.org/10.1139/e06-095}{10.1139/e06-095}, 2007.
+
+\bibitem[{Blackwell and Richards(2004)}]{Blackwell.Richards.2004}
+Blackwell,~D.~D. and Richards,~M.: Geothermal Map of North America, Am. Assoc. Petr. Geol., Tulsa, OK, 2004.
+
+\bibitem[{Booth et~al.(2003)Booth, Troost, Clague, and Waitt}]{Booth.etal.2003}
+Booth,~D.~B., Troost,~K.~G., Clague,~J.~J., and Waitt,~R.~B.: The Cordilleran ice sheet, in: The Quaternary Period in the United States, edited by: Gillespie,~A., Porter,~S., and Atwater,~B., vol.~1 of Dev.
+Quaternary Sci., Elsevier, Amsterdam, 17--43,
+doi:\href{http://dx.doi.org/10.1016/s1571-0866(03)01002-9}{10.1016/s1571-0866(03)01002-9}, 2003.
+
+\bibitem[{Boulton and Clark(1990)}]{Boulton.Clark.1990}
+Boulton,~G.~S. and Clark,~C.~D.: A highly mobile Laurentide ice sheet revealed by satellite images of glacial lineations, Nature, 346, 813--817,
+doi:\href{http://dx.doi.org/10.1038/346813a0}{10.1038/346813a0}, 1990.
+
+\bibitem[{Boulton et~al.(2001)Boulton, Dongelmans, Punkari, and Broadgate}]{Boulton.etal.2001}
+Boulton,~G.~S., Dongelmans,~P., Punkari,~M., and Broadgate,~M.: Palaeoglaciology of an ice sheet through a glacial cycle: the European ice sheet through the Weichselian, Quaternary Res., 20, 591--625,
+doi:\href{http://dx.doi.org/10.1016/s0277-3791(00)00160-8}{10.1016/s0277-3791(00)00160-8}, 2001.
+
+\bibitem[{Briner and Kaufman(2008)}]{Briner.Kaufman.2008}
+Briner,~J.~P. and Kaufman,~D.~S.: Late Pleistocene mountain glaciation in Alaska: key chronologies,~J. Quaternary Sci., 23, 659--670,
+doi:\href{http://dx.doi.org/10.1002/jqs.1196}{10.1002/jqs.1196}, 2008.
+
+\bibitem[{Bueler and Brown(2009)}]{Bueler.Brown.2009}
+Bueler,~E. and Brown,~J.: Shallow shelf approximation as a ``sliding law'' in a~thermodynamically coupled ice sheet model,~J. Geophys. Res., 114, F03008,
+doi:\href{http://dx.doi.org/10.1029/2008JF001179}{10.1029/2008JF001179}, 2009.
+
+\bibitem[{Bueler and van Pelt(2015)}]{Bueler.Pelt.2015}
+ Bueler,~E. and van~Pelt,~W.: Mass-conserving subglacial hydrology in the Parallel Ice Sheet Model version 0.6, Geosci. Model Dev., 8, 1613--1635,
+doi:\href{http://dx.doi.org/10.5194/gmd-8-1613-2015}{10.5194/gmd-8-1613-2015}, 2015.
+
+\bibitem[{Bueler et~al.(2007)Bueler, Lingle, and Brown}]{Bueler.etal.2007}
+Bueler,~E., Lingle,~C.~S., and Brown,~J.: Fast computation of a viscoelastic deformable Earth model for ice-sheet simulations, Ann. Glaciol., 46, 97--105,
+doi:\href{http://dx.doi.org/10.3189/172756407782871567}{10.3189/172756407782871567}, 2007.
+
+\bibitem[{Burke et~al.(2012{\natexlab{a}})Burke, Brennand, and Perkins}]{Burke.etal.2012}
+Burke,~M.~J., Brennand,~T.~A., and Perkins,~A.~J.: Transient subglacial hydrology of a thin ice sheet: insights from the Chasm esker, British Columbia, Canada, Quaternary Res., 58, 30--55,
+doi:\href{http://dx.doi.org/10.1016/j.quascirev.2012.09.004}{10.1016/j.quascirev.2012.09.004}, 2012{\natexlab{a}}.
+
+\bibitem[{Burke et~al.(2012{\natexlab{b}})Burke, Brennand, and Perkins}]{Burke.etal.2012a}
+Burke,~M.~J., Brennand,~T.~A., and Perkins,~A.~J.: Evolution of the subglacial hydrologic system beneath the rapidly decaying Cordilleran Ice Sheet caused by ice-dammed lake drainage: implications for meltwater-induced ice acceleration, Quaternary Res., 50, 125--140,
+doi:\href{http://dx.doi.org/10.1016/j.quascirev.2012.07.005}{10.1016/j.quascirev.2012.07.005}, 2012{\natexlab{b}}.
+
+\bibitem[{Calov and Greve(2005)}]{Calov.Greve.2005}
+Calov,~R. and Greve,~R.: A~semi-analytical solution for the positive degree-day model with stochastic temperature variations,~J. Glaciol., 51, 173--175,
+doi:\href{http://dx.doi.org/10.3189/172756505781829601}{10.3189/172756505781829601}, 2005.
+
+\bibitem[{Carlson and Clark(2012)}]{Carlson.Clark.2012}
+Carlson,~A.~E. and Clark,~P.~U.: Ice sheet sources of sea level rise and
+freshwater discharge during the last deglaciation, Rev. Geophys., 50, RG4007,
+doi:\href{http://dx.doi.org/10.1029/2011rg000371}{10.1029/2011rg000371},
+2012.
+
+\bibitem[{Carrara et~al.(1996)Carrara, Kiver, and Stradling}]{Carrara.etal.1996}
+Carrara,~P.~E., Kiver,~E.~P., and Stradling,~D.~F.: The southern limit of Cordilleran ice in the Colville and Pend Oreille valleys of northeastern Washington during the Late Wisconsin glaciation, Can.~J.
+Earth Sci., 33, 769--778,
+doi:\href{http://dx.doi.org/10.1139/e96-059}{10.1139/e96-059}, 1996.
+
+\bibitem[{Clague et~al.(1980)Clague, Armstrong, and Mathews}]{Clague.etal.1980}
+Clague,~J., Armstrong,~J., and Mathews,~W.: Advance of the late Wisconsin Cordilleran Ice Sheet in southern British Columbia since 22\,000~yr~BP, Quaternary Res., 13, 322--326,
+doi:\href{http://dx.doi.org/10.1016/0033-5894(80)90060-5}{10.1016/0033-5894(80)90060-5}, 1980.
+
+\bibitem[{Clague(1981)}]{Clague.1981}
+Clague,~J.~J.: Late Quaternary Geology and Geochronology of British Columbia Part 2: Summary and Discussion of Radiocarbon-Dated Quaternary History, Geol. Surv. of Can., Ottawa, ON, Paper 80-35,
+doi:\href{http://dx.doi.org/10.4095/119439}{10.4095/119439}, 1981.
+
+\bibitem[{Clague(1985)}]{Clague.1985}
+Clague,~J.~J.: Delaciation of the Prince Rupert -- Kitimat area, British Columbia, Can.~J. Earth Sci., 22, 256--265,
+doi:\href{http://dx.doi.org/10.1139/e85-022}{10.1139/e85-022}, 1985.
+
+\bibitem[{Clague(1986)}]{Clague.1986}
+Clague,~J.~J.: The Quaternary stratigraphic record of British Columbia --- evidence for episodic sedimentation and erosion controlled by glaciation, Can.~J. Earth Sci., 23, 885--894,
+doi:\href{http://dx.doi.org/10.1139/e86-090}{10.1139/e86-090}, 1986.
+
+\bibitem[{Clague(1989)}]{Clague.1989}
+Clague,~J.~J.: Character and distribution of Quaternary deposits (Canadian Cordillera), in: Quaternary Geology of Canada and Greenland, edited by: Fulton,~R.~J., vol.~1 of Geology of Canada, Geol.
+Surv.
+of Can., Ottawa, ON, 34--48,
+doi:\href{http://dx.doi.org/10.4095/127905}{10.4095/127905}, 1989.
+
+\bibitem[{Clague and James(2002)}]{Clague.James.2002}
+Clague,~J.~J. and James,~T.~S.: History and isostatic effects of the last ice sheet in southern British Columbia, Quaternary Res., 21, 71--87,
+doi:\href{http://dx.doi.org/10.1016/s0277-3791(01)00070-1}{10.1016/s0277-3791(01)00070-1}, 2002.
+
+\bibitem[{Clague and Ward(2011)}]{Clague.Ward.2011}
+Clague,~J.~J. and Ward,~B.: Pleistocene Glaciation of British
+Columbia, in:   Dev. Quaternary Sci., vol.~15, edited by: Ehlers,~J., Gibbard,~P.~L., and Hughes,~P.~D., Elsevier, Amsterdam, 563--573,
+doi:\href{http://dx.doi.org/10.1016/b978-0-444-53447-7.00044-1}{10.1016/b978-0-444-53447-7.00044-1}, 2011.
+
+\bibitem[{Clague et~al.(1997)Clague, Mathewes, Guilbault, Hutchinson, and Ricketts}]{Clague.etal.1997}
+Clague,~J.~J., Mathewes,~R.~W., Guilbault,~J.-P., Hutchinson,~I., and Ricketts,~B.~D.: Pre-Younger Dryas resurgence of the southwestern margin of the Cordilleran ice sheet, British Columbia, Canada, Boreas, 26, 261--278,
+doi:\href{http://dx.doi.org/10.1111/j.1502-3885.1997.tb00855.x}{10.1111/j.1502-3885.1997.tb00855.x}, 1997.
+
+\bibitem[{Clague et~al.(2005)Clague, Froese, Hutchinson, James, and Simon}]{Clague.etal.2005}
+Clague,~J.~J., Froese,~D., Hutchinson,~I., James,~T.~S., and Simon,~K.~M.: Early growth of the last Cordilleran ice sheet deduced from glacio-isostatic depression in southwest British Columbia, Canada, Quaternary Res., 63, 53--59,
+doi:\href{http://dx.doi.org/10.1016/j.yqres.2004.09.007}{10.1016/j.yqres.2004.09.007}, 2005.
+
+\bibitem[{Clark and Mix(2002)}]{Clark.Mix.2002}
+Clark,~P.~U. and Mix,~A.~C.: Ice sheets and sea level of the Last Glacial Maximum, Quaternary Res., 21, 1--7,
+doi:\href{http://dx.doi.org/10.1016/s0277-3791(01)00118-4}{10.1016/s0277-3791(01)00118-4}, 2002.
+
+\bibitem[{Clason et~al.(2012)Clason, Mair, Burgess, and Nienow}]{Clason.etal.2012}
+Clason,~C., Mair,~D.~W., Burgess,~D.~O., and Nienow,~P.~W.: Modelling the delivery of supraglacial meltwater to the ice/bed interface: application to southwest Devon Ice Cap, Nunavut, Canada,~J.
+Glaciol., 58, 361--374,
+doi:\href{http://dx.doi.org/10.3189/2012jog11j129}{10.3189/2012jog11j129}, 2012.
+
+\bibitem[{Clason et~al.(2014)Clason, Applegate, and Holmlund}]{Clason.etal.2014}
+Clason,~C., Applegate,~P., and Holmlund,~P.: Modelling Late Weichselian evolution of the Eurasian ice sheets forced by surface meltwater-enhanced basal sliding,~J. Glaciol., 60, 29--40,
+doi:\href{http://dx.doi.org/10.3189/2014jog13j037}{10.3189/2014jog13j037}, 2014.
+
+\bibitem[{Cosma et~al.(2008)Cosma, Hendy, and Chang}]{Cosma.etal.2008}
+Cosma,~T., Hendy,~I., and Chang,~A.: Chronological constraints on Cordilleran Ice Sheet glaciomarine sedimentation from core MD02-2496 off Vancouver Island (western Canada), Quaternary Sci.
+Rev., 27, 941--955,
+doi:\href{http://dx.doi.org/10.1016/j.quascirev.2008.01.013}{10.1016/j.quascirev.2008.01.013}, 2008.
+
+\bibitem[{Dansgaard et~al.(1993)Dansgaard, Johnsen, Clausen, Dahl-Jensen, Gundestrup, Hammer, Hvidberg, Steffensen, Sveinbj\"ornsdottir, Jouzel, and Bond}]{Dansgaard.etal.1993}
+Dansgaard,~W., Johnsen,~S.~J., Clausen,~H.~B., Dahl-Jensen,~D., Gundestrup,~N.~S., Hammer,~C.~U., Hvidberg,~C.~S., Steffensen,~J.~P., Sveinbj\"ornsdottir,~A.~E., Jouzel,~J., and Bond,~G.: Evidence for general instability of past climate from a 250-kyr ice-core record, Nature, 364, 218--220,
+doi:\href{http://dx.doi.org/10.1038/364218a0}{10.1038/364218a0}, data archived at the World Data Center for Paleoclimatology, Boulder, Colorado, USA., 1993.
+
+\bibitem[{Davies et~al.(2011)Davies, Mix, Stoner, Addison, Jaeger, Finney, and Wiest}]{Davies.etal.2011}
+Davies,~M.~H., Mix,~A.~C., Stoner,~J.~S., Addison,~J.~A., Jaeger,~J., Finney,~B., and Wiest,~J.: The deglacial transition on the southeastern Alaska Margin: Meltwater input, sea level rise, marine productivity, and sedimentary anoxia, Paleoceanography, 26, PA2223,
+doi:\href{http://dx.doi.org/10.1029/2010pa002051}{10.1029/2010pa002051}, 2011.
+
+\bibitem[{Davis and Mathews(1944)}]{Davis.Mathews.1944}
+Davis,~N.~F.~G. and Mathews,~W.~H.: Four phases of glaciation with illustrations from Southwestern British Columbia,~J. Geol., 52, 403--413,
+doi:\href{http://dx.doi.org/10.1086/625236}{10.1086/625236}, 1944.
+
+\bibitem[{Dawson(1888)}]{Dawson.1888}
+Dawson,~G.~M.: III. -- Recent observations on the glaciation of British Columbia and adjacent regions, Geol. Mag., 5, 347--350,
+doi:\href{http://dx.doi.org/10.1017/s0016756800182159}{10.1017/s0016756800182159}, 1888.
+
+\bibitem[{Demuro et~al.(2012)Demuro, Froese, Arnold, and Roberts}]{Demuro.etal.2012}
+Demuro,~M., Froese,~D.~G., Arnold,~L.~J., and Roberts,~R.~G.: Single-grain OSL dating of glaciofluvial quartz constrains Reid glaciation in NW Canada to MIS 6, Quaternary Res., 77, 305--316,
+doi:\href{http://dx.doi.org/10.1016/j.yqres.2011.11.009}{10.1016/j.yqres.2011.11.009}, 2012.
+
+\bibitem[{Duk-Rodkin(1999)}]{Duk-Rodkin.1999}
+Duk-Rodkin,~A.: Glacial limits map of Yukon Territory, Open File 3694, Geol. Surv. of Can., Ottawa, ON,
+doi:\href{http://dx.doi.org/10.4095/210739}{10.4095/210739}, 1999.
+
+\bibitem[{Duk-Rodkin et~al.(1996)Duk-Rodkin, Barendregt, Tarnocai, and Phillips}]{Duk-Rodkin.etal.1996}
+Duk-Rodkin,~A., Barendregt,~R.~W., Tarnocai,~C., and Phillips,~F.~M.: Late Tertiary to late Quaternary record in the Mackenzie Mountains, Northwest Territories, Canada: stratigraphy, paleosols, paleomagnetism, and chlorine-36, Can.~J.
+Earth Sci., 33, 875--895,
+doi:\href{http://dx.doi.org/10.1139/e96-066}{10.1139/e96-066}, 1996.
+
+\bibitem[{Dyke(2004)}]{Dyke.2004}
+Dyke,~A.~S.: An outline of North American deglaciation with emphasis
+on central and northern Canada, in: Dev. Quaternary Sci., vol.~2,
+edited by: Ehlers,~J. and Gibbard,~P.~L., Elsevier, Amsterdam, 373--424,
+doi:\href{http://dx.doi.org/10.1016/S1571-0866(04)80209-4}{10.1016/S1571-0866(04)80209-4}, 2004.
+
+\bibitem[{Dyke and Prest(1987)}]{Dyke.Prest.1987}
+Dyke,~A.~S. and Prest,~V.~K.: Late wisconsinan and holocene history of the laurentide ice sheet, G{e}ogr. Phys. Quatern., 41, 237--263,
+doi:\href{http://dx.doi.org/10.7202/032681ar}{10.7202/032681ar}, 1987.
+
+\bibitem[{Dyke et~al.(2003)Dyke, Moore, and Robertson}]{Dyke.etal.2003}
+Dyke,~A.~S., Moore,~A., and Robertson,~L.: Deglaciation of North America, Open File 1547, Geol. Surv. of Can., Ottawa, ON, 2003.
+
+\bibitem[{Ehlers and Gibbard(2004)}]{Ehlers.Gibbard.2004}
+Ehlers,~J. and Gibbard,~P.~L. (Eds.): Dev. Quaternary Sci., vol.~2, Elsevier, Amsterdam, 2004.
+
+\bibitem[{Ehlers et~al.(2011)Ehlers, Gibbard, and Hughes}]{Ehlers.etal.2011}
+Ehlers,~J., Gibbard,~P.~L., and Hughes,~P.~D. (Eds.):  Dev. Quaternary
+Sci., vol.~15, Elsevier, Amsterdam, 2011.
+
+\bibitem[{Fahnestock et~al.(2001)Fahnestock, Abdalati, Joughin, Brozena, and Gogineni}]{Fahnestock.etal.2001}
+Fahnestock,~M., Abdalati,~W., Joughin,~I., Brozena,~J., and Gogineni,~P.: High geothermal heat flow, basal melt, and the origin of rapid ice flow in central greenland, Sience, 294, 2338--2342,
+doi:\href{http://dx.doi.org/10.1126/science.1065370}{10.1126/science.1065370}, 2001.
+
+\bibitem[{Friele and Clague(2002{\natexlab{a}})}]{Friele.Clague.2002}
+Friele,~P.~A. and Clague,~J.~J.: Readvance of glaciers in the British Columbia Coast Mountains at the end of the last glaciation, Quatern. Int., 87, 45--58,
+doi:\href{http://dx.doi.org/10.1016/s1040-6182(01)00061-1}{10.1016/s1040-6182(01)00061-1}, 2002{\natexlab{a}}.
+
+\bibitem[{Friele and Clague(2002{\natexlab{b}})}]{Friele.Clague.2002a}
+Friele,~P.~A. and Clague,~J.~J.: Younger Dryas readvance in Squamish river valley, southern Coast mountains, British Columbia, Quaternary Res., 21, 1925--1933,
+doi:\href{http://dx.doi.org/10.1016/s0277-3791(02)00081-1}{10.1016/s0277-3791(02)00081-1}, 2002{\natexlab{b}}.
+
+\bibitem[{Fulton(1967)}]{Fulton.1967}
+Fulton,~R.~J.: Deglaciation studies in Kamloops region, an area of moderate relief, British Columbia, vol. 154 of Bull., Geol. Surv. of Can., Ottawa, ON,
+doi:\href{http://dx.doi.org/10.4095/101467}{10.4095/101467}, 1967.
+
+\bibitem[{Fulton(1991)}]{Fulton.1991}
+Fulton,~R.~J.: A Conceptual model for growth and decay of the Cordilleran Ice Sheet, G{e}ogr. Phys. Quatern., 45, 281--286,
+doi:\href{http://dx.doi.org/10.7202/032875ar}{10.7202/032875ar}, 1991.
+
+\bibitem[{Golledge et~al.(2012)Golledge, Mackintosh, Anderson, Buckley, Doughty, Barrell, Denton, Vandergoes, Andersen, and Schaefer}]{Golledge.etal.2012}
+Golledge,~N.~R., Mackintosh,~A.~N., Anderson,~B.~M., Buckley,~K.~M., Doughty,~A.~M., Barrell,~D.~J., Denton,~G.~H., Vandergoes,~M.~J., Andersen,~B.~G., and Schaefer,~J.~M.: Last Glacial Maximum climate in {N}ew {Z}ealand inferred from a modelled {S}outhern {A}lps icefield, Quaternary Res., 46, 30--45,
+doi:\href{http://dx.doi.org/10.1016/j.quascirev.2012.05.004}{10.1016/j.quascirev.2012.05.004}, 2012.
+
+\bibitem[{Herbert et~al.(2001)Herbert, Schuffert, Andreasen, Heusser, Lyle, Mix, Ravelo, Stott, and Herguera}]{Herbert.etal.2001}
+Herbert,~T.~D., Schuffert,~J.~D., Andreasen,~D., Heusser,~L., Lyle,~M., Mix,~A., Ravelo,~A.~C., Stott,~L.~D., and Herguera,~J.~C.: Collapse of the California current during glacial maxima linked to climate change on land, Sience, 293, 71--76,
+doi:\href{http://dx.doi.org/10.1126/science.1059209}{10.1126/science.1059209}, data archived at the World Data Center for Paleoclimatology, Boulder, Colorado, USA, 2001.
+
+\bibitem[{Hidy et~al.(2013)Hidy, Gosse, Froese, Bond, and Rood}]{Hidy.etal.2013}
+Hidy,~A.~J., Gosse,~J.~C., Froese,~D.~G., Bond,~J.~D., and Rood,~D.~H.: A latest Pliocene age for the earliest and most extensive Cordilleran Ice Sheet in northwestern Canada, Quaternary Res., 61, 77--84,
+doi:\href{http://dx.doi.org/10.1016/j.quascirev.2012.11.009}{10.1016/j.quascirev.2012.11.009}, 2013.
+
+\bibitem[{Hock(2003)}]{Hock.2003}
+Hock,~R.: Temperature index melt modelling in mountain areas,~J. Hydrol., 282, 104--115,
+doi:\href{http://dx.doi.org/10.1016/S0022-1694(03)00257-9}{10.1016/S0022-1694(03)00257-9}, 2003.
+
+\bibitem[{Imbrie et~al.(1989)Imbrie, McIntyre, and Mix}]{Imbrie.etal.1989}
+Imbrie,~J., McIntyre,~A., and Mix,~A.: Oceanic response to orbital forcing in the late quaternary: observational and experimental strategies, in: Climate and Geo-Sciences, edited by: Berger,~A., Schneider,~S., and Duplessy,~J., vol.
+285 of NATO ASI Series C, Kluwer, Norwell, MA, 121--164,
+doi:\href{http://dx.doi.org/10.1007/978-94-009-2446-8_7}{10.1007/978-94-009-2446-8\_7}, 1989.
+
+\bibitem[{Ivy-Ochs et~al.(1999)Ivy-Ochs, Schluchter, Kubik, and Denton}]{Ivy-Ochs.etal.1999}
+Ivy-Ochs,~S., Schluchter,~C., Kubik,~P.~W., and Denton,~G.~H.: Moraine
+exposure dates imply synchronous Younger Dryas Glacier advances in the
+European Alps and in the Southern Alps of New Zealand, Geogr. Ann. A, 81,
+313--323,
+doi:\href{http://dx.doi.org/10.1111/1468-0459.00060}{10.1111/1468-0459.00060},
+1999.
+
+\bibitem[{Jackson and Clague(1991)}]{Jackson.Clague.1991}
+Jackson,~L.~E. and Clague,~J.~J.: The Cordilleran ice sheet: one hundred and fifty years of exploration and discovery, G{e}ogr. Phys. Quatern., 45, 269--280,
+doi:\href{http://dx.doi.org/10.7202/032874ar}{10.7202/032874ar}, 1991.
+
+\bibitem[{Jackson et~al.(1997)Jackson, Phillips, Shimamura, and Little}]{Jackson.etal.1997}
+Jackson,~L.~E., Phillips,~F.~M., Shimamura,~K., and Little,~E.~C.: Cosmogenic $^{36}$Cl dating of the Foothills erratics train, Alberta, Canada, Geology, 25, 195,
+doi:\href{http://dx.doi.org/10.1130/0091-7613(1997)025<0195:ccdotf>2.3.co;2}{10.1130/0091-7613(1997)025\textless0195:ccdotf\textgreater2.3.co;2}, 1997.
+
+\bibitem[{James et~al.(2009)James, Gowan, Wada, and Wang}]{James.etal.2009}
+James,~T.~S., Gowan,~E.~J., Wada,~I., and Wang,~K.: Viscosity of the
+asthenosphere from glacial isostatic adjustment and subduction dynamics at
+the northern Cascadia subduction zone, British Columbia, Canada,~J. Geophys.
+Res., 114, B04405,
+doi:\href{http://dx.doi.org/10.1029/2008jb006077}{10.1029/2008jb006077},
+2009.
+
+\bibitem[{Johnsen et~al.(1995)Johnsen, Dahl-Jensen, Dansgaard, and Gundestrup}]{Johnsen.etal.1995}
+Johnsen,~S.~J., Dahl-Jensen,~D., Dansgaard,~W., and Gundestrup,~N.: Greenland palaeotemperatures derived from GRIP bore hole temperature and ice core isotope profiles, Tellus B, 47, 624--629,
+doi:\href{http://dx.doi.org/10.1034/j.1600-0889.47.issue5.9.x}{10.1034/j.1600-0889.47.issue5.9.x}, 1995.
+
+\bibitem[{Jouzel et~al.(2007)Jouzel, Masson-Delmotte, Cattani, Dreyfus, Falourd, Hoffmann, Minster, Nouet, Barnola, Chappellaz, Fischer, Gallet, Johnsen, Leuenberger, Loulergue, Luethi, Oerter, Parrenin, Raisbeck, Raynaud, Schilt, Schwander, Selmo, Souchez, Spahni, Stauffer, Steffensen, Stenni, Stocker, Tison, Werner, and Wolff}]{Jouzel.etal.2007}
+Jouzel,~J., Masson-Delmotte,~V., Cattani,~O., Dreyfus,~G., Falourd,~S., Hoffmann,~G., Minster,~B., Nouet,~J., Barnola,~J.~M., Chappellaz,~J., Fischer,~H., Gallet,~J.~C., Johnsen,~S., Leuenberger,~M., Loulergue,~L., Luethi,~D., Oerter,~H., Parrenin,~F., Raisbeck,~G., Raynaud,~D., Schilt,~A., Schwander,~J., Selmo,~E., Souchez,~R., Spahni,~R., Stauffer,~B., Steffensen,~J.~P., Stenni,~B., Stocker,~T.~F., Tison,~J.~L., Werner,~M., and Wolff,~E.~W.: Orbital and millennial antarctic climate variability over the past 800\,000~years, Sience, 317, 793--796,
+doi:\href{http://dx.doi.org/10.1126/science.1141038}{10.1126/science.1141038}, data archived at the World Data Center for Paleoclimatology, Boulder, Colorado, USA., 2007.
+
+\bibitem[{Kaufman and Manley(2004)}]{Kaufman.Manley.2004}
+Kaufman,~D.~S. and Manley,~W.~F.: Pleistocene maximum and late wisconsinan glacier extents across Alaska, {USA}, in: Dev. Quaternary Sci., vol.~2,
+edited by: Ehlers,~J. and Gibbard,~P.~L., Elsevier, Amsterdam, 9--27,
+doi:\href{http://dx.doi.org/10.1016/S1571-0866(04)80182-9}{10.1016/S1571-0866(04)80182-9}, 2004.
+
+\bibitem[{Kleman(1994)}]{Kleman.1994}
+Kleman,~J.: Preservation of landforms under ice sheets and ice caps, Geomorphology, 9, 19--32,
+doi:\href{http://dx.doi.org/10.1016/0169-555x(94)90028-0}{10.1016/0169-555x(94)90028-0}, 1994.
+
+\bibitem[{Kleman and Stroeven(1997)}]{Kleman.Stroeven.1997}
+Kleman,~J. and Stroeven,~A.~P.: Preglacial surface remnants and Quaternary glacial regimes in northwestern Sweden, Geomorphology, 19, 35--54,
+doi:\href{http://dx.doi.org/10.1016/s0169-555x(96)00046-3}{10.1016/s0169-555x(96)00046-3}, 1997.
+
+\bibitem[{Kleman et~al.(1997)Kleman, H{\"a}ttestrand, Borgstr{\"o}m, and Stroeven}]{Kleman.etal.1997}
+Kleman,~J., H{\"a}ttestrand,~C., Borgstr{\"o}m,~I., and Stroeven,~A.: Fennoscandian paleoglaciology reconstructed using a glacial geological inversion model,~J. Glaciol., 43, 283--299, 1997.
+
+\bibitem[{Kleman et~al.(2006)Kleman, H{\"a}ttestrand, Stroeven, Jansson, De~Angelis, and Borgstr{\"o}m}]{Kleman.etal.2006}
+Kleman,~J., H{\"a}ttestrand,~C., Stroeven,~A.~P., Jansson,~K.~N., De~Angelis,~H., and Borgstr{\"o}m,~I.: Reconstruction of palaeo-ice sheets -- inversion of their glacial geomorphological record, in: Glacier Science and Environmental Change, edited by: Knight,~P.~G., Blackwell, Malden, MA, 192--198,
+doi:\href{http://dx.doi.org/10.1002/9780470750636.ch38}{10.1002/9780470750636.ch38}, 2006.
+
+\bibitem[{Kleman et~al.(2008)Kleman, Stroeven, and Lundqvist}]{Kleman.etal.2008}
+Kleman,~J., Stroeven,~A.~P., and Lundqvist,~J.: Patterns of quaternary ice sheet erosion and deposition in Fennoscandia and a theoretical framework for explanation, Geomorphology, 97, 73--90,
+doi:\href{http://dx.doi.org/10.1016/j.geomorph.2007.02.049}{10.1016/j.geomorph.2007.02.049}, 2008.
+
+\bibitem[{Kleman et~al.(2010)Kleman, Jansson, De~Angelis, Stroeven, H\"{a}ttestrand, Alm, and Glasser}]{Kleman.etal.2010}
+Kleman,~J., Jansson,~K., De~Angelis,~H., Stroeven,~A., H\"{a}ttestrand,~C., Alm,~G., and Glasser,~N.: North American ice sheet build-up during the last glacial cycle, 115--21\,kyr, Quaternary Sci.
+Rev., 29, 2036--2051,
+doi:\href{http://dx.doi.org/10.1016/j.quascirev.2010.04.021}{10.1016/j.quascirev.2010.04.021}, 2010.
+
+\bibitem[{Kovanen(2002)}]{Kovanen.2002}
+Kovanen,~D.~J.: Morphologic and stratigraphic evidence for Aller{\o}d and Younger Dryas age glacier fluctuations of the Cordilleran Ice Sheet, British Columbia, Canada and Northwest Washington,~USA, Boreas, 31, 163--184,
+doi:\href{http://dx.doi.org/10.1111/j.1502-3885.2002.tb01064.x}{10.1111/j.1502-3885.2002.tb01064.x}, 2002.
+
+\bibitem[{Kovanen and Easterbrook(2002)}]{Kovanen.Easterbrook.2002}
+Kovanen,~D.~J.
+and Easterbrook,~D.~J.: Timing and extent of Aller{\o}d and Younger Dryas age (ca.
+12\,500--10\,000 $^{14}$C~yr~BP) oscillations of the Cordilleran Ice Sheet in the Fraser Lowland, Western North America, Quaternary Res., 57, 208--224,
+doi:\href{http://dx.doi.org/10.1006/qres.2001.2307}{10.1006/qres.2001.2307}, 2002.
+
+\bibitem[{Lakeman et~al.(2008)Lakeman, Clague, and Menounos}]{Lakeman.etal.2008}
+Lakeman,~T.~R., Clague,~J.~J., and Menounos,~B.: Advance of alpine glaciers during final retreat of the Cordilleran ice sheet in the Finlay River area, northern British Columbia, Canada, Quaternary Res., 69, 188--200,
+doi:\href{http://dx.doi.org/10.1016/j.yqres.2008.01.002}{10.1016/j.yqres.2008.01.002}, 2008.
+
+\bibitem[{Langen et~al.(2012)Langen, Solgaard, and Hvidberg}]{Langen.etal.2012}
+Langen,~P.~L., Solgaard,~A.~M., and Hvidberg,~C.~S.: Self-inhibiting growth of the Greenland ice sheet, Geophys. Res. Lett., 39, L12502,
+doi:\href{http://dx.doi.org/10.1029/2012GL051810}{10.1029/2012GL051810}, 2012.
+
+\bibitem[{Levermann et~al.(2012)Levermann, Albrecht, Winkelmann, Martin, Haseloff, and Joughin}]{Levermann.etal.2012}
+ Levermann,~A., Albrecht,~T., Winkelmann,~R., Martin,~M.~A., Haseloff,~M., and Joughin,~I.: Kinematic first-order calving law implies potential for  abrupt ice-shelf retreat, The Cryosphere, 6, 273--286,
+doi:\href{http://dx.doi.org/10.5194/tc-6-273-2012}{10.5194/tc-6-273-2012}, 2012.
+
+\bibitem[{Lingle and Clark(1985)}]{Lingle.Clark.1985}
+Lingle,~C.~S. and Clark,~J.~A.: A~numerical model of interactions between a~marine ice sheet and the Solid Earth: application to a~West Antarctic ice stream,~J. Geophys. Res., 90, 1100--1114,
+doi:\href{http://dx.doi.org/10.1029/JC090iC01p01100}{10.1029/JC090iC01p01100}, 1985.
+
+\bibitem[{Lisiecki and Raymo(2005)}]{Lisiecki.Raymo.2005}
+Lisiecki,~L.~E. and Raymo,~M.~E.: A Pliocene-Pleistocene stack of 57 globally distributed benthic $\delta^{18}$O records, Paleoceanography, 20, PA1003,
+doi:\href{http://dx.doi.org/10.1029/2004pa001071}{10.1029/2004pa001071}, 2005.
+
+\bibitem[{Lundqvist(1987)}]{Lundqvist.1987}
+Lundqvist,~J.: Glaciodynamics of the Younger Dryas Marginal zone in Scandinavia: implications of a revised glaciation model, Geogr. Ann. A, 69, 305,
+doi:\href{http://dx.doi.org/10.2307/521191}{10.2307/521191}, 1987.
+
+\bibitem[{Margold et~al.(2011)Margold, Jansson, Kleman, and Stroeven}]{Margold.etal.2011}
+Margold,~M., Jansson,~K.~N., Kleman,~J., and Stroeven,~A.~P.: Glacial meltwater landforms of central British Columbia,~J. Maps, 7, 486--506,
+doi:\href{http://dx.doi.org/10.4113/jom.2011.1205}{10.4113/jom.2011.1205}, 2011.
+
+\bibitem[{Margold et~al.(2013{\natexlab{a}})Margold, Jansson, Kleman, and Stroeven}]{Margold.etal.2013}
+Margold,~M., Jansson,~K.~N., Kleman,~J., and Stroeven,~A.~P.: Lateglacial ice dynamics of the Cordilleran Ice Sheet in northern British Columbia and southern Yukon Territory: retreat pattern of the Liard Lobe reconstructed from the glacial landform record,~J.
+Quaternary Sci., 28, 180--188,
+doi:\href{http://dx.doi.org/10.1002/jqs.2604}{10.1002/jqs.2604}, 2013{\natexlab{a}}.
+
+\bibitem[{Margold et~al.(2013{\natexlab{b}})Margold, Jansson, Kleman, Stroeven, and Clague}]{Margold.etal.2013a}
+Margold,~M., Jansson,~K.~N., Kleman,~J., Stroeven,~A.~P., and Clague,~J.~J.: Retreat pattern of the Cordilleran Ice Sheet in central British Columbia at the end of the last glaciation reconstructed from glacial meltwater landforms, Boreas, 42, 830--847,
+doi:\href{http://dx.doi.org/10.1111/bor.12007}{10.1111/bor.12007}, 2013{\natexlab{b}}.
+
+\bibitem[{Margold et~al.(2014)Margold, Stroeven, Clague, and Heyman}]{Margold.etal.2014}
+Margold,~M., Stroeven,~A.~P., Clague,~J.~J., and Heyman,~J.: Timing of terminal Pleistocene deglaciation at high elevations in southern and central British Columbia constrained by $^{10}$Be exposure dating, Quaternary Res., 99, 193--202,
+doi:\href{http://dx.doi.org/10.1016/j.quascirev.2014.06.027}{10.1016/j.quascirev.2014.06.027}, 2014.
+
+\bibitem[{Marshall et~al.(2000)Marshall, Tarasov, Clarke, and Peltier}]{Marshall.etal.2000}
+Marshall,~S.~J., Tarasov,~L., Clarke,~G.~K., and Peltier,~W.~R.: Glaciological reconstruction of the Laurentide Ice Sheet: physical processes and modelling challenges, Can.~J.
+Earth Sci., 37, 769--793,
+doi:\href{http://dx.doi.org/10.1139/e99-113}{10.1139/e99-113}, 2000.
+
+\bibitem[{Martin et~al.(2011)Martin, Winkelmann, Haseloff, Albrecht, Bueler, Khroulev, and Levermann}]{Martin.etal.2011}
+ Martin,~M.~A., Winkelmann,~R., Haseloff,~M., Albrecht,~T., Bueler,~E., Khroulev,~C., and Levermann,~A.: The Potsdam Parallel Ice Sheet Model (PISM-PIK) -- Part 2: Dynamic equilibrium simulation of the Antarctic ice sheet, The Cryosphere, 5, 727--740,
+doi:\href{http://dx.doi.org/10.5194/tc-5-727-2011}{10.5194/tc-5-727-2011}, 2011.
+
+\bibitem[{Menounos et~al.(2008)Menounos, Osborn, Clague, and Luckman}]{Menounos.etal.2008}
+Menounos,~B., Osborn,~G., Clague,~J.~J., and Luckman,~B.~H.: Latest Pleistocene and Holocene glacier fluctuations in western Canada, Quaternary Sci. Rev., 28, 2049--2074,
+doi:\href{http://dx.doi.org/10.1016/j.quascirev.2008.10.018}{10.1016/j.quascirev.2008.10.018}, 2008.
+
+\bibitem[{Mesinger et~al.(2006)Mesinger, DiMego, Kalnay, Mitchell, Shafran, Ebisuzaki, Jovi\'{c}, Woollen, Rogers, Berbery, Ek, Fan, Grumbine, Higgins, Li, Lin, Manikin, Parrish, and Shi}]{Mesinger.etal.2006}
+Mesinger,~F., DiMego,~G., Kalnay,~E., Mitchell,~K., Shafran,~P.~C., Ebisuzaki,~W., Jovi\'{c},~D., Woollen,~J., Rogers,~E., Berbery,~E.~H., Ek,~M.~B., Fan,~Y., Grumbine,~R., Higgins,~W., Li,~H., Lin,~Y., Manikin,~G., Parrish,~D., and Shi,~W.: North American regional reanalysis,~B.
+Am.
+Meteorol.
+Soc., 87, 343--360,
+doi:\href{http://dx.doi.org/10.1175/BAMS-87-3-343}{10.1175/BAMS-87-3-343}, 2006.
+
+\bibitem[{Osborn and Gerloff(1997)}]{Osborn.Gerloff.1997}
+Osborn,~G. and Gerloff,~L.: Latest pleistocene and early Holocene fluctuations of glaciers in the Canadian and northern American Rockies, Quatern. Int., 38--39, 7--19,
+doi:\href{http://dx.doi.org/10.1016/s1040-6182(96)00026-2}{10.1016/s1040-6182(96)00026-2}, 1997.
+
+\bibitem[{Patterson and Kelso(2015)}]{Patterson.Kelso.2015}
+Patterson,~T. and Kelso,~N.~V.: {N}atural {E}arth, {F}ree vector and raster
+map data, available at: \url{http://naturalearthdata.com} (last access:
+2015), 2015.
+
+\bibitem[{Perkins and Brennand(2014)}]{Perkins.Brennand.2014}
+Perkins,~A.~J.
+and Brennand,~T.~A.: Refining the pattern and style of Cordilleran Ice Sheet retreat: palaeogeography, evolution and implications of lateglacial ice-dammed lake systems on the southern Fraser Plateau, British Columbia, Canada, Boreas, 44, 319--342,
+doi:\href{http://dx.doi.org/10.1111/bor.12100}{10.1111/bor.12100}, 2014.
+
+\bibitem[{Perkins et~al.(2013)Perkins, Brennand, and Burke}]{Perkins.etal.2013}
+Perkins,~A.~J., Brennand,~T.~A., and Burke,~M.~J.: Genesis of an esker-like ridge over the southern Fraser Plateau, British Columbia: implications for paleo-ice sheet reconstruction based on geomorphic inversion, Geomorphology, 190, 27--39,
+doi:\href{http://dx.doi.org/10.1016/j.geomorph.2013.02.005}{10.1016/j.geomorph.2013.02.005}, 2013.
+
+\bibitem[{Petit et~al.(1999)Petit, Jouzel, Raynaud, Barkov, Barnola, Basile, Bender, Chappellaz, Davis, Delaygue, Delmotte, Kotlyakov, Legrand, Lipenkov, Lorius, P{\'e}pin, Ritz, Saltzman, and Stievenard}]{Petit.etal.1999}
+Petit,~J.~R., Jouzel,~J., Raynaud,~D., Barkov,~N.~I., Barnola,~J.-M., Basile,~I., Bender,~M., Chappellaz,~J., Davis,~M., Delaygue,~G., Delmotte,~M., Kotlyakov,~V.~M., Legrand,~M., Lipenkov,~V.~Y., Lorius,~C., P{\'e}pin,~L., Ritz,~C., Saltzman,~E., and Stievenard,~M.: Climate and atmospheric history of the past 420,000 years from the Vostok ice core, Antarctica, Nature, 399, 429--436,
+doi:\href{http://dx.doi.org/10.1038/20859}{10.1038/20859}, data archived at the World Data Center for Paleoclimatology, Boulder, Colorado, USA., 1999.
+
+\bibitem[{Porter(1989)}]{Porter.1989}
+Porter,~S.~C.: Some geological implications of average Quaternary glacial conditions, Quaternary Res., 32, 245--261,
+doi:\href{http://dx.doi.org/10.1016/0033-5894(89)90092-6}{10.1016/0033-5894(89)90092-6}, 1989.
+
+\bibitem[{Porter and Swanson(1998)}]{Porter.Swanson.1998}
+Porter,~S.~C.
+and Swanson,~T.~W.: Radiocarbon age constraints on rates of advance and retreat of the {P}uget {L}obe of the {C}ordilleran {I}ce {S}heet during the last glaciation, Quaternary Res., 50, 205--213,
+doi:\href{http://dx.doi.org/10.1006/qres.1998.2004}{10.1006/qres.1998.2004}, 1998.
+
+\bibitem[{Praetorius and Mix(2014)}]{Praetorius.Mix.2014}
+Praetorius,~S.~K. and Mix,~A.~C.: Synchronization of North Pacific and Greenland climates preceded abrupt deglacial warming, Sience, 345, 444--448,
+doi:\href{http://dx.doi.org/10.1126/science.1252000}{10.1126/science.1252000}, 2014.
+
+\bibitem[{Prest et~al.(1968)Prest, Grant, and Rampton}]{Prest.etal.1968}
+Prest,~V.~K., Grant,~D.~R., and Rampton,~V.~N.: Glacial map of Canada, ``A'' Series Map 1253A, Geol. Surv. of Can., Ottawa, ON,
+doi:\href{http://dx.doi.org/10.4095/108979}{10.4095/108979}, 1968.
+
+\bibitem[{Robert(1991)}]{Robert.1991}
+Robert,~B.~L.: Modeling the Cordilleran ice sheet, G{e}ogr. Phys. Quatern., 45, 287--299,
+doi:\href{http://dx.doi.org/10.7202/032876ar}{10.7202/032876ar}, 1991.
+
+\bibitem[{Rutter et~al.(2012)Rutter, Coronato, Helmens, Rabassa, and Z{\'a}rate}]{Rutter.etal.2012}
+Rutter,~N., Coronato,~A., Helmens,~K., Rabassa,~J., and Z{\'a}rate,~M.: Glaciations in North and South America from the Miocene to the Last Glacial Maximum: Comparisons, Linkages and Uncertainties, Springer Briefs in Earth System Sciences, Springer, Dordrecht,
+doi:\href{http://dx.doi.org/10.1007/978-94-007-4399-1}{10.1007/978-94-007-4399-1}, 2012.
+
+\bibitem[{Ryder et~al.(1991)Ryder, Fulton, and Clague}]{Ryder.etal.1991}
+Ryder,~J.~M., Fulton,~R.~J., and Clague,~J.~J.: The Cordilleran ice sheet and the glacial geomorphology of southern and central British Colombia, G{e}ogr. Phys. Quatern., 45, 365--377,
+doi:\href{http://dx.doi.org/10.7202/032882ar}{10.7202/032882ar}, 1991.
+
+\bibitem[{Seguinot(2013)}]{Seguinot.2013}
+Seguinot,~J.: Spatial and seasonal effects of temperature variability in a positive degree-day glacier surface mass-balance model,~J. Glaciol., 59, 1202--1204,
+doi:\href{http://dx.doi.org/10.3189/2013JoG13J081}{10.3189/2013JoG13J081}, 2013.
+
+\bibitem[{Seguinot et~al.(2014)Seguinot, Khroulev, Rogozhina, Stroeven, and Zhang}]{Seguinot.etal.2014}
+ Seguinot,~J., Khroulev,~C., Rogozhina,~I., Stroeven,~A.~P., and Zhang,~Q.: The effect of climate forcing on numerical simulations of the Cordilleran ice sheet at the Last Glacial Maximum, The Cryosphere, 8, 1087--1103,
+doi:\href{http://dx.doi.org/10.5194/tc-8-1087-2014}{10.5194/tc-8-1087-2014}, 2014.
+
+\bibitem[{Shea et~al.(2009)Shea, Moore, and Stahl}]{Shea.etal.2009}
+Shea,~J.~M., Moore,~R.~D., and Stahl,~K.: Derivation of melt factors from glacier mass-balance records in western Canada,~J. Glaciol., 55, 123--130,
+doi:\href{http://dx.doi.org/10.3189/002214309788608886}{10.3189/002214309788608886}, 2009.
+
+\bibitem[{Sissons(1979)}]{Sissons.1979}
+Sissons,~J.~B.: The Loch Lomond Stadial in the British Isles, Nature, 280, 199--203,
+doi:\href{http://dx.doi.org/10.1038/280199a0}{10.1038/280199a0}, 1979.
+
+\bibitem[{Stea et~al.(2011)Stea, Seaman, Pronk, Parkhill, Allard, and Utting}]{Stea.etal.2011}
+Stea,~R.~R., Seaman,~A.~A., Pronk,~T., Parkhill,~M.~A., Allard,~S., and Utting,~D.: The Appalachian Glacier Complex in Maritime Canada, in:   Dev.
+Quaternary Sci., vol.~15, edited by: Ehlers,~J., Gibbard,~P.~L., and Hughes,~P.~D., Elsevier, Amsterdam, 631--659,
+doi:\href{http://dx.doi.org/10.1016/b978-0-444-53447-7.00048-9}{10.1016/b978-0-444-53447-7.00048-9}, 2011.
+
+\bibitem[{Stroeven et~al.(2010)Stroeven, Fabel, Codilean, Kleman, Clague, Miguens-Rodriguez, and Xu}]{Stroeven.etal.2010}
+Stroeven,~A.~P., Fabel,~D., Codilean,~A.~T., Kleman,~J., Clague,~J.~J., Miguens-Rodriguez,~M., and Xu,~S.: Investigating the glacial history of the northern sector of the Cordilleran ice sheet with cosmogenic $^{10}$Be concentrations in quartz, Quaternary Sci.
+Rev., 29, 3630--3643,
+doi:\href{http://dx.doi.org/10.1016/j.quascirev.2010.07.010}{10.1016/j.quascirev.2010.07.010}, 2010.
+
+\bibitem[{Stroeven et~al.(2014)Stroeven, Fabel, Margold, Clague, and Xu}]{Stroeven.etal.2014}
+Stroeven,~A.~P., Fabel,~D., Margold,~M., Clague,~J.~J., and Xu,~S.: Investigating absolute chronologies of glacial advances in the {NW} sector of the Cordilleran ice sheet with terrestrial in situ cosmogenic nuclides, Quaternary Sci.
+Rev., 92, 429--443,
+doi:\href{http://dx.doi.org/10.1016/j.quascirev.2013.09.026}{10.1016/j.quascirev.2013.09.026}, 2014.
+
+\bibitem[{Stroeven et~al.(2015)Stroeven, H{\"a}ttestrand, Kleman, Heyman, Fabel, Fredin, Goodfellow, Harbor, Jansen, Olsen, Caffee, Fink, Lundqvist, Rosqvist, Str\"omberg, and Jansson}]{Stroeven.etal.inreview}
+Stroeven,~A.~P., H{\"a}ttestrand,~C., Kleman,~J., Heyman,~J., Fabel,~D.,
+Fredin,~O., Goodfellow,~B.~W., Harbor,~J.~M., Jansen,~J.~D., Olsen,~L.,
+Caffee,~M.~W., Fink,~D., Lundqvist,~J., Rosqvist,~G.~C., Str\"omberg,~B., and
+Jansson,~K.~N.: Deglaciation of Fennoscandia, Quaternary Sci. Rev., in
+review, 2015.
+
+\bibitem[{Stumpf et~al.(2000)Stumpf, Broster, and Levson}]{Stumpf.etal.2000}
+Stumpf,~A.~J., Broster,~B.~E., and Levson,~V.~M.: Multiphase flow of the late Wisconsinan Cordilleran ice sheet in western Canada, Geol. Soc. Am. Bull., 112, 1850--1863,
+doi:\href{http://dx.doi.org/10.1130/0016-7606(2000)112<1850:mfotlw>2.0.co;2}{10.1130/0016-7606(2000)112\textless1850:mfotlw\textgreater2.0.co;2}, 2000.
+
+\bibitem[{Taylor et~al.(2014)Taylor, Hendy, and Pak}]{Taylor.etal.2014}
+Taylor,~M., Hendy,~I., and Pak,~D.: Deglacial ocean warming and marine margin retreat of the Cordilleran Ice Sheet in the North Pacific Ocean, Earth Planet. Sc. Lett., 403, 89--98,
+doi:\href{http://dx.doi.org/10.1016/j.epsl.2014.06.026}{10.1016/j.epsl.2014.06.026}, 2014.
+
+\bibitem[{the PISM~authors(2015)}]{PISM-authors.2015}
+the PISM~authors: {PISM}, a~{P}arallel {I}ce {S}heet {M}odel, available at:
+\url{http://www.pism-docs.org} (last access: 2015), 2015.
+
+\bibitem[{Troost(2014)}]{Troost.2014}
+Troost,~K.~G.: The penultimate glaciation and mid-to late-pleistocene
+stratigraphy in the Central Puget Lowland, Washington, in: 2014 GSA Annual
+Meeting, Vancouver, 19--22 October 2014, abstract no. 138-9, 2014.
+
+\bibitem[{Tulaczyk et~al.(2000)Tulaczyk, Kamb, and Engelhardt}]{Tulaczyk.etal.2000}
+Tulaczyk,~S., Kamb,~W.~B., and Engelhardt,~H.~F.: Basal mechanics of Ice Stream B, west Antarctica: 1. Till mechanics,~J. Geophys. Res., 105, 463,
+doi:\href{http://dx.doi.org/10.1029/1999jb900329}{10.1029/1999jb900329}, 2000.
+
+\bibitem[{Turner et~al.(2013)Turner, Ward, Bond, Jensen, Froese, Telka, Zazula, and Bigelow}]{Turner.etal.2013}
+Turner,~D.~G., Ward,~B.~C., Bond,~J.~D., Jensen,~B.~J., Froese,~D.~G., Telka,~A.~M., Zazula,~G.~D., and Bigelow,~N.~H.: Middle to Late Pleistocene ice extents, tephrochronology and paleoenvironments of the White River area, southwest Yukon, Quaternary Res., 75, 59--77,
+doi:\href{http://dx.doi.org/10.1016/j.quascirev.2013.05.011}{10.1016/j.quascirev.2013.05.011}, 2013.
+
+\bibitem[{Ward et~al.(2007)Ward, Bond, and Gosse}]{Ward.etal.2007}
+Ward,~B.~C., Bond,~J.~D., and Gosse,~J.~C.: Evidence for a 55--50 ka (early Wisconsin) glaciation of the Cordilleran ice sheet, Yukon Territory, Canada, Quaternary Res., 68, 141--150,
+doi:\href{http://dx.doi.org/10.1016/j.yqres.2007.04.002}{10.1016/j.yqres.2007.04.002}, 2007.
+
+\bibitem[{Ward et~al.(2008)Ward, Bond, Froese, and Jensen}]{Ward.etal.2008}
+Ward,~B.~C., Bond,~J.~D., Froese,~D., and Jensen,~B.: Old Crow tephra ($140\pm10$\,ka) constrains penultimate Reid glaciation in central Yukon Territory, Quaternary Res., 27, 1909--1915,
+doi:\href{http://dx.doi.org/10.1016/j.quascirev.2008.07.012}{10.1016/j.quascirev.2008.07.012}, 2008.
+
+\bibitem[{Winkelmann et~al.(2011)Winkelmann, Martin, Haseloff, Albrecht, Bueler, Khroulev, and Levermann}]{Winkelmann.etal.2011}
+ Winkelmann,~R., Martin,~M.~A., Haseloff,~M., Albrecht,~T., Bueler,~E., Khroulev,~C., and Levermann,~A.: The Potsdam Parallel Ice Sheet Model (PISM-PIK) -- Part 1: Model description, The Cryosphere, 5, 715--726,
+doi:\href{http://dx.doi.org/10.5194/tc-5-715-2011}{10.5194/tc-5-715-2011}, 2011.
+
+\end{thebibliography}
 

### 2016-01-19 cordillera-cycle.tex
Add diff file to makefile. Done typesetting.
Fix misnamed figure labels.

@@
 \graphicspath{{figures/}}
+\hypersetup{hidelinks}
 
@@
       meridional ice divide over the western flank of the Rocky Mountains
-      (Figs.~\ref{fig:lr_maps}, lower panels and \ref{fig:mis2}). This
-      result appears to contrast with palaeoglaciological reconstructions
-      for central and southern British Columbia with ice divides in a~more
-      westerly position, over the western margin of the Interior Plateau
-      \citep{Ryder.etal.1991, Stumpf.etal.2000, Kleman.etal.2010,
-      Clague.Ward.2011, Margold.etal.2013a}. These indicate that
-      a~latitudinal saddle connected ice dispersal centres in the Columbia
-      Mountains with the main ice divide \citep{Ryder.etal.1991,
-      Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}. A~latitudinal
-      saddle does indeed feature in our modelling results, however, in an
-      inverse configuration between the main ice divide over the Columbia
-      Mountains and a~secondary divide over the southern Coast Mountains
-      (Fig.~\ref{fig:mis2}).
+      (Figs.~\ref{fig:lr_maps}, lower panels and \ref{fig:hr_maps_mis2}).
+      This result appears to contrast with palaeoglaciological
+      reconstructions for central and southern British Columbia with ice
+      divides in a~more westerly position, over the western margin of the
+      Interior Plateau \citep{Ryder.etal.1991, Stumpf.etal.2000,
+      Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}. These
+      indicate that a~latitudinal saddle connected ice dispersal centres in
+      the Columbia Mountains with the main ice divide
+      \citep{Ryder.etal.1991, Kleman.etal.2010, Clague.Ward.2011,
+      Margold.etal.2013a}. A~latitudinal saddle does indeed feature in our
+      modelling results, however, in an inverse configuration between the
+      main ice divide over the Columbia Mountains and a~secondary divide
+      over the southern Coast Mountains (Fig.~\ref{fig:hr_maps_mis2}).
 
@@
       the ice divide ensure significant ice drainage across the Rocky
-      Mountains at the LGM (Fig.~\ref{fig:mis2}).
+      Mountains at the LGM (Fig.~\ref{fig:hr_maps_mis2}).
 
@@
       forcing than those corresponding to ice volume maxima during MIS~2
-      (Figs.~\ref{fig:lr_maps}, upper panels and \ref{fig:mis4}). The GRIP
-      simulation (Fig.~\ref{fig:mis4}, left panel) results in a~modelled
-      maximum ice sheet extent that closely resembles that obtained during
-      MIS~2, with the only major difference of being slightly less extensive
-      across northern and eastern sectors. In contrast, the EPICA simulation
-      produces a~lower ice volume maximum (Fig.~\ref{fig:lr_ts}), which
-      translates in the modelled ice sheet geometry into a~significantly
-      reduced southern sector, more restricted ice cover in northern and
-      eastern sectors, and generally lower ice surface elevations in the
-      interior (Fig.~\ref{fig:mis4}, right panel). Thus, only the GRIP
-      simulation can explain the presence of MIS~4 glacial deposits in the
-      Puget Lowland \citep{Troost.2014} and that of ice-rafted debris in the
-      marine sediment record offshore Vancouver Island at
-      ca.~47\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008}.
+      (Figs.~\ref{fig:lr_maps}, upper panels and \ref{fig:hr_maps_mis4}).
+      The GRIP simulation (Fig.~\ref{fig:hr_maps_mis4}, left panel) results
+      in a~modelled maximum ice sheet extent that closely resembles that
+      obtained during MIS~2, with the only major difference of being
+      slightly less extensive across northern and eastern sectors. In
+      contrast, the EPICA simulation produces a~lower ice volume maximum
+      (Fig.~\ref{fig:lr_ts}), which translates in the modelled ice sheet
+      geometry into a~significantly reduced southern sector, more restricted
+      ice cover in northern and eastern sectors, and generally lower ice
+      surface elevations in the interior (Fig.~\ref{fig:hr_maps_mis4}, right
+      panel). Thus, only the GRIP simulation can explain the presence of
+      MIS~4 glacial deposits in the Puget Lowland \citep{Troost.2014} and
+      that of ice-rafted debris in the marine sediment record offshore
+      Vancouver Island at ca.~47\,\unit{\chem{^{14}C}\,cal\,ka}
+      \citep{Cosma.etal.2008}.
 
@@
       Modelled patterns of ice sheet retreat are relatively consistent
-      between the two simulations (Figs.~\ref{fig:hr_geom_deglacageshots}
+      between the two simulations (Figs.~\ref{fig:hr_maps_deglac}
       and~\ref{fig:hr_geom_deglacage}). The southern sector of the modelling
@@
       a~modelled ice divide of the LGM ice sheet being too far to the east
-      (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
+      (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:hr_maps_mis2};
       \citealp{Seguinot.etal.2014}). A~more westerly-located LGM ice divide
@@
       during MIS~2 in the GRIP and EPICA high-resolution simulations.}
-\label{fig:mis2}%
+\label{fig:hr_maps_mis2}%
 \end{figure*}%
@@
       during MIS~4 in the GRIP and EPICA high-resolution simulations.}
-\label{fig:mis4}%
+\label{fig:hr_maps_mis4}%
 \end{figure*}%
@@
       and~\ref{fig:hr_pf_epica}.}
-\label{fig:hr_geom_deglacageshots}%
+\label{fig:hr_maps_deglac}%
 \end{figure*}%
@@
       simulation, corresponding to the four transects indicated in
-      Figs.~\ref{fig:hr_geom_deglacageshots}
+      Figs.~\ref{fig:hr_maps_deglac}
       and~\ref{fig:hr_geom_deglacage}.}

### 2016-01-19 typeset.tex
Typeset: rewrap figure and table captions.
Fix misnamed figure labels.

@@
       a~modelled ice divide of the LGM ice sheet beingtoo far to the east
-      (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
+      (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:hr_maps_mis2};
       \citealp{Seguinot.etal.2014}). A~more westerly-located LGM ice divide
@@
 \begin{table}[t]
-\caption{Parameter values used in the ice sheet model.}
+\caption{%
+      Parameter values used in the ice sheet model.}
 %\scalebox{.85}[.85]
@@
       temperature offset time-series used to force the ice sheet model
-           through the last glacial cycle (Fig.~\ref{fig:lr_ts}). $f$
-           corresponds to the scaling factor adopted to yield last glacial
-           maximum ice limits in the vicinity of mapped end moraines, and
-           $T_{[32, 22]}$ refers to the resulting mean temperature anomaly
-           during the period $-$32 to~$-$22\,\unit{ka} after scaling.}
+      through the last glacial cycle (Fig.~\ref{fig:lr_ts}). $f$ corresponds
+      to the scaling factor adopted to yield last glacial maximum ice limits
+      in the vicinity of mapped end moraines, and $T_{[32, 22]}$ refers to
+      the resulting mean temperature anomaly during the period $-$32
+      to~$-$22\,\unit{ka} after scaling.}
 \label{tab:records}
@@
       sheet model (top panel), and modelled ice volume (bottom panel)
-           through the last 120\,\unit{ka}. Ice volumes are expressed in meters of sea
-           level equivalent (m\,s.l.e.). Gray fields indicate Marine Oxygen
-           Isotope Stage (MIS) boundaries for MIS~2 and MIS~4 according to
+      through the last 120\,\unit{ka}. Ice volumes are expressed in meters
+      of sea level equivalent (m\,s.l.e.). Gray fields indicate Marine
+      Oxygen Isotope Stage (MIS) boundaries for MIS~2 and MIS~4 according to
       a~global compilation of benthic \chem{\delta^{18}O} records
@@
 \caption{%
-Modelled surface topography (200\,\unit{m} contours) and surface velocity
-           (colour mapping) corresponding to the maximum ice volume during
-           MIS~2 in the GRIP and EPICA high-resolution simulations.}
-\label{fig:mis2}%
+      Modelled surface topography (200\,\unit{m} contours) and surface
+      velocity (colour mapping) corresponding to the maximum ice volume
+      during MIS~2 in the GRIP and EPICA high-resolution simulations.}
+\label{fig:hr_maps_mis2}%
 \end{figure}%
@@
 \caption{%
-Modelled surface topography (200\,\unit{m} contours) and surface velocity
-           (colour mapping) corresponding to the maximum ice volume during
-           MIS~4 in the GRIP and EPICA high-resolution simulations.}
-\label{fig:mis4}%
+      Modelled surface topography (200\,\unit{m} contours) and surface
+      velocity (colour mapping) corresponding to the maximum ice volume
+      during MIS~4 in the GRIP and EPICA high-resolution simulations.}
+\label{fig:hr_maps_mis4}%
 \end{figure}%
@@
 \caption{%
-Snapshots of modelled surface topography (200\,\unit{m} contours)
-           and surface velocity (colour mapping) during the last deglaciation
-           from the GRIP (top panels) and EPICA (bottom panels) 5\,\unit{km} simulations.
-           Dashed segments \textbf{(A--D)} indicate the location of profiles used in
-           Figs.~\ref{fig:hr_pf_grip} and~\ref{fig:hr_pf_epica}.}
-\label{fig:hr_geom_deglacageshots}%
+      Snapshots of modelled surface topography (200\,\unit{m} contours) and
+      surface velocity (colour mapping) during the last deglaciation from
+      the GRIP (top panels) and EPICA (bottom panels) 5\,\unit{km}
+      simulations. Dashed segments \textbf{(A--D)} indicate the location of
+      profiles used in Figs.~\ref{fig:hr_pf_grip}
+      and~\ref{fig:hr_pf_epica}.}
+\label{fig:hr_maps_deglac}%
 \end{figure}%
@@
       simulation, corresponding to the four transects indicated in
-           Figs.~\ref{fig:hr_geom_deglacageshots} and~\ref{fig:hr_geom_deglacage}.}
+      Figs.~\ref{fig:hr_maps_deglac}
+      and~\ref{fig:hr_geom_deglacage}.}
 \label{fig:hr_pf_grip}%
@@
       simulation, corresponding to the four transects indicated in
-           Figs.~\ref{fig:hr_geom_deglacageshots} and~\ref{fig:hr_geom_deglacage}.}
+      Figs.~\ref{fig:hr_maps_deglac}
+      and~\ref{fig:hr_geom_deglacage}.}
 \label{fig:hr_pf_epica}%

### 2016-01-19 cordillera-cycle.tex
Rename figure lr_ts_deglac to hr_ts_deglac.

@@
       forcing (Table~\ref{tab:extrema}; Figs.~\ref{fig:lr_ts}
-      and~\ref{fig:lr_ts_deglac}). Adopting the EPICA climate forcing yields
+      and~\ref{fig:hr_ts_deglac}). Adopting the EPICA climate forcing yields
       peak ice volume at 17.2\,\unit{ka} and an uninterrupted deglaciation
-      until about 9\,\unit{ka} (Fig.~\ref{fig:lr_ts_deglac}, lower panel,
+      until about 9\,\unit{ka} (Fig.~\ref{fig:hr_ts_deglac}, lower panel,
       red curves). On the contrary, the simulation driven by the GRIP
@@
       14.5\,\unit{ka}, and the second between 13.1 and 11.6\,\unit{ka}
-      (Fig.~\ref{fig:lr_ts_deglac}, lower panel, blue curve).
+      (Fig.~\ref{fig:hr_ts_deglac}, lower panel, blue curve).
 
@@
 \begin{figure}%
-\includegraphics{lr_ts_deglac}
+\includegraphics{hr_ts_deglac}
 \caption{%
@@
       (bottom panel).}
-\label{fig:lr_ts_deglac}%
+\label{fig:hr_ts_deglac}%
 \end{figure}%

### 2016-01-19 figures/hr_ts_deglac.py

@@ -53,4 +53,4 @@ ax2.set_xlabel('model age (ka)')
 ax2.legend(loc='upper right', bbox_to_anchor=(1, 1+2.5/32.5))
 ax1.grid(axis='y')
 ax2.grid(axis='y')
-fig.savefig('lr_ts_deglac')
+fig.savefig('hr_ts_deglac')

### 2016-01-19 cordillera-cycle.tex
Update record and extrema tables to 0.7.2 runs.
Also remove changebar in latexdiff which now take ages to compute.

@@
     \middlehline
-    GRIP     &  57.45 &  42.93 &  19.14
-             &   1.87 &   0.69 &   2.07
-             &   7.35 &   1.69 &   8.71 \\
-    NGRIP    &  60.25 &  45.86 &  22.85
-             &   2.11 &   0.74 &   2.10
-             &   8.84 &   1.82 &   8.76 \\
-    EPICA    &  61.89 &  45.30 &  17.23
-             &   1.51 &   1.01 &   2.09
-             &   5.11 &   2.74 &   8.91 \\
-    Vostok   &  60.86 &  41.25 &  16.84
-             &   1.51 &   1.03 &   2.01
-             &   5.15 &   2.88 &   8.40 \\
-    ODP 1012 &  56.46 &  47.39 &  23.19
-             &   1.38 &   0.90 &   2.07
-             &   4.39 &   2.35 &   8.62 \\
-    ODP 1020 &  60.23 &  52.97 &  20.56
-             &   1.25 &   0.72 &   2.09
-             &   3.84 &   1.75 &   8.81 \\
+    GRIP     &  57.59 &  42.91 &  19.14
+             &   1.93 &   0.67 &   2.09
+             &   7.43 &   1.54 &   8.62 \\
+    NGRIP    &  60.27 &  45.87 &  22.85
+             &   2.13 &   0.73 &   2.11
+             &   8.71 &   1.70 &   8.60 \\
+    EPICA    &  61.90 &  52.40 &  17.36
+             &   1.48 &   0.98 &   2.08
+             &   4.84 &   2.55 &   8.56 \\
+    Vostok   &  62.17 &  55.95 &  16.83
+             &   1.52 &   1.04 &   2.17
+             &   5.04 &   2.94 &   9.07 \\
+    ODP 1012 &  56.91 &  47.69 &  23.10
+             &   1.34 &   0.87 &   2.06
+             &   4.10 &   2.20 &   8.40 \\
+    ODP 1020 &  60.72 &  52.88 &  20.99
+             &   1.21 &   0.69 &   2.06
+             &   3.54 &   1.54 &   8.39 \\
     \cline{1-10}
-    Minimum  &  61.89 &  52.97 &  23.19
-             &   1.25 &   0.69 &   2.01
-             &   3.84 &   1.69 &   8.40 \\
-    Maximum  &  56.46 &  41.25 &  16.84
-             &   2.11 &   1.03 &   2.10
-             &   8.84 &   2.88 &   8.91 \\
+    Minimum  &  56.91 &  42.91 &  16.83
+             &   1.21 &   0.67 &   2.06
+             &   3.54 &   1.54 &   8.39 \\
+    Maximum  &  62.17 &  55.95 &  23.10
+             &   2.13 &   1.04 &   2.17
+             &   8.71 &   2.94 &   9.07 \\
     \bottomhline

### 2016-01-20 cordillera-cycle.tex
Actually update record table and change notation.
I had forgotten to do it in the last commit. Use bracket notation for mean temperature anomaly from 32 to 22 ka.

@@
       to the scaling factor adopted to yield last glacial maximum ice limits
-      in the vicinity of mapped end moraines, and $T_{[32, 22]}$ refers to
-      the resulting mean temperature anomaly during the period $-$32
-      to~$-$22\,\unit{ka} after scaling.}
+      in the vicinity of mapped end moraines, and
+      $[{\Delta}T_{\text{TS}}]_{32}^{22}$ refers to the resulting mean
+      temperature anomaly during the period 32 to~22\,\unit{ka} after
+      scaling.}
 \label{tab:records}
@@
     \tophline
-    Record & Latitude & Longitude & Elev. & Proxy & $f$ & $T_{[32, 22]}$
+    Record & Latitude & Longitude & Elev.
+           & Proxy & $f$ & $[{\Delta}T_{\text{TS}}]_{32}^{22}$
            & Reference\\
@@
              & \chem{\delta^{18}O}
-             & 0.37 & $-$5.8  % -16.4126 (before scaling)
+             & 0.38 & $-$6.2  % -16.4126 (before scaling)
              & \citet{Dansgaard.etal.1993} \\
@@
              & \chem{\delta^{18}O}
-             & 0.24 & $-$6.5  % -26.7098
+             & 0.25 & $-$6.6  % -26.7098
              & \citet{Andersen.etal.2004} \\
@@
              & \chem{\delta^{18}O}
-             & 0.74 & $-$5.9  % -7.9550
+             & 0.75 & $-$6.0  % -7.9550
              & \citet{Petit.etal.1999} \\
@@
              & \chem{U^{K'}_{37}}
-             & 1.20 & $-$6.0 % -5.0000
+             & 1.18 & $-$5.9  % -5.0000
              & \citet{Herbert.etal.2001} \\

### 2016-01-20 cordillera-cycle.tex
Update numbers in the text to 0.7.2 runs.

@@
 major glaciations during marine oxygen isotope stages 4
-(61.9--56.5\,\unit{ka}) and 2 (23.2--16.8\,\unit{ka}). The timing of
+(62.2--56.9\,\unit{ka}) and 2 (23.1--16.8\,\unit{ka}). The timing of
 glaciation is better reproduced using temperature reconstructions
@@
 Maximum and hosts the last remains of Cordilleran ice until the
-middle Holocene (6.6--6.2\,\unit{ka}).
+middle Holocene (6.7\,\unit{ka}).
 \end{abstract}
@@
       most of the glacial cycle, except during two major glacial events
-      which occur between 61.9 and 56.5\,\unit{ka} during MIS~4, and between
-      23.2 and 16.8\,\unit{ka} during MIS~2 (Fig.~\ref{fig:lr_ts}, lower
-      panel). An ice volume minimum is consistently reached between 53.0 and
-      41.3\,\unit{ka} during MIS~3. However, the magnitude and precise
+      which occur between 62.2 and 56.9\,\unit{ka} during MIS~4, and between
+      23.1 and 16.8\,\unit{ka} during MIS~2 (Fig.~\ref{fig:lr_ts}, lower
+      panel). An ice volume minimum is consistently reached between 56.0 and
+      42.9\,\unit{ka} during MIS~3. However, the magnitude and precise
       timing of these three events depend significantly on the choice of
@@
       deglaciation of the modelled area prior to 17\,\unit{ka}. The ODP~1020
-      simulation predicts an early maximum in ice volume at 23.2\,\unit{ka},
+      simulation predicts an early maximum in ice volume at 21.0\,\unit{ka},
       followed by slower deglaciation than modelled using the other
@@
       volumes also differ significantly between model runs, and vary between
-      3.84 and~8.84\,\unit{m} sea level equivalents (m\,s.l.e.;
+      3.54 and~8.71\,\unit{m} sea level equivalents (m\,s.l.e.;
       Table~\ref{tab:extrema}).
@@
       palaeo-temperature record. Modelled ice volume minima spread over
-      a~wide range between 1.69 and 2.88\,\unit{m}\,s.l.e.
+      a~wide range between 1.54 and 2.94\,\unit{m}\,s.l.e.
       (Table~\ref{tab:extrema}).
@@
       (Table~\ref{tab:records}), it is interesting to note that modelled
-      maximum ice volumes also fall within a~tight range of 8.40 to
-      8.91\,\unit{m}\,s.l.e. (Table~\ref{tab:extrema}).
+      maximum ice volumes also fall within a~tight range of 8.39 to
+      9.07\,\unit{m}\,s.l.e. (Table~\ref{tab:extrema}).
 
@@
       during the last glacial cycle. The first maximum configuration is
-      obtained during MIS~4 (61.9--56.5\,\unit{ka}) and the second during
-      MIS~2 (23.2--16.8\,\unit{ka}; Figs.~\ref{fig:lr_ts},
+      obtained during MIS~4 (62.2--56.9\,\unit{ka}) and the second during
+      MIS~2 (23.1--16.8\,\unit{ka}; Figs.~\ref{fig:lr_ts},
       \ref{fig:lr_maps}; Table~\ref{tab:extrema}). These events broadly
@@
       During MIS~2, the modelled total ice volume peaks at
-      8.01\,\unit{m}\,s.l.e. (19.1\,\unit{ka}) in the GRIP simulation and at
-      8.77\,\unit{m}\,s.l.e. (17.2\,\unit{ka}) in the EPICA simulation.
+      8.62\,\unit{m}\,s.l.e. (19.1\,\unit{ka}) in the GRIP simulation and at
+      8.56\,\unit{m}\,s.l.e. (17.4\,\unit{ka}) in the EPICA simulation.
 
@@
       During MIS~4, the modelled total ice volume peaks at
-      7.19\,\unit{m}\,s.l.e. (57.3\,\unit{ka}) in the GRIP simulation and at
-      5.04\,\unit{m}\,s.l.e. (61.9\,\unit{ka}) in the EPICA simulation,
-      corresponding to respectively 90 and 57\,\unit{\%} of modelled MIS~2
+      7.43\,\unit{m}\,s.l.e. (57.6\,\unit{ka}) in the GRIP simulation and at
+      4.84\,\unit{m}\,s.l.e. (61.9\,\unit{ka}) in the EPICA simulation,
+      corresponding to respectively 86 and 57\,\unit{\%} of modelled MIS~2
       ice volumes.
@@
       northeast to the Rocky Mountains in the southwest, is only in
-      operation for at most 32\,\unit{ka}, which is about a~third of the
+      operation for at most 34\,\unit{ka}, which is about a~third of the
       timespan of the last glacial cycle (Fig.~\ref{fig:hr_geom_duration},
@@
       The modelled total ice volume corresponding to these persistent
-      ice-dispersal centres attains a~minimum of 1.52\,\unit{m}\,s.l.e.
-      (42.9\,\unit{ka}) in the GRIP simulation and of 2.68\,\unit{m}\,s.l.e.
-      (51.8\,\unit{ka}) in the EPICA simulation, corresponding to
-      respectively 19 and 31\,\unit{\%} of the MIS~2 ice volumes.
+      ice-dispersal centres attains a~minimum of 1.54\,\unit{m}\,s.l.e.
+      (42.9\,\unit{ka}) in the GRIP simulation and of 2.55\,\unit{m}\,s.l.e.
+      (52.4\,\unit{ka}) in the EPICA simulation, corresponding to
+      respectively 18 and 30\,\unit{\%} of the MIS~2 ice volumes.
 
@@
       and~\ref{fig:hr_ts_deglac}). Adopting the EPICA climate forcing yields
-      peak ice volume at 17.2\,\unit{ka} and an uninterrupted deglaciation
+      peak ice volume at 17.4\,\unit{ka} and an uninterrupted deglaciation
       until about 9\,\unit{ka} (Fig.~\ref{fig:hr_ts_deglac}, lower panel,
       red curves). On the contrary, the simulation driven by the GRIP
-      palaeo-temperature record yields peak ice volume at 19.3\,\unit{ka}
+      palaeo-temperature record yields peak ice volume at 19.1\,\unit{ka}
       and a~deglaciation interrupted by two phases of regrowth until about
@@
       ice sheet finally disappear from the Skeena Mountains at
-      6.5\,\unit{ka} (GRIP) and 6.1\,\unit{ka} (EPICA).
+      6.7\,\unit{ka} in both simulations.
 
@@
       glacial cycle presented in this study consistently produce two glacial
-      maxima during MIS~4 (61.9--56.5\,\unit{ka},
-      3.8--8.8\,\unit{m}\,s.l.e.) and MIS~2 (23.2--16.8\,\unit{ka},
-      8.40--8.91\,\unit{m}\,s.l.e.), two periods corresponding to documented
+      maxima during MIS~4 (62.2--56.9\,\unit{ka},
+      3.5--8.7\,\unit{m}\,s.l.e.) and MIS~2 (23.1--16.8\,\unit{ka},
+      8.4--9.1\,\unit{m}\,s.l.e.), two periods corresponding to documented
       extensive glaciations. This result is independent of the
@@
       finally, the Skeena mountains, which hosts the last remnant of the ice
-      sheet during the middle Holocene (6.6--6.2\,\unit{ka}). Our results
+      sheet during the middle Holocene (6.7\,\unit{ka}). Our results
       identify the Skeena Mountains as a~key area to understanding glacial

### 2016-01-21 cordillera-cycle.tex
Rev. Jarosch: implement new outline.

@@
       unsaturation index records from Northwest Pacific ocean sediment cores
-      \citep{Herbert.etal.2001}. We then proceed to compare the model output
+      \citep{Herbert.etal.2001}. After testing the model sensitivity to
+      these climate forcing time-series and to some of the most influential
+      ice flow parameters, we then proceed to compare the model output
       to geological evidence and discuss the timing and extent of glaciation
@@
       is applied as a~constant value of 70\,\unit{mW\,m^{-2}} at
-      3\,\unit{km} depth (Sect.~\ref{sec:icedyn}). Surface mass balance is
+      3\,\unit{km} depth (Sect.~\ref{sec:icedyn}). Basal sliding follows
+      a pseudo-plastic law where the yield stress accounts for till
+      deconsolidation under high water pressure (Sect.~\ref{sec:sliding}).
+      Ice shelf calving is calculated based on ice thickness and principle
+      strain rates (Sect.~\ref{sec:sliding}). Surface mass balance is
       computed using a~positive degree-day (PDD) model
@@
 
-\subsection{Ice thermodynamics}
+\subsection{Ice rheology}
 \label{sec:icedyn}
@@
       of 50\,\unit{m}.
+
+\subsection{Basal sliding}
+\label{sec:sliding}
 
@@
 
+\subsection{Ice shelf calving}
+\label{sec:calving}
+
       Ice shelf calving is computed using a~double criterion. First,
@@
       9.07\,\unit{m}\,s.l.e. (Table~\ref{tab:extrema}).
+
+\section{Sensitivity to ice flow parameters}
+\label{sec:sens}
 
@@
       Selwyn and Mackenzie mountains ice dispersal centre
-      (Fig.~\ref{fig:hr_geom_duration}, SMKM), prior to the Late Pleistocene
       \citep[cf.][]{Ward.etal.2008, Demuro.etal.2012}.

### 2016-01-21 cordillera-cycle.tex
Rev. Jarosch: add references.

@@
 
+\bibitem[{Clarke et~al.(2013)Clarke, Anslow, Jarosch, Radi{\'{c}}, Menounos, Bolch, and Berthier}]{Clarke.etal.2013}
+Clarke, G.
+K.~C., Anslow, F.~S., Jarosch, A.~H., Radi{\'{c}}, V., Menounos, B., Bolch, T., and Berthier, E.: Ice Volume and Subglacial Topography for Western Canadian Glaciers from Mass Balance Fields, Thinning Rates, and a Bed Stress Model, J.
+Climate, 26, 4282--4303,
+doi:\href{http://dx.doi.org/10.1175/jcli-d-12-00513.1}{10.1175/jcli-d-12-00513.1}, 2013.
+
 \bibitem[{Clason et~al.(2014)Clason, Applegate, and Holmlund}]{Clason.etal.2014}
@@
 doi:\href{http://dx.doi.org/10.1016/j.quascirev.2008.01.013}{10.1016/j.quascirev.2008.01.013}, 2008.
+
+\bibitem[{Cuffey and Paterson(2010)}]{Cuffey.Paterson.2010}
+Cuffey, K.~M. and Paterson, W. S.~B.: The physics of glaciers, Elsevier, Amsterdam, 2010.
 
@@
 
+\bibitem[{Glen(1952)}]{Glen.1952}
+Glen, J.: Experiments on the deformation of ice, J. Glaciol., 2, 111--114, 1952.
+
 \bibitem[{Golledge et~al.(2012)Golledge, Mackintosh, Anderson, Buckley, Doughty, Barrell, Denton, Vandergoes, Andersen, and Schaefer}]{Golledge.etal.2012}
@@
 
+\bibitem[{Greve(1997)}]{Greve.1997}
+Greve, R.: A continuum-mechanical formulation for shallow polythermal ice sheets, Philos. T. R. Soc. A, 355, 921--974,
+doi:\href{http://dx.doi.org/10.1098/rsta.1997.0050}{10.1098/rsta.1997.0050}, 1997.
+
 \bibitem[{Herbert et~al.(2001)Herbert, Schuffert, Andreasen, Heusser, Lyle, Mix, Ravelo, Stott, and Herguera}]{Herbert.etal.2001}
@@
 doi:\href{http://dx.doi.org/10.1126/science.1059209}{10.1126/science.1059209}, data archived at the World Data Center for Paleoclimatology, Boulder, Colorado, USA, 2001.
+
+\bibitem[{Hijmans et~al.(2005)Hijmans, Cameron, Parra, Jones, and Jarvis}]{Hijmans.etal.2005}
+Hijmans, R.~J., Cameron, S.~E., Parra, J.~L., Jones, P.~G., and Jarvis, A.: Very high resolution interpolated climate surfaces for global land areas, Int. J. Climatol., 25, 1965--1978,
+doi:\href{http://dx.doi.org/10.1002/joc.1276}{10.1002/joc.1276}, 2005.
 
@@
 
+\bibitem[{Jarosch et~al.(2012)Jarosch, Anslow, and Clarke}]{Jarosch.etal.2012}
+Jarosch, A.~H., Anslow, F.~S., and Clarke, G. K.~C.: High-resolution precipitation and temperature downscaling for glacier models, Clim. Dynam., 38, 391--409,
+doi:\href{http://dx.doi.org/10.1007/s00382-010-0949-1}{10.1007/s00382-010-0949-1}, 2012.
+
 \bibitem[{Johnsen et~al.(1995)Johnsen, Dahl-Jensen, Dansgaard, and Gundestrup}]{Johnsen.etal.1995}
@@
 
+\bibitem[{Lliboutry and Duval(1985)}]{Lliboutry.Duval.1985}
+Lliboutry, L.~A. and Duval, P.: Various isotropic and anisotropic ices found in glaciers and polar ice caps and their corresponding rheologies, Ann. Geophys., 3, 207--224, 1985.
+
 \bibitem[{Lundqvist(1987)}]{Lundqvist.1987}
@@
 doi:\href{http://dx.doi.org/10.2307/521191}{10.2307/521191}, 1987.
+
+\bibitem[{L{\"u}thi et~al.(2002)L{\"u}thi, Funk, Iken, Gogineni, and Truffer}]{Luthi.etal.2002}
+L{\"u}thi, M., Funk, M., Iken, A., Gogineni, S., and Truffer, M.: Mechanisms of fast flow in {J}akobshavns {I}sbr{\ae}, {G}reenland; {P}art {III}: measurements of ice deformation, temperature and cross-borehole conductivity in boreholes to the bedrock, J.
+Glaciol., 48, 369--385,
+doi:\href{http://dx.doi.org/10.3189/172756502781831322}{10.3189/172756502781831322}, 2002.
 
@@
 
+\bibitem[{Nye(1953)}]{Nye.1953}
+Nye, J.~F.: The Flow Law of Ice from Measurements in Glacier Tunnels, Laboratory Experiments and the Jungfraufirn Borehole Experiment, Proc. R. Soc. London, Ser. A, 219, 477--489, 1953.
+
 \bibitem[{Osborn and Gerloff(1997)}]{Osborn.Gerloff.1997}
@@
 doi:\href{http://dx.doi.org/10.1016/s1040-6182(96)00026-2}{10.1016/s1040-6182(96)00026-2}, 1997.
+
+\bibitem[{Paterson and Budd(1982)}]{Paterson.Budd.1982}
+Paterson, W. S.~B. and Budd, W.~F.: Flow parameters for ice sheet modeling, Cold Reg. Sci. Technol., 6, 175--177, 1982.
 
@@
 doi:\href{http://dx.doi.org/10.4095/108979}{10.4095/108979}, 1968.
+
+\bibitem[{Rignot et~al.(2013)Rignot, Mouginot, Larsen, Gim, and Kirchner}]{Rignot.etal.2013}
+Rignot, E., Mouginot, J., Larsen, C.~F., Gim, Y., and Kirchner, D.: Low-frequency radar sounding of temperate ice masses in Southern Alaska, Geophys. Res. Lett., 40, 5399--5405,
+doi:\href{http://dx.doi.org/10.1002/2013gl057452}{10.1002/2013gl057452}, 2013.
 

### 2016-01-21 cordillera-cycle.tex
Rev. Jarosch: add new figures and tables.

@@
 
+\clearpage{}  % new page for tables
+
 \begin{table*}
@@
 \end{table*}
+
+\begin{table*}
+\caption{%
+      Parameter values used in the sensitivity test.}
+\label{tab:sens_params}
+%\centering\makebox[\textwidth]
+{\begin{tabular}{l|ccccc|cc|cc}
+    \tophline
+            & \multicolumn{5}{c|}{Rheology}
+            & \multicolumn{2}{c|}{Sliding}
+            & \multicolumn{2}{c}{GRIP scaling} \\
+    Config. & $A_c$ & $A_w$ & $Q_c$ & $Q_w$ & $E_{SIA}$
+            & $\delta$ & $W_{max}$ & $f$ & $T_{[32, 22]}$ \\
+            & \multicolumn{2}{c}{(\unit{Pa^{-3}\,s^{-1}})}
+            & \multicolumn{2}{c}{(\unit{J\,mol^{-1}})}
+            & & & (m) \\
+    \middlehline
+    Default$^1$  & $ 3.61\times 10^{-13}$
+                 & $ 1.73\times 10^3$
+                 & $   60\times 10^3$
+                 & $  139\times 10^3$
+                 & 1 & 0.02 & 2 & 0.38 & 6.2 \\
+    \cline{1-10}
+    Soft ice$^2$ & $2.847\times 10^{-13}$
+                 & $2.356\times 10^{-2}$
+                 & $   60\times 10^3$
+                 & $  115\times 10^3$
+                 & 5 & 0.02 & 2 & 0.40 & 6.6 \\
+    Hard ice$^2$ & $2.847\times 10^{-13}$
+                 & $2.356\times 10^{-2}$
+                 & $   60\times 10^3$
+                 & $  115\times 10^3$
+                 & 1 & 0.02 & 2 & 0.37 & 6.0 \\
+    \cline{1-10}
+    Soft bed     & $ 3.61\times 10^{-13}$
+                 & $ 1.73\times 10^3$
+                 & $   60\times 10^3$
+                 & $  139\times 10^3$
+                 & 1 & 0.01 & 1 & 0.40 & 6.5 \\
+    Hard bed     & $ 3.61\times 10^{-13}$
+                 & $ 1.73\times 10^3$
+                 & $   60\times 10^3$
+                 & $  139\times 10^3$
+                 & 1 & 0.05 & 5 & 0.36 & 5.9 \\
+    \bottomhline
+\end{tabular}}
+\belowtable{%
+      After $^1$\citet{Paterson.Budd.1982,Bueler.Pelt.2015};
+      and $^2$\citet{Cuffey.Paterson.2010}.}
+\end{table*}
+
+\clearpage{}  % avoid three tables on one page
 
@@
 
+\begin{table*}[p]
+\caption{%
+      Extremes in Cordilleran ice sheet volume and extent corresponding to
+      MIS~4, 3 and 2 using the GRIP paleo-climate forcing with each
+      parameter configuration (Fig.~3). Relative differences (R. diff.) give
+      rough error estimates related to varying selected ice rheology and
+      basal sliding parameters (Table~\ref{tab:sens_params}).}
+\label{tab:sens_extrema}
+%\centering\makebox[\textwidth]
+{\begin{tabular}{l*{3}{|ccc}}
+    \tophline
+             & \multicolumn{3}{c}{Age (ka)}
+             & \multicolumn{3}{c}{Ice extent (\unit{10^6\,km^2})}
+             & \multicolumn{3}{c}{Ice volume (m~s.l.e.)} \\
+    Config.  &  MIS~4 &  MIS~3 &  MIS~2
+             &  MIS~4 &  MIS~3 &  MIS~2
+             &  MIS~4 &  MIS~3 &  MIS~2 \\
+    \middlehline
+    Default  &  57.59 &  42.91 &  19.14
+             &   1.93 &   0.67 &   2.09
+             &   7.43 &   1.54 &   8.62 \\
+    \cline{1-10}
+    Soft ice &  58.89 &  49.97 &  21.57
+             &   1.96 &   0.54 &   2.08
+             &   6.58 &   1.03 &   6.88 \\
+    Hard ice &  57.32 &  42.90 &  19.14
+             &   1.90 &   0.75 &   2.12
+             &   7.83 &   1.91 &   9.46 \\
+    R. diff. &    3\% &   16\% &   13\%
+             &    3\% &   31\% &    2\%
+             &   17\% &   57\% &   30\% \\
+    \cline{1-10}
+    Soft bed &  58.90 &  49.21 &  19.53
+             &   1.88 &   0.55 &   2.05
+             &   6.46 &   1.03 &   7.52 \\
+    Hard bed &  57.31 &  42.91 &  19.14
+             &   1.93 &   0.96 &   2.13
+             &   7.99 &   2.89 &   9.31 \\
+    R. diff. &    3\% &   15\% &    2\%
+             &    3\% &   62\% &    4\%
+             &   21\% &  120\% &   21\% \\
+    \bottomhline
+\end{tabular}}
+\end{table*}
+
+\clearpage{}  % new page for figures and avoid too many unprocessed floats
+
 \begin{figure}%
@@
 \begin{figure}%
+\includegraphics{sens_plot_rheo}
+\caption{%
+      Ice softness parameter, $A$, as a function of pressure-adjusted
+      temperature, $T_{pa}$, for the default \citep{Paterson.Budd.1982},
+      hard ice \citep[with $E_{SIA}=1$]{Cuffey.Paterson.2010}, and soft ice
+      \citep[with $E_{SIA}=5$]{Cuffey.Paterson.2010} rheologies, using a
+      linear scale (top panel) and logarithmic scale (bottom panel). Figure
+      made using Eqn.~\ref{eqn:softness} with parameters from
+      Table~\ref{tab:sens_params}.}
+\label{fig:sens_plot_rheo}%
+\end{figure}%
+
+\begin{figure}%
+\includegraphics{sens_plot_ntil}
+\caption{%
+      Effective pressure, $N$, as a function of water content in the till,
+      $W$, for the default ($\delta=0.02$, $W_{max}=2$\,m), hard bed
+      ($\delta=0.05$, $W_{max}=5$\,m), and soft bed ($\delta=0.01$,
+      $W_{max}=1$\,m) sliding parametrisations, using a linear scale (top
+      panel) and a logarithmic scale (bottom panel). Calculations are made
+      for an ice thickness, $h$, of 1000\,m. Figure made using
+      Eqn.~\ref{eqn:ntil} with parameters from Table~\ref{tab:sens_params}
+      after \citet[Fig.~1]{Bueler.Pelt.2015}.}
+\label{fig:sens_plot_ntil}%
+\end{figure}%
+
+\begin{figure}%
 \includegraphics{atm}
@@
 
+\begin{figure*}
+\includegraphics{sens_ts}
+\caption{%
+      Modelled sea-level relevant ice volume through the last 120\,ka in the
+      simulation forced by the GRIP paleo-climate record, using default
+      parameters (black curves), different ice rheology parameters (top
+      panel), and different basal sliding parameters (bottom panel). Gray
+      fields indicate Marine Oxygen Isotope Stage (MIS) boundaries for MIS~2
+      and MIS~4 according to a global compilation of benthic
+      \chem{\delta^{18}O} records \citep{Lisiecki.Raymo.2005}.}
+\label{fig:sens_ts}
+\end{figure*}
+
 \begin{figure*}%

### 2016-01-21 cordillera-cycle.tex
Rev. Jarosch: detail ice rheology section.
Reworked slightly from the response. Format equations and units.

@@
       computed using a~positive degree-day (PDD) model
-      (Sect.~\ref{sec:surface}). Climate forcing is provided by a~monthly
+      (Sect.~\ref{sec:surface}). Default parameter values are given in
+      Table~\ref{tab:params}. Climate forcing is provided by a~monthly
       climatology averaged from 1979 to 2000 from the North American
@@
       by time-dependent offsets and lapse-rate temperature corrections
-      (Sect.~\ref{sec:atm}).
+      (Sect.~\ref{sec:atm}, Table~\ref{tab:records}).
 
@@
 
-      Ice rheology depends on temperature and water content through an
-      enthalpy formulation \citep{Aschwanden.etal.2012}. Surface air
+      Ice deformation is governed by the constitutive law for ice
+      \citep{Glen.1952, Nye.1953},
+%
+\begin{align}
+&\label{eqn:glenslaw}
+&\vec{\dot{\epsilon}} = A\,\tau_{\mathrm{e}}^{n-1}\,\vec{\tau} \,.
+\end{align}
+%
+      where $\vec{\dot{\epsilon}}$ is the the strain-rate tensor,
+      $\vec{\tau}$ the deviatoric stress tensor, and $\tau_{\mathrm{e}}$ the
+      equivalent stress defined by ${\tau_{\mathrm{e}}}^2 = \frac{1}{2}
+      \mathrm{tr}(\vec{\tau}^2)$. The ice softness coefficient, $A$, depends
+      on ice temperature, $T$, pressure, $p$, and water content, $\omega$,
+      through a piece-wise Arrhenius-type law
+      \citep[Eqs.~63--65]{Aschwanden.etal.2012},
+%
+\begin{align}
+&\label{eqn:softness}
+&A = E\cdot
+\begin{cases}
+A_{\mathrm{c}} \,e^\frac{-Q_{\mathrm{c}}}{RT_{\text{pa}}}            & \text{if}\ T_{\text{pa}}  <  T_{\mathrm{c}} \,, \\
+A_{\mathrm{w}} (1+f\omega)\,e^\frac{-Q_{\mathrm{w}}}{RT_{\text{pa}}} & \text{if}\ T_{\text{pa}} \ge T_{\mathrm{c}} \,,
+\end{cases}
+\end{align}
+%
+      where $T_{\text{pa}}$ is the pressure-adjusted ice temperature
+      calculated using the Clapeyron relation, ${T_{\text{pa}} = T - \beta
+      p}$. $R=8.31441$\,\unit{J\,mol^{-1}\,K^{-1}} is the ideal gas
+      constant, and $A_{\mathrm{c}}$, $A_{\mathrm{w}}$, $Q_{\mathrm{c}}$ and
+      $Q_{\mathrm{w}}$, are constant parameters corresponding to values
+      measured below and above a critical temperature threshold
+      $T_{\mathrm{c}}=-10$\,\unit{{\degree}C}
+      \citep[p.~72]{Paterson.Budd.1982,Cuffey.Paterson.2010}. The water
+      fraction, $\omega$, is capped at a maximum value of 0.01, above which
+      no measurements are available \citep[Eq.~5.7]{Lliboutry.Duval.1985,
+      Greve.1997}. Finally, $E$ is a non-dimensional enhancement factor
+      which can take different values, $E_{\text{SIA}}$, in the Shallow Ice
+      Approximation (SIA) and $E_{\text{SSA}}$, in the Shallow Shelf
+      Approximation (SSA).
+
+      In all our simulations, we set constant the power-law exponent, $n=3$,
+      according to \citet[p.~55--57]{Cuffey.Paterson.2010}, the Clapeyron
+      constant, $\beta=7.9\times 10^{-8}$\,\unit{K\,Pa^{-1}}, according to
+      \citet{Luthi.etal.2002}, the water fraction coefficient, $f=181.25$,
+      according to \citet{Lliboutry.Duval.1985}, and the SSA enhancement
+      factor, $E_{\text{SSA}}=1$, according to
+      \citet[p.~77]{Cuffey.Paterson.2010}. These fixed parameter values are
+      summarized in Table~\ref{tab:params}.
+
+      On the other hand, we test the model sensitivity
+      (Sect.~\ref{sec:sens}) to different values for the two creep
+      parameters, $A_{\mathrm{c}}$ and $A_{\mathrm{w}}$, the two activation
+      energies, $Q_{\mathrm{c}}$ and $Q_{\mathrm{w}}$, and the SIA
+      enhancement factor, $E_{\text{SIA}}$, as follow.
+%
+\begin{itemize}
+  \item Our \emph{default} configuration used in the control run and all
+    other simulations in the manuscript include rheological parameters,
+    $A_{\mathrm{c}}$, $A_{\mathrm{w}}$, $Q_{\mathrm{c}}$ and
+    $Q_{\mathrm{w}}$, derived from \citet{Paterson.Budd.1982} and given in
+    \citet[Eqn.~5]{Bueler.Brown.2009}, and $E_{\text{SIA}}=1$.
+  \item Our \emph{hard ice} configuration include rheological parameters,
+    $A_{\mathrm{c}}$, $A_{\mathrm{w}}$, $Q_{\mathrm{c}}$ and
+    $Q_{\mathrm{w}}$, derived from \citet[p.~72 and
+    76]{Cuffey.Paterson.2010}, and $E_{\text{SIA}}=1$, which correspond to a
+    stiffer rheology than that used in the control run.
+  \item Our \emph{soft ice} configuration include rheological parameters
+    from \citet{Cuffey.Paterson.2010}, and $E_{\text{SIA}}=5$, the
+    recommended value for ice age polar ice
+    \citep[p.~77]{Cuffey.Paterson.2010}.
+\end{itemize}
+%
+      An additional simulation using the ice rheology from
+      \citet{Cuffey.Paterson.2010} and $E_{\text{SIA}}=2$, the recommended
+      value for Holocene polar ice \citep[p.~77]{Cuffey.Paterson.2010} was
+      performed, but its results were very similar to that of our default
+      run and thus not presented here.
+
+      Actual parameter values for $A_{\mathrm{c}}$, $A_{\mathrm{w}}$,
+      $Q_{\mathrm{c}}$, $Q_{\mathrm{w}}$ and $E_{\text{SIA}}$ used in our
+      simulations are given in Table~\ref{tab:sens_params}, while the effect
+      of the three different parametrisations on temperature-dependent ice
+      softness, $A$, is illustrated in Fig.~\ref{fig:sens_plot_rheo}.
+
+      Surface air
       temperature derived from the climate forcing (Sect.~\ref{sec:atm})
@@
             & \multicolumn{2}{c}{GRIP scaling} \\
-    Config. & $A_c$ & $A_w$ & $Q_c$ & $Q_w$ & $E_{SIA}$
-            & $\delta$ & $W_{max}$ & $f$ & $T_{[32, 22]}$ \\
+    Config. & $A_{\mathrm{c}}$ & $A_{\mathrm{w}}$
+            & $Q_{\mathrm{c}}$ & $Q_{\mathrm{w}}$
+            & $E_{\text{SIA}}$ & $\delta$ & $W_{\text{max}}$
+            & $f$ & $T_{[32, 22]}$ \\
             & \multicolumn{2}{c}{(\unit{Pa^{-3}\,s^{-1}})}
             & \multicolumn{2}{c}{(\unit{J\,mol^{-1}})}
-            & & & (m) \\
+            & & & (\unit{m}) \\
     \middlehline
@@
       Ice softness parameter, $A$, as a function of pressure-adjusted
-      temperature, $T_{pa}$, for the default \citep{Paterson.Budd.1982},
-      hard ice \citep[with $E_{SIA}=1$]{Cuffey.Paterson.2010}, and soft ice
-      \citep[with $E_{SIA}=5$]{Cuffey.Paterson.2010} rheologies, using a
-      linear scale (top panel) and logarithmic scale (bottom panel). Figure
-      made using Eqn.~\ref{eqn:softness} with parameters from
+      temperature, $T_{\text{pa}}$, for the default
+      \citep{Paterson.Budd.1982}, hard ice \citep[with
+      $E_{\text{SIA}}=1$]{Cuffey.Paterson.2010}, and soft ice \citep[with
+      $E_{\text{SIA}}=5$]{Cuffey.Paterson.2010} rheologies, using a linear
+      scale (top panel) and logarithmic scale (bottom panel). Figure made
+      using Eqn.~\ref{eqn:softness} with parameters from
       Table~\ref{tab:sens_params}.}

### 2016-01-21 cordillera-cycle.tex
Rev. Jarosch: detail basal sliding section.
Similarily as in previous commit.

@@
 \begin{itemize}
-  \item Our \emph{default} configuration used in the control run and all
-    other simulations in the manuscript include rheological parameters,
+  \item Our \emph{default} configuration used in the control run of the
+    sensitivity study and all other simulations in the paper include
+    rheological parameters,
     $A_{\mathrm{c}}$, $A_{\mathrm{w}}$, $Q_{\mathrm{c}}$ and
@@
       A~pseudo-plastic sliding law,
+%
 \begin{align}
@@
 \end{align}
+%
       relates the bed-parallel shear stresses, $\vec{\tau}_{\mathrm{b}}$, to
-      the sliding velocity, $\vec{v}_{\mathrm{b}}$ (Table~\ref{tab:params}).
+      the sliding velocity, $\vec{v}_{\mathrm{b}}$.
       The yield stress, $\tau_{\mathrm{c}}$, is modelled using the
       Mohr--Coulomb criterion,
+%
 \begin{align}
@@
 \end{align}
-      where cohesion, $c_0$, is assumed to be zero. The friction angle,
-      $\phi$, varies from 15 to 45\unit{\degree} as a~piecewise-linear
-      function of modern bed elevation, with the lowest value occuring below
-      modern sea level (0\,\unit{m} above sea level, \unit{m\,a.s.l.}) and
-      the highest value occuring above the generalised elevation of the
-      highest shorelines
-      \citep[200\,\unit{m\,a.s.l.},][Fig.~5]{Clague.1981}, thus accounting
-      for a~weakening of till associated with the presence of marine
-      sediments (cf. \citealp{Martin.etal.2011};
-      \citealp[Supplement]{Aschwanden.etal.2013};
-      \citealp{PISM-authors.2015}). Effective pressure, $N$, is related to
-      the ice overburden stress, $\rho gh$, and the modelled amount of
+%
+      where cohesion, $c_0$, is assumed to be zero. Effective pressure, $N$,
+      is related to the ice overburden stress, $P_0=\rho gh$, and the
+      modelled amount of
       subglacial water, using a~formula derived from laboratory experiments
@@
       \citep{Tulaczyk.etal.2000, Bueler.Pelt.2015},
+%
 \begin{align}
-&N = \delta \rho gh \, 10^{(e_0/C_{\mathrm{c}}) (1 - (W/W_{\text{max}}))} \,,
+&\label{eqn:ntil}
+&N = \delta P_0 \, 10^{(e_0/C_{\mathrm{c}}) (1 - (W/W_{\text{max}}))} \,,
 \end{align}
-      where $\delta$ is chosen as 0.02, $e_0$ is a~measured reference void
+%
+      where $\delta$ sets the minimum ratio between the effective and
+      overburden pressures, $e_0$ is a~measured reference void
       ratio and $C_{\mathrm{c}}$ is a~measured compressibility coefficient
       (Table~\ref{tab:params}). The amount of water at the base, $W$, varies
-      from zero to $W_{\text{max}}=2$\,\unit{m}, a~threshold above which
+      from zero to $W_{\text{max}}$, a~threshold above which
       additional melt water is assumed to drain off instantaneously.
+
+      In all our sumlations, we set constant the pseudo-plastic sliding
+      exponent, $q=0.25$, and the threshold velocity,
+      $v_{\text{th}}=100$\,\unit{m\,a^{-1}}, according to values used by
+      \citet{Aschwanden.etal.2013}, the till cohesion, $c_0=0$, whose
+      measured values are consistently negligible
+      \citep[p.~268]{Tulaczyk.etal.2000, Cuffey.Paterson.2010}, the till
+      reference void ratio, $e_0=0.69$, and the till compressibility
+      coefficient, $C_{\mathrm{c}}=0.12$, according to the only measurements
+      available to our knowledge, published by \citep{Tulaczyk.etal.2000}.
+      These fixed parameter values are summarized in Table~\ref{tab:params}.
+
+      We also use a constant spatial distribution of the till friction
+      angle, $\phi$, whose values vary from 15 to 45\unit{\degree} as a
+      a~piecewise-linear function of modern bed elevation, with the lowest
+      value occuring below modern sea level (0\,\unit{m} above sea level,
+      \unit{m\,a.s.l.}) and the highest value occuring above the generalised
+      elevation of the highest shorelines
+      \citep[200\,\unit{m\,a.s.l.},][Fig.~5]{Clague.1981}. This range of
+      values span over the range of measured values for glacial till of 18
+      to 40\unit{\degree} \citep[p.~268]{Cuffey.Paterson.2010}. It accounts
+      for frictional basal conditions associated with discontinuous till
+      cover at high elevations, and for a~weakening of till associated with
+      the presence of marine sediments (cf. \citealp{Martin.etal.2011};
+      \citealp[Supplement]{Aschwanden.etal.2013};
+      \citealp{PISM-authors.2015}).
+
+      An additional simulation with a constant till friction angle,
+      $\phi=30$\unit{\degree}, corresponding to the average value in
+      \citet[p.~268]{Cuffey.Paterson.2010}, was actually performed, but the
+      induced variability was small and thus not presented here.
+
+      On the other hand, we test the model sensitivity
+      (Sect.~\ref{sec:sens}) to different values for the minimum ratio
+      between the effective and overburden pressures, $\delta$, and the
+      maximum water height in the till, $W_{\text{max}}$, as follow.
+%
+\begin{itemize}
+  \item Our \emph{default} configuration used in the control run of the
+    sensitivity study and all and all other simulations in the paper include
+    $\delta=0.02$ and $W_{\text{max}}=2$\,\unit{m} as in
+    \citet{Bueler.Pelt.2015}.
+  \item Our \emph{soft bed} configuration use $\delta=0.01$ and
+    $W_{\text{max}}=1$\,\unit{m}.
+  \item Our \emph{hard bed} configuration use $\delta=0.05$ and
+    $W_{\text{max}}=5$\,\unit{m}.
+\end{itemize}
+%
+      The effect of the three different parametrisations on the effective
+      pressure on the till, $N$, in response to water content, $W$, is
+      illustrated in Fig.~\ref{fig:sens_plot_ntil}. All parameter choices
+      are listed in Table~\ref{tab:sens_params}.
+
       Finally, the bedrock topography responds to ice load following
@@
       infinite half-space \citep{Lingle.Clark.1985,Bueler.etal.2007}.
-      A~relatively low viscosity value of $\nu_{m} =
+      A~relatively low viscosity value of $\nu_{\mathrm{m}} =
       1\times10^{19}$\,\unit{Pa\,s} is used for the astenosphere
@@
       while temperature reconstructions from Antarctic and oceanic cores
-      were provided as such. All records were scaled linearly
-      (Table~\ref{tab:records}) in order to simulate comparable ice extents
+      were provided as such. For each proxy record used and each of the
+      configurations used in the sensitivity tests, palaeo-temperature
+      anomalies were scaled linearly (Tables~\ref{tab:records}
+      and~\ref{tab:sens_params}) in order to simulate comparable ice extents
       at the LGM (Table~\ref{tab:extrema}) and realistic outlines
@@
       Effective pressure, $N$, as a function of water content in the till,
-      $W$, for the default ($\delta=0.02$, $W_{max}=2$\,m), hard bed
-      ($\delta=0.05$, $W_{max}=5$\,m), and soft bed ($\delta=0.01$,
-      $W_{max}=1$\,m) sliding parametrisations, using a linear scale (top
-      panel) and a logarithmic scale (bottom panel). Calculations are made
-      for an ice thickness, $h$, of 1000\,m. Figure made using
+      $W$, for the default ($\delta=0.02$, $W_{\text{max}}=2$\,m), hard bed
+      ($\delta=0.05$, $W_{\text{max}}=5$\,m), and soft bed ($\delta=0.01$,
+      $W_{\text{max}}=1$\,m) sliding parametrisations, using a linear scale
+      (top panel) and a logarithmic scale (bottom panel). Calculations are
+      made for an ice thickness, $h$, of 1000\,m. Figure made using
       Eqn.~\ref{eqn:ntil} with parameters from Table~\ref{tab:sens_params}

### 2016-01-21 cordillera-cycle.tex
Rev. Jarosch: add refs to clarify new outline.

@@
       \citep{Herbert.etal.2001}. After testing the model sensitivity to
-      these climate forcing time-series and to some of the most influential
-      ice flow parameters, we then proceed to compare the model output
+      these climate forcing time-series (Sect.~\ref{sec:results}) and to
+      some of the most influential ice flow parameters
+      (Sect.~\ref{sec:sens}), we then proceed to compare the model output
       to geological evidence and discuss the timing and extent of glaciation
@@
       applicability of different records to modelling the history of the
-      Cordilleran ice sheet.
+      Cordilleran ice sheet (Sect.~\ref{sec:discussion}).
 
@@
       by time-dependent offsets and lapse-rate temperature corrections
-      (Sect.~\ref{sec:atm}, Table~\ref{tab:records}).
+      (Sect.~\ref{sec:atm}, Table~\ref{tab:records}). A sensitivity study
+      to some of the most influential ice rheology (Sect~\ref{sec:icedyn})
+      and basal sliding (Sect.~\ref{sec:sliding}) parameters was performed
+      (Table~\ref{tab:sens_params}).
 

### 2016-01-21 cordillera-cycle.tex
Rev. Jarosch: add results of sensitivity tests.

@@
 \label{sec:sens}
+
+      Using the GRIP ice core palaeo-temperature record as climate forcing
+      time-series, the model shows a significant sensitivity to selected ice
+      rheologic and basal sliding parameters (Fig.~\ref{fig:sens_ts};
+      Table~\ref{tab:sens_extrema}) in terms of modelled ice extent, and
+      even more so in terms of modelled ice volumes.
+
+      As a direct result of the different scaling factors applied
+      (Table~\ref{tab:sens_params}), the resulting simulations show very
+      little difference in the modelled glaciated areas corresponding to
+      maximum ice volumes during MIS~2, but also during MIS~4
+      (Table~\ref{tab:sens_extrema}). However, such can not be said of the
+      modelled glaciated area corresponding to minimum ice volume during
+      MIS~3. In fact, the extent of the remnant ice cap which persists over
+      the Skeena Mountains during this stage shows a significant sensitivity
+      to ice rheology of 31\,\unit{\%}, and an even more important
+      sensitivity to basal sliding parameters of 62\,\unit{\%}
+      (Table~\ref{tab:sens_extrema}).
+
+      The modelled sea-level relevant ice volumes show more variability than
+      the modelled glaciated areas (Table~\ref{tab:sens_extrema},
+      Fig.~\ref{fig:sens_ts}). As one could expect, softer ice and weaker
+      till both result in a thinner ice sheet, while harder ice and stronger
+      till result in a thicker ice sheet. For instance, peak ice volume
+      during the MIS~2 (LGM) varies by 30\,\unit{\%} between the two
+      parametrisations of ice rheology used, and by 21\,\unit{\%} between
+      the two parametrisations of basal sliding used. The differences in
+      sea-level relevant ice volume are greatest during the MIS~3
+      (Table~\ref{tab:sens_extrema}, Fig.~\ref{fig:sens_ts}) where both the
+      areal and thickness contributions add up.
 

### 2016-01-21 figures/tab_sens.py
Enclose percents in unit commands in table.

@@ -32,13 +32,13 @@ footer=r'''    \bottomhline
 \end{document}
 '''
 
-# latex table line
+# latex table line (curly braces need to be doubled to escape)
 tabline = (' '*4 + '{title:8}' + ' &{:7.2f}'*3 + '\n'+
            ' '*12 + ' &{:7.2f}'*3 + '\n'+
            ' '*12 + ' &{:7.2f}'*3 + r' \\' + '\n')
-errline = (' '*4 + '{title:8}' + r' &{:5.0f}\%'*3 + '\n'+
-           ' '*12 + r' &{:5.0f}\%'*3 + '\n'+
-           ' '*12 + r' &{:5.0f}\%'*3 + r' \\' + '\n')
+errline = (' '*4 + '{title:8}' + r' &{:5.0f}\,\unit{{\%}}'*3 + '\n'+
+           ' '*12 + r' &{:5.0f}\,\unit{{\%}}'*3 + '\n'+
+           ' '*12 + r' &{:5.0f}\,\unit{{\%}}'*3 + r' \\' + '\n')
 
 # initialize array
 results = np.zeros((5, 9), dtype=float)

### 2016-01-21 cordillera-cycle.tex
Enclose percents in unit commands in table.
Rev. Jarosch: extend default parameter table.
Also make small changes in typeset to avoid extensive latexdiff markup and errors. One evening gone just to fix that...

@@
       \citep[SPECMAP,][]{Imbrie.etal.1989} time scale. Geothermal heat flux
-      is applied as a~constant value of 70\,\unit{mW\,m^{-2}} at
+      is applied as a~constant value of
+      $q_{\mathrm{G}}=70$\,\unit{mW\,m^{-2}} at
       3\,\unit{km} depth (Sect.~\ref{sec:icedyn}). Basal sliding follows
@@
 \caption{%
-      Parameter values used in the ice sheet model.}
+      Default parameter values used in the ice sheet model.}
 \label{tab:params}
-%\scalebox{.85}[.85]
-{\begin{tabular}{llrl}
+\scalebox{.85}[.85]
+{\begin{tabular}{llrll}
     \tophline
-    Not.    & Name & Value & Unit \\
+    Not.    & Name & Value & Unit & Source \\
     \middlehline
+    \multicolumn{2}{l}{{Ice rheology}} \\
+    \cline{1-5}
+
     $\rho$  & Ice density
             & 910
-            & \unit{kg\,m^{-3}} \\
+            & \unit{kg\,m^{-3}}
+            & \citet{Aschwanden.etal.2012} \\
 
@@
             & 9.81
-            & \unit{m\,s^{-2}} \\
-    \cline{1-4}
+            & \unit{m\,s^{-2}}
+            & \citet{Aschwanden.etal.2012} \\
+
+    $n$     & Glen exponent
+            & 3
+            & --
+            & \citet{Cuffey.Paterson.2010} \\
+
+    $A_{\mathrm{c}}$   & Ice hardness coefficient cold$^1$
+            & $3.61\times10^{-13}$
+            & \unit{Pa^{-3}\,s^{-1}}
+            & \citet{Paterson.Budd.1982} \\
+
+    $A_{\mathrm{w}}$   & Ice hardness coefficient warm$^1$
+            & $1.73\times10^3$
+            & \unit{Pa^{-3}\,s^{-1}}
+            & \citet{Paterson.Budd.1982} \\
+
+    $Q_{\mathrm{c}}$   & Flow law activation energy cold$^1$
+            & $6.0\times10^4$
+            & \unit{J\,mol^{-1}}
+            & \citet{Paterson.Budd.1982} \\
+
+    $Q_{\mathrm{w}}$   & Flow law activation energy warm$^1$
+            & $13.9\times10^4$
+            & \unit{J\,mol^{-1}}
+            & \citet{Paterson.Budd.1982} \\
+
+    $T_{\mathrm{c}}$   & Flow law critical temperature
+            & 263.15
+            & \unit{K}
+            & \citet{Paterson.Budd.1982} \\
+
+    $f$     & Flow law water fraction coeff.
+            & 181.25
+            & --
+            & \citet{Lliboutry.Duval.1985} \\
+
+    $R$     & Ideal gas constant
+            & 8.31441
+            & \unit{J\,mol^{-1}\,K^{-1}}
+            & -- \\
+
+    $\beta$ & Clapeyron constant
+            & $7.9\times10^{-8}$
+            & \unit{K\,Pa^{-1}}
+            & \citet{Luthi.etal.2002} \\
+
+    $c_{\mathrm{i}}$   & Ice specific heat capacity
+            & 2009
+            & \unit{J\,kg^{-1}\,K^{-1}}
+            & \citet{Aschwanden.etal.2012} \\
+
+    $c_{\mathrm{w}}$   & Water specific heat capacity
+            & 4170
+            & \unit{J\,kg^{-1}\,K^{-1}}
+            & \citet{Aschwanden.etal.2012} \\
+
+    $k$     & Ice thermal conductivity
+            & 2.10
+            & \unit{J\,m^{-1}\,K^{-1}\,s^{-1}}
+            & \citet{Aschwanden.etal.2012} \\
+
+    $L$     & Water latent heat of fusion
+            & $3.34\times10^5$
+            & \unit{J\,kg^{-1}\,K^{-1}}
+            & \citet{Aschwanden.etal.2012} \\
+
+    \cline{1-5}
     \multicolumn{2}{l}{{Basal sliding}} \\
-    \cline{1-4}
+    \cline{1-5}
+
     $q$     & Pseudo-plastic sliding exponent
             & 0.25
-            & -- \\
+            & --
+            & \citet{Aschwanden.etal.2013} \\
 
@@
             & 100.0
-            & \unit{m\,yr^{-1}} \\
+            & \unit{m\,yr^{-1}}
+            & \citet{Aschwanden.etal.2013} \\
 
@@
             & 0.0
-            & Pa \\
-
-    $\delta$& Effective pressure coefficient
-            & 0.02
-            & -- \\
-
+            & Pa
+            & \citet{Tulaczyk.etal.2000} \\
     $e_0$   & Till reference void ratio
             & 0.69
-            & -- \\
+            & --
+            & \citet{Tulaczyk.etal.2000} \\
 
@@
             & 0.12
+            & --
+            & \citet{Tulaczyk.etal.2000} \\
+
+    $\delta$& Minimum effective pressure ratio$^1$
+            & 0.02
+            & --
+            & \citet{Bueler.Pelt.2015} \\
+
+    $W_{\text{max}}$ & Maximal till water thickness$^1$
+            & 2.0
+            & m
+            & \citet{Bueler.Pelt.2015} \\
+
+    $b_0$   & Altitude of max. friction angle
+            & 0
+            & m
             & -- \\
 
-    $W_{\text{max}}$ & Maximal till water thickness
-            & 2.0
-            & m \\
-    \cline{1-4}
+    $b_1$   & Altitude of min. friction angle
+            & 200
+            & m
+            & \citet{Clague.1981} \\
+
+    $\phi_0$& Minimum friction angle
+            & 15
+            & \degree
+            & -- \\
+
+    $\phi_1$& Maximum friction angle
+            & 45
+            & \degree
+            & -- \\
+
+    \cline{1-5}
     \multicolumn{2}{l}{{Bedrock and lithosphere}} \\
-    \cline{1-4}
+    \cline{1-5}
+
+    $q_{\mathrm{G}}$   & Geothermal heat flux
+            & 70.0
+            & \unit{mW\,m^{-2}}
+            & -- \\
+
     $\rho_{\mathrm{b}}$& Bedrock density
             & 3300
-            & \unit{kg\,m^{-3}} \\
+            & \unit{kg\,m^{-3}}
+            & -- \\
 
@@
             & 1000
-            & \unit{J\,kg^{-1}\,K^{-1}} \\
+            & \unit{J\,kg^{-1}\,K^{-1}}
+            & -- \\
 
@@
             & 3.0
-            & \unit{J\,m^{-1}\,K^{-1}\,s^{-1}} \\
-
-    $\nu_m$ & Astenosphere viscosity
+            & \unit{J\,m^{-1}\,K^{-1}\,s^{-1}}
+            & -- \\
+
+    $\nu_{\mathrm{m}}$ & Astenosphere viscosity
             & $1\times10^{19}$
-            & \unit{Pa\,s} \\
-
-    $\rho_l$& Lithosphere density
+            & \unit{Pa\,s}
+            & \citet{James.etal.2009} \\
+
+    $\rho_{\mathrm{l}}$& Lithosphere density
             & 3300
-            & \unit{kg\,m^{-3}} \\
+            & \unit{kg\,m^{-3}}
+            & \citet{Lingle.Clark.1985} \\
 
@@
             & $5.0\times10^{24}$
-            & \unit{N} \\
-    \cline{1-4}
+            & \unit{N}
+            & \citet{Lingle.Clark.1985} \\
+
+    \cline{1-5}
     \multicolumn{2}{l}{{Surface and atmosphere}} \\
-    \cline{1-4}
+    \cline{1-5}
+
     $T_{\mathrm{s}}$   & Temperature of snow precipitation
             & 273.15
-            & \unit{K} \\
+            & \unit{K}
+            & -- \\
 
@@
             & 275.15
-            & \unit{K} \\
+            & \unit{K}
+            & -- \\
 
@@
             & $3.04\times10^{-3}$
-            & \unit{m\,K^{-1}\,day^{-1}} \\
+            & \unit{m\,K^{-1}\,day^{-1}}
+            & \citet{Shea.etal.2009} \\
 
@@
             & $4.59\times10^{-3}$
-            & \unit{m\,K^{-1}\,day^{-1}} \\
+            & \unit{m\,K^{-1}\,day^{-1}}
+            & \citet{Shea.etal.2009} \\
 
@@
             & $6\times10^{-3}$
-            & \unit{K\,m^{-1}} \\
+            & \unit{K\,m{-1}}
+            & -- \\
+
     \bottomhline
 \end{tabular}}
+\belowtable{%
+      $^1$Default value. Alternative values used in sensitivity tests are
+      given in Table~\ref{tab:sens_params}.}
 \end{table*}
@@
     \tophline
-             & \multicolumn{3}{c}{Age (ka)}
-             & \multicolumn{3}{c}{Ice extent (\unit{10^6\,km^2})}
-             & \multicolumn{3}{c}{Ice volume (m~s.l.e.)} \\
+             & \multicolumn{3}{c|}{Age (ka)}
+             & \multicolumn{3}{c|}{Ice extent (\unit{10^6\,km^2})}
+             & \multicolumn{3}{c}{Ice volume (m\,s.l.e.)} \\
     Config.  &  MIS~4 &  MIS~3 &  MIS~2
@@
              &   7.83 &   1.91 &   9.46 \\
-    R. diff. &    3\% &   16\% &   13\%
-             &    3\% &   31\% &    2\%
-             &   17\% &   57\% &   30\% \\
+    R. diff. &    3\,\unit{\%} &   16\,\unit{\%} &   13\,\unit{\%}
+             &    3\,\unit{\%} &   31\,\unit{\%} &    2\,\unit{\%}
+             &   17\,\unit{\%} &   57\,\unit{\%} &   30\,\unit{\%} \\
     \cline{1-10}
@@
              &   7.99 &   2.89 &   9.31 \\
-    R. diff. &    3\% &   15\% &    2\%
-             &    3\% &   62\% &    4\%
-             &   21\% &  120\% &   21\% \\
+    R. diff. &    3\,\unit{\%} &   15\,\unit{\%} &    2\,\unit{\%}
+             &    3\,\unit{\%} &   62\,\unit{\%} &    4\,\unit{\%}
+             &   21\,\unit{\%} &  120\,\unit{\%} &   21\,\unit{\%} \\
     \bottomhline

### 2016-01-21 cordillera-cycle.tex
Rev. Jarosch: add rel. errors to in discussion.
Moderate ice volume estimate by stating uncertainties from the sensitivity tests.

@@
       time-series, the model shows a significant sensitivity to selected ice
-      rheologic and basal sliding parameters (Fig.~\ref{fig:sens_ts};
+      rheological and basal sliding parameters (Fig.~\ref{fig:sens_ts};
       Table~\ref{tab:sens_extrema}) in terms of modelled ice extent, and
@@
       8.56\,\unit{m}\,s.l.e. (17.4\,\unit{ka}) in the EPICA simulation.
+      However, these numers are subject to uncertainties on physics
+      embedded in the model. The range of parameter values tested in our
+      sensitivity study on the lower-resolution GRIP simulation
+      (Sect.~\ref{sec:sens}) yielded relative errors of 30\,\unit{\%}
+      regarding ice rheological parameters and 21\,\unit{\%} regarding basal
+      sliding parameters (Fig.~\ref{fig:sens_ts};
+      Table~\ref{tab:sens_extrema}).
 
@@
       ice volumes.
+      These estimates appear little sensitive to the range of parameter
+      values tested in our sensitivity study (Sect.~\ref{sec:sens}), which
+      yielded relative errors of 17\,\unit{\%} to ice rheological parameters
+      and 21\,\unit{\%} to basal sliding parameters
+      (Fig.~\ref{fig:sens_ts}; Table~\ref{tab:sens_extrema}).
 
@@
       respectively 18 and 30\,\unit{\%} of the MIS~2 ice volumes.
+      However, these numbers should be considered with caution as our
+      sensitivity study (Sect.~\ref{sec:sens}) shows that minimum ice volume
+      during MIS~3 is highly sensitive to ice flow parameters, with relative
+      errors of 57\,\unit{\%} to the range of ice rheological parameters
+      tested and 120\,\unit{\%} to the range of basal sliding parameters
+      tested (Fig.~\ref{fig:sens_ts}; Table~\ref{tab:sens_extrema}).
 

### 2016-01-21 cordillera-cycle.tex
Restore a line that misteriously disappeared.

@@
       Selwyn and Mackenzie mountains ice dispersal centre
+      (Fig.~\ref{fig:hr_geom_duration}, SMKM), prior to the Late Pleistocene
       \citep[cf.][]{Ward.etal.2008, Demuro.etal.2012}.

### 2016-01-22 cordillera-cycle.tex
Rev. Jarosch: admit limits of NARR.
In intro, methods and discussion.

@@
       topography of the North American Cordillera also induces strong
-      geographic variations in temperature and precipitation, thus requiring
+      geographic variations in temperature and precipitation
+      \citep{Jarosch.etal.2012}, thus requiring
       the use of high-resolution climate forcing fields as an input to an
@@
       precipitation grids derived from a~present-day atmospheric reanalysis
-      \citep{Mesinger.etal.2006} that includes the steep precipitation
+      \citep{Mesinger.etal.2006} that was previously tested against
+      observational data and shown to best reproduce the steep precipitation
       gradients previously identified as necessary to model the LGM extent
       of the Cordilleran ice sheet in agreement with its geological imprint
+      among four atmospheric reanalyses available over the study area
       \citep{Seguinot.etal.2014}. To mimic climate evolution through the
@@
       seasonality, timing of the maximum annual precipitation, and daily
-      temperature variability (Fig.~\ref{fig:atm}). The use of NARR is
-      motivated by the need for an accurate, high-resolution precipitation
-      forcing, as identified in a~previous sensitivity study
-      \citep{Seguinot.etal.2014}.
+      temperature variability (Fig.~\ref{fig:atm}). Although the ability
+      of NARR to reproduce these steep climatic gradients is limited by its
+      spatial resolution of 32\,\unit{km} \citep{Jarosch.etal.2012}, it was
+      tested against observational data in a~previous sensitivity study and
+      identified as yielding modelled Cordilleran ice sheet LGM extent in
+      closer agreement to geological reconstructions than other atmospheric
+      reanalyses \citep{Seguinot.etal.2014}.
 
@@
       Such deviation from the geological inferences could reflect the fact
+      the NARR has difficulties with simulating orographic processes in some
+      areas of steep topography \citep{Jarosch.etal.2012}. In a previous
+      study \citep{Seguinot.etal.2014}, we have evaluated the performance of
+      NARR in forcing constant-climate simulations of the Cordilleran ice
+      sheet against that of an observation-based data set
+      \citep[WorldClim,][]{Hijmans.etal.2005}. Indeed, the use of NARR in
+      these simulations produced slightly different patterns of glaciation
+      relative to WorldClim, including more extensive ice cover on the
+      Columbia and Rocky mountains \citep[Figs.~6--7]{Seguinot.etal.2014}.
+      Our simulations have shown that these differences are mainly caused by
+      disparities in the precipitation fields of the two data sets
+      \citep[Figs.13--14]{Seguinot.etal.2014}. Over the southern part of our
+      model domain, it was shown that the implementation of temperature and
+      precipitation downscaling can address these limitations
+      \citep[e.g.,][]{Jarosch.etal.2012}. However, extending this
+      downscaling method to the entire model domain used in our study could
+      be challenging, because the northern part of the model domain is
+      characterized by stronger precipitation gradients (Fig.~\ref{fig:atm})
+      and includes much fewer weather stations than the south
+      \citep{Hijmans.etal.2005}.
+
+      The westerly position of the modelled ice divide and western ice sheet
+      margin may also reflect the fact
       that our model does not include feedback mechanisms between ice sheet

### 2016-01-22 cordillera-cycle.tex
Rev. Jarosch: clarify ETOPO1 bilinear interp.
Rev. Jarosch: address limits of ETOPO1 data.

@@
 
-      Basal topography is derived from the ETOPO1 combined topography and
-      bathymetry dataset with a~resolution of 1\,arc-min
-      \citep{Amante.Eakins.2009}. Sea level is lowered as a~function of time
-      based on the Spectral Mapping Project
+      Basal topography is bilinearly interpolated from the ETOPO1 combined
+      topography and bathymetry dataset with a~resolution of 1\,arc-min
+      \citep{Amante.Eakins.2009} to the model grids. Sea level is lowered as
+      a~function of time based on the Spectral Mapping Project
       \citep[SPECMAP,][]{Imbrie.etal.1989} time scale. Geothermal heat flux
@@
       Mountains (CM), the Columbia and Rocky mountains (CRM), the Skeena
-      Mountains (SM), and the Selwyn and Mackenzie mountains (SMKM). The
+      Mountains (SM), and the Selwyn and Mackenzie mountains (SMKM).
+
+      The location of modelled ice-dispersal centres is potentially biased
+      by present day ice volumes contained in the ETOPO1 basal topography
+      data. Although a reconstruction exist for the thickness of western
+      Canadian glaciers south of 60{\degree}N \citep{Clarke.etal.2013}, the
+      most problematic part of the model domain in this respect is by far
+      that of the Wrangell and Saint Elias mountains, where ice thicknesses
+      up to 1200\,m have been measured by low-frequency radar
+      \citep{Rignot.etal.2013}. In this area, located over the USA Canada
+      border just north of 60{\degree}N, temperate ice, surge dynamics and
+      deep subglacial depressions in the icefield interior pose a
+      fundamental challenge to reconstructing basal topography for the
+      entire ice cap \citep{Rignot.etal.2013}. Although this causes our
+      simulations to overestimate ice surface topography in this region, we
+      are not aware of bed topography reconstructions that could be used to
+      force the ice sheet model. However, this drawback seem to have
+      little effect on modelled Cordilleran ice sheet dynamics. In fact, the
       Wrangell and Saint Elias mountains, heavily glacierized at present,
@@
       ice cap does not appear to be a~major feed to the Cordilleran ice
-      sheet (Fig.~\ref{fig:hr_geom_duration}, WSEM). Although the Coast,
+      sheet (Fig.~\ref{fig:hr_geom_duration}, WSEM). Moreover, with this
+      exception of the Wrangell and Saint Elias mountains ice field,
+      present-day ice volumes are small relative to the ice volumes
+      concerned in our study.
+
+      Although the Coast,
       Skeena and Columbia and Rocky mountains (CM, SM, CRM) are covered by

### 2016-01-22 cordillera-cycle.tex
Rev. Jarosch: remove "SSA as a sliding law".

@@
       based on their predominant components. The Shallow Shelf Approximation
-      (SSA) is used as a ``sliding law'' for the Shallow Ice Approximation
+      (SSA) is combined with the Shallow Ice Approximation
       (SIA) by adding velocity solutions of the two approximations
-      \citep[Eqs.~7--9 and 15]{Bueler.Brown.2009, Winkelmann.etal.2011}.
+      \citep[Eqs.~7--9 and 15]{Winkelmann.etal.2011}.
 

### 2016-01-22 cordillera-cycle.tex
Rev. Jarosch: clarify where geoflux is applied.

@@
       a~function of time based on the Spectral Mapping Project
-      \citep[SPECMAP,][]{Imbrie.etal.1989} time scale. Geothermal heat flux
-      is applied as a~constant value of
-      $q_{\mathrm{G}}=70$\,\unit{mW\,m^{-2}} at
-      3\,\unit{km} depth (Sect.~\ref{sec:icedyn}). Basal sliding follows
+      \citep[SPECMAP,][]{Imbrie.etal.1989} time scale. Ice deformation
+      follows temperature and water-content dependent creep
+      (Sect.~\ref{sec:icedyn}). Basal sliding follows
       a pseudo-plastic law where the yield stress accounts for till
@@
       provides the upper boundary condition to the ice enthalpy
-      model. Temperature is computed subglacially to a~depth of
-      3\,\unit{km}, where it is conditioned by a~lower boundary geothermal
-      heat flux of 70\,\unit{mW\,m^{-2}}. Although this uniform value does
+      model. Temperature is computed in the ice and in the bedrock to
+      a~depth of 3\,\unit{km} below the ice-bedrock interface, where it is
+      conditioned by a~lower boundary geothermal heat flux of
+      $q_{\mathrm{G}}=70$\,\unit{mW\,m^{-2}}. Although this uniform value
+      does
       not account for the high spatial geothermal variability in the region

### 2016-01-22 cordillera-cycle.tex
Rev. Jarosch: clarify NARR bilinear interp.

@@
       \citep{Mesinger.etal.2006}, including variability associated with the
-      seasonal cycle (Fig.~\ref{fig:atm}). Degree-day factors for snow and
+      seasonal cycle, and bilinearly interpolated to the model grids
+      (Fig.~\ref{fig:atm}). Degree-day factors for snow and
       ice melt are derived from mass-balance measurements on contemporary
@@
 \end{align}
-      The present-day monthly climatology was computed from near-surface air
+      The present-day monthly climatology was bilinearly interpolated from
+      near-surface air
       temperature and precipitation rate fields from the NARR, averaged from

### 2016-01-22 cordillera-cycle.tex
Correct one typo.

@@
 %
-      where $\vec{\dot{\epsilon}}$ is the the strain-rate tensor,
+      where $\vec{\dot{\epsilon}}$ is the strain-rate tensor,
       $\vec{\tau}$ the deviatoric stress tensor, and $\tau_{\mathrm{e}}$ the

### 2016-01-23 cordillera-cycle.tex
Rev. Marshall: add references.

@@
 
+\bibitem[{Fisher et~al.(2008)Fisher, Osterberg, Dyke, Dahl-Jensen, Demuth, Zdanowicz, Bourgeois, Koerner, Mayewski, Wake, Kreutz, Steig, Zheng, Yalcin, Goto-Azuma, Luckman, and Rupper}]{Fisher.etal.2008}
+Fisher, D., Osterberg, E., Dyke, A., Dahl-Jensen, D., Demuth, M., Zdanowicz, C., Bourgeois, J., Koerner, R.~M., Mayewski, P., Wake, C., Kreutz, K., Steig, E., Zheng, J., Yalcin, K., Goto-Azuma, K., Luckman, B., and Rupper, S.: The Mt Logan Holocene--late Wisconsinan isotope record: tropical Pacific--Yukon connections, The Holocene, 18, 667--677,
+doi:\href{http://dx.doi.org/10.1177/0959683608092236}{10.1177/0959683608092236}, 2008.
+
+\bibitem[{Fisher et~al.(2004)Fisher, Wake, Kreutz, Yalcin, Steig, Mayewski, Anderson, Zheng, Rupper, Zdanowicz, Demuth, Waszkiewicz, Dahl-Jensen, Goto-Azuma, Bourgeois, Koerner, Sekerka, Osterberg, Abbott, Finney, and Burns}]{Fisher.etal.2004}
+Fisher, D.~A., Wake, C., Kreutz, K., Yalcin, K., Steig, E., Mayewski, P., Anderson, L., Zheng, J., Rupper, S., Zdanowicz, C., Demuth, M., Waszkiewicz, M., Dahl-Jensen, D., Goto-Azuma, K., Bourgeois, J.~B., Koerner, R.~M., Sekerka, J., Osterberg, E., Abbott, M.~B., Finney, B.~P., and Burns, S.~J.: Stable Isotope Records from Mount Logan, Eclipse Ice Cores and Nearby Jellybean Lake.
+Water Cycle of the North Pacific Over 2000 Years and Over Five Vertical Kilometres: Sudden Shifts and Tropical Connections, G\'{e}ogr.
+phys.
+Quatern., 58, 337,
+doi:\href{http://dx.doi.org/10.7202/013147ar}{10.7202/013147ar}, 2004.
+
 \bibitem[{Friele and Clague(2002{\natexlab{a}})}]{Friele.Clague.2002}
@@
 
+\bibitem[{Kienast and McKay(2001)}]{Kienast.McKay.2001}
+Kienast, S.~S. and McKay, J.~L.: Sea surface temperatures in the subarctic northeast Pacific reflect millennial-scale climate oscillations during the last 16 kyrs, Geophys. Res. Lett., 28, 1563--1566,
+doi:\href{http://dx.doi.org/10.1029/2000GL012543}{10.1029/2000GL012543}, 2001.
+
 \bibitem[{Kleman(1994)}]{Kleman.1994}
@@
 Lliboutry, L.~A. and Duval, P.: Various isotropic and anisotropic ices found in glaciers and polar ice caps and their corresponding rheologies, Ann. Geophys., 3, 207--224, 1985.
+
+\bibitem[{Luckman and Osborn(1979)}]{Luckman.Osborn.1979}
+Luckman, B. and Osborn, G.: Holocene glacier fluctuations in the middle Canadian Rocky Mountains, Quaternary Res., 11, 52--77,
+doi:\href{http://dx.doi.org/10.1016/0033-5894(79)90069-3}{10.1016/0033-5894(79)90069-3}, 1979.
 
@@
 
+\bibitem[{M\"{u}ller et~al.(1998)M\"{u}ller, Kirst, Ruhland, von Storch, and Rosell-Mel{\'{e}}}]{Muller.etal.1998}
+M\"{u}ller, P.~J., Kirst, G., Ruhland, G., von Storch, I., and Rosell-Mel{\'{e}}, A.: Calibration of the alkenone paleotemperature index $U^{K'}_{37}$ based on core-tops from the eastern South Atlantic and the global ocean (60$^\circ$N-60$^\circ$S), Geochim.
+Cosmochim.
+Ac., 62, 1757--1772,
+doi:\href{http://dx.doi.org/10.1016/s0016-7037(98)00097-0}{10.1016/s0016-7037(98)00097-0}, 1998.
+
 \bibitem[{Nye(1953)}]{Nye.1953}
@@
 doi:\href{http://dx.doi.org/10.1016/s1040-6182(96)00026-2}{10.1016/s1040-6182(96)00026-2}, 1997.
+
+\bibitem[{Osborn et~al.(1995)Osborn, Clapperton, Davis, Reasoner, Rodbell, Seltzer, and Zielinski}]{Osborn.etal.1995}
+Osborn, G., Clapperton, C., Davis, P., Reasoner, M., Rodbell, D.~T., Seltzer, G.~O., and Zielinski, G.: Potential glacial evidence for the younger dryas event in the Cordillera of North and South America, Quaternary Sci.
+Rev., 14, 823--832,
+doi:\href{http://dx.doi.org/10.1016/0277-3791(95)00064-x}{10.1016/0277-3791(95)00064-x}, 1995.
 
@@
 
+\bibitem[{Prahl and Wakeham(1987)}]{Prahl.Wakeham.1987}
+Prahl, F.~G. and Wakeham, S.~G.: Calibration of unsaturation patterns in long-chain ketone compositions for palaeotemperature assessment, Nature, 330, 367--369,
+doi:\href{http://dx.doi.org/10.1038/330367a0}{10.1038/330367a0}, 1987.
+
+\bibitem[{Prahl et~al.(1988)Prahl, Muehlhausen, and Zahnle}]{Prahl.etal.1988}
+Prahl, F.~G., Muehlhausen, L.~A., and Zahnle, D.~L.: Further evaluation of long-chain alkenones as indicators of paleoceanographic conditions, Geochim. Cosmochim. Ac., 52, 2303--2310,
+doi:\href{http://dx.doi.org/10.1016/0016-7037(88)90132-9}{10.1016/0016-7037(88)90132-9}, 1988.
+
 \bibitem[{Prest et~al.(1968)Prest, Grant, and Rampton}]{Prest.etal.1968}
@@
 
+\bibitem[{Reasoner et~al.(1994)Reasoner, Osborn, and Rutter}]{Reasoner.etal.1994}
+Reasoner, M.~A., Osborn, G., and Rutter, N.~W.: Age of the Crowfoot advance in the Canadian Rocky Mountains: A glacial event coeval with the Younger Dryas oscillation, Geology, 22, 439--442,
+doi:\href{http://dx.doi.org/10.1130/0091-7613(1994)022<0439:AOTCAI>2.3.CO;2}{10.1130/0091-7613(1994)022<0439:AOTCAI>2.3.CO;2}, 1994.
+
 \bibitem[{Rignot et~al.(2013)Rignot, Mouginot, Larsen, Gim, and Kirchner}]{Rignot.etal.2013}
@@
 doi:\href{http://dx.doi.org/10.5194/tc-8-1087-2014}{10.5194/tc-8-1087-2014}, 2014.
+
+\bibitem[{Shackleton(1967)}]{Shackleton.1967}
+Shackleton, N.: Oxygen Isotope Analyses and Pleistocene Temperatures Re-assessed, Nature, 215, 15--17,
+doi:\href{http://dx.doi.org/10.1038/215015a0}{10.1038/215015a0}, 1967.
 
@@
 doi:\href{http://dx.doi.org/10.1016/j.epsl.2014.06.026}{10.1016/j.epsl.2014.06.026}, 2014.
+
+\bibitem[{Taylor et~al.(2015)Taylor, Hendy, and Pak}]{Taylor.etal.2015}
+Taylor, M.~A., Hendy, I.~L., and Pak, D.~K.: The California Current System as a transmitter of millennial scale climate change on the northeastern Pacific margin from 10 to 50 ka, Paleoceanography, 30, 1168--1182,
+doi:\href{http://dx.doi.org/10.1002/2014pa002738}{10.1002/2014pa002738}, 2015.
 

### 2016-01-23 cordillera-cycle.tex
Rev. Marshall: adress minor comments.

@@
       all areas of the Cordilleran ice sheet \citep{Margold.etal.2013}.
-      Although solid evidence for late-glacial glacier re-advances have been
+      Although solid evidence for late-glacial glacier re-advances has been
       found in the Coast, Columbia and Rocky mountains
-      \citep{Clague.etal.1997, Friele.Clague.2002, Friele.Clague.2002a,
+      \citep{Reasoner.etal.1994, Osborn.Gerloff.1997,
+      Clague.etal.1997, Friele.Clague.2002, Friele.Clague.2002a,
       Kovanen.2002, Kovanen.Easterbrook.2002, Lakeman.etal.2008,
@@
       climatology averaged from 1979 to 2000 from the North American
-      Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006}, perturbated
+      Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006}, perturbed
       by time-dependent offsets and lapse-rate temperature corrections
@@
       Each simulation starts from assumed ice-free conditions at
-      12\,0000\,\unit{years} ago (120\,\unit{ka}), and runs to the
-      present. Our modelling domain of 1500 by 3000\,\unit{km} encompasses
+      120\,000\,\unit{years} ago (120\,\unit{ka}), and runs to the present.
+      Our modelling domain of 1\,500 by 3\,000\,\unit{km} encompasses
       the entire area covered by the Cordilleran ice sheet at the LGM
@@
       (SIA) by adding velocity solutions of the two approximations
-      \citep[Eqs.~7--9 and 15]{Winkelmann.etal.2011}.
+      \citep[Eqs.~7--9 and 15]{Winkelmann.etal.2011}. The computational
+      efficiency of this hybrid scheme nevertheless comes at the cost of
+      violating stress conservation in the transition zone where
+      gravitational stresses have a double contribution to computed SIA and
+      SSA velocities, leading to a positive bias of total ice velocity that
+      remains to be quantified.
 
@@
       up to 51~enthalpy layers in the ice sheet, corresponding to a~vertical
-      resolution of 100\,\unit{m}. The high-resolution simulations
+      resolution of 100\,\unit{m}. The high-resolution simulations use
       61~bedrock layers and up to 101~ice layers with a~vertical resolution
@@
       where $\delta$ sets the minimum ratio between the effective and
-      overburden pressures, $e_0$ is a~measured reference void
-      ratio and $C_{\mathrm{c}}$ is a~measured compressibility coefficient
-      (Table~\ref{tab:params}). The amount of water at the base, $W$, varies
-      from zero to $W_{\text{max}}$, a~threshold above which
+      overburden pressures, $e_0$ is a~reference void ratio and
+      $C_{\mathrm{c}}$ is the till compressibility coefficient
+      \citep{Tulaczyk.etal.2000}. The amount of water at the base, $W$,
+      varies from zero to $W_{\text{max}}$, a~threshold above which
       additional melt water is assumed to drain off instantaneously.
@@
       1500\,\unit{km}-long central divide above 3000\,\unit{m\,a.s.l.}
-      located along the spine of the Rocky Mountains. Although the
-      similarity of modelled ice extents is a~direct result of the choice of
-      scaling factors applied to different palaeo-temperature proxy records
-      (Table~\ref{tab:records}), it is interesting to note that modelled
+      located along the spine of the Rocky Mountains. As an indirect result
+      of the choice of scaling factors applied to different
+      palaeo-temperature proxy records (Table~\ref{tab:records}), modelled
       maximum ice volumes also fall within a~tight range of 8.39 to
@@
       record. To distinguish between different records, geological evidence
-      of former glaciations provide a~basis for validation of our runs,
+      of former glaciations provides a~basis for validation of our runs,
       while the results from numerical modelling can perhaps help to analyse
@@
       constraints from the southern margin. However, cosmogenic exposure
-      dating places the timing of maximum CIS extent during the McConnell
+      dating places the timing of maximum northern ice sheet margin extent
+      during the McConnell
       glaciation close to 17\,\unit{\chem{^{10}Be}\,ka}
@@
       mountain ranges of the North American Cordillera
-      (Fig.~\ref{fig:hr_geom_duration}, blue areas). A~2500\,\unit{km}-long
+      (Fig.~\ref{fig:hr_geom_duration}, blue areas). A~2\,500\,\unit{km}-long
       continuous expanse of ice, extending from the Alaska Range in the
-      northeast to the Rocky Mountains in the southwest, is only in
+      north-east to the Rocky Mountains in the south-west, is only in
       operation for at most 34\,\unit{ka}, which is about a~third of the
@@
       The possibility of late glacial readvances in the North American
-      Cordillera has been debated for some time \citep{Osborn.Gerloff.1997},
+      Cordillera has been debated for some time \citep{Luckman.Osborn.1979,
+      Reasoner.etal.1994, Osborn.Gerloff.1997, Osborn.etal.1995},
       and locally these have been reconstructed and dated. Radiocarbon-dated
@@
       includes moraines in the eastern Coast Mountains, and the Columbia and
-      Rocky mountains \citep{Osborn.Gerloff.1997, Menounos.etal.2008}.
+      Rocky mountains \citep{Reasoner.etal.1994, Osborn.Gerloff.1997,
+      Menounos.etal.2008}.
 
@@
 
-      Because a~general tenant in glacial geomorphology is that the majority
+      Because a~general conjecture in glacial geomorphology is that the majority
       of landforms (lineations and eskers) are part of the deglacial
@@
       indicate a~substantial eastwards flow component of the Cordilleran ice
-      sheet \citep{Prest.etal.1968, Kleman.etal.2010}. This Interior Plateau
-      lineation set could therefore present a~smoking gun for the
-      reliability of the presented model results. One explanation for the
+      sheet \citep{Prest.etal.1968, Kleman.etal.2010} not present in any of
+      the two simulations (Fig.~\ref{fig:hr_geom_lastflow}).
+      One explanation for the
       incongruent results could be that the missing feedback mechanisms
@@
       Interior Plateau. However, a~more westerly-positioned LGM ice divide
-      would certainly be associated with a~thinner ice sheet than that
+      would certainly be associated with an LGM ice sheet less extensive to
+      the west, and therefore thinner ice on the Interior Plateau during
+      deglaciation than
       modelled here. Decreased ice thickness would not promote warm-based

### 2016-01-23 cordillera-cycle.tex
Rev. Marshall: regional proxy records.

@@
       unsaturation index records from Northwest Pacific ocean sediment cores
-      \citep{Herbert.etal.2001}. After testing the model sensitivity to
+      \citep{Herbert.etal.2001},
+
+      Although these proxy records were all obtained outside the model
+      domain, more regional palaeo-temperature reconstructions spanning over
+      the last glacial cycle are lacking. For instance, the Mount Logan ice
+      core oxygen isotope record covers only the last 30\,000\,\unit{years}
+      (30\,\unit{ka}) and has been interpreted as a proxy for source region
+      rather than for palaeo-temperature \citep{Fisher.etal.2004,
+      Fisher.etal.2008}. Sea-surface temperatures have been reconstructed
+      offshore Vancouver Island from alkenone unsaturation indices over the
+      last 16\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Kienast.McKay.2001}, and
+      from the Mg/Ca ratio in planktonic foraminifera over the period from
+      10 to ca.~50\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Taylor.etal.2014,
+      Taylor.etal.2015}, but these records cover only parts of the last
+      glacial cycle.
+
+      After testing the model sensitivity to
       these climate forcing time-series (Sect.~\ref{sec:results}) and to
@@
       Each simulation starts from assumed ice-free conditions at
-      120\,000\,\unit{years} ago (120\,\unit{ka}), and runs to the present.
+      120\,\unit{ka}, and runs to the present.
       Our modelling domain of 1\,500 by 3\,000\,\unit{km} encompasses
@@
       affected by the weakening of the California current during the LGM.
+      Nevertheless, the fact that Cordilleran ice sheet dynamics are here
+      modelled in best agreement to geological reconstructions using
+      palaeo-temperature records from the distant Greenland and Antarctic
+      ice sheets, and the significant differences remaining between our two
+      preffered simulations highlight the need for more regional
+      palaeo-climate reconstructions of the last glacial cycle in and around
+      the North American Cordillera.
 
@@
 
-\bibitem[{M\"{u}ller et~al.(1998)M\"{u}ller, Kirst, Ruhland, von Storch, and Rosell-Mel{\'{e}}}]{Muller.etal.1998}
-M\"{u}ller, P.~J., Kirst, G., Ruhland, G., von Storch, I., and Rosell-Mel{\'{e}}, A.: Calibration of the alkenone paleotemperature index $U^{K'}_{37}$ based on core-tops from the eastern South Atlantic and the global ocean (60$^\circ$N-60$^\circ$S), Geochim.
-Cosmochim.
-Ac., 62, 1757--1772,
-doi:\href{http://dx.doi.org/10.1016/s0016-7037(98)00097-0}{10.1016/s0016-7037(98)00097-0}, 1998.
-
 \bibitem[{Nye(1953)}]{Nye.1953}
@@
 
-\bibitem[{Prahl and Wakeham(1987)}]{Prahl.Wakeham.1987}
-Prahl, F.~G. and Wakeham, S.~G.: Calibration of unsaturation patterns in long-chain ketone compositions for palaeotemperature assessment, Nature, 330, 367--369,
-doi:\href{http://dx.doi.org/10.1038/330367a0}{10.1038/330367a0}, 1987.
-
-\bibitem[{Prahl et~al.(1988)Prahl, Muehlhausen, and Zahnle}]{Prahl.etal.1988}
-Prahl, F.~G., Muehlhausen, L.~A., and Zahnle, D.~L.: Further evaluation of long-chain alkenones as indicators of paleoceanographic conditions, Geochim. Cosmochim. Ac., 52, 2303--2310,
-doi:\href{http://dx.doi.org/10.1016/0016-7037(88)90132-9}{10.1016/0016-7037(88)90132-9}, 1988.
-
 \bibitem[{Prest et~al.(1968)Prest, Grant, and Rampton}]{Prest.etal.1968}
@@
 doi:\href{http://dx.doi.org/10.5194/tc-8-1087-2014}{10.5194/tc-8-1087-2014}, 2014.
-
-\bibitem[{Shackleton(1967)}]{Shackleton.1967}
-Shackleton, N.: Oxygen Isotope Analyses and Pleistocene Temperatures Re-assessed, Nature, 215, 15--17,
-doi:\href{http://dx.doi.org/10.1038/215015a0}{10.1038/215015a0}, 1967.
 

### 2016-01-23 cordillera-cycle.tex
Rev. Marshall: fixed climate patterns.

@@
       The westerly position of the modelled ice divide and western ice sheet
-      margin may also reflect the fact
-      that our model does not include feedback mechanisms between ice sheet
-      topography and the regional climate. Firstly, during the build-up
+      margin may also reflect the fact that our model assumes fixed
+      modern-day spatial patterns for temperature and precipitation and thus
+      does not include feedback mechanisms between ice sheet topography and
+      the regional climate. Regarding temperature, it is reasonable to think
+      that the cooling was greater inland than near the coast, prohibiting
+      melt at the western margin. However, our simulations already produce
+      an excess of ice inland. Including such a temperature continentality
+      gradient in the model while keeping the precipitation pattern constant
+      would thus cause an even greater mismatch between the model results
+      and the geologically reconstructed Last Glacial Maximum (LGM) ice
+      margins.
+
+      Consequently, the mismatch we observe between the modelled and
+      reconstructed LGM ice margins let us think that the assumption of
+      fixed modern-day precipitation patterns is more critical than the
+      assumption of fixed modern-day temperature patterns.
+      Firstly, during the build-up
       phase preceding the LGM, rapid accumulation over the Coast Mountains
@@
       topography could have resulted in an inflow of potentially warmer air
-      over the eastern flank of the ice sheet, increasing melt along the
+      over the eastern flank of the ice sheet, counterbalancing the potential
+      continentality gradient discussed above through increasing melt along the
       advancing margin \citep[cf.][]{Langen.etal.2012}. Because these two

### 2016-01-24 cordillera-cycle.tex
Rev. Stumpf: update to a larger location map.

@@
 
-\begin{figure}%
+\begin{figure*}%
 \includegraphics{locmap}
 \caption{%
-      Relief map of northern North America showing a~reconstruction of the
-      areas once covered by the Cordilleran (CIS), Laurentide (LIS),
-      Innuitian (IIS), and Greenland (GIS) ice sheets during the last
-      18\,\unit{\chem{^{14}C}\,ka} \citep[21.4\,cal\,ka,][]{Dyke.2004}. The
-      rectangular box denotes the location of the modelling domain used in
-      this study. Major mountain ranges covered by the ice sheet include the
-      Alaska Range (AR), the Wrangell and Saint Elias mountains (WSEM), the
-      Selwyn and Mackenzie mountains (SMKM), the Skeena Mountains (SM), the
-      Coast Mountains (CM), the Columbia and Rocky Mountains (CRM), and the
-      North Cascades (NC). Major depressions include the Liard Lowland (LL),
-      the Interior Plateau of British Columbia (IP), and the Puget Lowland
-      (PL). The background map consists of ETOPO1 \citep{Amante.Eakins.2009}
-      and Natural Earth Data \citep{Patterson.Kelso.2015}.}
+      Relief map of the northern American Cordillera showing cumulative last
+      glacial maximum ice cover between 21.4 and
+      16.8\,\unit{\chem{^{14}C}\,cal\,ka} \citep[red line]{Dyke.2004}, and
+      the modelling domain used in this study (black rectangle). The
+      background map consists of ETOPO1 \citep{Amante.Eakins.2009} and
+      Natural Earth Data \citep{Patterson.Kelso.2015}.}
 \label{fig:locmap}%
-\end{figure}%
+\end{figure*}%
 

### 2016-01-24 figures/data/get-etopo1.sh

@@ -12,10 +12,13 @@ output=etopo1.nc
 # inflate archive
 [ -f $input ] || unzip $archive
 
-# reproject
-gdalwarp -s_srs EPSG:4326 -t_srs EPSG:3979 \
-         -te -3187500 -900000 3187500 3600000 \
-         -tr 7500 7500 \
+# reproject for the Cordillera
+#proj='+proj=lcc +lon_0=-95 +lat_0=49 +lat_1=79 +lat_2=77'  # cal
+proj='+proj=lcc +lon_0=-135 +lat_0=45 +lat_1=45 +lat_2=70'
+gdalwarp -s_srs EPSG:4326 -t_srs "$proj" \
+         -te -2250000 000000 2250000 3000000 \
+         -tr 2500 2500 \
+         -wm 512 -wo SOURCE_EXTRA=1000 \
          -srcnodata -2147483648 -dstnodata -2147483648 \
          -of netcdf -overwrite \
          $input $output

### 2016-01-24 figures/locmap.py

@@ -1,4 +1,4 @@
-#!/usr/bin/env python
+#!/usr/bin/env python2
 # coding: utf-8
 
 import sys
@@ -13,17 +13,19 @@ import cartopy.feature as cfeature
 from iceplotlib import cm as icm
 from netCDF4 import Dataset
 
-in2mm = 1/25.4
-pt2mm = 72*in2mm
 bwu = 0.5  # base width unit
 scale = '50m'
 
-# Canadian Atlas Lambert projection
+# Geographic projections
 ll = ccrs.PlateCarree()
-proj = ccrs.LambertConformal(
+cal = ccrs.LambertConformal(
     central_longitude=-95.0, central_latitude=49.0,
     false_easting=0.0, false_northing=0.0,
-    secant_latitudes=(49.0, 77.0), globe=None, cutoff=0)
+    standard_parallels=(49.0, 77.0), globe=None, cutoff=0)
+proj = ccrs.LambertConformal(
+    central_longitude=-135.0, central_latitude=45.0,
+    false_easting=0.0, false_northing=0.0,
+    standard_parallels=(45.0, 70.0), globe=None, cutoff=0)
 
 # ETOPO1 background topo
 def draw_etopo1(**kwargs):
@@ -61,8 +63,7 @@ def draw_lgm():
                     union = rec.geometry
                 else:
                     union = union.union(rec.geometry)
-    ax.add_geometries(union, proj, edgecolor='none', facecolor='#f5f4f2', alpha=0.75)
-    ax.add_geometries(union, proj, edgecolor='#0978ab', facecolor='none', lw=1.0*bwu)
+    ax.add_geometries(union, cal, edgecolor='#b00000', facecolor='none', lw=1.0*bwu)
 
 # Natural Earth elements
 def draw_rivers():
@@ -90,36 +91,50 @@ def draw_countries():
 
 def draw_graticules():
     ax.add_feature(cfeature.NaturalEarthFeature(
-        category='physical', name='graticules_15', scale=scale,
+        category='physical', name='graticules_5', scale=scale,
         edgecolor='#000000', facecolor='none', lw=0.25*bwu))
 
 # Geographic names
 def add_names():
[…]

### 2016-01-24 cordillera-cycle.tex
Rev. Stumpf: rework first sentences in abtract.

@@
 \begin{abstract}
-Despite more than a century of geological observations, the
-Cordilleran ice sheet of North America remains poorly understood in
-terms of its former extent, volume and dynamics. Although
-geomorphological evidence is abundant, its complexity is such that
-whole ice-sheet reconstructions of advance and retreat patterns are
+After more than a century of geological research, the Cordilleran
+ice sheet of North America remains among the least understood in
+terms of its former extent, volume, and dynamics. Because of the
+mountainous topography on which the ice sheet formed, geological
+studies have often had only local or regional relevance, and shown
+such a complexity that ice sheet-wide spatial reconstructions of
+advance and retreat patterns are
 lacking. Here we use a numerical ice sheet model calibrated against

### 2016-01-24 cordillera-cycle.tex
Correct date in Perkins.Brennand.2015.

@@
       pro-glacial lakes dammed by the retreating ice sheet
-      \citep{Perkins.Brennand.2014}. However, the two simulations differ in
+      \citep{Perkins.Brennand.2015}. However, the two simulations differ in
       the mode of retreat. The GRIP simulation yields an active retreat with
@@
 
-\bibitem[{Perkins and Brennand(2014)}]{Perkins.Brennand.2014}
+\bibitem[{Perkins and Brennand(2015)}]{Perkins.Brennand.2015}
 Perkins,~A.~J.
 and Brennand,~T.~A.: Refining the pattern and style of Cordilleran Ice Sheet retreat: palaeogeography, evolution and implications of lateglacial ice-dammed lake systems on the southern Fraser Plateau, British Columbia, Canada, Boreas, 44, 319--342,
-doi:\href{http://dx.doi.org/10.1111/bor.12100}{10.1111/bor.12100}, 2014.
+doi:\href{http://dx.doi.org/10.1111/bor.12100}{10.1111/bor.12100}, 2015.
 

### 2016-01-24 cordillera-cycle.tex
Rev. Stumpf: add ref to Ferbey et al. (2013).

@@
       indicate a~substantial eastwards flow component of the Cordilleran ice
-      sheet \citep{Prest.etal.1968, Kleman.etal.2010} not present in any of
-      the two simulations (Fig.~\ref{fig:hr_geom_lastflow}).
-      One explanation for the
+      sheet \citep{Prest.etal.1968, Kleman.etal.2010, Ferbey.etal.2013}
+      not present in any of the two simulations
+      (Fig.~\ref{fig:hr_geom_lastflow}). One explanation for the
       incongruent results could be that the missing feedback mechanisms
@@
 doi:\href{http://dx.doi.org/10.1126/science.1065370}{10.1126/science.1065370}, 2001.
+
+\bibitem[{Ferbey et~al.(2013)Ferbey, Arnold, and Hickin}]{Ferbey.etal.2013}
+Ferbey, T., Arnold, H., and Hickin, A.: Ice-flow indicator compilation, British Columbia., British Columbia Geol. Surv., Victoria, BC, Open-File 2013-06,
+\urlprefix\url{http://www.em.gov.bc.ca/Mining/Geoscience/PublicationsCatalogue/OpenFiles/2013/Pages/2013-06.aspx}, 2013.
 

### 2016-01-24 cordillera-cycle.tex
Rev. Stumpf: add two refs to Stumpf et al. (2000).

@@
       emanating in a~radial pattern from the centre of the mountain range
-      \citep[Fig.~2]{Kleman.etal.2010}. We suggest that persistent ice cover
+      \citep[Fig.~2]{Kleman.etal.2010} that appear to
+      predate the LGM phase of the Cordilleran ice sheet
+      \citep{Stumpf.etal.2000}. We suggest that persistent ice cover
       (Fig.~\ref{fig:hr_geom_duration}) associated with basal ice
@@
       \citep[Fig.~2]{Margold.etal.2013}.
+      Inside the retreating western margin, modelled south-westward flow
+      patterns emanating from the Skeena Mountains and running over the the
+      central Coast Mountains as are compatible with geological evidence
+      from this region \citep[Fig.~12]{Stumpf.etal.2000}.
 

### 2016-01-24 cordillera-cycle.tex
Put no-break spaces after 'a' in new text.

@@
       core oxygen isotope record covers only the last 30\,000\,\unit{years}
-      (30\,\unit{ka}) and has been interpreted as a proxy for source region
+      (30\,\unit{ka}) and has been interpreted as a~proxy for source region
       rather than for palaeo-temperature \citep{Fisher.etal.2004,
@@
       (Sect.~\ref{sec:icedyn}). Basal sliding follows
-      a pseudo-plastic law where the yield stress accounts for till
+      a~pseudo-plastic law where the yield stress accounts for till
       deconsolidation under high water pressure (Sect.~\ref{sec:sliding}).
@@
       by time-dependent offsets and lapse-rate temperature corrections
-      (Sect.~\ref{sec:atm}, Table~\ref{tab:records}). A sensitivity study
+      (Sect.~\ref{sec:atm}, Table~\ref{tab:records}). A~sensitivity study
       to some of the most influential ice rheology (Sect~\ref{sec:icedyn})
@@
       violating stress conservation in the transition zone where
-      gravitational stresses have a double contribution to computed SIA and
-      SSA velocities, leading to a positive bias of total ice velocity that
+      gravitational stresses have a~double contribution to computed SIA and
+      SSA velocities, leading to a~positive bias of total ice velocity that
       remains to be quantified.
@@
       on ice temperature, $T$, pressure, $p$, and water content, $\omega$,
-      through a piece-wise Arrhenius-type law
+      through a~piece-wise Arrhenius-type law
       \citep[Eqs.~63--65]{Aschwanden.etal.2012},
@@
       $Q_{\mathrm{w}}$, are constant parameters corresponding to values
-      measured below and above a critical temperature threshold
+      measured below and above a~critical temperature threshold
       $T_{\mathrm{c}}=-10$\,\unit{{\degree}C}
       \citep[p.~72]{Paterson.Budd.1982,Cuffey.Paterson.2010}. The water
-      fraction, $\omega$, is capped at a maximum value of 0.01, above which
+      fraction, $\omega$, is capped at a~maximum value of 0.01, above which
       no measurements are available \citep[Eq.~5.7]{Lliboutry.Duval.1985,
-      Greve.1997}. Finally, $E$ is a non-dimensional enhancement factor
+      Greve.1997}. Finally, $E$ is a~non-dimensional enhancement factor
       which can take different values, $E_{\text{SIA}}$, in the Shallow Ice
@@
     $Q_{\mathrm{w}}$, derived from \citet[p.~72 and
-    76]{Cuffey.Paterson.2010}, and $E_{\text{SIA}}=1$, which correspond to a
-    stiffer rheology than that used in the control run.
+    76]{Cuffey.Paterson.2010}, and $E_{\text{SIA}}=1$, which correspond to
+    a~stiffer rheology than that used in the control run.
   \item Our \emph{soft ice} configuration include rheological parameters
@@
 
-      We also use a constant spatial distribution of the till friction
-      angle, $\phi$, whose values vary from 15 to 45\unit{\degree} as a
+      We also use a~constant spatial distribution of the till friction
+      angle, $\phi$, whose values vary from 15 to 45\unit{\degree} as
       a~piecewise-linear function of modern bed elevation, with the lowest
@@
 
-      An additional simulation with a constant till friction angle,
+      An additional simulation with a~constant till friction angle,
       $\phi=30$\unit{\degree}, corresponding to the average value in
@@
       Using the GRIP ice core palaeo-temperature record as climate forcing
-      time-series, the model shows a significant sensitivity to selected ice
+      time-series, the model shows a~significant sensitivity to selected ice
       rheological and basal sliding parameters (Fig.~\ref{fig:sens_ts};
@@
 
-      As a direct result of the different scaling factors applied
+      As a~direct result of the different scaling factors applied
       (Table~\ref{tab:sens_params}), the resulting simulations show very
@@
       MIS~3. In fact, the extent of the remnant ice cap which persists over
-      the Skeena Mountains during this stage shows a significant sensitivity
+      the Skeena Mountains during this stage shows a~significant sensitivity
       to ice rheology of 31\,\unit{\%}, and an even more important
@@
       Fig.~\ref{fig:sens_ts}). As one could expect, softer ice and weaker
-      till both result in a thinner ice sheet, while harder ice and stronger
-      till result in a thicker ice sheet. For instance, peak ice volume
+      till both result in a~thinner ice sheet, while harder ice and stronger
+      till result in a~thicker ice sheet. For instance, peak ice volume
       during the MIS~2 (LGM) varies by 30\,\unit{\%} between the two
@@
       the NARR has difficulties with simulating orographic processes in some
-      areas of steep topography \citep{Jarosch.etal.2012}. In a previous
+      areas of steep topography \citep{Jarosch.etal.2012}. In a~previous
       study \citep{Seguinot.etal.2014}, we have evaluated the performance of
@@
       melt at the western margin. However, our simulations already produce
-      an excess of ice inland. Including such a temperature continentality
+      an excess of ice inland. Including such a~temperature continentality
       gradient in the model while keeping the precipitation pattern constant
@@
       by present day ice volumes contained in the ETOPO1 basal topography
-      data. Although a reconstruction exist for the thickness of western
+      data. Although a~reconstruction exist for the thickness of western
       Canadian glaciers south of 60{\degree}N \citep{Clarke.etal.2013}, the
@@
       border just north of 60{\degree}N, temperate ice, surge dynamics and
-      deep subglacial depressions in the icefield interior pose a
-      fundamental challenge to reconstructing basal topography for the
+      deep subglacial depressions in the icefield interior pose
+      a~fundamental challenge to reconstructing basal topography for the
       entire ice cap \citep{Rignot.etal.2013}. Although this causes our
@@
 \caption{%
-      Ice softness parameter, $A$, as a function of pressure-adjusted
+      Ice softness parameter, $A$, as a~function of pressure-adjusted
       temperature, $T_{\text{pa}}$, for the default
@@
       $E_{\text{SIA}}=1$]{Cuffey.Paterson.2010}, and soft ice \citep[with
-      $E_{\text{SIA}}=5$]{Cuffey.Paterson.2010} rheologies, using a linear
+      $E_{\text{SIA}}=5$]{Cuffey.Paterson.2010} rheologies, using a~linear
       scale (top panel) and logarithmic scale (bottom panel). Figure made
@@
 \caption{%
-      Effective pressure, $N$, as a function of water content in the till,
+      Effective pressure, $N$, as a~function of water content in the till,
       $W$, for the default ($\delta=0.02$, $W_{\text{max}}=2$\,m), hard bed
       ($\delta=0.05$, $W_{\text{max}}=5$\,m), and soft bed ($\delta=0.01$,
-      $W_{\text{max}}=1$\,m) sliding parametrisations, using a linear scale
-      (top panel) and a logarithmic scale (bottom panel). Calculations are
+      $W_{\text{max}}=1$\,m) sliding parametrisations, using a~linear scale
+      (top panel) and a~logarithmic scale (bottom panel). Calculations are
       made for an ice thickness, $h$, of 1000\,m. Figure made using
@@
       fields indicate Marine Oxygen Isotope Stage (MIS) boundaries for MIS~2
-      and MIS~4 according to a global compilation of benthic
+      and MIS~4 according to a~global compilation of benthic
       \chem{\delta^{18}O} records \citep{Lisiecki.Raymo.2005}.}

### 2016-01-25 cordillera-cycle.tex
Clarify grounded areas and s.l.-relevant volumes.

@@
       features that can be observed across the range of forcing data used.
-      In all simulations, modelled ice volumes remain relatively low during
+      In all simulations, modelled sea-level relevant ice volumes remain
+      relatively low during
       most of the glacial cycle, except during two major glacial events
@@
       23.1 and 16.8\,\unit{ka} during MIS~2 (Fig.~\ref{fig:lr_ts}, lower
-      panel). An ice volume minimum is consistently reached between 56.0 and
+      panel). An sea-level relevant ice volume minimum is consistently
+      reached between 56.0 and
       42.9\,\unit{ka} during MIS~3. However, the magnitude and precise
@@
       Simulations forced by the Greenland ice core palaeo-temperature
-      records (GRIP, NGRIP) produce the highest variability in modelled ice
+      records (GRIP, NGRIP) produce the highest variability in modelled
+      sea-level relevant ice
       volume throughout the last glacial cycle. In contrast, simulations
@@
       While simulations driven by the GRIP and the two Antarctic
-      palaeo-temperature records attain a~last ice volume maximum between
+      palaeo-temperature records attain a~last sea-level relevant ice volume
+      maximum between
       19.1 and 16.8\,\unit{ka}, those informed by the NGRIP and the two
@@
       thousands of years earlier. Moreover, the ODP~1012 run yields a~rapid
-      deglaciation of the modelled area prior to 17\,\unit{ka}. The ODP~1020
+      deglaciation of the modelled grounded ice extent prior to
+      17\,\unit{ka}. The ODP~1020
       simulation predicts an early maximum in ice volume at 21.0\,\unit{ka},
@@
       attained during MIS~2 (Table~\ref{tab:extrema};
-      Fig.~\ref{fig:lr_maps}, upper panels). Corresponding maximum ice
-      volumes also differ significantly between model runs, and vary between
+      Fig.~\ref{fig:lr_maps}, upper panels). Corresponding maximum sea level
+      potentials
+      also differ significantly between model runs, and vary between
       3.54 and~8.71\,\unit{m} sea level equivalents (m\,s.l.e.;
@@
       dimensions depend sensitively on the choice of the applied
-      palaeo-temperature record. Modelled ice volume minima spread over
+      palaeo-temperature record. Modelled sea-level relevant ice volume
+      minima spread over
       a~wide range between 1.54 and 2.94\,\unit{m}\,s.l.e.
@@
       palaeo-temperature proxy records (Table~\ref{tab:records}), modelled
-      maximum ice volumes also fall within a~tight range of 8.39 to
+      maximum sea level potentials also fall within a~tight range of 8.39 to
       9.07\,\unit{m}\,s.l.e. (Table~\ref{tab:extrema}).
@@
       rheological and basal sliding parameters (Fig.~\ref{fig:sens_ts};
-      Table~\ref{tab:sens_extrema}) in terms of modelled ice extent, and
-      even more so in terms of modelled ice volumes.
+      Table~\ref{tab:sens_extrema}) in terms of modelled grounded ice
+      extent, and even more so in terms of modelled sea-level relevant ice
+      volumes.
 
@@
       little difference in the modelled glaciated areas corresponding to
-      maximum ice volumes during MIS~2, but also during MIS~4
+      the maximum stage during MIS~2, but also during MIS~4
       (Table~\ref{tab:sens_extrema}). However, such can not be said of the
-      modelled glaciated area corresponding to minimum ice volume during
+      modelled glaciated area corresponding to minimum stage during
       MIS~3. In fact, the extent of the remnant ice cap which persists over
@@
 
-      The modelled sea-level relevant ice volumes show more variability than
+      The modelled sea-level potentials show more variability than
       the modelled glaciated areas (Table~\ref{tab:sens_extrema},
@@
       till both result in a~thinner ice sheet, while harder ice and stronger
-      till result in a~thicker ice sheet. For instance, peak ice volume
+      till result in a~thicker ice sheet. For instance, peak sea-level
+      relevant ice volume
       during the MIS~2 (LGM) varies by 30\,\unit{\%} between the two
@@
       the two parametrisations of basal sliding used. The differences in
-      sea-level relevant ice volume are greatest during the MIS~3
+      ice volume are greatest during the MIS~3
       (Table~\ref{tab:sens_extrema}, Fig.~\ref{fig:sens_ts}) where both the
@@
 
-      During MIS~2, the modelled total ice volume peaks at
+      During MIS~2, the modelled sea level potential peaks at
       8.62\,\unit{m}\,s.l.e. (19.1\,\unit{ka}) in the GRIP simulation and at
@@
 
-      During MIS~4, the modelled total ice volume peaks at
+      During MIS~4, the modelled sea level potential peaks at
       7.43\,\unit{m}\,s.l.e. (57.6\,\unit{ka}) in the GRIP simulation and at
@@
 
-      The modelled total ice volume corresponding to these persistent
+      The modelled sea level potential corresponding to these persistent
       ice-dispersal centres attains a~minimum of 1.54\,\unit{m}\,s.l.e.
@@
 \caption{%
-      Extremes in Cordilleran ice sheet volume and extent corresponding to
+      Extremes in Cordilleran ice sheet grounded ice extent and sea-level
+      relevant ice volume corresponding to
       MIS~4, 3 and 2 for each of the six low-resolution simulations
@@
 \caption{%
-      Extremes in Cordilleran ice sheet volume and extent corresponding to
+      Extremes in Cordilleran ice sheet grounded ice extent and sea-level
+      relevant ice volume corresponding to
       MIS~4, 3 and 2 using the GRIP paleo-climate forcing with each
@@
       (Table~\ref{tab:records}) used as palaeo-climate forcing for the ice
-      sheet model (top panel), and modelled ice volume (bottom panel)
-      through the last 120\,\unit{ka}. Ice volumes are expressed in meters
+      sheet model (top panel), and modelled sea-level relevant ice volume
+      (bottom panel) through the last 120\,\unit{ka}, expressed in meters
       of sea level equivalent (\unit{m\,s.l.e.}). Gray fields indicate Marine
@@
       Snapshots of modelled surface topography (500\,\unit{m} contours)
-      corresponding to the ice volume extremes indicated on
-      Fig.~\ref{fig:lr_ts}. An ice cap persists over the Skeena Mountains
+      corresponding to the sea-level relevant ice volume extremes indicated
+      on Fig.~\ref{fig:lr_ts}. An ice cap persists over the Skeena Mountains
       (SM) during MIS~3. Note the occurence of spatial similarities despite
@@
       Modelled surface topography (200\,\unit{m} contours) and surface
-      velocity (colour mapping) corresponding to the maximum ice volume
+      velocity (colour mapping) corresponding to the maximum ice
+      sea-level relevant volume
       during MIS~2 in the GRIP and EPICA high-resolution simulations.}
@@
       Modelled surface topography (200\,\unit{m} contours) and surface
-      velocity (colour mapping) corresponding to the maximum ice volume
+      velocity (colour mapping) corresponding to the maximum ice
+      sea-level relevant volume
       during MIS~4 in the GRIP and EPICA high-resolution simulations.}
@@
       Temperature offset time-series from the GRIP and EPICA ice core
-      records (Table~\ref{tab:records}) (top panel), and modelled ice volume
-      during the deglaciation, expressed in meters of sea-level equivalent
-      (bottom panel).}
+      records (Table~\ref{tab:records}) (top panel), and modelled sea-level
+      relevant ice volume during the deglaciation (bottom panel).}
 \label{fig:hr_ts_deglac}%

### 2016-01-25 cordillera-cycle.tex
Update acknowledgements.

@@
       grids, using a~lower horizontal resolution of 10\,\unit{km}, and
-      a~higher horizontal resolution of 5\,\unit{km}. These computations
-      were performed on 16 to 128 computing cores at the Swedish National
-      Supercomputing Centre.
+      a~higher horizontal resolution of 5\,\unit{km}.
 
@@
       and Andy Aschwanden for providing constant help and development with
-      PISM. This work was supported by the Swedish Research Council~(VR)
+      PISM. We thank Shawn Marshall, Alexander Jarosch and Andrew Stumpf for
+      their constructive and complementary reviews on the first manuscript.
+      This work was supported by the Swedish Research Council~(VR)
       grant no.~2008-3449 to A.~P.~Stroeven and by the German Academic
@@
       allocation no.~2013/1-159 and~2014/1-159 to A.~P.~Stroeven at the
-      National Supercomputing Center (NSC).
+      National Supercomputing Center (NSC) and by the Swiss National
+      Supercomputing Centre (CSCS) allocation no.~s573 to J.~Seguinot.
 \end{acknowledgements}

### 2016-01-25 cordillera-cycle.tex
Rev. Marshall: rework SIA+SSA limits statement.

@@
       (SIA) by adding velocity solutions of the two approximations
-      \citep[Eqs.~7--9 and 15]{Winkelmann.etal.2011}. The computational
-      efficiency of this hybrid scheme nevertheless comes at the cost of
-      violating stress conservation in the transition zone where
-      gravitational stresses have a~double contribution to computed SIA and
-      SSA velocities, leading to a~positive bias of total ice velocity that
-      remains to be quantified.
+      \citep[Eqs.~7--9 and 15]{Winkelmann.etal.2011}. Although this
+      heuristic imply errors in the transition zone where gravitational
+      stresses intervene both in the SIA and SSA velocity computation, this
+      hybrid scheme is computationally much more efficient than a fully
+      three-dimensional model with which the simulations presented here
+      would not be feasible.
 

### 2016-01-25 figures/lr_misareas.py

@@ -6,7 +6,7 @@ import numpy as np
 import iceplotlib.plot as iplt
 
 res = '10km'
-offsets = np.arange(5.7, 7.0, 0.1)
+offsets = np.arange(5.7, 7.0, 0.05)
 target = 2.1
 
 # prepare empty array

### 2016-01-25 cordillera-cycle.tex
Update tables to 0.05 K simulations.

@@
              & \chem{\delta^{18}O}
-             & 0.75 & $-$6.0  % -7.9550
+             & 0.75 & $-$5.95  % -7.9550
              & \citet{Petit.etal.1999} \\
@@
              & \chem{U^{K'}_{37}}
-             & 1.61 & $-$6.1  % -3.7889
+             & 1.62 & $-$6.15  % -3.7889
              & \citet{Herbert.etal.2001} \\
@@
              & \chem{U^{K'}_{37}}
-             & 1.18 & $-$5.9  % -5.0000
+             & 1.21 & $-$6.05  % -5.0000
              & \citet{Herbert.etal.2001} \\
@@
              &   4.84 &   2.55 &   8.56 \\
-    Vostok   &  62.17 &  55.95 &  16.83
-             &   1.52 &   1.04 &   2.17
-             &   5.04 &   2.94 &   9.07 \\
-    ODP 1012 &  56.91 &  47.69 &  23.10
-             &   1.34 &   0.87 &   2.06
-             &   4.10 &   2.20 &   8.40 \\
-    ODP 1020 &  60.72 &  52.88 &  20.99
-             &   1.21 &   0.69 &   2.06
-             &   3.54 &   1.54 &   8.39 \\
+    Vostok   &  62.21 &  55.87 &  16.86
+             &   1.50 &   1.01 &   2.09
+             &   4.94 &   2.81 &   8.57 \\
+    ODP 1012 &  56.88 &  47.46 &  23.21
+             &   1.36 &   0.90 &   2.13
+             &   4.18 &   2.30 &   8.75 \\
+    ODP 1020 &  60.37 &  52.72 &  20.41
+             &   1.25 &   0.73 &   2.09
+             &   3.66 &   1.65 &   8.62 \\
     \cline{1-10}
-    Minimum  &  56.91 &  42.91 &  16.83
-             &   1.21 &   0.67 &   2.06
-             &   3.54 &   1.54 &   8.39 \\
-    Maximum  &  62.17 &  55.95 &  23.10
-             &   2.13 &   1.04 &   2.17
-             &   8.71 &   2.94 &   9.07 \\
+    Minimum  &  56.88 &  42.91 &  16.86
+             &   1.25 &   0.67 &   2.08
+             &   3.66 &   1.54 &   8.56 \\
+    Maximum  &  62.21 &  55.87 &  23.21
+             &   2.13 &   1.01 &   2.13
+             &   8.71 &   2.81 &   8.75 \\
     \bottomhline

### 2016-01-25 cordillera-cycle.tex
Update text to 0.05 K simulations.

@@
 major glaciations during marine oxygen isotope stages 4
-(62.2--56.9\,\unit{ka}) and 2 (23.1--16.8\,\unit{ka}). The timing of
+(62.2--56.9\,\unit{ka}) and 2 (23.2--16.9\,\unit{ka}). The timing of
 glaciation is better reproduced using temperature reconstructions
@@
       which occur between 62.2 and 56.9\,\unit{ka} during MIS~4, and between
-      23.1 and 16.8\,\unit{ka} during MIS~2 (Fig.~\ref{fig:lr_ts}, lower
+      23.2 and 16.9\,\unit{ka} during MIS~2 (Fig.~\ref{fig:lr_ts}, lower
       panel). An sea-level relevant ice volume minimum is consistently
-      reached between 56.0 and
+      reached between 55.9 and
       42.9\,\unit{ka} during MIS~3. However, the magnitude and precise
@@
       maximum between
-      19.1 and 16.8\,\unit{ka}, those informed by the NGRIP and the two
+      19.1 and 16.9\,\unit{ka}, those informed by the NGRIP and the two
       oceanic palaeo-temperature records attain their maximum ice volumes
@@
       17\,\unit{ka}. The ODP~1020
-      simulation predicts an early maximum in ice volume at 21.0\,\unit{ka},
+      simulation predicts an early maximum in ice volume at 20.4\,\unit{ka},
       followed by slower deglaciation than modelled using the other
@@
       also differ significantly between model runs, and vary between
-      3.54 and~8.71\,\unit{m} sea level equivalents (m\,s.l.e.;
+      3.66 and~8.71\,\unit{m} sea level equivalents (m\,s.l.e.;
       Table~\ref{tab:extrema}).
@@
       minima spread over
-      a~wide range between 1.54 and 2.94\,\unit{m}\,s.l.e.
+      a~wide range between 1.54 and 2.81\,\unit{m}\,s.l.e.
       (Table~\ref{tab:extrema}).
@@
       palaeo-temperature proxy records (Table~\ref{tab:records}), modelled
-      maximum sea level potentials also fall within a~tight range of 8.39 to
-      9.07\,\unit{m}\,s.l.e. (Table~\ref{tab:extrema}).
+      maximum sea level potentials also fall within a~tight range of 8.56 to
+      8.75\,\unit{m}\,s.l.e. (Table~\ref{tab:extrema}).
 
@@
       obtained during MIS~4 (62.2--56.9\,\unit{ka}) and the second during
-      MIS~2 (23.1--16.8\,\unit{ka}; Figs.~\ref{fig:lr_ts},
+      MIS~2 (23.2--16.9\,\unit{ka}; Figs.~\ref{fig:lr_ts},
       \ref{fig:lr_maps}; Table~\ref{tab:extrema}). These events broadly
@@
       maxima during MIS~4 (62.2--56.9\,\unit{ka},
-      3.5--8.7\,\unit{m}\,s.l.e.) and MIS~2 (23.1--16.8\,\unit{ka},
-      8.4--9.1\,\unit{m}\,s.l.e.), two periods corresponding to documented
+      3.6--8.7\,\unit{m}\,s.l.e.) and MIS~2 (23.2--16.9\,\unit{ka},
+      8.6--8.8\,\unit{m}\,s.l.e.), two periods corresponding to documented
       extensive glaciations. This result is independent of the

### 2016-01-25 cordillera-cycle.tex
Rescale param table to fit new copernicus layout.
Last small changes before sending to co-authors.

@@
       The simulations presented here were run using the Parallel Ice Sheet
-      Model (PISM, development version~8ff7cbe), an open source, finite
+      Model (PISM, version~0.7.2), an open source, finite
       difference, shallow ice sheet model \citep{PISM-authors.2015}. The
@@
       PISM. We thank Shawn Marshall, Alexander Jarosch and Andrew Stumpf for
-      their constructive and complementary reviews on the first manuscript.
+      their constructive and complementary reviews.
       This work was supported by the Swedish Research Council~(VR)
@@
 \label{tab:params}
-\scalebox{.85}[.85]
+\scalebox{.75}[.75]
 {\begin{tabular}{llrll}

### 2016-01-25 figures/locmap.py

@@ -115,7 +115,7 @@ def add_names():
 
     # add names of intermontane plateaus
     ax.text(-127, 60, 'Liard\nLowland', **txtkwa)
-    ax.text(-122, 52.5, 'Interior\nPlateau', **txtkwa)
+    ax.text(-122.5, 52, 'Interior\nPlateau', **txtkwa)
     #ax.text(-124, 54, 'Fraser\nPlateau', **txtkwa)
     #ax.text(-123, 48, 'Puget\nLowland', **txtkwa)
 

### 2016-01-25 figures/sens_plot_ntil.py

@@ -86,7 +86,7 @@ if __name__ == '__main__':
     grid[0].legend(loc='upper right')
 
     # log scale on bottom axes
-    grid[1].set_xlabel('till content W (m)')
+    grid[1].set_xlabel('till water content W (m)')
     grid[1].set_ylim(2e-1, 2e2)
     grid[1].set_yscale('log')
 

### 2016-01-26 cordillera-cycle.tex
Update sensitivity tests to 0.05 K simulations.

@@
       the Skeena Mountains during this stage shows a~significant sensitivity
-      to ice rheology of 31\,\unit{\%}, and an even more important
-      sensitivity to basal sliding parameters of 62\,\unit{\%}
+      to ice rheology of 25\,\unit{\%}, and an even more important
+      sensitivity to basal sliding parameters of 57\,\unit{\%}
       (Table~\ref{tab:sens_extrema}).
@@
       relevant ice volume
-      during the MIS~2 (LGM) varies by 30\,\unit{\%} between the two
-      parametrisations of ice rheology used, and by 21\,\unit{\%} between
+      during the MIS~2 (LGM) varies by 25\,\unit{\%} between the two
+      parametrisations of ice rheology used, and by 14\,\unit{\%} between
       the two parametrisations of basal sliding used. The differences in
@@
       sensitivity study on the lower-resolution GRIP simulation
-      (Sect.~\ref{sec:sens}) yielded relative errors of 30\,\unit{\%}
-      regarding ice rheological parameters and 21\,\unit{\%} regarding basal
+      (Sect.~\ref{sec:sens}) yielded relative errors of 25\,\unit{\%}
+      regarding ice rheological parameters and 14\,\unit{\%} regarding basal
       sliding parameters (Fig.~\ref{fig:sens_ts};
@@
       values tested in our sensitivity study (Sect.~\ref{sec:sens}), which
-      yielded relative errors of 17\,\unit{\%} to ice rheological parameters
-      and 21\,\unit{\%} to basal sliding parameters
+      yielded relative errors of 12\,\unit{\%} to ice rheological parameters
+      and 19\,\unit{\%} to basal sliding parameters
       (Fig.~\ref{fig:sens_ts}; Table~\ref{tab:sens_extrema}).
@@
       during MIS~3 is highly sensitive to ice flow parameters, with relative
-      errors of 57\,\unit{\%} to the range of ice rheological parameters
-      tested and 120\,\unit{\%} to the range of basal sliding parameters
+      errors of 45\,\unit{\%} to the range of ice rheological parameters
+      tested and 109\,\unit{\%} to the range of basal sliding parameters
       tested (Fig.~\ref{fig:sens_ts}; Table~\ref{tab:sens_extrema}).
@@
                  & $  115\times 10^3$
-                 & 5 & 0.02 & 2 & 0.40 & 6.6 \\
+                 & 5 & 0.02 & 2 & 0.41 & 6.65 \\
     Hard ice$^2$ & $2.847\times 10^{-13}$
@@
                  & $  115\times 10^3$
-                 & 1 & 0.02 & 2 & 0.37 & 6.0 \\
+                 & 1 & 0.02 & 2 & 0.36 & 5.95 \\
     \cline{1-10}
@@
                  & $  139\times 10^3$
-                 & 1 & 0.01 & 1 & 0.40 & 6.5 \\
+                 & 1 & 0.01 & 1 & 0.40 & 6.55 \\
     Hard bed     & $ 3.61\times 10^{-13}$
@@
                  & $  139\times 10^3$
-                 & 1 & 0.05 & 5 & 0.36 & 5.9 \\
+                 & 1 & 0.05 & 5 & 0.36 & 5.85 \\
     \bottomhline
@@
     \cline{1-10}
-    Soft ice &  58.89 &  49.97 &  21.57
-             &   1.96 &   0.54 &   2.08
-             &   6.58 &   1.03 &   6.88 \\
-    Hard ice &  57.32 &  42.90 &  19.14
-             &   1.90 &   0.75 &   2.12
-             &   7.83 &   1.91 &   9.46 \\
+    Soft ice &  58.89 &  49.95 &  21.57
+             &   1.99 &   0.54 &   2.11
+             &   6.69 &   1.04 &   7.09 \\
+    Hard ice &  57.31 &  42.91 &  19.14
+             &   1.85 &   0.71 &   2.09
+             &   7.56 &   1.74 &   9.22 \\
     R. diff. &    3\,\unit{\%} &   16\,\unit{\%} &   13\,\unit{\%}
-             &    3\,\unit{\%} &   31\,\unit{\%} &    2\,\unit{\%}
-             &   17\,\unit{\%} &   57\,\unit{\%} &   30\,\unit{\%} \\
+             &    7\,\unit{\%} &   25\,\unit{\%} &    1\,\unit{\%}
+             &   12\,\unit{\%} &   45\,\unit{\%} &   25\,\unit{\%} \\
     \cline{1-10}
-    Soft bed &  58.90 &  49.21 &  19.53
-             &   1.88 &   0.55 &   2.05
-             &   6.46 &   1.03 &   7.52 \\
-    Hard bed &  57.31 &  42.91 &  19.14
-             &   1.93 &   0.96 &   2.13
-             &   7.99 &   2.89 &   9.31 \\
-    R. diff. &    3\,\unit{\%} &   15\,\unit{\%} &    2\,\unit{\%}
-             &    3\,\unit{\%} &   62\,\unit{\%} &    4\,\unit{\%}
-             &   21\,\unit{\%} &  120\,\unit{\%} &   21\,\unit{\%} \\
+    Soft bed &  58.90 &  49.70 &  21.57
+             &   1.88 &   0.55 &   2.12
+             &   6.42 &   1.08 &   7.89 \\
+    Hard bed &  57.32 &  42.95 &  19.14
+             &   1.90 &   0.93 &   2.10
+             &   7.84 &   2.77 &   9.10 \\
+    R. diff. &    3\,\unit{\%} &   16\,\unit{\%} &   13\,\unit{\%}
+             &    1\,\unit{\%} &   57\,\unit{\%} &    1\,\unit{\%}
+             &   19\,\unit{\%} &  109\,\unit{\%} &   14\,\unit{\%} \\
     \bottomhline

### 2016-01-26 figures/util/sens.py

@@ -9,5 +9,5 @@ configs = ['ccyc4+till1545',
            'ccyc4+d010+wmax01m+till1545',
            'ccyc4+d050+wmax05m+till1545']
 labels = ['Default', 'Soft ice', 'Hard ice', 'Soft bed', 'Hard bed']
-offsets = [6.2, 6.6, 6.0, 6.5, 5.9]
+offsets = [6.2, 6.65, 5.95, 6.55, 5.85]
 colors = ['0.25', '#cab2d6', '#6a3d9a', '#fdbf6f', '#ff7f00']

### 2016-01-29 cordillera-cycle.tex
Add correction from Arjen.
Add corrections from Martin.

@@
       such as those available for the Fennoscandian and Laurentide ice
-      sheets \citep{Boulton.etal.2001, Dyke.Prest.1987, Dyke.etal.2003,
+      sheets \citep{Dyke.Prest.1987, Boulton.etal.2001, Dyke.etal.2003,
       Kleman.etal.1997, Kleman.etal.2010, Stroeven.etal.inreview}. Here, we
@@
       unsaturation index records from Northwest Pacific ocean sediment cores
-      \citep{Herbert.etal.2001},
+      \citep{Herbert.etal.2001}.
 
@@
       to geological evidence and discuss the timing and extent of glaciation
-      and the patterns of deglaciation, based on which we use the
-      applicability of different records to modelling the history of the
-      Cordilleran ice sheet (Sect.~\ref{sec:discussion}).
+      and the patterns of deglaciation (Sect.~\ref{sec:discussion}).
 
@@
       Ice shelf calving is calculated based on ice thickness and principle
-      strain rates (Sect.~\ref{sec:sliding}). Surface mass balance is
+      strain rates (Sect.~\ref{sec:calving}). Surface mass balance is
       computed using a~positive degree-day (PDD) model
@@
   \item Our \emph{default} configuration used in the control run of the
-    sensitivity study and all and all other simulations in the paper include
+    sensitivity study and all other simulations in the paper include
     $\delta=0.02$ and $W_{\text{max}}=2$\,\unit{m} as in
@@
       23.2 and 16.9\,\unit{ka} during MIS~2 (Fig.~\ref{fig:lr_ts}, lower
-      panel). An sea-level relevant ice volume minimum is consistently
+      panel). A sea-level relevant ice volume minimum is consistently
       reached between 55.9 and
@@
 
-      The westerly position of the modelled ice divide and western ice sheet
-      margin may also reflect the fact that our model assumes fixed
-      modern-day spatial patterns for temperature and precipitation and thus
-      does not include feedback mechanisms between ice sheet topography and
-      the regional climate. Regarding temperature, it is reasonable to think
+      The easterly position of the modelled ice divide and eastern ice sheet
+      margin may also reflect the fact that our model does not include
+      feedback mechanisms between ice sheet topography and the regional
+      climate since it assumes fixed modern-day spatial patterns for
+      temperature and precipitation. In fact, it is reasonable to think
       that the cooling was greater inland than near the coast, prohibiting
-      melt at the western margin. However, our simulations already produce
+      melt at the eastern margin. However, our simulations already produce
       an excess of ice inland. Including such a~temperature continentality
@@
       A~notable exception to the transient character of the maximum extent
-      of Cordilleran ice sheet is the northern slope of the Alaska Range,
+      of the Cordilleran ice sheet is the northern slope of the Alaska Range,
       where modelled glaciers are confined to its foothills during the
@@
       patterns emanating from the Skeena Mountains and running over the the
-      central Coast Mountains as are compatible with geological evidence
+      central Coast Mountains are compatible with geological evidence
       from this region \citep[Fig.~12]{Stumpf.etal.2000}.

### 2016-01-31 reviews/response-editor.tex
Resp. editor: list changes made.

@@
 
+\bigskip
+
+Changes not related to the reviews:
+
+\begin{itemize}
+
+    \item We have updated all simulations to a more recent and stable version
+    of PISM (0.7.2).
+
+    \item We have rescaled palaeo-temperature proxy records with a higher
+    precision than in the discussion version.
+
+    \item We have clarified that given ice volumes concern only sea-level
+    relevant ice and areas only grounded ice.
+
+\end{itemize}
+
+Changes following the review by S.~J.~Marshall:
+
+\begin{itemize}
+
+    \item \textbf{Sect.~1 (Introduction):}
+    We have added references regarding geological evidence for late-glacial
+    readvances.
+
+    \item \textbf{Sect.~1 (Introduction):}
+    We have added a short discussion of more regional palaeo-temperature proxy
+    records which we deemed less adequate to force the model in order to
+    justify the choice of less regional proxy records used in our study.
+
+    \item \textbf{Sect.~2.2 (Ice rheology):}
+    We have clarified the advantage and the main limitation of the hybrid SIA
+    and SSA modelling approach.
+
+    \item \textbf{Sect.~5.1.2 (Ice configuration during MIS 2):}
+    In the light of modelled Last Glacial Maximum ice sheet extent, we have
+    discussed further the limitations of forcing the model with fixed
+    present-day temperature and precipitations patterns.
+
+    \item \textbf{Sect.~6 (Conclusions):}
+    We highlighted that although the GRIP and EPICA forcing yield best results
+    in our study, more regional palaeo-climate reconstructions are needed.
+
+    \item \textbf{Various places:}
+    We have applied grammar corrections.
+
+\end{itemize}
+
+Changes following the review by A.~H.~Jarosch:
+
+\begin{itemize}
+
+    \item \textbf{Sect.~2.1 (Model overview):}
+    We have clarified that ETOPO1 data were bilinearly interpolated to the
+    model grids.
+
+    \item \textbf{Sect.~2.1 (Model overview):}
+    We have explained the new subdivisions of Sect.~2 (Model setup).
+
+    \item \textbf{Sect.~2.2 (Ice rheology):}
+    In this new subsection, we have added much detail on the model physics and
+    detail our choice of parameter for the sensitivity study to ice rheological
+    parameters.
+
+    \item \textbf{Sect.~2.2 (Ice rheology):}
+    We removed the formulation ``SSA as a sliding law for the SIA'' which could
+    induce the reader in error.
+
+    \item \textbf{Sect.~2.2 (Ice rheology):}
+    We have made clear that temperatures are computed in the bedrock to a depth
+    of 3\,km below the ice sheet base where geothermal heat flux is applied.
+
+    \item \textbf{Sect.~2.3 (Basal sliding):}
+    In this new subsection, we have added much detail on the model physics and
+    detail our choice of parameter for the sensitivity study to basal sliding
+    parameters.
+
+    \item \textbf{Sect.~2.6 (Climate forcing):}
+    We have clarified that NARR temperature and precipitation fields were
+    bilinearly interpolated to the model grids and admitted their limitations
+    in reproducing steep climatic gradients characteristic of the study area.
+
+    \item \textbf{Sect.~4 (Sensitivity to ice flow parameters):}
+    In this new section we present results of the new sensitivity study to ice
+    rheological and basal sliding parameters
+
+    \item \textbf{Sect.~5.1.2 (Ice configuration during MIS 2):}
+    We have recognized the limitations of the NARR with simulating orographic
+    processes in some areas of steep topography and acknowledged that using
+    downscaling methods could indeed reduce some of the inconsistence between
+    model results and geological reconstructions.
+
+    \item \textbf{Sect.~5.1.2 (Ice configuration during MIS 2):}
+    We have balanced the final statement concerning modelled ice volume maxima
+    using the uncertainty estimates obtained in the new Sect.~4.
+
+    \item \textbf{Sect.~5.1.3 (Ice configuration during MIS 4):}
+    We have balanced the final statement concerning modelled ice volume maxima
+    using the uncertainty estimates obtained in the new Sect.~4.
+
+    \item \textbf{Sect.~5.2.2 (Major ice-dispersal centres):}
+    In the light of modelled accumulation areas, we have shown that present-day
+    ice volumes contained in the ETOPO1 basal topography data does not affect
+    the conclusions of our study.
+
+    \item \textbf{Sect.~5.2.2 (Major ice-dispersal centres):}
+    We have balanced the final statement concerning modelled ice volume minima
+    using the uncertainty estimates obtained in the new Sect.~4.
+
+    \item \textbf{Table~1:}
+    The default parameter table was extended to include more parameter values
+    and add references to their source.
+
+    \item \textbf{Table~3:}
+    This new table contains ice rheological and basal sliding parameter values
+    used in the sensitivity study.
+
+    \item \textbf{Table~5:}
+    This new table contains modelled ice volume and area extremes corresponding
+    to the results of the sensitivity study.
+
+    \item \textbf{Fig.~2:}
+    This new figure illustrates the different parametrisations of ice rheology
+    used in the sensitivity study.
+
+    \item \textbf{Fig.~3:}
+    This new figure illustrates the different parametrisations of basal sliding
+    used in the sensitivity study.
+
+    \item \textbf{Fig.~5:}
+    This new figure contains modelled sea-level relevant ice volume time series
+    resulting from the sensitivity study.
+
+    \item \textbf{Various places:}
+    We have applied grammar corrections.
+
+\end{itemize}
+
+Changes following the review by A.~Stumpf:
+
+\begin{itemize}
+
+    \item \textbf{Abstract:}
+    We have reworked the first two sentences to clarify that the Cordilleran
+    ice sheet is less understood than its counterparts not by lack of previous
+    studies but due to its complexity.
+
+    \item \textbf{Fig.~1:}
+    We have enlarged the location map to include full geographic names instead
+    of abbreviations.
+
+    \item \textbf{Figs.~4, 6, 8--12, 14--16:}
+    We have added rivers, lakes and graticules on all model output maps.
+
+    \item \textbf{Sect.~5.2.3 (Erosional imprint on the landscape):}
+    We have added a reference to Stumpf et al. (2000).
+
+    \item \textbf{Sect.~5.3.3 (Deglacial flow directions):}
+    We have added a reference to Stumpf et al. (2000).
+
+    \item \textbf{Sect.~5.3.3 (Deglacial flow directions):}
+    We have added a reference to Ferbey et al. (2013).
+
+\end{itemize}
+
 \end{document}

### 2016-01-31 reviews/response-editor.tex
Resp. editor: rework introductory words.

@@
 Dear Hilmar Gudmundsson,
-(cc co-authors and Copernicus publications)
 
-As you probably noticed, Copernicus stopped peer review of our manuscript
-"Numerical simulations of the Cordilleran ice sheet through the last glacial
-cycle" when we were nearly ready to submit our responses. Since the deadline
-coincided with new year celebrations, I could not obtain feedback from my
-co-authors on time to finalize the responses.
+We greatly apologize for delays accumulated during the peer review of our
+manuscript. To some extent, the delay has been caused by our wish to address
+the reviews as well as possible. In addition to a change of the first author's
+working environment, setting up and running the sensitivity study asked by
+Alexander Jarosch on a different supercomputer, using a new allocation that was
+meant for a different project and became effective only in October, took us
+considerable time.
 
-We greatly apologize for this additional delay and ask for your assistance to
-continue with the peer review that has required a large time investment from
-our side and from the side of the reviewers. To some extent, the delay has been
-caused by our wish to address the reviews as well as possible. In addition to a
-change of my working environment, setting up and running the sensitivity study
-asked by Alexander Jarosch on a different supercomputer, using a new allocation
-that was meant for a different project and became effective only in October,
-took us considerable time.
+We believe that we have addressed all points raised by the reviews and hereby
+submit our revised manuscript to The Cryosphere. Please find hereafter a
+summary of changes made to the manuscript since its first publication in The
+Cryosphere Discussion, and a marked-up file showing all changes made.
 
-I contacted Svenja Lange prior to the submission deadline to inquire after the
-possibility of a short-term extension but later realized that Copernicus
-offices were closed for the new year celebration until yesterday. I have tried
-to contact Natascha T\"{o}pfer yesterday (as suggested in the following email from
-Copernicus) but then realised on the Cryosphere website that she is on a leave.
-Is there a chance that the on-line submission system could be re-opened for us
-to submit our responses?
+With best regards,
 
-Thank you very much in advance for your understanding,
-
-Best regards,
-
-Julien.
+Julien Seguinot on behalf of all authors.
 

### 2016-02-01 cordillera-cycle.tex
Add corrections from Irina.

@@
       precipitation grids derived from a~present-day atmospheric reanalysis
-      \citep{Mesinger.etal.2006} that was previously tested against
+      \citep{Mesinger.etal.2006}. These were previously tested against
       observational data and shown to best reproduce the steep precipitation
@@
       follows temperature and water-content dependent creep
-      (Sect.~\ref{sec:icedyn}). Basal sliding follows
+      (Sect.~\ref{sec:icedyn}). Default parameter values are given in
+      Table~\ref{tab:params}. Basal sliding follows
       a~pseudo-plastic law where the yield stress accounts for till
       deconsolidation under high water pressure (Sect.~\ref{sec:sliding}).
-      Ice shelf calving is calculated based on ice thickness and principle
-      strain rates (Sect.~\ref{sec:calving}). Surface mass balance is
+      Ice shelf calving is simulated using an ice thickness threshold and
+      principle components of the strain rate tensor
+      (Sect.~\ref{sec:calving}). Surface mass balance is
       computed using a~positive degree-day (PDD) model
-      (Sect.~\ref{sec:surface}). Default parameter values are given in
-      Table~\ref{tab:params}. Climate forcing is provided by a~monthly
+      (Sect.~\ref{sec:surface}). Climate forcing is provided by a~monthly
       climatology averaged from 1979 to 2000 from the North American
@@
       Greve.1997}. Finally, $E$ is a~non-dimensional enhancement factor
-      which can take different values, $E_{\text{SIA}}$, in the Shallow Ice
-      Approximation (SIA) and $E_{\text{SSA}}$, in the Shallow Shelf
-      Approximation (SSA).
+      which can take different values, $E_{\text{SIA}}$, in the SIA and
+      $E_{\text{SSA}}$, in the SSA.
 
@@
   \item Our \emph{default} configuration used in the control run of the
-    sensitivity study and all other simulations in the paper include
+    sensitivity study and all other simulations in the paper includes
     rheological parameters,
@@
     \citet[Eqn.~5]{Bueler.Brown.2009}, and $E_{\text{SIA}}=1$.
-  \item Our \emph{hard ice} configuration include rheological parameters,
+  \item Our \emph{hard ice} configuration includes rheological parameters,
     $A_{\mathrm{c}}$, $A_{\mathrm{w}}$, $Q_{\mathrm{c}}$ and
@@
     a~stiffer rheology than that used in the control run.
-  \item Our \emph{soft ice} configuration include rheological parameters
+  \item Our \emph{soft ice} configuration includes rheological parameters
     from \citet{Cuffey.Paterson.2010}, and $E_{\text{SIA}}=5$, the
@@
 
-      In all our sumlations, we set constant the pseudo-plastic sliding
+      In all our simulations, we set constant the pseudo-plastic sliding
       exponent, $q=0.25$, and the threshold velocity,
@@
       between the effective and overburden pressures, $\delta$, and the
-      maximum water height in the till, $W_{\text{max}}$, as follow.
+      maximum water height in the till, $W_{\text{max}}$, as follows.
 %
@@
   \item Our \emph{default} configuration used in the control run of the
-    sensitivity study and all other simulations in the paper include
+    sensitivity study and all other simulations in the paper includes
     $\delta=0.02$ and $W_{\text{max}}=2$\,\unit{m} as in

### 2016-02-02 cordillera-cycle.tex
Add more corrections from Irina.

@@
       would thus cause an even greater mismatch between the model results
-      and the geologically reconstructed Last Glacial Maximum (LGM) ice
-      margins.
-
-      Consequently, the mismatch we observe between the modelled and
-      reconstructed LGM ice margins let us think that the assumption of
-      fixed modern-day precipitation patterns is more critical than the
-      assumption of fixed modern-day temperature patterns.
+      and the geologically reconstructed ice margins during the LGM.
+
+      Consequently, the mismatch between the modelled and
+      reconstructed LGM ice margins is likely due to the assumption of the
+      fixed modern-day precipitation patterns rather than the
+      assumption of the fixed modern-day temperature patterns.
       Firstly, during the build-up
@@
       topography could have resulted in an inflow of potentially warmer air
-      over the eastern flank of the ice sheet, counterbalancing the potential
+      over the eastern flank of the ice sheet, thereby counterbalancing the potential
       continentality gradient discussed above through increasing melt along the
@@
       8.56\,\unit{m}\,s.l.e. (17.4\,\unit{ka}) in the EPICA simulation.
-      However, these numers are subject to uncertainties on physics
-      embedded in the model. The range of parameter values tested in our
-      sensitivity study on the lower-resolution GRIP simulation
+      However, these numbers are subject to significant uncertainties in the
+      ice flow parameters embedded in the reference model setup. The range
+      of parameter values tested in our sensitivity study
       (Sect.~\ref{sec:sens}) yielded relative errors of 25\,\unit{\%}
-      regarding ice rheological parameters and 14\,\unit{\%} regarding basal
-      sliding parameters (Fig.~\ref{fig:sens_ts};
+      with regard to ice rheological parameters and 14\,\unit{\%} regarding
+      basal sliding parameters (Fig.~\ref{fig:sens_ts};
       Table~\ref{tab:sens_extrema}).
@@
       ice volumes.
-      These estimates appear little sensitive to the range of parameter
+      These estimates show little sensitivity to the range of parameter
       values tested in our sensitivity study (Sect.~\ref{sec:sens}), which
@@
 
-      The location of modelled ice-dispersal centres is potentially biased
-      by present day ice volumes contained in the ETOPO1 basal topography
-      data. Although a~reconstruction exist for the thickness of western
-      Canadian glaciers south of 60{\degree}N \citep{Clarke.etal.2013}, the
+      The location of the modelled ice-dispersal centres is potentially biased
+      by the present day ice volumes contained in the ETOPO1 basal topography
+      data. Although a~recent reconstruction of the thickness of the western
+      Canadian glaciers south of 60{\degree}N \citep{Clarke.etal.2013} shows
+      that the
       most problematic part of the model domain in this respect is by far
       that of the Wrangell and Saint Elias mountains, where ice thicknesses
-      up to 1200\,m have been measured by low-frequency radar
-      \citep{Rignot.etal.2013}. In this area, located over the USA Canada
-      border just north of 60{\degree}N, temperate ice, surge dynamics and
-      deep subglacial depressions in the icefield interior pose
-      a~fundamental challenge to reconstructing basal topography for the
+      of up to 1200\,m have been measured by a low-frequency radar
+      \citep{Rignot.etal.2013}. In this area, located over the USA-Canada
+      border just north of 60{\degree}N, temperate ice, glacier surging
+      dynamics and deep subglacial depressions in the icefield interior pose
+      a~fundamental challenge to the reconstruction basal topography for the
       entire ice cap \citep{Rignot.etal.2013}. Although this causes our
       simulations to overestimate ice surface topography in this region, we
-      are not aware of bed topography reconstructions that could be used to
-      force the ice sheet model. However, this drawback seem to have
+      are not aware of bed topography reconstructions.
+      However, this drawback seem to have
       little effect on modelled Cordilleran ice sheet dynamics. In fact, the
@@
       ice cap does not appear to be a~major feed to the Cordilleran ice
-      sheet (Fig.~\ref{fig:hr_geom_duration}, WSEM). Moreover, with this
+      sheet (Fig.~\ref{fig:hr_geom_duration}, WSEM). With this
       exception of the Wrangell and Saint Elias mountains ice field,
@@
       However, these numbers should be considered with caution as our
-      sensitivity study (Sect.~\ref{sec:sens}) shows that minimum ice volume
+      sensitivity study (Sect.~\ref{sec:sens}) shows that the minimum ice volume
       during MIS~3 is highly sensitive to ice flow parameters, with relative
@@
       Inside the retreating western margin, modelled south-westward flow
-      patterns emanating from the Skeena Mountains and running over the the
+      patterns emanating from the Skeena Mountains and running over the
       central Coast Mountains are compatible with geological evidence
@@
       indicate a~substantial eastwards flow component of the Cordilleran ice
-      sheet \citep{Prest.etal.1968, Kleman.etal.2010, Ferbey.etal.2013}
-      not present in any of the two simulations
+      sheet \citep{Prest.etal.1968, Kleman.etal.2010, Ferbey.etal.2013},
+      which is not present in any of the two simulations
       (Fig.~\ref{fig:hr_geom_lastflow}). One explanation for the
@@
       affected by the weakening of the California current during the LGM.
-      Nevertheless, the fact that Cordilleran ice sheet dynamics are here
-      modelled in best agreement to geological reconstructions using
-      palaeo-temperature records from the distant Greenland and Antarctic
-      ice sheets, and the significant differences remaining between our two
-      preffered simulations highlight the need for more regional
-      palaeo-climate reconstructions of the last glacial cycle in and around
-      the North American Cordillera.
+      Nevertheless, the fact that the use of paleo-temperature
+      reconstructions from remote sites in Greenland and Antarctica produces
+      the best agreement between modelled Cordilleran ice sheet dynamics and
+      available geological evidence, and the significant differences
+      remaining between the two preferred simulations highlight the need for
+      more regional palaeo-climate reconstructions of the last glacial cycle
+      in and around the North American Cordillera.
 

### 2016-02-02 reviews/response-editor.tex
Resp. Editor: final polish.

@@
 \input{response-header}
+\linespread{1.0}
 
@@
 
+\bigskip
+
 Changes following the review by S.~J.~Marshall:
@@
 
+\bigskip
+
 Changes following the review by A.~H.~Jarosch:
@@
     In this new subsection, we have added much detail on the model physics and
-    detail our choice of parameter for the sensitivity study to ice rheological
+    on our choice of parameter for the sensitivity study to ice rheological
     parameters.
@@
     We removed the formulation ``SSA as a sliding law for the SIA'' which could
-    induce the reader in error.
+    induce in error.
 
@@
     In this new subsection, we have added much detail on the model physics and
-    detail our choice of parameter for the sensitivity study to basal sliding
+    on our choice of parameter for the sensitivity study to basal sliding
     parameters.
@@
     \item \textbf{Sect.~4 (Sensitivity to ice flow parameters):}
-    In this new section we present results of the new sensitivity study to ice
+    In this new section, we present results of the new sensitivity study to ice
     rheological and basal sliding parameters
@@
     We have balanced the final statement concerning modelled ice volume maxima
-    using the uncertainty estimates obtained in the new Sect.~4.
+    using the uncertainty estimates obtained in the new Sect.~4 (Sensitivity to
+    ice flow parameters).
 
@@
     In the light of modelled accumulation areas, we have shown that present-day
-    ice volumes contained in the ETOPO1 basal topography data does not affect
-    the conclusions of our study.
+    ice volumes contained in the ETOPO1 basal topography data likely does not
+    affect the conclusions of our study.
 
@@
     We have balanced the final statement concerning modelled ice volume minima
-    using the uncertainty estimates obtained in the new Sect.~4.
+    using the uncertainty estimates obtained in the new Sect.~4 (Sensitivity to
+    ice flow parameters).
 
@@
     The default parameter table was extended to include more parameter values
-    and add references to their source.
+    and references to their source.
 
@@
     This new table contains modelled ice volume and area extremes corresponding
-    to the results of the sensitivity study.
+    to the results of the sensitivity study (similarly to Table~4).
 
@@
 
-    \item \textbf{Fig.~5:}
+    \item \textbf{Fig.~7:}
     This new figure contains modelled sea-level relevant ice volume time series
-    resulting from the sensitivity study.
+    resulting from the sensitivity study (similarly to Fig.~5).
 
@@
 \end{itemize}
+
+\bigskip
 
@@
     \item \textbf{Sect.~5.2.3 (Erosional imprint on the landscape):}
-    We have added a reference to Stumpf et al. (2000).
+    We have added a reference to Stumpf et~al. (2000).
 
     \item \textbf{Sect.~5.3.3 (Deglacial flow directions):}
-    We have added a reference to Stumpf et al. (2000).
-
-    \item \textbf{Sect.~5.3.3 (Deglacial flow directions):}
-    We have added a reference to Ferbey et al. (2013).
+    We have added references to Stumpf et~al. (2000) and to Ferbey et~al. (2013).
 

### 2016-02-02 cordillera-cycle.tex
Add even more corrections from Irina.

@@
       23.2 and 16.9\,\unit{ka} during MIS~2 (Fig.~\ref{fig:lr_ts}, lower
-      panel). A sea-level relevant ice volume minimum is consistently
+      panel). An ice volume minimum is consistently
       reached between 55.9 and
@@
       Simulations forced by the Greenland ice core palaeo-temperature
-      records (GRIP, NGRIP) produce the highest variability in modelled
-      sea-level relevant ice
+      records (GRIP, NGRIP) produce the highest variability in modelled ice
       volume throughout the last glacial cycle. In contrast, simulations
@@
       While simulations driven by the GRIP and the two Antarctic
-      palaeo-temperature records attain a~last sea-level relevant ice volume
-      maximum between
+      palaeo-temperature records attain a~last ice volume maximum between
       19.1 and 16.9\,\unit{ka}, those informed by the NGRIP and the two
@@
 
-      In the MIS~3 ice volume minimum reconstructions, a~central ice cap
+      During MIS~3 ice volume minimum reconstructions, a~central ice cap
       persists over the Skeena Mountains (Fig.~\ref{fig:lr_maps}, middle
@@
       the maximum stage during MIS~2, but also during MIS~4
-      (Table~\ref{tab:sens_extrema}). However, such can not be said of the
-      modelled glaciated area corresponding to minimum stage during
+      (Table~\ref{tab:sens_extrema}). However, this cannot be said of the
+      modelled glaciated area corresponding to the minimum stage during
       MIS~3. In fact, the extent of the remnant ice cap which persists over
       the Skeena Mountains during this stage shows a~significant sensitivity
-      to ice rheology of 25\,\unit{\%}, and an even more important
+      to ice rheology of 25\,\unit{\%}, and an even higher
       sensitivity to basal sliding parameters of 57\,\unit{\%}
@@
 
-      The modelled sea-level potentials show more variability than
+      The modelled sea-level potentials show a stronger variability than
       the modelled glaciated areas (Table~\ref{tab:sens_extrema},
@@
       till both result in a~thinner ice sheet, while harder ice and stronger
-      till result in a~thicker ice sheet. For instance, peak sea-level
-      relevant ice volume
-      during the MIS~2 (LGM) varies by 25\,\unit{\%} between the two
+      till result in a~thicker ice sheet. For instance, the modelled peak
+      sea-level relevant ice volume
+      during MIS~2 (LGM) varies by 25\,\unit{\%} between the two
       parametrisations of ice rheology used, and by 14\,\unit{\%} between
       the two parametrisations of basal sliding used. The differences in
-      ice volume are greatest during the MIS~3
+      ice volume are greatest during MIS~3
       (Table~\ref{tab:sens_extrema}, Fig.~\ref{fig:sens_ts}) where both the
@@
       a~more in-depth comparison with geological evidence for the timing of
-      maximum Cordilleran ice sheet extent. In the Puget Lowland
+      the maximum Cordilleran ice sheet extent. In the Puget Lowland
       (Fig.~\ref{fig:locmap}), the LGM advance of the southern Cordilleran
@@
       constraints from the southern margin. However, cosmogenic exposure
-      dating places the timing of maximum northern ice sheet margin extent
-      during the McConnell
+      dating places the timing of the maximum northern ice sheet margin
+      extent during the McConnell
       glaciation close to 17\,\unit{\chem{^{10}Be}\,ka}
@@
       Such deviation from the geological inferences could reflect the fact
-      the NARR has difficulties with simulating orographic processes in some
-      areas of steep topography \citep{Jarosch.etal.2012}. In a~previous
-      study \citep{Seguinot.etal.2014}, we have evaluated the performance of
-      NARR in forcing constant-climate simulations of the Cordilleran ice
-      sheet against that of an observation-based data set
+      that the NARR has difficulties with simulating orographic processes in
+      some areas of steep topography \citep{Jarosch.etal.2012}. In our
+      previous study \citep{Seguinot.etal.2014}, we have evaluated an
+      applicability of the climate forcing bilinearly interpolated from the
+      NARR to constant-climate simulations of the Cordilleran ice sheet
+      during the LGM against that of an observation-based data set
       \citep[WorldClim,][]{Hijmans.etal.2005}. Indeed, the use of NARR in
       these simulations produced slightly different patterns of glaciation
-      relative to WorldClim, including more extensive ice cover on the
+      relative to WorldClim, including a more extensive ice cover on the
       Columbia and Rocky mountains \citep[Figs.~6--7]{Seguinot.etal.2014}.
       Our simulations have shown that these differences are mainly caused by
-      disparities in the precipitation fields of the two data sets
-      \citep[Figs.13--14]{Seguinot.etal.2014}. Over the southern part of our
-      model domain, it was shown that the implementation of temperature and
-      precipitation downscaling can address these limitations
+      disparities between the precipitation fields of the two data sets
+      \citep[Figs.13--14]{Seguinot.etal.2014}. It has been shown that over
+      the southern part of our model domain that temperature and
+      precipitation downscaling can potentially address these limitations
       \citep[e.g.,][]{Jarosch.etal.2012}. However, extending this
-      downscaling method to the entire model domain used in our study could
-      be challenging, because the northern part of the model domain is
+      downscaling method to the entire model domain used in our study is
+      challenging, because the northern part of the model domain is
       characterized by stronger precipitation gradients (Fig.~\ref{fig:atm})
-      and includes much fewer weather stations than the south
-      \citep{Hijmans.etal.2005}.
-
-      The easterly position of the modelled ice divide and eastern ice sheet
-      margin may also reflect the fact that our model does not include
-      feedback mechanisms between ice sheet topography and the regional
-      climate since it assumes fixed modern-day spatial patterns for
-      temperature and precipitation. In fact, it is reasonable to think
-      that the cooling was greater inland than near the coast, prohibiting
+      and fewer weather stations than the southern part where the previous
+      analysis has been performed \citep{Hijmans.etal.2005}.
+
+      Since the model does not include feedback mechanisms between the ice
+      sheet topography and the regional climate, the modelled easterly
+      positions of the ice divide and eastern ice sheet margin may also be
+      sensitive to the assumption of fixed modern-day spatial patterns of
+      air temperature and precipitation. In fact, it is reasonable
+      to think that the cooling during the last glacial cycle was greater
+      inland than near the coast, prohibiting
       melt at the eastern margin. However, our simulations already produce

### 2016-02-02 cordillera-cycle.tex
Add yet more corrections from Irina...

@@
       topography and bathymetry dataset with a~resolution of 1\,arc-min
-      \citep{Amante.Eakins.2009} to the model grids. Sea level is lowered as
+      \citep{Amante.Eakins.2009} to the model grids. It responds to ice load
+      following a~bedrock deformation model that includes local isostasy,
+      elastic lithosphere flexure and viscous astenosphere deformation in an
+      infinite half-space \citep{Lingle.Clark.1985,Bueler.etal.2007}.
+      A~relatively low viscosity value of $\nu_{\mathrm{m}} =
+      1\times10^{19}$\,\unit{Pa\,s} is used for the astenosphere
+      (Table~\ref{tab:params}) in accordance with the results from regional
+      glacial isostatic adjustment modelling at the northern Cascadia
+      subduction zone \citep{James.etal.2009}. Sea level is lowered as
       a~function of time based on the Spectral Mapping Project
-      \citep[SPECMAP,][]{Imbrie.etal.1989} time scale. Ice deformation
+      \citep[SPECMAP,][]{Imbrie.etal.1989} time scale.
+
+      Ice deformation
       follows temperature and water-content dependent creep
@@
       to some of the most influential ice rheology (Sect~\ref{sec:icedyn})
-      and basal sliding (Sect.~\ref{sec:sliding}) parameters was performed
+      and basal sliding (Sect.~\ref{sec:sliding}) parameters is performed
       (Table~\ref{tab:sens_params}).
@@
       energies, $Q_{\mathrm{c}}$ and $Q_{\mathrm{w}}$, and the SIA
-      enhancement factor, $E_{\text{SIA}}$, as follow.
+      enhancement factor, $E_{\text{SIA}}$, as follows.
 %
@@
       \citet{Cuffey.Paterson.2010} and $E_{\text{SIA}}=2$, the recommended
-      value for Holocene polar ice \citep[p.~77]{Cuffey.Paterson.2010} was
-      performed, but its results were very similar to that of our default
-      run and thus not presented here.
+      value for Holocene polar ice \citep[p.~77]{Cuffey.Paterson.2010} has
+      been performed, but is not presented here, since its results were very
+      similar to that of our default run.
 
@@
       coefficient, $C_{\mathrm{c}}=0.12$, according to the only measurements
-      available to our knowledge, published by \citep{Tulaczyk.etal.2000}.
+      available to our knowledge, published by \citet{Tulaczyk.etal.2000}.
       These fixed parameter values are summarized in Table~\ref{tab:params}.
@@
       a~piecewise-linear function of modern bed elevation, with the lowest
-      value occuring below modern sea level (0\,\unit{m} above sea level,
+      value occuring below the modern sea level (0\,\unit{m} above sea level,
       \unit{m\,a.s.l.}) and the highest value occuring above the generalised
@@
       $\phi=30$\unit{\degree}, corresponding to the average value in
-      \citet[p.~268]{Cuffey.Paterson.2010}, was actually performed, but the
-      induced variability was small and thus not presented here.
+      \citet[p.~268]{Cuffey.Paterson.2010}, has been performed, but is not
+      presented here, since the induced variability was small.
 
@@
       are listed in Table~\ref{tab:sens_params}.
-
-      Finally, the bedrock topography responds to ice load following
-      a~bedrock deformation model that includes local isostasy, elastic
-      lithosphere flexure and viscous astenosphere deformation in an
-      infinite half-space \citep{Lingle.Clark.1985,Bueler.etal.2007}.
-      A~relatively low viscosity value of $\nu_{\mathrm{m}} =
-      1\times10^{19}$\,\unit{Pa\,s} is used for the astenosphere
-      (Table~\ref{tab:params}) in accordance with the results from regional
-      glacial isostatic adjustment modelling at the northern Cascadia
-      subduction zone \citep{James.etal.2009}.
 
@@
       temperature variability (Fig.~\ref{fig:atm}). Although the ability
-      of NARR to reproduce these steep climatic gradients is limited by its
-      spatial resolution of 32\,\unit{km} \citep{Jarosch.etal.2012}, it was
-      tested against observational data in a~previous sensitivity study and
-      identified as yielding modelled Cordilleran ice sheet LGM extent in
-      closer agreement to geological reconstructions than other atmospheric
-      reanalyses \citep{Seguinot.etal.2014}.
+      of the NARR to reproduce the steep climatic gradients is limited by its
+      spatial resolution of 32\,\unit{km} \citep{Jarosch.etal.2012}, it has
+      been tested against observational data in our previous sensitivity
+      study and identified as yielding a closer fit between the modelled LGM
+      extent of the Cordilleran ice sheet and the geological evidence than
+      other atmospheric reanalyses \citep{Seguinot.etal.2014}.
 
@@
       were provided as such. For each proxy record used and each of the
-      configurations used in the sensitivity tests, palaeo-temperature
+      parameter setup used in the sensitivity tests, palaeo-temperature
       anomalies were scaled linearly (Tables~\ref{tab:records}

### 2016-02-02 cordillera-cycle.tex
One last small correction.

@@
       \citep[Eqs.~7--9 and 15]{Winkelmann.etal.2011}. Although this
-      heuristic imply errors in the transition zone where gravitational
+      heuristic approach implies errors in the transition zone where gravitational
       stresses intervene both in the SIA and SSA velocity computation, this

### 2016-02-19 cordillera-cycle.tex
Add references to Matthias' reconstructions.

@@
       by the present day ice volumes contained in the ETOPO1 basal topography
-      data. Although a~recent reconstruction of the thickness of the western
-      Canadian glaciers south of 60{\degree}N \citep{Clarke.etal.2013} shows
-      that the
-      most problematic part of the model domain in this respect is by far
+      data. The most problematic part of the model domain in this respect is
       that of the Wrangell and Saint Elias mountains, where ice thicknesses
@@
       a~fundamental challenge to the reconstruction basal topography for the
-      entire ice cap \citep{Rignot.etal.2013}. Although this causes our
-      simulations to overestimate ice surface topography in this region, we
-      are not aware of bed topography reconstructions.
+      entire ice cap \citep{Rignot.etal.2013}. Recent ice thickness
+      reconstructions available for all glaciers around the globe
+      \citep{Huss.Farinotti.2012}, for Canadian glaciers south of 60{\degree}N
+      \citep{Clarke.etal.2013}, and for Alaska tidewater glaciers
+      \citep{McNabb.etal.2015} have not been implemented in our simulations.
       However, this drawback seem to have
@@
 
+\bibitem[{Huss and Farinotti(2012)}]{Huss.Farinotti.2012}
+Huss, M. and Farinotti, D.: Distributed ice thickness and volume of all glaciers around the globe, J. Geophys. Res. Earth Surf., 117, F04\,010,
+doi:\href{http://dx.doi.org/10.1029/2012JF002523}{10.1029/2012JF002523}, 2012.
+
 \bibitem[{Imbrie et~al.(1989)Imbrie, McIntyre, and Mix}]{Imbrie.etal.1989}
@@
 doi:\href{http://dx.doi.org/10.5194/tc-5-727-2011}{10.5194/tc-5-727-2011}, 2011.
+
+\bibitem[{McNabb et~al.(2015)McNabb, Hock, and Huss}]{McNabb.etal.2015}
+McNabb, R.~W., Hock, R., and Huss, M.: Variations in Alaska tidewater glacier frontal ablation, 1985--2013, J. Geophys. Res. Earth Surf., 120, 120--136,
+doi:\href{http://dx.doi.org/10.1002/2014JF003276}{10.1002/2014JF003276}, 2015.
 

### 2016-02-22 cordillera-cycle.tex
Add last technical corrections from Jarosch.

@@
       $\vec{\tau}$ the deviatoric stress tensor, and $\tau_{\mathrm{e}}$ the
-      equivalent stress defined by ${\tau_{\mathrm{e}}}^2 = \frac{1}{2}
+      effective stress defined in our case by ${\tau_{\mathrm{e}}}^2 = \frac{1}{2}
       \mathrm{tr}(\vec{\tau}^2)$. The ice softness coefficient, $A$, depends
       on ice temperature, $T$, pressure, $p$, and water content, $\omega$,
-      through a~piece-wise Arrhenius-type law
-      \citep[Eqs.~63--65]{Aschwanden.etal.2012},
+      through a~piece-wise Arrhenius-type law,
 %

### 2016-03-03 figures/sens_dist_vel.py
Change 'yr' to 'a' for consistence.

@@ -56,7 +56,7 @@ for i, ax in enumerate(grid):
 ax.set_ylim(-0.5, 5.5)
 ax.set_xticks(range(1,6))
 ax.set_xticklabels(ut.sens.labels)
-grid[1].set_ylabel(r'log of surface velocity ($m\,yr^{-1}$)')
+grid[1].set_ylabel(r'log of surface velocity ($m\,a^{-1}$)')
 
 # save
 fig.savefig('sens_dist_vel')

### 2016-03-03 figures/util/pl.py

@@ -132,7 +132,7 @@ def fig_hr_maps_mis(mis):
     # add colorbar and return figure
     cb = fig.colorbar(im, cax, extend='both', orientation='horizontal',
                       format='%i', ticks=np.logspace(1, 3.5, 6))
-    cb.set_label(r'surface velocity ($m\,yr^{-1}$)', labelpad=-2.0)
+    cb.set_label(r'surface velocity ($m\,a^{-1}$)', labelpad=-2.0)
     return fig
 
 

### 2016-11-04 reviews/reviews-response.tex

\documentclass[10pt]{article}

\usepackage[T1]{fontenc}
\usepackage[pdftex]{color}
\usepackage{geometry}

\sloppy
\setlength{\parskip}{1.0ex}
\setlength{\parindent}{0em}

% ----------------------------------------------------------------------

\begin{document}
\textbf{Authors' response to the reviews}
\bigskip

\def\msquote#1{\begin{quote}\textit{#1}\end{quote}}

Dear Frank Pattyn,

We believe that we have addressed all points raised by the reviews and hereby submit our revised manuscript to The Cryosphere. Our changes are described in detail in our public responses. We list them here by order of appearance, including references to line numbers in the revised manuscript.

\begin{description}

\item[Throughout the manuscript:]
  Following the comment of Referee~{\#}1, we have adopted negative values for temperature offsets in the revised text and figures.

\item[Sect.~2.1, l.~122--129:]
  We have clarified that the till friction angle parameter is not affected by elevation changes related to isostatic effects:

  \msquote{It is computed once at the beginning of the run [...] Because the distribution of marine sediments is assumed constant throughout the run, the till friction angle $\phi$ is not affected by changes in bed elevation.}

\item[Sect.~2.1, l.~132--137:]
   Following the request of Referee~{\#}2, we have added a paragraph on the implementation of isostatic effects in the model:

  \msquote{Sea level is lowered by 120\,m and basal topography responds to ice load following a bedrock deformation model that includes point-wise isostasy, elastic lithosphere flexure and viscous mantle deformation in a~semi-infinite half-space. It uses a lithosphere density of 3300\,kg\,m$^{-3}$, a flexural rigidity of $5 \times 10^{24}$\,N\,m and a mantle viscosity of $1 \times 10^{21}$\,Pa\,s. Due to the high mantle viscosity, there exists a time lag between ice sheet growth and isostatic bedrock response.}

\item[Sect.~5.1, l.~269--276:]
  Following a question of Referee~{\#}1, we have improved Figs.~8--11 (see below) and added a few sentences to the main text concerning the new figures:

  \msquote{Because much of the disparity between reanalyses and observational temperature data is caused by topographical detail at scales unresolved in the reanalyses, we apply a lapse-rate correction prior to this comparison, using the same lapse rate of 6\,$^\circ$C\,km$^{-1}$ as in the simulations. [...] The spatial distribution of temperature differences between reanalyses and WorldClim data shows partly consistent patterns, such as positive anomalies in the northernmost part of the modelling domain. These may relate to weaknesses of WorldClim data in areas of sparse observational coverage (Fig. 9). However, significant disparity among reanalyses exists, too.}

\item[Sect.~5.4, l.~352--355:]
   Following the comment of Referee~{\#}2, we have emphasized that most of our simulations do not reach steady-state:

\msquote{Most of the simulations presented in this study, and more particularly those that reproduce the LGM ice margin reconstruction more closely (Fig.~15) do not reach a steady-state. Instead, rates of growth remain high throughout the run, and bed elevation does not come close to equilibrium with the ice load.}

\item[Sect.~5.5, l.~375--386:]
  Following the request of Referee~{\#}1, we have included a discussion of potential effects of precipitation changes on model results:

  \msquote{As previously discussed, potential effects of the growing ice-sheet on precipitation changes are not included in our model. These changes likely consisted of a reduction of precipitation in continental regions and in the ice sheet interior, and an increase of precipitation along part of the margin where the presence of ice imposed ascending winds. They could result in a more westerly-centred ice sheet than modelled here, with lower ice thickness in its interior. In addition, the final position of the eastern ice margin is largely controlled by its advance rate through the run. Therefore, the precipitation shadowing effects may have resulted in more similar ice-sheet configurations if they were included in the model. Although using a GCM of intermediate complexity may represent a first step towards including ice sheet feedback on climate, their spatial resolution does not allow for accurate modelling of orographic precipitation changes in a region as mountainous as the North American Cordillera.}

\item[Sect.~5.5, l.~398--403:]
  Following the request of Referee~{\#}2, we have added a discussion of isostatic effects on model response:

\msquote{The bedrock deformation model uses homogeneous lithospheric properties and a semi-infinite mantle of constant viscosity. Thus, it does not represent lateral variations of crustal and mantle properties characteristic of tectonically active margins such as the American Cordillera, nor does it include the influence from the neighbouring Laurentide ice sheet. These limitations should be acknowledged when interpreting ice surface elevation and volume reconstructions presented in this study.}

\item[Fig.~8:]
  Temperature density plots now use bilinear interpolation, lapse-rate correction and a new colour scale. The caption has been updated accordingly.

\item[Fig.~9:]
  Temperature difference maps now use bilinear interpolation, lapse-rate correction and a new colour scale. The caption has been updated accordingly.

\item[Fig.~10:]
  Precipitation density plots now use bilinear interpolation and a new colour scale. The caption has been updated accordingly.

\item[Fig.~11:]
  Temperature difference maps now use bilinear interpolation. The caption has been updated accordingly.

\end{description}

We hope that you will consider the revised manuscript for publication in The Cryosphere and look forward to hearing from you.

\flushright
With Best Regards,\\
J.~Seguinot~et~al.

\end{document}

### 2016-11-04 climate/disc-ac1.tex
Move Cordillera climate paper into a subdirectory.

--- 2014-04-03 reviews/disc-ac1.tex
@@
     \includegraphics[width=70mm]{cordillera-climate-tempdiff+nolr}
-    \caption{Summer temperature difference maps, as in paper Fig.~9.}
+    \caption{Summer temperature difference maps, as in paper Fig. 9, with a new colour scale.}
 \end{figure}
@@
     \includegraphics[width=70mm]{cordillera-climate-tempdiff}
-    \caption{Summer temperature difference maps, after applying a lapse-rate correction of 6\,{$^\circ$}C\,km$^{-1}$.}
+    \caption{Summer temperature difference maps, after applying a lapse-rate correction.}
 \end{figure}

### 2016-11-04 cycle/response-stumpf.tex
Move Cordillera cycle paper into subdirectory.

--- 2016-01-04 reviews/response-stumpf.tex
@@
     non-glacial conditions existed during MIS (Olympia Nonglacial Interval;
-    \citealp[e.g.,][]{Plouffe.Jette.1997}). I applaud them for undertaking of
+    e.g., Plouffe and Jette, 1997). I applaud them for undertaking of
     such a difficult task, and notifying the reader where input data is sparse
@@
     advanced our understanding of the CIS, and could be an important dataset to
-    test against the modeling. \citet{Stumpf.etal.2014} provides a list of some
+    test against the modeling. Stumpf et~al. (2014) provides a list of some
     of these studies.}
@@
     3) Was the model tested against regional-scale ground-based data
-    \citep[e.g.,][]{Ferbey.etal.2013} to constrain ice divide positions, ice
+    (e.g., Ferbey et~al., 2013) to constrain ice divide positions, ice
     flow direction, and ice sheet thickness?}
@@
 \referee{%
-    4) For the central sector of the CIS, \citet{Stumpf.etal.2000} provides
+    4) For the central sector of the CIS, Stumpf et~al. (2000) provides
     some insight into the chronology and effectiveness of glacial erosional
@@
 \referee{%
-    5) \citet{Stumpf.etal.2000} was the first study to extensively document a
+    5) Stumpf et~al. (2000) was the first study to extensively document a
     predominant westerly directed ice flow across high elevations in the Skeena

### 2016-11-04 cycle/abstract.tex
Move Cordillera cycle paper into subdirectory.

--- 2015-06-21 abstract.tex
@@
 % Affiliations
-\textsuperscript{1}{Laboratory of Hydraulics, Hydrology and Glaciology, ETH Zürich,
-          Zürich, Switzerland}\\
-\textsuperscript{2}{Department of Physical Geography and the Bolin Centre for Climate
-          Research, Stockholm University, Stockholm, Sweden}\\
-\textsuperscript{3}{Helmholtz Centre Potsdam, GFZ German Research Centre for Geosciences,
+\textsuperscript{1}{Laboratory of Hydraulics, Hydrology and Glaciology,
+                    ETH Z\"{u}rich, Z\"{u}rich, Switzerland}\\
+\textsuperscript{2}{Department of Physical Geography and the
+                    Bolin Centre for Climate Research,
+                    Stockholm University, Stockholm, Sweden}\\
+\textsuperscript{3}{Helmholtz Centre Potsdam,
+                    GFZ German Research Centre for Geosciences,
                     Potsdam, Germany}\\
@@
 
-  Despite more than a century of geological observations, the Cordilleran ice
-  sheet of North America remains poorly understood in terms of its former
-  extent, volume and dynamics. Although geomorphological evidence is abundant,
-  its complexity is such that whole ice-sheet reconstructions of advance and
-  retreat patterns are lacking. Here we use a numerical ice sheet model
-  calibrated against field-based evidence to attempt a quantitative
-  reconstruction of the Cordilleran ice sheet history through the last glacial
-  cycle. A series of simulations is driven by time-dependent temperature
-  offsets from six proxy records located around the globe. Although this
-  approach reveals large variations in model response to evolving climate
-  forcing, all simulations produce two major glaciations during
-  marine oxygen isotope stages~4 (61.9--56.5\,ka) and~2
-  (23.2--16.8\,ka). The timing of glaciation is
-  better reproduced using temperature reconstructions from Greenland and
-  Antarctic ice cores than from regional oceanic sediment cores. During most of
-  the last glacial cycle, the modelled ice cover is discontinuous and
-  restricted to high mountain areas. However, widespread precipitation over the
-  Skeena Mountains favours the persistence of a central ice dome throughout the
-  glacial cycle. It acts as a nucleation centre before the Last Glacial Maximum
-  and hosts the last remains of Cordilleran ice until the middle Holocene
-  (6.6--6.2\,ka).
+After more than a century of geological research, the Cordilleran
+ice sheet of North America remains among the least understood in
+terms of its former extent, volume, and dynamics. Because of the
+mountainous topography on which the ice sheet formed, geological
+studies have often had only local or regional relevance, and shown
+such a complexity that ice sheet-wide spatial reconstructions of
+advance and retreat patterns are
+lacking. Here we use a numerical ice sheet model calibrated against
+field-based evidence to attempt a quantitative reconstruction of the
+Cordilleran ice sheet history through the last glacial
+cycle. A series of simulations is driven by time-dependent
+temperature offsets from six proxy records located around the
+globe. Although this approach reveals large variations in model
+response to evolving climate forcing, all simulations produce two
+major glaciations during marine oxygen isotope stages 4
+(62.2--56.9\,ka) and 2 (23.2--16.9\,ka). The timing of
+glaciation is better reproduced using temperature reconstructions
+from Greenland and Antarctic ice cores than from regional oceanic
+sediment cores. During most of the last glacial cycle, the modelled
+ice cover is discontinuous and restricted to high mountain
+areas. However, widespread precipitation over the Skeena Mountains
+favours the persistence of a central ice dome throughout the glacial
+cycle. It acts as a nucleation centre before the Last Glacial
+Maximum and hosts the last remains of Cordilleran ice until the
+middle Holocene (6.7\,ka).
 

### 2017-01-29 articles/cycle/response-jarosch.tex
Move articles in articles directory.
To follow the new canonical structure.

--- 2016-01-05 reviews/response-jarosch.tex
@@
 
-% Replacements for Copernicus commands
+% Replacements for Copernicus commands that need to be here
 \newcommand{\doi}[1]{doi:\allowbreak\href{http://dx.doi.org/#1}{#1}}
+\newcommand{\degree}[0]{\ensuremath{^{\circ}}}
 \newcommand{\tophline}{\noalign{\vspace{1ex}}\hline\noalign{\vspace{1mm}}}
@@
 \newcommand{\belowtable}[1]{\par\vspace{1ex}\raggedleft{\scriptsize#1}}
+
+% Bold vectors and matrices
+\renewcommand{\vec}[1]{\ensuremath{\mbox{\boldmath$#1$}}}
 
@@
         A_c \,e^\frac{-Q_c}{RT_{pa}}
-            & \mathsf{if}\ T_{pa} < T_c \,, \\
+            & \mathrm{if}\ T_{pa} < T_c \,, \\
         A_w (1+f\omega)\,e^\frac{-Q_w}{RT_{pa}}
@@
 
-A fixed temperature lapse rate, $\gamma=6{\degree}C$, is certainly another
+A fixed temperature lapse rate, $\gamma=6$\,{\degree}C, is certainly another
 coarse approximation in our model setup. Although it is clear that $\gamma$
@@
     and wish them success for their future endeavours.}
+
+Thank you very much. We think that the manuscript is now stronger indeed.
 

### 2017-01-29 articles/cycle/response-marshall.tex
Move articles in articles directory.
To follow the new canonical structure.

--- 2016-01-04 reviews/response-marshall.tex
@@
 
-\newpage
-\sechead{1 \quad Specific comments}
+\sechead{2 \quad Specific comments}
 
@@
     resolution (5 to 10\,km) and ice physics, carefully calibrated `control'
-    climatology \citep[published in][]{Seguinot.etal.2014}), and a good
+    climatology (published in Seguinot et~al., 2014), and a good
     exploration of different paleoclimate time series histories in this
@@
 sea-surface temperatures \citep{Prahl.Wakeham.1987, Prahl.etal.1988,
-Muller.etal.1998}. However, we do not use $\delta^{18}O$ records from benthic
+Muller.etal.1998}. However, we do not use \chem{\delta^{18}O} records from benthic
 foraminifera, which have been evaluated worldwide \citep{Lisiecki.Raymo.2005},
@@
 pachyderma} and \emph{G. bulloides}) between 12 and
-21\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Taylor.etal.2014}. Very recently, this
+21\,\chem{^{14}C}\,cal\,ka \citep{Taylor.etal.2014}. Very recently, this
 planktonic foraminifera record has been extended to cover the period from 10 to
-ca.~50\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Taylor.etal.2015}, and thus it
+ca.~50\,\chem{^{14}C}\,cal\,ka \citep{Taylor.etal.2015}, and thus it
 could perhaps be used to model the growth and decay of the Cordilleran ice
@@
 \end{thebibliography}
+
 % ----------------------------------------------------------------------

### 2017-05-15 figures/ciscyc_tab_hr.py
Rename figures by subproject.
Update Makefile accordingly, but not article files.

@@ -5,7 +5,6 @@ import util as ut
 import numpy as np
 
 res = '10km'
-rec = 'grip'
 
 # latex header and footer
 header=r'''\documentclass[border=2.5mm]{standalone}
@@ -21,7 +20,7 @@ header=r'''\documentclass[border=2.5mm]{standalone}
              & \multicolumn{3}{c|}{Age (ka)}
              & \multicolumn{3}{c|}{Ice extent (\unit{10^6\,km^2})}
              & \multicolumn{3}{c}{Ice volume (m\,s.l.e.)} \\
-    Config.  &  MIS~4 &  MIS~3 &  MIS~2
+    Record   &  MIS~4 &  MIS~3 &  MIS~2
              &  MIS~4 &  MIS~3 &  MIS~2
              &  MIS~4 &  MIS~3 &  MIS~2 \\
     \middlehline
@@ -32,33 +31,30 @@ footer=r'''    \bottomhline
 \end{document}
 '''
 
-# latex table line (curly braces need to be doubled to escape)
+# latex table line
 tabline = (' '*4 + '{title:8}' + ' &{:7.2f}'*3 + '\n'+
            ' '*12 + ' &{:7.2f}'*3 + '\n'+
            ' '*12 + ' &{:7.2f}'*3 + r' \\' + '\n')
-errline = (' '*4 + '{title:8}' + r' &{:5.0f}\,\unit{{\%}}'*3 + '\n'+
-           ' '*12 + r' &{:5.0f}\,\unit{{\%}}'*3 + '\n'+
-           ' '*12 + r' &{:5.0f}\,\unit{{\%}}'*3 + r' \\' + '\n')
 
 # initialize array
-results = np.zeros((5, 9), dtype=float)
+results = np.zeros((6, 9), dtype=float)
 
 # open latex file to write in
-with open('tab_sens.tex', 'w') as f:
+with open('ciscyc_tab_hr.tex', 'w') as f:
 
     # write header
     f.write(header)
 
     # loop on records
-    for i, conf in enumerate(ut.sens.configs):
-        dt = ut.sens.offsets[i]
-        label = ut.sens.labels[i]
+    for i, rec in enumerate(ut.hr.records):
+        dt = ut.hr.offsets[i]
+        label = ut.hr.labels[i]
 
         # get MIS times
-        mis_idces, mis_times = ut.io.get_mis_times(res, rec, dt, config=conf)
[…]

### 2017-05-15 figures/ciscyc_tab_lr.py

@@ -40,15 +40,15 @@ tabline = (' '*4 + '{title:8}' + ' &{:7.2f}'*3 + '\n'+
 results = np.zeros((6, 9), dtype=float)
 
 # open latex file to write in
-with open('tab_hr.tex', 'w') as f:
+with open('ciscyc_tab_lr.tex', 'w') as f:
 
     # write header
     f.write(header)
 
     # loop on records
-    for i, rec in enumerate(ut.hr.records):
-        dt = ut.hr.offsets[i]
-        label = ut.hr.labels[i]
+    for i, rec in enumerate(ut.lr.records):
+        dt = ut.lr.offsets[i]
+        label = ut.lr.labels[i]
 
         # get MIS times
         mis_idces, mis_times = ut.io.get_mis_times(res, rec, dt)
@@ -73,6 +73,15 @@ with open('tab_hr.tex', 'w') as f:
         # write info in table
         results[i] = np.concatenate((-mis_times/1e3, mis_gareas, mis_slvols))
         f.write(tabline.format(title=label, *results[i]))
+
+    # compute extrema
+    minima = results.min(axis=0)
+    maxima = results.max(axis=0)
+
+    # write info in table
+    f.write('    \cline{1-10}\n')
+    f.write(tabline.format(title='Minimum', *minima))
+    f.write(tabline.format(title='Maximum', *maxima))
 
     # write footer
     f.write(footer)

### 2017-05-15 figures/ciscyc_tab_sens.py

@@ -5,6 +5,7 @@ import util as ut
 import numpy as np
 
 res = '10km'
+rec = 'grip'
 
 # latex header and footer
 header=r'''\documentclass[border=2.5mm]{standalone}
@@ -20,7 +21,7 @@ header=r'''\documentclass[border=2.5mm]{standalone}
              & \multicolumn{3}{c|}{Age (ka)}
              & \multicolumn{3}{c|}{Ice extent (\unit{10^6\,km^2})}
              & \multicolumn{3}{c}{Ice volume (m\,s.l.e.)} \\
-    Record   &  MIS~4 &  MIS~3 &  MIS~2
+    Config.  &  MIS~4 &  MIS~3 &  MIS~2
              &  MIS~4 &  MIS~3 &  MIS~2
              &  MIS~4 &  MIS~3 &  MIS~2 \\
     \middlehline
@@ -31,30 +32,33 @@ footer=r'''    \bottomhline
 \end{document}
 '''
 
-# latex table line
+# latex table line (curly braces need to be doubled to escape)
 tabline = (' '*4 + '{title:8}' + ' &{:7.2f}'*3 + '\n'+
            ' '*12 + ' &{:7.2f}'*3 + '\n'+
            ' '*12 + ' &{:7.2f}'*3 + r' \\' + '\n')
+errline = (' '*4 + '{title:8}' + r' &{:5.0f}\,\unit{{\%}}'*3 + '\n'+
+           ' '*12 + r' &{:5.0f}\,\unit{{\%}}'*3 + '\n'+
+           ' '*12 + r' &{:5.0f}\,\unit{{\%}}'*3 + r' \\' + '\n')
 
 # initialize array
-results = np.zeros((6, 9), dtype=float)
+results = np.zeros((5, 9), dtype=float)
 
 # open latex file to write in
-with open('tab_lr.tex', 'w') as f:
+with open('ciscyc_tab_sens.tex', 'w') as f:
 
     # write header
     f.write(header)
 
     # loop on records
-    for i, rec in enumerate(ut.lr.records):
-        dt = ut.lr.offsets[i]
-        label = ut.lr.labels[i]
+    for i, conf in enumerate(ut.sens.configs):
+        dt = ut.sens.offsets[i]
+        label = ut.sens.labels[i]
 
         # get MIS times
-        mis_idces, mis_times = ut.io.get_mis_times(res, rec, dt)
[…]

### 2017-05-15 figures/cisghf_timestamps.py

@@ -1,6 +1,7 @@
 #!/usr/bin/env python2
 # coding: utf-8
 
+import util as ut
 import numpy as np
 import matplotlib.pyplot as plt
 import netCDF4

### 2017-05-15 figures/util/pl.py

@@ -195,5 +195,5 @@ def savefig(fig=None):
     """Save figure to script filename."""
     import sys
     fig = fig or gcf()
-    res = fig.savefig(ut.io.os.path.splitext(sys.argv[0])[0])
+    res = fig.savefig(os.path.splitext(sys.argv[0])[0])
     return res
